### 2021-02-19 Analysis/01_data_preparation.Rmd
Analysis INIT

---
title: "Data preparation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Data preparation

 - [ ] Selecting the data from the subclade and their subtrees.
 - [ ] Running the procrustes superimposition

### 2021-02-19 Analysis/02-ancestral_states_estimations.Rmd
Analysis INIT

---
title: "Ancestral traits estimations"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Ancestral traits estimations

 - [ ] Run the ancestral states estimations on different trees to have a distribution rather than point estimates. Three options here to build the fuzzy trait space:
    * [ ] **Fuzzy average**: run the ace on multiple trees and use their multiple averages as the ace distribution.
    * [ ] **Random fuzzy**: run the ace on multiple trees and randomly sample one value from the estimation
    * [ ] **Fuzzy distribution**: run the ace on multiple trees and sample *n* random values from the estimation
 - [ ] Combine the data together into `dispRity` objects

### 2021-02-19 Analysis/03-elaboration_exploration_analyses.Rmd
Analysis INIT

---
title: "03-elaboration exploration analyses"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Elaboration/exploration analyses

## Elaboration/exploration metrics

Measure the different elaboration vs exploration metrics with different vectors:

 - [ ] MRCA trajectory: the vector between the root of the tree and the OTU's descendant
 - [ ] MRCA average trajectory: the vector between the centroid of the group's nodes and the root of the tree
 - [ ] Evolutionary living average trajectory: the vector between the root of the tree and the centroid of the living taxa
 - [ ] Evolutionary overall average trajectory: the vector between the root of the tree and the centroid of the tips and nodes
 - [ ] Maximum living trajectory: the vector between the two most distant OTUs
 - [ ] Maximum evolutionary trajectory: the vector between the root and the OTU most distant from the root

## Elaboration/exploration in the present

Apply these metrics on both clades to get two scores for each taxa (elaboration vs. exploration plots).

## Elaboration/exploration through time

Do some time-slicing and run these metrics through time.

### 2021-02-19 Analysis/01_data_preparation.Rmd
Update 01_data_preparation.Rmd

@@
 
+```{r}
+library(ape)
+library(dispRity)
+```
+
 # Data preparation
 
- - [ ] Selecting the data from the subclade and their subtrees.
- - [ ] Running the procrustes superimposition
+Here I'm going to focus on the Psittaciformes (parrots and stuff) and Charadriiformes (sea gulls and stuff).
+
+ - [x] Selecting the data from the subclade and their subtrees.
+ - [x] Selecting the shape data
+
+## Tree data
+
+For this pilote analysis, I will only select 10 random trees between both Ericson and Hackett backbone trees from Jetz et al big trees (5 each).
+To find the mrca to get the subtrees, we will use the pairs `"Micropsitta"` and `"  Nestor"` for the parrots and `"Ichthyaetus"` and `"Burhinus"` for the sea gulls
+
+```{r, eval = FALSE}
+## Get the pairs of birds for both clades
+psita_clade <- c("Micropsitta_meeki", "Nestor_notabilis")
+chara_clade <- c("Burhinus_oedicnemus", "Larus_saundersi")
+
+## Number of trees to select
+n_trees <- 10
+
+## Reading the trees
+ericson_tree_1_1000 <- read.tree("../Data/Raw/Tree_data/AllBirdsEricson1.tre")
+hackett_tree_1_1000 <- read.tree("../Data/Raw/Tree_data/AllBirdsHackett1.tre")
+
+## Selecting n/2 random trees for both tree
+selected_trees <- replicate(2, sample(1:1000, n_trees/2))
+
+## Sampling the trees
+ericson_tree_5 <- ericson_tree_1_1000[selected_trees[,1]]
+hackett_tree_5 <- hackett_tree_1_1000[selected_trees[,2]]
+## Combine the trees
+all_birds_10 <- c(ericson_tree_5, hackett_tree_5)
+
+## Extracting the different clades from the different trees
+wrap.extract.clade <- function(tree, tips) {
+    ## Extract the clade
+    return(extract.clade(tree, node = getMRCA(tree, tip = tips)))
+}
+
+## Get the two trees
+chara_trees_10 <- lapply(all_birds_10, wrap.extract.clade, tip = chara_clade)
+psita_trees_10 <- lapply(all_birds_10, wrap.extract.clade, tip = psita_clade)
+class(chara_trees_10) <- class(psita_trees_10) <- "multiPhylo"
+
+## Save as a list of trees
+trees_list <- list("charadriiformes" = chara_trees_10,
+                   "psittaciformes"  = psita_trees_10)
+save(trees_list, file = "../Data/Processed/trees_list.rda")
+```
+
+## Beak data
+
+```{r}
+## Loading the tree data (to have the species list)
+load("../Data/Processed/trees_list.rda")
+```
+
+Note that here we are extracting the sets of birds (orders) from the global (full) shape space.
+This means that for the following analyses we're only looking at sub-parts in that space, i.e. the PCA does not contain all the data.
+If this is undersirable, I guess we can always do the whole pipeline again (select species > procrustes per order > pca per order).
+
+```{r}
+## Loading the whole shapespace
+shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")
+
+## Select the subspecies per shapespace
+psita_space <- shapespace$PCscores[rownames(shapespace$PCscores) %in% trees_list$psittaciformes[[1]]$tip.label, ]
+chara_space <- shapespace$PCscores[rownames(shapespace$PCscores) %in% trees_list$charadriiformes[[1]]$tip.label, ]
+
+## Clean the tree lists and the dataset
+cleaned_psita <- clean.data(psita_space, trees_list$psittaciformes)
+cleaned_chara <- clean.data(chara_space, trees_list$charadriiformes)
+
+## Re-saving the trees and saving the spaces
+spaces_list <- list(psittaciformes  = cleaned_psita$data,
+                    charadriiformes = cleaned_chara$data)
+trees_list  <- list(psittaciformes  = cleaned_psita$tree,
+                    charadriiformes = cleaned_chara$tree)
+save(trees_list,  file = "../Data/Processed/trees_list.rda")
+save(spaces_list, file = "../Data/Processed/spaces_list.rda")
+```

### 2021-02-19 Analysis/01_data_preparation.Rmd
Added tip names

@@
 To find the mrca to get the subtrees, we will use the pairs `"Micropsitta"` and `"  Nestor"` for the parrots and `"Ichthyaetus"` and `"Burhinus"` for the sea gulls
+
+> Note
 
@@
                     charadriiformes = cleaned_chara$data)
-trees_list  <- list(psittaciformes  = cleaned_psita$tree,
-                    charadriiformes = cleaned_chara$tree)
+
+## Adding the node labels for each selected trees
+psita_trees <- lapply(cleaned_psita$tree, function(tree) {tree$node.label <- paste0("n", 1:Nnode(tree)); return(tree)})
+chara_trees <- lapply(cleaned_chara$tree, function(tree) {tree$node.label <- paste0("n", 1:Nnode(tree)); return(tree)})
+class(psita_trees) <- class(chara_trees) <- "multiPhylo"
+
+trees_list  <- list(psittaciformes  = psita_trees,
+                    charadriiformes = chara_trees)
+
 save(trees_list,  file = "../Data/Processed/trees_list.rda")

### 2021-02-19 Analysis/02-ancestral_states_estimations.Rmd
Added utility functions for the ancestral states estimations

@@
 
+```{r}
+library(ape)
+library(castor)
+```
+
 # Ancestral traits estimations
@@
  - [ ] Combine the data together into `dispRity` objects
+
+## Loading the data
+
+```{r}
+load(file = "../Data/Processed/trees_list.rda")
+load(file = "../Data/Processed/spaces_list.rda")
+```
+
+## Running the ancestral states
+
+We can estimate the ancestral characters using the `ape::ace` function.
+Here we define "ancestral characters" as "the position of a node of the tree in the ordinated beak shape space".
+We will assume that position in shape space evolves under a Brownian Motion that is only dependent on time.
+I.e.
+the longer the evolutionary time between two elements (tips or nodes), the more variance there is in position in shape space.
+
+> Note that this is not the only assumption we can make.
+One other correct one would be to scale the branch length by amount of changes (cf.
+by evolutionary time).
+This way, the longer the branch (the longer the molecular changes) between two elements, the more variance.
+
+We will also assume that the position in the trait space are independent (i.e. position on the first axis of the trait space is not related to the position on the second axis, etc...).
+
+> This assumption can also be relaxed pretty easily using a multi trait BM.
+However, it is rather defendable to not do here since the ordination by definition makes the axis orthogonal and thus non-correlated.
+
+From each ancestral character estimation, we will extract either:
+ * [ ] the character value with the highest likelihood
+ * [ ] one random character value from within the 95% confidence interval
+ * [ ] ten random character values from the 95% CI
+
+```{r}
+# ##DEBUG
+# matrix <- matrix(rnorm(100), 10, 5)
+# trees <- rmtree(7, 10)
+# trees <- lapply(trees, function(tree) {tree$node.label <- paste0("n", 1:Nnode(tree)); return(tree)})
+# class(trees) <- "multiPhylo"
+# tree <- trees[[1]]
+# rownames(matrix) <- tree$tip.label
+
+## Function for running the ace on the whole matrix (i.e. for all the traits)
+one.ace <- function(traits, tree) {
+    cat(".")
+    results <- ace(traits, tree)
+    results$labels <- tree$node.label
+    return(results)
+}
+run.full.ace <- function(tree, matrix) {
+    cat("Estimating one matrix:")
+    ace_results <- apply(matrix, 2, one.ace, tree)
+    cat("Done\n")
+    return(ace_results)
+}
+
+# ##DEBUG
+# all_results <- lapply(trees, run.full.ace, matrix)
+# ace_result <- all_results[[1]]
+
+## Function for extracting the ace values of interest
+#' @param ace the output from run.full.ace
+#' @param what either "max" for the most likely value, or any integer for a number of random values (e.g. 1 or 10)
+#' @param sampling the sampling distribution can be "uniform" (default) or "normal"0
+#' 
+## Extracting random values from the ace_results 95% (uniform or normal)
+fun.uniform <- function(ci95, what) {
+    return(runif(what, min = ci95[1], max = ci95[2]))
+}
+fun.normal <- function(ci95, what) {
+    return(rnorm(what, mean = mean(x), sd = sd(x)/3))
+}
+extract.ace <- function(ace_result, what, sampling = "uniform", matrix) {
+    ## Switching between both type of results
+    if(what == "max") {
+        ## Extracting the most likely states
+        node_states <- do.call(cbind, lapply(ace_result, function(x) return(x$ace)))
+        rownames(node_states) <- ace_result[[1]]$labels
+        return(rbind(matrix, node_states))
+    } else {
+        ## Toggle the sampling function
+        fun.sample <- ifelse(sampling == "uniform", fun.uniform, fun.normal)
+        trait_results <- lapply(ace_result, function(trait, fun.sample, what) apply(trait$CI95, 1, fun.sample, what = what), fun.sample, what)
+
+        if(what == 1) {
+            ## Return single samples
+            node_states <- do.call(cbind, trait_results)
+            rownames(node_states) <- ace_result[[1]]$labels
+            return(rbind(matrix, node_states))            
+        } else {
+            ## Return multiple samples
+            node_states <- list()
+            for(i in 1:what) {
+                node_states[[i]] <- do.call(cbind, lapply(trait_results, function(X, i) return(X[i, ]), i))
+                rownames(node_states[[i]]) <- ace_result[[1]]$labels
+            }
+            return(lapply(node_states, function(X, matrix) rbind(matrix, X), matrix))
+        }
+    }
+}
+# ## DEBUG
+# #all_trees_max_lik <- lapply(all_results, extract.ace, what = "max", matrix = matrix)
+```
+
+```{r, eval = TRUE}
+## Running the ace for all the characters for both spaces
+psita_ace <- lapply(trees_list$psittaciformes, run.full.ace,
+                    matrix = spaces_list$psittaciformes)
+chara_ace <- lapply(trees_list$charadriiformes, run.full.ace,
+                    matrix = spaces_list$charadriiformes)
+save(psita_ace, file = "../Data/Processed/psita_ace.rda")
+save(chara_ace, file = "../Data/Processed/chara_ace.rda")
+```
+
+## Make the dispRity objects

### 2021-02-26 Functions/ancestral.states.fun.R
Isolated functions per script

@@ -1,0 +1,70 @@
+
+## Function for running the ace on the whole matrix (i.e. for all the traits)
+one.ace <- function(traits, tree) {
+    cat(".")
+    results <- ace(traits, tree)
+    results$labels <- tree$node.label
+    return(results)
+}
+run.full.ace <- function(tree, matrix) {
+    cat("Estimating one matrix:")
+    ace_results <- apply(matrix, 2, one.ace, tree)
+    cat("Done\n")
+    return(ace_results)
+}
+
+## Function for extracting the ace values of interest
+#' @param ace the output from run.full.ace
+#' @param what either "max" for the most likely value, or any integer for a number of random values (e.g. 1 or 10)
+#' @param sampling the sampling distribution can be "uniform" (default) or "normal"0
+#' 
+## Extracting random values from the ace_results 95% (uniform or normal)
+fun.uniform <- function(ci95, what) {
+    return(runif(what, min = ci95[1], max = ci95[2]))
+}
+fun.normal <- function(ci95, what) {
+    return(rnorm(what, mean = mean(ci95), sd = diff(ci95)/4))
+}
+extract.ace <- function(ace_result, what, sampling = "uniform", matrix) {
+    ## Switching between both type of results
+    if(what == "max") {
+        ## Extracting the most likely states
+        node_states <- do.call(cbind, lapply(ace_result, function(x) return(x$ace)))
+        rownames(node_states) <- ace_result[[1]]$labels
+        return(rbind(matrix, node_states))
+    } else {
+        ## Toggle the sampling function
+        fun.sample <- ifelse(sampling == "uniform", fun.uniform, fun.normal)
+        trait_results <- lapply(ace_result, function(trait, fun.sample, what) apply(trait$CI95, 1, fun.sample, what = what), fun.sample, what)
+
+        if(what == 1) {
+            ## Return single samples
+            node_states <- do.call(cbind, trait_results)
+            rownames(node_states) <- ace_result[[1]]$labels
+            return(rbind(matrix, node_states))            
+        } else {
+            ## Return multiple samples
+            node_states <- list()
+            for(i in 1:what) {
+                node_states[[i]] <- do.call(cbind, lapply(trait_results, function(X, i) return(X[i, ]), i))
+                rownames(node_states[[i]]) <- ace_result[[1]]$labels
[…]

### 2021-02-26 Analysis/02-ancestral_states_estimations.Rmd
FInished ancestral states script
Missing add tree

--- 2021-02-19 Analysis/02-ancestral_states_estimations.Rmd
@@
 
+```{r}
+library(ape)
+library(dispRity)
+source("../FUnctions/ancestral.states.fun.R")
+```
+
 # Ancestral traits estimations
@@
  - [ ] Combine the data together into `dispRity` objects
+
+## Loading the data
+
+```{r}
+load(file = "../Data/Processed/trees_list.rda")
+load(file = "../Data/Processed/spaces_list.rda")
+```
+
+## Running the ancestral states
+
+We can estimate the ancestral characters using the `ape::ace` function.
+Here we define "ancestral characters" as "the position of a node of the tree in the ordinated beak shape space".
+We will assume that position in shape space evolves under a Brownian Motion that is only dependent on time.
+I.e.
+the longer the evolutionary time between two elements (tips or nodes), the more variance there is in position in shape space.
+
+> Note that this is not the only assumption we can make.
+One other correct one would be to scale the branch length by amount of changes (cf.
+by evolutionary time).
+This way, the longer the branch (the longer the molecular changes) between two elements, the more variance.
+
+We will also assume that the position in the trait space are independent (i.e. position on the first axis of the trait space is not related to the position on the second axis, etc...).
+
+> This assumption can also be relaxed pretty easily using a multi trait BM.
+However, it is rather defendable to not do here since the ordination by definition makes the axis orthogonal and thus non-correlated.
+
+From each ancestral character estimation, we will extract either:
+ * [ ] the character value with the highest likelihood
+ * [ ] one random character value from within the 95% confidence interval
+ * [ ] ten random character values from the 95% CI
+
+```{r, eval = FALSE}
+## Running the ace for all the characters for both spaces
+psita_ace <- lapply(trees_list$psittaciformes, run.full.ace,
+                    matrix = spaces_list$psittaciformes)
+chara_ace <- lapply(trees_list$charadriiformes, run.full.ace,
+                    matrix = spaces_list$charadriiformes)
+save(psita_ace, file = "../Data/Processed/psita_ace.rda")
+save(chara_ace, file = "../Data/Processed/chara_ace.rda")
+```
+
+### Extracting the ancestral estimates
+
+Here we will select the ancestral traits estimations two ways:
+ * First by selecting the most likely estimations (i.e. highest lik log) for each node;
+ * Second by randomly sampling 10 values using a normal distribution around the 95% estimation interval (mean = mean(ci95); sd = diff(ci95)/4; the diff(ci95) is because there are roughly 4 standard deviations in the 95% range of a normal distribution).
+
+```{r}
+## Loading the precaculated results
+load("../Data/Processed/psita_ace.rda")
+load("../Data/Processed/chara_ace.rda")
+
+## Building the matrices
+psita_max_ace <- lapply(psita_ace, extract.ace, what = "max",
+                         matrix = spaces_list$psittaciformes)
+psita_10n_ace <- lapply(psita_ace, extract.ace, what = 10, sampling = "normal",
+                         matrix = spaces_list$psittaciformes)
+chara_max_ace <- lapply(chara_ace, extract.ace, what = "max",
+                         matrix = spaces_list$charadriiformes)
+chara_10n_ace <- lapply(psita_ace, extract.ace, what = 10, sampling = "normal",
+                         matrix = spaces_list$psittaciformes)
+```
+
+## Make the dispRity objects
+
+> TODO: add the trees to these matrices for the next dispRitree package!
+
+```{r}
+## Converting the lists into dispRity objects
+disparity_matrices <- list(
+    "psita_max" = make.dispRity(data = psita_max_ace),
+    "psita_10n" = make.dispRity(data = unlist(psita_10n_ace, recursive = FALSE)),
+    "chara_max" = make.dispRity(data = chara_max_ace),
+    "chara_10n" = make.dispRity(data = unlist(chara_10n_ace, recursive = FALSE)))
+save(disparity_matrices, file = "../Data/Processed/disparity_matrices.rda")
+```

### 2021-03-22 Analysis/02-ancestral_states_estimations.Rmd
added visualisation for ancestral shape spaces

@@
 library(dispRity)
-source("../FUnctions/ancestral.states.fun.R")
+source("../Functions/ancestral.states.fun.R")
 ```
@@
 
+> TODO: improve the ACE speed exploring castor::asr_independent_contrasts, castor::asr_squared_change_parsimony or castor::asr_subtree_averaging
+
 ### Extracting the ancestral estimates
@@
                          matrix = spaces_list$charadriiformes)
-chara_10n_ace <- lapply(psita_ace, extract.ace, what = 10, sampling = "normal",
-                         matrix = spaces_list$psittaciformes)
+chara_10n_ace <- lapply(chara_ace, extract.ace, what = 10, sampling = "normal",
+                         matrix = spaces_list$charadriiformes)
 ```
@@
 ```
+
+## Visualisation
+
+```{r, fig.height = 18, fig.width = 12}
+par(mfrow = c(3, 2))
+
+## Loading the full space
+plot.dispRity(spaces_list$psittaciformes, col = "blue", main = "Psittaciformes section of the shape space", xlab = "PC1", ylab = "PC2")
+plot.dispRity(spaces_list$charadriiformes, col = "orange", main = "Charadriiformes section of the shape space", xlab = "PC1", ylab = "PC2")
+
+## Full space with max ancestor
+plot(custom.subsets(psita_max_ace[[1]], group = list("Nodes" = 342:681, "Tips" = 1:341)), col = c("grey", "blue"), pch = c(19, 19), main = "Psittaciformes with nodes (max.
+lik.; tree 1)", xlab = "PC1", ylab = "PC2")
+plot(custom.subsets(chara_max_ace[[1]], group = list("Nodes" = 360:717, "Tips" = 1:359)), col = c("grey", "orange"), pch = c(19, 19), main = "Charadriiformes with nodes (max.
+lik.; tree 1)", xlab = "PC1", ylab = "PC2")
+
+## Full space with random ancestor
+plot(custom.subsets(psita_10n_ace[[1]], group = list("Nodes" = 342:681, "Tips" = 1:341)), col = c("grey", "blue"), pch = c(19, 19), main = "Psittaciformes with nodes (random; tree 1)", xlab = "PC1", ylab = "PC2")
+plot(custom.subsets(chara_10n_ace[[1]], group = list("Nodes" = 360:717, "Tips" = 1:359)), col = c("grey", "orange"), pch = c(19, 19), main = "Charadriiformes with nodes (random; tree 1)", xlab = "PC1", ylab = "PC2")
+```

### 2021-03-22 Functions/ancestral.states.fun.R
Update ancestral.states.fun.R

@@ -26,14 +26,14 @@ fun.normal <- function(ci95, what) {
     return(rnorm(what, mean = mean(ci95), sd = diff(ci95)/4))
 }
 extract.ace <- function(ace_result, what, sampling = "uniform", matrix) {
-    ## Switching between both type of results
+    ## Switching between both type of results
     if(what == "max") {
         ## Extracting the most likely states
         node_states <- do.call(cbind, lapply(ace_result, function(x) return(x$ace)))
         rownames(node_states) <- ace_result[[1]]$labels
         return(rbind(matrix, node_states))
     } else {
-        ## Toggle the sampling function
+        ## Toggle the sampling function
         fun.sample <- ifelse(sampling == "uniform", fun.uniform, fun.normal)
         trait_results <- lapply(ace_result, function(trait, fun.sample, what) apply(trait$CI95, 1, fun.sample, what = what), fun.sample, what)
 

### 2021-03-22 Analysis/04-elaboration_exploration_analyses.Rmd
evo elavoration.exploration analyses

---
title: "04-elaboration exploration analyses"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Elaboration/exploration analyses

```{r}
## Loading the packages and data
library(dispRity)
load(file = "../Data/Processed/disparity_matrices.rda")
load(file = "../Data/Processed/trees_list.rda")
load(file = "../Data/Processed/spaces_list.rda")
load(file = "../Data/Processed/tmp_simple_matrices.rda")
```

We can add some evo perspective to the basic Endler's elaboration/exploration idea by making the base vector "evolutionary" and by looking at elaboration/exploration through time!

> TODO: I haven't done implementing all the fuzzy space thingies in `dispRity` so for now it's just going to be a naive 1 tree 1 ace 1 answer version of it

## Elaboration/exploration evolutionary metrics

Measure the different elaboration vs exploration metrics with different vectors:

 - [ ] MRCA trajectory: the vector between the root of the tree and the OTU's descendant (`"root-anc"`)
 - [ ] MRCA average trajectory: the vector between the centroid of the group's nodes and the root of the tree (`"cent-anc"`)
 - [ ] Evolutionary living average trajectory: the vector between the root of the tree and the centroid of the living taxa (`"root-tip"`)
 - [ ] Evolutionary overall average trajectory: the vector between the root of the tree and the centroid of the tips and nodes (`"root-cent"`)
 - [ ] Maximum living trajectory: the vector between the two most distant OTUs (`"tip-span"`)
 - [ ] Maximum evolutionary trajectory: the vector between the root and the OTU most nodes away from the root (`"root-nested"`)

```{r}
## Utility functions
project.reject <- function(data, vector, absolute = TRUE) {

    ## Get the projection
    projection <- projections.tree(
          matrix = data$matrix[[1]], 
          tree = data$tree[[1]],
          measure = "position", type = vector)

    ## Get the rejection
    rejection <- projections.tree(
          matrix = data$matrix[[1]], 
          tree = data$tree[[1]],
          measure = "distance", type = vector)

    ## Combine and out
    if(absolute) {
      return(cbind(abs(projection), rejection))
    } else {
      return(cbind(projection, rejection))
    }
}
plot.project.reject <- function(results, data, xlab = "Elaboration", ylab = "Exploration", col = "black", main = "Project/Reject") {

    pch <- c(19, 21)
    col <- c("grey", col)

    ## Find the first node to separate nodes/tips
    root <- which(rownames(data$matrix[[1]]) == "n1")
    ## Split the plot data
    tips_block <- c(1:root-1)
    nodes_block <- c(root:nrow((data$matrix[[1]])))
    ## Split the colours and pch
    colours <- c(rep(col[1], length(nodes_block)),
                 rep(col[2], length(tips_block)))
    pchs <- c(rep(pch[1], length(nodes_block)),
              rep(pch[2], length(tips_block)))
    ## plot the results
    plot(results[c(nodes_block, tips_block), ], xlab = xlab, ylab = ylab, col = colours, pch = pchs, main = main)
}
plot.project.reject.log <- function(results, data, xlab = "Elaboration (log)", ylab = "Exploration (log)", col = "black", main = "Project/Reject") {

    plot.project.reject(log(results), data, xlab, ylab, col, main)
}

```

```{r, fig.height = 12, fig.width = 12}
## MRCA trajectory 
psita_results <- project.reject(tmp_simple_matrices$psita,
                                 vector = c("root", "ancestor"))
chara_results <- project.reject(tmp_simple_matrices$chara,
                                 vector = c("root", "ancestor"))

par(mfrow = c(2,2))
plot.project.reject(psita_results, tmp_simple_matrices$psita, col = "blue", main = "Psittaciformes - root/ancestor")
plot.project.reject(chara_results, tmp_simple_matrices$chara, col = "orange", main = "Charadriiformes - root/ancestor")
plot.project.reject.log(psita_results, tmp_simple_matrices$psita, col = "blue", main = "Psittaciformes - root/ancestor")
plot.project.reject.log(chara_results, tmp_simple_matrices$chara, col = "orange", main = "Charadriiformes - root/ancestor")
```

```{r, fig.height = 6, fig.width = 12}
## Centroid (tips) ancestor trajectory 
psita_results <- project.reject(tmp_simple_matrices$psita,
                                 vector = c("tips", "ancestor"))
chara_results <- project.reject(tmp_simple_matrices$chara,
                                 vector = c("tips", "ancestor"))

par(mfrow = c(1,2))
plot.project.reject.log(psita_results, tmp_simple_matrices$psita, col = "blue", main = "Psittaciformes - tips/ancestor")
plot.project.reject.log(chara_results, tmp_simple_matrices$chara, col = "orange", main = "Charadriiformes - tips/ancestor")
```

```{r, fig.height = 6, fig.width = 12}
## Centroid (nodes) ancestor trajectory 
psita_results <- project.reject(tmp_simple_matrices$psita,
                                 vector = c("nodes", "ancestor"))
chara_results <- project.reject(tmp_simple_matrices$chara,
                                 vector = c("nodes", "ancestor"))

par(mfrow = c(1,2))
plot.project.reject.log(psita_results, tmp_simple_matrices$psita, col = "blue", main = "Psittaciformes - nodes/ancestor")
plot.project.reject.log(chara_results, tmp_simple_matrices$chara, col = "orange", main = "Charadriiformes - nodes/ancestor")
```

```{r, fig.height = 6, fig.width = 12}
## Centroid (root) tips trajectory 
psita_results <- project.reject(tmp_simple_matrices$psita,
                                 vector = c("root", "tips"))
chara_results <- project.reject(tmp_simple_matrices$chara,
                                 vector = c("root", "tips"))

par(mfrow = c(1,2))
plot.project.reject.log(psita_results, tmp_simple_matrices$psita, col = "blue", main = "Psittaciformes - root/tips")
plot.project.reject.log(chara_results, tmp_simple_matrices$chara, col = "orange", main = "Charadriiformes - root/tips")
```

## Elaboration/exploration in the present

Apply these metrics on both clades to get two scores for each taxa (elaboration vs. exploration plots).

## Elaboration/exploration through time

Do some time-slicing and run these metrics through time.

<!-- ## Annex: some potentially interesting ratio thingy

```{r, fig.height = 6, fig.width = 6, eval = FALSE, print = FALSE}
## Loading the whole shape space
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
## Getting the vector from that space
space_vector <- apply(shapespace, 2, quantile, prob = c(0.25, 0.75))
psita_space_proj <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ])
psita_space_reje <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
chara_space_proj <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ])
chara_space_reje <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
## Groups
group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
## Full space plot
plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
arrows(x0 = space_vector[1,1], y0 = space_vector[1,2],
       x1 = space_vector[2,1], y1 = space_vector[2,2], col = "black", lwd = 4)
```

```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
par(mfrow = c(2, 2))
## The elaboration/exploration profiles
plot(psita_space_proj, psita_space_reje, pch = 19, col = "blue",
    xlab = "Elaboration", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
plot(chara_space_proj, chara_space_reje, pch = 19, col = "orange",
    xlab = "Elaboration", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")

## The elaboration/exploration profiles
plot(abs(psita_space_proj), psita_space_reje, pch = 19, col = "blue",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
plot(abs(chara_space_proj), chara_space_reje, pch = 19, col = "orange",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
```

Meh... But we can also do ratios: the one between the within group elaboration/exploration and the total elaboration/exploration.
This can help us find the big time explorers/elaborators? (here: the one in the top 99% CI)

```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
## Get the ratio
reje_ratio <- psita_reje/psita_space_reje
proj_ratio <- abs(psita_proj)/abs(psita_space_proj)

## Find the 99% top explorers/elaborators
top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
top_explo_names <- rownames(spaces_list$psittaciformes)[top_explo]
top_elab_names <- rownames(spaces_list$psittaciformes)[top_elab]

plot(proj_ratio, reje_ratio, pch = 19, col = "blue",
    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Psittaciformes")
text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
```

```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
## Get the ratio
reje_ratio <- chara_reje/chara_space_reje
proj_ratio <- abs(chara_proj)/abs(chara_space_proj)

## Find the 99% top explorers/elaborators
top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
top_explo_names <- rownames(spaces_list$charadriiformes)[top_explo]
top_elab_names <- rownames(spaces_list$charadriiformes)[top_elab]

plot(proj_ratio, reje_ratio, pch = 19, col = "orange",
    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Charadriiformes")
text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
```

Maybe it's picking up some weird birds...
 -->

### 2021-03-22 Analysis/02-elaboration_exploration_basics.Rmd
basics

---
title: "02-elaboration exploration basics"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Elaboration/exploration analyses

```{r}
## Loading the packages and data
library(dispRity)
load(file = "../Data/Processed/trees_list.rda")
load(file = "../Data/Processed/spaces_list.rda")
```

## Endler's style (basics)

<img title="Elaboration/Projection" alt="Elaboration (projection) vs. exploration (rejection)" src="https://upload.wikimedia.org/wikipedia/commons/9/98/Projection_and_rejection.png">

[image source](https://en.wikipedia.org/wiki/Vector_projection#/media/File:Projection_and_rejection.png)

The basic idea is to measure the elaboration as the projection of observations (species) on a base vector (e.g. ***a1*** the projection of ***a*** on ***b***) and the exploration as the rejection from the base vector (e.g. ***a2*** the rejection of ***a*** from ***b***).
Each observation can then be described in 2D with a elaboration score (the higher the absolute value of the score, the more the species is an elaborator - in both directions of ***b***) and a exploration score (the higher, the more they explore).
Of course we do that in any number of dimensions (not just 2!) and we're going to play with the meaning of vector ***b*** (the base vector).

We can play around with it with these two spaces by defining a vector that is spanning from the 25%th quantile to the 75% quantile of all data (the 50% "average"ish axis).

```{r}
## Getting the base vectors for both groups
psita_vector <- apply(spaces_list$psittaciformes, 2, quantile, prob = c(0.25, 0.75))
chara_vector <- apply(spaces_list$charadriiformes, 2, quantile, prob = c(0.25, 0.75))

## Calculating the projections/rejections for both groups
psita_proj <- projections(spaces_list$psittaciformes, point1 = psita_vector[1,], point2 = psita_vector[2, ])
psita_reje <- projections(spaces_list$psittaciformes, point1 = psita_vector[1,], point2 = psita_vector[2, ], measure = "distance")
chara_proj <- projections(spaces_list$charadriiformes, point1 = chara_vector[1,], point2 = chara_vector[2, ])
chara_reje <- projections(spaces_list$charadriiformes, point1 = chara_vector[1,], point2 = chara_vector[2, ], measure = "distance")
```

```{r, fig.height = 12, fig.width = 8}
## A function for identifying and plotting the top explorer/elaborators names (for checking)
identify.tops <- function(scores, data, prob = 0.99, plot = FALSE, ...) {
    ## Get the tops
    tops <- which(scores >= quantile(scores, prob = prob))
    ## Get their names
    top_names <- rownames(data)[tops]
    if(plot) {
        text(data[tops,], labels = top_names, ...)
    }
    return(top_names)
}

par(mfrow = c(3, 2))
plot(spaces_list$psittaciformes[, 1:2], col = "blue", main = "Psittaciformes section of the shape space", xlab = "PC1 (unscaled)", ylab = "PC2 (unscaled)", pch = 19)

arrows(x0 = psita_vector[1,1], y0 = psita_vector[1,2],
       x1 = psita_vector[2,1], y1 = psita_vector[2,2], col = "black", lwd = 4)

plot(spaces_list$charadriiformes[, 1:2], col = "orange", main = "Psittaciformes section of the shape space", xlab = "PC1 (unscaled)", ylab = "PC2 (unscaled)", pch = 19)

arrows(x0 = chara_vector[1,1], y0 = chara_vector[1,2],
       x1 = chara_vector[2,1], y1 = chara_vector[2,2], col = "black", lwd = 4)

## The elaboration/exploration profiles
plot(psita_proj, psita_reje, pch = 19, col = "blue",
    xlab = "Elaboration", ylab = "Exploration", main = "Psittaciformes explo/elaboration profile")
plot(chara_proj, chara_reje, pch = 19, col = "orange",
    xlab = "Elaboration", ylab = "Exploration", main = "Charadriiformes explo/elaboration profile")

## The elaboration/exploration profiles
plot(abs(psita_proj), psita_reje, pch = 19, col = "blue",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Psittaciformes explo/elaboration profile")

plot(abs(chara_proj), chara_reje, pch = 19, col = "orange",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Charadriiformes explo/elaboration profile")
```

```{r, print = FALSE}
## Making a top table
psita_tops_proj <- identify.tops(score = abs(psita_proj), data = spaces_list$psittaciformes)
psita_tops_reje <- identify.tops(score = psita_reje, data = spaces_list$psittaciformes)
chara_tops_proj <- identify.tops(score = abs(chara_proj), data = spaces_list$charadriiformes)
chara_tops_reje <- identify.tops(score = chara_reje, data = spaces_list$charadriiformes)

## Making a top table
psita_top_proj <- identify.tops(score = abs(psita_proj), data = spaces_list$psittaciformes, prob = 1)
psita_top_reje <- identify.tops(score = psita_reje, data = spaces_list$psittaciformes, prob = 1)
chara_top_proj <- identify.tops(score = abs(chara_proj), data = spaces_list$charadriiformes, prob = 1)
chara_top_reje <- identify.tops(score = chara_reje, data = spaces_list$charadriiformes, prob = 1)
```

The top 99% elaborators are:

 * parrots: `r psita_tops_proj`

 * sea gulls: `r chara_tops_proj`

The top 99% explorers are:

 * parrots: `r psita_tops_reje`

 * sea gulls: `r chara_tops_reje`

With our tops for each category:

<img title="Top elaborator parrot" src="https://upload.wikimedia.org/wikipedia/commons/4/43/Probosciger_aterrimus-20030511B.jpg">

<img title="Top explorer parrot" src="https://upload.wikimedia.org/wikipedia/commons/thumb/1/13/Kea_on_rock_while_snowing.jpg/1200px-Kea_on_rock_while_snowing.jpg">

<img title="Top elaborator sea gull" src="https://upload.wikimedia.org/wikipedia/commons/6/62/Giant_Snipe%2C_Gallinago_undulata.jpg">

<img title="Top elaborator sea gull" src="https://www.wildlifenorthamerica.com/image/Bird/horned-puffin.jpg">

Does this kind of works?
Also this is nothing major, different or special compared to just picking species based on PC scores.
The minor bonus here are:

 * it can make the data slightly more interpretable if the base vector is well defined (e.g. in colour space this vector can be the "pure" color vector, etc...).
 * it's a useful flattening of a multidimensional space: although the first 2 PC in this dataset illustrate 85.1% of the variance, it's still missing 228 other dimensions (including the "duck axis"!); also in palaeo data where PC1 and 2 bearly make up 1% of the data that's a much better tool for visualising everything.

I guess the key thing is determining the exploration/elaboration vector.
We can add some evo perspective to all of these by making the base vector "evolutionary" and by looking at elaboration/exploration through time!

### 2021-03-25 Analysis/05-mcmcglmm_analyses.Rmd
Create 05-mcmcglmm_analyses.Rmd

---
title: "05 - MCMCglmm analyses"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Elaboration/exploration analyses

```{r}
## Loading the packages and data
library(dispRity)
load(file = "../Data/Processed/disparity_matrices.rda")
load(file = "../Data/Processed/trees_list.rda")
load(file = "../Data/Processed/spaces_list.rda")
load(file = "../Data/Processed/tmp_simple_matrices.rda")
```

The whole discussion from last meeting was on how to determine the axis of elaboration to measure the elaboration/exploration metrics.

Here I am going to explore the major axis from the var-covar matrix from (`MCMCglmm::...()$VCV)` and do this TODO list:

 - [ ] rerun part 1 to 4 using the beak PCA *with* centroid size (`"2020_08_07_MMB_MORPHO_FORMSPACE_FULL.rds"`)
 - [ ] Use the 95% axis from the mcmcglmm vcv matrix
 - [ ] Use the 95% axis from a vcv matrix from a MANOVA (h-matrix/e-matrix)
 - [ ] Use the x axis from the Phylo-PCA or a Phylo aligned PCA (`geomorph::gm.prcomp()`; see section 3 and 3(4) here https://cran.r-project.org/web/packages/geomorph/vignettes/geomorph.PCA.html)

### 2021-03-25 Analysis/05-mcmcglmm_analyses.Rmd
Update 05-mcmcglmm_analyses.Rmd

@@
  - [ ] rerun part 1 to 4 using the beak PCA *with* centroid size (`"2020_08_07_MMB_MORPHO_FORMSPACE_FULL.rds"`)
- - [ ] Use the 95% axis from the mcmcglmm vcv matrix
+ - [ ] Use the 95% axis from the mcmcglmm vcv matrix (check [here](https://github.com/andbeck/mcmc-plus-tensor))
  - [ ] Use the 95% axis from a vcv matrix from a MANOVA (h-matrix/e-matrix)
- - [ ] Use the x axis from the Phylo-PCA or a Phylo aligned PCA (`geomorph::gm.prcomp()`; see section 3 and 3(4) here https://cran.r-project.org/web/packages/geomorph/vignettes/geomorph.PCA.html)
+ - [ ] Use the x axis from the Phylo-PCA or a Phylo aligned PCA (`geomorph::gm.prcomp()`; see section 3 and 3(4) [here](https://cran.r-project.org/web/packages/geomorph/vignettes/geomorph.PCA.html))

### 2021-03-29 Analysis/05-mcmcglmm_analyses.Rmd
added different alignements PC analyses
Needs to add MCMglmm

@@
  - [ ] Use the x axis from the Phylo-PCA or a Phylo aligned PCA (`geomorph::gm.prcomp()`; see section 3 and 3(4) [here](https://cran.r-project.org/web/packages/geomorph/vignettes/geomorph.PCA.html))
+
+# Testing with the plethodon dataset
+
+## Normal PCA
+
+```{r}
+library(geomorph)
+
+## Loading the landmarks and the tree
+data(plethspecies)
+
+## For the sake of visualisation I'm just going to rename everything from A to I
+dimnames(plethspecies$land)[[3]] <- plethspecies$phy$tip.label <- LETTERS[1:9]
+
+## Procrustes blabla
+procrustes <- gpagen(plethspecies$land, print.progress = F)
+tree <- plethspecies$phy
+```
+
+```{r}
+## PCA (with plotted tree)
+pca <- gm.prcomp(procrustes$coords, phy = tree)
+plot(pca, phylo = TRUE, main = "PCA trait space\n(tree and nodes are just plotted)")
+```
+
+This is just the normal PCA, nothing special here.
+
+## Phylo corrected PCA
+
+```{r}
+## Phylo PCA
+GLS_pca <- gm.prcomp(procrustes$coords, phy = tree, GLS = TRUE)
+plot(GLS_pca, phylo = TRUE, main = "PCA trait space\n(with phylo correction)")
+```
+
+This one is following [Revell 2009](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2009.00804.x) procedure that is basically applying a phylogenetic correction to the data while ordinating it.
+I think this is the one that doesn't make too much sense (see the story of doing "phylo corrected log").
+But I need to double check.
+
+## Phylo aligned PCA
+
+```{r}
+## Phylo aligned PCA
+Pa_pca <- gm.prcomp(procrustes$coords, phy = tree, align.to.phy = TRUE)
+plot(Pa_pca, phylo = TRUE, main = "PCA trait space\n(with phylo aligned)")
+```
+
+This one is the new [Collyer and Adams 2020](https://www.researchgate.net/profile/Michael-Collyer-4/publication/344900621_Phylogenetically_Aligned_Component_Analysis/links/5fa4123b458515157bec1bcb/Phylogenetically-Aligned-Component-Analysis.pdf) method where the phylogenetic signal is "distilled in the first PC axes".
+I need to double check that one as well.
+
+## MCMCglmm
+
+```{r}
+print("ha")
+```
+
+Here we used [Robinson and Beckerman 2013](https://onlinelibrary.wiley.com/doi/abs/10.1111/ele.12047) to get the VCV matrix that we can use to get the base vector.
+
+# Comparing the projection rejection profiles
+
+For each method we can measure the projection and rejections using the following base vectors:
+
+ - [ ] The 95% quantile of the PCA
+ - [ ] The MCMCglmm main axis on the PCA
+ - [ ] The first axis (range unit-scaled) of the PCA
+ - [ ] The first axis (range unit-scaled) of the GLS PCA
+ - [ ] The first axis (range unit-scaled) of the Pa PCA
+
+```{r}
+## Utility functions
+project.reject <- function(data, vector, absolute = FALSE) {
+
+    ## Get the projection
+    projection <- projections(
+          matrix = data, 
+          point1 = vector[1,],
+          point2 = vector[2,],
+          measure = "position")
+
+    ## Get the rejection
+    rejection <- projections(
+          matrix = data, 
+          point1 = vector[[1]],
+          point2 = vector[[2]],
+          measure = "distance")
+
+    ## Combine and out
+    if(absolute) {
+      results <- cbind(abs(projection), rejection)
+    } else {
+      results <- cbind(projection, rejection)
+    }
+    rownames(results) <- rownames(data)
+    return(results)
+}
+plot.project.reject <- function(results, xlab = "Elaboration", ylab = "Exploration", col = "black", main = "Project/Reject") {
+
+    pchs <- 19
+    colours <- "black"
+
+    plot(results, xlab = xlab, ylab = ylab, col = colours, pch = pchs, main = main, cex = 0.2)
+    text(results, rownames(results), pos = 2)
+}
+plot.ord <- function(ordination, main, vector) {
+    plot(ordination[, c(1:2)], pch = 19, cex = 0.2, main = main)
+    text(ordination[, c(1:2)], rownames(ordination), pos = 2)
+    lines(vector[, c(1:2)], lwd = 2)
+}
+```
+
+```{r, fig.height = 6, fig.width = 12}
+par(mfrow = c(1,2))
+## The 95% quantile of the PCA
+vector <- apply(pca$x, 2, quantile, prob = c(0.025, 0.975))
+results <- project.reject(pca$x, vector)
+
+plot.ord(pca$x, vector, main = "PCA")
+plot.project.reject(results, main = "Vector: pca 95%")
+```
+
+```{r}
+# par(mfrow = c(1,2))
+# ## The MCMCglmm main axis on the PCA
+# vector <- rbind("point1" = rep(0, ncol(Pa_pca$x)),
+#                 "point2" = c(max(Pa_pca$x[,1]),
+#                              rep(0, ncol(Pa_pca$x)-1)))
+# results <- project.reject(Pa_pca$x, vector)
+
+# plot.ord(Pa_pca$x, vector, main = "Pa PCA")
+# plot.project.reject(results, main = "Vector: Pa pca PC1")
+
+print("ha")
+```
+
+```{r, fig.height = 6, fig.width = 12}
+par(mfrow = c(1,2))
+## The first axis (range unit-scaled) of the PCA
+vector <- rbind("point1" = rep(0, ncol(pca$x)),
+                "point2" = c(max(pca$x[,1]),
+                             rep(0, ncol(pca$x)-1)))
+results <- project.reject(pca$x, vector)
+
+plot.ord(pca$x, vector, main = "PCA")
+plot.project.reject(results, main = "Vector: PC1")
+```
+
+```{r}
+par(mfrow = c(1,2))
+## The first axis (range unit-scaled) of the GLS PCA
+vector <- rbind("point1" = rep(0, ncol(GLS_pca$x)),
+                "point2" = c(max(GLS_pca$x[,1]),
+                             rep(0, ncol(GLS_pca$x)-1)))
+results <- project.reject(GLS_pca$x, vector)
+
+plot.ord(GLS_pca$x, vector, main = "GLS PCA")
+plot.project.reject(results, main = "Vector: GLS pca PC1")
+```
+
+```{r}
+par(mfrow = c(1,2))
+## The first axis (range unit-scaled) of the Pa PCA
+vector <- rbind("point1" = rep(0, ncol(Pa_pca$x)),
+                "point2" = c(max(Pa_pca$x[,1]),
+                             rep(0, ncol(Pa_pca$x)-1)))
+results <- project.reject(Pa_pca$x, vector)
+
+plot.ord(Pa_pca$x, vector, main = "Pa PCA")
+plot.project.reject(results, main = "Vector: Pa pca PC1")
+```

### 2021-04-01 Analysis/05-mcmcglmm_analyses.Rmd
Update 05-mcmcglmm_analyses.Rmd

@@
 tree <- plethspecies$phy
+## Making the tree ultrametric
+tree$edge.length[which(tree$edge[,2] == 7)] <- tree$edge.length[which(tree$edge[,2] == 7)] + 0.000001
 ```
@@
 
-```{r}
+```{r, eval = FALSE, print = FALSE}
+library(MCMCglmm)
+## Running a MCMCglmm model on with all the traits in the PCA
+morpho <- as.data.frame(pca$x)
+mcmcglmm_data <- data.frame(pca$x, animal = rownames(pca$x))
+tree_inverse <- inverseA(tree)$Ainv
+priors <- list(R = list(V = 1, nu = 0.002), G = list(G1 = list(V = 1, nu = 0.002)))
+
+phenotype <- pca$x[,1, drop = FALSE]
+
+# simulate phylogenetic and residual effects with unit variance
+test.data <- data.frame(phenotype = phenotype, animal = row.names(phenotype))
+
+model1 <- MCMCglmm(fixed = Comp1 ~ 1, random = ~animal, pedigree = tree,
+ data = test.data, prior = priors, verbose = TRUE, nitt = 1300, burnin = 300, thin = 1)
+
+test <- MCMCglmm(fixed = cbind(morpho$Comp1, morpho$Comp2) ~ Comp-1, random = ~species, pedigree = tree, prior = priors, data = morpho, verbose = TRUE, nitt = 1300, burnin = 300, thin = 1)
+
+data(bird.families)
+
+phylo.effect <- rbv(bird.families, 1, nodes="TIPS") 
+phenotype <- phylo.effect + rnorm(dim(phylo.effect)[1], 0, 1)
+
+plot(model2$VCV)
+
+pca$x ~ pca$x[,1]-1, random = ~
+family = c(rep("gaussian"), ncol(pca$x))
+
+formula = 
+MCMCglmm::MCMCglmm(fixed = formula, random = mulTree.data$random.terms, pedigree = mulTree.data$phy[[tree]], prior = priors, data = mulTree.data$data, verbose = FALSE, nitt = parameters[1], thin = parameters[2], burnin = parameters[3])
+
+model1.1<-MCMCglmm(cbind(trait1, trait2, trait3, trait4, trait5)  ~ trait-1, random=~us(trait):sire, rcov=~us(trait):units, family=c("gaussian","gaussian","gaussian","gaussian", "gaussian"), prior=prior1, data=data1[[1]], verbose=FALSE, burnin=50000, thin=100, nitt=150000)
+
 print("ha")

### 2021-04-05 Analysis/05-mcmcglmm_analyses.Rmd
Update 05-mcmcglmm_analyses.Rmd

--- 2021-03-29 Analysis/05-mcmcglmm_analyses.Rmd
@@
 
-```{r}
-print("ha")
+```{r, eval = TRUE, print = TRUE, echo = FALSE}
+library(MCMCglmm)
+
+## Add the animal and clade (which is more like a clade + a grade) column to data
+mcmcglmm_data <- as.data.frame(pca$x)
+mcmcglmm_data <- cbind(mcmcglmm_data,
+                       clade = as.factor(c(rep("clade", 5), rep("grade", 4))),
+                       animal = rownames(pca$x))
+## Selecting the traits that have a non zero variance (0 =< 1e-6: for some reason that's what MCMCglmm likes: TODO:fix: see bugs with Choleski decomposition and rounding estimates errors in MCMCglmm)
+select_traits <- apply(pca$x, 2, var) > 2e-5
+
+## TEST WITH ONLY THREE TRAITS
+select_traits <- c(1,1)
+
+## MCMC params
+MCMCparams <- list(nitt   = 110000,
+                   burnin = 10000,
+                   thin   = 100)
+
+## Making the fixed terms formula
+#TODO: do some wrapping to get all the variable names automatically
+formula_noclade = cbind(Comp1, Comp2) ~ trait-1
+formula_clade = cbind(Comp1, Comp2) ~ trait:clade-1
+
+## Base model
+#GT: The simplest model simply estimates the overal covariance matrix for the whole data set
+## The base model residuals
+base_residuals <- ~ us(trait):units
+## Running the base model
+model_base <- MCMCglmm(fixed  = formula_noclade,
+                       family = rep("gaussian", sum(select_traits)),
+                       rcov   = base_residuals,
+                       data   = mcmcglmm_data,
+                       nitt   = MCMCparams$nitt,
+                       burnin = MCMCparams$burnin,
+                       thin   = MCMCparams$thin)
+
+## Clade model
+#GT: The next model, still without phylogeny, involves modelling clade effect.
+Note that we are using clade both in the fixed effects and in the random effects.
+The random effects relate to covariances and main effects to means
+## The residuals for the clade model
+#TODO: generate automatically from the levels of the data
+clade_residuals <- ~ us(at.level(clade,1):trait):units + us(at.level(clade,2):trait):units
+## The clade model
+model_clade <- MCMCglmm(fixed  = formula_clade,
+                        family = rep("gaussian", sum(select_traits)),
+                        rcov   = clade_residuals,
+                        data   = mcmcglmm_data,
+                        nitt   = MCMCparams$nitt,
+                        burnin = MCMCparams$burnin,
+                        thin   = MCMCparams$thin)
+
+## Phylogeny model
+#GT: Now we will add phylogeny.
+To begin with we will not include the clade (gull, plover, sandpiper) effect and instead just partition covariances into phylogenetic and non-phylogenetic effects.
+At this point the model is becoming more complex and we need to specific a prior.
+## Priors for the phylo model
+#TODO: select the priors
+nu <- 0.002
+prior_phylo <- list(R = list(R1 = list(V = diag(sum(select_traits)), nu = nu)),
+                    G = list(G1 = list(V = diag(sum(select_traits)), nu = nu)))
+## The random terms for the phylo model
+random_phylo <- ~ us(trait):animal
+## The phylo model
+#TODO: this is prone to crash due to random samples in the MCMC (I think it's rounding problems with the Cholesky decomposition)
+model_phylo <- MCMCglmm(fixed = formula_noclade,
+                        family = rep("gaussian", sum(select_traits)),
+                        random = random_phylo,
+                        rcov   = base_residuals,
+                        prior  = prior_phylo,
+                        pedigree = tree,
+                        data   = mcmcglmm_data,
+                        nitt   = MCMCparams$nitt,
+                        burnin = MCMCparams$burnin,
+                        thin   = MCMCparams$thin)
+
+## Phylo and clade model
+#GT: Now we will add phylogeny with the clade (gull, plover, sandpiper) effect.
+This model has a single overall phylogenetic effects and clade specific non-phylogenetic effects.
+Note the changes to the prior specification as well as the main model.
+This is the model that I would suggest using to test for changes in clade covariances relative to phylogeny to test which clades innovate/elaborate
+
+#TODO: make the priors choice
+prior_phylo_clade <- list(R = list(R1 = list(V = diag(sum(select_traits)), nu = nu),
+                                   R2 = list(V = diag(sum(select_traits)), nu = nu)), 
+                          G = list(G1 = list(V = diag(sum(select_traits)), nu = nu)))
+
+## The phylo+clade model
+model_phy_clade <- MCMCglmm(fixed  = formula_clade,
+                            family = rep("gaussian", sum(select_traits)),
+                            random = random_phylo,
+                            rcov   = clade_residuals,
+                            prior  = prior_phylo_clade,
+                            pedigree = tree,
+                            data   = mcmcglmm_data,
+                            nitt   = MCMCparams$nitt,
+                            burnin = MCMCparams$burnin,
+                            thin   = MCMCparams$thin)
+
+## Phylo_clade and clade
+#GT: Now we will add phylogeny with phylogenetic and non-phylogenetic clade (gull, plover, sandpiper) effects.
+This model has a single overall phylogenetic effects and clade specific non-phylogenetic effects.
+Note the changes to the prior specification as well as the main model.
+This is the model that I would suggest using to test for changes in clade covariances relative to phylogeny to test which clades innovate/elaborate
+
+## The phylo_clade random terms
+random_phylo_clade <- ~ us(at.level(clade,1):trait):animal + us(at.level(clade,2):trait):animal
+
+#TODO: make these priors automatic (again!)
+prior_phy_cla_cla <- list(R = list(R1 = list(V = diag(sum(select_traits)), nu = nu),
+                                   R2 = list(V = diag(sum(select_traits)), nu = nu)), 
+                          G = list(G1 = list(V = diag(sum(select_traits)), nu = nu),
+                                   G2 = list(V = diag(sum(select_traits)), nu = nu)))
+
+## The phylo_clade + clade model
+model_phy_clade <- MCMCglmm(fixed  = formula_clade,
+                            family = rep("gaussian", sum(select_traits)),
+                            random = random_phylo_clade,
+                            rcov   = clade_residuals,
+                            prior  = prior_phy_cla_cla,
+                            pedigree = tree,
+                            data   = mcmcglmm_data,
+                            nitt   = MCMCparams$nitt,
+                            burnin = MCMCparams$burnin,
+                            thin   = MCMCparams$thin)
 ```
@@
 
-```{r}
-# par(mfrow = c(1,2))
-# ## The MCMCglmm main axis on the PCA
-# vector <- rbind("point1" = rep(0, ncol(Pa_pca$x)),
-#                 "point2" = c(max(Pa_pca$x[,1]),
-#                              rep(0, ncol(Pa_pca$x)-1)))
-# results <- project.reject(Pa_pca$x, vector)
-
-# plot.ord(Pa_pca$x, vector, main = "Pa PCA")
-# plot.project.reject(results, main = "Vector: Pa pca PC1")
-
-print("ha")
-```
-
 ```{r, fig.height = 6, fig.width = 12}
@@
 ```
+
+```{r}
+# par(mfrow = c(1,2))
+# ## The MCMCglmm main axis on the PCA
+# vector <- rbind("point1" = rep(0, ncol(Pa_pca$x)),
+#                 "point2" = c(max(Pa_pca$x[,1]),
+#                              rep(0, ncol(Pa_pca$x)-1)))
+# results <- project.reject(Pa_pca$x, vector)
+
+# plot.ord(Pa_pca$x, vector, main = "Pa PCA")
+# plot.project.reject(results, main = "Vector: Pa pca PC1")
+
+print("ha")
+```

### 2021-04-06 Analysis/get.covar.R
added simplify option

@@ -5,6 +5,7 @@
 #' @param data The MCMCglmm object
 #' @param n The number of samples
 # @param levels Optional, the level(s) to get the covariance from (if missing, all levels are returned)
+#' @param simplify logical, whether to output the results as a matrix of lists (simplify = FALSE) where rows are the levels and columns are the replicates or whether to return it as a list containing n levels elements each containing the replicates (simplify = TRUE).
 #' 
 #' @examples
 #'
@@ -13,34 +14,18 @@
 #' @author Thomas Guillerme, Gavin Thomas
 #' @export
 
-get.covar <- function(data, n){ #, levels) { #TG: levels not handled yet: returns everything
+get.covar <- function(data, n, simplify = TRUE){ #, levels) { #TG: levels not handled yet: returns everything
     
     ## The number of traits
     n_traits <- length(strsplit(paste(data$Fixed$formula)[2], ",")[[1]])  
-    ## The number of levels
+    ## The number of levels
     n_levels <- c(
       ## The number of random effects
       "random" = length(grep("animal", colnames(data$VCV)))/(n_traits^2),
       ## The number of groups (residuals)
       "residual" = length(grep("units", colnames(data$VCV)))/(n_traits^2))
 
-    ## Sample n covar matrices
-    covar_matrices <- replicate(n, get.one.covar(data, n_levels, n_traits), simplify = TRUE)
-
-    ## Function for recombining the "rows" in the output from replicate() (into a list of lists)
-    combine.replicates <- function(replicates) {
-        return(list(
-            ## Get all the VCV
-            VCV = lapply(replicates, function(x) return(x$VCV)),
-            ## Get all the Sol
-            Sol = lapply(replicates, function(x) return(x$Sol))
-        ))
-    }
-
-    ## Recombine all the replicates
-    covar_matrices <- apply(covar_matrices, 1, combine.replicates)
-
-    ## Add the names of the levels
+    ## Get the names of the levels
     param_names <- character()
     if(n_levels["random"] != 0) {
         if(n_levels["random"] > 1) {
@@ -52,12 +37,39 @@ get.covar <- function(data, n){ #, levels) { #TG: levels not handled yet: return
     if(n_levels["residual"] != 0) {
        param_names <- c(param_names, unique(unlist(lapply(strsplit(colnames(data$Sol), ":"), `[[`, 2))))
     }
-    ## Naming the list
-    if(length(param_names) == length(covar_matrices)) {
-        names(covar_matrices) <- param_names
-    }
 
-    return(covar_matrices)
[…]

### 2021-04-06 Analysis/get.ellipse.R
Plotting functions

@@ -1,11 +1,12 @@
-#' @title Get ellipse
+#' @title Plot ellipses
 #'
-#' @description Get the coordinates on n ellipses out of an MCMCglmm object 
+#' @description Plot n ellipses out of an MCMCglmm object 
 #'
 #' @param data the MCMCglmm data
 #' @param n the number of ellipses (default is 1)
+#' @param dimensions in which dimensions to plot the ellipses
 #' @param levels which level to get the ellipses from
-#' @param points the number of points to draw from the ellipse (default is 5)
+#' @param points the number of points to draw from the ellipse (default is 50)
 #' @param centre
 #' 
 #' @examples
@@ -32,17 +33,84 @@
 #' @author Thomas Guillerme, Gavin Thomas
 #' @export
 
-my_fun <- function() {
+plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.transparent = TRUE, ...) {
 
-    return()
+    ## Get the covariance matrices
+    covar_matrices <- get.covar(data, n = n, simplify = FALSE)
+
+    ## Get all the ellipses
+    all_ellipses <- apply(covar_matrices,1, get.all.ellipses, dimensions, npoints)
+
+    ## Get the plot arguments
+    plot_args <- list(x = NULL, y = NULL, ...)
+    if(is.null(plot_args$xlab)) {
+        plot_args$xlab  <- paste("Dim.", dimensions[1])
+    }
+    if(is.null(plot_args$ylab)) {
+        plot_args$ylab  <- paste("Dim.", dimensions[2])
+    }
+    if(is.null(plot_args$lwd)) {
+        plot_args$lwd <- 1
+    }
+    if(is.null(plot_args$lty)) {
+        plot_args$lty <- 1
+    }
+    if(is.null(plot_args$xlim)) {
+        plot_args$xlim <- range(unlist(all_ellipses))
+    }
+    if(is.null(plot_args$ylim)) {
+        plot_args$ylim <- range(unlist(all_ellipses))
+    }
+    if(is.null(plot_args$main)) {
+        plot_args$main <- "Correlation matrices"
+    }
+
+    ## Handle colours
+    if(missing(col)) {
+        all_cols <- rainbow(length(all_ellipses))
+    } else {
+        if(missing_col <- length(all_ellipses) - length(col) > 0) {
+            all_cols <- rep(col, missing_col)
+        } else {
+            all_cols <- col
+        }
+    }
+    if(use.transparent) {
[…]

### 2021-04-06 Analysis/sandbox.R
TODO

@@ -1,0 +1,9 @@
+
+#TODO:
+list_of_matrices <- get.covar(data, n)
+plot.ellipses(list_of_matrices)
+
+compare.matrices() # comparing the VCV matrices Robinson and Beckerman style
+
+get.major.axis.coords(list_of_matrices, axis = 1) # Getting the list of major axis (get the eigen value from it?)
+

### 2021-04-07 Functions/plot.ellipses.R
rearranged files

@@ -1,0 +1,116 @@
+#' @title Plot ellipses
+#'
+#' @description Plot n ellipses out of an MCMCglmm object 
+#'
+#' @param data the MCMCglmm data
+#' @param n the number of ellipses (default is 1)
+#' @param dimensions in which dimensions to plot the ellipses
+#' @param levels which level to get the ellipses from
+#' @param points the number of points to draw from the ellipse (default is 50)
+#' @param centre
+#' 
+#' @examples
+#' \dontrun{
+#' ## Some MCMCglmm object
+#' data <- ...
+#' 
+#' ## Get the coordinates of one ellipse (defining the major and minor axes)
+#' ellipse_coords <- get.ellipse(data)
+#' major_axis <- 
+#' minor_axis
+#' 
+#' ## Get 100 random ellipses from the data (of 50 points each)
+#' ellipses_coords <- get.ellipse(data, n = 100, points = 50)
+#' 
+#' ## Plot the first ellipse
+#' plot(ellipses_coords[[1]], type = "l")
+#' ## And the subsequent ones
+#' do.call(lines, ellipses_coords[-1])
+#' }
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme, Gavin Thomas
+#' @export
+
+plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.transparent = TRUE, ...) {
+
+    ## Get the covariance matrices
+    covar_matrices <- get.covar(data, n = n, simplify = FALSE)
+
+    ## Get all the ellipses
+    all_ellipses <- apply(covar_matrices,1, get.all.ellipses, dimensions, npoints)
+
+    ## Get the plot arguments
+    plot_args <- list(x = NULL, y = NULL, ...)
+    if(is.null(plot_args$xlab)) {
+        plot_args$xlab  <- paste("Dim.", dimensions[1])
+    }
+    if(is.null(plot_args$ylab)) {
+        plot_args$ylab  <- paste("Dim.", dimensions[2])
+    }
+    if(is.null(plot_args$lwd)) {
+        plot_args$lwd <- 1
+    }
+    if(is.null(plot_args$lty)) {
+        plot_args$lty <- 1
+    }
+    if(is.null(plot_args$xlim)) {
+        plot_args$xlim <- range(unlist(all_ellipses))
+    }
+    if(is.null(plot_args$ylim)) {
+        plot_args$ylim <- range(unlist(all_ellipses))
+    }
+    if(is.null(plot_args$main)) {
[…]

### 2021-04-07 Analysis/06-post_mcmcglmm_analyses.Rmd
post mcmc analyses init

---
title: "Post MCMCglmm analyses"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

In this script I'll be looking at how to analyse the outputs from the MCMCglmm models to get the something about the elaboration/exploration stories.

I'm going to look at it using two approaches:
 * the **blob-wise** one where I'll look at the differences between whole blobs (ellipses) in terms of elaboration/exploration (Robinson & Beckerman style);
 * and the **tip-wise** one where I'll look at the elaboration/exploration score of the tips on the tree (cf. their blobs) relative to the whole phylogeny and to their respective blobs (clades).

## Data

I'm going to focus on the data from Gavin and especially the models 5 and 6 that are:

 * model 6: `model_phylo1_clade3` with one phylogenetic random effect (across the whole tree) and three clade residual effects.
 * model 7: `model_phylo3_clade3` with three phylogenetic random effect (one for each clade) and three clade residual effects.

```{r}
## Loading the correct models
load("../Data/Processed/model_list.rda")
model_phylo1_clade3 <- model_list[[5]]
model_phylo3_clade3 <- model_list[[6]]
```

## Blob-wise approach

Here we're basically comparing multidimensional ellipses (here 3D ones but I think we should push it to more dimensions!).

```{r}

```

## Clade-wise approach

```{r}

```

```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
```

### 2021-04-07 Functions/plot.ellipses.R
centre fixes

@@ -5,9 +5,13 @@
 #' @param data the MCMCglmm data
 #' @param n the number of ellipses (default is 1)
 #' @param dimensions in which dimensions to plot the ellipses
-#' @param levels which level to get the ellipses from
-#' @param points the number of points to draw from the ellipse (default is 50)
-#' @param centre
+#' @param col a series of colours by levels
+#' @param use.transparent whether to make the colours semi transparent (TRUE; default) or fully opaque (FALSE)
+#' @param npoints the number of points to draw from the ellipse (default is 50)
+#' @param centre either "zero" to centre all the ellipses on 0 (default), "level" to centre the ellipses on their average centre per level, "none" to not centre the ellipses.
+#' 
+#' logical whether to centre the ellipses (TRUE, default) or not (FALSE) or a list of centring coordinates
+#' @param add logical whether to add the ellipses to an existing plot (TRUE) or not (FALSE, default)
 #' 
 #' @examples
 #' \dontrun{
@@ -33,13 +37,17 @@
 #' @author Thomas Guillerme, Gavin Thomas
 #' @export
 
-plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.transparent = TRUE, ...) {
+plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.transparent = TRUE, centre = "zero", add = FALSE, ...) {
 
     ## Get the covariance matrices
     covar_matrices <- get.covar(data, n = n, simplify = FALSE)
 
     ## Get all the ellipses
-    all_ellipses <- apply(covar_matrices,1, get.all.ellipses, dimensions, npoints)
+    # all_ellipses <- list()
+    # for(level in 1:nrow(covar_matrices)) {
+    #     all_ellipses[[level]] <- get.all.ellipses(covar_matrices[level, ], dimensions, npoints, centre = centres[[level]])
+    # }
+    all_ellipses <- apply(covar_matrices, 1, get.all.ellipses, dimensions, npoints, centre)
 
     ## Get the plot arguments
     plot_args <- list(x = NULL, y = NULL, ...)
@@ -82,8 +90,10 @@ plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.t
     }
 
     ## Empty plot
-    empty_plot <- plot_args
-    do.call(plot, empty_plot)
+    if(!add) {
+        empty_plot <- plot_args
+        do.call(plot, empty_plot)
[…]

### 2021-04-07 Analysis/06-post_mcmcglmm_analyses.Rmd
basic analyses

@@
   html_document:
-    fig_width: 12
-    fig_height: 6
+    fig_width: 8
+    fig_height: 8
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
+```
 
@@
 load("../Data/Processed/model_list.rda")
-model_phylo1_clade3 <- model_list[[5]]
+model_phylo1_clade3 <- model_list[[4]]
 model_phylo3_clade3 <- model_list[[6]]
 ```
+
+And here's what the trait space looks like:
+
+```{r}
+## Loading the data
+load("../Data/Processed/morphdat.rda")
+## Plotting it
+colour_vector <- c("orange", "blue", "darkgreen")
+plot(morphdat[, c(1,2)], pch = 19,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+legend("topleft", legend = levels(morphdat$clade), col = colour_vector, pch = 19)
+```
+
+Note that the PC% are from the whole PC in Gavin's example.
 
@@
 
+### Phylo + clade model
+
+First we can visualise some of these ellipses (100, randomly drawn from the posterior):
+
 ```{r}
-
+source("../Functions/plot.ellipses.R")
+source("../Functions/get.covar.R")
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+plot.ellipses(model_phylo1_clade3, n = 100,
+              centre = "none", add = TRUE, col = c("grey",colour_vector))
 ```
 
-## Clade-wise approach
+Because this is really messy, we can centre these ellipses on the groups ellipses average centres:
 
 ```{r}
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+plot.ellipses(model_phylo1_clade3, n = 100,
+              centre = "level", add = TRUE, col = c("grey",colour_vector))
+```
+<!-- ## Get the ellipses centres
+centres <- list(
+    "phylo" = colMeans(morphdat[, c(1,2)]),
+    "gulls" = colMeans(morphdat[which(morphdat$clade == "gulls"), c(1,2)]),
+    "plovers" = colMeans(morphdat[which(morphdat$clade == "plovers"), c(1,2)]),
+    "sandpipers" = colMeans(morphdat[which(morphdat$clade == "sandpipers"), c(1,2)]))
 
+ -->
+
+### Phylo clade + clade model
+
+```{r, warn = FALSE}
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+plot.ellipses(model_phylo3_clade3, n = 100,
+              centre = "none", add = TRUE, col = c("black", colour_vector, colour_vector))
 ```
 
-```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
+This looks much more messier (and also probably not scaled correctly?)...
+
+```{r}
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+plot.ellipses(model_phylo3_clade3, n = 100,
+              centre = "level", add = TRUE, col = c("black", colour_vector, colour_vector))
 ```
+
+## Tip-wise approach

### 2021-04-08 Functions/get.axes.R
draft

@@ -1,0 +1,71 @@
+#' @title Get ellipses axes
+#'
+#' @description Get the coordinates of any axes from an nD ellipse
+#'
+#' @param covar A covariance matrix describing the ellipse
+#' @param axis Which axis to get the coordinates from (default is 1 for the major axis).
+#' @param centre The centre of the ellipse (by default this is 0).
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+get.axes <- function(covar, axis = 1, centre = 0) {
+
+    ## Get the right size for the centre
+    if((diff <- dim(covar)[1]) - length(centre) != 0) {
+        centre <- c(centre, rep(centre[1], abs(diff)-1))
+    }
+
+    ## Get the major axis
+    lines(eigen(covar[c(1,2), c(1,2)])$vectors)
+
+    eigen.decomp <- eigen(covar[c(1,2), c(1,2)])
+
+
+    test <- eigen.decomp$vectors %*% sqrt(eigen.decomp$values)
+
+    lines(eigen.decomp$vectors * eigen.decomp$values)
+
+    # https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care
+
+    ## Solution:
+        # A <- matrix(c(20.43, -8.59,-8.59, 24.03), nrow = 2)
+
+        # E <- eigen(A, symmetric = TRUE)  ## symmetric eigen decomposition
+        # U <- E[[2]]  ## eigen vectors, i.e., rotation matrix
+        # D <- sqrt(E[[1]])  ## root eigen values, i.e., scaling factor
+
+        # r <- 1.44  ## radius of original circle
+        # Z <- rbind(c(r, 0), c(0, r), c(-r, 0), c(0, -r))  ## original vertices on major / minor axes
+        # Z <- tcrossprod(Z * rep(D, each = 4), U)  ## transformed vertices on major / minor axes
+
+        # #          [,1]      [,2]
+        # #[1,] -5.055136  6.224212
+        # #[2,] -4.099908 -3.329834
+        # #[3,]  5.055136 -6.224212
+        # #[4,]  4.099908  3.329834
+
+        # C0 <- c(-0.05, 0.09)  ## new centre
[…]

### 2021-04-08 Analysis/06-post_mcmcglmm_analyses.Rmd
expectations

@@
 
+We can then measure different stuff from these ellipses (e.g. going full statistics from Robinson & Beckerman).
+
+However, here I'm just going to focus on the elaboration/exploration of these ellipses. For the three clades, I'm going to measure:
+ * the angle of the clade's ellipse to the general one (which will be their degree of exploration);
+ * their scaled projection (where the ellipses all start from the point 0,0): showing how much the clade elaborates on the major phylo axis;
+ * and their scaled rejection (how much they explore).
+
+For these three metrics I expect:
+ * the gulls and plovers to elaborate quiet a lot (angle ~90) compared to the sandpipers (angle ~90)
+ * the gulls to explore the more followed by the plovers and then the sandpipers
+ * the sandpipers to ellaborate the more followed by the gulls and then the plovers
+
+Note that the three metrics are measured independently of the ellipse position in space and are measured in 3D.
+
+```{r}
+
+```
+
 ### Phylo clade + clade model

### 2021-04-09 Analysis/06-post_mcmcglmm_analyses.Rmd
Updated analyses (added axes bits)

@@
 source("../Functions/get.covar.R")
+
+## Selecting 100 random covariance matrices from the MCMCglmm
+covar_matrices <- get.covar(model_phylo1_clade3,
+                            n = 100,
+                            simplify = FALSE)
+
+## Plotting the results
 plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
@@
      ylab = "PC1 (6.86%)")
-plot.ellipses(model_phylo1_clade3, n = 100,
-              centre = "none", add = TRUE, col = c("grey",colour_vector))
+plot.ellipses(covar_matrices, centre = "none",
+              col = c("grey",colour_vector),
+              add = TRUE)
 ```
@@
      ylab = "PC1 (6.86%)")
-plot.ellipses(model_phylo1_clade3, n = 100,
-              centre = "level", add = TRUE, col = c("grey",colour_vector))
+plot.ellipses(covar_matrices, centre = "level",
+              col = c("grey",colour_vector),
+              add = TRUE)
 ```
-<!-- ## Get the ellipses centres
-centres <- list(
-    "phylo" = colMeans(morphdat[, c(1,2)]),
-    "gulls" = colMeans(morphdat[which(morphdat$clade == "gulls"), c(1,2)]),
-    "plovers" = colMeans(morphdat[which(morphdat$clade == "plovers"), c(1,2)]),
-    "sandpipers" = colMeans(morphdat[which(morphdat$clade == "sandpipers"), c(1,2)]))
-
- -->
 
@@
 ```{r}
+source("../Functions/get.axes.R")
+## Get all the phylo main axes
+level_centred_major_axes <- get.axes(covar_matrices, centre = "level")
+uncentered_major_axes <- get.axes(covar_matrices, centre = "none")
+```
 
+And this is what it looks like for the uncentred ones:
+
+```{r}
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+plot.ellipses(covar_matrices, centre = "none",
+              col = c("grey",colour_vector),
+              add = TRUE)
+plot.all.axes(uncentered_major_axes,
+         col = c("grey",colour_vector),
+         add = TRUE)
+```
+
+And for the centred ones:
+
+```{r}
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)")
+plot.ellipses(covar_matrices, centre = "level",
+              col = c("grey",colour_vector),
+              add = TRUE)
+plot.all.axes(level_centred_major_axes,
+         col = c("grey",colour_vector),
+         add = TRUE)
+```
+
+Or without all the points n stuff
+
+```{r, fig.height = 4, fig.width = 8}
+par(mfrow = c(1,2))
+plot.all.axes(uncentered_major_axes,
+         col = c("grey",colour_vector),
+         add = FALSE, main = "Major axes (un-centred)")
+plot.all.axes(uncentered_major_axes,
+         col = c("grey",colour_vector),
+         add = FALSE, main = "Major axes (level-centred)")
 ```

### 2021-04-09 Analysis/get.ellipse.R
reorganised functions

@@ -1,116 +1,0 @@
-#' @title Plot ellipses
-#'
-#' @description Plot n ellipses out of an MCMCglmm object 
-#'
-#' @param data the MCMCglmm data
-#' @param n the number of ellipses (default is 1)
-#' @param dimensions in which dimensions to plot the ellipses
-#' @param levels which level to get the ellipses from
-#' @param points the number of points to draw from the ellipse (default is 50)
-#' @param centre
-#' 
-#' @examples
-#' \dontrun{
-#' ## Some MCMCglmm object
-#' data <- ...
-#' 
-#' ## Get the coordinates of one ellipse (defining the major and minor axes)
-#' ellipse_coords <- get.ellipse(data)
-#' major_axis <- 
-#' minor_axis
-#' 
-#' ## Get 100 random ellipses from the data (of 50 points each)
-#' ellipses_coords <- get.ellipse(data, n = 100, points = 50)
-#' 
-#' ## Plot the first ellipse
-#' plot(ellipses_coords[[1]], type = "l")
-#' ## And the subsequent ones
-#' do.call(lines, ellipses_coords[-1])
-#' }
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme, Gavin Thomas
-#' @export
-
-plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.transparent = TRUE, ...) {
-
-    ## Get the covariance matrices
-    covar_matrices <- get.covar(data, n = n, simplify = FALSE)
-
-    ## Get all the ellipses
-    all_ellipses <- apply(covar_matrices,1, get.all.ellipses, dimensions, npoints)
-
-    ## Get the plot arguments
-    plot_args <- list(x = NULL, y = NULL, ...)
-    if(is.null(plot_args$xlab)) {
-        plot_args$xlab  <- paste("Dim.", dimensions[1])
-    }
-    if(is.null(plot_args$ylab)) {
-        plot_args$ylab  <- paste("Dim.", dimensions[2])
-    }
-    if(is.null(plot_args$lwd)) {
-        plot_args$lwd <- 1
-    }
-    if(is.null(plot_args$lty)) {
-        plot_args$lty <- 1
-    }
-    if(is.null(plot_args$xlim)) {
-        plot_args$xlim <- range(unlist(all_ellipses))
-    }
-    if(is.null(plot_args$ylim)) {
-        plot_args$ylim <- range(unlist(all_ellipses))
-    }
-    if(is.null(plot_args$main)) {
[…]

### 2021-04-09 Functions/plot.ellipses.R

@@ -2,8 +2,8 @@
 #'
 #' @description Plot n ellipses out of an MCMCglmm object 
 #'
-#' @param data the MCMCglmm data
-#' @param n the number of ellipses (default is 1)
+#' @param data the MCMCglmm data or a matrix or list of covar matrtices
+#' @param n if data is MCMCglmm, the number of ellipses (default is 1)
 #' @param dimensions in which dimensions to plot the ellipses
 #' @param col a series of colours by levels
 #' @param use.transparent whether to make the colours semi transparent (TRUE; default) or fully opaque (FALSE)
@@ -39,14 +39,16 @@
 
 plot.ellipses <- function(data, n, dimensions = c(1,2), npoints = 50, col, use.transparent = TRUE, centre = "zero", add = FALSE, ...) {
 
-    ## Get the covariance matrices
-    covar_matrices <- get.covar(data, n = n, simplify = FALSE)
+    if(is(data, "MCMCglmm")) {
+        ## Get the covariance matrices
+        covar_matrices <- get.covar(data, n = n, simplify = FALSE)
+    } else {
+        ## TODO: handle different input formats
+        covar_matrices <- data
+        n <- dim(covar_matrices)[2]
+    }
 
     ## Get all the ellipses
-    # all_ellipses <- list()
-    # for(level in 1:nrow(covar_matrices)) {
-    #     all_ellipses[[level]] <- get.all.ellipses(covar_matrices[level, ], dimensions, npoints, centre = centres[[level]])
-    # }
     all_ellipses <- apply(covar_matrices, 1, get.all.ellipses, dimensions, npoints, centre)
 
     ## Get the plot arguments

### 2021-04-09 Functions/get.axes.R
get.axes functions

@@ -5,6 +5,7 @@
 #' @param covar A covariance matrix describing the ellipse
 #' @param axis Which axis to get the coordinates from (default is 1 for the major axis).
 #' @param centre The centre of the ellipse (by default this is 0).
+#' @param level The confidence interval (default it 0.95).
 #' 
 #' @examples
 #'
@@ -12,60 +13,143 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-
-get.axes <- function(covar, axis = 1, centre = 0) {
+# 
+#
+# DEBUG
+# data <- model_phylo1_clade3
+# n <- 1
+# covar_matrices <- get.covar(data, n = n, simplify = FALSE)
+# covar <- covar_matrices[1,][[1]]$VCV
+get.one.axes <- function(covar, axis = 1, centre = 0, level = 0.95) {
 
     ## Get the right size for the centre
     if((diff <- dim(covar)[1]) - length(centre) != 0) {
         centre <- c(centre, rep(centre[1], abs(diff)-1))
     }
+    
+    ## Get the data dimensionality
+    dims <- dim(covar)[1]
 
-    ## Get the major axis
-    lines(eigen(covar[c(1,2), c(1,2)])$vectors)
+    ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
+    unit_hypersphere1 <- unit_hypersphere2 <- matrix(0, ncol = dims, nrow = dims)
+    ## The "front" (e.g. "top", "right") units
+    diag(unit_hypersphere1) <- 1
+    ## The "back" (e.g. "bottom", "left") units
+    diag(unit_hypersphere2) <- -1
+    unit_hypersphere <- rbind(unit_hypersphere1, unit_hypersphere2)
+    ## Scale the hypersphere (where level is the confidence interval)
+    unit_hypersphere <- unit_hypersphere * sqrt(qchisq(level, 2))
 
-    eigen.decomp <- eigen(covar[c(1,2), c(1,2)])
+    ## Do the eigen decomposition (symmetric - faster)
+    eigen_decomp <- eigen(covar, symmetric = TRUE)
 
+    ## Re-scaling the unit hypersphere
+    scaled_edges <- unit_hypersphere * rep(sqrt(eigen_decomp$values), each = dims*2)
+    ## Rotating the edges coordinates
+    edges <- tcrossprod(scaled_edges, eigen_decomp$vectors)
 
-    test <- eigen.decomp$vectors %*% sqrt(eigen.decomp$values)
[…]

### 2021-04-09 Analysis/06-post_mcmcglmm_analyses.Rmd
typo

@@
          add = FALSE, main = "Major axes (un-centred)")
-plot.all.axes(uncentered_major_axes,
+plot.all.axes(level_centred_major_axes,
          col = c("grey",colour_vector),

### 2021-04-09 Functions/get.axes.R
Note and example + link to page

@@ -20,16 +20,21 @@
 # n <- 1
 # covar_matrices <- get.covar(data, n = n, simplify = FALSE)
 # covar <- covar_matrices[1,][[1]]$VCV
-get.one.axes <- function(covar, axis = 1, centre = 0, level = 0.95) {
+get.one.axes <- function(covar, axis = 1, centre = 0, level = 0.95, dimensions) {
 
-    ## Get the right size for the centre
-    if((diff <- dim(covar)[1]) - length(centre) != 0) {
-        centre <- c(centre, rep(centre[1], abs(diff)-1))
-    }
-    
+    # The magic: https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331
+
+    ## Select the right dimensions
+    covar <- covar[dimensions, dimensions]
+
     ## Get the data dimensionality
     dims <- dim(covar)[1]
 
+    ## Get the right size for the centre
+    if((diff <- dims) - length(centre) != 0) {
+        centre <- c(centre, rep(centre[1], abs(diff)-1))
+    }
+    
     ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
     unit_hypersphere1 <- unit_hypersphere2 <- matrix(0, ncol = dims, nrow = dims)
     ## The "front" (e.g. "top", "right") units
@@ -51,30 +56,58 @@ get.one.axes <- function(covar, axis = 1, centre = 0, level = 0.95) {
     ## Move the matrix around
     edges <- edges + rep(centre, each = dims*2)
 
+    ## Get the longest axes with dist
+    distances <- apply(edges, 1, dist)[1:dims]
+    axis <- which(distances == sort(distances, decreasing = TRUE)[axis])
+
+    # stop("DEBUG")
+    # ## Weird bug with asp?
+    # plot(ellipse(covar), type = "l", asp = 1)
+    # points(edges)
+    # lines(edges[c(1,3),], lty = 2)
+    # lines(edges[c(2,4),], lty = 2)
+
+
     ## Get the edges coordinates
     return(edges[c(axis, axis+dims), ])
 }
 
 
+
+
 #TODO: document!
 #' @param data see plot.ellipses
 #' @param centre see plot.ellipses
-get.axes <- function(data, centre) {
+get.axes <- function(data, centre, n = 1, dimensions) {
     if(is(data, "MCMCglmm")) {
[…]

### 2021-04-09 Analysis/06-post_mcmcglmm_analyses.Rmd
Update 06-post_mcmcglmm_analyses.Rmd

@@
 
-Or without all the points n stuff
+Note that the major axes look slightly off for some of the ellipses.
+I think this is due to the difference in aspect ratio in both plots that exagerates vertical positions and deforms the ellipses:
+
+```{r}
+plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
+     col = colour_vector[morphdat$clade],
+     xlab = "PC1 (90.5%)",
+     ylab = "PC1 (6.86%)",
+     main = "Major axes of the ellipses\n(aspect ratio = 1)",xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))
+plot.ellipses(covar_matrices, centre = "level",
+              col = c("grey",colour_vector),
+              add = TRUE)
+plot.all.axes(level_centred_major_axes,
+         col = c("grey",colour_vector),
+         add = TRUE)
+```
+
+This is an example with just the major axes (projected in 2D):
 

### 2021-04-12 Analysis/06-post_mcmcglmm_analyses.Rmd
Update 06-post_mcmcglmm_analyses.Rmd

@@
 ## Tip-wise approach
+
+For the tip wise approach, the method is pretty straightforward from what I've done previously: we just measure the projection/rejection (elaboration/exploration) on each for each axis that we've drawn above.
+We thus end up with 100 projection/rejection score for each element (tip) in the tree for:
+
+ * Each element against the phylo axes
+ * Each element against their clade axes
+
+```{r}
+library(dispRity)
+## Wrapping function
+lapply.projections <- function(one_axis, trait_space) {
+    return(list(
+        projection = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,]),
+        rejection  = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,], measure = "distance")))
+}
+## Plot rejection/projection profiles
+plot.proj.rej <- function(point_list, cols, data, main) {
+    ## Colour vector
+    colour_vector <- cols[morphdat$clade]
+    colour_vector <- adjustcolor(colour_vector, alpha.f = 1/10)
+    ## lims
+    lim <- range(unlist(point_list))
+    plot(NULL, xlim = lim, ylim = lim, xlab = "projection", ylab = "rejection", main = main)
+    ## Add each points
+    plot.points <- function(one_set, colour_vector) {
+        points_args <- list()
+        points_args$pch <- 19
+        points_args$col <- colour_vector
+        points_args$x <- one_set$projection
+        points_args$y <- one_set$rejection
+        do.call(points, points_args)
+    }
+
+    lapply(point_list, plot.points, colour_vector)
+    return(invisible())
+}
+
+## Getting the results
+phylo_results_uncentered <- lapply(uncentered_major_axes$phylogeny, lapply.projections, trait_space = morphdat[, c(1,2,3)])
+
+## Plot results
+plot.proj.rej(phylo_results_uncentered, cols = colour_vector, data = morphdat, main = "Phylo elaboration/rejection")
+```
+
+But maybe it's easier to boxplot that to get the info by clade
+
+```{r}
+boxplot.proj.rej <- function(point_list, data, cols) {
+    ## Get the classifier
+    classifier <- data$clade    
+    sort.clade.rej <- function(data, classifier) {
+        return(cbind(data$rejection, classifier))
+    }
+    sort.clade.proj <- function(data, classifier) {
+        return(cbind(data$projection, classifier))
+    }
+    ## Sort the data per clade
+    rejections <- do.call(rbind, lapply(point_list, sort.clade.rej, classifier))
+    projections <- do.call(rbind, lapply(point_list, sort.clade.proj, classifier))
+    colnames(rejections)[1] <- colnames(projections)[1] <- "points"
+    par(mfrow = c(2,1))
+    boxplot(points ~ classifier, data = projections, col = cols, main = "Projections")
+    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Rejections")
+}
+
+boxplot.proj.rej(phylo_results_uncentered, cols = colour_vector, data = morphdat)
+```

### 2021-04-12 Analysis/06-post_mcmcglmm_analyses.Rmd
MCMC speed up idea

@@
 However, here I'm just going to focus on the elaboration/exploration of these ellipses. For the three clades, I'm going to measure:
+
  * the angle of the clade's ellipse to the general one (which will be their degree of exploration);
@@
 For these three metrics I expect:
+
  * the gulls and plovers to elaborate quiet a lot (angle ~90) compared to the sandpipers (angle ~90)
@@
 ```
+
+TODO:
+
+ - run the blob wise and tip wise models
+
+# Ideas for the MCMC
+
+ * Use the model7 with two nested clade levels?
+
+ * For speeding up the MCMC:
+   1. Check on a big global model: a) how long it takes to get 1000 gen. (or so) and b) how long the burnin phase takes.
+   2.
+Use info 1a and 1b to make design "mini chains" (i.e.
+chains that just go past the burnin and run for an extra ~1000 gen.); the conservative assumption is that post-burnin, the chain has reach a local optimum that exists within the distribution of the global (true) optimum.
+   3.
+Run 3 mini chains per tree on 100 trees (resulting in 300 mini chains) to get phylogenetic uncertainty into account.
+For each triplet of models, discard any that's way off (Natalie thesis style for morpho measurements).
+This results in N models that have reach N local optimums while taking into account phylo uncertainty.
+   4.
+Discard any of the N models that is way off (outliers).
+We assume that these outliers were chains stuck on local optimums that were outside to global optimum distribution.
+We then end up with M (M<=N) models that are the most conservative representation of the model because they take into account phylo uncertainty AND optimum uncertainty (i.e.
+if running one big model, you have to make the assumption that the local optimum reached equal the global optimum, here we can skip that assumption by saying that our distribution of local optimums contains the global optimum).
+   5.
+Combine the results of all the MCMC into one BIG posterior distribution; randomly sample n (=Mx5?) samples (or any other number) from that distribution and work from there!
+(if we don't discard any models M = 300 mini chains of >=1000 samples each, thus the sample represents ~5% of all the data).

### 2021-04-12 Analysis/06-post_mcmcglmm_analyses.Rmd
vector diff function

@@
  * model 6: `model_phylo1_clade3` with one phylogenetic random effect (across the whole tree) and three clade residual effects.
- * model 7: `model_phylo3_clade3` with three phylogenetic random effect (one for each clade) and three clade residual effects.
+ * model 7: `model_phylo4` with three phylogenetic random effect (one for each clade) and three clade residual effects.
 
@@
 model_phylo1_clade3 <- model_list[[4]]
-model_phylo3_clade3 <- model_list[[6]]
+model_phylo4 <- model_list[[7]]
 ```
@@
 
-```{r, fig.height = 4, fig.width = 8}
-par(mfrow = c(1,2))
+```{r}
 plot.all.axes(uncentered_major_axes,
@@
          add = FALSE, main = "Major axes (un-centred)")
-plot.all.axes(level_centred_major_axes,
-         col = c("grey",colour_vector),
-         add = FALSE, main = "Major axes (level-centred)")
 ```
 
-### Phylo clade + clade model
+We can then compare these vectors per blobs.
+We in order to facilitate the vectors comparisons, we will translate the groups vectors (clade ones) on the origin of the phylo vector.
+
+```{r}
+source("../Functions/vector.diff.R")
+## Get all the angles and stuff
+gull_results <- mapply(vector.diff, uncentered_major_axes$cladegulls, uncentered_major_axes$phylo)
+plovers_results <- mapply(vector.diff, uncentered_major_axes$cladeplovers, uncentered_major_axes$phylo)
+sandpipers_results <- mapply(vector.diff, uncentered_major_axes$cladesandpipers, uncentered_major_axes$phylo)
+
+## Sort the results per type
+angles <- cbind(gull_results["angle", ],
+                plovers_results["angle", ],
+                sandpipers_results["angle", ])
+projections <- cbind(gull_results["projection", ],
+                     plovers_results["projection", ],
+                     sandpipers_results["projection", ])
+rejections <- cbind(gull_results["rejection", ],
+                    plovers_results["rejection", ],
+                    sandpipers_results["rejection", ])
+colnames(angles) <- colnames(projections) <- colnames(rejections) <- levels(morphdat$clade)
+```
+
+```{r, fig.height = 18, fig.heigh = 6}
+## Plot the blob wise metrics
+par(mfrow = c(3, 1))
+boxplot(angles, col = colour_vector, ylab = "angle")
+boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
+boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
+```
+
+Same but with absolute values and angles modulo 90:
+
+```{r, fig.height = 18, fig.heigh = 6}
+## Plot the blob wise metrics
+par(mfrow = c(3, 1))
+boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
+boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
+boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
+```
+
+### Phylo clade model
+
+For this example, the MCMCglmm model has four random (phylo) effects (one for the whole phylo and one per clade) rather than just the one for the whole phylogeny.
+The model seems to struggle converging and there are things off with the centring (they seem to be all centred no matter what) but that last bit should not be a problem for comparing the vectors since they are translated anyways.
 
@@
      ylab = "PC1 (6.86%)")
-plot.ellipses(model_phylo3_clade3, n = 100,
-              centre = "none", add = TRUE, col = c("black", colour_vector, colour_vector))
+plot.ellipses(model_phylo4, n = 100,
+              centre = "none", add = TRUE, col = c("grey",colour_vector, "pink"))
 ```
 
-This looks much more messier (and also probably not scaled correctly?)...
+The pink ellipse is representing the residuals.
+
+And we can still do the whole axis stuff, same way as above:
 
 ```{r}
-plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
-     col = colour_vector[morphdat$clade],
-     xlab = "PC1 (90.5%)",
-     ylab = "PC1 (6.86%)")
-plot.ellipses(model_phylo3_clade3, n = 100,
-              centre = "level", add = TRUE, col = c("black", colour_vector, colour_vector))
+## Get the covariance matrices
+covar_matrices2 <- get.covar(model_phylo4,
+                            n = 100,
+                            simplify = FALSE)
+## Drop the residuals
+covar_matrices2 <- covar_matrices2[-5, ]
+
+## Get all the phylo main axes
+uncentered_major_axes2 <- get.axes(covar_matrices2, centre = "none")
+names(uncentered_major_axes2) <- c("phylo", "cladegulls", "cladeplovers", "cladesandpipers")
+
+```
+
+Here's what the axes look like without the ellipses and stuff:
+
+```{r}
+plot.all.axes(uncentered_major_axes2,
+         col = c("grey",colour_vector),
+         add = FALSE, main = "Major axes (centred)")
+```
+
+```{r}
+source("../Functions/vector.diff.R")
+## Get all the angles and stuff
+gull_results <- mapply(vector.diff, uncentered_major_axes2$cladegulls, uncentered_major_axes2$phylo)
+plovers_results <- mapply(vector.diff, uncentered_major_axes2$cladeplovers, uncentered_major_axes2$phylo)
+sandpipers_results <- mapply(vector.diff, uncentered_major_axes2$cladesandpipers, uncentered_major_axes2$phylo)
+
+## Sort the results per type
+angles <- cbind(gull_results["angle", ],
+                plovers_results["angle", ],
+                sandpipers_results["angle", ])
+projections <- cbind(gull_results["projection", ],
+                     plovers_results["projection", ],
+                     sandpipers_results["projection", ])
+rejections <- cbind(gull_results["rejection", ],
+                    plovers_results["rejection", ],
+                    sandpipers_results["rejection", ])
+colnames(angles) <- colnames(projections) <- colnames(rejections) <- levels(morphdat$clade)
+```
+
+```{r, fig.height = 18, fig.heigh = 6}
+## Plot the blob wise metrics
+par(mfrow = c(3, 1))
+boxplot(angles, col = colour_vector, ylab = "angle")
+boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
+boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
+```
+
+```{r, fig.height = 18, fig.heigh = 6}
+## Plot the blob wise metrics
+par(mfrow = c(3, 1))
+boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
+boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
+boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
 ```
@@
     sort.clade.rej <- function(data, classifier) {
-        return(cbind(data$rejection, classifier))
+        return(data.frame("points" = data$rejection,
+                          "classifier" = classifier))
     }
     sort.clade.proj <- function(data, classifier) {
-        return(cbind(data$projection, classifier))
+        return(data.frame("points" = data$projection,
+                          "classifier" = classifier))
     }
@@
     projections <- do.call(rbind, lapply(point_list, sort.clade.proj, classifier))
-    colnames(rejections)[1] <- colnames(projections)[1] <- "points"
     par(mfrow = c(2,1))
-    boxplot(points ~ classifier, data = projections, col = cols, main = "Projections")
-    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Rejections")
+    boxplot(points ~ classifier, data = projections, col = cols, main = "Phylo projections\n(projected on the whole phylo effect)")
+    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Phylo rejections")
 }
@@
 
-TODO:
+And now we can see the differences for the within clades (i.e. the elaboration/exploration within each clade).
+The idea here is to distinguish between extra special taxa.
+For example, in mammals the platypus and the naked mole rat are special, however, the platypus is actually not special among platypuses whereas the naked mole rat is special among rodents!
 
- - run the blob wise and tip wise models
+```{r}
+## The within clades explo/elabo
+gulls_results_uncentered <- lapply(uncentered_major_axes$cladegulls, lapply.projections, trait_space = morphdat[which(morphdat$clade == "gulls"), c(1,2,3)])
+plover_results_uncentered <- lapply(uncentered_major_axes$cladeplovers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "plovers"), c(1,2,3)])
+sandpipers_results_uncentered <- lapply(uncentered_major_axes$cladesandpipers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "sandpipers"), c(1,2,3)])
+
+projection_list <- list("gulls"   = unlist(lapply(gulls_results_uncentered, `[[`, "projection")),
+             "plovers" = unlist(lapply(plover_results_uncentered, `[[`, "projection")),
+             "sandpipers" = unlist(lapply(sandpipers_results_uncentered, `[[`, "projection")))
+
+rejection_list <- list("gulls"   = unlist(lapply(gulls_results_uncentered, `[[`, "rejection")),
+             "plovers" = unlist(lapply(plover_results_uncentered, `[[`, "rejection")),
+             "sandpipers" = unlist(lapply(sandpipers_results_uncentered, `[[`, "rejection")))
+par(mfrow = c(2,1))
+boxplot(projection_list, main = "projection within clades\n(projected on the clade specific effect)", col  = colour_vector)
+boxplot(rejection_list, main = "rejection within clades", col  = colour_vector)
+```
+
+They seem a bit off as well.
+Maybe it'll be worth centering their axis on their trait space centroid.
+
+TODO.
 

### 2021-04-12 Functions/vector.diff.R

@@ -1,0 +1,150 @@
+#' @title vector differences
+#'
+#' @description Comparing two vectors
+#'
+#' @param vector the vector to compare to the base (also a vector)
+#' @param base the base vector
+#' @param return what to return: c("angle", "projection", "rejection")
+#' @param scale whether to scale the base vector to be the unit (TRUE)
+#' 
+#' @details
+#' This is a modification from dispRity::projections to align the vector with the base and output all the three elements (angle, projection and rejection) rather than just one.
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+# DEBUG
+# vector <- level_centred_major_axes$cladegulls[[1]]
+# base <- level_centred_major_axes$phylo[[1]]
+# vector.diff(vector, base)
+
+vector.diff <- function(vector, base, return = c("angle", "projection", "rejection"), scaled = TRUE) {
+
+    # warning("DEBUG: visual test vector.diff")
+    # lims <- range(c(base, vector))
+    # plot(NULL, type = "l", xlim = lims, ylim = lims)
+
+    ## Translating the inputs to correspond to dispRity::projections
+    matrix <- vector
+    point1 <- base[1,]
+    point2 <- base[2,]
+
+    # arrows(x0 = matrix[1, 1],
+    #        y0 = matrix[1, 2],
+    #        x1 = matrix[2, 1],
+    #        y1 = matrix[2, 2],
+    #        col = "orange", length = 0.1)
+
+    # arrows(x0 = point1[1],
+    #        y0 = point1[2],
+    #        x1 = point2[1],
+    #        y1 = point2[2],
+    #        col = "black", length = 0.1)
+
+    ## Get the translation vector to point1
+    translation_vector <- point1 - matrix[1,]
+
+    # arrows(x0 = 0,
+    #        y0 = 0,
+    #        x1 = translation_vector[1],
+    #        y1 = translation_vector[2],
+    #        col = "red", length = 0.1)
+
+    ## Align the matrix with point1
+    matrix <- rbind(matrix[1,] + translation_vector,
+                    matrix[2,] + translation_vector)
+
+    # arrows(x0 = matrix[1,1],
+    #        y0 = matrix[1,2],
+    #        x1 = matrix[2,1],
+    #        y1 = matrix[2,2],
[…]

### 2021-04-12 Functions/get.axes.R
removed actual axis measuring

@@ -57,8 +57,8 @@ get.one.axes <- function(covar, axis = 1, centre = 0, level = 0.95, dimensions) 
     edges <- edges + rep(centre, each = dims*2)
 
     ## Get the longest axes with dist
-    distances <- apply(edges, 1, dist)[1:dims]
-    axis <- which(distances == sort(distances, decreasing = TRUE)[axis])
+    # distances <- apply(edges, 1, dist)[1:dims]
+    # axis <- which(distances == sort(distances, decreasing = TRUE)[axis])
 
     # stop("DEBUG")
     # ## Weird bug with asp?

### 2021-04-26 Analysis/06-post_mcmcglmm_analyses.Rmd
plot.ellipses

@@
 
+We can do all that with the internal package `"beer"` that can be install via github (if you have access to the repo):
+
+```{r, eval = FALSE}
+devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+```
+
+```{r, eval = TRUE}
+library(beer)
+```
+
 ## Data
@@
 ## Loading the correct models
-load("../Data/Processed/model_list.rda")
+data("model_list")
 model_phylo1_clade3 <- model_list[[4]]
@@
 ## Loading the data
-load("../Data/Processed/morphdat.rda")
+data("morphdat")
 ## Plotting it
 colour_vector <- c("orange", "blue", "darkgreen")
-plot(morphdat[, c(1,2)], pch = 19,
-     col = colour_vector[morphdat$clade],
+plot.space(morphdat,
+           col = colour_vector,
+           levels = morphdat$clade,
            xlab = "PC1 (90.5%)",
@@
 ```{r}
-source("../Functions/plot.ellipses.R")
-source("../Functions/get.covar.R")
-
 ## Selecting 100 random covariance matrices from the MCMCglmm
-covar_matrices <- get.covar(model_phylo1_clade3,
-                            n = 100,
-                            simplify = FALSE)
+covar_matrices <- get.covar(model_phylo1_clade3, n = 100)
 
 ## Plotting the results
-plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
-     col = colour_vector[morphdat$clade],
+plot.space(morphdat,
+           col = colour_vector,
+           levels = morphdat$clade,
            xlab = "PC1 (90.5%)",
            ylab = "PC1 (6.86%)")
-plot.ellipses(covar_matrices, centre = "none",
-              col = c("grey",colour_vector),
+plot.ellipses(covar_matrices, col = c("grey",colour_vector),
               add = TRUE)
@@
 ```{r}
-plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
-     col = colour_vector[morphdat$clade],
+plot.space(morphdat,
+           col = colour_vector,
+           levels = morphdat$clade,
            xlab = "PC1 (90.5%)",
            ylab = "PC1 (6.86%)")
-plot.ellipses(covar_matrices, centre = "level",
+plot.ellipses(covar_matrices, centre = mean,
               col = c("grey",colour_vector),

### 2021-04-26 beer/R/plot.ellipses.R

@@ -3,18 +3,89 @@
 #' @description Plotting axes from an ellipse
 #'
 #' @param beer the \code{"beer"} object.
-#' @param ... any plotting options to be passed to \code{\link{graphics}{lines}}
+#' @param dimensions which dimensions to plot (default is \code{c(1,2)}).
+#' @param npoints the number of points per ellipses (for smoothing; default is \code{50}).
+#' @param centre optional, the centre of the ellipses (see details)
+#' @param col a colour vector.
 #' @param add logical, whether to add the lines to an existing plot (\code{TRUE}, default) or not (\code{FALSE}).
+#' @param scale.axes logical, whether to set both axes on the same scale (\code{TRUE}) or not (\code{FALSE}, default).
+#' @param use.transparent logical, whether to make the colours more transparent for overlaps (default is \code{TRUE}).
+#' @param ... any plotting options to be passed to \code{\link{graphics}{lines}}
+#' 
+#' @details
+#' \code{centre} can be one of the following:
+#' \itemize{
+#'      \item \code{"intercept"} to use the estimated intercept is used (i.e. \code{MCMCglmm$Sol});
+#'      \item a \code{numeric} of one or 2 elements to be used as the x/y coordinates of the centre
+#'      \item a \code{function} to estimate the central tendency of the estimated intercepts (e.g. \code{mean})
+#' }
+#' 
 #' 
 #' @examples
 #'
 #' @seealso
 #' 
 #' @author Thomas Guillerme
-#' @importFrom graphics lines
 #' @export
+plot.ellipses <- function(beer, dimensions = c(1,2), npoints = 50, centre = "intercept", col, add = TRUE, legend, use.transparent = TRUE, ...) {
+    ## Get the ellipses
+    all_ellipses <- lapply(beer, level.ellipses, dimensions, npoints, centre)
+    
+    ## Get the plot arguments
+    plot_args <- plot.dots(data = all_ellipses, dimensions = dimensions, ...)
 
-plot.ellipses <- function(beer, ..., add) {
-    # graphics::lines(1,1)
-    return("prototype")
+    ## Get the colours
+    if(missing(col)) {
+        col <- grDevices::rainbow(length(beer))
+    }
+    if(use.transparent) {
+        adjust <- 1/length(beer) + 1/length(all_ellipses)
+        col <- grDevices::adjustcolor(col, alpha.f = adjust)
+    }
+
+    ## Do the base plot
+    if(!add) {
+       do.call(plot, plot_args) 
+    }
+
[…]

### 2021-04-26 beer/tests/testthat/test-plot.ellipses.R

@@ -1,4 +1,10 @@
 ## Test
+data(model_list)
+data(morphdat)
 test_that("plot.ellipses works", {
-    expect_equal(plot.ellipses(), "prototype")
+    beer <- get.covar(model_list[[7]], n = 7)
+    plot.space(morphdat, levels = morphdat$clade)
+    expect_null(plot.ellipses(beer, use.transparent = TRUE))
+    expect_null(plot.ellipses(beer, centre = -1.5, add = FALSE))
+    expect_null(plot.ellipses(beer, centre = median, add = FALSE))
 })

### 2021-04-26 beer/R/get.covar.R
get covar ups

@@ -2,7 +2,7 @@
 #'
 #' @description Get a sub-sample of the covariance matrices from a MCMCglmm object and the associated MME posterior intercept (solution) sorted by levels
 #'
-#' @param data The MCMCglmm object
+#' @param MCMCglmm The MCMCglmm object
 #' @param n The number of samples
 #' 
 #' @return 
@@ -15,20 +15,27 @@
 #' 
 #' @author Thomas Guillerme, Gavin Thomas
 #' @export
-get.covar <- function(data, n = 1){   
+get.covar <- function(MCMCglmm, n = 1){   
     ## The number of traits
-    traits <- traits.MCMCglmm(data)
+    traits <- traits.MCMCglmm(MCMCglmm)
     n_traits <- length(traits)
     ## The number of levels
-    levels <- levels.MCMCglmm(data)
+    levels <- levels.MCMCglmm(MCMCglmm)
     n_levels <- length(levels)
 
     ## Sample n covar matrices
-    covar_matrices <- replicate(n, get.one.covar(data, levels, traits), simplify = FALSE)
+    covar_matrices <- unlist(replicate(n, get.one.covar(MCMCglmm, levels, traits), simplify = FALSE), recursive = FALSE)
+
+    ## Rearrange the list per levels
+    results_out <- list()
+    for(one_level in 1:n_levels) {
+        results_out[[one_level]] <- unname(covar_matrices[which(names(covar_matrices) == levels[one_level])])
+        names(results_out)[one_level] <- levels[one_level]
+    } 
 
     ## Set the class to beer
-    class(covar_matrices) <- "beer"
-    return(covar_matrices)
+    class(results_out) <- "beer"
+    return(results_out)
 }
 
 ## Internal function to get one covariance matrix

### 2021-04-26 beer/tests/testthat/test-get.covar.R

@@ -57,9 +57,10 @@ test_that("get.one.covar/get.covar works", {
     ## replicate bit works
     test <- get.covar(model_list[[1]], n = 7)
     expect_is(test, "beer")
-    expect_equal(length(test), 7)
-
+    expect_equal(length(test), length(levels.MCMCglmm(model_list[[1]])))
+    expect_equal(length(test[[1]]), 7)
     test <- get.covar(model_list[[7]], n = 7)
     expect_is(test, "beer")
-    expect_equal(length(test), 7)
+    expect_equal(length(test), length(levels.MCMCglmm(model_list[[7]])))
+    expect_equal(length(test[[1]]), 7)
 })

### 2021-04-26 beer/R/plot.space.R
plot space up

@@ -5,9 +5,10 @@
 #' @param data a trait space.
 #' @param levels the levels (categories, eg. clades)
 #' @param dimensions which dimensions to plot (default is \code{c(1,2)}).
-#' @param ... any plotting options to be passed to \code{\link{graphics}{lines}}
-#' @param scale.axes logical, whether to set both axes on the same scale (\code{TRUE}) or not (\code{FALSE}, default).
+#' @param col a colour vector.
 #' @param add logical, whether to add the lines to an existing plot (\code{TRUE}, default) or not (\code{FALSE}).
+#' @param scale.axes logical, whether to set both axes on the same scale (\code{TRUE}) or not (\code{FALSE}, default).
+#' @param ... any plotting options to be passed to \code{\link{graphics}{points}}
 #' 
 #' @examples
 #' data(morphdat)
@@ -20,7 +21,7 @@
 #' @importFrom graphics lines
 #' @export
 
-plot.space <- function(data, levels, dimensions = c(1,2), ..., col, add = FALSE, legend, scale.axes = FALSE) {
+plot.space <- function(data, levels, dimensions = c(1,2), col, add = FALSE, scale.axes = FALSE, ...) {
 
     ## Get the plotting arguments
     if(!scale.axes) {
@@ -38,6 +39,7 @@ plot.space <- function(data, levels, dimensions = c(1,2), ..., col, add = FALSE,
             col <- grDevices::rainbow(length(unique(levels)))
         }
     }
+
     ## Assigning the colour (by levels or not)
     plot_args$col <- col
     if(!missing(levels)) {

### 2021-04-26 beer/R/plot.axes.R
Update plot.axes.R

@@ -49,6 +49,6 @@ plot.axes <- function(data, dimensions = c(1,2), centre = "intercept", col, add 
     line_args <- plot_args
     for(one_level in 1:length(data)) {
         line_args$col <- col[one_level]
-        lapply(axes[[one_level]], function(data, line_args) {line_args$x <- data ; do.call(lines, line_args)}, line_args)
+        lapply(data[[one_level]], function(data, line_args) {line_args$x <- data ; do.call(lines, line_args)}, line_args)
     }
 }

### 2021-04-26 Analysis/06-post_mcmcglmm_analyses.Rmd
Update 06-post_mcmcglmm_analyses.Rmd

@@
 ```{r}
-source("../Functions/get.axes.R")
 ## Get all the phylo main axes
-level_centred_major_axes <- get.axes(covar_matrices, centre = "level")
-uncentered_major_axes <- get.axes(covar_matrices, centre = "none")
+level_centred_major_axes <- get.axes(covar_matrices, centre = mean)
+uncentered_major_axes <- get.axes(covar_matrices, centre = "intercept")
 ```
@@
      ylab = "PC1 (6.86%)")
-plot.ellipses(covar_matrices, centre = "none",
+plot.ellipses(covar_matrices, centre = "intercept",
               col = c("grey",colour_vector),
               add = TRUE)
-plot.all.axes(uncentered_major_axes,
+plot.axes(uncentered_major_axes,
          col = c("grey",colour_vector),
@@
 ```{r}
-plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
-     col = colour_vector[morphdat$clade],
+plot.space(morphdat,
+           col = colour_vector,
+           levels = morphdat$clade,
            xlab = "PC1 (90.5%)",
            ylab = "PC1 (6.86%)")
-plot.ellipses(covar_matrices, centre = "level",
+plot.ellipses(covar_matrices, centre = mean,
               col = c("grey",colour_vector),
               add = TRUE)
-plot.all.axes(level_centred_major_axes,
+plot.axes(level_centred_major_axes,
          col = c("grey",colour_vector),
@@
 ```{r}
-plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
-     col = colour_vector[morphdat$clade],
+plot.space(morphdat,
+           col = colour_vector,
+           levels = morphdat$clade,
            xlab = "PC1 (90.5%)",
            ylab = "PC1 (6.86%)",
-     main = "Major axes of the ellipses\n(aspect ratio = 1)",xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))
-plot.ellipses(covar_matrices, centre = "level",
+           main = "Major axes of the ellipses\n(aspect ratio = 1)",
+           xlim = c(-1.5, 1.5),
+           ylim = c(-1.5, 1.5))
+plot.ellipses(covar_matrices, centre = mean,
               col = c("grey",colour_vector),
               add = TRUE)
-plot.all.axes(level_centred_major_axes,
+plot.axes(level_centred_major_axes,
          col = c("grey",colour_vector),
@@
 ```{r}
-plot.all.axes(uncentered_major_axes,
+plot.axes(uncentered_major_axes,
          col = c("grey",colour_vector),
@@
 
-```{r}
+```{r, eval = FALSE}
 source("../Functions/vector.diff.R")
@@
 
-```{r, fig.height = 18, fig.heigh = 6}
+```{r, fig.height = 18, fig.heigh = 6, eval = FALSE}
 ## Plot the blob wise metrics
@@
 
-### Phylo clade model
+# Methods bits draft
 
-For this example, the MCMCglmm model has four random (phylo) effects (one for the whole phylo and one per clade) rather than just the one for the whole phylogeny.
-The model seems to struggle converging and there are things off with the centring (they seem to be all centred no matter what) but that last bit should not be a problem for comparing the vectors since they are translated anyways.
+## The model
 
-```{r, warn = FALSE}
-plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
-     col = colour_vector[morphdat$clade],
-     xlab = "PC1 (90.5%)",
-     ylab = "PC1 (6.86%)")
-plot.ellipses(model_phylo4, n = 100,
-              centre = "none", add = TRUE, col = c("grey",colour_vector, "pink"))
-```
+We run a MCMCglmm with different levels.
 
-The pink ellipse is representing the residuals.
+## The mini chains
 
-And we can still do the whole axis stuff, same way as above:
+We used a mini chain approach to:
 
-```{r}
-## Get the covariance matrices
-covar_matrices2 <- get.covar(model_phylo4,
-                            n = 100,
-                            simplify = FALSE)
-## Drop the residuals
-covar_matrices2 <- covar_matrices2[-5, ]
+ * accounted for phylogenetic uncertainty [healy.et.al]
+ * accounted for global parameter estimates uncertainty (local vs. global peak)
+ * greatly speeding the calculations through high parallelisation
 
-## Get all the phylo main axes
-uncentered_major_axes2 <- get.axes(covar_matrices2, centre = "none")
-names(uncentered_major_axes2) <- c("phylo", "cladegulls", "cladeplovers", "cladesandpipers")
+We then combine the mini chains
 
-```
+## Exploration and elaboration analyses
 
-Here's what the axes look like without the ellipses and stuff:
+For each model, we randomly draw $N$ variance covariance matrices estimates.
+We then used them to analyse the exploration/elaboration component at the species level and at the clade level indepently.
 
-```{r}
-plot.all.axes(uncentered_major_axes2,
-         col = c("grey",colour_vector),
-         add = FALSE, main = "Major axes (centred)")
-```
+We extracted the major axes of each selected covariance matrices using blablbalba
 
-```{r}
-source("../Functions/vector.diff.R")
-## Get all the angles and stuff
-gull_results <- mapply(vector.diff, uncentered_major_axes2$cladegulls, uncentered_major_axes2$phylo)
-plovers_results <- mapply(vector.diff, uncentered_major_axes2$cladeplovers, uncentered_major_axes2$phylo)
-sandpipers_results <- mapply(vector.diff, uncentered_major_axes2$cladesandpipers, uncentered_major_axes2$phylo)
+### At the clade level
 
-## Sort the results per type
-angles <- cbind(gull_results["angle", ],
-                plovers_results["angle", ],
-                sandpipers_results["angle", ])
-projections <- cbind(gull_results["projection", ],
-                     plovers_results["projection", ],
-                     sandpipers_results["projection", ])
-rejections <- cbind(gull_results["rejection", ],
-                    plovers_results["rejection", ],
-                    sandpipers_results["rejection", ])
-colnames(angles) <- colnames(projections) <- colnames(rejections) <- levels(morphdat$clade)
-```
+We then translated each set of major axes so that they align with each other (i.e. translated the clade major axes origin onto the global major axes origin).
 
-```{r, fig.height = 18, fig.heigh = 6}
-## Plot the blob wise metrics
-par(mfrow = c(3, 1))
-boxplot(angles, col = colour_vector, ylab = "angle")
-boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
-boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
-```
+We then calculated the angle between the clade major axes and the global major axes as well as their projection and rejections.
 
-```{r, fig.height = 18, fig.heigh = 6}
-## Plot the blob wise metrics
-par(mfrow = c(3, 1))
-boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
-boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
-boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
-```
+### At the species level
+
+Here we only projected the species coordinates on either:
+
+ * the global major axes to measure the global elaboration/exploration
+ * the clade major axes to measure the nested elaboration/exploration
 
@@
 
-```{r}
+```{r, eval = FALSE}
 library(dispRity)
@@
 
-```{r}
+```{r, eval = FALSE}
 boxplot.proj.rej <- function(point_list, data, cols) {
@@
 
-```{r}
+```{r, eval = FALSE}
 ## The within clades explo/elabo

### 2021-04-28 beer/R/analyses.tip.R
basic plots

@@ -21,8 +21,13 @@ analyses.tip <- function(data, axes) {
 
     ## Run the analyses
     results <- lapply(axes, lapply.projections, data)
-    projections <- lapply(results, `[[`, "projection")
-    rejections <- lapply(results, `[[`, "rejection")
+    projections <- do.call(cbind, lapply(results, `[[`, "projection"))
+    rejections <- do.call(cbind, lapply(results, `[[`, "rejection"))
+    if(!is.null(rownames(data))) {
+        rownames(projections) <- rownames(rejections) <- rownames(data)
+    }
+
+
     return(list(projections = projections, rejections = rejections))
 }
 

### 2021-04-28 beer/R/plot.analyses.tip.R

@@ -1,0 +1,137 @@
+#' @title Plot analyses tips
+#'
+#' @description Plotting the results of the analyses per tips
+#'
+#' @param results the results of an analyses per group (typically from \code{\link{analyses.group}}).
+#' @param what which results to plot. By default everything is plotted (e.g. \code{c("projections", "rejections")}).
+#' @param col a colour vector (for the levels).
+#' @param group optional, either a list factors for grouping the data or a phylogenetic tree.
+#' @param ... any plotting options to be passed to \code{\link{graphics}{boxplot}}
+
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+plot.analyses.tip <- function(results, what, col, group, ...) {
+    
+    ## Get the graphical options
+    box_args <- plot.dots(...)
+
+    ## Grouping
+    if(missing(group)) {
+        group <- list(rownames(results[[1]]))
+        names(group) <- "data"
+    }
+
+    ## Missing colours
+    if(missing(col)) {
+        if(length(group) == 1) {
+            col <- "white"    
+        } else {
+            col <- grDevices::rainbow(length(group))
+        }
+    }
+    box_args$col <- col
+
+    ## Missing what
+    if(missing(what)) {
+        what <- names(results)
+    }
+
+    ## Transform the results into a table for boxplotting
+    table <- list()
+    for(one_result in 1:length(what)) {
+        table[[one_result]] <- lapply(group, function(one_group, one_result) c(one_result[one_group, ]), one_result = results[what][[one_result]])
+        names(table[[one_result]]) <- names(group)
+    }
+
+    ## Plot the results
+    par(mfrow = c(length(what),1))
+    for(one_stat in 1:length(what)) {
+        ## Setting the box arguments
+        one_box <- box_args
+        one_box$ylab <- what[one_stat]
+
+        ## Setting the box data
+        selected_data <- table[[one_stat]]
+        one_box$x <- selected_data
+
+        ## Removing the labels
+        if(one_stat != length(what)) {
+            one_box$xaxt <- "n"
+        }
+        ## Removing the title
+        if(one_stat != 1) {
+            one_box$main <- NULL
+        }
+
+        ## Plotting the box
+        do.call(boxplot, one_box)
+    }
+    return(invisible())
+}
+
+
[…]

### 2021-04-28 beer/R/plot.tree.R

@@ -1,21 +1,0 @@
-#' @title plot.tree
-#'
-#' @description Plotting axes from an ellipse
-#'
-#' @param beer the \code{"beer"} object.
-#' @param tree one or more trees.
-#' @param ... any plotting options to be passed to \code{\link{graphics}{lines}}
-#' @param add logical, whether to add the lines to an existing plot (\code{TRUE}, default) or not (\code{FALSE}).
-#' 
-#' @examples
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme
-#' @importFrom graphics lines
-#' @export
-
-plot.tree <- function(beer, tree, ..., add) {
-    # graphics::lines(1,1)
-    return("prototype")
-}

### 2021-04-28 beer/tests/testthat/test-analyses.tip.R

@@ -5,10 +5,10 @@ test_that("tip.results works", {
     set.seed(1)
     all_axes <- get.axes(get.covar(model_list[[7]], n = 4))
     data <- morphdat
-    test <- analyses.tip(data, all_axes[[1]])
+    test <- analyses.tip(data, axes = all_axes[[1]])
     expect_equal(names(test), c("projections", "rejections"))
-    expect_equal(length(test[[1]]), 4)
-    expect_equal(length(test[[1]][[1]]), dim(data)[1])
-    expect_equal(length(test[[2]]), 4)
-    expect_equal(length(test[[2]][[1]]), dim(data)[1])
+    expect_equal(dim(test[[1]]), c(359, 4))
+    expect_equal(rownames(test[[1]]), rownames(data))
+    expect_equal(dim(test[[2]]), c(359, 4))
+    expect_equal(rownames(test[[2]]), rownames(data))
 })

### 2021-04-28 beer/tests/testthat/test-plot.analyses.tip.R

@@ -1,0 +1,15 @@
+## Test
+data(model_list)
+data(morphdat)
+test_that("plot.analyses.tip works", {
+    set.seed(1)
+    all_axes <- get.axes(get.covar(model_list[[7]], n = 4))
+    data <- morphdat
+    results <- analyses.tip(data, axes = all_axes[[1]])
+    ## Basic
+    expect_null(plot.analyses.tip(results))
+    expect_null(plot.analyses.tip(results, what = "projections", main = "title", ylim = c(0, 10), col = "blue"))
+    expect_null(plot.analyses.tip(results, group = split(rownames(morphdat), f = morphdat$clade)))
+
+    ## With a tree
+})

### 2021-04-28 beer/tests/testthat/test-plot.tree.R

@@ -1,4 +1,0 @@
-## Test
-test_that("plot.tree works", {
-    expect_equal(plot.tree(), "prototype")
-})

### 2021-04-28 Analysis/06-post_mcmcglmm_analyses.Rmd
Update 06-post_mcmcglmm_analyses.Rmd

@@
 
-```{r, eval = FALSE}
-source("../Functions/vector.diff.R")
+```{r}
 ## Get all the angles and stuff
-gull_results <- mapply(vector.diff, uncentered_major_axes$cladegulls, uncentered_major_axes$phylo)
-plovers_results <- mapply(vector.diff, uncentered_major_axes$cladeplovers, uncentered_major_axes$phylo)
-sandpipers_results <- mapply(vector.diff, uncentered_major_axes$cladesandpipers, uncentered_major_axes$phylo)
-
-## Sort the results per type
-angles <- cbind(gull_results["angle", ],
-                plovers_results["angle", ],
-                sandpipers_results["angle", ])
-projections <- cbind(gull_results["projection", ],
-                     plovers_results["projection", ],
-                     sandpipers_results["projection", ])
-rejections <- cbind(gull_results["rejection", ],
-                    plovers_results["rejection", ],
-                    sandpipers_results["rejection", ])
-colnames(angles) <- colnames(projections) <- colnames(rejections) <- levels(morphdat$clade)
+group_results <- analyses.group(uncentered_major_axes, base = "animal")
+names(group_results) <- levels(morphdat$clade)
 ```
 
-```{r, fig.height = 18, fig.heigh = 6, eval = FALSE}
+```{r, fig.height = 18, fig.heigh = 6}
 ## Plot the blob wise metrics
-par(mfrow = c(3, 1))
-boxplot(angles, col = colour_vector, ylab = "angle")
-boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
-boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
-```
-
-Same but with absolute values and angles modulo 90:
-
-```{r, fig.height = 18, fig.heigh = 6, eval = FALSE}
-## Plot the blob wise metrics
-par(mfrow = c(3, 1))
-boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
-boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
-boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
+plot.analyses.group(group_results, col = colour_vector)
 ```
@@
 ```{r, eval = FALSE}
-library(dispRity)
-## Wrapping function
-lapply.projections <- function(one_axis, trait_space) {
-    return(list(
-        projection = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,]),
-        rejection  = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,], measure = "distance")))
-}
-## Plot rejection/projection profiles
-plot.proj.rej <- function(point_list, cols, data, main) {
-    ## Colour vector
-    colour_vector <- cols[morphdat$clade]
-    colour_vector <- adjustcolor(colour_vector, alpha.f = 1/10)
-    ## lims
-    lim <- range(unlist(point_list))
-    plot(NULL, xlim = lim, ylim = lim, xlab = "projection", ylab = "rejection", main = main)
-    ## Add each points
-    plot.points <- function(one_set, colour_vector) {
-        points_args <- list()
-        points_args$pch <- 19
-        points_args$col <- colour_vector
-        points_args$x <- one_set$projection
-        points_args$y <- one_set$rejection
-        do.call(points, points_args)
-    }
-
-    lapply(point_list, plot.points, colour_vector)
-    return(invisible())
-}
-
-## Getting the results
-phylo_results_uncentered <- lapply(uncentered_major_axes$phylogeny, lapply.projections, trait_space = morphdat[, c(1,2,3)])
+## Global per tip analyses
+tip_results_global <- analyses.tip(data = morphdat, axes = uncentered_major_axes$phylogeny)
 
 ## Plot results
-plot.proj.rej(phylo_results_uncentered, cols = colour_vector, data = morphdat, main = "Phylo elaboration/rejection")
+plot.analyses.tip(tip_results_global, cols = colour_vector, group = split(rownames(morphdat), f = morphdat$clade)))
 ```

### 2021-04-28 Analysis/06-post_mcmcglmm_analyses.Rmd
Update 06-post_mcmcglmm_analyses.Rmd

@@
 
-```{r, eval = FALSE}
+```{r}
 ## Global per tip analyses
-tip_results_global <- analyses.tip(data = morphdat, axes = uncentered_major_axes$phylogeny)
+tip_results_global <- analyses.tip(data = morphdat[, c(1:3)], axes = uncentered_major_axes$animal)
 
 ## Plot results
-plot.analyses.tip(tip_results_global, cols = colour_vector, group = split(rownames(morphdat), f = morphdat$clade)))
+plot.analyses.tip(tip_results_global, cols = colour_vector, group = split(rownames(morphdat), f = morphdat$clade))
 ```

### 2021-05-21 Analysis/01-Data_preparation.Rmd
Reorganised analyses

---
title: "Bird beak's elaboration and exploration: data preparation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

# Data preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### 2021-05-21 Analysis/02-MCMCglmm_mini_chains.Rmd
Reorganised analyses

---
title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

# MCMCglmm mini chains analyses

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### 2021-05-21 Analysis/03-Elaboration_and_exploration.Rmd
Reorganised analyses

--- 2021-04-28 Analysis/06-post_mcmcglmm_analyses.Rmd
@@
 ---
-title: "Post MCMCglmm analyses"
+title: title: "Bird beak's elaboration and exploration"
 author: "Thomas Guillerme"
@@
 ---
+
+# Measuring elaboration and exploration
+
+## Per clade
+
+## Per species
 

### 2021-05-21 Analysis/Older_ideas/01_data_preparation.Rmd
Reorganised analyses

--- 2021-02-19 Analysis/01_data_preparation.Rmd
@@
 ```
+
+## Space visualisation
+
+The full birdy space:
+
+```{r, fig.height = 12, fig.width = 12}
+## Loading the full space
+shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
+## Groups
+group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
+              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
+              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
+## Full space plot
+plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
+```

### 2021-05-21 Analysis/Older_ideas/02-elaboration_exploration_basics.Rmd
Reorganised analyses

--- 2021-03-22 Analysis/02-elaboration_exploration_basics.Rmd
@@
 ## Getting the base vectors for both groups
-psita_vector <- apply(spaces_list$psittaciformes, 2, quantile, prob = c(0.25, 0.75))
-chara_vector <- apply(spaces_list$charadriiformes, 2, quantile, prob = c(0.25, 0.75))
+psita_vector <- apply(spaces_list$psittaciformes, 2, quantile, prob = c(0.025, 0.975))
+chara_vector <- apply(spaces_list$charadriiformes, 2, quantile, prob = c(0.025, 0.975))
 
@@
 
-```{r, print = FALSE}
+```{r, print = TRUE}
 ## Making a top table
@@
 We can add some evo perspective to all of these by making the base vector "evolutionary" and by looking at elaboration/exploration through time!
+
+## Annex: some potentially interesting ratio thingy
+
+```{r, fig.height = 6, fig.width = 6, eval = TRUE, print = TRUE}
+## Loading the whole shape space
+shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
+## Getting the vector from that space
+space_vector <- apply(shapespace, 2, quantile, prob = c(0.025, 0.975))
+psita_space_proj <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ])
+psita_space_reje <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
+chara_space_proj <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ])
+chara_space_reje <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
+## Groups
+group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
+              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
+              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
+## Full space plot
+plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
+arrows(x0 = space_vector[1,1], y0 = space_vector[1,2],
+       x1 = space_vector[2,1], y1 = space_vector[2,2], col = "black", lwd = 4)
+```
+
+```{r, fig.height = 8, fig.width = 8, eval = TRUE, print = TRUE}
+par(mfrow = c(2, 2))
+## The elaboration/exploration profiles
+plot(psita_space_proj, psita_space_reje, pch = 19, col = "blue",
+    xlab = "Elaboration", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
+plot(chara_space_proj, chara_space_reje, pch = 19, col = "orange",
+    xlab = "Elaboration", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
+
+## The elaboration/exploration profiles
+plot(abs(psita_space_proj), psita_space_reje, pch = 19, col = "blue",
+    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
+plot(abs(chara_space_proj), chara_space_reje, pch = 19, col = "orange",
+    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
+```
+
+Meh... But we can also do ratios: the one between the within group elaboration/exploration and the total elaboration/exploration.
+This can help us find the big time explorers/elaborators? (here: the one in the top 99% CI)
+
+```{r, fig.height = 8, fig.width = 8, eval = TRUE, print = TRUE}
+## Get the ratio
+reje_ratio <- psita_reje/psita_space_reje
+proj_ratio <- abs(psita_proj)/abs(psita_space_proj)
+
+## Find the 99% top explorers/elaborators
+top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
+top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
+top_explo_names <- rownames(spaces_list$psittaciformes)[top_explo]
+top_elab_names <- rownames(spaces_list$psittaciformes)[top_elab]
+
+plot(proj_ratio, reje_ratio, pch = 19, col = "blue",
+    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Psittaciformes")
+text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
+text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
+```
+
+```{r, fig.height = 8, fig.width = 8, eval = TRUE, print = TRUE}
+## Get the ratio
+reje_ratio <- chara_reje/chara_space_reje
+proj_ratio <- abs(chara_proj)/abs(chara_space_proj)
+
+## Find the 99% top explorers/elaborators
+top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
+top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
+top_explo_names <- rownames(spaces_list$charadriiformes)[top_explo]
+top_elab_names <- rownames(spaces_list$charadriiformes)[top_elab]
+
+plot(proj_ratio, reje_ratio, pch = 19, col = "orange",
+    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Charadriiformes")
+text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
+text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
+```
+
+Maybe it's picking up some weird birds...

### 2021-05-21 Analysis/Older_ideas/03-ancestral_states_estimations.Rmd
Reorganised analyses

--- 2021-03-22 Analysis/02-ancestral_states_estimations.Rmd
@@
 ```{r}
+## TEMPORARY analyses: one matrix, one ace and one tree:
+tmp_simple_matrices <- list(
+    "psita" = make.dispRity(data = psita_max_ace[[1]], tree = trees_list$psittaciformes[[1]]),
+    "chara" = make.dispRity(data = chara_max_ace[[1]], tree = trees_list$charadriiformes[[1]]))
+save(tmp_simple_matrices, file = "../Data/Processed/tmp_simple_matrices.rda")
+```
+
+```{r}
 ## Converting the lists into dispRity objects

### 2021-05-21 Analysis/Older_ideas/04-elaboration_exploration_analyses.Rmd
Reorganised analyses

--- 2021-03-22 Analysis/04-elaboration_exploration_analyses.Rmd
@@
 
+```{r, fig.height = 6, fig.width = 12}
+## Centroid (tips) ancestor trajectory 
+psita_results <- project.reject(tmp_simple_matrices$psita,
+                                 vector = c("root", colMeans))
+chara_results <- project.reject(tmp_simple_matrices$chara,
+                                 vector = c("tips", "ancestor"))
+
+par(mfrow = c(1,2))
+plot.project.reject.log(psita_results, tmp_simple_matrices$psita, col = "blue", main = "Psittaciformes - tips/ancestor")
+plot.project.reject.log(chara_results, tmp_simple_matrices$chara, col = "orange", main = "Charadriiformes - tips/ancestor")
+```
+
 ## Elaboration/exploration in the present
@@
 Apply these metrics on both clades to get two scores for each taxa (elaboration vs. exploration plots).
+
+Measure the exploration/elaboration for all clades in the whole bird trees and see which ones are particularly low/high?
 
@@
 
-<!-- ## Annex: some potentially interesting ratio thingy
+## Ideas for getting the axis of "elaboration"
 
-```{r, fig.height = 6, fig.width = 6, eval = FALSE, print = FALSE}
-## Loading the whole shape space
-shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
-## Getting the vector from that space
-space_vector <- apply(shapespace, 2, quantile, prob = c(0.25, 0.75))
-psita_space_proj <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ])
-psita_space_reje <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
-chara_space_proj <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ])
-chara_space_reje <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
-## Groups
-group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
-              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
-              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
-## Full space plot
-plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
-arrows(x0 = space_vector[1,1], y0 = space_vector[1,2],
-       x1 = space_vector[2,1], y1 = space_vector[2,2], col = "black", lwd = 4)
+```{r, fig.height = 8, fig.width = 8}
+library(dads)
+
+## Simulate a pure-bd tree with a BM trait
+data <- dads(stop.rule = list(max.living = 300), bd.params = list(speciation = 2), traits = make.traits(n = 3))
+plot(data, main = "One trait through time")
+
+## The node ages
+ages <- tree.age(data$tree)
+node_ages <- ages[which(ages$elements %in% data$tree$node.label),]
+
+## A model of trait through time [TODO: use some time regression or something line that]
+# time_trait <- cbind(time = rev(node_ages$ages), trait = data$data[data$tree$node.label, ])
+# model <- lm(data$data[data$tree$node.label,] ~ 0 + node_ages$ages)
+arrows(y0 = 0, x0 = max(node_ages$ages), y1 = mean(data$data[data$tree$tip.label, ]), x1 = 0, lwd = 3)
 ```
 
-```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
-par(mfrow = c(2, 2))
-## The elaboration/exploration profiles
-plot(psita_space_proj, psita_space_reje, pch = 19, col = "blue",
-    xlab = "Elaboration", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
-plot(chara_space_proj, chara_space_reje, pch = 19, col = "orange",
-    xlab = "Elaboration", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
+```{r}
+library(rgl)
+open3d()
+plot(data, trait = c(1:2), use.3D = TRUE)
+time_trait_arrow <- matrix(NA, 2, 3)
+time_trait_arrow[1,] <- c(max(node_ages$ages),
+                          0,
+                          0)
+time_trait_arrow[2,] <- c(0,
+                          mean(data$data[data$tree$tip.label, 1]),
+                          mean(data$data[data$tree$tip.label, 2]))
+segments3d(time_trait_arrow, lwd = 3)
 
-## The elaboration/exploration profiles
-plot(abs(psita_space_proj), psita_space_reje, pch = 19, col = "blue",
-    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
-plot(abs(chara_space_proj), chara_space_reje, pch = 19, col = "orange",
-    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
+## 3D version
+open3d()
+plot(data, trait = c(1,2,3), use.3D = TRUE)
+data_trait_arrow <- matrix(NA, 2, 3)
+data_trait_arrow[1,] <- c(0,
+                          0,
+                          0)
+data_trait_arrow[2,] <- c(mean(data$data[data$tree$tip.label, 1]),
+                          mean(data$data[data$tree$tip.label, 2]),
+                          mean(data$data[data$tree$tip.label, 3]))
+segments3d(data_trait_arrow, lwd = 3)
+
 ```
-
-Meh... But we can also do ratios: the one between the within group elaboration/exploration and the total elaboration/exploration.
-This can help us find the big time explorers/elaborators? (here: the one in the top 99% CI)
-
-```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
-## Get the ratio
-reje_ratio <- psita_reje/psita_space_reje
-proj_ratio <- abs(psita_proj)/abs(psita_space_proj)
-
-## Find the 99% top explorers/elaborators
-top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
-top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
-top_explo_names <- rownames(spaces_list$psittaciformes)[top_explo]
-top_elab_names <- rownames(spaces_list$psittaciformes)[top_elab]
-
-plot(proj_ratio, reje_ratio, pch = 19, col = "blue",
-    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Psittaciformes")
-text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
-text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
-```
-
-```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
-## Get the ratio
-reje_ratio <- chara_reje/chara_space_reje
-proj_ratio <- abs(chara_proj)/abs(chara_space_proj)
-
-## Find the 99% top explorers/elaborators
-top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
-top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
-top_explo_names <- rownames(spaces_list$charadriiformes)[top_explo]
-top_elab_names <- rownames(spaces_list$charadriiformes)[top_elab]
-
-plot(proj_ratio, reje_ratio, pch = 19, col = "orange",
-    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Charadriiformes")
-text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
-text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
-```
-
-Maybe it's picking up some weird birds...
- -->

### 2021-05-21 Analysis/02-MCMCglmm_mini_chains.Rmd
Mini-chains methods blurb

@@
 date: "`r Sys.Date()`"
+bibliography: references.bib
 output:
@@
 
+All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
+Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
+The following code snippets are used to illustrate the implementation of the functions.
+See §§§ for running the actual analyses.
+
+```{r install_beer, eval = FALSE}
+## Installing beer
+devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+```
+
+```{r load_beer, eval = TRUE}
+## Loading beer
+library(beer)
+
+## Loading the data
+data(morphdat)
+
+## Loading the trees
+data(trees)
+```
+
 # MCMCglmm mini chains analyses
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
+To estimate the variance covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed model (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
+
+## Model
+
+We ran a general multi-response model with `§§§N_traits` trait of the trait space being the the fixed effects (using a gaussian trait distribution) and and the residual covariance structure.
+We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).
+
+## Mini chains
+
+In order to improve the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
+This approach is not too dissimilar from @mulTree as it runs multiple MCMCglmm analyses on multiple trees and pull the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty.
+However, the "mini chains" approach differ by the size of the chains that is optimised for speed.
+
+### Mini-chains parametrization
+
+We first ran three independent MCMC chains with the model and data described above using three random trees and flat priors.
+We ran these models (for `§§§N_generations` with sampling every `§§§N_nitt` and no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis).
+From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
+We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
+For these triplets of values (burnin and list of priors), we applied Cooper §§§' method <!-- Citing Natalie's thesis (or paper?) using this method --> to obtain an average while discarding any outlier estimate (we computed the average and standard deviation for each sample, discarded samples §§§ standard deviations from the average and recalculated this average).
+We then used these parameters (burnin and priors) for the mini chains.
+
+```{r, parametrise_the_model}
+## Set up the parametrising chains models
+
 ```
+
+### Running the mini-chains
+
+We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 1000 sampled iterations (`nitt = burnin + thin * 1000`).
+Each mini-chain thus provided us with a 1000 exploitable posterior data points.
+The two main advantages of this mini-chains is that 1) they are much faster to run since no diagnoses is necessary and the chains are only ran for a relatively short time (which allow several chains to crash/fail without loosing all the data); and 2) they take into account tree uncertainty whitout having to run the complete MCMCglmm on all trees (c.f.
+@mulTree).
+
+```{r, parametrise_the_model}
+## Set up the mini-chains models
+
+```
+
+To make sure we reached consistent results, we run the a set of N mini-chains were N was determined using the following variance optimisation algorithm
+
+ * 1. run an initial $N = 20$ chains;
+ * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
+ * 3. update N to N = N * 1.05 and run an additional 5% chains
+ * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
+ * 5. calculate $\Delta_V$ the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
+ * 6.
+If $\Delta_V < 0.05$ (meaning an increase of 5% of the replicates resulted in an increase of the parameters of less than 5%), the desired number of mini-chains is reached.
+Else, set $V_{1} = V_{2}$ and repeat from step 3.
+
+After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "normal" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.
+
+## References

### 2021-05-21 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd

@@
 Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
+Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
+Don't use it while driving.
 The following code snippets are used to illustrate the implementation of the functions.
@@
 
-## Loading the data
-data(morphdat)
-
-## Loading the trees
-data(trees)
+## Loading the demo data
+data(demo_data)
+data(demo_trees)
 ```
@@
 ## Set up the parametrising chains models
+param_MCMCglmm <- make.mini.chains(data         = demo_data,
+                                   dimensions   = c(1:4),
+                                   tree         = demo_trees,
+                                   trait.family = "gaussian",
+                                   randoms      = c("global", "clade"),
+                                   residuals    = "global",
+                                   priors       = template,
+                                   verbose      = TRUE,
+                                   parameters   = list(
+                                       nitt   = 10000,
+                                       thin   = 100,
+                                       burnin = 0))
 
+## Running the three MCMCglmm models
+param_chains <- run.mini.chains(param_MCMCglmm, replicates = 3)
 ```

### 2021-05-21 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
 
+All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
+Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
+The following code snippets are used to illustrate the implementation of the functions.
+See §§§ for running the actual analyses.
+
+```{r install_beer, eval = FALSE}
+## Installing beer
+devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+```
+
+```{r load_beer, eval = TRUE}
+## Loading beer
+library(beer)
+
+## Loading the demo data
+data(demo_data)
+data(demo_trees)
+```
+
 # Data preparation
+
+TODO:
+
+ * [ ] what's the data, where is it from
+ * [ ] Procrustes
+ * [ ] Building the trait space
+   - [ ] form space version?
+   - [ ] shape space version?
+ * [ ] Clade data and grouping
+
+Trees were manipulated in `R` [@Rcore] using the package `ape` [@ape].
 
@@
 ```
+
+## Demo data
+
+This is what the demo data looks like
+
+```{r, plot_the_data}
+## Plotting it
+colour_vector <- c("orange", "blue", "darkgreen")
+plot.space(demo_data,
+           col = colour_vector,
+           levels = demo_data$clade,
+           xlab = "PC1 (90.5%)",
+           ylab = "PC1 (6.86%)")
+legend("topleft", legend = levels(demo_data$clade), col = colour_vector, pch = 19)
+```
+
+## References

### 2021-05-21 Analysis/03-Elaboration_and_exploration.Rmd
INIT

@@
 # Measuring elaboration and exploration
+
+<!-- We can do all that with the internal package `"beer"` that can be install via github (if you have access to the repo): -->
+
+```{r, eval = FALSE, echo = FALSE}
+devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+```
+
+```{r, eval = TRUE, echo = FALSE}
+library(beer)
+```
+
+## Extracting the variance covariance matrices from the MCMCglmm mini-chains
 
@@
  * and the **tip-wise** one where I'll look at the elaboration/exploration score of the tips on the tree (cf. their blobs) relative to the whole phylogeny and to their respective blobs (clades).
-
-We can do all that with the internal package `"beer"` that can be install via github (if you have access to the repo):
-
-```{r, eval = FALSE}
-devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
-```
-
-```{r, eval = TRUE}
-library(beer)
-```
 
@@
 
-# Ideas for the MCMC
-
- * Use the model7 with two nested clade levels?
-
- * For speeding up the MCMC:
-   1. Check on a big global model: a) how long it takes to get 1000 gen. (or so) and b) how long the burnin phase takes.
-   2.
-Use info 1a and 1b to make design "mini chains" (i.e.
-chains that just go past the burnin and run for an extra ~1000 gen.); the conservative assumption is that post-burnin, the chain has reach a local optimum that exists within the distribution of the global (true) optimum.
-   3.
-Run 3 mini chains per tree on 100 trees (resulting in 300 mini chains) to get phylogenetic uncertainty into account.
-For each triplet of models, discard any that's way off (Natalie thesis style for morpho measurements).
-This results in N models that have reach N local optimums while taking into account phylo uncertainty.
-   4.
-Discard any of the N models that is way off (outliers).
-We assume that these outliers were chains stuck on local optimums that were outside to global optimum distribution.
-We then end up with M (M<=N) models that are the most conservative representation of the model because they take into account phylo uncertainty AND optimum uncertainty (i.e.
-if running one big model, you have to make the assumption that the local optimum reached equal the global optimum, here we can skip that assumption by saying that our distribution of local optimums contains the global optimum).
-   5.
-Combine the results of all the MCMC into one BIG posterior distribution; randomly sample n (=Mx5?) samples (or any other number) from that distribution and work from there!
-(if we don't discard any models M = 300 mini chains of >=1000 samples each, thus the sample represents ~5% of all the data).
+## References

### 2021-05-21 beer/R/make.mini.chains.R
Typo

@@ -52,7 +52,7 @@ make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", 
     }
 
     ## Check the dimensionality of the data
-    if(!all(dim_check <- apply(morphdat[, dimensions], 2, is, "numeric"))) {
+    if(!all(dim_check <- apply(data[, dimensions], 2, is, "numeric"))) {
         stop(paste0("Invalid dimensions (not numeric?): ", paste(names(which(!dim_check)), collapse = ", "), "."))
     }
     

### 2021-05-21 beer/R/extract.parameters.R
extracting parameters function (UNTESTED)

@@ -1,0 +1,101 @@
+#' @title Extract parameters
+#'
+#' @description Extracting parameters from prior MCMCglmm chains
+#'
+#' @param chains The prior MCMCglmm chains
+#' @param parameters The parameters to extract (see details).
+#' @param buffer The buffer for the burnin (how many itteration to include past the first median estimate)
+#' @param credence The credence level: the degree of belief parameter (nu) for the priors.
+#' 
+#' @details
+#' The parameters that can be extracted are:
+#' \itemize{
+#'      \item \code{"burnin"}: the global burnin length
+#'      \item \code{"priors"}: a list of priors (from the posteriors)
+#' }
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+extract.parameters <- function(chains, parameters = c("burnin", "priors"), buffer = 0.25, credence = 0.05) {
+    return()
+}
+
+extract.one.chain <- function(chain, parameters) {
+
+    ## Get the required parameters
+    parameters <- list()
+    if("burnin" %in% parameters) {
+        parameters$burnin <- get.burnin(chain)
+    }
+
+    if("priors" %in% parameters) {
+        # 
+    }
+
+
+
+
+
+
+
+    return(parameters)
+}
+
+get.burnin <- function(chain, buffer) {
+    ## Find the median for each MCMC
+    find.median <- function(estimate, buffer) {
+        ## Get the first estimate past the median
+        return(ceiling(which(estimate > median(estimate))[1] * (1+buffer)))
+    }
+
+    burnin_points <- apply(chain$Sol, 2, find.median, buffer)
+    return(max(burnin_points * attr(chain$Sol, "mcpar")[3]))
+}
+
+get.prior <- function(chain, credence = 0.05) {
+
+    ## Get the set of parameters
+    traits <- traits.MCMCglmm(chain)
+    n_traits <- length(traits)
+    ## The number of levels
+    levels <- levels.MCMCglmm(chain)
+    n_levels <- length(levels)
+    n_ran <- sum(names(levels) %in% "random")
+    n_res <- sum(names(levels) %in% "residual")
+
+    ## Create the prior template
+    template <- flat.prior(ntraits = n_traits, residuals = n_res, randoms = n_ran), nu = credence)
+
[…]

### 2021-05-21 beer/tests/testthat/test-extract.parameters.R

@@ -1,0 +1,4 @@
+## Test
+test_that("extract.parameters works", {
+
+})

### 2021-05-24 beer/R/extract.parameters.R
extract.parameters (UNFINISHED)

@@ -5,7 +5,7 @@
 #' @param chains The prior MCMCglmm chains
 #' @param parameters The parameters to extract (see details).
 #' @param buffer The buffer for the burnin (how many itteration to include past the first median estimate)
-#' @param credence The credence level: the degree of belief parameter (nu) for the priors.
+#' @param nu The degree of belief parameter (nu) for the priors.
 #' 
 #' @details
 #' The parameters that can be extracted are:
@@ -21,32 +21,50 @@
 #' @author Thomas Guillerme
 #' @export
 
-extract.parameters <- function(chains, parameters = c("burnin", "priors"), buffer = 0.25, credence = 0.05) {
+extract.parameters <- function(chains, parameters = c("burnin", "priors"), buffer = 0.25, nu = 0.05) {
+
+    ## Get the required parameters
+    param_out <- list()
+    if("burnin" %in% parameters) {
+        param_out$burnin <- max(unlist(lapply(chains, get.burnin, buffer = buffer)))
+    }
+    if("priors" %in% parameters) {
+        ## Check for burnin
+        if(!is.null(param_out$burnin)) {
+            burnin <- param_out$burnin/attr(chains[[1]]$Sol, "mcpar")[3]
+        } else {
+            burnin <- max(unlist(lapply(chains, get.burnin, buffer = buffer)))/attr(chains[[1]]$Sol, "mcpar")[3]
+        }
+        ## Get the list of priors
+        priors_lists <- lapply(chains, get.prior, nu = nu, burnin = burnin)
+
+    }
+
     return()
 }
 
-extract.one.chain <- function(chain, parameters) {
+# ## Get a corrected median value
+# corrected.median <- function(x) {
+    
+#     x <- sort(x)
 
-    ## Get the required parameters
-    parameters <- list()
-    if("burnin" %in% parameters) {
-        parameters$burnin <- get.burnin(chain)
-    }
+#     if(length(x) > 2) {
+#         ## Get all the differences
+#         range <- diff(range(x))
+#         differences <- diff(x)
 
-    if("priors" %in% parameters) {
-        # 
-    }
+#         ## Remove outliers
+#         differences/range > 0.5
+#         x <- x[-(outliers+1)]
+#     }
 
+#     ## Calculate the median
+#     return(median(x))
+# }
 
-
-
-
-
-
-    return(parameters)
-}
-
[…]

### 2021-05-24 beer/tests/testthat/test-extract.parameters.R

@@ -1,4 +1,40 @@
 ## Test
+load("MCMCM_chains_test.rda")
+chains <- MCMCM_chains_test
+test_that("get.burnin works", {
+    test <- get.burnin(chains[[1]], buffer = 0)
+    expect_equal(test, 700)
+    test <- get.burnin(chains[[1]], buffer = 0.25)
+    expect_equal(test, 900)
+    test <- get.burnin(chains[[2]], buffer = 0)
+    expect_equal(test, 400)
+    test <- get.burnin(chains[[3]], buffer = 0)
+    expect_equal(test, 600)
+})
+
 test_that("extract.parameters works", {
+
+    test <- get.prior(chains[[1]])
+    expect_equal(names(test), c("R", "G"))
+    expect_equal(names(test[[1]]), c("R1"))
+    expect_equal(names(test[[2]]), c("G1", "G2", "G3", "G4"))
+    expect_equal(names(test[[1]]$R1), c("V", "nu"))
+    expect_equal(names(test[[2]]$G1), c("V", "nu"))
+    expect_equal(dim(test[[1]]$R1$V), c(4,4))
+    expect_equal(test[[1]]$R1$nu, 0.05)
+    expect_equal(round(mean(test[[1]]$R1$V), 6), 0.001468)
+    expect_equal(round(mean(test[[2]]$G1$V), 6), 0.029164)
+    test <- get.prior(chains[[2]])
+    expect_equal(round(mean(test[[1]]$R1$V), 6), 0.001529)
+    expect_equal(round(mean(test[[2]]$G1$V), 6), 0.030054)
+})
+
+
+
+
+test_that("extract.parameters works", {
+
+
+
 
 })

### 2021-05-24 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd

@@
 We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
-For these triplets of values (burnin and list of priors), we applied Cooper §§§' method <!-- Citing Natalie's thesis (or paper?) using this method --> to obtain an average while discarding any outlier estimate (we computed the average and standard deviation for each sample, discarded samples §§§ standard deviations from the average and recalculated this average).
-We then used these parameters (burnin and priors) for the mini chains.
+For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
 
@@
                                    residuals    = "global",
-                                   priors       = template,
+                                   priors       = 0.02,
                                    verbose      = TRUE,

### 2021-05-26 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd

@@
 ```{r load_beer, eval = TRUE}
+set.seed(42)
 ## Loading beer
@@
 data(demo_trees)
+data(demo_tree)
+
+## Setting the consensus tree
+consensus_tree <- demo_tree[[1]]
 ```
@@
 
-We first ran three independent MCMC chains with the model and data described above using three random trees and flat priors.
+We first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
 We ran these models (for `§§§N_generations` with sampling every `§§§N_nitt` and no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis).
-From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
-We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
-For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
 
@@
                                    dimensions   = c(1:4),
-                                   tree         = demo_trees,
+                                   tree         = consensus_tree,
                                    trait.family = "gaussian",
@@
 
+From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
+We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
+For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
+
+```{r, get_model_parameters}
+## Extract the parameters from the parameter chains
+est_params <- extract.parameters(param_chains)
+```
+
 ### Running the mini-chains
 
-We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 1000 sampled iterations (`nitt = burnin + thin * 1000`).
-Each mini-chain thus provided us with a 1000 exploitable posterior data points.
+We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
+Each mini-chain thus provided us with a 100 exploitable posterior data points.
 The two main advantages of this mini-chains is that 1) they are much faster to run since no diagnoses is necessary and the chains are only ran for a relatively short time (which allow several chains to crash/fail without loosing all the data); and 2) they take into account tree uncertainty whitout having to run the complete MCMCglmm on all trees (c.f.
@@
 ```{r, parametrise_the_model}
+## The thining parameter
+thin <- 100
+## The number of samples per chains
+samples <- 100
+
 ## Set up the mini-chains models
+mini_chains <- make.mini.chains(data         = demo_data,
+                                dimensions   = c(1:4),
+                                tree         = demo_trees,
+                                trait.family = "gaussian",
+                                randoms      = c("global", "clade"),
+                                residuals    = "global",
+                                priors       = est_params$priors,
+                                verbose      = TRUE,
+                                parameters   = list(
+                                    nitt   = est_params$burnin + thin * samples,
+                                    thin   = thin,
+                                    burnin = est_params$burnin))
+```
 
+We then run the MCMCglmm over $N_trees$ to get an estimate of the variance co-variance matrices including phylogenetic uncertainty.
+
+```{r, running_replicates}
+## The number of replicates
+n_reps <- 10
+
+## Run the required chains
+results <- run.mini.chains(mini_chains, replicates = n_reps, path = "../Data/Processed/MCMC/")
+```
+
+#### Number of runs optimisation (TO BE DEVELOPPED?)
+
+```{r, eval = FALSE}
+## Function for running one mini chain on a random tree (and saving the results)
+fun.mini.chains <- function() {
+    run.mini.chains(mini_chains, replicates = 1, path = "../Data/Processed/MCMC/", file.name = "mini.chains")
+}
 ```
@@
 
+```{r, optimised_replicates, eval = FALSE}
+## The summary function
+
+```
+
 After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "normal" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.

### 2021-05-26 beer/R/extract.parameters.R
extract params tessted

@@ -20,7 +20,6 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-
 extract.parameters <- function(chains, parameters = c("burnin", "priors"), buffer = 0.25, nu = 0.05) {
 
     ## Get the required parameters
@@ -36,31 +35,43 @@ extract.parameters <- function(chains, parameters = c("burnin", "priors"), buffe
             burnin <- max(unlist(lapply(chains, get.burnin, buffer = buffer)))/attr(chains[[1]]$Sol, "mcpar")[3]
         }
         ## Get the list of priors
-        priors_lists <- lapply(chains, get.prior, nu = nu, burnin = burnin)
-
+        priors_list <- lapply(chains, get.prior, nu = nu, burnin = burnin)  
+        ## Merging the priors
+        param_out$priors <- merge.V(priors_list)
     }
 
-    return()
+    return(param_out)
 }
 
-# ## Get a corrected median value
-# corrected.median <- function(x) {
+## Merge the lists of priors V matrices
+merge.V <- function(priors_list) {
+    ## Get the number of levels
+    n_ran <- length(priors_list[[1]]$G)
+    n_res <- length(priors_list[[1]]$R)
+    n_traits <- dim(priors_list[[1]]$R$R1$V)[1]
+    nu <- priors_list[[1]]$R$R1$nu
+
+    ## Get the template to fill
+    template <- flat.prior(ntraits = n_traits, residuals = n_res, randoms = n_ran, nu = nu)
     
-#     x <- sort(x)
+    ## Get the length of the list
+    n_chains <- length(priors_list)
 
-#     if(length(x) > 2) {
-#         ## Get all the differences
-#         range <- diff(range(x))
-#         differences <- diff(x)
+    ## Fill the list for G
+    for(one_ran in 1:n_ran) {
+        ## Get the list of V matrices
+        V_list <- lapply(lapply(priors_list, `[[`, "G"), function(X, i) return(X[[i]]$V), i = one_ran)
+        template$G[[one_ran]]$V <- apply(array(do.call(cbind, V_list), dim = c(dim(V_list[[1]]), length(V_list))), c(1,2), median)
+    }
 
-#         ## Remove outliers
-#         differences/range > 0.5
-#         x <- x[-(outliers+1)]
-#     }
-
-#     ## Calculate the median
-#     return(median(x))
-# }
+    ## Fill the list for R
+    for(one_res in 1:n_res) {
[…]

### 2021-05-26 beer/tests/testthat/test-extract.parameters.R

@@ -12,7 +12,7 @@ test_that("get.burnin works", {
     expect_equal(test, 600)
 })
 
-test_that("extract.parameters works", {
+test_that("get.prior works", {
 
     test <- get.prior(chains[[1]])
     expect_equal(names(test), c("R", "G"))
@@ -29,12 +29,14 @@ test_that("extract.parameters works", {
     expect_equal(round(mean(test[[2]]$G1$V), 6), 0.030054)
 })
 
-
-
-
 test_that("extract.parameters works", {
-
-
-
-
+    test <- extract.parameters(chains)
+    expect_is(test, "list")
+    expect_equal(names(test), c("burnin", "priors"))
+    expect_equal(test$burnin, 900)
+    expect_equal(sum(round(test$priors$G$G1$V, 1)), 0.3)
+    expect_equal(sum(round(test$priors$G$G2$V, 1)), 0.2)
+    expect_equal(sum(round(test$priors$G$G3$V, 1)), 0.2)
+    expect_equal(sum(round(test$priors$G$G4$V, 1)), 0.3)
+    expect_equal(sum(round(test$priors$R$R1$V, 2)), 0.02)
 })

### 2021-05-26 beer/R/run.mini.chains.R
Update run.mini.chains.R

@@ -28,9 +28,8 @@ run.mini.chains <- function(mini.chains, replicates, parallel, path, file.name =
     ## Saving the business
     if(!missing(path)) {
         save(output, file = paste0(c(path, file.name, "_", replicates, ".rda"), collapse = "")) #, "_", format(Sys.time(), "%Y-%m-%d-%H%M%S")
-    } else {
-        return(output)
     }
+    return(output)
 }
 
 run.a.chain <- function(mini.chains) {

### 2021-05-26 Analysis/02-MCMCglmm_mini_chains.Rmd
reproducible pipeline!

@@
 library(beer)
+library(MCMCglmm)
 
@@
 ## Setting the consensus tree
-consensus_tree <- demo_tree[[1]]
+consensus_tree <- demo_tree
 ```
@@
 
-```{r, parametrise_the_model}
+```{r, parametrise_the_model, eval = TRUE}
 ## Set up the parametrising chains models
@@
 
-```{r, get_model_parameters}
+```{r, get_model_parameters, eval = TRUE}
 ## Extract the parameters from the parameter chains
@@
 
-```{r, parametrise_the_model}
+```{r, set_the_mini_chains, eval = TRUE}
 ## The thining parameter
@@
 
-```{r, running_replicates}
+```{r, running_replicates, eval = TRUE}
 ## The number of replicates
@@
 results <- run.mini.chains(mini_chains, replicates = n_reps, path = "../Data/Processed/MCMC/")
+
+## Combining the results as one big MCMC
+combined_results <- combine.mini.chains(results)
+
+## Saving the results
+save(combined_results, file = "../Data/processed/combined_results.rda")
 ```

### 2021-05-27 Analysis/03-elaboration_exploration_analyses.Rmd
loading data for elaboration part
Major axes equation

@@
 ---
-title: "03-elaboration exploration analyses"
+title: "Bird beak's elaboration and exploration"
 author: "Thomas Guillerme"
 date: "`r Sys.Date()`"
+bibliography: references.bib
 output:
   html_document:
-    fig_width: 12
-    fig_height: 6
+    fig_width: 8
+    fig_height: 8
 ---
 
-# Elaboration/exploration analyses
+All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
+Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
+Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
+Don't use it while driving.
+The following code snippets are used to illustrate the implementation of the functions.
+See <!-- §§§run_replicates_script --> for running the actual analyses.
 
-## Elaboration/exploration metrics
+# Elaboration and exploration analyses
 
-Measure the different elaboration vs exploration metrics with different vectors:
+```{r, eval = FALSE, echo = FALSE}
+library(beer)
+## Loading the MCMCglmm
+load("../Data/processed/combined_results.rda")
+## Loading the tree and data
+data(demo_data)
+data(demo_tree)
+```
 
- - [ ] MRCA trajectory: the vector between the root of the tree and the OTU's descendant
- - [ ] MRCA average trajectory: the vector between the centroid of the group's nodes and the root of the tree
- - [ ] Evolutionary living average trajectory: the vector between the root of the tree and the centroid of the living taxa
- - [ ] Evolutionary overall average trajectory: the vector between the root of the tree and the centroid of the tips and nodes
- - [ ] Maximum living trajectory: the vector between the two most distant OTUs
- - [ ] Maximum evolutionary trajectory: the vector between the root and the OTU most distant from the root
+## Extracting the variance covariance matrices from the MCMCglmm mini-chains
 
-## Elaboration/exploration in the present
+To analyse the exploration and elaboration aspects of both clades and tips, we first selected a random 1000 covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.
 
-Apply these metrics on both clades to get two scores for each taxa (elaboration vs. exploration plots).
+```{r, eval = FALSE}
+set.seed(42)
+## Selecting 1000 random covariance matrices from the MCMCglmm
+covar_matrices <- get.covar(combined_results, n = 100)
+```
 
-## Elaboration/exploration through time
+We will use these covariance matrices as the basis of the phylogenetic main axes of variation.
+These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different levels: 
+ * The whole bird phylogeny level ("phylogeny level");
+ * And the named node levels ("clade levels").
 
-Do some time-slicing and run these metrics through time.
+These major axes where calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e. the major axes of the 95% CI ellipse of matrix).
+In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigen values and the variance-covariance matrix's eigen vectors.
+
+### Details for getting the major axes from the variance co-variance matrices
+
+For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and of radius being composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
+
+\begin{equation*}
+O_{n} = 
+    \begin{pmatrix}
+        1 & 0 & \cdots & 0 \\
+        0 & 1 & \cdots & 0 \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        0 & 0 & \cdots & 1 \\
+        -1 & 0 & \cdots & 0 \\
+        0 & -1 & \cdots & 0 \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        0 & 0 & \cdots & -1 \\
+    \end{pmatrix}
+\end{equation*}
+
+(in other words $O_{n}$ is the matrix representing each edge point of the unit hyper sphere of $n$ dimensions and of radius $1$);
+
+And $O'_{n}$ the scaled matrix hyper sphere to the right confidence interval size using the $\chi^2$ distribution:
+
+\begin{equation}
+    O'_{n} = O_{n} \sqrt{\chi^2(0.95)}
+\end{equation}
+
+For the variance-covariance matrix $VCV_{n}$ of $n$ dimensions
+
+\begin{equation*}
+VCV_{n} = 
+    \begin{pmatrix}
+        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
+        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
+    \end{pmatrix}
+\end{equation*}
+
+and the eigenvectors **v** and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
+
+\begin{equation}
+    VCV_{n} \bold{v} = \lambda \bold{v}
+\end{equation}
+
+We can get $M_{n}$ matrix containing all the edges' coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
+
+\begin{equation}
+    M_{n} = [(O'_{n}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
+\end{equation}
+
+Where $M_{1,n}$ is the major axes of the 0.95 CI hyper ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second second axes (minor axes in 2D), etc...
+
+The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axis` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
+
+## References

### 2021-05-27 Functions/test.rotation.R
Function for elaboration/exploration plotting in all it's glory

@@ -1,0 +1,186 @@
+plot.base <- function(matrix, base_vector, lim) {
+
+    if(missing(lim)) {
+        lim <- ceiling(max(abs(range(matrix))))
+        lim <- c(-lim, lim)
+    }
+
+    if(ncol(matrix) == 3) {
+        plot3d(NULL, xlim = lim, ylim =lim, zlim = lim, xlab = "", ylab = "", zlab = "")
+        segments3d(x = lim, y = c(0, 0), z = c(0, 0), col = "grey")
+        text3d(x = min(lim)*0.9, y = 0, z = 0, texts = "X", col = "grey")
+        segments3d(x = c(0,0), y = lim, z = c(0, 0), col = "grey")
+        text3d(x = 0, y = min(lim)*0.9, z = 0, texts = "Y", col = "grey")
+        segments3d(x = c(0,0), y = c(0,0), z = lim, col = "grey")
+        text3d(x = 0, y = 0, z = min(lim)*0.9, texts = "Z", col = "grey")
+        ## Add the original points
+        text3d(matrix, texts = rownames(matrix), col = "grey")
+        segments3d(base_vector, col = "grey")
+    } else {
+        plot(NULL, xlim = lim, ylim = lim, xlab = "x", ylab = "y")
+        abline(v = 0, col = "grey")
+        abline(h = 0, col = "grey")
+        text(matrix[,1:2], labels = rownames(matrix), col = "grey")
+        lines(base_vector, col = "grey")
+    }
+}
+plot.recentred <- function(matrix, base_vector) {
+    if(ncol(matrix) == 3) {
+        text3d(matrix, texts = rownames(matrix), col = "black")
+        for(i in 1:nrow(matrix)) {
+            segments3d(rbind(c(0,0,0), matrix[i,]), col ="grey")
+        }
+        segments3d(base_vector, col = "black", lwd = 2)
+    } else {
+        text(matrix, labels = rownames(matrix), col = "black")
+        for(i in 1:nrow(matrix)) {
+            lines(rbind(c(0,0), matrix[i,]), col = "grey")
+        }
+        lines(base_vector, col = "black")
+    }
+}
+plot.projections <- function(matrix, projections, rejections) {
+    if(ncol(matrix) == 3) {
+        for(i in 1:nrow(matrix)) {
+            ## Plot the projections
+            segments3d(rbind(c(0,0,0), projections[i,]), col = "red", lwd = 2)
+            ## Plot the rejection
[…]

### 2021-05-27 Analysis/major.axis.tex
major axis latex formulas

\documentclass[a4paper,11pt]{article}

\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{lastpage}
\usepackage{url}
\usepackage{hyperref}
\pagenumbering{arabic}
\linespread{1.66}

\begin{document}

\begin{flushright}
Version dated: \today
\end{flushright}
\begin{center}

%Title
\noindent{\Large{\bf{Major axes estimation}}}\
\bigskip
%Author
Thomas Guillerme\\href{mailto:t.guillerme@imperial.ac.uk}{t.guillerme@imperial.ac.uk}

For \bold{O_{n}}, the unit hypersphere matrix of \textit{n} dimensions and of radius being composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

\begin{equation*}
O_{n} = 
    \begin{pmatrix}
        1 & 0 & \cdots & 0 \\
        0 & 1 & \cdots & 0 \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        0 & 0 & \cdots & 1 \\
        -1 & 0 & \cdots & 0 \\
        0 & -1 & \cdots & 0 \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        0 & 0 & \cdots & -1 \\
    \end{pmatrix}
\end{equation*}

(i.e $O_{n}$ is the matrix representing each edge point of the unit hyper sphere of \textit{n} dimensions and of radius 1);

And \bodel{O'_{n}} the scaled matrix hyper sphere to the right confidence interval size using the $\chi^2$ distribution:

\begin{equation}
    \bold{O'_{n}} = \bold{O_{n}} \sqrt{\chi^2(0.95)}
\end{equation}

For the variance-covariance matrix \textit{VCV_{n}} of \textit{n} dimensions

\begin{equation*}
VCV_{n} = 
    \begin{pmatrix}
        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
    \end{pmatrix}
\end{equation*}

and the eigenvectors \bold{v} and the eigenvalues \lambda satisfying the following eigen decomposition:

\begin{equation}
    VCV_{n}\bold{v} = \lambda \bold{v}
\end{equation}

We can get $M_{n}$ matrix containing all the edges' coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors \bold{v} and the product of the scale 0.95 CI unit sphere \bold{O'_{n}} and the eigenvalues \lambda:

\begin{equation}
    M_{n} = [(\bold{O'_{n}}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
\end{equation}

Where M_{1,n} is the major axes of the 0.95 CI hyper ellipse fitting the variance-covariance matrix.

\end{center}

\end{document}

### 2021-05-27 Functions/test.rotation.R
bugs + colours

@@ -43,17 +43,17 @@ plot.projections <- function(matrix, projections, rejections) {
     if(ncol(matrix) == 3) {
         for(i in 1:nrow(matrix)) {
             ## Plot the projections
-            segments3d(rbind(c(0,0,0), projections[i,]), col = "red", lwd = 2)
+            segments3d(rbind(c(0,0,0), projections[i,]), col = "blue", lwd = 2)
             ## Plot the rejection
-            segments3d(rbind(projections[i, ], rejections[i,]+projections[i, ]), col = "green", lwd = 2)
+            segments3d(rbind(projections[i, ], rejections[i,]+projections[i, ]), col = "orange", lwd = 2)
         }
 
     } else {
         for(i in 1:nrow(matrix)) {
             ## Plot the projections
-            lines(rbind(c(0,0), projections[i,]), col = "red")
+            lines(rbind(c(0,0), projections[i,]), col = "blue")
             ## Plot the rejection
-            lines(rbind(projections[i, ], rejections[i,]+projections[i, ]), col = "green")
+            lines(rbind(projections[i, ], rejections[i,]+projections[i, ]), col = "orange")
         }        
     }
 }
@@ -157,15 +157,8 @@ projections.debug <- function(matrix, point1 = 0, point2 = colMeans(matrix), mea
     return(values)
 }
 
-library(rgl)
-matrix <- point1 <- point2 <- base_vector <- projections <- rejections <- base_angl <- NULL
-set.seed(2) #1,4
-for(i in 1:10) {
-    test(seed = i, n = 2) #2, 4, 7
-    Sys.sleep(0.5)
-}
 
-test <- function(seed, n) {
+test.fun <- function(seed, n) {
     set.seed(seed)
     if(n == 3) {
         matrix <- matrix(rnorm(15), 5, 3)
@@ -179,6 +172,15 @@ test <- function(seed, n) {
     ## Test the algorithm visualy
     projections.debug(matrix, point1 = point1, point2 = point2, measure = "position", scaled = TRUE)
 }
+
+library(rgl)
+matrix <- point1 <- point2 <- base_vector <- projections <- rejections <- base_angl <- NULL
+set.seed(2) #1,4
+#for(i in 1:10) {
+i = 10
+    test.fun(seed = i, n = 2) #2, 4, 7
+    Sys.sleep(0.5)
+# }
 
 
 

### 2021-05-27 Functions/test.rotation.R
Update test.rotation.R

@@ -17,11 +17,12 @@ plot.base <- function(matrix, base_vector, lim) {
         text3d(matrix, texts = rownames(matrix), col = "grey")
         segments3d(base_vector, col = "grey")
     } else {
+        par(bty = "n")
         plot(NULL, xlim = lim, ylim = lim, xlab = "x", ylab = "y")
         abline(v = 0, col = "grey")
         abline(h = 0, col = "grey")
-        text(matrix[,1:2], labels = rownames(matrix), col = "grey")
-        lines(base_vector, col = "grey")
+        text(matrix[,1:2], labels = rownames(matrix), col = "grey", cex = 1.5)
+        lines(base_vector, col = "grey", lwd = 1.5)
     }
 }
 plot.recentred <- function(matrix, base_vector) {
@@ -32,11 +33,11 @@ plot.recentred <- function(matrix, base_vector) {
         }
         segments3d(base_vector, col = "black", lwd = 2)
     } else {
-        text(matrix, labels = rownames(matrix), col = "black")
+        text(matrix, labels = rownames(matrix), col = "black", cex = 1.5)
         for(i in 1:nrow(matrix)) {
-            lines(rbind(c(0,0), matrix[i,]), col = "grey")
+            lines(rbind(c(0,0), matrix[i,]), col = "grey", lwd = 1.5)
         }
-        lines(base_vector, col = "black")
+        lines(base_vector, col = "black", lwd = 1.5)
     }
 }
 plot.projections <- function(matrix, projections, rejections) {
@@ -51,9 +52,9 @@ plot.projections <- function(matrix, projections, rejections) {
     } else {
         for(i in 1:nrow(matrix)) {
             ## Plot the projections
-            lines(rbind(c(0,0), projections[i,]), col = "blue")
+            lines(rbind(c(0,0), projections[i,]), col = "blue", lwd = 1.5)
             ## Plot the rejection
-            lines(rbind(projections[i, ], rejections[i,]+projections[i, ]), col = "orange")
+            lines(rbind(projections[i, ], rejections[i,]+projections[i, ]), col = "orange", lwd = 1.5)
         }        
     }
 }

### 2021-05-27 Analysis/03-elaboration_exploration_analyses.Rmd
Illustration of proj/rej

@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 library(beer)
@@
 data(demo_tree)
+## Setting the colour vector
+colour_vector <- c("orange", "blue", "darkgreen")
 ```
@@
 
-```{r, eval = FALSE}
+```{r, eval = TRUE}
 set.seed(42)
@@
 These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different levels:
+
  * The whole bird phylogeny level ("phylogeny level");
@@
 
-### Details for getting the major axes from the variance co-variance matrices
+## Major axes calculation
 
@@
 
-The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axis` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
+> The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axis` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
+
+```{r}
+## Calculating the major axes
+major_axes <- get.axes(covar_matrices, centre = "intercept")
+```
+
+One can visualise these ellipses and major axes in 2D using the `beer::plot.axes` and `beer::plot.ellipses` functions (based on @ellipse):
+
+```{r}
+## Plotting the ellipses
+plot.ellipses(covar_matrices, col = c("grey",colour_vector),
+              add = FALSE)
+plot.axes(major_axes, col = c("grey",colour_vector),
+              add = TRUE)
+legend("topleft", legend = c("all_clades", levels(demo_data$clade)), col = c("grey", colour_vector), pch = 19)
+```
+
+We can then compare these axes either between each other or relative to the tips positions in the trait space using elaboration and exploration scores:
+
+## Elaboration and exploration scores calculations
+
+From these major axes, we can use linear algebra in $n$ dimensions to estimate clades or tips' exploration and elaboration scores.
+These scores are a direct interpretation of @endler2005's elaboration and innovation (here exploration) patterns (Fig 1. in @endler2005):
+Given a major axis of evolutionary "trajectory" (in the sense of evolutionary trend, not an evolutionary direction), on can calculate the elaboration score as the projection value of any vector defined by a clade or a tip on the major axis and the exploration (or "innovation" in @endler2005) as that vector's rejection.
+
+```{r, echo = FALSE, eval = TRUE}
+source("../Functions/test.rotation.R")
+```
+Figure: in a space with 5 elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
+the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration elaboration scores for either the elements (e.g.
+element A's projection in blue and its rejection in orange) or from any other axes (e.g.
+the major phylogenetic effect axes from the MCMCglmm from a specific clade).
+
+To measure the projection and rejection scores, we used the following algorithm:
+
+ * Rotate
+ * Rescale
+ * Project/reject
 

### 2021-05-28 Analysis/equations.tex
equation draft board

--- 2021-05-27 Analysis/major.axis.tex
@@
 
+For any vectors $\vec{a}$ and $\vec{b}$, defined either as one set or a pair of sets of coordinates in $n$ dimensions:
+
+\begin{equation}
+    \vec{a} = 
+    \begin{bmatrix}
+    x \\
+    y \\
+    ... \\
+    n \\
+    \end{bmatrix}
+\end{equation}
+
+\begin{equation}
+    \vec{a} = 
+    \begin{bmatrix}
+    x_{1} & x_{2} \\
+    y_{1} & y_{2} \\
+    ... & ... \\
+    n_{1} & n_{2} \\
+    \end{bmatrix}
+\end{equation}
+
+We can calculate $\vec{a_{1}}$, the orthogonal projection of $\vec{a}$ onto $\vec{b}$ using:
+
+\begin{equation}
+    \vec{a_{1}} = \frac{\vec{a} \cdot \vec{b}}{\|\vec{b}\|}
+\end{equation}
+
+With $\|\vec{b}\| = \sqrt{\vec{b} \cdot \vec{b}}$ being the norm of $\vec{b}$.
+
+ And $\vec{a_{2}}$, the rejection of $\vec{a}$ onto $\vec{b}$:
+
+\begin{equation}
+    \vec{a_{2}} = \vec{a} - \vec{a_{1}}
+\end{equation}
+
+And $\theta$, the angle formed by $\vec{a}$ and $\vec{b}$:
+
+\begin{equation}
+    \theta = acos \left(\frac{\vec{a} \cdot \vec{b}}{\|\vec{a}\| \cdot \|\vec{b}\|} \right)
+\end{equation}
+
+Given a $n$ dimensional trait space $TS_{m,n}$ with $m$ elements defined as:
+
+\begin{equation*}
+TS_{m,n} = 
+    \begin{bmatrix}
+        x_{1} & x_{2} & \cdots & x_{m} \\
+        y_{1} & y_{2} & \cdots & y_{m} \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        n_{1} & n_{2} & \cdots & n_{m} \\
+    \end{bmatrix}
+\end{equation*}
+
+And a base vector $B_{n}$ defined as:
+
+\begin{equation*}
+B_{n} = 
+    \begin{bmatrix}
+        Bx_{1} & Bx_{2}\\
+        By_{1} & By_{2}\\
+        \vdots  & \vdots \\
+        Bn_{1} & Bn_{2}\\
+    \end{bmatrix}
+\end{equation*}
+
+\begin{equation}
+        cTS = TS - [Bx_{1}, By_{1}, \cdots, Bn_{1}]^{\text{T}}
+\end{equation}
+
+\begin{equation}
+I_{\vec{B}} - \vec{B}\vec{B}^\text{T} - \vec{\hat{\imath}}\vec{\hat{\imath}}^\text{T} + [\vec{B} \vec{\hat{\imath}}]     \begin{bmatrix}
+        cos(\theta) & -sin(\theta)\\
+        sin(\theta) & cos(\theta)\\
+    \end{bmatrix} [\vec{B} \vec{\hat{\imath}}]^\text{T}
+\end{equation}
+
+Where $\theta$ is:
+
+\begin{equation}
+    \theta = acos \left(\frac{\vec{B} \cdot \vec{\hat{\imath}}}{\|\vec{B}\| \cdot \|\vec{\hat{\imath}}\|} \right)
+\end{equation}
+
+Or $\theta = acos (B_x)$ since both \|\vec{B}\| and \|\vec{\hat{\imath}}\| are equal to 1 and \|\vec{\hat{\imath}}\| is the unit vector on the first axis.
+
 \end{document}

### 2021-05-28 Analysis/03-elaboration_exploration_analyses.Rmd
added all the math!

@@
 
-\begin{equation*}
+\begin{equation}
 O_{n} = 
@@
     \end{pmatrix}
-\end{equation*}
+\end{equation}
 
@@
 
-\begin{equation*}
+\begin{equation}
 VCV_{n} = 
@@
 
-To measure the projection and rejection scores, we used the following algorithm:
+#### Measuring projection and rejection
 
- * Rotate
- * Rescale
- * Project/reject
+For any vectors $\vec{a}$ and $\vec{b}$, defined either as one set or a pair of sets of coordinates in $n$ dimensions:
+
+\begin{equation}
+    \vec{a} = 
+    \begin{bmatrix}
+    x \\
+    y \\
+    \cdots \\
+    n \\
+    \end{bmatrix}
+\end{equation}
+
+\begin{equation}
+    \vec{a} = 
+    \begin{bmatrix}
+    x_{1} & x_{2} \\
+    y_{1} & y_{2} \\
+    \cdots & \cdots \\
+    n_{1} & n_{2} \\
+    \end{bmatrix}
+\end{equation}
+
+We can calculate $\vec{a_{1}}$, the orthogonal projection of $\vec{a}$ onto $\vec{b}$ using:
+
+\begin{equation}
+    \vec{a_{1}} = \frac{\vec{a} \cdot \vec{b}}{\|\vec{b}\|}
+\end{equation}
+
+With $\|\vec{b}\| = \sqrt{\vec{b} \cdot \vec{b}}$ being the norm of $\vec{b}$.
+And $\vec{a_{2}}$, the rejection of $\vec{a}$ onto $\vec{b}$:
+
+\begin{equation}
+    \vec{a_{2}} = \vec{a} - \vec{a_{1}}
+\end{equation}
+
+#### Generalisation of projection onto any vector in a set space
+
+Using this, we can calculate the projection and rejection for any element within a trait space $TS_{m,n}$:
+
+\begin{equation}
+    TS_{m,n} = 
+    \begin{bmatrix}
+    x_{1} & x_{2} & \cdots & x_{m} \\
+    y_{1} & y_{2} & \cdots & y_{m} \\
+    \vdots  & \vdots  & \ddots & \vdots \\
+    n_{1} & n_{2} & \cdots & n_{m} \\
+    \end{bmatrix}
+\end{equation}
+
+And any base vector $\vec{B}$ defined as:
+
+\begin{equation}
+    B = 
+    \begin{bmatrix}
+    x_{1} & x_{2}\\
+    y_{1} & y_{2}\\
+    \vdots  & \vdots  \\
+    n_{1} & n_{2} \\
+    \end{bmatrix}
+\end{equation}
+
+By using the linear transformation $f_{\vec{B}}$ of the trait space $TS$ moving $\vec{B}$ onto $TS$'s first axes unit vector $\vec{\hat{\imath}}$:
+
+\begin{equation}
+    f_{\vec{B}}(TS) = \left( \frac{TS - [Bx_{1}, By_{1}, \cdots, Bn_{1}]^{\text{T}}}{\|\vec{B}\|} \right) \cdot R_{\vec{B}}
+\end{equation}
+
+With R_{\vec{B}} being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:
+
+\begin{equation}
+R_{\vec{B}} = I_{\vec{B}} - \vec{B}\vec{B}^\text{T} - \vec{\hat{\imath}}\vec{\hat{\imath}}^\text{T} + [\vec{B} \vec{\hat{\imath}}]     \begin{bmatrix}
+        cos(\theta) & -sin(\theta)\\
+        sin(\theta) & cos(\theta)\\
+    \end{bmatrix} [\vec{B} \vec{\hat{\imath}}]^\text{T}
+\end{equation}
+
+Where $\theta$ is:
+
+\begin{equation}
+    \theta = acos \left(\frac{\vec{B} \cdot \vec{\hat{\imath}}}{\|\vec{B}\| \cdot \|\vec{\hat{\imath}}\|} \right)
+\end{equation}
+
+Or $\theta = acos (B_x)$ since both $\|\vec{B}\|$ and $\|\vec{\hat{\imath}}\|$ are equal to 1 and $\|\vec{\hat{\imath}}\|$ is the unit vector on the first axis.
+
+#### Algorithm for calculating the projection/rejection of any element in a defined space
+
+In practice we followed [this procedure](https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions) and applied a modification of [this implementation](https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385) (see @aguilera2004 for the formal generalisation of this algorithm) using the following algorithm:
+
+ 1. In the trait space, define $\vec{B}$ as the base vector (typically $\vec{B}$ is defined as the pair of coordinates set from the major axis described above).
+ 2. Centre the trait space on the origin of $\vec{B}$ so that the first set of coordinates of $\vec{B}$ are 0.
+ 3. Scale the trait space to the norm of $\vec{B}$ so that the norm of $\vec{B}$ is now 1.
+ 4.
+Rotate the trait space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \arrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait space - typically the x axes unit vector).
+ 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now it's projection onto $\vec{B}$.
 

### 2021-05-28 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

@@
 
+Following the ways to:
+ 1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
+ 2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
+ 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes;
+
+We can measure these scores for each clade and each species separately.
+
+## Elaboration and exploration by clade
+
+To run the analyses per clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).
+
+To do so, for randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
+For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).
+
+```{r}
+## Get all the angles and stuff
+group_results <- analyses.group(major_axes[1:4], base = "animal")
+names(group_results) <- levels(demo_data$clade)
+```
+
+The results are reported as boxplots of **log values** for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.
+
+```{r, fig.height = 18, fig.heigh = 6}
+## Plot the blob wise metrics
+plot.analyses.group(lapply(group_results, log), col = colour_vector)
+```
+
+## Elaboration and exploration by species
+
+Similarly to the analyses described above, we ran the analyses per species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results per groups.
+
+```{r}
+## Global analyses
+tip_results_global <- analyses.tip(data = demo_data[, c(1:4)], axes = major_axes$animal)
+```
+
+And here are the **log values** plotted by group:
+
+```{r}
+## Plot results
+plot.analyses.tip(lapply(tip_results_global, log), col = colour_vector, group = split(rownames(demo_data), f = demo_data$clade))
+```
+
+Note that here only the tips positions projected on the global phylogenetic axes.
+
+We also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.
+
+TODO!
+
 ## References

### 2021-06-02 Analysis/03-elaboration_exploration_analyses.Rmd
analyses up

@@
 ## Selecting 1000 random covariance matrices from the MCMCglmm
-covar_matrices <- get.covar(combined_results, n = 100)
+covar_matrices <- get.covar(combined_results, n = 1000)[1:4]
 ```
@@
 
-> The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axis` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
+Finally, the centred the matrix $M_{1,m}$ on the centroid of each clade.
+
+> The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axes` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
 
 ```{r}
+## Get the centre of each clade
+group_centres <- list(
+    "gulls"      = colMeans(demo_data[demo_data$clade == "gulls", c(1:4)]),
+    "plovers"    = colMeans(demo_data[demo_data$clade == "plovers", c(1:4)]),
+    "sandpipers" = colMeans(demo_data[demo_data$clade == "sandpipers", c(1:4)]),
+    "phylogeny"  = colMeans(demo_data[, c(1:4)]))
+
 ## Calculating the major axes
-major_axes <- get.axes(covar_matrices, centre = "intercept")
+major_axes <- get.axes(covar_matrices, centre = group_centres)
 ```
@@
 ```{r}
+set.seed(51)
+## Sub-selecting 5 random VCV matrices for readibility
+sub_sample_matrices <- get.covar(combined_results, 5)[1:4]
+sub_sample_major_axes <- get.axes(sub_sample_matrices, centre = "intercept")
+
 ## Plotting the ellipses
-plot.ellipses(covar_matrices, col = c("grey",colour_vector),
-              add = FALSE)
-plot.axes(major_axes, col = c("grey",colour_vector),
-              add = TRUE)
-legend("topleft", legend = c("all_clades", levels(demo_data$clade)), col = c("grey", colour_vector), pch = 19)
+plot.ellipses(sub_sample_matrices, col = c("grey",colour_vector),
+              add = FALSE, transparent.scale = 0.7)
+plot.axes(sub_sample_major_axes, col = c("grey",colour_vector),
+              add = TRUE, transparent.scale = 0.7)
+legend("topleft", legend = c("phylogeny", levels(demo_data$clade)), col = c("grey", colour_vector), lty = 1)
+```
+
+We can also visualise all the major axes in relation to the elements in the space.
+
+```{r, fig.height = 18, fig.width = 18}
+## Plotting it
+plot.space(demo_data,
+           col = colour_vector,
+           levels = demo_data$clade,
+           xlab = "PC1 (90.5%)",
+           ylab = "PC1 (6.86%)",
+           xlim = c(-2, 2),
+           ylim = c(-1, 1))
+points(do.call(rbind,group_centres)[,c(1,2)],
+       pch = 13, col = c(colour_vector, "grey"))
+plot.axes(major_axes["animal"], col = "grey",
+          add = TRUE, transparent.scale = 0.01,
+          centre = 0)
+plot.axes(major_axes["animal:clade1"], col = colour_vector[1],
+          add = TRUE, transparent.scale = 0.01,
+          centre = colMeans(demo_data[demo_data$clade == "gulls", c(1,2)]))
+plot.axes(major_axes["animal:clade2"], col = colour_vector[2],
+          add = TRUE, transparent.scale = 0.01,
+          centre = colMeans(demo_data[demo_data$clade == "plovers", c(1,2)]))
+plot.axes(major_axes["animal:clade3"], col = colour_vector[3],
+          add = TRUE, transparent.scale = 0.01,
+          centre = colMeans(demo_data[demo_data$clade == "sandpipers", c(1,2)]))
+legend("topleft", legend = c("phylogeny", levels(demo_data$clade)), col = c("grey", colour_vector), lty = c(1,1,1,1), pch = c(NA, 19, 19, 19))
 ```
@@
 ```
+
 Figure: in a space with 5 elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
@@
 ## Get all the angles and stuff
-group_results <- analyses.group(major_axes[1:4], base = "animal")
+group_results <- analyses.group(major_axes, base = "animal")
 names(group_results) <- levels(demo_data$clade)
@@
 
-The results are reported as boxplots of **log values** for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.
+The results are reported as boxplots for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.
 
-```{r, fig.height = 18, fig.heigh = 6}
+```{r, fig.height = 10, fig.heigh = 10}
 ## Plot the blob wise metrics
-plot.analyses.group(lapply(group_results, log), col = colour_vector)
+plot.analyses.group(group_results, col = colour_vector)
 ```
@@
 
-And here are the **log values** plotted by group:
+And here are the values plotted by group:
 
@@
 ## Plot results
-plot.analyses.tip(lapply(tip_results_global, log), col = colour_vector, group = split(rownames(demo_data), f = demo_data$clade))
+plot.analyses.tip(tip_results_global, col = colour_vector, group = split(rownames(demo_data), f = demo_data$clade))
 ```

### 2021-06-02 beer/R/get.axes.R
up package

@@ -12,7 +12,7 @@
 #' \code{centre} can be one of the following:
 #' \itemize{
 #'      \item \code{"intercept"} to use the estimated intercept is used (i.e. \code{MCMCglmm$Sol});
-#'      \item a \code{numeric} of one or 2 elements to be used as the x/y coordinates of the centre
+#'      \item a list of \code{numeric} coordinates to be applied to each level.
 #'      \item a \code{function} to estimate the central tendency of the estimated intercepts (e.g. \code{mean})
 #' }
 #' 
@@ -29,8 +29,12 @@ get.axes <- function(beer, axis = 1, level = 0.95, dimensions, centre = "interce
         dimensions <- 1:dim(beer[[1]][[1]]$VCV)[1]
     }
 
-    ## Recentreing the levels
-    recentred_beer <- lapply(beer, recentre.levels, dimensions, centre)
+    ## Recentreing the levels
+    if(is(centre, "list")) {
+        recentred_beer <- mapply(recentre.levels, beer, centre, MoreArgs = list(dimensions = dimensions), SIMPLIFY = FALSE)
+    } else {
+        recentred_beer <- lapply(beer, recentre.levels, centre, dimensions)
+    }
 
     ## Get all the axes per level
     return(lapply(recentred_beer, lapply, get.one.axis, axis, level, dimensions))
@@ -47,7 +51,7 @@ get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {
     ## Get the data dimensionality
     dims <- dim(data$VCV)[1]
 
-    ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
+    ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
     unit_hypersphere1 <- unit_hypersphere2 <- matrix(0, ncol = dims, nrow = dims)
     ## The "front" (e.g. "top", "right") units
     diag(unit_hypersphere1) <- 1

### 2021-06-02 beer/R/plot.ellipses.R

@@ -66,10 +66,10 @@ replace.intercept <- function(level_sample, value, dimensions) {
     lapply(level_sample, function(X) {X$Sol[dimensions] <- value[dimensions]; return(X)})
 }
 ## Internal: changing the intercept wrapper
-recentre.levels <- function(level_sample, dimensions, centre) {
+recentre.levels <- function(level_sample, centre, dimensions) {
     ## Centre the ellipse
     if(is(centre, "function")) {
-        ## Get the central tendency (as a function)
+        ## Get the central tendency (as a function)
         centre_values <- apply(do.call(rbind, lapply(level_sample, `[[`, "Sol")), 2, centre)
         ## Recentre the intercepts
         level_sample <- replace.intercept(level_sample, value = centre_values, dimensions)
@@ -88,8 +88,8 @@ recentre.levels <- function(level_sample, dimensions, centre) {
 ## Internal: making a list of ellipses for the level
 level.ellipses <- function(level_sample, dimensions, npoints, centre) {
 
-    ## Recentreing the levels
-    level_sample <- recentre.levels(level_sample, dimensions, centre)
+    ## Recentreing the levels
+    level_sample <- recentre.levels(level_sample, centre, dimensions)
 
     ## Get the ellipses for the level
     return(lapply(level_sample, make.ellipse, dimensions, npoints))

### 2021-06-03 beer/R/analyses.tip.R
Updated analyses wrapper

@@ -3,7 +3,8 @@
 #' @description Get results per tip
 #'
 #' @param data the data set to make the projection/rejections
-#' @param axes the main axes to on which to project the data
+#' @param axes the main axes to on which to project the data or a list of main axes corresponding to the groups (if group is not missing)
+#' @param group optional, a list of groups to run the analyses within groups
 #' 
 #' 
 #' @examples
@@ -12,21 +13,41 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-analyses.tip <- function(data, axes) {
-    ## Scale the data to the same number of dimensions than in the axes
-    dimensions <- dim(axes[[1]])[2]
-    if(dim(data)[2] > dimensions) {
-        data <- data[1:dimensions]
-    }
+analyses.tip <- function(data, axes, group) {
 
-    ## Run the analyses
-    results <- lapply(axes, lapply.projections, data)
-    projections <- do.call(cbind, lapply(results, `[[`, "projection"))
-    rejections <- do.call(cbind, lapply(results, `[[`, "rejection"))
-    if(!is.null(rownames(data))) {
-        rownames(projections) <- rownames(rejections) <- rownames(data)
-    }
+    ## Global analyses
+    if(missing(group)) {
+        ## Scale the data to the same number of dimensions than in the axes
+        dimensions <- dim(axes[[1]])[2]
+        if(dim(data)[2] > dimensions) {
+            data <- data[1:dimensions]
+        }
 
+        ## Run the analyses
+        results <- lapply(axes, lapply.projections, data)
+        projections <- do.call(cbind, lapply(results, `[[`, "projection"))
+        rejections <- do.call(cbind, lapply(results, `[[`, "rejection"))
+        if(!is.null(rownames(data))) {
+            rownames(projections) <- rownames(rejections) <- rownames(data)
+        }
+
+    } else {
+        if(length(group) !=  length(axes)) {
+            stop(paste0(length(group), " groups are used as a factor but ", length(axes), " axes groups are provided."))
+        }
+        ## Analyses per groups
+        dimensions <- dim(axes[[1]][[1]])[2]
+        ## Split the data
+        data <- lapply(group, function(x, data) return(data[x, ]), data)
+
+        ## Calculating the results per axes
+        results <- mapply(mapply.projections, axes, data, SIMPLIFY = FALSE)
+
+        ## Combining the results
[…]

### 2021-06-03 Analysis/03-elaboration_exploration_analyses.Rmd
Updated analyses with within clade proj/rej

@@
 plot.ellipses(sub_sample_matrices, col = c("grey",colour_vector),
-              add = FALSE, transparent.scale = 0.7)
+              add = FALSE, transparent.scale = 0.5)
 plot.axes(sub_sample_major_axes, col = c("grey",colour_vector),
               add = TRUE, transparent.scale = 0.7)
-legend("topleft", legend = c("phylogeny", levels(demo_data$clade)), col = c("grey", colour_vector), lty = 1)
+legend("topleft",
+       legend = c("phylogeny", levels(demo_data$clade)),
+       col = c("grey", colour_vector), lty = 1)
 ```
@@
 
-```{r, fig.height = 18, fig.width = 18}
+```{r, fig.height = 10, fig.width = 10}
 ## Plotting it
@@
           centre = colMeans(demo_data[demo_data$clade == "sandpipers", c(1,2)]))
-legend("topleft", legend = c("phylogeny", levels(demo_data$clade)), col = c("grey", colour_vector), lty = c(1,1,1,1), pch = c(NA, 19, 19, 19))
+legend("topleft",
+       legend = c("phylogeny", levels(demo_data$clade)),
+       col = c("grey", colour_vector),
+       lty = c(1,1,1,1),
+       pch = c(NA, 19, 19, 19))
 ```
@@
 
-In practice we followed [this procedure](https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions) and applied a modification of [this implementation](https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385) (see @aguilera2004 for the formal generalisation of this algorithm) using the following algorithm:
+In practice we followed [this procedure](https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions) and applied a modification of [this implementation](https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385) (see @aguilera2004 for the formal generalisation of this algorithm) using the following algorithm implemented in `dispRity::projections` (@disprity):
 
@@
 
-```{r, fig.height = 10, fig.heigh = 10}
+```{r}
 ## Plot the blob wise metrics
@@
 ## Global analyses
-tip_results_global <- analyses.tip(data = demo_data[, c(1:4)], axes = major_axes$animal)
+tip_results_global <- analyses.tip(data = demo_data[, c(1:4)],
+                                   axes = major_axes$animal)
 ```
@@
 ## Plot results
-plot.analyses.tip(tip_results_global, col = colour_vector, group = split(rownames(demo_data), f = demo_data$clade))
+plot.analyses.tip(tip_results_global,
+                  col = colour_vector,
+                  group = split(rownames(demo_data), f = demo_data$clade),
+                  main = "Tips vs. whole phylogeny")
 ```
@@
 
-We also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.
+Finaly, we also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.
 
-TODO!
+```{r}
+## Results within groups
+tip_results_within <- analyses.tip(
+                    data = demo_data[, c(1:4)],
+                    axes = major_axes[-length(major_axes)],
+                    group = split(rownames(demo_data), f = demo_data$clade))
+```
+
+```{r}
+## Plot results
+plot.analyses.tip(tip_results_within,
+                  col = colour_vector,
+                  group = split(rownames(demo_data), f = demo_data$clade),
+                  main = "Tips vs. clade")
+```
 

### 2021-06-04 Analysis/03-elaboration_exploration_analyses.Rmd
Up

@@
 
-Finaly, we also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.
+Finally, we also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.
 
@@
 
+But this is probably best to look at in a species by species fashion:
+
+TODO
+
 ## References

### 2021-06-07 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd
analyses up
Update 02-MCMCglmm_mini_chains.Rmd
Small edits for clarity and typos

@@
 All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
-Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
-Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
+Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
+Some analyses in `beer` do take some time to run so we advise you do not operate heavy machinery while using it.
 Don't use it while driving.
 The following code snippets are used to illustrate the implementation of the functions.
-See §§§ for running the actual analyses.
+See <!-- §§§run_replicates_script --> for running the actual analyses.
 
@@
 
-```{r load_beer, eval = TRUE}
+```{r load_beer, eval = TRUE, message = FALSE}
 set.seed(42)
@@
 
-To estimate the variance covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed model (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
+To estimate the variance covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
 
@@
 
-We ran a general multi-response model with `§§§N_traits` trait of the trait space being the the fixed effects (using a gaussian trait distribution) and and the residual covariance structure.
+We ran a general multi-response model with `§§§N_traits` of the trait space as fixed effects (using a Gaussian distribution) and and the residual covariance structure.
 We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).
+% NC: I know this is in the data prep, but can you specify an example MCMCglmm model here to make it clearer what you mean?
 
@@
 
-In order to improve the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
-This approach is not too dissimilar from @mulTree as it runs multiple MCMCglmm analyses on multiple trees and pull the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty.
-However, the "mini chains" approach differ by the size of the chains that is optimised for speed.
+To increase the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
+It runs multiple MCMCglmm analyses on multiple trees and pulls the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty (similar to @mulTree - ADD LINK?).
+The size of the chains are optimised for speed. % NC: Editted as multree is not familiar to everyone, so keeping it general and linking to multree is better
 
@@
 We first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
-We ran these models (for `§§§N_generations` with sampling every `§§§N_nitt` and no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis).
+We ran these models (for `§§§N_generations` sampling every `§§§N_nitt` with no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis). 
+% NC: I thought in MCMCglmm the nitt parameter was total number of iterations, and thin gives how often the chain is sampled?
+I think you've called nitt generations here but better to be consistent with MCMCglmm?
 
@@
 
-We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
-Each mini-chain thus provided us with a 100 exploitable posterior data points.
-The two main advantages of this mini-chains is that 1) they are much faster to run since no diagnoses is necessary and the chains are only ran for a relatively short time (which allow several chains to crash/fail without loosing all the data); and 2) they take into account tree uncertainty whitout having to run the complete MCMCglmm on all trees (c.f.
+We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree, and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
+Each mini-chain thus provided us with 100 posterior data points.
+The two main advantages of this mini-chain approach is that 1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and 2) they take into account tree uncertainty without having to run the complete MCMCglmm on all trees (c.f.
 @mulTree).
@@
 ```{r, set_the_mini_chains, eval = TRUE}
-## The thining parameter
+## The thinning parameter
 thin <- 100
@@
 
-We then run the MCMCglmm over $N_trees$ to get an estimate of the variance co-variance matrices including phylogenetic uncertainty.
+We then run the MCMCglmm over $N_trees$ to get an estimate of the variance covariance matrices including phylogenetic uncertainty.
 
-```{r, running_replicates, eval = TRUE}
+```{r, running_replicates, eval = FALSE}
 ## The number of replicates
@@
 
-#### Number of runs optimisation (TO BE DEVELOPPED?)
+<!--
+
+#### Number of runs optimisation (TO BE DEVELOPED?)
 
@@
 
-To make sure we reached consistent results, we run the a set of N mini-chains were N was determined using the following variance optimisation algorithm
+To make sure we reached consistent results, we run the set of N mini-chains were N was determined using the following variance optimisation algorithm.
 
@@
  * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
- * 3. update N to N = N * 1.05 and run an additional 5% chains
+ * 3. update N to N = N * 1.05 and run an additional 5% chains;
  * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
- * 5. calculate $\Delta_V$ the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
+ * 5. calculate $\Delta_V$, the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
  * 6.
@@
 
-After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "normal" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.
+After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "standard" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.
+
+-->
 

### 2021-06-07 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd
Minore edits and typos

@@
 All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
-Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
-Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
+Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
+Some analyses in `beer` do take some time to run so we advise users not to operate heavy machinery while using it.
 Don't use it while driving.
@@
 
-To analyse the exploration and elaboration aspects of both clades and tips, we first selected a random 1000 covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.
+To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.
 
@@
 
-We will use these covariance matrices as the basis of the phylogenetic main axes of variation.
-These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different levels:
+We use these covariance matrices as the basis of the phylogenetic main axes of variation.
+These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different phylogenetic levels:
 
@@
 
-These major axes where calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e. the major axes of the 95% CI ellipse of matrix).
-In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigen values and the variance-covariance matrix's eigen vectors.
+These major axes were calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e.
+the major axes of the 95% CI ellipse of matrix).
+% NC: A small figure here would help
+In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors.
 
 ## Major axes calculation
+% NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!
 
-For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and of radius being composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
+For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
 
@@
 
-(in other words $O_{n}$ is the matrix representing each edge point of the unit hyper sphere of $n$ dimensions and of radius $1$);
+(in other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$);
 
-And $O'_{n}$ the scaled matrix hyper sphere to the right confidence interval size using the $\chi^2$ distribution:
+And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:
 
@@
 
-We can get $M_{n}$ matrix containing all the edges' coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
+We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
 
@@
 
-Where $M_{1,n}$ is the major axes of the 0.95 CI hyper ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second second axes (minor axes in 2D), etc...
+Where $M_{1,n}$ is the major axis of the 0.95 CI hyper-ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second axis (minor axis in 2D), etc. etc.
 
-Finally, the centred the matrix $M_{1,m}$ on the centroid of each clade.
+Finally, we centred the matrix $M_{1,m}$ on the centroid of each clade.
 
@@
 
-We can then compare these axes either between each other or relative to the tips positions in the trait space using elaboration and exploration scores:
+We can then compare these axes either with each other or relative to the tip positions in the trait space using elaboration and exploration scores as described below.
 
@@
 
-From these major axes, we can use linear algebra in $n$ dimensions to estimate clades or tips' exploration and elaboration scores.
+From the major axes calculated above, we can use linear algebra in $n$ dimensions to estimate exploration and elaboration scores for clades or tips of the tree.
 These scores are a direct interpretation of @endler2005's elaboration and innovation (here exploration) patterns (Fig 1. in @endler2005):
-Given a major axis of evolutionary "trajectory" (in the sense of evolutionary trend, not an evolutionary direction), on can calculate the elaboration score as the projection value of any vector defined by a clade or a tip on the major axis and the exploration (or "innovation" in @endler2005) as that vector's rejection.
+Given a major axis of evolutionary "trajectory" (in the sense of evolutionary trend, not an evolutionary direction), we can calculate the elaboration score as the projection value of any vector defined by a clade or a tip on the major axis and the exploration (or "innovation" in @endler2005) as that vector's rejection.
 
@@
 
-Figure: in a space with 5 elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
-the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration elaboration scores for either the elements (e.g.
+Figure: in a space with five elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
+the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration and elaboration scores for either the elements (e.g.
 element A's projection in blue and its rejection in orange) or from any other axes (e.g.
-the major phylogenetic effect axes from the MCMCglmm from a specific clade).
+the major phylogenetic effect axes from the MCMCglmm for a specific clade).
 
@@
 Rotate the trait space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \arrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait space - typically the x axes unit vector).
- 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now it's projection onto $\vec{B}$.
+ 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now its projection onto $\vec{B}$.
 
-Following the ways to:
+Following the ways to: %NC: Not sure what you mean here
  1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
  2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
- 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes;
+ 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes.
 
@@
 
-To run the analyses per clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).
+To run the analyses for each clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).
 
-To do so, for randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
+To do so, for a randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
 For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).
@@
 
-Similarly to the analyses described above, we ran the analyses per species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results per groups.
+Similarly to the analyses described above, we ran the analyses for each species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results for each groups.
 
@@
 
-Note that here only the tips positions projected on the global phylogenetic axes.
+Note that here only the tip positions are projected on the global phylogenetic axes.
 
-Finally, we also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.
+Finally, we also looked at the exploration elaboration profiles within each clade, by replacing the projection of the elements onto $\vec{B}$ for each clade to the projection onto their respective clade major axes $\vec{a}$.
 

### 2021-06-07 Analysis/02-MCMCglmm_mini_chains.Rmd
changed var covar to var-covar

@@
 
-To estimate the variance covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
+To estimate the variance-covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
 
@@
 We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).
-% NC: I know this is in the data prep, but can you specify an example MCMCglmm model here to make it clearer what you mean?
+<!-- % NC: I know this is in the data prep, but can you specify an example MCMCglmm model here to make it clearer what you mean? -->
 
@@
 It runs multiple MCMCglmm analyses on multiple trees and pulls the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty (similar to @mulTree - ADD LINK?).
-The size of the chains are optimised for speed. % NC: Editted as multree is not familiar to everyone, so keeping it general and linking to multree is better
+The size of the chains are optimised for speed. <!-- % NC: Editted as multree is not familiar to everyone, so keeping it general and linking to multree is better -->
 
@@
 We ran these models (for `§§§N_generations` sampling every `§§§N_nitt` with no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis). 
-% NC: I thought in MCMCglmm the nitt parameter was total number of iterations, and thin gives how often the chain is sampled?
+<!-- % NC: I thought in MCMCglmm the nitt parameter was total number of iterations, and thin gives how often the chain is sampled?
 I think you've called nitt generations here but better to be consistent with MCMCglmm?
-
+ -->
 ```{r, parametrise_the_model, eval = TRUE}
@@
 
-We then run the MCMCglmm over $N_trees$ to get an estimate of the variance covariance matrices including phylogenetic uncertainty.
+We then run the MCMCglmm over $N_trees$ to get an estimate of the variance-covariance matrices including phylogenetic uncertainty.
 

### 2021-06-07 Analysis/03-elaboration_exploration_analyses.Rmd
changed var covar to var-covar plus added major axis figure

@@
 
-## Extracting the variance covariance matrices from the MCMCglmm mini-chains
+## Extracting the variance-covariance matrices from the MCMCglmm mini-chains
 
@@
 the major axes of the 95% CI ellipse of matrix).
-% NC: A small figure here would help
+<!-- % NC: A small figure here would help -->
+
+```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
+set.seed(42)
+## Space plot
+cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
+space <- space.maker(50, 2, rnorm, cor.matrix = cor_matrix)
+lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
+plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim)
+
+## Plotting the ellipse
+lines(ellipse::ellipse(cor_matrix), col = "blue", lwd = 3)
+## Plotting the major axis
+lines(get.axes(list(list(list(VCV = cor_matrix, Sol = c(0,0)))))[[1]][[1]], lwd = 3, col = "orange")
+```
+
+Figure: the main axis of variation (in orange) of the 95% CI ellipse of the variance-covariance matrix (in blue) of 20 elements in 2 dimensions (in black).
+
 In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors.
@@
 ## Major axes calculation
-% NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!
-
+<!-- % NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!
+ -->
 For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
@@
 
-Following the ways to: %NC: Not sure what you mean here
+Following the ways to: <!-- %NC: Not sure what you mean here -->
  1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);

### 2021-06-07 Analysis/04-full_analyses_run.Rmd
full analyses INIT

---
title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This script can be used to repeat the full analyses.
You can read about the methods overview [here](00-methods_overview.Rmd) or find details about the full analyses in the respective detailed descriptions of the [data](01-Data_preparation.Rmd), the [MCMCglmm mini chains](02-MCMCglmm_mini_chains.Rmd) and the [elaboration exploration analyses](03-elaboration_exploration_analyses.Rmd).
You can reproduce the data preparation, models estimations and calculations of the elaboration and exploration from the [publication](§§§pub) with the [running script](04-full_analyses_run.Rmd).

TODO

### 2021-06-07 Analysis/00-methods_overview.Rmd
Methods overview INIT

--- 2021-06-07 Analysis/03-elaboration_exploration_analyses.Rmd
@@
 
-# Elaboration and exploration analyses
+The following contains an overview of the methods used in the paper.
 
-```{r, eval = TRUE, echo = FALSE}
-library(beer)
-## Loading the MCMCglmm
-load("../Data/processed/combined_results.rda")
-## Loading the tree and data
-data(demo_data)
-data(demo_tree)
-## Setting the colour vector
-colour_vector <- c("orange", "blue", "darkgreen")
-```
+# Data
 
-## Extracting the variance covariance matrices from the MCMCglmm mini-chains
+The detailed and repeatable version of this section can be found [here](01-Data_preparation.Rmd).
 
-To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.
+We used data.
 
-```{r, eval = TRUE}
-set.seed(42)
-## Selecting 1000 random covariance matrices from the MCMCglmm
-covar_matrices <- get.covar(combined_results, n = 1000)[1:4]
-```
+# MCMCglmm
 
-We use these covariance matrices as the basis of the phylogenetic main axes of variation.
-These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different phylogenetic levels:
+The detailed and repeatable version of this section can be found [here](02-MCMCglmm_mini_chains.Rmd).
 
- * The whole bird phylogeny level ("phylogeny level");
- * And the named node levels ("clade levels").
+# Elaboration and exploration
 
-These major axes were calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e.
-the major axes of the 95% CI ellipse of matrix).
-% NC: A small figure here would help
-In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors.
-
-## Major axes calculation
-% NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!
-
-For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
-
-\begin{equation}
-O_{n} = 
-    \begin{pmatrix}
-        1 & 0 & \cdots & 0 \\
-        0 & 1 & \cdots & 0 \\
-        \vdots  & \vdots  & \ddots & \vdots  \\
-        0 & 0 & \cdots & 1 \\
-        -1 & 0 & \cdots & 0 \\
-        0 & -1 & \cdots & 0 \\
-        \vdots  & \vdots  & \ddots & \vdots  \\
-        0 & 0 & \cdots & -1 \\
-    \end{pmatrix}
-\end{equation}
-
-(in other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$);
-
-And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:
-
-\begin{equation}
-    O'_{n} = O_{n} \sqrt{\chi^2(0.95)}
-\end{equation}
-
-For the variance-covariance matrix $VCV_{n}$ of $n$ dimensions
-
-\begin{equation}
-VCV_{n} = 
-    \begin{pmatrix}
-        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
-        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
-        \vdots  & \vdots  & \ddots & \vdots  \\
-        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
-    \end{pmatrix}
-\end{equation}
-
-and the eigenvectors **v** and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
-
-\begin{equation}
-    VCV_{n} \bold{v} = \lambda \bold{v}
-\end{equation}
-
-We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
-
-\begin{equation}
-    M_{n} = [(O'_{n}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
-\end{equation}
-
-Where $M_{1,n}$ is the major axis of the 0.95 CI hyper-ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second axis (minor axis in 2D), etc. etc.
-
-Finally, we centred the matrix $M_{1,m}$ on the centroid of each clade.
-
-> The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axes` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
-
-```{r}
-## Get the centre of each clade
-group_centres <- list(
-    "gulls"      = colMeans(demo_data[demo_data$clade == "gulls", c(1:4)]),
-    "plovers"    = colMeans(demo_data[demo_data$clade == "plovers", c(1:4)]),
-    "sandpipers" = colMeans(demo_data[demo_data$clade == "sandpipers", c(1:4)]),
-    "phylogeny"  = colMeans(demo_data[, c(1:4)]))
-
-## Calculating the major axes
-major_axes <- get.axes(covar_matrices, centre = group_centres)
-```
-
-One can visualise these ellipses and major axes in 2D using the `beer::plot.axes` and `beer::plot.ellipses` functions (based on @ellipse):
-
-```{r}
-set.seed(51)
-## Sub-selecting 5 random VCV matrices for readibility
-sub_sample_matrices <- get.covar(combined_results, 5)[1:4]
-sub_sample_major_axes <- get.axes(sub_sample_matrices, centre = "intercept")
-
-## Plotting the ellipses
-plot.ellipses(sub_sample_matrices, col = c("grey",colour_vector),
-              add = FALSE, transparent.scale = 0.5)
-plot.axes(sub_sample_major_axes, col = c("grey",colour_vector),
-              add = TRUE, transparent.scale = 0.7)
-legend("topleft",
-       legend = c("phylogeny", levels(demo_data$clade)),
-       col = c("grey", colour_vector), lty = 1)
-```
-
-We can also visualise all the major axes in relation to the elements in the space.
-
-```{r, fig.height = 10, fig.width = 10}
-## Plotting it
-plot.space(demo_data,
-           col = colour_vector,
-           levels = demo_data$clade,
-           xlab = "PC1 (90.5%)",
-           ylab = "PC1 (6.86%)",
-           xlim = c(-2, 2),
-           ylim = c(-1, 1))
-points(do.call(rbind,group_centres)[,c(1,2)],
-       pch = 13, col = c(colour_vector, "grey"))
-plot.axes(major_axes["animal"], col = "grey",
-          add = TRUE, transparent.scale = 0.01,
-          centre = 0)
-plot.axes(major_axes["animal:clade1"], col = colour_vector[1],
-          add = TRUE, transparent.scale = 0.01,
-          centre = colMeans(demo_data[demo_data$clade == "gulls", c(1,2)]))
-plot.axes(major_axes["animal:clade2"], col = colour_vector[2],
-          add = TRUE, transparent.scale = 0.01,
-          centre = colMeans(demo_data[demo_data$clade == "plovers", c(1,2)]))
-plot.axes(major_axes["animal:clade3"], col = colour_vector[3],
-          add = TRUE, transparent.scale = 0.01,
-          centre = colMeans(demo_data[demo_data$clade == "sandpipers", c(1,2)]))
-legend("topleft",
-       legend = c("phylogeny", levels(demo_data$clade)),
-       col = c("grey", colour_vector),
-       lty = c(1,1,1,1),
-       pch = c(NA, 19, 19, 19))
-```
-
-We can then compare these axes either with each other or relative to the tip positions in the trait space using elaboration and exploration scores as described below.
-
-## Elaboration and exploration scores calculations
-
-From the major axes calculated above, we can use linear algebra in $n$ dimensions to estimate exploration and elaboration scores for clades or tips of the tree.
-These scores are a direct interpretation of @endler2005's elaboration and innovation (here exploration) patterns (Fig 1. in @endler2005):
-Given a major axis of evolutionary "trajectory" (in the sense of evolutionary trend, not an evolutionary direction), we can calculate the elaboration score as the projection value of any vector defined by a clade or a tip on the major axis and the exploration (or "innovation" in @endler2005) as that vector's rejection.
-
-```{r, echo = FALSE, eval = TRUE}
-source("../Functions/test.rotation.R")
-```
-
-Figure: in a space with five elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
-the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration and elaboration scores for either the elements (e.g.
-element A's projection in blue and its rejection in orange) or from any other axes (e.g.
-the major phylogenetic effect axes from the MCMCglmm for a specific clade).
-
-#### Measuring projection and rejection
-
-For any vectors $\vec{a}$ and $\vec{b}$, defined either as one set or a pair of sets of coordinates in $n$ dimensions:
-
-\begin{equation}
-    \vec{a} = 
-    \begin{bmatrix}
-    x \\
-    y \\
-    \cdots \\
-    n \\
-    \end{bmatrix}
-\end{equation}
-
-\begin{equation}
-    \vec{a} = 
-    \begin{bmatrix}
-    x_{1} & x_{2} \\
-    y_{1} & y_{2} \\
-    \cdots & \cdots \\
-    n_{1} & n_{2} \\
-    \end{bmatrix}
-\end{equation}
-
-We can calculate $\vec{a_{1}}$, the orthogonal projection of $\vec{a}$ onto $\vec{b}$ using:
-
-\begin{equation}
-    \vec{a_{1}} = \frac{\vec{a} \cdot \vec{b}}{\|\vec{b}\|}
-\end{equation}
-
-With $\|\vec{b}\| = \sqrt{\vec{b} \cdot \vec{b}}$ being the norm of $\vec{b}$.
-And $\vec{a_{2}}$, the rejection of $\vec{a}$ onto $\vec{b}$:
-
-\begin{equation}
-    \vec{a_{2}} = \vec{a} - \vec{a_{1}}
-\end{equation}
-
-#### Generalisation of projection onto any vector in a set space
-
-Using this, we can calculate the projection and rejection for any element within a trait space $TS_{m,n}$:
-
-\begin{equation}
-    TS_{m,n} = 
-    \begin{bmatrix}
-    x_{1} & x_{2} & \cdots & x_{m} \\
-    y_{1} & y_{2} & \cdots & y_{m} \\
-    \vdots  & \vdots  & \ddots & \vdots \\
-    n_{1} & n_{2} & \cdots & n_{m} \\
-    \end{bmatrix}
-\end{equation}
-
-And any base vector $\vec{B}$ defined as:
-
-\begin{equation}
-    B = 
-    \begin{bmatrix}
-    x_{1} & x_{2}\\
-    y_{1} & y_{2}\\
-    \vdots  & \vdots  \\
-    n_{1} & n_{2} \\
-    \end{bmatrix}
-\end{equation}
-
-By using the linear transformation $f_{\vec{B}}$ of the trait space $TS$ moving $\vec{B}$ onto $TS$'s first axes unit vector $\vec{\hat{\imath}}$:
-
-\begin{equation}
-    f_{\vec{B}}(TS) = \left( \frac{TS - [Bx_{1}, By_{1}, \cdots, Bn_{1}]^{\text{T}}}{\|\vec{B}\|} \right) \cdot R_{\vec{B}}
-\end{equation}
-
-With R_{\vec{B}} being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:
-
-\begin{equation}
-R_{\vec{B}} = I_{\vec{B}} - \vec{B}\vec{B}^\text{T} - \vec{\hat{\imath}}\vec{\hat{\imath}}^\text{T} + [\vec{B} \vec{\hat{\imath}}]     \begin{bmatrix}
-        cos(\theta) & -sin(\theta)\\
-        sin(\theta) & cos(\theta)\\
-    \end{bmatrix} [\vec{B} \vec{\hat{\imath}}]^\text{T}
-\end{equation}
-
-Where $\theta$ is:
-
-\begin{equation}
-    \theta = acos \left(\frac{\vec{B} \cdot \vec{\hat{\imath}}}{\|\vec{B}\| \cdot \|\vec{\hat{\imath}}\|} \right)
-\end{equation}
-
-Or $\theta = acos (B_x)$ since both $\|\vec{B}\|$ and $\|\vec{\hat{\imath}}\|$ are equal to 1 and $\|\vec{\hat{\imath}}\|$ is the unit vector on the first axis.
-
-#### Algorithm for calculating the projection/rejection of any element in a defined space
-
-In practice we followed [this procedure](https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions) and applied a modification of [this implementation](https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385) (see @aguilera2004 for the formal generalisation of this algorithm) using the following algorithm implemented in `dispRity::projections` (@disprity):
-
- 1. In the trait space, define $\vec{B}$ as the base vector (typically $\vec{B}$ is defined as the pair of coordinates set from the major axis described above).
- 2. Centre the trait space on the origin of $\vec{B}$ so that the first set of coordinates of $\vec{B}$ are 0.
- 3. Scale the trait space to the norm of $\vec{B}$ so that the norm of $\vec{B}$ is now 1.
- 4.
-Rotate the trait space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \arrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait space - typically the x axes unit vector).
- 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now its projection onto $\vec{B}$.
-
-Following the ways to: %NC: Not sure what you mean here
- 1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
- 2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
- 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes.
-
-We can measure these scores for each clade and each species separately.
-
-## Elaboration and exploration by clade
-
-To run the analyses for each clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).
-
-To do so, for a randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
-For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).
-
-```{r}
-## Get all the angles and stuff
-group_results <- analyses.group(major_axes, base = "animal")
-names(group_results) <- levels(demo_data$clade)
-```
-
-The results are reported as boxplots for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.
-
-```{r}
-## Plot the blob wise metrics
-plot.analyses.group(group_results, col = colour_vector)
-```
-
-## Elaboration and exploration by species
-
-Similarly to the analyses described above, we ran the analyses for each species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results for each groups.
-
-```{r}
-## Global analyses
-tip_results_global <- analyses.tip(data = demo_data[, c(1:4)],
-                                   axes = major_axes$animal)
-```
-
-And here are the values plotted by group:
-
-```{r}
-## Plot results
-plot.analyses.tip(tip_results_global,
-                  col = colour_vector,
-                  group = split(rownames(demo_data), f = demo_data$clade),
-                  main = "Tips vs. whole phylogeny")
-```
-
-Note that here only the tip positions are projected on the global phylogenetic axes.
-
-Finally, we also looked at the exploration elaboration profiles within each clade, by replacing the projection of the elements onto $\vec{B}$ for each clade to the projection onto their respective clade major axes $\vec{a}$.
-
-```{r}
-## Results within groups
-tip_results_within <- analyses.tip(
-                    data = demo_data[, c(1:4)],
-                    axes = major_axes[-length(major_axes)],
-                    group = split(rownames(demo_data), f = demo_data$clade))
-```
-
-```{r}
-## Plot results
-plot.analyses.tip(tip_results_within,
-                  col = colour_vector,
-                  group = split(rownames(demo_data), f = demo_data$clade),
-                  main = "Tips vs. clade")
-```
-
-But this is probably best to look at in a species by species fashion:
-
-TODO
+The detailed and repeatable version of this section can be found [here](03-elaboration_exploration_analyses.Rmd).
 

### 2021-06-07 Analysis/00-methods_overview.Rmd
fleshed out manuscript friendly methods

@@
 
-All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
-Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
-Some analyses in `beer` do take some time to run so we advise users not to operate heavy machinery while using it.
-Don't use it while driving.
-The following code snippets are used to illustrate the implementation of the functions.
-See <!-- §§§run_replicates_script --> for running the actual analyses.
+<!-- %TG: this is written for copy/pasting in the methods part of the paper -->
+
+```{r, echo = FALSE, message = FALSE}
+library(dispRity)
+library(beer)
+```
 
@@
 
-The detailed and repeatable version of this section can be found [here](01-Data_preparation.Rmd).
+> The detailed and repeatable version of this section can be found [here](01-Data_preparation.Rmd).
 
-We used data.
+TODO: We used data.
 
@@
 
-The detailed and repeatable version of this section can be found [here](02-MCMCglmm_mini_chains.Rmd).
+> The detailed and repeatable version of this section can be found [here](02-MCMCglmm_mini_chains.Rmd).
+
+To estimate the variance-covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
+
+We ran a general multi-response model with `§§§N_traits` as fixed effects and the residual covariance structure.
+We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade):
+
+```
+Data ~ residuals + whole phylogeny + phylogeny per clade
+```
+
+To increase the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
+It runs multiple small MCMCglmm analyses on multiple trees and pulls the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty (similar to @mulTree).
+
+To do so, we first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
+We ran these models for §§§N_iterations sampling every §§§N generations with no burnin (resulting in an effective sample size of §§§ and a convergence of §§§ measured as the ASDS§§§).
+We then extract the burnin periods from these three chains (defined as the generation when the chain reaches the median likelihood value times 1.25) and used the resulting residuals terms and random terms structure as priors for the mini-chains.
+
+We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree, and 3) the estimated priors and burnin and run the chains to reach only 100 sample after the burnin.
+We ran §§§n_chains mini-chains, thus providing us with `§§§n_chains * 100` posterior data points.
+The two main advantages of this mini-chain approach is that 1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and 2) they take into account tree uncertainty without having to run the complete MCMCglmm on all trees (c.f.
+@mulTree).
+
+![Mini-chains overview](../Manuscript/Figures/mini-chains_diagram.png)
+
+Figure: Mini chains overview.
 
@@
 
-The detailed and repeatable version of this section can be found [here](03-elaboration_exploration_analyses.Rmd).
+> The detailed and repeatable version of this section can be found [here](03-elaboration_exploration_analyses.Rmd).
+
+To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random variance-covariance matrices from the combined mini chains MCMCglmm results.
+This resulted in 1000 $\times$ `§§§N_clades+1` (1000 matrices for each random effect).
+For each of these matrices, we measured the main axis of the 95% confidence interval ellipsoid in `§§§N_dimensions` which corresponds to the main axis of variation for each random effect.
+We then centred each axes on their respective clade's centroid coordinates.
+
+```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
+set.seed(42)
+## Space plot
+cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
+space <- space.maker(50, 2, rnorm, cor.matrix = cor_matrix)
+lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
+plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim)
+
+## Plotting the ellipse
+lines(ellipse::ellipse(cor_matrix), col = "blue", lwd = 3)
+## Plotting the major axis
+lines(get.axes(list(list(list(VCV = cor_matrix, Sol = c(0,0)))))[[1]][[1]], lwd = 3, col = "orange")
+```
+
+Figure: the main axis of variation (in orange) of the 95% CI ellipse of the variance-covariance matrix (in blue) of 20 elements in 2 dimensions (in black).
+
+We then calculated the elaboration and exploration relative to each of these axes using projection and regression from linear algebra.
+The elaboration of a clade or a species is calculated as the linear algebra's projection of the species' coordinates or the clade's major axes onto the main axis of reference.
+The exploration ("innovation" in @endler2005) of a clade or a speciies is calculated as the linear algebra's rejection of it from the main axis of reference.
+In other words the elaboration corresponds to the where does a species or a clade fits on the major evolutionary axis and the exploration corresponds to how far does a species or a clade deviates from that axis.
+
+```{r, echo = FALSE, eval = TRUE}
+source("../Functions/test.rotation.R")
+```
+
+Figure: in a space with five elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
+the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration and elaboration scores for either the elements (e.g.
+element A's projection in blue and its rejection in orange) or from any other axes (e.g.
+the major phylogenetic effect axes from the MCMCglmm for a specific clade).
+
+Using such a way to measure the exploration and elaboration scores for clades or species has the advantage of being generalised to any number of dimensions and to be relative to any main axis of reference.
+In our case we use the major axis of the variance-covariance of the random terms of a MCMCglmm model but this reference could also be an arbitrary axis defined in the space (e.g.
+the "duckness" axis or the "blueness" axis - REF Frane?).
+
+We measured this elaboration and exploration scores for 1) each clade's main axes compared to the overall phylogenetic main axis and 2) each species compared to the overall phylogenetic main axis and their respective clade main axes.
+For the first approach (clade's elaboration and exploration) we also measured the angle between the two axes (because of linear algebra, this measurement is redundant yet sometimes more intuitive). 
+In both approaches, high value of exploration indicates that the clade or the species deviates more from the main axis of reference and elaboration values away from 0.5 means that the clade or species is elaboration on the major axis more than the central/"average" shape.
+
+# Running and summarising the analyses
+
+To fully reproduce the results of the [publication](§§§pub), you can use the [running](04-full_analyses_run.Rmd) and [summarising](05-full_analyses_summary.Rmd).
 

### 2021-06-13 Analysis/04-full_analyses_run.Rmd
Test run for other machine

@@
 
-TODO
+```{r, load pkg and data}
+install.packages(c("devtools", "dispRity", "MCMCglmm"))
+devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+library(beer)
+library(MCMCglmm)
+data(chara_data)
+data(chara_tree)
+```
+
+```{r, run param chains}
+## Set up the parametrising chains models
+param_MCMCglmm <- make.mini.chains(data         = chara_data,
+                                   dimensions   = c(1:17),
+                                   tree         = chara_tree,
+                                   trait.family = "gaussian",
+                                   randoms      = c("global", "clade"),
+                                   residuals    = "global",
+                                   priors       = 0.02,
+                                   verbose      = TRUE,
+                                   parameters   = list(
+                                       nitt   = 5000000,
+                                       thin   = 1000,
+                                       burnin = 0))
+
+## Running the three MCMCglmm models
+param_chains <- run.mini.chains(param_MCMCglmm, replicates = 3)
+save(param_chains, file = "param_chains.rda")
+```
+
+```{r, get params}
+## Extract the parameters from the parameter chains
+est_params <- extract.parameters(param_chains)
+```
+
+```{r, set_the_mini_chains, eval = TRUE}
+## The thinning parameter
+thin <- 1000
+## The number of samples per chains
+samples <- 100
+
+## Set up the mini-chains models
+mini_chains <- make.mini.chains(data         = chara_data[, 1:17],
+                                dimensions   = c(1:17),
+                                tree         = chara_tree,
+                                trait.family = "gaussian",
+                                randoms      = c("global", "clade"),
+                                residuals    = "global",
+                                priors       = est_params$priors,
+                                verbose      = TRUE,
+                                parameters   = list(
+                                    nitt   = est_params$burnin + thin * samples,
+                                    thin   = thin,
+                                    burnin = est_params$burnin))
+```
+
+```{r, run the mini_chains}
+## The number of replicates
+n_reps <- 1000
+
+## Run the required chains
+results <- run.mini.chains(mini_chains, replicates = n_reps, path = "../Data/Processed/MCMC/")
+
+## Combining the results as one big MCMC
+combined_results <- combine.mini.chains(results)
+
+## Saving the results
+save(combined_results, file = "../Data/processed/combined_results.rda")
+```

### 2021-07-14 Analysis/07-benchmarking.Rmd
benchmarking script
TODO

---
title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

Benchmarking `MCMCglmm` running times

```{r, load pkg and data}
library(beer)
library(MCMCglmm)
data(chara_data)
data(chara_tree)
```

Measure:
 * `size` the object size
 * `d_size` the object size difference compared to the first test
 * `r_size` the object size increase rate (estimated as `d_size/iteration_differences`)
 * `time` the elapsed time
 * `d_time` the elapsed time difference compared to the first test
 * `r_time` the elapsed time increase rate (estimated as `d_time/iteration_differences`)

Variables:
 * number of dimensions
 * thinning rate
 * saving parameters from `MCMCglmm`

Saving parameters:
 * `pr = TRUE/FALSE`: save the posteriors of the random effects (FALSE?)
 * `pl = TRUE/FALSE`: save the posteriors of the latent variable (TRUE?)
 * `DIC = TRUE/FALSE`: deviance and deviance information criterion are calculated
 * `saveX = TRUE/FALSE`: save fixed effect design matrix
 * `saveZ = TRUE/FALSE`: save random effect design matrix
 * `saveXL = TRUE/FALSE`: save structural parameter design matrix

Protocol:
 1. Run one chain with the set number of parameters for 10 iterations (thinning = 1)
 2. Run it again for 20 iterations (thinning = 1)
 3. Run it again for 30 iterations (thinning = 1)
 4. Run it again for 40 iterations (thinning = 1)
 5. Run it again for 50 iterations (thinning = 1)

```{r, run param chains}
## Set up the parametrising chains models
param_10g_2d <- make.mini.chains(data         = chara_data,
                                   dimensions   = c(1:2),
                                   tree         = chara_tree,
                                   trait.family = "gaussian",
                                   randoms      = c("global", "clade"),
                                   residuals    = "global",
                                   priors       = 0.02,
                                   verbose      = TRUE,
                                   parameters   = list(
                                       nitt   = 10,
                                       thin   = 1,
                                       burnin = 0))
## Set up the parametrising chains models
param_20g_2d <- make.mini.chains(data         = chara_data,
                                   dimensions   = c(1:2),
                                   tree         = chara_tree,
                                   trait.family = "gaussian",
                                   randoms      = c("global", "clade"),
                                   residuals    = "global",
                                   priors       = 0.02,
                                   verbose      = TRUE,
                                   parameters   = list(
                                       nitt   = 20,
                                       thin   = 1,
                                       burnin = 0))
## Set up the parametrising chains models
param_30g_2d <- make.mini.chains(data         = chara_data,
                                   dimensions   = c(1:2),
                                   tree         = chara_tree,
                                   trait.family = "gaussian",
                                   randoms      = c("global", "clade"),
                                   residuals    = "global",
                                   priors       = 0.02,
                                   verbose      = TRUE,
                                   parameters   = list(
                                       nitt   = 30,
                                       thin   = 1,
                                       burnin = 0))
## Set up the parametrising chains models
param_40g_2d <- make.mini.chains(data         = chara_data,
                                   dimensions   = c(1:2),
                                   tree         = chara_tree,
                                   trait.family = "gaussian",
                                   randoms      = c("global", "clade"),
                                   residuals    = "global",
                                   priors       = 0.02,
                                   verbose      = TRUE,
                                   parameters   = list(
                                       nitt   = 40,
                                       thin   = 1,
                                       burnin = 0))
## Set up the parametrising chains models
param_50g_2d <- make.mini.chains(data         = chara_data,
                                   dimensions   = c(1:2),
                                   tree         = chara_tree,
                                   trait.family = "gaussian",
                                   randoms      = c("global", "clade"),
                                   residuals    = "global",
                                   priors       = 0.02,
                                   verbose      = TRUE,
                                   parameters   = list(
                                       nitt   = 50,
                                       thin   = 1,
                                       burnin = 0))
param_list <- list(param_10g_2d,
                   param_20g_2d,
                   param_30g_2d,
                   param_40g_2d,
                   param_50g_2d)
```

```{r }
## Recording the results
system(paste("echo \"thin, iterations, dimensions, time, d_time, r_time, size, d_size, r_size\" > recording.csv"))

## First iteration
timer_base <- system.time(param_chains <- run.mini.chains(param_list[[1]], replicates = 1))["elapsed"]
chain_size_base <- object.size(param_chains)

## Saving the results
parameters <- paste(1, 10, 2, sep = ", ")
data <- paste(timer_base, 0, timer_base/10, chain_size_base, 0, chain_size_base/10, sep = ", ")
system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))

## Other iterations
## Difference in iterations
d_itt <- 10
timer <- system.time(param_chains <- run.mini.chains(param_list[[2]], replicates = 1))["elapsed"]
chain_size <- object.size(param_chains)
parameters <- paste(1, 10 + d_itt, 2, sep = ", ")
data <- paste(timer, timer-timer_base, (timer-timer_base)/d_itt, chain_size, chain_size-chain_size_base, (chain_size-chain_size_base)/d_itt, sep = ", ")
system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))

```

### 2021-07-14 beer/R/make.mini.chains.R
added ... arg

@@ -11,6 +11,7 @@
 #' @param parameters the named list of parameters for the MCMCglmm (if missing, it is set to \code{list(nitt=1000, burnin = 100, thin = 100)} for a fast test).
 #' @param priors A list of priors or a value of the overal nu parameter for generating flat priors for the \code{\link[MCMCglmm]{MCMCglmm}}.
 #' @param verbose whether to make the \code{\link[MCMCglmm]{MCMCglmm}} verbose (\code{TRUE}; default) or not (\code{FALSE}).
+#' @param ... any other parameters to be passed to \code{\link[MCMCglmm]{MCMCglmm}}.
 #' 
 #' @details
 #' The types of model for the residuals and random terms can be:
@@ -27,7 +28,7 @@
 #' 
 #' @author Thomas Guillerme
 #' @export
-make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", residuals = "global", randoms = "global", parameters, priors = 0.02, verbose = TRUE) {
+make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", residuals = "global", randoms = "global", parameters, priors = 0.02, verbose = TRUE, ...) {
 
     ## Setting the fixed effect model
     ## Is there any clade effect in the model?
@@ -142,7 +143,7 @@ make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", 
     }
 
     ## Setting the tree(s)
-    output <- lapply(tree, function(tree, fixed, random, rvoc, family, data, priors, verbose, parameters)
+    output <- lapply(tree, function(tree, fixed, random, rvoc, family, data, priors, verbose, parameters, ...)
        return(list(data = data,
                    tree = tree,
                    ## The MCMCglmm function
@@ -156,8 +157,9 @@ make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", 
                                               verbose  = verbose,
                                               burnin   = parameters$burnin,
                                               nitt     = parameters$nitt,
-                                              thin     = parameters$thin)))
-                    , fixed, random, rvoc, family, data, priors, verbose, parameters)
+                                              thin     = parameters$thin,
+                                              ...)))
+                    , fixed, random, rvoc, family, data, priors, verbose, parameters, ...)
 
     class(output) <- c("beer", "mini.chains")
     return(output)

### 2021-07-14 Analysis/07-benchmarking.Rmd
Update 07-benchmarking.Rmd

@@
 
-Saving parameters:
- * `pr = TRUE/FALSE`: save the posteriors of the random effects (FALSE?)
- * `pl = TRUE/FALSE`: save the posteriors of the latent variable (TRUE?)
- * `DIC = TRUE/FALSE`: deviance and deviance information criterion are calculated
- * `saveX = TRUE/FALSE`: save fixed effect design matrix
- * `saveZ = TRUE/FALSE`: save random effect design matrix
- * `saveXL = TRUE/FALSE`: save structural parameter design matrix
+> Here I'm not going to test the saving parameters since it's pretty much a constant RAM saving. By removing all the un-necessary stuff to save, we have 4.2 times smaller output!
 
@@
 
-```{r, run param chains}
+```{r}
+## Function for setting up the parameters
+set.params <- function(dimensions, thin, nitt) {
   ## Set up the parametrising chains models
-param_10g_2d <- make.mini.chains(data         = chara_data,
-                                   dimensions   = c(1:2),
+  param1 <- make.mini.chains(data         = chara_data,
+                             dimensions   = c(1:dimensions),
                              tree         = chara_tree,
@@
                              parameters   = list(
-                                       nitt   = 10,
-                                       thin   = 1,
+                                 nitt   = nitt*1,
+                                 thin   = thin,
                                  burnin = 0))
   ## Set up the parametrising chains models
-param_20g_2d <- make.mini.chains(data         = chara_data,
-                                   dimensions   = c(1:2),
+  param2 <- make.mini.chains(data         = chara_data,
+                             dimensions   = c(1:dimensions),
                              tree         = chara_tree,
@@
                              parameters   = list(
-                                       nitt   = 20,
-                                       thin   = 1,
+                                 nitt   = nitt*2,
+                                 thin   = thin,
                                  burnin = 0))
   ## Set up the parametrising chains models
-param_30g_2d <- make.mini.chains(data         = chara_data,
-                                   dimensions   = c(1:2),
+  param3 <- make.mini.chains(data         = chara_data,
+                             dimensions   = c(1:dimensions),
                              tree         = chara_tree,
@@
                              parameters   = list(
-                                       nitt   = 30,
-                                       thin   = 1,
+                                 nitt   = nitt*3,
+                                 thin   = thin,
                                  burnin = 0))
   ## Set up the parametrising chains models
-param_40g_2d <- make.mini.chains(data         = chara_data,
-                                   dimensions   = c(1:2),
+  param4 <- make.mini.chains(data         = chara_data,
+                             dimensions   = c(1:dimensions),
                              tree         = chara_tree,
@@
                              parameters   = list(
-                                       nitt   = 40,
-                                       thin   = 1,
+                                 nitt   = nitt*4,
+                                 thin   = thin,
                                  burnin = 0))
   ## Set up the parametrising chains models
-param_50g_2d <- make.mini.chains(data         = chara_data,
-                                   dimensions   = c(1:2),
+  param5 <- make.mini.chains(data         = chara_data,
+                             dimensions   = c(1:dimensions),
                              tree         = chara_tree,
@@
                              parameters   = list(
-                                       nitt   = 50,
-                                       thin   = 1,
+                                 nitt   = nitt*5,
+                                 thin   = thin,
                                  burnin = 0))
-param_list <- list(param_10g_2d,
-                   param_20g_2d,
-                   param_30g_2d,
-                   param_40g_2d,
-                   param_50g_2d)
-```
-
-```{r }
-## Recording the results
-system(paste("echo \"thin, iterations, dimensions, time, d_time, r_time, size, d_size, r_size\" > recording.csv"))
+  return(list(param1,
+              param2,
+              param3,
+              param4,
+              param5))
+}
+## Function for running the test
+test.params <- function(dimensions, thin, nitt) {
+  ## Setting up the parameters
+  param_list <- set.params(dimensions = dimensions, thin = thin, nitt = nitt)
 
@@
 
-## Saving the results
-parameters <- paste(1, 10, 2, sep = ", ")
-data <- paste(timer_base, 0, timer_base/10, chain_size_base, 0, chain_size_base/10, sep = ", ")
+  ## Preping the results
+  parameters <- paste(thin, nitt, dimensions, sep = ", ")
+  data <- paste(timer_base, 0, timer_base/nitt, chain_size_base, 0, chain_size_base/nitt, sep = ", ")
+  ## Appending the file
   system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))
 
-## Other iterations
-## Difference in iterations
-d_itt <- 10
+  ## Second run
   timer <- system.time(param_chains <- run.mini.chains(param_list[[2]], replicates = 1))["elapsed"]
   chain_size <- object.size(param_chains)
-parameters <- paste(1, 10 + d_itt, 2, sep = ", ")
-data <- paste(timer, timer-timer_base, (timer-timer_base)/d_itt, chain_size, chain_size-chain_size_base, (chain_size-chain_size_base)/d_itt, sep = ", ")
+
+  ## Preping the results
+  parameters <- paste(thin, nitt*2, dimensions, sep = ", ")
+  data <- paste(timer, timer-timer_base, (timer-timer_base)/nitt*1, chain_size, chain_size-chain_size_base, (chain_size-chain_size_base)/nitt*1, sep = ", ")
+  ## Appending the file
   system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))
 
+  ## Third run
+  timer <- system.time(param_chains <- run.mini.chains(param_list[[3]], replicates = 1))["elapsed"]
+  chain_size <- object.size(param_chains)
+
+  ## Preping the results
+  parameters <- paste(thin, nitt*3, dimensions, sep = ", ")
+  data <- paste(timer, timer-timer_base, (timer-timer_base)/nitt*2, chain_size, chain_size-chain_size_base, (chain_size-chain_size_base)/nitt*2, sep = ", ")
+  ## Appending the file
+  system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))
+
+  ## Fourth run
+  timer <- system.time(param_chains <- run.mini.chains(param_list[[4]], replicates = 1))["elapsed"]
+  chain_size <- object.size(param_chains)
+
+  ## Preping the results
+  parameters <- paste(thin, nitt*4, dimensions, sep = ", ")
+  data <- paste(timer, timer-timer_base, (timer-timer_base)/nitt*3, chain_size, chain_size-chain_size_base, (chain_size-chain_size_base)/nitt*3, sep = ", ")
+  ## Appending the file
+  system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))
+
+  ## Fifth run
+  timer <- system.time(param_chains <- run.mini.chains(param_list[[5]], replicates = 1))["elapsed"]
+  chain_size <- object.size(param_chains)
+
+  ## Preping the results
+  parameters <- paste(thin, nitt*5, dimensions, sep = ", ")
+  data <- paste(timer, timer-timer_base, (timer-timer_base)/nitt*4, chain_size, chain_size-chain_size_base, (chain_size-chain_size_base)/nitt*4, sep = ", ")
+  ## Appending the file
+  system(paste("echo \"", paste(parameters, data, sep = ", "), "\" >> recording.csv"))
+
+  return(NULL)
+}
 ```
+
+## Running the tests for 1k generations (sampling every 100)
+
+```{r, eval = TRUE}
+## Initialising the results result table
+system(paste("echo \"thin, iterations, dimensions, time, d_time, r_time, size, d_size, r_size\" > recording.csv"))
+
+## Running the test
+for(dim in 2:17) {
+  test.params(dimensions = dim, thin = 100, nitt = 1000)
+}
+```
+
+```{r}
+## Upload the table
+rable <- read.csv("recording.csv")
+
+## Log values
+run_time <- log(rable[, "time"])
+run_size <- log(rable[, "size"])
+cols <- rainbow(16)
+
+par(mfrow = c(2,1))
+plot(NULL, xlim = c(1, 5), ylim = range(run_time), main = "Run time", xlab = "10s of samples (per 1k iterations)", ylab = "log elapsed seconds")
+for(i in 1:(nrow(rable)/5)) {
+  lines(run_time[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+}
+
+plot(NULL, xlim = c(1, 5), ylim = range(run_size), main = "Object size", xlab = "10s of samples (per 1k iterations)", ylab = "log bytes")
+for(i in 1:(nrow(rable)/5)) {
+  lines(run_size[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+}
+```
+
+## Running the tests for 10k generations (sampling every 1000)
+
+```{r, eval = TRUE}
+## Initialising the results result table
+system(paste("echo \"thin, iterations, dimensions, time, d_time, r_time, size, d_size, r_size\" > recording2.csv"))
+
+## Running the test
+for(dim in 2:17) {
+  test.params(dimensions = dim, thin = 1000, nitt = 10000)
+}
+```
+
+```{r}
+## Upload the table
+rable <- read.csv("recording2.csv")
+
+## Log values
+run_time <- log(rable[, "time"])
+run_size <- log(rable[, "size"])
+cols <- rainbow(16)
+
+par(mfrow = c(2,1))
+plot(NULL, xlim = c(1, 5), ylim = range(run_time), main = "Run time", xlab = "10s of samples (per 10k iterations)", ylab = "log elapsed seconds")
+for(i in 1:(nrow(rable)/5)) {
+  lines(run_time[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+}
+
+plot(NULL, xlim = c(1, 5), ylim = range(run_size), main = "Object size", xlab = "10s of samples (per 10k iterations)", ylab = "log bytes")
+for(i in 1:(nrow(rable)/5)) {
+  lines(run_size[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+}
+```

### 2021-07-20 Analysis/08-Number_of_dimensions.Rmd
Create 08-Number_of_dimensions.Rmd

---
title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

Checking the ideal number of dimensions for running the model.

```{r, load pkg and data}
library(beer)
library(MCMCglmm)
data(chara_data)
data(chara_tree)
```

```{r}
## Get the groups
beaks <- readRDS("MCMCglmm_covar/Data/2020_06_19_MMB_LANDMARKS_ALL.rds")
order_groups <- lapply(as.list(unique(beaks$Order)), function(X) unique(beaks$TipLabel[beaks$Order %in% X]))
names(order_groups) <- unique(beaks$Order)
family_groups <- lapply(as.list(unique(beaks$Family)), function(X) unique(beaks$TipLabel[beaks$Family %in% X]))
names(family_groups) <- unique(beaks$Family)

## Get the PCA
load("MCMCglmm_covar/full_PCAs.rda")

## Selecting axis function
# @params strict whether to have less than the threshold (strict) or at least the threshold (false)
select.axis <- function(ordination, threshold = 0.99, group, details = FALSE, strict = FALSE) {
    ## Set the group
    if(missing(group)) {
        group <- list(rownames(ordination))
    }

    ## Cum sum scaled variance
    get.cum.var <- function(one_group, matrix) {
        if(length(one_group) != 1) {
            vars <- apply(matrix[one_group, ], 2, var)
            return(cumsum(vars/sum(vars)))
        } else {
            return(c(0, rep(1, ncol(matrix)-1)))
        }
    }

    ## Get the variances for each group
    group_vars <- lapply(group, get.cum.var, matrix = ordination)

    select.threshold <- function(x, threshold, strict) {
        if(strict) {
            return(which(x < threshold))
        } else {
            return(c(0, which(x < threshold)) + 1)
        }
    }

    ## Get the axis per threshold
    selected_dims <- lapply(group_vars, select.threshold, threshold = threshold, strict = strict)

    ## Return the results
    if(!details) {
        return(unique(unlist(selected_dims)))
    } else {
        return(selected_dims)
    }
}

## Preparing the result table
thresholds <- seq(from = 0.9, to = 0.99, by = 0.01)
results <- matrix(0, 6, length(thresholds))
rownames(results) <- c("sizefree-all", "sizefree-order", "sizefree-family", "sizeform-all", "sizeform-order", "sizeform-family")
colnames(results) <- thresholds

for(i in 1:length(thresholds)) {
  results[1,i] <- length(select.axis(full_PCAs$sizefree, threshold = thresholds[i]))
  results[2,i] <- length(select.axis(full_PCAs$sizefree, threshold = thresholds[i], group = order_groups))
  results[3,i] <- length(select.axis(full_PCAs$sizefree, threshold = thresholds[i], group = family_groups))
  results[4,i] <- length(select.axis(full_PCAs$sizeform, threshold = thresholds[i]))
  results[5,i] <- length(select.axis(full_PCAs$sizeform, threshold = thresholds[i], group = order_groups))
  results[6,i] <- length(select.axis(full_PCAs$sizeform, threshold = thresholds[i], group = family_groups))
}

## Colour vector
colours <- c("purple", "red", "orange",
             "blue", "cyan", "green")

## Plotting the results
plot(NULL, xaxt = "n", xlim = c(1,10), ylim = range(results), ylab = "Dimensions", xlab = "Thresholds")
axis(1, at = c(1:10), labels = thresholds)
abline(h = 7, col = "lightgrey")
for(i in 1:6) {
  lines(results[i,], col = colours[i])
}
legend("topleft", legend = rownames(results), lty = 1, col = colours)
```

### 2021-07-21 Analysis/08-Number_of_dimensions.Rmd
Updated PC axis analyses

@@
 
-Checking the ideal number of dimensions for running the model.
-
-```{r, load pkg and data}
+```{r, echo = FALSE, print = FALSE}
 library(beer)
@@
 
-```{r}
+## Number of dimensions required to capture at least X% of variance per group
+
+```{r, echo = FALSE}
 ## Get the groups
@@
     select.threshold <- function(x, threshold, strict) {
-        if(strict) {
+        # if(strict) {
             return(which(x < threshold))
-        } else {
-            return(c(0, which(x < threshold)) + 1)
-        }
+        # } else {
+        #     return(c(0, which(x < threshold)) + 1)
+        # }
     }
@@
 ## Colour vector
-colours <- c("purple", "red", "orange",
-             "blue", "cyan", "green")
+colours <- c("orange", "yellow", "pink",
+             "lightblue", "lightgreen", "cyan")
 
@@
 ```
+
+This is the number of dimensions required per PCA contain X% of variance per group.
+The groups being either the full PCA (`all` - the classic way to measure the variance threshold) or each order (`order`) or family (`family`).
+The number of dimensions per group is measured either for the PCA with the size component removed (`sizefree`) and for the PCA with the size component (`sizeform`).
+
+## Variance per specific axis for the 37 first dimensions
+
+```{r, echo = FALSE, print = FALSE}
+## Calculating the variance for each dimension
+get.dim.vars <- function(ordination, group, n_dims) {
+    ## Set the group
+    if(missing(group)) {
+        group <- list(rownames(ordination))
+    }
+    if(missing(n_dims)) {
+        n_dims <- ncol(ordination)
+    }
+
+    ## Cum sum scaled variance
+    get.scaled.var <- function(one_group, matrix, n_dims) {
+        if(length(one_group) != 1) {
+            vars <- apply(matrix[one_group, ], 2, var)
+            return(c(vars/sum(vars))[1:n_dims])
+        } else {
+            return(c(1, rep(0, ncol(matrix)-1))[1:n_dims])
+        }
+    }
+    ## Get the variances for each group
+    return(lapply(group, get.scaled.var, matrix = ordination, n_dims = n_dims))
+}
+## Plotting results
+barplot.results <- function(result, main, threshold = c(0.95, 0.99), col = "orange") {
+  barplot(result, col = col, main = main, ylab = "Scaled variance")
+  abline(v = length(which(cumsum(result) < threshold[1])), lwd = 2)
+  abline(v = length(which(cumsum(result) < threshold[2])), lwd = 2, col = "grey")
+}
+
+## Get all the detailed results
+## Full PCA
+sizefree_fullpca <- get.dim.vars(full_PCAs$sizefree,
+                                 n_dims = 37)
+names(sizefree_fullpca) <- c("Sizefree: all species")
+## Orders
+sizefree_order   <- get.dim.vars(full_PCAs$sizefree,
+                                 group = order_groups,
+                                 n_dims = 37)
+names(sizefree_order) <- paste("Sizefree:", names(order_groups))
+sizefree_order <- sizefree_order[match(sort(names(sizefree_order)), names(sizefree_order))]
+## Family
+sizefree_family  <- get.dim.vars(full_PCAs$sizefree,
+                                 group = family_groups,
+                                 n_dims = 37)
+names(sizefree_family) <- paste("Sizefree:", names(family_groups))
+sizefree_family <- sizefree_family[match(sort(names(sizefree_family)), names(sizefree_family))]
+
+## Full PCA
+sizeform_fullpca <- get.dim.vars(full_PCAs$sizeform,
+                                 n_dims = 37)
+names(sizeform_fullpca) <- c("Sizeform: all species")
+## Orders
+sizeform_order   <- get.dim.vars(full_PCAs$sizeform,
+                                 group = order_groups,
+                                 n_dims = 37)
+names(sizeform_order) <- paste("Sizeform:", names(order_groups))
+sizeform_order <- sizeform_order[match(sort(names(sizeform_order)), names(sizeform_order))]
+## Family
+sizeform_family  <- get.dim.vars(full_PCAs$sizeform,
+                                 group = family_groups,
+                                 n_dims = 37)
+names(sizeform_family) <- paste("Sizeform:", names(family_groups))
+sizeform_family <- sizeform_family[match(sort(names(sizeform_family)), names(sizeform_family))]
+
+```
+
+Here I will display the scaled variance per axis for each group (as mentioned above).
+For example, for the full PCA, the results are pretty common and easy to grasp:
+
+```{r, fig.height = 4, fig.width = 4, echo = FALSE, print = FALSE, result = FALSE}
+## Plotting everything
+silent <- mapply(barplot.results, sizefree_fullpca, as.list(names(sizefree_fullpca)))
+## Plotting everything
+silent <- mapply(barplot.results, sizeform_fullpca, as.list(names(sizeform_fullpca)), MoreArgs = list(col = colours[4]))
+```
+
+When considering all the data, the first three axis contain 95% of the variance (black vertical line) and PC1 to 6 for 99% of the variance (grey vertical line).
+For the PCA with size, the first axes contains 95% of the variance and the PC1 to 3 contains 99%.
+The rest of the variance is distributed pretty log-normally as expected.
+
+We can then measure the same be measuring things per family or per order (i.e. within the PCA, how many axis are required to represent 95% of the variance of order/family X).
+The results vary much more depending on the order/family! For example, in Strigiformes, the first PC does not contain much variance at all compared to the other axes!
+Note that the order with just one bar at 100% just have one species so they don't count.
+
+```{r, fig.height = 4, fig.width = 4, echo = FALSE, print = FALSE, result = FALSE}
+## Plotting everything
+silent <- mapply(barplot.results, sizefree_order, as.list(names(sizefree_order)), MoreArgs = list(col = colours[2]))
+```
+
+```{r, fig.height = 4, fig.width = 4, echo = FALSE, print = FALSE, result = FALSE}
+## Plotting everything
+silent <- mapply(barplot.results, sizefree_family, as.list(names(sizefree_family)), MoreArgs = list(col = colours[3]))
+```
+
+```{r, fig.height = 4, fig.width = 4, echo = FALSE, print = FALSE, result = FALSE}
+## Plotting everything
+silent <- mapply(barplot.results, sizeform_order, as.list(names(sizeform_order)), MoreArgs = list(col = colours[5]))
+```
+
+```{r, fig.height = 4, fig.width = 4, echo = FALSE, print = FALSE, result = FALSE}
+## Plotting everything
+silent <- mapply(barplot.results, sizeform_family, as.list(names(sizeform_family)), MoreArgs = list(col = colours[6]))
+```

### 2021-08-03 Analysis/08.1-Number_of_dimensions_old.Rmd
Updated dimensions analyses

--- 2021-07-21 Analysis/08-Number_of_dimensions.Rmd
@@
 ```{r, echo = FALSE, print = FALSE}
+library(dispRity)
 library(beer)
@@
 load("MCMCglmm_covar/full_PCAs.rda")
+```
 
+```{r}
 ## Selecting axis function

### 2021-08-10 Analysis/02-MCMCglmm_mini_chains.Rmd
typos

@@
  -->
+
 ```{r, parametrise_the_model, eval = TRUE}

### 2021-08-10 Analysis/03-elaboration_exploration_analyses.Rmd
typos

@@
            xlab = "PC1 (90.5%)",
-           ylab = "PC1 (6.86%)",
+           ylab = "PC2 (6.86%)",
            xlim = c(-2, 2),

### 2021-08-10 Analysis/07-benchmarking.Rmd
Update 07-benchmarking.Rmd

@@
 
-```{r, eval = TRUE}
+```{r, eval = FALSE}
 ## Initialising the results result table
@@
 
-```{r, eval = TRUE}
-## Initialising the results result table
-system(paste("echo \"thin, iterations, dimensions, time, d_time, r_time, size, d_size, r_size\" > recording2.csv"))
-
+```{r, eval = FALSE}
 ## Running the test
@@
 ## Upload the table
-rable <- read.csv("recording2.csv")
+rable <- read.csv("recording.csv")
+rable_1k <- rable[1:80, ]
+rable_10k <- rable[-c(1:80), ]
 
 ## Log values
-run_time <- log(rable[, "time"])
-run_size <- log(rable[, "size"])
+run_time_1k <- rable_1k[, "time"]
+run_size_1k <- rable_1k[, "size"]
+run_time_10k <- rable_10k[, "time"]
+run_size_10k <- rable_10k[, "size"]
+
 cols <- rainbow(16)
 
-par(mfrow = c(2,1))
-plot(NULL, xlim = c(1, 5), ylim = range(run_time), main = "Run time", xlab = "10s of samples (per 10k iterations)", ylab = "log elapsed seconds")
-for(i in 1:(nrow(rable)/5)) {
-  lines(run_time[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+par(mfrow = c(2,2))
+plot(NULL, xlim = c(1, 5), ylim = range(run_time_1k), main = "Run time", xlab = "iterations (thin = 100)", ylab = "elapsed seconds", xaxt = "n")
+axis(1, at = c(1:5), labels = seq(from = 1000, to = 5000, by = 1000))
+for(i in 1:(nrow(rable_1k)/5)) {
+  lines(run_time_1k[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
 }
 
-plot(NULL, xlim = c(1, 5), ylim = range(run_size), main = "Object size", xlab = "10s of samples (per 10k iterations)", ylab = "log bytes")
-for(i in 1:(nrow(rable)/5)) {
-  lines(run_size[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+plot(NULL, xlim = c(1, 5), ylim = range(run_time_10k), main = "Run time", xlab = "iterations (thin = 1000)", ylab = "elapsed seconds", xaxt = "n")
+axis(1, at = c(1:5), labels = seq(from = 10000, to = 50000, by = 10000))
+for(i in 1:(nrow(rable_10k)/5)) {
+  lines(run_time_10k[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+}
+
+plot(NULL, xlim = c(1, 5), ylim = range(run_size_1k), main = "Object size", xlab = "iterations (thin = 100)", ylab = "bytes", xaxt = "n")
+axis(1, at = c(1:5), labels = seq(from = 1000, to = 5000, by = 1000))
+for(i in 1:(nrow(rable_1k)/5)) {
+  lines(run_size_1k[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
+}
+
+plot(NULL, xlim = c(1, 5), ylim = range(run_size_10k), main = "Object size", xlab = "iterations (thin = 1000)", ylab = "bytes", xaxt = "n")
+axis(1, at = c(1:5), labels = seq(from = 10000, to = 50000, by = 10000))
+for(i in 1:(nrow(rable_10k)/5)) {
+  lines(run_size_10k[(1+5*(i-1)):(5+5*(i-1))], col = cols[i])
 }

### 2021-08-23 beer/R/sauron.plot.R
added add option

@@ -13,6 +13,7 @@
 #' @param transparent.scale if multiple major axes/ellipses are plotted, a scaling factor for the transparency. If left empty, the transparency is set to \code{1/n} or \code{0.1} (whichever is higher).
 #' @param legend whether to add the automatic legend (\code{TRUE}) or not (\code{FALSE}; default).
 #' @param legend.pos the coordinates of where to add the legend (if \code{legend = TRUE}). Can a numeric vector or a character string like \code{"topleft"} (default), \code{"bottomright"}, etc..
+#' @param add logical, whether to add the plot to an existing plot (\code{TRUE}) or not (\code{FALSE}; default).
 #' @param ... graphic parameters
 #' 
 #' @details
@@ -33,7 +34,7 @@
 #' @author Thomas Guillerme
 #' @export
 
-sauron.plot <- function(data, n, points = TRUE, major.axes = NULL, ellipses = NULL, level = 0.95, dimensions = c(1,2), centres = colMeans, transparent.scale, legend = FALSE, legend.pos = "topleft", ...) {
+sauron.plot <- function(data, n, points = TRUE, major.axes = NULL, ellipses = NULL, level = 0.95, dimensions = c(1,2), centres = colMeans, transparent.scale, legend = FALSE, legend.pos = "topleft", add = FALSE, ...) {
 
     ## Some sanitizing to happen in dispRity on data
 
@@ -191,7 +192,9 @@ sauron.plot <- function(data, n, points = TRUE, major.axes = NULL, ellipses = NU
     }
 
     ## Plotting the background
-    do.call(plot, plot_args)
+    if(!add) {
+        do.call(plot, plot_args)
+    }
 
     ## Adding the points
     if(points) {

### 2021-08-23 Analysis/03-elaboration_exploration_analyses.Rmd
Updated first part of the analyses

@@
 ## Setting the colour vector
-colour_vector <- c("orange", "blue", "darkgreen")
+colour_vector <- c("orange", "blue", "darkgreen", "grey")
 ```
@@
 ```{r, eval = TRUE}
-set.seed(42)
-## Selecting 1000 random covariance matrices from the MCMCglmm
-covar_matrices <- get.covar(combined_results, n = 1000)[1:4]
+covar_matrices <- MCMCglmm.dispRity(
+                  data          = demo_data,
+                  posteriors    = combined_results,
+                  group         = MCMCglmm.levels(combined_results)[1:4],
+                  rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
+```
+
+```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
+## Plotting just the space
+sauron.plot(covar_matrices,
+            col = colour_vector,
+            main = "demo space",
+            legend = TRUE)
 ```
@@
 
-These major axes were calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e.
-the major axes of the 95% CI ellipse of matrix).
-<!-- % NC: A small figure here would help -->
+These major axes were calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e. the major axes of the 95% CI ellipse of matrix).
 
@@
 
-```{r}
-## Get the centre of each clade
-group_centres <- list(
-    "gulls"      = colMeans(demo_data[demo_data$clade == "gulls", c(1:4)]),
-    "plovers"    = colMeans(demo_data[demo_data$clade == "plovers", c(1:4)]),
-    "sandpipers" = colMeans(demo_data[demo_data$clade == "sandpipers", c(1:4)]),
-    "phylogeny"  = colMeans(demo_data[, c(1:4)]))
+One can visualise these ellipses and major axes in 2D using the `sauron.plot`functions with the `major.axis = mean` and `ellipses = mean` options (based on @ellipse) to display the mean major.axis and the mean ellipses:
 
-## Calculating the major axes
-major_axes <- get.axes(covar_matrices, centre = group_centres)
+```{r, fig.height = 6, fig.width = 6}
+sauron.plot(covar_matrices,
+            col = colour_vector,
+            main = "Average ellipses and\nmajor axes",
+            major.axes = mean,
+            ellipses = mean,
+            legend = TRUE)
 ```
 
-One can visualise these ellipses and major axes in 2D using the `beer::plot.axes` and `beer::plot.ellipses` functions (based on @ellipse):
+We can also visualise 100 random major axes and 100 random ellipses:
 
-```{r}
-set.seed(51)
-## Sub-selecting 5 random VCV matrices for readibility
-sub_sample_matrices <- get.covar(combined_results, 5)[1:4]
-sub_sample_major_axes <- get.axes(sub_sample_matrices, centre = "intercept")
-
-## Plotting the ellipses
-plot.ellipses(sub_sample_matrices, col = c("grey",colour_vector),
-              add = FALSE, transparent.scale = 0.5)
-plot.axes(sub_sample_major_axes, col = c("grey",colour_vector),
-              add = TRUE, transparent.scale = 0.7)
-legend("topleft",
-       legend = c("phylogeny", levels(demo_data$clade)),
-       col = c("grey", colour_vector), lty = 1)
-```
-
-We can also visualise all the major axes in relation to the elements in the space.
-
-```{r, fig.height = 10, fig.width = 10}
-## Plotting it
-plot.space(demo_data,
+```{r, fig.height = 12, fig.width = 6}
+par(mfrow = c(2,1))
+## Plotting 100 axes
+sauron.plot(covar_matrices,
+            n = 100,
             col = colour_vector,
-           levels = demo_data$clade,
-           xlab = "PC1 (90.5%)",
-           ylab = "PC2 (6.86%)",
+            main = "100 random major axes",
+            major.axes = TRUE,
+            legend = TRUE,
+            points = FALSE,
             xlim = c(-2, 2),
-           ylim = c(-1, 1))
-points(do.call(rbind,group_centres)[,c(1,2)],
-       pch = 13, col = c(colour_vector, "grey"))
-plot.axes(major_axes["animal"], col = "grey",
-          add = TRUE, transparent.scale = 0.01,
-          centre = 0)
-plot.axes(major_axes["animal:clade1"], col = colour_vector[1],
-          add = TRUE, transparent.scale = 0.01,
-          centre = colMeans(demo_data[demo_data$clade == "gulls", c(1,2)]))
-plot.axes(major_axes["animal:clade2"], col = colour_vector[2],
-          add = TRUE, transparent.scale = 0.01,
-          centre = colMeans(demo_data[demo_data$clade == "plovers", c(1,2)]))
-plot.axes(major_axes["animal:clade3"], col = colour_vector[3],
-          add = TRUE, transparent.scale = 0.01,
-          centre = colMeans(demo_data[demo_data$clade == "sandpipers", c(1,2)]))
-legend("topleft",
-       legend = c("phylogeny", levels(demo_data$clade)),
-       col = c("grey", colour_vector),
-       lty = c(1,1,1,1),
-       pch = c(NA, 19, 19, 19))
+            ylim = c(-2, 2))
+## Plotting 100 ellipses
+sauron.plot(covar_matrices,
+            n = 100,
+            col = colour_vector,
+            main = "100 random ellipses",
+            ellipses = TRUE,
+            legend = TRUE,
+            points = FALSE,
+            xlim = c(-2, 2),
+            ylim = c(-2, 2))
 ```

### 2021-08-27 Analysis/01-Data_preparation.Rmd
analyses up
Update 01-Data_preparation.Rmd

@@
 
+The method described in the next three markdown is essentially a mini chains with pseudo-metropolis coupling markov chain generalised linear mixed model method to determine clades and tips elaboration and exploration (`mcmcpmcmcmcglmmmctee`) but we will refrain using this acronym and method name.
+
 ```{r install_beer, eval = FALSE}
@@
 ```{r load_beer, eval = TRUE}
-## Loading beer
+## Loading packages
+library(dispRity)
 library(beer)
-
-## Loading the demo data
-data(demo_data)
-data(demo_trees)
+library(knitr)
 ```
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
+ * The trait space
+
+```{r}
+## Reading the different trait spaces
+shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")
+shapespace <- shapespace$PCscores
+formspace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_FORMSPACE_FULL.rds")
+formspace <- formspace$PCscores
 ```
 
-## Demo data
+```{r, echo = FALSE, fig.height = 16, fig.width = 8}
+par(mfrow = c(2,1))
+plot.dispRity(shapespace, type = "preview", pch = 21, main = "Shape space")
+plot.dispRity(formspace, type = "preview", pch = 21, main = "Form space")
+```
 
-This is what the demo data looks like
+ * The tree(s)
 
-```{r, plot_the_data}
-## Plotting it
-colour_vector <- c("orange", "blue", "darkgreen")
-plot.space(demo_data,
-           col = colour_vector,
-           levels = demo_data$clade,
-           xlab = "PC1 (90.5%)",
-           ylab = "PC1 (6.86%)")
-legend("topleft", legend = levels(demo_data$clade), col = colour_vector, pch = 19)
+```{r}
+## Reading the consensus tree
+consensus_tree <- read.nexus("../Data/Raw/Tree_data/9993taxa_1_tree_Global_MCC_CA.tre")
+
+## Selecting only the tips in the trait space
+matching_tips <- consensus_tree$tip.label %in% rownames(shapespace)
+consensus_tree <- drop.tip(consensus_tree, tip = consensus_tree$tip.label[!matching_tips])
+```
+
+ * The clade data
+
+```{r}
+## Reading the clades list
+clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
+## Remove the absent species
+clades_list <- clades_list[rownames(clades_list) %in% rownames(shapespace), ]
+
+#@param table the clade list data frame
+#@param level the ID of the column to use as a level
+#@param threshold drop levels with less than the threshold
+make.level.list <- function(table, level, threshold = 20) {
+  ## Select the level
+  table_level <- table[, level]
+  ## Remove the empties
+  table_level <- as.factor(as.character(table_level[-which(table_level == "")]))
+  ## Select the levels below the threshold
+  remove_levels <- which(table(table_level) < threshold)
+  ## Remove the levels
+  while(length(remove_levels) > 0) {
+    table_level <- as.factor(as.character(table_level[-which(table_level == names(remove_levels)[1])]))
+    remove_levels <- remove_levels[-1]
+  }
+  ## Convert into a list
+  factors <- levels(table_level)
+  levels_list <- list()
+  while(length(factors) > 0) {
+    levels_list[[length(levels_list) + 1]] <- which(table_level == factors[1])
+    factors <- factors[-1]
+  }
+  ## Name the elements correctly
+  names(levels_list) <- levels(table_level)
+  return(levels_list)
+}
+
+## Reading the clades list
+groups_level1 <- make.level.list(clades_list, level = 1, threshold = 20)
+groups_level2 <- make.level.list(clades_list, level = 2, threshold = 20)
+all_groups <- c(groups_level1, groups_level2)
+```
+
+### Selecting the number of axes
+
+How many dimensions do we need to represent at least 95% of the variance in each group
+
+```{r}
+## Make a dispRity object
+shapespace <- custom.subsets(shapespace, group = all_groups)
+formspace  <- custom.subsets(formspace, group = all_groups)
+shape_axes <- select.axes(shapespace)
+form_axes  <- select.axes(formspace)
+```
+
+```{r, echo = FALSE}
+kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X1")
+```
+
+```{r, echo = FALSE}
+kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X2")
+```
+
+```{r, echo = FALSE, fig.height = 18, fig.width = 18}
+plot(shape_axes)
+```
+Fig X1: variance per axes for the Shape space
+
+```{r, echo = FALSE, fig.height = 18, fig.width = 18}
+plot(form_axes)
+```
+Fig X1: variance per axes for the Form space
+
+By analysing the distribution of the variance within each group, we will need `r length(shape_axes$dimensions)` dimensions for the shape space and `r length(form_axes$dimensions)` for the form space.
+
+### Saving the data
+
+```{r}
+## Save the data to be analysed in Data/Processed
+form_shape_space_1tree_2levels <- list(
+    "consensus_tree" = consensus_tree,
+    "shape_space"    = shapespace,
+    "form_space"     = formspace,
+    "groups_list"    = all_groups)
+save(form_shape_space_1tree_2levels, file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
 ```

### 2021-08-27 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
  * The clade data
+We used the different clade levels (the 11 magnificent 7 + the families). For each level/group, we removed any groups with less than 20 species.
 
@@
   table_level <- table[, level]
+  ## Find the empty level
+  empty_level <- which(levels(table_level) == "")
+  if(length(empty_level) != 0) {
       ## Remove the empties
-  table_level <- as.factor(as.character(table_level[-which(table_level == "")]))
+      level_list <- as.factor(as.character(table_level[-which(table_level == "")]))
+  }
+  empty_level <- ""
   ## Select the levels below the threshold
-  remove_levels <- which(table(table_level) < threshold)
+  remove_levels <- which(table(level_list) < threshold)
   ## Remove the levels
   while(length(remove_levels) > 0) {
-    table_level <- as.factor(as.character(table_level[-which(table_level == names(remove_levels)[1])]))
+    level_list <- as.factor(as.character(level_list[-which(level_list == names(remove_levels)[1])]))
+    table_level[which(table_level == names(remove_levels)[1])] <- empty_level
+    table_level <- as.factor(as.character(table_level))
     remove_levels <- remove_levels[-1]
@@
   ## Convert into a list
-  factors <- levels(table_level)
-  levels_list <- list()
+  factors <- levels(level_list)
+  list_levels <- list()
   while(length(factors) > 0) {
-    levels_list[[length(levels_list) + 1]] <- which(table_level == factors[1])
+    list_levels[[length(list_levels) + 1]] <- which(level_list == factors[1])
     factors <- factors[-1]
@@
   ## Name the elements correctly
-  names(levels_list) <- levels(table_level)
-  return(levels_list)
+  names(list_levels) <- levels(level_list)
+
+  ## Also returns the new column
+  return(list(col = table_level, list = list_levels))
 }
@@
 groups_level2 <- make.level.list(clades_list, level = 2, threshold = 20)
-all_groups <- c(groups_level1, groups_level2)
+all_groups <- c(groups_level1$list, groups_level2$list)
 ```

### 2021-08-27 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
 ```{r}
+## Add the clade columns and the animal column to the spaces
+shapespace$level1 <- groups_level1$col
+shapespace$level2 <- groups_level2$col
+shapespace$animal <- rownames(shapespace)
+formspace$level1 <- groups_level1$col
+formspace$level2 <- groups_level2$col
+formspace$animal <- rownames(formspace)
+
 ## Save the data to be analysed in Data/Processed
@@
     "groups_list"    = all_groups)
+
 save(form_shape_space_1tree_2levels, file = "../Data/Processed/form_shape_space_1tree_2levels.rda")

### 2021-08-27 Analysis/01-Data_preparation.Rmd
final version for testing data

--- 2021-08-27 Analysis/01-Data_preparation.Rmd
@@
 ```{r}
-## Make a dispRity object
-shapespace <- custom.subsets(shapespace, group = all_groups)
-formspace  <- custom.subsets(formspace, group = all_groups)
-shape_axes <- select.axes(shapespace)
-form_axes  <- select.axes(formspace)
+## Make a dispRity object and get axes data
+shape_axes <- select.axes(custom.subsets(shapespace, group = all_groups))
+form_axes  <- select.axes(custom.subsets(formspace, group = all_groups))
 ```
@@
 ```{r}
+## Add the clade columns and the animal column to the spaces
+shapespace <- as.data.frame(shapespace)
+shapespace$level1 <- groups_level1$col
+shapespace$level2 <- groups_level2$col
+shapespace$animal <- rownames(shapespace)
+formspace <- as.data.frame(formspace)
+formspace$level1 <- groups_level1$col
+formspace$level2 <- groups_level2$col
+formspace$animal <- rownames(formspace)
+
 ## Save the data to be analysed in Data/Processed
@@
     "shape_space"    = shapespace,
+    "shape_dim"      = shape_axes$dimensions,
     "form_space"     = formspace,
+    "form_dim"       = form_axes$dimensions,
     "groups_list"    = all_groups)
+
 save(form_shape_space_1tree_2levels, file = "../Data/Processed/form_shape_space_1tree_2levels.rda")

### 2021-08-27 Analysis/02-MCMCglmm_mini_chains.Rmd
Ready for compute!

@@
 library(MCMCglmm)
+library(dispRity)
 
-## Loading the demo data
-data(demo_data)
-data(demo_trees)
-data(demo_tree)
-
-## Setting the consensus tree
-consensus_tree <- demo_tree
+## Loading the parametrisation data
+load(file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
 ```
@@
 ```{r, parametrise_the_model, eval = TRUE}
+## Select the data
+shapespace <- form_shape_space_1tree_2levels$shape_space
+groups_list <- form_shape_space_1tree_2levels$groups_list
+selected_dim <- form_shape_space_1tree_2levels$shape_dim
+consensus_tree <- form_shape_space_1tree_2levels$consensus_tree
+
 ## Set up the parametrising chains models
-param_MCMCglmm <- make.mini.chains(data         = demo_data,
-                                   dimensions   = c(1:4),
+param_MCMCglmm <- make.mini.chains(data         = shapespace,
+                                   dimensions   = selected_dim,
                                    tree         = consensus_tree,
                                    trait.family = "gaussian",
-                                   randoms      = c("global", "clade"),
+                                   randoms      = c("global", "level1", "level2"),
                                    residuals    = "global",
@@
                                    parameters   = list(
-                                       nitt   = 10000,
-                                       thin   = 100,
+                                       nitt   = 1000000,
+                                       thin   = 500,
                                        burnin = 0))
@@
 ## Running the three MCMCglmm models
-param_chains <- run.mini.chains(param_MCMCglmm, replicates = 3)
+params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(param_chains, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_31levels_chain1.rda")
 ```
@@
 
-```{r, get_model_parameters, eval = TRUE}
+```{r, get_model_parameters, eval = FALSE}
 ## Extract the parameters from the parameter chains
@@
 
-```{r, set_the_mini_chains, eval = TRUE}
+```{r, set_the_mini_chains, eval = FALSE}
 ## The thinning parameter

### 2021-08-31 Analysis/02-MCMCglmm_mini_chains.Rmd
100k chains

@@
                                    parameters   = list(
-                                       nitt   = 1000000,
+                                       nitt   = 100000,
                                        thin   = 500,
@@
 ## Running the three MCMCglmm models
-params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(param_chains, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_31levels_chain1.rda")
+100k_params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(param_chains, file = "../Data/Processed/MCMCglmm/100k_params_shapespace_8D_31levels_chain1.rda")
 ```

### 2021-09-01 Analysis/03-elaboration_exploration_analyses.Rmd
pipeline rework

@@
 ```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
+get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {
+
+    # The magic: https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331
+
+    ## Select the right dimensions
+    data$VCV <- data$VCV[dimensions, dimensions, drop = FALSE]
+
+    ## Get the data dimensionality
+    dims <- dim(data$VCV)[1]
+
+    ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
+    unit_hypersphere1 <- unit_hypersphere2 <- matrix(0, ncol = dims, nrow = dims)
+    ## The "front" (e.g. "top", "right") units
+    diag(unit_hypersphere1) <- 1
+    ## The "back" (e.g. "bottom", "left") units
+    diag(unit_hypersphere2) <- -1
+    unit_hypersphere <- rbind(unit_hypersphere1, unit_hypersphere2)
+    ## Scale the hypersphere (where level is the confidence interval)
+    unit_hypersphere <- unit_hypersphere * sqrt(qchisq(level, 2))
+
+    ## Do the eigen decomposition (symmetric - faster)
+    eigen_decomp <- eigen(data$VCV, symmetric = TRUE)
+
+    ## Re-scaling the unit hypersphere
+    scaled_edges <- unit_hypersphere * rep(sqrt(eigen_decomp$values), each = dims*2)
+    ## Rotating the edges coordinates
+    edges <- tcrossprod(scaled_edges, eigen_decomp$vectors)
+
+    ## Move the matrix around
+    edges <- edges + rep(data$Sol[dimensions, drop = FALSE], each = dims*2)
+
+    ## Get the edges coordinates
+    return(edges[c(axis, axis+dims), , drop = FALSE])
+}
+
 set.seed(42)
@@
 ## Plotting the major axis
-lines(get.axes(list(list(list(VCV = cor_matrix, Sol = c(0,0)))))[[1]][[1]], lwd = 3, col = "orange")
+lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0))), lwd = 3, col = "orange")
 ```
@@
             ellipses = mean,
-            legend = TRUE)
+            legend = TRUE,
+            cex = 0.5)
 ```
@@
 
-Figure: in a space with five elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g.
-the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration and elaboration scores for either the elements (e.g.
-element A's projection in blue and its rejection in orange) or from any other axes (e.g.
-the major phylogenetic effect axes from the MCMCglmm for a specific clade).
+Figure: consider a space with five elements named "a" to "e" (in grey) and one focal axes in this space defined as the vector from point1: "d" to point2: "e" (the thin solid grey line linking the grey "d" and "e").
+We can then define all points from "d" (point1) as vectors (the dashed grey lines).
+We can then rotate the whole space so that the vector "d" to "e" becomes the reference coordinate vector (0,1) (the thick grey line) and rotate and translate all the other vectors accordingly (the solid grey lines linking the coloured letters "a" to "c").
+We can then project each of these vectors on the reference vector and measure their distance *on* the reference vector (the projection; the coloured solid lines) and their distance *from* the reference vector (the rejection; the coloured dotted lines).
+We can then also measure the angles between each vector and their respective projections (the angles between a grey solid line to a coloured letter and the corresponding coloured vector).
+Note that if this angle is equal to 90 degrees, the projection is equal to 0 (point "c") or if that angle is > 90 degrees, the projection is negative (point "b").
 
@@
 
-```{r}
-## Get all the angles and stuff
-group_results <- analyses.group(major_axes, base = "animal")
-names(group_results) <- levels(demo_data$clade)
+```{r, fig.height = 6, fig.width = 18}
+## Projecting each group's major axes on the phylogenetic major axes
+group_elaboration <- dispRity.covar(covar_matrices,
+               metric = projections.covar,
+               between.groups = list(c(1,4), c(2,4), c(3,4)),
+               measure = "position")
+group_exploration <- dispRity.covar(covar_matrices,
+               metric = projections.covar,
+               between.groups = list(c(1,4), c(2,4), c(3,4)),
+               measure = "distance")
+group_angle <- dispRity.covar(covar_matrices,
+               metric = projections.covar,
+               between.groups = list(c(1,4), c(2,4), c(3,4)),
+               measure = "degree")
+par(mfrow = c(1,3))
+plot(group_elaboration, ylab = "elaboration", col = colour_vector[-4])
+plot(group_exploration, ylab = "exploration", col = colour_vector[-4])
+plot(group_angle, ylab = "angle", col = colour_vector[-4])
 ```
@@
 The results are reported as boxplots for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.
-
-```{r}
-## Plot the blob wise metrics
-plot.analyses.group(group_results, col = colour_vector)
-```
 
@@
 
-```{r}
+```{r, eval = FALSE}
+## Get the mean axes for each level
+major_axes <- axis.covar(covar_matrices, sample = mean)
+
+## Projecting each group's major axes on the phylogenetic major axes
+phy_group_elaboration <- dispRity(covar_matrices,
+               dimensions = covar_matrices$call$dimensions,
+               metric  = dispRity::projections,
+               measure = "position",
+               point1  = major_axes$phylogeny[[1]][1,],
+               point2  = major_axes$phylogeny[[1]][2,])
+phy_group_exploration <- dispRity(covar_matrices,
+               dimensions = covar_matrices$call$dimensions,
+               metric  = dispRity::projections,
+               measure = "distance",
+               point1  = major_axes$phylogeny[[1]][1,],
+               point2  = major_axes$phylogeny[[1]][2,])
+phy_group_angle <- dispRity(covar_matrices,
+               dimensions = covar_matrices$call$dimensions,
+               metric  = dispRity::projections,
+               measure = "degree",
+               point1  = major_axes$phylogeny[[1]][1,],
+               point2  = major_axes$phylogeny[[1]][2,])
+
+## TODO: add option in dispRity plot to only select some groups
+par(mfrow = c(1,3))
+plot(phy_group_elaboration, col = colour_vector)
+plot(phy_group_exploration, ylab = "exploration", col = colour_vector)
+plot(phy_group_angle, ylab = "angle", col = colour_vector)
+
 ## Global analyses
@@
 
-```{r}
+```{r, eval = FALSE}
 ## Plot results
@@
 
-```{r}
+```{r, eval = FALSE}
 ## Results within groups

### 2021-09-10 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
 ## Selecting only the tips in the trait space
-matching_tips <- consensus_tree$tip.label %in% rownames(shapespace)
-consensus_tree <- drop.tip(consensus_tree, tip = consensus_tree$tip.label[!matching_tips])
+clean.tree <- function(tree, space) {
+    matching_tips <- tree$tip.label %in% rownames(space)
+    return(drop.tip(tree, tip = tree$tip.label[!matching_tips]))
+}
+consensus_tree <- clean.tree(consensus_tree, shapespace)
+
+## Reading the distribution of trees
+trees_list <- read.tree("../Data/raw/Tree_data/AllBirdsHackett1.tre")
+trees_list <- lapply(trees_list, clean.tree, shapespace)
+class(trees_list) <- "multiPhylo"
+save(trees_list, file = "../Data/Processed/trees_list.rda")
 ```

### 2021-09-10 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd

@@
 ## Loading beer
-library(beer)
+library(mcmcmcglmmm)
 library(MCMCglmm)
@@
 load(file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
+
+## Select the data
+shapespace <- form_shape_space_1tree_2levels$shape_space
+groups_list <- form_shape_space_1tree_2levels$groups_list
+selected_dim <- form_shape_space_1tree_2levels$shape_dim
+consensus_tree <- form_shape_space_1tree_2levels$consensus_tree
+
 ```
@@
 
-```{r, parametrise_the_model, eval = TRUE}
-## Select the data
-shapespace <- form_shape_space_1tree_2levels$shape_space
-groups_list <- form_shape_space_1tree_2levels$groups_list
-selected_dim <- form_shape_space_1tree_2levels$shape_dim
-consensus_tree <- form_shape_space_1tree_2levels$consensus_tree
-
+```{r, parametrise_the_model, eval = FALSE}
 ## Set up the parametrising chains models
@@
                                    parameters   = list(
-                                       nitt   = 100000,
+                                       nitt   = 20000,
                                        thin   = 500,
@@
 ```{r, get_model_parameters, eval = FALSE}
+## Load the computed models
+param_chains <- list()
+for(i in 1:3) {
+    load(paste0("../Data/Cluster/allbirds_ 8748t_8Dshapespace_31levels/param_chains/params_shapespace_8D_31levels_20k_chain", i, ".rda"))
+    param_chains[[i]] <- params_shapespace_8D_31levels[[1]]
+}
 ## Extract the parameters from the parameter chains
@@
 
-```{r, set_the_mini_chains, eval = FALSE}
+```{r, set_the_mini_chains, eval = TRUE}
+## Get the distribution of trees
+load(file = "../Data/Processed/trees_list.rda")
+
 ## The thinning parameter
-thin <- 100
+thin <- 200
 ## The number of samples per chains
-samples <- 100
+samples <- 10
 
 ## Set up the mini-chains models
-mini_chains <- make.mini.chains(data         = demo_data,
-                                dimensions   = c(1:4),
-                                tree         = demo_trees,
+shapespace_8748t_8D_32level <- make.mini.chains(
+                                data         = shapespace,
+                                dimensions   = selected_dim,
+                                tree         = consensus_tree,
                                 trait.family = "gaussian",
-                                randoms      = c("global", "clade"),
+                                randoms      = c("global", "level1", "level2"),
                                 residuals    = "global",
@@
                                     burnin = est_params$burnin))
+## Save the mini.chains
+save(shapespace_8748t_8D_32level, file = "shapespace_8748t_8D_32level.mini.chains")
 ```

### 2021-09-10 Data/Cluster/make.minichains.jobs.sh
Updated template

@@ -1,0 +1,66 @@
+##########################
+# Script for initialising the minichains
+##########################
+#SYNTAX: sh make.minichains.jobs.sh -c <chain_name> -r <replicates> -p <path (optional)>
+#with:
+#-c <chain_name> the name of the chain
+#-r <number_of_replicates> the number of replicates
+#-p <path> optional, the path to the chain
+##########################
+#guillert(at)tcd.ie - 2021/09/10
+##########################
+
+#INPUT
+## Input values
+while [[ $# -gt 1 ]]
+do
+key="$1"
+
+case $key in
+    -c|--chain)
+        CHAINNAME="$2"
+        shift
+        ;;
+    -r|--replicates)
+        REPLICATES="$2"
+        shift
+        ;;        
+    -p|--path)
+        LOCALPATH="$2"
+        ;;
+        *)
+        ;;
+esac
+shift
+done
+
+## Creating the batch job template
+cp job_template.job ${LOCALPATH}${CHAINNAME}_template.job
+echo "" >> ${LOCALPATH}${CHAINNAME}_template.job
+echo "## Run the script" >> ${LOCALPATH}${CHAINNAME}_template.job
+echo "Rscript ${LOCALPATH}${CHAINNAME}_<REPLICATE>.R" >> ${LOCALPATH}${CHAINNAME}_template.job
+## Bug on this line!!!!
+
+
+## Creating the Rscript template
+echo ".libPaths(\"~/homelib\")" > ${LOCALPATH}${CHAINNAME}_template.R
+echo "library(MCMCglmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "library(mcmcmcglmmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "load(\"${LOCALPATH}${CHAINNAME}.mini.chains\")" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "model <- run.mini.chains(${CHAINNAME}, replicates = 1, record.tree = TRUE)" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "save(model, file = \"${LOCALPATH}${CHAINNAME}_<REPLICATE>.rda\")" >> ${LOCALPATH}${CHAINNAME}_template.R
+
+## Looping through each templates to create the replicates
[…]

### 2021-09-10 mcmcmcglmmm/R/make.mini.chains.R

@@ -86,6 +86,9 @@ make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", 
     if(!all(dim_check <- apply(data[, dimensions], 2, is, "numeric"))) {
         stop(paste0("Invalid dimensions (not numeric?): ", paste(names(which(!dim_check)), collapse = ", "), "."))
     }
+    ## Reduce the dataset size
+    non_data <- unlist(lapply(as.list(1:ncol(data)), function(col, data) return(class(data[,col])), data = data))
+    data_reduce <- data[, c(dimensions, which(non_data != "numeric"))]
     
     ## Fixed effect
     ## Setting the fixed formula (initialising)
@@ -178,15 +181,15 @@ make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", 
 
     ## Setting the tree(s)
     output <- lapply(tree, function(tree, fixed, random, rvoc, family, data, priors, verbose, parameters, ...)
-       return(list(data = data,
-                   tree = tree,
+       return(list(#data = data,
+                   #tree = tree,
                    ## The MCMCglmm function
                    run  = function() MCMCglmm(fixed    = fixed,
                                               random   = random,
                                               rcov     = rcov,
                                               family   = family,
                                               pedigree = tree,
-                                              data     = data,
+                                              data     = data_reduce,
                                               prior    = priors,
                                               verbose  = verbose,
                                               burnin   = parameters$burnin,

### 2021-09-10 mcmcmcglmmm/tests/testthat/test-mini.chains.logic.R

@@ -54,7 +54,7 @@ test_that("make.mini.chains works", {
     test <- make.mini.chains(data = morphdat, tree = tree_list, dimensions = c(1,2), verbose = FALSE)
     expect_is(test, c("beer", "mini.chains"))
     expect_equal(length(test), 3)
-    expect_equal(names(test[[1]]), c("data", "tree", "run"))
+    expect_equal(names(test[[1]]), c("run"))
     ## Run!
     tust <- test[[1]]$run()
     expect_is(tust, "MCMCglmm")
@@ -69,7 +69,7 @@ test_that("make.mini.chains works", {
     test <- make.mini.chains(data = morphdat, tree = tree_list, dimensions = c(1,2), verbose = FALSE, residuals = "global")
     expect_is(test, c("beer", "mini.chains"))
     expect_equal(length(test), 3)
-    expect_equal(names(test[[1]]), c("data", "tree", "run"))
+    expect_equal(names(test[[1]]), c("run"))
     ## Run!
     tust <- test[[1]]$run()
     expect_is(tust, "MCMCglmm")
@@ -85,7 +85,7 @@ test_that("make.mini.chains works", {
     test <- make.mini.chains(data = morphdat, tree = tree_list, dimensions = c(1,2), verbose = FALSE, random = "global")
     expect_is(test, c("beer", "mini.chains"))
     expect_equal(length(test), 3)
-    expect_equal(names(test[[1]]), c("data", "tree", "run"))
+    expect_equal(names(test[[1]]), c("run"))
     ## Run!
     tust <- test[[1]]$run()
     expect_is(tust, "MCMCglmm")
@@ -100,7 +100,7 @@ test_that("make.mini.chains works", {
     test <- make.mini.chains(data = morphdat, tree = tree_list, dimensions = c(1,2), verbose = FALSE, random = "global", residuals = "global")
     expect_is(test, c("beer", "mini.chains"))
     expect_equal(length(test), 3)
-    expect_equal(names(test[[1]]), c("data", "tree", "run"))
+    expect_equal(names(test[[1]]), c("run"))
     ## Run!
     tust <- test[[1]]$run()
     expect_is(tust, "MCMCglmm")
@@ -116,7 +116,7 @@ test_that("make.mini.chains works", {
     test <- make.mini.chains(data = morphdat, tree = tree, dimensions = c(1:3), verbose = FALSE, residuals = "clade", priors = priors_list, randoms = "global")
[…]

### 2021-09-10 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd

@@
 for(i in 1:3) {
-    load(paste0("../Data/Cluster/allbirds_ 8748t_8Dshapespace_31levels/param_chains/params_shapespace_8D_31levels_20k_chain", i, ".rda"))
+    load(paste0("../Data/Cluster/allbirds_8748t_8Dshapespace_31levels/param_chains/params_shapespace_8D_31levels_20k_chain", i, ".rda"))
     param_chains[[i]] <- params_shapespace_8D_31levels[[1]]
@@
                                 dimensions   = selected_dim,
-                                tree         = consensus_tree,
+                                tree         = trees_list[sample(1:1000, 20)],
                                 trait.family = "gaussian",
@@
 ## Save the mini.chains
-save(shapespace_8748t_8D_32level, file = "shapespace_8748t_8D_32level.mini.chains")
+save(shapespace_8748t_8D_32level, file = "../Data/Cluster/allbirds_8748t_8Dshapespace_31levels/shapespace_8748t_8D_32level.mini.chains")
 ```

### 2021-09-14 beer/R/covar.projections.wrapper.R
covar.projections.wrapper done

@@ -98,7 +98,7 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
         wrap.dispRity.within <- function(measure, data, axes, verbose) {
             ## Calculate the measure
             if(verbose) message(".")
-            results <- lapply(axes, function(axes, data) dispRity(data, dimensions = data$call$dimensions, metric = projections, point1 = axes[1, ], point2 = axes[2, ])$disparity, data = data)
+            results <- lapply(axes, function(axes, data) dispRity::dispRity(data, dimensions = data$call$dimensions, metric = dispRity::projections, point1 = axes[1, ], point2 = axes[2, ])$disparity, data = data)
             ## Combine all the results together
             results <- lapply(1:length(results[[1]]), function(u) do.call(cbind, lapply(results, `[[`, u)))
             return(lapply(results, function(res) do.call(cbind, res)))
@@ -138,12 +138,18 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
     }
 
     if(dispRity.out) {
+        ## Select the data subset for correct disparity display
+        if(type == "between") {
+            sub_data <- get.subsets(data, subsets = unique(unlist(list_of_pairs)))
+        } else {
+            sub_data <- get.subsets(data, subsets = non_base_id)
+        }
         ## Make into a dispRity object
         output <- list()
         for(one_measure in 1:length(measure)) {
-            output[[one_measure]] <- dispRitize(results[[one_measure]], data,
+            output[[one_measure]] <- dispRitize(results[[one_measure]], sub_data,
                                                name = measure[[one_measure]],
-                                               fun = ifelse(type == "between", projections.covar, projections),
+                                               fun = ifelse(type == "between", projections.covar, dispRity::projections),
                                                type = type)
         }
         names(output) <- measure

### 2021-09-14 beer/R/covar.projections.wrapper_fun.R

@@ -1,9 +1,13 @@
-dispRitize <- function(one_result, data, name, fun, type) {
+dispRitize <- function(one_result, data, name, fun, type, cent.tend = median) {
     ## Copying the data
     output <- data
     ## Adding the disparity results
     output$disparity <- lapply(one_result, function(X) list(elements = X, X))
-    ## Adding the call
+    # ## Summarising the results for the "elements" part and adding rownames
+    # for(i in 1:n.subsets(data)) {
+    #     output$disparity[[i]] <- matrix(apply(output$disparity[[i]]$elements, 1, cent.tend), ncol = 1, dimnames = list(c(rownames(data$matrix[[1]])[c(data$subsets[[i]]$elements)])))    
+    # }
+        ## Adding the call
     output$call$bootstrap <- list(ncol(one_result[[1]]), "covar", NULL)
     output$call$disparity$metrics$name[[1]] <- name
     output$call$disparity$metrics$fun[[1]]  <- fun

### 2021-09-14 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

@@
 
-All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
-Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
+<!-- TG: The analyses will be eventually run in dispRity so we will miss out the following notices: 
 Some analyses in `beer` do take some time to run so we advise users not to operate heavy machinery while using it.
-Don't use it while driving.
+Don't use it while driving. -->
 The following code snippets are used to illustrate the implementation of the functions.
-See <!-- §§§run_replicates_script --> for running the actual analyses.
 
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE, print = FALSE}
 library(beer)
@@
 
-To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.
+To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices represented in this space.
 
-```{r, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE}
 covar_matrices <- MCMCglmm.dispRity(
@@
 
-```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
+```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, echo = FALSE}
 ## Plotting just the space
@@
 
-We use these covariance matrices as the basis of the phylogenetic main axes of variation.
-These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different phylogenetic levels:
+From this space and the mini chains run before, we can extract the posterior variance covariance matrices to calculate the major axes of variation in the `r dim(length(covar_matrices$call$dimensions))` dimensions trait space at different levels:
 
@@
 
-```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
+```{r, echp = FALSE, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, echo = FALSE}
 get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {
@@
 Figure: the main axis of variation (in orange) of the 95% CI ellipse of the variance-covariance matrix (in blue) of 20 elements in 2 dimensions (in black).
+Note that when estimated using a `mcmcmcglmmm` approach there are as many 95% CI ellipses and corresponding main axes as they are posterior samples.
 
-In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors.
-
-## Major axes calculation
-<!-- % NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!
- -->
-For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
-
-\begin{equation}
-O_{n} = 
-    \begin{pmatrix}
-        1 & 0 & \cdots & 0 \\
-        0 & 1 & \cdots & 0 \\
-        \vdots  & \vdots  & \ddots & \vdots  \\
-        0 & 0 & \cdots & 1 \\
-        -1 & 0 & \cdots & 0 \\
-        0 & -1 & \cdots & 0 \\
-        \vdots  & \vdots  & \ddots & \vdots  \\
-        0 & 0 & \cdots & -1 \\
-    \end{pmatrix}
-\end{equation}
-
-(in other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$);
-
-And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:
-
-\begin{equation}
-    O'_{n} = O_{n} \sqrt{\chi^2(0.95)}
-\end{equation}
-
-For the variance-covariance matrix $VCV_{n}$ of $n$ dimensions
-
-\begin{equation}
-VCV_{n} = 
-    \begin{pmatrix}
-        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
-        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
-        \vdots  & \vdots  & \ddots & \vdots  \\
-        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
-    \end{pmatrix}
-\end{equation}
-
-and the eigenvectors **v** and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
-
-\begin{equation}
-    VCV_{n} \bold{v} = \lambda \bold{v}
-\end{equation}
-
-We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
-
-\begin{equation}
-    M_{n} = [(O'_{n}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
-\end{equation}
-
-Where $M_{1,n}$ is the major axis of the 0.95 CI hyper-ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second axis (minor axis in 2D), etc. etc.
-
-Finally, we centred the matrix $M_{1,m}$ on the centroid of each clade.
-
-> The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axes` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
+In practice, we calculated the coordinates of these axes end points  directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors (see details 1).
 
@@
 
-```{r, fig.height = 6, fig.width = 6}
+```{r, echo = FALSE, fig.height = 6, fig.width = 6}
 sauron.plot(covar_matrices,
@@
 
-```{r, fig.height = 12, fig.width = 6}
+```{r, echo = FALSE, fig.height = 12, fig.width = 6}
 par(mfrow = c(2,1))
@@
 
-#### Measuring projection and rejection
+We measured the exploration and elaboration using two main approaches:
+ * By clades (i.e exploration/elaboration **between** clades): where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes.
+ * By species (i.e exploration/elaboration **within** clades): where we calculated the projections of each species onto 1) the global phylogenetic major axes and 2) the phylogenetic major axes of their respective clade.
+
+## Elaboration and exploration by clade
+
+To run the analyses for each clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).
+
+For each posterior of the `mcmcmcglmmm` model, we projected each group's major axes onto the global phylogenetic major axis of the specific posterior sample.
+In practice, for the projection to work, we translated and centred each group's major axes onto the origin of the global phylogenetic major axis (so that both origins have the coordinates 0) and then measure the projection and rejection as specified in Details 2.
+
+```{r, fig.height = 6, fig.width = 18, echo = FALSE}
+## Run the analyses between groups
+results_between <- covar.projections.wrapper(covar_matrices,
+                                             type = "between",
+                                             base = "phylogeny",
+                                             dispRity.out = TRUE)
+## Plot the results
+wrap.plot <- function(results, col, main) {
+    ## How many sub plots?
+    n <- length(results)
+    ## Which will be the centre plot?
+    centre <- ceiling(n/2)
+    ## Plot everything
+    par(mfrow = c(1, n))
+    for(i in 1:n) {
+        plot(results[[i]], col = col, main = ifelse(i == centre, main, ""))
+    }
+}
+wrap.plot(results_between,
+          col = colour_vector[-4],
+          main = "Groups difference")
+```
+Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes.
+For each panel, the highest the score the more the clade is elaborating, exploring, exploring perpendicular to the main axes for the position, distance and degree respectively.
+
+## Elaboration and exploration by species
+
+Similarly to the analyses described above, we ran the analyses for each species.
+Here, instead of projecting the major axes of the groups onto the global phylogenetic major axis we projected each species onto the phylogenetic major axis (for each posterior distribution).
+
+```{r, fig.height = 6, fig.width = 18, echo = FALSE, eval = TRUE}
+## Run the analyses of groups on the phylogeny
+results_phylo <- covar.projections.wrapper(covar_matrices,
+                                            type = "within",
+                                            base = "phylogeny",
+                                            dispRity.out = TRUE)
+## Plot the results
+wrap.plot(results_phylo,
+          col = colour_vector[-4],
+          main = "Projection on the\nglobal phylogeny axes")
+```
+Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.
+
+Finally, we ran the same analyses but instead of the projection of the species onto the global phylogenetic major axes, we projected them onto their respective group's phylogenetic major axes.
+
+```{r, fig.height = 6, fig.width = 18, echo = FALSE, eval = TRUE}
+## Run the analyses groups on themselves
+results_within <- covar.projections.wrapper(covar_matrices,
+                                            type = "within",
+                                            dispRity.out = TRUE)
+## Plot the results
+wrap.plot(results_within,
+          col = colour_vector,
+          main = "Projection on the\ngroup's phylogeny axes")
+```
+Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).
+
+>TG: TODO: do this plot on a phylogeny
+
+# Supplementary details
+
+## Details 1: Major axes calculation
+
+For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
+
+\begin{equation}
+O_{n} = 
+    \begin{pmatrix}
+        1 & 0 & \cdots & 0 \\
+        0 & 1 & \cdots & 0 \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        0 & 0 & \cdots & 1 \\
+        -1 & 0 & \cdots & 0 \\
+        0 & -1 & \cdots & 0 \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        0 & 0 & \cdots & -1 \\
+    \end{pmatrix}
+\end{equation}
+
+In other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$;
+
+And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:
+
+\begin{equation}
+    O'_{n} = O_{n} \sqrt{\chi^2(0.95)}
+\end{equation}
+
+For the variance-covariance matrix $VCV_{n}$ of $n$ dimensions
+
+\begin{equation}
+VCV_{n} = 
+    \begin{pmatrix}
+        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
+        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
+    \end{pmatrix}
+\end{equation}
+
+and the eigenvectors **v** and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
+
+\begin{equation}
+    VCV_{n} \bold{v} = \lambda \bold{v}
+\end{equation}
+
+We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scaled 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
+
+\begin{equation}
+    M_{n} = [(O'_{n}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
+\end{equation}
+
+Where $M_{1,n}$ is the major axis of the 0.95 CI hyper-ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second axis (minor axis in 2D), etc. etc.
+
+Finally, we centred the matrix $M_{1,m}$ on the centroid of each clade.
+
+> The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axes` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).
+
+## Details 2: Measuring projection and rejection
 
@@
 
-#### Generalisation of projection onto any vector in a set space
+### Generalisation of projection onto any vector in a set space
 
@@
 
-Following the ways to: <!-- %NC: Not sure what you mean here -->
- 1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
- 2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
- 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes.
-
-We can measure these scores for each clade and each species separately.
-
-## Elaboration and exploration by clade
-
-To run the analyses for each clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).
-
-To do so, for a randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
-For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).
-
-```{r, fig.height = 6, fig.width = 18}
-## Projecting each group's major axes on the phylogenetic major axes
-group_elaboration <- dispRity.covar(covar_matrices,
-               metric = projections.covar,
-               between.groups = list(c(1,4), c(2,4), c(3,4)),
-               measure = "position")
-group_exploration <- dispRity.covar(covar_matrices,
-               metric = projections.covar,
-               between.groups = list(c(1,4), c(2,4), c(3,4)),
-               measure = "distance")
-group_angle <- dispRity.covar(covar_matrices,
-               metric = projections.covar,
-               between.groups = list(c(1,4), c(2,4), c(3,4)),
-               measure = "degree")
-par(mfrow = c(1,3))
-plot(group_elaboration, ylab = "elaboration", col = colour_vector[-4])
-plot(group_exploration, ylab = "exploration", col = colour_vector[-4])
-plot(group_angle, ylab = "angle", col = colour_vector[-4])
-```
-
-The results are reported as boxplots for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.
-
-## Elaboration and exploration by species
-
-Similarly to the analyses described above, we ran the analyses for each species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results for each groups.
-
-```{r, eval = FALSE}
-## Get the mean axes for each level
-major_axes <- axis.covar(covar_matrices, sample = mean)
-
-## Projecting each group's major axes on the phylogenetic major axes
-phy_group_elaboration <- dispRity(covar_matrices,
-               dimensions = covar_matrices$call$dimensions,
-               metric  = dispRity::projections,
-               measure = "position",
-               point1  = major_axes$phylogeny[[1]][1,],
-               point2  = major_axes$phylogeny[[1]][2,])
-phy_group_exploration <- dispRity(covar_matrices,
-               dimensions = covar_matrices$call$dimensions,
-               metric  = dispRity::projections,
-               measure = "distance",
-               point1  = major_axes$phylogeny[[1]][1,],
-               point2  = major_axes$phylogeny[[1]][2,])
-phy_group_angle <- dispRity(covar_matrices,
-               dimensions = covar_matrices$call$dimensions,
-               metric  = dispRity::projections,
-               measure = "degree",
-               point1  = major_axes$phylogeny[[1]][1,],
-               point2  = major_axes$phylogeny[[1]][2,])
-
-## TODO: add option in dispRity plot to only select some groups
-par(mfrow = c(1,3))
-plot(phy_group_elaboration, col = colour_vector)
-plot(phy_group_exploration, ylab = "exploration", col = colour_vector)
-plot(phy_group_angle, ylab = "angle", col = colour_vector)
-
-## Global analyses
-tip_results_global <- analyses.tip(data = demo_data[, c(1:4)],
-                                   axes = major_axes$animal)
-```
-
-And here are the values plotted by group:
-
-```{r, eval = FALSE}
-## Plot results
-plot.analyses.tip(tip_results_global,
-                  col = colour_vector,
-                  group = split(rownames(demo_data), f = demo_data$clade),
-                  main = "Tips vs. whole phylogeny")
-```
-
-Note that here only the tip positions are projected on the global phylogenetic axes.
-
-Finally, we also looked at the exploration elaboration profiles within each clade, by replacing the projection of the elements onto $\vec{B}$ for each clade to the projection onto their respective clade major axes $\vec{a}$.
-
-```{r, eval = FALSE}
-## Results within groups
-tip_results_within <- analyses.tip(
-                    data = demo_data[, c(1:4)],
-                    axes = major_axes[-length(major_axes)],
-                    group = split(rownames(demo_data), f = demo_data$clade))
-```
-
-```{r, eval = FALSE}
-## Plot results
-plot.analyses.tip(tip_results_within,
-                  col = colour_vector,
-                  group = split(rownames(demo_data), f = demo_data$clade),
-                  main = "Tips vs. clade")
-```
-
-But this is probably best to look at in a species by species fashion:
-
-TODO
-
 ## References

### 2021-09-14 beer/R/covar.projections.wrapper.R
bug fix

@@ -50,7 +50,7 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
     ## 1 - get major axis
     if(verbose) message("calculating the major axis:...")
     major_axes <- axis.covar(data, sample = sample, n = n, axis = major.axis, level = level)
-    # major_axes <- axis.covar(data, n = n)
+    # major_axes <- axis.covar(data)
     if(verbose) message("Done.")
 
     ## 2 - get the data
@@ -81,7 +81,7 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
             list_of_pairs <- unlist(apply(combn(1:n.subsets(data), 2), 2, list), recursive = FALSE)
         } else {
             base_id <- which(names(size.subsets(data)) == base)
-            list_of_pairs <- lapply(as.list(1:(n.subsets(data)-1)), function(x,y) c(x, y), y = base_id)
+            list_of_pairs <- lapply(as.list((1:n.subsets(data))[-base_id]), function(x,y) c(x, y), y = base_id)
         }
 
         ## Get all results

### 2021-09-17 Analysis/03-elaboration_exploration_analyses.Rmd
typo
Update 03-elaboration_exploration_analyses.Rmd

@@
 
->TG: TODO: do this plot on a phylogeny
+```{r, echo = FALSE, eval = TRUE}
+library(phytools)
+source("~/Packaging/phytools/R/plotTree.wBars.R")
+
+get.disparity.matrix <- function(data, col) {
+    ## Get the matrices
+    matrices <- lapply(data$disparity, `[[`, 1)
+    ## Add the rownames
+    list_of_names <- list()
+    for(one_group in 1:length(matrices)) {
+        rownames(matrices[[one_group]]) <- list_of_names[[one_group]] <- rownames(data$matrix[[1]])[c(data$subsets[[one_group]]$elements)]
+    }
+    ## Name the groups
+    names(list_of_names) <- names(data$subsets)
+
+    ## Matrix
+    matrix <- do.call(rbind, matrices)
+
+    ## Colours
+    if(!missing(col)) {
+        name.cols <- function(names, col) {
+            out <- rep(col, length(names))
+            names(out) <- names
+            return(out)
+        }
+        col <- mapply(name.cols, list_of_names, as.list(col))
+        names(col) <- NULL
+        col <- unlist(col)
+    }
+
+    ## Output
+    if(missing(col)) {
+        return(list(data = matrix, names = list_of_names))
+    } else {
+        return(list(data = matrix, names = list_of_names, col = col))
+    }
+}
+
+## Get the different elements
+phylo_elaboration  <- get.disparity.matrix(results_phylo$position,
+                                          col = colour_vector[-4])
+within_elaboration <- get.disparity.matrix(
+    get.subsets(results_within$position, subsets = c(1,2,3)),
+                                          col = colour_vector[-4])
+phylo_exploration  <- get.disparity.matrix(results_phylo$distance,
+                                          col = colour_vector[-4])
+within_exploration <- get.disparity.matrix(
+    get.subsets(results_within$distance, subsets = c(1,2,3)),
+                                          col = colour_vector[-4])
+phylo_angle        <- get.disparity.matrix(results_phylo$degree,
+                                          col = colour_vector[-4])
+within_angle       <- get.disparity.matrix(
+    get.subsets(results_within$degree, subsets = c(1,2,3)),
+                                          col = colour_vector[-4])
+```
+
+## Projections (on the main phylogeny)
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, phylo_elaboration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Elaboration", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
+```
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, phylo_exploration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Exploration", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
+```
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, phylo_angle$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Angle", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
+```
+
+## Projections (onto each clade's phylogeny)
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, within_elaboration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Elaboration (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
+```
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, within_exploration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Exploration (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
+```
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, within_angle$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Angle (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
+```
+
+These graphs can also be plotted as circular for the big trees later.
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+## Plot per species
+plotTree.wBars_modif(demo_tree, within_angle$data, type = "fan", col = phylo_elaboration$col, tip.label = FALSE, main = "Angle (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.grid.scale = TRUE))
+```
+
+## Color PCA according to mean proj/rej?
+
+And there's also this way of looking at it. But bof. Kinda ugly and probably too much info. (also it seems something is not working correctly).
+
+```{r,echo = FALSE, eval = TRUE}
+## Plotting fun
+plot.gradient.space <- function(data, values, cent.tend = median, col, gradient.fun = heat.colors, ...) {
+
+    get.colour.gradient <- function(values, col.fun, col.range, get.range = FALSE) {
+
+        ## Sort the data by range
+        histo <- hist(c(col.range, values), plot = FALSE)
+        n_col <- length(histo$counts)
+
+        ## Get the gradient
+        avail_cols <- rev(col.fun(n_col))
+
+        ## Attribute the colours
+        col_out <- rep(avail_cols[n_col], length(values))
+        range <- avail_cols[c(n_col, 1)]
+
+        ## Attribute the colour by range
+        for(colour in rev(1:n_col)) {
+          col_out[(values <= histo$breaks[colour])] <- avail_cols[colour]
+        }
+
+        if(get.range) {
+            return(list(range = range, col = col_out))
+        } else {
+            return(col_out)
+        }
+    }
+
+    ## Get the colour gradient
+    gradient_values <- apply(values, 1, cent.tend)
+    gradient_cols <- get.colour.gradient(gradient_values, col.fun = gradient.fun, col.range = range(gradient_values), get.range = TRUE)
+    colour_gradient <- gradient_cols$col
+    names(colour_gradient) <- rownames(values)
+
+    ## Plot the points
+    sauron.plot(data,
+                col = col,
+                major.axes = mean,
+                ellipses = mean,
+                legend = TRUE,
+                points = TRUE,
+                pch = 19,
+                cex = 1.5
+                , ...)
+    points(data$matrix[[1]], pch = 19, col = colour_gradient)
+    legend("bottomright", pch = 19, col = gradient_cols$range, legend = round(range(gradient_values), 2))
+}
+```
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+plot.gradient.space(data = covar_matrices, values = phylo_elaboration$data, col = colour_vector, main = "Elaboration gradient")
+```
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+plot.gradient.space(data = covar_matrices, values = phylo_exploration$data, col = colour_vector, main = "Exploration gradient")
+```
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+plot.gradient.space(data = covar_matrices, values = phylo_angle$data, col = colour_vector, main = "Angle gradient")
+```
+
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+plot.gradient.space(data = covar_matrices, values = within_elaboration$data, col = colour_vector, main = "Elaboration gradient (within)")
+```
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+plot.gradient.space(data = covar_matrices, values = within_exploration$data, col = colour_vector, main = "Exploration gradient (within)")
+```
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
+plot.gradient.space(data = covar_matrices, values = within_angle$data, col = colour_vector, main = "Angle gradient (within)")
+```
 
@@
 
-With R_{\vec{B}} being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:
+With $R_{\vec{B}}$ being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:
 

### 2021-09-20 mcmcmcglmmm/R/combine.mini.chains.R
Update combine.mini.chains.R

@@ -18,7 +18,7 @@ combine.mini.chains <- function(mini.chains) {
         mini.chains <- output
     }
 
-    ## Get the start end and thin values
+    ## Get the start end and thin values
     mcpar <- do.call(rbind, lapply(mini.chains, function(x) attr(x$Sol, "mcpar")))
     iterations_total <- sum(apply(mcpar[, c(1,2)], 1, diff))
     mcpar <- mcpar[1, ]

### 2021-09-20 Analysis/02.1-MCMCglmm_mini_chains_cluster.Rmd
Cluster script

--- 2021-09-10 Analysis/02-MCMCglmm_mini_chains.Rmd
@@
 
-All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
-Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
-Some analyses in `beer` do take some time to run so we advise you do not operate heavy machinery while using it.
-Don't use it while driving.
-The following code snippets are used to illustrate the implementation of the functions.
-See <!-- §§§run_replicates_script --> for running the actual analyses.
+This script is a working script to run the chains on the SHARC Sheffield cluster.
+Refer to `02-MCMCglmm_mini_chains.Rmd` for more details and reproducibility
 
@@
 
-# MCMCglmm mini chains analyses
+# All birds shapespace analyses
 
-To estimate the variance-covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].
-
-## Model
-
-We ran a general multi-response model with `§§§N_traits` of the trait space as fixed effects (using a Gaussian distribution) and and the residual covariance structure.
-We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).
-<!-- % NC: I know this is in the data prep, but can you specify an example MCMCglmm model here to make it clearer what you mean? -->
-
-## Mini chains
-
-To increase the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
-It runs multiple MCMCglmm analyses on multiple trees and pulls the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty (similar to @mulTree - ADD LINK?).
-The size of the chains are optimised for speed. <!-- % NC: Editted as multree is not familiar to everyone, so keeping it general and linking to multree is better -->
-
-### Mini-chains parametrization
-
-We first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
-We ran these models (for `§§§N_generations` sampling every `§§§N_nitt` with no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis). 
-<!-- % NC: I thought in MCMCglmm the nitt parameter was total number of iterations, and thin gives how often the chain is sampled?
-I think you've called nitt generations here but better to be consistent with MCMCglmm?
- -->
+## Mini-chains parametrization
 
@@
 
-We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree, and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
-Each mini-chain thus provided us with 100 posterior data points.
-The two main advantages of this mini-chain approach is that 1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and 2) they take into account tree uncertainty without having to run the complete MCMCglmm on all trees (c.f.
-@mulTree).
-
 ```{r, set_the_mini_chains, eval = TRUE}
@@
 
-We then run the MCMCglmm over $N_trees$ to get an estimate of the variance-covariance matrices including phylogenetic uncertainty.
+The chains are then run on the cluster
+TODO: details
 
-```{r, running_replicates, eval = FALSE}
-## The number of replicates
-n_reps <- 10
+### Combining the model data
 
-## Run the required chains
-results <- run.mini.chains(mini_chains, replicates = n_reps, path = "../Data/Processed/MCMC/")
+```{r}
+## Preparing the model files
+model_list <- list()
+class(model_list) <- c("beer", "mini.chains")
+models_path <- "../Data/Cluster/allbirds_8748t_8Dshapespace_31levels/cluster_out/"
+## Loading all models
+for(file in list.files(models_path)) {
+    load(paste0(models_path, file, collapse = "/"))
+    model_list[[length(model_list)+1]] <- model[[1]]
+}
 
-## Combining the results as one big MCMC
-combined_results <- combine.mini.chains(results)
+## Combining the model files
+allbirds_8748t_8Dshapespace_31levels <- combine.mini.chains(model_list)
 
 ## Saving the results
-save(combined_results, file = "../Data/processed/combined_results.rda")
+save(allbirds_8748t_8Dshapespace_31levels, file = "../Data/processed/allbirds_8748t_8Dshapespace_31levels.rda")
 ```

### 2021-09-21 Analysis/01-Data_preparation.Rmd
Save the groups

@@
 all_groups <- c(groups_level1$list, groups_level2$list)
+group_names_level_1_2 <- all_groups
+save(group_names_level_1_2, file = "../Data/Processed/group_names_level_1_2.rda")
 ```

### 2021-09-21 Data/Cluster/make.minichains.jobs.sh
moved to Functions

@@ -1,66 +1,0 @@
-##########################
-# Script for initialising the minichains
-##########################
-#SYNTAX: sh make.minichains.jobs.sh -c <chain_name> -r <replicates> -p <path (optional)>
-#with:
-#-c <chain_name> the name of the chain
-#-r <number_of_replicates> the number of replicates
-#-p <path> optional, the path to the chain
-##########################
-#guillert(at)tcd.ie - 2021/09/10
-##########################
-
-#INPUT
-## Input values
-while [[ $# -gt 1 ]]
-do
-key="$1"
-
-case $key in
-    -c|--chain)
-        CHAINNAME="$2"
-        shift
-        ;;
-    -r|--replicates)
-        REPLICATES="$2"
-        shift
-        ;;        
-    -p|--path)
-        LOCALPATH="$2"
-        ;;
-        *)
-        ;;
-esac
-shift
-done
-
-## Creating the batch job template
-cp job_template.job ${LOCALPATH}${CHAINNAME}_template.job
-echo "" >> ${LOCALPATH}${CHAINNAME}_template.job
-echo "## Run the script" >> ${LOCALPATH}${CHAINNAME}_template.job
-echo "Rscript ${LOCALPATH}${CHAINNAME}_<REPLICATE>.R" >> ${LOCALPATH}${CHAINNAME}_template.job
-## Bug on this line!!!!
-
-
-## Creating the Rscript template
-echo ".libPaths(\"~/homelib\")" > ${LOCALPATH}${CHAINNAME}_template.R
-echo "library(MCMCglmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
-echo "library(mcmcmcglmmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
-echo "load(\"${LOCALPATH}${CHAINNAME}.mini.chains\")" >> ${LOCALPATH}${CHAINNAME}_template.R
-echo "model <- run.mini.chains(${CHAINNAME}, replicates = 1, record.tree = TRUE)" >> ${LOCALPATH}${CHAINNAME}_template.R
-echo "save(model, file = \"${LOCALPATH}${CHAINNAME}_<REPLICATE>.rda\")" >> ${LOCALPATH}${CHAINNAME}_template.R
-
-## Looping through each templates to create the replicates
[…]

### 2021-09-21 Analysis/01-Data_preparation.Rmd
updated data preparation pipeline

@@
 
-## Selecting only the tips in the trait space
-clean.tree <- function(tree, space) {
-    matching_tips <- tree$tip.label %in% rownames(space)
-    return(drop.tip(tree, tip = tree$tip.label[!matching_tips]))
-}
-consensus_tree <- clean.tree(consensus_tree, shapespace)
-
 ## Reading the distribution of trees
 trees_list <- read.tree("../Data/raw/Tree_data/AllBirdsHackett1.tre")
-trees_list <- lapply(trees_list, clean.tree, shapespace)
-class(trees_list) <- "multiPhylo"
-save(trees_list, file = "../Data/Processed/trees_list.rda")
 ```
@@
 
+Here we are going to group the spaces, trees and levels for subsequent levels (nested).
+There are 5 levels: Super order (1), Order (2), Sub-order (3), Super families (4) and families (5).
+The script above creates the model for the two first levels, the subsequent models create the mini-chains for the following overlapping pairs of models (order + sub-order; sub-order + super families; super families + families).
+
 ```{r}
-## Add the clade columns and the animal column to the spaces
-shapespace <- as.data.frame(shapespace)
-shapespace$level1 <- groups_level1$col
-shapespace$level2 <- groups_level2$col
-shapespace$animal <- rownames(shapespace)
-formspace <- as.data.frame(formspace)
-formspace$level1 <- groups_level1$col
-formspace$level2 <- groups_level2$col
-formspace$animal <- rownames(formspace)
+source("../Functions/prep.data.R")
+## Getting the data ready for all birds
+shapespace_allbirds_lvl_superorder_order <- prep.data(
+                                          level = 0, lvl.inc = c(1,2),
+                                          clades    = clades_list,
+                                          space     = shapespace,
+                                          dim       = shape_axes$dimensions,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_superorder_order.rda")
 
-## Save the data to be analysed in Data/Processed
-form_shape_space_1tree_2levels <- list(
-    "consensus_tree" = consensus_tree,
-    "shape_space"    = shapespace,
-    "shape_dim"      = shape_axes$dimensions,
-    "form_space"     = formspace,
-    "form_dim"       = form_axes$dimensions,
-    "groups_list"    = all_groups)
+## Getting the data ready for each bird super order
+shapespace_superorder_lvl_order_suborder <- prep.data(
+                                          level = 1, lvl.inc = c(2,3),
+                                          clades    = clades_list,
+                                          space     = shapespace,
+                                          dim       = shape_axes$dimensions,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_superorder_lvl_order_suborder, file = "../Data/Processed/shapespace_superorder_lvl_order_suborder.rda")
 
-save(form_shape_space_1tree_2levels, file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
+## Getting the data ready for each bird order
+shapespace_order_lvl_suborder_superfam <- prep.data(
+                                          level = 2, lvl.inc = c(3,4),
+                                          clades    = clades_list,
+                                          space     = shapespace,
+                                          dim       = shape_axes$dimensions,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_order_lvl_suborder_superfam, file = "../Data/Processed/shapespace_order_lvl_suborder_superfam.rda")
+
+## Getting the data ready for each bird sub order
+shapespace_suborder_lvl_superfam_fam <- prep.data(
+                                          level = 3, lvl.inc = c(4,5),
+                                          clades    = clades_list,
+                                          space     = shapespace,
+                                          dim       = shape_axes$dimensions,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_suborder_lvl_superfam_fam, file = "../Data/Processed/shapespace_suborder_lvl_superfam_fam.rda")
+
+## Getting the data ready for each bird super family
+shapespace_superfam_lvl_fam <- prep.data(
+                                          level = 4, lvl.inc = 5,
+                                          clades    = clades_list,
+                                          space     = shapespace,
+                                          dim       = shape_axes$dimensions,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_superfam_lvl_fam, file = "../Data/Processed/shapespace_superfam_lvl_fam.rda")
 ```

### 2021-09-21 Functions/prep.data.R

@@ -7,7 +7,7 @@
 #@param consensus the consensus tree
 #@param trees the list of trees
 #@param min the minimum of species to include by level (default is 15)
-prep.data <- function(level, lvl.inc, clades, space, dim, consensus, trees, min = 15) {
+prep.data <- function(level, lvl.inc, clades, space, dim, consensus, trees, min = 15, verbose) {
 
   ## Selecting the right level(s)
   if(level == 0) {

### 2021-09-22 Analysis/01-Data_preparation.Rmd
Data  preparation update

@@
                                           verbose   = TRUE)
-save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_superorder_order.rda")
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
 
@@
 save(shapespace_superfam_lvl_fam, file = "../Data/Processed/shapespace_superfam_lvl_fam.rda")
+
+## Get a list of all selected levels for the record (and number of elements per levels)
+super_orders     <- shapespace_allbirds_lvl_superorder_order[[1]]$levels[[1]]
+orders           <- shapespace_allbirds_lvl_superorder_order[[1]]$levels[[2]]
+sub_orders       <- lapply(shapespace_order_lvl_suborder_superfam, function(x) x$levels[[1]])
+names(sub_orders) <- NULL
+sub_orders       <- unlist(sub_orders, recursive = FALSE)
+super_families   <- lapply(shapespace_order_lvl_suborder_superfam, function(x) x$levels[[2]])
+names(super_families) <- NULL
+super_families   <- unlist(super_families, recursive = FALSE)
+families         <- lapply(shapespace_superfam_lvl_fam, function(x) x$levels[[1]])
+names(families) <- NULL
+families         <- unlist(families, recursive = FALSE)
+
+## Get the list of levels
+levels_list <- list("super_orders"   = unlist(lapply(super_orders, length)),
+                    "orders"         = unlist(lapply(orders, length)),
+                    "sub_orders"     = unlist(lapply(sub_orders, length)),
+                    "super_families" = unlist(lapply(super_families, length)),
+                    "families"       = unlist(lapply(families, length)))
+save(levels_list, file = "../Data/Processed/levels_list.rda")
 ```

### 2021-09-22 Analysis/02-MCMCglmm_mini_chains.Rmd
Mini chains explanations

--- 2021-08-31 Analysis/02-MCMCglmm_mini_chains.Rmd
@@
 
-## Loading the parametrisation data
-load(file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
+## Loading the demo data and trees
+data(demo_data)
+data(demo_tree)
+traitspace <- demo_data
+consensus_tree <- demo_tree
+trees_list <- list(demo_tree, demo_tree, demo_tree)
+class(trees_list) <- "multiPhylo"
 ```
@@
 
-## Model
+## Models
 
-We ran a general multi-response model with `§§§N_traits` of the trait space as fixed effects (using a Gaussian distribution) and and the residual covariance structure.
-We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).
-<!-- % NC: I know this is in the data prep, but can you specify an example MCMCglmm model here to make it clearer what you mean? -->
+```{r, echo = FALSE}
+## Reading the clades list
+load("../Data/Processed/levels_list.rda")
+```
+
+We ran a general multi-response model on multiple dimensions with one global residual term, a one global random term (phylogeny) and variable clade specific random terms (one per taxonomic level comprising more than 15 species).
+We ran these models across five taxonomic levels: the super orders (`r length(levels_list$super_orders)`), the orders (`r length(levels_list$orders)`), the sub-orders (`r length(levels_list$sub_orders)`), the super families (`r length(levels_list$super_families)`) and the families (`r length(levels_list$families)`).
+Because of the high number of resulting total random terms (`r sum(unlist(lapply(levels_list, length))) + 1`!) we ran multiple models in an overlapping nested fashion with each taxonomic level including two sub-levels.
+For example, one model for all birds including the super order and the order levels as random terms, `r length(levels_list$super_orders)` models at the super order level including the order and the sub order levels as random terms, etc...
+In total we thus ran `r sum(unlist(lapply(levels_list[1:4], length)))+1` models (see table below).
+
+```{r, echo = FALSE}
+## Get the number of traits
+load("../Data/Processed/shapespace_superfam_lvl_fam.rda")
+dimensions <- shapespace_superfam_lvl_fam[[1]]$dimensions
+```
+
+Each model was run on `r length(dimensions)` dimensions from the same PCA shapespace (see Data section) <!-- TG:TODO --> using a Gaussian distribution for the fixed effects.
+Each model followed this generic syntax:
+
+data = traits + random terms + residuals terms
+
+Or in `MCMCglmm` pseudo code:
+
+```
+formula   = PC1:8 ~ trait:clade-1
+random    = ~ us(at.level(clade):trait):animal + us(trait):animal
+residuals = ~ us(trait):units
+```
+<!-- TG:TODO: explain that much better) -->
 
@@
 
-To increase the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
-It runs multiple MCMCglmm analyses on multiple trees and pulls the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty (similar to @mulTree - ADD LINK?).
-The size of the chains are optimised for speed. <!-- % NC: Editted as multree is not familiar to everyone, so keeping it general and linking to multree is better -->
+To increase the speed of the analysis and take phylogenetic uncertainty into account, we used a highly parallelisable "mini chains" approach.
+It runs multiple short `MCMCglmm` analyses on multiple trees and pulls the results together into one larger `MCMCglmm` that contains more variation due to phylogenetic uncertainty (similar to `mulTree` @mulTree).
+The size of the chains are optimised for speed.
+
+In brief the method works as follows:
+
+ 1- run three models without burnin on a consensus tree with flat priors: these are the parametrisation chains.
+
+ 2- extract the burnin length and the posteriors to be used as priors from the parametrisation chains.
+
+ 3- run multiple models with 10 samplings past the burnin using as priors the posteriors from the parametrisation chains: these are the mini chains.
+
+ 4- combine all the post burning 10 samples from the mini chains into one big chain to be used as the posterior from the model.
+
+Each step are described and demonstrated in more details below.
 
@@
 We first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
-We ran these models (for `§§§N_generations` sampling every `§§§N_nitt` with no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis). 
-<!-- % NC: I thought in MCMCglmm the nitt parameter was total number of iterations, and thin gives how often the chain is sampled?
-I think you've called nitt generations here but better to be consistent with MCMCglmm?
- -->
+We ran these chains for 20k iterations, sampling every 500 iterations (with no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis). 
+<!-- TG: TODO! -->
 
-```{r, parametrise_the_model, eval = TRUE}
-## Select the data
-shapespace <- form_shape_space_1tree_2levels$shape_space
-groups_list <- form_shape_space_1tree_2levels$groups_list
-selected_dim <- form_shape_space_1tree_2levels$shape_dim
-consensus_tree <- form_shape_space_1tree_2levels$consensus_tree
-
-## Set up the parametrising chains models
-param_MCMCglmm <- make.mini.chains(data         = shapespace,
-                                   dimensions   = selected_dim,
+```{r, parametrise_the_model, eval = FALSE}
+## Set up the parametrising chains models on the demo data
+param_MCMCglmm <- make.mini.chains(data         = traitspace,
+                                   dimensions   = 1:3,
                                    tree         = consensus_tree,
                                    trait.family = "gaussian",
-                                   randoms      = c("global", "level1", "level2"),
+                                   randoms      = c("global", "clade"),
                                    residuals    = "global",
@@
 ## Running the three MCMCglmm models
-100k_params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(param_chains, file = "../Data/Processed/MCMCglmm/100k_params_shapespace_8D_31levels_chain1.rda")
+parametrization <- run.mini.chains(param_MCMCglmm, replicates = 3)
 ```
 
-From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
+From these three chains, we first extracted the three burnin periods (defined as the number of iterations when the chain reaches the median likelihood value times 1.1).
 We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
@@
 ## Extract the parameters from the parameter chains
-est_params <- extract.parameters(param_chains)
+estimated_params <- extract.parameters(parametrization)
 ```
@@
 
-We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree, and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
-Each mini-chain thus provided us with 100 posterior data points.
-The two main advantages of this mini-chain approach is that 1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and 2) they take into account tree uncertainty without having to run the complete MCMCglmm on all trees (c.f.
-@mulTree).
+We set up a mini-chain to be a MCMCglmm running with:
+  1) the model described above,
+  2) a random tree, and
+  3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 10 sampled iterations (`nitt = burnin + thin * 10`).
+Each mini-chain thus provided us with 10 posterior exploitable data points.
+The two main advantages of this mini-chain approach is that
+  1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and
+  2) they take into account tree uncertainty without having to run the complete `MCMCglmm` on all trees (c.f. `mulTree` @mulTree).
 
 ```{r, set_the_mini_chains, eval = FALSE}
+
 ## The thinning parameter
-thin <- 100
+thin <- 500
 ## The number of samples per chains
-samples <- 100
+samples <- 10
 
 ## Set up the mini-chains models
-mini_chains <- make.mini.chains(data         = demo_data,
-                                dimensions   = c(1:4),
-                                tree         = demo_trees,
+parametrisation_chain <- make.mini.chains(
+                                data         = traitspace,
+                                dimensions   = 1:3,
+                                tree         = trees_list,
                                 trait.family = "gaussian",
@@
 
-We then run the MCMCglmm over $N_trees$ to get an estimate of the variance-covariance matrices including phylogenetic uncertainty.
+We then run the `MCMCglmm` over 100 randomly selected trees to get an estimate of the variance-covariance matrices including phylogenetic uncertainty.
 
@@
 ## The number of replicates
-n_reps <- 10
+n_reps <- 100
 
 ## Run the required chains
-results <- run.mini.chains(mini_chains, replicates = n_reps, path = "../Data/Processed/MCMC/")
+results <- run.mini.chains(mini_chains, replicates = n_reps)
 
@@
 combined_results <- combine.mini.chains(results)
-
-## Saving the results
-save(combined_results, file = "../Data/processed/combined_results.rda")
 ```

### 2021-09-22 beer/R/covar.projections.wrapper.R
updated wrapper (working)

@@ -48,7 +48,7 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
     ## Check method measure = c("position", "distance", "degree")
 
     ## 1 - get major axis
-    if(verbose) message("calculating the major axis:...")
+    if(verbose) message("calculating the major axis:...", appendLF = FALSE)
     major_axes <- axis.covar(data, sample = sample, n = n, axis = major.axis, level = level)
     # major_axes <- axis.covar(data)
     if(verbose) message("Done.")
@@ -93,21 +93,11 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
     }
 
     ## B - Type within
-    if(type == "within") {
+    if(type == "within") {        
         ## Wrapper for within function
-        wrap.dispRity.within <- function(measure, data, axes, verbose) {
-            ## Calculate the measure
-            if(verbose) message(".")
-            results <- lapply(axes, function(axes, data) dispRity::dispRity(data, dimensions = data$call$dimensions, metric = dispRity::projections, point1 = axes[1, ], point2 = axes[2, ])$disparity, data = data)
-            ## Combine all the results together
-            results <- lapply(1:length(results[[1]]), function(u) do.call(cbind, lapply(results, `[[`, u)))
-            return(lapply(results, function(res) do.call(cbind, res)))
-        }
-        
-        ## FAST VERSION
-        wrap.dispRity.within2 <- function(data, measure, axes, verbose) {
-            ## Calculate the measure
-            if(verbose) message(".")
+        wrap.dispRity.within <- function(data, axes, measure, verbose) {
+
+            if(verbose) message(".", appendLF = FALSE)
 
             ## Generating the function for fast disparity
             make.fun <- function(measure, axes) {
@@ -117,12 +107,9 @@ covar.projections.wrapper <- function(data, type, sample, n, base, average, majo
                 metric <- make.fun(measure, axis)
                 return(dispRity.fast(group, space, metric))
             }
-            lapply.group <- function(group, axes, measure, space) {
-                return(lapply(axes, lapply.axes, measure, group, space))
-            }
             clean.results <- function(results, measure) {
-                output <- lapply(1:length(results[[1]]), function(u) do.call(cbind, lapply(results, `[[`, u)))
-                names(output) <- one_measure
+                output <- lapply(1:length(results[[1]]), function(out) do.call(cbind, lapply(results, `[[`, out)))
+                names(output) <- measure
[…]

### 2021-09-22 beer/R/covar.projections.wrapper_fun.R

@@ -22,13 +22,8 @@ dispRitize <- function(one_result, data, name, fun, type, cent.tend = median) {
 
 ## Run projections faster!
 ## Angle between two vectors
-vector.angle <- function(v1, v2){#, degree = TRUE) {
-    angle <- acos(geometry::dot(v1, v2, d = 1) / (sqrt(sum(v1^2))*sqrt(sum(v2^2))))
-    # if(degree) {
-        return(angle *180/pi)
-    # } else {
-    #     angle
-    # }
+vector.angle <- function(v1, v2){
+    return(acos(geometry::dot(v1, v2, d = 1) / (sqrt(sum(v1^2))*sqrt(sum(v2^2)))) *180/pi)
 }
 ## Rotate a matrix along one axis (y)
 get.rotation.matrix <- function(x, y){
@@ -93,9 +88,9 @@ projections.fast <- function(matrix, point1 = 0, point2 = colMeans(matrix), meas
     ## Project the vectors
     projections <- t(apply(matrix, 1, geometry::dot, y = base_vector[2,], d = 2))
     ## Calculate the angles
-    if(measure == "degree" || measure == "radian") {
-        mes_angles <- t(t(apply(matrix, 1, vector.angle, base_vector[2,])))
-        mes_angles <- ifelse(is.nan(mes_angles), 0, mes_angles)
+    if("degree" %in% measure) {
+        angles_val <- t(t(apply(matrix, 1, vector.angle, base_vector[2,])))
+        angles_val <- ifelse(is.nan(angles_val), 0, angles_val)
     }
 
     # "position" #distance on
@@ -111,7 +106,7 @@ projections.fast <- function(matrix, point1 = 0, point2 = colMeans(matrix), meas
         values[["distance"]] <- apply(matrix - projections, 1, function(row) sqrt(sum(row^2)))
     }
     if("degree" %in% measure) {
-        values[["degree"]] <- c(mes_angles)
+        values[["degree"]] <- angles_val[,1]
     }
     return(values)
 }

### 2021-09-22 Functions/set.parametrisation.R
Create set.parametrisation.R

@@ -1,0 +1,59 @@
+#' @title Set up parametersisation scripts
+#'
+#' @description 
+#'
+#' @param 
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+
+set.parametrisation <- function(data, nitt = 20000, thin = 500, replicates = 3) {
+    ## Get the data name
+    data_name <- match.call()$data
+
+    ## Loop through each element
+    for(i in 1:length(data)) {
+        ## Set the random levels
+        param_chain <- make.mini.chains(
+            data         = data[[i]]$space,
+            dimensions   = data[[i]]$dimensions,
+            tree         = data[[i]]$consensus_tree,
+            trait.family = "gaussian",
+            randoms      = c("global", colnames(data[[i]]$space)[1:length(data[[1]]$levels) + length(data[[i]]$dimensions)]),
+            residuals    = "global",
+            priors       = 0.02,
+            verbose      = TRUE,
+            parameters   = list(nitt   = nitt,
+                                thin   = thin,
+                                burnin = 0))
+        ## Get the chain name
+        chain_name <- names(data)[i]
+        if(is.null(chain_name)) {
+            chain_name <- ""
+        } else {
+            chain_name <- paste0(chain_name, "_")
+        }
+        chain_name <- paste0(data_name, "_", chain_name, "param")
+        ## Save the parametrisation
+        save(param_chain, file = paste0(chain_name, ".minichains"))
+
+        ## Save the running script
+        system("echo \"library(mcmcmcglmmm)\" > script")
+        system("echo \"## Loading parametrisation\" >> script")
+        system(paste0("echo \"load('", chain_name,".minichains')\" >> script"))
+        system("echo \"## Running parametrisation\" >> script")
+        system(paste0("echo \"", chain_name, " <- run.mini.chains(param_chain, replicates = ", replicates, ")\" >> script"))
+        system("echo \"## Save the results\" >> script")
+        system(paste0("echo \"save(", chain_name, ", file = '",chain_name, ".rda')\" >> script"))
+        system(paste0("mv script ", paste0(data_name, "_", chain_name, "param.R")))
+    }
+    return(NULL)
+}
+
+#load("../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
+#set.parametrisation(shapespace_allbirds_lvl_superorder_order)

### 2021-09-23 Functions/set.parametrisation.R
Update set.parametrisation.R

@@ -19,8 +19,9 @@ set.parametrisation <- function(data, nitt = 50000, thin = 500, replicates = 3) 
     ## Loop through each element
     for(i in 1:length(data)) {
         ## Set the random levels (remove empty ones)
+        random_cols <- 1:length(data[[i]]$levels)
         if(length(empties <- which(unlist(lapply(data[[i]]$levels, length)) == 0)) > 0) {
-            data[[i]]$levels[[empties]] <- NULL
+            random_cols <- random_cols[-empties]
         }
 
         ## Make the param chains
@@ -29,7 +30,7 @@ set.parametrisation <- function(data, nitt = 50000, thin = 500, replicates = 3) 
             dimensions   = data[[i]]$dimensions,
             tree         = data[[i]]$consensus_tree,
             trait.family = "gaussian",
-            randoms      = c("global", colnames(data[[i]]$space)[1:length(data[[i]]$levels) + length(data[[i]]$dimensions)]),
+            randoms      = c("global", colnames(data[[i]]$space)[random_cols + length(data[[i]]$dimensions)]),
             residuals    = "global",
             priors       = 0.02,
             verbose      = TRUE,

### 2021-09-23 Analysis/02-MCMCglmm_mini_chains.Rmd
Added models table

@@
 
-We ran a general multi-response model on multiple dimensions with one global residual term, a one global random term (phylogeny) and variable clade specific random terms (one per taxonomic level comprising more than 15 species).
+We ran a general multi-response model on multiple dimensions with one global residual term, a one global random term (phylogeny) and variable clade specific random terms (one per taxonomic level).
 We ran these models across five taxonomic levels: the super orders (`r length(levels_list$super_orders)`), the orders (`r length(levels_list$orders)`), the sub-orders (`r length(levels_list$sub_orders)`), the super families (`r length(levels_list$super_families)`) and the families (`r length(levels_list$families)`).
@@
 For example, one model for all birds including the super order and the order levels as random terms, `r length(levels_list$super_orders)` models at the super order level including the order and the sub order levels as random terms, etc...
-In total we thus ran `r sum(unlist(lapply(levels_list[1:4], length)))+1` models (see table below).
+
+Furthermore for each nested model, we only ran models if the taxonomic level contained more than 16 species (two times the number of dimensions, see below) and contained at least two nested sub taxonomic levels with at least 16 species each.
+
+```{r, echo = FALSE}
+library(knitr)
+model_table <- read.csv("../Data/Processed/models_list.csv")
+```
+
+This resulted in using the following `r nrow(model_table)` models:
+
+```{r, echo = FALSE}
+kable(model_table)
+```
+
+> Note that we do acknowledge that the different taxonomic levels used in this analyses do not have a strong biological justification (i.e.
+they are just nodes named by biologists) but we found using them nonetheless useful since they refer to commonly known groups in birds (and that alternative choices would be equally arbitrary).
 

### 2021-09-24 Analysis/02-MCMCglmm_mini_chains.Rmd
added mini chains figure

@@
 
+![mini chain diagram](../Manuscript/Figures/mini-chains_diagram.png "Mini chains diagram")
+
 In brief the method works as follows:
@@
                                    parameters   = list(
-                                       nitt   = 20000,
+                                       nitt   = 50000,
                                        thin   = 500,

### 2021-09-24 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

@@
 
-and the eigenvectors **v** and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
+and the eigenvectors *v* and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
 
 \begin{equation}
-    VCV_{n} \bold{v} = \lambda \bold{v}
+    VCV_{n} v = \lambda v
 \end{equation}
 
-We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scaled 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
+We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors *v* and the product of the scaled 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
 
 \begin{equation}
-    M_{n} = [(O'_{n}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
+    M_{n} = [(O'_{n}\sqrt{\lambda}) \times v]^{\text{T}}
 \end{equation}

### 2021-09-28 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

@@
           main = "Projection on the\nglobal phylogeny axes")
+
+source("../Functions/circular.plot.R")
+
+circular.plot.phytools(data = results_phylo$position, tree = demo_tree, gradient = c("blue", "orange"), clade.names = TRUE, clades = colour_vector[-4], lwd = 3)
+
 ```

### 2021-10-11 Analysis/03.1-results.Rmd
isolated results computations

---
title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_exploration_analyses.Rmd.

# Elaboration and exploration analyses

```{r, eval = TRUE, echo = FALSE, print = FALSE}
library(beer)
```

# All birds shapespace analyses

```{r, eval = TRUE, echo = FALSE}
## Load the covar
load(file = "../Data/Processed/allbirds_8748t_8Dshapespace_31levels.rda")
## Load the data
load(file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
shapespace <- form_shape_space_1tree_2levels$shape_space
## Loading the group names
load("../Data/Processed/group_names_level_1_2.rda")

superorder_shapespace <- MCMCglmm.dispRity(
      data          = shapespace,
      posteriors    = allbirds_8748t_8Dshapespace_31levels,
      group         = MCMCglmm.levels(allbirds_8748t_8Dshapespace_31levels)[c(1:8, 32)],
      rename.groups = c(names(group_names_level_1_2), "phylogeny")[c(1:8, 32)])

order_shapespace <- MCMCglmm.dispRity(
      data          = shapespace,
      posteriors    = allbirds_8748t_8Dshapespace_31levels,
      group         = MCMCglmm.levels(allbirds_8748t_8Dshapespace_31levels)[-c(1:8, 33)],
      rename.groups = c(names(group_names_level_1_2), "phylogeny")[-c(1:8,33)])
```

```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, echo = FALSE}
## Plotting just the space
sauron.plot(superorder_shapespace,
            main = "Shapespace",
            cex = 0.5,
            legend = FALSE)
```
The global Shape space

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
par(mfrow = c(2,2))
## Plotting mean axes
sauron.plot(superorder_shapespace,
            col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
            main = "Average ellipses by super order",
            ellipses = mean,
            legend = FALSE,
            points = FALSE,
            cex = 0.5)
## Plotting mean ellipses
sauron.plot(superorder_shapespace,
            col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
            main = "Average major axes by super order",
            major.axes = mean,
            legend = TRUE,
            points = FALSE,
            cex = 0.5)
## Plotting mean axes
sauron.plot(order_shapespace,
            col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
            main = "Average ellipses by order",
            ellipses = mean,
            legend = FALSE,
            points = FALSE,
            cex = 0.5)
## Plotting mean ellipses
sauron.plot(order_shapespace,
            col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
            main = "Average major axes by order",
            major.axes = mean,
            legend = TRUE,
            points = FALSE,
            cex = 0.5)
```
Main axes and ellipses visualisation

## Calculate the elaborations/explorations

```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}
## Run the analyses between groups
super_results_between <- covar.projections.wrapper(superorder_shapespace,
                                             type = "between",
                                             base = "phylogeny",
                                             dispRity.out = TRUE)
## Run the analyses between groups
order_results_between <- covar.projections.wrapper(order_shapespace,
                                             type = "between",
                                             base = "phylogeny",
                                             dispRity.out = TRUE)
## Run the analyses of groups on the phylogeny
super_results_phylo <- covar.projections.wrapper(superorder_shapespace,
                                            type = "within",
                                            base = "phylogeny",
                                            incl.base = TRUE,
                                            dispRity.out = TRUE,
                                            verbose = TRUE)
## Run the analyses of groups on the phylogeny
order_results_phylo <- covar.projections.wrapper(order_shapespace,
                                            type = "within",
                                            base = "phylogeny",
                                            incl.base = TRUE,
                                            dispRity.out = TRUE,
                                            verbose = TRUE)
## Run the analyses groups on themselves
super_results_within <- covar.projections.wrapper(superorder_shapespace,
                                            type = "within",
                                            dispRity.out = TRUE,
                                            verbose = TRUE)
## Run the analyses groups on themselves
order_results_within <- covar.projections.wrapper(order_shapespace,
                                            type = "within",
                                            dispRity.out = TRUE,
                                            verbose = TRUE)
## Save all
allbirds_8748t_8Dshapespace_31levels_results_list <- list("super_results_between" = super_results_between,
                     "order_results_between" = order_results_between,
                     "super_results_phylo"   = super_results_phylo,
                     "order_results_phylo"   = order_results_phylo,
                     "super_results_within"  = super_results_within,
                     "order_results_within"  = order_results_within)
save(allbirds_8748t_8Dshapespace_31levels_results_list, file = "../Data/Processed/allbirds_8748t_8Dshapespace_31levels_results_list.rda")
```

```{r, echo = FALSE}
## Quick load
load(file = "../Data/Processed/allbirds_8748t_8Dshapespace_31levels_results_list.rda")
results <- allbirds_8748t_8Dshapespace_31levels_results_list
```

```{r, echo = FALSE}
## Plot the results
wrap.plot <- function(results, col, main) {
    ## How many sub plots?
    n <- length(results)
    ## Edit names for cleaning
    # for(i in 1:n) {
    #     names(results[[i]]$disparity) <- gsub(":phylogeny", "", names(results[[i]]$disparity))
    # }

    ## Which will be the centre plot?
    centre <- 1
    ## Plot everything
    op <- par(mfrow = c(n, 1), mar = c(7, 4, 4, 2) + 0.1)
    for(i in 1:n) {
        plot(results[[i]], col = col, main = ifelse(i == centre, main, ""), las = 2, xlab = "")
    }
    par(op)
}
wrap.plot(super_results_between,
          col = rainbow(n.subsets(superorder_shapespace)-1),
          main = "Superorders differences")
```
Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes. For each panel, the highest the score the more the clade is elaborating, exploring, exploring perpendicular to the main axes for the position, distance and degree respectively.

```{r, fig.height = 20, fig.width = 10, echo = FALSE}
wrap.plot(order_results_between,
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "Superorders differences")
```

## Elaboration and exploration by species

Similarly to the analyses described above, we ran the analyses for each species.
Here, instead of projecting the major axes of the groups onto the global phylogenetic major axis we projected each species onto the phylogenetic major axis (for each posterior distribution).

```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}

## Plot the results
wrap.plot(super_results_phylo,
          col = rainbow(n.subsets(superorder_shapespace)-1),
          main = "Projections per superorder\non the global phylogeny axes")
```
Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.

```{r, fig.height = 20, fig.width = 10, echo = FALSE}

## Plot the results
wrap.plot(order_results_phylo,
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "Projections per order\non the global phylogeny axes")
```

Finally, we ran the same analyses but instead of the projection of the species onto the global phylogenetic major axes, we projected them onto their respective group's phylogenetic major axes.

```{r, fig.height = 20, fig.width = 10, echo = FALSE}

## Plot the results
wrap.plot(super_results_within,
          col = rainbow(n.subsets(superorder_shapespace)-1),
          main = "Projections per superorder\nonto each superorder")
```
Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).

```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}

## Plot the results
wrap.plot(order_results_within,
          col = rainbow(n.subsets(order_shapespace)-1),
          main = "Projections per order\nonto each order")
```

### 2021-10-11 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE}
 ## Run the analyses between groups
@@
 ## Save all
-allbirds_8748t_8Dshapespace_31levels_results_list <- list("super_results_between" = super_results_between,
+allbirds_8748t_8Dshapespace_31levels_results_list <- list(
+        "super_results_between" = super_results_between,
         "order_results_between" = order_results_between,
@@
 }
-wrap.plot(super_results_between,
+wrap.plot(results$super_results_between,
           col = rainbow(n.subsets(superorder_shapespace)-1),
@@
 ```{r, fig.height = 20, fig.width = 10, echo = FALSE}
-wrap.plot(order_results_between,
+wrap.plot(results$order_results_between,
           col = rainbow(n.subsets(order_shapespace)-1),
@@
 ```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}
-
 ## Plot the results
-wrap.plot(super_results_phylo,
+wrap.plot(results$super_results_phylo,
           col = rainbow(n.subsets(superorder_shapespace)-1),
@@
 ## Plot the results
-wrap.plot(order_results_phylo,
+wrap.plot(results$order_results_phylo,
           col = rainbow(n.subsets(order_shapespace)-1),
@@
 ## Plot the results
-wrap.plot(super_results_within,
+wrap.plot(results$super_results_within,
           col = rainbow(n.subsets(superorder_shapespace)-1),
@@
 ## Plot the results
-wrap.plot(order_results_within,
+wrap.plot(results$order_results_within,
           col = rainbow(n.subsets(order_shapespace)-1),
@@
 ```
+
+## Different scores distributions per species
+
+```{r, eval = TRUE, echo = FALSE, eval = FALSE}
+library(hrdcde)
+
+#@param results the results list
+#@param type which type (position, distance or angle)
+#@param cent.tend which central tendency fun (default is median)
+#@param use.hdr whether to use kernel density estimate for the outputs (default is TRUE)
+#@param prob which quantiles to use (default is c(50, 95, 99))
+#@param ... plot options
+
+get.distributions <- function(results, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), ...) {
+
+    ## Extracting all the results
+    disparity <- lapply(results[[type]]$disparity, function(X, type) return(X$elements))
+
+    ## Recentre the results
+    if(type == "position") {
+        disparity <- lapply(disparity, function(X) return(X-0.5))
+    }
+
+    ## Measure the central tendencies
+    centrals <- lapply(disparity, function(X, fun) apply(X, 1, fun), fun = cent.tend)
+    ## Number of groups
+    n_groups <- length(centrals)
+
+    ## Calculate and plot the hdrcde
+    op_tmp <- par(mfrow = c(ceiling(sqrt(n_groups)), round(sqrt(n_groups))))
+    densities <- list()
+    for(i in 1:length(centrals)) {
+        if(use.hdr) {
+            densities[[i]] <- hdr.den(centrals[[i]], prob = prob, main = names(centrals)[i], ...)
+        } else {
+            hist(centrals[[i]], main = names(centrals)[i], ...)
+        }
+    }
+    par(op_tmp)
+
+    ## Find the extreme species
+    get.extremes.hdr <- function(density, central) {
+        ## Get the minimums and maximums
+        mins <- which(central <= density$hdr[1,1])
+        maxs <- which(central >= density$hdr[1,2])
+        ## Get their values
+        return(c(central[names(mins)], central[names(maxs)]))
+    }
+    CI.converter <- function(CI) {
+        sort(c(50-CI/2, 50+CI/2)/100)
+    }
+    get.extremes.quantile <- function(central, prob) {
+        ## Get the quantiles
+        quants <- quantile(central, probs = CI.converter(max(prob)))
+        ## Get the minimums and maximums
+        mins <- which(central <= quants[1])
+        maxs <- which(central >= quants[2])
+        ## Get their values
+        return(c(central[names(mins)], central[names(maxs)]))
+    }
+
+    if(use.hdr) {
+        output <- mapply(get.extremes.hdr, densities, centrals, SIMPLIFY = FALSE)
+        names(output) <- names(centrals)
+    } else {
+        output <- lapply(centrals, get.extremes.quantile, prob = prob)
+    }
+
+    ## Return the as a table
+    for(i in 1:length(output)) {
+        output[[i]] <- data.frame("level" = rep(names(output)[i], length(output[[i]])),
+                                  "species" = names(output[[i]]),
+                                  "score"   = unname(output[[i]]))
+    }
+    return(output)
+}
+```
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = FALSE}
+get.distributions(results$super_results_between, type = "position")
+```

### 2021-10-11 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
-```{r, echo = FALSE, fig.height = 18, fig.width = 18}
+```{r, echo = FALSE, fig.height = 18, fig.width = 18, eval = FALSE}
 par(mfrow = c(2,2))
@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE}
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}
 ## Run the analyses between groups
@@
 
-```{r, eval = TRUE, echo = FALSE, eval = FALSE}
-library(hrdcde)
+```{r, eval = TRUE, echo = FALSE, eval = TRUE}
+library(hdrcde)
 
-#@param results the results list
+#@param res the results list
 #@param type which type (position, distance or angle)
@@
 #@param prob which quantiles to use (default is c(50, 95, 99))
+#@param exclude.phylo remove the phylogeny subset (default = FALSE)
 #@param ... plot options
 
-get.distributions <- function(results, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), ...) {
+get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, ...) {
 
+    ## Remove the phylo one (if needed)
+    if(exclude.phylo) {
+        select <- names(size.subsets(res[[type]]))
+        select <- select[-which(select %in% "phylogeny")]
+        res[[type]] <- get.subsets(res[[type]], subsets = select)
+    }
     ## Extracting all the results
-    disparity <- lapply(results[[type]]$disparity, function(X, type) return(X$elements))
+    disparity <- lapply(res[[type]]$disparity, function(X, type) return(X$elements))
 
@@
         if(use.hdr) {
-            densities[[i]] <- hdr.den(centrals[[i]], prob = prob, main = names(centrals)[i], ...)
+            densities[[i]] <- hdr.den(centrals[[i]], prob = prob, main = names(centrals)[i])#, ...)
         } else {
@@
         mins <- which(central <= density$hdr[1,1])
-        maxs <- which(central >= density$hdr[1,2])
+        maxs <- which(central >= density$hdr[1, ncol(density$hdr)])
         ## Get their values
@@
     for(i in 1:length(output)) {
+        if(length(output[[i]]) == 0) {
+            output[[i]] <- data.frame("level" = names(output)[i], "species" = NA, "score" = NA)
+        } else {
             output[[i]] <- data.frame("level" = rep(names(output)[i], length(output[[i]])),
@@
         }
+    }
+
     return(output)
+}
+combine.results <- function(res1, res2) {
+    ## Combine the results
+    res_out <- rbind(res1[,1:2], res2[, 1:2])
+    res_out$elaboration <- c(res1[, 3], rep(NA, nrow(res2)))
+    res_out$exploration <- c(rep(NA, nrow(res1)), res2[, 3])
+    ## Find any species in commons
+    commons <- intersect(res1$species, res2$species)
+    if(length(commons) > 0 && is.na(commons)) {
+        if(is.na(res1$species)) {
+            if(is.na(res2$species)) {
+                ## Both are NA (keep just the first one)
+                res_out <- res_out[-1, , drop = FALSE]
+            } else {
+                ## Remove res1
+                res_out <- res_out[-1, , drop = FALSE]
+            }
+        } else {
+            ## Remove res2
+            res_out <- res_out[1:length(res1$species), , drop = FALSE]
+        }
+    } else {
+        ## Recursively remove the ones in common
+        while(length(commons) > 0) {
+            duplicates <- which(res_out$species == commons[1])
+            res_out[duplicates[1], 4] <- res_out[duplicates[2], 4]
+            res_out <- res_out[-duplicates[2], ]
+            commons <- commons[-1]
+        }
+    }
+    return(res_out)
 }
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = FALSE}
-get.distributions(results$super_results_between, type = "position")
+### Extreme species per super order (based on phylogeny)
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position")
 ```
+Figure: elaborations distributions per super orders
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+super_distances_phylo <- get.distributions(results$super_results_phylo, type = "distance")
+```
+Figure: explorations distributions per super orders
+
+```{r, echo = FALSE}
+library(knitr)
+## Combining the results
+super_phylo_table <- do.call(rbind, mapply(combine.results, super_positions_phylo, super_distances_phylo, SIMPLIFY = FALSE))
+rownames(super_phylo_table) <- NULL
+kable(super_phylo_table, digits = 3)
+```
+
+### Extreme species per order (based on phylogeny)
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 24, fig.width = 24, eval = TRUE}
+order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
+```
+Figure: elaborations distributions per order
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 24, fig.width = 24, eval = TRUE}
+order_distances_phylo <- get.distributions(results$order_results_phylo, type = "distance", exclude.phylo = TRUE)
+```
+Figure: explorations distributions per order
+
+```{r, echo = FALSE}
+## Combining the results
+order_phylo_table <- do.call(rbind, mapply(combine.results, order_positions_phylo, order_distances_phylo, SIMPLIFY = FALSE))
+rownames(order_phylo_table) <- NULL
+kable(order_phylo_table, digits = 3)
+```
+
+### Extreme species per super order (within)
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE)
+```
+Figure: elaborations distributions within super orders
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+super_distances_within <- get.distributions(results$super_results_within, type = "distance", exclude.phylo = TRUE)
+```
+Figure: explorations distributions within super orders
+
+```{r, echo = FALSE}
+## Combining the results
+super_phylo_table_within <- do.call(rbind, mapply(combine.results, super_positions_within, super_distances_within, SIMPLIFY = FALSE))
+rownames(super_phylo_table_within) <- NULL
+kable(super_phylo_table_within, digits = 3)
+```
+
+### Extreme species per order (within)
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+order_positions__within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE)
+```
+Figure: elaborations distributions within orders
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+order_distances_within <- get.distributions(results$order_results_within, type = "distance", exclude.phylo = TRUE)
+```
+Figure: explorations distributions within orders
+
+```{r, echo = FALSE}
+## Combining the results
+order_phylo_table_within <- do.call(rbind, mapply(combine.results, order_positions_within, order_distances_within, SIMPLIFY = FALSE))
+rownames(order_phylo_table_within) <- NULL
+kable(order_phylo_table_within, digits = 3)
+```

### 2021-10-11 Analysis/03.1-results.Rmd
typo

@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions__within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE)
+order_positions_within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE)
 ```

### 2021-10-12 Analysis/01-Data_preparation.Rmd
fixed levels names (to match the ones in the dataset!)

@@
 ```{r install_beer, eval = FALSE}
-## Installing beer
-devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+## Installing the dispRity branch
+devtools::install_github("TGuillerme/dispRity@MCMCglmm")
 ```
@@
 ## Loading packages
-library(dispRity)
-library(beer)
-library(knitr)
+library(dispRity) #v>=1.7
 ```
@@
 clades_list <- clades_list[rownames(clades_list) %in% rownames(shapespace), ]
+## Sorting the list by rownames from the shapespace
+clades_list <- clades_list[match(rownames(shapespace), rownames(clades_list)), -c(6, 7)]
+## Renaming the levels
+colnames(clades_list) <- c("superorder", "order", "suborder", "superfamily", "family")
 
-#@param table the clade list data frame
-#@param level the ID of the column to use as a level
-#@param threshold drop levels with less than the threshold
-make.level.list <- function(table, level, threshold = 20) {
-  ## Select the level
-  table_level <- table[, level]
-  ## Find the empty level
-  empty_level <- which(levels(table_level) == "")
-  if(length(empty_level) != 0) {
-      ## Remove the empties
-      level_list <- as.factor(as.character(table_level[-which(table_level == "")]))
-  }
-  empty_level <- ""
-  ## Select the levels below the threshold
-  remove_levels <- which(table(level_list) < threshold)
-  ## Remove the levels
-  while(length(remove_levels) > 0) {
-    level_list <- as.factor(as.character(level_list[-which(level_list == names(remove_levels)[1])]))
-    table_level[which(table_level == names(remove_levels)[1])] <- empty_level
-    table_level <- as.factor(as.character(table_level))
-    remove_levels <- remove_levels[-1]
-  }
-  ## Convert into a list
-  factors <- levels(level_list)
-  list_levels <- list()
-  while(length(factors) > 0) {
-    list_levels[[length(list_levels) + 1]] <- which(level_list == factors[1])
-    factors <- factors[-1]
-  }
-  ## Name the elements correctly
-  names(list_levels) <- levels(level_list)
-
-  ## Also returns the new column
-  return(list(col = table_level, list = list_levels))
+## Cleaning the levels (removing any level with less than 15 elements)
+update.level <- function(level, min = 15) {
+    to_remove <- which(table(level) < min)
+    remove_table <- data.frame("clade" = names(to_remove), "species" =  c(table(level)[to_remove]))
+    level[which(level %in% c(names(to_remove)))] <-  ""
+    return(list(level = as.factor(as.character(level)), removed = remove_table))
 }
 
-## Reading the clades list
-groups_level1 <- make.level.list(clades_list, level = 1, threshold = 20)
-groups_level2 <- make.level.list(clades_list, level = 2, threshold = 20)
-all_groups <- c(groups_level1$list, groups_level2$list)
-group_names_level_1_2 <- all_groups
-save(group_names_level_1_2, file = "../Data/Processed/group_names_level_1_2.rda")
+## Updated list
+update_clade_list <- lapply(as.list(clades_list), update.level, min = 15)
+clade_list <- data.frame(lapply(update_clade_list, `[[`, 1))
+removed_table <- do.call(rbind, lapply(update_clade_list, function(x) {rownames(x$removed) <- NULL; return(x$removed)}))
+
+## Saving the clade list
+save(clade_list, file = "../Data/Processed/clade_list.rda")
+```
+
+```{r, echo = FALSE}
+library(knitr)
+kable(removed_table, caption = "List of removed clades by level and number of species in that clade in the dataset.", label = "Table X1")
 ```
@@
 ```{r}
+## Temporary change the empty levels to noclade
+clades <- clades_list[, c(1)]
+levels(clades)[1] <- "noclade"
+
 ## Make a dispRity object and get axes data
-shape_axes <- select.axes(custom.subsets(shapespace, group = all_groups))
-form_axes  <- select.axes(custom.subsets(formspace, group = all_groups))
+shape_axes <- select.axes(custom.subsets(shapespace, group = clades))
+form_axes  <- select.axes(custom.subsets(formspace, group = clades))
 ```
@@
 ```{r, echo = FALSE}
-kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X1")
+kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X2")
 ```
@@
 ```{r, echo = FALSE}
-kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X2")
+kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X3")
 ```
@@
 ```
-Fig X1: variance per axes for the Shape space
+Fig X1: variance per axes per superorder and order in the shapespace
 
@@
 ```
-Fig X1: variance per axes for the Form space
+Fig X1: variance per axes per superorder and order in the formspace
 

### 2021-10-12 Functions/prep.data.R

@@ -50,21 +50,20 @@ prep.data <- function(level, lvl.inc, clades, space, dim, consensus, trees, min 
   get.level.list <- function(one_list, clades, lvl.inc, min, verbose) {
 
     ## Separate the data by levels
-    get.levels <- function(data, min) {
-      ## Make the data as factor
-      factors <- as.factor(as.character(data))
+    get.levels <- function(data, min) {      
+      ## Separate the data by levels
+      levels_list <- lapply(as.list(levels(data[,1])), function(level, data) which(data == level), data = data)
+      ## Add the species names
+      levels_list <- lapply(levels_list, function(one_level, data) {names(one_level) <- rownames(data)[one_level]; return(one_level)}, data = data)
+      names(levels_list) <- levels(data[,1])
 
-      ## Remove empty levels
-      if(any(empty <- levels(factors) == "")) {
-        factors <- as.factor(as.character(data[-which(data == "")]))
+      ## Remove "" clade
+      if(any(empty <- which(names(levels_list) == ""))) {
+        levels_list <- levels_list[-empty]
       }
 
-      ## Separate the data by levels
-      levels_list <- lapply(as.list(levels(factors)), function(level, data) which(data == level), data = data)
-      names(levels_list) <- levels(factors)
-
-      ## Remove lists to smal
-      if(length(to_small <- which(unlist(lapply(levels_list, length)) <= min)) > 0) {
+      ## Remove lists to small
+      if(length(to_small <- which(unlist(lapply(levels_list, length)) < min)) > 0) {
         levels_list <- levels_list[-to_small]
       }
       return(levels_list)
@@ -74,7 +73,7 @@ prep.data <- function(level, lvl.inc, clades, space, dim, consensus, trees, min 
     levels_out <- list()
     for(i in 1:length(lvl.inc)) {
       ## Get the levels for this set of species
-      data <- clades[one_list$species, lvl.inc[i]]
+      data <- clades[one_list$species, lvl.inc[i], drop = FALSE]
       levels_out[[i]] <- get.levels(data, min)
     }
     return(levels_out)
@@ -93,11 +92,13 @@ prep.data <- function(level, lvl.inc, clades, space, dim, consensus, trees, min 
     
     ## Adding the levels
     level_columns <- replicate(length(levels), rep("", nrow(sub_space)))
+    rownames(level_columns) <- rownames(sub_space)
+    
     ## Filling in the levels
     for(i in 1:length(levels)) {
[…]

### 2021-10-19 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

--- 2021-09-24 Analysis/03-elaboration_exploration_analyses.Rmd
@@
 ```{r, eval = TRUE, echo = FALSE, print = FALSE}
-library(beer)
-## Loading the MCMCglmm
-load("../Data/processed/combined_results.rda")
+library(dispRity)
 ## Loading the tree and data
-data(demo_data)
-data(demo_tree)
+data(charadriiformes)
 ## Setting the colour vector
@@
 ```{r, eval = TRUE, echo = FALSE}
-covar_matrices <- MCMCglmm.dispRity(
-                  data          = demo_data,
-                  posteriors    = combined_results,
-                  group         = MCMCglmm.levels(combined_results)[1:4],
+covar_matrices <- MCMCglmm.subsets(
+                  data          = charadriiformes$data,
+                  posteriors    = charadriiformes$posteriors,
+                  group         = MCMCglmm.levels(charadriiformes$posteriors)[1:4],
                   rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
@@
 ## Plotting just the space
-sauron.plot(covar_matrices,
+covar.plot(covar_matrices,
             col = colour_vector,
@@
 ```{r, echo = FALSE, fig.height = 6, fig.width = 6}
-sauron.plot(covar_matrices,
+covar.plot(covar_matrices,
             col = colour_vector,
@@
 ## Plotting 100 axes
-sauron.plot(covar_matrices,
+covar.plot(covar_matrices,
             n = 100,
@@
 ## Plotting 100 ellipses
-sauron.plot(covar_matrices,
+covar.plot(covar_matrices,
             n = 100,
@@
 ## Run the analyses between groups
-results_between <- covar.projections.wrapper(covar_matrices,
-                                             type = "between",
-                                             base = "phylogeny",
-                                             dispRity.out = TRUE)
+results_between <- dispRity.covar.projections(covar_matrices,
+                                             type = "groups",
+                                             base = "phylogeny")
 ## Plot the results
@@
 ## Run the analyses of groups on the phylogeny
-results_phylo <- covar.projections.wrapper(covar_matrices,
-                                            type = "within",
-                                            base = "phylogeny",
-                                            dispRity.out = TRUE)
+results_phylo <- dispRity.covar.projections(covar_matrices,
+                                            type = "elements",
+                                            base = "phylogeny")
 ## Plot the results
@@
           main = "Projection on the\nglobal phylogeny axes")
+# source("../Functions/circular.plot.R")
+
+# circular.plot.phytools(data = results_phylo$position, tree = demo_tree, gradient = c("blue", "orange"), clade.names = TRUE, clades = colour_vector[-4], lwd = 3)
 ```
@@
 ## Run the analyses groups on themselves
-results_within <- covar.projections.wrapper(covar_matrices,
-                                            type = "within",
-                                            dispRity.out = TRUE)
+results_within <- dispRity.covar.projections(covar_matrices,
+                                            type = "elements")
 ## Plot the results
@@
 Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).
-
-```{r, echo = FALSE, eval = TRUE}
-library(phytools)
-source("~/Packaging/phytools/R/plotTree.wBars.R")
-
-get.disparity.matrix <- function(data, col) {
-    ## Get the matrices
-    matrices <- lapply(data$disparity, `[[`, 1)
-    ## Add the rownames
-    list_of_names <- list()
-    for(one_group in 1:length(matrices)) {
-        rownames(matrices[[one_group]]) <- list_of_names[[one_group]] <- rownames(data$matrix[[1]])[c(data$subsets[[one_group]]$elements)]
-    }
-    ## Name the groups
-    names(list_of_names) <- names(data$subsets)
-
-    ## Matrix
-    matrix <- do.call(rbind, matrices)
-
-    ## Colours
-    if(!missing(col)) {
-        name.cols <- function(names, col) {
-            out <- rep(col, length(names))
-            names(out) <- names
-            return(out)
-        }
-        col <- mapply(name.cols, list_of_names, as.list(col))
-        names(col) <- NULL
-        col <- unlist(col)
-    }
-
-    ## Output
-    if(missing(col)) {
-        return(list(data = matrix, names = list_of_names))
-    } else {
-        return(list(data = matrix, names = list_of_names, col = col))
-    }
-}
-
-## Get the different elements
-phylo_elaboration  <- get.disparity.matrix(results_phylo$position,
-                                          col = colour_vector[-4])
-within_elaboration <- get.disparity.matrix(
-    get.subsets(results_within$position, subsets = c(1,2,3)),
-                                          col = colour_vector[-4])
-phylo_exploration  <- get.disparity.matrix(results_phylo$distance,
-                                          col = colour_vector[-4])
-within_exploration <- get.disparity.matrix(
-    get.subsets(results_within$distance, subsets = c(1,2,3)),
-                                          col = colour_vector[-4])
-phylo_angle        <- get.disparity.matrix(results_phylo$degree,
-                                          col = colour_vector[-4])
-within_angle       <- get.disparity.matrix(
-    get.subsets(results_within$degree, subsets = c(1,2,3)),
-                                          col = colour_vector[-4])
-```
 

### 2021-10-19 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 ```{r, eval = TRUE, echo = FALSE, print = FALSE}
-library(beer)
+library(dispRity)
 ```
@@
 
-superorder_shapespace <- MCMCglmm.dispRity(
+superorder_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.dispRity(
+order_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 24, fig.width = 24, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
 order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 24, fig.width = 24, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
 order_distances_phylo <- get.distributions(results$order_results_phylo, type = "distance", exclude.phylo = TRUE)
@@
 ```
+
+## Correlations between elaboration and exploration?
+
+```{r, echo = FALSE}
+get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE, ...) {
+
+    ## Remove the phylo one (if needed)
+    if(exclude.phylo) {
+        select_pos <- names(size.subsets(res$position))
+        select_dis <- names(size.subsets(res$distance))
+        select_pos <- select_pos[-which(select_pos %in% "phylogeny")]
+        select_dis <- select_dis[-which(select_dis %in% "phylogeny")]
+        res$position <- get.subsets(res$position, subsets = select_pos)
+        res$distance <- get.subsets(res$distance, subsets = select_dis)
+    }
+
+    ## Extracting all the results
+    disparity_pos <- lapply(res$position$disparity, function(X, type) return(X$elements))
+    disparity_dis <- lapply(res$distance$disparity, function(X, type) return(X$elements))
+
+    ## Recentre the results
+    disparity_pos <- lapply(disparity_pos, function(X) return(X-0.5))
+
+    ## Measure the central tendencies
+    centrals_pos <- lapply(disparity_pos, function(X, fun) apply(X, 1, fun), fun = cent.tend)
+    centrals_dis <- lapply(disparity_dis, function(X, fun) apply(X, 1, fun), fun = cent.tend)
+
+    ## Number of groups
+    n_groups <- length(centrals_pos)
+
+    ## Calculate and plot the hdrcde
+    op_tmp <- par(mfrow = c(ceiling(sqrt(n_groups)), round(sqrt(n_groups))))
+    for(i in 1:n_groups) {
+        ## Testing the correlation
+        cor_test <- cor.test(centrals_pos[[i]], centrals_dis[[i]])
+        plot_main <- paste0(names(centrals_pos)[i], ": ", unname(round(cor_test$estimate, 3)), " (p=", unname(round(cor_test$p.value, 3)), ")")
+        ## Plotting the results
+        plot(centrals_pos[[i]], centrals_dis[[i]], pch = 19, xlab = "elaboration", ylab = "exploration", main = plot_main, cex = 0.5)
+    }
+    par(op_tmp)
+    return(invisible())
+}
+```
+
+### Correlations per super order (based on phylo)
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+get.correlations(results$super_results_phylo)
+```
+
+### Correlations within super orders
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+get.correlations(results$super_results_within, exclude.phylo = TRUE)
+```
+
+### Correlations per order (based on phylo)
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+get.correlations(results$order_results_phylo, exclude.phylo = TRUE)
+```
+
+### Correlations within orders
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
+get.correlations(results$order_results_within, exclude.phylo = TRUE)
+```

### 2021-10-27 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Create 02.1-shapespace_allbirds_lvl_superorder_order.Rmd

--- 2021-09-20 Analysis/02.1-MCMCglmm_mini_chains_cluster.Rmd
@@
 
-```{r install_beer, eval = FALSE}
-## Installing beer
-devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
-```
-
-```{r load_beer, eval = TRUE, message = FALSE}
+```{r load_packages, eval = TRUE, message = FALSE}
 set.seed(42)
@@
 ## Loading the parametrisation data
-load(file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
 
 ## Select the data
-shapespace <- form_shape_space_1tree_2levels$shape_space
-groups_list <- form_shape_space_1tree_2levels$groups_list
-selected_dim <- form_shape_space_1tree_2levels$shape_dim
-consensus_tree <- form_shape_space_1tree_2levels$consensus_tree
-
+shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
+selected_dim <- shapespace_allbirds_lvl_superorder_order[[1]]$dimensions
+consensus_tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
+trees_list <- shapespace_allbirds_lvl_superorder_order[[1]]$trees_list
 ```
@@
 for(i in 1:3) {
-    load(paste0("../Data/Cluster/allbirds_8748t_8Dshapespace_31levels/param_chains/params_shapespace_8D_31levels_20k_chain", i, ".rda"))
-    param_chains[[i]] <- params_shapespace_8D_31levels[[1]]
+    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/shapespace_allbirds_lvl_superorder_order_param", i, ".rda"))
+    param_chains[[i]] <- shapespace_allbirds_lvl_superorder_order_param[[1]]
 }
@@
 ```{r, set_the_mini_chains, eval = TRUE}
-## Get the distribution of trees
-load(file = "../Data/Processed/trees_list.rda")
-
 ## The thinning parameter
@@
 ## Set up the mini-chains models
-shapespace_8748t_8D_32level <- make.mini.chains(
+shapespace_allbirds_lvl_superorder_order <- make.mini.chains(
                                 data         = shapespace,
                                 dimensions   = selected_dim,
-                                tree         = trees_list[sample(1:1000, 20)],
+                                tree         = trees_list[sample(1:1000, 1000)],
                                 trait.family = "gaussian",
@@
 ## Save the mini.chains
-save(shapespace_8748t_8D_32level, file = "../Data/Cluster/allbirds_8748t_8Dshapespace_31levels/shapespace_8748t_8D_32level.mini.chains")
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order.mini.chains")
 ```
@@
 class(model_list) <- c("beer", "mini.chains")
-models_path <- "../Data/Cluster/allbirds_8748t_8Dshapespace_31levels/cluster_out/"
+models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster_out/"
 ## Loading all models
@@
 ## Combining the model files
-allbirds_8748t_8Dshapespace_31levels <- combine.mini.chains(model_list)
+shapespace_allbirds_lvl_superorder_order <- combine.mini.chains(model_list)
+
+## Checking the diagnosis
 
 ## Saving the results
-save(allbirds_8748t_8Dshapespace_31levels, file = "../Data/processed/allbirds_8748t_8Dshapespace_31levels.rda")
-```
-
-<!--
-
-#### Number of runs optimisation (TO BE DEVELOPED?)
-
-```{r, eval = FALSE}
-## Function for running one mini chain on a random tree (and saving the results)
-fun.mini.chains <- function() {
-    run.mini.chains(mini_chains, replicates = 1, path = "../Data/Processed/MCMC/", file.name = "mini.chains")
-}
-```
-
-To make sure we reached consistent results, we run the set of N mini-chains were N was determined using the following variance optimisation algorithm.
-
- * 1. run an initial $N = 20$ chains;
- * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
- * 3. update N to N = N * 1.05 and run an additional 5% chains;
- * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
- * 5. calculate $\Delta_V$, the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
- * 6.
-If $\Delta_V < 0.05$ (meaning an increase of 5% of the replicates resulted in an increase of the parameters of less than 5%), the desired number of mini-chains is reached.
-Else, set $V_{1} = V_{2}$ and repeat from step 3.
-
-```{r, optimised_replicates, eval = FALSE}
-## The summary function
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/processed/shapespace_allbirds_lvl_superorder_order.rda")
 
 ```
-
-After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "standard" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.
-
--->
-
-## References

### 2021-11-01 mcmcmcglmmm/R/combine.mini.chains.R
removed non utf

@@ -32,11 +32,11 @@ combine.mini.chains <- function(mini.chains) {
     output$Sol <- as.mcmc(do.call(rbind, lapply(mini.chains, `[[`, "Sol")), mcpar)
     output$VCV <- as.mcmc(do.call(rbind, lapply(mini.chains, `[[`, "VCV")), mcpar)
 
-    ## Get the deviances
+    ## Get the deviances
     output$Deviance <- as.mcmc(do.call(c, lapply(mini.chains, `[[`, "Deviance")), mcpar)
     output$DIC <- mean(do.call(c, lapply(mini.chains, `[[`, "DIC")))
 
-    ## Return the whole bunch
+    ## Return the whole bunch
     class(output) <- class(mini.chains[[1]])
     return(output)
 }

### 2021-11-01 beer/R/MCMCglmm.dispRity.R
removed beer (now handled by dispRity@MCMCglmm)

@@ -1,141 +1,0 @@
-#' @title MCMCglmm.dispRity
-#'
-#' @description Creating a dispRity object from a MCMCglmm posterior output
-#'
-#' @param data The data used for the MCMCglmm model
-#' @param posteriors A MCMCglmm object, the posteriors of the model
-#' @param group Optional, a named vector of which group to include from the posteriors (if left empty the random and residual terms are used). See details.
-#' @param tree Optional, the tree(s) used in the MCMCglmm analyses.
-#' @param rename.groups optional, a vector of group names for renaming them. See details.
-#' 
-#' @details
-#' \itemize{
-#'      \item For the \code{group} option, the group names must be ones found in the \code{posteriors} formula in the format \emph{<Type = Term:FactorLevel>} as returned by \code{MCMCglmm.levels(posteriors)}. For example, for returning two random effect, the phylogenetic one (\code{"animal"}) and one for a specific clade (say the 2nd clade) as well as two residual terms for a specific factor (say level 1 and 4) you can use \code{group = c(random = "animal", random = "animal:clade2", residual = "units:myfactor1", residual = "units:myfactor4")}.
-#'      \item For the \code{rename.groups} option, the vector must be of class \code{"character"} and must of the same length as the number of random and residual terms in \code{posteriors} or of \code{group} argument (if used). If the \code{group} argument is left empty, the groups are extracted from the \code{posteriors} in the following order: the random terms first then the residual terms as specified in the \code{posteriors} object formulas (respectively \code{posteriors$Random$formula} and \code{posteriors$Residual$formula}).
-#' 
-#' 
-#' }
-#' 
-#' @examples
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme
-#' @export
-
-MCMCglmm.dispRity <- function(data, posteriors, group, tree, rename.groups) {
-
-    ## Cleaning then checking the data (i.e. only removing the values)
-    #dispRity_export in: data_class <- check.class(data, c("data.frame", "matrix"))
-    data_class <- class(data)[[1]] #dispRity_export out:
-    if(data_class == "data.frame") {
-        ## Remove potential non-numeric columns
-        col_classes <- sapply(1:ncol(data), function(col, dat) class(data[,col]), dat = data)
-        numerics <- (col_classes %in% c("numeric", "integer"))
-        ## Check for non-numerics
-        if(!any(numerics)) {
[…]

### 2021-11-01 beer/R/as.covar.R

@@ -1,30 +1,0 @@
-#' @title Use covar
-#'
-#' @description Adapts a dispRity.metric to use a covar element from a dispRity object
-#'
-#' @param metric The metric to adapt as a covar one.
-#' 
-#' @examples
-#' data(BeckLee_mat50)
-#' ## Using the centroids metric on a normal disparity object
-#' dispRity(BeckLee_mat50, metric = centroids)
-#' ## Same for the mean centroids
-#' dispRity(BeckLee_mat50, metric = c(mean, centroids))
-#' 
-#' ## Using it on a dispRity object containing a covar part
-#' dispRity(BeckLee_mat50, metric = as.covar(centroids))
-#' ## Same for the mean centroids
-#' dispRity(BeckLee_mat50, metric = c(mean, as.covar(centroids)))
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme
-#' @export
-
-as.covar <- function(metric, ...) {
-    if("matrix2" %in% names(formals(metric))) {
-        return(function(matrix, matrix2, ..) metric(matrix = matrix$VCV, matrix2 = matrix2$VCV, ...))
-    } else {
-        return(function(matrix, ..) metric(matrix = matrix$VCV, ...))
-    }
-}

### 2021-11-01 beer/R/covar.projections.wrapper.R

@@ -1,191 +1,0 @@
-#' @title Covar projection analyses wrapper
-#'
-#' @description Wrapper function for a covar projection analyses
-#'
-#' @param data a \code{dispRity} object containing a covar component
-#' @param type either \code{"within"} for the projection of the group onto itself or \code{"between"} for the projections of the groups onto the \code{base} (see details)
-#' @param sample optional, one or more specific posterior sample IDs (is ignored if n is used) or a function to summarise all axes.
-#' @param n optional, a random number of covariance matrices to sample (if left empty, all are used).
-#' @param base optional, if \code{type = "between"}, a \code{character} string for the name of the subset in data to use as a baseline (e.g. the phylogeny) or if \code{type = "within"} the major axis to use as a base (see details).
-#' @param incl.base optional, if \code{base} is provided and \code{type = "within"}, whether to include the comparison with the base itself (\code{TRUE})or not (default; \code{FALSE}).
-#' @param major.axis which major axis to use (default is \code{1})
-#' @param level the confidence interval to estimate the major axis (default is \code{0.95})
-#' @param measure which parameters to measure from the projection. Can be any of the following \code{"c(position, distance, degree)"} (default) for respectively the distance on the projection, the distance from the projection and the angle of the projection.
-#' @param verbose Whether to be verbose (\code{TRUE}) or not (\code{FALSE}, default).
-#' @param dispRity.out Whether to make the output into a list of dispRity objects (\code{TRUE}) or not (\code{FALSE}, default).
-#' 
-#' @details
-#' Effectively, the wrapper runs either of the following function (simplified here):
-#' \itemize{
-#'      \item if \code{type = "between"}: \code{dispRity(data, metric = projections.covar, between.groups = TRUE)} for the pairwise projections between each subset in \code{data}.
-#'      \item if \code{type = "within"}: \code{dispRity(data, metric = projections, point1 = axis[1,], point2 = axis[1,])} for the projections between each subset in \code{data} onto an axis. 
-#' }
-#' If \code{base} is specified:
-#' \itemize{
-#'      \item \code{type = "between"} will run pairs between each subset and \code{base} (instead of the full pairwise analyses).
-#'      \item \code{type = "within"} will run the projection of each subset onto the major axis from \code{base} rather than its own.
-#' }
-#' 
-#' @returnssss
[…]

### 2021-11-01 beer/R/covar.projections.wrapper_fun.R

@@ -1,112 +1,0 @@
-dispRitize <- function(one_result, data, name, fun, type, cent.tend = median) {
-    ## Copying the data
-    output <- data
-    ## Adding the disparity results
-    output$disparity <- lapply(one_result, function(X) list(elements = X, X))
-    # ## Summarising the results for the "elements" part and adding rownames
-    # for(i in 1:n.subsets(data)) {
-    #     output$disparity[[i]] <- matrix(apply(output$disparity[[i]]$elements, 1, cent.tend), ncol = 1, dimnames = list(c(rownames(data$matrix[[1]])[c(data$subsets[[i]]$elements)])))    
-    # }
-        ## Adding the call
-    output$call$bootstrap <- list(ncol(one_result[[1]]), "covar", NULL)
-    output$call$disparity$metrics$name[[1]] <- name
-    output$call$disparity$metrics$fun[[1]]  <- fun
-    output$call$disparity$metrics$between.groups <- (type == "between")
-    return(output)
-}
-
-## Between type
-# dispRity.covar(data, metric = projections.covar, between.groups = list_of_pairs, measure = measure)$disparity
-
-
-
-## Run projections faster!
-## Angle between two vectors
-vector.angle <- function(v1, v2){
-    return(acos(geometry::dot(v1, v2, d = 1) / (sqrt(sum(v1^2))*sqrt(sum(v2^2)))) *180/pi)
-}
-## Rotate a matrix along one axis (y)
-get.rotation.matrix <- function(x, y){
-    ## This magic comes from https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385
-    ## following: https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions
-    ## Also this: http://wscg.zcu.cz/wscg2004/Papers_2004_Short/N29.pdf
-    u <- x/sqrt(sum(x^2))
-
-    v <- y-sum(u*y)*u
-    v <- v/sqrt(sum(v^2))
-
-    cost <- sum(x*y)/sqrt(sum(x^2))/sqrt(sum(y^2))
-    sint <- sqrt(1-cost^2);
-
-    return(diag(length(x)) - u %*% t(u) - v %*% t(v) + cbind(u,v) %*% matrix(c(cost,-sint,sint,cost), 2) %*% t(cbind(u,v)))
-}
-## Projection of elements on an axis
[…]

### 2021-11-01 beer/R/covar.utilities.R

@@ -1,78 +1,0 @@
-#' @name get.covar
-#' @aliases axes.covar ellipse.covar
-#' @title Utilities for a dispRity object with covariance matrices
-#'
-#' @description Different utility functions to extract aspects of a \code{MCMCglmm} object.
-#'
-#' @usage get.covar(data, n)
-#' @usage axis.covar(data, n, level = 0.95, axis = 1)
-#'
-#' @param data a \code{dispRity} object with a \code{covar} element.
-#' @param sample optional, one or more specific posterior sample IDs (is ignored if n is used) or a function to summarise all axes.
-#' @param n optional, a random number of covariance matrices to sample (if left empty, all are used).
-#' @param level which confidence interval level to use (default is \code{0.95}).
-#' @param axis which major axis to calculate (default is \code{1}, the first one).
-
-#' 
-#' @examples
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme
-#' @export
-get.covar <- function(data, sample, n) {
-
-    ## Some sanitizing on data, sample, n
-
-    ## Just return everything!
-    if(missing(sample) && missing(n)) {
-        return(sample.n(data$MCMCglmm$covars, n = length(data$MCMCglmm$covars[[1]])))
-    }
-
-    ## Just return the n random samples
-    if(!missing(n)) {
-        if(!missing(sample)) {
-            warning("sample argument is ignored since n = ", n, " random samples are asked for.")
-        }
-        return(sample.n(data$MCMCglmm$covars, n = n))
-    }
-
-    ## Return specific samples
-    if(!missing(sample)) {
-        if(is(sample, "function") || is(sample, "standardGeneric")) {
-            ## Summarise the results
-            return(lapply(lapply(sample.n(data$MCMCglmm$covars, n = length(data$MCMCglmm$covars[[1]])), summarise.fun, fun = sample), list))
-        } else {
-            ## Return specific samples
-            return(sample.n(data$MCMCglmm$covars, selected_n = sample))
-        }
-    }
-}
-
-axis.covar <- function(data, sample, n, dimensions, level = 0.95, axis = 1) {
-    
-    ## Sanitizing on level and axis
-
-    ## Selecting the dimensions (all by default)
-    if(missing(dimensions)) {
-        dimensions <- data$call$dimensions
-    }
-
-    ## Handle sample
-    if(!missing(sample) && (is(sample, "function") || is(sample, "standardGeneric"))) {
-        return(lapply(
[…]

### 2021-11-01 beer/R/covar.utilities_fun.R

@@ -1,69 +1,0 @@
-## Internal for sauron.plot and covar.utilties
-## Selecting n elements from a covar object
-sample.n <- function(covar, n, selected_n) {
-    ## Get the posterior size
-    n_post <- length(covar[[1]])
-    if(!missing(n) && n == n_post) {
-        return(covar)
-    } else {
-        ## Sample n
-        if(missing(selected_n)) {
-            selected_n <- sample(1:n_post, n, replace = n > n_post)
-        }
-        ## Return n for each group
-        return(lapply(covar, function(group, n) group[n], n = selected_n))
-    }
-}
-
-## Internal: get the coordinates of one axes
-get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {
-
-    # The magic: https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331
-
-    ## Select the right dimensions
-    data$VCV <- data$VCV[dimensions, dimensions, drop = FALSE]
-
-    ## Get the data dimensionality
-    dims <- dim(data$VCV)[1]
-
-    ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
-    unit_hypersphere1 <- unit_hypersphere2 <- matrix(0, ncol = dims, nrow = dims)
-    ## The "front" (e.g. "top", "right") units
-    diag(unit_hypersphere1) <- 1
-    ## The "back" (e.g. "bottom", "left") units
-    diag(unit_hypersphere2) <- -1
-    unit_hypersphere <- rbind(unit_hypersphere1, unit_hypersphere2)
-    ## Scale the hypersphere (where level is the confidence interval)
-    unit_hypersphere <- unit_hypersphere * sqrt(qchisq(level, 2))
-
-    ## Do the eigen decomposition (symmetric - faster)
-    eigen_decomp <- eigen(data$VCV, symmetric = TRUE)
-
-    ## Re-scaling the unit hypersphere
-    scaled_edges <- unit_hypersphere * rep(sqrt(eigen_decomp$values), each = dims*2)
-    ## Rotating the edges coordinates
-    edges <- tcrossprod(scaled_edges, eigen_decomp$vectors)
-
-    ## Move the matrix around
-    edges <- edges + rep(data$Sol[dimensions, drop = FALSE], each = dims*2)
-
-    ## Get the edges coordinates
[…]

### 2021-11-01 beer/R/dispRity.covar.R

@@ -1,56 +1,0 @@
-#' @title Temporary wrapper for dispRity
-#'
-#' @description TO BE PROPERLY INCLUDED IN dispRity
-#'
-#' @param data the data with covar
-#' @param metric the metric
-#' @param between.groups (for group analyses or tip analyses)
-#' 
-#' @examples
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme
-#' @export
-
-dispRity.covar <- function(data, metric, between.groups = FALSE, ...) {
-   
-
-   if(is(between.groups, "logical") && !between.groups) {
-      ## Do the per tip analyses
-      disparity <- lapply(lapply(data$MCMCglmm$covars, lapply, metric, ...), unlist)
-
-
-
-   } else {
-      ## Do the per group analyses
-      if(is(between.groups, "logical") && between.groups) {
-         list_of_pairs <- unlist(apply(combn(1:n.subsets(data), 2), 2, list), recursive = FALSE)
-      } else {
-         list_of_pairs <- between.groups
-         between.groups <- TRUE
-      }
-      ## Get the list of pairs names
-      names(list_of_pairs) <- lapply(list_of_pairs, function(one_pair, data) return(paste0(names(data$subsets)[one_pair[1]], ":", names(data$subsets)[one_pair[2]])), data)
-
-      ## Make the list of pairs (super rough - done properly in dispRity)
-      covar_pairs <- lapply(list_of_pairs, function(one_pair, data) return(list(data$MCMCglmm$covars[[one_pair[1]]], data$MCMCglmm$covars[[one_pair[2]]])), data = data)
-
-      ## Mapply all that
-      mapply.one.pair <- function(one_pair, metric, ...) {
-         values <- unlist(mapply(metric, one_pair[[1]], one_pair[[2]], MoreArgs = list(...), SIMPLIFY = FALSE))
-         return(list("elements" = mean(values), matrix(values, nrow = 1)))
-      }
-      ## Calculate disparity
-      disparity <- lapply(covar_pairs, mapply.one.pair, metric, ...)
-      #disparity <- lapply(covar_pairs, mapply.one.pair, metric)
-   }
-
-   ## Updating the object   
-   data$disparity <- disparity
-   data$call$subsets <- "custom" #TO BE covars
-   data$call$bootstrap <- list(length(data$MCMCglmm$covars[[1]]), "covars", FALSE) # TO be handled normally
[…]

### 2021-11-01 beer/R/dispRity.fast.R

@@ -1,52 +1,0 @@
-#' @title Fast dispRity
-#'
-#' @description Fast dispRity calculations (but less safe than dispRity, see details)
-#'
-#' @param group a logical vector for grouping
-#' @param space a matrix
-#' @param metric a metric dispRity style (up to two levels)
-#' 
-#' @details
-#' \emph{USE \code{\link{dispRity}} INSTEAD OF THIS FUNCTION.}
-#' This function should only be used in very specific cases requiring advanced optimisation.
-#' This function is simply applying \code{metric(space[group,])} for each group and returns a list of results.
-#' It does not check the validity of the data, metric and groups.
-#' It does not handle specific data (e.g. multiple matrices), specific metrics (e.g. no optional arguments), does not produce meaningful error messages and does not intake nor returns a dispRity object.
-#' 
-#' @examples
-#' ## A random space
-#' space <- matrix(rnorm(25), 5, 5)
-#' 
-#' ## A metric
-#' metric <- c(sum, variances)
-#' 
-#' ## A group of four observations
-#' group <- c(TRUE, TRUE, TRUE, TRUE, FALSE)
-#' 
-#' ## The disparity
-#' dispRity.fast(group, space, metric)
-#'
-#' 
-#' @author Thomas Guillerme
-
-dispRity.fast <- function(group, space, metric) {
-    ## Setting up the default args
-    args <- list(matrix = space[group, ])
-    ## Simple level 1 metric
-    if(length(metric) == 1) {
-        return(do.call(metric, args))
-    } 
-    ## Simple level 2 + 1 metric
-    if(is.null(names(metric))) {
-        return(metric[[1]](do.call(metric[[2]], args)))
-    }
-    ## Handle the named arguments
-    args <- c(args, metric[-1])
-
-    ## Level 1 metric + args
-    if(length(metric[[1]]) == 1) {
-        return(do.call(metric[[1]], args))
-    } 
-    ## Level 2 + 1 metric + args
-    return(metric[[1]][[1]](do.call(metric[[1]][[2]], args)))
-}

### 2021-11-01 beer/R/optim.replicate.R

@@ -1,207 +1,0 @@
-#' @title Optimised replicates
-#'
-#' @description A wrapper for replicating a based on results variance
-#' 
-#' @param input.fun The function to execute. The function should output a single value or a vector or a list of values.
-#' @param diagnose The diagnosis function should intake a vector of values.
-#' @param summarise Optional, if the output of \code{input.fun()} is complex, a function to summarise that output and pass it to \code{diagnose()}.
-#' @param minimum The minimum of replicates to run (default is 20)
-#' @param maximum Optional, a maximum of iterations (default is Inf).
-#' @param stop.variance The percentage of variance change for stopping (default is 0.05).
-#' @param increment The percentage of increment (if left empty, the value from stop.variance is used).
-#' @param verbose Whether to be verbose (\code{TRUE}) or not (\code{FALSE}; default)
-#' @param parallel Whether to use the \code{\link[future.apply]{future_replicate}} function for parallelisation (\code{TRUE}) or not (\code{FALSE}; default).
-#' @param par.plan If \code{parallel = TRUE}, which plan to use (see \code{\link[future]{plan}}; by default, the plan is "future::multisession").
-#' @param bkp.path Optional, a \code{"character"} string to be the path to a backup file. See details.
-#' @param bkp.name Optional, if \code{bkp.path} is not missing, a \code{"character"} for the name of the file to save (or to load). See details.
-#' @param ... Any optional arguments to be passed to the function.
-#' 
-#' @details When \code{bkp} is specified, the algorithm first looks if a file with \code{bkp.name} exists (by default, the name is \code{"optim.replicate.bkp"}). If not, the results will be saved after each series of replicates into this file. If it exist, the initial steps are skipped and the algorithm continues from the last backup point. \emph{NOTE} that this option slows down the simulations.
-#' 
-#' @return
-#' A list of:
-#' \itemize{
-#'      \item{results}: a \code{matrix} with the summarised output from \code{fun()} for every set of iterations;
-#'      \item{diagnoses}: a \code{matrix} with changes in variances after each diagnosis;
-#'      \item{outputs}: a \code{list} of the outputs of each iteration of \code{fun()}.
-#' }
-#' 
-#' @examples
-#'
-#' @seealso
-#' 
[…]

### 2021-11-01 beer/R/projections.covar.R

@@ -1,31 +1,0 @@
-#To export in dispRity.metrics
-#' @param ... the projections arguments
-#' @param axis which major axes to use
-#' @param level which confidence interval to use for calculating the major axis
-#' 
-#' 
-projections.covar <- function(matrix, matrix2, measure = "position", scaled = TRUE, axis = 1, level = 0.95) {
-
-    ## Get the main axes from the VCV matrices
-    # source("covar.utilities_fun.R")
-    base_vector  <- get.one.axis(matrix2, axis, level, dimensions = 1:nrow(matrix2$VCV))
-    projected_vector <- get.one.axis(matrix, axis, level, dimensions = 1:nrow(matrix$VCV))
-
-    ## Translating into projections format
-    matrix <- projected_vector
-    point1 <- base_vector[1,]
-    point2 <- base_vector[2,]
-
-    ## Moving the two axes so that there origins are the same:
-    ## 1 - Get the translation vector to point1
-    translation_vector <- point1 - matrix[1,]
- 
-    ## 2 - Align the matrix with point1
-    matrix <- rbind(matrix[1,] + translation_vector,
-                    matrix[2,] + translation_vector)
-
-    ## Measure the projection
-    return(projections(matrix, point1 = point1, point2 = point2, measure = measure, scaled = scaled)[2])
-}
-
-

### 2021-11-01 beer/R/sauron.plot.R

@@ -1,252 +1,0 @@
-#' @title sauron.plot
-#'
-#' @description plot a MCMCglmm.disparity object
-#'
-#' @param data an MCMCglmm dispRity object
-#' @param n only selecting a random subset of the posteriors (if left missing, all the posteriors are used)
-#' @param points logical, whether to plot the points (\code{TRUE}; default) or not (\code{FALSE}).
-#' @param major.axes can be \code{NULL} or \code{FALSE} for no axes, \code{"all"} or \code{TRUE} for displaying the \code{n} posteriors or a \code{function} for summarising the major axes (e.g. \code{major.axes = mean}.
-#' @param ellipses can be \code{NULL} or \code{FALSE} for no ellipses, \code{"all"} or \code{TRUE} for displaying the \code{n} posteriors or a \code{function} for summarising the ellipses (e.g. \code{ellipses = mean}.
-#' @param level the confidence interval level of the major axes and ellipses (default is \code{0.95});
-#' @param dimensions which dimensions (default is 1:2), could be 3D.
-#' @param centres optional, a way to determine ellipses or major axes centres. Can be either a function (default is \code{colMeans}), a vector or a list of coordinates vectors or "intercept". See details.
-#' @param transparent.scale if multiple major axes/ellipses are plotted, a scaling factor for the transparency. If left empty, the transparency is set to \code{1/n} or \code{0.1} (whichever is higher).
-#' @param legend whether to add the automatic legend (\code{TRUE}) or not (\code{FALSE}; default).
-#' @param legend.pos the coordinates of where to add the legend (if \code{legend = TRUE}). Can a numeric vector or a character string like \code{"topleft"} (default), \code{"bottomright"}, etc..
-#' @param add logical, whether to add the plot to an existing plot (\code{TRUE}) or not (\code{FALSE}; default).
-#' @param ... graphic parameters
-#' 
-#' @details
-#' The argument \code{centres} allows to determine how to calculate the centre of each ellipses or major axes. The argument can be either:
-#' \itemize{
-#'      \item A function to calculate the centre from a group like the default \code{colMeans} function that calculates the centroid coordinates of each group;
[…]

### 2021-11-01 beer/R/sauron.plot_fun.R

@@ -1,50 +1,0 @@
-## Internal: recentring the covar matrices (changing their Sol)
-recentre <- function(one_group, one_centre, dimensions) {
-    recentre.Sol <- function(covar, centre, dim) {
-        covar$Sol[dim] <- centre[dim]
-        return(covar)
-    }
-    return(lapply(one_group, recentre.Sol, centre = one_centre, dim = dimensions))
-}
-
-## Internal: making one ellipse
-make.ellipse <- function(one_sample, dimensions, npoints){
-    return(ellipse::ellipse(x       = one_sample$VCV[dimensions, dimensions],
-                            centre  = one_sample$Sol[dimensions],
-                            npoints = npoints))
-}
-
-## Internal: making a list of ellipses for the level
-level.ellipses <- function(level_sample, dimensions, npoints, centre) {
-
-    ## Recentreing the levels
-    level_sample <- recentre.levels(level_sample, centre, dimensions)
-
-    ## Get the ellipses for the level
-    return(lapply(level_sample, make.ellipse, dimensions, npoints))
-}
-
-## Internal: changing the intercept ($Sol)
-replace.intercept <- function(level_sample, value, dimensions) {
-    lapply(level_sample, function(X) {X$Sol[dimensions] <- value[dimensions]; return(X)})
-}
-
-
-## Internal: changing the intercept wrapper
-recentre.levels <- function(level_sample, centre, dimensions) {
-    ## Centre the ellipse
-    if(is(centre, "function")) {
-        ## Get the central tendency (as a function)
-        centre_values <- apply(do.call(rbind, lapply(level_sample, `[[`, "Sol")), 2, centre)
-        ## Recentre the intercepts
-        level_sample <- replace.intercept(level_sample, value = centre_values, dimensions)
-    }
-    if(is(centre, "numeric") || is(centre, "integer")) {
-        if((diff <- length(level_sample[[1]]$Sol) - length(centre)) > 0) {
-            centre <- c(centre, rep(centre, diff))
-        }
-        ## Manually recentre the intercepts
-        level_sample <- replace.intercept(level_sample, value = centre, dimensions)
-    }
-    return(level_sample)
-}

### 2021-11-01 beer/tests/testthat.R

@@ -1,4 +1,0 @@
-library(testthat)
-library(beer)
-
-test_check("beer")

### 2021-11-01 beer/tests/testthat/for_export/make.MCMCglmm.covar.data.R

@@ -1,159 +1,0 @@
-
-# ## Loading data and packages
-# load(file = "covar_char_data.rda")
-# load(file = "covar_tree_data.rda")
-# library(MCMCglmm)
-
-# ################################################################################################
-# # Step 3 - MCMCglmm !!!!
-# # There are many ways we can model covariances in MCMCglmm, including how we separate 
-# # phylogenetic and non-phylogenetic effects. Below we will try a range of models, starting simply with
-# # no phylogenetic effects and building up to more complex models.
-# #
-# ################################################################################################
-
-# ntraits <- 3 #this refers to the number of traits in the covariance matrix. Here we are using 3 througout
-
-
-# # The simplest model simply estimates the overal covariance matrix for the whole data set
-# model1 <- MCMCglmm(cbind(PC1, PC2, PC3) ~ trait-1, family = rep("gaussian", ntraits),
-#                       rcov=
-#                        ~ us(trait):units,
-#                       data = covar_char_data, nitt = 110000, burnin = 10000, thin = 100) 
-
-
-# # The next model, still without phylogeny, involves modelling clade effect
-# # Note that we are using clade both in the fixed effects and in the random effects
-# # The random effects relate to covariances and main effects to means
-# model2<- MCMCglmm(cbind(PC1, PC2, PC3) ~ trait:clade-1, family = rep("gaussian", ntraits),
-#                       rcov=
-#                        ~ us(at.level(clade,1):trait):units
-#                       + us(at.level(clade,2):trait):units
-#                       + us(at.level(clade,3):trait):units,
-#                       data = covar_char_data, nitt = 110000, burnin = 10000, thin = 100) 
-
-
-# # Now we will add phylogeny. To begin with we will not include the clade (gull, plover, sandpiper) effect
-# # and instead just partition covariances into phylogenetic and non-phylogenetic effects.
-# # At this point the model is becoming more complex and we need to specific a prior.
-# priorRG1 <- list(R=list(
-#   R1 = list(V = diag(ntraits), nu = 0.002)),
-#   G=list(
-#     G1 = list(V = diag(ntraits), nu = 0.002)))
-
-
-# model3 <- MCMCglmm(cbind(PC1, PC2, PC3) ~ trait-1, family = rep("gaussian", ntraits),
-#                        random =
-#                          ~ us(trait):animal,
[…]

### 2021-11-01 beer/tests/testthat/test-MCMCglmm.dispRity.R

@@ -1,89 +1,0 @@
-## Test
-test_that("MCMCglmm.dispRity and sauron.plot works", {
-    ## Testing the mini chains pipeline
-    load("covar_model_list.Rda")
-    load("covar_char_data.Rda")
-    load("covar_tree_data.Rda")
-
-    ## Sanitizing
-    # data class
-    # posteriors class
-    # group (optional)
-    # tree (optional)
-    # rename.groups (optional)
-
-    ## Model 1: 1 group residual
-    test <- MCMCglmm.dispRity(data = covar_char_data, posteriors =covar_model_list[[1]])
-    expect_is(test, "dispRity")
-    expect_equal(length(test), 5)
-    expect_equal(names(test), c("matrix", "tree", "call", "subsets", "MCMCglmm"))
-    expect_equal(dim(test$matrix[[1]]), c(359, 17))
-    expect_equal(test$call$dimensions, c(1,2,3))
-    expect_equal(test$call$subsets, "MCMCglmm")
-    expect_null(test$tree[[1]])
-    expect_equal(names(test$subsets), "units")
-    expect_equal(names(test$subsets[[1]]), "elements")
-    expect_equal(names(test$MCMCglmm), c("formula", "covars"))
-    expect_equal(names(test$MCMCglmm$formula), c("Fixed", "Random", "Residual"))
-    expect_equal(paste0(as.character(test$MCMCglmm$formula$Fixed), collapse = ""), "~cbind(PC1, PC2, PC3)trait - 1")
-    expect_null(test$MCMCglmm$formula$Random)
-    expect_equal(paste0(as.character(test$MCMCglmm$formula$Residual), collapse = ""), "~us(trait):units")
-    expect_equal(length(test$MCMCglmm$covars), 1)
-    expect_equal(length(test$MCMCglmm$covars[[1]]), 1000)
-    expect_equal(names(test$MCMCglmm$covars[[1]][[1]]), c("VCV", "Sol"))
-
-    ## Default plot
-    expect_null(sauron.plot(test))
-    expect_null(sauron.plot(test, n = 3, ellipses = TRUE, main = "something", level = 0.5, legend = TRUE, legend.pos = c(0.5, -0.5), pch = 12))
-
-    ## Model 2: 3 group residual
-    test <- MCMCglmm.dispRity(data = covar_char_data, posteriors = covar_model_list[[2]])
-    expect_equal(length(test$MCMCglmm$covars), 3)
[…]

### 2021-11-01 beer/tests/testthat/test-MCMCglmm.utilities.R

@@ -1,84 +1,0 @@
-## Test
-require(MCMCglmm)
-data(model_list)
-set.seed(1)
-data(PlodiaPO)
-model_simple<-MCMCglmm(PO~1, random=~FSfamily, data=PlodiaPO, verbose=FALSE,
-nitt=1300, burnin=300, thin=1)
-
-test_that("MCMCglmm.traits works", {
-    ## Simple example
-    expect_equal(MCMCglmm.traits(model_simple), "PO")
-
-    ## Test data example
-    for(i in 1:7) {
-        expect_equal(MCMCglmm.traits(model_list[[i]]), c("PC1", "PC2", "PC3"))
-    }
-})
-
-test_that("MCMCglmm.levels works", {
-
-    ## Simple example
-    expect_equal(MCMCglmm.levels(model_simple), c("random" = "FSfamily", "residual" = "units"))
-    expect_equal(MCMCglmm.levels(model_simple, convert = FALSE), c("random" = "FSfamily", "residual" = "units"))
-
-    ## Test data example
-    expect_equal(MCMCglmm.levels(model_list[[1]]), c("residual" = "units"))
-    expect_equal(MCMCglmm.levels(model_list[[2]]), c("residual" = "units:clade_1", "residual" = "units:clade_2", "residual" = "units:clade_3"))
-    expect_equal(MCMCglmm.levels(model_list[[3]]), c("random" = "animal", "residual" = "units"))
-    expect_equal(MCMCglmm.levels(model_list[[3]], convert = FALSE), c("random" = "us(trait):animal", "residual" = "us(trait):units"))
-    expect_equal(MCMCglmm.levels(model_list[[4]]), c("random" = "animal", "residual" = "units:clade_1", "residual" = "units:clade_2", "residual" = "units:clade_3"))
-    expect_equal(MCMCglmm.levels(model_list[[5]]), c("random" = "animal:clade_1", "random" = "animal:clade_2", "random" = "animal:clade_3", "residual" = "units:clade_1", "residual" = "units:clade_2", "residual" = "units:clade_3"))
-    expect_equal(MCMCglmm.levels(model_list[[6]]), c("random" = "animal:clade_1", "random" = "animal:clade_2", "random" = "animal:clade_3", "random" = "animal", "residual" = "units:clade_1", "residual" = "units:clade_2", "residual" = "units:clade_3"))
[…]

### 2021-11-01 beer/tests/testthat/test-covar.projections.wrapper.R

@@ -1,84 +1,0 @@
-## Test
-test_that("covar.projections.wrapper works", {
-    ## Loading the MCMCglmm
-    load("../../../Data/processed/combined_results.rda")
-    ## Loading the tree and data
-    data(demo_data)
-    data(demo_tree)
-
-    data <- MCMCglmm.dispRity(
-                      data          = demo_data,
-                      posteriors    = combined_results,
-                      group         = MCMCglmm.levels(combined_results)[1:4],
-                      rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
-    
-    ## Test between no base
-    test <- covar.projections.wrapper(data, type = "between", n = 7, verbose = FALSE)
-    expect_equal(names(test), c("position", "distance", "degree"))
-    expect_equal(names(test[[1]]), c("gulls:plovers", "gulls:sandpipers",  "gulls:phylogeny", "plovers:sandpipers",  "plovers:phylogeny", "sandpipers:phylogeny"))
-    expect_equal(dim(test[[1]][[1]]), c(1,7))
-    
-    ## Test between average phylo base
-    test <- covar.projections.wrapper(data, type = "between", sample = mean, base = "phylogeny")
-    expect_equal(names(test), c("position", "distance", "degree"))
-    expect_equal(names(test[[1]]), c("gulls:phylogeny", "plovers:phylogeny", "sandpipers:phylogeny"))
-    expect_equal(dim(test[[1]][[1]]), c(1,1))
-
-    ## Test between all phylo bases
-    test <- covar.projections.wrapper(data, type = "between", sample = c(1,2,3,4,5), base = "phylogeny")
-    expect_equal(names(test), c("position", "distance", "degree"))
-    expect_equal(names(test[[1]]), c("gulls:phylogeny", "plovers:phylogeny", "sandpipers:phylogeny"))
-    expect_equal(dim(test[[1]][[1]]), c(1,5))
-
-    ## Test between average phylo base dispRity style
-    test <- covar.projections.wrapper(data, type = "between", n = 1, base = "phylogeny", dispRity.out = TRUE)
-    expect_equal(names(test), c("position", "distance", "degree"))
-    expect_equal(unique(unlist(lapply(test, class))), "dispRity")
-    
-    ## Test within no base
-    test <- covar.projections.wrapper(data, type = "within", n = 5, measure = c("degree", "distance"))
-    expect_equal(names(test), c("degree", "distance"))
[…]

### 2021-11-01 beer/tests/testthat/test-covar.utilities.R

@@ -1,93 +1,0 @@
-## Test
-require(MCMCglmm)
-set.seed(1)
-data(PlodiaPO)
-model_simple<-MCMCglmm(PO~1, random=~FSfamily, data=PlodiaPO, verbose=FALSE,
-nitt=1300, burnin=300, thin=1)
-
-test_that("get.covar works", {
-    ## Works on a dispRity object
-    data_test <- MCMCglmm.dispRity(data = PlodiaPO, posteriors = model_simple)
-
-    ## works on a MCMCglmm object (basically calls MCMCglmm.covars)
-    test <- get.covar(data_test)
-    expect_equal(length(test), length(MCMCglmm.levels(model_simple)))
-    expect_equal(length(test[[1]]), 1000)
-    ## n argument works
-    test <- get.covar(data_test, n = 7)
-    expect_equal(length(test), length(MCMCglmm.levels(model_simple)))
-    expect_equal(length(test[[1]]), 7)
-    ## sample argument works
-    test <- get.covar(data_test, sample = c(42, 5))
-    expect_equal(length(test), length(MCMCglmm.levels(model_simple)))
-    expect_equal(length(test[[1]]), 2)
-    ## sample argument works as a function
-    test <- get.covar(data_test, sample = median)
-    expect_equal(length(test), length(MCMCglmm.levels(model_simple)))
-    expect_equal(length(test[[1]]), 1)
-    ## Only takes the n argument
-    warn <- capture_warnings(test <- get.covar(data_test, sample = 42, n = 7))
-    expect_equal(warn[[1]], "sample argument is ignored since n = 7 random samples are asked for.")
-    expect_equal(length(test), length(get.covar(data_test)))
-    expect_equal(length(test[[1]]), 7)
-})
-
-
-load("covar_char_data.Rda")
-load("covar_model_list.Rda")
-
-## Test
-test_that("get.one.axis works", {
-    set.seed(1)
-    beer <- MCMCglmm.covars(covar_model_list[[7]], n = 1)
-    ## Right output
-    expect_is(get.one.axis(beer[[1]][[1]]), "matrix")
-    expect_equal(dim(get.one.axis(beer[[1]][[1]])), c(2,3))
-    ## Selects the right axes
-    test1 <- get.one.axis(beer[[1]][[1]], axis = 1)
-    test2 <- get.one.axis(beer[[1]][[1]], axis = 2)
[…]

### 2021-11-01 beer/tests/testthat/test-dispRity.fast.R

@@ -1,6 +1,0 @@
-## Test
-
-test_that("fast.dispRity works", {
-    ## Works on a dispRity object
-    
-})

### 2021-11-01 beer/tests/testthat/test-optim.replicate.R

@@ -1,94 +1,0 @@
-## Test
-test_that("verbose function works", {
-
-    test <- function() mean(c(1,2,3))
-    message <- capture_message(res <- test())
-    expect_equal(res, 2)
-    expect_null(message)
-    test2 <- eval.verbose(test)
-    message <- capture_message(res <- test2())
-    expect_equal(res, 2)
-    expect_equal(message[[1]], ".")
-    test3 <- eval.verbose(test, msg = "what?")
-    message <- capture_message(res <- test3())
-    expect_equal(res, 2)
-    expect_equal(message[[1]], "what?")
-
-    set.seed(1)
-    test <- function() rnorm(1)
-    message <- capture_message(res <- test())
-    expect_equal(round(res, 2), -0.63)
-    expect_null(message)
-    test2 <- eval.verbose.count(test, count = 0)
-    message <- capture_message(res <- test2())
-    expect_equal(message[[1]], "1.")
-    message <- capture_message(res <- test2())
-    expect_equal(message[[1]], "2.")
-    test3 <- eval.verbose.count(test, count = 41)
-    message <- capture_message(res <- test3())
-    expect_equal(message[[1]], "42.")
-})
-
-test_that("optim.replicate works", {
-    set.seed(1)
-    fun1 <- function() return(sample(1:100, 1))
-    fun2 <- function() return(c("a" = sample(1:100, 1), "b" = sample(1:100, 1)))
-    diagnose.fun <- function(x) var(x)
-
-    ## Verbose test
-    messages <- capture_messages(test <- optim.replicate(fun2, diagnose.fun, minimum = 100, verbose = TRUE, stop.variance = 0.001, increment = .05))
-    expect_equal(length(messages), 1518)
-    expect_equal(messages[[1]], "Running the initial 100 replicates:")
-    expect_equal(messages[[2]], "1.")
-    expect_equal(messages[[102]], "Done.")
-    expect_equal(messages[[103]], "\nRunning an additional 5 replicates:")
-    expect_equal(messages[[104]], "101.")
-    expect_equal(messages[[110]], "\nDiagnosis change:0.011, -0.022")
-    expect_equal(messages[[111]], "\nRunning an additional 6 replicates:")
-    expect_equal(messages[[1518]], "\nResults converged after 1359 iterations: additional variances (-0.001, 0) all < 0.001.")
[…]

### 2021-11-01 beer/tests/visualise.projections.covar.R

@@ -1,188 +1,0 @@
-plot.base <- function(matrix, base_vector, lim) {
-
-    if(missing(lim)) {
-        lim <- ceiling(max(abs(range(matrix))))
-        lim <- c(-lim, lim)
-    }
-
-    if(ncol(matrix) == 3) {
-        plot3d(NULL, xlim = lim, ylim =lim, zlim = lim, xlab = "", ylab = "", zlab = "")
-        segments3d(x = lim, y = c(0, 0), z = c(0, 0), col = "grey")
-        text3d(x = min(lim)*0.9, y = 0, z = 0, texts = "X", col = "grey")
-        segments3d(x = c(0,0), y = lim, z = c(0, 0), col = "grey")
-        text3d(x = 0, y = min(lim)*0.9, z = 0, texts = "Y", col = "grey")
-        segments3d(x = c(0,0), y = c(0,0), z = lim, col = "grey")
-        text3d(x = 0, y = 0, z = min(lim)*0.9, texts = "Z", col = "grey")
-        ## Add the original points
-        text3d(matrix, texts = rownames(matrix), col = "grey")
-        segments3d(base_vector, col = "grey")
-    } else {
-        par(bty = "n")
-        plot(NULL, xlim = lim, ylim = lim, xlab = "x", ylab = "y")
-        abline(v = 0, col = "grey")
-        abline(h = 0, col = "grey")
-        text(matrix[,1:2], labels = rownames(matrix), col = "grey", cex = 1.5)
-        lines(base_vector, col = "grey", lwd = 1.5)
-    }
-}
-plot.recentred <- function(matrix, base_vector, col = c("black", "grey")) {
-    if(ncol(matrix) == 3) {
-        text3d(matrix, texts = rownames(matrix), col = col[1])
-        for(i in 1:nrow(matrix)) {
-            segments3d(rbind(c(0,0,0), matrix[i,]), col = col[2])
-        }
-        segments3d(base_vector, col = col[1], lwd = 2)
-    } else {
-        text(matrix, labels = rownames(matrix), col = col[1], cex = 2)
-        for(i in 1:nrow(matrix)) {
-            lines(rbind(c(0,0), matrix[i,]), col = col[2], lwd = 2)
-        }
-        lines(base_vector, col = col[1], lwd = 3)
-    }
-}
-plot.projections <- function(matrix, projections, rejections, col = c("black", "grey")) {
-    if(ncol(matrix) == 3) {
-        for(i in 1:nrow(matrix)) {
-            ## Plot the projections
[…]

### 2021-11-01 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Update 02.1-shapespace_allbirds_lvl_superorder_order.Rmd

@@
 class(model_list) <- c("beer", "mini.chains")
-models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster_out/"
+models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster/"
 ## Loading all models

### 2021-11-01 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
removed unused data

@@
 models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster/"
-## Loading all models
-for(file in list.files(models_path)) {
-    load(paste0(models_path, file, collapse = "/"))
+## Getting all the model files
+model_files <- list.files(models_path, pattern = ".rda")
+## Get the first model
+load(paste0(models_path, model_files[1], collapse = "/"))
+print(model_files[1])
+## Clean the bits that are not going to be used (but are huge)
+model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+## Update the model list
 model_list[[length(model_list)+1]] <- model[[1]]
+model <- NULL
+model_files <- model_files[-1]
+
+## Get the subsequent models and merge them as we go
+while(length(model_files) != 0) {
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
+    model_list[[length(model_list)+1]] <- model[[1]]
+    ## Merge the models (overwrite the model list)
+    model_list <- list(combine.mini.chains(model_list))
+    ## DEBUG
+    print("model list size:")
+    print(format(object.size(model_list), units = "MB"))
+    model <- NULL
+    model_files <- model_files[-1]
 }
@@
 ## Combining the model files
-shapespace_allbirds_lvl_superorder_order <- combine.mini.chains(model_list)
+shapespace_allbirds_lvl_superorder_order <- model_list[[1]]
 
@@
 ## Saving the results
-save(shapespace_allbirds_lvl_superorder_order, file = "../Data/processed/shapespace_allbirds_lvl_superorder_order.rda")
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
 

### 2021-11-01 Analysis/03.1-results.Rmd
updated to dispRity syntax

@@
 ```{r, eval = TRUE, echo = FALSE}
+## Load the data
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
+## Select the data
+shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
 ## Load the covar
-load(file = "../Data/Processed/allbirds_8748t_8Dshapespace_31levels.rda")
-## Load the data
-load(file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
-shapespace <- form_shape_space_1tree_2levels$shape_space
-## Loading the group names
-load("../Data/Processed/group_names_level_1_2.rda")
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
 
@@
       data          = shapespace,
-      posteriors    = allbirds_8748t_8Dshapespace_31levels,
-      group         = MCMCglmm.levels(allbirds_8748t_8Dshapespace_31levels)[c(1:8, 32)],
-      rename.groups = c(names(group_names_level_1_2), "phylogeny")[c(1:8, 32)])
+      posteriors    = shapespace_allbirds_lvl_superorder_order,
+      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(1:8, 32)],
+      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
 
@@
       data          = shapespace,
-      posteriors    = allbirds_8748t_8Dshapespace_31levels,
-      group         = MCMCglmm.levels(allbirds_8748t_8Dshapespace_31levels)[-c(1:8, 33)],
-      rename.groups = c(names(group_names_level_1_2), "phylogeny")[-c(1:8,33)])
+      posteriors    = shapespace_allbirds_lvl_superorder_order,
+      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[-c(1:8, 33)],
+      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
 ```
@@
 ## Plotting just the space
-sauron.plot(superorder_shapespace,
+covar.plot(superorder_shapespace,
            main = "Shapespace",
@@
 ## Plotting mean axes
-sauron.plot(superorder_shapespace,
+covar.plot(superorder_shapespace,
            col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
@@
 ## Plotting mean ellipses
-sauron.plot(superorder_shapespace,
+covar.plot(superorder_shapespace,
            col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
@@
 ## Plotting mean axes
-sauron.plot(order_shapespace,
+covar.plot(order_shapespace,
            col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
@@
 ## Plotting mean ellipses
-sauron.plot(order_shapespace,
+covar.plot(order_shapespace,
            col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
@@
 ## Run the analyses between groups
-super_results_between <- covar.projections.wrapper(superorder_shapespace,
-                                             type = "between",
+super_results_between <- dispRity.covar.projections(superorder_shapespace,
+                                             type = "groups",
+                                             base = "phylogeny")
+## Run the analyses between groups
+order_results_between <- dispRity.covar.projections(order_shapespace,
+                                             type = "groups",
+                                             base = "phylogeny")
+## Run the analyses of groups on the phylogeny
+super_results_phylo <- dispRity.covar.projections(superorder_shapespace,
+                                            type = "elements",
                                             base = "phylogeny",
-                                             dispRity.out = TRUE)
-## Run the analyses between groups
-order_results_between <- covar.projections.wrapper(order_shapespace,
-                                             type = "between",
-                                             base = "phylogeny",
-                                             dispRity.out = TRUE)
-## Run the analyses of groups on the phylogeny
-super_results_phylo <- covar.projections.wrapper(superorder_shapespace,
-                                            type = "within",
-                                            base = "phylogeny",
-                                            incl.base = TRUE,
-                                            dispRity.out = TRUE,
+                                            inc.base = TRUE,
                                             verbose = TRUE)
 ## Run the analyses of groups on the phylogeny
-order_results_phylo <- covar.projections.wrapper(order_shapespace,
-                                            type = "within",
+order_results_phylo <- dispRity.covar.projections(order_shapespace,
+                                            type = "elements",
                                             base = "phylogeny",
-                                            incl.base = TRUE,
-                                            dispRity.out = TRUE,
+                                            inc.base = TRUE,
                                             verbose = TRUE)
 ## Run the analyses groups on themselves
-super_results_within <- covar.projections.wrapper(superorder_shapespace,
-                                            type = "within",
-                                            dispRity.out = TRUE,
+super_results_within <- dispRity.covar.projections(superorder_shapespace,
+                                            type = "elements",
                                             verbose = TRUE)
 ## Run the analyses groups on themselves
-order_results_within <- covar.projections.wrapper(order_shapespace,
-                                            type = "within",
-                                            dispRity.out = TRUE,
+order_results_within <- dispRity.covar.projections(order_shapespace,
+                                            type = "elements",
                                             verbose = TRUE)
 ## Save all
-allbirds_8748t_8Dshapespace_31levels_results_list <- list(
+shapespace_allbirds_lvl_superorder_order_results_list <- list(
         "super_results_between" = super_results_between,
@@
         "order_results_within"  = order_results_within)
-save(allbirds_8748t_8Dshapespace_31levels_results_list, file = "../Data/Processed/allbirds_8748t_8Dshapespace_31levels_results_list.rda")
+save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
 ```
@@
 ## Quick load
-load(file = "../Data/Processed/allbirds_8748t_8Dshapespace_31levels_results_list.rda")
-results <- allbirds_8748t_8Dshapespace_31levels_results_list
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+results <- shapespace_allbirds_lvl_superorder_order_results_list
 ```

### 2021-11-02 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
+## Get the levels
+phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
+super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
+order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
+
+## Creating the subsets
 superorder_shapespace <- MCMCglmm.subsets(
@@
       posteriors    = shapespace_allbirds_lvl_superorder_order,
-      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(1:8, 32)],
+      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, phylo_lvl)],
       rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
@@
       posteriors    = shapespace_allbirds_lvl_superorder_order,
-      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[-c(1:8, 33)],
+      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(order_lvl, phylo_lvl)],
       rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
@@
 
-```{r, echo = FALSE, fig.height = 18, fig.width = 18, eval = FALSE}
+```{r, echo = FALSE, fig.height = 18, fig.width = 18, eval = TRUE}
 par(mfrow = c(2,2))

### 2021-11-04 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 ## Plot the results
-wrap.plot <- function(results, col, main) {
+wrap.plot <- function(results, col, main, names) {
+
+    ## Adding one plot
+    one.plot <- function(one_result, xlab, col, zero = TRUE, main, ...) {
+        values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
+        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main, ...)
+        if(zero) {
+            abline(v = 0, col = "black", lty = 3, lwd = 1)
+        }
+    }
+
     ## How many sub plots?
     n <- length(results)
-    ## Edit names for cleaning
-    # for(i in 1:n) {
-    #     names(results[[i]]$disparity) <- gsub(":phylogeny", "", names(results[[i]]$disparity))
-    # }
+
+    ## Recycle labels?
+    if(missing(names)) {
+        names_list <-names(results)
+    } else {
+        names_list <- names
+    }
 
     ## Which will be the centre plot?
-    centre <- 1
+    centre <- ceiling(n/2)
     ## Plot everything
-    op <- par(mfrow = c(n, 1), mar = c(7, 4, 4, 2) + 0.1)
+    op <- par(mfrow = c(1, n+1))
+    ## Plot the first panel
+    par(bty = "n", mar = c(5, 5, 4, 0) + 0.1)
+    empty <- matrix(0, ncol = length(results[[1]]$disparity))
+    colnames(empty) <- names(results[[1]]$disparity)
+    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
     for(i in 1:n) {
-        plot(results[[i]], col = col, main = ifelse(i == centre, main, ""), las = 2, xlab = "")
+        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
+        one.plot(results[[i]], col = col, main = ifelse(i == centre, main, ""), xlab = names_list, zero = (names(results)[[i]] == "position"))
     }
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "Superorders differences")
+          main = "Superorders differences",
+          names = c("elaboration", "innovation", "angle"))
 ```
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "Superorders differences")
+          main = "Superorders differences",
+          names = c("elaboration", "innovation", "angle"))
 ```
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "Projections per superorder\non the global phylogeny axes")
+          main = "Projections per superorder\non the global phylogeny axes",
+          names = c("elaboration", "innovation", "angle"))
 ```
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "Projections per order\non the global phylogeny axes")
+          main = "Projections per order\non the global phylogeny axes",
+          names = c("elaboration", "innovation", "angle"))
 ```
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "Projections per superorder\nonto each superorder")
+          main = "Projections per superorder\nonto each superorder",
+          names = c("elaboration", "innovation", "angle"))
 ```
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "Projections per order\nonto each order")
+          main = "Projections per order\nonto each order",
+          names = c("elaboration", "innovation", "angle"))
 ```

### 2021-11-04 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

--- 2021-11-02 Analysis/03.1-results.Rmd
@@
 ---
-title: "Bird beak's elaboration and exploration"
+title: "Bird beak's elaboration and innovation"
 author: "Thomas Guillerme"
@@
 
-This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_exploration_analyses.Rmd.
-
-# Elaboration and exploration analyses
+This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_innovation_analyses.Rmd.
 
@@
 
-```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, echo = FALSE}
+First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
 ## Plotting just the space
@@
            cex = 0.5,
-           legend = FALSE)
+           legend = TRUE,
+           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
+           legend.args = list(x = "bottomright", cex = 2/3))
 ```
-The global Shape space
 
-```{r, echo = FALSE, fig.height = 18, fig.width = 18, eval = TRUE}
+Figure: The global shape space
+
+We can then check the distribution of the variance covariance matrices for each super order and each order.
+The VCV matrices are represented as average ellipses or average axis.
+
+```{r, echo = FALSE, fig.height = 14, fig.width = 14, eval = TRUE}
 par(mfrow = c(2,2))
@@
 ```
-Main axes and ellipses visualisation
 
-## Calculate the elaborations/explorations
+Main axes and ellipses visualisation for each superorders and orders.
+
+### Are these 2D average ellipses and/or axes an accurate representation of what's going on? 
+<details>
+  <summary>[Click to expand] TL;DR: It depends...
+</summary>
+
+A couple of things have to be taken into account:
+
+ * First, the averaging of the VCV (and it's ellipse and/or maon axis) can result into weird "middle" VCV that are not representing the individual VCV (e.g.
+if half VCV are negative, and half positive, the average will be probably 0; which does not exist).
+ * Second, the ellipses in nD projected onto a 2D space will be deformed (unless the VCV is a real pancake on the first 2D, i.e. with no variance on the other dimensions).
+
+Note that both problems are only visual though since: 1) we don't use averages anywhere (but the full distributions all the time) and 2) the linear algebra bit with the projections and stuff works in any number of dimensions and transforms the results into 1D results (i.e.
+the projection/rejection of groups/species on the major axes in 8D space).
+
+Here's a visualisation of how the variance in VCVs affects the averaging artefact:
+
+```{r, eval = TRUE, echo = FALSE}
+visualisation.group <- function(data, subset, dimensions, n = 100) {
+
+    ## Get the subset
+    sub <- get.subsets(data, subsets = subset)
+
+    ## Make 2D
+    if(!missing(dimensions)) {
+        sub$matrix[[1]] <- sub$matrix[[1]][,dimensions]
+        sub$call$dimensions <- dimensions
+        fun2d <- function(covar) {
+            covar$VCV <- covar$VCV[dimensions, dimensions]
+            covar$loc <- covar$loc[dimensions]
+            return(covar)
+        }
+        sub$covar <- lapply(sub$covar, lapply, fun2d)
+        main_D <- "2D"
+    } else {
+        main_D <- "8D"
+    }
+
+    ## Make the tile
+    main <- paste0(n, " random ", subset, " VCV (", main_D, ")")
+
+    ## Plot
+    covar.plot(sub, major.axes = TRUE, ellipses = TRUE, points = FALSE, legend = FALSE, main = main)
+    covar.plot(sub, major.axes = mean, ellipses = mean, points = FALSE, col = "red", add = TRUE)
+}
+
+visualisation.group(superorder_shapespace, "Mirandornithes")
+```
+
+```{r, echo = FALSE, eval = TRUE}
+visualisation.group(superorder_shapespace, "Strisores")
+```
+
+Or even more obvious (with an direction that does not really exist (due to averaging)):
+
+```{r, echo = FALSE, eval = TRUE}
+visualisation.group(superorder_shapespace, "Paleognathae")
+```
+
+For the 2Dness, the effect is a bit less obvious (because the two first dimensions already have so much variance).
+Here is an example for the Columbimorphae.
+(to check different groups, you can easily modify the source code of the Rmd.
+TODO: this would be good in some shiny app innovation thingy).
+
+```{r, echo = FALSE, eval = TRUE, fig.height = 6, fig.width = 12}
+par(mfrow = c(1,2))
+visualisation.group(superorder_shapespace, "Columbimorphae")
+visualisation.group(superorder_shapespace, "Columbimorphae", dimensions = c(1,2))
+```
+
+Back to the main story.
+</details>
+
+## Calculate the elaborations/innovations
+
+We then calculate the projections/rejections as explained in more details in the [03-elaboration_exploration_analyses.Rmd]() script.
+Really briefly we calculate three types of projections/rejections abbreviated as follows below:
+
+ * `between`: the projection of each group's VCV (major axis) on the phylogenetic's VCV (major axis)
+ * `phylo`: the projection of each element within a group (species) onto the phylogenetic's VCV (major axis)
+ * `within`: the projection of each element within a group (species) onto its own VCV (major axis)
 
@@
 
-```{r, echo = FALSE}
+### Projections between groups:
+
+```{r, echo = FALSE, fig.height = 8, fig.width = 8}
 ## Plot the results
-wrap.plot <- function(results, col, main) {
+wrap.plot <- function(results, col, main, names, centre.position = TRUE, absolute = FALSE) {
+
+    ## Adding one plot
+    one.plot <- function(one_result, xlab, col, zero = TRUE, main, centre, absolute, ...) {
+        values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
+        if(centre) {
+            values <- lapply(values, function(X) {X-0.5})
+        }
+        if(absolute) {
+            values <- lapply(values, function(X) {abs(X-0.5)})
+        }
+        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main = main, ...)
+        if(zero) {
+            abline(v = 0, col = "black", lty = 3, lwd = 1)
+        }
+    }
+
     ## How many sub plots?
     n <- length(results)
-    ## Edit names for cleaning
-    # for(i in 1:n) {
-    #     names(results[[i]]$disparity) <- gsub(":phylogeny", "", names(results[[i]]$disparity))
-    # }
+
+    ## Recycle labels?
+    if(missing(names)) {
+        names_list <-names(results)
+    } else {
+        names_list <- names
+    }
 
     ## Which will be the centre plot?
-    centre <- 1
+    centre <- ceiling(n/2)
     ## Plot everything
-    op <- par(mfrow = c(n, 1), mar = c(7, 4, 4, 2) + 0.1)
+    op <- par(mfrow = c(1, n+1))
+    ## Plot the first panel
+    par(bty = "n", mar = c(5, 14, 4, 0) + 0.1)
+    empty <- matrix(0, ncol = length(results[[1]]$disparity))
+    colnames(empty) <- names(results[[1]]$disparity)
+    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
     for(i in 1:n) {
-        plot(results[[i]], col = col, main = ifelse(i == centre, main, ""), las = 2, xlab = "")
+        ## Is position?
+        is_position <- names(results)[[i]] == "position"
+        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
+        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = is_position, absolute = absolute)
     }
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "Superorders differences")
+          main = "[between] superorders VCV\nonto the phylogenetic VCV",
+          names = c("elaboration", "innovation", "angle"))
 ```
+
 Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes.
@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE}
+<details>
+  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+</summary>
+
+```{r, echo = FALSE, fig.height = 8, fig.width = 8}
+wrap.plot(results$super_results_between,
+          col = rainbow(n.subsets(superorder_shapespace)-1),
+          main = "[between] superorders VCV\nonto the phylogenetic VCV",
+          names = c("absolute elaboration", "innovation", "angle"),
+          centre.position = FALSE, absolute = TRUE)
+```
+</details>
+
+```{r, fig.height = 16, fig.width = 8, echo = FALSE}
 wrap.plot(results$order_results_between,
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "Superorders differences")
+          main = "[between] orders VCV\nonto the phylogenetic VCV",
+          names = c("elaboration", "innovation", "angle"))
 ```
 
-## Elaboration and exploration by species
+<details>
+  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+</summary>
+
+```{r, echo = FALSE, fig.height = 16, fig.width = 8}
+wrap.plot(results$order_results_between,
+          col = rainbow(n.subsets(order_shapespace)-1),
+          main = "[between] orders VCV\nonto the phylogenetic VCV",
+          names = c("absolute elaboration", "innovation", "angle"),
+          centre.position = FALSE, absolute = TRUE)
+```
+</details>
+
+### Projections of the species on the phylogenetic VCV
 
@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}
+```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
 ## Plot the results
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "Projections per superorder\non the global phylogeny axes")
+          main = "[phylo] superorders species\nonto the phylogenetic VCV",
+          names = c("elaboration", "innovation", "angle"))
 ```
+
+<details>
+  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+</summary>
+
+```{r, echo = FALSE, fig.height = 8, fig.width = 8}
+wrap.plot(results$super_results_phylo,
+          col = rainbow(n.subsets(superorder_shapespace)-1),
+          main = "[phylo] superorders species\nonto the phylogenetic VCV",
+          names = c("absolute elaboration", "innovation", "angle"),
+          centre.position = FALSE, absolute = TRUE)
+```
+</details>
+
 Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE}
+```{r, fig.height = 16, fig.width = 8, echo = FALSE}
 
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "Projections per order\non the global phylogeny axes")
+          main = "[phylo] orders species\nonto the phylogenetic VCV",
+          names = c("elaboration", "innovation", "angle"))
 ```
+
+<details>
+  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+</summary>
+
+```{r, echo = FALSE, fig.height = 16, fig.width = 8}
+wrap.plot(results$order_results_phylo,
+          col = rainbow(n.subsets(order_shapespace)-1),
+          main = "[phylo] orders species\nonto the phylogenetic VCV",
+          names = c("absolute elaboration", "innovation", "angle"),
+          centre.position = FALSE, absolute = TRUE)
+```
+</details>
+
+### Projections of the species on their own group's VCV
 
@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE}
+```{r, fig.height = 8, fig.width = 8, echo = FALSE}
 
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "Projections per superorder\nonto each superorder")
+          main = "[within] superorders species\nonto their group's VCV",
+          names = c("elaboration", "innovation", "angle"))
 ```
+
+<details>
+  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+</summary>
+
+```{r, echo = FALSE, fig.height = 8, fig.width = 8}
+wrap.plot(results$super_results_within,
+          col = rainbow(n.subsets(superorder_shapespace)-1),
+          main = "[within] superorders species\nonto their group's VCV",
+          names = c("absolute elaboration", "innovation", "angle"),
+          centre.position = FALSE, absolute = TRUE)
+```
+</details>
+
 Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE}
+```{r, fig.height = 16, fig.width = 8, echo = FALSE}
 
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "Projections per order\nonto each order")
+          main = "[within] oreders species\nonto their group's VCV",
+          names = c("elaboration", "innovation", "angle"))
 ```
 
-## Different scores distributions per species
+<details>
+  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+</summary>
+
+```{r, echo = FALSE, fig.height = 16, fig.width = 8}
+wrap.plot(results$order_results_within,
+          col = rainbow(n.subsets(order_shapespace)-1),
+          main = "[within] oreders species\nonto their group's VCV",
+          names = c("absolute elaboration", "innovation", "angle"),
+          centre.position = FALSE, absolute = TRUE)
+```
+</details>
+
+So that's the story so far...
+
+# Different scores distributions per species
 
@@
 #@param exclude.phylo remove the phylogeny subset (default = FALSE)
+#@param absolute whether to make the position values absolute or not
 #@param ... plot options
 
-get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, ...) {
+get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, absolute = FALSE, ...) {
 
@@
     if(type == "position") {
+        if(!absolute) {
             disparity <- lapply(disparity, function(X) return(X-0.5))
+        } else {
+            disparity <- lapply(disparity, function(X) return(abs(X-0.5)))
+        }
     }
@@
     res_out$elaboration <- c(res1[, 3], rep(NA, nrow(res2)))
-    res_out$exploration <- c(rep(NA, nrow(res1)), res2[, 3])
+    res_out$innovation <- c(rep(NA, nrow(res1)), res2[, 3])
     ## Find any species in commons
@@
 
+Furthermore, we can look at things species per species to find which species species have the most extreme values (in general, or within their groups).
+
 ### Extreme species per super order (based on phylogeny)
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position")
+super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position", absolute = TRUE)
 ```
@@
 ```
-Figure: explorations distributions per super orders
+Figure: innovations distributions per super orders
+
+<details>
+  <summary>[Click to expand] Table of species extreme values projected onto the phylogenetic VCV.
+</summary>
 
@@
 ```
+
+</details>
 
@@
 ```
-Figure: explorations distributions per order
+Figure: innovations distributions per order
+
+<details>
+  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
+</summary>
 
@@
 ```
+</details>
 
@@
 ```
-Figure: explorations distributions within super orders
+Figure: innovations distributions within super orders
+
+<details>
+  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
+</summary>
 
@@
 ```
-Figure: explorations distributions within orders
+Figure: innovations distributions within orders
+
+<details>
+  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
+</summary>
 
@@
 ```
+</details>
 
-## Correlations between elaboration and exploration?
+## Correlations between elaboration and innovation?
 
 ```{r, echo = FALSE}
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE, ...) {
+get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE, absolute = TRUE, ...) {
 
@@
     ## Recentre the results
+    if(!absolute) {
         disparity_pos <- lapply(disparity_pos, function(X) return(X-0.5))
+    } else {
+        disparity_pos <- lapply(disparity_pos, function(X) return(abs(X-0.5)))
+    }
 
@@
         ## Plotting the results
-        plot(centrals_pos[[i]], centrals_dis[[i]], pch = 19, xlab = "elaboration", ylab = "exploration", main = plot_main, cex = 0.5)
+        plot(centrals_pos[[i]], centrals_dis[[i]], pch = 19, xlab = "elaboration", ylab = "innovation", main = plot_main, cex = 0.5)
     }

### 2021-11-08 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
-```{r, eval = TRUE, echo = FALSE, print = FALSE}
+```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
 library(dispRity)
@@
 
- * `between`: the projection of each group's VCV (major axis) on the phylogenetic's VCV (major axis)
- * `phylo`: the projection of each element within a group (species) onto the phylogenetic's VCV (major axis)
- * `within`: the projection of each element within a group (species) onto its own VCV (major axis)
+ * `group on phylogeny`: the projection of each group's VCV (major axis) on the phylogenetic's VCV (major axis)
+ * `species on phylogeny`: the projection of each element within a group (species) onto the phylogenetic's VCV (major axis)
+ * `species on group`: the projection of each element within a group (species) onto its own VCV (major axis)
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE}
+<!-- TODO: add the pictures of the different tests here -->
+
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
 ## Run the analyses between groups
@@
 
-### Projections between groups:
+### Projections of groups on the phylogeny:
 
-```{r, echo = FALSE, fig.height = 8, fig.width = 8}
+```{r "starting with boxplots", echo = FALSE, fig.height = 8, fig.width = 8}
 ## Plot the results
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[between] superorders VCV\nonto the phylogenetic VCV",
+          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
           names = c("elaboration", "innovation", "angle"))
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[between] superorders VCV\nonto the phylogenetic VCV",
+          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
           names = c("absolute elaboration", "innovation", "angle"),
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[between] orders VCV\nonto the phylogenetic VCV",
+          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
           names = c("elaboration", "innovation", "angle"))
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[between] orders VCV\nonto the phylogenetic VCV",
+          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
           names = c("absolute elaboration", "innovation", "angle"),
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[phylo] superorders species\nonto the phylogenetic VCV",
+          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
           names = c("elaboration", "innovation", "angle"))
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[phylo] superorders species\nonto the phylogenetic VCV",
+          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
           names = c("absolute elaboration", "innovation", "angle"),
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[phylo] orders species\nonto the phylogenetic VCV",
+          main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
           names = c("elaboration", "innovation", "angle"))
@@
 
-```{r, echo = FALSE, fig.height = 16, fig.width = 8}
+```{r "hehehe", echo = FALSE, fig.height = 16, fig.width = 8, eval = FALSE}
 wrap.plot(results$order_results_phylo,
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[phylo] orders species\nonto the phylogenetic VCV",
+          main = "[species on phylogeny] orders\nspecies onto the phylogenetic VCV",
           names = c("absolute elaboration", "innovation", "angle"),
           centre.position = FALSE, absolute = TRUE)
-```
+````
 </details>
@@
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[within] superorders species\nonto their group's VCV",
+          main = "[species on group] superorders species\nonto their group's VCV",
           names = c("elaboration", "innovation", "angle"))
@@
 
-```{r, echo = FALSE, fig.height = 8, fig.width = 8}
+```{r "beforebefore last", echo = FALSE, fig.height = 8, fig.width = 8}
 wrap.plot(results$super_results_within,
           col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[within] superorders species\nonto their group's VCV",
+          main = "[species on group] superorders species\nonto their group's VCV",
           names = c("absolute elaboration", "innovation", "angle"),
@@
 
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
+```{r "before last", fig.height = 16, fig.width = 8, echo = FALSE}
 
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[within] oreders species\nonto their group's VCV",
+          main = "[species on group] oreders species\nonto their group's VCV",
           names = c("elaboration", "innovation", "angle"))
@@
 
-```{r, echo = FALSE, fig.height = 16, fig.width = 8}
+```{r "done with boxplots", echo = FALSE, fig.height = 16, fig.width = 8}
 wrap.plot(results$order_results_within,
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[within] oreders species\nonto their group's VCV",
+          main = "[species on group] oreders species\nonto their group's VCV",
           names = c("absolute elaboration", "innovation", "angle"),
@@
 
-### Extreme species per super order (based on phylogeny)
+### Extreme species per super order (species on phylogeny)
 
@@
 
-### Extreme species per order (based on phylogeny)
+### Extreme species per order (species on phylogeny)
 
@@
 
-### Extreme species per super order (within)
+### Extreme species per super order (species on group)
 
@@
 
-### Extreme species per order (within)
+### Extreme species per order (species on group)
 
@@
 ```{r, echo = FALSE}
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE, absolute = TRUE, ...) {
+## Calculating all the correlations
+get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE, absolute = TRUE) {
 
@@
 
-    ## Calculate and plot the hdrcde
-    op_tmp <- par(mfrow = c(ceiling(sqrt(n_groups)), round(sqrt(n_groups))))
-    for(i in 1:n_groups) {
-        ## Testing the correlation
-        cor_test <- cor.test(centrals_pos[[i]], centrals_dis[[i]])
-        plot_main <- paste0(names(centrals_pos)[i], ": ", unname(round(cor_test$estimate, 3)), " (p=", unname(round(cor_test$p.value, 3)), ")")
+    ## Prepare the results
+    prep.cor.results <- function(pos, dis, name) {
+        ## Run the correlation test
+        cor_test <- cor.test(pos, dis)
+        ## Prepare the plot title
+        plot_main <- paste0(name, ": ", unname(round(cor_test$estimate, 3)), " (p=", unname(round(cor_test$p.value, 3)), ")")
+        ## Return the plotting parameters
+        return(list(x = pos, y = dis, main = plot_main))
+    }
+
+    return(mapply(prep.cor.results, centrals_pos, centrals_dis, as.list(names(centrals_pos)), SIMPLIFY = FALSE))
+}
+
+## Plotting one correlation 
+plot.correlations <- function(cor.results, col.sub, col.fun = rainbow, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ...) {
+    dots <- list(...)
+
+    if(missing(col.sub)) {
+        col.sub <- NULL
+    }
+
+    ## Subbing the colours
+    if(!is.null(col.sub)) {
+        ## Reset the levels
+        col_sub <- as.factor(as.character(col.sub))
+        ## Make the empty level grey
+        if(any(levels(col_sub) == "")) {
+            col_avail <- c("grey", col.fun(length(levels(col_sub))-1))
+        } else {
+            col_avail <- col.fun(length(levels(col_sub)))
+        }
+        ## Create the colour vector
+        col_vector <- col_avail[as.numeric(col_sub)]
+
+    } else {
+        if(is.null(dots$col)) {
+            col_vector <- "black"
+        } else {
+            col_vector <- dots$col
+        }
+    }
+
     ## Plotting the results
-        plot(centrals_pos[[i]], centrals_dis[[i]], pch = 19, xlab = "elaboration", ylab = "innovation", main = plot_main, cex = 0.5)
+    plot(cor.results$x, cor.results$y, pch = 19, xlab = "elaboration", ylab = "innovation", main = cor.results$main, cex = pt.cex, col = col_vector)
+
+    if(!is.null(col.sub)) {
+        ## Adding the legend
+        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
     }
-    par(op_tmp)
     return(invisible())
@@
 
-### Correlations per super order (based on phylo)
+### Correlations per super order (species on phylogeny)
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-get.correlations(results$super_results_phylo)
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+## Getting the subsets for the colouring
+col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
+## Calculate the correlations
+super_phylo_cor <- get.correlations(results$super_results_phylo)
+## Set the vector of legend positions
+legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
+for(i in 1:length(super_phylo_cor)) {
+    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i])
+}
 ```
 
-### Correlations within super orders
+### Correlations within super orders (species on group)
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-get.correlations(results$super_results_within, exclude.phylo = TRUE)
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+## Getting the subsets for the colouring
+col_subs <- split(shapespace$level2, f = shapespace$level1)[-1]
+## Calculate the correlations
+super_within_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
+## Set the vector of legend positions
+legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
+for(i in 1:length(super_within_cor)) {
+    plot.correlations(super_within_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i])
+}
 ```
 
-### Correlations per order (based on phylo)
+### Correlations per order (species on phylogeny)
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-get.correlations(results$order_results_phylo, exclude.phylo = TRUE)
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+## Loading the sub-order levels
+load(file = "../Data/Processed/shapespace_superorder_lvl_order_suborder.rda")
+## Select the data
+shapespaces <- lapply(shapespace_superorder_lvl_order_suborder, function(X) X$space)
+## Create a level 3
+order_space <- do.call(rbind, shapespaces)
+
+## Calculate the correlations
+order_phylo_cor <- get.correlations(results$order_results_phylo, exclude.phylo = TRUE)
+
+## Getting the subsets for the colouring
+split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
+col_subs <- replicate(length(order_phylo_cor), NULL)
+col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
+
+## Set the vector of legend positions
+for(i in 1:length(order_phylo_cor)) {
+    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]])
+}
 ```
 
-### Correlations within orders
+### Correlations within orders (species on group)
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-get.correlations(results$order_results_within, exclude.phylo = TRUE)
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+## Calculate the correlations
+order_within_cor <- get.correlations(results$order_results_within, exclude.phylo = TRUE)
+
+## Getting the subsets for the colouring
+split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
+col_subs <- replicate(length(order_within_cor), NULL)
+col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
+
+## Set the vector of legend positions
+for(i in 1:length(order_within_cor)) {
+    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]])
+}
 ```

### 2021-11-16 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
                                              type = "groups",
-                                             base = "phylogeny")
+                                             base = "phylogeny",
+                                             output = c("position", "distance"))
 ## Run the analyses between groups
@@
                                              type = "groups",
-                                             base = "phylogeny")
+                                             base = "phylogeny",
+                                             output = c("position", "distance"))
 ## Run the analyses of groups on the phylogeny
@@
                                             inc.base = TRUE,
-                                            verbose = TRUE)
+                                            verbose = TRUE,
+                                            output = c("position", "distance"))
 ## Run the analyses of groups on the phylogeny
@@
                                             inc.base = TRUE,
-                                            verbose = TRUE)
+                                            verbose = TRUE,
+                                            output = c("position", "distance"))
 ## Run the analyses groups on themselves
@@
                                             type = "elements",
-                                            verbose = TRUE)
+                                            verbose = TRUE,
+                                            output = c("position", "distance"))
 ## Run the analyses groups on themselves
@@
                                             type = "elements",
-                                            verbose = TRUE)
+                                            verbose = TRUE,
+                                            output = c("position", "distance"))
 ## Save all
@@
 ## Plot the results
-wrap.plot <- function(results, col, main, names, centre.position = TRUE, absolute = FALSE) {
+wrap.plot <- function(results, col, main, names, centre.position = TRUE, absolute = TRUE, order = TRUE, xlim = NULL) {
+
+    ## Ranking the results
+    if(order) {
+        ## Get the ranking
+        orders <- do.call(cbind, lapply(results, get.disparity, concatenate = FALSE))
+        ## Get the mean values
+        orders <- apply(orders, c(1,2), function(x) mean(abs(unlist(x))))
+        ## Sum them and rank them
+        rank <- order(apply(orders, 1, sum), decreasing = FALSE)
+    } else {
+        rank <- 1:length(results[[1]]$disparity)
+    }
 
     ## Adding one plot
-    one.plot <- function(one_result, xlab, col, zero = TRUE, main, centre, absolute, ...) {
+    one.plot <- function(one_result, xlab, col, zero = TRUE, main, centre, absolute, rank = NULL, xlim = NULL, ...) {
         values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
@@
         }
-        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main = main, ...)
+        if(!is.null(rank)) {
+            values <- values[rank]
+        }
+        if(!is.null(xlim)) {
+            xlim <- xlim
+        } else {
+            xlim <- range(values)
+        }
+
+        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main = main, ylim = xlim, ...)
         if(zero) {
@@
     empty <- matrix(0, ncol = length(results[[1]]$disparity))
-    colnames(empty) <- names(results[[1]]$disparity)
+    colnames(empty) <- names(results[[1]]$disparity)[rank]
     boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
@@
         par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
-        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = is_position, absolute = absolute)
+        one.plot(results[[i]], col = col[rank], centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
     }
@@
           main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation", "angle"))
+          names = c("elaboration", "innovation"))
 ```
@@
 <details>
-  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+  <summary>[Click to expand] Same but zooming on the (0,5) interval (to get less of the flamingo effect)
 </summary>
@@
 wrap.plot(results$super_results_between,
+          xlim = c(0,6),
           col = rainbow(n.subsets(superorder_shapespace)-1),
           main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
-          names = c("absolute elaboration", "innovation", "angle"),
-          centre.position = FALSE, absolute = TRUE)
+          names = c("elaboration", "innovation"),
+          centre.position = TRUE, absolute = TRUE)
 ```
@@
           main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation", "angle"))
+          names = c("elaboration", "innovation"))
 ```
@@
 <details>
-  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
+  <summary>[Click to expand] Again, same but with some zoom (0,15)
 </summary>
@@
 wrap.plot(results$order_results_between,
+          xlim = c(0, 15),
           col = rainbow(n.subsets(order_shapespace)-1),
           main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
-          names = c("absolute elaboration", "innovation", "angle"),
-          centre.position = FALSE, absolute = TRUE)
+          names = c("elaboration", "innovation"),
+          centre.position = TRUE, absolute = TRUE)
 ```
@@
           main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation", "angle"))
+          names = c("elaboration", "innovation"))
 ```
-
-<details>
-  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
-</summary>
-
-```{r, echo = FALSE, fig.height = 8, fig.width = 8}
-wrap.plot(results$super_results_phylo,
-          col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("absolute elaboration", "innovation", "angle"),
-          centre.position = FALSE, absolute = TRUE)
-```
-</details>
 
@@
           main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation", "angle"))
+          names = c("elaboration", "innovation"))
 ```
-
-<details>
-  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
-</summary>
-
-```{r "hehehe", echo = FALSE, fig.height = 16, fig.width = 8, eval = FALSE}
-wrap.plot(results$order_results_phylo,
-          col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[species on phylogeny] orders\nspecies onto the phylogenetic VCV",
-          names = c("absolute elaboration", "innovation", "angle"),
-          centre.position = FALSE, absolute = TRUE)
-````
-</details>
 
@@
 ```{r, fig.height = 8, fig.width = 8, echo = FALSE}
-
 ## Plot the results
@@
           main = "[species on group] superorders species\nonto their group's VCV",
-          names = c("elaboration", "innovation", "angle"))
+          names = c("elaboration", "innovation"))
 ```
-
-<details>
-  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
-</summary>
-
-```{r "beforebefore last", echo = FALSE, fig.height = 8, fig.width = 8}
-wrap.plot(results$super_results_within,
-          col = rainbow(n.subsets(superorder_shapespace)-1),
-          main = "[species on group] superorders species\nonto their group's VCV",
-          names = c("absolute elaboration", "innovation", "angle"),
-          centre.position = FALSE, absolute = TRUE)
-```
-</details>
 
@@
 ```{r "before last", fig.height = 16, fig.width = 8, echo = FALSE}
-
 ## Plot the results
@@
           main = "[species on group] oreders species\nonto their group's VCV",
-          names = c("elaboration", "innovation", "angle"))
+          names = c("elaboration", "innovation"))
 ```
-
-<details>
-  <summary>[Click to expand] The same but using absolute values for the projections (elaboration)
-</summary>
-
-```{r "done with boxplots", echo = FALSE, fig.height = 16, fig.width = 8}
-wrap.plot(results$order_results_within,
-          col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[species on group] oreders species\nonto their group's VCV",
-          names = c("absolute elaboration", "innovation", "angle"),
-          centre.position = FALSE, absolute = TRUE)
-```
-</details>
 

### 2021-11-22 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Update 02.1-shapespace_allbirds_lvl_superorder_order.Rmd

@@
 
-```{r}
+```{r, eval = FALSE}
 ## Preparing the model files
@@
 
-## Checking the diagnosis
-
 ## Saving the results
 save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+```
+
+## Checking the mini chains' ESSs
+
+```{r, eval = TRUE}
 

### 2021-11-23 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

@@
  * By species (i.e exploration/elaboration **within** clades): where we calculated the projections of each species onto 1) the global phylogenetic major axes and 2) the phylogenetic major axes of their respective clade.
+
+These three analyses will be called respectively **group on phylogeny**, **species on phylogeny** and **species on group** respectively.
+
+TODO: add about projection correction: absolute value - 0.5 (because 1) we're not interested in which direction the elaboration goes and 2) centre of the projection is 0.5 not 0)
+
+```{r}
+
+set.seed(42)
+## Space plot
+cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
+space <- space.maker(10, 2, rnorm, cor.matrix = cor_matrix)
+lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
+
+par(mfrow = c(2, 2))
+
+## Plot 1
+plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim, col = "black", main = "Original space")
+
+## Plotting the ellipse
+lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "blue", lwd = 1)
+
+## Plotting the major axis
+lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "blue", lty = 1)
+
+## Plotting a phylogenetic major axis
+lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
+
+legend("topleft", pch = c(19, NA, NA, NA), lty = c(NA, 1,1,1), lwd = c(NA, 1, 3, 3), legend = c("Group of species", "95% CI ellipse for the group", "Major axis for the group", "Major axis for the phylogeny"), col = c("black", "blue", "blue", "orange"))
+
+## Translate the major axis
+
+plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim, col = "grey", main = "Group on phylogeny")
+
+## Plotting the ellipse
+lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "grey", lwd = 1)
+
+## Plotting the major axis
+lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "grey", lty = 1)
+
+## Plotting a phylogenetic major axis
+lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
+
+## Translating the axis
+
+plot(2, main = "species on phylogeny")
+plot(3, main = "species on group")
+```
 

### 2021-11-24 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Update 02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Updated model reading (to avoid re-reading everything each time)

--- 2021-11-01 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
@@
 
-```{r}
+```{r, eval = FALSE}
+## Set up the variables
+models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster/"
+## Getting all the model files
+model_files <- list.files(models_path, pattern = ".rda")
+
+## Check if the model exists already
+if(!("shapespace_allbirds_lvl_superorder_order.MCMCglmm" %in% list.files("../Data/Processed/"))) {
     ## Preparing the model files
@@
     class(model_list) <- c("beer", "mini.chains")
-models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster/"
-## Loading all models
-for(file in list.files(models_path)) {
-    load(paste0(models_path, file, collapse = "/"))
+    ## Get the first model
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    ## Move it to the storage
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/"))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
     model_list[[length(model_list)+1]] <- model[[1]]
+    model <- NULL
+    model_files <- model_files[-1]
+} else {
+    ## Updating the model
+    load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+    model_list <- shapespace_allbirds_lvl_superorder_order
+}
+
+## Get the subsequent models and merge them as we go
+while(length(model_files) != 0) {
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    ## Move it to storage
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
+    model_list[[length(model_list)+1]] <- model[[1]]
+    ## Merge the models (overwrite the model list)
+    model_list <- list(combine.mini.chains(model_list))
+    ## DEBUG
+    print("model list size:")
+    print(format(object.size(model_list), units = "MB"))
+    model <- NULL
+    model_files <- model_files[-1]
 }
@@
 ## Combining the model files
-shapespace_allbirds_lvl_superorder_order <- combine.mini.chains(model_list)
-
-## Checking the diagnosis
+shapespace_allbirds_lvl_superorder_order <- model_list
 
 ## Saving the results
-save(shapespace_allbirds_lvl_superorder_order, file = "../Data/processed/shapespace_allbirds_lvl_superorder_order.rda")
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+```
 
+## Checking the mini chains' ESSs
+
+```{r, eval = TRUE}
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order)
 ```

### 2021-11-24 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
+tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
+
 ## Load the covar
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
 
@@
 
+TODO: centre and absolute directly in `dispRity.covar.projections`
+
 ```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
@@
         ## Get the ranking
-        orders <- do.call(cbind, lapply(results, get.disparity, concatenate = FALSE))
+        orders <- lapply(results, get.disparity, concatenate = FALSE)
         ## Get the mean values
-        orders <- apply(orders, c(1,2), function(x) mean(abs(unlist(x))))
+        if(absolute) {
+            ## Centre and absolute values
+            orders_pos <- lapply(orders$position, function(X) abs(X-0.5))
+        } else {
+            if(centre.position) {
+                ## Just centre
+                orders_pos <- lapply(orders$position, function(X) X-0.5)
+            } else {
+                orders_pos <- orders$position
+            }
+        }
+        orders_dis <- orders$distance
+
+        ## Get the mean for each and combine into a table
+        cent.tend <- mean
+        means <- cbind(unlist(lapply(orders_pos, cent.tend)), unlist(lapply(orders_dis, cent.tend)))
+
         ## Sum them and rank them
-        rank <- order(apply(orders, 1, sum), decreasing = FALSE)
+        sum_means <- apply(means, 1, sum)
+        rank <- order(sum_means, decreasing = FALSE)
     } else {
@@
         par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
-        one.plot(results[[i]], col = col[rank], centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
+        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
     }
@@
           main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
+          names = c("elaboration", "innovation"),
+          centre.position = TRUE, absolute = TRUE)
 ```
@@
 ```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-
 ## Plot the results
@@
 
-```{r "before last", fig.height = 16, fig.width = 8, echo = FALSE}
+```{r, fig.height = 16, fig.width = 8, echo = FALSE}
 ## Plot the results
@@
 #@param absolute whether to make the position values absolute or not
+#@param add.relative whether to add the relative results (as rel.pos and rel.dis) or not
 #@param ... plot options
 
-get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, absolute = FALSE, ...) {
+get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, absolute = FALSE, add.relative = FALSE, ...) {
 
@@
 
+    ## Add the relative results
+    if(add.relative) {
+        make.relative <- function(X) {
+            cbind(X, "rel" = ifelse(is.na(X$score), NA, X$score/max(X$score)))
+        }
+        output <- lapply(output, make.relative)
+    }
     return(output)
 }
-combine.results <- function(res1, res2) {
+combine.results <- function(res1, res2, sort.results = TRUE) {
     ## Combine the results
@@
     res_out$elaboration <- c(res1[, 3], rep(NA, nrow(res2)))
+    if("rel" %in% colnames(res1)) {
+        res_out$rel.elab <- c(res1[, 4], rep(NA, nrow(res2)))
+    }
     res_out$innovation <- c(rep(NA, nrow(res1)), res2[, 3])
+    if("rel" %in% colnames(res1)) {
+        res_out$rel.inno <- c(rep(NA, nrow(res1)), res2[, 4])
+    }
+
+    # Remove NA species
+    if(any(is_na <- is.na(res_out$species))) {
+        if(all(is_na)) {
+            return(res_out)
+        } else {
+            res_out <- res_out[!is_na, , drop = FALSE]
+        }
+    }
+
     ## Find any species in commons
     commons <- intersect(res1$species, res2$species)
-    if(length(commons) > 0 && is.na(commons)) {
+    if(length(commons) > 0) {
+        if(any(is.na(commons))) {
             if(is.na(res1$species)) {
@@
                 duplicates <- which(res_out$species == commons[1])
-            res_out[duplicates[1], 4] <- res_out[duplicates[2], 4]
+                columns <- if("rel" %in% colnames(res1)){c(5,6)} else {4}
+                res_out[duplicates[1], columns] <- res_out[duplicates[2], columns]
                 res_out <- res_out[-duplicates[2], ]
@@
         }
-    return(res_out)
+    }
+
+    if(sort.results) {
+        sums <- apply(res_out[, -c(1,2)], 1, sum, na.rm = TRUE)
+        rank <- match(sort(sums, decreasing = TRUE), sums)
+    } else {
+        rank <- 1:nrow(res_out)
+    }
+
+    return(res_out[rank,])
 }
@@
 
+```{r, "printing results in text", echo = FALSE, eval = TRUE}
+## Combining the results
+super_phylo_table <- do.call(rbind, mapply(combine.results, super_positions_phylo, super_distances_phylo, SIMPLIFY = FALSE))
+## Get the top 10%
+get.top <- function(data, what, top = 0.9, relative = FALSE) {
+    if(any(data[, "level"] == "phylogeny")) {
+        data_top <- data[which(data[, "level"] != "phylogeny"), ]
+    } else {
+        data_top <- data
+    }
+
+    ## Get the relative column
+    if(relative) {
+        what <- switch(what,
+                       "elaboration" = "rel.elab",
+                       "innovation"  = "rel.inno")
+    }
+    ## Get the top 10(ish) species
+    data_top_10 <- data_top[which(data_top[, what] >= quantile(data_top[, what], prob = top, na.rm = TRUE)), ]
+    return(data_top_10[order(data_top_10[, what], decreasing = TRUE), ])
+}
+top_elabo <- get.top(super_phylo_table, what = "elaboration")
+top_innov <- get.top(super_phylo_table, what = "innovation")
+```
+
+> And here are the top 10% of the top 1% elaborators and innovators by superorder:
+The 10% top elaborators by superorder are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
+The 10% top innovators by superorder are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
+
 <details>
@@
 library(knitr)
-## Combining the results
-super_phylo_table <- do.call(rbind, mapply(combine.results, super_positions_phylo, super_distances_phylo, SIMPLIFY = FALSE))
 rownames(super_phylo_table) <- NULL
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
+order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE, absolute = TRUE)
 ```
@@
 
+```{r, echo = FALSE}
+order_phylo_table <- do.call(rbind, mapply(combine.results, order_positions_phylo, order_distances_phylo, SIMPLIFY = FALSE))
+top_elabo <- get.top(order_phylo_table, what = "elaboration")
+top_innov <- get.top(order_phylo_table, what = "innovation")
+```
+
+> And here are the top 10% of the top 1% elaborators and innovators by order:
+The 10% top ellaborators by order are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
+The 10% top innovators by order are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
+
 <details>
@@
 ## Combining the results
-order_phylo_table <- do.call(rbind, mapply(combine.results, order_positions_phylo, order_distances_phylo, SIMPLIFY = FALSE))
 rownames(order_phylo_table) <- NULL
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE)
+super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE, absolute = TRUE, add.relative = TRUE)
 ```
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_within <- get.distributions(results$super_results_within, type = "distance", exclude.phylo = TRUE)
+super_distances_within <- get.distributions(results$super_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
 ```
 Figure: innovations distributions within super orders
+
+```{r, echo = FALSE}
+super_phylo_table_within <- do.call(rbind, mapply(combine.results, super_positions_within, super_distances_within, SIMPLIFY = FALSE))
+top_elabo <- get.top(super_phylo_table_within, what = "elaboration", relative = TRUE)
+top_innov <- get.top(super_phylo_table_within, what = "innovation", relative = TRUE)
+```
+
+> And here are the top 10% of the top 1% elaborators and innovators by superorder relative to each superorder (i.e. the best one(s) in each order):
+```{r, echo = FALSE}
+combined <- rbind(top_elabo, top_innov)
+rownames(combined) <- NULL  
+kable(combined, digits = 3)
+```
 
@@
 ## Combining the results
-super_phylo_table_within <- do.call(rbind, mapply(combine.results, super_positions_within, super_distances_within, SIMPLIFY = FALSE))
 rownames(super_phylo_table_within) <- NULL
@@
 
+```{r, echo = FALSE}
+order_phylo_table_within <- do.call(rbind, mapply(combine.results, order_positions_within, order_distances_within, SIMPLIFY = FALSE))
+top_elabo <- get.top(order_phylo_table_within, what = "elaboration", relative = TRUE)
+top_innov <- get.top(order_phylo_table_within, what = "innovation", relative = TRUE)
+```
+
+> And here are the top 10% of the top 1% elaborators and innovators by order relative to each order (i.e. the best one(s) in each order):
+```{r, echo = FALSE}
+combined <- rbind(top_elabo, top_innov)
+rownames(combined) <- NULL  
+kable(combined, digits = 3)
+```
+
 <details>
@@
 ## Combining the results
-order_phylo_table_within <- do.call(rbind, mapply(combine.results, order_positions_within, order_distances_within, SIMPLIFY = FALSE))
 rownames(order_phylo_table_within) <- NULL
@@
 ## Plotting one correlation 
-plot.correlations <- function(cor.results, col.sub, col.fun = rainbow, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ...) {
+plot.correlations <- function(cor.results, col.sub, col.fun = rainbow, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
     dots <- list(...)
@@
 
+    if(!missing(tree)) {
+        op <- par(mfrow = c(1,2))
+        #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1
+        par(mar = c(5, 4, 4, 0)+0.1)   
+    }
+
     ## Plotting the results
@@
     }
+
+    if(!missing(tree)) {
+        par(mar = c(5, 0, 4, 2) + 0.1)
+
+        ## Recycle the colour vector
+        named_col_vector <- col_vector
+        names(named_col_vector) <- names(cor.results$x)
+
+        ## Reduce the tree
+        if(!full.tree) {
+            tree <- drop.tip(tree, tip = tree$tip.label[which(!(tree$tip.label %in% names(named_col_vector)))])
+        }
+
+        ## Matching edges and colours
+        match.edge.colours <- function(tree, named_col_vector, background = "grey") {
+            ## Get the background colour
+            edge_cols <- rep(background, Nedge(tree))
+            ## Get the clades
+            clades <- unique(named_col_vector)
+            ## Ignore the greys
+            if(any(which_grey <- clades == "grey")) {
+                clades <- clades[!which_grey]
+            }
+            ## Find the edges for each clade
+            for(clade in 1:length(clades)) {
+                ## Get the tips for the clade
+                tips <- which(named_col_vector == clades[clade])
+                ## Get the MRCA node
+                mrca_node <- getMRCA(tree, tips)
+                stop_edge <- which(tree$edge[, 2] %in% mrca_node)
+                ## Get the tips ID
+                tip_ids <- which(tree$tip.label %in% names(tips))
+                ## Get the vector of edges
+                edges <- integer()
+                ## Get the first nodes (and edges)
+                new_edges <- which(tree$edge[, 2] %in% tip_ids)
+                edges <- c(edges, new_edges)
+                nodes <- tree$edge[new_edges, 1]
+
+                # test.col <- c("red", "blue", "orange", "green", "pink", "yellow")
+                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
+
+                ## Recursively loop through each node up until the last edge
+                while(!(stop_edge %in% edges)) {
+                    ## Get the node edges
+                    new_edges <- which(tree$edge[, 2] %in% nodes)
+
+                    # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
+
+                    edges <- c(edges, new_edges)
+                    ## Update the nodes
+                    nodes <- tree$edge[new_edges, 1]
+                }
+
+                ## Do the last recursion
+                new_edges <- which(tree$edge[, 2] %in% nodes)
+                edges <- c(edges, new_edges)
+
+                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
+
+                ## Colour these edges
+                edge_cols[edges] <- clades[clade]
+            }
+            return(edge_cols)
+        }
+        ## Match the colour to the correct edge
+        edge_cols <- match.edge.colours(tree, named_col_vector)
+
+        ## Add the tree
+        plot(tree, show.tip.label = FALSE, edge.color = edge_cols)
+    }
+
     return(invisible())
@@
 for(i in 1:length(super_phylo_cor)) {
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i])
+    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
+
+    # bugged for all but 1
+
 }
@@
 for(i in 1:length(super_within_cor)) {
-    plot.correlations(super_within_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i])
+    plot.correlations(super_within_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
 }
@@
 for(i in 1:length(order_phylo_cor)) {
-    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]])
+    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]], tree = tree)
 }
@@
 for(i in 1:length(order_within_cor)) {
-    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]])
+    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]], tree = tree)
 }

### 2021-11-24 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Update 02.1-shapespace_allbirds_lvl_superorder_order.Rmd

@@
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order)
+diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order[[1]])
 ```

### 2021-11-25 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = TRUE, echo = TRUE}
 ## Load the data
@@
                                              base = "phylogeny",
-                                             output = c("position", "distance"))
+                                             verbose = TRUE,
+                                             output = c("position", "distance"),centre.proj = FALSE)
 ## Run the analyses between groups
@@
                                              base = "phylogeny",
-                                             output = c("position", "distance"))
+                                             verbose = TRUE,
+                                             output = c("position", "distance"),centre.proj = FALSE)
 ## Run the analyses of groups on the phylogeny
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"))
+                                            output = c("position", "distance"),centre.proj = FALSE)
 ## Run the analyses of groups on the phylogeny
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"))
+                                            output = c("position", "distance"),centre.proj = FALSE)
 ## Run the analyses groups on themselves
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"))
+                                            output = c("position", "distance"),centre.proj = FALSE)
 ## Save all
@@
         "order_results_within"  = order_results_within)
-save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
 ```
@@
 ## Quick load
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
+results_cent <- shapespace_allbirds_lvl_superorder_order_results_list
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
@@
 ## Plot the results
-wrap.plot <- function(results, col, main, names, centre.position = TRUE, absolute = TRUE, order = TRUE, xlim = NULL) {
+wrap.plot <- function(results, col, main, names, order = TRUE, xlim = NULL) {
 
@@
         ## Get the mean values
-        if(absolute) {
-            ## Centre and absolute values
-            orders_pos <- lapply(orders$position, function(X) abs(X-0.5))
-        } else {
-            if(centre.position) {
-                ## Just centre
-                orders_pos <- lapply(orders$position, function(X) X-0.5)
-            } else {
         orders_pos <- orders$position
-            }
-        }
         orders_dis <- orders$distance
@@
         values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
-        if(centre) {
-            values <- lapply(values, function(X) {X-0.5})
-        }
-        if(absolute) {
-            values <- lapply(values, function(X) {abs(X-0.5)})
-        }
         if(!is.null(rank)) {
@@
           main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"),
-          centre.position = TRUE, absolute = TRUE)
+          names = c("elaboration", "innovation"))
 ```
@@
           main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"),
-          centre.position = TRUE, absolute = TRUE)
+          names = c("elaboration", "innovation"))
 ```
@@
           main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"),
-          centre.position = TRUE, absolute = TRUE)
+          names = c("elaboration", "innovation"))
 ```
@@
           col = rainbow(n.subsets(order_shapespace)-1),
-          main = "[species on group] oreders species\nonto their group's VCV",
+          main = "[species on group] orders species\nonto their group's VCV",
           names = c("elaboration", "innovation"))
@@
 #@param exclude.phylo remove the phylogeny subset (default = FALSE)
-#@param absolute whether to make the position values absolute or not
 #@param add.relative whether to add the relative results (as rel.pos and rel.dis) or not
 #@param ... plot options
-
-get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, absolute = FALSE, add.relative = FALSE, ...) {
+get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, add.relative = FALSE, ...) {
 
@@
     disparity <- lapply(res[[type]]$disparity, function(X, type) return(X$elements))
-
-    ## Recentre the results
-    if(type == "position") {
-        if(!absolute) {
-            disparity <- lapply(disparity, function(X) return(X-0.5))
-        } else {
-            disparity <- lapply(disparity, function(X) return(abs(X-0.5)))
-        }
-    }
 
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position", absolute = TRUE)
+super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position")
 ```
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE, absolute = TRUE)
+order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
 ```
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE, absolute = TRUE, add.relative = TRUE)
+super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE,  add.relative = TRUE)
 ```
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE)
+order_positions_within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE, add.relative = TRUE)
 ```
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_within <- get.distributions(results$order_results_within, type = "distance", exclude.phylo = TRUE)
+order_distances_within <- get.distributions(results$order_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
 ```
@@
 ## Calculating all the correlations
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE, absolute = TRUE) {
+get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE) {
 
@@
     disparity_dis <- lapply(res$distance$disparity, function(X, type) return(X$elements))
-
-    ## Recentre the results
-    if(!absolute) {
-        disparity_pos <- lapply(disparity_pos, function(X) return(X-0.5))
-    } else {
-        disparity_pos <- lapply(disparity_pos, function(X) return(abs(X-0.5)))
-    }
 
@@
                 mrca_node <- getMRCA(tree, tips)
-                stop_edge <- which(tree$edge[, 2] %in% mrca_node)
+                # stop_edge <- which(tree$edge[, 2] %in% mrca_node)
+
+                ## Handle non-monophyletic groups
+                clade_tree <- extract.clade(tree, node = mrca_node)
+                if(Ntip(clade_tree) > length(tips)) {
+                    clade_tree <- drop.tip(clade_tree, tip = clade_tree$tip.label[!(clade_tree$tip.label %in% names(tips))])
+                }
+                n_edges_target <- Nedge(clade_tree)
+
                 ## Get the tips ID
@@
                 new_edges <- which(tree$edge[, 2] %in% tip_ids)
-                edges <- c(edges, new_edges)
+                edges <- unique(c(edges, new_edges))
                 nodes <- tree$edge[new_edges, 1]
 
-                # test.col <- c("red", "blue", "orange", "green", "pink", "yellow")
+                # plot(tree, show.tip.label = FALSE)
+                # test.col <- c(replicate(10, c("red", "blue", "orange", "green", "pink", "yellow")))
                 # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
@@
                 ## Recursively loop through each node up until the last edge
-                while(!(stop_edge %in% edges)) {
+                # while(!(stop_edge %in% edges)) {
+                while(length(edges) <= n_edges_target) {
                     ## Get the node edges
@@
 
-                    edges <- c(edges, new_edges)
+                    edges <- unique(c(edges, new_edges))
                     ## Update the nodes
@@
 
-                ## Do the last recursion
-                new_edges <- which(tree$edge[, 2] %in% nodes)
-                edges <- c(edges, new_edges)
+                # ## Do the last recursion
+                # new_edges <- which(tree$edge[, 2] %in% nodes)
+                # edges <- unique(c(edges, new_edges))
 
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
 ## Getting the subsets for the colouring
@@
     plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
-
-    # bugged for all but 1
-
 }
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
 ## Getting the subsets for the colouring
 col_subs <- split(shapespace$level2, f = shapespace$level1)[-1]
+
 ## Calculate the correlations
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
 ## Loading the sub-order levels
@@
 for(i in 1:length(order_phylo_cor)) {
-    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]], tree = tree)
+    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]])
 }
@@
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
 ## Calculate the correlations
@@
 for(i in 1:length(order_within_cor)) {
-    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]], tree = tree)
+    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]])
 }

### 2021-11-26 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
+## Clean garbage
+rm(shapespace_allbirds_lvl_superorder_order)
 
@@
       rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+## Clean garbage
+rm(shapespace_allbirds_lvl_superorder_order)
 ```
@@
            cex = 0.5)
-## Plotting mean axes
 covar.plot(order_shapespace,
@@
     covar.plot(sub, major.axes = mean, ellipses = mean, points = FALSE, col = "red", add = TRUE)
+    rm(sub)
 }
@@
 ## Quick load
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
-results_cent <- shapespace_allbirds_lvl_superorder_order_results_list
+# load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
+# results_cent <- shapespace_allbirds_lvl_superorder_order_results_list
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
 results <- shapespace_allbirds_lvl_superorder_order_results_list
+rm(shapespace_allbirds_lvl_superorder_order_results_list)
 ```
@@
     }
+    rm(disparity)
     return(output)

### 2021-11-30 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
-```{r, eval = TRUE, echo = TRUE}
+```{r, eval = TRUE, echo = FALSE}
 ## Load the data
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(
+superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(
+order_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
                                              verbose = TRUE,
-                                             output = c("position", "distance"),centre.proj = FALSE)
+                                             output = c("position", "distance"), centre.proj = TRUE)
 ## Run the analyses between groups
@@
                                              verbose = TRUE,
-                                             output = c("position", "distance"),centre.proj = FALSE)
+                                             output = c("position", "distance"), centre.proj = TRUE)
 ## Run the analyses of groups on the phylogeny
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"),centre.proj = FALSE)
+                                            output = c("position", "distance"), centre.proj = TRUE)
 ## Run the analyses of groups on the phylogeny
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"),centre.proj = FALSE)
+                                            output = c("position", "distance"), centre.proj = TRUE)
 ## Run the analyses groups on themselves
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"),centre.proj = FALSE)
-## Save all
+                                            output = c("position", "distance"), centre.proj = TRUE)
 shapespace_allbirds_lvl_superorder_order_results_list <- list(
@@
         "order_results_within"  = order_results_within)
-save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
+save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_centred.rda")
 ```
@@
 # results_cent <- shapespace_allbirds_lvl_superorder_order_results_list
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_centred.rda")
 results <- shapespace_allbirds_lvl_superorder_order_results_list
@@
 > And here are the top 10% of the top 1% elaborators and innovators by superorder:
+
 The 10% top elaborators by superorder are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
+
 The 10% top innovators by superorder are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.

### 2021-12-02 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 ```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
+## The ggplot color panel
+gg.color.hue <- function(n) {
+    grDevices::hcl(h = seq(15, 375, length = n + 1), l = 65, c = 100)[1:n]
+}
+
+# colour.palette <- rainbow
+colour.palette <- gg.color.hue
+
 ## Plotting just the space
@@
            legend = TRUE,
-           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
            legend.args = list(x = "bottomright", cex = 2/3))
@@
 covar.plot(superorder_shapespace,
-           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
            main = "Average ellipses by super order",
@@
            points = FALSE,
-           cex = 0.5)
+           cex = 0.5, apply.to.VCV = TRUE)
 ## Plotting mean ellipses
 covar.plot(superorder_shapespace,
-           col = c(rainbow(n.subsets(superorder_shapespace)-1), "grey"),
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
            main = "Average major axes by super order",
@@
            points = FALSE,
-           cex = 0.5)
+           cex = 0.5, apply.to.VCV = TRUE)
 covar.plot(order_shapespace,
-           col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
+           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
            main = "Average ellipses by order",
@@
            points = FALSE,
-           cex = 0.5)
+           cex = 0.5, apply.to.VCV = TRUE)
 ## Plotting mean ellipses
 covar.plot(order_shapespace,
-           col = c(rainbow(n.subsets(order_shapespace)-1), "grey"),
+           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
            main = "Average major axes by order",
@@
            points = FALSE,
-           cex = 0.5)
+           cex = 0.5, apply.to.VCV = TRUE)
 ```
@@
 wrap.plot(results$super_results_between,
-              col = rainbow(n.subsets(superorder_shapespace)-1),
+              col = colour.palette(n.subsets(superorder_shapespace)-1),
               main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
@@
           xlim = c(0,6),
-          col = rainbow(n.subsets(superorder_shapespace)-1),
+          col = colour.palette(n.subsets(superorder_shapespace)-1),
           main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
@@
 wrap.plot(results$order_results_between,
-          col = rainbow(n.subsets(order_shapespace)-1),
+          col = colour.palette(n.subsets(order_shapespace)-1),
           main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
@@
           xlim = c(0, 15),
-          col = rainbow(n.subsets(order_shapespace)-1),
+          col = colour.palette(n.subsets(order_shapespace)-1),
           main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
@@
 wrap.plot(results$super_results_phylo,
-          col = rainbow(n.subsets(superorder_shapespace)-1),
+          col = colour.palette(n.subsets(superorder_shapespace)-1),
           main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
@@
 wrap.plot(results$order_results_phylo,
-          col = rainbow(n.subsets(order_shapespace)-1),
+          col = colour.palette(n.subsets(order_shapespace)-1),
           main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
@@
 wrap.plot(results$super_results_within,
-          col = rainbow(n.subsets(superorder_shapespace)-1),
+          col = colour.palette(n.subsets(superorder_shapespace)-1),
           main = "[species on group] superorders species\nonto their group's VCV",
@@
 wrap.plot(results$order_results_within,
-          col = rainbow(n.subsets(order_shapespace)-1),
+          col = colour.palette(n.subsets(order_shapespace)-1),
           main = "[species on group] orders species\nonto their group's VCV",
@@
 ## Plotting one correlation 
-plot.correlations <- function(cor.results, col.sub, col.fun = rainbow, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
+plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
     dots <- list(...)
@@
 
+<details>
+  <summary>[Click to expand] Finding corner cases groups.
+</summary>
+
+So here's how you remove exploration and elaboration: by centering these three species on the centre of the phylo axis:
+
+Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
+colour.palette <- rainbow
+posteriors <- shapespace_allbirds_lvl_superorder_order
+
+# plot(superorder_shapespace$matrix[[1]][, 1:2], pch = 19, cex = 0.5)
+# identify(superorder_shapespace$matrix[[1]][, 1:2])
+# shapespace[c(268, 286, 80), c(9, 10)]
+
+## Moving these three species around: Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
+fixed_points <- c(268, 286, 408)
+skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
+move_around <- c(8749, 8750, 8751)
+
+## Get the major axis
+phylo_axis <- axis.covar(superorder_shapespace, sample = median)$phylogeny[[1]]
+## Point with no elaboration/exploration
+no_nothing <- 0
+
+## Example with no exploration/elaboration
+skewed_space[move_around, c(1:8)] <- no_nothing
+
+## Recreate the space
+superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny"))
+```
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Visualise all that
+par(mfrow = c(1,2))
+## Plot the space
+covar.plot(superorder_skewed_space,
+           cex = 0.5,
+           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
+           main = "elements without\nelaboration innovation")
+## Add the phylogenetic axis
+lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
+lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
+## Adding the points
+points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+
+## Recalculate the disparity
+results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
+## Getting the correlation plot
+col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
+cor_results_skewed <- get.correlations(results_skewed)
+plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
+data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
+points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+```
+
+So easy right? We just need to change them along the PC axis to move them around in space.
+
+No, it's not fucking easy because it's a 8 dimensional problem...
+Even though the two first axes cumulate 86.27% of the variance, fixing the three elements on the two elements and centering them on the 6 other dimensions (13.73% of the variance) results in BIG changes:
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Reinitialising the space
+skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
+
+## Example with no exploration/elaboration
+skewed_space[move_around, c(3:8)] <- 0
+
+## Recreate the space
+superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
+```
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Visualise all that
+par(mfrow = c(1,2))
+## Plot the space
+covar.plot(superorder_skewed_space,
+           cex = 0.5,
+           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
+           main = "Centring only on dimensions 3:8")
+## Add the phylogenetic axis
+lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
+lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
+## Adding the points
+points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+
+## Recalculate the disparity
+results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
+## Getting the correlation plot
+col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
+cor_results_skewed <- get.correlations(results_skewed)
+plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
+data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
+points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+```
+
+For example, the blue element, Anas_falcata is moved to no elaboration/exploration at all if we do that ignoring data past the second axis.
+There's a duck axis lurking in the depth... (spoiler, it's the fourth dimension).
+
+Let's flatten the problem in 2D for simplification:
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Reinitialising the space
+skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
+
+## Recreate the space
+superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
+
+## 2Ding the space
+dimensions <- c(1,2)
+superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
+superorder_skewed_space$call$dimensions <- dimensions
+fun2d <- function(covar) {
+    covar$VCV <- covar$VCV[dimensions, dimensions]
+    covar$loc <- covar$loc[dimensions]
+    return(covar)
+}
+superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
+```
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Visualise all that
+par(mfrow = c(1,2))
+## Plot the space
+covar.plot(superorder_skewed_space,
+           cex = 0.5,
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
+           main = "Flattening the results")
+## Add the phylogenetic axis
+lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
+lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
+## Adding the points
+points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+
+## Recalculate the disparity
+results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
+## Getting the correlation plot
+col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
+cor_results_skewed <- get.correlations(results_skewed)
+plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
+data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
+points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+abline(a = 0, b = 30, col = "grey")
+abline(a = 0, b = 0.03, col = "grey")
+```
+
+So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.03 and 30).
+Let's see if it's something birdy or not by modifying these elements position in trait space to have elaboration of 0 and innovation of 1 and the opposite.
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Reinitialising the space
+skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
+
+## Move around the first species to have elaboration = 0 and innovation = 0
+skewed_space[move_around[1], c(1:2)] <- 0
+## Move around the second species to have elaboration = 0 and innovation = 1
+skewed_space[move_around[2], c(1:2)] <- c(0, 0.4)
+## Move around the third species to have elaboration = 1 and innovation = 0
+skewed_space[move_around[3], c(1:2)] <- c(0.4, 0)
+
+## Recreate the space
+superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
+
+## 2Ding the space
+dimensions <- c(1,2)
+superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
+superorder_skewed_space$call$dimensions <- dimensions
+fun2d <- function(covar) {
+    covar$VCV <- covar$VCV[dimensions, dimensions]
+    covar$loc <- covar$loc[dimensions]
+    return(covar)
+}
+superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
+```
+
+```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
+## Visualise all that
+par(mfrow = c(1,2))
+## Plot the space
+covar.plot(superorder_skewed_space,
+           cex = 0.5,
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
+           main = "Flattening the results")
+## Add the phylogenetic axis
+lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
+lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
+## Adding the points
+points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+
+## Recalculate the disparity
+results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
+## Getting the correlation plot
+col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
+cor_results_skewed <- get.correlations(results_skewed)
+plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
+data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
+points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
+points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
+abline(a = 0, b = 30, col = "grey")
+abline(a = 0, b = 0.03, col = "grey")
+```
+
+TODO: fix this!
+
+</details>
+
 ### Correlations within super orders (species on group)
+
+Regardless, here are the results when projecting the species on their own group's VCV
 
@@
 ```
+
+<!-- THIS BIT IS IGNORED FOR NOW -->
 

### 2021-12-02 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
-colour.palette <- rainbow
+# colour.palette <- rainbow
 posteriors <- shapespace_allbirds_lvl_superorder_order
@@
 fixed_points <- c(268, 286, 408)
+move_around <- c(8749, 8750, 8751)
 skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-move_around <- c(8749, 8750, 8751)
 
@@
 points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-abline(a = 0, b = 30, col = "grey")
-abline(a = 0, b = 0.03, col = "grey")
-```
-
-So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.03 and 30).
+abline(a = 0, b = 7, col = "grey")
+abline(a = 0, b = 0.14, col = "grey")
+```
+
+So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.14 and 7).
 Let's see if it's something birdy or not by modifying these elements position in trait space to have elaboration of 0 and innovation of 1 and the opposite.
@@
 ## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
+lines(phylo_axis[,1:2], lty = 1, lwd = 4, col = "grey")
+lines(phylo_axis[,1:2], lty = 1, lwd = 2.5, col = "black")
 ## Adding the points
@@
 points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-abline(a = 0, b = 30, col = "grey")
-abline(a = 0, b = 0.03, col = "grey")
 ```

### 2022-01-07 mcmcmcglmmm/inst/MCMCglmm_mini_chains.Rmd
Create MCMCglmm_mini_chains.Rmd

---
title: "mini chains MCMCglmm method"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

```{r, echo = FALSE}
## For a fancy vignette
library(knitr)
```

This vignette explains the details of how the mini chains MCMCglmm method works and how to use it's implementation in the `mcmcmcglmmm` package.
Note that this method is entirely based on Jarrod Hadfield's [`MCMCglmm` package](https://cran.r-project.org/web/packages/MCMCglmm/index.html) which fits multivariate generalised linear mixed models using markov chain monte carlo techniques [@hadfield2010]. 
The `MCMCglmm` method **will not be explained in details here**.
If you need to know more about this method, please refer to the excellent vignettes provided by the `MCMCglmm` package (a brief [overview](https://cran.r-project.org/web/packages/MCMCglmm/vignettes/Overview.pdf) and a more advanced set of [course notes](https://cran.r-project.org/web/packages/MCMCglmm/vignettes/CourseNotes.pdf)).

## Installation and requirements

To install the `mcmcmcglmmm` package, you can do it by directly downloading the latest version on GitHub using `devtools`:

```{r install_mcmcmcglmmm, eval = FALSE}
## Installing mcmcmcglmmm
devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/mcmcmcglmmm")
```

It also relies on the latest version of the `dispRity` package (>=1.6.8) for some specific background functions.
The package will automatically install the latest version of `MCMCglmm` and `dispRity`.

```{r load_mcmcmcglmmm, eval = TRUE, message = FALSE}
set.seed(42)
## Loading mcmcmcglmmm
library(mcmcmcglmmm)

## Loading the demo data and trees
data(demo_data)
data(demo_tree)
```

The data loaded by `demo_data` and `demo_tree` is a small dataset from @cooney2017.
It contains 55 species data frame containing 5 dimensions, species names and clade names and one corresponding phylogenetic tree for the 55 species.

Here is what the dataset looks like:

```{r, echo = FALSE}
kable(head(demo_data))
```
Note that the `"animal"` column is required for `MCMCglmm` (but please refer to the function manuals referenced above for more details).

And here's what the tree looks like:

```{r, echo = FALSE}
plot(ladderize(demo_tree), cex = 0.5)
```

### Using your own data

For simplicity further down the line, or if you want to follow this vignette with your own data, we are going to name the `demo_data` `trait_space` (here an ordinated morphospace) and create a list of three `demo_tree` (that are identical) to demonstrate how `mcmcmcglmmm` works on multiple trees.
Feel free to replace `demo_data` and `demo_tree` in the following snippet by your own data for more fun (the result of the tutorial will use the object names `trait_space` `consensus_tree` and `trees_list`).

```{r, eval = TRUE}
## Renaming the data
trait_space <- demo_data
consensus_tree <- demo_tree
trees_list <- list(demo_tree, demo_tree, demo_tree)
class(trees_list) <- "multiPhylo"
```

# The mini chains MCMCglmm method

To estimate the variance-covariance matrix for the phylogeny and each clade we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@hadfield2010].

## The model

For this demo, we are going to use a mixed model with two nested random levels, one for the whole phylogeny and one for each different clade on our response variable: the first 3 PC axes from the trait space.
In brief the model we're using has the following structure:

> model = traits + random terms + residuals terms

Where `traits` is the 3 dimensional ordination values for each species, `random terms` are the two nested phylogenetic effects and `residuals` are the residuals on the dataset. In `MCMCglmm` pseudo code, the model looks like this (for each `formula`, `random` and `residuals` arguments):

```
formula   = PC1:8 ~ trait:clade-1
random    = ~ us(at.level(clade):trait):animal + us(trait):animal
residuals = ~ us(trait):units
```

In this demo case, the model would actually run pretty fast on each of the demo trees.
If you have a small dataset like presented here, you might be interested in looking at the older `mulTree` [@mulTree] method that will simply run this model on the three trees in a parallelisable way.

## `mcmcmcglmmm`

However, on much larger datasets, this becomes very quickly untractable computationally.
For example, for a dataset of  8 dimensions, 8k species and 3 nested levels it takes approximately 2 weeks and a whopping 512 GB of RAM to barely get results for one tree!

Thus, to increase the speed of these analyses, while taking phylogenetic uncertainty into account, we used a highly parallelisable "mini chains" approach.
In brief, it runs multiple short `MCMCglmm` analyses on multiple trees and pulls the results together into one larger `MCMCglmm` that contains more variation due to phylogenetic uncertainty (where the size of the chains are optimised for speed and low RAM usage).

[mini chain diagram](mini-chains_diagram.png "Mini chains diagram")

The method works as follows:

 1- run three models without burnin on a consensus tree with flat priors: these are the parametrisation chains.

 2- extract the burnin length and the posteriors to be used as priors from the parametrisation chains.

 3- run multiple models with 10 samplings past the burnin using as priors the posteriors from the parametrisation chains: these are the mini chains.

 4- combine all the post burning 10 samples from the mini chains into one big chain to be used as the posterior from the model.

Each step are described and demonstrated in more details below.

Using this method allowed us to run the big model described above in approximately 40 hours and 8GB of RAM per tree!

### Mini-chains parametrization

First we run three independent MCMC chains with the model and data described above using the consensus tree and flat priors (with a belief parameter of 0.02).
We ran these chains for 50k iterations, sampling every 500 iterations (with no burnin).

To do so, we can use the `make.mini.chains` function that creates a `mini.chains` object that contains all the information (data, model and trees) for further analysis.

```{r, parametrise_the_model, eval = TRUE}
## Set up the parametrising chains models on the demo data
param_MCMCglmm <- make.mini.chains(data         = trait_space,
                                   dimensions   = 1:3,
                                   tree         = consensus_tree,
                                   trait.family = "gaussian",
                                   randoms      = c("global", "clade"),
                                   residuals    = "global",
                                   priors       = 0.02,
                                   verbose      = TRUE,
                                   parameters   = list(
                                       nitt   = 50000,
                                       thin   = 500,
                                       burnin = 0))
## What is this new object?
class(param_MCMCglmm)
```

The main arguments here are:

 * `data`: the dataset as described above. If you use the phylogeny or clades as a random term, it needs to have one column called `"animal"` and another one named by the grouping factor (here `"clade"`).
 * `dimensions`: which dimensions from `data` to include. If the argument is a numeric value (like `1:3`) it automatically picks the corresponding columns that have numeric values (here `"PC1"`, `"PC2"` and `"PC3"`). However, you can also directly specify the name of the column (e.g. `c("PC1", "PC2", "PC3")` would five the same results).
 * `tree`: the tree to use. Here we use only one tree (the consensus one) but later we are going to use a tree distribution.
 * `trait.family`: which corresponds to the `family` argument in `MCMCglmm`. Note that if you specify only one family (here `"gaussian"`) it is applied to all the traits (here `"PC1"`, `"PC2"` and `"PC3"`). But you can also specify a vector of families (e.g. `c("gaussian", "poisson", "gaussian")`).
 * `randoms`: which random terms to use. Here `"global"` is the generic term for applying a random term to the whole dataset (the phylogeny) and `"clade` allows to apply another nested random terms to the elements described in the column `"clade"` in the dataset. Note that you can play around with many more options for the random terms by looking at the function manual (`?make.mini.chains`).
 * `residuals`: which residual terms to use. This is the same as the `randoms` argument. The `"global"` term is the generic term for applying the residuals to the whole dataset (traits) and no other nested term has been provided.
 * `priors`: setting up the prior. Here you can provide a list of priors in the proper `MCMCglmm` format (see below) or, as we are doing here, simply a flat prior for each term. To provide a flat prior, you just need to give the belief parameter value (here 2%) and the function will generate the correct flat priors attached to this belief.

The other arguments here are either self evident (e.g `verbose`) or the same as for `MCMCglmm` (here `parameters` is just a list of arguments with the correct names to be passed to `MCMCglmm`).

Once the mini.chain is ready, you can run it using the `run.mini.chains` function given the mini chain (`param_MCMCglmm`) and a number of replicates (here 3).

> This step in this example should take anywhere around 1 to 5 minutes.

```{r, eval = FALSE}
## Running the three MCMCglmm models
parametrization <- run.mini.chains(param_MCMCglmm, replicates = 3)

## Saving the model
```

From these three chains, we first extracted the three burnin periods (defined as the number of iterations when the chain reaches the median likelihood value times 1.1).
We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains. 
For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.

```{r, get_model_parameters, eval = FALSE}
## Extract the parameters from the parameter chains
estimated_params <- extract.parameters(parametrization)
```

### Running the mini-chains

We set up a mini-chain to be a MCMCglmm running with:
  1) the model described above,
  2) a random tree, and
  3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 10 sampled iterations (`nitt = burnin + thin * 10`).
Each mini-chain thus provided us with 10 posterior exploitable data points.
The two main advantages of this mini-chain approach is that
  1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and
  2) they take into account tree uncertainty without having to run the complete `MCMCglmm` on all trees (c.f. `mulTree` @mulTree).

```{r, set_the_mini_chains, eval = FALSE}

## The thinning parameter
thin <- 500
## The number of samples per chains
samples <- 10

## Set up the mini-chains models
parametrisation_chain <- make.mini.chains(
                                data         = trait_space,
                                dimensions   = 1:3,
                                tree         = trees_list,
                                trait.family = "gaussian",
                                randoms      = c("global", "clade"),
                                residuals    = "global",
                                priors       = est_params$priors,
                                verbose      = TRUE,
                                parameters   = list(
                                    nitt   = est_params$burnin + thin * samples,
                                    thin   = thin,
                                    burnin = est_params$burnin))
```

We then run the `MCMCglmm` over 100 randomly selected trees to get an estimate of the variance-covariance matrices including phylogenetic uncertainty.

```{r, running_replicates, eval = FALSE}
## The number of replicates
n_reps <- 100

## Run the required chains
results <- run.mini.chains(mini_chains, replicates = n_reps)

## Combining the results as one big MCMC
combined_results <- combine.mini.chains(results)
```

<!--

#### Number of runs optimisation (TO BE DEVELOPED?)

```{r, eval = FALSE}
## Function for running one mini chain on a random tree (and saving the results)
fun.mini.chains <- function() {
    run.mini.chains(mini_chains, replicates = 1, path = "../Data/Processed/MCMC/", file.name = "mini.chains")
}
```

To make sure we reached consistent results, we run the set of N mini-chains were N was determined using the following variance optimisation algorithm.

 * 1. run an initial $N = 20$ chains;
 * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
 * 3. update N to N = N * 1.05 and run an additional 5% chains;
 * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
 * 5. calculate $\Delta_V$, the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
 * 6. If $\Delta_V < 0.05$ (meaning an increase of 5% of the replicates resulted in an increase of the parameters of less than 5%), the desired number of mini-chains is reached. Else, set $V_{1} = V_{2}$ and repeat from step 3.

```{r, optimised_replicates, eval = FALSE}
## The summary function

```

After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "standard" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.

-->

## References

### 2022-01-07 mcmcmcglmmm/R/mcmcmcglmmm-package.R
package name and data formals

@@ -12,75 +12,24 @@
 #'
 # @import MCMCglmm
 # @import dispRity
-
 NULL
 
 
-
-
-
-# #' @title Charadriiformes
-# #' @name charadriiformes
-# #'
-# #' @description An example of a \code{\link[MCMCglmm]{MCMCglmm}} model.
-# #'
-# #' @details This dataset is based on a random subset of 359 Charadriiformes (gulls, plovers and sandpipers) from Cooney et al 2017.
-# #' It contains:
-# #' \itemize{
-# #'   \item \code{data} A \code{"data.frame"} .
-# #'   \item \code{tree} A consensus tree of 359 charadriiformes species (\code{"phylo"}).
-# #'   \item \code{posteriors} The posteriors from a \code{"MCMCglmm"} model (see example below).
-# #' }
-# #'
-# #' @format one \code{data.frame}, one \code{phylo} and one \code{MCMCglmm}.
-# #'
-# #' @references Cooney CR, Bright JA, Capp EJ, Chira AM, Hughes EC, Moody CJ, Nouri LO, Varley ZK, Thomas GH. Mega-evolutionary dynamics of the adaptive radiation of birds. Nature. 2017 Feb;542(7641):344-7.
-# #' 
-# #' @examples
-# # set.seed(42)
-# #' \dontrun{
-# #' ## Reproducing the MCMCglmm model
-# #' require(MCMCglmm)
-# #' data(charadriiformes)
-# #' 
-# #' ## Setting up the model parameters:
-# #' ## 1 - The formula (the first three PC axes)
-# #' model_formula <- cbind(PC1, PC2, PC3) ~ trait:clade-1
-# #' ## 2 - The residual term
-# #' model_residuals <- ~us(trait):units
-# #' ## 3 - The random terms
-# #' ## (one per clade and one for the whole phylogeny)
-# #' model_randoms <- ~ us(at.level(clade,1):trait):animal
-# #'                  + us(at.level(clade,2):trait):animal
-# #'                  + us(at.level(clade,3):trait):animal
-# #'                  + us(trait):animal
-# #' 
-# #' ## Flat priors for the residuals and random terms
-# #' flat_priors <- list(
-# #'      ## The residuals priors
-# #'      R = list(
-# #'          R1 = list(V = diag(3), nu = 0.002)), 
-# #'      ## The random priors (the phylogenetic terms)
-# #'      G = list(
[…]

### 2022-01-07 mcmcmcglmmm/R/make.mini.chains.R
class update: removed beer

@@ -211,7 +211,7 @@ make.mini.chains <- function(data, dimensions, tree, trait.family = "gaussian", 
                                               ...)))
                     , fixed, random, rvoc, family, data, priors, verbose, parameters, ...)
 
-    class(output) <- c("beer", "mini.chains")
+    class(output) <- c("mini.chains")
     return(output)
 }
 

### 2022-01-18 Analysis/02.1-shapespace_allbirds_lvl_superorder_order.Rmd
Update 02.1-shapespace_allbirds_lvl_superorder_order.Rmd

@@
     ## Move it to the storage
-    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/"))
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
     print(model_files[1])
@@
 save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+
 ```

### 2022-01-31 Analysis/03.1-results.Rmd
shiny updated results

@@
                                              verbose = TRUE,
-                                             output = c("position", "distance"), centre.proj = TRUE)
+                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
 ## Run the analyses between groups
@@
                                              verbose = TRUE,
-                                             output = c("position", "distance"), centre.proj = TRUE)
+                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
 ## Run the analyses of groups on the phylogeny
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"), centre.proj = TRUE)
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
 ## Run the analyses of groups on the phylogeny
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"), centre.proj = TRUE)
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
 ## Run the analyses groups on themselves
@@
                                             verbose = TRUE,
-                                            output = c("position", "distance"), centre.proj = TRUE)
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
 shapespace_allbirds_lvl_superorder_order_results_list <- list(
@@
           col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
+          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV (ZOOM)",
           names = c("elaboration", "innovation"))
@@
           col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
+          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV (ZOOM)",
           names = c("elaboration", "innovation"))
@@
     if(!missing(tree)) {
-        op <- par(mfrow = c(1,2))
+        # op <- par(mfrow = c(1,2))
         #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1

### 2022-02-04 Analysis/03.1-results.Rmd
Update 03.1-results.Rmd

@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
 ## Run the analyses between groups
@@
                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
+
 ## Run the analyses of groups on the phylogeny
@@
         "order_results_within"  = order_results_within)
-save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_centred.rda")
+save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+```
+
+```{r, echo = FALSE, eval = FALSE}
+## Getting the plotting data
+ploting_data <- list("median_elaborations" = apply(get.disparity(get.subsets(super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "median_innovations"  = apply(get.disparity(get.subsets(super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "distances_to_centre" =  get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
+                     "consensus_tree"      = tree)
+
+## Save all that as a list
+save(ploting_data, file = "ploting_data.rda")
 ```
@@
 ## Quick load
-# load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
-# results_cent <- shapespace_allbirds_lvl_superorder_order_results_list
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_centred.rda")
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
 results <- shapespace_allbirds_lvl_superorder_order_results_list
@@
     if(!missing(tree)) {
-        # op <- par(mfrow = c(1,2))
+        op <- par(mfrow = c(1,2))
         #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1

### 2022-02-04 Analysis/03.1-results.Rmd
adding the snippet for the plotting data

@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
 ## Run the analyses between groups
@@
 
-```{r, echo = FALSE, eval = FALSE}
-## Getting the plotting data
-ploting_data <- list("median_elaborations" = apply(get.disparity(get.subsets(super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
-                     "median_innovations"  = apply(get.disparity(get.subsets(super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
-                     "distances_to_centre" =  get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
-                     "consensus_tree"      = tree)
-
-## Save all that as a list
-save(ploting_data, file = "ploting_data.rda")
-```
-
 ```{r, echo = FALSE}
@@
 rm(shapespace_allbirds_lvl_superorder_order_results_list)
+```
+
+```{r, echo = FALSE, eval = FALSE}
+## Getting the plotting data
+ploting_data <- list("median_elaborations" = apply(get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "median_innovations"  = apply(get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "distances_to_centre" =  get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
+                     "consensus_tree"      = tree)
+
+## Save all that as a list
+save(ploting_data, file = "ploting_data.rda")
 ```

### 2022-02-10 Analysis/09-Figure3.Rmd
Create 09-Figure3.Rmd

--- 2022-02-04 Analysis/03.1-results.Rmd
@@
 
-First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):
-
 ```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
@@
 colour.palette <- gg.color.hue
-
-## Plotting just the space
-covar.plot(superorder_shapespace,
-           main = "Shapespace",
-           cex = 0.5,
-           legend = TRUE,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           legend.args = list(x = "bottomright", cex = 2/3))
 ```
 
-Figure: The global shape space
-
-We can then check the distribution of the variance covariance matrices for each super order and each order.
-The VCV matrices are represented as average ellipses or average axis.
-
-```{r, echo = FALSE, fig.height = 14, fig.width = 14, eval = TRUE}
-par(mfrow = c(2,2))
-## Plotting mean axes
-covar.plot(superorder_shapespace,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Average ellipses by super order",
-           ellipses = mean,
-           legend = FALSE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-## Plotting mean ellipses
-covar.plot(superorder_shapespace,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Average major axes by super order",
-           major.axes = mean,
-           legend = TRUE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-covar.plot(order_shapespace,
-           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
-           main = "Average ellipses by order",
-           ellipses = mean,
-           legend = FALSE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-## Plotting mean ellipses
-covar.plot(order_shapespace,
-           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
-           main = "Average major axes by order",
-           major.axes = mean,
-           legend = TRUE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-```
-
-Main axes and ellipses visualisation for each superorders and orders.
-
-### Are these 2D average ellipses and/or axes an accurate representation of what's going on? 
-<details>
-  <summary>[Click to expand] TL;DR: It depends...
-</summary>
-
-A couple of things have to be taken into account:
-
- * First, the averaging of the VCV (and it's ellipse and/or maon axis) can result into weird "middle" VCV that are not representing the individual VCV (e.g.
-if half VCV are negative, and half positive, the average will be probably 0; which does not exist).
- * Second, the ellipses in nD projected onto a 2D space will be deformed (unless the VCV is a real pancake on the first 2D, i.e. with no variance on the other dimensions).
-
-Note that both problems are only visual though since: 1) we don't use averages anywhere (but the full distributions all the time) and 2) the linear algebra bit with the projections and stuff works in any number of dimensions and transforms the results into 1D results (i.e.
-the projection/rejection of groups/species on the major axes in 8D space).
-
-Here's a visualisation of how the variance in VCVs affects the averaging artefact:
-
-```{r, eval = TRUE, echo = FALSE}
-visualisation.group <- function(data, subset, dimensions, n = 100) {
-
-    ## Get the subset
-    sub <- get.subsets(data, subsets = subset)
-
-    ## Make 2D
-    if(!missing(dimensions)) {
-        sub$matrix[[1]] <- sub$matrix[[1]][,dimensions]
-        sub$call$dimensions <- dimensions
-        fun2d <- function(covar) {
-            covar$VCV <- covar$VCV[dimensions, dimensions]
-            covar$loc <- covar$loc[dimensions]
-            return(covar)
-        }
-        sub$covar <- lapply(sub$covar, lapply, fun2d)
-        main_D <- "2D"
-    } else {
-        main_D <- "8D"
-    }
-
-    ## Make the tile
-    main <- paste0(n, " random ", subset, " VCV (", main_D, ")")
-
-    ## Plot
-    covar.plot(sub, major.axes = TRUE, ellipses = TRUE, points = FALSE, legend = FALSE, main = main)
-    covar.plot(sub, major.axes = mean, ellipses = mean, points = FALSE, col = "red", add = TRUE)
-    rm(sub)
-}
-
-visualisation.group(superorder_shapespace, "Mirandornithes")
-```
-
-```{r, echo = FALSE, eval = TRUE}
-visualisation.group(superorder_shapespace, "Strisores")
-```
-
-Or even more obvious (with an direction that does not really exist (due to averaging)):
-
-```{r, echo = FALSE, eval = TRUE}
-visualisation.group(superorder_shapespace, "Paleognathae")
-```
-
-For the 2Dness, the effect is a bit less obvious (because the two first dimensions already have so much variance).
-Here is an example for the Columbimorphae.
-(to check different groups, you can easily modify the source code of the Rmd.
-TODO: this would be good in some shiny app innovation thingy).
-
-```{r, echo = FALSE, eval = TRUE, fig.height = 6, fig.width = 12}
-par(mfrow = c(1,2))
-visualisation.group(superorder_shapespace, "Columbimorphae")
-visualisation.group(superorder_shapespace, "Columbimorphae", dimensions = c(1,2))
-```
-
-Back to the main story.
-</details>
-
-## Calculate the elaborations/innovations
-
-We then calculate the projections/rejections as explained in more details in the [03-elaboration_exploration_analyses.Rmd]() script.
-Really briefly we calculate three types of projections/rejections abbreviated as follows below:
-
- * `group on phylogeny`: the projection of each group's VCV (major axis) on the phylogenetic's VCV (major axis)
- * `species on phylogeny`: the projection of each element within a group (species) onto the phylogenetic's VCV (major axis)
- * `species on group`: the projection of each element within a group (species) onto its own VCV (major axis)
-
-<!-- TODO: add the pictures of the different tests here -->
-
-TODO: centre and absolute directly in `dispRity.covar.projections`
-
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
-## Run the analyses between groups
-super_results_between <- dispRity.covar.projections(superorder_shapespace,
-                                             type = "groups",
-                                             base = "phylogeny",
-                                             verbose = TRUE,
-                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses between groups
-order_results_between <- dispRity.covar.projections(order_shapespace,
-                                             type = "groups",
-                                             base = "phylogeny",
-                                             verbose = TRUE,
-                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses of groups on the phylogeny
-super_results_phylo <- dispRity.covar.projections(superorder_shapespace,
-                                            type = "elements",
-                                            base = "phylogeny",
-                                            inc.base = TRUE,
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-
-## Run the analyses of groups on the phylogeny
-order_results_phylo <- dispRity.covar.projections(order_shapespace,
-                                            type = "elements",
-                                            base = "phylogeny",
-                                            inc.base = TRUE,
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses groups on themselves
-super_results_within <- dispRity.covar.projections(superorder_shapespace,
-                                            type = "elements",
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses groups on themselves
-order_results_within <- dispRity.covar.projections(order_shapespace,
-                                            type = "elements",
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-shapespace_allbirds_lvl_superorder_order_results_list <- list(
-        "super_results_between" = super_results_between,
-        "order_results_between" = order_results_between,
-        "super_results_phylo"   = super_results_phylo,
-        "order_results_phylo"   = order_results_phylo,
-        "super_results_within"  = super_results_within,
-        "order_results_within"  = order_results_within)
-save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
-```
-
-```{r, echo = FALSE, eval = FALSE}
-## Getting the plotting data
-ploting_data <- list("median_elaborations" = apply(get.disparity(get.subsets(super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
-                     "median_innovations"  = apply(get.disparity(get.subsets(super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
-                     "distances_to_centre" =  get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
-                     "consensus_tree"      = tree)
-
-## Save all that as a list
-save(ploting_data, file = "ploting_data.rda")
-```
+Loading the elaboration exploration results (from 03_results)
 
@@
 
-### Projections of groups on the phylogeny:
+## Figure 3
 
-```{r "starting with boxplots", echo = FALSE, fig.height = 8, fig.width = 8}
-## Plot the results
-wrap.plot <- function(results, col, main, names, order = TRUE, xlim = NULL) {
-
-    ## Ranking the results
-    if(order) {
-        ## Get the ranking
-        orders <- lapply(results, get.disparity, concatenate = FALSE)
-        ## Get the mean values
-        orders_pos <- orders$position
-        orders_dis <- orders$distance
-
-        ## Get the mean for each and combine into a table
-        cent.tend <- mean
-        means <- cbind(unlist(lapply(orders_pos, cent.tend)), unlist(lapply(orders_dis, cent.tend)))
-
-        ## Sum them and rank them
-        sum_means <- apply(means, 1, sum)
-        rank <- order(sum_means, decreasing = FALSE)
-    } else {
-        rank <- 1:length(results[[1]]$disparity)
-    }
-
-    ## Adding one plot
-    one.plot <- function(one_result, xlab, col, zero = TRUE, main, centre, absolute, rank = NULL, xlim = NULL, ...) {
-        values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
-        if(!is.null(rank)) {
-            values <- values[rank]
-        }
-        if(!is.null(xlim)) {
-            xlim <- xlim
-        } else {
-            xlim <- range(values)
-        }
-
-        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main = main, ylim = xlim, ...)
-        if(zero) {
-            abline(v = 0, col = "black", lty = 3, lwd = 1)
-        }
-    }
-
-    ## How many sub plots?
-    n <- length(results)
-
-    ## Recycle labels?
-    if(missing(names)) {
-        names_list <-names(results)
-    } else {
-        names_list <- names
-    }
-
-    ## Which will be the centre plot?
-    centre <- ceiling(n/2)
-    ## Plot everything
-    op <- par(mfrow = c(1, n+1))
-    ## Plot the first panel
-    par(bty = "n", mar = c(5, 14, 4, 0) + 0.1)
-    empty <- matrix(0, ncol = length(results[[1]]$disparity))
-    colnames(empty) <- names(results[[1]]$disparity)[rank]
-    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
-    for(i in 1:n) {
-        ## Is position?
-        is_position <- names(results)[[i]] == "position"
-        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
-        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
-    }
-    par(op)
-}
-wrap.plot(results$super_results_between,
-              col = colour.palette(n.subsets(superorder_shapespace)-1),
-              main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
-              names = c("elaboration", "innovation"))
-```
-
-Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes.
-For each panel, the highest the score the more the clade is elaborating, exploring, exploring perpendicular to the main axes for the position, distance and degree respectively.
-
-<details>
-  <summary>[Click to expand] Same but zooming on the (0,5) interval (to get less of the flamingo effect)
-</summary>
-
-```{r, echo = FALSE, fig.height = 8, fig.width = 8}
-wrap.plot(results$super_results_between,
-          xlim = c(0,6),
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV (ZOOM)",
-          names = c("elaboration", "innovation"))
-```
-</details>
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-wrap.plot(results$order_results_between,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-<details>
-  <summary>[Click to expand] Again, same but with some zoom (0,15)
-</summary>
-
-```{r, echo = FALSE, fig.height = 16, fig.width = 8}
-wrap.plot(results$order_results_between,
-          xlim = c(0, 15),
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV (ZOOM)",
-          names = c("elaboration", "innovation"))
-```
-</details>
-
-### Projections of the species on the phylogenetic VCV
-
-Similarly to the analyses described above, we ran the analyses for each species.
-Here, instead of projecting the major axes of the groups onto the global phylogenetic major axis we projected each species onto the phylogenetic major axis (for each posterior distribution).
-
-```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
-## Plot the results
-wrap.plot(results$super_results_phylo,
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$order_results_phylo,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-### Projections of the species on their own group's VCV
-
-Finally, we ran the same analyses but instead of the projection of the species onto the global phylogenetic major axes, we projected them onto their respective group's phylogenetic major axes.
-
-```{r, fig.height = 8, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$super_results_within,
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[species on group] superorders species\nonto their group's VCV",
-          names = c("elaboration", "innovation"))
-```
-
-Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$order_results_within,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[species on group] orders species\nonto their group's VCV",
-          names = c("elaboration", "innovation"))
-```
-
-So that's the story so far...
-
-# Different scores distributions per species
-
-```{r, eval = TRUE, echo = FALSE, eval = TRUE}
-library(hdrcde)
-
-#@param res the results list
-#@param type which type (position, distance or angle)
-#@param cent.tend which central tendency fun (default is median)
-#@param use.hdr whether to use kernel density estimate for the outputs (default is TRUE)
-#@param prob which quantiles to use (default is c(50, 95, 99))
-#@param exclude.phylo remove the phylogeny subset (default = FALSE)
-#@param add.relative whether to add the relative results (as rel.pos and rel.dis) or not
-#@param ... plot options
-get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, add.relative = FALSE, ...) {
-
-    ## Remove the phylo one (if needed)
-    if(exclude.phylo) {
-        select <- names(size.subsets(res[[type]]))
-        select <- select[-which(select %in% "phylogeny")]
-        res[[type]] <- get.subsets(res[[type]], subsets = select)
-    }
-    ## Extracting all the results
-    disparity <- lapply(res[[type]]$disparity, function(X, type) return(X$elements))
-
-    ## Measure the central tendencies
-    centrals <- lapply(disparity, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-    ## Number of groups
-    n_groups <- length(centrals)
-
-    ## Calculate and plot the hdrcde
-    op_tmp <- par(mfrow = c(ceiling(sqrt(n_groups)), round(sqrt(n_groups))))
-    densities <- list()
-    for(i in 1:length(centrals)) {
-        if(use.hdr) {
-            densities[[i]] <- hdr.den(centrals[[i]], prob = prob, main = names(centrals)[i])#, ...)
-        } else {
-            hist(centrals[[i]], main = names(centrals)[i], ...)
-        }
-    }
-    par(op_tmp)
-
-    ## Find the extreme species
-    get.extremes.hdr <- function(density, central) {
-        ## Get the minimums and maximums
-        mins <- which(central <= density$hdr[1,1])
-        maxs <- which(central >= density$hdr[1, ncol(density$hdr)])
-        ## Get their values
-        return(c(central[names(mins)], central[names(maxs)]))
-    }
-    CI.converter <- function(CI) {
-        sort(c(50-CI/2, 50+CI/2)/100)
-    }
-    get.extremes.quantile <- function(central, prob) {
-        ## Get the quantiles
-        quants <- quantile(central, probs = CI.converter(max(prob)))
-        ## Get the minimums and maximums
-        mins <- which(central <= quants[1])
-        maxs <- which(central >= quants[2])
-        ## Get their values
-        return(c(central[names(mins)], central[names(maxs)]))
-    }
-
-    if(use.hdr) {
-        output <- mapply(get.extremes.hdr, densities, centrals, SIMPLIFY = FALSE)
-        names(output) <- names(centrals)
-    } else {
-        output <- lapply(centrals, get.extremes.quantile, prob = prob)
-    }
-
-    ## Return the as a table
-    for(i in 1:length(output)) {
-        if(length(output[[i]]) == 0) {
-            output[[i]] <- data.frame("level" = names(output)[i], "species" = NA, "score" = NA)
-        } else {
-            output[[i]] <- data.frame("level" = rep(names(output)[i], length(output[[i]])),
-                                      "species" = names(output[[i]]),
-                                      "score"   = unname(output[[i]]))
-        }
-    }
-
-    ## Add the relative results
-    if(add.relative) {
-        make.relative <- function(X) {
-            cbind(X, "rel" = ifelse(is.na(X$score), NA, X$score/max(X$score)))
-        }
-        output <- lapply(output, make.relative)
-    }
-    rm(disparity)
-    return(output)
-}
-combine.results <- function(res1, res2, sort.results = TRUE) {
-    ## Combine the results
-    res_out <- rbind(res1[,1:2], res2[, 1:2])
-    res_out$elaboration <- c(res1[, 3], rep(NA, nrow(res2)))
-    if("rel" %in% colnames(res1)) {
-        res_out$rel.elab <- c(res1[, 4], rep(NA, nrow(res2)))
-    }
-    res_out$innovation <- c(rep(NA, nrow(res1)), res2[, 3])
-    if("rel" %in% colnames(res1)) {
-        res_out$rel.inno <- c(rep(NA, nrow(res1)), res2[, 4])
-    }
-
-    # Remove NA species
-    if(any(is_na <- is.na(res_out$species))) {
-        if(all(is_na)) {
-            return(res_out)
-        } else {
-            res_out <- res_out[!is_na, , drop = FALSE]
-        }
-    }
-
-    ## Find any species in commons
-    commons <- intersect(res1$species, res2$species)
-    if(length(commons) > 0) {
-        if(any(is.na(commons))) {
-            if(is.na(res1$species)) {
-                if(is.na(res2$species)) {
-                    ## Both are NA (keep just the first one)
-                    res_out <- res_out[-1, , drop = FALSE]
-                } else {
-                    ## Remove res1
-                    res_out <- res_out[-1, , drop = FALSE]
-                }
-            } else {
-                ## Remove res2
-                res_out <- res_out[1:length(res1$species), , drop = FALSE]
-            }
-        } else {
-            ## Recursively remove the ones in common
-            while(length(commons) > 0) {
-                duplicates <- which(res_out$species == commons[1])
-                columns <- if("rel" %in% colnames(res1)){c(5,6)} else {4}
-                res_out[duplicates[1], columns] <- res_out[duplicates[2], columns]
-                res_out <- res_out[-duplicates[2], ]
-                commons <- commons[-1]
-            }
-        }
-    }
-
-    if(sort.results) {
-        sums <- apply(res_out[, -c(1,2)], 1, sum, na.rm = TRUE)
-        rank <- match(sort(sums, decreasing = TRUE), sums)
-    } else {
-        rank <- 1:nrow(res_out)
-    }
-
-    return(res_out[rank,])
-}
-```
-
-Furthermore, we can look at things species per species to find which species species have the most extreme values (in general, or within their groups).
-
-### Extreme species per super order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position")
-```
-Figure: elaborations distributions per super orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_phylo <- get.distributions(results$super_results_phylo, type = "distance")
-```
-Figure: innovations distributions per super orders
-
-```{r, "printing results in text", echo = FALSE, eval = TRUE}
-## Combining the results
-super_phylo_table <- do.call(rbind, mapply(combine.results, super_positions_phylo, super_distances_phylo, SIMPLIFY = FALSE))
-## Get the top 10%
-get.top <- function(data, what, top = 0.9, relative = FALSE) {
-    if(any(data[, "level"] == "phylogeny")) {
-        data_top <- data[which(data[, "level"] != "phylogeny"), ]
-    } else {
-        data_top <- data
-    }
-
-    ## Get the relative column
-    if(relative) {
-        what <- switch(what,
-                       "elaboration" = "rel.elab",
-                       "innovation"  = "rel.inno")
-    }
-    ## Get the top 10(ish) species
-    data_top_10 <- data_top[which(data_top[, what] >= quantile(data_top[, what], prob = top, na.rm = TRUE)), ]
-    return(data_top_10[order(data_top_10[, what], decreasing = TRUE), ])
-}
-top_elabo <- get.top(super_phylo_table, what = "elaboration")
-top_innov <- get.top(super_phylo_table, what = "innovation")
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by superorder:
-
-The 10% top elaborators by superorder are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
-
-The 10% top innovators by superorder are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto the phylogenetic VCV.
-</summary>
-
-```{r, echo = FALSE}
-library(knitr)
-rownames(super_phylo_table) <- NULL
-kable(super_phylo_table, digits = 3)
-```
-
-</details>
-
-### Extreme species per order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
-```
-Figure: elaborations distributions per order
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_phylo <- get.distributions(results$order_results_phylo, type = "distance", exclude.phylo = TRUE)
-```
-Figure: innovations distributions per order
-
-```{r, echo = FALSE}
-order_phylo_table <- do.call(rbind, mapply(combine.results, order_positions_phylo, order_distances_phylo, SIMPLIFY = FALSE))
-top_elabo <- get.top(order_phylo_table, what = "elaboration")
-top_innov <- get.top(order_phylo_table, what = "innovation")
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by order:
-The 10% top ellaborators by order are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
-The 10% top innovators by order are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(order_phylo_table) <- NULL
-kable(order_phylo_table, digits = 3)
-```
-</details>
-
-### Extreme species per super order (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE,  add.relative = TRUE)
-```
-Figure: elaborations distributions within super orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_within <- get.distributions(results$super_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: innovations distributions within super orders
-
-```{r, echo = FALSE}
-super_phylo_table_within <- do.call(rbind, mapply(combine.results, super_positions_within, super_distances_within, SIMPLIFY = FALSE))
-top_elabo <- get.top(super_phylo_table_within, what = "elaboration", relative = TRUE)
-top_innov <- get.top(super_phylo_table_within, what = "innovation", relative = TRUE)
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by superorder relative to each superorder (i.e. the best one(s) in each order):
-```{r, echo = FALSE}
-combined <- rbind(top_elabo, top_innov)
-rownames(combined) <- NULL  
-kable(combined, digits = 3)
-```
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(super_phylo_table_within) <- NULL
-kable(super_phylo_table_within, digits = 3)
-```
-
-</details>
-
-### Extreme species per order (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: elaborations distributions within orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_within <- get.distributions(results$order_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: innovations distributions within orders
-
-```{r, echo = FALSE}
-order_phylo_table_within <- do.call(rbind, mapply(combine.results, order_positions_within, order_distances_within, SIMPLIFY = FALSE))
-top_elabo <- get.top(order_phylo_table_within, what = "elaboration", relative = TRUE)
-top_innov <- get.top(order_phylo_table_within, what = "innovation", relative = TRUE)
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by order relative to each order (i.e. the best one(s) in each order):
-```{r, echo = FALSE}
-combined <- rbind(top_elabo, top_innov)
-rownames(combined) <- NULL  
-kable(combined, digits = 3)
-```
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(order_phylo_table_within) <- NULL
-kable(order_phylo_table_within, digits = 3)
-```
-</details>
-
-## Correlations between elaboration and innovation?
+Density plot of the correlation of the median elaboration and median exploration for all species.
 
@@
 
-### Correlations per super order (species on phylogeny)
+Correlations per super order (species on phylogeny):
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
+```{r, eval = TRUE, echo = FALSE}
+## Calculate the correlations for the projections onto the phylo axis
+super_phylo_cor <- get.correlations(results$super_results_phylo)
+
+## Calculate the correlations for the projections onto the group' axes
+super_group_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
+```
+
+### Figure 3a - gg_version
+
+Density plot of the correlation of the median elaboration and median exploration for all species.
+
+```{r, echo = FALSE, eval = FALSE}
+## Getting the data to plot
+global_data <- super_phylo_cor$phylogeny
+
+gg.density.plot <- function(data, palette = "viridis") {
+    ggplot(data.frame("elaboration" = data$x, "exploration" = data$y), aes(x=elaboration, y=exploration) ) +
+      geom_hex() +
+      scale_fill_continuous(type = palette) +
+      theme_bw()
+}
+
+gg.density.plot(global_data)
+```
+
+## Figure 3b
+
+```{r, echo = FALSE, eval = FALSE}
+
+one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, ...) {
+
+    dots <- list(...)
+
+    if(missing(col.sub)) {
+        col.sub <- NULL
+    }
+
+    ## Subbing the colours
+    if(!is.null(col.sub)) {
+        ## Reset the levels
+        col_sub <- as.factor(as.character(col.sub))
+        ## Make the empty level grey
+        if(any(levels(col_sub) == "")) {
+            col_avail <- c("grey", col.fun(length(levels(col_sub))-1))
+        } else {
+            col_avail <- col.fun(length(levels(col_sub)))
+        }
+        ## Create the colour vector
+        col_vector <- col_avail[as.numeric(col_sub)]
+    } else {
+        if(is.null(dots$col)) {
+            col_vector <- "black"
+        } else {
+            col_vector <- dots$col
+        }
+    }
+
+    # if(!missing(tree)) {
+    #     op <- par(mfrow = c(1,2))
+    #     #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1
+    #     par(mar = c(5, 4, 4, 0)+0.1)   
+    # }
+
+    ## Plotting the results
+    plot(cor.results$x, cor.results$y, pch = 19, xlab = ifelse(), ylab = ifelse(ID == 1 || ID == 5, "exploration", ""), main = cor.results$main, cex = pt.cex, col = col_vector)
+
+    if(!is.null(col.sub)) {
+        ## Adding the legend
+        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
+    }
+
+    return(invisible())
+}
+
+par(mfrow = c(2, 4))
+lims <- list(x = range(global_data$x),
+             y = range(global_data$y))
+
 ## Getting the subsets for the colouring
 col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
-## Calculate the correlations
-super_phylo_cor <- get.correlations(results$super_results_phylo)
+
 ## Set the vector of legend positions
 legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
-for(i in 1:length(super_phylo_cor)) {
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
+
+for(i in 1:8) {
+    plot.correlations(super_phylo_cor[[i]], xlim = lims$x, ylim = lims$x, tree = NULL, col.sub = col_subs[[i]], legend.pos = legends_pos[i])
 }
+
 ```
-
-<details>
-  <summary>[Click to expand] Finding corner cases groups.
-</summary>
-
-So here's how you remove exploration and elaboration: by centering these three species on the centre of the phylo axis:
-
-Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
-# colour.palette <- rainbow
-posteriors <- shapespace_allbirds_lvl_superorder_order
-
-# plot(superorder_shapespace$matrix[[1]][, 1:2], pch = 19, cex = 0.5)
-# identify(superorder_shapespace$matrix[[1]][, 1:2])
-# shapespace[c(268, 286, 80), c(9, 10)]
-
-## Moving these three species around: Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
-fixed_points <- c(268, 286, 408)
-move_around <- c(8749, 8750, 8751)
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Get the major axis
-phylo_axis <- axis.covar(superorder_shapespace, sample = median)$phylogeny[[1]]
-## Point with no elaboration/exploration
-no_nothing <- 0
-
-## Example with no exploration/elaboration
-skewed_space[move_around, c(1:8)] <- no_nothing
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny"))
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
-           main = "elements without\nelaboration innovation")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-So easy right? We just need to change them along the PC axis to move them around in space.
-
-No, it's not fucking easy because it's a 8 dimensional problem...
-Even though the two first axes cumulate 86.27% of the variance, fixing the three elements on the two elements and centering them on the 6 other dimensions (13.73% of the variance) results in BIG changes:
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Example with no exploration/elaboration
-skewed_space[move_around, c(3:8)] <- 0
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
-           main = "Centring only on dimensions 3:8")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-For example, the blue element, Anas_falcata is moved to no elaboration/exploration at all if we do that ignoring data past the second axis.
-There's a duck axis lurking in the depth... (spoiler, it's the fourth dimension).
-
-Let's flatten the problem in 2D for simplification:
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-
-## 2Ding the space
-dimensions <- c(1,2)
-superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
-superorder_skewed_space$call$dimensions <- dimensions
-fun2d <- function(covar) {
-    covar$VCV <- covar$VCV[dimensions, dimensions]
-    covar$loc <- covar$loc[dimensions]
-    return(covar)
-}
-superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Flattening the results")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-abline(a = 0, b = 7, col = "grey")
-abline(a = 0, b = 0.14, col = "grey")
-```
-
-So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.14 and 7).
-Let's see if it's something birdy or not by modifying these elements position in trait space to have elaboration of 0 and innovation of 1 and the opposite.
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Move around the first species to have elaboration = 0 and innovation = 0
-skewed_space[move_around[1], c(1:2)] <- 0
-## Move around the second species to have elaboration = 0 and innovation = 1
-skewed_space[move_around[2], c(1:2)] <- c(0, 0.4)
-## Move around the third species to have elaboration = 1 and innovation = 0
-skewed_space[move_around[3], c(1:2)] <- c(0.4, 0)
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-
-## 2Ding the space
-dimensions <- c(1,2)
-superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
-superorder_skewed_space$call$dimensions <- dimensions
-fun2d <- function(covar) {
-    covar$VCV <- covar$VCV[dimensions, dimensions]
-    covar$loc <- covar$loc[dimensions]
-    return(covar)
-}
-superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Flattening the results")
-## Add the phylogenetic axis
-lines(phylo_axis[,1:2], lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis[,1:2], lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-TODO: fix this!
-
-</details>
 

### 2022-02-10 Analysis/09-Figure3.Rmd
ugh...

@@
 ```{r, echo = FALSE, eval = FALSE}
-
 one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, ...) {
@@
     ## Plotting the results
-    plot(cor.results$x, cor.results$y, pch = 19, xlab = ifelse(), ylab = ifelse(ID == 1 || ID == 5, "exploration", ""), main = cor.results$main, cex = pt.cex, col = col_vector)
+
+    ## Selecting the x and y axis base on plot ID
+    select.x <- function(ID, yes = "", no = "") {
+        ifelse(ID %in% c(5:8), yes, no)
+    }
+    select.y <- function(ID, yes = "", no = "") {
+        ifelse(ID == 1 || ID == 5, yes, no)
+    }
+
+    par(mar = c(select.x(ID, 4, 1), select.y(ID, 4, 1), 3, 1)+0.1)
+    plot(cor.results$x, cor.results$y, pch = 19,
+        xlab = select.x(ID, "elaboration"),
+        ylab = select.y(ID, "explortation"),
+        main = strsplit(cor.results$main, split = ":")[[1]][1],
+        cex = pt.cex, col = col_vector,
+        ...)
 
     if(!is.null(col.sub)) {
+        legend_text <- levels(col_sub)
+        if(length(empty <- which(legend_text == "")) > 0) {
+            legend_text[empty] <- "other"
+        }
         ## Adding the legend
-        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
+        legend(x = legend.pos, legend = legend_text, pch = 19, cex = legend.cex, col = col_avail, bg = "white")
     }
@@
 par(mfrow = c(2, 4))
-lims <- list(x = range(global_data$x),
-             y = range(global_data$y))
-
 ## Getting the subsets for the colouring
@@
 for(i in 1:8) {
-    plot.correlations(super_phylo_cor[[i]], xlim = lims$x, ylim = lims$x, tree = NULL, col.sub = col_subs[[i]], legend.pos = legends_pos[i])
+    one.plot.correlations(super_phylo_cor[[i]],col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.8, legend.cex = 0.8)
 }
+```
 
+## Figure 3c
+
+```{r, echo = FALSE, eval = FALSE}
+par(mfrow = c(2, 4))
+## Getting the subsets for the colouring
+col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
+
+## Set the vector of legend positions
+legends_pos <- c("bottomright", "topright", "topright", "topleft", "topleft", "topright", "bottomright", "bottomright")
+
+for(i in 1:8) {
+    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.8, legend.cex = 0.8)
+}
 ```

### 2022-02-11 Analysis/10-Figure1.Rmd
Create 10-Figure1.Rmd

--- 2022-01-31 Analysis/03.1-results.Rmd
@@
 colour.palette <- gg.color.hue
-
-## Plotting just the space
-covar.plot(superorder_shapespace,
-           main = "Shapespace",
-           cex = 0.5,
-           legend = TRUE,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           legend.args = list(x = "bottomright", cex = 2/3))
 ```
 
-Figure: The global shape space
+Plotting the tree
 
-We can then check the distribution of the variance covariance matrices for each super order and each order.
-The VCV matrices are represented as average ellipses or average axis.
+```{r}
+tree <- ladderize(tree)
+## Get the tip_values
+orders_lvl <- shapespace$level2
+superorders_lvl <- shapespace$level1
 
-```{r, echo = FALSE, fig.height = 14, fig.width = 14, eval = TRUE}
-par(mfrow = c(2,2))
-## Plotting mean axes
-covar.plot(superorder_shapespace,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Average ellipses by super order",
-           ellipses = mean,
-           legend = FALSE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-## Plotting mean ellipses
-covar.plot(superorder_shapespace,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Average major axes by super order",
-           major.axes = mean,
-           legend = TRUE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-covar.plot(order_shapespace,
-           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
-           main = "Average ellipses by order",
-           ellipses = mean,
-           legend = FALSE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
-## Plotting mean ellipses
-covar.plot(order_shapespace,
-           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
-           main = "Average major axes by order",
-           major.axes = mean,
-           legend = TRUE,
-           points = FALSE,
-           cex = 0.5, apply.to.VCV = TRUE)
+## Replace the order levels as colours
+coloured_levels <- orders_lvl
+levels(coloured_levels) <- colour.palette(length(levels(coloured_levels)))
+
+plot(tree, show.tip.label = FALSE, edge.color =  match.tip.edge(as.character(coloured_levels), tree, replace.na = "grey"))
 ```
 
-Main axes and ellipses visualisation for each superorders and orders.
+Plotting the ellipses
 
-### Are these 2D average ellipses and/or axes an accurate representation of what's going on? 
-<details>
-  <summary>[Click to expand] TL;DR: It depends...
-</summary>
+```{r}
+plot.one.ellipse <- function(data, ID, colours) {
 
-A couple of things have to be taken into account:
-
- * First, the averaging of the VCV (and it's ellipse and/or maon axis) can result into weird "middle" VCV that are not representing the individual VCV (e.g.
-if half VCV are negative, and half positive, the average will be probably 0; which does not exist).
- * Second, the ellipses in nD projected onto a 2D space will be deformed (unless the VCV is a real pancake on the first 2D, i.e. with no variance on the other dimensions).
-
-Note that both problems are only visual though since: 1) we don't use averages anywhere (but the full distributions all the time) and 2) the linear algebra bit with the projections and stuff works in any number of dimensions and transforms the results into 1D results (i.e.
-the projection/rejection of groups/species on the major axes in 8D space).
-
-Here's a visualisation of how the variance in VCVs affects the averaging artefact:
-
-```{r, eval = TRUE, echo = FALSE}
-visualisation.group <- function(data, subset, dimensions, n = 100) {
-
-    ## Get the subset
-    sub <- get.subsets(data, subsets = subset)
-
-    ## Make 2D
-    if(!missing(dimensions)) {
-        sub$matrix[[1]] <- sub$matrix[[1]][,dimensions]
-        sub$call$dimensions <- dimensions
-        fun2d <- function(covar) {
-            covar$VCV <- covar$VCV[dimensions, dimensions]
-            covar$loc <- covar$loc[dimensions]
-            return(covar)
-        }
-        sub$covar <- lapply(sub$covar, lapply, fun2d)
-        main_D <- "2D"
-    } else {
-        main_D <- "8D"
 }
 
-    ## Make the tile
-    main <- paste0(n, " random ", subset, " VCV (", main_D, ")")
-
-    ## Plot
-    covar.plot(sub, major.axes = TRUE, ellipses = TRUE, points = FALSE, legend = FALSE, main = main)
-    covar.plot(sub, major.axes = mean, ellipses = mean, points = FALSE, col = "red", add = TRUE)
-    rm(sub)
-}
-
-visualisation.group(superorder_shapespace, "Mirandornithes")
 ```
-
-```{r, echo = FALSE, eval = TRUE}
-visualisation.group(superorder_shapespace, "Strisores")
-```
-
-Or even more obvious (with an direction that does not really exist (due to averaging)):
-
-```{r, echo = FALSE, eval = TRUE}
-visualisation.group(superorder_shapespace, "Paleognathae")
-```
-
-For the 2Dness, the effect is a bit less obvious (because the two first dimensions already have so much variance).
-Here is an example for the Columbimorphae.
-(to check different groups, you can easily modify the source code of the Rmd.
-TODO: this would be good in some shiny app innovation thingy).
-
-```{r, echo = FALSE, eval = TRUE, fig.height = 6, fig.width = 12}
-par(mfrow = c(1,2))
-visualisation.group(superorder_shapespace, "Columbimorphae")
-visualisation.group(superorder_shapespace, "Columbimorphae", dimensions = c(1,2))
-```
-
-Back to the main story.
-</details>
-
-## Calculate the elaborations/innovations
-
-We then calculate the projections/rejections as explained in more details in the [03-elaboration_exploration_analyses.Rmd]() script.
-Really briefly we calculate three types of projections/rejections abbreviated as follows below:
-
- * `group on phylogeny`: the projection of each group's VCV (major axis) on the phylogenetic's VCV (major axis)
- * `species on phylogeny`: the projection of each element within a group (species) onto the phylogenetic's VCV (major axis)
- * `species on group`: the projection of each element within a group (species) onto its own VCV (major axis)
-
-<!-- TODO: add the pictures of the different tests here -->
-
-TODO: centre and absolute directly in `dispRity.covar.projections`
-
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
-## Run the analyses between groups
-super_results_between <- dispRity.covar.projections(superorder_shapespace,
-                                             type = "groups",
-                                             base = "phylogeny",
-                                             verbose = TRUE,
-                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses between groups
-order_results_between <- dispRity.covar.projections(order_shapespace,
-                                             type = "groups",
-                                             base = "phylogeny",
-                                             verbose = TRUE,
-                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses of groups on the phylogeny
-super_results_phylo <- dispRity.covar.projections(superorder_shapespace,
-                                            type = "elements",
-                                            base = "phylogeny",
-                                            inc.base = TRUE,
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses of groups on the phylogeny
-order_results_phylo <- dispRity.covar.projections(order_shapespace,
-                                            type = "elements",
-                                            base = "phylogeny",
-                                            inc.base = TRUE,
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses groups on themselves
-super_results_within <- dispRity.covar.projections(superorder_shapespace,
-                                            type = "elements",
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-## Run the analyses groups on themselves
-order_results_within <- dispRity.covar.projections(order_shapespace,
-                                            type = "elements",
-                                            verbose = TRUE,
-                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
-shapespace_allbirds_lvl_superorder_order_results_list <- list(
-        "super_results_between" = super_results_between,
-        "order_results_between" = order_results_between,
-        "super_results_phylo"   = super_results_phylo,
-        "order_results_phylo"   = order_results_phylo,
-        "super_results_within"  = super_results_within,
-        "order_results_within"  = order_results_within)
-save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_centred.rda")
-```
-
-```{r, echo = FALSE}
-## Quick load
-# load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_uncentred.rda")
-# results_cent <- shapespace_allbirds_lvl_superorder_order_results_list
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_centred.rda")
-results <- shapespace_allbirds_lvl_superorder_order_results_list
-rm(shapespace_allbirds_lvl_superorder_order_results_list)
-```
-
-### Projections of groups on the phylogeny:
-
-```{r "starting with boxplots", echo = FALSE, fig.height = 8, fig.width = 8}
-## Plot the results
-wrap.plot <- function(results, col, main, names, order = TRUE, xlim = NULL) {
-
-    ## Ranking the results
-    if(order) {
-        ## Get the ranking
-        orders <- lapply(results, get.disparity, concatenate = FALSE)
-        ## Get the mean values
-        orders_pos <- orders$position
-        orders_dis <- orders$distance
-
-        ## Get the mean for each and combine into a table
-        cent.tend <- mean
-        means <- cbind(unlist(lapply(orders_pos, cent.tend)), unlist(lapply(orders_dis, cent.tend)))
-
-        ## Sum them and rank them
-        sum_means <- apply(means, 1, sum)
-        rank <- order(sum_means, decreasing = FALSE)
-    } else {
-        rank <- 1:length(results[[1]]$disparity)
-    }
-
-    ## Adding one plot
-    one.plot <- function(one_result, xlab, col, zero = TRUE, main, centre, absolute, rank = NULL, xlim = NULL, ...) {
-        values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
-        if(!is.null(rank)) {
-            values <- values[rank]
-        }
-        if(!is.null(xlim)) {
-            xlim <- xlim
-        } else {
-            xlim <- range(values)
-        }
-
-        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main = main, ylim = xlim, ...)
-        if(zero) {
-            abline(v = 0, col = "black", lty = 3, lwd = 1)
-        }
-    }
-
-    ## How many sub plots?
-    n <- length(results)
-
-    ## Recycle labels?
-    if(missing(names)) {
-        names_list <-names(results)
-    } else {
-        names_list <- names
-    }
-
-    ## Which will be the centre plot?
-    centre <- ceiling(n/2)
-    ## Plot everything
-    op <- par(mfrow = c(1, n+1))
-    ## Plot the first panel
-    par(bty = "n", mar = c(5, 14, 4, 0) + 0.1)
-    empty <- matrix(0, ncol = length(results[[1]]$disparity))
-    colnames(empty) <- names(results[[1]]$disparity)[rank]
-    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
-    for(i in 1:n) {
-        ## Is position?
-        is_position <- names(results)[[i]] == "position"
-        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
-        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
-    }
-    par(op)
-}
-wrap.plot(results$super_results_between,
-              col = colour.palette(n.subsets(superorder_shapespace)-1),
-              main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
-              names = c("elaboration", "innovation"))
-```
-
-Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes.
-For each panel, the highest the score the more the clade is elaborating, exploring, exploring perpendicular to the main axes for the position, distance and degree respectively.
-
-<details>
-  <summary>[Click to expand] Same but zooming on the (0,5) interval (to get less of the flamingo effect)
-</summary>
-
-```{r, echo = FALSE, fig.height = 8, fig.width = 8}
-wrap.plot(results$super_results_between,
-          xlim = c(0,6),
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV (ZOOM)",
-          names = c("elaboration", "innovation"))
-```
-</details>
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-wrap.plot(results$order_results_between,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-<details>
-  <summary>[Click to expand] Again, same but with some zoom (0,15)
-</summary>
-
-```{r, echo = FALSE, fig.height = 16, fig.width = 8}
-wrap.plot(results$order_results_between,
-          xlim = c(0, 15),
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV (ZOOM)",
-          names = c("elaboration", "innovation"))
-```
-</details>
-
-### Projections of the species on the phylogenetic VCV
-
-Similarly to the analyses described above, we ran the analyses for each species.
-Here, instead of projecting the major axes of the groups onto the global phylogenetic major axis we projected each species onto the phylogenetic major axis (for each posterior distribution).
-
-```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
-## Plot the results
-wrap.plot(results$super_results_phylo,
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$order_results_phylo,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-### Projections of the species on their own group's VCV
-
-Finally, we ran the same analyses but instead of the projection of the species onto the global phylogenetic major axes, we projected them onto their respective group's phylogenetic major axes.
-
-```{r, fig.height = 8, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$super_results_within,
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[species on group] superorders species\nonto their group's VCV",
-          names = c("elaboration", "innovation"))
-```
-
-Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$order_results_within,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[species on group] orders species\nonto their group's VCV",
-          names = c("elaboration", "innovation"))
-```
-
-So that's the story so far...
-
-# Different scores distributions per species
-
-```{r, eval = TRUE, echo = FALSE, eval = TRUE}
-library(hdrcde)
-
-#@param res the results list
-#@param type which type (position, distance or angle)
-#@param cent.tend which central tendency fun (default is median)
-#@param use.hdr whether to use kernel density estimate for the outputs (default is TRUE)
-#@param prob which quantiles to use (default is c(50, 95, 99))
-#@param exclude.phylo remove the phylogeny subset (default = FALSE)
-#@param add.relative whether to add the relative results (as rel.pos and rel.dis) or not
-#@param ... plot options
-get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, add.relative = FALSE, ...) {
-
-    ## Remove the phylo one (if needed)
-    if(exclude.phylo) {
-        select <- names(size.subsets(res[[type]]))
-        select <- select[-which(select %in% "phylogeny")]
-        res[[type]] <- get.subsets(res[[type]], subsets = select)
-    }
-    ## Extracting all the results
-    disparity <- lapply(res[[type]]$disparity, function(X, type) return(X$elements))
-
-    ## Measure the central tendencies
-    centrals <- lapply(disparity, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-    ## Number of groups
-    n_groups <- length(centrals)
-
-    ## Calculate and plot the hdrcde
-    op_tmp <- par(mfrow = c(ceiling(sqrt(n_groups)), round(sqrt(n_groups))))
-    densities <- list()
-    for(i in 1:length(centrals)) {
-        if(use.hdr) {
-            densities[[i]] <- hdr.den(centrals[[i]], prob = prob, main = names(centrals)[i])#, ...)
-        } else {
-            hist(centrals[[i]], main = names(centrals)[i], ...)
-        }
-    }
-    par(op_tmp)
-
-    ## Find the extreme species
-    get.extremes.hdr <- function(density, central) {
-        ## Get the minimums and maximums
-        mins <- which(central <= density$hdr[1,1])
-        maxs <- which(central >= density$hdr[1, ncol(density$hdr)])
-        ## Get their values
-        return(c(central[names(mins)], central[names(maxs)]))
-    }
-    CI.converter <- function(CI) {
-        sort(c(50-CI/2, 50+CI/2)/100)
-    }
-    get.extremes.quantile <- function(central, prob) {
-        ## Get the quantiles
-        quants <- quantile(central, probs = CI.converter(max(prob)))
-        ## Get the minimums and maximums
-        mins <- which(central <= quants[1])
-        maxs <- which(central >= quants[2])
-        ## Get their values
-        return(c(central[names(mins)], central[names(maxs)]))
-    }
-
-    if(use.hdr) {
-        output <- mapply(get.extremes.hdr, densities, centrals, SIMPLIFY = FALSE)
-        names(output) <- names(centrals)
-    } else {
-        output <- lapply(centrals, get.extremes.quantile, prob = prob)
-    }
-
-    ## Return the as a table
-    for(i in 1:length(output)) {
-        if(length(output[[i]]) == 0) {
-            output[[i]] <- data.frame("level" = names(output)[i], "species" = NA, "score" = NA)
-        } else {
-            output[[i]] <- data.frame("level" = rep(names(output)[i], length(output[[i]])),
-                                      "species" = names(output[[i]]),
-                                      "score"   = unname(output[[i]]))
-        }
-    }
-
-    ## Add the relative results
-    if(add.relative) {
-        make.relative <- function(X) {
-            cbind(X, "rel" = ifelse(is.na(X$score), NA, X$score/max(X$score)))
-        }
-        output <- lapply(output, make.relative)
-    }
-    rm(disparity)
-    return(output)
-}
-combine.results <- function(res1, res2, sort.results = TRUE) {
-    ## Combine the results
-    res_out <- rbind(res1[,1:2], res2[, 1:2])
-    res_out$elaboration <- c(res1[, 3], rep(NA, nrow(res2)))
-    if("rel" %in% colnames(res1)) {
-        res_out$rel.elab <- c(res1[, 4], rep(NA, nrow(res2)))
-    }
-    res_out$innovation <- c(rep(NA, nrow(res1)), res2[, 3])
-    if("rel" %in% colnames(res1)) {
-        res_out$rel.inno <- c(rep(NA, nrow(res1)), res2[, 4])
-    }
-
-    # Remove NA species
-    if(any(is_na <- is.na(res_out$species))) {
-        if(all(is_na)) {
-            return(res_out)
-        } else {
-            res_out <- res_out[!is_na, , drop = FALSE]
-        }
-    }
-
-    ## Find any species in commons
-    commons <- intersect(res1$species, res2$species)
-    if(length(commons) > 0) {
-        if(any(is.na(commons))) {
-            if(is.na(res1$species)) {
-                if(is.na(res2$species)) {
-                    ## Both are NA (keep just the first one)
-                    res_out <- res_out[-1, , drop = FALSE]
-                } else {
-                    ## Remove res1
-                    res_out <- res_out[-1, , drop = FALSE]
-                }
-            } else {
-                ## Remove res2
-                res_out <- res_out[1:length(res1$species), , drop = FALSE]
-            }
-        } else {
-            ## Recursively remove the ones in common
-            while(length(commons) > 0) {
-                duplicates <- which(res_out$species == commons[1])
-                columns <- if("rel" %in% colnames(res1)){c(5,6)} else {4}
-                res_out[duplicates[1], columns] <- res_out[duplicates[2], columns]
-                res_out <- res_out[-duplicates[2], ]
-                commons <- commons[-1]
-            }
-        }
-    }
-
-    if(sort.results) {
-        sums <- apply(res_out[, -c(1,2)], 1, sum, na.rm = TRUE)
-        rank <- match(sort(sums, decreasing = TRUE), sums)
-    } else {
-        rank <- 1:nrow(res_out)
-    }
-
-    return(res_out[rank,])
-}
-```
-
-Furthermore, we can look at things species per species to find which species species have the most extreme values (in general, or within their groups).
-
-### Extreme species per super order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position")
-```
-Figure: elaborations distributions per super orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_phylo <- get.distributions(results$super_results_phylo, type = "distance")
-```
-Figure: innovations distributions per super orders
-
-```{r, "printing results in text", echo = FALSE, eval = TRUE}
-## Combining the results
-super_phylo_table <- do.call(rbind, mapply(combine.results, super_positions_phylo, super_distances_phylo, SIMPLIFY = FALSE))
-## Get the top 10%
-get.top <- function(data, what, top = 0.9, relative = FALSE) {
-    if(any(data[, "level"] == "phylogeny")) {
-        data_top <- data[which(data[, "level"] != "phylogeny"), ]
-    } else {
-        data_top <- data
-    }
-
-    ## Get the relative column
-    if(relative) {
-        what <- switch(what,
-                       "elaboration" = "rel.elab",
-                       "innovation"  = "rel.inno")
-    }
-    ## Get the top 10(ish) species
-    data_top_10 <- data_top[which(data_top[, what] >= quantile(data_top[, what], prob = top, na.rm = TRUE)), ]
-    return(data_top_10[order(data_top_10[, what], decreasing = TRUE), ])
-}
-top_elabo <- get.top(super_phylo_table, what = "elaboration")
-top_innov <- get.top(super_phylo_table, what = "innovation")
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by superorder:
-
-The 10% top elaborators by superorder are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
-
-The 10% top innovators by superorder are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto the phylogenetic VCV.
-</summary>
-
-```{r, echo = FALSE}
-library(knitr)
-rownames(super_phylo_table) <- NULL
-kable(super_phylo_table, digits = 3)
-```
-
-</details>
-
-### Extreme species per order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
-```
-Figure: elaborations distributions per order
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_phylo <- get.distributions(results$order_results_phylo, type = "distance", exclude.phylo = TRUE)
-```
-Figure: innovations distributions per order
-
-```{r, echo = FALSE}
-order_phylo_table <- do.call(rbind, mapply(combine.results, order_positions_phylo, order_distances_phylo, SIMPLIFY = FALSE))
-top_elabo <- get.top(order_phylo_table, what = "elaboration")
-top_innov <- get.top(order_phylo_table, what = "innovation")
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by order:
-The 10% top ellaborators by order are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
-The 10% top innovators by order are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(order_phylo_table) <- NULL
-kable(order_phylo_table, digits = 3)
-```
-</details>
-
-### Extreme species per super order (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE,  add.relative = TRUE)
-```
-Figure: elaborations distributions within super orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_within <- get.distributions(results$super_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: innovations distributions within super orders
-
-```{r, echo = FALSE}
-super_phylo_table_within <- do.call(rbind, mapply(combine.results, super_positions_within, super_distances_within, SIMPLIFY = FALSE))
-top_elabo <- get.top(super_phylo_table_within, what = "elaboration", relative = TRUE)
-top_innov <- get.top(super_phylo_table_within, what = "innovation", relative = TRUE)
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by superorder relative to each superorder (i.e. the best one(s) in each order):
-```{r, echo = FALSE}
-combined <- rbind(top_elabo, top_innov)
-rownames(combined) <- NULL  
-kable(combined, digits = 3)
-```
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(super_phylo_table_within) <- NULL
-kable(super_phylo_table_within, digits = 3)
-```
-
-</details>
-
-### Extreme species per order (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: elaborations distributions within orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_within <- get.distributions(results$order_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: innovations distributions within orders
-
-```{r, echo = FALSE}
-order_phylo_table_within <- do.call(rbind, mapply(combine.results, order_positions_within, order_distances_within, SIMPLIFY = FALSE))
-top_elabo <- get.top(order_phylo_table_within, what = "elaboration", relative = TRUE)
-top_innov <- get.top(order_phylo_table_within, what = "innovation", relative = TRUE)
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by order relative to each order (i.e. the best one(s) in each order):
-```{r, echo = FALSE}
-combined <- rbind(top_elabo, top_innov)
-rownames(combined) <- NULL  
-kable(combined, digits = 3)
-```
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(order_phylo_table_within) <- NULL
-kable(order_phylo_table_within, digits = 3)
-```
-</details>
-
-## Correlations between elaboration and innovation?
-
-```{r, echo = FALSE}
-## Calculating all the correlations
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE) {
-
-    ## Remove the phylo one (if needed)
-    if(exclude.phylo) {
-        select_pos <- names(size.subsets(res$position))
-        select_dis <- names(size.subsets(res$distance))
-        select_pos <- select_pos[-which(select_pos %in% "phylogeny")]
-        select_dis <- select_dis[-which(select_dis %in% "phylogeny")]
-        res$position <- get.subsets(res$position, subsets = select_pos)
-        res$distance <- get.subsets(res$distance, subsets = select_dis)
-    }
-
-    ## Extracting all the results
-    disparity_pos <- lapply(res$position$disparity, function(X, type) return(X$elements))
-    disparity_dis <- lapply(res$distance$disparity, function(X, type) return(X$elements))
-
-    ## Measure the central tendencies
-    centrals_pos <- lapply(disparity_pos, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-    centrals_dis <- lapply(disparity_dis, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-
-    ## Number of groups
-    n_groups <- length(centrals_pos)
-
-    ## Prepare the results
-    prep.cor.results <- function(pos, dis, name) {
-        ## Run the correlation test
-        cor_test <- cor.test(pos, dis)
-        ## Prepare the plot title
-        plot_main <- paste0(name, ": ", unname(round(cor_test$estimate, 3)), " (p=", unname(round(cor_test$p.value, 3)), ")")
-        ## Return the plotting parameters
-        return(list(x = pos, y = dis, main = plot_main))
-    }
-
-    return(mapply(prep.cor.results, centrals_pos, centrals_dis, as.list(names(centrals_pos)), SIMPLIFY = FALSE))
-}
-
-## Plotting one correlation 
-plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
-    dots <- list(...)
-
-    if(missing(col.sub)) {
-        col.sub <- NULL
-    }
-
-    ## Subbing the colours
-    if(!is.null(col.sub)) {
-        ## Reset the levels
-        col_sub <- as.factor(as.character(col.sub))
-        ## Make the empty level grey
-        if(any(levels(col_sub) == "")) {
-            col_avail <- c("grey", col.fun(length(levels(col_sub))-1))
-        } else {
-            col_avail <- col.fun(length(levels(col_sub)))
-        }
-        ## Create the colour vector
-        col_vector <- col_avail[as.numeric(col_sub)]
-
-    } else {
-        if(is.null(dots$col)) {
-            col_vector <- "black"
-        } else {
-            col_vector <- dots$col
-        }
-    }
-
-    if(!missing(tree)) {
-        # op <- par(mfrow = c(1,2))
-        #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1
-        par(mar = c(5, 4, 4, 0)+0.1)   
-    }
-
-    ## Plotting the results
-    plot(cor.results$x, cor.results$y, pch = 19, xlab = "elaboration", ylab = "innovation", main = cor.results$main, cex = pt.cex, col = col_vector)
-
-    if(!is.null(col.sub)) {
-        ## Adding the legend
-        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
-    }
-
-    if(!missing(tree)) {
-        par(mar = c(5, 0, 4, 2) + 0.1)
-
-        ## Recycle the colour vector
-        named_col_vector <- col_vector
-        names(named_col_vector) <- names(cor.results$x)
-
-        ## Reduce the tree
-        if(!full.tree) {
-            tree <- drop.tip(tree, tip = tree$tip.label[which(!(tree$tip.label %in% names(named_col_vector)))])
-        }
-
-        ## Matching edges and colours
-        match.edge.colours <- function(tree, named_col_vector, background = "grey") {
-            ## Get the background colour
-            edge_cols <- rep(background, Nedge(tree))
-            ## Get the clades
-            clades <- unique(named_col_vector)
-            ## Ignore the greys
-            if(any(which_grey <- clades == "grey")) {
-                clades <- clades[!which_grey]
-            }
-            ## Find the edges for each clade
-            for(clade in 1:length(clades)) {
-                ## Get the tips for the clade
-                tips <- which(named_col_vector == clades[clade])
-                ## Get the MRCA node
-                mrca_node <- getMRCA(tree, tips)
-                # stop_edge <- which(tree$edge[, 2] %in% mrca_node)
-
-                ## Handle non-monophyletic groups
-                clade_tree <- extract.clade(tree, node = mrca_node)
-                if(Ntip(clade_tree) > length(tips)) {
-                    clade_tree <- drop.tip(clade_tree, tip = clade_tree$tip.label[!(clade_tree$tip.label %in% names(tips))])
-                }
-                n_edges_target <- Nedge(clade_tree)
-
-                ## Get the tips ID
-                tip_ids <- which(tree$tip.label %in% names(tips))
-                ## Get the vector of edges
-                edges <- integer()
-                ## Get the first nodes (and edges)
-                new_edges <- which(tree$edge[, 2] %in% tip_ids)
-                edges <- unique(c(edges, new_edges))
-                nodes <- tree$edge[new_edges, 1]
-
-                # plot(tree, show.tip.label = FALSE)
-                # test.col <- c(replicate(10, c("red", "blue", "orange", "green", "pink", "yellow")))
-                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                ## Recursively loop through each node up until the last edge
-                # while(!(stop_edge %in% edges)) {
-                while(length(edges) <= n_edges_target) {
-                    ## Get the node edges
-                    new_edges <- which(tree$edge[, 2] %in% nodes)
-
-                    # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                    edges <- unique(c(edges, new_edges))
-                    ## Update the nodes
-                    nodes <- tree$edge[new_edges, 1]
-                }
-
-                # ## Do the last recursion
-                # new_edges <- which(tree$edge[, 2] %in% nodes)
-                # edges <- unique(c(edges, new_edges))
-
-                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                ## Colour these edges
-                edge_cols[edges] <- clades[clade]
-            }
-            return(edge_cols)
-        }
-        ## Match the colour to the correct edge
-        edge_cols <- match.edge.colours(tree, named_col_vector)
-
-        ## Add the tree
-        plot(tree, show.tip.label = FALSE, edge.color = edge_cols)
-    }
-
-    return(invisible())
-}
-```
-
-### Correlations per super order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
-## Getting the subsets for the colouring
-col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
-## Calculate the correlations
-super_phylo_cor <- get.correlations(results$super_results_phylo)
-## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
-for(i in 1:length(super_phylo_cor)) {
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
-}
-```
-
-<details>
-  <summary>[Click to expand] Finding corner cases groups.
-</summary>
-
-So here's how you remove exploration and elaboration: by centering these three species on the centre of the phylo axis:
-
-Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
-# colour.palette <- rainbow
-posteriors <- shapespace_allbirds_lvl_superorder_order
-
-# plot(superorder_shapespace$matrix[[1]][, 1:2], pch = 19, cex = 0.5)
-# identify(superorder_shapespace$matrix[[1]][, 1:2])
-# shapespace[c(268, 286, 80), c(9, 10)]
-
-## Moving these three species around: Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
-fixed_points <- c(268, 286, 408)
-move_around <- c(8749, 8750, 8751)
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Get the major axis
-phylo_axis <- axis.covar(superorder_shapespace, sample = median)$phylogeny[[1]]
-## Point with no elaboration/exploration
-no_nothing <- 0
-
-## Example with no exploration/elaboration
-skewed_space[move_around, c(1:8)] <- no_nothing
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny"))
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
-           main = "elements without\nelaboration innovation")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-So easy right? We just need to change them along the PC axis to move them around in space.
-
-No, it's not fucking easy because it's a 8 dimensional problem...
-Even though the two first axes cumulate 86.27% of the variance, fixing the three elements on the two elements and centering them on the 6 other dimensions (13.73% of the variance) results in BIG changes:
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Example with no exploration/elaboration
-skewed_space[move_around, c(3:8)] <- 0
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
-           main = "Centring only on dimensions 3:8")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-For example, the blue element, Anas_falcata is moved to no elaboration/exploration at all if we do that ignoring data past the second axis.
-There's a duck axis lurking in the depth... (spoiler, it's the fourth dimension).
-
-Let's flatten the problem in 2D for simplification:
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-
-## 2Ding the space
-dimensions <- c(1,2)
-superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
-superorder_skewed_space$call$dimensions <- dimensions
-fun2d <- function(covar) {
-    covar$VCV <- covar$VCV[dimensions, dimensions]
-    covar$loc <- covar$loc[dimensions]
-    return(covar)
-}
-superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Flattening the results")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-abline(a = 0, b = 7, col = "grey")
-abline(a = 0, b = 0.14, col = "grey")
-```
-
-So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.14 and 7).
-Let's see if it's something birdy or not by modifying these elements position in trait space to have elaboration of 0 and innovation of 1 and the opposite.
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Move around the first species to have elaboration = 0 and innovation = 0
-skewed_space[move_around[1], c(1:2)] <- 0
-## Move around the second species to have elaboration = 0 and innovation = 1
-skewed_space[move_around[2], c(1:2)] <- c(0, 0.4)
-## Move around the third species to have elaboration = 1 and innovation = 0
-skewed_space[move_around[3], c(1:2)] <- c(0.4, 0)
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-
-## 2Ding the space
-dimensions <- c(1,2)
-superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
-superorder_skewed_space$call$dimensions <- dimensions
-fun2d <- function(covar) {
-    covar$VCV <- covar$VCV[dimensions, dimensions]
-    covar$loc <- covar$loc[dimensions]
-    return(covar)
-}
-superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Flattening the results")
-## Add the phylogenetic axis
-lines(phylo_axis[,1:2], lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis[,1:2], lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-TODO: fix this!
-
-</details>
-
-### Correlations within super orders (species on group)
-
-Regardless, here are the results when projecting the species on their own group's VCV
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
-## Getting the subsets for the colouring
-col_subs <- split(shapespace$level2, f = shapespace$level1)[-1]
-
-## Calculate the correlations
-super_within_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
-## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
-for(i in 1:length(super_within_cor)) {
-    plot.correlations(super_within_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
-}
-```
-
-<!-- THIS BIT IS IGNORED FOR NOW -->
-
-### Correlations per order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
-## Loading the sub-order levels
-load(file = "../Data/Processed/shapespace_superorder_lvl_order_suborder.rda")
-## Select the data
-shapespaces <- lapply(shapespace_superorder_lvl_order_suborder, function(X) X$space)
-## Create a level 3
-order_space <- do.call(rbind, shapespaces)
-
-## Calculate the correlations
-order_phylo_cor <- get.correlations(results$order_results_phylo, exclude.phylo = TRUE)
-
-## Getting the subsets for the colouring
-split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
-col_subs <- replicate(length(order_phylo_cor), NULL)
-col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
-
-## Set the vector of legend positions
-for(i in 1:length(order_phylo_cor)) {
-    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]])
-}
-```
-
-### Correlations within orders (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
-## Calculate the correlations
-order_within_cor <- get.correlations(results$order_results_within, exclude.phylo = TRUE)
-
-## Getting the subsets for the colouring
-split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
-col_subs <- replicate(length(order_within_cor), NULL)
-col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
-
-## Set the vector of legend positions
-for(i in 1:length(order_within_cor)) {
-    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]])
-}
-```

### 2022-02-11 Analysis/10-Figure1.Rmd
Update 10-Figure1.Rmd

@@
 
-Plotting the tree
+```{r, echo = FALSE}
+## Quick load
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+results <- shapespace_allbirds_lvl_superorder_order_results_list
+rm(shapespace_allbirds_lvl_superorder_order_results_list)
+```
+
+## Phylogeny window (detailed)
 
 ```{r}
+## Plotting the ellipses
+plot.one.ellipse <- function(data, ID, colours, x = TRUE, y = TRUE, main = "", legend = FALSE) {
+
+    ## Get the PC %ages
+    if(x || y) {
+        vars <- dispRity::variances(data$matrix[[1]])
+        pc_var <- round((vars/sum(vars)*100), 2)
+    }
+
+    ## Do the empty plot
+    plot(NULL, xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1), xaxt = "n", yaxt = "n", xlab = ifelse(x, paste0("PC1 (", pc_var[1], "%)"), ""), ylab = ifelse(y, paste0("PC2 (", pc_var[2], "%)"), ""), main = main)
+    if(x) axis(1)
+    if(y) axis(2)
+    abline(v = 0, col = "grey", lwd = 0.5)
+    abline(h = 0, col = "grey", lwd = 0.5)
+
+    covar.plot(get.subsets(data, subsets = names(data$subsets)[ID]),
+           col = colours[ID],
+           ellipses = mean,
+           legend = FALSE,
+           points = FALSE,
+           cex = 0.5, apply.to.VCV = TRUE,
+           lwd = 4,
+           add = TRUE,
+           legend = legend)
+}
+## Plot the projection/rejection
+plot.one.proj.rej <- function(data, ID, colours, x = TRUE, y = TRUE, main = "") {
+
+    ## Get the proj/rej
+    if(ID == 1) {select <- 2} else {select <- ID-1}
+    elaboration <- c(get.disparity(data$position, concatenate = FALSE)[[select]])
+    exploration <- c(get.disparity(data$distance, concatenate = FALSE)[[select]])
+
+    ## Change the colours
+    col <- c(colours[[ID]], adjustcolor(colours[[ID]], alpha.f = 0.5))
+    ## Plot the results
+    par(mar = c(2, 6, 2, 2)+0.1, bty = "n")
+    boxplot(cbind(elaboration, exploration), horizontal = TRUE, ylim = c(0,15), col = col, las = 1, xaxt = ifelse(x, "l", "n"), yaxt = ifelse(y, "l", "n"))
+}
+
+pdf(file = "Figure_1a.pdf", height = 10, width = 5)
+## Plot the phylo ellipse
+grey_palette <- c(colour.palette(8), "grey")
+par(mfrow = c(2,1), bty = "n")
+plot.one.ellipse(superorder_shapespace, ID = c(1,9), colours = grey_palette, x = TRUE, y = TRUE, main = "A")
+legend("topright", lwd = 5, col = grey_palette[c(1,9)], lty = 1, legend = c("Aequornithes' median variance-covariance", "Overall phylogenetic's median variance-covariance"), bg = "white")
+plot.one.proj.rej(results$super_results_between, ID = 1, colours = grey_palette)
+```
+
+```{r}
+# pdf(file = "Figure_1b.pdf", height = 16, width = 14)
+## Global window
+layout_matrix <- layout(cbind(matrix(1, ncol = 3, nrow = 8),
+                              matrix(2:((8*2)+1), ncol = 2, nrow = 8, byrow = TRUE)))
+# layout.show(layout_matrix)
+
+## Plotting the tree
+
 tree <- ladderize(tree)
@@
 ## Replace the order levels as colours
-coloured_levels <- orders_lvl
-levels(coloured_levels) <- colour.palette(length(levels(coloured_levels)))
+coloured_levels <- superorders_lvl
+colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+par(mar = c(1, 1, 1, 1)+0.1)
+plot(tree, show.tip.label = FALSE, edge.color =  match.tip.edge(as.character(coloured_levels), tree, replace.na = "grey"))
 
-plot(tree, show.tip.label = FALSE, edge.color =  match.tip.edge(as.character(coloured_levels), tree, replace.na = "grey"))
-```
+colour_palette <- levels(coloured_levels)
 
-Plotting the ellipses
+## Plotting the clade names
+## Finding the ladderized clade names
+# dev.new()
+# plot(x = rep(1, 9), y = c(1:9), pch = 19, col = colour_palette)
+# text(x = rep(1, 9), y = c(1:9), label = paste(1:9, levels_ID), col = colour_palette)
 
-```{r}
-plot.one.ellipse <- function(data, ID, colours) {
+levels_ID <- levels(superorders_lvl)
+ladder_order <- c("Paleognathae" = 7, "Galloanserae" = 4, "Mirandornithes" = 5, "Columbiomorphae" = 3, "Otidimorphae" = 6, "Aequornithes" = 2, "Strisores" = 8, "Telluraves" = 9)
+
+# par(mfrow = c(length(levels(orders_lvl))), 1)
+# par(mfrow = c(length(levels(coloured_levels)), 1), mar = c(2,2,2,0), bty = "n")
+
+#TODO: order the levels to fit the ladderised tree
+
+for(i in 1:8) {
+
+    ## Plot the clade name
+    # par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    # plot(NULL, xaxt = "n", yaxt = "n", xlim = c(-1, 1), ylim = c(-1, 1), xlab = "", ylab = "")
+
+    ## Plot the ellipse
+    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    plot.one.ellipse(superorder_shapespace, ID = ladder_order[i], colours = colour_palette, x = FALSE, y = FALSE)
+    text(0,1, names(ladder_order)[i], cex = 2, col = colour_palette[][ladder_order[i]])
+
+    ## Plot the elaboration/exploration
+    par(bty = "n", mar = c(3, 1, 1, 1)+0.1)
+    plot.one.proj.rej(results$super_results_between, ID = ladder_order[i], colours = colour_palette, x = TRUE, y = FALSE)
 

### 2022-02-11 Analysis/10-Figure1.Rmd
Update 10-Figure1.Rmd

@@
 
-pdf(file = "Figure_1a.pdf", height = 10, width = 5)
+pdf(file = "Figure_1a.pdf", height = 11, width = 5.5)
 ## Plot the phylo ellipse

### 2022-02-14 Analysis/09-Figure3.Rmd
Update 09-Figure3.Rmd

--- 2022-02-10 Analysis/09-Figure3.Rmd
@@
 ```{r, echo = FALSE, eval = FALSE}
+library(ggplot2)
 ## Getting the data to plot
@@
 
+pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4, width = 4)
 gg.density.plot(global_data)
+dev.off()
+```
+
+Can't seem to integrate the density plot to the base R layout so I'll just do it in Inkscape for now
+
+```{r, echo = FALSE, eval = FALSE}
+pdf("../Manuscript/Figures/Figure3_a.pdf", height = 6, width = 3)
+## Plotting the projection cartoon
+library(mcmcmcglmmm)
+## Get projection lines (from a centred matrix)
+get.proj.line <- function(point, base) {
+  point <- t(point)
+  base <- t(base)
+  proj_x <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[1];
+  proj_y <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[2];
+  return(rbind(cbind(proj_x, proj_y), point))
+}
+
+set.seed(42)
+## Space plot
+cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
+space <- space.maker(10, 2, rnorm, cor.matrix = cor_matrix)
+lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
+
+par(mfrow = c(2, 1))
+## Projections per species
+## Translate the major axis
+par(bty = "n", mar = c(1,1,1,1)+0.1)
+plot(space, pch = 19, xlab = "", ylab = "", xlim = lim, ylim = lim, col = "grey", main = "Panels B and C: species'\nprojection on the phylogeny", xaxt = "n", yaxt = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+## Plotting the ellipse
+lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "grey", lwd = 1)
+## Plotting the major axis
+lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "grey", lty = 1)
+lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
+# legend("bottomright", legend = "C")
+
+## Do the projection
+matrix <- space
+point1 <- c(-1.2, 0)
+point2 <- c(2, 0.75)
+base_vector <- rbind(point1, point2)
+
+## Get all the space (with the two last rows being the base vectors)
+space_cent <- rbind(matrix, base_vector)
+## Centre all the space
+space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
+## Re-attribute the centred variables
+matrix_cent <- space_cent[1:nrow(matrix), ]
+base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]
+
+## Plot the projections
+for(i in 1:nrow(matrix)) {
+  cent_proj <- get.proj.line(matrix_cent[i,], base_vector_cent[2, ])
+  cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
+  lines(cent_proj, col = "darkgreen")
+}
+
+## Species projection onto their own axis
+par(bty = "n", mar = c(1,1,1,1)+0.1)
+plot(space, pch = 19, xlab = "", ylab = "", xlim = lim, ylim = lim, col = "grey", main = "Panels D: species'\nprojection on their own group", xaxt = "n", yaxt = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "grey")
+## Plotting the ellipse
+lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "blue", lwd = 1)
+## Plotting the major axis
+lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "blue", lty = 1)
+# legend("bottomright", legend = "D")
+
+## Do the projection
+matrix <- space
+base_vector <- get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75)
+point1 <- base_vector[1,]
+point2 <- base_vector[2,]
+
+## Get all the space (with the two last rows being the base vectors)
+space_cent <- rbind(matrix, base_vector)
+## Centre all the space
+space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
+## Re-attribute the centred variables
+matrix_cent <- space_cent[1:nrow(matrix), ]
+base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]
+## Extend the base vector a little bit
+extend_base <- base_vector
+extend_base[1, ] <- extend_base[1,1] + 0.05
+lines(extend_base, lwd = 1, col = "blue", lty = 2)
+
+## Plot the projections
+for(i in 1:nrow(matrix_cent)) {
+  cent_proj <- get.proj.line(matrix_cent[i,], base_vector_cent[2, ])
+  cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
+  cent_proj[, 2] <- (cent_proj[, 2] + point1[2])
+  lines(cent_proj, col = "darkgreen")
+}
+dev.off()
 ```
@@
 ```{r, echo = FALSE, eval = FALSE}
+
+pdf("../Manuscript/Figures/Figure3_cd.pdf", height = 12, width = 8)
+## Great the plot layout
+layout_matrix <- cbind(matrix(0,2,2), matrix(0,2,2))
+layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8, 2, 4)), matrix(9:16, 2, 4))
+plot_layout <- layout(layout_matrix)
+# layout.show(plot_layout)
 
@@
     ## Plotting the results
-    plot(cor.results$x, cor.results$y, pch = 19, xlab = ifelse(), ylab = ifelse(ID == 1 || ID == 5, "exploration", ""), main = cor.results$main, cex = pt.cex, col = col_vector)
+
+    ## Selecting the x and y axis base on plot ID
+    select.x <- function(ID, yes = "", no = "") {
+        # ifelse(ID %in% c(5:8), yes, no)
+        return(no)
+    }
+    select.y <- function(ID, yes = "", no = "") {
+        # ifelse(ID == 1 || ID == 5, yes, no)
+        return(no)
+    }
+
+    par(mar = c(select.x(ID, 4, 2), select.y(ID, 4, 2), 3, 1)+0.1)
+    plot(cor.results$x, cor.results$y, pch = 19,
+        xlab = select.x(ID, "elaboration"),
+        ylab = select.y(ID, "explortation"),
+        main = strsplit(cor.results$main, split = ":")[[1]][1],
+        cex = pt.cex, col = col_vector,
+        ...)
 
     if(!is.null(col.sub)) {
+        legend_text <- levels(col_sub)
+        if(length(empty <- which(legend_text == "")) > 0) {
+            legend_text[empty] <- "other"
+        }
         ## Adding the legend
-        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
+        legend(x = legend.pos, legend = legend_text, pch = 19, cex = legend.cex, col = col_avail, bg = "white")
     }
@@
 
-par(mfrow = c(2, 4))
-lims <- list(x = range(global_data$x),
-             y = range(global_data$y))
-
+# par(mfrow = c(2, 4))
 ## Getting the subsets for the colouring
@@
 for(i in 1:8) {
-    plot.correlations(super_phylo_cor[[i]], xlim = lims$x, ylim = lims$x, tree = NULL, col.sub = col_subs[[i]], legend.pos = legends_pos[i])
+    one.plot.correlations(super_phylo_cor[[i]],col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.8, legend.cex = 0.8)
 }
 
+## Fig 3c
+
+# par(mfrow = c(2, 4))
+## Getting the subsets for the colouring
+col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
+
+## Set the vector of legend positions
+legends_pos <- c("bottomright", "topright", "topright", "topleft", "topleft", "topright", "bottomright", "bottomright")
+
+for(i in 1:8) {
+    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.8, legend.cex = 0.8)
+}
+dev.off()
 ```

### 2022-02-21 Analysis/10-Figure1.Rmd
Update 10-Figure1.Rmd

--- 2022-02-11 Analysis/10-Figure1.Rmd
@@
 
-Plotting the tree
+```{r, echo = FALSE}
+## Quick load
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+results <- shapespace_allbirds_lvl_superorder_order_results_list
+rm(shapespace_allbirds_lvl_superorder_order_results_list)
+```
+
+## Phylogeny window (detailed)
 
 ```{r}
+## Plotting the ellipses
+plot.one.ellipse <- function(data, name.col, x = TRUE, y = TRUE, main = "", legend = FALSE, with.phylo = FALSE) {
+
+    ## Get the PC %ages
+    if(x || y) {
+        vars <- dispRity::variances(data$matrix[[1]])
+        pc_var <- round((vars/sum(vars)*100), 2)
+    }
+
+    ## Do the empty plot
+    plot(NULL, xlim = c(-1.25, 1.25), ylim = c(-1.25, 1.25), xaxt = "n", yaxt = "n", xlab = ifelse(x, paste0("PC1 (", pc_var[1], "%)"), ""), ylab = ifelse(y, paste0("PC2 (", pc_var[2], "%)"), ""), main = main)
+    if(x) axis(1)
+    if(y) axis(2)
+    abline(v = 0, col = "grey", lwd = 0.5)
+    abline(h = 0, col = "grey", lwd = 0.5)
+
+    if(!with.phylo) {
+        subsets <- names(name.col)
+        cols <- name.col
+    } else {
+        subsets <- c("phylogeny", names(name.col))
+        cols <- c("grey", name.col)
+    }
+    covar.plot(get.subsets(data, subsets = subsets),
+           col = cols,
+           ellipses = mean,
+           legend = FALSE,
+           points = FALSE,
+           cex = 0.5, apply.to.VCV = TRUE,
+           lwd = 4,
+           add = TRUE,
+           legend = legend)
+}
+## Plot the projection/rejection
+plot.one.proj.rej <- function(data, name.col, x = TRUE, y = TRUE, main = "") {
+
+    ## Get the proj/rej
+    elaboration <- get.disparity(data$position, concatenate = FALSE)
+    exploration <- get.disparity(data$distance, concatenate = FALSE)
+
+    ## Get the correct group
+    if(length(ID <- which(names(test) %in% names(name.col))) < 1) {
+        ID <- which(names(test) %in% paste0(names(name.col), ":phylogeny"))
+    }
+    elaboration <- c(elaboration[[ID]])
+    exploration <- c(exploration[[ID]])
+
+    ## Change the colours
+    col <- c(name.col, adjustcolor(name.col, alpha.f = 0.5))
+
+    ## Set the limits
+    quant_95 <- quantile(c(elaboration, exploration), prob = 0.975)
+    lim_up <- c(ceiling(quant_95/5)*5)
+
+    ## Plot the results
+    boxplot(cbind(elaboration, exploration), horizontal = TRUE, ylim = c(0,lim_up), col = col, las = 1, xaxt = "n", yaxt = "n")
+    if(x) {
+        axis(1, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5))
+    }
+    if(y) {
+        axis(2, at = c(1, 2), labels = c("exploration", "innovation"), las = 2)
+    }
+}
+
+# pdf(file = "Figure_1a.pdf", height = 11, width = 5.5)
+# ## Plot the phylo ellipse
+# grey_palette <- c(colour.palette(8), "grey")
+# par(mfrow = c(2,1), bty = "n")
+# plot.one.ellipse(superorder_shapespace, ID = c(1,9), colours = grey_palette, x = TRUE, y = TRUE, main = "A")
+# legend("topright", lwd = 5, col = grey_palette[c(1,9)], lty = 1, legend = c("Aequornithes' median variance-covariance", "Overall phylogenetic's median variance-covariance"), bg = "white")
+# plot.one.proj.rej(results$super_results_between, ID = 1, colours = grey_palette)
+```
+
+```{r}
+box_size <- 1.5
+pdf(file = "../Manuscript/Figures/Figure_1b.pdf", height = 2*5, width = box_size*12)
+## Global window
+
+layout_matrix <- layout(cbind(
+                        rbind(matrix(1, ncol = 2, nrow = 3),
+                              matrix(2:5, ncol = 2, nrow = 2, byrow = TRUE)),
+                        matrix(6:((27*2)+1), ncol = 10, nrow = 5, byrow = TRUE)
+                        ))
+
+# layout.show(layout_matrix)
+
+# layout_matrix <- layout(cbind(matrix(1, ncol = 1, nrow = 27),
+#                               matrix(2:((27*2)+1), ncol = 2, nrow = 27, byrow = TRUE)))
+# layout.show(layout_matrix)
+
+## Plotting the tree
+
 tree <- ladderize(tree)
@@
 
+## Select the level
+selected_level <- coloured_levels <- orders_lvl
+
 ## Replace the order levels as colours
-coloured_levels <- orders_lvl
-levels(coloured_levels) <- colour.palette(length(levels(coloured_levels)))
+colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+# par(mar = c(1, 1, 1, 1)+0.1)
+# plot(tree, show.tip.label = FALSE, edge.color =  match.tip.edge(as.character(coloured_levels), tree, replace.na = "grey"))
 
-plot(tree, show.tip.label = FALSE, edge.color =  match.tip.edge(as.character(coloured_levels), tree, replace.na = "grey"))
-```
-
-Plotting the ellipses
-
-```{r}
-plot.one.ellipse <- function(data, ID, colours) {
-
+## Shorten the tree
+selected_tips <- character()
+for(one_level in 2:length(levels(selected_level))) {
+    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
+    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
 }
 
+## Shorten the tree
+tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
+tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
+
+## Setting the tips edges colours
+tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
+## Rainbow the tip edges
+tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
+## Plotting the tree
+par(mar = c(0, 0, 0, 0)+0.1)
+plot(tree_short, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
+# tiplabels()
+
+## Matching the order and the colours to the ladderised tree
+tip_order <- c(
+    "Tinamiformes"= 1,
+    "Galliformes"= 3,
+    "Anseriformes"= 2,
+    "Otidiformes"= 7,
+    "Cuculiformes" = 8,
+    "Gruiformes" = 9, 
+    "Musophagiformes" = 10,
+    "Procellariiformes" = 12,
+    "Sphenisciformes" = 11,
+    "Ciconiiformes" = 13,
+    "Pelecaniformes"= 15,
+    "Suliformes" = 14,
+    "Podicipediformes" = 4,
+    "Columbiformes" = 6,
+    "Pterocliformes" = 5,
+    "Charadriiformes" = 18,
+    "Trogoniformes" = 21,
+    "Bucerotiformes" = 22,
+    "Coraciiformes" = 23,
+    "Piciformes" = 24,
+    "Strigiformes" = 20,
+    "Accipitriformes" = 19, 
+    "Falconiformes" = 25,
+    "Psittaciformes" = 26,
+    "Passeriformes" = 27,
+    "Caprimulgiformes" = 16, 
+    "Apodiformes" = 17)
+tip_colours <- colour.palette(Ntip(tree_short))
+names(tip_colours) <- names(sort(tip_order))
+# par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+# plot(rep(1,27), 1:27, col = "white")
+# labs <- names(tip_colours)
+# cols <- tip_colours
+# text(x = rep(1,27), y = 1:27, label = rev(labs), col = rev(cols))
+
+for(i in 1:length(tip_order)) {
+    ## Select the colour and name
+    name.col <- tip_colours[match(sort(names(tip_colours)), names(tip_colours))][i]
+
+    ## Plot the ellipse
+    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    plot.one.ellipse(order_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
+    text(0,1.2, names(name.col), cex = 1.5, col = name.col)
+
+    ## Plot the elaboration/exploration
+    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
+    plot.one.proj.rej(results$order_results_between, name.col, x = TRUE, y = FALSE)
+}
+dev.off()
+
 ```

### 2022-02-21 Analysis/10-Figure1.Rmd
Update 10-Figure1.Rmd

@@
     ## Get the correct group
-    if(length(ID <- which(names(test) %in% names(name.col))) < 1) {
-        ID <- which(names(test) %in% paste0(names(name.col), ":phylogeny"))
+    if(length(ID <- which(names(elaboration) %in% names(name.col))) < 1) {
+        ID <- which(names(elaboration) %in% paste0(names(name.col), ":phylogeny"))
     }
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure_1b.pdf", height = 2*5, width = box_size*12)
+pdf(file = "../Manuscript/Figures/Figure_1.pdf", height = box_size*8, width = box_size*10)
+
 ## Global window
+lay_mat_order <- rbind(
+        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 1
+        matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
+        matrix(c(1,1, 18:25), ncol = 10, nrow = 1), # line 3
+        matrix(c(26:35), ncol = 10, nrow = 1), # line 4
+        matrix(c(36:45), ncol = 10, nrow = 1), # line 5
+        matrix(c(46:55), ncol = 10, nrow = 1) # line 7
+    )
+lay_mat_super <- rbind(
+        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 7
+        matrix(c(1,1, 10:17), ncol = 10, nrow = 1) # line 8
+    )
 
-layout_matrix <- layout(cbind(
-                        rbind(matrix(1, ncol = 2, nrow = 3),
-                              matrix(2:5, ncol = 2, nrow = 2, byrow = TRUE)),
-                        matrix(6:((27*2)+1), ncol = 10, nrow = 5, byrow = TRUE)
-                        ))
+layout_matrix <- layout(rbind(lay_mat_super, (lay_mat_order+17)))
 
@@
 
-## Plotting the tree
-
+## PLOTTING THE SUPER ORDERS
 tree <- ladderize(tree)
@@
 ## Select the level
-selected_level <- coloured_levels <- orders_lvl
+selected_level <- coloured_levels <- superorders_lvl
 
@@
 tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
+## Adding the number of tips
+tree_plot <- tree_short
+n_species <- table(selected_level)
+for(i in 1:Ntip(tree_plot)) {
+    ## Get the number of species
+    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
+    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
+}
+
 ## Plotting the tree
 par(mar = c(0, 0, 0, 0)+0.1)
-plot(tree_short, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
+plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
 # tiplabels()
+
+## Matching the order and the colours to the ladderised tree
+tip_order <- c(
+    "Paleognathae" = 1,
+    "Galloanserae" = 2,
+    "Otidimorphae" = 5,
+    "Aequornithes" = 6,
+    "Mirandornithes" = 3,
+    "Columbimorphae" = 4,
+    "Telluraves" = 8,
+    "Strisores" = 7)
+tip_colours <- colour.palette(Ntip(tree_short))
+names(tip_colours) <- names(sort(tip_order))
+# par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+# plot(rep(1,27), 1:27, col = "white")
+# labs <- names(tip_colours)
+# cols <- tip_colours
+# text(x = rep(1,27), y = 1:27, label = rev(labs), col = rev(cols))
+
+for(i in 1:length(tip_order)) {
+    ## Select the colour and name
+    name.col <- tip_colours[match(sort(names(tip_colours)), names(tip_colours))][i]
+
+    ## Plot the ellipse
+    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    plot.one.ellipse(superorder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
+    text(0,1.2, names(name.col), cex = 1.5, col = name.col)
+
+    ## Plot the elaboration/exploration
+    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
+    plot.one.proj.rej(results$super_results_between, name.col, x = TRUE, y = FALSE)
+}
+
+## PLOTTING THE ORDERS
+tree <- ladderize(tree)
+## Get the tip_values
+orders_lvl <- shapespace$level2
+superorders_lvl <- shapespace$level1
+
+## Select the level
+selected_level <- coloured_levels <- orders_lvl
+
+## Replace the order levels as colours
+colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+
+## Shorten the tree
+selected_tips <- character()
+for(one_level in 2:length(levels(selected_level))) {
+    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
+    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
+}
+
+## Shorten the tree
+tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
+tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
+
+## Setting the tips edges colours
+tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
+## Rainbow the tip edges
+tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
+
+## Adding the number of tips
+tree_plot <- tree_short
+n_species <- table(selected_level)
+for(i in 1:Ntip(tree_plot)) {
+    ## Get the number of species
+    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
+    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
+}
+
+## Plotting the tree
+par(mar = c(0, 0, 0, 0)+0.1)
+plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
 
@@
 names(tip_colours) <- names(sort(tip_order))
-# par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
-# plot(rep(1,27), 1:27, col = "white")
-# labs <- names(tip_colours)
-# cols <- tip_colours
-# text(x = rep(1,27), y = 1:27, label = rev(labs), col = rev(cols))
 

### 2022-02-22 Analysis/09-Figure3.Rmd
Update 09-Figure3.Rmd

@@
 layout_matrix <- cbind(matrix(0,2,2), matrix(0,2,2))
-layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8, 2, 4)), matrix(9:16, 2, 4))
+layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8, 2, 4, byrow = TRUE)), matrix(9:16, 2, 4,  byrow = TRUE))
 plot_layout <- layout(layout_matrix)
@@
 
-one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, ...) {
-
-    dots <- list(...)
+one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, legend_counter) {
 
@@
 
-    # if(!missing(tree)) {
-    #     op <- par(mfrow = c(1,2))
-    #     #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1
-    #     par(mar = c(5, 4, 4, 0)+0.1)   
-    # }
+    ## Sort the data by group size
+    all_data <- cbind(x = cor.results$x, y = cor.results$y)
+    ## Classify the groups per numbers of species
+    all_data <- data.frame(all_data, col = col_vector)
 
-    ## Plotting the results
+    ## Sort the data per group size
+    group_order <- sort(table(all_data$col), decreasing = TRUE)
 
-    ## Selecting the x and y axis base on plot ID
-    select.x <- function(ID, yes = "", no = "") {
-        # ifelse(ID %in% c(5:8), yes, no)
-        return(no)
-    }
-    select.y <- function(ID, yes = "", no = "") {
-        # ifelse(ID == 1 || ID == 5, yes, no)
-        return(no)
+    ## Set the plotting data
+    plot_data <- data.frame()
+    # col_adjust <- seq(from = 1/length(group_order), to = 1, by = 0.1)*3
+    for(i in 1:length(group_order)) {
+        ## Get the data from the biggest group
+        selected_data <- all_data[which(all_data$col == names(group_order[i])), ]
+        ## Transparantise the colours (for the background)
+        # selected_data$col <- adjustcolor(selected_data$col, alpha = col_adjust[i])
+        ## Combine the plotting data
+        plot_data <- rbind(plot_data, selected_data)
     }
 
-    par(mar = c(select.x(ID, 4, 2), select.y(ID, 4, 2), 3, 1)+0.1)
-    plot(cor.results$x, cor.results$y, pch = 19,
-        xlab = select.x(ID, "elaboration"),
-        ylab = select.y(ID, "explortation"),
-        main = strsplit(cor.results$main, split = ":")[[1]][1],
-        cex = pt.cex, col = col_vector,
-        ...)
+    ## Plot the points
+    plot(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = plot_data$col, xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = strsplit(cor.results$main, split = ":")[[1]][1])
+    axis(1, at = pretty(plot_data$x, n = 4))
+    axis(2, at = pretty(plot_data$y, n = 4), las = 2)
 
     if(!is.null(col.sub)) {
-        legend_text <- levels(col_sub)
-        if(length(empty <- which(legend_text == "")) > 0) {
-            legend_text[empty] <- "other"
+        ## See if there is any others
+        levels <- levels(col_sub)
+        if(length(other <- which(levels == "")) > 0) {
+            legend_text <- c(1, 1:(length(levels(col_sub))-1)+legend_counter)
+        } else {
+            legend_text <- c(1:(length(levels(col_sub)))+legend_counter)
         }
+        ## Update the legend counter
+        legend_counter <<- max(legend_text)
         ## Adding the legend
@@
 ## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
+legends_pos <- c("topleft", "topright", "topright", "topright",
+                 "topleft", "topright", "topleft", "topright")
+legend_counter <- 1
 
 for(i in 1:8) {
-    one.plot.correlations(super_phylo_cor[[i]],col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.8, legend.cex = 0.8)
+    par(mar = c(2,2.5,2,1.5))
+    one.plot.correlations(super_phylo_cor[[i]],col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter)
 }
@@
 ## Set the vector of legend positions
-legends_pos <- c("bottomright", "topright", "topright", "topleft", "topleft", "topright", "bottomright", "bottomright")
+legends_pos <- c("bottomright", "topright", "topleft", "topleft",
+                 "topleft", "topleft", "topright", "bottomright")
+legend_counter <- 1
 
 for(i in 1:8) {
-    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.8, legend.cex = 0.8)
+    par(mar = c(2,2.5,2,1.5))
+    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter)
 }

### 2022-02-23 Analysis/10-Figure1.Rmd
renamed figure1

@@
 
-# pdf(file = "Figure_1a.pdf", height = 11, width = 5.5)
+# pdf(file = "Figure1a.pdf", height = 11, width = 5.5)
 # ## Plot the phylo ellipse
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure_1.pdf", height = box_size*8, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)
 

### 2022-02-24 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd

@@
 ## Installing beer
-devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
+devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/mcmcmcglmmm")
 ```
@@
 
-<!--
-
-#### Number of runs optimisation (TO BE DEVELOPED?)
-
-```{r, eval = FALSE}
-## Function for running one mini chain on a random tree (and saving the results)
-fun.mini.chains <- function() {
-    run.mini.chains(mini_chains, replicates = 1, path = "../Data/Processed/MCMC/", file.name = "mini.chains")
-}
-```
-
-To make sure we reached consistent results, we run the set of N mini-chains were N was determined using the following variance optimisation algorithm.
-
- * 1. run an initial $N = 20$ chains;
- * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
- * 3. update N to N = N * 1.05 and run an additional 5% chains;
- * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
- * 5. calculate $\Delta_V$, the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
- * 6.
-If $\Delta_V < 0.05$ (meaning an increase of 5% of the replicates resulted in an increase of the parameters of less than 5%), the desired number of mini-chains is reached.
-Else, set $V_{1} = V_{2}$ and repeat from step 3.
-
-```{r, optimised_replicates, eval = FALSE}
-## The summary function
-
-```
-
-After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "standard" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.
-
--->
-
 ## References

### 2022-02-24 Analysis/02-mcmcmcglmmm_analysis.Rmd
Vignettes INIT (big)

---
title: "Projection analysis to measure biological elaboration and innovation."
author: "Thomas Guillerme, Natalie Cooper, Andrew P Beckerman,  Gavin H Thomas"
bibliography: references.bib
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
    keep_tex: true
  self_contained: true
---

```{r header, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
library(mcmcmcglmmm)
library(dispRity)
```

# Mini chains MCMCglmm method: an efficient way to calculate variance-covariance matrices on big data with phylogenetic uncertainty

One commonly used method to estimate variance-covariance matrices from phylogenetic datasets is to use multivariate generalised linear mixed models.
This is often done in a Bayesian way using the `MCMCglmm` package (@hadfield2010; cited nearly 4000 times in the last decade).
The MCMCglmm method allows to run a generalised linear mixed model on a multidimensional dataset with an underlying phylogenetic structure.
<!-- Expand that from the longevity paper -->.

Using this method we can run a nested phylogenetic model on a multidimensional dataset (here the Charadriiformes PCA shapespace).
This model will have a the multidimensional trait values as a response and the classic residual terms as an error term.
Additionally though, we can add a random error term as the phylogenetic signal.
This random error term can be expanded into four independent nested terms: one for the entire phylogenetic structure and three for each individual clade's phylogenetic structure.

$$\text{data} = \text{traits} + \text{residuals} + (\text{phylogenetic term} + \text{clade term})$$

Or in `MCMCglmm` pseudo code:

```
formula   = PC1:8 ~ trait:clade-1
random    = ~ us(at.level(clade):trait):animal + us(trait):animal
residuals = ~ us(trait):units
```
<!-- TG:TODO: explain that much better) -->

Although this method is perfectly efficiently implemented on small to relatively big datasets, the major hurdle is the calculation of the variance-covariance matrix that increases following a power law with both the number of species and the number of dimensions.
Therefore, such method becomes easily untractable on large datasets (>8k species and >3 dimensions).
Furthermore, this method's phylogenetic correction is based on a single input tree (usually the consensus tree).
This can also be problematic on big trees (i.e. > 5k species) where the notion of a consensus tree being a good representation of phylogenetic is at best misleading and at worth incorrect.
In fact, the bigger the tree, the more there is variance in topological relation between species (especially in terms of branch length towards deep branches).
Therefore, there has been a push to use tree distributions rather than consensus tree (i.e. point estimates) in using the MCMCglmm method (@healy2014; @mulTree).
This however, leads to yet another untractable computer problem: it requires running the large dataset much more than once on multiple trees (e.g. if using @mulTree) thus linearly increasing computational time.

Thus, to increase the speed of these analyses, while taking phylogenetic uncertainty into account, we used a highly parallelisable "mini chains" approach.
In brief, it runs multiple short `MCMCglmm` analyses on multiple trees and pulls the results together into one larger `MCMCglmm` that contains more variation due to phylogenetic uncertainty (where the size of the chains are optimised for speed and low RAM usage).

## Mini chains

The mini chains `MCMCglmm` method (`mcmcmcglmmm`) effectively works by running many very short parallel `MCMCglmm` on as many different tree topologies.
Effectively each mini chain generates a very small number of samples post-burnin (e.g. 10) and combine them across many replicates (e.g. 1000) into a chain containing effectively many samples (e.g. 10000), each ran independently on a different tree toplogy.

![Figure 2](../Figures/mini-chains_diagram.png)

##### Figure 2: {#fig2}
Mini chains diagram: the numerical values displayed here (number of chains, priors, etc...) are used as examples and can be tuned to fit specific questions and datasets.

In more details, the method works as follows:

 1. Running a small number of "parametrisation chains": these chains are run with the traits data and the model of interest on a consensus tree from the tree distribution, along with flat priors with a low belief parameter (effectively not putting any important weight on the priors). 
 2. Extracting parameters from the "parametrisation chains": this step allows to extract the parameters of interest from the posteriors of the "parametrisation chains". The parameters of interest are the conservative burnin average time (here defined as highest number of iterations required to reach the median posterior likelihood across the parametrisation chains with an additional 10% extra generations) and the mini chains priors (the median posterior model results - ignoring the previously estimated burnin - with a slightly higher belief parameter than for the previous chains).
 3. Running the mini chains: these chains use the same data and model as for the "parametrisation chains" but use a random tree from the tree distribution rather than the consensus one. Furthermore, they use the parameters previously estimated (i.e. the priors and the burnin phase). These chains are run just for a small number of itterations past the burnin phase to extract a fixed number of samples (e.g. if the sampling rate is every 100 generations and the requested number of samples, the chain runs for $\text{burnin iterations} + \text{requested_samples} \times \text{sampling rate}$).
 4. Combining the mini chains: finally we can combine the post-burnin samples from all the mini chains back into a single classical `MCMCglmm` chain (with no burnin).

Some note about diagnosis.

### Implementation

The mini chain approach is implemented on the github package `mcmcmcglmmm` (link).
The package comes with a demo vignette explaining in details how to reproduce each step in the procedure described above.

### Effectiveness

In a practical implication of the mini-chains, we ran for another project a similar model on 8748 species with 8 dimensions and 31 nested levels.
Using this method, it took effectively 20 parallel cores with 8GB of RAM 2 months to complete the model with each individual mini-chain running for around 40 hours (using the SHARC cluster from Sheffield: CITE).
We estimated that the same model using the normal `MCMCglmm` implementation would have required 2.1 years and 4TB of RAM.

This method thus allows to run multivariate generalised linear mixed effects models on large datasets and across a tree distribution.
The posteriors of this analyses contain a number of variance-covariance matrices (`model$VCV` in a `MCMCglmm` output) and their location in the trait space (`model$Sol` in a `MCMCglmm` output).
We can then use these distributions of variance-covariance matrices as major axes in the multidimensional space on which to run the elaboration and exploration analyses.

# Reproducible example

Here we illustrate the BEER pipeline by applying it to a data set from @cooney2017 focused on the birds charadriiformes order.

## Data

This order contains 359 species divided into three clades: gulls, sandpipers and plovers and sandpipers that have respectively 159, 102 and 98 species each.
For each of these species, we used the 3D beak dataset from @cooney2017 that is an ordination of the shapes of the beaks of 8@@@ birds (the resulting shape space used here does not contain any information about the beak size - i.e. centroid size).
We extracted the 359 charadriiformes species from this space where the first three dimensions account for more than 99% of the variance in the dataset.
For the rest of the example analysis here, we used as a traitspace of these 359 species with 3 dimensions (hereafter the shapespace).

```{r, echo = FALSE, eval = TRUE}
## Loading the charadriiformes data
data(charadriiformes) 
## Extracting the tree
tree <- charadriiformes$tree
## Extracting the data column that contains the clade assignments
data <- charadriiformes$data[, "clade"]
## Changing the levels names (the clade names) to colours
levels(data) <- c("orange", "blue", "darkgreen")
data <- as.character(data)
## Matching the data rownames to the tip order in the tree
data <- data[match(ladderize(tree)$tip.label, rownames(charadriiformes$data))]

## Matching the tip colours (labels) to their descending edges in the tree
## (and making the non-match edges grey)
clade_edges <- match.tip.edge(data, tree, replace.na = "grey")

## Plotting the results
plot(ladderize(tree), show.tip.label = FALSE, edge.color = clade_edges, main = "Charadriiformes")
legend("bottomleft", lty = c(1,1,1), col = c("blue", "darkgreen", "orange"), legend = c("plovers", "sandpipers", "gulls"))
axisPhylo()
```

##### Figure 5: {#fig5}
The Charadriiformes data used in this pipeline illustration. The axis units are millions of years ago.

## Mini-chains MCMCglmm method

To illustrate the BEER pipeline, we analysed the variance-covariance of the shapespace using a generalised linear mixed model with two different levels of random terms: one for the whole charadriiformes phylogeny and one different one for each different clade (gulls, plovers and sandpipers).
We first run three parametrisation chains for 50000 generations with no burnin and a sampling every 500 generations.
From this chain we extracted: 1) the maximum burnin time as 1.1 times the number of generations required to reach the median likelihood value and 2) the median posteriors of the model (discarding the burnin) as priors for our mini chains models with a believe parameter value of 5%.
We then ran 50 minichains for 10 samples each using the previous sampling rate (500 generations) and the extracted parameters form the parametrisation chain (the burnin and priors described above).

```{r, echo = FALSE, eval = FALSE}
library(mcmcmcglmmm)

## Making the parametrisation chains
parametrisation_chains <- make.mini.chains(data = charadriiformes$data,
                                           dimensions = c("PC1", "PC2", "PC3"),
                                           randoms = c("global", "clade"),
                                           residuals = "global",
                                           tree = charadriiformes$tree,
                                           parameters = list(nitt = 50000, burnin = 0, thin = 500))

## Running the parametrisation chains
param_results <- run.mini.chains(parametrisation_chains, replicates = 3)
save(param_results, file = "param_results.rda")
load("param_results.rda")

diagnosis <- diagnose.mini.chains(param_results)
## Extracting the parameters
params <- extract.parameters(param_results)

## The sampling rate:
sampling <- 500
## The required number of samples per tree
samples <- 10

## Making the mini chains
parametrisation_chains <- make.mini.chains(data = charadriiformes$data,
                                           dimensions = c("PC1", "PC2", "PC3"),
                                           randoms = c("global", "clade"),
                                           residuals = "global",
                                           priors = params$priors,
                                           tree = charadriiformes$tree_distribution,
                                           parameters = list(
                                            nitt = params$burnin + samples*sampling,
                                            burning = params$burnin,
                                            thin = sampling))
## Running the mini chains
results <- run.mini.chains(parametrisation_chains, replicates = 100)
posteriors <- combine.mini.chains(results)
save(posteriors, file = "posteriors.rda")
```

This resulted in obtaining 1000 exploitable posterior variance-covariance matrices for each residual terms of our model: the overall phylogenetic effect and the three specific phylogenetic effects for each of the three clades.

```{r, echo = FALSE, eval = TRUE}
load("posteriors.rda")
silent <- diagnose.mini.chains(posteriors, plot = TRUE)
```

<!-- Do some developments about this part -->

### 2022-02-24 Analysis/03-projection_analysis.Rmd
Vignettes INIT (big)

---
title: "Projection analysis to measure biological elaboration and innovation."
author: "Thomas Guillerme, Natalie Cooper, Andrew P Beckerman,  Gavin H Thomas"
bibliography: references.bib
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
    keep_tex: true
  self_contained: true
---

# Projection analyses

With the variance-covariance matrices obtained from the `mcmcmcglmmm` analyses, we can use linear algebra to make a mathematical interpretation of @endler2005's elaboration and innovation (here exploration) patterns (Fig 1. in @endler2005).
In fact, if we can interpret the variance-covariance matrix as estimated evolutionary space in the trait space that contains all the observed traits of a group and takes evolutionary history into account, we can then use the first eigen vector of that variance-covariance matrix as the major evolutionary axis, i.e. the axis in multidimensional space along which most trait values co-evolve.
In other words, if the variance-covariance matrix is a multidimensional ellipse, it's major axis is its major evolutionary axis.
Then, if species evolve parallel to this axis, they are elaborating on traits combinations; conversely, if species evolve orthogonaly to this axis, they are exploring new traits combinations (innovating).

Mathematically we can use linear algebra to project and reject each species in the trait-space along this major axis of evolution (Figure 4).
The detailed operations are described in the [supplementary materials](#projections-supplementary).

```{r, fig.height = 6, fig.width = 6, echo = FALSE, eval = TRUE}
source("../Functions/test.rotation.R")
```

##### Figure 3: {#fig3}
Consider a space with five elements named \"a\" to \"e\" (in grey) and one focal axes in this space defined as the vector from point1: \"d\" to point2: \"e\" (the thin solid grey line linking the grey \"d\" and \"e\"). We can then define all points from \"d\" (point1) as vectors (the dashed grey lines). We can then rotate the whole space so that the vector \"d\" to \"e\" becomes the reference coordinate vector (0,1) (the thick grey line) and rotate and translate all the other vectors accordingly (the solid grey lines linking the coloured letters \"a\" to \"c\"). We can then project each of these vectors on the reference vector and measure their distance *on* the reference vector (the projection; the coloured solid lines) and their distance *from* the reference vector (the rejection; the coloured dotted lines). We can then also measure the angles between each vector and their respective projections (the angles between a grey solid line to a coloured letter and the corresponding coloured vector). Note that if this angle is equal to 90 degrees, the projection is equal to 0 (point \"c\") or if that angle is > 90 degrees, the projection is negative (point \"b\").

In practice, we can use these projections and rejections of vectors or points onto a specific other vector as a way to measure the elaboration and exploration species within a group or of entire groups.
The first instance, measuring the elaboration and exploration of species within a group, is rather straightforward from [Figure 3](#fig3): the idea is to project each points (defined as the dashed vectors in figure 4) onto a vector of interest (e.g. either the group's own main axis of variance-covariance or the entire phylogeny's one).

We measured the exploration and elaboration using two main approaches (see [Figure 4](#fig4)):

 * By clades (i.e exploration/elaboration **between** clades), where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes.
 * By species (i.e exploration/elaboration **within** clades), where we calculated the projections of each species onto 1) the global phylogenetic major axes and 2) the phylogenetic major axes of their respective clade.

```{r, echo = FALSE, eval = TRUE}
## Get projection lines (from a centred matrix)
get.proj.line <- function(point, base) {
  point <- t(point)
  base <- t(base)
  proj_x <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[1];
  proj_y <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[2];
  return(rbind(cbind(proj_x, proj_y), point))
}

set.seed(42)
## Space plot
cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
space <- space.maker(10, 2, rnorm, cor.matrix = cor_matrix)
lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))

par(mfrow = c(2, 2))

## Plot original space
plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim, col = "black", main = "Original space")
abline(v = 0, col = "grey", lwd = 0.5)
abline(h = 0, col = "grey", lwd = 0.5)
## Plotting the ellipse
lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "blue", lwd = 1)
## Plotting the major axis
lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "blue", lty = 1)
## Plotting a phylogenetic major axis
lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
# legend("topright", pch = c(19, NA, NA, NA), lty = c(NA, 1,1,1), lwd = c(NA, 1, 3, 3), legend = c("Group of species", "95% CI ellipse for the group", "Major axis for the group", "Major axis for the phylogeny"), col = c("black", "blue", "blue", "orange"))
legend("bottomright", legend = "A")

## Projections of the groups
## Translate the major axis
plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim, col = "grey", main = "Groups' projection\non the phylogeny")
abline(v = 0, col = "grey", lwd = 0.5)
abline(h = 0, col = "grey", lwd = 0.5)
## Plotting the ellipse
lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "grey", lwd = 1)
## Plotting the major axis
lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "grey", lty = 1)
lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
legend("bottomright", legend = "B")

## Do the translation
major_axis <- get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75)
point_to_translate <- c(-1.2, 0)
# lines(rbind(major_axis[2,], point_to_translate), lty = 3, col = "grey")
arrows(x0 = major_axis[2,1], y0 = major_axis[2,2], x1 = point_to_translate[1], y1 = point_to_translate[2], col = "black", lty = 3, length = 0.1)

move_x <- point_to_translate[1] - major_axis[2, 1]
move_y <- point_to_translate[2] - major_axis[2, 2]
axis_trans <- major_axis
axis_trans[,1] <- axis_trans[,1] + move_x
axis_trans[,2] <- axis_trans[,2] + move_y
lines(axis_trans, lwd = 3, col = "blue", lty = 1)

## Project the translation
matrix <- axis_trans
point1 <- c(-1.2, 0)
point2 <- c(2, 0.75)
base_vector <- rbind(point1, point2)
## Get all the space (with the two last rows being the base vectors)
space_cent <- rbind(matrix, base_vector)
## Centre all the space
space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
## Re-attribute the centred variables
matrix_cent <- space_cent[1:nrow(matrix), ]
base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]
## Projected translation
cent_proj <- get.proj.line(matrix_cent[1,], base_vector_cent[2, ])
cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
lines(cent_proj, col = "darkgreen")
## Extend the phylogeny line
extend_base <- rbind(base_vector[1, ], cent_proj[1, ])
lines(extend_base, lwd = 1, col = "orange", lty = 2)

## Projections per species
## Translate the major axis
plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim, col = "grey", main = "Species' projection\non the phylogeny")
abline(v = 0, col = "grey", lwd = 0.5)
abline(h = 0, col = "grey", lwd = 0.5)
## Plotting the ellipse
lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "grey", lwd = 1)
## Plotting the major axis
lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "grey", lty = 1)
lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
legend("bottomright", legend = "C")

## Do the projection
matrix <- space
point1 <- c(-1.2, 0)
point2 <- c(2, 0.75)
base_vector <- rbind(point1, point2)

## Get all the space (with the two last rows being the base vectors)
space_cent <- rbind(matrix, base_vector)
## Centre all the space
space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
## Re-attribute the centred variables
matrix_cent <- space_cent[1:nrow(matrix), ]
base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]

## Plot the projections
for(i in 1:nrow(matrix)) {
  cent_proj <- get.proj.line(matrix_cent[i,], base_vector_cent[2, ])
  cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
  lines(cent_proj, col = "darkgreen")
}

## Species projection onto their own axis
plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim, col = "grey", main = "Species' projection\non their own axis")
abline(v = 0, col = "grey", lwd = 0.5)
abline(h = 0, col = "grey", lwd = 0.5)
lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "grey")
## Plotting the ellipse
lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "blue", lwd = 1)
## Plotting the major axis
lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "blue", lty = 1)
legend("bottomright", legend = "D")

## Do the projection
matrix <- space
base_vector <- get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75)
point1 <- base_vector[1,]
point2 <- base_vector[2,]

## Get all the space (with the two last rows being the base vectors)
space_cent <- rbind(matrix, base_vector)
## Centre all the space
space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
## Re-attribute the centred variables
matrix_cent <- space_cent[1:nrow(matrix), ]
base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]
## Extend the base vector a little bit
extend_base <- base_vector
extend_base[1, ] <- extend_base[1,1] + 0.05
lines(extend_base, lwd = 1, col = "blue", lty = 2)

## Plot the projections
for(i in 1:nrow(matrix_cent)) {
  cent_proj <- get.proj.line(matrix_cent[i,], base_vector_cent[2, ])
  cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
  cent_proj[, 2] <- (cent_proj[, 2] + point1[2])
  lines(cent_proj, col = "darkgreen")
}

```

##### Figure 4: {#fig4}
Different types of elaboration exploration analyses for a group of species. A) Given a group of 10 species (in black) with two traits, we can calculate the 95% confidence interval variance-covariance ellipse and its major axis (in blue). Say we also calculated another vector (in orange) that could be for example the major axis of another variance-covariance matrix (e.g. from another group or the whole phylogeny). B) we can calculate the projection of the group's major axis on the other major axis by translating it (black doted arrow) and calculate it's elaboration relative to the orange vector (orange dashed line) or its relative exploration (green line). C) We can also calculate the elaboration and exploration scores for each individual species relative to the orange major axis (e.g. the phylognetic major axis); D) or relative to their own major axis.

<!-- Scaling and interpretation -->
Note that the examples depicted in [Figure 4](#fig4) are not scaled for clarity.
In the algorithm implemented in `dispRity`, the results are always effectively scaled as described in [Figure 3](#fig3) and in the [detailed algorithm](#projections-supplementary).
In practice, this makes the vector onto which elements are projected of length one.
Furthermore, because the projections happen based on the first point of the major axis of interest (which can be arbitrary), we advice to centre the elaboration results on 0.5 and rescaled (effectively removing 0.5 from the scaled projection value and rescaling it to 1) and to make them absolute.
This makes the elaboration and exploration results more intuitive to interpret: an elaboration score comprised in $[0, 1]$ is distributed onto the major axis, a score of $[1, \infty)$ is beyond the major axis (in any _n_ direction).
If the major axis is for example the 95% confidence interval from the phylogenetic variance-covariance matrix, this denotes a species that is especially elaborative.
Similarly for exploration, any exploration score comprised in $[0, 1]$ indicates an exploration that is within the major axis' variance-covariance 95% confidence interval; a score of $[1, \infty)$ denotes a particularly explorative species.

# Reproducible example

Here we illustrate the BEER pipeline by applying it to a data set from @cooney2017 focused on the birds charadriiformes order.

## Data

This order contains 359 species divided into three clades: gulls, sandpipers and plovers and sandpipers that have respectively 159, 102 and 98 species each.
For each of these species, we used the 3D beak dataset from @cooney2017 that is an ordination of the shapes of the beaks of 8@@@ birds (the resulting shape space used here does not contain any information about the beak size - i.e. centroid size).
We extracted the 359 charadriiformes species from this space where the first three dimensions account for more than 99% of the variance in the dataset.
For the rest of the example analysis here, we used as a traitspace of these 359 species with 3 dimensions (hereafter the shapespace).

```{r, echo = FALSE, eval = TRUE}
## Loading the charadriiformes data
data(charadriiformes) 
## Extracting the tree
tree <- charadriiformes$tree
## Extracting the data column that contains the clade assignments
data <- charadriiformes$data[, "clade"]
## Changing the levels names (the clade names) to colours
levels(data) <- c("orange", "blue", "darkgreen")
data <- as.character(data)
## Matching the data rownames to the tip order in the tree
data <- data[match(ladderize(tree)$tip.label, rownames(charadriiformes$data))]

## Matching the tip colours (labels) to their descending edges in the tree
## (and making the non-match edges grey)
clade_edges <- match.tip.edge(data, tree, replace.na = "grey")

## Plotting the results
plot(ladderize(tree), show.tip.label = FALSE, edge.color = clade_edges, main = "Charadriiformes")
legend("bottomleft", lty = c(1,1,1), col = c("blue", "darkgreen", "orange"), legend = c("plovers", "sandpipers", "gulls"))
axisPhylo()
```

##### Figure 5: {#fig5}
The Charadriiformes data used in this pipeline illustration. The axis units are millions of years ago.

## Projections analysis

We then used these results to understand measure which of the three charadriiformes is the most elaborative and the most exploring.
We tested that both at 1) the entire clade level, by projecting the 1000 posterior major axis of each clade's phylogenetic residuals onto the global phylogeny's residuals; and 2) the species level to see which clade presented the most elaborative and exploring species.

```{r, echo = FALSE, eval = TRUE}
covar_data <- MCMCglmm.subsets(data = charadriiformes$data,
                               posteriors = posteriors,
                               group = MCMCglmm.levels(posteriors)[1:4],
                               rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
par(mfrow = c(2,2))
covar.plot(covar_data, points = TRUE, points.cex = 0.5, col = c("orange", "blue", "darkgreen"))
legend("topleft", legend = "A")
covar.plot(covar_data, points = TRUE, points.cex = 0.5, ellipses = mean, major.axes = mean, legend = TRUE, col = c("orange", "blue", "darkgreen", "grey"), apply.to.VCV = TRUE, legend.x = "bottomright")
legend("topleft", legend = "B")
covar.plot(covar_data, points = FALSE, ellipses = TRUE, n = 100, col = c("orange", "blue", "darkgreen", "grey"))
legend("topleft", legend = "C")
covar.plot(covar_data, points = FALSE, major.axes = TRUE, n = 100, col = c("orange", "blue", "darkgreen", "grey"))
legend("topleft", legend = "D")
```

##### Figure 6: {#fig.6}
Visualisation of the posteriors of the model: A) the first two dimensions of the inpu trait space; B) the same space with the average ellipses and average major.axis plotted; C) the distribution of 100 random ellipses; D) the distribution of 100 random major axes.")

```{r, echo = FALSE, eval = FALSE}
## Measure the group differences
group_differences <- dispRity.covar.projections(covar_data,
                                                type = "groups",
                                                base = "phylogeny",
                                                output = c("position", "distance"))
## Measure the individual differences
species_differences <- dispRity.covar.projections(covar_data,
                                                  type = "elements",
                                                  base = "phylogeny",
                                                  output = c("position", "distance"))

```

# References

# Supplementary details {#projections-supplementary}

## Detailed projection and rejection operations

In details, we can define the major axes from the variance-covariance matrices and then project and reject each element of interest in the space onto this axis.
The following steps are generalised to $n$ dimensions and detailed below (as well as the algorithm used in `dispRity` to perform the transformations:

### Major axes definition

For $O_{n}$, the unit hypersphere matrix of $n$ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

\begin{equation}
O_{n} = 
    \begin{pmatrix}
        1 & 0 & \cdots & 0 \\
        0 & 1 & \cdots & 0 \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        0 & 0 & \cdots & 1 \\
        -1 & 0 & \cdots & 0 \\
        0 & -1 & \cdots & 0 \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        0 & 0 & \cdots & -1 \\
    \end{pmatrix}
\end{equation}

In other words $O_{n}$ is the matrix representing the hypersphere of $n$ dimensions and of radius $1$ that fits in the centre of the trait space;

And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:

$$O'_{n} = O_{n} \sqrt{\chi^2(0.95)}$$

Then, for the variance-covariance matrix $VCV_{n}$ of $n$ dimensions obtained from the posterior distribution of the `mcmcmcglmmmm`:

\begin{equation}
VCV_{n} = 
    \begin{pmatrix}
        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
    \end{pmatrix}
\end{equation}

and the eigenvectors *v* and the eigenvalues $\lambda$ satisfying the following eigen decomposition:

$$VCV_{n} v = \lambda v$$

We can get $M_{n}$, the matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors *v* and the product of the scaled 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:

$$M_{n} = [(O'_{n}\sqrt{\lambda}) \times v]^{\text{T}}$$

Finally, we can centre the matrix $M_{n}$ on the estimated solution of each generalised linear mixed model (`model$Sol`, in `MCMCglmm`) corresponding to the estimation of the position of the variance-covariance matrix in the trait space.
$M_{n}$ then contains all the major axis of the 0.95 hyper-ellipse fitting the variance-covariance matrix.
We can then define the first row of the matrix, $M_{1,n}$, as the major axis, the second row, $M_{2,n}$, as the second major axis (the minor axis in a 2D ellipse), etc.

The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `dispRity::axis.covar`.

### Measuring projection and rejection

Once we have defined a major axis, we can project any elements in the trait space onto that axis.
Specifically, for any elements in the trait space, we can define it as the vector $\vec{a}$ with one set of coordinates in $n$ dimensions:

\begin{equation}
    \vec{a} = 
    \begin{bmatrix}
    x \\
    y \\
    \cdots \\
    n \\
    \end{bmatrix}
\end{equation}

And for any major axis that we can define as a vector $\vec{b}$ as a set of pairs of coordinates in $n$ dimensions:

\begin{equation}
    \vec{b} = 
    \begin{bmatrix}
    x_{1} & x_{2} \\
    y_{1} & y_{2} \\
    \cdots & \cdots \\
    n_{1} & n_{2} \\
    \end{bmatrix}
\end{equation}

We can then calculate $\vec{a_{1}}$, the orthogonal projection of $\vec{a}$ onto $\vec{b}$ using:

\begin{equation}
    \vec{a_{1}} = \frac{\vec{a} \cdot \vec{b}}{\|\vec{b}\|}
\end{equation}

With $\|\vec{b}\| = \sqrt{\vec{b} \cdot \vec{b}}$ being the norm of $\vec{b}$.
And $\vec{a_{2}}$, the rejection of $\vec{a}$ onto $\vec{b}$:

\begin{equation}
    \vec{a_{2}} = \vec{a} - \vec{a_{1}}
\end{equation}

#### Generalisation of projection onto any vector in a set space

Using this, we can generalise the procedure so as to calculate the projection and rejection for any element within a trait space $TS_{m,n}$:

\begin{equation}
    TS_{m,n} = 
    \begin{bmatrix}
    x_{1} & x_{2} & \cdots & x_{m} \\
    y_{1} & y_{2} & \cdots & y_{m} \\
    \vdots  & \vdots  & \ddots & \vdots \\
    n_{1} & n_{2} & \cdots & n_{m} \\
    \end{bmatrix}
\end{equation}

And any major axis defined as a vector $\vec{B}$:

\begin{equation}
    B = 
    \begin{bmatrix}
    x_{1} & x_{2}\\
    y_{1} & y_{2}\\
    \vdots  & \vdots  \\
    n_{1} & n_{2} \\
    \end{bmatrix}
\end{equation}

By using the linear transformation $f_{\vec{B}}$ of the trait space $TS$ moving $\vec{B}$ onto $TS$'s first axes unit vector $\vec{\hat{\imath}}$:

$$f_{\vec{B}}(TS) = \left( \frac{TS - [Bx_{1}, By_{1}, \cdots, Bn_{1}]^{\text{T}}}{\|\vec{B}\|} \right) \cdot R_{\vec{B}}$$

With $R_{\vec{B}}$ being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:

\begin{equation}
R_{\vec{B}} = I_{\vec{B}} - \vec{B}\vec{B}^\text{T} - \vec{\hat{\imath}}\vec{\hat{\imath}}^\text{T} + [\vec{B} \vec{\hat{\imath}}]
    \begin{bmatrix}
        cos(\theta) & -sin(\theta)\\
        sin(\theta) & cos(\theta)\\
    \end{bmatrix} [\vec{B} \vec{\hat{\imath}}]^\text{T}
\end{equation}

Where $\theta$ is:

\begin{equation}
    \theta = acos \left(\frac{\vec{B} \cdot \vec{\hat{\imath}}}{\|\vec{B}\| \cdot \|\vec{\hat{\imath}}\|} \right)
\end{equation}

Or $\theta = acos (B_x)$ since both $\|\vec{B}\|$ and $\|\vec{\hat{\imath}}\|$ are equal to 1 and $\|\vec{\hat{\imath}}\|$ is the unit vector on the first axis.

### Algorithm for calculating the projection/rejection of any element in a defined space

In practice we followed [this procedure](https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions) and applied a modification of [this implementation](https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385) (see @aguilera2004 for the formal generalisation of this algorithm in $n$ dimensions) using the following algorithm implemented in `dispRity::projections` (@disprity):

 1. In the trait space, define $\vec{B}$ as the base vector (typically $\vec{B}$ is defined as the pair of coordinates set from the major axis described above).
 2. Centre the trait space on the origin of $\vec{B}$ so that the first set of coordinates of $\vec{B}$ are 0.
 3. Scale the trait space to the norm of $\vec{B}$ so that the norm of $\vec{B}$ is now 1.
 4. Rotate the trait space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \rightarrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait space - typically the x axes unit vector). 
 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now its projection onto $\vec{B}$.

### 2022-02-25 Analysis/02-mcmcmcglmmm_analysis.Rmd
adding header

@@
 
-```{r header, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
+This vignette provides advanced details on the mini-chains MCMCglmm pipeline and provides code and detail to reproduce the results presented in the paper associated with this vignette.
+Due to the amount of data and computational time required to reproduce the results in the paper (2 CPU years and 350GB of data) we only demonstrate the method on the Charadriiformes order.
+
+> If you use this vignette in published work, please cite @PAPER and @mcmcmcglmmmDOI.
+
+```{r header, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
+## Installing the dispRity package from the CRAN
+install.packages("dispRity")
+## Installing the mcmcmcglmmm package from github
+if(!require(devtools)) install.packages("devtools")
+devtools:install_github("TGuillerme/mcmcmcglmmm")
+## Loading the packagges
 library(mcmcmcglmmm)

### 2022-02-25 Analysis/03-projection_analysis.Rmd
adding header

@@
 ---
+
+This vignette provides advanced details on the mini-chains MCMCglmm pipeline and provides code and detail to reproduce the results presented in the paper associated with this vignette.
+Due to the amount of data and computational time required to reproduce the results in the paper (2 CPU years and 350GB of data) we only demonstrate the method on the Charadriiformes order.
+
+> If you use this vignette in published work, please cite @PAPER and @dispRity.
+
+```{r header, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
+## Installing the dispRity package from the CRAN
+install.packages("dispRity")
+## Loading the packages
+library(dispRity)
+```
 

### 2022-03-01 Analysis/10-Figure1.Rmd
Update 10-Figure1.Rmd

@@
     ## Do the empty plot
-    plot(NULL, xlim = c(-1.25, 1.25), ylim = c(-1.25, 1.25), xaxt = "n", yaxt = "n", xlab = ifelse(x, paste0("PC1 (", pc_var[1], "%)"), ""), ylab = ifelse(y, paste0("PC2 (", pc_var[2], "%)"), ""), main = main)
+    lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+    plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = ifelse(x, paste0("PC1 (", pc_var[1], "%)"), ""), ylab = ifelse(y, paste0("PC2 (", pc_var[2], "%)"), ""), main = main)
     if(x) axis(1)
@@
     }
+    if(!with.phylo) {
         covar.plot(get.subsets(data, subsets = subsets),
@@
                legend = legend)
+    } else {
+        covar.plot(get.subsets(data, subsets = subsets),
+               col = cols,
+               ellipses = mean,
+               legend = FALSE,
+               points = FALSE,
+               cex = 0.5, apply.to.VCV = TRUE,
+               lwd = 4,
+               add = TRUE,
+               legend = legend,
+               scale = "phylogeny")        
+    }
 }
@@
 
+    ## Get the quantiles
+    elaboration_quantiles <- quantile(elaboration, prob = c(0.025, 0.25, 0.50, 0.975))
+    exploration_quantiles <- quantile(exploration, prob = c(0.025, 0.25, 0.50, 0.975))
+
+    ## Empty plot
+    par(bty = "n")
+    plot(NULL, xlim = c(0,1.5), ylim = c(0,lim_up), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+
+    ## Add the lines
+    lines(lty = 2, x = c(0,0), y = elaboration_quantiles[c(1,4)], col = col[1], lwd = 1)
+    lines(lty = 1, x = c(0,0), y = elaboration_quantiles[c(2,3)], col = col[1], lwd = 3)
+    lines(lty = 2, x = c(1,1), y = exploration_quantiles[c(1,4)], col = col[2], lwd = 1)
+    lines(lty = 1, x = c(1,1), y = exploration_quantiles[c(2,3)], col = col[2], lwd = 3)
+    points(0, median(elaboration), pch = 19, col = col[1])
+    points(1, median(exploration), pch = 19, col = col[2])
+
+    ## Add the y axis
+    axis(4, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5), las = 2)
+
     ## Plot the results
-    boxplot(cbind(elaboration, exploration), horizontal = TRUE, ylim = c(0,lim_up), col = col, las = 1, xaxt = "n", yaxt = "n")
-    if(x) {
-        axis(1, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5))
-    }
-    if(y) {
-        axis(2, at = c(1, 2), labels = c("exploration", "innovation"), las = 2)
-    }
+    # boxplot(cbind(elaboration, exploration), horizontal = TRUE, ylim = c(0,lim_up), col = col, las = 1, xaxt = "n", yaxt = "n")
+    # if(x) {
+    #     axis(1, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5))
+    # }
+    # if(y) {
+    #     axis(2, at = c(1, 2), labels = c("exploration", "innovation"), las = 2)
+    # }
 }
@@
     )
+
+# sequence <- c(c(rep(2, 3), 3), c(rep(4, 3), 5), c(rep(6, 3), 7), c(rep(8, 3), 9))
 
@@
     "Strisores" = 7)
+
 tip_colours <- colour.palette(Ntip(tree_short))
 names(tip_colours) <- names(sort(tip_order))
+
 # par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
@@
     ## Select the colour and name
-    name.col <- tip_colours[match(sort(names(tip_colours)), names(tip_colours))][i]
+    # name.col <- tip_colours[match(sort(names(tip_colours)), names(tip_colours))][i]
+    name.col <- tip_colours[i]
 
@@
     plot.one.ellipse(superorder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-    text(0,1.2, names(name.col), cex = 1.5, col = name.col)
+    text(0,0.3, names(name.col), cex = 1.5, col = name.col)
 
@@
     "Apodiformes" = 17)
+
+# tip_order <- sort(tip_order)
+
 tip_colours <- colour.palette(Ntip(tree_short))
@@
     plot.one.ellipse(order_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-    text(0,1.2, names(name.col), cex = 1.5, col = name.col)
+    text(0,0.3, names(name.col), cex = 1.5, col = name.col)
 

### 2022-03-02 Analysis/10-Figure1.Rmd
Update 10-Figure1.Rmd
Looks alright

@@
     par(bty = "n")
-    plot(NULL, xlim = c(0,1.5), ylim = c(0,lim_up), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+    plot(NULL, xlim = c(-0.5,1.5), ylim = c(0,lim_up), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
 
@@
     ## Add the y axis
-    axis(4, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5), las = 2)
+    axis(2, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5), las = 2)
 
@@
         matrix(c(36:45), ncol = 10, nrow = 1), # line 5
-        matrix(c(46:55), ncol = 10, nrow = 1) # line 7
+        matrix(c(46:55), ncol = 10, nrow = 1) # line 6
     )
@@
 
-# sequence <- c(c(rep(2, 3), 3), c(rep(4, 3), 5), c(rep(6, 3), 7), c(rep(8, 3), 9))
-
-layout_matrix <- layout(rbind(lay_mat_super, (lay_mat_order+17)))
+heights_vals <- rep(1, 9)
+widths_vals <- rep(c(1, 1/3), 5)
+layout_matrix <- layout(rbind(lay_mat_super, (lay_mat_order+17)), widths = widths_vals, heights = heights_vals, respect = TRUE)
 

### 2022-03-03 Analysis/10-Figure1.Rmd
added dimensions hist

@@
 }
+## Plot the variation per dimension
+add.dims <- function(data, name.col, n = 1000) {
+    covars_cent_tend <- lapply(sample.n(get.subsets(data, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
+    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+    dim_var <- dim_var/max(dim_var)/10
+
+    ## Select the quadrant
+    if(colMeans(data$matrix[[1]][c(data$subsets[[names(name.col)]]$elements), c(1,2)])[1] > 0.1) {
+        x_start <- -0.2
+    } else {
+        x_start <- 0.2
+    }
+
+    y_start <- -0.1
+
+    buffer <- 0
+    for(one_dim in 1:length(dim_var)) {
+        lines(x = c(x_start, x_start+dim_var[one_dim]), y = rep(y_start+buffer, 2), lwd = 3, col = name.col)
+        buffer <- buffer - 0.02
+    }
+}
 
@@
 ```{r}
+
 box_size <- 1.5
+pdf(file = "../Manuscript/Figures/Figure1_phylo_Dplot.pdf", height = box_size*1, width = box_size*1)
+
+lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = "")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+
+covars_cent_tend <- lapply(sample.n(get.subsets(data, subset = "phylogeny")$covar, n), VCV.cent.tend, mean)
+dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+dim_var <- dim_var/max(dim_var)/10
+x_start <- 0.2
+y_start <- -0.1
+buffer <- 0
+for(one_dim in 1:length(dim_var)) {
+    lines(x = c(x_start, x_start+dim_var[one_dim]), y = rep(y_start+buffer, 2), lwd = 3, col = "grey")
+    buffer <- buffer - 0.02
+}
+
+dev.off()
+
 pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)
@@
     plot.one.ellipse(superorder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-    text(0,0.3, names(name.col), cex = 1.5, col = name.col)
+
+    ## Calculate which dimension contains the most variance
+    add.dims(superorder_shapespace, name.col)
+    n <- 1000
+    covars_cent_tend <- lapply(sample.n(get.subsets(superorder_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
+    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+    text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
+    # text(0,0.3, paste0(names(name.col)), cex = 1.2, col = name.col)
 
@@
     plot.one.ellipse(order_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-    text(0,0.3, names(name.col), cex = 1.5, col = name.col)
+
+    ## Calculate which dimension contains the most variance
+    add.dims(order_shapespace, name.col)
+    n <- 1000
+    covars_cent_tend <- lapply(sample.n(get.subsets(order_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
+    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+    text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
+    # text(0,0.3, paste0(names(name.col)), cex = 1.2, col = name.col)
 

### 2022-03-08 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 
-```{r header, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
+```{r header, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, eval = FALSE}
 ## Installing the dispRity package from the CRAN
@@
 if(!require(devtools)) install.packages("devtools")
-devtools:install_github("TGuillerme/mcmcmcglmmm")
+devtools::install_github("TGuillerme/mcmcmcglmmm")
+```
+
+```{r, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
 ## Loading the packagges
@@
 
-```{r, echo = FALSE, eval = FALSE}
-library(mcmcmcglmmm)
-
+```{r, echo = FALSE, eval = FALSE, eval = TRUE}
 ## Making the parametrisation chains

### 2022-03-10 Analysis/04-Figure1.Rmd
rearranged Analysis folder

--- 2022-03-03 Analysis/10-Figure1.Rmd
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure1_phylo_Dplot.pdf", height = box_size*1, width = box_size*1)
+pdf(file = "../Manuscript/Figures/Figure1_phylo_Dplot.pdf", height = box_size*2, width = box_size*2)
 
@@
     dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
-    # text(0,0.3, paste0(names(name.col)), cex = 1.2, col = name.col)
+    # text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
+    text(0,0.3, paste0(names(name.col)), cex = 1.2, col = name.col)
 

### 2022-03-10 Analysis/06-Supplementary_figures_tables.Rmd
rearranged Analysis folder

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

```{r, echo = FALSE}
library(dispRity)
library(knitr)
library(mcmcmcglmmm)
```

## PC axis selection

```{r, echo = FALSE}
## Loading the shape space
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
## Loading the clade list
load(file = "../Data/Processed/clade_list.rda")

## Temporary change the empty levels to noclade
clades <- clade_list[, c(1)]
levels(clades)[1] <- "noclade"

## Make a dispRity object and get axes data
shape_axes <- select.axes(custom.subsets(shapespace, group = clades))

pdf(file = "../Manuscript/Figures/Supplementary_figure_axis_selection.pdf", height = 8, width = 12)
plot(shape_axes)
dev.off()
```

```{r, fig.heigth = 8, fig.width = 12, echo = FALSE}
plot(shape_axes)
```

Supplementary figure: axis selection: cumulative variance (in light grey) and variance per dimension (in dark grey) for each bird super order and the whole space. The horizontal dotted bar is the threshold value to explain 95% of the variance of the group and the vertical dotted bar is the number of dimensions required to reach that threshold value.

```{r, echo = FALSE}
kable(summary(shape_axes)[, 1:18], caption = "Variance per axis (PCn.var) and cumulative variance (PCn.sum) for each super order for the 9 first dimensions of the shape space. The \"noclade\" group contains all the species from superorders that contain less than 15 species each. Note that although the first five dimensions are enough to explain more than 95% of the whole space variance, this is not true for the Aequornithes, Columbimorphae, Galloanserae, Mirandornithes and Paleognathae.")
```

## Random terms sizes

```{r, echo = FALSE}
## Load the data
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Creating the subsets
superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

order_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Get the subset sizes into a table
super_orders <- size.subsets(superorder_shapespace)
orders <- size.subsets(order_shapespace)

## The table
data_table <- data.frame("rank" = c("class", rep("super order", length(super_orders)-1), rep("order", (length(orders)-1))), "species" = c(super_orders[length(super_orders)], super_orders[-length(super_orders)], orders[-length(orders)]))
rownames(data_table)[1] <- "Aves"

kable(data_table, caption = "Number of species for each group used as a random term in the analysis.")
```

## mcmcmcmcglmmm convergence diagnosis

```{r, echo = FALSE, fig.height = 9, fig.width = 6}
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")

## The diagnosis
diagnose <- diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order[[1]])

kable(do.call(rbind, lapply(diagnose, quantile)), caption = paste0("Distribution of the Effective sample for each terms in the model. The solutions (traits as explanatory variables) contained ", unlist(lapply(diagnose, length))[1], " terms; the Gcovariances (random terms, i.e. the phylogenetic components) contained ", unlist(lapply(diagnose, length))[2], " terms; and the Rcovariances (residual terms) contained ", unlist(lapply(diagnose, length))[3], " terms across 4000 posterior samples."))

pdf(file = "../Manuscript/Figures/Supplementary_figure_mcmc_ESS.pdf", height = 6, width = 3)
silent <- diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order[[1]])
dev.off()
```

### 2022-03-10 Analysis/Older_ideas/04-full_analyses_run.Rmd
rearranged Analysis folder

--- 2021-06-13 Analysis/04-full_analyses_run.Rmd
@@
 
+## Check the following options:
+
+ * saving parameters
+ * number of dimensions
+ * numbers of iterations (especially rate)
+ * check options in brms
+
 ```{r, run param chains}
@@
                                        nitt   = 5000000,
-                                       thin   = 1000,
+                                       thin   = 5000,
                                        burnin = 0))

### 2022-03-10 Analysis/sandbox.R

@@ -1,9 +1,0 @@
-
-#TODO:
-list_of_matrices <- get.covar(data, n)
-plot.ellipses(list_of_matrices)
-
-compare.matrices() # comparing the VCV matrices Robinson and Beckerman style
-
-get.major.axis.coords(list_of_matrices, axis = 1) # Getting the list of major axis (get the eigen value from it?)
-

### 2022-03-10 Analysis/05-Figure2.Rmd
Figure name files UP

---
title: "Bird beak's elaboration and innovation"
author: "Gavin Thomas"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

TODO!

### 2022-03-10 Analysis/04-Figure1.Rmd
Added lambdas?

@@
 
+## Calculate the lambdas for the whole tree and nested clades
+
+```{r}
+## Loading the clade info
+load("../Data/Processed/group_species_names.rda")
+
+## Loading the plotting data
+load("../Data/Processed/ploting_data.rda")
+
+## Function for measuring Pagel's lambda from a sub-tree
+#' @param species a vector of species names to extract from tree
+#' @param variable the variable (a named vector with the correct species names)
+#' @param tree the full tree
+#' @param method the method to be passed to motmot::transformPhylo.ML (default is "lambda")
+#' @param ... any other arguments to be passed to motmot::transformPhylo.ML
+get.p.lambda <- function(species, variable, tree, method = "lambda", ...) {
+    ## Selected species
+    to_keep <- tree$tip.label %in% species
+
+    ## Trim the tree
+    sub_tree <- drop.tip(tree, tip = tree$tip.label[!to_keep])
+
+    ## Trim the variable
+    sub_var <- variable[species]
+
+    ## Measure lambda
+    return(motmot::transformPhylo.ML(phy = sub_tree, y = matrix(sub_var, ncol = 1, dimnames = list(names(sub_var))), model = method, ...))
+}
+
+## Get the lambdas for all levels for the elaborations
+overall_lambda_elaboration <- get.p.lambda(species = ploting_data$consensus_tree$tip.label, variable = ploting_data$median_elaborations, tree = ploting_data$consensus_tree, modelCIs = FALSE)
+all_lambdas_elaboration <- lapply(group_species_names, lapply, function(x, tree, variable, ...) {cat(".") ; get.p.lambda(names(x), variable, tree, ...)},
+                                  tree = ploting_data$consensus_tree,
+                                  variable = ploting_data$median_elaborations,
+                                  modelCIs = FALSE)
+overall_lambda_innovation <- get.p.lambda(species = ploting_data$consensus_tree$tip.label, variable = ploting_data$median_innovations, tree = ploting_data$consensus_tree, modelCIs = FALSE)
+all_lambdas_innovation <- lapply(group_species_names, lapply, function(x, tree, variable, ...) {cat(".") ; get.p.lambda(names(x), variable, tree, ...)},
+                                  tree = ploting_data$consensus_tree,
+                                  variable = ploting_data$median_innovations,
+                                  modelCIs = FALSE)
+## Make into a table
+lambda_elaboration <- list(overall_lambda_elaboration$Lambda, lapply(all_lambdas_elaboration, lapply, function(X) return(X$Lambda)))
+lambda_innovation <- list(overall_lambda_innovation$Lambda, lapply(all_lambdas_innovation, lapply, function(X) return(X$Lambda)))
+lambda_table <- cbind(unlist(lambda_elaboration), unlist(lambda_innovation))
+lambdas_superorder <- lambda_table[c(2: 9), ]
+lambdas_order <- lambda_table[-c(1: 9), ]
+```
+
 ```{r}
@@
     # text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
-    text(0,0.3, paste0(names(name.col)), cex = 1.2, col = name.col)
+    text_main <- paste0(names(name.col))#, "\n", "λe = ", round(lambdas_superorder[paste0("superorders.", names(name.col)),][1], 2), "; λi = ", round(lambdas_superorder[paste0("superorders.", names(name.col)),][2], 2))
+    text(0,0.3, text_main, cex = 1.2, col = name.col)
 
@@
     # text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
-    text(0,0.3, paste0(names(name.col)), cex = 1.2, col = name.col)
+    text_main <- paste0(names(name.col))#, "\n", "λe = ", round(lambdas_order[paste0("orders.", names(name.col)),][1], 2), "; λi = ", round(lambdas_order[paste0("orders.", names(name.col)),][2], 2))
+    text(0,0.3, text_main, cex = 1.2, col = name.col)
 

### 2022-03-11 Analysis/05-Figure2.Rmd
Moved the lambda bit to figure 2

@@
 title: "Bird beak's elaboration and innovation"
-author: "Gavin Thomas"
+author: "Gavin Thomas & Thomas Guillerme"
 date: "`r Sys.Date()`"
@@
 
-TODO!
+## Calculate the lambdas for the whole tree and nested clades
+
+```{r}
+## Loading the clade info
+load("../Data/Processed/group_species_names.rda")
+
+## Loading the plotting data
+load("../Data/Processed/ploting_data.rda")
+
+## Function for measuring Pagel's lambda from a sub-tree
+#' @param species a vector of species names to extract from tree
+#' @param variable the variable (a named vector with the correct species names)
+#' @param tree the full tree
+#' @param method the method to be passed to motmot::transformPhylo.ML (default is "lambda")
+#' @param ... any other arguments to be passed to motmot::transformPhylo.ML
+get.p.lambda <- function(species, variable, tree, method = "lambda", ...) {
+    ## Selected species
+    to_keep <- tree$tip.label %in% species
+
+    ## Trim the tree
+    sub_tree <- drop.tip(tree, tip = tree$tip.label[!to_keep])
+
+    ## Trim the variable
+    sub_var <- variable[species]
+
+    ## Measure lambda
+    return(motmot::transformPhylo.ML(phy = sub_tree, y = matrix(sub_var, ncol = 1, dimnames = list(names(sub_var))), model = method, ...))
+}
+
+## Get the lambdas for all levels for the elaborations
+overall_lambda_elaboration <- get.p.lambda(species = ploting_data$consensus_tree$tip.label, variable = ploting_data$median_elaborations, tree = ploting_data$consensus_tree, modelCIs = FALSE)
+all_lambdas_elaboration <- lapply(group_species_names, lapply, function(x, tree, variable, ...) {cat(".") ; get.p.lambda(names(x), variable, tree, ...)},
+                                  tree = ploting_data$consensus_tree,
+                                  variable = ploting_data$median_elaborations,
+                                  modelCIs = FALSE)
+overall_lambda_innovation <- get.p.lambda(species = ploting_data$consensus_tree$tip.label, variable = ploting_data$median_innovations, tree = ploting_data$consensus_tree, modelCIs = FALSE)
+all_lambdas_innovation <- lapply(group_species_names, lapply, function(x, tree, variable, ...) {cat(".") ; get.p.lambda(names(x), variable, tree, ...)},
+                                  tree = ploting_data$consensus_tree,
+                                  variable = ploting_data$median_innovations,
+                                  modelCIs = FALSE)
+## Make into a table
+lambda_elaboration <- list(overall_lambda_elaboration$Lambda, lapply(all_lambdas_elaboration, lapply, function(X) return(X$Lambda)))
+lambda_innovation <- list(overall_lambda_innovation$Lambda, lapply(all_lambdas_innovation, lapply, function(X) return(X$Lambda)))
+lambda_table <- cbind(unlist(lambda_elaboration), unlist(lambda_innovation))
+lambdas_superorder <- lambda_table[c(2: 9), ]
+lambdas_order <- lambda_table[-c(1: 9), ]
+```

### 2022-03-15 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
 ---
+
+<!-- Compilation time: 30 minutes -->
 
@@
 ## Reading the distribution of trees
-trees_list <- read.tree("../Data/raw/Tree_data/AllBirdsHackett1.tre")
+trees_list <- read.tree("../Data/Raw/Tree_data/AllBirdsHackett1.tre")
 ```
@@
 
+## Passeriformes focus
+
+```{r}
+## Making a Passeriformes only list
+clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
+clades_list_passeriformes <- clades_list[which(clades_list$level2_order == "Passeriformes"), c(3, 5)]
+clades_list_passeriformes[,1] <- as.factor(as.character(clades_list_passeriformes[,1]))
+clades_list_passeriformes[,2] <- as.factor(as.character(clades_list_passeriformes[,2]))
+
+## Shapespace passeriformes
+shapespace_passeriformes <- shapespace[rownames(shapespace) %in% rownames(clades_list_passeriformes), ]
+
+## Do the prep data from here (with all the trees! ouch!)
+shapsepace_passeriformes_suborder_family <- prep.data(
+                                          level = 0, lvl.inc = c(1,2),
+                                          clades    = clades_list_passeriformes,
+                                          space     = shapespace_passeriformes,
+                                          dim       = 1:8,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapsepace_passeriformes_suborder_family, file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+```
+
 ## References

### 2022-03-15 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

--- 2022-02-24 Analysis/02-mcmcmcglmmm_analysis.Rmd
@@
 ---
-title: "Projection analysis to measure biological elaboration and innovation."
-author: "Thomas Guillerme, Natalie Cooper, Andrew P Beckerman,  Gavin H Thomas"
+title: "Bird beak's elaboration and exploration"
+author: "Thomas Guillerme"
+date: "`r Sys.Date()`"
 bibliography: references.bib
-date: "`r Sys.Date()`"
 output:
@@
     fig_height: 8
-    keep_tex: true
-  self_contained: true
 ---
 
-```{r header, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, eval = TRUE}
+This script is a working script to run the chains on the SHARC Sheffield cluster.
+Refer to `02-MCMCglmm_mini_chains.Rmd` for more details and reproducibility
+
+```{r load_packages, eval = TRUE, message = FALSE}
+set.seed(42)
+## Loading beer
 library(mcmcmcglmmm)
+library(MCMCglmm)
 library(dispRity)
+
+## Loading the parametrisation data
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
+
+## Select the data
+shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
+selected_dim <- shapespace_allbirds_lvl_superorder_order[[1]]$dimensions
+consensus_tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
+trees_list <- shapespace_allbirds_lvl_superorder_order[[1]]$trees_list
 ```
 
-# Mini chains MCMCglmm method: an efficient way to calculate variance-covariance matrices on big data with phylogenetic uncertainty
+# All birds shapespace analyses
 
-One commonly used method to estimate variance-covariance matrices from phylogenetic datasets is to use multivariate generalised linear mixed models.
-This is often done in a Bayesian way using the `MCMCglmm` package (@hadfield2010; cited nearly 4000 times in the last decade).
-The MCMCglmm method allows to run a generalised linear mixed model on a multidimensional dataset with an underlying phylogenetic structure.
-<!-- Expand that from the longevity paper -->.
+## Mini-chains parametrization
 
-Using this method we can run a nested phylogenetic model on a multidimensional dataset (here the Charadriiformes PCA shapespace).
-This model will have a the multidimensional trait values as a response and the classic residual terms as an error term.
-Additionally though, we can add a random error term as the phylogenetic signal.
-This random error term can be expanded into four independent nested terms: one for the entire phylogenetic structure and three for each individual clade's phylogenetic structure.
+```{r, parametrise_the_model, eval = FALSE}
+## Set up the parametrising chains models
+param_MCMCglmm <- make.mini.chains(data         = shapespace,
+                                   dimensions   = selected_dim,
+                                   tree         = consensus_tree,
+                                   trait.family = "gaussian",
+                                   randoms      = c("global", "level1", "level2"),
+                                   residuals    = "global",
+                                   priors       = 0.02,
+                                   verbose      = TRUE,
+                                   parameters   = list(
+                                       nitt   = 20000,
+                                       thin   = 500,
+                                       burnin = 0))
 
-$$\text{data} = \text{traits} + \text{residuals} + (\text{phylogenetic term} + \text{clade term})$$
+## Running the three MCMCglmm models
+100k_params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(param_chains, file = "../Data/Processed/MCMCglmm/100k_params_shapespace_8D_31levels_chain1.rda")
+```
 
-Or in `MCMCglmm` pseudo code:
+From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
+We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
+For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
+
+```{r, get_model_parameters, eval = FALSE}
+## Load the computed models
+param_chains <- list()
+for(i in 1:3) {
+    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/shapespace_allbirds_lvl_superorder_order_param", i, ".rda"))
+    param_chains[[i]] <- shapespace_allbirds_lvl_superorder_order_param[[1]]
+}
+## Extract the parameters from the parameter chains
+est_params <- extract.parameters(param_chains)
+```
+
+### Running the mini-chains
+
+```{r, set_the_mini_chains, eval = FALSE}
+## The thinning parameter
+thin <- 200
+## The number of samples per chains
+samples <- 10
+
+## Set up the mini-chains models
+shapespace_allbirds_lvl_superorder_order <- make.mini.chains(
+                                data         = shapespace,
+                                dimensions   = selected_dim,
+                                tree         = trees_list[sample(1:1000, 1000)],
+                                trait.family = "gaussian",
+                                randoms      = c("global", "level1", "level2"),
+                                residuals    = "global",
+                                priors       = est_params$priors,
+                                verbose      = TRUE,
+                                parameters   = list(
+                                    nitt   = est_params$burnin + thin * samples,
+                                    thin   = thin,
+                                    burnin = est_params$burnin))
+## Save the mini.chains
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order.mini.chains")
+```
+
+The chains are then run on the cluster
+TODO: details
+
+### Combining the model data
+
+```{r, eval = FALSE}
+## Set up the variables
+models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster/"
+## Getting all the model files
+model_files <- list.files(models_path, pattern = ".rda")
+
+## Check if the model exists already
+if(!("shapespace_allbirds_lvl_superorder_order.MCMCglmm" %in% list.files("../Data/Processed/"))) {
+    ## Preparing the model files
+    model_list <- list()
+    class(model_list) <- c("beer", "mini.chains")
+    ## Get the first model
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    ## Move it to the storage
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
+    model_list[[length(model_list)+1]] <- model[[1]]
+    model <- NULL
+    model_files <- model_files[-1]
+} else {
+    ## Updating the model
+    load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+    model_list <- shapespace_allbirds_lvl_superorder_order
+}
+
+## Get the subsequent models and merge them as we go
+while(length(model_files) != 0) {
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    ## Move it to storage
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
+    model_list[[length(model_list)+1]] <- model[[1]]
+    ## Merge the models (overwrite the model list)
+    model_list <- list(combine.mini.chains(model_list))
+    ## DEBUG
+    print("model list size:")
+    print(format(object.size(model_list), units = "MB"))
+    model <- NULL
+    model_files <- model_files[-1]
+}
+
+## Combining the model files
+shapespace_allbirds_lvl_superorder_order <- model_list
+
+## Saving the results
+save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
 
 ```
-formula   = PC1:8 ~ trait:clade-1
-random    = ~ us(at.level(clade):trait):animal + us(trait):animal
-residuals = ~ us(trait):units
+
+## Checking the mini chains' ESSs
+
+```{r, eval = TRUE}
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order[[1]])
 ```
-<!-- TG:TODO: explain that much better) -->
-
-Although this method is perfectly efficiently implemented on small to relatively big datasets, the major hurdle is the calculation of the variance-covariance matrix that increases following a power law with both the number of species and the number of dimensions.
-Therefore, such method becomes easily untractable on large datasets (>8k species and >3 dimensions).
-Furthermore, this method's phylogenetic correction is based on a single input tree (usually the consensus tree).
-This can also be problematic on big trees (i.e. > 5k species) where the notion of a consensus tree being a good representation of phylogenetic is at best misleading and at worth incorrect.
-In fact, the bigger the tree, the more there is variance in topological relation between species (especially in terms of branch length towards deep branches).
-Therefore, there has been a push to use tree distributions rather than consensus tree (i.e. point estimates) in using the MCMCglmm method (@healy2014; @mulTree).
-This however, leads to yet another untractable computer problem: it requires running the large dataset much more than once on multiple trees (e.g.
-if using @mulTree) thus linearly increasing computational time.
-
-Thus, to increase the speed of these analyses, while taking phylogenetic uncertainty into account, we used a highly parallelisable "mini chains" approach.
-In brief, it runs multiple short `MCMCglmm` analyses on multiple trees and pulls the results together into one larger `MCMCglmm` that contains more variation due to phylogenetic uncertainty (where the size of the chains are optimised for speed and low RAM usage).
-
-## Mini chains
-
-The mini chains `MCMCglmm` method (`mcmcmcglmmm`) effectively works by running many very short parallel `MCMCglmm` on as many different tree topologies.
-Effectively each mini chain generates a very small number of samples post-burnin (e.g.
-10) and combine them across many replicates (e.g.
-1000) into a chain containing effectively many samples (e.g.
-10000), each ran independently on a different tree toplogy.
-
-![Figure 2](../Figures/mini-chains_diagram.png)
-
-##### Figure 2: {#fig2}
-Mini chains diagram: the numerical values displayed here (number of chains, priors, etc...) are used as examples and can be tuned to fit specific questions and datasets.
-
-In more details, the method works as follows:
-
- 1.
-Running a small number of "parametrisation chains": these chains are run with the traits data and the model of interest on a consensus tree from the tree distribution, along with flat priors with a low belief parameter (effectively not putting any important weight on the priors).
- 2.
-Extracting parameters from the "parametrisation chains": this step allows to extract the parameters of interest from the posteriors of the "parametrisation chains".
-The parameters of interest are the conservative burnin average time (here defined as highest number of iterations required to reach the median posterior likelihood across the parametrisation chains with an additional 10% extra generations) and the mini chains priors (the median posterior model results - ignoring the previously estimated burnin - with a slightly higher belief parameter than for the previous chains).
- 3.
-Running the mini chains: these chains use the same data and model as for the "parametrisation chains" but use a random tree from the tree distribution rather than the consensus one.
-Furthermore, they use the parameters previously estimated (i.e.
-the priors and the burnin phase).
-These chains are run just for a small number of itterations past the burnin phase to extract a fixed number of samples (e.g.
-if the sampling rate is every 100 generations and the requested number of samples, the chain runs for $\text{burnin iterations} + \text{requested_samples} \times \text{sampling rate}$).
- 4. Combining the mini chains: finally we can combine the post-burnin samples from all the mini chains back into a single classical `MCMCglmm` chain (with no burnin).
-
-Some note about diagnosis.
-
-### Implementation
-
-The mini chain approach is implemented on the github package `mcmcmcglmmm` (link).
-The package comes with a demo vignette explaining in details how to reproduce each step in the procedure described above.
-
-### Effectiveness
-
-In a practical implication of the mini-chains, we ran for another project a similar model on 8748 species with 8 dimensions and 31 nested levels.
-Using this method, it took effectively 20 parallel cores with 8GB of RAM 2 months to complete the model with each individual mini-chain running for around 40 hours (using the SHARC cluster from Sheffield: CITE).
-We estimated that the same model using the normal `MCMCglmm` implementation would have required 2.1 years and 4TB of RAM.
-
-This method thus allows to run multivariate generalised linear mixed effects models on large datasets and across a tree distribution.
-The posteriors of this analyses contain a number of variance-covariance matrices (`model$VCV` in a `MCMCglmm` output) and their location in the trait space (`model$Sol` in a `MCMCglmm` output).
-We can then use these distributions of variance-covariance matrices as major axes in the multidimensional space on which to run the elaboration and exploration analyses.
-
-# Reproducible example
-
-Here we illustrate the BEER pipeline by applying it to a data set from @cooney2017 focused on the birds charadriiformes order.
-
-## Data
-
-This order contains 359 species divided into three clades: gulls, sandpipers and plovers and sandpipers that have respectively 159, 102 and 98 species each.
-For each of these species, we used the 3D beak dataset from @cooney2017 that is an ordination of the shapes of the beaks of 8@@@ birds (the resulting shape space used here does not contain any information about the beak size - i.e.
-centroid size).
-We extracted the 359 charadriiformes species from this space where the first three dimensions account for more than 99% of the variance in the dataset.
-For the rest of the example analysis here, we used as a traitspace of these 359 species with 3 dimensions (hereafter the shapespace).
-
-```{r, echo = FALSE, eval = TRUE}
-## Loading the charadriiformes data
-data(charadriiformes) 
-## Extracting the tree
-tree <- charadriiformes$tree
-## Extracting the data column that contains the clade assignments
-data <- charadriiformes$data[, "clade"]
-## Changing the levels names (the clade names) to colours
-levels(data) <- c("orange", "blue", "darkgreen")
-data <- as.character(data)
-## Matching the data rownames to the tip order in the tree
-data <- data[match(ladderize(tree)$tip.label, rownames(charadriiformes$data))]
-
-## Matching the tip colours (labels) to their descending edges in the tree
-## (and making the non-match edges grey)
-clade_edges <- match.tip.edge(data, tree, replace.na = "grey")
-
-## Plotting the results
-plot(ladderize(tree), show.tip.label = FALSE, edge.color = clade_edges, main = "Charadriiformes")
-legend("bottomleft", lty = c(1,1,1), col = c("blue", "darkgreen", "orange"), legend = c("plovers", "sandpipers", "gulls"))
-axisPhylo()
-```
-
-##### Figure 5: {#fig5}
-The Charadriiformes data used in this pipeline illustration. The axis units are millions of years ago.
-
-## Mini-chains MCMCglmm method
-
-To illustrate the BEER pipeline, we analysed the variance-covariance of the shapespace using a generalised linear mixed model with two different levels of random terms: one for the whole charadriiformes phylogeny and one different one for each different clade (gulls, plovers and sandpipers).
-We first run three parametrisation chains for 50000 generations with no burnin and a sampling every 500 generations.
-From this chain we extracted: 1) the maximum burnin time as 1.1 times the number of generations required to reach the median likelihood value and 2) the median posteriors of the model (discarding the burnin) as priors for our mini chains models with a believe parameter value of 5%.
-We then ran 50 minichains for 10 samples each using the previous sampling rate (500 generations) and the extracted parameters form the parametrisation chain (the burnin and priors described above).
-
-```{r, echo = FALSE, eval = FALSE}
-library(mcmcmcglmmm)
-
-## Making the parametrisation chains
-parametrisation_chains <- make.mini.chains(data = charadriiformes$data,
-                                           dimensions = c("PC1", "PC2", "PC3"),
-                                           randoms = c("global", "clade"),
-                                           residuals = "global",
-                                           tree = charadriiformes$tree,
-                                           parameters = list(nitt = 50000, burnin = 0, thin = 500))
-
-## Running the parametrisation chains
-param_results <- run.mini.chains(parametrisation_chains, replicates = 3)
-save(param_results, file = "param_results.rda")
-load("param_results.rda")
-
-diagnosis <- diagnose.mini.chains(param_results)
-## Extracting the parameters
-params <- extract.parameters(param_results)
-
-## The sampling rate:
-sampling <- 500
-## The required number of samples per tree
-samples <- 10
-
-## Making the mini chains
-parametrisation_chains <- make.mini.chains(data = charadriiformes$data,
-                                           dimensions = c("PC1", "PC2", "PC3"),
-                                           randoms = c("global", "clade"),
-                                           residuals = "global",
-                                           priors = params$priors,
-                                           tree = charadriiformes$tree_distribution,
-                                           parameters = list(
-                                            nitt = params$burnin + samples*sampling,
-                                            burning = params$burnin,
-                                            thin = sampling))
-## Running the mini chains
-results <- run.mini.chains(parametrisation_chains, replicates = 100)
-posteriors <- combine.mini.chains(results)
-save(posteriors, file = "posteriors.rda")
-```
-
-This resulted in obtaining 1000 exploitable posterior variance-covariance matrices for each residual terms of our model: the overall phylogenetic effect and the three specific phylogenetic effects for each of the three clades.
-
-```{r, echo = FALSE, eval = TRUE}
-load("posteriors.rda")
-silent <- diagnose.mini.chains(posteriors, plot = TRUE)
-```
-
-<!-- Do some developments about this part -->

### 2022-03-18 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

--- 2022-02-04 Analysis/03.1-results.Rmd
@@
 
-This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_innovation_analyses.Rmd.
+This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
+This script take around 6 minutes to compile.
 
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+superorder_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+order_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 
-TODO: centre and absolute directly in `dispRity.covar.projections`
-
 ```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}

### 2022-03-18 Analysis/04-Figure1.Rmd
Update 04-Figure1.Rmd

@@
 ```{r}
-
+source("../Functions/covar.utilities.R")
 box_size <- 1.5

### 2022-03-25 Analysis/02-mcmcmcglmmm_analysis.Rmd
Added passerines bits

@@
 ```
+
+## Passeriformes focus
+
+```{r load_packages, eval = TRUE, message = FALSE}
+set.seed(42)
+## Loading beer
+library(mcmcmcglmmm)
+library(MCMCglmm)
+library(dispRity)
+
+## Loading the parametrisation data
+load(file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+
+## Select the data
+shapespace <- shapsepace_passeriformes_suborder_family[[1]]$space
+selected_dim <- shapsepace_passeriformes_suborder_family[[1]]$dimensions
+consensus_tree <- shapsepace_passeriformes_suborder_family[[1]]$consensus_tree
+trees_list <- shapsepace_passeriformes_suborder_family[[1]]$trees_list
+```
+
+# All birds shapespace analyses
+
+## Mini-chains parametrization
+
+```{r, parametrise_the_model, eval = FALSE}
+## Set up the parametrising chains models
+param_MCMCglmm <- make.mini.chains(data         = shapespace,
+                                   dimensions   = selected_dim,
+                                   tree         = consensus_tree,
+                                   trait.family = "gaussian",
+                                   randoms      = c("global", "level1", "level2"),
+                                   residuals    = "global",
+                                   priors       = 0.02,
+                                   verbose      = TRUE,
+                                   parameters   = list(
+                                       nitt   = 100000,
+                                       thin   = 500,
+                                       burnin = 0))
+
+## Running the three MCMCglmm models
+save(param_chains, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_29levels_5k_chain1.rda")
+```
+
+From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
+We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
+For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
+
+```{r, get_model_parameters, eval = FALSE}
+## Load the computed models
+param_chains <- list()
+for(i in 1:3) {
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_shapespace_passeriformes_chain", i, ".rda"))
+    param_chains[[i]] <- params_shapespace_passeriformes[[1]]
+}
+## Extract the parameters from the parameter chains
+est_params <- extract.parameters(param_chains)
+```
+
+### Running the mini-chains
+
+```{r, set_the_mini_chains, eval = FALSE}
+## The thinning parameter
+thin <- 200
+## The number of samples per chains
+samples <- 10
+
+## Set up the mini-chains models
+shapespace_passeriformes <- make.mini.chains(
+                                data         = shapespace,
+                                dimensions   = selected_dim,
+                                tree         = trees_list[sample(1:1000, 1000)],
+                                trait.family = "gaussian",
+                                randoms      = c("global", "level1", "level2"),
+                                residuals    = "global",
+                                priors       = est_params$priors,
+                                verbose      = TRUE,
+                                parameters   = list(
+                                    nitt   = est_params$burnin + thin * samples,
+                                    thin   = thin,
+                                    burnin = est_params$burnin))
+## Save the mini.chains
+save(shapespace_passeriformes, file = "../Data/Cluster/shapespace_passeriformes/shapespace_passeriformes.mini.chains")
+```

### 2022-04-21 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
Renamed figure Rmds to contain also analyses bits

--- 2022-03-18 Analysis/04-Figure1.Rmd
@@
 ## Quick load
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_4k.rda")
 results <- shapespace_allbirds_lvl_superorder_order_results_list
@@
 ```{r}
+source("../Functions/covar.utilities.R")
 ## Plotting the ellipses
@@
 ```{r}
-source("../Functions/covar.utilities.R")
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure1_phylo_Dplot.pdf", height = box_size*2, width = box_size*2)
+# pdf(file = "../Manuscript/Figures/Figure1_phylo_Dplot.pdf", height = box_size*2, width = box_size*2)
 
-lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
-plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = "")
-abline(v = 0, col = "grey", lwd = 0.5)
-abline(h = 0, col = "grey", lwd = 0.5)
+# lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+# plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = "")
+# abline(v = 0, col = "grey", lwd = 0.5)
+# abline(h = 0, col = "grey", lwd = 0.5)
 
-covars_cent_tend <- lapply(sample.n(get.subsets(data, subset = "phylogeny")$covar, n), VCV.cent.tend, mean)
-dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-dim_var <- dim_var/max(dim_var)/10
-x_start <- 0.2
-y_start <- -0.1
-buffer <- 0
-for(one_dim in 1:length(dim_var)) {
-    lines(x = c(x_start, x_start+dim_var[one_dim]), y = rep(y_start+buffer, 2), lwd = 3, col = "grey")
-    buffer <- buffer - 0.02
-}
+# covars_cent_tend <- lapply(sample.n(get.subsets(data, subset = "phylogeny")$covar, n), VCV.cent.tend, mean)
+# dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+# dim_var <- dim_var/max(dim_var)/10
+# x_start <- 0.2
+# y_start <- -0.1
+# buffer <- 0
+# for(one_dim in 1:length(dim_var)) {
+#     lines(x = c(x_start, x_start+dim_var[one_dim]), y = rep(y_start+buffer, 2), lwd = 3, col = "grey")
+#     buffer <- buffer - 0.02
+# }
 
-dev.off()
+# dev.off()
 

### 2022-04-21 Analysis/05-Figure2_with_lambdas.Rmd
Renamed figure Rmds to contain also analyses bits

--- 2022-03-11 Analysis/05-Figure2.Rmd
@@
 lambda_table <- cbind(unlist(lambda_elaboration), unlist(lambda_innovation))
-lambdas_superorder <- lambda_table[c(2: 9), ]
-lambdas_order <- lambda_table[-c(1: 9), ]
+lambda_table <- data.frame(rbind(c("class", "aves"), do.call(rbind, strsplit(rownames(lambda_table), split = "\\."))), lambda_table)
+colnames(lambda_table) <- c("level", "name", "elaboration", "innovation")
+rownames(lambda_table) <- lambda_table$name
 ```

### 2022-04-21 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 
-# All birds shapespace analyses
-
 ## Mini-chains parametrization
@@
                                    parameters   = list(
-                                       nitt   = 100000,
+                                       nitt   = 50000,
                                        thin   = 500,

### 2022-04-21 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 names(tip_colours) <- names(sort(tip_order))
+tip_colours_super_orders <- tip_colours
+save(tip_colours_super_orders, file = "../Data/Processed/tip_colours_super_orders.rda")
 
@@
 names(tip_colours) <- names(sort(tip_order))
+tip_colours_orders <- tip_colours
+save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
 

### 2022-04-21 Analysis/05-Figure2_with_lambdas.Rmd
Update 05-Figure2_with_lambdas.Rmd

@@
 
-```{r}
+```{r, echo = FALSE}
 ## Loading the clade info
@@
 ```
+
+```{r}
+rownames(lambda_table) <- NULL
+## Save the lambda table:
+knitr::kable(lambda_table)
+```
+
+The problem here is that lambdas are single point estimates on single point estimates (median elaboration/innovation mapped on the consensus tree).
+It doesn't give much nuanced information and kind of defeats the `mcmcmcglmmmm` approach!
+Here's an approach getting the lambda for each tree.
+
+```{r, echo = FALSE, eval = FALSE}
+## Results loading
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+results <- shapespace_allbirds_lvl_superorder_order_results_list
+rm(shapespace_allbirds_lvl_superorder_order_results_list)
+## Trees loading
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
+## Select the trees
+trees <- shapespace_allbirds_lvl_superorder_order[[1]]$trees_list[sample(1:1000, 400)]
+rm(shapespace_allbirds_lvl_superorder_order)
+
+## Extract the relevant bits
+species_results <- list(
+      "all_birds_global" = lapply(results$super_results_phylo, get.subsets, subsets = "phylogeny"),
+      "super_global" = lapply(results$super_results_phylo, get.subsets, subsets = names(size.subsets(results$super_results_phylo[[1]]))[1:8]),
+      "super_within" = lapply(results$super_results_within, get.subsets, subsets = names(size.subsets(results$super_results_within[[1]]))[1:8]),
+      "order_global" = lapply(results$order_results_phylo, get.subsets, subsets = names(size.subsets(results$order_results_phylo[[1]]))[-28]),
+      "order_within" = lapply(results$order_results_within, get.subsets, subsets = names(size.subsets(results$order_results_within[[1]]))[-28])
+      )
+
+## Subsample 20 measurements for each
+get.subsamples <- function(X, n = 20) {
+  return(lapply(get.disparity(X, concatenate = FALSE), function(x, n) return(x[, sample(1:ncol(x), n), drop = FALSE]), n = n))
+}
+species_results <- lapply(species_results, lapply, get.subsamples)
+
+## Measure lambdas for each groups
+get.lambdas <- function(x, trees, verbose = TRUE) {
+
+    if(verbose) {
+      cat(paste0("Getting lambda for ", dim(x)[1], " tips:..."))
+    }
+
+    ## Get the species
+    species <- rownames(x)
+
+    ## Selected species
+    to_keep <- trees[[1]]$tip.label %in% species
+
+    ## Trim the tree
+    sub_trees <- lapply(trees[sample(1:length(trees), ncol(x))], drop.tip, tip = trees[[1]]$tip.label[!to_keep])
+    class(sub_trees) <- "multiPhylo"
+
+    ## Listify the variables
+    variables <- apply(x, 2, list)
+
+    ## Measure lambda
+    mapply.motmot <- function(data, tree) {
+      return(motmot::transformPhylo.ML(phy = tree, y = matrix(data[[1]], ncol = 1, dimnames = list(names(data[[1]]))), model = "lambda")$Lambda)
+    }
+    out <- t(mapply(mapply.motmot, data = variables, tree = sub_trees))
+    colnames(out) <- c("MLLambda", "LowerCI", "UpperCI")
+
+    if(verbose) {
+      cat(paste0("Done.\n"))
+    }
+
+    ## Measure lambda
+    return(out)
+}
+
+## Get the distributions of lambdas (20 per groups)
+lambda_distributions <- lapply(species_results, lapply, lapply, get.lambdas, trees = trees)
+save(lambda_distributions, file = "../Data/Processed/lambda_distributions.rda")
+```
+
+```{r, echo = FALSE, eval = TRUE, fig.height = 12, fig.width = 6}
+## Summarising the lambdas
+load(file = "../Data/Processed/lambda_distributions.rda")
+
+## Function for plotting the lambda lines
+plot.one.lambda <- function(loc = 1, one_lambda, col = "black") {
+  ## Add the lambda range
+  lines(y = rep(loc, 2), x = range(one_lambda), lwd = 1, col = col, lty = 3)
+  ## Add the median range
+  lines(y = rep(loc, 2), x = c(median(one_lambda[,"LowerCI"]), median(one_lambda[, "UpperCI"])), lty = 2, lwd = 1.5, col = col)
+  ## Add the mean range
+  lines(y = rep(loc, 2), x = range(one_lambda[, "MLLambda"]), lty = 1, lwd = 4, col = col)
+  ## Add the median point estimate
+  points(y = loc, x = median(one_lambda[, "MLLambda"]), pch = 19, col = col)
+}
+
+## List of elements to plot
+y_locs <- 1:length(c("Aves", names(lambda_distributions$super_global[[1]]), names(lambda_distributions$order_global[[1]])))
+names(y_locs) <- c("Aves", names(lambda_distributions$super_global[[1]]), names(lambda_distributions$order_global[[1]]))
+
+## List of colours
+## Colour palette is from figure 4
+load("../Data/Processed/tip_colours_orders.rda")
+load("../Data/Processed/tip_colours_super_orders.rda")
+## Set the background colours
+col <- rep("grey", length(y_locs))
+names(col) <- names(y_locs)
+col[names(tip_colours_super_orders)] <- tip_colours_super_orders
+col[names(tip_colours_orders)] <- tip_colours_orders
+
+## Selecting the right data and group ID
+get.data.ID <- function(one_group, type = "global") {
+  if(one_group == 1) {
+    ## All birds
+    return("all_birds_global")
+  }
+  if(one_group < 10) {
+    ## Super order level
+    if(type == "global") {
+        return("super_global")
+    } else {
+        return("super_within")
+    }
+  }
+  ## Order level
+  if(type == "global") {
+      return("order_global")
+  } else {
+      return("order_within")
+  }  
+}
+get.group.ID <- function(one_group) {
+  ## All birds
+  if(one_group == 1) {
+    return(1)
+  }
+  ## Super orders
+  if(one_group < 10) {
+    return(one_group - 1)
+  }
+  ## orders
+  return(one_group - 9)
+}
+
+par(mfrow = c(1,2))
+## Elaboration results
+par(mar = c(4, 8, 4, 0) + 0.1) # c(bottom, left, top, right) 
+plot(NULL, xlim = c(0,1), xlab = "Pagel's Lambda", ylim = rev(range(min(y_locs)-0.5, max(y_locs))), main = "Elaboration", ylab = "", yaxt = "n")
+axis(2, labels = names(y_locs), at = y_locs-0.25, las = 2)
+
+## Super orders
+for(one_group in y_locs) {
+    ## Adding the lambda distributions global
+    abline(h = y_locs[one_group]+0.25, col = "grey", lwd = 0.5)
+    plot_data <- lambda_distributions[[get.data.ID(one_group, type = "global")]][["position"]][[get.group.ID(one_group)]]
+    plot.one.lambda(loc = y_locs[one_group]-0.5, plot_data, col = col[[one_group]])
+    ## Adding the position within
+      plot_data <- lambda_distributions[[get.data.ID(one_group, type = "within")]][["position"]][[get.group.ID(one_group)]]
+    plot.one.lambda(loc = y_locs[one_group], plot_data, col = adjustcolor(col[[one_group]], alpha = 0.5))
+}
+
+## Innovation results
+par(mar = c(4, 0, 4, 8) + 0.1)
+plot(NULL, xlim = c(0,1), xlab = "Pagel's Lambda", ylim = rev(range(min(y_locs)-0.5, max(y_locs))), main = "Innovation", ylab = "", yaxt = "n")
+for(one_group in y_locs) {
+    ## Adding the lambda distributions global
+    abline(h = y_locs[one_group]+0.25, col = "grey", lwd = 0.5)
+    plot_data <- lambda_distributions[[get.data.ID(one_group, type = "global")]][["distance"]][[get.group.ID(one_group)]]
+    plot.one.lambda(loc = y_locs[one_group]-0.5, plot_data, col = col[[one_group]])
+    ## Adding the position within
+      plot_data <- lambda_distributions[[get.data.ID(one_group, type = "within")]][["distance"]][[get.group.ID(one_group)]]
+    plot.one.lambda(loc = y_locs[one_group], plot_data, col = adjustcolor(col[[one_group]], alpha = 0.5))
+}
+
+```

### 2022-04-25 Analysis/03-Elaboration_innovation_analysis.Rmd
Clade info for tree ploting

@@
 save(ploting_data, file = "ploting_data.rda")
+
+## Group info
+super_orders <- lapply(results$super_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$super_results_phylo$position$matrix[[1]])
+orders <- lapply(results$order_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$order_results_phylo$position$matrix[[1]])
+## Adding the colours
+load("../Data/Processed/tip_colours_super_orders.rda")
+load("../Data/Processed/tip_colours_orders.rda")
+add.col <- function(one_order, one_col) {
+    one_order$col <- one_col
+    return(one_order)
+}
+super_orders <- mapply(add.col, one_order = super_orders[1:8], one_col = as.list(tip_colours_super_orders), SIMPLIFY = FALSE)
+orders <- mapply(add.col, one_order = orders[1:27], one_col = as.list(tip_colours_orders), SIMPLIFY = FALSE)
+
+## Big list:
+clade_info <- list(super_orders = super_orders,
+                   orders = orders)
+## Save all that data
+save(clade_info, file = "../Data/Processed/clade_info.rda")
+
+## Get the names
+names(clade_info$orders)
+
+## Get the colours
+unlist(lapply(clade_info$orders, `[[`, "col"))
+
+## Get the species in one specific group
+clade_info$orders$Cuculiformes$species
+
 ```

### 2022-04-27 Analysis/05-Figure2_with_lambdas.Rmd
Update 05-Figure2_with_lambdas.Rmd

@@
 
-```{r, echo = FALSE, eval = TRUE, fig.height = 12, fig.width = 6}
+```{r, echo = FALSE, eval = TRUE, fig.height = 18, fig.width = 12}
 ## Summarising the lambdas

### 2022-04-27 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
updated figure 4 with thermos

@@
     ## Empty plot
-    par(bty = "n")
+    par(bty = "o")
     plot(NULL, xlim = c(-0.5,1.5), ylim = c(0,lim_up), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
@@
     ## Plot the elaboration/exploration
-    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
+    par(mar = c(2, 1, 1, 1)+0.1)
     plot.one.proj.rej(results$super_results_between, name.col, x = TRUE, y = FALSE)

### 2022-05-16 Analysis/06-Figure3_with_correlations.Rmd
sync

--- 2022-02-22 Analysis/09-Figure3.Rmd
@@
 ```{r, echo = FALSE, eval = FALSE}
+## Loading the distances from centre for each species
+load("../Data/Processed/ploting_data.rda")
+centroids <- ploting_data$distances_to_centre[[1]]
 
@@
 # layout.show(plot_layout)
-
-one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, legend_counter) {
+one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, legend_counter, contour = NULL) {
 
@@
 
-    ## Plot the points
-    plot(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = plot_data$col, xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = strsplit(cor.results$main, split = ":")[[1]][1])
+    ## Empty plot
+    plot(NULL, xlim = range(plot_data$x), ylim = range(plot_data$y), xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = strsplit(cor.results$main, split = ":")[[1]][1])
     axis(1, at = pretty(plot_data$x, n = 4))
     axis(2, at = pretty(plot_data$y, n = 4), las = 2)
+
+    ## Adding contours
+    if(!is.null(contour)) {
+        ## Select the right contour data
+        contour_data <- contour[match(rownames(plot_data), names(contour))]        
+        ## Calculate the contour
+        contour_z <- MASS::kde2d(contour_data/max(contour_data)*max(plot_data$x), contour_data/max(contour_data)*max(plot_data$y), n = 50)
+
+        contour_z <- matrix(0, nrow = length(contour_data), ncol = length(contour_data))
+        diag(contour_z) <-  contour_data
+        # contour_z <- outer(contour_data/max(contour_data)*max(plot_data$x), contour_data/max(contour_data)*max(plot_data$y))
+
+        ## Adding the contour in the background
+        contour(contour_z, lwd = 1, add = TRUE, col = hcl.colors(12, palette = "viridis"), labels = pretty(contour_data))
+    }
+
+    # ## col test for contour
+    # color.gradient <- function(x, colors=c("red","yellow","green"), colsteps=100) {
+    #     return( colorRampPalette(colors) (colsteps) [ findInterval(x, seq(min(x),max(x), length.out=colsteps)) ] )
+    # }
+
+    ## Plot the points
+    points(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = plot_data$col)
+    # points(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = color.gradient(contour_data))
 
@@
     par(mar = c(2,2.5,2,1.5))
-    one.plot.correlations(super_phylo_cor[[i]],col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter)
+    one.plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = centroids)
 }

### 2022-05-16 Analysis/Older_ideas/08-Number_of_dimensions.Rmd
sync

--- 2021-07-20 Analysis/08-Number_of_dimensions.Rmd
@@
 
-Checking the ideal number of dimensions for running the model.
-
-```{r, load pkg and data}
+```{r, echo = FALSE, print = FALSE}
+library(dispRity)
 library(beer)
@@
 
-## Selecting axis function
-# @params strict whether to have less than the threshold (strict) or at least the threshold (false)
-select.axis <- function(ordination, threshold = 0.99, group, details = FALSE, strict = FALSE) {
-    ## Set the group
-    if(missing(group)) {
-        group <- list(rownames(ordination))
-    }
+## Get the results for each group with > 10 species
+min_sp <- 10
+test_sizefree_order <- select.axes(full_PCAs$sizefree, threshold = 0.95, group = order_groups[-which(unlist(lapply(order_groups, length)) < min_sp)])
+test_sizefree_family <- select.axes(full_PCAs$sizefree, threshold = 0.95, group = family_groups[-which(unlist(lapply(family_groups, length)) < min_sp)])
+test_sizeform_order <- select.axes(full_PCAs$sizeform, threshold = 0.95, group = order_groups[-which(unlist(lapply(order_groups, length)) < min_sp)])
+test_sizeform_family <- select.axes(full_PCAs$sizeform, threshold = 0.95, group = family_groups[-which(unlist(lapply(family_groups, length)) < min_sp)])
+```
 
-    ## Cum sum scaled variance
-    get.cum.var <- function(one_group, matrix) {
-        if(length(one_group) != 1) {
-            vars <- apply(matrix[one_group, ], 2, var)
-            return(cumsum(vars/sum(vars)))
-        } else {
-            return(c(0, rep(1, ncol(matrix)-1)))
-        }
-    }
+## Size free PCA loadings per order
 
-    ## Get the variances for each group
-    group_vars <- lapply(group, get.cum.var, matrix = ordination)
+```{r, fig.height = 14, fig.width = 12}
+plot(test_sizefree_order)
+```
 
-    select.threshold <- function(x, threshold, strict) {
-        if(strict) {
-            return(which(x < threshold))
-        } else {
-            return(c(0, which(x < threshold)) + 1)
-        }
-    }
+## Size free PCA loadings per family
 
-    ## Get the axis per threshold
-    selected_dims <- lapply(group_vars, select.threshold, threshold = threshold, strict = strict)
+```{r, fig.height = 28, fig.width = 24}
+plot(test_sizefree_family)
+```
 
-    ## Return the results
-    if(!details) {
-        return(unique(unlist(selected_dims)))
-    } else {
-        return(selected_dims)
-    }
-}
+## Size + form PCA loadings per order
 
-## Preparing the result table
-thresholds <- seq(from = 0.9, to = 0.99, by = 0.01)
-results <- matrix(0, 6, length(thresholds))
-rownames(results) <- c("sizefree-all", "sizefree-order", "sizefree-family", "sizeform-all", "sizeform-order", "sizeform-family")
-colnames(results) <- thresholds
+```{r, fig.height = 14, fig.width = 12}
+plot(test_sizeform_order)
+```
 
-for(i in 1:length(thresholds)) {
-  results[1,i] <- length(select.axis(full_PCAs$sizefree, threshold = thresholds[i]))
-  results[2,i] <- length(select.axis(full_PCAs$sizefree, threshold = thresholds[i], group = order_groups))
-  results[3,i] <- length(select.axis(full_PCAs$sizefree, threshold = thresholds[i], group = family_groups))
-  results[4,i] <- length(select.axis(full_PCAs$sizeform, threshold = thresholds[i]))
-  results[5,i] <- length(select.axis(full_PCAs$sizeform, threshold = thresholds[i], group = order_groups))
-  results[6,i] <- length(select.axis(full_PCAs$sizeform, threshold = thresholds[i], group = family_groups))
-}
+## Size + form PCA loadings per family
 
-## Colour vector
-colours <- c("purple", "red", "orange",
-             "blue", "cyan", "green")
-
-## Plotting the results
-plot(NULL, xaxt = "n", xlim = c(1,10), ylim = range(results), ylab = "Dimensions", xlab = "Thresholds")
-axis(1, at = c(1:10), labels = thresholds)
-abline(h = 7, col = "lightgrey")
-for(i in 1:6) {
-  lines(results[i,], col = colours[i])
-}
-legend("topleft", legend = rownames(results), lty = 1, col = colours)
+```{r, fig.height = 28, fig.width = 24}
+plot(test_sizeform_family)
 ```

### 2022-05-16 Functions/Cluster/make.minichains.jobs.sh

@@ -39,8 +39,6 @@ cp job_template.job ${LOCALPATH}${CHAINNAME}_template.job
 echo "" >> ${LOCALPATH}${CHAINNAME}_template.job
 echo "## Run the script" >> ${LOCALPATH}${CHAINNAME}_template.job
 echo "Rscript ${LOCALPATH}${CHAINNAME}_<REPLICATE>.R" >> ${LOCALPATH}${CHAINNAME}_template.job
-## Bug on this line!!!!
-
 
 ## Creating the Rscript template
 echo ".libPaths(\"~/homelib\")" > ${LOCALPATH}${CHAINNAME}_template.R
@@ -48,7 +46,8 @@ echo "library(MCMCglmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
 echo "library(mcmcmcglmmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
 echo "load(\"${LOCALPATH}${CHAINNAME}.mini.chains\")" >> ${LOCALPATH}${CHAINNAME}_template.R
 echo "model <- run.mini.chains(${CHAINNAME}, replicates = 1, record.tree = TRUE)" >> ${LOCALPATH}${CHAINNAME}_template.R
-echo "save(model, file = \"${LOCALPATH}${CHAINNAME}_<REPLICATE>.rda\")" >> ${LOCALPATH}${CHAINNAME}_template.R
+# echo "save(model, file = \"${LOCALPATH}${CHAINNAME}_<REPLICATE>.rda\")" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "save(model, file = \"/data/bo1tgu/shapespace_allbirds_lvl_superorder_order_tmp_store/${CHAINNAME}_<REPLICATE>.rda\")" >> ${LOCALPATH}${CHAINNAME}_template.R
 
 
 ## Get the number of digits in REPLICATES

### 2022-05-16 Functions/test.rotation.R

@@ -19,13 +19,16 @@ plot.base <- function(matrix, base_vector, lim) {
     } else {
         par(bty = "n")
         plot(NULL, xlim = lim, ylim = lim, xlab = "x", ylab = "y")
-        abline(v = 0, col = "grey")
-        abline(h = 0, col = "grey")
+        abline(v = 0, col = "grey", lwd = 0.5)
+        abline(h = 0, col = "grey", lwd = 0.5)
         text(matrix[,1:2], labels = rownames(matrix), col = "grey", cex = 1.5)
+        for(i in 1:nrow(matrix)) {
+            lines(rbind(base_vector[1,], matrix[i,]), col = "grey", lwd = 1, lty = 2)
+        }
         lines(base_vector, col = "grey", lwd = 1.5)
     }
 }
-plot.recentred <- function(matrix, base_vector) {
+plot.recentred <- function(matrix, base_vector, col = c("orange", "blue", "darkgreen", "black", "black")) {
     if(ncol(matrix) == 3) {
         text3d(matrix, texts = rownames(matrix), col = "black")
         for(i in 1:nrow(matrix)) {
@@ -33,14 +36,14 @@ plot.recentred <- function(matrix, base_vector) {
         }
         segments3d(base_vector, col = "black", lwd = 2)
     } else {
-        text(matrix, labels = rownames(matrix), col = "black", cex = 1.5)
         for(i in 1:nrow(matrix)) {
-            lines(rbind(c(0,0), matrix[i,]), col = "grey", lwd = 1.5)
+            lines(rbind(c(0,0), matrix[i,]), col = "grey", lwd = 1,)
         }
-        lines(base_vector, col = "black", lwd = 1.5)
+        lines(base_vector, col = "grey", lwd = 4)
+        text(matrix, labels = rownames(matrix), col = col, cex = 1.5)
     }
 }
-plot.projections <- function(matrix, projections, rejections) {
+plot.projections <- function(matrix, projections, rejections, col = c("orange", "blue", "darkgreen", "black", "black")) {
     if(ncol(matrix) == 3) {
         for(i in 1:nrow(matrix)) {
             ## Plot the projections
@@ -50,15 +53,22 @@ plot.projections <- function(matrix, projections, rejections) {
         }
 
     } else {
-        for(i in 1:nrow(matrix)) {
+        ## Sort projections in increasing orders
[…]

### 2022-05-20 Analysis/06-Figure3_with_correlations.Rmd
Update 06-Figure3_with_correlations.Rmd

@@
 ```{r, echo = FALSE, eval = FALSE}
-library(ggplot2)
-## Getting the data to plot
-global_data <- super_phylo_cor$phylogeny
+pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4.5, width = 4)
 
-gg.density.plot <- function(data, palette = "viridis") {
-    ggplot(data.frame("elaboration" = data$x, "exploration" = data$y), aes(x=elaboration, y=exploration) ) +
-      geom_hex() +
-      scale_fill_continuous(type = palette) +
-      theme_bw()
+data <- super_phylo_cor$phylogeny
+scale.rgb <- function(x, y, max.x, max.y, alpha = 1) {
+    return(rgb(red = 255-((x/max.x)*255), blue = 255-((y/max.y)*255), green = 255, maxColorValue = 255, alpha = alpha*255))
+}
+add.quadrant <- function(data) {
+    max.x = max(range(pretty(data$x)))
+    max.y = max(range(pretty(data$y)))
+
+    ## Bottom left
+    polygon(x = c(0, 0, mean(range(pretty(data$x))), mean(range(pretty(data$x)))),
+            y = c(0, mean(range(pretty(data$y))), mean(range(pretty(data$y))), 0),
+            col = scale.rgb(0, 0, max.x, max.y, alpha = 1/3), border = "white")
+    ## Bottom right
+    polygon(x = c(mean(range(pretty(data$x))), mean(range(pretty(data$x))), max(range(pretty(data$x))), max(range(pretty(data$x)))),
+            y = c(0, mean(range(pretty(data$y))), mean(range(pretty(data$y))), 0),
+            col = scale.rgb(max(range(pretty(data$x))), 0, max.x, max.y, alpha = 1/3), border = "white")
+    ## Top left
+    polygon(x = c(0, 0, mean(range(pretty(data$x))), mean(range(pretty(data$x)))),
+            y = c(mean(range(pretty(data$y))), max(range(pretty(data$y))), max(range(pretty(data$y))), mean(range(pretty(data$y)))),
+            col = scale.rgb(0, max(range(pretty(data$y))), max.x, max.y, alpha = 1/3), border = "white")
+    ## Top right
+    polygon(x = c(mean(range(pretty(data$x))), mean(range(pretty(data$x))), max(range(pretty(data$x))), max(range(pretty(data$x)))),
+            y = c(mean(range(pretty(data$y))), max(range(pretty(data$y))), max(range(pretty(data$y))), mean(range(pretty(data$y)))),
+            col = scale.rgb(max(range(pretty(data$x))), max(range(pretty(data$y))), max.x, max.y, alpha = 1/3), border = "white")
+
+    abline(v = mean(range(pretty(data$x))), lwd = 0.5, lty = 2)
+    abline(h = mean(range(pretty(data$y))), lwd = 0.5, lty = 2)
+}
+quadrant.color <- function(elaboration, innovation, exact.color = FALSE, alpha = 1/3) {
+
+    ## Max values (from the whole dataset)
+    max.x <- 2.5
+    max.y <- 1.4
+
+    if(exact.color) {
+        return(rgb(red = 255-((elaboration/max.x)*255), blue = 255-((innovation/max.y)*255), green = 255, maxColorValue = 255, alpha = alpha*255))
     }
 
-pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4, width = 4)
-gg.density.plot(global_data)
+    ## Get the position in the quadrant
+    quadrant <- as.integer(c(elaboration < (max.x/2), innovation < (max.y/2)))
+
+    ## Get the position name
+    if(all(quadrant == c(1, 1))) {
+        return(rgb(red = 255, blue = 255, green = 255, maxColorValue = 255, alpha = alpha*255))
+    } else {
+        if(all(quadrant == c(1, 0))) {
+            return(rgb(red = 0, blue = 255, green = 255, maxColorValue = 255, alpha = alpha*255))
+        } else {
+            if(all(quadrant == c(0, 1))) {
+                return(rgb(red = 255, blue = 0, green = 255, maxColorValue = 255, alpha = alpha*255))
+            } else {
+                if(all(quadrant == c(0, 0))) {
+                    return(rgb(red = 0, blue = 0, green = 255, maxColorValue = 255, alpha = alpha*255))
+                }
+            }
+        }
+    }
+}
+
+max.x = max(range(pretty(data$x)))
+max.y = max(range(pretty(data$y)))
+
+par(mar = c(5, 4, 4, 2) + 0.1)
+plot(NULL, xlab = "elaboration", ylab = "innovation", xlim = range(pretty(data$x)), ylim = range(pretty(data$y)))
+add.quadrant(data)
+# points(data$x, data$y, pch = 21, col = scale.rgb(data$x, data$y, max.x = max.x, max.y = max.y))
+colour_densities <- densCols(cbind(data$x, data$y), colramp = scales::viridis_pal(alpha = 1, begin = 0, end = 1, direction = 1, option = "D"))
+points(cbind(data$x, data$y), col = colour_densities, pch = 20)
+
 dev.off()
@@
 # layout.show(plot_layout)
-one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, legend_counter, contour = NULL) {
+one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, legend_counter, contour = NULL, with.quadrant = FALSE) {
 
@@
     axis(2, at = pretty(plot_data$y, n = 4), las = 2)
+
+    if(with.quadrant) {
+        add.quadrant(plot_data)
+    }
 
@@
     par(mar = c(2,2.5,2,1.5))
-    one.plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = centroids)
+    one.plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
 }
@@
     par(mar = c(2,2.5,2,1.5))
-    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter)
+    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = TRUE)
 }

### 2022-05-20 Analysis/06-Figure3_with_correlations.Rmd
Added contour estimations test

@@
     if(!is.null(contour)) {
-        ## Select the right contour data
+
+        ## Get the contour data
         contour_data <- contour[match(rownames(plot_data), names(contour))]
-        ## Calculate the contour
-        contour_z <- MASS::kde2d(contour_data/max(contour_data)*max(plot_data$x), contour_data/max(contour_data)*max(plot_data$y), n = 50)
 
-        contour_z <- matrix(0, nrow = length(contour_data), ncol = length(contour_data))
-        diag(contour_z) <-  contour_data
-        # contour_z <- outer(contour_data/max(contour_data)*max(plot_data$x), contour_data/max(contour_data)*max(plot_data$y))
-
-        ## Adding the contour in the background
-        contour(contour_z, lwd = 1, add = TRUE, col = hcl.colors(12, palette = "viridis"), labels = pretty(contour_data))
+        ## This function is copied/modified from ContourFunctions
+        estimate.contour <- function(x, y, z) {
+            X <- data.frame(x, y, z)
+            ## Set the model for estiamtions
+            lfmod <- locfit::locfit(z ~ x + y, data=X, family=family)
+            ## Predicting function
+            pred.func <- function(xx) {
+              predict(lfmod, data.frame(x=xx[,1], y=xx[,2]))
+            }
+            ## Calculating the contour matrix
+            return(ContourFunctions::eval_over_grid_with_batch(x = seq(min(x), max(x), length.out = 100),
+                                                               y = seq(min(y), max(y), length.out = 100),
+                                                               fn = pred.func,
+                                                               batchmax = 500))
         }
 
+        ## Plot the estimated contour
+        contour(estimate.contour(x = plot_data$x, y = plot_data$y, z = unname(contour_data)),
+                lwd = 1, add = TRUE,
+                col = hcl.colors(12, palette = "viridis"),
+                labels = pretty(contour_data))
+
+        ## TODO: scale and rescale the data (so that it's beyond c(0,1))
+        ## TODO: check bug with Strisores (i = 7)
+
+        ## TESTING
         # ## col test for contour
@@
 
+        # plot(NULL, xlim = range(plot_data$x), ylim = range(plot_data$y), xlab = "", ylab = "", main = strsplit(cor.results$main, split = ":")[[1]][1])
+        # points(plot_data$x, plot_data$y, pch = 19, cex = 1, col = color.gradient(contour_data))
+    }
+
     ## Plot the points
     points(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = plot_data$col)
-    # points(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = color.gradient(contour_data))
 
@@
     par(mar = c(2,2.5,2,1.5))
-    one.plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
+    one.plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = centroids, with.quadrant = TRUE)
 }

### 2022-05-25 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 ```
+
+### Combining the model data
+
+```{r, eval = FALSE}
+## Set up the variables
+models_path <- "../Data/Cluster/shapespace_passeriformes/cluster/"
+## Getting all the model files
+model_files <- list.files(models_path, pattern = ".rda")
+
+## Check if the model exists already
+if(!("shapespace_passeriformes.MCMCglmm" %in% list.files("../Data/Processed/"))) {
+    ## Preparing the model files
+    model_list <- list()
+    class(model_list) <- c("mini.chains")
+    ## Get the first model
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    ## Move it to the storage
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
+    model_list[[length(model_list)+1]] <- model[[1]]
+    model <- NULL
+    model_files <- model_files[-1]
+} else {
+    ## Updating the model
+    load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+    model_list <- shapespace_allbirds_lvl_superorder_order
+}
+
+## Get the subsequent models and merge them as we go
+while(length(model_files) != 0) {
+    load(paste0(models_path, model_files[1], collapse = "/"))
+    ## Move it to storage
+    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
+    print(model_files[1])
+    ## Clean the bits that are not going to be used (but are huge)
+    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
+    ## Update the model list
+    model_list[[length(model_list)+1]] <- model[[1]]
+    ## Merge the models (overwrite the model list)
+    model_list <- list(combine.mini.chains(model_list))
+    ## DEBUG
+    print("model list size:")
+    print(format(object.size(model_list), units = "MB"))
+    model <- NULL
+    model_files <- model_files[-1]
+}
+
+## Combining the model files
+shapespace_passeriformes <- model_list
+
+## Saving the results
+save(shapespace_passeriformes, file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+
+```

### 2022-05-25 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 ```
+
+# All birds shapespace analyses
+
+```{r, eval = TRUE, echo = FALSE}
+## Load the data
+load(file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+## Select the data
+shapespace <- shapsepace_passeriformes_suborder_family[[1]]$space
+tree <- shapsepace_passeriformes_suborder_family[[1]]$consensus_tree
+## Clean garbage
+rm(shapsepace_passeriformes_suborder_family)
+
+## Load the covar
+load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+shapespace_passeriformes <- shapespace_passeriformes[[1]]
+
+## Get the levels
+phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
+suborder_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
+family_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))
+
+## Creating the subsets
+suborder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+      data          = shapespace,
+      posteriors    = shapespace_passeriformes,
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(suborder_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
+
+family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+      data          = shapespace,
+      posteriors    = shapespace_passeriformes,
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(family_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+## Clean garbage
+rm(shapespace_passeriformes)
+```
+
+## Calculate the elaborations/innovations
+
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
+## Run the analyses between groups
+suborder_results_between <- dispRity.covar.projections(suborder_shapespace,
+                                             type = "groups",
+                                             base = "phylogeny",
+                                             verbose = TRUE,
+                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
+## Run the analyses between groups
+family_results_between <- dispRity.covar.projections(family_shapespace,
+                                             type = "groups",
+                                             base = "phylogeny",
+                                             verbose = TRUE,
+                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
+## Run the analyses of groups on the phylogeny
+suborder_results_phylo <- dispRity.covar.projections(suborder_shapespace,
+                                            type = "elements",
+                                            base = "phylogeny",
+                                            inc.base = TRUE,
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+
+## Run the analyses of groups on the phylogeny
+family_results_phylo <- dispRity.covar.projections(family_shapespace,
+                                            type = "elements",
+                                            base = "phylogeny",
+                                            inc.base = TRUE,
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+## Run the analyses groups on themselves
+suborder_results_within <- dispRity.covar.projections(suborder_shapespace,
+                                            type = "elements",
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+## Run the analyses groups on themselves
+family_results_within <- dispRity.covar.projections(family_shapespace,
+                                            type = "elements",
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+shapespace_passeriformes_suborder_family_results_list <- list(
+        "suborder_results_between" = suborder_results_between,
+        "family_results_between"   = family_results_between,
+        "suborder_results_phylo"   = suborder_results_phylo,
+        "family_results_phylo"     = family_results_phylo,
+        "suborder_results_within"  = suborder_results_within,
+        "family_results_within"    = family_results_within)
+save(shapespace_passeriformes_suborder_family_results_list, file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
+```

### 2022-05-25 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
Updated passeriformes (with TODOS for functionalisation)

@@
 ```
+
+# Passeriformes bit
+
+```{r, eval = TRUE, echo = FALSE}
+## Load the data
+load(file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+## Select the data
+shapespace <- shapsepace_passeriformes_suborder_family[[1]]$space
+tree <- shapsepace_passeriformes_suborder_family[[1]]$consensus_tree
+## Clean garbage
+rm(shapsepace_passeriformes_suborder_family)
+
+## Load the covar
+load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+shapespace_passeriformes <- shapespace_passeriformes[[1]]
+
+## Get the levels
+phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
+subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
+famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))
+
+## Creating the subsets
+suborder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+      data          = shapespace,
+      posteriors    = shapespace_passeriformes,
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
+
+family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+      data          = shapespace,
+      posteriors    = shapespace_passeriformes,
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+## Clean garbage
+rm(shapespace_passeriformes)
+```
+
+First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):
+
+```{r, echo = FALSE}
+## Quick load
+load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
+results <- shapespace_passeriformes_suborder_family_results_list
+rm(shapespace_passeriformes_suborder_family_results_list)
+```
+
+## Phylogeny window (detailed)
+
+```{r}
+
+source("../Functions/covar.utilities.R")
+source("../Functions/ploting.utilities.R")
+
+box_size <- 1.5
+pdf(file = "../Manuscript/Figures/Figure1_passeriformes.pdf", height = box_size*7, width = box_size*10)
+
+## Global window
+lay_mat_family <- rbind(
+        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 1
+        matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
+        matrix(c(1,1, 18:25), ncol = 10, nrow = 1), # line 3
+        matrix(c(26:35), ncol = 10, nrow = 1), # line 4
+        matrix(c(36:45), ncol = 10, nrow = 1), # line 5
+        matrix(c(46:55), ncol = 10, nrow = 1) # line 6
+    )
+lay_mat_sub <- rbind(
+        matrix(c(1,1, 2:5, rep(0, 4)), ncol = 10, nrow = 1) # line 7
+    )
+
+heights_vals <- rep(1, 8)
+widths_vals <- rep(c(1, 1/3), 5)
+layout_matrix <- layout(rbind(lay_mat_sub, (lay_mat_family+5)), widths = widths_vals, heights = heights_vals, respect = TRUE)
+
+# layout.show(layout_matrix)
+
+## TODO: FUNCTIONALISE THE TREE PLOTTING: START
+
+## PLOTTING THE SUBORDERS
+tree <- ladderize(tree)
+## Get the tip_values
+families_lvl <- shapespace$level2
+suborders_lvl <- shapespace$level1
+
+## Select the level
+selected_level <- coloured_levels <- suborders_lvl
+
+## Replace the order levels as colours
+colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+
+## Shorten the tree
+selected_tips <- character()
+for(one_level in 2:length(levels(selected_level))) {
+    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
+    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
+}
+
+## Shorten the tree
+tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
+tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
+
+## Setting the tips edges colours
+tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
+## Rainbow the tip edges
+tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
+## Adding the number of tips
+tree_plot <- tree_short
+n_species <- table(selected_level)
+for(i in 1:Ntip(tree_plot)) {
+    ## Get the number of species
+    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
+    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
+}
+
+## Plotting the tree
+par(mar = c(0, 0, 0, 0)+0.1)
+plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
+
+## TODO: FUNCTIONALISE THE TREE PLOTTING: END
+
+## Matching the order and the colours to the ladderised tree
+tip_order <- c(
+    "Tyranni" = 1,
+    "Passeri" = 2)
+
+tip_colours <- colour.palette(Ntip(tree_short))
+names(tip_colours) <- names(sort(tip_order))
+tip_colours_suborders <- tip_colours
+save(tip_colours_suborders, file = "../Data/Processed/tip_colours_suborders.rda")
+
+## Plotting the ellipses for each suborder
+for(i in 1:length(tip_order)) {
+
+    ## TODO: FUNCTIONALISE THE ELLIPSE PLOTTING: START
+
+    ## Select the colour and name
+    name.col <- tip_colours[i]
+
+    ## Plot the ellipse
+    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    plot.one.ellipse(suborder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
+
+    ## Calculate which dimension contains the most variance
+    add.dims(suborder_shapespace, name.col)
+    n <- 1000
+    covars_cent_tend <- lapply(sample.n(get.subsets(suborder_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
+    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+    text_main <- paste0(names(name.col))
+    text(0,0.3, text_main, cex = 1.2, col = name.col)
+
+    ## Plot the elaboration/exploration
+    par(mar = c(2, 1, 1, 1)+0.1)
+    plot.one.proj.rej(results$suborder_results_between, name.col, x = TRUE, y = FALSE)
+
+        ## TODO: FUNCTIONALISE THE ELLIPSE PLOTTING: END
+}
+
+## PLOTTING THE FAMILIES
+tree <- ladderize(tree)
+## Select the level
+selected_level <- coloured_levels <- families_lvl
+
+## Replace the order levels as colours
+colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+
+## Shorten the tree
+selected_tips <- character()
+for(one_level in 2:length(levels(selected_level))) {
+    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
+    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
+}
+
+## Shorten the tree
+tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
+tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
+
+## Setting the tips edges colours
+tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
+## Rainbow the tip edges
+tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
+
+## Adding the number of tips
+tree_plot <- tree_short
+n_species <- table(selected_level)
+for(i in 1:Ntip(tree_plot)) {
+    ## Get the number of species
+    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
+    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
+}
+
+## Plotting the tree
+par(mar = c(0, 0, 0, 0)+0.1)
+plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
+
+## Matching the order and the colours to the ladderised tree
+tip_order <- c(
+    "Pittidae" = 2,
+    "Eurylaimidae" = 1,
+    "Tyrannidae" = 5,
+    "Pipridae" = 3,
+    "Cotingidae" = 4,
+    "Thamnophilidae" = 6,
+    "Furnaridae" = 9,
+    "Rhinocryptidae" = 8,
+    "Grallariidae" = 7,
+    "Ptilonorhynchidae" = 10,
+    "Maluridae" = 11,
+    "Meliphagidae" = 13,
+    "Acanthizidae" = 12,
+    "Vireonidae" = 20,
+    "Pachycephalidae" = 18,
+    "Oriolidae" = 19,
+    "Laniidae" = 25,
+    "Corvidae" = 24,
+    "Paradisaeidae" = 23,
+    "Monarchidae" = 22,
+    "Dicruridae" = 21,
+    "Campephagidae" = 14,
+    "Malaconotidae" = 16,
+    "Vangidae" = 15,
+    "Platysteiridae" = 17,
+    "Troglodytidae" = 27,
+    "Sittidae" = 26)
+
+tip_colours <- colour.palette(Ntip(tree_short))
+names(tip_colours) <- names(sort(tip_order))
+tip_colours_families <- tip_colours
+save(tip_colours_families, file = "../Data/Processed/tip_colours_families.rda")
+
+for(i in 1:length(tip_order)) {
+    ## Select the colour and name
+    name.col <- tip_colours[i]
+
+    ## Plot the ellipse
+    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    plot.one.ellipse(family_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
+
+    ## Calculate which dimension contains the most variance
+    add.dims(family_shapespace, name.col)
+    n <- 1000
+    covars_cent_tend <- lapply(sample.n(get.subsets(family_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
+    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+    text_main <- paste0(names(name.col))
+    text(0,0.3, text_main, cex = 1.2, col = name.col)
+
+    ## Plot the elaboration/exploration
+    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
+    plot.one.proj.rej(results$family_results_between, name.col, x = TRUE, y = FALSE)
+}
+dev.off()
+
+```

### 2022-05-25 Functions/ploting.utilities.R
Started isolating the ploting functions

@@ -1,0 +1,152 @@
+#' @title gg.color.hue
+#'
+#' @description Create the ggplot hue palette.
+#' 
+#' @param n the number of colours
+#' 
+## The ggplot color panel
+gg.color.hue <- function(n) {
+    grDevices::hcl(h = seq(15, 375, length = n + 1), l = 65, c = 100)[1:n]
+}
+
+
+#' @title plot.one.ellipse
+#'
+#' @description Plots a single ellipse (typically for the ellipse figure)
+#' 
+#' @param data the dispRity object with the covar components
+#' @param name.col the colour name
+#' @param x logical, whether to plot the x axis
+#' @param y logical, whether to plot the y axis
+#' @param main the plot title
+#' @param legend logical, whether to add the legend
+#' @param with.phylo logical, whether to plot the phylo ellipse
+#' 
+## Plotting the ellipses
+plot.one.ellipse <- function(data, name.col, x = TRUE, y = TRUE, main = "", legend = FALSE, with.phylo = FALSE) {
+
+    ## Get the PC %ages
+    if(x || y) {
+        vars <- dispRity::variances(data$matrix[[1]])
+        pc_var <- round((vars/sum(vars)*100), 2)
+    }
+
+    ## Do the empty plot
+    lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+    plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = ifelse(x, paste0("PC1 (", pc_var[1], "%)"), ""), ylab = ifelse(y, paste0("PC2 (", pc_var[2], "%)"), ""), main = main)
+    if(x) axis(1)
+    if(y) axis(2)
+    abline(v = 0, col = "grey", lwd = 0.5)
+    abline(h = 0, col = "grey", lwd = 0.5)
+
+    if(!with.phylo) {
+        subsets <- names(name.col)
+        cols <- name.col
+    } else {
+        subsets <- c("phylogeny", names(name.col))
+        cols <- c("grey", name.col)
+    }
+    if(!with.phylo) {
+        covar.plot(get.subsets(data, subsets = subsets),
+               col = cols,
+               ellipses = mean,
+               legend = FALSE,
+               points = FALSE,
+               cex = 0.5, apply.to.VCV = TRUE,
+               lwd = 4,
+               add = TRUE,
+               legend = legend)
+    } else {
[…]

### 2022-05-25 Functions/covar.utilities.R
Create covar.utilities.R

@@ -1,0 +1,81 @@
+## Get the VCV central tendency
+VCV.cent.tend <- function(one_covar, fun) {
+    VCVs <- lapply(one_covar, `[[`, "VCV")
+    locs <- lapply(one_covar, `[[`, "loc")
+    return(list(VCV = apply(array(do.call(cbind, VCVs), dim = c(dim(VCVs[[1]]), length(VCVs))), c(1,2), fun),
+                loc = apply(do.call(rbind, locs), 2, fun)))
+}
+
+## Selecting n elements from a covar object
+sample.n <- function(covar, n, selected_n, dimensions = NULL) {
+    ## Get the posterior size
+    n_post <- length(covar[[1]])
+    if(!missing(n) && n == n_post) {
+        if(is.null(dimensions)) {
+            ## Short cut
+            return(covar)    
+        } else {
+            return(lapply(covar, lapply, redimension, dimensions = dimensions))
+        }
+    } else {
+        ## Sample n
+        if(missing(selected_n)) {
+            selected_n <- sample.int(n_post, n, replace = n > n_post)
+        }
+        ## Return n for each group
+        if(is.null(dimensions)) {
+            return(lapply(covar, function(group, n) group[n], n = selected_n))
+        } else {
+            return(lapply(lapply(covar, function(group, n) group[n], n = selected_n), lapply, redimension, dimensions = dimensions))
+        }
+    }
+}
+
+## Internal: get the coordinates of one axes
+get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {
+
+    # The magic: https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331
+
+    ## VCVing the matrix
+    if(!is(data, "list")) {
+        data <- list(VCV = data)
+    } else {
+        if(is.null(data$VCV)) {
+            data$VCV <- data
+        }
+    }
+    ## adding a loc
+    if(is.null(data$loc)) {
+        data$loc <- rep(0, nrow(data$VCV))
+    }
+
+    ## Select the right dimensions
+    data$VCV <- data$VCV[dimensions, dimensions, drop = FALSE]
+
+    ## Get the data dimensionality
+    dims <- length(diag(data$VCV))
+
[…]

### 2022-05-25 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
Cleaning script for plotting

@@
 library(dispRity)
+source("../Functions/ploting.utilities.R")
+source("../Functions/covar.utilities.R")
 ```
@@
 rm(shapespace_allbirds_lvl_superorder_order)
-```
 
-First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
-## The ggplot color panel
-gg.color.hue <- function(n) {
-    grDevices::hcl(h = seq(15, 375, length = n + 1), l = 65, c = 100)[1:n]
-}
-
-# colour.palette <- rainbow
-colour.palette <- gg.color.hue
-```
-
-```{r, echo = FALSE}
-## Quick load
+## Loading the results
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_4k.rda")
@@
 
-## Phylogeny window (detailed)
-
-```{r}
-source("../Functions/covar.utilities.R")
-## Plotting the ellipses
-plot.one.ellipse <- function(data, name.col, x = TRUE, y = TRUE, main = "", legend = FALSE, with.phylo = FALSE) {
-
-    ## Get the PC %ages
-    if(x || y) {
-        vars <- dispRity::variances(data$matrix[[1]])
-        pc_var <- round((vars/sum(vars)*100), 2)
-    }
-
-    ## Do the empty plot
-    lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
-    plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = ifelse(x, paste0("PC1 (", pc_var[1], "%)"), ""), ylab = ifelse(y, paste0("PC2 (", pc_var[2], "%)"), ""), main = main)
-    if(x) axis(1)
-    if(y) axis(2)
-    abline(v = 0, col = "grey", lwd = 0.5)
-    abline(h = 0, col = "grey", lwd = 0.5)
-
-    if(!with.phylo) {
-        subsets <- names(name.col)
-        cols <- name.col
-    } else {
-        subsets <- c("phylogeny", names(name.col))
-        cols <- c("grey", name.col)
-    }
-    if(!with.phylo) {
-        covar.plot(get.subsets(data, subsets = subsets),
-               col = cols,
-               ellipses = mean,
-               legend = FALSE,
-               points = FALSE,
-               cex = 0.5, apply.to.VCV = TRUE,
-               lwd = 4,
-               add = TRUE,
-               legend = legend)
-    } else {
-        covar.plot(get.subsets(data, subsets = subsets),
-               col = cols,
-               ellipses = mean,
-               legend = FALSE,
-               points = FALSE,
-               cex = 0.5, apply.to.VCV = TRUE,
-               lwd = 4,
-               add = TRUE,
-               legend = legend,
-               scale = "phylogeny")        
-    }
-}
-## Plot the projection/rejection
-plot.one.proj.rej <- function(data, name.col, x = TRUE, y = TRUE, main = "") {
-
-    ## Get the proj/rej
-    elaboration <- get.disparity(data$position, concatenate = FALSE)
-    exploration <- get.disparity(data$distance, concatenate = FALSE)
-
-    ## Get the correct group
-    if(length(ID <- which(names(elaboration) %in% names(name.col))) < 1) {
-        ID <- which(names(elaboration) %in% paste0(names(name.col), ":phylogeny"))
-    }
-    elaboration <- c(elaboration[[ID]])
-    exploration <- c(exploration[[ID]])
-
-    ## Change the colours
-    col <- c(name.col, adjustcolor(name.col, alpha.f = 0.5))
-
-    ## Set the limits
-    quant_95 <- quantile(c(elaboration, exploration), prob = 0.975)
-    lim_up <- c(ceiling(quant_95/5)*5)
-
-    ## Get the quantiles
-    elaboration_quantiles <- quantile(elaboration, prob = c(0.025, 0.25, 0.50, 0.975))
-    exploration_quantiles <- quantile(exploration, prob = c(0.025, 0.25, 0.50, 0.975))
-
-    ## Empty plot
-    par(bty = "o")
-    plot(NULL, xlim = c(-0.5,1.5), ylim = c(0,lim_up), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
-
-    ## Add the lines
-    lines(lty = 2, x = c(0,0), y = elaboration_quantiles[c(1,4)], col = col[1], lwd = 1)
-    lines(lty = 1, x = c(0,0), y = elaboration_quantiles[c(2,3)], col = col[1], lwd = 3)
-    lines(lty = 2, x = c(1,1), y = exploration_quantiles[c(1,4)], col = col[2], lwd = 1)
-    lines(lty = 1, x = c(1,1), y = exploration_quantiles[c(2,3)], col = col[2], lwd = 3)
-    points(0, median(elaboration), pch = 19, col = col[1])
-    points(1, median(exploration), pch = 19, col = col[2])
-
-    ## Add the y axis
-    axis(2, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5), las = 2)
-
-    ## Plot the results
-    # boxplot(cbind(elaboration, exploration), horizontal = TRUE, ylim = c(0,lim_up), col = col, las = 1, xaxt = "n", yaxt = "n")
-    # if(x) {
-    #     axis(1, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5))
-    # }
-    # if(y) {
-    #     axis(2, at = c(1, 2), labels = c("exploration", "innovation"), las = 2)
-    # }
-}
-## Plot the variation per dimension
-add.dims <- function(data, name.col, n = 1000) {
-    covars_cent_tend <- lapply(sample.n(get.subsets(data, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
-    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    dim_var <- dim_var/max(dim_var)/10
-
-    ## Select the quadrant
-    if(colMeans(data$matrix[[1]][c(data$subsets[[names(name.col)]]$elements), c(1,2)])[1] > 0.1) {
-        x_start <- -0.2
-    } else {
-        x_start <- 0.2
-    }
-
-    y_start <- -0.1
-
-    buffer <- 0
-    for(one_dim in 1:length(dim_var)) {
-        lines(x = c(x_start, x_start+dim_var[one_dim]), y = rep(y_start+buffer, 2), lwd = 3, col = name.col)
-        buffer <- buffer - 0.02
-    }
-}
-
-# pdf(file = "Figure1a.pdf", height = 11, width = 5.5)
-# ## Plot the phylo ellipse
-# grey_palette <- c(colour.palette(8), "grey")
-# par(mfrow = c(2,1), bty = "n")
-# plot.one.ellipse(superorder_shapespace, ID = c(1,9), colours = grey_palette, x = TRUE, y = TRUE, main = "A")
-# legend("topright", lwd = 5, col = grey_palette[c(1,9)], lty = 1, legend = c("Aequornithes' median variance-covariance", "Overall phylogenetic's median variance-covariance"), bg = "white")
-# plot.one.proj.rej(results$super_results_between, ID = 1, colours = grey_palette)
-```
-
 ```{r}
 box_size <- 1.5
-# pdf(file = "../Manuscript/Figures/Figure1_phylo_Dplot.pdf", height = box_size*2, width = box_size*2)
-
-# lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
-# plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = "")
-# abline(v = 0, col = "grey", lwd = 0.5)
-# abline(h = 0, col = "grey", lwd = 0.5)
-
-# covars_cent_tend <- lapply(sample.n(get.subsets(data, subset = "phylogeny")$covar, n), VCV.cent.tend, mean)
-# dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-# dim_var <- dim_var/max(dim_var)/10
-# x_start <- 0.2
-# y_start <- -0.1
-# buffer <- 0
-# for(one_dim in 1:length(dim_var)) {
-#     lines(x = c(x_start, x_start+dim_var[one_dim]), y = rep(y_start+buffer, 2), lwd = 3, col = "grey")
-#     buffer <- buffer - 0.02
-# }
-
-# dev.off()
-
 pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)
 
-## Global window
+## Global window parameters
 lay_mat_order <- rbind(
@@
     )
-
 heights_vals <- rep(1, 9)
@@
 
-# layout.show(layout_matrix)
-
-# layout_matrix <- layout(cbind(matrix(1, ncol = 1, nrow = 27),
-#                               matrix(2:((27*2)+1), ncol = 2, nrow = 27, byrow = TRUE)))
-# layout.show(layout_matrix)
-
-## PLOTTING THE SUPER ORDERS
-tree <- ladderize(tree)
-## Get the tip_values
-orders_lvl <- shapespace$level2
-superorders_lvl <- shapespace$level1
-
-## Select the level
-selected_level <- coloured_levels <- superorders_lvl
-
-## Replace the order levels as colours
-colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-# par(mar = c(1, 1, 1, 1)+0.1)
-# plot(tree, show.tip.label = FALSE, edge.color =  match.tip.edge(as.character(coloured_levels), tree, replace.na = "grey"))
-
-## Shorten the tree
-selected_tips <- character()
-for(one_level in 2:length(levels(selected_level))) {
-    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
-    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
-}
-
-## Shorten the tree
-tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
-tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
-
-## Setting the tips edges colours
-tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
-## Rainbow the tip edges
-tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
-## Adding the number of tips
-tree_plot <- tree_short
-n_species <- table(selected_level)
-for(i in 1:Ntip(tree_plot)) {
-    ## Get the number of species
-    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
-    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
-}
-
-## Plotting the tree
-par(mar = c(0, 0, 0, 0)+0.1)
-plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
-# tiplabels()
-
-## Matching the order and the colours to the ladderised tree
+## PLOTTING THE SUPERORDERS
+## Getting the tip order for the ladderised super order tree
 tip_order <- c(
@@
 
-tip_colours <- colour.palette(Ntip(tree_short))
-names(tip_colours) <- names(sort(tip_order))
-tip_colours_super_orders <- tip_colours
+## Plotting the tree and getting the tip orders
+tip_colours_super_orders <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
 save(tip_colours_super_orders, file = "../Data/Processed/tip_colours_super_orders.rda")
 
-# par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
-# plot(rep(1,27), 1:27, col = "white")
-# labs <- names(tip_colours)
-# cols <- tip_colours
-# text(x = rep(1,27), y = 1:27, label = rev(labs), col = rev(cols))
-
+## Plotting the ellipses
 for(i in 1:length(tip_order)) {
     ## Select the colour and name
-    # name.col <- tip_colours[match(sort(names(tip_colours)), names(tip_colours))][i]
     name.col <- tip_colours[i]
@@
     dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    # text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
-    text_main <- paste0(names(name.col))#, "\n", "λe = ", round(lambdas_superorder[paste0("superorders.", names(name.col)),][1], 2), "; λi = ", round(lambdas_superorder[paste0("superorders.", names(name.col)),][2], 2))
+    text_main <- paste0(names(name.col))
     text(0,0.3, text_main, cex = 1.2, col = name.col)
@@
 ## PLOTTING THE ORDERS
-tree <- ladderize(tree)
-## Get the tip_values
-orders_lvl <- shapespace$level2
-superorders_lvl <- shapespace$level1
-
-## Select the level
-selected_level <- coloured_levels <- orders_lvl
-
-## Replace the order levels as colours
-colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-
-## Shorten the tree
-selected_tips <- character()
-for(one_level in 2:length(levels(selected_level))) {
-    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
-    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
-}
-
-## Shorten the tree
-tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
-tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
-
-## Setting the tips edges colours
-tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
-## Rainbow the tip edges
-tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
-
-## Adding the number of tips
-tree_plot <- tree_short
-n_species <- table(selected_level)
-for(i in 1:Ntip(tree_plot)) {
-    ## Get the number of species
-    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
-    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
-}
-
-## Plotting the tree
-par(mar = c(0, 0, 0, 0)+0.1)
-plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
-
 ## Matching the order and the colours to the ladderised tree
@@
 
-# tip_order <- sort(tip_order)
-
-tip_colours <- colour.palette(Ntip(tree_short))
-names(tip_colours) <- names(sort(tip_order))
-tip_colours_orders <- tip_colours
+## Plotting the tree and getting the tip orders
+tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
 save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
@@
     ## Select the colour and name
-    # name.col <- tip_colours[match(sort(names(tip_colours)), names(tip_colours))][i]
     name.col <- tip_colours[i]
@@
     dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    # text(0,0.3, paste0(names(name.col), " (", which(dim_var == max(dim_var)), ")"), cex = 1, col = name.col)
-    text_main <- paste0(names(name.col))#, "\n", "λe = ", round(lambdas_order[paste0("orders.", names(name.col)),][1], 2), "; λi = ", round(lambdas_order[paste0("orders.", names(name.col)),][2], 2))
+    text_main <- paste0(names(name.col))
     text(0,0.3, text_main, cex = 1.2, col = name.col)

### 2022-05-25 Functions/ploting.utilities.R
Update ploting.utilities.R

@@ -150,3 +150,61 @@ plot.one.proj.rej <- function(data, name.col, x = TRUE, y = TRUE, main = "") {
     ## Add the y axis
     axis(2, at = seq(from = 0, to = lim_up, by = 5), labels = seq(from = 0, to = lim_up, by = 5), las = 2)
 }
+
+#' @title plot.short.tree
+#'
+#' @description plots the shorten tree (with tip numbers) and outputs the group colour vector (typically for the ellipse figure)
+#' 
+#' @param tree the complete tree
+#' @param shapespace the shapespace data frame (with levels)
+#' @param level which level to plot the short tree for
+#' @param colour.palette the colour palette to draw from
+#' @param tip.order the order of the tips in the tree (can be left missing)
+#'
+plot.short.tree <- function(tree, shapespace, level, colour.palette = gg.color.hue, tip.order) {
+
+    ## Ladderizing the tree
+    tree <- ladderize(tree)
+
+    ## Getting the correct level
+    selected_level <- coloured_levels <- shapespace[, level]
+
+    ## Replace the order levels as colours
+    colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+    levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
+
+    ## Shorten the tree
+    selected_tips <- character()
+    for(one_level in 2:length(levels(selected_level))) {
+        selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
+        names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
+    }
+
+    ## Shorten the tree
+    tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
+    tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
+
+    ## Setting the tips edges colours
+    tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
+    ## Rainbow the tip edges
+    tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
+    ## Adding the number of tips
+    tree_plot <- tree_short
+    n_species <- table(selected_level)
+    for(i in 1:Ntip(tree_plot)) {
[…]

### 2022-05-26 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 
-```{r}
+```{r, eval = TRUE, echo = FALSE}
 box_size <- 1.5
 pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)
+
+## SETTING THE WINDOW PARAMETERS
 
@@
 ## PLOTTING THE SUPERORDERS
+
 ## Getting the tip order for the ladderised super order tree
@@
 for(i in 1:length(tip_order)) {
-    ## Select the colour and name
-    name.col <- tip_colours[i]
-
-    ## Plot the ellipse
-    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
-    plot.one.ellipse(superorder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-
-    ## Calculate which dimension contains the most variance
-    add.dims(superorder_shapespace, name.col)
-    n <- 1000
-    covars_cent_tend <- lapply(sample.n(get.subsets(superorder_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
-    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    text_main <- paste0(names(name.col))
-    text(0,0.3, text_main, cex = 1.2, col = name.col)
-
-    ## Plot the elaboration/exploration
-    par(mar = c(2, 1, 1, 1)+0.1)
-    plot.one.proj.rej(results$super_results_between, name.col, x = TRUE, y = FALSE)
+    ## Plot each pannel
+    wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between)
 }
@@
 ## PLOTTING THE ORDERS
+
 ## Matching the order and the colours to the ladderised tree
@@
 
+## Plotting the ellipses
 for(i in 1:length(tip_order)) {
-    ## Select the colour and name
-    name.col <- tip_colours[i]
-
-    ## Plot the ellipse
-    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
-    plot.one.ellipse(order_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-
-    ## Calculate which dimension contains the most variance
-    add.dims(order_shapespace, name.col)
-    n <- 1000
-    covars_cent_tend <- lapply(sample.n(get.subsets(order_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
-    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    text_main <- paste0(names(name.col))
-    text(0,0.3, text_main, cex = 1.2, col = name.col)
-
-    ## Plot the elaboration/exploration
-    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
-    plot.one.proj.rej(results$order_results_between, name.col, x = TRUE, y = FALSE)
+    ## Plot each pannel
+    wrap.plot.ellipses(i, tip_colours_orders, order_shapespace, results$order_results_between)
 }
@@
 
-# Passeriformes bit
+# Passeriformes shapespace analyses
 
@@
 rm(shapespace_passeriformes)
-```
 
-First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):
-
-```{r, echo = FALSE}
-## Quick load
+## Loading the results
 load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
@@
 ```{r}
-
-source("../Functions/covar.utilities.R")
-source("../Functions/ploting.utilities.R")
-
 box_size <- 1.5
@@
     )
-
 heights_vals <- rep(1, 8)
@@
 
-# layout.show(layout_matrix)
+## PLOTTING THE SUBORDERS
 
-## TODO: FUNCTIONALISE THE TREE PLOTTING: START
-
-## PLOTTING THE SUBORDERS
-tree <- ladderize(tree)
-## Get the tip_values
-families_lvl <- shapespace$level2
-suborders_lvl <- shapespace$level1
-
-## Select the level
-selected_level <- coloured_levels <- suborders_lvl
-
-## Replace the order levels as colours
-colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-
-## Shorten the tree
-selected_tips <- character()
-for(one_level in 2:length(levels(selected_level))) {
-    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
-    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
-}
-
-## Shorten the tree
-tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
-tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
-
-## Setting the tips edges colours
-tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
-## Rainbow the tip edges
-tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
-## Adding the number of tips
-tree_plot <- tree_short
-n_species <- table(selected_level)
-for(i in 1:Ntip(tree_plot)) {
-    ## Get the number of species
-    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
-    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
-}
-
-## Plotting the tree
-par(mar = c(0, 0, 0, 0)+0.1)
-plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
-
-## TODO: FUNCTIONALISE THE TREE PLOTTING: END
-
-## Matching the order and the colours to the ladderised tree
+## Getting the tip order for the ladderised super order tree
 tip_order <- c(
@@
 
-tip_colours <- colour.palette(Ntip(tree_short))
-names(tip_colours) <- names(sort(tip_order))
-tip_colours_suborders <- tip_colours
+## Plotting the tree and getting the tip orders
+tip_colours_suborders <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
 save(tip_colours_suborders, file = "../Data/Processed/tip_colours_suborders.rda")
 
-## Plotting the ellipses for each suborder
+## Plotting the ellipses
 for(i in 1:length(tip_order)) {
-
-    ## TODO: FUNCTIONALISE THE ELLIPSE PLOTTING: START
-
-    ## Select the colour and name
-    name.col <- tip_colours[i]
-
-    ## Plot the ellipse
-    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
-    plot.one.ellipse(suborder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-
-    ## Calculate which dimension contains the most variance
-    add.dims(suborder_shapespace, name.col)
-    n <- 1000
-    covars_cent_tend <- lapply(sample.n(get.subsets(suborder_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
-    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    text_main <- paste0(names(name.col))
-    text(0,0.3, text_main, cex = 1.2, col = name.col)
-
-    ## Plot the elaboration/exploration
-    par(mar = c(2, 1, 1, 1)+0.1)
-    plot.one.proj.rej(results$suborder_results_between, name.col, x = TRUE, y = FALSE)
-
-        ## TODO: FUNCTIONALISE THE ELLIPSE PLOTTING: END
+    ## Plot each pannel
+    wrap.plot.ellipses(i, tip_colours_suborders, suborder_shapespace, results$suborder_results_between)
 }
@@
 ## PLOTTING THE FAMILIES
-tree <- ladderize(tree)
-## Select the level
-selected_level <- coloured_levels <- families_lvl
-
-## Replace the order levels as colours
-colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
-
-## Shorten the tree
-selected_tips <- character()
-for(one_level in 2:length(levels(selected_level))) {
-    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
-    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
-}
-
-## Shorten the tree
-tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
-tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]
-
-## Setting the tips edges colours
-tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
-## Rainbow the tip edges
-tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
-
-## Adding the number of tips
-tree_plot <- tree_short
-n_species <- table(selected_level)
-for(i in 1:Ntip(tree_plot)) {
-    ## Get the number of species
-    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
-    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
-}
-
-## Plotting the tree
-par(mar = c(0, 0, 0, 0)+0.1)
-plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)
 
@@
 
-tip_colours <- colour.palette(Ntip(tree_short))
-names(tip_colours) <- names(sort(tip_order))
-tip_colours_families <- tip_colours
+## Plotting the tree and getting the tip orders
+tip_colours_families <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
 save(tip_colours_families, file = "../Data/Processed/tip_colours_families.rda")
 
+## Plotting the ellipses
 for(i in 1:length(tip_order)) {
-    ## Select the colour and name
-    name.col <- tip_colours[i]
-
-    ## Plot the ellipse
-    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
-    plot.one.ellipse(family_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
-
-    ## Calculate which dimension contains the most variance
-    add.dims(family_shapespace, name.col)
-    n <- 1000
-    covars_cent_tend <- lapply(sample.n(get.subsets(family_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
-    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
-    text_main <- paste0(names(name.col))
-    text(0,0.3, text_main, cex = 1.2, col = name.col)
-
-    ## Plot the elaboration/exploration
-    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
-    plot.one.proj.rej(results$family_results_between, name.col, x = TRUE, y = FALSE)
+    ## Plot each pannel
+    wrap.plot.ellipses(i, tip_colours_families, family_shapespace, results$family_results_between)
 }

### 2022-05-26 Functions/ploting.utilities.R
added wrapping plot

@@ -208,3 +208,34 @@ plot.short.tree <- function(tree, shapespace, level, colour.palette = gg.color.h
         return(tip_colours)
     }
 }
+
+#' @name wrap.plot.ellipses
+#'    
+#' @description plots the ellipses, dimensionality and projection/rejection thermos
+#'
+#' @param i the index of the element to plot (usually along tip_order)
+#' @param tip.colours the named vector with colours and group names
+#' @param shapespace the shapespace as a dispRity object
+#' @param results the list of dispRity objects containing the results (typically output from dispRity.covar.projections)
+#
+## Plotting the ellipses, the dimensions and the projection/rejection
+wrap.plot.ellipses <- function(i, tip.colours, shapespace, results) {
+    ## Select the colour and name
+    name.col <- tip.colours[i]
+
+    ## Plot the ellipse
+    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
+    plot.one.ellipse(shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)
+
+    ## Calculate which dimension contains the most variance
+    add.dims(shapespace, name.col)
+    n <- 1000
+    covars_cent_tend <- lapply(sample.n(get.subsets(shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
+    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
+    text_main <- paste0(names(name.col))
+    text(0,0.3, text_main, cex = 1.2, col = name.col)
+
+    ## Plot the elaboration/exploration
+    par(mar = c(2, 1, 1, 1)+0.1)
+    plot.one.proj.rej(results, name.col, x = TRUE, y = FALSE)
+}

### 2022-05-26 Analysis/06-Figure3_with_correlations.Rmd
Update 06-Figure3_with_correlations.Rmd

--- 2022-05-16 Analysis/06-Figure3_with_correlations.Rmd
@@
 library(dispRity)
+source("../Functions/ploting.utilities.R")
+source("../Functions/correlations.utilities.R")
 ```
@@
 rm(shapespace_allbirds_lvl_superorder_order)
-```
 
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
-## The ggplot color panel
-gg.color.hue <- function(n) {
-    grDevices::hcl(h = seq(15, 375, length = n + 1), l = 65, c = 100)[1:n]
-}
-
-# colour.palette <- rainbow
-colour.palette <- gg.color.hue
-```
-
-Loading the elaboration exploration results (from 03_results)
-
-```{r, echo = FALSE}
-## Quick load
+## Loading the projection/rejections results
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
@@
 
-## Figure 3
-
-Density plot of the correlation of the median elaboration and median exploration for all species.
-
-```{r, echo = FALSE}
-## Calculating all the correlations
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE) {
-
-    ## Remove the phylo one (if needed)
-    if(exclude.phylo) {
-        select_pos <- names(size.subsets(res$position))
-        select_dis <- names(size.subsets(res$distance))
-        select_pos <- select_pos[-which(select_pos %in% "phylogeny")]
-        select_dis <- select_dis[-which(select_dis %in% "phylogeny")]
-        res$position <- get.subsets(res$position, subsets = select_pos)
-        res$distance <- get.subsets(res$distance, subsets = select_dis)
-    }
-
-    ## Extracting all the results
-    disparity_pos <- lapply(res$position$disparity, function(X, type) return(X$elements))
-    disparity_dis <- lapply(res$distance$disparity, function(X, type) return(X$elements))
-
-    ## Measure the central tendencies
-    centrals_pos <- lapply(disparity_pos, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-    centrals_dis <- lapply(disparity_dis, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-
-    ## Number of groups
-    n_groups <- length(centrals_pos)
-
-    ## Prepare the results
-    prep.cor.results <- function(pos, dis, name) {
-        ## Run the correlation test
-        cor_test <- cor.test(pos, dis)
-        ## Prepare the plot title
-        plot_main <- paste0(name, ": ", unname(round(cor_test$estimate, 3)), " (p=", unname(round(cor_test$p.value, 3)), ")")
-        ## Return the plotting parameters
-        return(list(x = pos, y = dis, main = plot_main))
-    }
-
-    return(mapply(prep.cor.results, centrals_pos, centrals_dis, as.list(names(centrals_pos)), SIMPLIFY = FALSE))
-}
-
-## Plotting one correlation 
-plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
-    dots <- list(...)
-
-    if(missing(col.sub)) {
-        col.sub <- NULL
-    }
-
-    ## Subbing the colours
-    if(!is.null(col.sub)) {
-        ## Reset the levels
-        col_sub <- as.factor(as.character(col.sub))
-        ## Make the empty level grey
-        if(any(levels(col_sub) == "")) {
-            col_avail <- c("grey", col.fun(length(levels(col_sub))-1))
-        } else {
-            col_avail <- col.fun(length(levels(col_sub)))
-        }
-        ## Create the colour vector
-        col_vector <- col_avail[as.numeric(col_sub)]
-
-    } else {
-        if(is.null(dots$col)) {
-            col_vector <- "black"
-        } else {
-            col_vector <- dots$col
-        }
-    }
-
-    if(!missing(tree)) {
-        op <- par(mfrow = c(1,2))
-        #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1
-        par(mar = c(5, 4, 4, 0)+0.1)   
-    }
-
-    ## Plotting the results
-    plot(cor.results$x, cor.results$y, pch = 19, xlab = "elaboration", ylab = "innovation", main = cor.results$main, cex = pt.cex, col = col_vector)
-
-    if(!is.null(col.sub)) {
-        ## Adding the legend
-        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
-    }
-
-    if(!missing(tree)) {
-        par(mar = c(5, 0, 4, 2) + 0.1)
-
-        ## Recycle the colour vector
-        named_col_vector <- col_vector
-        names(named_col_vector) <- names(cor.results$x)
-
-        ## Reduce the tree
-        if(!full.tree) {
-            tree <- drop.tip(tree, tip = tree$tip.label[which(!(tree$tip.label %in% names(named_col_vector)))])
-        }
-
-        ## Matching edges and colours
-        match.edge.colours <- function(tree, named_col_vector, background = "grey") {
-            ## Get the background colour
-            edge_cols <- rep(background, Nedge(tree))
-            ## Get the clades
-            clades <- unique(named_col_vector)
-            ## Ignore the greys
-            if(any(which_grey <- clades == "grey")) {
-                clades <- clades[!which_grey]
-            }
-            ## Find the edges for each clade
-            for(clade in 1:length(clades)) {
-                ## Get the tips for the clade
-                tips <- which(named_col_vector == clades[clade])
-                ## Get the MRCA node
-                mrca_node <- getMRCA(tree, tips)
-                # stop_edge <- which(tree$edge[, 2] %in% mrca_node)
-
-                ## Handle non-monophyletic groups
-                clade_tree <- extract.clade(tree, node = mrca_node)
-                if(Ntip(clade_tree) > length(tips)) {
-                    clade_tree <- drop.tip(clade_tree, tip = clade_tree$tip.label[!(clade_tree$tip.label %in% names(tips))])
-                }
-                n_edges_target <- Nedge(clade_tree)
-
-                ## Get the tips ID
-                tip_ids <- which(tree$tip.label %in% names(tips))
-                ## Get the vector of edges
-                edges <- integer()
-                ## Get the first nodes (and edges)
-                new_edges <- which(tree$edge[, 2] %in% tip_ids)
-                edges <- unique(c(edges, new_edges))
-                nodes <- tree$edge[new_edges, 1]
-
-                # plot(tree, show.tip.label = FALSE)
-                # test.col <- c(replicate(10, c("red", "blue", "orange", "green", "pink", "yellow")))
-                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                ## Recursively loop through each node up until the last edge
-                # while(!(stop_edge %in% edges)) {
-                while(length(edges) <= n_edges_target) {
-                    ## Get the node edges
-                    new_edges <- which(tree$edge[, 2] %in% nodes)
-
-                    # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                    edges <- unique(c(edges, new_edges))
-                    ## Update the nodes
-                    nodes <- tree$edge[new_edges, 1]
-                }
-
-                # ## Do the last recursion
-                # new_edges <- which(tree$edge[, 2] %in% nodes)
-                # edges <- unique(c(edges, new_edges))
-
-                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                ## Colour these edges
-                edge_cols[edges] <- clades[clade]
-            }
-            return(edge_cols)
-        }
-        ## Match the colour to the correct edge
-        edge_cols <- match.edge.colours(tree, named_col_vector)
-
-        ## Add the tree
-        plot(tree, show.tip.label = FALSE, edge.color = edge_cols)
-    }
-
-    return(invisible())
-}
-```
-
-Correlations per super order (species on phylogeny):
+Extract the median correlations from the results (for the groups ion the phylogeny and the groups on themselves)
 
@@
 
-### Figure 3a - gg_version
+### Figure b): the overall density plot
 
@@
 ```{r, echo = FALSE, eval = FALSE}
-library(ggplot2)
-## Getting the data to plot
-global_data <- super_phylo_cor$phylogeny
+pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4.5, width = 4)
 
-gg.density.plot <- function(data, palette = "viridis") {
-    ggplot(data.frame("elaboration" = data$x, "exploration" = data$y), aes(x=elaboration, y=exploration) ) +
-      geom_hex() +
-      scale_fill_continuous(type = palette) +
-      theme_bw()
-}
+## Plotting the densities with the quadrant
+par(mar = c(5, 4, 4, 2) + 0.1)
+plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)
 
-pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4, width = 4)
-gg.density.plot(global_data)
 dev.off()
@@
 
-Can't seem to integrate the density plot to the base R layout so I'll just do it in Inkscape for now
+### Figure a): explanations
+
+This might have to be removed.
 
@@
 
-## Figure 3b
+## Figure c) and d): the correlation plots for each group
 
 ```{r, echo = FALSE, eval = FALSE}
-## Loading the distances from centre for each species
+## Loading the distances from centre for each species (if adding outlines)
 load("../Data/Processed/ploting_data.rda")
@@
 
+## Setting the colour pallette
+colour.palette <- gg.color.hue
+
 pdf("../Manuscript/Figures/Figure3_cd.pdf", height = 12, width = 8)
-## Great the plot layout
+
+## Make the plot layout
 layout_matrix <- cbind(matrix(0,2,2), matrix(0,2,2))
@@
 # layout.show(plot_layout)
-one.plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, ID, legend_counter, contour = NULL) {
 
-    if(missing(col.sub)) {
-        col.sub <- NULL
-    }
-
-    ## Subbing the colours
-    if(!is.null(col.sub)) {
-        ## Reset the levels
-        col_sub <- as.factor(as.character(col.sub))
-        ## Make the empty level grey
-        if(any(levels(col_sub) == "")) {
-            col_avail <- c("grey", col.fun(length(levels(col_sub))-1))
-        } else {
-            col_avail <- col.fun(length(levels(col_sub)))
-        }
-        ## Create the colour vector
-        col_vector <- col_avail[as.numeric(col_sub)]
-    } else {
-        if(is.null(dots$col)) {
-            col_vector <- "black"
-        } else {
-            col_vector <- dots$col
-        }
-    }
-
-    ## Sort the data by group size
-    all_data <- cbind(x = cor.results$x, y = cor.results$y)
-    ## Classify the groups per numbers of species
-    all_data <- data.frame(all_data, col = col_vector)
-
-    ## Sort the data per group size
-    group_order <- sort(table(all_data$col), decreasing = TRUE)
-
-    ## Set the plotting data
-    plot_data <- data.frame()
-    # col_adjust <- seq(from = 1/length(group_order), to = 1, by = 0.1)*3
-    for(i in 1:length(group_order)) {
-        ## Get the data from the biggest group
-        selected_data <- all_data[which(all_data$col == names(group_order[i])), ]
-        ## Transparantise the colours (for the background)
-        # selected_data$col <- adjustcolor(selected_data$col, alpha = col_adjust[i])
-        ## Combine the plotting data
-        plot_data <- rbind(plot_data, selected_data)
-    }
-
-    ## Empty plot
-    plot(NULL, xlim = range(plot_data$x), ylim = range(plot_data$y), xaxt = "n", yaxt = "n", xlab = "", ylab = "", main = strsplit(cor.results$main, split = ":")[[1]][1])
-    axis(1, at = pretty(plot_data$x, n = 4))
-    axis(2, at = pretty(plot_data$y, n = 4), las = 2)
-
-    ## Adding contours
-    if(!is.null(contour)) {
-        ## Select the right contour data
-        contour_data <- contour[match(rownames(plot_data), names(contour))]        
-        ## Calculate the contour
-        contour_z <- MASS::kde2d(contour_data/max(contour_data)*max(plot_data$x), contour_data/max(contour_data)*max(plot_data$y), n = 50)
-
-        contour_z <- matrix(0, nrow = length(contour_data), ncol = length(contour_data))
-        diag(contour_z) <-  contour_data
-        # contour_z <- outer(contour_data/max(contour_data)*max(plot_data$x), contour_data/max(contour_data)*max(plot_data$y))
-
-        ## Adding the contour in the background
-        contour(contour_z, lwd = 1, add = TRUE, col = hcl.colors(12, palette = "viridis"), labels = pretty(contour_data))
-    }
-
-    # ## col test for contour
-    # color.gradient <- function(x, colors=c("red","yellow","green"), colsteps=100) {
-    #     return( colorRampPalette(colors) (colsteps) [ findInterval(x, seq(min(x),max(x), length.out=colsteps)) ] )
-    # }
-
-    ## Plot the points
-    points(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = plot_data$col)
-    # points(plot_data$x, plot_data$y, pch = 19, cex = pt.cex, col = color.gradient(contour_data))
-
-    if(!is.null(col.sub)) {
-        ## See if there is any others
-        levels <- levels(col_sub)
-        if(length(other <- which(levels == "")) > 0) {
-            legend_text <- c(1, 1:(length(levels(col_sub))-1)+legend_counter)
-        } else {
-            legend_text <- c(1:(length(levels(col_sub)))+legend_counter)
-        }
-        ## Update the legend counter
-        legend_counter <<- max(legend_text)
-        ## Adding the legend
-        legend(x = legend.pos, legend = legend_text, pch = 19, cex = legend.cex, col = col_avail, bg = "white")
-    }
-
-    return(invisible())
-}
-
-# par(mfrow = c(2, 4))
 ## Getting the subsets for the colouring
@@
                  "topleft", "topright", "topleft", "topright")
+## Creat the vector for counting the legend numbers
 legend_counter <- 1
@@
     par(mar = c(2,2.5,2,1.5))
-    one.plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = centroids)
+    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = FALSE)
 }
@@
     par(mar = c(2,2.5,2,1.5))
-    one.plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter)
+    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
 }
@@
 
-### Correlations within super orders (species on group)
+# Correlations for passeriformes
 
-Regardless, here are the results when projecting the species on their own group's VCV
+```{r, eval = TRUE, echo = FALSE}
+## Load the data
+load(file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+## Select the data
+shapespace <- shapsepace_passeriformes_suborder_family[[1]]$space
+tree <- shapsepace_passeriformes_suborder_family[[1]]$consensus_tree
+## Clean garbage
+rm(shapsepace_passeriformes_suborder_family)
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
-## Getting the subsets for the colouring
-col_subs <- split(shapespace$level2, f = shapespace$level1)[-1]
+## Load the covar
+load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+shapespace_passeriformes <- shapespace_passeriformes[[1]]
 
-## Calculate the correlations
-super_within_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
-## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
-for(i in 1:length(super_within_cor)) {
-    plot.correlations(super_within_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
-}
+## Get the levels
+phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
+subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
+famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))
+
+## Creating the subsets
+suborder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+      data          = shapespace,
+      posteriors    = shapespace_passeriformes,
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
+
+family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+      data          = shapespace,
+      posteriors    = shapespace_passeriformes,
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+## Clean garbage
+rm(shapespace_passeriformes)
+
+## Loading the results
+load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
+results <- shapespace_passeriformes_suborder_family_results_list
+rm(shapespace_passeriformes_suborder_family_results_list)
 ```
 
-<!-- THIS BIT IS IGNORED FOR NOW -->
+```{r, eval = TRUE, echo = FALSE}
+## Calculate the correlations for the projections onto the phylo axis
+suborder_phylo_cor <- get.correlations(results$suborder_results_phylo)
 
-### Correlations per order (species on phylogeny)
+## Calculate the correlations for the projections onto the group' axes
+suborder_group_cor <- get.correlations(results$suborder_results_within, exclude.phylo = TRUE)
+```
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
-## Loading the sub-order levels
-load(file = "../Data/Processed/shapespace_superorder_lvl_order_suborder.rda")
-## Select the data
-shapespaces <- lapply(shapespace_superorder_lvl_order_suborder, function(X) X$space)
-## Create a level 3
-order_space <- do.call(rbind, shapespaces)
+```{r, echo = FALSE, eval = FALSE}
+pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4.5, width = 4)
 
-## Calculate the correlations
-order_phylo_cor <- get.correlations(results$order_results_phylo, exclude.phylo = TRUE)
+## Plotting the densities with the quadrant
+par(mar = c(5, 4, 4, 2) + 0.1)
+plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)
+
+dev.off()
+```
+
+## Figure 3b
+
+```{r, echo = FALSE, eval = FALSE}
+pdf("../Manuscript/Figures/Figure3_cd_passeriformes_1.pdf", height = 8, width = 8)
+par(mfrow = c(2,2))
 
 ## Getting the subsets for the colouring
-split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
-col_subs <- replicate(length(order_phylo_cor), NULL)
-col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
+col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
 
 ## Set the vector of legend positions
-for(i in 1:length(order_phylo_cor)) {
-    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]])
+legends_pos <- c("topleft", "topright", "topright", "topright",
+                 "topleft", "topright", "topleft", "topright")
+legend_counter <- 1
+
+for(i in 1:2) {
+    par(mar = c(2,2.5,2,1.5))
+    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = FALSE)
 }
-```
 
-### Correlations within orders (species on group)
+## Fig 3c
 
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
-## Calculate the correlations
-order_within_cor <- get.correlations(results$order_results_within, exclude.phylo = TRUE)
-
+# par(mfrow = c(2, 4))
 ## Getting the subsets for the colouring
-split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
-col_subs <- replicate(length(order_within_cor), NULL)
-col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
+col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
 
 ## Set the vector of legend positions
-for(i in 1:length(order_within_cor)) {
-    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]])
+legends_pos <- c("bottomright", "topright", "topleft", "topleft",
+                 "topleft", "topleft", "topright", "bottomright")
+legend_counter <- 1
+
+for(i in 1:2) {
+    par(mar = c(2,2.5,2,1.5))
+    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = TRUE, contour = NULL)
 }
+dev.off()
 ```

### 2022-05-26 Functions/covar.utilities.R
Update covar.utilities.R

@@ -79,3 +79,28 @@ get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {
     ## Get the edges coordinates
     return(edges[c(axis, axis+dims), , drop = FALSE])
 }
+
+## Scale a VCV matrix to another one
+scale.VCV <- function(VCV1, VCV2) {
+    ## Dividing both VCVs
+    
+    ## Getting the off diagonal (the scaling ratio)
+    # VCV <- VCV1$VCV/VCV2$VCV
+    # ratio <- abs(VCV[(nrow(VCV))^2-(1:nrow(VCV))*(nrow(VCV)-1)])
+
+    ## Getting the scaling ratio based on the major axis length
+    ratio <- dist(get.one.axis(VCV1))[1]/dist(get.one.axis(VCV2))[1]
+
+    ## Scaling VCV1
+    VCV1$VCV <- VCV1$VCV/ratio^2
+    return(VCV1)
+}
+
+## Internal: recentring the covar matrices (changing their loc)
+recentre <- function(one_group, one_centre, dimensions) {
+    recentre.loc <- function(covar, centre, dim) {
+        covar$loc[dim] <- centre[dim]
+        return(covar)
+    }
+    return(lapply(one_group, recentre.loc, centre = one_centre, dim = dimensions))
+}

### 2022-05-26 Analysis/08-Figure_key.Rmd
Create 08-Figure_key.Rmd

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_innovation_analyses.Rmd.

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
source("../Functions/ploting.utilities.R")
source("../Functions/covar.utilities.R")

```

# All birds shapespace analyses

```{r, eval = TRUE, echo = FALSE}
## Load the data
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Creating the subsets
superorder_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

```

# Extract only the Mirandornithes

```{r, eval = TRUE, echo = FALSE}
## Selecting the paleo data
space_paleo <- get.subsets(superorder_shapespace, subsets = c("phylogeny", "Paleognathae"))

## PANEL 1

## General plot
pdf("../Manuscript/Figures/figure_key_space_plot.pdf")
plot(shapespace[, c(1,2)], col = "lightgrey", xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5))
points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)
covar.plot(space_paleo,
      col = c("grey", gg.color.hue(1)),
      ellipses = mean,
      major.axes = mean,
      legend = FALSE,
      points = FALSE,
      cex = 0.5, apply.to.VCV = TRUE,
      lwd = 4,
      add = TRUE,
      scale = 2)
dev.off()

## Plotting the projection cartoon

## Do the projection
matrix <- as.matrix(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)])
rownames(matrix) <- NULL
## Get the base vectors
covars <- sample.n(space_paleo$covar, 1000)
centres <- lapply(unlist(space_paleo$subsets, recursive = FALSE) ,function(group, data, fun) fun(data[c(group), ]), data = space_paleo$matrix[[1]], fun = colMeans)
covars <- mapply(recentre, covars, centres, MoreArgs = list(dimensions = c(1,2)), SIMPLIFY = FALSE)
covars <- lapply(covars, function(one_covar, scale) mapply(scale.VCV, one_covar, scale, SIMPLIFY = FALSE), scale = covars[[2]])
covars_cent_tend <- lapply(covars, VCV.cent.tend, mean)
base_vectors <- lapply(lapply(covars_cent_tend, get.one.axis, axis = 1, level = 0.95, dimensions = c(1,2)), list)

## Get projection lines
get.proj.line <- function(point, base) {
  point <- t(point)
  base <- t(base)
  proj_x <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[1];
  proj_y <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[2];
  return(rbind(cbind(proj_x, proj_y), point))
}

## Projection onto the phylo line
pdf("../Manuscript/Figures/figure_key_projection_on_phylo.pdf")
plot(NULL, xlab = "PC1", ylab = "PC2", xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5))
# lines(base_vector)
for(i in 1:nrow(matrix)) {
      lines(get.proj.line(matrix[i, ], base_vectors$phylogeny[[1]]))
}
points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)
covar.plot(space_paleo,
      col = c("grey", gg.color.hue(1)),
      ellipses = mean,
      major.axes = mean,
      legend = FALSE,
      points = FALSE,
      cex = 0.5, apply.to.VCV = TRUE,
      lwd = 4,
      add = TRUE,
      scale = 2)
dev.off()

## Projection onto the group line
pdf("../Manuscript/Figures/figure_key_projection_on_group.pdf")
plot(NULL, xlab = "PC1", ylab = "PC2", xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5))
points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)

base <- base_vectors$Paleognathae[[1]]
base <- matrix(c(base)[c(1, 1, 4, 4)], 2, 2, byrow = FALSE) # or c(1, 4, 3, 2)
for(i in 1:nrow(matrix)) {
      lines(get.proj.line(matrix[i, ], test))
}

points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)
covar.plot(space_paleo,
      col = c("grey", gg.color.hue(1)),
      ellipses = mean,
      major.axes = mean,
      legend = FALSE,
      points = FALSE,
      cex = 0.5, apply.to.VCV = TRUE,
      lwd = 4,
      add = TRUE,
      scale = 2)
dev.off()

## Phylo plot
tree_paleo <- drop.tip(tree, tip = rownames(shapespace)[which(shapespace$level1 != "Paleognathae")])
plot(ladderize(tree_paleo), show.tip.label = FALSE)

## Loading the distances from centre for each species (if adding outlines)
load("../Data/Processed/ploting_data.rda")
centroids <- ploting_data$distances_to_centre[[1]]

```

### 2022-05-27 Analysis/08-Figure_key.Rmd
added tree

@@
 ## Phylo plot
+## Get the tree
 tree_paleo <- drop.tip(tree, tip = rownames(shapespace)[which(shapespace$level1 != "Paleognathae")])
-plot(ladderize(tree_paleo), show.tip.label = FALSE)
+## Ladderize
+tree_paleo <- ladderize(tree_paleo)
 
-## Loading the distances from centre for each species (if adding outlines)
+## Get the data (in the right order)
 load("../Data/Processed/ploting_data.rda")
-centroids <- ploting_data$distances_to_centre[[1]]
+elaborations <- ploting_data$median_elaborations[tree_paleo$tip.label]
+innovations <- ploting_data$median_innovations[tree_paleo$tip.label]
+
+## Transform them into a colour gradient
+col.grad <- colorRamp(c("blue","orange"))
+cols_elab <- col.grad(elaborations/max(elaborations))
+cols_innov <- col.grad(innovations/max(innovations))
+
+## Plot the tree
+pdf("../Manuscript/Figures/figure_key_tree_example.pdf")
+plot(tree_paleo, show.tip.label = FALSE, type = "fan")
+tiplabels("", frame = "none", pch = 19, cex = 2, col = rgb(cols_elab, maxColorValue = 256), offset = -1)
+tiplabels("", frame = "none", pch = 19, cex = 2, col = rgb(cols_innov, maxColorValue = 256), offset = 4)
+dev.off()
 

### 2022-05-27 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 
+## Getting the plotting data super_order_level
+ploting_data_superorder <- list(
+                     "median_elaborations" = apply(get.disparity(get.subsets(results$super_results_within$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "median_innovations"  = apply(get.disparity(get.subsets(results$super_results_within$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "distances_to_centre" = get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
+                     "consensus_tree"      = tree)
+
+## Getting the plotting data super_order_level
+ploting_data_order <- list(
+                     "median_elaborations" = apply(get.disparity(get.subsets(results$order_results_within$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "median_innovations"  = apply(get.disparity(get.subsets(results$order_results_within$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "distances_to_centre" = get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
+                     "consensus_tree"      = tree)
+
 ## Save all that as a list
-save(ploting_data, file = "ploting_data.rda")
+save(ploting_data, file = "../Data/Processed/ploting_data.rda")
+save(ploting_data_superorder, file = "../Data/Processed/ploting_data_superorder.rda")
+save(ploting_data_order, file = "../Data/Processed/ploting_data_order.rda")
 

### 2022-05-31 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

--- 2021-10-12 Analysis/01-Data_preparation.Rmd
@@
 
+## Passeriformes focus
+
+```{r}
+## Making a Passeriformes only list
+clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
+clades_list_passeriformes <- clades_list[which(clades_list$level2_order == "Passeriformes"), c(3, 6)]
+clades_list_passeriformes[,1] <- as.factor(as.character(clades_list_passeriformes[,1]))
+clades_list_passeriformes[,2] <- as.factor(as.character(clades_list_passeriformes[,2]))
+
+## Shapespace passeriformes
+shapespace_passeriformes <- shapespace[rownames(shapespace) %in% rownames(clades_list_passeriformes), ]
+
+## Do the prep data from here (with all the trees! ouch!)
+shapespace_passeriformes_suborder_family <- prep.data(
+                                          level = 0, lvl.inc = c(1,2),
+                                          clades    = clades_list_passeriformes,
+                                          space     = shapespace_passeriformes,
+                                          dim       = 1:8,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_passeriformes_suborder_family, file = "../Data/Processed/shapespace_passeriformes_suborder_family.rda")
+```
+
 ## References

### 2022-05-31 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 ## Loading the parametrisation data
-load(file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+load(file = "../Data/Processed/shapespace_passeriformes_suborder_family.rda")
 
 ## Select the data
-shapespace <- shapsepace_passeriformes_suborder_family[[1]]$space
-selected_dim <- shapsepace_passeriformes_suborder_family[[1]]$dimensions
-consensus_tree <- shapsepace_passeriformes_suborder_family[[1]]$consensus_tree
-trees_list <- shapsepace_passeriformes_suborder_family[[1]]$trees_list
+shapespace <- shapespace_passeriformes_suborder_family[[1]]$space
+selected_dim <- shapespace_passeriformes_suborder_family[[1]]$dimensions
+consensus_tree <- shapespace_passeriformes_suborder_family[[1]]$consensus_tree
+trees_list <- shapespace_passeriformes_suborder_family[[1]]$trees_list
 ```
@@
 ## Set up the parametrising chains models
-param_MCMCglmm <- make.mini.chains(data         = shapespace,
+param_chains <- make.mini.chains(data         = shapespace,
                                    dimensions   = selected_dim,
@@
 
-## Running the three MCMCglmm models
-save(param_chains, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_29levels_5k_chain1.rda")
+# Running the three MCMCglmm models
+save(param_chains, file = "../Data/Processed/params_shapespace_8D_50levels_50k_chain1.rda")
 ```

### 2022-06-07 Analysis/06-Figure2_with_correlations.Rmd
renamed figure correlation (now 2)

--- 2022-05-26 Analysis/06-Figure3_with_correlations.Rmd
@@
 
-This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_innovation_analyses.Rmd.
+This markdown contains the code to reproduce the figures containing the elaboration and innovation per species for the all birds and the passeriformes analyses.
+For the details on how to get the data objects called here, refer to `03-elaboration_innovation_analyses.Rmd`.
 
@@
 
-# All birds shapespace analyses
+# All birds analysis
 
-```{r, eval = TRUE, echo = FALSE}
+## Loading the data
+
+```{r, eval = TRUE, echo = TRUE}
 ## Load the data
@@
 
-Extract the median correlations from the results (for the groups ion the phylogeny and the groups on themselves)
+Extract the median elaboration and innovation from the results for the mega level (projection on the phylogeny) and the macro level (projection on each group).
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = TRUE, echo = TRUE}
 ## Calculate the correlations for the projections onto the phylo axis
@@
 
-### Figure b): the overall density plot
+## Ploting the figure
+
+The following snippets plots the entire figure minus the panel IDs:
+ * Panel A: the overal density figure
+ * Panel B: the correlation plots for the projection on the mega level
+ * Panel C: the correlation plots for the projection on the macro level
 
@@
 
-```{r, echo = FALSE, eval = FALSE}
-pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4.5, width = 4)
+```{r, echo = TRUE, eval = TRUE}
+## Set the pdf parameters
+pdf("../Manuscript/Figures/Figure2_correlations.pdf", height = 12, width = 8)
 
-## Plotting the densities with the quadrant
+## Set up the plotting layout
+layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
+layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
+plot_layout <- layout(layout_matrix)
+# layout.show(plot_layout)
+
+## Setting the colour pallette for the subsequent plots
+colour.palette <- gg.color.hue
+
+## --------
+## PANEL A
+## --------
+
+## Plotting the densities (with the quadrants) in the first panel
 par(mar = c(5, 4, 4, 2) + 0.1)
@@
 
-dev.off()
-```
-
-### Figure a): explanations
-
-This might have to be removed.
-
-```{r, echo = FALSE, eval = FALSE}
-pdf("../Manuscript/Figures/Figure3_a.pdf", height = 6, width = 3)
-## Plotting the projection cartoon
-library(mcmcmcglmmm)
-## Get projection lines (from a centred matrix)
-get.proj.line <- function(point, base) {
-  point <- t(point)
-  base <- t(base)
-  proj_x <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[1];
-  proj_y <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[2];
-  return(rbind(cbind(proj_x, proj_y), point))
-}
-
-set.seed(42)
-## Space plot
-cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
-space <- space.maker(10, 2, rnorm, cor.matrix = cor_matrix)
-lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
-
-par(mfrow = c(2, 1))
-## Projections per species
-## Translate the major axis
-par(bty = "n", mar = c(1,1,1,1)+0.1)
-plot(space, pch = 19, xlab = "", ylab = "", xlim = lim, ylim = lim, col = "grey", main = "Panels B and C: species'\nprojection on the phylogeny", xaxt = "n", yaxt = "n")
-abline(v = 0, col = "grey", lwd = 0.5)
-abline(h = 0, col = "grey", lwd = 0.5)
-## Plotting the ellipse
-lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "grey", lwd = 1)
-## Plotting the major axis
-lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "grey", lty = 1)
-lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "orange")
-# legend("bottomright", legend = "C")
-
-## Do the projection
-matrix <- space
-point1 <- c(-1.2, 0)
-point2 <- c(2, 0.75)
-base_vector <- rbind(point1, point2)
-
-## Get all the space (with the two last rows being the base vectors)
-space_cent <- rbind(matrix, base_vector)
-## Centre all the space
-space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
-## Re-attribute the centred variables
-matrix_cent <- space_cent[1:nrow(matrix), ]
-base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]
-
-## Plot the projections
-for(i in 1:nrow(matrix)) {
-  cent_proj <- get.proj.line(matrix_cent[i,], base_vector_cent[2, ])
-  cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
-  lines(cent_proj, col = "darkgreen")
-}
-
-## Species projection onto their own axis
-par(bty = "n", mar = c(1,1,1,1)+0.1)
-plot(space, pch = 19, xlab = "", ylab = "", xlim = lim, ylim = lim, col = "grey", main = "Panels D: species'\nprojection on their own group", xaxt = "n", yaxt = "n")
-abline(v = 0, col = "grey", lwd = 0.5)
-abline(h = 0, col = "grey", lwd = 0.5)
-lines(x = c(-1.2, 2), y = c(0, 0.75), lwd = 3, col = "grey")
-## Plotting the ellipse
-lines(ellipse::ellipse(cor_matrix, level = 0.75), col = "blue", lwd = 1)
-## Plotting the major axis
-lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75), lwd = 3, col = "blue", lty = 1)
-# legend("bottomright", legend = "D")
-
-## Do the projection
-matrix <- space
-base_vector <- get.one.axis(list(VCV = cor_matrix, Sol = c(0,0)), level = 0.75)
-point1 <- base_vector[1,]
-point2 <- base_vector[2,]
-
-## Get all the space (with the two last rows being the base vectors)
-space_cent <- rbind(matrix, base_vector)
-## Centre all the space
-space_cent <- space_cent - rep(point1, rep.int(nrow(space_cent), ncol(space_cent)))
-## Re-attribute the centred variables
-matrix_cent <- space_cent[1:nrow(matrix), ]
-base_vector_cent <- space_cent[-c(1:nrow(matrix)), ]
-## Extend the base vector a little bit
-extend_base <- base_vector
-extend_base[1, ] <- extend_base[1,1] + 0.05
-lines(extend_base, lwd = 1, col = "blue", lty = 2)
-
-## Plot the projections
-for(i in 1:nrow(matrix_cent)) {
-  cent_proj <- get.proj.line(matrix_cent[i,], base_vector_cent[2, ])
-  cent_proj[, 1] <- (cent_proj[, 1] + point1[1])
-  cent_proj[, 2] <- (cent_proj[, 2] + point1[2])
-  lines(cent_proj, col = "darkgreen")
-}
-dev.off()
-```
-
-## Figure c) and d): the correlation plots for each group
-
-```{r, echo = FALSE, eval = FALSE}
-## Loading the distances from centre for each species (if adding outlines)
-load("../Data/Processed/ploting_data.rda")
-centroids <- ploting_data$distances_to_centre[[1]]
-
-## Setting the colour pallette
-colour.palette <- gg.color.hue
-
-pdf("../Manuscript/Figures/Figure3_cd.pdf", height = 12, width = 8)
-
-## Make the plot layout
-layout_matrix <- cbind(matrix(0,2,2), matrix(0,2,2))
-layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8, 2, 4, byrow = TRUE)), matrix(9:16, 2, 4,  byrow = TRUE))
-plot_layout <- layout(layout_matrix)
-# layout.show(plot_layout)
+## --------
+## PANEL B
+## --------
 
@@
                  "topleft", "topright", "topleft", "topright")
-## Creat the vector for counting the legend numbers
+## Create the vector for counting the legend numbers
 legend_counter <- 1
 
+## Plotting each individual correlation plots
 for(i in 1:8) {
@@
 
-## Fig 3c
+## --------
+## PANEL C
+## --------
 
-# par(mfrow = c(2, 4))
 ## Getting the subsets for the colouring
@@
                  "topleft", "topleft", "topright", "bottomright")
+## Resetting the legend_counter
 legend_counter <- 1
 
+## Plotting each individual correlation plots
 for(i in 1:8) {
@@
 
-# Correlations for passeriformes
+Extract the median elaboration and innovation from the results for the macro level (projection on the passeriformes) and the macro level-1 (projection on each group).
 
-```{r, eval = TRUE, echo = FALSE}
+# Passeriformes analysis
+
+## Loading the data
+
+```{r, eval = TRUE, echo = TRUE}
 ## Load the data
@@
 
-```{r, echo = FALSE, eval = FALSE}
-pdf("../Manuscript/Figures/Figure3_b.pdf", height = 4.5, width = 4)
+## Ploting the figure
 
-## Plotting the densities with the quadrant
+The following snippets plots the entire figure minus the panel IDs:
+ * Panel A: the overal density figure
+ * Panel B: the correlation plots for the projection on the mega level
+ * Panel C: the correlation plots for the projection on the macro level
+
+Density plot of the correlation of the median elaboration and median exploration for all species.
+
+```{r, echo = TRUE, eval = TRUE}
+## Set the pdf parameters
+pdf("../Manuscript/Figures/Figure2_correlations.pdf", height = 12, width = 8)
+
+## Set up the plotting layout
+layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
+layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
+plot_layout <- layout(layout_matrix)
+# layout.show(plot_layout)
+
+## Setting the colour pallette for the subsequent plots
+colour.palette <- gg.color.hue
+
+## --------
+## PANEL A
+## --------
+
+## Plotting the densities (with the quadrants) in the first panel
 par(mar = c(5, 4, 4, 2) + 0.1)
-plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)
+plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)
 
-dev.off()
-```
-
-## Figure 3b
-
-```{r, echo = FALSE, eval = FALSE}
-pdf("../Manuscript/Figures/Figure3_cd_passeriformes_1.pdf", height = 8, width = 8)
-par(mfrow = c(2,2))
+## --------
+## PANEL B
+## --------
 
@@
                  "topleft", "topright", "topleft", "topright")
+## Create the vector for counting the legend numbers
 legend_counter <- 1
 
-for(i in 1:2) {
+## Plotting each individual correlation plots
+for(i in 1:8) {
     par(mar = c(2,2.5,2,1.5))
-    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = FALSE)
+    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = FALSE)
 }
 
-## Fig 3c
+## --------
+## PANEL C
+## --------
 
-# par(mfrow = c(2, 4))
 ## Getting the subsets for the colouring
@@
                  "topleft", "topleft", "topright", "bottomright")
+## Resetting the legend_counter
 legend_counter <- 1
 
-for(i in 1:2) {
+## Plotting each individual correlation plots
+for(i in 1:8) {
     par(mar = c(2,2.5,2,1.5))
-    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = TRUE, contour = NULL)
+    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
 }

### 2022-06-10 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd
fixes to data

--- 2022-05-25 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 library(dispRity)
+source("../Functions/ploting.utilities.R")
 ```
@@
 ```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
-## The ggplot color panel
-gg.color.hue <- function(n) {
-    grDevices::hcl(h = seq(15, 375, length = n + 1), l = 65, c = 100)[1:n]
-}
-
 # colour.palette <- rainbow
@@
 ```{r, eval = TRUE, echo = FALSE}
-visualisation.group <- function(data, subset, dimensions, n = 100) {
-
-    ## Get the subset
-    sub <- get.subsets(data, subsets = subset)
-
-    ## Make 2D
-    if(!missing(dimensions)) {
-        sub$matrix[[1]] <- sub$matrix[[1]][,dimensions]
-        sub$call$dimensions <- dimensions
-        fun2d <- function(covar) {
-            covar$VCV <- covar$VCV[dimensions, dimensions]
-            covar$loc <- covar$loc[dimensions]
-            return(covar)
-        }
-        sub$covar <- lapply(sub$covar, lapply, fun2d)
-        main_D <- "2D"
-    } else {
-        main_D <- "8D"
-    }
-
-    ## Make the tile
-    main <- paste0(n, " random ", subset, " VCV (", main_D, ")")
-
-    ## Plot
-    covar.plot(sub, major.axes = TRUE, ellipses = TRUE, points = FALSE, legend = FALSE, main = main)
-    covar.plot(sub, major.axes = mean, ellipses = mean, points = FALSE, col = "red", add = TRUE)
-    rm(sub)
-}
-
 visualisation.group(superorder_shapespace, "Mirandornithes")
@@
 
+## Get the data at the macro level (with no missing data)
+get.macro.projection <- function(data, what) {
+    if(what == "centroids") {
+        ## Measuring the centroid distances
+        data <- dispRity(data, metric = dispRity::centroids)
+    } else {
+        data <- data[[what]]
+    }
+    ## Getting all the species
+    all_species <- apply(get.disparity(get.subsets(data, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median)
+    ## Getting the species that have a projection on their own group
+    macro_level <- apply(do.call(rbind, get.disparity(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)]), concatenate = FALSE)), 1, median)
+    ## Replacing the species that have their own group
+    all_species[match(names(macro_level), names(all_species))] <- macro_level
+}
+
+## Getting the plotting data super_order_level
+ploting_data_superorder <- list(
+                     "median_elaborations" = get.macro.projection(results$super_results_within, what = "position"),
+                     "median_innovations"  = get.macro.projection(results$super_results_within, what = "distance"),
+                     "distances_to_centre" = get.macro.projection(superorder_shapespace, what = "centroids"),
+                     "consensus_tree"      = tree)
+
+## Getting the plotting data super_order_level
+ploting_data_order <- list(
+                     "median_elaborations" = get.macro.projection(results$order_results_within, what = "position"),
+                     "median_innovations"  = get.macro.projection(results$order_results_within, what = "distance"),
+                     "distances_to_centre" = get.macro.projection(order_shapespace, what = "centroids"),
+                     "consensus_tree"      = tree)
+
 ## Save all that as a list
-save(ploting_data, file = "ploting_data.rda")
+save(ploting_data, file = "../Data/Processed/ploting_data.rda")
+save(ploting_data_superorder, file = "../Data/Processed/ploting_data_superorder.rda")
+save(ploting_data_order, file = "../Data/Processed/ploting_data_order.rda")
 
@@
 }
-super_orders <- mapply(add.col, one_order = super_orders[1:8], one_col = as.list(tip_colours_super_orders), SIMPLIFY = FALSE)
-orders <- mapply(add.col, one_order = orders[1:27], one_col = as.list(tip_colours_orders), SIMPLIFY = FALSE)
+super_orders <- mapply(add.col, one_order = super_orders[1:8][match(names(super_orders[1:8]), names(tip_colours_super_orders))], one_col = as.list(tip_colours_super_orders), SIMPLIFY = FALSE)
+orders <- mapply(add.col, one_order = orders[1:27][match(names(orders[1:27]), names(tip_colours_orders))], one_col = as.list(tip_colours_orders), SIMPLIFY = FALSE)
 
@@
 
-# All birds shapespace analyses
+# Passeriformes shapespace analyses
 

### 2022-06-10 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
                      "median_innovations"  = apply(get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
-                     "distances_to_centre" = get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = centroids, centroid = 0)),
+                     "distances_to_centre" = get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = dispRity::centroids, centroid = 0))[[1]],
                      "consensus_tree"      = tree)
@@
     all_species <- apply(get.disparity(get.subsets(data, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median)
+    names(all_species) <- rownames(get.subsets(data, subsets = "phylogeny")$matrix[[1]])
     ## Getting the species that have a projection on their own group
     macro_level <- apply(do.call(rbind, get.disparity(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)]), concatenate = FALSE)), 1, median)
+    names(macro_level) <- unlist(lapply(lapply(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)])$subsets, `[[`, "elements"), function(X, matrix) rownames(matrix)[X], matrix = data$matrix[[1]]))
     ## Replacing the species that have their own group
     all_species[match(names(macro_level), names(all_species))] <- macro_level
+    return(all_species)
 }

### 2022-06-10 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 param_chains <- list()
-for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_shapespace_passeriformes_chain", i, ".rda"))
-    param_chains[[i]] <- params_shapespace_passeriformes[[1]]
-}
+# for(i in 1:3) {
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 1, ".rda"))
+    param_chains[[1]] <- param_chain_passeriformes_50levels_chain1[[1]]
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 2, ".rda"))
+    param_chains[[2]] <- param_chain_passeriformes_50levels_chain1[[1]]
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 3, ".rda"))
+    param_chains[[3]] <- param_chain_passeriformes_50levels_chain1[[1]]
+# }
 ## Extract the parameters from the parameter chains

### 2022-06-10 Analysis/09-novelty_predictions.Rmd
fast piggles done

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This markdown contains the code to reproduce the figures containing the elaboration and innovation per species for the all birds and the passeriformes analyses.
For the details on how to get the data objects called here, refer to `03-elaboration_innovation_analyses.Rmd`.

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
library(phylolm)
source("../Functions/ploting.utilities.R")
source("../Functions/correlations.utilities.R")
```

# All birds analysis

## PGLS tables

This table contains the results of the model `centroids ~ elaboration*innovation` where `centroids` is a proxy for novelty.
Results at different levels are for that level (i.e. or the Anseriformes results, it's their distance from their centroid and their elaboration/projection on their VCV major axis).

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/ploting_data.rda")
load("../Data/Processed/ploting_data_superorder.rda")
load("../Data/Processed/ploting_data_order.rda")
load("../Data/Processed/clade_info.rda")

## Running one pgls
run.one.pgls <- function(group, data, clade_info) {
      cat(".")
      if(group != "all_birds") {
            ## Extract the specific group
            if(group %in% names(clade_info[[1]])) {
                  ## Is super order
                  clade_info <- clade_info[[1]]
            } else {
                  ## Is order
                  clade_info <- clade_info[[2]]
            }

            ## Get the species in the clade
            sp_list <- clade_info[[which(names(clade_info) == group)]]$species

            ## Clean the tree
            short_tree <- drop.tip(data$consensus_tree,
                                   tip = data$consensus_tree$tip.label[!(data$consensus_tree$tip.label %in% sp_list)])
      } else {
            ## Doing it on all birds
            sp_list <- names(data$median_elaborations)
            short_tree <- data$consensus_tree
      }

      ## Set up the data
      short_data <- data.frame("centroids" = data$distances_to_centre[sp_list],
                               "elaboration" = data$median_elaborations[sp_list],
                               "innovation" = data$median_innovation[sp_list])

      ## Run the piggles
      return(phylolm::phylolm(centroids ~ elaboration*innovation, data = short_data, phy = short_tree, model = "lambda"))
}

## Get the mega level pgls
all_birds_pgls <- run.one.pgls("all_birds", ploting_data, clade_info)

## Get the macro level pgls
superorder_pgls <- lapply(as.list(names(clade_info[[1]])), run.one.pgls, ploting_data_superorder, clade_info)
order_pgls <- lapply(as.list(names(clade_info[[2]])), run.one.pgls, ploting_data_order, clade_info)

## Summarise that?
extract.summary <- function(model) {
      sum_mod <- summary(model)
      coefficients <- c(t(sum_mod$coefficients[-1, c(1, 4)]))
      names(coefficients) <- c("elaboration", "p_elab", "innovation", "p_innov", "interaction", "p_interact")
      c("lambda" = sum_mod$optpar, coefficients)
}

## Combine these results
all_results <- as.data.frame(rbind(extract.summary(all_birds_pgls),
                     do.call(rbind, lapply(superorder_pgls, extract.summary)),
                     do.call(rbind, lapply(order_pgls, extract.summary))))
## Add the labels
labels <- data.frame("name"  = c("All_birds", names(clade_info[[1]]), names(clade_info[[2]])),
                     "level" = c("all_birds", rep("super_order", length(clade_info[[1]])),  rep("order", length(clade_info[[2]]))))

knitr::kable(cbind(labels, round(all_results, 3)))
```

### 2022-06-13 Analysis/10-sensitivity_analysis.Rmd
Create 10-sensitivity_analysis.Rmd

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This markdown contains the script for runing the sensitivty analyses.
Due to computational time, the sensitivity analysis are only run on a subset of the entire trait space, the charadriiformes.

```{r}
library(dispRity)
library(mcmcmcglmmm)
```

## Observed results for charadriiformes

For this section we are just running the normal analysis (as in the main text) but on the charadriiformes species only.
The charadriiformes dataset is a present as an example dataset in the dispRity package

```{r, echo = FALSE, eval = TRUE}
## Loading the Charadriiformes data
data(charadriiformes)
## Extracting the tree
tree <- charadriiformes$tree
## Extracting the data column that contains the clade assignments
data <- charadriiformes$data[, "clade"]
## Changing the levels names (the clade names) to colours
levels(data) <- c("orange", "blue", "darkgreen")
data <- as.character(data)
## Matching the data rownames to the tip order in the tree
data <- data[match(ladderize(tree)$tip.label, rownames(charadriiformes$data))]

## Matching the tip colours (labels) to their descending edges in the tree
## (and making the non-match edges grey)
clade_edges <- match.tip.edge(data, tree, replace.na = "grey")

## Plotting the results
plot(ladderize(tree), show.tip.label = FALSE, edge.color = clade_edges, main = "Charadriiformes")
legend("bottomleft", lty = c(1,1,1), col = c("blue", "darkgreen", "orange"), legend = c("plovers", "sandpipers", "gulls"))
axisPhylo()
```
Fgure 1: the charadriiformes consensus tree

```{r, echo = FALSE, eval = TRUE}
## Creating a dispRity object from the charadriiformes model
plot(shapespace[, c(1,2)], xlab = "PC1", ylab = "PC2", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4), main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
legend("bottomrigh", pch = 19, legend = c("gulls", "plovers", "sandpipers"), col= c("orange", "blue", "darkgreen"))
```
Figure 2: the charadriiformes shapespace

## Ball shapesapce

In this sensitivity analysis we transform the shapespace to make each dimension the same size (i.e. each dimension explains the same amount of variance).
We do that by simply scaling the values on each dimensions.

## No phylogenetic signal

In this sensitivity we are aiming to "remove" the phylogenetic signal by randomising the attribution of each species traits.
We keep the phylogenetic structure as presented above (i.e. the charadriiformes tree) but randomly attribute a beak shape value to each species to dilute or "remove" the phylogenetic signal in each clade.

### Preparing the data

```{r, eval = FALSE, echo = FALSE}
source("../Functions/prep.data.R")
## Reading the different trait spaces
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")
shapespace <- shapespace$PCscores
## Reading the consensus tree
consensus_tree <- read.nexus("../Data/Raw/Tree_data/9993taxa_1_tree_Global_MCC_CA.tre")
## Reading the distribution of trees
trees_list <- read.tree("../Data/Raw/Tree_data/AllBirdsHackett1.tre")

## Making a charadriiformes only list
data(charadriiformes)
clade_list <- charadriiformes$data[,18, drop = FALSE]
colnames(clade_list) <- "level1"

## Shapespace passeriformes
shapespace_charadriiformes <- shapespace[match(rownames(clade_list),rownames(shapespace)), ]

## Do the prep data from here (with all the trees)
shapespace_charadriiformes_family <- prep.data(
                                          level = 0, lvl.inc = c(1),
                                          clades    = clade_list,
                                          space     = shapespace_charadriiformes,
                                          dim       = 1:8,
                                          consensus = consensus_tree,
                                          trees     = trees_list,
                                          verbose   = TRUE)
save(shapespace_charadriiformes_family, file = "../Data/Processed/shapespace_charadriiformes_family.rda")
```

### Setting the mini-chains

```{r, eval = TRUE, message = FALSE}
## Loading the parametrisation data
load(file = "../Data/Processed/shapespace_charadriiformes_family.rda")

## Select the data
shapespace <- shapespace_charadriiformes_family[[1]]$space
selected_dim <- shapespace_charadriiformes_family[[1]]$dimensions
consensus_tree <- shapespace_charadriiformes_family[[1]]$consensus_tree
trees_list <- shapespace_charadriiformes_family[[1]]$trees_list
```

## Mini-chains parametrization

```{r, eval = TRUE}
## Set up the parametrising chains models
param_chains <- make.mini.chains(data         = shapespace,
                                 dimensions   = selected_dim,
                                 tree         = consensus_tree,
                                 trait.family = "gaussian",
                                 randoms      = c("global", "level1"),
                                 residuals    = "global",
                                 priors       = 0.02,
                                 verbose      = TRUE,
                                 parameters   = list(
                                       nitt   = 20000,
                                       thin   = 500,
                                       burnin = 0))

# Running the three MCMCglmm models
save(param_chains, file = "../Data/Processed/params_shapespace_charadriiformes.rda")
```

```{r, eval = TRUE}
## Run the parametrisation chains (local)
params <- run.mini.chains(param_chains, replicates = 3)

## Get the estimated parameters
est_params <- extract.parameters(params)
```

## Set up the mini-chain

```{r, eval = TRUE}
## The thinning parameter
thin <- 200
## The number of samples per chains
samples <- 10
## Set up the mini-chains models
shapespace_charadriiformes <- make.mini.chains(
                                data         = shapespace,
                                dimensions   = selected_dim,
                                tree         = trees_list[sample(1:1000, 400)],
                                trait.family = "gaussian",
                                randoms      = c("global", "level1"),
                                residuals    = "global",
                                priors       = est_params$priors,
                                verbose      = TRUE,
                                parameters   = list(
                                      nitt   = est_params$burnin + thin * samples,
                                      thin   = thin,
                                      burnin = est_params$burnin))
## Save the mini.chains
save(shapespace_charadriiformes, file = "../Data/Cluster/shapespace_charadriiformes/shapespace_charadriiformes.mini.chains")
```

# Randomspace

```{r}
## Get the scaling the variance of each dimension
randospace <- shapespace
randomise.space <- function(space, dims) {
      rando <- space
      rando[, dims] <- space[sample(1:nrow(space), nrow(space)), dims]
      return(rando)
}

## Comparing D1 and D3 for the contrast
par(mfrow = c(2,1))
plot(shapespace[, c(1, 2)], xlab = "PC1", ylab = "PC2", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4),
      main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
plot(randomise.space(shapespace, selected_dim)[, c(1, 2)], xlab = "PC1", ylab = "PC2", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4),
      main = "randomised shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
```

## Mini-chains (using the parameters from the normal chain above)

```{r, eval = TRUE}
## The thinning parameter
thin <- 200
## The number of samples per chains
samples <- 10

## Setting up 20 random replicates
for(i in 1:25) {
      ## Set up the mini-chains models
      randospace_charadriiformes_ <- make.mini.chains(
                                      data         = randomise.space(shapespace, selected_dim),
                                      dimensions   = selected_dim,
                                      tree         = trees_list[sample(1:1000, 400)],
                                      trait.family = "gaussian",
                                      randoms      = c("global", "level1"),
                                      residuals    = "global",
                                      priors       = est_params$priors,
                                      verbose      = TRUE,
                                      parameters   = list(
                                            nitt   = est_params$burnin + thin * samples,
                                            thin   = thin,
                                            burnin = est_params$burnin))
      ## Save the mini.chains
      save(randospace_charadriiformes_, file = paste0("../Data/Cluster/shapespace_charadriiformes/randospace_charadriiformes_", i, ".mini.chains"))
}
```

# Ballspace

```{r}
## Get the scaling the variance of each dimension
ballspace <- shapespace
ballspace[, selected_dim] <- apply(shapespace[, selected_dim], 2, scale)

## Comparing D1 and D3 for the contrast
par(mfrow = c(2,1))
plot(shapespace[, c(1,3)], xlab = "PC1", ylab = "PC3", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4),
      main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
plot(ballspace[, c(1,3)], xlab = "PC1", ylab = "PC3", xlim = c(-3, 3), ylim = c(-3, 3), ,
      main = "balled shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
```

## Mini-chains parametrization

```{r, eval = TRUE}
## Set up the parametrising chains models
param_chains <- make.mini.chains(data         = ballspace,
                                 dimensions   = selected_dim,
                                 tree         = consensus_tree,
                                 trait.family = "gaussian",
                                 randoms      = c("global", "level1"),
                                 residuals    = "global",
                                 priors       = 0.02,
                                 verbose      = TRUE,
                                 parameters   = list(
                                       nitt   = 20000,
                                       thin   = 500,
                                       burnin = 0))

# Running the three MCMCglmm models
save(param_chains, file = "../Data/Processed/params_ballspace_charadriiformes.rda")
```

```{r, eval = TRUE}
## Run the parametrisation chains (local)
params <- run.mini.chains(param_chains, replicates = 3)

## Get the estimated parameters
est_params <- extract.parameters(params)
```

## Set up the mini-chain

```{r, eval = TRUE}
## The thinning parameter
thin <- 200
## The number of samples per chains
samples <- 10
## Set up the mini-chains models
ballspace_charadriiformes <- make.mini.chains(
                                data         = ballspace,
                                dimensions   = selected_dim,
                                tree         = trees_list[sample(1:1000, 400)],
                                trait.family = "gaussian",
                                randoms      = c("global", "level1"),
                                residuals    = "global",
                                priors       = est_params$priors,
                                verbose      = TRUE,
                                parameters   = list(
                                      nitt   = est_params$burnin + thin * samples,
                                      thin   = thin,
                                      burnin = est_params$burnin))
## Save the mini.chains
save(ballspace_charadriiformes, file = "../Data/Cluster/shapespace_charadriiformes/ballspace_charadriiformes.mini.chains")
```

### 2022-06-14 Analysis/10-sensitivity_analysis.Rmd
Update 10-sensitivity_analysis.Rmd

@@
 
-## Observed results for charadriiformes
+# Null models
 
-For this section we are just running the normal analysis (as in the main text) but on the charadriiformes species only.
-The charadriiformes dataset is a present as an example dataset in the dispRity package
+From this space we simulated three null model scenarios to test the impact of specific parameters in our approach:
+
+ * the `randomspace` model to test the effect of phylogenetic signal on our results: we can test the effect of phylogenetic correlation and phylogenetic signal on our results by randomly assigning beaks to different species to "dilute" the phylogenetic signal: i.e.
+so that there is no assumption that two closely related species have similar beaks.
+We do that be reasigning beaks shape in the shapespace randomly for each species in each step of the `mcmcmcglmmm` analysis.
+
+ * the `ballspace` model to test the effect of decreasing variance on each dimension of the shapespace: in fact, our main findings (elaboration is more common than innovation at the species level) might be due to the nature of the shapespace where the first dimension (PC1) has more variance than the other ones.
+This can pull the major axis of the global phylogenetic VCV to be aligned with PC1 where there is more opportunity for each individiual species to have higher values of elaboration (i.e.
+more changes aligned with PC1) that innovation (i.e.
+less changes on the other PCs).
+We can test this by rescaling the variance of each PC axes to be equal.
+
+ * the `BMspace` model to test the deviation from a base Brownian Motion trait evolution model: for this model we simulate a series of shapespaces with the same number of species and the same phylogeny using a Brownian motion process for each 8 dimensions with a decreasing variance on each dimensions to match the observed variance in per dimension in the shapespace.
+
+For each of these null models, we resimulated a shapespace with the modifications highlighted above and rerun the whole `mcmcmcglmmm` and projection pipeline as for in the main study.
+For calculation speed reasons we only ran these models on a subset of the global dataset, the charadriiformes containing 359 sp divided into three groups (159 gulls, 98 plovers and 102 sandpipers).
+
+# Simulating the different spaces
+
+## Original data
+
+```{r, eval = FALSE, echo = FALSE}
+source("../Functions/prep.data.R")
+## Reading the different trait spaces
+shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")
+shapespace <- shapespace$PCscores
+## Reading the consensus tree
+consensus_tree <- read.nexus("../Data/Raw/Tree_data/9993taxa_1_tree_Global_MCC_CA.tre")
+## Reading the distribution of trees
+trees_list <- read.tree("../Data/Raw/Tree_data/AllBirdsHackett1.tre")
+
+## Making a charadriiformes only list
+data(charadriiformes)
+clade_list <- charadriiformes$data[,18, drop = FALSE]
+colnames(clade_list) <- "level1"
+
+## Shapespace passeriformes
+shapespace_charadriiformes <- shapespace[match(rownames(clade_list),rownames(shapespace)), ]
+
+## Do the prep data from here (with all the trees)
+shapespace_charadriiformes_family <- prep.data(
+                                          level = 0, lvl.inc = c(1),
+                                          clades    = clade_list,
+                                          space     = shapespace_charadriiformes,
+                                          dim       = 1:8,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_charadriiformes_family, file = "../Data/Processed/shapespace_charadriiformes_family.rda")
+```
+
+```{r, eval = TRUE, echo = FALSE}
+## Loading the parametrisation data
+load(file = "../Data/Processed/shapespace_charadriiformes_family.rda")
+
+## Select the data
+shapespace <- shapespace_charadriiformes_family[[1]]$space
+selected_dim <- shapespace_charadriiformes_family[[1]]$dimensions
+consensus_tree <- shapespace_charadriiformes_family[[1]]$consensus_tree
+trees_list <- shapespace_charadriiformes_family[[1]]$trees_list
+```
 
@@
 
-## Ball shapesapce
+## The randomspace
 
-In this sensitivity analysis we transform the shapespace to make each dimension the same size (i.e. each dimension explains the same amount of variance).
-We do that by simply scaling the values on each dimensions.
+To generate the `randomspace` we shuffle each beak to random species (i.e. reshuffling rows in the shapespace):
 
-## No phylogenetic signal
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12}
+## Randomise each species
+randospace <- shapespace
+randomise.space <- function(space, dims) {
+      rando <- space
+      rando[, dims] <- space[sample(1:nrow(space), nrow(space)), dims]
+      return(rando)
+}
 
-In this sensitivity we are aiming to "remove" the phylogenetic signal by randomising the attribution of each species traits.
-We keep the phylogenetic structure as presented above (i.e.
-the charadriiformes tree) but randomly attribute a beak shape value to each species to dilute or "remove" the phylogenetic signal in each clade.
+## Comparing D1 and D3 for the contrast
+par(mfrow = c(1,2))
+plot(shapespace[, c(1, 2)], xlab = write.pc(1, shape_var_prop), ylab = write.pc(2, shape_var_prop),
+      xlim = range.axis(shapespace, selected_dim), ylim = range.axis(shapespace, selected_dim),
+      main = "Original shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+plot(randomise.space(shapespace, selected_dim)[, c(1, 2)], xlab = write.pc(1, shape_var_prop), ylab = write.pc(2, shape_var_prop),
+      xlim = range.axis(shapespace, selected_dim), ylim = range.axis(shapespace, selected_dim),
+      main = "Randomised shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+```
 
-### Preparing the data
+## The ballspace
+
+To generate the `ballspace` we scaled the data on each dimension to have same variance:
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12}
+## Get the scaling the variance of each dimension
+ballspace <- shapespace
+ballspace[, selected_dim] <- apply(shapespace[, selected_dim], 2, scale)
+
+shape_var_prop <-  apply(shapespace[, selected_dim], 2, var)
+shape_var_prop <- shape_var_prop/sum(shape_var_prop)*100
+ball_var_prop <-  apply(ballspace[, selected_dim], 2, var)
+ball_var_prop <- ball_var_prop/sum(ball_var_prop)*100
+
+## Plotting fun
+write.pc <- function(axis, space) {
+      paste0("PC", axis, " (", round(var[axis], 2), "%)")
+}
+range.axis <- function(space, dims) {
+      return(range(pretty(unlist(c(space[, dims])))))
+}
+
+## Comparing D1 and D3 for the contrast
+par(mfrow = c(1,2))
+plot(shapespace[, c(1,3)], xlab = write.pc(1, shape_var_prop), ylab = write.pc(3, shape_var_prop),
+      xlim = range.axis(shapespace, selected_dim), ylim = range.axis(shapespace, selected_dim),
+      main = "Original shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+plot(ballspace[, c(1,3)], xlab = write.pc(1, ball_var_prop), ylab = write.pc(3, ball_var_prop),
+      xlim = range.axis(ballspace, selected_dim), ylim = range.axis(ballspace, selected_dim),
+      main = "Balled shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+```
+
+## The BMspace
+
+To generate the `BMspace` we simulate 8 uncorrelated Brownian Motions on the consensus tree with the variance for each BM being the observed one for each dimension of the shapespace:
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12}
+## Get the scaling the variance of each dimension
+BMise.space <- function(space, tree, dims) {
+      ## Get the space dimenions variance
+      space_var_prop <-  apply(space[, dims], 2, var)
+      space_var_prop <- space_var_prop/sum(space_var_prop)*100
+
+      ## Simulate the space
+      return(sapply(space_var_prop, function(var, tree) fastBM(tree = tree, sig2 = var), tree))
+}
+
+## Comparing D1 and D3 for the contrast
+par(mfrow = c(1,2))
+plot(shapespace[, c(1, 2)], xlab = write.pc(1, shape_var_prop), ylab = write.pc(2, shape_var_prop),
+      xlim = range.axis(shapespace, selected_dim), ylim = range.axis(shapespace, selected_dim),
+      main = "Original shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+BMspace <- BMise.space(shapespace, tree, selected_dim)
+BM_var_prop <-  apply(BMspace[, selected_dim], 2, var)
+BM_var_prop <- BM_var_prop/sum(BM_var_prop)*100
+plot(BMspace[, c(1, 2)], xlab = write.pc(1, BM_var_prop), ylab = write.pc(2, BM_var_prop),
+      xlim = range.axis(BMspace, selected_dim), ylim = range.axis(BMspace, selected_dim),
+      main = "Randomised shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+```
+
+# Setting the mini-chains
+
+## Original, BM space and randomised spaces (same parameters)
+
+We used the same mini chains parameters for the BMspace and the randomspace since both are stochastic spaces (i.e.
+each space is a different itteration) making it to intensive computationally to estimate the starting parameters every iteration.
+We therefore used the conservative parameters from the original space.
+We used three chains of 20000 iterations for the parameterisation chains.
 
 ```{r, eval = FALSE, echo = FALSE}
-source("../Functions/prep.data.R")
-## Reading the different trait spaces
-shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")
-shapespace <- shapespace$PCscores
-## Reading the consensus tree
-consensus_tree <- read.nexus("../Data/Raw/Tree_data/9993taxa_1_tree_Global_MCC_CA.tre")
-## Reading the distribution of trees
-trees_list <- read.tree("../Data/Raw/Tree_data/AllBirdsHackett1.tre")
-
-## Making a charadriiformes only list
-data(charadriiformes)
-clade_list <- charadriiformes$data[,18, drop = FALSE]
-colnames(clade_list) <- "level1"
-
-## Shapespace passeriformes
-shapespace_charadriiformes <- shapespace[match(rownames(clade_list),rownames(shapespace)), ]
-
-## Do the prep data from here (with all the trees)
-shapespace_charadriiformes_family <- prep.data(
-                                          level = 0, lvl.inc = c(1),
-                                          clades    = clade_list,
-                                          space     = shapespace_charadriiformes,
-                                          dim       = 1:8,
-                                          consensus = consensus_tree,
-                                          trees     = trees_list,
-                                          verbose   = TRUE)
-save(shapespace_charadriiformes_family, file = "../Data/Processed/shapespace_charadriiformes_family.rda")
-```
-
-### Setting the mini-chains
-
-```{r, eval = TRUE, message = FALSE}
-## Loading the parametrisation data
-load(file = "../Data/Processed/shapespace_charadriiformes_family.rda")
-
-## Select the data
-shapespace <- shapespace_charadriiformes_family[[1]]$space
-selected_dim <- shapespace_charadriiformes_family[[1]]$dimensions
-consensus_tree <- shapespace_charadriiformes_family[[1]]$consensus_tree
-trees_list <- shapespace_charadriiformes_family[[1]]$trees_list
-```
-
-## Mini-chains parametrization
-
-```{r, eval = TRUE}
 ## Set up the parametrising chains models
@@
 
-```{r, eval = TRUE}
+```{r, eval = FALSE, echo = FALSE}
 ## Run the parametrisation chains (local)
@@
 
-## Set up the mini-chain
-
-```{r, eval = TRUE}
+```{r, eval = FALSE, echo = FALSE}
 ## The thinning parameter
@@
 
-# Randomspace
-
-```{r}
-## Get the scaling the variance of each dimension
-randospace <- shapespace
-randomise.space <- function(space, dims) {
-      rando <- space
-      rando[, dims] <- space[sample(1:nrow(space), nrow(space)), dims]
-      return(rando)
-}
-
-## Comparing D1 and D3 for the contrast
-par(mfrow = c(2,1))
-plot(shapespace[, c(1, 2)], xlab = "PC1", ylab = "PC2", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4),
-      main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
-plot(randomise.space(shapespace, selected_dim)[, c(1, 2)], xlab = "PC1", ylab = "PC2", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4),
-      main = "randomised shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
-```
-
-## Mini-chains (using the parameters from the normal chain above)
-
-```{r, eval = TRUE}
+```{r, eval = FALSE, echo = FALSE}
 ## The thinning parameter
@@
 
-# Ballspace
+```{r, eval = FALSE, echo = FALSE}
+## The thinning parameter
+thin <- 200
+## The number of samples per chains
+samples <- 10
 
-```{r}
-## Get the scaling the variance of each dimension
-ballspace <- shapespace
-ballspace[, selected_dim] <- apply(shapespace[, selected_dim], 2, scale)
-
-## Comparing D1 and D3 for the contrast
-par(mfrow = c(2,1))
-plot(shapespace[, c(1,3)], xlab = "PC1", ylab = "PC3", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4),
-      main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
-plot(ballspace[, c(1,3)], xlab = "PC1", ylab = "PC3", xlim = c(-3, 3), ylim = c(-3, 3), ,
-      main = "balled shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+## Setting up 20 random replicates
+for(i in 1:25) {
+      ## Set up the mini-chains models
+      randospace_charadriiformes_ <- make.mini.chains(
+                                      data         = BMise.space(shapespace, tree, selected_dim),
+                                      dimensions   = selected_dim,
+                                      tree         = trees_list[sample(1:1000, 400)],
+                                      trait.family = "gaussian",
+                                      randoms      = c("global", "level1"),
+                                      residuals    = "global",
+                                      priors       = est_params$priors,
+                                      verbose      = TRUE,
+                                      parameters   = list(
+                                            nitt   = est_params$burnin + thin * samples,
+                                            thin   = thin,
+                                            burnin = est_params$burnin))
+      ## Save the mini.chains
+      save(randospace_charadriiformes_, file = paste0("../Data/Cluster/shapespace_charadriiformes/randospace_charadriiformes_", i, ".mini.chains"))
+}
 ```
 
-## Mini-chains parametrization
+## Ball space
 
-```{r, eval = TRUE}
+Since this space is not stochastic, we used the normal procedure for setting up the mini-chains.
+We used three chains of 20000 iterations for the parameterisation chains.
+
+```{r, eval = FALSE, echo = FALSE}
 ## Set up the parametrising chains models
@@
 ```
+
+## Running the mini-chains
+
+The mini-chains were run for 100 replicates to have 1000 exploitable posterior VCV results for each model.
+The following table summarises the modeel parameters, number of trees and spaces sampled (because the randomspace and the BMspace are stochastic we generated 25 of them) and number of explotable VCVs
+
+Model | parameters | replicates | trees sampled | spaces sampled | posterior VCVs |
+------|------------|------------|---------------|----------------|----------------|
+shapespace | from shapespace | 100 | 1 | 1000
+randomspace | from shapespace | 100 | 25 | 1000
+BMspace | from shapespace | 100 | 25 | 1000
+ballspace | from ballspace | 100 | 1 | 1000

### 2022-06-14 Analysis/10-sensitivity_analysis.Rmd
Update 10-sensitivity_analysis.Rmd

@@
 ```{r}
+set.seed(42)
 library(dispRity)
@@
 ```
+
 Fgure 1: the charadriiformes consensus tree
@@
 ```{r, echo = FALSE, eval = TRUE}
+## Shapespace var
+shape_var_prop <-  apply(shapespace[, selected_dim], 2, var)
+shape_var_prop <- shape_var_prop/sum(shape_var_prop)*100
+
+## Plotting fun
+write.pc <- function(axis, var) {
+      paste0("PC", axis, " (", round(var[axis], 2), "%)")
+}
+range.axis <- function(space, dims) {
+      return(range(pretty(unlist(c(space[, dims])))))
+}
+
 ## Creating a dispRity object from the charadriiformes model
-plot(shapespace[, c(1,2)], xlab = "PC1", ylab = "PC2", xlim = c(-0.4, 0.4), ylim = c(-0.4, 0.4), main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+plot(shapespace[, c(1,2)], xlab = write.pc(1, shape_var_prop), ylab = write.pc(2, shape_var_prop),
+    xlim = range.axis(shapespace, selected_dim), ylim = range.axis(shapespace, selected_dim),
+    main = "shapespace", pch = 19, col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
 legend("bottomrigh", pch = 19, legend = c("gulls", "plovers", "sandpipers"), col= c("orange", "blue", "darkgreen"))
 ```
+
 Figure 2: the charadriiformes shapespace
@@
 ## Randomise each species
-randospace <- shapespace
 randomise.space <- function(space, dims) {
@@
 
+Figure 3: difference between the shapespace and the randomspace
+
 ## The ballspace
@@
 
-shape_var_prop <-  apply(shapespace[, selected_dim], 2, var)
-shape_var_prop <- shape_var_prop/sum(shape_var_prop)*100
+## Ballspace var
 ball_var_prop <-  apply(ballspace[, selected_dim], 2, var)
 ball_var_prop <- ball_var_prop/sum(ball_var_prop)*100
-
-## Plotting fun
-write.pc <- function(axis, space) {
-      paste0("PC", axis, " (", round(var[axis], 2), "%)")
-}
-range.axis <- function(space, dims) {
-      return(range(pretty(unlist(c(space[, dims])))))
-}
 
@@
 
+Figure 4: difference between the shapespace and the ballspace
+
 ## The BMspace
@@
       ## Simulate the space
-      return(sapply(space_var_prop, function(var, tree) fastBM(tree = tree, sig2 = var), tree))
+      return(sapply(space_var_prop, function(var, tree) phytools::fastBM(tree = tree, sig2 = var), tree))
 }
@@
       xlim = range.axis(BMspace, selected_dim), ylim = range.axis(BMspace, selected_dim),
-      main = "Randomised shapespace", pch = 19,
+      main = "BM simulated shapespace", pch = 19,
       col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
 ```
+
+Figure 5: difference between the shapespace and the BMspace
 

### 2022-06-14 Analysis/10-sensitivity_analysis.Rmd
Update 10-sensitivity_analysis.Rmd

@@
 Model       | parameters      | replicates | trees sampled | spaces sampled | posterior VCVs |
-------|------------|------------|---------------|----------------|----------------|
-shapespace | from shapespace | 100 | 1 | 1000
-randomspace | from shapespace | 100 | 25 | 1000
-BMspace | from shapespace | 100 | 25 | 1000
-ballspace | from ballspace | 100 | 1 | 1000
+------------|-----------------|------------|---------------|----------------|----------------|
+shapespace  | from shapespace | 100        | 100           | 1              | 1000
+randomspace | from shapespace | 100        | 100           | 25             | 1000
+BMspace     | from shapespace | 100        | 100           | 25             | 1000
+ballspace   | from ballspace  | 100        | 100           | 1              | 1000

### 2022-06-14 Analysis/10-sensitivity_analysis.Rmd
Update 10-sensitivity_analysis.Rmd

--- 2022-06-14 Analysis/10-sensitivity_analysis.Rmd
@@
       ## Simulate the space
-      return(sapply(space_var_prop, function(var, tree) phytools::fastBM(tree = tree, sig2 = var), tree))
+      BMspace <- as.data.frame(sapply(space_var_prop, function(var, tree) phytools::fastBM(tree = tree, sig2 = var), tree))
+
+      ## Match the species with the correct levels
+      BMspace <- BMspace[match(space$animal, rownames(BMspace)), ]
+      BMspace$level1 <- space$level1
+      BMspace$animal <- space$animal
+      return(BMspace)
 }
@@
 ```{r, eval = FALSE, echo = FALSE}
+## Loading the parametrisation data
+load(file = "../Data/Processed/shapespace_charadriiformes_family.rda")
+
+## Select the data
+shapespace <- shapespace_charadriiformes_family[[1]]$space
+selected_dim <- shapespace_charadriiformes_family[[1]]$dimensions
+consensus_tree <- shapespace_charadriiformes_family[[1]]$consensus_tree
+trees_list <- shapespace_charadriiformes_family[[1]]$trees_list
+
+for(i in 1:25) {
+      ## Create the space
+      BM_space <- BMise.space(shapespace, tree, selected_dim)
+
+      ## Set up the parametrising chains models
+      param_chains <- make.mini.chains(data         = BM_space,
+                                       dimensions   = selected_dim,
+                                       tree         = consensus_tree,
+                                       trait.family = "gaussian",
+                                       randoms      = c("global", "level1"),
+                                       residuals    = "global",
+                                       priors       = 0.02,
+                                       verbose      = TRUE,
+                                       parameters   = list(
+                                             nitt   = 20000,
+                                             thin   = 500,
+                                             burnin = 0))
+
+      ## Run the parametrisation chains (local)
+      params <- run.mini.chains(param_chains, replicates = 3)
+
+      ## Get the estimated parameters
+      est_params <- extract.parameters(params)
+
       ## The thinning parameter
@@
       ## Setting up 20 random replicates
-for(i in 1:25) {
       ## Set up the mini-chains models
-      randospace_charadriiformes_ <- make.mini.chains(
-                                      data         = BMise.space(shapespace, tree, selected_dim),
+      BMspace_charadriiformes_ <- make.mini.chains(
+                                      data         = BM_space,
                                       dimensions   = selected_dim,
@@
       ## Save the mini.chains
-      save(randospace_charadriiformes_, file = paste0("../Data/Cluster/shapespace_charadriiformes/randospace_charadriiformes_", i, ".mini.chains"))
+      save(BMspace_charadriiformes_, file = paste0("../Data/Cluster/shapespace_charadriiformes/BMspace_charadriiformes_", i, ".mini.chains"))
 }
@@
 Model       | parameters      | replicates | trees sampled | spaces sampled | posterior VCVs |
-------|------------|------------|---------------|----------------|----------------|
-shapespace | from shapespace | 100 | 1 | 1000
-randomspace | from shapespace | 100 | 25 | 1000
-BMspace | from shapespace | 100 | 25 | 1000
-ballspace | from ballspace | 100 | 1 | 1000
+------------|-----------------|------------|---------------|----------------|----------------|
+shapespace  | from shapespace | 100        | 100           | 1              | 1000 |
+randomspace | from shapespace | 100        | 100           | 25             | 1000 |
+BMspace     | from shapespace | 100        | 100           | 25             | 1000 |
+ballspace   | from ballspace  | 100        | 100           | 1              | 1000 |
+
+# Analysing the results

### 2022-06-14 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

--- 2021-10-12 Analysis/01-Data_preparation.Rmd
@@
 
+## Passeriformes focus
+
+```{r}
+## Get the traitspace for all passeriformes
+load("../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
+clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
+shapespace_passeriformes <- shapespace_allbirds_lvl_superorder_order[[1]]$space[(clades_list[, "level2_order"] == "Passeriformes"), ]
+
+## Loading the Passeriformes clades
+load("../Data/Processed/clade.info_passerines.rda")
+## Clean the levels
+shapespace_passeriformes$level1 <- rep("", nrow(shapespace_passeriformes))
+shapespace_passeriformes$level2 <- rep("", nrow(shapespace_passeriformes))
+
+## Fill in the level1
+for(clade in 1:length(clade_info_passerines$passerine_big_clades)) {
+  clade_name <- names(clade_info_passerines$passerine_big_clades)[clade]
+  shapespace_passeriformes[clade_info_passerines$passerine_big_clades[[clade]], "level1"] <- clade_name
+}
+## Fill in the level2
+for(clade in 1:length(clade_info_passerines$passerine_small_clades)) {
+  clade_name <- names(clade_info_passerines$passerine_small_clades)[clade]
+  shapespace_passeriformes[clade_info_passerines$passerine_small_clades[[clade]], "level2"] <- clade_name
+}
+shapespace_passeriformes[, "level1"] <- as.factor(shapespace_passeriformes[, "level1"])
+shapespace_passeriformes[, "level2"] <- as.factor(shapespace_passeriformes[, "level2"])
+
+## Do the prep data from here (with all the trees! ouch!)
+shapespace_passeriformes_two_levels <- prep.data(
+                                          level = 0, lvl.inc = c(1,2),
+                                          clades    = shapespace_passeriformes[, c("level1", "level2")],
+                                          space     = shapespace_passeriformes,
+                                          dim       = 1:8,
+                                          consensus = consensus_tree,
+                                          trees     = trees_list,
+                                          verbose   = TRUE)
+save(shapespace_passeriformes_two_levels, file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
+```
+
 ## References

### 2022-06-14 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 ## Loading the parametrisation data
-load(file = "../Data/Processed/shapespace_passeriformes_suborder_family.rda")
+load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
 
 ## Select the data
-shapespace <- shapespace_passeriformes_suborder_family[[1]]$space
-selected_dim <- shapespace_passeriformes_suborder_family[[1]]$dimensions
-consensus_tree <- shapespace_passeriformes_suborder_family[[1]]$consensus_tree
-trees_list <- shapespace_passeriformes_suborder_family[[1]]$trees_list
+shapespace <- shapespace_passeriformes_two_levels[[1]]$space
+selected_dim <- shapespace_passeriformes_two_levels[[1]]$dimensions
+consensus_tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
+trees_list <- shapespace_passeriformes_two_levels[[1]]$trees_list
 ```
@@
 # Running the three MCMCglmm models
-save(param_chains, file = "../Data/Processed/params_shapespace_8D_50levels_50k_chain1.rda")
+save(param_chains, file = "../Data/Processed/params_passerines_shapespace_8D_28levels_50k.rda")
 ```

### 2022-06-16 Analysis/10-sensitivity_analysis.Rmd
Update 10-sensitivity_analysis.Rmd

@@
 
-## Original, BM space and randomised spaces (same parameters)
+## Original and randomised spaces (same parameters)
 
@@
 ```
+
+## BM space
+
+For the BM space, the parameters are estimated for each simulation individually to avoid corner case bugs (e.g. wrong Choleski decompositions!).
+We used three chains of 20000 iterations for the parameterisation chains.
 
@@
 randomspace | from shapespace | 100        | 100           | 25             | 1000 |
-BMspace     | from shapespace | 100        | 100           | 25             | 1000 |
+BMspace     | from each BMspace | 100        | 100           | 25             | 1000 |
 ballspace   | from ballspace  | 100        | 100           | 1              | 1000 |
 
+## Combining and diagnosing the results
+
+```{r, eval = TRUE, echo = FALSE}
+## Set up the variables
+models_path <- "../Data/Cluster/shapespace_charadriiformes/"
+model_files <- list.files(models_path, pattern = ".rda")
+
+## Original space
+shapespace_models_names <- model_files[grep("shapespace", model_files)]
+## Models placeholder
+shapespace_models <- list()
+
+## Load the models
+for(i in 1:length(shapespace_models_names)) {
+      load(paste0(models_path, shapespace_models_names[i], collapse = "/"))
+      shapespace_models <- c(shapespace_models, shapespace_20)
+}
+
+## Combine and diagnose
+shapespace_charadriiformes_posteriors <- combine.mini.chains(shapespace_models)
+silent <- diagnose.mini.chains(shapespace_charadriiformes_posteriors)
+save(shapespace_charadriiformes_posteriors, file = "../Data/Processed/shapespace_charadriiformes_posteriors.MCMCglmm")
+
+```
+
+Figure 6: Original shapespace mcmcmcglmmm diagnosis
+
+```{r, eval = TRUE, echo = FALSE}
+## Ball space
+ballspace_models_names <- model_files[grep("ballspace", model_files)]
+## Models placeholder
+ballspace_models <- list()
+
+## Load the models
+for(i in 1:length(ballspace_models_names)) {
+      load(paste0(models_path, ballspace_models_names[i], collapse = "/"))
+      ballspace_models <- c(ballspace_models, ballspace_20)
+}
+
+## Combine and diagnose
+ballspace_charadriiformes_posteriors <- combine.mini.chains(ballspace_models)
+silent <- diagnose.mini.chains(ballspace_charadriiformes_posteriors)
+save(ballspace_charadriiformes_posteriors, file = "../Data/Processed/ballspace_charadriiformes_posteriors.MCMCglmm")
+```
+
+Figure 7: Balled shapespace mcmcmcglmmm diagnosis
+
+```{r, eval = TRUE, echo = FALSE}
+## Random space
+randospace_models_names <- model_files[grep("randospace", model_files)]
+## Models placeholder
+randospace_models <- list()
+
+## Load the models
+for(i in 1:length(randospace_models_names)) {
+      load(paste0(models_path, randospace_models_names[i], collapse = "/"))
+      randospace_models <- c(randospace_models, unlist(randospace_20, recursive = FALSE))
+}
+
+## Combine and diagnose
+randospace_charadriiformes_posteriors <- combine.mini.chains(randospace_models)
+silent <- diagnose.mini.chains(randospace_charadriiformes_posteriors)
+save(randospace_charadriiformes_posteriors, file = "../Data/Processed/randospace_charadriiformes_posteriors.MCMCglmm")
+```
+
+Figure 8: Randomised shapespace mcmcmcglmmm diagnosis
+
+```{r, eval = FALSE, echo = FALSE}
+## Random space
+BMspace_models_names <- model_files[grep("BMspace", model_files)]
+## Models placeholder
+BMspace_models <- list()
+
+## Load the models
+for(i in 1:length(BMspace_models_names)) {
+      load(paste0(models_path, BMspace_models_names[i], collapse = "/"))
+      BMspace_models <- c(BMspace_models, unlist(BMspace_20, recursive = FALSE))
+}
+
+## Combine and diagnose
+BMspace_charadriiformes_posteriors <- combine.mini.chains(BMspace_models)
+silent <- diagnose.mini.chains(BMspace_charadriiformes_posteriors)
+save(BMspace_charadriiformes_posteriors, file = "../Data/Processed/BMspace_charadriiformes_posteriors.MCMCglmm")
+```
+
+<!-- Figure 9: BMised shapespace mcmcmcglmmm diagnosis -->
+
 # Analysing the results
+
+## Creating the dispRity objects
+
+```{r, eval = TRUE, echo = FALSE}
+## Get the shapespace
+load(file = "../Data/Processed/params_shapespace_charadriiformes.rda")
+shapespace <- param_chains$data
+## Get the posteriors
+load(file = "../Data/Processed/shapespace_charadriiformes_posteriors.MCMCglmm")
+
+## Make the dispRity object
+chara_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+                                     data          = shapespace,
+                                     posteriors    = shapespace_charadriiformes_posteriors,
+                                     group         = MCMCglmm.levels(shapespace_charadriiformes_posteriors)[1:4],
+                                     rename.groups = c(levels(shapespace$level1), "phylogeny"))
+covar.plot(chara_shapespace, ellipses = TRUE, major.axes = TRUE, n = 100, col = c("orange", "blue", "darkgreen", "grey"), legend = TRUE)
+```
+
+```{r, eval = TRUE, echo = FALSE}
+## Get the shapespace
+load(file = "../Data/Processed/params_ballspace_charadriiformes.rda")
+shapespace <- param_chains$data
+## Get the posteriors
+load(file = "../Data/Processed/ballspace_charadriiformes_posteriors.MCMCglmm")
+
+## Make the dispRity object
+chara_ballspace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+                                    data          = shapespace,
+                                    posteriors    = ballspace_charadriiformes_posteriors,
+                                    group         = MCMCglmm.levels(ballspace_charadriiformes_posteriors)[1:4],
+                                    rename.groups = c(levels(shapespace$level1), "phylogeny"))
+covar.plot(chara_ballspace, ellipses = TRUE, major.axes = TRUE, n = 100, col = c("orange", "blue", "darkgreen", "grey"), legend = TRUE)
+```
+
+```{r, eval = TRUE, echo = FALSE}
+## Get the shapespace
+load(file = "../Data/Cluster/shapespace_charadriiformes/randospace_charadriiformes_1.mini.chains")
+shapespace <- randospace_charadriiformes_$data
+## Get the posteriors
+load(file = "../Data/Processed/randospace_charadriiformes_posteriors.MCMCglmm")
+
+## Make the dispRity object
+chara_randospace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+                                     data          = shapespace,
+                                     posteriors    = randospace_charadriiformes_posteriors,
+                                     group         = MCMCglmm.levels(randospace_charadriiformes_posteriors)[1:4],
+                                     rename.groups = c(levels(shapespace$level1), "phylogeny"))
+covar.plot(chara_randospace, ellipses = TRUE, major.axes = TRUE, n = 100, col = c("orange", "blue", "darkgreen", "grey"), legend = TRUE)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+## Get the shapespace
+load(file = "../Data/Cluster/shapespace_charadriiformes/BMspace_charadriiformes_1.mini.chains")
+shapespace <- BMspace_charadriiformes_$data
+## Get the posteriors
+load(file = "../Data/Processed/BMspace_charadriiformes_posteriors.MCMCglmm")
+
+## Make the dispRity object
+chara_BMspace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+                                  data          = shapespace,
+                                  posteriors    = BMspace_charadriiformes_posteriors,
+                                  group         = MCMCglmm.levels(BMspace_charadriiformes_posteriors)[1:4],
+                                  rename.groups = c(levels(shapespace$level1), "phylogeny"))
+covar.plot(chara_BMspace, ellipses = TRUE, major.axes = TRUE, n = 100, col = c("orange", "blue", "darkgreen", "grey"), legend = TRUE)
+```

### 2022-06-23 Analysis/02-mcmcmcglmmm_analysis.Rmd
Up todo

@@
 # for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 1, ".rda"))
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_passerines_shapespace_8D_28levels_50k_chain", 1, ".rda"))
     param_chains[[1]] <- param_chain_passeriformes_50levels_chain1[[1]]
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 2, ".rda"))
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_passerines_shapespace_8D_28levels_50k_chain", 2, ".rda"))
     param_chains[[2]] <- param_chain_passeriformes_50levels_chain1[[1]]
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 3, ".rda"))
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_passerines_shapespace_8D_28levels_50k_chain", 3, ".rda"))
     param_chains[[3]] <- param_chain_passeriformes_50levels_chain1[[1]]

### 2022-06-27 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

--- 2022-06-14 Analysis/02-mcmcmcglmmm_analysis.Rmd
@@
 ## Load the computed models
-param_chains <- list()
-# for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 1, ".rda"))
-    param_chains[[1]] <- param_chain_passeriformes_50levels_chain1[[1]]
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 2, ".rda"))
-    param_chains[[2]] <- param_chain_passeriformes_50levels_chain1[[1]]
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/param_chain_passeriformes_50levels_chain", 3, ".rda"))
-    param_chains[[3]] <- param_chain_passeriformes_50levels_chain1[[1]]
-# }
+param_chains_out <- list()
+for(i in 1:3) {
+    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_passerines_shapespace_8D_28levels_50k_chain", i, ".rda"))
+    param_chains_out[[i]] <- param_chains[[1]]
+}
 ## Extract the parameters from the parameter chains
-est_params <- extract.parameters(param_chains)
+est_params <- extract.parameters(param_chains_out)
 ```

### 2022-06-30 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
     ## Updating the model
-    load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-    model_list <- shapespace_allbirds_lvl_superorder_order
+    load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+    model_list <- shapespace_passeriformes
 }

### 2022-06-30 Analysis/01-Data_preparation.Rmd
removed NAs

@@
 
+## Check what's happening with Neodrepanis_hypoxantha
+
 ## Loading the Passeriformes clades
@@
 for(clade in 1:length(clade_info_passerines$passerine_big_clades)) {
+  ## Get the clade name
   clade_name <- names(clade_info_passerines$passerine_big_clades)[clade]
+  ## Attribute the species/clade
   shapespace_passeriformes[clade_info_passerines$passerine_big_clades[[clade]], "level1"] <- clade_name
+  ## Remove species not present originally in the shapespace
+  if(length(nas <- which(apply(shapespace_passeriformes[, 1:8], 1, function(row) any(is.na(row))))) > 0) {
+      shapespace_passeriformes <- shapespace_passeriformes[-nas, ]
+  }
 }
@@
 for(clade in 1:length(clade_info_passerines$passerine_small_clades)) {
+  ## Get the clade name
   clade_name <- names(clade_info_passerines$passerine_small_clades)[clade]
+  ## Attribute the species/clade
   shapespace_passeriformes[clade_info_passerines$passerine_small_clades[[clade]], "level2"] <- clade_name
+  ## Remove species not present originally in the shapespace
+  if(length(nas <- which(apply(shapespace_passeriformes[, 1:8], 1, function(row) any(is.na(row))))) > 0) {
+      shapespace_passeriformes <- shapespace_passeriformes[-nas, ]
+  }
 }

### 2022-06-30 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Figure 3 sorted
passeriformes preliminary analyses

--- 2022-05-26 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure3_ellipses.pdf", height = box_size*8, width = box_size*10)
 
@@
 ## Load the data
-load(file = "../Data/Processed/shapsepace_passeriformes_suborder_family.rda")
+load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
 ## Select the data
-shapespace <- shapsepace_passeriformes_suborder_family[[1]]$space
-tree <- shapsepace_passeriformes_suborder_family[[1]]$consensus_tree
+shapespace <- shapespace_passeriformes_two_levels[[1]]$space
+tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
 ## Clean garbage
-rm(shapsepace_passeriformes_suborder_family)
+rm(shapespace_passeriformes_two_levels)
 
@@
 phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
-subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
-famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))
+suborder_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
+family_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))
 
 ## Creating the subsets
-suborder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
       posteriors    = shapespace_passeriformes,
-      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(suborder_lvl, phylo_lvl)],
       rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
 
-family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
       posteriors    = shapespace_passeriformes,
-      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
+      group         = MCMCglmm.levels(shapespace_passeriformes)[c(family_lvl, phylo_lvl)],
       rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure1_passeriformes.pdf", height = box_size*7, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure3_passeriformes.pdf", height = box_size*7, width = box_size*10)
 
@@
         matrix(c(36:45), ncol = 10, nrow = 1), # line 5
-        matrix(c(46:55), ncol = 10, nrow = 1) # line 6
+        matrix(c(46:47, rep(-13,8)), ncol = 10, nrow = 1) # line 6
     )
 lay_mat_sub <- rbind(
-        matrix(c(1,1, 2:5, rep(0, 4)), ncol = 10, nrow = 1) # line 7
+        matrix(c(1,1, 2:9, c(1,1), 10:13, rep(0, 4)), ncol = 10, nrow = 2, byrow = TRUE) # line 7
     )
@@
 widths_vals <- rep(c(1, 1/3), 5)
-layout_matrix <- layout(rbind(lay_mat_sub, (lay_mat_family+5)), widths = widths_vals, heights = heights_vals, respect = TRUE)
+layout_matrix <- layout(rbind(lay_mat_sub, (lay_mat_family+13)), widths = widths_vals, heights = heights_vals, respect = TRUE)
+# layout.show(layout_matrix)
 
@@
 tip_order <- c(
-    "Tyranni" = 1,
-    "Passeri" = 2)
+    "Suboscines" = 1,
+    "Meliphagoidea" = 2,
+    "Corvides" = 3,
+    "Sylviida" = 4,
+    "Muscicapida" = 5,
+    "Passerida" = 6)
 
 ## Plotting the tree and getting the tip orders
-tip_colours_suborders <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
-save(tip_colours_suborders, file = "../Data/Processed/tip_colours_suborders.rda")
+tip_colours_passeriformes_level1 <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
+save(tip_colours_passeriformes_level1, file = "../Data/Processed/tip_colours_passeriformes_level1.rda")
 
@@
     ## Plot each pannel
-    wrap.plot.ellipses(i, tip_colours_suborders, suborder_shapespace, results$suborder_results_between)
+    wrap.plot.ellipses(i, tip_colours_passeriformes_level1, suborder_shapespace, results$suborder_results_between)
 }
@@
 tip_order <- c(
-    "Pittidae" = 2,
-    "Eurylaimidae" = 1,
-    "Tyrannidae" = 5,
-    "Pipridae" = 3,
-    "Cotingidae" = 4,
-    "Thamnophilidae" = 6,
-    "Furnaridae" = 9,
-    "Rhinocryptidae" = 8,
-    "Grallariidae" = 7,
-    "Ptilonorhynchidae" = 10,
-    "Maluridae" = 11,
-    "Meliphagidae" = 13,
-    "Acanthizidae" = 12,
-    "Vireonidae" = 20,
-    "Pachycephalidae" = 18,
-    "Oriolidae" = 19,
-    "Laniidae" = 25,
-    "Corvidae" = 24,
-    "Paradisaeidae" = 23,
-    "Monarchidae" = 22,
-    "Dicruridae" = 21,
-    "Campephagidae" = 14,
-    "Malaconotidae" = 16,
-    "Vangidae" = 15,
-    "Platysteiridae" = 17,
-    "Troglodytidae" = 27,
-    "Sittidae" = 26)
+    "Aegithaloidea"= 12,
+    "Bombycilloidea"= 17,
+    "Cisticolidae"= 14,
+    "Corvoidea"= 7,
+    "Emberizoidea"= 23,
+    "Eurylaimides"= 1,
+    "Fringillidae"= 10,
+    "Furnariida"= 3,
+    "Hirundinidae"= 11,
+    "Locustelloidea"= 15,
+    "Malaconotoidea"= 5,
+    "Meliphagoidea"= 4,
+    "Motacillidae"= 22,
+    "Muscicapoidea"= 18,
+    "Nectariniidae"= 19,
+    "Orioloidea"= 6,
+    "Paridae"= 9,
+    "Passeridae"= 21,
+    "Petroicidae"= 8,
+    "PloceidaeEstrildidae"= 20,
+    "Pycnonotidae"= 13,
+    "Sylvioidea"= 16,
+    "Tyrannida" = 2)
 
 ## Plotting the tree and getting the tip orders
-tip_colours_families <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
-save(tip_colours_families, file = "../Data/Processed/tip_colours_families.rda")
+tip_colours_passeriformes_level2 <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+save(tip_colours_passeriformes_level2, file = "../Data/Processed/tip_colours_passeriformes_level2.rda")
 
@@
     ## Plot each pannel
-    wrap.plot.ellipses(i, tip_colours_families, family_shapespace, results$family_results_between)
+    wrap.plot.ellipses(i, tip_colours_passeriformes_level2, family_shapespace, results$family_results_between)
 }

### 2022-06-30 Analysis/06-Figure2_with_correlations.Rmd
Update 06-Figure2_with_correlations.Rmd
passeriformes preliminary analyses

@@
     par(mar = c(2,2.5,2,1.5))
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = FALSE)
+    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
 }
@@
 ## --------
-
-## Getting the subsets for the colouring
-col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
 
@@
 ## Creating the subsets
-suborder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 
-family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 ## Set the pdf parameters
-pdf("../Manuscript/Figures/Figure2_correlations.pdf", height = 12, width = 8)
+pdf("../Manuscript/Figures/Figure2_correlations_passeriformes.pdf", height = 12, width = 6)
 
 ## Set up the plotting layout
-layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
-layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
+layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,1))
+layout_matrix <- rbind(rbind(layout_matrix, matrix(1:6+1, 2, 3, byrow = TRUE)), matrix(7:12+1, 2, 3,  byrow = TRUE))
 plot_layout <- layout(layout_matrix)
@@
 par(mar = c(5, 4, 4, 2) + 0.1)
-plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)
+plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)
 
@@
 ## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright",
-                 "topleft", "topright", "topleft", "topright")
+legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
 ## Create the vector for counting the legend numbers
@@
 ## Plotting each individual correlation plots
-for(i in 1:8) {
+for(i in 1:6) {
     par(mar = c(2,2.5,2,1.5))
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = FALSE)
+    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
 }
@@
 
-## Getting the subsets for the colouring
-col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
+## Set the vector of legend positions
+legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
 
-## Set the vector of legend positions
-legends_pos <- c("bottomright", "topright", "topleft", "topleft",
-                 "topleft", "topleft", "topright", "bottomright")
 ## Resetting the legend_counter
@@
 ## Plotting each individual correlation plots
-for(i in 1:8) {
+for(i in 1:6) {
     par(mar = c(2,2.5,2,1.5))
-    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
+    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
 }
@@
 ```
+
+Legend numbers are:
+1 - others (groups with < 15 species)
+2 - Corvoidea
+3 - Malaconotoidea
+4 - Orioloidea
+5 - Meliphagoidea
+6 - Bombycilloidea
+7 - Muscicapoidea
+8 - Emberizoidea
+9 - Motacillidae
+10 - Nectariniidae 
+11 - Passeridae
+12 - PloceidaeEstrildidae
+13 - Eurylaimides
+14 - Furnariida
+15 - Tyrannida 
+16 - Aegithaloidea
+17 - Cisticolidae
+18 - Fringillidae
+19 - Hirundinidae
+20 - Locustelloidea
+21 - Paridae
+22 - Pycnonotidae
+23 - Sylvioidea

### 2022-07-13 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 
+## Overlapping ellipses plot
+
+```{r}
+load("../Data/Processed/tip_colours_super_orders.rda")
+load("../Data/Processed/tip_colours_orders.rda")
+
+## Matching colours (fast utility)
+match.col <- function(data, col) {
+    return(c(col[match(names(size.subsets(data)), names(col))[-length(size.subsets(data))]], "grey"))
+}
+
+pdf(file = "../Manuscript/Figures/Figure_supplementary_aligned_ellipses.pdf", height = 6, width = 12)
+par(mfrow = c(1,2))
+## Plotting the aligned and scale ellipses for the superorders
+covar.plot(superorder_shapespace, ellipses = mean,
+           col = match.col(superorder_shapespace, tip_colours_super_orders),
+           legend = TRUE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
+           main = "Scaled and centred average super order VCV")
+
+## Plotting the aligned and scale ellipses for the superorders
+covar.plot(order_shapespace, ellipses = mean,
+           col = match.col(order_shapespace, tip_colours_orders),
+           legend = FALSE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(0.5, length(tip_colours_orders)), 3),
+           main = "Scaled and centred average order VCV")
+dev.off()
+```
+
 # Passeriformes shapespace analyses

### 2022-07-14 Analysis/03-Elaboration_innovation_analysis.Rmd
Removed old analyses
FIND BACK OLD STUFF (inc. top 5% species) IN THIS COMMIT

@@
 
+## Saving plotting data (median values)
+
 ```{r, echo = FALSE, eval = FALSE}
@@
 
-```
-
-### Projections of groups on the phylogeny:
-
-```{r "starting with boxplots", echo = FALSE, fig.height = 8, fig.width = 8}
-## Plot the results
-wrap.plot <- function(results, col, main, names, order = TRUE, xlim = NULL) {
-
-    ## Ranking the results
-    if(order) {
-        ## Get the ranking
-        orders <- lapply(results, get.disparity, concatenate = FALSE)
-        ## Get the mean values
-        orders_pos <- orders$position
-        orders_dis <- orders$distance
-
-        ## Get the mean for each and combine into a table
-        cent.tend <- mean
-        means <- cbind(unlist(lapply(orders_pos, cent.tend)), unlist(lapply(orders_dis, cent.tend)))
-
-        ## Sum them and rank them
-        sum_means <- apply(means, 1, sum)
-        rank <- order(sum_means, decreasing = FALSE)
-    } else {
-        rank <- 1:length(results[[1]]$disparity)
-    }
-
-    ## Adding one plot
-    one.plot <- function(one_result, xlab, col, zero = TRUE, main, centre, absolute, rank = NULL, xlim = NULL, ...) {
-        values <- lapply(get.disparity(one_result, concatenate = FALSE), c)
-        if(!is.null(rank)) {
-            values <- values[rank]
-        }
-        if(!is.null(xlim)) {
-            xlim <- xlim
-        } else {
-            xlim <- range(values)
-        }
-
-        boxplot(values, horizontal = TRUE, yaxt = "n", col = col, xlab = xlab, main = main, ylim = xlim, ...)
-        if(zero) {
-            abline(v = 0, col = "black", lty = 3, lwd = 1)
-        }
-    }
-
-    ## How many sub plots?
-    n <- length(results)
-
-    ## Recycle labels?
-    if(missing(names)) {
-        names_list <-names(results)
-    } else {
-        names_list <- names
-    }
-
-    ## Which will be the centre plot?
-    centre <- ceiling(n/2)
-    ## Plot everything
-    op <- par(mfrow = c(1, n+1))
-    ## Plot the first panel
-    par(bty = "n", mar = c(5, 14, 4, 0) + 0.1)
-    empty <- matrix(0, ncol = length(results[[1]]$disparity))
-    colnames(empty) <- names(results[[1]]$disparity)[rank]
-    boxplot(empty, horizontal = TRUE, las = 2, col = "white", ylim = c(0,0), xaxt = "n", border = "white")
-    for(i in 1:n) {
-        ## Is position?
-        is_position <- names(results)[[i]] == "position"
-        par(mar = c(5, 0, 4, 0) + 0.1, bty = "o")
-        one.plot(results[[i]], col = col, centre = is_position && centre.position, main = ifelse(i == centre, main, ""), xlab = names_list[i], zero = FALSE, absolute = absolute, rank = rank, xlim = xlim)
-    }
-    par(op)
-}
-wrap.plot(results$super_results_between,
-              col = colour.palette(n.subsets(superorder_shapespace)-1),
-              main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV",
-              names = c("elaboration", "innovation"))
-```
-
-Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes.
-For each panel, the highest the score the more the clade is elaborating, exploring, exploring perpendicular to the main axes for the position, distance and degree respectively.
-
-<details>
-  <summary>[Click to expand] Same but zooming on the (0,5) interval (to get less of the flamingo effect)
-</summary>
-
-```{r, echo = FALSE, fig.height = 8, fig.width = 8}
-wrap.plot(results$super_results_between,
-          xlim = c(0,6),
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[groups on phylogeny] superorders VCV\nonto the phylogenetic VCV (ZOOM)",
-          names = c("elaboration", "innovation"))
-```
-</details>
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-wrap.plot(results$order_results_between,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-<details>
-  <summary>[Click to expand] Again, same but with some zoom (0,15)
-</summary>
-
-```{r, echo = FALSE, fig.height = 16, fig.width = 8}
-wrap.plot(results$order_results_between,
-          xlim = c(0, 15),
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[groups on phylogeny] orders VCV\nonto the phylogenetic VCV (ZOOM)",
-          names = c("elaboration", "innovation"))
-```
-</details>
-
-### Projections of the species on the phylogenetic VCV
-
-Similarly to the analyses described above, we ran the analyses for each species.
-Here, instead of projecting the major axes of the groups onto the global phylogenetic major axis we projected each species onto the phylogenetic major axis (for each posterior distribution).
-
-```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
-## Plot the results
-wrap.plot(results$super_results_phylo,
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[species on phylogeny] superorders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$order_results_phylo,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[species on phylogeny] orders species\nonto the phylogenetic VCV",
-          names = c("elaboration", "innovation"))
-```
-
-### Projections of the species on their own group's VCV
-
-Finally, we ran the same analyses but instead of the projection of the species onto the global phylogenetic major axes, we projected them onto their respective group's phylogenetic major axes.
-
-```{r, fig.height = 8, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$super_results_within,
-          col = colour.palette(n.subsets(superorder_shapespace)-1),
-          main = "[species on group] superorders species\nonto their group's VCV",
-          names = c("elaboration", "innovation"))
-```
-
-Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).
-
-```{r, fig.height = 16, fig.width = 8, echo = FALSE}
-## Plot the results
-wrap.plot(results$order_results_within,
-          col = colour.palette(n.subsets(order_shapespace)-1),
-          main = "[species on group] orders species\nonto their group's VCV",
-          names = c("elaboration", "innovation"))
-```
-
-So that's the story so far...
-
-# Different scores distributions per species
-
-```{r, eval = TRUE, echo = FALSE, eval = TRUE}
-library(hdrcde)
-
-#@param res the results list
-#@param type which type (position, distance or angle)
-#@param cent.tend which central tendency fun (default is median)
-#@param use.hdr whether to use kernel density estimate for the outputs (default is TRUE)
-#@param prob which quantiles to use (default is c(50, 95, 99))
-#@param exclude.phylo remove the phylogeny subset (default = FALSE)
-#@param add.relative whether to add the relative results (as rel.pos and rel.dis) or not
-#@param ... plot options
-get.distributions <- function(res, type, cent.tend = median, use.hdr = TRUE, prob = c(50, 95, 99), exclude.phylo = FALSE, add.relative = FALSE, ...) {
-
-    ## Remove the phylo one (if needed)
-    if(exclude.phylo) {
-        select <- names(size.subsets(res[[type]]))
-        select <- select[-which(select %in% "phylogeny")]
-        res[[type]] <- get.subsets(res[[type]], subsets = select)
-    }
-    ## Extracting all the results
-    disparity <- lapply(res[[type]]$disparity, function(X, type) return(X$elements))
-
-    ## Measure the central tendencies
-    centrals <- lapply(disparity, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-    ## Number of groups
-    n_groups <- length(centrals)
-
-    ## Calculate and plot the hdrcde
-    op_tmp <- par(mfrow = c(ceiling(sqrt(n_groups)), round(sqrt(n_groups))))
-    densities <- list()
-    for(i in 1:length(centrals)) {
-        if(use.hdr) {
-            densities[[i]] <- hdr.den(centrals[[i]], prob = prob, main = names(centrals)[i])#, ...)
-        } else {
-            hist(centrals[[i]], main = names(centrals)[i], ...)
-        }
-    }
-    par(op_tmp)
-
-    ## Find the extreme species
-    get.extremes.hdr <- function(density, central) {
-        ## Get the minimums and maximums
-        mins <- which(central <= density$hdr[1,1])
-        maxs <- which(central >= density$hdr[1, ncol(density$hdr)])
-        ## Get their values
-        return(c(central[names(mins)], central[names(maxs)]))
-    }
-    CI.converter <- function(CI) {
-        sort(c(50-CI/2, 50+CI/2)/100)
-    }
-    get.extremes.quantile <- function(central, prob) {
-        ## Get the quantiles
-        quants <- quantile(central, probs = CI.converter(max(prob)))
-        ## Get the minimums and maximums
-        mins <- which(central <= quants[1])
-        maxs <- which(central >= quants[2])
-        ## Get their values
-        return(c(central[names(mins)], central[names(maxs)]))
-    }
-
-    if(use.hdr) {
-        output <- mapply(get.extremes.hdr, densities, centrals, SIMPLIFY = FALSE)
-        names(output) <- names(centrals)
-    } else {
-        output <- lapply(centrals, get.extremes.quantile, prob = prob)
-    }
-
-    ## Return the as a table
-    for(i in 1:length(output)) {
-        if(length(output[[i]]) == 0) {
-            output[[i]] <- data.frame("level" = names(output)[i], "species" = NA, "score" = NA)
-        } else {
-            output[[i]] <- data.frame("level" = rep(names(output)[i], length(output[[i]])),
-                                      "species" = names(output[[i]]),
-                                      "score"   = unname(output[[i]]))
-        }
-    }
-
-    ## Add the relative results
-    if(add.relative) {
-        make.relative <- function(X) {
-            cbind(X, "rel" = ifelse(is.na(X$score), NA, X$score/max(X$score)))
-        }
-        output <- lapply(output, make.relative)
-    }
-    rm(disparity)
-    return(output)
-}
-combine.results <- function(res1, res2, sort.results = TRUE) {
-    ## Combine the results
-    res_out <- rbind(res1[,1:2], res2[, 1:2])
-    res_out$elaboration <- c(res1[, 3], rep(NA, nrow(res2)))
-    if("rel" %in% colnames(res1)) {
-        res_out$rel.elab <- c(res1[, 4], rep(NA, nrow(res2)))
-    }
-    res_out$innovation <- c(rep(NA, nrow(res1)), res2[, 3])
-    if("rel" %in% colnames(res1)) {
-        res_out$rel.inno <- c(rep(NA, nrow(res1)), res2[, 4])
-    }
-
-    # Remove NA species
-    if(any(is_na <- is.na(res_out$species))) {
-        if(all(is_na)) {
-            return(res_out)
-        } else {
-            res_out <- res_out[!is_na, , drop = FALSE]
-        }
-    }
-
-    ## Find any species in commons
-    commons <- intersect(res1$species, res2$species)
-    if(length(commons) > 0) {
-        if(any(is.na(commons))) {
-            if(is.na(res1$species)) {
-                if(is.na(res2$species)) {
-                    ## Both are NA (keep just the first one)
-                    res_out <- res_out[-1, , drop = FALSE]
-                } else {
-                    ## Remove res1
-                    res_out <- res_out[-1, , drop = FALSE]
-                }
-            } else {
-                ## Remove res2
-                res_out <- res_out[1:length(res1$species), , drop = FALSE]
-            }
-        } else {
-            ## Recursively remove the ones in common
-            while(length(commons) > 0) {
-                duplicates <- which(res_out$species == commons[1])
-                columns <- if("rel" %in% colnames(res1)){c(5,6)} else {4}
-                res_out[duplicates[1], columns] <- res_out[duplicates[2], columns]
-                res_out <- res_out[-duplicates[2], ]
-                commons <- commons[-1]
-            }
-        }
-    }
-
-    if(sort.results) {
-        sums <- apply(res_out[, -c(1,2)], 1, sum, na.rm = TRUE)
-        rank <- match(sort(sums, decreasing = TRUE), sums)
-    } else {
-        rank <- 1:nrow(res_out)
-    }
-
-    return(res_out[rank,])
-}
-```
-
-Furthermore, we can look at things species per species to find which species species have the most extreme values (in general, or within their groups).
-
-### Extreme species per super order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_phylo <- get.distributions(results$super_results_phylo, type = "position")
-```
-Figure: elaborations distributions per super orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_phylo <- get.distributions(results$super_results_phylo, type = "distance")
-```
-Figure: innovations distributions per super orders
-
-```{r, "printing results in text", echo = FALSE, eval = TRUE}
-## Combining the results
-super_phylo_table <- do.call(rbind, mapply(combine.results, super_positions_phylo, super_distances_phylo, SIMPLIFY = FALSE))
-## Get the top 10%
-get.top <- function(data, what, top = 0.9, relative = FALSE) {
-    if(any(data[, "level"] == "phylogeny")) {
-        data_top <- data[which(data[, "level"] != "phylogeny"), ]
-    } else {
-        data_top <- data
-    }
-
-    ## Get the relative column
-    if(relative) {
-        what <- switch(what,
-                       "elaboration" = "rel.elab",
-                       "innovation"  = "rel.inno")
-    }
-    ## Get the top 10(ish) species
-    data_top_10 <- data_top[which(data_top[, what] >= quantile(data_top[, what], prob = top, na.rm = TRUE)), ]
-    return(data_top_10[order(data_top_10[, what], decreasing = TRUE), ])
-}
-top_elabo <- get.top(super_phylo_table, what = "elaboration")
-top_innov <- get.top(super_phylo_table, what = "innovation")
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by superorder:
-
-The 10% top elaborators by superorder are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
-
-The 10% top innovators by superorder are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto the phylogenetic VCV.
-</summary>
-
-```{r, echo = FALSE}
-library(knitr)
-rownames(super_phylo_table) <- NULL
-kable(super_phylo_table, digits = 3)
-```
-
-</details>
-
-### Extreme species per order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_phylo <- get.distributions(results$order_results_phylo, type = "position", exclude.phylo = TRUE)
-```
-Figure: elaborations distributions per order
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_phylo <- get.distributions(results$order_results_phylo, type = "distance", exclude.phylo = TRUE)
-```
-Figure: innovations distributions per order
-
-```{r, echo = FALSE}
-order_phylo_table <- do.call(rbind, mapply(combine.results, order_positions_phylo, order_distances_phylo, SIMPLIFY = FALSE))
-top_elabo <- get.top(order_phylo_table, what = "elaboration")
-top_innov <- get.top(order_phylo_table, what = "innovation")
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by order:
-The 10% top ellaborators by order are: `r paste(paste(top_elabo[, "species"], round(top_elabo[, "elaboration"], 3), sep = ": "), collapse = ", ")`.
-The 10% top innovators by order are: `r paste(paste(top_innov[, "species"], round(top_innov[, "innovation"], 3), sep = ": "), collapse = ", ")`.
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(order_phylo_table) <- NULL
-kable(order_phylo_table, digits = 3)
-```
-</details>
-
-### Extreme species per super order (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_positions_within <- get.distributions(results$super_results_within, type = "position", exclude.phylo = TRUE,  add.relative = TRUE)
-```
-Figure: elaborations distributions within super orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-super_distances_within <- get.distributions(results$super_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: innovations distributions within super orders
-
-```{r, echo = FALSE}
-super_phylo_table_within <- do.call(rbind, mapply(combine.results, super_positions_within, super_distances_within, SIMPLIFY = FALSE))
-top_elabo <- get.top(super_phylo_table_within, what = "elaboration", relative = TRUE)
-top_innov <- get.top(super_phylo_table_within, what = "innovation", relative = TRUE)
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by superorder relative to each superorder (i.e. the best one(s) in each order):
-```{r, echo = FALSE}
-combined <- rbind(top_elabo, top_innov)
-rownames(combined) <- NULL  
-kable(combined, digits = 3)
-```
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(super_phylo_table_within) <- NULL
-kable(super_phylo_table_within, digits = 3)
-```
-
-</details>
-
-### Extreme species per order (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_positions_within <- get.distributions(results$order_results_within, type = "position", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: elaborations distributions within orders
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 12, fig.width = 12, eval = TRUE}
-order_distances_within <- get.distributions(results$order_results_within, type = "distance", exclude.phylo = TRUE, add.relative = TRUE)
-```
-Figure: innovations distributions within orders
-
-```{r, echo = FALSE}
-order_phylo_table_within <- do.call(rbind, mapply(combine.results, order_positions_within, order_distances_within, SIMPLIFY = FALSE))
-top_elabo <- get.top(order_phylo_table_within, what = "elaboration", relative = TRUE)
-top_innov <- get.top(order_phylo_table_within, what = "innovation", relative = TRUE)
-```
-
-> And here are the top 10% of the top 1% elaborators and innovators by order relative to each order (i.e. the best one(s) in each order):
-```{r, echo = FALSE}
-combined <- rbind(top_elabo, top_innov)
-rownames(combined) <- NULL  
-kable(combined, digits = 3)
-```
-
-<details>
-  <summary>[Click to expand] Table of species extreme values projected onto their own group's VCV.
-</summary>
-
-```{r, echo = FALSE}
-## Combining the results
-rownames(order_phylo_table_within) <- NULL
-kable(order_phylo_table_within, digits = 3)
-```
-</details>
-
-## Correlations between elaboration and innovation?
-
-```{r, echo = FALSE}
-## Calculating all the correlations
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE) {
-
-    ## Remove the phylo one (if needed)
-    if(exclude.phylo) {
-        select_pos <- names(size.subsets(res$position))
-        select_dis <- names(size.subsets(res$distance))
-        select_pos <- select_pos[-which(select_pos %in% "phylogeny")]
-        select_dis <- select_dis[-which(select_dis %in% "phylogeny")]
-        res$position <- get.subsets(res$position, subsets = select_pos)
-        res$distance <- get.subsets(res$distance, subsets = select_dis)
-    }
-
-    ## Extracting all the results
-    disparity_pos <- lapply(res$position$disparity, function(X, type) return(X$elements))
-    disparity_dis <- lapply(res$distance$disparity, function(X, type) return(X$elements))
-
-    ## Measure the central tendencies
-    centrals_pos <- lapply(disparity_pos, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-    centrals_dis <- lapply(disparity_dis, function(X, fun) apply(X, 1, fun), fun = cent.tend)
-
-    ## Number of groups
-    n_groups <- length(centrals_pos)
-
-    ## Prepare the results
-    prep.cor.results <- function(pos, dis, name) {
-        ## Run the correlation test
-        cor_test <- cor.test(pos, dis)
-        ## Prepare the plot title
-        plot_main <- paste0(name, ": ", unname(round(cor_test$estimate, 3)), " (p=", unname(round(cor_test$p.value, 3)), ")")
-        ## Return the plotting parameters
-        return(list(x = pos, y = dis, main = plot_main))
-    }
-
-    return(mapply(prep.cor.results, centrals_pos, centrals_dis, as.list(names(centrals_pos)), SIMPLIFY = FALSE))
-}
-
-## Plotting one correlation 
-plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, legend.pos = "topleft", pt.cex = 0.5, legend.cex = 0.5, tree, full.tree = FALSE, ...) {
-    dots <- list(...)
-
-    if(missing(col.sub)) {
-        col.sub <- NULL
-    }
-
-    ## Subbing the colours
-    if(!is.null(col.sub)) {
-        ## Reset the levels
-        col_sub <- as.factor(as.character(col.sub))
-        ## Make the empty level grey
-        if(any(levels(col_sub) == "")) {
-            col_avail <- c("grey", col.fun(length(levels(col_sub))-1))
-        } else {
-            col_avail <- col.fun(length(levels(col_sub)))
-        }
-        ## Create the colour vector
-        col_vector <- col_avail[as.numeric(col_sub)]
-
-    } else {
-        if(is.null(dots$col)) {
-            col_vector <- "black"
-        } else {
-            col_vector <- dots$col
-        }
-    }
-
-    if(!missing(tree)) {
-        op <- par(mfrow = c(1,2))
-        #c(bottom, left, top, right); c(5, 4, 4, 2) + 0.1
-        par(mar = c(5, 4, 4, 0)+0.1)   
-    }
-
-    ## Plotting the results
-    plot(cor.results$x, cor.results$y, pch = 19, xlab = "elaboration", ylab = "innovation", main = cor.results$main, cex = pt.cex, col = col_vector)
-
-    if(!is.null(col.sub)) {
-        ## Adding the legend
-        legend(x = legend.pos, legend = levels(col_sub), pch = 19, cex = legend.cex, col = col_avail)
-    }
-
-    if(!missing(tree)) {
-        par(mar = c(5, 0, 4, 2) + 0.1)
-
-        ## Recycle the colour vector
-        named_col_vector <- col_vector
-        names(named_col_vector) <- names(cor.results$x)
-
-        ## Reduce the tree
-        if(!full.tree) {
-            tree <- drop.tip(tree, tip = tree$tip.label[which(!(tree$tip.label %in% names(named_col_vector)))])
-        }
-
-        ## Matching edges and colours
-        match.edge.colours <- function(tree, named_col_vector, background = "grey") {
-            ## Get the background colour
-            edge_cols <- rep(background, Nedge(tree))
-            ## Get the clades
-            clades <- unique(named_col_vector)
-            ## Ignore the greys
-            if(any(which_grey <- clades == "grey")) {
-                clades <- clades[!which_grey]
-            }
-            ## Find the edges for each clade
-            for(clade in 1:length(clades)) {
-                ## Get the tips for the clade
-                tips <- which(named_col_vector == clades[clade])
-                ## Get the MRCA node
-                mrca_node <- getMRCA(tree, tips)
-                # stop_edge <- which(tree$edge[, 2] %in% mrca_node)
-
-                ## Handle non-monophyletic groups
-                clade_tree <- extract.clade(tree, node = mrca_node)
-                if(Ntip(clade_tree) > length(tips)) {
-                    clade_tree <- drop.tip(clade_tree, tip = clade_tree$tip.label[!(clade_tree$tip.label %in% names(tips))])
-                }
-                n_edges_target <- Nedge(clade_tree)
-
-                ## Get the tips ID
-                tip_ids <- which(tree$tip.label %in% names(tips))
-                ## Get the vector of edges
-                edges <- integer()
-                ## Get the first nodes (and edges)
-                new_edges <- which(tree$edge[, 2] %in% tip_ids)
-                edges <- unique(c(edges, new_edges))
-                nodes <- tree$edge[new_edges, 1]
-
-                # plot(tree, show.tip.label = FALSE)
-                # test.col <- c(replicate(10, c("red", "blue", "orange", "green", "pink", "yellow")))
-                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                ## Recursively loop through each node up until the last edge
-                # while(!(stop_edge %in% edges)) {
-                while(length(edges) <= n_edges_target) {
-                    ## Get the node edges
-                    new_edges <- which(tree$edge[, 2] %in% nodes)
-
-                    # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                    edges <- unique(c(edges, new_edges))
-                    ## Update the nodes
-                    nodes <- tree$edge[new_edges, 1]
-                }
-
-                # ## Do the last recursion
-                # new_edges <- which(tree$edge[, 2] %in% nodes)
-                # edges <- unique(c(edges, new_edges))
-
-                # edgelabels(".", cex = 0.5, new_edges, bg = test.col[1]) ; test.col <- test.col[-1]
-
-                ## Colour these edges
-                edge_cols[edges] <- clades[clade]
-            }
-            return(edge_cols)
-        }
-        ## Match the colour to the correct edge
-        edge_cols <- match.edge.colours(tree, named_col_vector)
-
-        ## Add the tree
-        plot(tree, show.tip.label = FALSE, edge.color = edge_cols)
-    }
-
-    return(invisible())
-}
-```
-
-### Correlations per super order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
-## Getting the subsets for the colouring
-col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
-## Calculate the correlations
-super_phylo_cor <- get.correlations(results$super_results_phylo)
-## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
-for(i in 1:length(super_phylo_cor)) {
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
-}
-```
-
-<details>
-  <summary>[Click to expand] Finding corner cases groups.
-</summary>
-
-So here's how you remove exploration and elaboration: by centering these three species on the centre of the phylo axis:
-
-Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
-# colour.palette <- rainbow
-posteriors <- shapespace_allbirds_lvl_superorder_order
-
-# plot(superorder_shapespace$matrix[[1]][, 1:2], pch = 19, cex = 0.5)
-# identify(superorder_shapespace$matrix[[1]][, 1:2])
-# shapespace[c(268, 286, 80), c(9, 10)]
-
-## Moving these three species around: Mitu_tuberosum and Penelope_argyrotis (normal ish) and Anas_falcata
-fixed_points <- c(268, 286, 408)
-move_around <- c(8749, 8750, 8751)
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Get the major axis
-phylo_axis <- axis.covar(superorder_shapespace, sample = median)$phylogeny[[1]]
-## Point with no elaboration/exploration
-no_nothing <- 0
-
-## Example with no exploration/elaboration
-skewed_space[move_around, c(1:8)] <- no_nothing
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny"))
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
-           main = "elements without\nelaboration innovation")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-So easy right? We just need to change them along the PC axis to move them around in space.
-
-No, it's not fucking easy because it's a 8 dimensional problem...
-Even though the two first axes cumulate 86.27% of the variance, fixing the three elements on the two elements and centering them on the 6 other dimensions (13.73% of the variance) results in BIG changes:
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Example with no exploration/elaboration
-skewed_space[move_around, c(3:8)] <- 0
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_skewed_space)-1), "grey"),
-           main = "Centring only on dimensions 3:8")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-For example, the blue element, Anas_falcata is moved to no elaboration/exploration at all if we do that ignoring data past the second axis.
-There's a duck axis lurking in the depth... (spoiler, it's the fourth dimension).
-
-Let's flatten the problem in 2D for simplification:
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-
-## 2Ding the space
-dimensions <- c(1,2)
-superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
-superorder_skewed_space$call$dimensions <- dimensions
-fun2d <- function(covar) {
-    covar$VCV <- covar$VCV[dimensions, dimensions]
-    covar$loc <- covar$loc[dimensions]
-    return(covar)
-}
-superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Flattening the results")
-## Add the phylogenetic axis
-lines(phylo_axis, lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis, lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-abline(a = 0, b = 7, col = "grey")
-abline(a = 0, b = 0.14, col = "grey")
-```
-
-So here there seems to be some kind of very shy limit (arbitrarily found as a slope of 0.14 and 7).
-Let's see if it's something birdy or not by modifying these elements position in trait space to have elaboration of 0 and innovation of 1 and the opposite.
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Reinitialising the space
-skewed_space <- rbind(shapespace, shapespace[fixed_points, ])
-
-## Move around the first species to have elaboration = 0 and innovation = 0
-skewed_space[move_around[1], c(1:2)] <- 0
-## Move around the second species to have elaboration = 0 and innovation = 1
-skewed_space[move_around[2], c(1:2)] <- c(0, 0.4)
-## Move around the third species to have elaboration = 1 and innovation = 0
-skewed_space[move_around[3], c(1:2)] <- c(0.4, 0)
-
-## Recreate the space
-superorder_skewed_space <- MCMCglmm.subsets(n = 10, data = skewed_space, posteriors = posteriors, group = MCMCglmm.levels(posteriors)[c(super_lvl, phylo_lvl)], rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #
-
-## 2Ding the space
-dimensions <- c(1,2)
-superorder_skewed_space$matrix[[1]] <- superorder_skewed_space$matrix[[1]][,dimensions]
-superorder_skewed_space$call$dimensions <- dimensions
-fun2d <- function(covar) {
-    covar$VCV <- covar$VCV[dimensions, dimensions]
-    covar$loc <- covar$loc[dimensions]
-    return(covar)
-}
-superorder_skewed_space$covar <- lapply(superorder_skewed_space$covar, lapply, fun2d)
-```
-
-```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 16, echo = FALSE}
-## Visualise all that
-par(mfrow = c(1,2))
-## Plot the space
-covar.plot(superorder_skewed_space,
-           cex = 0.5,
-           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
-           main = "Flattening the results")
-## Add the phylogenetic axis
-lines(phylo_axis[,1:2], lty = 1, lwd = 4, col = "grey")
-lines(phylo_axis[,1:2], lty = 1, lwd = 2.5, col = "black")
-## Adding the points
-points(skewed_space[fixed_points, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(skewed_space[move_around, 1:2], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-
-## Recalculate the disparity
-results_skewed <- dispRity.covar.projections(superorder_skewed_space, type = "elements", base = "phylogeny", inc.base = TRUE, verbose = TRUE, output = c("position", "distance"), centre.proj = TRUE)
-## Getting the correlation plot
-col_subs <- c(split(skewed_space$level2, f = skewed_space$level1)[-1], "phylogeny" = list(skewed_space$level1))
-cor_results_skewed <- get.correlations(results_skewed)
-plot.correlations(cor_results_skewed[[9]], col.sub = col_subs[[9]])
-data <- cbind(cor_results_skewed[[9]]$x, cor_results_skewed[[9]]$y)
-points(data[which(rownames(data) %in% rownames(skewed_space)[fixed_points]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 2)
-points(data[which(rownames(data) %in% rownames(skewed_space)[move_around]), ], pch = 21, cex = 2, col = c("black", "darkred", "blue"), lwd = 4)
-```
-
-TODO: fix this!
-
-</details>
-
-### Correlations within super orders (species on group)
-
-Regardless, here are the results when projecting the species on their own group's VCV
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 12, eval = TRUE}
-## Getting the subsets for the colouring
-col_subs <- split(shapespace$level2, f = shapespace$level1)[-1]
-
-## Calculate the correlations
-super_within_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
-## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright", "topleft", "topright", "topleft", "topright", "topright")
-for(i in 1:length(super_within_cor)) {
-    plot.correlations(super_within_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], tree = tree)
-}
-```
-
-<!-- THIS BIT IS IGNORED FOR NOW -->
-
-### Correlations per order (species on phylogeny)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
-## Loading the sub-order levels
-load(file = "../Data/Processed/shapespace_superorder_lvl_order_suborder.rda")
-## Select the data
-shapespaces <- lapply(shapespace_superorder_lvl_order_suborder, function(X) X$space)
-## Create a level 3
-order_space <- do.call(rbind, shapespaces)
-
-## Calculate the correlations
-order_phylo_cor <- get.correlations(results$order_results_phylo, exclude.phylo = TRUE)
-
-## Getting the subsets for the colouring
-split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
-col_subs <- replicate(length(order_phylo_cor), NULL)
-col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
-
-## Set the vector of legend positions
-for(i in 1:length(order_phylo_cor)) {
-    plot.correlations(order_phylo_cor[[i]], col.sub = col_subs[[i]])
-}
-```
-
-### Correlations within orders (species on group)
-
-```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 6, eval = FALSE}
-## Calculate the correlations
-order_within_cor <- get.correlations(results$order_results_within, exclude.phylo = TRUE)
-
-## Getting the subsets for the colouring
-split_tmp <- split(order_space$level2, f = order_space$level1)[-1] ## "Charadriiformes" "Galliformes" not included?
-col_subs <- replicate(length(order_within_cor), NULL)
-col_subs[-c(6,12)] <- split_tmp[match(sort(names(split_tmp)), names(split_tmp))]
-
-## Set the vector of legend positions
-for(i in 1:length(order_within_cor)) {
-    plot.correlations(order_within_cor[[i]], col.sub = col_subs[[i]])
-}
 ```

### 2022-07-14 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Added mushed ellipses

@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_ellipses.pdf", height = box_size*8, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure3_ellipses_2.pdf", height = box_size*8, width = box_size*10)
 
@@
         matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
-        matrix(c(1,1, 18:25), ncol = 10, nrow = 1), # line 3
+        matrix(c(57,58, 18:25), ncol = 10, nrow = 1), # line 3
         matrix(c(26:35), ncol = 10, nrow = 1), # line 4
@@
         matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 7
-        matrix(c(1,1, 10:17), ncol = 10, nrow = 1) # line 8
+        matrix(c(73,0, 10:17), ncol = 10, nrow = 1) # line 8
     )
@@
 
-## PLOTTING THE ORDERS
+## Matching colours (fast utility)
+match.col <- function(data, col) {
+    return(c(col[match(names(size.subsets(data)), names(col))[-length(size.subsets(data))]], "grey"))
+}
 
@@
 }
+
+## Plotting the aligned and scale ellipses for the superorders
+par(bty = "n")
+lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "n", ylab = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+covar.plot(superorder_shapespace, ellipses = mean,
+           col = match.col(superorder_shapespace, tip_colours_super_orders),
+           legend = FALSE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
+           add = TRUE, n = 50)
+
+## PLOTTING THE ORDERS
+
+load("../Data/Processed/tip_colours_orders.rda")
+par(bty = "n")
+lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "n", ylab = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+covar.plot(order_shapespace, ellipses = mean,
+           col = match.col(order_shapespace, tip_colours_orders),
+           legend = FALSE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(0.5, length(tip_colours_orders)), 3),
+           add = TRUE, n =50)
+
 dev.off()

### 2022-07-14 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 
+## Measuring ellipse's statistics
+
+For each group, we can calculate:
+
+ * It's orientation (angle) relative to the whole phylogeny
+ * It's (dis)alignment relative to the the phylogeny
+ * It's distance from the centroid of the phylogeny
+
+We can also calculate for each group:
+
+ * It's angle internal variance (as a proxy for internal variance)
+ * It's axis significance (Andrew's bit)
+
+### Groups compared to the phylogeny
+
+```{r}
+## Measuring fun
+ellipse.stats <- function(data) {
+    ## Get the groups list
+    groups_list <- unlist(apply(matrix(c(1:(n.subsets(data)-1), rep(n.subsets(data),(n.subsets(data)-1))), ncol = 2), 1, list), recursive = FALSE)
+
+    ## Renanme the phylogeny subset
+    names(data$subsets)[n.subsets(data)] <- "phy"
+
+    ## Measure the disparities
+    results <- list(
+        angles     = dispRity(data, metric = as.covar(projections.between), measure = "degree", between.groups = groups_list),
+        alignments = dispRity(data, metric = as.covar(disalignment, VCV = c(FALSE, TRUE), loc = c(TRUE, FALSE)), between.groups = groups_list),
+        distances  = dispRity(data, metric = group.dist, probs = c(0.5), between.groups = groups_list)
+    )
+    return(results)
+}
+
+## General measurement
+super_orders_stats <- ellipse.stats(superorder_shapespace)
+orders_stats <- ellipse.stats(order_shapespace)
+
+load("../Data/Processed/tip_colours_super_orders.rda")
+col <- tip_colours_super_orders
+results <- super_orders_stats
+## Ploting fun
+ploting.stats <- function(results, col) {
+    par(mfrow = c(1,3))
+    ##
+    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
+    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
+    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
+}
+
+```
+
+### Groups compared in cherries
+
+### Groups own variance
+
 # Passeriformes shapespace analyses

### 2022-07-15 Analysis/03-Elaboration_innovation_analysis.Rmd
added passeriformes results

@@
 ```
+
+## Saving plotting data (median values)
+
+```{r, echo = FALSE, eval = FALSE}
+load("../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
+results <- shapespace_passeriformes_suborder_family_results_list
+rm(shapespace_passeriformes_suborder_family_results_list)
+## Getting the plotting data
+ploting_data_passeriformes <- list(
+                     "median_elaborations" = apply(get.disparity(get.subsets(results$suborder_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "median_innovations"  = apply(get.disparity(get.subsets(results$suborder_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
+                     "distances_to_centre" = get.disparity(dispRity(get.subsets(suborder_shapespace, subsets = "phylogeny"), metric = dispRity::centroids, centroid = 0))[[1]],
+                     "consensus_tree"      = tree)
+
+## Get the data at the macro level (with no missing data)
+get.macro.projection <- function(data, what) {
+    if(what == "centroids") {
+        ## Measuring the centroid distances
+        data <- dispRity(data, metric = dispRity::centroids)
+    } else {
+        data <- data[[what]]
+    }
+    ## Getting all the species
+    all_species <- apply(get.disparity(get.subsets(data, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median)
+    names(all_species) <- rownames(get.subsets(data, subsets = "phylogeny")$matrix[[1]])
+    ## Getting the species that have a projection on their own group
+    macro_level <- apply(do.call(rbind, get.disparity(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)]), concatenate = FALSE)), 1, median)
+    names(macro_level) <- unlist(lapply(lapply(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)])$subsets, `[[`, "elements"), function(X, matrix) rownames(matrix)[X], matrix = data$matrix[[1]]))
+    ## Replacing the species that have their own group
+    all_species[match(names(macro_level), names(all_species))] <- macro_level
+    return(all_species)
+}
+
+## Getting the plotting data super_order_level
+ploting_data_passeriformes_suborder <- list(
+                     "median_elaborations" = get.macro.projection(results$suborder_results_within, what = "position"),
+                     "median_innovations"  = get.macro.projection(results$suborder_results_within, what = "distance"),
+                     "distances_to_centre" = get.macro.projection(suborder_shapespace, what = "centroids"),
+                     "consensus_tree"      = tree)
+
+## Getting the plotting data super_order_level
+ploting_data_passeriformes_family <- list(
+                     "median_elaborations" = get.macro.projection(results$family_results_within, what = "position"),
+                     "median_innovations"  = get.macro.projection(results$family_results_within, what = "distance"),
+                     "distances_to_centre" = get.macro.projection(family_shapespace, what = "centroids"),
+                     "consensus_tree"      = tree)
+
+## Save all that as a list
+save(ploting_data_passeriformes, file = "../Data/Processed/ploting_data_passeriformes.rda")
+## The projections within the suborders
+save(ploting_data_passeriformes_suborder, file = "../Data/Processed/ploting_data_passeriformes_suborder.rda")
+## The projections within the families
+save(ploting_data_passeriformes_family, file = "../Data/Processed/ploting_data_passeriformes_family.rda")
+
+## Group info
+suborders <- lapply(results$suborder_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$suborder_results_phylo$position$matrix[[1]])
+families <- lapply(results$family_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$family_results_phylo$position$matrix[[1]])
+## Adding the colours
+load("../Data/Processed/tip_colours_suborders.rda")
+load("../Data/Processed/tip_colours_families.rda")
+add.col <- function(one_order, one_col) {
+    one_order$col <- one_col
+    return(one_order)
+}
+suborders <- mapply(add.col, one_order = suborders[1:8][match(names(suborders[1:8]), names(tip_colours_suborders))], one_col = as.list(tip_colours_suborders), SIMPLIFY = FALSE)
+families <- mapply(add.col, one_order = families[1:27][match(names(families[1:27]), names(tip_colours_families))], one_col = as.list(tip_colours_families), SIMPLIFY = FALSE)
+
+## Big list:
+clade_info_passeriformes <- list(suborders = suborders,
+                                 families = families)
+## Save all that data
+save(clade_info_passeriformes, file = "../Data/Processed/clade_info_passeriformes.rda")
+```

### 2022-07-15 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Fancier ellipse figure

@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_ellipses_2.pdf", height = box_size*8, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure3_ellipses_2.pdf", height = box_size*10, width = box_size*10)
 
 ## SETTING THE WINDOW PARAMETERS
+## Global window parameters
+lay_mat_super <- rbind(
+        matrix(c(1,1, rep(73, 3), 0, 2:5), ncol = 10, nrow = 1), # line 7
+        matrix(c(1,1, rep(73, 3), 0, 6:9), ncol = 10, nrow = 1), # line 8
+        matrix(c(78,78, 10:17), ncol = 10, nrow = 1) # line 8
+    )
+lay_mat_order <- rbind(
+        matrix(c(1,1, rep(57, 3), 60, 2:5), ncol = 10, nrow = 1), 
+        matrix(c(1,1, rep(57, 3), 60, 6:9), ncol = 10, nrow = 1), 
+        matrix(c(1,1, 10:17), ncol = 10, nrow = 1),
+        matrix(c(18:27), ncol = 10, nrow = 1),
+        matrix(c(28:37), ncol = 10, nrow = 1), 
+        matrix(c(38:47), ncol = 10, nrow = 1),
+        matrix(c(48:55, 58:59), ncol = 10, nrow = 1)
+    )
 
-## Global window parameters
-lay_mat_order <- rbind(
-        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 1
-        matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
-        matrix(c(57,58, 18:25), ncol = 10, nrow = 1), # line 3
-        matrix(c(26:35), ncol = 10, nrow = 1), # line 4
-        matrix(c(36:45), ncol = 10, nrow = 1), # line 5
-        matrix(c(46:55), ncol = 10, nrow = 1) # line 6
-    )
-lay_mat_super <- rbind(
-        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 7
-        matrix(c(73,0, 10:17), ncol = 10, nrow = 1) # line 8
-    )
 heights_vals <- rep(1, 9)
@@
 
-load("../Data/Processed/tip_colours_orders.rda")
 par(bty = "n")
@@
 ## Plotting the tree and getting the tip orders
-tip_colours_passeriformes_level1 <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
-save(tip_colours_passeriformes_level1, file = "../Data/Processed/tip_colours_passeriformes_level1.rda")
+tip_colours_passeriformes_suborder <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
+save(tip_colours_passeriformes_suborder, file = "../Data/Processed/tip_colours_passeriformes_suborder.rda")
 
@@
     ## Plot each pannel
-    wrap.plot.ellipses(i, tip_colours_passeriformes_level1, suborder_shapespace, results$suborder_results_between)
+    wrap.plot.ellipses(i, tip_colours_passeriformes_suborder, suborder_shapespace, results$suborder_results_between)
 }
@@
 ## Plotting the tree and getting the tip orders
-tip_colours_passeriformes_level2 <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
-save(tip_colours_passeriformes_level2, file = "../Data/Processed/tip_colours_passeriformes_level2.rda")
+tip_colours_passeriformes_family <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+save(tip_colours_passeriformes_family, file = "../Data/Processed/tip_colours_passeriformes_family.rda")
 
@@
     ## Plot each pannel
-    wrap.plot.ellipses(i, tip_colours_passeriformes_level2, family_shapespace, results$family_results_between)
+    wrap.plot.ellipses(i, tip_colours_passeriformes_family, family_shapespace, results$family_results_between)
 }

### 2022-07-15 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_ellipses_2.pdf", height = box_size*10, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure3_ellipses.pdf", height = box_size*10, width = box_size*10)
 
@@
            lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
-           add = TRUE, n = 50)
+           add = TRUE)
 
@@
            lines.lwd = c(rep(0.5, length(tip_colours_orders)), 3),
-           add = TRUE, n =50)
+           add = TRUE)
 
@@
 
-```
-
-## Overlapping ellipses plot
-
-```{r}
-load("../Data/Processed/tip_colours_super_orders.rda")
-load("../Data/Processed/tip_colours_orders.rda")
-
-## Matching colours (fast utility)
-match.col <- function(data, col) {
-    return(c(col[match(names(size.subsets(data)), names(col))[-length(size.subsets(data))]], "grey"))
-}
-
-pdf(file = "../Manuscript/Figures/Figure_supplementary_aligned_ellipses.pdf", height = 6, width = 12)
-par(mfrow = c(1,2))
-## Plotting the aligned and scale ellipses for the superorders
-covar.plot(superorder_shapespace, ellipses = mean,
-           col = match.col(superorder_shapespace, tip_colours_super_orders),
-           legend = TRUE, points = FALSE, scale = "phylogeny",
-           lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
-           main = "Scaled and centred average super order VCV")
-
-## Plotting the aligned and scale ellipses for the superorders
-covar.plot(order_shapespace, ellipses = mean,
-           col = match.col(order_shapespace, tip_colours_orders),
-           legend = FALSE, points = FALSE, scale = "phylogeny",
-           lines.lwd = c(rep(0.5, length(tip_colours_orders)), 3),
-           main = "Scaled and centred average order VCV")
-dev.off()
 ```
@@
 }
+
+## Plotting the aligned and scale ellipses for the superorders
+par(bty = "n")
+lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "n", ylab = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+covar.plot(suborder_shapespace, ellipses = mean,
+           col = match.col(suborder_shapespace, tip_colours_suborders),
+           legend = FALSE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(1, length(tip_colours_suborders)), 3),
+           add = TRUE, n = 50)
+
+## PLOTTING THE ORDERS
+
+par(bty = "n")
+lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "n", ylab = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+covar.plot(family_shapespace, ellipses = mean,
+           col = match.col(family_shapespace, tip_colours_families),
+           legend = FALSE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(0.5, length(tip_colours_families)), 3),
+           add = TRUE, n =50)
 dev.off()

### 2022-07-15 Analysis/03-Elaboration_innovation_analysis.Rmd
added ellipses stats

@@
 
-## Measuring ellipse's statistics
-
-For each group, we can calculate:
-
- * It's orientation (angle) relative to the whole phylogeny
- * It's (dis)alignment relative to the the phylogeny
- * It's distance from the centroid of the phylogeny
-
-We can also calculate for each group:
-
- * It's angle internal variance (as a proxy for internal variance)
- * It's axis significance (Andrew's bit)
-
-### Groups compared to the phylogeny
-
-```{r}
-## Measuring fun
-ellipse.stats <- function(data) {
-    ## Get the groups list
-    groups_list <- unlist(apply(matrix(c(1:(n.subsets(data)-1), rep(n.subsets(data),(n.subsets(data)-1))), ncol = 2), 1, list), recursive = FALSE)
-
-    ## Renanme the phylogeny subset
-    names(data$subsets)[n.subsets(data)] <- "phy"
-
-    ## Measure the disparities
-    results <- list(
-        angles     = dispRity(data, metric = as.covar(projections.between), measure = "degree", between.groups = groups_list),
-        alignments = dispRity(data, metric = as.covar(disalignment, VCV = c(FALSE, TRUE), loc = c(TRUE, FALSE)), between.groups = groups_list),
-        distances  = dispRity(data, metric = group.dist, probs = c(0.5), between.groups = groups_list)
-    )
-    return(results)
-}
-
-## General measurement
-super_orders_stats <- ellipse.stats(superorder_shapespace)
-orders_stats <- ellipse.stats(order_shapespace)
-
-load("../Data/Processed/tip_colours_super_orders.rda")
-col <- tip_colours_super_orders
-results <- super_orders_stats
-## Ploting fun
-ploting.stats <- function(results, col) {
-    par(mfrow = c(1,3))
-    ##
-    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
-    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
-    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
-}
-
-```
-
-### Groups compared in cherries
-
-### Groups own variance
-
 # Passeriformes shapespace analyses
@@
 ```
+
+## Measuring ellipse's statistics
+
+For each group, we can calculate:
+
+ * It's orientation (angle) relative to the whole phylogeny
+ * It's (dis)alignment relative to the the phylogeny
+ * It's distance from the centroid of the phylogeny
+
+We can also calculate for each group:
+
+ * It's angle internal variance (as a proxy for internal variance)
+ * It's axis significance (Andrew's bit)
+
+### Groups compared to the phylogeny
+
+```{r, eval = FALSE, echo = FALSE}
+## Measuring fun
+ellipse.stats <- function(data) {
+    ## Get the groups list
+    groups_list <- unlist(apply(matrix(c(1:(n.subsets(data)-1), rep(n.subsets(data),(n.subsets(data)-1))), ncol = 2), 1, list), recursive = FALSE)
+
+    ## Renanme the phylogeny subset
+    names(data$subsets)[n.subsets(data)] <- "phy"
+
+    ## Measure the disparities
+    results <- list(
+        angles     = dispRity(data, metric = as.covar(projections.between), measure = "degree", between.groups = groups_list),
+        alignments = dispRity(data, metric = as.covar(disalignment, VCV = c(FALSE, TRUE), loc = c(TRUE, FALSE)), between.groups = groups_list),
+        distances  = dispRity(data, metric = group.dist, probs = c(0.5), between.groups = groups_list)
+    )
+    return(results)
+}
+
+## General measurement
+super_orders_stats <- ellipse.stats(superorder_shapespace)
+orders_stats <- ellipse.stats(order_shapespace)
+suborder_stats <- ellipse.stats(suborder_shapespace)
+family_stats <- ellipse.stats(family_shapespace)
+
+ellipse_stats <- list(super_orders_stats = super_orders_stats, 
+                      orders_stats = orders_stats,
+                      suborder_stats = suborder_stats,
+                      family_stats = family_stats)
+save(ellipse_stats, file = "../Data/Processed/ellipse_stats.rda")
+```
+
+```{r, eval = TRUE, echo = FALSE}
+load("../Data/Processed/ellipse_stats.rda")
+## Ploting fun
+ploting.stats <- function(results, col) {
+
+    col <- col[match(names(size.subsets(results[[1]])), names(col))]
+
+    layout_matrix <- layout(matrix(1:3, nrow = 1), widths = c(4.2, 3, 3), heights = n.subsets(results[[1]])/2)
+    par(mar = c(5, 10, 2, 1) + 0.1)
+    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
+    par(mar = c(5, 1, 2, 1) + 0.1)
+    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
+    par(mar = c(5, 1, 2, 1) + 0.1)
+    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
+}
+
+## Ploting fun
+table.stats <- function(results, class) {
+    ## Table results
+    table <- cbind(summary(results[[1]])[, -c(1:3)], summary(results[[2]])[, -c(1:3)], summary(results[[3]])[, -c(1:3)])
+    rownames(table) <- gsub(":phy", "", as.vector(summary(results[[1]])[,1]))
+    colnames(table)[c(1,6,11)] <- c("angle", "disalignment", "distance")
+    if(!missing(class)) {
+        table <- cbind(class = class, table)
+    }
+    return(table)
+}
+
+all_results <- rbind(table.stats(ellipse_stats$super_orders_stats, class = "superorder"),
+                     table.stats(ellipse_stats$orders_stats, class = "orders"),
+                     table.stats(ellipse_stats$suborder_stats, class = "suborder"),
+                     table.stats(ellipse_stats$family_stats, class = "family"))
+knitr::kable(all_results)
+```
+
+```{r, eval = TRUE, echo = FALSE}
+load("../Data/Processed/tip_colours_super_orders.rda")
+ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
+```
+
+```{r, eval = TRUE, echo = FALSE}
+load("../Data/Processed/tip_colours_orders.rda")
+ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
+```
+
+```{r, eval = TRUE, echo = FALSE}
+load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
+ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
+```
+
+```{r, eval = TRUE, echo = FALSE}
+load("../Data/Processed/tip_colours_passeriformes_family.rda")
+ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
+```
+
+### Groups compared in cherries
+
+### Groups own variance

### 2022-07-15 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
added passeriformes

@@
            lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
-           add = TRUE)
+           add = TRUE, n = 100)
 
@@
            lines.lwd = c(rep(0.5, length(tip_colours_orders)), 3),
-           add = TRUE)
+           add = TRUE, n = 100)
 
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_passeriformes.pdf", height = box_size*7, width = box_size*10)
+pdf(file = "../Manuscript/Figures/Figure3_passeriformes.pdf", height = box_size*9, width = box_size*10)
 
 ## Global window
+x <- 61
+lay_mat_sub <- rbind(
+        matrix(c(1,1, rep(x, 3), 0, 2:5), ncol = 10, nrow = 1), # line 7
+        matrix(c(1,1, rep(x, 3), 0, 6:9), ncol = 10, nrow = 1), # line 8
+        matrix(c(x+2,x+2, 10:13, rep(x+3, 4)), ncol = 10, nrow = 1) # line 8
+    )
+y <- x - 13
 lay_mat_family <- rbind(
-        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 1
-        matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
-        matrix(c(1,1, 18:25), ncol = 10, nrow = 1), # line 3
-        matrix(c(26:35), ncol = 10, nrow = 1), # line 4
-        matrix(c(36:45), ncol = 10, nrow = 1), # line 5
-        matrix(c(46:47, rep(-13,8)), ncol = 10, nrow = 1) # line 6
-    )
-lay_mat_sub <- rbind(
-        matrix(c(1,1, 2:9, c(1,1), 10:13, rep(0, 4)), ncol = 10, nrow = 2, byrow = TRUE) # line 7
+        matrix(c(1,1, rep(y+1, 3), y+4, 2:5), ncol = 10, nrow = 1), 
+        matrix(c(1,1, rep(y+1, 3), y+4, 6:9), ncol = 10, nrow = 1), 
+        matrix(c(1,1, 10:17), ncol = 10, nrow = 1),
+        matrix(c(18:27), ncol = 10, nrow = 1),
+        matrix(c(28:37), ncol = 10, nrow = 1), 
+        matrix(c(38:47), ncol = 10, nrow = 1)
     )
@@
 covar.plot(suborder_shapespace, ellipses = mean,
-           col = match.col(suborder_shapespace, tip_colours_suborders),
+           col = match.col(suborder_shapespace, tip_colours_passeriformes_suborder),
            legend = FALSE, points = FALSE, scale = "phylogeny",
-           lines.lwd = c(rep(1, length(tip_colours_suborders)), 3),
-           add = TRUE, n = 50)
+           lines.lwd = c(rep(1, length(tip_colours_passeriformes_suborder)), 3),
+           add = TRUE, n = 100)
 
 ## PLOTTING THE ORDERS
-
 par(bty = "n")
@@
 covar.plot(family_shapespace, ellipses = mean,
-           col = match.col(family_shapespace, tip_colours_families),
+           col = match.col(family_shapespace, tip_colours_passeriformes_family),
            legend = FALSE, points = FALSE, scale = "phylogeny",
-           lines.lwd = c(rep(0.5, length(tip_colours_families)), 3),
-           add = TRUE, n =50)
+           lines.lwd = c(rep(0.5, length(tip_colours_passeriformes_family)), 3),
+           add = TRUE, n = 100)
 dev.off()

### 2022-07-19 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 
- * It's orientation (angle) relative to the whole phylogeny
  * It's (dis)alignment relative to the the phylogeny
  * It's distance from the centroid of the phylogeny
-
-We can also calculate for each group:
-
- * It's angle internal variance (as a proxy for internal variance)
- * It's axis significance (Andrew's bit)
+ * It's angles sd
+ * It's orientation (angle) relative to the whole phylogeny
+ * Whether that orientation is significant
 
@@
 ```{r, eval = FALSE, echo = FALSE}
-## Measuring fun
-ellipse.stats <- function(data) {
-    ## Get the groups list
-    groups_list <- unlist(apply(matrix(c(1:(n.subsets(data)-1), rep(n.subsets(data),(n.subsets(data)-1))), ncol = 2), 1, list), recursive = FALSE)
-
-    ## Renanme the phylogeny subset
-    names(data$subsets)[n.subsets(data)] <- "phy"
-
-    ## Measure the disparities
-    results <- list(
-        angles     = dispRity(data, metric = as.covar(projections.between), measure = "degree", between.groups = groups_list),
-        alignments = dispRity(data, metric = as.covar(disalignment, VCV = c(FALSE, TRUE), loc = c(TRUE, FALSE)), between.groups = groups_list),
-        distances  = dispRity(data, metric = group.dist, probs = c(0.5), between.groups = groups_list)
-    )
-    return(results)
-}
-
 ## General measurement
@@
 ```
-
-### Groups compared in cherries
-
-### Groups own variance

### 2022-07-19 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 ```{r, eval = FALSE, echo = FALSE}
+source("../Functions/stats.utilities.R")
 ## General measurement

### 2022-07-19 Analysis/03-Elaboration_innovation_analysis.Rmd
Isolated statsy bits

--- 2022-07-19 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 ```
-
-## Measuring ellipse's statistics
-
-For each group, we can calculate:
-
- * It's (dis)alignment relative to the the phylogeny
- * It's distance from the centroid of the phylogeny
- * It's angles sd
- * It's orientation (angle) relative to the whole phylogeny
- * Whether that orientation is significant
-
-### Groups compared to the phylogeny
-
-```{r, eval = FALSE, echo = FALSE}
-## General measurement
-super_orders_stats <- ellipse.stats(superorder_shapespace)
-orders_stats <- ellipse.stats(order_shapespace)
-suborder_stats <- ellipse.stats(suborder_shapespace)
-family_stats <- ellipse.stats(family_shapespace)
-
-ellipse_stats <- list(super_orders_stats = super_orders_stats, 
-                      orders_stats = orders_stats,
-                      suborder_stats = suborder_stats,
-                      family_stats = family_stats)
-save(ellipse_stats, file = "../Data/Processed/ellipse_stats.rda")
-```
-
-```{r, eval = TRUE, echo = FALSE}
-load("../Data/Processed/ellipse_stats.rda")
-## Ploting fun
-ploting.stats <- function(results, col) {
-
-    col <- col[match(names(size.subsets(results[[1]])), names(col))]
-
-    layout_matrix <- layout(matrix(1:3, nrow = 1), widths = c(4.2, 3, 3), heights = n.subsets(results[[1]])/2)
-    par(mar = c(5, 10, 2, 1) + 0.1)
-    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
-    par(mar = c(5, 1, 2, 1) + 0.1)
-    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
-    par(mar = c(5, 1, 2, 1) + 0.1)
-    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
-}
-
-## Ploting fun
-table.stats <- function(results, class) {
-    ## Table results
-    table <- cbind(summary(results[[1]])[, -c(1:3)], summary(results[[2]])[, -c(1:3)], summary(results[[3]])[, -c(1:3)])
-    rownames(table) <- gsub(":phy", "", as.vector(summary(results[[1]])[,1]))
-    colnames(table)[c(1,6,11)] <- c("angle", "disalignment", "distance")
-    if(!missing(class)) {
-        table <- cbind(class = class, table)
-    }
-    return(table)
-}
-
-all_results <- rbind(table.stats(ellipse_stats$super_orders_stats, class = "superorder"),
-                     table.stats(ellipse_stats$orders_stats, class = "orders"),
-                     table.stats(ellipse_stats$suborder_stats, class = "suborder"),
-                     table.stats(ellipse_stats$family_stats, class = "family"))
-knitr::kable(all_results)
-```
-
-```{r, eval = TRUE, echo = FALSE}
-load("../Data/Processed/tip_colours_super_orders.rda")
-ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
-```
-
-```{r, eval = TRUE, echo = FALSE}
-load("../Data/Processed/tip_colours_orders.rda")
-ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
-```
-
-```{r, eval = TRUE, echo = FALSE}
-load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
-ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
-```
-
-```{r, eval = TRUE, echo = FALSE}
-load("../Data/Processed/tip_colours_passeriformes_family.rda")
-ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
-```

### 2022-07-19 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Isolated statsy bits

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
This script take around 6 minutes to compile.

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
source("../Functions/ploting.utilities.R")
```

# Measuring ellipse's statistics

For each group, we can calculate:

 * It's (dis)alignment relative to the the phylogeny
 * It's distance from the centroid of the phylogeny
 * It's angles sd
 * It's orientation (angle) relative to the whole phylogeny
 * Whether that orientation is significant

# Loading the data

```{r, eval = TRUE, echo = FALSE}
## Load the data for all birds
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Create the subset for all birds
all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

# Load the data for the passeriformes
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Create the subset for all birds
all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

```

### Groups compared to the phylogeny

```{r, eval = FALSE, echo = FALSE}
source("../Functions/stats.utilities.R")
## General measurement
super_orders_stats <- ellipse.stats(superorder_shapespace)
orders_stats <- ellipse.stats(order_shapespace)
suborder_stats <- ellipse.stats(suborder_shapespace)
family_stats <- ellipse.stats(family_shapespace)

ellipse_stats <- list(super_orders_stats = super_orders_stats, 
                      orders_stats = orders_stats,
                      suborder_stats = suborder_stats,
                      family_stats = family_stats)
save(ellipse_stats, file = "../Data/Processed/ellipse_stats.rda")
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/ellipse_stats.rda")
## Ploting fun
ploting.stats <- function(results, col) {

    col <- col[match(names(size.subsets(results[[1]])), names(col))]

    layout_matrix <- layout(matrix(1:3, nrow = 1), widths = c(4.2, 3, 3), heights = n.subsets(results[[1]])/2)
    par(mar = c(5, 10, 2, 1) + 0.1)
    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
    par(mar = c(5, 1, 2, 1) + 0.1)
    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
    par(mar = c(5, 1, 2, 1) + 0.1)
    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
}

## Ploting fun
table.stats <- function(results, class) {
    ## Table results
    table <- cbind(summary(results[[1]])[, -c(1:3)], summary(results[[2]])[, -c(1:3)], summary(results[[3]])[, -c(1:3)])
    rownames(table) <- gsub(":phy", "", as.vector(summary(results[[1]])[,1]))
    colnames(table)[c(1,6,11)] <- c("angle", "disalignment", "distance")
    if(!missing(class)) {
        table <- cbind(class = class, table)
    }
    return(table)
}

all_results <- rbind(table.stats(ellipse_stats$super_orders_stats, class = "superorder"),
                     table.stats(ellipse_stats$orders_stats, class = "orders"),
                     table.stats(ellipse_stats$suborder_stats, class = "suborder"),
                     table.stats(ellipse_stats$family_stats, class = "family"))
knitr::kable(all_results)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_super_orders.rda")
ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_orders.rda")
ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_family.rda")
ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
```

### 2022-07-19 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
calcualting the stats for all ellipses

@@
 library(dispRity)
-source("../Functions/ploting.utilities.R")
 ```
@@
 ## Select the data
-shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
-tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
+shapespace_allbirds <- shapespace_allbirds_lvl_superorder_order[[1]]$space
 ## Clean garbage
@@
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
+shapespace_allbirds_posteriors <- shapespace_allbirds_lvl_superorder_order[[1]]
+rm(shapespace_allbirds_lvl_superorder_order)
 
 ## Get the levels
-phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
-super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
-order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
+phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_posteriors) == "animal")
+super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_posteriors))
+order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_posteriors))
 
@@
 all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
-      data          = shapespace,
-      posteriors    = shapespace_allbirds_lvl_superorder_order,
-      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, order_lvl, phylo_lvl)],
-      rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+      data          = shapespace_allbirds,
+      posteriors    = shapespace_allbirds_posteriors,
+      group         = MCMCglmm.levels(shapespace_allbirds_posteriors)[c(super_lvl, order_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace_allbirds$level1)[-1], levels(shapespace_allbirds$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
 ## Clean garbage
-rm(shapespace_allbirds_lvl_superorder_order)
-
-# Load the data for the passeriformes
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
-## Select the data
-shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
-tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
-## Clean garbage
-rm(shapespace_allbirds_lvl_superorder_order)
+rm(shapespace_allbirds_posteriors)
 
 ## Load the covar
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
+load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+shapespace_passeriformes_posteriors <- shapespace_passeriformes[[1]]
+rm(shapespace_passeriformes)
+## Load the data for the passeriformes
+load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
+## Select the data
+shapespace_passeriformes <- shapespace_passeriformes_two_levels[[1]]$space
+## Clean garbage
+rm(shapespace_passeriformes_two_levels)
 
 ## Get the levels
-phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
-super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
-order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
+phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes_posteriors) == "animal")
+suborder_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes_posteriors))
+family_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes_posteriors))
 
 ## Create the subset for all birds
-all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
-      data          = shapespace,
-      posteriors    = shapespace_allbirds_lvl_superorder_order,
-      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, order_lvl, phylo_lvl)],
-      rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+passeriformes_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
+      data          = shapespace_passeriformes,
+      posteriors    = shapespace_passeriformes_posteriors,
+      group         = MCMCglmm.levels(shapespace_passeriformes_posteriors)[c(suborder_lvl, family_lvl, phylo_lvl)],
+      rename.groups = c(levels(shapespace_passeriformes$level1)[-1], levels(shapespace_passeriformes$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
 ## Clean garbage
-rm(shapespace_allbirds_lvl_superorder_order)
+rm(shapespace_passeriformes_posteriors)
+```
 
+```{r}
+## Creating the nested levels subgroups
+make.nested.subset <- function(group, data, group.data) {
+    ## Get the subsets within the group
+    level2_groups <- as.character(unique(group.data$level2[group.data$level1 == group]))
+    if(any(empty <- level2_groups == "")) {
+        level2_groups <- level2_groups[-which(empty)]
+    }
+    if(length(level2_groups) < 1) {
+        return(NULL)
+    }
+    ## Create a subset
+    return(get.subsets(data, subsets = c(level2_groups, group)))
+}
+
+## Get the subgroups for all_birds
+all_birds_shapespace_subgroups <- lapply(as.list(levels(shapespace_allbirds$level1)[-1]), make.nested.subset, data = all_birds_shapespace, group.data = shapespace_allbirds)
+names(all_birds_shapespace_subgroups) <- levels(shapespace_allbirds$level1)[-1]
+## Get the subgroups for the passeriformes
+passeriformes_shapespace_subgroups <- lapply(as.list(levels(shapespace_passeriformes$level1)[-1]), make.nested.subset, data = passeriformes_shapespace, group.data = shapespace_passeriformes)
+names(passeriformes_shapespace_subgroups) <- levels(shapespace_passeriformes$level1)[-1]
 ```
@@
 source("../Functions/stats.utilities.R")
+
+## Fucked up thing happening here:
+cd("~/Packaging/dispRity/")
+load_all()
+
 ## General measurement
-super_orders_stats <- ellipse.stats(superorder_shapespace)
-orders_stats <- ellipse.stats(order_shapespace)
-suborder_stats <- ellipse.stats(suborder_shapespace)
-family_stats <- ellipse.stats(family_shapespace)
+all_birds_all_levels <- ellipse.stats(all_birds_shapespace)
+passeriformes_all_levels <- ellipse.stats(passeriformes_shapespace)
 
-ellipse_stats <- list(super_orders_stats = super_orders_stats, 
-                      orders_stats = orders_stats,
-                      suborder_stats = suborder_stats,
-                      family_stats = family_stats)
-save(ellipse_stats, file = "../Data/Processed/ellipse_stats.rda")
+## Nested measurements
+all_birds_nested_levels <- lapply(all_birds_shapespace_subgroups, ellipse.stats)
+passeriformes_nested_levels <- lapply(passeriformes_shapespace_subgroups, ellipse.stats)
+
+## Big measurements list
+ellipses_stats <- list("all_birds" = all_birds_all_levels)
+ellipses_stats <- c(ellipses_stats, all_birds_nested_levels)
+ellipses_stats$Passeriformes <- passeriformes_all_levels
+ellipses_stats <- c(ellipses_stats, passeriformes_nested_levels)
+
+save(ellipses_stats, file = "../Data/Processed/ellipses_stats.rda")
 ```

### 2022-07-19 Functions/stats.utilities.R
Update stats.utilities.R

@@ -131,9 +131,15 @@ run.ellipse.test <- function(data, base = "phylogeny", p.value.method = "Bayesia
     if(verbose) cat("Testing the ellipses differences (from the base):")
 
     ## Setting up the base
-    base <- data$covar[[base]]
+    if(!(base %in% names(data$covar))) {
+        ## By default set the last subset as the base
+        base_name <- names(data$covar)[length((data$covar))]
+    } else {
+        base_name <- base
+    }
+    base <- data$covar[[base_name]]
     ## Setting up the other sets
-    posteriors <- data$covar
+    posteriors <- data$covar[names(data$covar) != base_name]
     
     ## Measuring everything
     all_results <- lapply(posteriors, covar.ellipse.test, y = base, p.value.method = p.value.method, measure = measure, major.axis = major.axis, verbose = verbose)
@@ -177,14 +183,15 @@ ellipse.stats <- function(data, verbose = TRUE) {
     results$distances  <- dispRity(data, metric = group.dist, probs = c(0.5), between.groups = groups_list, verbose = verbose)
 
     if(verbose) cat("Measuring sd:")
-    results$distances  <- unlist(lapply(lapply(get.disparity(dispRity(data, metric = as.covar(angles.base), verbose = verbose), concatenate = FALSE), c), sd))
+    covar.angles <- as.covar(angles.base)
+    results$sd  <- unlist(lapply(lapply(get.disparity(dispRity(data, metric = covar.angles, verbose = verbose), concatenate = FALSE), c), sd))
 
     if(verbose) cat("Measuring alignments:")
-    results$alignments <- dispRity(data, metric = as.covar(disalignment, VCV = c(FALSE, TRUE), loc = c(TRUE, FALSE)), between.groups = groups_list, verbose = verbose)
+    covar.disalignment <- as.covar(disalignment, VCV = c(FALSE, TRUE), loc = c(TRUE, FALSE))
+    results$alignments <- dispRity(data, metric = covar.disalignment, between.groups = groups_list, verbose = verbose)
 
     if(verbose) cat("Measuring orthogonality:")
     results$angles     <- dispRity(data, metric = as.covar(projections.between), measure = "orthogonality", between.groups = groups_list, verbose = verbose)
-
     if(verbose) cat("Testing orthogonality:")
     results$tests      <- run.ellipse.test(data)
 

### 2022-07-20 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 ## Load the data for all birds
@@
 ## Create the subset for all birds
-all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
+all_birds_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 500 matrices for now
       data          = shapespace_allbirds,
@@
 ## Create the subset for all birds
-passeriformes_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
+passeriformes_shapespace <- MCMCglmm.subsets( # Just run it on 500 matrices for now
       data          = shapespace_passeriformes,
@@
       group         = MCMCglmm.levels(shapespace_passeriformes_posteriors)[c(suborder_lvl, family_lvl, phylo_lvl)],
-      rename.groups = c(levels(shapespace_passeriformes$level1)[-1], levels(shapespace_passeriformes$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
+      rename.groups = c(levels(shapespace_passeriformes$level1)[-1], levels(shapespace_passeriformes$level2)[-1], "Passeriformes")) #[-1] is because the first level is ""
 ## Clean garbage
@@
 
-```{r}
+```{r, eval = FALSE, echo = FALSE}
 ## Creating the nested levels subgroups
@@
 
+And here are the full results! Here's what's in each column:
+
+ * **Group**: the name of the group (a random term of the model), the focal VCV
+ * **Comparison**: to which group are the VCV compared (the group is either one or two levels higher)
+ * **n**: number of species
+ * **distance**: the distance between the focal and the comparison group's centers
+ * **ellipse sd**: the standard deviation of the angles in the the focal VCV distribution
+ * **disalignment**: the disalignment of the group compared to the focal group (0 = aligned), with the 95% CI
+ * **orthogonality**: the orthogonality amount between both groups (0 = parallel, 1 = orthogonal), with the 95% CI
+ * **Post. prob.**: the posterior probability of the orthogonality in the focal group being different from the comparison one (with the 95% CI estimated from the randomised posterior probabilities)
+
 ```{r, eval = TRUE, echo = FALSE}
-load("../Data/Processed/ellipse_stats.rda")
-## Ploting fun
-ploting.stats <- function(results, col) {
-
-    col <- col[match(names(size.subsets(results[[1]])), names(col))]
-
-    layout_matrix <- layout(matrix(1:3, nrow = 1), widths = c(4.2, 3, 3), heights = n.subsets(results[[1]])/2)
-    par(mar = c(5, 10, 2, 1) + 0.1)
-    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
-    par(mar = c(5, 1, 2, 1) + 0.1)
-    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
-    par(mar = c(5, 1, 2, 1) + 0.1)
-    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
-}
+load("../Data/Processed/ellipses_stats.rda")
 
 ## Ploting fun
-table.stats <- function(results, class) {
+table.stats <- function(results) {
     ## Table results
-    table <- cbind(summary(results[[1]])[, -c(1:3)], summary(results[[2]])[, -c(1:3)], summary(results[[3]])[, -c(1:3)])
-    rownames(table) <- gsub(":phy", "", as.vector(summary(results[[1]])[,1]))
-    colnames(table)[c(1,6,11)] <- c("angle", "disalignment", "distance")
-    if(!missing(class)) {
-        table <- cbind(class = class, table)
-    }
+
+    ## Names, comparison and size
+    header <- summary(results$distances)[,c(1,2), drop = FALSE]
+    head_1 <- as.data.frame(do.call(rbind, strsplit(header[,1], split = ":")))
+    header <- cbind(head_1, "n" = header[, 2])
+    colnames(header)[c(1,2)] <- c("Group", "Comparison")
+
+    ## Adding the statistics
+    table <- cbind(header, "distance" = summary(results$distances)[,4])
+    table <- cbind(table, "ellipse sd" = results$sd[-length(results$sd)])
+    table <- cbind(table, summary(results$alignments, quantiles = c(95))[, -c(1:3)])
+    colnames(table)[ncol(table)-2] <- "disalignment"
+    table <- cbind(table, summary(results$angles, quantiles = c(95))[, -c(1:3)])
+    colnames(table)[ncol(table)-2] <- "orthogonality"
+    table <- cbind(table, results$test)
+    colnames(table)[ncol(table)-2] <- "Post. prob."
+    rownames(table) <- NULL
     return(table)
@@
 
-all_results <- rbind(table.stats(ellipse_stats$super_orders_stats, class = "superorder"),
-                     table.stats(ellipse_stats$orders_stats, class = "orders"),
-                     table.stats(ellipse_stats$suborder_stats, class = "suborder"),
-                     table.stats(ellipse_stats$family_stats, class = "family"))
+## Sorting all results
+all_results <- do.call(rbind, lapply(ellipses_stats[-12], table.stats)) ## Droping the Meliphagoidea (single group)
+rownames(all_results) <- NULL
+
+write.csv(all_results, "../Data/Processed/ellipses.stats.csv")
 knitr::kable(all_results)
@@
 
-```{r, eval = TRUE, echo = FALSE}
+<!-- length(which(all_results[which(all_results[,"Post. prob."] >= 0.95), ]$orthogonality > 0.5)) -->
+
+```{r, eval = TRUE, echo = FALSE, fig.height = 20, fig.width = 8}
+# pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 6, height = 15)
+source("../Functions/ploting.utilities.R")
+## Plot the orthogonality into context
+orthogonality <- all_results[, c(1:2, 9:14)]
+
+## Group per level
+super_orders <- orthogonality[c(1:8), ]
+orders <- orthogonality[c(36:60), ]
+orders_n2 <- orthogonality[c(9:35), ]
+sub_orders <- orthogonality[c(61:66), ]
+families <- orthogonality[c(90:110), ]
+families_n2 <- orthogonality[c(67:89), ]
+
+## Sort each level by orthogonality score
+ortho.sort <- function(data) {
+    return(data[order(data$orthogonality, decreasing = TRUE), , drop = FALSE])
+}
+## Plotting table (with the sorted orthogonality)
+plot_table <- rbind(ortho.sort(super_orders),
+                    ortho.sort(orders),
+                    ortho.sort(sub_orders),
+                    ortho.sort(families))
+colours <- c(rep(gg.color.hue(4)[1], nrow(super_orders)),
+             rep(gg.color.hue(4)[2], nrow(orders)),
+             rep(gg.color.hue(4)[3], nrow(sub_orders)),
+             rep(gg.color.hue(4)[4], nrow(families)))
+
+## Adding starts here and there
+get.signif <- function(prob) {
+    if(prob >= 0.99) {
+        return("***")
+    } else {
+        if(prob >= 0.95) {
+            return("**")
+        } else {
+            if(prob >= 0.90) {
+                return("*")
+            } else {
+                if(prob >= 0.80) {
+                    return(".")
+                } else {
+                    return("")
+                }
+            }
+        }
+    }
+}
+add.stars <- function(data, counter, shift = 0, total, ...) {
+    text(data[3] + 0.03, (total+1)-counter + shift + 0.2, labels = get.signif(data[6]), cex = 1, ...)   
+}
+add.line <- function(data, counter, shift = 0, total, ...) {
+    lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), ...)
+    points(data[3], (total+1)-counter + shift, pch = 19, ...)
+}
+
+## Plot frame
+par(mar = c(5, 9, 4, 2))
+plot(NULL, yaxt = "n", xaxt = "n", ylab = "", xlab = "orthogonality", ylim = c(1,nrow(plot_table)), xlim = c(0,1))
+abline(v = 0.5, lwd = 0.5, col = "grey")
+abline(v = 0.75, lty = 2, col = "grey", lwd = 0.5)
+axis(2, at = nrow(plot_table):1, labels = plot_table$Group, las = 2)
+axis(1, at = c(0, 0.5, 1), labels = c(0, 0.5, 1))
+## Adding the results
+
+for(i in 1:nrow(plot_table)) {
+    add.line(plot_table[i, ], counter = i, col = colours[i], total = nrow(plot_table))
+    add.stars(plot_table[i, ], counter = i, col = colours[i], total = nrow(plot_table))
+}
+
+## Adding the n2 results
+orders_n2_plot <- orders_n2[match(orders[order(orders$orthogonality, decreasing = TRUE), ]$Group, orders_n2$Group), ]
+families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
+for(i in 1:nrow(orders_n2_plot)) {
+    add.line(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 1/3), total = nrow(plot_table))
+    add.stars(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 1/3), total = nrow(plot_table))
+}
+for(i in 1:nrow(orders_n2)) {
+    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
+    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
+}
+
+legend("topleft", legend = c("Super orders (vs.
+phylogeny)", "Orders (vs.
+super orders)", "Orders (vs.
+phylogeny)", "Suborders (vs.
+passeriformes)", "Families (vs.
+suborders)", "Families (vs.
+passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
+# dev.off()
+```
+
+Distribution of the VCV ellipses' orthogonality for all the groups compared to the ellipse of their parent clade (or n-2). 
+For the posterior probabilities: . >= 0.8; * >= 0.9; ** >= 0.95; *** >= 0.99.
+
+```{r, eval = FALSE, echo = FALSE}
 load("../Data/Processed/tip_colours_super_orders.rda")
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 load("../Data/Processed/tip_colours_orders.rda")
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 load("../Data/Processed/tip_colours_passeriformes_family.rda")

### 2022-07-21 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Isolated table function

@@
 ## Create the subset for all birds
-all_birds_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 500 matrices for now
+all_birds_shapespace <- MCMCglmm.subsets(
       data          = shapespace_allbirds,
@@
 ## Create the subset for all birds
-passeriformes_shapespace <- MCMCglmm.subsets( # Just run it on 500 matrices for now
+passeriformes_shapespace <- MCMCglmm.subsets(
       data          = shapespace_passeriformes,
@@
 
-## Fucked up thing happening here:
-cd("~/Packaging/dispRity/")
-load_all()
+# ## Fucked up thing happening here:
+# cd("~/Packaging/dispRity/")
+# load_all()
 
@@
 load("../Data/Processed/ellipses_stats.rda")
-
-## Ploting fun
-table.stats <- function(results) {
-    ## Table results
-
-    ## Names, comparison and size
-    header <- summary(results$distances)[,c(1,2), drop = FALSE]
-    head_1 <- as.data.frame(do.call(rbind, strsplit(header[,1], split = ":")))
-    header <- cbind(head_1, "n" = header[, 2])
-    colnames(header)[c(1,2)] <- c("Group", "Comparison")
-
-    ## Adding the statistics
-    table <- cbind(header, "distance" = summary(results$distances)[,4])
-    table <- cbind(table, "ellipse sd" = results$sd[-length(results$sd)])
-    table <- cbind(table, summary(results$alignments, quantiles = c(95))[, -c(1:3)])
-    colnames(table)[ncol(table)-2] <- "disalignment"
-    table <- cbind(table, summary(results$angles, quantiles = c(95))[, -c(1:3)])
-    colnames(table)[ncol(table)-2] <- "orthogonality"
-    table <- cbind(table, results$test)
-    colnames(table)[ncol(table)-2] <- "Post. prob."
-    rownames(table) <- NULL
-    return(table)
-}
 
@@
 For the posterior probabilities: . >= 0.8; * >= 0.9; ** >= 0.95; *** >= 0.99.
-
-```{r, eval = FALSE, echo = FALSE}
-load("../Data/Processed/tip_colours_super_orders.rda")
-ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
-```
-
-```{r, eval = FALSE, echo = FALSE}
-load("../Data/Processed/tip_colours_orders.rda")
-ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
-```
-
-```{r, eval = FALSE, echo = FALSE}
-load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
-ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
-```
-
-```{r, eval = FALSE, echo = FALSE}
-load("../Data/Processed/tip_colours_passeriformes_family.rda")
-ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
-```

### 2022-07-21 Functions/stats.utilities.R

@@ -201,3 +201,29 @@ ellipse.stats <- function(data, verbose = TRUE) {
 
     return(results)
 }
+
+#' @name table.stats 
+#' @description Summarising the statistics into a table
+#'
+#' @param results the results output from ellipse.stats
+table.stats <- function(results) {
+    ## Table results
+
+    ## Names, comparison and size
+    header <- summary(results$distances)[,c(1,2), drop = FALSE]
+    head_1 <- as.data.frame(do.call(rbind, strsplit(header[,1], split = ":")))
+    header <- cbind(head_1, "n" = header[, 2])
+    colnames(header)[c(1,2)] <- c("Group", "Comparison")
+
+    ## Adding the statistics
+    table <- cbind(header, "distance" = summary(results$distances)[,4])
+    table <- cbind(table, "ellipse sd" = results$sd[-length(results$sd)])
+    table <- cbind(table, summary(results$alignments, quantiles = c(95))[, -c(1:3)])
+    colnames(table)[ncol(table)-2] <- "disalignment"
+    table <- cbind(table, summary(results$angles, quantiles = c(95))[, -c(1:3)])
+    colnames(table)[ncol(table)-2] <- "orthogonality"
+    table <- cbind(table, results$test)
+    colnames(table)[ncol(table)-2] <- "Post. prob."
+    rownames(table) <- NULL
+    return(table)
+}

### 2022-07-21 Analysis/03-Elaboration_innovation_analysis.Rmd
EI at species vs group level figure

@@
 
+```{r}
+## Distribution of elaboration and innovation at the species level
+elaboration_species      <- get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]]
+innovation_species       <- get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]]
+
+## Distribution of elaboration and innovation at the group level
+scale.group <- function(X) return(X/max(X))
+elaboration_super_orders <- lapply(get.disparity(results$super_results_between$position, concatenate = FALSE), scale.group)
+innovation_super_orders  <- lapply(get.disparity(results$super_results_between$distance, concatenate = FALSE), scale.group)
+elaboration_orders       <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
+innovation_orders        <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
+
+## Plot all that
+plot.density <- function(species, group, main, scale.x = TRUE, scale.y = TRUE, col = c("blue", "orange")) {
+    scale.xy <- function(density, x = FALSE, y = TRUE) {
+        if(x) {
+            density$x <- density$x/max(density$x)
+        }
+        if(y) {
+            density$y <- density$y/max(density$y)
+        }
+        return(density)
+    }
+    den_sp    <- scale.xy(density(species), x = scale.x, y = scale.y)
+    den_group <- scale.xy(density(group), x = scale.x, y = scale.y)
+
+    plot(NULL, xlim = range(c(den_sp$x, den_group$x)), ylim = range(c(den_sp$y, den_group$y)), xlab = main, ylab = "density")
+    lines(den_sp, col = col[1])
+    lines(den_group, col = col[2])
+}
+
+pdf("../Manuscript/Figures/Relative_EI.pdf", height = 12, width = 6)
+par(mfrow = c(2,1))
+plot.density(species = c(elaboration_species),
+             group   = c(unlist(elaboration_super_orders), unlist(elaboration_orders)),
+             main    = "relative elaboration", scale.x = TRUE)
+plot.density(species = c(innovation_species),
+             group   = c(unlist(innovation_super_orders), unlist(innovation_orders)),
+             main    = "relative innovation", scale.x = TRUE)
+legend("topright", lty = 1, legend = c("species", "groups"), col = c("blue", "orange"))
+dev.off()
+```
+
 # Passeriformes shapespace analyses

### 2022-07-22 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 load("../Data/Processed/ellipses_stats.rda")
+source("../Functions/stats.utilities.R")
 
@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 20, fig.width = 8}
-# pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 6, height = 15)
+#TODO: make the groups nested in the fable
+
+pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 6, height = 15)
 source("../Functions/ploting.utilities.R")
 ## Plot the orthogonality into context
-orthogonality <- all_results[, c(1:2, 9:14)]
+orthogonality <- all_results[, c(1:2, 9:14, 3, 5)]
 
@@
 
+## Add the number of species and sd to the group names
+convert.spaces <- function(x, max) {
+    shifts <- max-nchar(as.character(x))
+    if(shifts > 0) {
+        return(paste(c(x, rep(" ", shifts)), collapse = ""))
+    } else {
+        return(x)
+    }
+}
+group_names <- paste0(plot_table$Group, "(n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
+
+plot_order <- 1:60
+## Nestedness bit (not finished)
+# plot_order <- integer()
+# ## Super-orders
+# for(i in 1:6) {
+#     plot_order <- c(plot_order, i) 
+#     sub_groups <- which(plot_table$Comparison == plot_table$Group[plot_order[length(plot_order)]])
+#     if(length(sub_groups)) {
+#         plot_order <- c(plot_order, sub_groups)
+#     }
+# }
+# ## Sub-orders
+# for(i in 34:39) {
+#     plot_order <- c(plot_order, i) 
+#     sub_groups <- which(plot_table$Comparison == plot_table$Group[plot_order[length(plot_order)]])
+#     if(length(sub_groups)) {
+#         plot_order <- c(plot_order, sub_groups)
+#     }
+# }
+
 ## Plot frame
-par(mar = c(5, 9, 4, 2))
+par(mar = c(5, 17, 4, 2))
 plot(NULL, yaxt = "n", xaxt = "n", ylab = "", xlab = "orthogonality", ylim = c(1,nrow(plot_table)), xlim = c(0,1))
@@
 abline(v = 0.75, lty = 2, col = "grey", lwd = 0.5)
-axis(2, at = nrow(plot_table):1, labels = plot_table$Group, las = 2)
+axis(2, at = nrow(plot_table):1, labels = group_names[plot_order], las = 2)
 axis(1, at = c(0, 0.5, 1), labels = c(0, 0.5, 1))
 ## Adding the results
-
 for(i in 1:nrow(plot_table)) {
-    add.line(plot_table[i, ], counter = i, col = colours[i], total = nrow(plot_table))
-    add.stars(plot_table[i, ], counter = i, col = colours[i], total = nrow(plot_table))
+    add.line (plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
+    add.stars(plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
 }
@@
 passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
-# dev.off()
+dev.off()
 ```

### 2022-07-22 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Added curves comparisons

@@
 For the posterior probabilities: . >= 0.8; * >= 0.9; ** >= 0.95; *** >= 0.99.
+
+```{r, echo = FALSE, eval = TRUE}
+## Quick load
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
+results <- shapespace_allbirds_lvl_superorder_order_results_list
+rm(shapespace_allbirds_lvl_superorder_order_results_list)
+
+## Distribution of elaboration and innovation at the species level
+elaboration_species      <- get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]]
+innovation_species       <- get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]]
+
+## Distribution of elaboration and innovation at the group level
+scale.group <- function(X) return(X/max(X))
+elaboration_super_orders <- lapply(get.disparity(results$super_results_between$position, concatenate = FALSE), scale.group)
+innovation_super_orders  <- lapply(get.disparity(results$super_results_between$distance, concatenate = FALSE), scale.group)
+elaboration_orders       <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
+innovation_orders        <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
+elaboration_group <- rbind(do.call(rbind, elaboration_super_orders), do.call(rbind, elaboration_orders))
+innovation_group <- rbind(do.call(rbind, innovation_super_orders), do.call(rbind, innovation_orders))
+
+## Measuring the area under the curves
+elaboration_species_auc <- apply(elaboration_species, 2, area.density)
+innovation_species_auc <- apply(innovation_species, 2, area.density)
+elaboration_group_auc <- apply(elaboration_group, 2, area.density)
+innovation_group_auc <- apply(innovation_group, 2, area.density)
+
+pdf("../Manuscript/Figures/Relative_EI.pdf", height = 8, width = 8)
+par(mfrow = c(2,2))
+plot.density(species = c(elaboration_species),
+             group   = c(unlist(elaboration_super_orders), unlist(elaboration_orders)),
+             xlab    = "relative elaboration", scale.x = TRUE,
+             ylab    = "density",
+             main    = "A - Total posterior elaboration per group (scaled)\nand per species")
+legend("topright", lty = 1, legend = c("species", "groups"), col = c("blue", "orange"))
+plot.density(species = c(innovation_species),
+             group   = c(unlist(innovation_super_orders), unlist(innovation_orders)),
+             xlab    = "relative innovation", scale.x = TRUE,
+             ylab    = "density",
+             main    = "B - Total posterior innovation per group (scaled)\nand per species")
+plot.density(elaboration_species_auc, elaboration_group_auc, scale.x = FALSE, col = c("blue", "orange"), ylab = "density",
+             xlab = "Posterior area under the curve (elaboration)",
+             main = paste0("C - Difference in amounts of elaboration\nper group or per species\nBhattacharyya coefficient = ",
+                           round(bhatt.coeff(elaboration_species_auc, elaboration_group_auc), 3)))
+plot.density(innovation_species_auc, innovation_group_auc, scale.x = FALSE, col = c("blue", "orange"), ylab = "density",
+             xlab = "Posterior area under the curve (innovation)",
+             main = paste0("D - Difference in amounts of innovation\nper group or per species\nBhattacharyya coefficient = ",
+                           bhatt.coeff(innovation_species_auc, innovation_group_auc)))
+dev.off()
+```

### 2022-07-22 Functions/stats.utilities.R
Update stats.utilities.R

@@ -273,3 +273,43 @@ table.stats <- function(results) {
     rownames(table) <- NULL
     return(table)
 }
+
+#' @name scale.xy 
+#' @description Scales a density curve
+#'
+#' @param density a density curve
+#' @param x,y logical, whether to scale the x and or the y axis
+scale.xy <- function(density, x = TRUE, y = TRUE) {
+    if(x) {
+        density$x <- density$x/max(density$x)
+    }
+    if(y) {
+        density$y <- density$y/max(density$y)
+    }
+    return(density)
+}
+
+#' @name plot.density 
+#' @description Wrapper for plotting a density curve
+#'
+#' @param species the data per species
+#' @param group the data per group
+#' @param scale.x,scale.y logical, whether to scale the x and or the y axis
+plot.density <- function(species, group, scale.x = TRUE, scale.y = TRUE, col = c("blue", "orange"), ...) {
+    den_sp    <- scale.xy(density(species), x = scale.x, y = scale.y)
+    den_group <- scale.xy(density(group), x = scale.x, y = scale.y)
+
+    plot(NULL, xlim = range(c(den_sp$x, den_group$x)), ylim = range(c(den_sp$y, den_group$y)), ...)
+    lines(den_sp, col = col[1])
+    lines(den_group, col = col[2])
+}
+
+#' @name area.density
+#' @description area under the curve for density
+#'
+#' @param data calculate the area under a density curve
+#' @param scale.x,scale.y logical, whether to scale the x and or the y axis
+area.density <- function(data, scale.x = TRUE, scale.y = TRUE) {
+    density <- scale.xy(density(data), x = scale.x, y = scale.y)
+    area_under_curve <- sum(diff(density$x[order(density$x)]) * zoo::rollmean(density$y[order(density$x)], 2))
+}

### 2022-07-22 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 
-pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 6, height = 15)
+pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 9, height = 18)
 source("../Functions/ploting.utilities.R")
@@
              ylab    = "density",
-             main    = "A - Total posterior elaboration per group (scaled)\nand per species")
+             main    = "A - Total posterior elaboration per group\n(scaled) and per species")
 legend("topright", lty = 1, legend = c("species", "groups"), col = c("blue", "orange"))
@@
              ylab    = "density",
-             main    = "B - Total posterior innovation per group (scaled)\nand per species")
+             main    = "B - Total posterior innovation per group\n(scaled) and per species")
 plot.density(elaboration_species_auc, elaboration_group_auc, scale.x = FALSE, col = c("blue", "orange"), ylab = "density",

### 2022-08-01 Analysis/03-Elaboration_innovation_analysis.Rmd
Added curves comparisons
Update 03-Elaboration_innovation_analysis.Rmd

--- 2022-07-19 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 ## Adding the colours
-load("../Data/Processed/tip_colours_suborders.rda")
-load("../Data/Processed/tip_colours_families.rda")
+load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
+load("../Data/Processed/tip_colours_passeriformes_family.rda")
 add.col <- function(one_order, one_col) {
@@
 }
-suborders <- mapply(add.col, one_order = suborders[1:8][match(names(suborders[1:8]), names(tip_colours_suborders))], one_col = as.list(tip_colours_suborders), SIMPLIFY = FALSE)
-families <- mapply(add.col, one_order = families[1:27][match(names(families[1:27]), names(tip_colours_families))], one_col = as.list(tip_colours_families), SIMPLIFY = FALSE)
+suborders <- mapply(add.col, one_order = suborders[1:8][match(names(suborders[1:8]), names(tip_colours_passeriformes_suborder))], one_col = as.list(tip_colours_suborders), SIMPLIFY = FALSE)
+families <- mapply(add.col, one_order = families[1:27][match(names(families[1:27]), names(tip_colours_passeriformes_family))], one_col = as.list(tip_colours_families), SIMPLIFY = FALSE)
 

### 2022-08-01 Poster/Poster_plots.R
Update Poster_plots.R

@@ -30,7 +30,7 @@ dev.off()
 
 
 ## Ploting the tree
-pdf(file = "tree_orders.pdf", height = 40, width = 26)
+pdf(file = "tree_orders_col.pdf", height = 40, width = 26)
 
 tree <- ladderize(ploting_data$consensus_tree, right = FALSE)
 
@@ -54,6 +54,26 @@ plot(tree, show.tip.label = FALSE, edge.color = as.character(edge_colors), edge.
 tiplabels("", frame = "none", pch = 15, col = rgb(cols_elab, maxColorValue = 255), offset = 2)
 tiplabels("", frame = "none", pch = 15, col = rgb(cols_innov, maxColorValue = 255), offset = 4)
 dev.off()
+
+
+## Tiplabels is clade name
+clade_names <- rep("grey", Ntip(tree))
+names(clade_names) <- tree$tip.label
+load("../Data/Processed/clade_info.rda")
+load("../Data/Processed/clade_info_passeriformes.rda")
+
+for(i in 1:27) {
+    clade_names[clade_info$orders[[i]]$species] <- clade_info$orders[[i]]$col
+}
+for(i in 1:23) {
+    clade_names[clade_info_passeriformes$families[[i]]$species] <- clade_info_passeriformes$families[[i]]$col
+}
+
+length(unique(clade_names))
+
+plot(tree, show.tip.label = FALSE, edge.color = as.character(edge_colors), edge.width = 3, edge.lty = 1)
+tiplabels("", frame = "none", pch = 15, col = clade_names, offset = 2)
+
 
 # 164.760
 # 210.800

### 2022-08-01 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

--- 2022-07-19 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 }
-suborders <- mapply(add.col, one_order = suborders[1:8][match(names(suborders[1:8]), names(tip_colours_suborders))], one_col = as.list(tip_colours_suborders), SIMPLIFY = FALSE)
-families <- mapply(add.col, one_order = families[1:27][match(names(families[1:27]), names(tip_colours_families))], one_col = as.list(tip_colours_families), SIMPLIFY = FALSE)
+suborders <- mapply(add.col, one_order = suborders[1:6][match(names(suborders[1:6]), names(tip_colours_passeriformes_suborder))], one_col = as.list(tip_colours_passeriformes_suborder), SIMPLIFY = FALSE)
+families <- mapply(add.col, one_order = families[1:23][match(names(families[1:23]), names(tip_colours_passeriformes_family))], one_col = as.list(tip_colours_passeriformes_family), SIMPLIFY = FALSE)
 

### 2022-08-03 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 
-pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 9, height = 18)
+pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 9, height = 9)
 source("../Functions/ploting.utilities.R")
@@
 plot_table <- rbind(ortho.sort(super_orders),
-                    ortho.sort(orders),
-                    ortho.sort(sub_orders),
-                    ortho.sort(families))
+                    ortho.sort(orders))
+                    # ortho.sort(sub_orders),
+                    # ortho.sort(families))
 colours <- c(rep(gg.color.hue(4)[1], nrow(super_orders)),
-             rep(gg.color.hue(4)[2], nrow(orders)),
-             rep(gg.color.hue(4)[3], nrow(sub_orders)),
-             rep(gg.color.hue(4)[4], nrow(families)))
+             rep(gg.color.hue(4)[2], nrow(orders)))
+             # rep(gg.color.hue(4)[3], nrow(sub_orders)),
+             # rep(gg.color.hue(4)[4], nrow(families)))
 
@@
 }
-add.line <- function(data, counter, shift = 0, total, ...) {
-    lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), ...)
-    points(data[3], (total+1)-counter + shift, pch = 19, ...)
+add.line <- function(data, counter, shift = 0, total, pch = 19, lty = 1, ...) {
+    lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), lty = lty, ...)
+    points(data[3], (total+1)-counter + shift, pch = pch,  ...)
 }
@@
 convert.spaces <- function(x, max) {
-    shifts <- max-nchar(as.character(x))
-    if(shifts > 0) {
-        return(paste(c(x, rep(" ", shifts)), collapse = ""))
-    } else {
+    # shifts <- max-nchar(as.character(x))
+    # if(shifts > 0) {
+    #     return(paste(c(x, rep(" ", shifts)), collapse = ""))
+    # } else {
+    #     return(x)
+    # }
     return(x)
 }
-}
-group_names <- paste0(plot_table$Group, "(n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
+group_names <- paste0(plot_table$Group, " (n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
 
-plot_order <- 1:60
-## Nestedness bit (not finished)
-# plot_order <- integer()
-# ## Super-orders
-# for(i in 1:6) {
-#     plot_order <- c(plot_order, i) 
-#     sub_groups <- which(plot_table$Comparison == plot_table$Group[plot_order[length(plot_order)]])
-#     if(length(sub_groups)) {
-#         plot_order <- c(plot_order, sub_groups)
-#     }
-# }
-# ## Sub-orders
-# for(i in 34:39) {
-#     plot_order <- c(plot_order, i) 
-#     sub_groups <- which(plot_table$Comparison == plot_table$Group[plot_order[length(plot_order)]])
-#     if(length(sub_groups)) {
-#         plot_order <- c(plot_order, sub_groups)
-#     }
-# }
+plot_order <- 1:length(group_names)
 
@@
 orders_n2_plot <- orders_n2[match(orders[order(orders$orthogonality, decreasing = TRUE), ]$Group, orders_n2$Group), ]
-families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
+# families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
 for(i in 1:nrow(orders_n2_plot)) {
-    add.line(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 1/3), total = nrow(plot_table))
-    add.stars(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 1/3), total = nrow(plot_table))
+    add.line(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 2/3), total = nrow(plot_table), lty = 2, pch = 21)
+    add.stars(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 2/3), total = nrow(plot_table))
 }
-for(i in 1:nrow(orders_n2)) {
-    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
-    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
-}
+# for(i in 1:nrow(families_n2)) {
+#     add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
+#     add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
+# }
 
 legend("topleft", legend = c("Super orders (vs.
+phylogeny)", "Orders (vs.
+super orders)", "Orders (vs.
+phylogeny)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
+
+# legend("topleft", legend = c("Super orders (vs.
 phylogeny)", "Orders (vs.
@@
 passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
+
 dev.off()

### 2022-08-04 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
 library(dispRity) #v>=1.7
+library(xtable)
+library(knitr)
+source("../Functions/stats.utilities.R")
+
 ```
@@
 clades <- clades_list[, c(1)]
-levels(clades)[1] <- "noclade"
+levels(clades)[1] <- "no clade"
 
@@
 kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X2")
+to_export <- summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))]
+rownames(to_export)[nrow(to_export)] <- "Entire space"
+silent <- make.xtable(to_export, caption = "Variance per axis and per group in the shape space.
+The \"no clade\" group contains birds not attributed to any of the other groups (i.e.
+from a clade with less than 15 species).", label = "tab_variance_per_axis", path = "../Manuscript/Tables/", include.rownames = TRUE)
 ```

### 2022-08-04 Functions/stats.utilities.R
Update stats.utilities.R

@@ -313,3 +313,68 @@ area.density <- function(data, scale.x = TRUE, scale.y = TRUE) {
     density <- scale.xy(density(data), x = scale.x, y = scale.y)
     area_under_curve <- sum(diff(density$x[order(density$x)]) * zoo::rollmean(density$y[order(density$x)], 2))
 }
+
+
+
+
+#' @title Making the xtable for publication
+#'
+#' @description Making the xtable for publication
+#'
+#' @param table the table to print
+#' @param digit the digit rounding
+#' @param caption the caption
+#' @param label the label (in LaTeX: \ref{label})
+#' @param longtable whether to use longtable
+#' @param path optional, a path for saving the table
+#' @param include.rownames logical, whether to include the rownames (default = FALSE)
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+make.xtable <- function(table, digit = 3, caption, label, longtable = FALSE, path, include.rownames = FALSE) {
+
+    ## Rounding
+    table[ ,which(apply(table, 2, class)  == "numeric")] <- round(table[, which(apply(table, 2, class)  == "numeric")], digits = digit)
+
+    # ## Add significance values
+    # if(all(!is.na(match(colnames(table), c("bhatt.coeff", "statistic", "p.value"))))){
+    #     for(row in 1:nrow(table)) {
+    #         if(as.numeric(table$bhatt.coeff[row]) >= 0.95 || as.numeric(table$bhatt.coeff[row]) <= 0.05) {
+    #             table$bhatt.coeff[row] <- paste0("BOLD",table$bhatt.coeff[row])
+    #         }
+    #         if(as.numeric(table$p.value[row]) <= 0.05) {
+    #             table$p.value[row] <- paste0("BOLD",table$p.value[row])
+    #         }
+    #     }
+    # }
+
+    # ## Bold cells function
+    bold.cells <- function(x) gsub('BOLD(.*)', paste0('\\\\textbf{\\1', '}'), x)
+
+    ## convert into xtable format
+    textable <- xtable(table, digit = digit, caption = caption, label = label)
+
+    # ## Change attributes
+    # if(all(colnames(table) == c("metric", "test", "bhatt.coeff", "statistic", "p.value"))){
+    #     ## Add a horizontal line
+    #     attr(textable, "align")[4] <- "r|"
+    # }
+
+    if(!missing(path)) {
+        if(longtable == TRUE) {
[…]

### 2022-08-04 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+pdf(file = "../Manuscript/Figures/parameters_ESS_all_birds.pdf", height = 9, width = 4)
 diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order[[1]])
+dev.off()
 ```
@@
 ```
+
+```{r, eval = TRUE}
+load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+pdf(file = "../Manuscript/Figures/parameters_ESS_passeriformes.pdf", height = 9, width = 4)
+diagnose.mini.chains(shapespace_passeriformes[[1]])
+dev.off()
+```

### 2022-08-04 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 knitr::kable(all_results)
+```
+
+Median orthogonality for all super-orders and orders to the phylogeny (paragraph 4)
+
+```{r}
+get.angles <- function(X) {
+    unlist(get.disparity(X$angles, concatenate = FALSE))
+}
+## Median angle
+median(get.angles(ellipses_stats$all_birds)) * 90
+## Median angle CI
+quantile(get.angles(ellipses_stats$all_birds), probs = c(0.025, 0.975)) * 90
+```
+
+Median orthogonality for all super-orders and orders to the phylogeny (paragraph 4)
+
+```{r}
+## Median angle
+median(unlist(lapply(ellipses_stats[2:9], get.angles))) * 90
+## Median angle CI
+quantile(unlist(lapply(ellipses_stats[2:9], get.angles)), probs = c(0.025, 0.975)) * 90
 ```

### 2022-08-04 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 ```{r, eval = TRUE, echo = FALSE}
+## Get the bar plots for the loadings of the phylogeny component
+# pdf(file = "../Manuscript/Figures/phylo_loadings.pdf", height = 4, width = 4)
+# plot(NULL, xlim = range(superorder_shapespace$matrix), ylim = range(superorder_shapespace$matrix))
+# add.dims(superorder_shapespace, name.col = c(phylogeny = "grey"), n = 1000)
+# dev.off()
+
 box_size <- 1.5
@@
 ```{r}
+## Adding the bar plot for the phylogeny component
+# pdf(file = "../Manuscript/Figures/phylo_loadings.pdf", height = 4, width = 4)
+# plot(NULL, xlim = range(suborder_shapespace$matrix), ylim = range(suborder_shapespace$matrix))
+# add.dims(suborder_shapespace, name.col = c(phylogeny = "grey"), n = 1000)
+# dev.off()
+
 box_size <- 1.5

### 2022-08-04 Manuscript/Supplementaries.tex
Create Supplementaries.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%======================================================%%
%% to compile with pdflatex/xelatex use pdflatex option %%
%%======================================================%%

%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
\documentclass[sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
%%\documentclass[default]{sn-jnl}% Default
%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout

%%%% Standard Packages
%%<additional latex packages if required can be included here>
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

\jyear{2021}%

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

% Landscape pages for tables
\usepackage{lscape}

\begin{document}

\title[Innovation and elaboration on the avian tree of life]{Innovation and elaboration on the avian tree of life - supplementary materials}

%%=============================================================%%
%% Prefix   -> \pfx{Dr}
%% GivenName    -> \fnm{Joergen W.}
%% Particle -> \spfx{van der} -> surname prefix
%% FamilyName   -> \sur{Ploeg}
%% Suffix   -> \sfx{IV}
%% NatureName   -> \tanm{Poet Laureate} -> Title after name
%% Degrees  -> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Thomas} \sur{Guillerme}}\email{guillert@tcd.ie}

\author[2]{\fnm{Natalie} \sur{Cooper}}\email{natalie.cooper@nhm.ac.uk}

\author[1]{\fnm{Andrew P.} \sur{Beckerman}}\email{a.beckerman@sheffield.ac.uk}

\author[1]{\fnm{Gavin H.} \sur{Thomas}}\email{gavin.thomas@sheffield.ac.uk}

\affil*[1]{\orgdiv{School of Biosciences}, \orgname{University of Sheffield}, \orgaddress{\city{Sheffield}, \postcode{S10 2TN}, \country{UK}}}

\affil[2]{\orgname{Natural History Museum}, \orgaddress{\street{Cromwell Road}, \city{London}, \postcode{SW75BD}, \country{UK}}}

%%==================================%%
%% sample for unstructured abstract %%
%%==================================%%

\abstract{The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.}

%%================================%%
%% Sample for structured abstract %%
%%================================%%

% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}

\keywords{keyword1, Keyword2, Keyword3, Keyword4}

\maketitle

\section{Supplementary material 1}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_cheat_sheet.pdf}
\caption{Relation between the figures in the main text.
The top left panel represents the two first dimensions of the 8 dimensional trait space with all bird beaks represented as grey circles and the Palaeognathes as red ones.
The grey ellipse represents the average posterior variance covariance matrix (VCV) from the general phylogenetic random term of the MCMCglmm model (i.e. the overall phylogenetic axis of main variation).
Species aligned with this axis tend to be elaborators at the mega-evolutionary level and species going away from that line are innovators at the mega-evolutionary level.
The red ellipse and axis represents the average posterior VCV for the Palaeognathes only.
Species position on this axis elaborate and innovate at the macro-evolutionary level.
On the top right panel (Figure 1 in the main text), we measured the elaboration and innovation at the entire clade level (cf. at the species level below).
The ellipses are a scaled and centered in space representation of the group’s VCV ellipse relative to the overall phylogenetic one with a the relative length of the ellipse on each dimension represented on the barplots.
The clade's innovation and elaboration scores are measured by projecting the clade's major axis onto the phylogenetic one and measuring the projection (elaboration) and rejection (innovation) as mentioned above.
The distribution of innovation and elaboration scores comes from the projection of the 4000 pairs of posterior VCVs.
On the bottom right panel (Figure 3 in the main text), we measured the position of each species in terms of elaboration and innovation.
A species elaboration score is their projection onto the phylogenetics' VCV major axis (i.e. their position on that axis) whereas their innovation score is their rejection from it (i.e. their distance away from that axis).}
\label{Fig:cheat_sheet}
\end{figure}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/Supplementary_figure_axis_selection.pdf}
\caption{variance and cumulative variance for each 8 axis per super-order.}
\label{Fig:axes_variance}
\end{figure}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/mini-chains_diagram.pdf}
\caption{mcmcmcglmmm: mini-chains MCMCglmm method diagram}
\label{Fig:fig_mcmcmcglmm}
\end{figure}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/parameters_ESS_all_birds.pdf}
\caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for all birds for the 4000 posteriors. \textbf{solution} is the solution of the model; \textbf{Gcovariance} are the covariances for the random terms (the phylogeny and the clades); \textbf{Rcovariance} are the residuals error term.}
\label{Fig:model_ess_all_birds}
\end{figure}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/parameters_ESS_passeriformes.pdf}
\caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for the passeriformes for the 1870 posteriors. \textbf{solution} is the solution of the model; \textbf{Gcovariance} are the covariances for the random terms (the phylogeny and the clades); \textbf{Rcovariance} are the residuals error term.}
\label{Fig:model_ess_passeriformes}
\end{figure}

%TODO:TG: update the passeriformes parameters estimate (with more models)

\begin{landscape}
\input{Tables/Tab_variance_per_axis.tex}
\end{landscape}

\section{Supplementary material 2: Passeriformes results}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/Supplementary_figure_axis_selection.pdf}
\caption{variance and cumulative variance for each 8 axis per super-order.}
\label{fig_ellipses_passeriformes}
\end{figure}

Figure {#fig_ellipses}: elaboration and innovation at the group level. Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses). We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions). The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse. Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.. The inset boxplots correspond to the elaboration (E) and innovation (I) scores  for all 4000 posterior samples. The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50% and 95% confidence intervals respectively. Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.

Figure {#fig_orthogonality}. Posterior results showing the orientation of phenotypic evolution (orthogonality)of each group compared to their parent clade or to their parent’s parent clade. Orthogonality scales from 0 (modulo of 0°) to 1 (modulo of 90°). For each clade we indicate the number of species (n) and the standard deviation of ellipse orientation within the clade (sd; expressed in degrees). The lines represent the 95% credibility interval of the 4000 variance-covariance ellipses and the dots represent their median orthogonality value. Opaque lines represent the comparisons between the group’s ellipse and their parent clade and semi-transparent lines represent their comparison to their parent’s parent clade. The background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of 45°) and 0.75 (modulo of 67.5°) Note that some clades are both similarly orthogonal to their parent and parent’s parent clade (e.g. Pterocliformes) while some are not (e.g. Columbiformes).

The avian phylogeny (n = 8748 species) coloured by speciesbeak shape innovation and elaboration level. Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration. Innovations and elaboration are displayed as: a)comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands). The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors’ position in that space. A companion plot showing further nested structure within the Passeriformes is shown in Fig {supplementary_figure_passerines}.

\bibliography{sn-bibliography}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl

%% Default %%
%%\input sn-sample-bib.tex%

\end{document}

### 2022-08-04 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Updated figure ellipse fig for passeriformes

@@
 
+## csv results
 write.csv(all_results, "../Data/Processed/ellipses.stats.csv")
+
+## md results
 knitr::kable(all_results)
+
+## tex results
+make.xtable(all_results[, -c(4, 6:8)], digit = 3, caption = "Posterior variance-covariance ellipses results for each clade compared to their parent clade or their parent parents clade (Comparison).
+n = the number of species per group.
+sd = the standard deviation of the ellipses orientation (across the posterior distribution).
+orthogonality = the degree of right angle for each group compared to their parent or parent's group (0 = parellel, 1 = orthogonal).
+Post.
+prob = the posterior probability of the orthogonality in the focal group being different from the comparison one (the 95% CI is from the randomised posterior probabilities).", label = "tab_ortho_results", longtable = FALSE, path = "../Manuscript/Tables/", include.rownames = FALSE)
 ```
@@
 
-pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 9, height = 9)
+pdf("../Manuscript/Figures/orthogonality_results.pdf", width = 9, height = 9)
 source("../Functions/ploting.utilities.R")
@@
 
+```{r, eval = TRUE, echo = FALSE, fig.height = 20, fig.width = 8}
+## Same but for the passeriformes
+pdf("../Manuscript/Figures/orthogonality_results_passeriformes.pdf", width = 9, height = 9)
+source("../Functions/ploting.utilities.R")
+## Plotting table (with the sorted orthogonality)
+plot_table <- rbind(ortho.sort(sub_orders),
+                    ortho.sort(families))
+colours <- c(rep(gg.color.hue(4)[3], nrow(sub_orders)),
+             rep(gg.color.hue(4)[4], nrow(families)))
+group_names <- paste0(plot_table$Group, " (n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
+plot_order <- 1:length(group_names)
+
+## Plot frame
+par(mar = c(5, 17, 4, 2))
+plot(NULL, yaxt = "n", xaxt = "n", ylab = "", xlab = "orthogonality", ylim = c(1,nrow(plot_table)), xlim = c(0,1))
+abline(v = 0.5, lwd = 0.5, col = "grey")
+abline(v = 0.75, lty = 2, col = "grey", lwd = 0.5)
+axis(2, at = nrow(plot_table):1, labels = group_names[plot_order], las = 2)
+axis(1, at = c(0, 0.5, 1), labels = c(0, 0.5, 1))
+## Adding the results
+for(i in 1:nrow(plot_table)) {
+    add.line (plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
+    add.stars(plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
+}
+
+## Adding the n2 results
+families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
+for(i in 1:nrow(families_n2_plot)) {
+    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+nrow(sub_orders), col = adjustcolor(colours[i+nrow(sub_orders)], alpha.f = 2/3), total = nrow(plot_table), lty = 2, pch = 21)
+    add.stars(families_n2_plot[i, ], shift = -1/3, counter = i+nrow(sub_orders), col = adjustcolor(colours[i+nrow(sub_orders)], alpha.f = 2/3), total = nrow(plot_table))
+}
+
+legend("topleft", legend = c("Sub-orders (vs.
+passeriformes)", "Families (vs.
+sub-orders)", "Families (vs.
+passeriformes)"), col = c(gg.color.hue(4)[2], gg.color.hue(4)[3], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
+
+# legend("topleft", legend = c("Super orders (vs.
+phylogeny)", "Orders (vs.
+super orders)", "Orders (vs.
+phylogeny)", "Suborders (vs.
+passeriformes)", "Families (vs.
+suborders)", "Families (vs.
+passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
+
+dev.off()
+```
+
 ```{r, echo = FALSE, eval = TRUE}

### 2022-08-04 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_ellipses.pdf", height = box_size*10, width = box_size*10)
+pdf(file = "../Manuscript/Figures/ellipses.pdf", height = box_size*10, width = box_size*10)
 
@@
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_passeriformes.pdf", height = box_size*9, width = box_size*10)
+pdf(file = "../Manuscript/Figures/ellipses_passeriformes.pdf", height = box_size*9, width = box_size*10)
 

### 2022-08-04 Analysis/01-Data_preparation.Rmd
Cleaned up figure names

@@
 form_axes  <- select.axes(custom.subsets(formspace, group = clades))
+
+pdf(file = "../Manuscript/Figures/axis_selection.pdf", height = 8, width = 12)
+plot(shape_axes)
+dev.off()
 ```

### 2022-08-04 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Cleaned up figure names

@@
 ## tex results
-make.xtable(all_results[, -c(4, 6:8)], digit = 3, caption = "Posterior variance-covariance ellipses results for each clade compared to their parent clade or their parent parents clade (Comparison).
+to_export <- all_results[, -c(4, 6:8)]
+colnames(to_export)[c(6,9)] <- "2.5\\%"
+colnames(to_export)[c(7,10)] <- "97.5\\%"
+make.xtable(to_export, digit = 3, caption = "Posterior variance-covariance ellipses results for each clade compared to their parent clade or their parent parents clade (Comparison).
 n = the number of species per group.
@@
 Post.
-prob = the posterior probability of the orthogonality in the focal group being different from the comparison one (the 95% CI is from the randomised posterior probabilities).", label = "tab_ortho_results", longtable = FALSE, path = "../Manuscript/Tables/", include.rownames = FALSE)
+prob = the posterior probability of the orthogonality in the focal group being different from the comparison one (the 95\\% CI is from the randomised posterior probabilities).", label = "tab_ortho_results", longtable = FALSE, path = "../Manuscript/Tables/", include.rownames = FALSE)
 ```
@@
 
-pdf("../Manuscript/Figures/Relative_EI.pdf", height = 8, width = 8)
+pdf("../Manuscript/Figures/relative_EI.pdf", height = 8, width = 8)
 par(mfrow = c(2,2))

### 2022-08-04 Analysis/06-Figure2_with_correlations.Rmd
Cleaned up figure names

@@
 ## Set the pdf parameters
-pdf("../Manuscript/Figures/Figure2_correlations.pdf", height = 12, width = 8)
+pdf("../Manuscript/Figures/correlations.pdf", height = 12, width = 8)
 
@@
 ## Set the pdf parameters
-pdf("../Manuscript/Figures/Figure2_correlations_passeriformes.pdf", height = 12, width = 6)
+pdf("../Manuscript/Figures/correlations_passeriformes.pdf", height = 12, width = 6)
 

### 2022-08-04 Manuscript/Supplementaries.tex
Update Supplementaries.tex

@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/figure_cheat_sheet.pdf}
+   \includegraphics[width=0.9\textwidth]{Figures/cheat_sheet.pdf}
 \caption{Relation between the figures in the main text.
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/Supplementary_figure_axis_selection.pdf}
+   \includegraphics[width=0.9\textwidth]{Figures/axis_selection.pdf}
 \caption{variance and cumulative variance for each 8 axis per super-order.}
@@
 \begin{landscape}
-\input{Tables/Tab_variance_per_axis.tex}
+\input{Tables/tab_variance_per_axis.tex}
 \end{landscape}
+
+\input{Tables/tab_ortho_results.tex}
 
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/Supplementary_figure_axis_selection.pdf}
-\caption{variance and cumulative variance for each 8 axis per super-order.}
+   \includegraphics[width=1\textwidth]{Figures/ellipses_passeriformes.pdf}
+\caption{Elaboration and innovation at the group level for each passeriformes sub-orders and families. See the caption of figure 1 in the main text for details.}
 \label{fig_ellipses_passeriformes}
@@
 
-Figure {#fig_ellipses}: elaboration and innovation at the group level.
-Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses).
-We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
-The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
-Subsequent inset plots show the results for each superorder.
-Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace..
-The inset boxplots correspond to the elaboration (E) and innovation (I) scores  for all 4000 posterior samples.
-The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50% and 95% confidence intervals respectively.
-Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot.
-Panel B) shows  the results for each order.
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results_passeriformes.pdf}
+\caption{Posterior results showing the orientation of phenotypic evolution (orthogonality) for each passeriformes sub-orders and families. See the caption of figure 2 in the main text for details.}
+\label{fig_orthogonality_passeriformes}
+\end{figure}
 
-Figure {#fig_orthogonality}.
-Posterior results showing the orientation of phenotypic evolution (orthogonality)of each group compared to their parent clade or to their parent’s parent clade.
-Orthogonality scales from 0 (modulo of 0°) to 1 (modulo of 90°).
-For each clade we indicate the number of species (n) and the standard deviation of ellipse orientation within the clade (sd; expressed in degrees).
-The lines represent the 95% credibility interval of the 4000 variance-covariance ellipses and the dots represent their median orthogonality value.
-Opaque lines represent the comparisons between the group’s ellipse and their parent clade and semi-transparent lines represent their comparison to their parent’s parent clade.
-The background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of 45°) and 0.75 (modulo of 67.5°) Note that some clades are both similarly orthogonal to their parent and parent’s parent clade (e.g.
-Pterocliformes) while some are not (e.g.
-Columbiformes).
-
-The avian phylogeny (n = 8748 species) coloured by speciesbeak shape innovation and elaboration level.
-Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
-Innovations and elaboration are displayed as: a)comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
-The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors’ position in that space.
-A companion plot showing further nested structure within the Passeriformes is shown in Fig {supplementary_figure_passerines}.
+\begin{figure}[!htbp]
+\centering
+   % \includegraphics[width=0.9\textwidth]{Figures/.pdf}
+\caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 3 in the main text for details.}
+\label{fig_phylogeny_passeriformes}
+\end{figure}
 

### 2022-08-04 Manuscript/Supplementaries.tex
References not working
Need to make my own template. This one is just shite...

@@
 \usepackage{lscape}
+\usepackage{natbib}
 
@@
 
-\section{Supplementary material 1}
+\section{Supplementary material 1: additional results}
 
@@
 
-\bibliography{sn-bibliography}% common bib file
+\newpage
+
+\section{Supplementary materials 3: detailed projection and rejection operations}
+
+The following section contains the detailed definition and procedure of how we measured elaboration and innovation
+
+We can define the major axis from the variance-covariance matrices and then project and reject each element of interest in the space onto this axis.
+The following steps are generalised to $n$ dimensions and detailed below (as well as the algorithm used in \texttt{dispRity} \citep{dispRity} to perform the transformations):
+
+\subsection{Defining the major axis}
+
+For $O_{n}$, the unit hypersphere matrix of $n$ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:
+
+\begin{equation}
+O_{n} = 
+    \begin{pmatrix}
+        1 & 0 & \cdots & 0 \\
+        0 & 1 & \cdots & 0 \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        0 & 0 & \cdots & 1 \\
+        -1 & 0 & \cdots & 0 \\
+        0 & -1 & \cdots & 0 \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        0 & 0 & \cdots & -1 \\
+    \end{pmatrix}
+\end{equation}
+
+In other words, $O_{n}$ is the matrix representing the hypersphere of $n$ dimensions and of radius $1$ that fits in the centre of the trait-space;
+
+And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:
+
+$$O'_{n} = O_{n} \sqrt{\chi^2(0.95)}$$
+
+Then, for the variance-covariance matrix $VCV_{n}$ of $n$ dimensions obtained from the posterior distribution of the \texttt{mcmcmcglmmmm}:
+
+\begin{equation}
+VCV_{n} = 
+    \begin{pmatrix}
+        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
+        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
+        \vdots  & \vdots  & \ddots & \vdots  \\
+        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
+    \end{pmatrix}
+\end{equation}
+
+and the eigenvectors \textbf{v} and the eigenvalues $\lambda$ satisfying the following eigen decomposition:
+
+$$VCV_{n} \textbf{v} = \lambda \textbf{v}$$
+
+We can get $M_{n}$, the matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors \textbf{v} and the product of the scaled 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:
+
+$$M_{n} = [(O'_{n}\sqrt{\lambda}) \times v]^{\text{T}}$$
+
+Finally, we can centre the matrix $M_{n}$ on the estimated solution of each GLMM (\texttt{model\$Sol}, in \texttt{MCMCglmm} \citep{MCMCglmm}) corresponding to the estimation of the position of the variance-covariance matrix in the trait-space.
+$M_{n}$ then contains all the major axes of the 0.95 hyper-ellipse fitting the variance-covariance matrix.
+We can then define the first row of the matrix, $M_{1,n}$, as the major axis, the second row, $M_{2,n}$, as the second major axis (the minor axis in a 2D ellipse), etc.
+
+The detailed procedure was adapted from \href{Zheyuan Li's}{https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331} post on Stack Overflow and implemented in \texttt{dispRity::axis.covar}.
+
+\subsection{Measuring projection and rejection}
+
+Once we have defined a major axis, we can project any elements in the trait-space onto that axis.
+Specifically, for any elements in the trait-space, we can define it as the vector $\vec{a}$ with one set of coordinates in $n$ dimensions:
+
+\begin{equation}
+    \vec{a} = 
+    \begin{bmatrix}
+    x \\
+    y \\
+    \cdots \\
+    n \\
+    \end{bmatrix}
+\end{equation}
+
+And for any major axis that we can define as a vector $\vec{b}$ as a set of pairs of coordinates in $n$ dimensions:
+
+\begin{equation}
+    \vec{b} = 
+    \begin{bmatrix}
+    x_{1} & x_{2} \\
+    y_{1} & y_{2} \\
+    \cdots & \cdots \\
+    n_{1} & n_{2} \\
+    \end{bmatrix}
+\end{equation}
+
+We can then calculate $\vec{a_{1}}$, the orthogonal projection of $\vec{a}$ onto $\vec{b}$ using:
+
+\begin{equation}
+    \vec{a_{1}} = \frac{\vec{a} \cdot \vec{b}}{\|\vec{b}\|}
+\end{equation}
+
+With $\|\vec{b}\| = \sqrt{\vec{b} \cdot \vec{b}}$ being the norm of $\vec{b}$.
+And $\vec{a_{2}}$, the rejection of $\vec{a}$ onto $\vec{b}$:
+
+\begin{equation}
+    \vec{a_{2}} = \vec{a} - \vec{a_{1}}
+\end{equation}
+
+\subsubsection{Generalisation of projection onto any vector in a set space}
+
+Using this, we can generalise the procedure so as to calculate the projection and rejection for any element within a trait-space $TS_{m,n}$:
+
+\begin{equation}
+    TS_{m,n} = 
+    \begin{bmatrix}
+    x_{1} & x_{2} & \cdots & x_{m} \\
+    y_{1} & y_{2} & \cdots & y_{m} \\
+    \vdots  & \vdots  & \ddots & \vdots \\
+    n_{1} & n_{2} & \cdots & n_{m} \\
+    \end{bmatrix}
+\end{equation}
+
+And any major axis defined as a vector $\vec{B}$:
+
+\begin{equation}
+    B = 
+    \begin{bmatrix}
+    x_{1} & x_{2}\\
+    y_{1} & y_{2}\\
+    \vdots  & \vdots  \\
+    n_{1} & n_{2} \\
+    \end{bmatrix}
+\end{equation}
+
+By using the linear transformation $f_{\vec{B}}$ of the trait-space $TS$ moving $\vec{B}$ onto $TS$'s first axis unit vector $\vec{\hat{\imath}}$:
+
+$$f_{\vec{B}}(TS) = \left( \frac{TS - [Bx_{1}, By_{1}, \cdots, Bn_{1}]^{\text{T}}}{\|\vec{B}\|} \right) \cdot R_{\vec{B}}$$
+
+With $R_{\vec{B}}$ being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:
+
+\begin{equation}
+R_{\vec{B}} = I_{\vec{B}} - \vec{B}\vec{B}^\text{T} - \vec{\hat{\imath}}\vec{\hat{\imath}}^\text{T} + [\vec{B} \vec{\hat{\imath}}]
+    \begin{bmatrix}
+        cos(\theta) & -sin(\theta)\\
+        sin(\theta) & cos(\theta)\\
+    \end{bmatrix} [\vec{B} \vec{\hat{\imath}}]^\text{T}
+\end{equation}
+
+Where $\theta$ is:
+
+\begin{equation}
+    \theta = acos \left(\frac{\vec{B} \cdot \vec{\hat{\imath}}}{\|\vec{B}\| \cdot \|\vec{\hat{\imath}}\|} \right)
+\end{equation}
+
+Or $\theta = acos (B_x)$ since both $\|\vec{B}\|$ and $\|\vec{\hat{\imath}}\|$ are equal to 1 and $\|\vec{\hat{\imath}}\|$ is the unit vector on the first axis.
+
+\subsection{Algorithm for calculating the projection/rejection of any element in a defined space}
+
+In practice we followed this \href{procedure}{https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions} and applied a modification of this  \href{implementation}{https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385} (see \cite{aguilera2004} for the formal generalisation of this algorithm in $n$ dimensions) using the following algorithm implemented in \texttt{dispRity::projections}:
+
+\begin{enumerate}
+ \item In the trait-space, define $\vec{B}$ as the base vector (typically $\vec{B}$ is defined as the pair of coordinates from the major axis described above).
+ \item Centre the trait-space on the origin of $\vec{B}$ so that the first set of coordinates of $\vec{B}$ are 0.
+ \item Scale the trait-space to the norm of $\vec{B}$ so that the norm of $\vec{B}$ is now 1.
+ \item Rotate the trait-space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \rightarrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait-space - typically the x axis unit vector).
+ \item Project/reject every element in the trait-space on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now its projection onto $\vec{B}$.
+\end{enumerate}
+
+\bibliographystyle{sysbio}
+\bibliography{References}% common bib file
 %% if required, the content of .bbl file can be included here once bbl is generated

### 2022-08-05 Manuscript/main.tex
Updated manuscript and supplementaries

\documentclass[12pt,letterpaper]{article}

%Packages
% \usepackage{textcomp}
% \usepackage{latexsym}
% \usepackage{url}
% \usepackage{epsfig}
% \usepackage{graphicx}
% \usepackage{amssymb}
% \usepackage{amsmath}
% \usepackage{mathtools}
% \usepackage{bm}
% \usepackage{array}
% \usepackage[version=3]{mhchem}
% \usepackage{ifthen}
% \usepackage{caption}
% \usepackage{amsthm}
% \usepackage{amstext}
% \usepackage{enumerate}
% \usepackage[osf]{mathpazo}
% \usepackage{dcolumn}

\usepackage{hyperref}
\usepackage{lineno}
\usepackage{pdflscape}
\usepackage{mathtools}
\usepackage[osf]{mathpazo}
\usepackage{fullpage}
\usepackage{float}

\pagenumbering{arabic}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}
%Running head
\begin{flushright}
Version dated: \today
\end{flushright}

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf Innovation and elaboration on the avian tree of life}
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1}$}\\
\noindent {\small \it 
$^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}

\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\ 
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\noindent (Keywords: )\\

\vspace{1.5in}

\newpage

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
 blabalbal \cite{dispRity}

\begin{figure}
\centering
   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
\caption{.}
\label{Fig:ellipses}
\end{figure}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
\caption{.}
\label{Fig:orthogonality}
\end{figure}

\begin{figure}[!htbp]
\centering
   % \includegraphics[width=0.9\textwidth]{Figures/phylogenys.pdf}
\caption{.}
\label{Fig:phylogeny}
\end{figure}

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/relative_EI.pdf}
\caption{.}
\label{Fig:relative_EI}
\end{figure}

\bibliographystyle{naturemag}
\bibliography{References}

\end{document}

### 2022-08-05 Manuscript/supplementaries.tex
Updated manuscript and supplementaries

--- 2022-08-04 Manuscript/Supplementaries.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[12pt,letterpaper]{article}
 
-%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing
+%Packages
+% \usepackage{textcomp}
+% \usepackage{latexsym}
+% \usepackage{url}
+% \usepackage{epsfig}
+% \usepackage{graphicx}
+% \usepackage{amssymb}
+% \usepackage{amsmath}
+% \usepackage{mathtools}
+% \usepackage{bm}
+% \usepackage{array}
+% \usepackage[version=3]{mhchem}
+% \usepackage{ifthen}
+% \usepackage{caption}
+% \usepackage{amsthm}
+% \usepackage{amstext}
+% \usepackage{enumerate}
+% \usepackage[osf]{mathpazo}
+% \usepackage{dcolumn}
 
-%%=======================================================%%
-%% to print line numbers in the margin use lineno option %%
-%%=======================================================%%
+\usepackage{hyperref}
+\usepackage{lineno}
+\usepackage{pdflscape}
+\usepackage{mathtools}
+\usepackage[osf]{mathpazo}
+\usepackage{fullpage}
+% \usepackage{float}
 
-%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+\pagenumbering{arabic}
 
-%%======================================================%%
-%% to compile with pdflatex/xelatex use pdflatex option %%
-%%======================================================%%
-
-%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
-
-%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
-\documentclass[sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
-%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
-%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
-%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style
-%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
-%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
-%%\documentclass[default]{sn-jnl}% Default
-%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
-
-%%%% Standard Packages
-%%<additional latex packages if required can be included here>
-%%%%
-
-%%%%%=============================================================================%%%%
-%%%%  Remarks: This template is provided to aid authors with the preparation
-%%%%  of original research articles intended for submission to journals published 
-%%%%  by Springer Nature. The guidance has been prepared in partnership with 
-%%%%  production teams to conform to Springer Nature technical requirements. 
-%%%%  Editorial and presentation requirements differ among journal portfolios and 
-%%%%  research disciplines. You may find sections in this template are irrelevant 
-%%%%  to your work and are empowered to omit any such section if allowed by the 
-%%%%  journal you intend to submit to. The submission guidelines and policies 
-%%%%  of the journal take precedence. A detailed User Manual is available in the 
-%%%%  template package for technical guidance.
-%%%%%=============================================================================%%%%
-
-\jyear{2021}%
-
-%% as per the requirement new theorem styles can be included as shown below
-\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}% 
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}%
-
-\raggedbottom
-%%\unnumbered% uncomment this for unnumbered level heads
-
-% Landscape pages for tables
-\usepackage{lscape}
-\usepackage{natbib}
+%---------------------------------------------
+%
+%       START
+%
+%---------------------------------------------
 
 \begin{document}
+%Running head
+\begin{flushright}
+Version dated: \today
+\end{flushright}
 
-\title[Innovation and elaboration on the avian tree of life]{Innovation and elaboration on the avian tree of life - supplementary materials}
+\bigskip
+\medskip
+\begin{center}
 
-%%=============================================================%%
-%% Prefix   -> \pfx{Dr}
-%% GivenName    -> \fnm{Joergen W.}
-%% Particle -> \spfx{van der} -> surname prefix
-%% FamilyName   -> \sur{Ploeg}
-%% Suffix   -> \sfx{IV}
-%% NatureName   -> \tanm{Poet Laureate} -> Title after name
-%% Degrees  -> \dgr{MSc, PhD}
-%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
-%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
-%%=============================================================%%
+\noindent{\Large \bf Innovation and elaboration on the avian tree of life - supplementary materials}
+\bigskip
 
-\author*[1]{\fnm{Thomas} \sur{Guillerme}}\email{guillert@tcd.ie}
+\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1}$}\\
+\noindent {\small \it 
+$^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
+$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}
 
-\author[2]{\fnm{Natalie} \sur{Cooper}}\email{natalie.cooper@nhm.ac.uk}
+\end{center}
+\medskip
+\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\ 
+\vspace{1in}
 
-\author[1]{\fnm{Andrew P.} \sur{Beckerman}}\email{a.beckerman@sheffield.ac.uk}
+%Line numbering
+\modulolinenumbers[1]
+\linenumbers
 
-\author[1]{\fnm{Gavin H.} \sur{Thomas}}\email{gavin.thomas@sheffield.ac.uk}
+\noindent (Keywords: )\\
 
-\affil*[1]{\orgdiv{School of Biosciences}, \orgname{University of Sheffield}, \orgaddress{\city{Sheffield}, \postcode{S10 2TN}, \country{UK}}}
+\vspace{1.5in}
 
-\affil[2]{\orgname{Natural History Museum}, \orgaddress{\street{Cromwell Road}, \city{London}, \postcode{SW75BD}, \country{UK}}}
+\newpage
 
-%%==================================%%
-%% sample for unstructured abstract %%
-%%==================================%%
-
-\abstract{The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.}
-
-%%================================%%
-%% Sample for structured abstract %%
-%%================================%%
-
-% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
-As a guide the abstract should not exceed 200 words.
-Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+%---------------------------------------------
 %
-% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
-As a guide the abstract should not exceed 200 words.
-Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+%       INTRODUCTION
 %
-% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
-As a guide the abstract should not exceed 200 words.
-Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
-% 
-% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
-As a guide the abstract should not exceed 200 words.
-Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}
-
-\keywords{keyword1, Keyword2, Keyword3, Keyword4}
-
-\maketitle
+%---------------------------------------------
 
@@
    \includegraphics[width=0.9\textwidth]{Figures/cheat_sheet.pdf}
-\caption{Relation between the figures in the main text.
+\caption{\scriptsize Relation between the figures in the main text.
 The top left panel represents the two first dimensions of the 8 dimensional trait space with all bird beaks represented as grey circles and the Palaeognathes as red ones.
@@
 
-%TODO:TG: update the passeriformes parameters estimate (with more models)
-
 \begin{landscape}
@@
 We can define the major axis from the variance-covariance matrices and then project and reject each element of interest in the space onto this axis.
-The following steps are generalised to $n$ dimensions and detailed below (as well as the algorithm used in \texttt{dispRity} \citep{dispRity} to perform the transformations):
+The following steps are generalised to $n$ dimensions and detailed below (as well as the algorithm used in \texttt{dispRity} \cite{dispRity} to perform the transformations):
 
@@
 
-Finally, we can centre the matrix $M_{n}$ on the estimated solution of each GLMM (\texttt{model\$Sol}, in \texttt{MCMCglmm} \citep{MCMCglmm}) corresponding to the estimation of the position of the variance-covariance matrix in the trait-space.
+Finally, we can centre the matrix $M_{n}$ on the estimated solution of each GLMM (\texttt{model\$Sol}, in \texttt{MCMCglmm} \cite{MCMCglmm}) corresponding to the estimation of the position of the variance-covariance matrix in the trait-space.
 $M_{n}$ then contains all the major axes of the 0.95 hyper-ellipse fitting the variance-covariance matrix.
@@
 
-The detailed procedure was adapted from \href{Zheyuan Li's}{https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331} post on Stack Overflow and implemented in \texttt{dispRity::axis.covar}.
+The detailed procedure was adapted from \href{https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331}{Zheyuan Li's} post on Stack Overflow and implemented in \texttt{dispRity::axis.covar}.
 
@@
 
-In practice we followed this \href{procedure}{https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions} and applied a modification of this  \href{implementation}{https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385} (see \cite{aguilera2004} for the formal generalisation of this algorithm in $n$ dimensions) using the following algorithm implemented in \texttt{dispRity::projections}:
+In practice we followed \href{https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions}{this procedure} and applied a modification of \href{https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385}{this implementation} (see \cite{aguilera2004} for the formal generalisation of this algorithm in $n$ dimensions) using the following algorithm implemented in \texttt{dispRity::projections}:
 
@@
 
-\bibliographystyle{sysbio}
-\bibliography{References}% common bib file
-%% if required, the content of .bbl file can be included here once bbl is generated
-%%\input sn-article.bbl
-
-%% Default %%
-%%\input sn-sample-bib.tex%
+\bibliographystyle{naturemag}
+\bibliography{References}
 

### 2022-08-05 Manuscript/main.tex
updated up to paragraph 4

@@
 \usepackage{float}
+\usepackage{xr} %linking to supplementaries
+\externaldocument{supplementaries}
 
@@
 
+\begin{abstract}
+Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
+A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
+At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
+The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance.
+Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+Our analyses show that elaboration is a common feature at all scales, consistent with theory.
+We also find that innovation is a common and major contributor to avian morphological diversity among clades.
+Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
+These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+\end{abstract}
+
 \vspace{1.5in}
@@
 
-%---------------------------------------------
-%
-%       INTRODUCTION
-%
-%---------------------------------------------
- blabalbal \cite{dispRity}
+Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow lines of genetic least resistance \cite{schluter1996adaptive}.
+Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape.
+Indeed, evidence suggests that genetic constraints can extend over timescales spanning tens of millions of years leading to remarkable macroevolutionary stability in the orientation of phenotypic evolution \cite{mcglothlin2018adaptive}.
+However, if most organisms evolve incrementally along stable lines of least resistance then how can we explain the staggering diversity of species morphologies across the tree of life?
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the evolvability of phenotypes across scales.
+This apparent tension implies deep-time reorientation of phenotypic lines of least resistance and points to alternative routes to the exploration of phenotypic space by evolving either along, or away from, the dominant phenotypic axis.
+We refer to these alternatives as elaboration - evolution along the phylogenetically constrained phenotypic major axis - and innovation - evolution away from the phylogenetically constrained phenotypic major axis.
+The contributions of these two multivariate pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales.
+Here we estimate and compare major axes of beak shape evolution in birds across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.
+
+Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
+1) large, multivariate datasets to allow exploration of trait covariances at different scales,
+2) reliable and efficient computational methods to estimate the lines of least resistance in a nested fashion,
+and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
+Here, we propose a novel analytical pipeline for measuring elaboration and innovation at the species- and higher taxonomic levels (Fig S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
+All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
+We fitted a phylogenetic generalised linear mixed effect model (pGLMM with beak shape as an eight-dimensional response variable andtreated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods).
+We regard the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
+The nested partitions of multivariate trait space provide the basis for all subsequent analyses.
+
+Adaptive radiation theory implies that phenotypic divergence arises primarily along a dominant phenotypic axis (the line of least evolutionary resistance; \cite{marroig2005size}).
+However, this contrasts with observations of major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
+We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different groups compared to the global phylogenetic major axis.
+We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis (Fig.
+\ref{Fig:ellipses} - Fig.
+\ref{Fig:orthogonality}).
+Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis and each super order (Fig.
+\ref{Fig:ellipses}a) and order (Fig.
+\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Figure \ref{Fig:orthogonality}).
+The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ (95\% CI: $22.83^\circ$-$89.09^\circ$).
+Comparisons of orientation among subclades (i.e.
+orders within superorders) show similar differences (median = $47.17^\circ$; 95\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
+This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
+These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
 
@@
 
+\section{Methods}
+
 \bibliographystyle{naturemag}
-\bibliography{References}
+\bibliography{references}
 

### 2022-08-05 Manuscript/main.tex
main text up

@@
 % \usepackage{url}
-% \usepackage{epsfig}
-% \usepackage{graphicx}
 % \usepackage{amssymb}
@@
 % \usepackage{ifthen}
-% \usepackage{caption}
 % \usepackage{amsthm}
@@
 % \usepackage{enumerate}
-% \usepackage[osf]{mathpazo}
 % \usepackage{dcolumn}
 
+\usepackage{epsfig}
+\usepackage{graphicx}
+\usepackage{caption}
 \usepackage{hyperref}
@@
 \noindent (Keywords: )\\
-
-\begin{abstract}
-Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
+P
+atterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
 A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
@@
 These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
-\end{abstract}
-
-\vspace{1.5in}
-
-\newpage
 
@@
 and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
-Here, we propose a novel analytical pipeline for measuring elaboration and innovation at the species- and higher taxonomic levels (Fig S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
+Here, we propose a novel analytical pipeline for measuring elaboration and innovation at the species- and higher taxonomic levels (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
 All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
@@
 \ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Figure \ref{Fig:orthogonality}).
-The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ (95\% CI: $22.83^\circ$-$89.09^\circ$).
+The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
 Comparisons of orientation among subclades (i.e.
-orders within superorders) show similar differences (median = $47.17^\circ$; 95\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
+orders within superorders) show similar differences (median = $47.17^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
 This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
@@
 
-\begin{figure}
+\begin{figure}[!htbp]
 \centering
@@
 \end{figure}
+
+\noindent \textbf{Figure \ref{Fig:ellipses}:}
+Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
+We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
+Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
+The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
+Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
 
@@
 
+\noindent \textbf{Figure \ref{Fig:orthogonality}:}
+Orthogonality of each clade's major axis compared to their parent or parent parent's clade.
+Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+Dots represent the clades median orthogonality and the lines their 95\% CI.
+No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at keast half orthogonal to their super-order (green lines and circles).
+These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
+Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
+The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp > 0.99; ** = pp > 0.95; * = pp > 0.9; . = pp > 0.8).
+Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp > 0.99.
+This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+
+However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
+While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
+To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
+As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis.
+Clades (or species) can be both elaborators and innovators.
+To measure innovation and elaboration we translated the major axes of each group's (i.e.
+superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
+We then used multilinear algebra to project a focal group's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheett}).
+We interpret the measured projection (distance onto the phylogenetic major axis) as the group's elaboration score and the measured rejection (distance from the major axis) as the group's innovation score.
+These scores were measured for each of 4000 posterior pairs of group vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
+We found that for most groups the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
+Remarkably, more than half of the assessed groups (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
+However, there is substantial variation in terms of elaboration and innovation across groups including relatively low elaboration and innovation (e.g.
+Cuculiformes; Fig.
+\ref{Fig:ellipses}), low elaboration and high innovation (e.g.
+Galliformes; Fig.
+\ref{Fig:ellipses}), high elaboration and low innovation (e.g.
+Coraciiformes; Fig.
+\ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g.
+Podicipediformes; Fig.
+\ref{Fig:ellipses}).
+These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig.S\ref{fig_ellipses_passeriformes}).
+This suggests that divergence away from line of least evolutionary resistance (i.e.
+the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
+
+This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
+Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), align with PC1 in shapespace Fig.
+\ref{Fig:ellipses}-bar plots).
+These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
+S\ref{fig_ellipses_passeriformes}).
+Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
+For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
+However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
+
+Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
+This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
+To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shape-space onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
+We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire groups.
+We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
+However, we also found various combinations of elaboration and innovation across birds.
+For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
+Other groups such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
+Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: add to supp?
+The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
+\ref{Fig:phylogeny} and ),%{#fig_correlations}),%TODO: add to supp?
+ and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
+The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
+
 \begin{figure}[!htbp]
@@
 \end{figure}
+\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by speciesbeak shape innovation and elaboration level.
+Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
+Innovations and elaboration are displayed as: a)comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
+The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors’ position in that space.
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
+
+The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
+We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
+We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
+We then measured the difference between both areas and found that there was a significant difference in innovation between species and groups but no clear difference in elaboration (Fig.
+\ref{Fig:relative_EI}).
+This difference confirms our observation that innovation is more common at the group level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
+However, we suggest that rather than a paradox the reorientation of trait space can instead be described as a gradual process occuring in response to shifting selection pressures acting on species evolving in similar adaptive zones.
+This implies that gradual directional evolution \cite{pagel2022general} rather than megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple} may be sufficient to explain diversity in avian beak morphology.
 
@@
 
+\noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among groups (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
+A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 groups.
+The scores for each group is scaled by the maximum score within each group (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
+This allows us to compare both types (species vs.
+groups) despite their different sample sizes.
+B - same as A but for the innovation score.
+C - density curve of the area under the curve for each posterior sample relative elaboration score as a proxy for the overall amount of elaboration per sample and per type (species or group - respectively blues and orange).
+The Bhattacharyya coefficient indicates the probability of overlap between the posterior distributions.
+D - same as C but for innovation.
+We can see a clear difference in the amount of innovation across the posterior samples between species or groups.
+
+Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation along the megaevolutionary line of least resistance is much more common.
+This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
+However, individual specie-level variation in the past could also have led to a shift in a group’s line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape, Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
+Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
+
 \section{Methods}

### 2022-08-05 Manuscript/main.tex
Full text first pass

@@
 Innovations and elaboration are displayed as: a)comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
-The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors’ position in that space.
+The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
@@
 This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
-However, individual specie-level variation in the past could also have led to a shift in a group’s line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape, Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
+However, individual specie-level variation in the past could also have led to a shift in a group's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape, Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
 Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
@@
 \section{Methods}
+
+\textbf{Beak shape data.}
+We used the dataset from \cite{cooney2017mega,chira2020signature,hughes2022global} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
+The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,chira2020signature,hughes2022global} for detailed descriptions of data collection and processing).
+We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the first eight first dimensions of the resulting PCA.
+These eight dimensions contained at least 95\% of variance in each superorder, and 98.7\% of the variance in the whole dataset.
+We used a threshold of $>95$\% of variance explained for each superorder because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
+$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each superorder.
+For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
+This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each superorder (see Fig.
+S\ref{Fig:axes_variance} and Table \ref{tab_variance_per_axis}).
+This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
+
+\textbf{Phylogenetic data.}
+We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
+We pruned these trees to contain only the 8748 species present in our beak shapespace.
+
+\textbf{Phylogenetic GLMM analysis.}
+We ran a phylogenetic Bayesian generalised linear mixed effects model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
+
+\begin{equation}
+\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)} (equation 1)
+\end{equation}
+
+This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself.
+These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
+We ran two separate nested models.
+One model using one random term for the whole phylogeny and 35 nested random terms for each superorder and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
+And one model on a subset of the dataset containing only the 5966 passeriformes species with 30 nested random terms.
+This resulted in two multidimensional GLMM.
+One with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
+And one with a $5966 \times 8$ response variable, one global residual term and 30 random terms.
+
+To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
+Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2.1 CPU years and 4TB of RAM for the whole tree model and @@@ for the passeriformes model).
+%TODO: passeriformes model size
+Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package mcmcmglmmm @@@(cite?).%TODO: cite mcmcmcglmmm
+
+\textbf{mcmcmcglmm.}
+This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
+This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
+1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
+2) extracting the following parameters from the parameterisation chains:
+a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
+b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
+We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
+This effectively resulted in getting 10 exploitable posterior samples per tree.
+We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
+S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
+Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
+The whole procedure is reproducible in % TODO: add vignette @@@vignette name and implemented in the mcmcmcglmmm standalone package https://github.com/TGuillerme/mcmcmcglmmm.
+
+\textbf{Projection/rejection analyses.}
+We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
+In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group (see Fig.
+S).
+%TODO: add figure projection
+In brief we can interpret where a species or a group falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
+We calculated the elaboration and innovation scores using two main approaches:
+1) by clades (i.e.
+to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes (Fig.
+\ref{Fig:ellipses});
+and 2) by species (i.e.
+to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes and b) the phylogenetic major axes of their respective clade (Fig.
+\ref{Fig:phylogeny}).
+To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
+This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
+The entire mathematical procedure is described in details in % TODO: link to: @@@supplementary text@@@.
+The whole procedure applied to our dataset is reproducible in % TODO: link to: @@@vignette name
+and implemented in the dispRity package \cite{dispRity}.
+
+\textbf{Groups' orthogonality.}
+We measured the orthogonality of each group (i.e. random term in our models compared to their parent group and their parent parent's (e.g.
+the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).To do so we measured the angles between the major axes of the focal group and their parent group for each posterior variance-covariance matrix as well as within each group by randomly comparing major axes within a group.
+We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
+We then measured the posterior probability of the angles within a group being different than the angles between groups.
+These results are available in Fig. \ref{Fig:orthogonality}.
 

### 2022-08-08 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 
+## Get the median elaboration/innovation for the groups
+groups_median_EI <- list(
+    "super_orders" = list("elaboration" = unlist(lapply(get.disparity(results$super_results_between$position, concatenate = FALSE), function(X) median(c(X)))),
+                          "innovation"  = unlist(lapply(get.disparity(results$super_results_between$distance, concatenate = FALSE), function(X) median(c(X))))) ,
+    "orders"       = list("elaboration" = unlist(lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), function(X) median(c(X)))),
+                          "innovation"  = unlist(lapply(get.disparity(results$order_results_between$distance, concatenate = FALSE), function(X) median(c(X)))))
+)
+save(groups_median_EI, file = "../Data/Processed/groups_median_EI.rda")
 ```
@@
 save(clade_info_passeriformes, file = "../Data/Processed/clade_info_passeriformes.rda")
+
+## Get the median elaboration/innovation for the groups
+groups_median_EI_passeriformes <- list(
+    "sub_orders" = list("elaboration" = unlist(lapply(get.disparity(results$suborder_results_between$position, concatenate = FALSE), function(X) median(c(X)))),
+                          "innovation"  = unlist(lapply(get.disparity(results$suborder_results_between$distance, concatenate = FALSE), function(X) median(c(X))))) ,
+    "families"       = list("elaboration" = unlist(lapply(get.disparity(results$family_results_between$position, concatenate = FALSE), function(X) median(c(X)))),
+                          "innovation"  = unlist(lapply(get.disparity(results$family_results_between$distance, concatenate = FALSE), function(X) median(c(X)))))
+)
+save(groups_median_EI_passeriformes, file = "../Data/Processed/groups_median_EI_passeriformes.rda")
 ```

### 2022-08-08 Manuscript/main.tex
manuscript up (NOW MOVING TO OVERLEAF

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1}$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
 \noindent {\small \it 
 $^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}
+$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\
+$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
 
@@
 \noindent (Keywords: )\\
-P
-atterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
+
+Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
 A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
@@
 Other groups such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
-Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: add to supp?
+Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
 The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
-\ref{Fig:phylogeny} and ),%{#fig_correlations}),%TODO: add to supp?
+\ref{Fig:phylogeny}).
+%{#fig_correlations}),%TODO: maybe add to supp?
  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
@@
 To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
-Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2.1 CPU years and 4TB of RAM for the whole tree model and @@@ for the passeriformes model).
-%TODO: passeriformes model size
-Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package mcmcmglmmm @@@(cite?).%TODO: cite mcmcmcglmmm
+Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the whole tree and the passeriformes ones).
+Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
 
@@
 Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
-The whole procedure is reproducible in % TODO: add vignette @@@vignette name and implemented in the mcmcmcglmmm standalone package https://github.com/TGuillerme/mcmcmcglmmm.
+The whole procedure is reproducible in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
 
@@
 We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
-In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group (see Fig.
-S).
-%TODO: add figure projection
+In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group.
 In brief we can interpret where a species or a group falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
@@
 This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
-The entire mathematical procedure is described in details in % TODO: link to: @@@supplementary text@@@.
-The whole procedure applied to our dataset is reproducible in % TODO: link to: @@@vignette name
+The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
+The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/MCMCglmm/inst/vignettes/model_test_vignette.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
 and implemented in the dispRity package \cite{dispRity}.

### 2022-08-12 Analysis/05-Figure_with_phylogeny.Rmd
Added phylo figure repro

---
title: "Bird beak's elaboration and innovation"
author: "Gavin Thomas"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

## Plot the median elaboration/innovation on the phylogeny

```{r, echo = FALSE}
library(ggtree)
library(ggtreeExtra)
library(treeio)
library(treeplyr)
library(ggplot2)
library(phytools)
library(motmot)
library(viridis)
library(RColorBrewer)
library(colorspace)
library(ggnewscale)

### All birds plot - main text

# Load data for all phylo, superorder, and order comparisons
load("../Data/Processed/ploting_data.rda")
load("../Data/Processed/ploting_data_superorder.rda")
load("../Data/Processed/ploting_data_order.rda")

# Get clade info data (includes colours for clade labelling)
load("../Data/Processed/clade_info.rda")
load("../Data/Processed/tip_colours_orders.rda")

tree <- ploting_data$consensus_tree
tree <- ladderize(tree)

# Set up data frame - with some missing data will need to use full_join
beak_all <-  data.frame(species = names(ploting_data$median_elaborations), elab=ploting_data$median_elaborations, inov=ploting_data$median_innovations, phylogeny=ploting_data$distances_to_centre)
beak_superorder <- data.frame(species = names(ploting_data_superorder$median_elaborations), elabsuperorder=ploting_data_superorder$median_elaborations,  inovsuperorder=ploting_data_superorder$median_innovations)
beak_order <- data.frame(species = names(ploting_data_order$median_elaborations), elaborder=ploting_data_order$median_elaborations,  inovorder=ploting_data_order$median_innovations)

beak <- left_join(beak_all, beak_superorder, by="species")
beak <- left_join(beak, beak_order, by="species")

# Combine data and tree to plot estimated ancestral states for distance to centroid per species
beak_tree <- make.treedata(tree=tree, data= beak)
dcen_plot <- getVector(beak_tree, phylogeny)
fit <- phytools::fastAnc(beak_tree$phy, dcen_plot, vars=TRUE,CI=FALSE)

# Combine ancestral state estimates with tip values and names 
td <- data.frame(node = ggtree::nodeid(beak_tree$phy, names(dcen_plot)),
               dist_cent = dcen_plot)
nd <- data.frame(node = names(fit$ace), dist_cent = fit$ace)
d <- rbind(td, nd)
d$node <- as.numeric(d$node)

# Join data frame with tree
birdelabtree <- full_join(beak_tree$phy, d, by = "node")
birdelabtree@phylo <- ladderize(birdelabtree@phylo)

# The basic plotting call (1) including angle (5) 
innovelab_plot <- ggtree(birdelabtree, aes(color= dist_cent), layout="fan",
                      continuous = TRUE, size=0.5, ladderize=TRUE, open.angle=20) +
  scale_color_gradientn(colours=cividis(100))  +
  theme(legend.position="none", legend.box="horizontal")

# Set up data in long format with fudge to get variable names to plot correctly
heatmapdat <- beak[,c(1:3,5:8)]
colnames(heatmapdat) <- c("species", "Elaboration (all)", "Innovation (all)", 
                      "Elab (superorder)", "Innov (superorder)",
                      "Elab (order)", "Innov (order)")

heatmapdat <- reshape2::melt(heatmapdat, id.vars="species", variable.name = "type", value.name="score", factorsAsStrings=FALSE)

# Make data frame with node labels and node ids
n <- length(clade_info[[2]])
order_labels <- data.frame(order_label=rep(NA, n), node=rep(NA, n), colour_bar=rep(NA, n))

for (i in 1:n) {
  order_labels[i,1] <- names(clade_info[[2]][i])
  order_labels[i,2] <- MRCA(birdelabtree, clade_info[[2]][[i]][[1]])
}

# Order data frame alphabetically according to the taxon labels
# Necessary because geom_cladelabel adds labels alphabetically but plots colours 
# in the order of the provided vector (alphatising ensures plotting order is correct)
order_labels <-arrange(order_labels, order_label)
order_labels[,3] <- arrange(data.frame(colour=tip_colours_orders, orders=names(tip_colours_orders)), orders)[,1]

## Second part of plot - add heatmap for elaboration and innovation scores
innovelab_plot <- innovelab_plot +
  new_scale_fill() + 
  geom_fruit(data=heatmapdat, geom=geom_tile, 
             mapping=aes(y=species, x=type, fill=score),
             axis.params=list(
               axis="x",
               text.angle = -90,
               text.size = 4.5,
               line.size = 0,
               vjust = 0.5,
               hjust=0),
             offset = 0.02,size = 0.02) +
  scale_fill_continuous_divergingx(palette = 'Spectral', rev=TRUE,
                                   mid = 1, na.value="grey80") + 
  theme(legend.position=c(0.52,0.43), legend.key.size = unit(0.5, "cm"), 
        legend.key.width=unit(0.6, "cm")) + 
  labs(fill= "Elab & Innov") +
  labs(colour= "Distance")

## Third part of plot - add clade labels
innovelab_plot <- innovelab_plot +
  new_scale_colour() + 
  geom_cladelab(data=order_labels, 
                mapping=aes(node=node, 
                            label=order_label,
                            colour=order_label),
                textcolour="black",
                hjust=0,
                angle="auto",
                barsize=4,
                horizontal=TRUE, 
                fontsize=4,
                offset=30,
                offset.text=2,
                show.legend=FALSE
  ) +
  scale_colour_manual(
    values=order_labels$colour_bar, 
    guide="none"
    )

## Save as png and pdf
ggsave("../Manuscript/Figures/InnovElabTree_main_text.png", innovelab_plot_final, width=12, height=12)
ggsave("../Manuscript/Figures/InnovElabTree_main_text.pdf", innovelab_plot_final, width=12, height=12)

#####################################################################
#####################################################################

### Passerines supplementary plot

# Load data for all phylo, superorder, and order comparisons
load("../Data/Processed/ploting_data_passeriformes.rda")
load("../Data/Processed/ploting_data_passeriformes_suborder.rda")
load("../Data/Processed/ploting_data_passeriformes_family.rda")

# Get clade info data (includes colours for clade labelling)
load("../Data/Processed/clade_info_passeriformes.rda")
load("../Data/Processed/tip_colours_passeriformes_family.rda")

clades <- read.csv("../Data/Processed/BirdClades.csv")
passerines <- clades$TipLabel[which(clades$level2_order=="Passeriformes")]

tree_pass <- ploting_data_passeriformes$consensus_tree
tree_pass <- ladderize(tree_pass)

# Set up data frame - with some missing data will need to use full_join
beak_pass <-  data.frame(species = names(ploting_data_passeriformes$median_elaborations), elab=ploting_data_passeriformes$median_elaborations, inov=ploting_data_passeriformes$median_innovations, phylogeny=ploting_data_passeriformes$distances_to_centre)
beak_pass_suborder <- data.frame(species = names(ploting_data_passeriformes_suborder$median_elaborations), elabsuperorder=ploting_data_passeriformes_suborder$median_elaborations,  inovsuperorder=ploting_data_passeriformes_suborder$median_innovations)
beak_pass_fam <- data.frame(species = names(ploting_data_passeriformes_family$median_elaborations), elaborder=ploting_data_passeriformes_family$median_elaborations,  inovorder=ploting_data_passeriformes_family$median_innovations)

beak_pass <- left_join(beak_pass, beak_pass_suborder, by="species")
beak_pass <- left_join(beak_pass, beak_pass_fam, by="species")

# Ensure only passerines included
beak_pass <- subset(beak_pass, species %in% passerines)

# Combine data and tree to plot estimated ancestral states for distance to centroid per species
beak_tree_pass <- make.treedata(tree=tree_pass, data= beak_pass)
dcen_plot_pass <- getVector(beak_tree_pass, phylogeny)
fit_pass <- phytools::fastAnc(beak_tree_pass$phy, dcen_plot_pass, vars=TRUE,CI=FALSE)

# Combine ancestral state estimates with tip values and names 
td_pass <- data.frame(node = nodeid(beak_tree_pass$phy, names(dcen_plot_pass)),
                 dist_cent = dcen_plot_pass)
nd_pass <- data.frame(node = names(fit_pass$ace), dist_cent = fit_pass$ace)
d_pass <- rbind(td_pass, nd_pass)
d_pass$node <- as.numeric(d_pass$node)

# Join data frame with tree
birdelabtree_pass <- full_join(beak_tree_pass$phy, d_pass, by = "node")
birdelabtree_pass@phylo <- ladderize(birdelabtree_pass@phylo)

# The basic plotting call (1) including angle (5) 
innovelab_pass_plot <- ggtree(birdelabtree_pass, aes(color= dist_cent), layout="fan",
                         continuous = TRUE, size=0.5, ladderize=TRUE, open.angle=25) +
  scale_color_gradientn(colours=cividis(100))  +
  theme(legend.position="none", legend.box="horizontal")

# Set up data in long format with fudge to get variable names to plot correctly
heatmapdat_pass <- beak_pass[,c(1:3,5:8)]
colnames(heatmapdat_pass) <- c("species", "Elaboration (all)", "Innovation (all)", 
                          "Elab (suborder)", "Innov (suborder)",
                          "Elab (family)", "Innov (family)")

heatmapdat_pass <- reshape2::melt(heatmapdat_pass, id.vars="species", variable.name = "type", value.name="score", factorsAsStrings=FALSE)

# Make data frame with node labels and node ids
n_pass <- length(clade_info_passeriformes[[2]])
family_labels_pass <- data.frame(family_label=rep(NA, n_pass), node=rep(NA, n_pass), colour_bar=rep(NA, n_pass))

for (i in 1:n_pass) {
  family_labels_pass[i,1] <- names(clade_info_passeriformes[[2]][i])
  family_labels_pass[i,2] <- ggtree::MRCA(birdelabtree_pass, clade_info_passeriformes[[2]][[i]][[1]])
  family_labels_pass[i,3] <- clade_info_passeriformes[[2]][[i]][[2]]
}

# Order data frame alphabetically according to the taxon labels
# Necessary because geom_cladelabel adds labels alphabetically but plots colours 
# in the order of the provided vector (alphatising ensures plotting order is correct)
# Order data frame alphabetically according to the taxon labels
# Necessary because geom_cladelabel adds labels alphabetically but plots colours 
# in the order of the provided vector (alphatising ensures plotting order is correct)
family_labels_pass <-arrange(family_labels_pass, family_label)
family_labels_pass[,3] <- arrange(data.frame(colour=tip_colours_passeriformes_family, family_label=names(tip_colours_passeriformes_family)), family_label)[,1]

## Second part of plot - add heatmap for elaboration and innovation scores
innovelab_pass_plot <- innovelab_pass_plot +
  new_scale_fill() + 
  geom_fruit(data=heatmapdat_pass, geom=geom_tile, 
             mapping=aes(y=species, x=type, fill=score),
             axis.params=list(
               axis="x",
               text.angle = -90,
               text.size = 4.5,
               line.size = 0,
               vjust = 0.5,
               hjust=0),
             offset = 0.02,size = 0.02) +
  scale_fill_continuous_divergingx(palette = 'Spectral', rev=TRUE,
                                   mid = 1, na.value="grey80") + 
  theme(legend.position=c(0.56,0.43), legend.key.size = unit(0.5, "cm"), 
        legend.key.width=unit(0.6, "cm")) + 
  labs(fill= "Elab & Innov") +
  labs(color= "Distance")

## Third part of plot - add clade labels
innovelab_pass_plot <- innovelab_pass_plot +
  new_scale_colour() + 
  geom_cladelab(data=family_labels_pass, 
                mapping=aes(node=node, 
                            label=family_label,
                            colour=family_label),
                textcolour="black",
                hjust=0,
                angle="auto",
                barsize=4,
                horizontal=TRUE, 
                fontsize=4,
                offset=18.5,
                offset.text=1.5,
                show.legend=FALSE
  ) +
  scale_colour_manual(
    values=family_labels_pass$colour_bar, 
    guide="none"
  )

## Save as png and pdf
ggsave("../Manuscript/Figures/InnovElabTree_passerine_supplement.png", innovelab_pass_plot, width=12, height=12)
ggsave("../Manuscript/Figures/InnovElabTree_passerine_supplement.pdf", innovelab_pass_plot, width=12, height=12)

```

### 2022-08-12 Manuscript/main.tex
Updated MS

@@
 \bigskip
-\medskip
+\bigskip
 \begin{center}
@@
 \noindent {\small \it 
-$^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\
+$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
+$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
 $^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
@@
 \end{center}
-\medskip
+\bigskip
 \noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\ 
@@
 
-\noindent (Keywords: )\\
+\noindent (Keywords: macroevolution, variance-covariance matrices, line of least resistance, GLMM)\\
 
@@
 At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
-The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance.
+The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance \cite{endler2005animal, renaud2006conserved}.
 Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
@@
 Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow lines of genetic least resistance \cite{schluter1996adaptive}.
-Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape.
+Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape \cite{jones2004evolution}.
 Indeed, evidence suggests that genetic constraints can extend over timescales spanning tens of millions of years leading to remarkable macroevolutionary stability in the orientation of phenotypic evolution \cite{mcglothlin2018adaptive}.
@@
 This apparent tension implies deep-time reorientation of phenotypic lines of least resistance and points to alternative routes to the exploration of phenotypic space by evolving either along, or away from, the dominant phenotypic axis.
-We refer to these alternatives as elaboration - evolution along the phylogenetically constrained phenotypic major axis - and innovation - evolution away from the phylogenetically constrained phenotypic major axis.
+We refer to these alternatives as elaboration - evolution along the phylogenetically constrained phenotypic major axis - and innovation - evolution away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
 The contributions of these two multivariate pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales.
@@
 All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-We fitted a phylogenetic generalised linear mixed effect model (pGLMM with beak shape as an eight-dimensional response variable andtreated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods).
+We fitted a phylogenetic generalised linear mixed effect model (pGLMM with beak shape as an eight-dimensional response variable andtreated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
 We regard the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
@@
 \end{figure}
+\bigskip
 
@@
 
+\bigskip
+
 \begin{figure}[!htbp]
@@
 \end{figure}
+
+\bigskip
 
@@
 For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
-The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp > 0.99; ** = pp > 0.95; * = pp > 0.9; . = pp > 0.8).
-Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp > 0.99.
+The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
+Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
 This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+
+\bigskip
 
@@
 To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
-As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis.
+As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
 Clades (or species) can be both elaborators and innovators.
@@
 superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
-We then used multilinear algebra to project a focal group's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheett}).
+We then used multilinear algebra to project a focal group's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
 We interpret the measured projection (distance onto the phylogenetic major axis) as the group's elaboration score and the measured rejection (distance from the major axis) as the group's innovation score.
@@
 \centering
-   % \includegraphics[width=0.9\textwidth]{Figures/phylogenys.pdf}
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
 \caption{.}
@@
 \end{figure}
+
+\bigskip
+
 \noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by speciesbeak shape innovation and elaboration level.
@@
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
+
+\bigskip
 
@@
 
+\bigskip
+
 \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among groups (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
@@
 We can see a clear difference in the amount of innovation across the posterior samples between species or groups.
+
+\bigskip
 
@@
 
-This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself.
+This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
 These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).

### 2022-08-12 Manuscript/supplementaries.tex
updated up to paragraph 4
manuscript up (NOW MOVING TO OVERLEAF
Updated MS

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1}$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
 \noindent {\small \it 
-$^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}
+$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
+$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
+$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
 
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/axis_selection.pdf}
-\caption{variance and cumulative variance for each 8 axis per super-order.}
-\label{Fig:axes_variance}
+   \includegraphics[width=0.9\textwidth]{Figures/mini-chains_diagram.pdf}
+\caption{mcmcmcglmmm: mini-chains MCMCglmm method diagram}
+\label{Fig:mcmcmcglmm}
 \end{figure}
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/mini-chains_diagram.pdf}
-\caption{mcmcmcglmmm: mini-chains MCMCglmm method diagram}
-\label{Fig:fig_mcmcmcglmm}
+   \includegraphics[width=0.9\textwidth]{Figures/axis_selection.pdf}
+\caption{variance and cumulative variance for each 8 axis per super-order.}
+\label{Fig:axes_variance}
 \end{figure}
@@
 \centering
-   % \includegraphics[width=0.9\textwidth]{Figures/.pdf}
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement.pdf}
 \caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 3 in the main text for details.}
@@
 \section{Supplementary materials 3: detailed projection and rejection operations}
+\label{supp_projection}
 
@@
 \bibliographystyle{naturemag}
-\bibliography{References}
+\bibliography{references}
 

### 2022-08-12 Manuscript/main.tex
added funding + data

@@
 
-\noindent (Keywords: macroevolution, variance-covariance matrices, line of least resistance, GLMM)\\
+\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
 
@@
 The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
-The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/MCMCglmm/inst/vignettes/model_test_vignette.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
+The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
 and implemented in the dispRity package \cite{dispRity}.
@@
 
+\textbf{Reproducibility and repeatability.}
+The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
+The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
+
+\textbf{Funding.}
+This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
+
 \bibliographystyle{naturemag}

### 2022-08-23 Analysis/01-Data_preparation.Rmd
TODO!
Update 01-Data_preparation.Rmd

@@
 
-All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
-Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
-The following code snippets are used to illustrate the implementation of the functions.
-See §§§ for running the actual analyses.
+All the custom code and custom functions called in this document can be downloaded and installed in `R`.
 
-The method described in the next three markdown is essentially a mini chains with pseudo-metropolis coupling markov chain generalised linear mixed model method to determine clades and tips elaboration and exploration (`mcmcpmcmcmcglmmmctee`) but we will refrain using this acronym and method name.
-
-```{r install_beer, eval = FALSE}
-## Installing the dispRity branch
-devtools::install_github("TGuillerme/dispRity@MCMCglmm")
-```
-
-```{r load_beer, eval = TRUE}
+```{r packages, eval = TRUE}
 ## Loading packages
@@
 source("../Functions/stats.utilities.R")
-
 ```
@@
 # Data preparation
-
-TODO:
-
- * [ ] what's the data, where is it from
- * [ ] Procrustes
- * [ ] Building the trait space
-   - [ ] form space version?
-   - [ ] shape space version?
- * [ ] Clade data and grouping
 
@@
  * The clade data
-We used the different clade levels (the 11 magnificent 7 + the families). For each level/group, we removed any groups with less than 20 species.
+We used the different clade levels (super orders, orders, sub-orders and families). For each level/group, we removed any groups with less than 20 species.
 
@@
 ```{r}
+source("../Functions/prep.data.R")
 ## Get the traitspace for all passeriformes
@@
 clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
-shapespace_passeriformes <- shapespace_allbirds_lvl_superorder_order[[1]]$space[(clades_list[, "level2_order"] == "Passeriformes"), ]
-
-## Check what's happening with Neodrepanis_hypoxantha
+shapespace_passeriformes <- shapespace_allbirds_lvl_superorder_order[[1]]$space[shapespace_allbirds_lvl_superorder_order[[1]]$space$level2 == "Passeriformes", ]
 
@@
 }
+
 ## Fill in the level2
@@
 
-## Do the prep data from here (with all the trees! ouch!)
+## Do the prep data from here
 shapespace_passeriformes_two_levels <- prep.data(

### 2022-08-29 Analysis/01-Data_preparation.Rmd
removed formspace

@@
 shapespace <- shapespace$PCscores
-formspace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_FORMSPACE_FULL.rds")
-formspace <- formspace$PCscores
 ```
@@
 ```{r, echo = FALSE, fig.height = 16, fig.width = 8}
-par(mfrow = c(2,1))
 plot.dispRity(shapespace, type = "preview", pch = 21, main = "Shape space")
-plot.dispRity(formspace, type = "preview", pch = 21, main = "Form space")
 ```

### 2022-09-02 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 ## Load the covar
-load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
+load(file = "../Data/Processed/shapespace_passeriformes_5996t.MCMCglmm")
 shapespace_passeriformes_posteriors <- shapespace_passeriformes[[1]]
@@
 source("../Functions/stats.utilities.R")
-
-# ## Fucked up thing happening here:
-# cd("~/Packaging/dispRity/")
-# load_all()
 
@@
 ```
+
+## Testing whether the passeriformes ellipses are similar
+
+Basically whether the level 2 passeriformes random terms VCVs in the all birds models are similar to the level 0 passeriformes random terms VCVs in the passeriformes only model.
+We don't expect them to be identical due to randomness of the selected trees and random seeds but they should be at least very similar.
+
+```{r}
+## Get the VCVs from each model
+model1 <- all_birds_shapespace$covar$Passeriformes
+model2 <- passeriformes_shapespace$covar$Passeriformes
+allbirds <- all_birds_shapespace$covar$phylogeny
+passerida <- passeriformes_shapespace$covar$Passerida
+
+## Testing the differences between passeriformes in both spaces
+passeriformes_model1 <- covar.ellipse.test(model1, model2, n = 2000, measure = "orthogonality")
+passeriformes_model2 <- covar.ellipse.test(model2, model1, n = 2000, measure = "orthogonality")
+
+## Testing the differences between passeriformes in both spaces
+passeriformes_model1_vs_allbirds <- covar.ellipse.test(model1, allbirds, n = 2000, measure = "orthogonality")
+passeriformes_model2_vs_allbirds <- covar.ellipse.test(model2, allbirds, n = 2000, measure = "orthogonality")
+passeriformes_model1_vs_passerida <- covar.ellipse.test(model1, passerida, n = 2000, measure = "orthogonality")
+passeriformes_model2_vs_passerida <- covar.ellipse.test(model2, passerida, n = 2000, measure = "orthogonality")
+
+## Merging the results
+passeriformes_comparisons <- rbind(passeriformes_model1,
+                                   passeriformes_model1_vs_allbirds,
+                                   passeriformes_model1_vs_passerida,
+                                   passeriformes_model2,
+                                   passeriformes_model2_vs_allbirds,
+                                   passeriformes_model2_vs_passerida)
+
+## Visualising the differences
+pdf(file = "../Manuscript/Figures/Passeriformes_differences.pdf", width = 9, height = 5)
+par(mar = c(5, 12, 2, 2))
+plot(NULL, xlim = c(0.5, 1), ylim = c(6,1), xlab = "Posterior probability of orthogonality being different\n(comparing variance of orthogonality within vs. between)", yaxt = "n", ylab = "")
+axis(2, las = 2, at = c(1:6), labels = c("Model 2 (sd = 13.65)", "Whole phylogeny (sd = 3.4)", "Passerida (sd = 19.89)", "Model 1 (sd = 11.64)", "Whole phylogeny (sd = 3.4)", "
+    Passerida (sd = 19.89)"))
+legend("topleft", col = c("orange", "blue"), pch =  19, lty =  1, legend = c("Model 1 (whole tree; sd = 11.64) vs.", "Model 2 (passeriformes; sd = 13.65) vs."))
+
+col <- c(rep("orange", 3), rep("blue", 3))
+for(line in 1:6) {
+    lines(matrix(c(passeriformes_comparisons[line, c(2,3)], rep(line, 2)), 2, 2), col = col[line])
+    points(passeriformes_comparisons[line, 1], line, pch = 19, col = col[line])
+}
+dev.off()
+```

### 2022-09-06 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 This script is a working script to run the chains on the SHARC Sheffield cluster.
-Refer to `02-MCMCglmm_mini_chains.Rmd` for more details and reproducibility
 
@@
 save(shapespace_passeriformes, file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
-
 ```

### 2022-09-06 Manuscript/main.tex
Update main.tex
Minor typos all over

@@
 
+\section{Abstract}
+
 Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
@@
 These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+
+\section{Main}
 
@@
 All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-We fitted a phylogenetic generalised linear mixed effect model (pGLMM with beak shape as an eight-dimensional response variable andtreated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
+We fitted a phylogenetic generalised linear mixed effect model (pGLMM) with beak shape as an eight-dimensional response variable and treated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
 We regard the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
@@
 \ref{Fig:ellipses}a) and order (Fig.
-\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Figure \ref{Fig:orthogonality}).
+\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
+\ref{Fig:orthogonality}).
 The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
@@
 Dots represent the clades median orthogonality and the lines their 95\% CI.
-No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at keast half orthogonal to their super-order (green lines and circles).
+No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
 These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
@@
 \ref{Fig:ellipses}).
-These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig.S\ref{fig_ellipses_passeriformes}).
+These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
 This suggests that divergence away from line of least evolutionary resistance (i.e.
@@
 Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), align with PC1 in shapespace Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig.
 \ref{Fig:ellipses}-bar plots).
@@
 This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
-To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shape-space onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
+To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shapespace onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
 We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire groups.
@@
 The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
-\ref{Fig:phylogeny}).
-%{#fig_correlations}),%TODO: maybe add to supp?
+\ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
@@
 
-\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by speciesbeak shape innovation and elaboration level.
+\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
 Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
-Innovations and elaboration are displayed as: a)comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
+Innovations and elaboration are displayed as: a) comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
 The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
@@
 
-Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation along the megaevolutionary line of least resistance is much more common.
+Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.
 This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
-However, individual specie-level variation in the past could also have led to a shift in a group's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape, Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
+However, individual species-level variation in the past could also have led to a shift in a group's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
+Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
 Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
@@
 The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,chira2020signature,hughes2022global} for detailed descriptions of data collection and processing).
-We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the first eight first dimensions of the resulting PCA.
+We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
 These eight dimensions contained at least 95\% of variance in each superorder, and 98.7\% of the variance in the whole dataset.
@@
 \begin{equation}
-\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)} (equation 1)
+\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
 \end{equation}
@@
 One model using one random term for the whole phylogeny and 35 nested random terms for each superorder and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
-And one model on a subset of the dataset containing only the 5966 passeriformes species with 30 nested random terms.
+And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
 This resulted in two multidimensional GLMM.
 One with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
-And one with a $5966 \times 8$ response variable, one global residual term and 30 random terms.
+And one with a $5229 \times 8$ response variable, one global residual term and 30 random terms.
 
@@
 The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
-and implemented in the dispRity package \cite{dispRity}.
+and implemented in the \texttt{dispRity} package \cite{dispRity}.
 
 \textbf{Groups' orthogonality.}
-We measured the orthogonality of each group (i.e. random term in our models compared to their parent group and their parent parent's (e.g.
-the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).To do so we measured the angles between the major axes of the focal group and their parent group for each posterior variance-covariance matrix as well as within each group by randomly comparing major axes within a group.
+We measured the orthogonality of each random term in our models (clades) compared to their parent group and their parent parent's.
+For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
+To do so we measured the angles between the major axes of the focal group and their parent group for each posterior variance-covariance matrix as well as within each group by randomly comparing major axes within a group.
 We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).

### 2022-09-08 Manuscript/main.tex
added comments from last meeting and removed "group" from the paper (now clade)

@@
 %---------------------------------------------
+% QUESTIONS/narrative
+% How does phenotypic diversity arise?
+% "Disparity" through time = how is disparity within clades partitioned wrt overall disparity. Diversity accumulates in a boring way, but doesn't tell us about mechanisms, 
+% Lines of evolutionary least resistance theory. Very little testing 
+% Test this using new method: elaboration/innovation
+% How does this vary at different hierarchical levels - mega vs macro evolutionary?
+% Species-level analyses
+% Passerines
+
+% Bits from Andrew
+% A first insight into the evolutionary scale at which phenotypic diversity has arisen in birds comes from a simple exploration of the variance among major axes at different scales.
+If variance in the major axis is highest at the phylogenetic (mega) scale rather than the clade (macro) scale, it is evidence that diversity is very old.
+
+% ANSWER
+
+% Supporting this, we find overwhelming evidence that deep time evolutionary history does not constrain phenotypic variation.
+This is evidenced by the orientation of evolution within clades being markedly different from and frequently orthogonal to the major phylogenetic axis, indicating that deep evolutionary history does not constrain diversity.
+
+% DETAILS
+
+% Against this backdrop, we then estimated the relative importance of elaboration or innovation to driving this pattern….
+We made 3 specific investigations to do this....(Needs to be more precise about question, how we answer and answer - e.g.
+is elaboration most common at all evolutionary scales?
+Is innovation isolated to evolutionary scales?).
+
+% Trick to embed question in 'result'??
 
@@
 However, this contrasts with observations of major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
-We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different groups compared to the global phylogenetic major axis.
+We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different clades compared to the global phylogenetic major axis.
 We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis (Fig.
@@
 Clades (or species) can be both elaborators and innovators.
-To measure innovation and elaboration we translated the major axes of each group's (i.e.
+To measure innovation and elaboration we translated the major axes of each clade's (i.e.
 superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
-We then used multilinear algebra to project a focal group's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
-We interpret the measured projection (distance onto the phylogenetic major axis) as the group's elaboration score and the measured rejection (distance from the major axis) as the group's innovation score.
-These scores were measured for each of 4000 posterior pairs of group vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-We found that for most groups the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
-Remarkably, more than half of the assessed groups (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
-However, there is substantial variation in terms of elaboration and innovation across groups including relatively low elaboration and innovation (e.g.
+We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
+We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
+These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
+We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
+Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
+However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
 Cuculiformes; Fig.
@@
 To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shapespace onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
-We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire groups.
+We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire clades.
 We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
@@
 For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
-Other groups such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
+Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
 Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
@@
 We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
-We then measured the difference between both areas and found that there was a significant difference in innovation between species and groups but no clear difference in elaboration (Fig.
+We then measured the difference between both areas and found that there was a significant difference in innovation between species and clades but no clear difference in elaboration (Fig.
 \ref{Fig:relative_EI}).
-This difference confirms our observation that innovation is more common at the group level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
+This difference confirms our observation that innovation is more common at the clade level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
 However, we suggest that rather than a paradox the reorientation of trait space can instead be described as a gradual process occuring in response to shifting selection pressures acting on species evolving in similar adaptive zones.
@@
 
-\noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among groups (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
-A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 groups.
-The scores for each group is scaled by the maximum score within each group (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
-This allows us to compare both types (species vs.
-groups) despite their different sample sizes.
+\noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
+A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 clades.
+The scores for each clade is scaled by the maximum score within each clade (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
+This allows us to compare both types (species vs. clades) despite their different sample sizes.
 B - same as A but for the innovation score.
-C - density curve of the area under the curve for each posterior sample relative elaboration score as a proxy for the overall amount of elaboration per sample and per type (species or group - respectively blues and orange).
+C - density curve of the area under the curve for each posterior sample relative elaboration score as a proxy for the overall amount of elaboration per sample and per type (species or clade - respectively blues and orange).
 The Bhattacharyya coefficient indicates the probability of overlap between the posterior distributions.
 D - same as C but for innovation.
-We can see a clear difference in the amount of innovation across the posterior samples between species or groups.
+We can see a clear difference in the amount of innovation across the posterior samples between species or clades.
 
@@
 This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
-However, individual species-level variation in the past could also have led to a shift in a group's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
+However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
 Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
@@
 We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
-In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group.
-In brief we can interpret where a species or a group falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
+In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that clade.
+In brief we can interpret where a species or a clade falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
 We calculated the elaboration and innovation scores using two main approaches:
@@
 \textbf{Groups' orthogonality.}
-We measured the orthogonality of each random term in our models (clades) compared to their parent group and their parent parent's.
+We measured the orthogonality of each random term in our models (clades) compared to their parent clade and their parent parent's.
 For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
-To do so we measured the angles between the major axes of the focal group and their parent group for each posterior variance-covariance matrix as well as within each group by randomly comparing major axes within a group.
+To do so we measured the angles between the major axes of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade by randomly comparing major axes within a clade.
 We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
-We then measured the posterior probability of the angles within a group being different than the angles between groups.
+We then measured the posterior probability of the angles within a clade being different than the angles between clades.
 These results are available in Fig. \ref{Fig:orthogonality}.

### 2022-09-08 Manuscript/supplementaries.tex
Glossary draft

@@
 
-\section{Supplementary material 1: additional results}
+\section{Supplementary material 1: glossary}
+
+\begin{table}[ht]
+\centering
+\scriptsize
+\begin{tabular}{llll}
+  \hline
+Term in this paper          & evolutionary biology definition & mathematics definition \\
+  \hline
+clade                       & a monophyletic group of species & a distance matrix associated with a subset of the \textbf{shapespace} \\ 
+clade's major axis          & the \textbf{line of least resistance} specific to a clade & the first \textbf{eigenvector} of a \textbf{clade}'s \textbf{variance-covariance} matrix &  \\ 
+eigenvector                 & one evolutionary \textbf{line of least resistance} for a clade & one of the vectors (direction) defining a \textbf{variance-covariance} matrix \\
+elaboration                 & morphological score of a \textbf{species} or a \textbf{clade} in relation to a \textbf{line of least resistance} of reference & the \textbf{projection} of a \textbf{species} or a \textbf{clade} onto the first \textbf{eigenvector} of a \textbf{variance-covariance} matrix \\
+family                      & a specific \textbf{clade} level (nested within \textbf{sub-orders}) & a distance matrix \\
+higher taxa                 & a higher \textbf{clade} level & a distance matrix associated with a subset of the \textbf{shapespace} \\
+innovation                  & morphological score of a \textbf{species} or a \textbf{clade} in relation to a \textbf{line of least resistance} of reference & the \textbf{rejection} of a \textbf{species} or a \textbf{clade} away from the first \textbf{eigenvector} of a \textbf{variance-covariance} matrix \\
+lines of least resistance   & the major evolutionary ``direction'' of traits co-evolution & the first \textbf{eigenvector} or a \textbf{variance-covariance} matrix \\
+macroevolution              & evolution above the species level & analyses using a portion of the entire shapespace \\
+major axis                  & one of the ``directions'' of traits co-evolution & an \textbf{eigenvector} or a \textbf{variance-covariance} matrix; this can be represented as the longest orthogonal lines of an ellipse \\
+megaevolution               & evolution above the \textbf{clade} level - here arbitrarily defined at the whole \textbf{phylogeny} level (i.e. all the birds) & analyses using the entire shapespace \\
+order                       & a specific \textbf{clade} level (nested within \textbf{super-orders}) & a distance matrix associated with a subset of the \textbf{shapespace} \\
+phylogenetic major axis     & the \textbf{line of least resistance} for the whole \textbf{phylogeny} & the first \textbf{eigenvector} of the whole \textbf{phylogeny}'s \textbf{variance-covariance} matrix \\
+phylogeny                   & the relationships between all the \textbf{species} and their respective \textbf{clades} used in this analysis & a distance matrix associated with the whole \textbf{shapespace} \\
+projection                  & the \textbf{elaboration} score of \textbf{species} or a \textbf{clade} & the position of a \textbf{species} or a \textbf{clade} onto an \textbf{eigenvector}\\
+rejection                   & the \textbf{innovation} score of \textbf{species} or a \textbf{clade} & the distance of a \textbf{species} or a \textbf{clade} away from an \textbf{eigenvector} \\
+residual                    & the variation in the \textbf{shapespace} not explained by the \textbf{phylogeny} or a \textbf{clade} & the error term in the pGLMM not explained by the random error terms (\textbf{phylogeny} and \textbf{clade}) \\
+shapespace                  & all the observed morphological combination of the 8748 \textbf{species} of birds & a matrix where each columns are traits (or a transformation thereof, here they are principal components) and each rows are observations (here \textbf{species}).
+\\
+species                     & the lowest evolutionary units in our analysis & rows in the \textbf{shapespace} matrix \\
+sub-orders                  & a specific \textbf{clade} level (nested within \textbf{orders}) & a distance matrix associated with a subset of the \textbf{shapespace} \\
+super-order                 & a specific \textbf{clade} level (the first nested level within the whole \textbf{phylogeny}) & a distance matrix associated with a subset of the \textbf{shapespace} \\
+variance-covariance         & the traits co variation in the \textbf{shapespace} & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) \\
+
+\end{tabular}
+\caption{Glossary of some specific terms in this manuscript}
+\label{tab_glossary}
+\end{table}
+
+\section{Supplementary material 2: additional results}
 
@@
 
-\section{Supplementary material 2: Passeriformes results}
+\section{Supplementary material 3: Passeriformes results}
 
@@
 
-\section{Supplementary materials 3: detailed projection and rejection operations}
+\section{Supplementary materials 4: detailed projection and rejection operations}
 \label{supp_projection}

### 2022-09-09 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 \scriptsize
-\begin{tabular}{llll}
+\begin{tabular}{p{0.2\textwidth}p{0.4\textwidth}p{0.4\textwidth}}
   \hline
@@
 clade                       & a monophyletic group of species & a distance matrix associated with a subset of the \textbf{shapespace} \\ 
-clade's major axis          & the \textbf{line of least resistance} specific to a clade & the first \textbf{eigenvector} of a \textbf{clade}'s \textbf{variance-covariance} matrix &  \\ 
+clade's major axis          & the \textbf{line of least resistance} specific to a clade & the first \textbf{eigenvector} of a \textbf{clade}'s \textbf{variance-covariance} matrix \\ 
 eigenvector                 & one evolutionary \textbf{line of least resistance} for a clade & one of the vectors (direction) defining a \textbf{variance-covariance} matrix \\

### 2022-09-09 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]
+
+## Get the variance for each level
+variance_per_levels <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order)
+## Rename the columns
+colnames(variance_per_levels) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
+
+## Plot the results
+pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf")
+par(mar = c(5,10,2,2))
+boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by term", xlab = "Scaled variance (sum of each VCV)")
+dev.off()
 

### 2022-09-13 Analysis/03-Elaboration_innovation_analysis.Rmd
added variance and se calculations

--- 2022-08-08 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 
+## Get the variance for each level
+variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = TRUE)
+## Rename the columns
+colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
+reorder_clades <- c(9:35,1:8, 36, 37)
+## Pull by term
+variance_per_levels<- cbind("orders" = apply(variance_per_clades[, 9:35], 1, sum),
+                            "super-orders" = apply(variance_per_clades[, 1:8], 1, sum),
+                            "phylogeny" = variance_per_clades[, 36],
+                            "residuals" = variance_per_clades[, 37])
+variance_per_terms <- cbind("clades" = apply(variance_per_clades[, 1:35], 1, sum),
+                            "phylogeny" = variance_per_clades[, 36],
+                            "residuals" = variance_per_clades[, 37])
+## Plot the results
+pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf", height = 8, width = 14)
+par(mar = c(5,8,2,2), mfrow = c(1, 3))
+boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+dev.off()
+
 ## Get the levels
@@
 rm(shapespace_allbirds_lvl_superorder_order)
+
+## Get the number of observations per groups
+variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = FALSE)
+colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
+
+## Converting into standard errors
+get.se <- function(var, n) {
+    sqrt(var[[1]])/sqrt(n)
+}
+## Get the se per group
+se_per_clades <- mapply(get.se, apply(variance_per_clades, 2, list), as.list(c(size.subsets(superorder_shapespace)[-9], size.subsets(order_shapespace), "residuals" = 8748)))
+## Scale that
+se_per_clades <- se_per_clades/rowSums(se_per_clades)
+## Get the other levels
+se_per_levels<- cbind("orders" = apply(se_per_clades[, 9:35], 1, sum),
+                      "super-orders" = apply(se_per_clades[, 1:8], 1, sum),
+                      "phylogeny" = se_per_clades[, 36],
+                      "residuals" = se_per_clades[, 37])
+se_per_terms <- cbind("clades" = apply(se_per_clades[, 1:35], 1, sum),
+                      "phylogeny" = se_per_clades[, 36],
+                      "residuals" = se_per_clades[, 37])
+
+## Plot the results
+pdf(file = "../Manuscript/Figures/se_per_levels_whole_tree.pdf", height = 8, width = 14)
+par(mar = c(5,8,2,2), mfrow = c(1, 3))
+boxplot(se_per_terms, horizontal = TRUE, las = 1, main = "Scaled standard error by terms", xlab = "Scaled se ")
+boxplot(se_per_levels, horizontal = TRUE, las = 1, main = "Scaled standard error by levels", xlab = "Scaled se")
+boxplot(se_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled standard error by clades", xlab = "Scaled variance")
+dev.off()
 ```

### 2022-10-07 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
            legend.args = list(x = "bottomright", cex = 2/3))
+
+## Saving as a pdf
+pdf(file = "../Manuscript/Figures/traitspace_plot.pdf")
+covar.plot(superorder_shapespace,
+           main = "Shapespace",
+           cex = 0.5,
+           legend = TRUE,
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
+           legend.args = list(x = "bottomright", cex = 2/3))
+dev.off()
 ```
@@
            cex = 0.5, apply.to.VCV = TRUE)
+
+pdf(file = "../Manuscript/Figures/unscaled_ellipses.pdf", height = 8, width = 8)
+par(mfrow = c(2,2))
+## Plotting mean axes
+covar.plot(superorder_shapespace,
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
+           main = "Average ellipses by super order",
+           ellipses = mean,
+           legend = FALSE,
+           points = FALSE,
+           cex = 0.5, apply.to.VCV = TRUE)
+## Plotting mean ellipses
+covar.plot(superorder_shapespace,
+           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
+           main = "Average major axes by super order",
+           major.axes = mean,
+           legend = TRUE,
+           points = FALSE,
+           cex = 0.5, apply.to.VCV = TRUE)
+covar.plot(order_shapespace,
+           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
+           main = "Average ellipses by order",
+           ellipses = mean,
+           legend = FALSE,
+           points = FALSE,
+           cex = 0.5, apply.to.VCV = TRUE)
+## Plotting mean ellipses
+covar.plot(order_shapespace,
+           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
+           main = "Average major axes by order",
+           major.axes = mean,
+           legend = TRUE,
+           points = FALSE,
+           cex = 0.5, apply.to.VCV = TRUE)
+dev.off()
 ```

### 2022-10-07 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 ## Load the covar
-load(file = "../Data/Processed/shapespace_passeriformes_5996t.MCMCglmm")
+load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
 shapespace_passeriformes_posteriors <- shapespace_passeriformes[[1]]
@@
 source("../Functions/stats.utilities.R")
+library(xtable)
 

### 2022-10-07 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update figure EI comparison

@@
 #TODO: make the groups nested in the fable
-
 pdf("../Manuscript/Figures/orthogonality_results.pdf", width = 9, height = 9)
@@
 sub-orders)", "Families (vs.
-passeriformes)"), col = c(gg.color.hue(4)[2], gg.color.hue(4)[3], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
+passeriformes)"), col = c(gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
 
@@
 ```
+
+## Testing whether clades innovate more than species
 
@@
 
-pdf("../Manuscript/Figures/relative_EI.pdf", height = 8, width = 8)
-par(mfrow = c(2,2))
+pdf("../Manuscript/Figures/relative_EI.pdf", height = 6, width = 12)
+par(mfrow = c(1,2))
 plot.density(species = c(elaboration_species),
@@
              ylab    = "density",
-             main    = "A - Total posterior elaboration per group\n(scaled) and per species")
-legend("topright", lty = 1, legend = c("species", "groups"), col = c("blue", "orange"))
+             main    = "A - Total posterior elaboration per clade\n(scaled) and per species")
+legend("topright", lty = 1, legend = c("species", "clades"), col = c("blue", "orange"))
+text(x = 0.7, y = 0.5, paste0("Posterior area under the curve overlap\n(Bhattacharyya coefficient): ", round(bhatt.coeff(elaboration_species_auc, elaboration_group_auc), 3)))
 plot.density(species = c(innovation_species),
@@
              ylab    = "density",
-             main    = "B - Total posterior innovation per group\n(scaled) and per species")
-plot.density(elaboration_species_auc, elaboration_group_auc, scale.x = FALSE, col = c("blue", "orange"), ylab = "density",
-             xlab = "Posterior area under the curve (elaboration)",
-             main = paste0("C - Difference in amounts of elaboration\nper group or per species\nBhattacharyya coefficient = ",
-                           round(bhatt.coeff(elaboration_species_auc, elaboration_group_auc), 3)))
-plot.density(innovation_species_auc, innovation_group_auc, scale.x = FALSE, col = c("blue", "orange"), ylab = "density",
-             xlab = "Posterior area under the curve (innovation)",
-             main = paste0("D - Difference in amounts of innovation\nper group or per species\nBhattacharyya coefficient = ",
-                           bhatt.coeff(innovation_species_auc, innovation_group_auc)))
+             main    = "B - Total posterior innovation per clade\n(scaled) and per species")
+text(x = 0.7, y = 0.5, paste0("Posterior area under the curve overlap\n(Bhattacharyya coefficient): ", round(bhatt.coeff(innovation_species_auc, innovation_group_auc), 3)))
 dev.off()

### 2022-10-07 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Updated passeriformes figure

@@
 ## Testing the differences between passeriformes in both spaces
-passeriformes_model1 <- covar.ellipse.test(model1, model2, n = 2000, measure = "orthogonality")
-passeriformes_model2 <- covar.ellipse.test(model2, model1, n = 2000, measure = "orthogonality")
+passeriformes_model1 <- covar.ellipse.test(model1, model2, n = 4000, measure = "orthogonality")
+passeriformes_model2 <- covar.ellipse.test(model2, model1, n = 4000, measure = "orthogonality")
 
 ## Testing the differences between passeriformes in both spaces
-passeriformes_model1_vs_allbirds <- covar.ellipse.test(model1, allbirds, n = 2000, measure = "orthogonality")
-passeriformes_model2_vs_allbirds <- covar.ellipse.test(model2, allbirds, n = 2000, measure = "orthogonality")
-passeriformes_model1_vs_passerida <- covar.ellipse.test(model1, passerida, n = 2000, measure = "orthogonality")
-passeriformes_model2_vs_passerida <- covar.ellipse.test(model2, passerida, n = 2000, measure = "orthogonality")
+passeriformes_model1_vs_allbirds <- covar.ellipse.test(model1, allbirds, n = 4000, measure = "orthogonality")
+passeriformes_model2_vs_allbirds <- covar.ellipse.test(model2, allbirds, n = 4000, measure = "orthogonality")
+passeriformes_model1_vs_passerida <- covar.ellipse.test(model1, passerida, n = 4000, measure = "orthogonality")
+passeriformes_model2_vs_passerida <- covar.ellipse.test(model2, passerida, n = 4000, measure = "orthogonality")
 

### 2022-10-07 Manuscript/main.tex
Update main.tex

@@
 Comparisons of orientation among subclades (i.e.
-orders within superorders) show similar differences (median = $47.17^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
+orders within superorders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
 This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
@@
 This allows us to compare both types (species vs. clades) despite their different sample sizes.
+The Bhattacharryya Coefficient represents the overlap of the area under the curve for species and clades (a BC of 0 represents no overlap).
 B - same as A but for the innovation score.
-C - density curve of the area under the curve for each posterior sample relative elaboration score as a proxy for the overall amount of elaboration per sample and per type (species or clade - respectively blues and orange).
-The Bhattacharyya coefficient indicates the probability of overlap between the posterior distributions.
-D - same as C but for innovation.
 We can see a clear difference in the amount of innovation across the posterior samples between species or clades.

### 2022-10-07 VCVtools/tests/testthat/test-make.VCV.R
Delete test-make.VCV.R

@@ -1,37 +1,0 @@
-context("make.ellipse")
-
-## Test
-test_that("make.ellipse works", {
-
-    ## All the lambda/shape business
-
-    ## Visual examples
-    ## Run the example with 100D
-    input_values <- seq(from = 0, to = 1, length.out = 11)
-    shape_list <- get.lambda(input_values)
-    plot(NULL, xlim = c(0,1), ylim = c(0,1), xlab = "Dimensions (proportional)", ylab = "Axis length (proportional)", main = "Axis lengths with in 100D")
-    cols <- rainbow(length(shape_list))
-    for(i in 1:length(shape_list)) {
-        lines(generate.shape(dimensions = 100, lambda = shape_list[i]), col = cols[i])
-    }
-    legend(x = 0.5, y = 0.75, legend = paste(paste0("shape = ", round(input_values, 1)), paste0("lambda = ", round(shape_list, 2)), sep = "; "), lty = 1, col = cols, cex = 0.7)
-
-    ## Run the example with 3D and minimum = 0.1
-    input_values <- seq(from = 0, to = 1, length.out = 11)
-    shape_list <- get.lambda(input_values)
-    plot(NULL, xlim = c(0,1), ylim = c(0,1), xlab = "Dimensions (proportional)", ylab = "Axis length (proportional)", main = "Axis lengths with in 3D\nwith min.thick = 0.01")
-    cols <- rainbow(length(shape_list))
-    for(i in 1:length(shape_list)) {
-        lines(generate.shape(dimensions = 3, lambda = shape_list[i], min.thick = 0.01), col = cols[i])
-    }
-    legend(x = 0.5, y = 0.75, legend = paste(paste0("shape = ", round(input_values, 1)), paste0("lambda = ", round(shape_list, 2)), sep = "; "), lty = 1, col = cols, cex = 0.7)
-
-
-    ## Testing make.VCV
-    ## Default is a circle
-    test <- make.VCV()
-    expect_equal(names(test), c("VCV", "loc"))
-    expect_null(plot.VCV(test, VCV.vectors = TRUE, eigen.vectors = TRUE, eigen.values = TRUE))
-    ## Example with options
-    expect_null(plot.VCV(make.VCV(shape = 0.7, covariance = 0.5), VCV.vectors = TRUE, eigen.vectors = TRUE, eigen.values = TRUE, col.eigen.vectors = "red", lty.VCV.vectors = 3, lwd = 2))
-})

### 2022-10-11 Analysis/02-mcmcmcglmmm_analysis_variance_tests.Rmd
Create 02-mcmcmcglmmm_analysis_variance_tests.Rmd

--- 2022-06-30 Analysis/02-mcmcmcglmmm_analysis.Rmd
@@
 
+Making a 2D model with just the super order random terms
+
 ```{r, parametrise_the_model, eval = FALSE}
@@
                                        burnin = 0))
+save(param_MCMCglmm, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
+```
 
+## Running the parametrisation chains (for the small model)
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
 ## Running the three MCMCglmm models
-100k_params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(param_chains, file = "../Data/Processed/MCMCglmm/100k_params_shapespace_8D_31levels_chain1.rda")
+start <- Sys.time()
+params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(params_shapespace_2D_9levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_2D_9levels_20k_chain1.rda")
+end <- Sys.time()
+end - start
+```
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
+## Running the three MCMCglmm models
+start <- Sys.time()
+params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(params_shapespace_2D_9levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_2D_9levels_20k_chain2.rda")
+end <- Sys.time()
+end - start
+```
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
+## Running the three MCMCglmm models
+start <- Sys.time()
+params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(params_shapespace_2D_9levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_2D_9levels_20k_chain3.rda")
+end <- Sys.time()
+end - start
+```
+
+## (for the big model)
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k.rda")
+## Running the three MCMCglmm models
+start <- Sys.time()
+params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(params_shapespace_8D_32levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain1.rda")
+end <- Sys.time()
+end - start
+```
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k.rda")
+## Running the three MCMCglmm models
+start <- Sys.time()
+params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(params_shapespace_8D_32levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain2.rda")
+end <- Sys.time()
+end - start
+```
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k.rda")
+## Running the three MCMCglmm models
+start <- Sys.time()
+params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(params_shapespace_8D_32levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain3.rda")
+end <- Sys.time()
+end - start
 ```
@@
 ## Set up the mini-chains models
-shapespace_allbirds_lvl_superorder_order <- make.mini.chains(
+shapespace_allbirds_lvl_superorder_order_2D <- make.mini.chains(
                                 data         = shapespace,
-                                dimensions   = selected_dim,
+                                dimensions   = 1,
                                 tree         = trees_list[sample(1:1000, 1000)],
                                 trait.family = "gaussian",
-                                randoms      = c("global", "level1", "level2"),
+                                randoms      = c("global", "level1"),
                                 residuals    = "global",
@@
 ## Save the mini.chains
-save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order.mini.chains")
+save(shapespace_allbirds_lvl_superorder_order_2D, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order_2D.mini.chains")
 ```
@@
 ```
-
-## Checking the mini chains' ESSs
-
-```{r, eval = TRUE}
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-diagnose.mini.chains(shapespace_allbirds_lvl_superorder_order[[1]])
-```
-
-## Passeriformes focus
-
-```{r load_packages, eval = TRUE, message = FALSE}
-set.seed(42)
-## Loading beer
-library(mcmcmcglmmm)
-library(MCMCglmm)
-library(dispRity)
-
-## Loading the parametrisation data
-load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
-
-## Select the data
-shapespace <- shapespace_passeriformes_two_levels[[1]]$space
-selected_dim <- shapespace_passeriformes_two_levels[[1]]$dimensions
-consensus_tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
-trees_list <- shapespace_passeriformes_two_levels[[1]]$trees_list
-```
-
-## Mini-chains parametrization
-
-```{r, parametrise_the_model, eval = FALSE}
-## Set up the parametrising chains models
-param_chains <- make.mini.chains(data         = shapespace,
-                                 dimensions   = selected_dim,
-                                 tree         = consensus_tree,
-                                 trait.family = "gaussian",
-                                 randoms      = c("global", "level1", "level2"),
-                                 residuals    = "global",
-                                 priors       = 0.02,
-                                 verbose      = TRUE,
-                                 parameters   = list(
-                                       nitt   = 50000,
-                                       thin   = 500,
-                                       burnin = 0))
-
-# Running the three MCMCglmm models
-save(param_chains, file = "../Data/Processed/params_passerines_shapespace_8D_28levels_50k.rda")
-```
-
-From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
-We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
-For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
-
-```{r, get_model_parameters, eval = FALSE}
-## Load the computed models
-param_chains_out <- list()
-for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_passeriformes/param_chains/params_passerines_shapespace_8D_28levels_50k_chain", i, ".rda"))
-    param_chains_out[[i]] <- param_chains[[1]]
-}
-## Extract the parameters from the parameter chains
-est_params <- extract.parameters(param_chains_out)
-```
-
-### Running the mini-chains
-
-```{r, set_the_mini_chains, eval = FALSE}
-## The thinning parameter
-thin <- 200
-## The number of samples per chains
-samples <- 10
-
-## Set up the mini-chains models
-shapespace_passeriformes <- make.mini.chains(
-                                data         = shapespace,
-                                dimensions   = selected_dim,
-                                tree         = trees_list[sample(1:1000, 1000)],
-                                trait.family = "gaussian",
-                                randoms      = c("global", "level1", "level2"),
-                                residuals    = "global",
-                                priors       = est_params$priors,
-                                verbose      = TRUE,
-                                parameters   = list(
-                                    nitt   = est_params$burnin + thin * samples,
-                                    thin   = thin,
-                                    burnin = est_params$burnin))
-## Save the mini.chains
-save(shapespace_passeriformes, file = "../Data/Cluster/shapespace_passeriformes/shapespace_passeriformes.mini.chains")
-```
-
-### Combining the model data
-
-```{r, eval = FALSE}
-## Set up the variables
-models_path <- "../Data/Cluster/shapespace_passeriformes/cluster/"
-## Getting all the model files
-model_files <- list.files(models_path, pattern = ".rda")
-
-## Check if the model exists already
-if(!("shapespace_passeriformes.MCMCglmm" %in% list.files("../Data/Processed/"))) {
-    ## Preparing the model files
-    model_list <- list()
-    class(model_list) <- c("mini.chains")
-    ## Get the first model
-    load(paste0(models_path, model_files[1], collapse = "/"))
-    ## Move it to the storage
-    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
-    print(model_files[1])
-    ## Clean the bits that are not going to be used (but are huge)
-    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
-    ## Update the model list
-    model_list[[length(model_list)+1]] <- model[[1]]
-    model <- NULL
-    model_files <- model_files[-1]
-} else {
-    ## Updating the model
-    load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
-    model_list <- shapespace_passeriformes
-}
-
-## Get the subsequent models and merge them as we go
-while(length(model_files) != 0) {
-    load(paste0(models_path, model_files[1], collapse = "/"))
-    ## Move it to storage
-    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
-    print(model_files[1])
-    ## Clean the bits that are not going to be used (but are huge)
-    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
-    ## Update the model list
-    model_list[[length(model_list)+1]] <- model[[1]]
-    ## Merge the models (overwrite the model list)
-    model_list <- list(combine.mini.chains(model_list))
-    ## DEBUG
-    print("model list size:")
-    print(format(object.size(model_list), units = "MB"))
-    model <- NULL
-    model_files <- model_files[-1]
-}
-
-## Combining the model files
-shapespace_passeriformes <- model_list
-
-## Saving the results
-save(shapespace_passeriformes, file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
-
-```

### 2022-10-12 Analysis/03-Elaboration_innovation_analysis.Rmd
tests for variance (to be removed)

@@
 par(mar = c(5,8,2,2), mfrow = c(1, 3))
-boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
 boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+boxplot(log(variance_per_clades[,reorder_clades])/log(sample_sizes), horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing sample size effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+boxplot(log(variance_per_clades[,reorder_clades])/log(observed_variance)[names(variances)], horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing variance effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
 dev.off()
+
+## Getting the variances per clades
+variances <- apply(variance_per_clades[,reorder_clades], 2, mean)
+
+## Getting the sample sizes per clades (the one for the residuals is the whole tree size)
+sample_sizes <- c("residuals" = size.subsets(order_shapespace)[28],
+                  size.subsets(superorder_shapespace),
+                  size.subsets(order_shapespace))[names(variances)]
+
+## Measuring the variance for each clade and for the residuals
+## the obs variance of the residuals is just the variance in the whole dataset (phylogeny)
+observed_variance <- c("residuals" = unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[28], 
+                       unlist(get.disparity(dispRity(superorder_shapespace, metric = sd))),
+                       unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[-28])^2
+
+plot(log(variances)/log(sample_sizes), log(sample_sizes), xlab = "log variance / log n", ylab = "log n")
+
+x <- rnorm(100)
+y <- rnorm(100)
+z <- rnorm(100)
+
+cor(x, y)
+cor(x/y, y)
 

### 2022-10-12 Analysis/03-Elaboration_innovation_analysis.Rmd
New narrative?

--- 2022-10-07 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 reorder_clades <- c(9:35,1:8, 36, 37)
-## Pull by term
-variance_per_levels<- cbind("orders" = apply(variance_per_clades[, 9:35], 1, sum),
-                            "super-orders" = apply(variance_per_clades[, 1:8], 1, sum),
-                            "phylogeny" = variance_per_clades[, 36],
-                            "residuals" = variance_per_clades[, 37])
-variance_per_terms <- cbind("clades" = apply(variance_per_clades[, 1:35], 1, sum),
-                            "phylogeny" = variance_per_clades[, 36],
-                            "residuals" = variance_per_clades[, 37])
-## Plot the results
-pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf", height = 8, width = 14)
-par(mar = c(5,8,2,2), mfrow = c(1, 3))
-boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-dev.off()
+# ## Pull by term
+# variance_per_levels<- cbind("orders" = apply(variance_per_clades[, 9:35], 1, sum),
+#                             "super-orders" = apply(variance_per_clades[, 1:8], 1, sum),
+#                             "phylogeny" = variance_per_clades[, 36],
+#                             "residuals" = variance_per_clades[, 37])
+# variance_per_terms <- cbind("clades" = apply(variance_per_clades[, 1:35], 1, sum),
+#                             "phylogeny" = variance_per_clades[, 36],
+#                             "residuals" = variance_per_clades[, 37])
+# ## Plot the results
+# pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf", height = 8, width = 14)
+# par(mar = c(5,8,2,2), mfrow = c(1, 3))
+# # boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# # boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(log(variance_per_clades[,reorder_clades])/log(sample_sizes), horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing sample size effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(log(variance_per_clades[,reorder_clades])/log(observed_variance)[names(variances)], horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing variance effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# dev.off()
+
+# ## Getting the variances per clades
+# variances <- apply(variance_per_clades[,reorder_clades], 2, mean)
+
+# ## Getting the sample sizes per clades (the one for the residuals is the whole tree size)
+# sample_sizes <- c("residuals" = size.subsets(order_shapespace)[28],
+#                   size.subsets(superorder_shapespace),
+#                   size.subsets(order_shapespace))[names(variances)]
+
+# ## Measuring the variance for each clade and for the residuals
+# ## the obs variance of the residuals is just the variance in the whole dataset (phylogeny)
+# observed_variance <- c("residuals" = unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[28], 
+#                        unlist(get.disparity(dispRity(superorder_shapespace, metric = sd))),
+#                        unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[-28])^2
+
+# plot(log(variances)/log(sample_sizes), log(sample_sizes), xlab = "log variance / log n", ylab = "log n")
 
@@
 
+```{r}
+## Plotting the novelty for the phylo picture:
+load("../Data/Processed/ploting_data.rda")
+load("../Data/Processed/ploting_data_superorder.rda")
+load("../Data/Processed/ploting_data_order.rda")
+
+## Get the values
+elaborations_tree  <- ploting_data$median_elaborations
+innovations_tree   <- ploting_data$median_innovations
+elaborations_super <- ploting_data_superorder$median_elaborations
+innovations_super  <- ploting_data_superorder$median_innovations
+elaborations_order <- ploting_data_order$median_elaborations
+innovations_order  <- ploting_data_order$median_innovations
+
+pdf("../Manuscript/Figures/median_novelty_distributions.pdf", height = 4, width = 8)
+par(mfrow = c(1,2), bty = "n")
+## Same but boxplot version
+boxplot(cbind("Whole tree level"  = c(elaborations_tree, innovations_tree),
+              "Super-order level" = c(elaborations_super, innovations_super),
+              "Order level"       = c(elaborations_order, innovations_order)),
+        xlab = "Novelty per nested levels", ylab = "Novelty", col = colour.palette(5)[1:3])
+boxplot(cbind("Elaboration" = c(elaborations_tree, elaborations_super, elaborations_order),
+              "Innovation"  = c(innovations_tree, innovations_super, innovations_order)),
+        xlab = "Novelty per type", ylab = "Novelty", col = colour.palette(5)[4:5])
+dev.off() 
+```
+
 # Passeriformes shapespace analyses

### 2022-10-12 Manuscript/main_flipped_narrative.tex
New narrative?

--- 2022-10-07 Manuscript/main.tex
@@
 
+% TODO in abstract, add the "paradox" bit: However, if most organisms evolve incrementally along stable lines of least resistance then how can we explain the staggering diversity of species morphologies across the tree of life?
+
 \section{Main}
 
+% paragraph 1:  introducing the major axis of variation + novelty and innovation
 Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow lines of genetic least resistance \cite{schluter1996adaptive}.
-Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape \cite{jones2004evolution}.
+Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,Rombaut2022} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape \cite{jones2004evolution}.
 Indeed, evidence suggests that genetic constraints can extend over timescales spanning tens of millions of years leading to remarkable macroevolutionary stability in the orientation of phenotypic evolution \cite{mcglothlin2018adaptive}.
-However, if most organisms evolve incrementally along stable lines of least resistance then how can we explain the staggering diversity of species morphologies across the tree of life?
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the evolvability of phenotypes across scales.
-This apparent tension implies deep-time reorientation of phenotypic lines of least resistance and points to alternative routes to the exploration of phenotypic space by evolving either along, or away from, the dominant phenotypic axis.
-We refer to these alternatives as elaboration - evolution along the phylogenetically constrained phenotypic major axis - and innovation - evolution away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
-The contributions of these two multivariate pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales.
-Here we estimate and compare major axes of beak shape evolution in birds across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.
+This evolutionary major axis of phenotypic variation allows to describe the novelty of phenotypic traits made of two components: elaboration and innovation \cite{endler2005animal}.
+We can define phenotypic changes along the major axis of variation as elaboration and changes away from that line as innovation.
+And thus by definition, we would suspect that elaboration is a major driver of morphological evolution across the time (because it's equal to evolution along that line).
+In this paper we will test this hypothesis and measure whether elaboration is more common across the tree of life.
+%This is basically a fake hypothesis - although we're testing that, the plot twist happens later!
 
+% paragraph 3: how we measure this (glmm) on the birds data + nestedness
 Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
@@
 All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-We fitted a phylogenetic generalised linear mixed effect model (pGLMM) with beak shape as an eight-dimensional response variable and treated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
+We define the major axis of variation as the major axis of traits variance-covariance matrices for these bird beaks.
+And to test our hypothesis (is elaboration more common across the tree of life) we need to  measure bird beaks' novelty at different nested phylogenetic levels.
+We will this calculate this variance-covariances matrices in a nested way across all birds: first for all birds, then for all super-orders and then for each order.
+To measure these variance-covariance matrices, we fitted a phylogenetic generalised linear mixed effect model (pGLMM) with beak shape as an eight-dimensional response variable and treated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
 We regard the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
 The nested partitions of multivariate trait space provide the basis for all subsequent analyses.
+We then measure innovation and elaboration as the multilinear algebra projection of each species onto their nested major axis of variation.
 
-Adaptive radiation theory implies that phenotypic divergence arises primarily along a dominant phenotypic axis (the line of least evolutionary resistance; \cite{marroig2005size}).
-However, this contrasts with observations of major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
+% paragraph 4: results for elaboration and innovation for species
+Here are the results for each species. \ref{Fig:phylogeny}
+There seems to be a nested structure of the variation in novelty %TODO: needs to introduce nestedness; %TODO: needs to introduce novelty.
+There is more variation the deeper you baseline comparison in the tree is, suggesting that novelty is relative to the baseline comparison.
+Also, the majority of that novelty is through elaboration.
+We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
+However, we also found various combinations of elaboration and innovation across birds.
+For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
+Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
+Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
+The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
+\ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
+ and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
+The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
+\textbf{But there is a paradox here:}
+If most organisms evolve incrementally along major axis of variation then how can we explain the staggering diversity of species morphologies across the tree of life?
+
+\begin{figure}[!htbp]
+\centering
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
+\caption{.}
+\label{Fig:phylogeny}
+\end{figure}
+
+\begin{figure}[!htbp]
+\centering
+    \includegraphics[width=0.9\textwidth]{Figures/median_novelty_distributions.pdf}
+\caption{THIS FIGURE SHOULD BE EMBEDED IN THE ANUS FIGURE: showing the novelty grouped into two different ways}
+\label{Fig:phylogeny_2}
+\end{figure}
+
+\bigskip
+
+\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
+Blue and yellow colors highlight species with relatively low novelty whereas orange to red colours indicate high novelty.
+The novelty is represented as innovation (upper circle) and elaboration (lower circle) concentrically represented in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing novelty for each species at the order levels then at the super-order level and finally at the entire phylogeny level (@@@ THIS SHOULD POINT TO THE ADDITIONAL PLOTS TO BE ADDED IN THE FIGURE).
+We can observe an increase of novelty at deeper phylogenetic with novelty consistently composed of more elaboration than innovation.
+%TODO: add density plots of the innovations and elaboration at different levels.
+The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
+
+% Paragraph 5: paradox + methods for comparing ellipses
+Although the adaptive radiation theory implies that phenotypic divergence arises primarily along the major axis of variation \cite{marroig2005size}, there has been many contrasting observations in the literature highlighting major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
 We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different clades compared to the global phylogenetic major axis.
@@
 These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
+
+% Paragraph 5+: Some elluipses stuff with bird names.
+However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
+While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
+To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
+As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
+Clades (or species) can be both elaborators and innovators.
+To measure innovation and elaboration we translated the major axes of each clade's (i.e.
+superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
+We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
+We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
+These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
+We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
+Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
+However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
+Cuculiformes; Fig.
+\ref{Fig:ellipses}), low elaboration and high innovation (e.g.
+Galliformes; Fig.
+\ref{Fig:ellipses}), high elaboration and low innovation (e.g.
+Coraciiformes; Fig.
+\ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g.
+Podicipediformes; Fig.
+\ref{Fig:ellipses}).
+These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
+This suggests that divergence away from line of least evolutionary resistance (i.e.
+the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
+
+This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
+Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig.
+\ref{Fig:ellipses}-bar plots).
+These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
+S\ref{fig_ellipses_passeriformes}).
+Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
+For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
+However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
 
@@
 
-\bigskip
-
-However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
-While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
-To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
-As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
-Clades (or species) can be both elaborators and innovators.
-To measure innovation and elaboration we translated the major axes of each clade's (i.e.
-superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
-We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
-We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
-These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
-Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
-However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
-Cuculiformes; Fig.
-\ref{Fig:ellipses}), low elaboration and high innovation (e.g.
-Galliformes; Fig.
-\ref{Fig:ellipses}), high elaboration and low innovation (e.g.
-Coraciiformes; Fig.
-\ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g.
-Podicipediformes; Fig.
-\ref{Fig:ellipses}).
-These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
-This suggests that divergence away from line of least evolutionary resistance (i.e.
-the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
-
-This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
-Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig.
-\ref{Fig:ellipses}-bar plots).
-These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
-S\ref{fig_ellipses_passeriformes}).
-Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
-For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
-However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
-
-Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
-This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
-To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shapespace onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
-We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire clades.
-We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
-However, we also found various combinations of elaboration and innovation across birds.
-For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
-Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
-Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
-The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
-\ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
- and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
-
-\begin{figure}[!htbp]
-\centering
-    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
-\caption{.}
-\label{Fig:phylogeny}
-\end{figure}
-
-\bigskip
-
-\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
-Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
-Innovations and elaboration are displayed as: a) comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
-The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
-A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
-
-\bigskip
+% Paragraph 7: putting these results together
 
@@
 
-\bigskip
-
+% Paragraph 8: conculsion
 Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.

### 2022-10-17 Analysis/02-mcmcmcglmmm_analysis_variance_tests.Rmd
Update 02-mcmcmcglmmm_analysis_variance_tests.Rmd

@@
 params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_2D_9levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_2D_9levels_20k_chain1.rda")
+save(params_shapespace_2D_9levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain1.rda")
 end <- Sys.time()
@@
 params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_2D_9levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_2D_9levels_20k_chain2.rda")
+save(params_shapespace_2D_9levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain2.rda")
 end <- Sys.time()
@@
 params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_2D_9levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_2D_9levels_20k_chain3.rda")
+save(params_shapespace_2D_9levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain3.rda")
 end <- Sys.time()
@@
 params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_8D_32levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain1.rda")
+save(params_shapespace_8D_32levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain1.rda")
 end <- Sys.time()
@@
 params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_8D_32levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain2.rda")
+save(params_shapespace_8D_32levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain2.rda")
 end <- Sys.time()
@@
 params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_8D_32levels_20k, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain3.rda")
+save(params_shapespace_8D_32levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain3.rda")
 end <- Sys.time()
@@
 for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/shapespace_allbirds_lvl_superorder_order_param", i, ".rda"))
-    param_chains[[i]] <- shapespace_allbirds_lvl_superorder_order_param[[1]]
+    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain", i, ".rda"))
+    param_chains[[i]] <- params_shapespace_2D_9levels_20k[[1]]
 }
@@
                                 data         = shapespace,
-                                dimensions   = 1,
+                                dimensions   = 1:2,
                                 tree         = trees_list[sample(1:1000, 1000)],
@@
 TODO: details
+
+```{r}
+library(mcmcmcglmmm)
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order_2D.mini.chains")
+## Running the three MCMCglmm models
+start <- Sys.time()
+shapespace_allbirds_lvl_superorder_order_2D <- run.mini.chains(shapespace_allbirds_lvl_superorder_order_2D, replicates = 20)
+save(shapespace_allbirds_lvl_superorder_order_2D, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain1_20.rda")
+end <- Sys.time()
+end - start
+```
 

### 2022-10-17 Analysis/02-mcmcmcglmmm_analysis.Rmd
Update 02-mcmcmcglmmm_analysis.Rmd

@@
 ## Running the three MCMCglmm models
-100k_params_shapespace_8D_31levels <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(param_chains, file = "../Data/Processed/MCMCglmm/100k_params_shapespace_8D_31levels_chain1.rda")
+params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
+save(param_chains, file = "../Data/Processed/MCMCglmm/params_shapespace_8D_32levels_20k_chain1.rda")
 ```
@@
 for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/shapespace_allbirds_lvl_superorder_order_param", i, ".rda"))
-    param_chains[[i]] <- shapespace_allbirds_lvl_superorder_order_param[[1]]
+    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain", i, ".rda"))
+    param_chains[[i]] <- params_shapespace_8D_32levels_20k[[1]]
 }
@@
 
-The chains are then run on the cluster
-TODO: details
+The chains are then run on the cluster for 400 replicates.
+
+```{r, eval = FALSE}
+## Don't run:
+run.mini.chains(shapespace_allbirds_lvl_superorder_order, replicates = 400)
+```
+
+Note that the previous line is running the 400 replicates in a serial way, it is much more efficient to run each one separately in parallel across 400 CPUs.
+Each individual replicate runs for around 50 hours on a 2.4Ghz CPU and requires around 9GB of RAM.
+The whole process was completed in around 2 CPU years.
 
 ### Combining the model data
+
+The following script was used to combine the 400 mini chains into the combined chain used for further analysis.
+Note that this script is relying on relative paths that might need to be set up before running and are probably dependent on different clusters' architectures.
 
@@
 
-## Passeriformes focus
+# Passeriformes focus
+
+The following section repeats every step as highlighted above but applied to the passeriformes as a subset of the entire dataset.
 
@@
 ```
-
-From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
-We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
-For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
 

### 2022-10-17 Analysis/03-Elaboration_innovation_analysis.Rmd
Fromaly added a variance test part

--- 2022-10-07 Analysis/03-Elaboration_innovation_analysis.Rmd
@@
 
-## Get the variance for each level
-variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = TRUE)
-## Rename the columns
-colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
-reorder_clades <- c(9:35,1:8, 36, 37)
-## Pull by term
-variance_per_levels<- cbind("orders" = apply(variance_per_clades[, 9:35], 1, sum),
-                            "super-orders" = apply(variance_per_clades[, 1:8], 1, sum),
-                            "phylogeny" = variance_per_clades[, 36],
-                            "residuals" = variance_per_clades[, 37])
-variance_per_terms <- cbind("clades" = apply(variance_per_clades[, 1:35], 1, sum),
-                            "phylogeny" = variance_per_clades[, 36],
-                            "residuals" = variance_per_clades[, 37])
-## Plot the results
-pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf", height = 8, width = 14)
-par(mar = c(5,8,2,2), mfrow = c(1, 3))
-boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-dev.off()
-
 ## Get the levels
@@
 rm(shapespace_allbirds_lvl_superorder_order)
-
-## Get the number of observations per groups
-variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = FALSE)
-colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
-
-## Converting into standard errors
-get.se <- function(var, n) {
-    sqrt(var[[1]])/sqrt(n)
-}
-## Get the se per group
-se_per_clades <- mapply(get.se, apply(variance_per_clades, 2, list), as.list(c(size.subsets(superorder_shapespace)[-9], size.subsets(order_shapespace), "residuals" = 8748)))
-## Scale that
-se_per_clades <- se_per_clades/rowSums(se_per_clades)
-## Get the other levels
-se_per_levels<- cbind("orders" = apply(se_per_clades[, 9:35], 1, sum),
-                      "super-orders" = apply(se_per_clades[, 1:8], 1, sum),
-                      "phylogeny" = se_per_clades[, 36],
-                      "residuals" = se_per_clades[, 37])
-se_per_terms <- cbind("clades" = apply(se_per_clades[, 1:35], 1, sum),
-                      "phylogeny" = se_per_clades[, 36],
-                      "residuals" = se_per_clades[, 37])
-
-## Plot the results
-pdf(file = "../Manuscript/Figures/se_per_levels_whole_tree.pdf", height = 8, width = 14)
-par(mar = c(5,8,2,2), mfrow = c(1, 3))
-boxplot(se_per_terms, horizontal = TRUE, las = 1, main = "Scaled standard error by terms", xlab = "Scaled se ")
-boxplot(se_per_levels, horizontal = TRUE, las = 1, main = "Scaled standard error by levels", xlab = "Scaled se")
-boxplot(se_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled standard error by clades", xlab = "Scaled variance")
-dev.off()
 ```

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Fromaly added a variance test part

--- 2022-10-17 Analysis/02-mcmcmcglmmm_analysis_variance_tests.Rmd
@@
 ```
+
+# Analysing the variance per terms
+
+```{r}
+## Get the variance for each level
+variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = TRUE)
+## Rename the columns
+colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
+reorder_clades <- c(9:35,1:8, 36, 37)
+# ## Pull by term
+# variance_per_levels<- cbind("orders" = apply(variance_per_clades[, 9:35], 1, sum),
+#                             "super-orders" = apply(variance_per_clades[, 1:8], 1, sum),
+#                             "phylogeny" = variance_per_clades[, 36],
+#                             "residuals" = variance_per_clades[, 37])
+# variance_per_terms <- cbind("clades" = apply(variance_per_clades[, 1:35], 1, sum),
+#                             "phylogeny" = variance_per_clades[, 36],
+#                             "residuals" = variance_per_clades[, 37])
+# ## Plot the results
+# pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf", height = 8, width = 14)
+# par(mar = c(5,8,2,2), mfrow = c(1, 3))
+# # boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# # boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(log(variance_per_clades[,reorder_clades])/log(sample_sizes), horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing sample size effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# boxplot(log(variance_per_clades[,reorder_clades])/log(observed_variance)[names(variances)], horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing variance effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
+# dev.off()
+
+# ## Getting the variances per clades
+# variances <- apply(variance_per_clades[,reorder_clades], 2, mean)
+
+# ## Getting the sample sizes per clades (the one for the residuals is the whole tree size)
+# sample_sizes <- c("residuals" = size.subsets(order_shapespace)[28],
+#                   size.subsets(superorder_shapespace),
+#                   size.subsets(order_shapespace))[names(variances)]
+
+# ## Measuring the variance for each clade and for the residuals
+# ## the obs variance of the residuals is just the variance in the whole dataset (phylogeny)
+# observed_variance <- c("residuals" = unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[28], 
+#                        unlist(get.disparity(dispRity(superorder_shapespace, metric = sd))),
+#                        unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[-28])^2
+
+# plot(log(variances)/log(sample_sizes), log(sample_sizes), xlab = "log variance / log n", ylab = "log n")
+
+```
+
+## Standard error per terms
+
+```{r}
+
+## Get the number of observations per groups
+variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = FALSE)
+colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
+
+## Converting into standard errors
+get.se <- function(var, n) {
+    sqrt(var[[1]])/sqrt(n)
+}
+## Get the se per group
+se_per_clades <- mapply(get.se, apply(variance_per_clades, 2, list), as.list(c(size.subsets(superorder_shapespace)[-9], size.subsets(order_shapespace), "residuals" = 8748)))
+## Scale that
+se_per_clades <- se_per_clades/rowSums(se_per_clades)
+## Get the other levels
+se_per_levels<- cbind("orders" = apply(se_per_clades[, 9:35], 1, sum),
+                      "super-orders" = apply(se_per_clades[, 1:8], 1, sum),
+                      "phylogeny" = se_per_clades[, 36],
+                      "residuals" = se_per_clades[, 37])
+se_per_terms <- cbind("clades" = apply(se_per_clades[, 1:35], 1, sum),
+                      "phylogeny" = se_per_clades[, 36],
+                      "residuals" = se_per_clades[, 37])
+
+## Plot the results
+pdf(file = "../Manuscript/Figures/se_per_levels_whole_tree.pdf", height = 8, width = 14)
+par(mar = c(5,8,2,2), mfrow = c(1, 3))
+boxplot(se_per_terms, horizontal = TRUE, las = 1, main = "Scaled standard error by terms", xlab = "Scaled se ")
+boxplot(se_per_levels, horizontal = TRUE, las = 1, main = "Scaled standard error by levels", xlab = "Scaled se")
+boxplot(se_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled standard error by clades", xlab = "Scaled variance")
+dev.off()
+```

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

@@
 ---
+
+This script explores the meaning of the variance explained in big multidimensional MCMCglmm with variable sample sizes.
+
+# Variance explained in a simple model
+
+# More complex model with equal sample size
+
+# Model with different sample sizes
+
+# Model used in this paper (8D + 32 terms)
+
+Here the model is highly dimensional and the terms sample sizes vary by two orders of magnitude.
 

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

@@
 
+```{r}
+library(dispRity)
+library(mcmcmcglmmm)
+library(dads)
+```
+
 # Variance explained in a simple model
 
+Here I'm just doing a one dimensional model of a one dimensional trait (BM) and two clades (`"s"` and `"t"`).
+
+```{r}
+## Setting the data
+set.seed(0)
+data_1 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
+data_2 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
+
+## Renaming the tips and data in data_2
+rownames(data_2$data)[match(data_2$tree$tip.label, rownames(data_2$data))] <- gsub("t", "s", data_2$tree$tip.label)
+data_2$tree$tip.label <- gsub("t", "s", data_2$tree$tip.label)
+data_1$tree$node.label <- data_2$tree$node.label <- NULL
+
+## Combine the trees
+ultrametric_tree <- chronos(multi2di(bind.tree(data_1$tree, data_2$tree)))
+class(ultrametric_tree) <- "phylo"
+tree <- slide.nodes(ultrametric_tree, node = 42, slide = 0.2)
+
+## Combine the data
+data <- c(data_1$data[match(data_1$tree$tip.label, rownames(data_1$data)), ], data_2$data[match(data_2$tree$tip.label, rownames(data_2$data)), ])
+data <- as.data.frame(matrix(data, ncol = 1, dimnames = list(c(names(data)), c("variable"))))
+data <- cbind(data, "var2" = rnorm(40), "clade" = as.factor(c(rep("t", 20), rep("s", 20))), "animal" = rownames(data))
+```
+
+We will now run it as the following model:
+
+```
+data ~ residuals + random.phylo + random.clade_s + random.clade_t
+```
+
+```{r}
+## Setting up the model
+first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))
+
+## running the first model
+first_model_run <- run.mini.chains(first_model, replicates = 1)[[1]]
+```
+
+And this is what it looks like graphically
+
+```{r, echo = FALSE}
+## Model object
+first_model_results <- MCMCglmm.subsets(
+      data          = data,
+      posteriors    = first_model_run,
+      group         = MCMCglmm.levels(first_model_run)[c(1:3)],
+      rename.groups = c("clade_s", "clade_t", "phylogeny"))
+
+## Set the clade colours
+clade_cols <- c("clade_s" = "blue", "clade_t" = "orange", "phylogeny" = "grey")
+
+## Plotting the model
+covar.plot(first_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
+```
+We can then extract the variance of the model:
+
+```{r}
+plot.decompose.var <- function(data, posteriors, group.cols, obs.col = "red", relative = FALSE, ...) {
+    ## Get the model variance
+    model_variance <- MCMCglmm.variance(posteriors, scale = relative)
+    colnames(model_variance) <- c(name.subsets(data), "residuals")
+    ## Get the data variance
+    if(!relative) {
+        obs_var <- sd(unlist(data$matrix[[1]][, c(data$call$dimensions)]))^2
+    }
+    var_name <- ifelse(relative, "relative variance", "variance")
+    if(missing(group.cols)) {
+        group_cols <- rep("grey", ncol(model_variance))
+    }
+
+    ## Plotting all these with logs
+    par(mfrow = c(2,2))
+
+    ## Overall results
+    boxplot(model_variance, main = paste0(var_name, " per terms"), xlab = var_name, col = group.cols, horizontal = TRUE, las = 1)
+    if(!relative) {
+        abline(v = obs_var, col = obs.col, lwd = 2)
+    }
+    legend("topright", legend = "data variance", col = obs.col, lty = 1, lwd = 2)
+
+    ## Logged results
+    boxplot(apply(model_variance, 2, log), main = paste0(var_name, " per terms (logged)"), xlab = paste0(var_name, "(logged)"), col = group.cols, horizontal = TRUE, las = 1)
+    if(!relative) {
+        abline(v = log(obs_var), col = obs.col, lwd = 2)
+    }
+
+    ## Results by sample size
+    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
+    median_variances <- apply(model_variance, 2, median)    
+    plot(y = median_variances, x = sample_sizes, ylab = paste0("median ", var_name), xlab = "sample size", pch = 19, col =  group.cols)
+    if(!relative) {
+        points(y = obs_var, x = sample_sizes[length(sample_sizes)], col = "red", pch = 19)
+    }
+    model <- lm(median_variances ~ sample_sizes)
+    abline(model$coefficients)
+    fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
+    legend("topright", legend = fit, lty = 1)
+
+    ## Results log logged
+    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
+    median_variances <- apply(model_variance, 2, median)    
+    plot(y = log(median_variances), x = log(sample_sizes), ylab = paste0("median ", var_name, " (log)"), xlab = "sample size (log)", pch = 19, col =  group.cols)
+    if(!relative) {
+        points(y = log(obs_var), x = log(sample_sizes[length(sample_sizes)]), col = "red", pch = 19)
+    }
+    model <- lm(log(median_variances) ~ log(sample_sizes))
+    abline(model$coefficients)
+    fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
+    legend("topright", legend = fit, lty = 1)
+}
+
+plot.decompose.var(first_model_results, first_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), relative = FALSE)
+```
+
 # More complex model with equal sample size
+
+For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (but equal-ish) but based on biological data.
+
+```{r}
+data(charadriiformes)
+## Same but selecting only the three first random terms
+chara_data <- MCMCglmm.subsets(data       = charadriiformes$data,
+                               posteriors = charadriiformes$posteriors,
+                               tree       = charadriiformes$tree,
+                               group      = MCMCglmm.levels(charadriiformes$posteriors)[1:4],
+                               rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
+```
+
+This is what it looks like:
+
+```{r}
+## Set up the clade colours
+clade_cols <- c("gulls" = "blue", "plovers" = "orange", "sandpipers" = "darkgreen", "phylogeny" = "grey")
+## Plot the model
+covar.plot(chara_data, col = clade_cols, ellipses = mean, legend = TRUE, lwd = 2, points.cex = 0.5)
+```
+
+And this is what the variances look like:
+
+```{r}
+plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"))
+```
 

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

--- 2022-10-17 Analysis/11-variance_partitioning.Rmd
@@
 
+```{r}
+library(dispRity)
+library(mcmcmcglmmm)
+library(dads)
+```
+
 # Variance explained in a simple model
+
+Here I'm just doing a one dimensional model of a one dimensional trait (BM) and two clades (`"s"` and `"t"`).
+
+```{r, echo = FALSE}
+## Setting the data
+set.seed(0)
+data_1 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
+data_2 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
+
+## Renaming the tips and data in data_2
+rownames(data_2$data)[match(data_2$tree$tip.label, rownames(data_2$data))] <- gsub("t", "s", data_2$tree$tip.label)
+data_2$tree$tip.label <- gsub("t", "s", data_2$tree$tip.label)
+data_1$tree$node.label <- data_2$tree$node.label <- NULL
+
+## Combine the trees
+ultrametric_tree <- chronos(multi2di(bind.tree(data_1$tree, data_2$tree)))
+class(ultrametric_tree) <- "phylo"
+tree <- slide.nodes(ultrametric_tree, node = 42, slide = 0.2)
+
+## Combine the data
+data <- c(data_1$data[match(data_1$tree$tip.label, rownames(data_1$data)), ], data_2$data[match(data_2$tree$tip.label, rownames(data_2$data)), ])
+data <- as.data.frame(matrix(data, ncol = 1, dimnames = list(c(names(data)), c("variable"))))
+data <- cbind(data, "var2" = rnorm(40), "clade" = as.factor(c(rep("t", 20), rep("s", 20))), "animal" = rownames(data))
+```
+
+We will now run it as the following model:
+
+```
+data ~ residuals + random.phylo + random.clade_s + random.clade_t
+```
+
+```{r, echo = FALSE, message = FALSE}
+## Setting up the model
+first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))
+
+## running the first model
+first_model_run <- run.mini.chains(first_model, replicates = 1)[[1]]
+```
+
+And this is what it looks like graphically
+
+```{r, echo = FALSE}
+## Model object
+first_model_results <- MCMCglmm.subsets(
+      data          = data,
+      posteriors    = first_model_run,
+      group         = MCMCglmm.levels(first_model_run)[c(1:3)],
+      rename.groups = c("clade_s", "clade_t", "phylogeny"))
+
+## Set the clade colours
+clade_cols <- c("clade_s" = "blue", "clade_t" = "orange", "phylogeny" = "grey")
+
+## Plotting the model
+covar.plot(first_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
+```
+We can then extract the variance of the model:
+
+```{r, echo = FALSE}
+#@ name plot.decompose.var
+#@ description plotting the variance of a model in different ways
+#@ param data a dispRity object
+#@ param posteriors a MCMCglmm object
+#@ param group.cols the colours for each term
+#@ param obs.col the colour for the observed variance
+#@ param relative whether to make the variance relative or absolute
+#@ param plot which panels to plot. Can be any of the followings (default is all):
+# "var boxplot" = variance per group;
+# "var by n" = variance per sample size;
+# "log var boxplot" = logged variance per group;
+# "log var by n" = logged variance per sample size;
+# "corrected var boxplot" = logged variance divided by logged sample sizel;
+# "corrected var by n" = logged variance per logged sample size divided by logged sample size;
+# "standard error" = variance per group divided by sqrt(n);
+# "log standard error" = logged standard error;
+
+plot.decompose.var <- function(data, posteriors, group.cols, obs.col = "red", relative = FALSE, plot = c("var boxplot", "var by n", "log var boxplot", "log var by n", "corrected var boxplot", "corrected var by n", "standard error", "log standard error"), ...) {
+    ## Get the model variance
+    model_variance <- MCMCglmm.variance(posteriors, scale = relative)
+    colnames(model_variance) <- c(name.subsets(data), "residuals")
+    ## Get the data variance
+    if(!relative) {
+        obs_var <- sd(unlist(data$matrix[[1]][, c(data$call$dimensions)]))^2
+    }
+    var_name <- ifelse(relative, "relative variance", "variance")
+    if(missing(group.cols)) {
+        group_cols <- rep("grey", ncol(model_variance))
+    }
+    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
+    median_variances <- apply(model_variance, 2, median)
+
+    ## Plotting all these with logs
+    if(length(plot) > 1) {
+        par(mfrow = c(ceiling(length(plot)/2), 2))
+    }
+
+    ## Overall results
+    if("var boxplot" %in% plot) {
+        boxplot(model_variance, main = paste0(var_name, " per terms"), xlab = var_name, col = group.cols, horizontal = TRUE, las = 1)
+        if(!relative) {
+            abline(v = obs_var, col = obs.col, lwd = 2)
+        }
+        legend("topright", legend = "data variance", col = obs.col, lty = 1, lwd = 2)
+    }
+
+    ## Results by sample size
+    if("var by n" %in% plot) {
+        plot(y = median_variances, x = sample_sizes, ylab = paste0("median ", var_name), xlab = "sample size", pch = 19, col =  group.cols, main = paste0(var_name, " by sample size"))
+        if(!relative) {
+            points(y = obs_var, x = sample_sizes[length(sample_sizes)], col = obs.col, pch = 21, cex = 2)
+        }
+        model <- lm(median_variances ~ sample_sizes)
+        fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
+        legend("topright", legend = fit, lty = 1)
+    }
+
+    ## Logged results
+    if("log var boxplot" %in% plot) {
+        boxplot(apply(model_variance, 2, log), main = paste0(var_name, " per terms (logged)"), xlab = paste0(var_name, "(logged)"), col = group.cols, horizontal = TRUE, las = 1)
+        if(!relative) {
+            abline(v = log(obs_var), col = obs.col, lwd = 2)
+        }
+    }
+
+    ## Results log logged
+    if("log var by n" %in% plot) {
+        plot(y = log(median_variances), x = log(sample_sizes), ylab = paste0("median ", var_name, " (log)"), xlab = "sample size (log)", pch = 19, col =  group.cols, main = paste0("log ", var_name, " by sample size"))
+        if(!relative) {
+            points(y = log(obs_var), x = log(sample_sizes[length(sample_sizes)]), col = obs.col, pch = 21, cex = 2)
+        }
+        model <- lm(log(median_variances) ~ log(sample_sizes))
+        abline(model$coefficients)
+        fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
+        legend("topright", legend = fit, lty = 1)
+    }
+
+    ## Boxplots of log logged sample size corrected results
+    if("corrected var boxplot" %in% plot) {
+        boxplot(apply(model_variance, 2, log)/log(sample_sizes), main = paste0(var_name, " per terms (logged)\ndivided by log(sample size)"), xlab = paste0(var_name, "(logged)/log(n)"), col = group.cols, horizontal = TRUE, las = 1)
+        if(!relative) {
+            abline(v = log(obs_var)/log(sample_sizes[length(sample_sizes)]), col = obs.col, lwd = 2)
+        }
+    }
+
+    ## Results log logged / log sample size
+    if("corrected var by n" %in% plot) {
+        plot(y = log(median_variances)/log(sample_sizes), x = log(sample_sizes), ylab = paste0("median ", var_name, " (log) / log(sample size)"), xlab = "sample size (log)", pch = 19, col =  group.cols, main = paste0("corrected", var_name, " by sample size"))
+        if(!relative) {
+            points(y = log(obs_var)/log(sample_sizes[length(sample_sizes)]), x = log(sample_sizes[length(sample_sizes)]), col = obs.col, pch = 21, cex = 2)
+        }
+        model <- lm(log(median_variances)/log(sample_sizes) ~ log(sample_sizes))
+        abline(model$coefficients)
+        fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
+        legend("topright", legend = fit, lty = 1)
+    }
+
+    ## Boxplot of standard error
+    if("standard error" %in% plot) {
+        boxplot(model_variance/sqrt(sample_sizes), main = "Standard error per terms", xlab = "standard error", col = group.cols, horizontal = TRUE, las = 1)
+        if(!relative) {
+            abline(v = obs_var/sqrt(sample_sizes[length(sample_sizes)]), col = obs.col, lwd = 2)
+        }
+    }
+
+    if("log standard error" %in% plot) {
+        boxplot(log(model_variance/sqrt(sample_sizes)), main = "Standard error per terms (logged)", xlab = "log standard error", col = group.cols, horizontal = TRUE, las = 1)
+        if(!relative) {
+            abline(v = log(obs_var/sqrt(sample_sizes[length(sample_sizes)])), col = obs.col, lwd = 2)
+        }
+    }
+}
+```
+
+```{r, echo = FALSE}
+## Plotting all the variance elements
+plot.decompose.var(first_model_results, first_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = 1:8, relative = FALSE)
+```
 
@@
 
+For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (but equal-ish) but based on biological data.
+
+```{r, echo = FALSE}
+data(charadriiformes)
+## Same but selecting only the three first random terms
+chara_data <- MCMCglmm.subsets(data       = charadriiformes$data,
+                               posteriors = charadriiformes$posteriors,
+                               tree       = charadriiformes$tree,
+                               group      = MCMCglmm.levels(charadriiformes$posteriors)[1:4],
+                               rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
+```
+
+This is what it looks like:
+
+```{r, echo = FALSE}
+## Set up the clade colours
+clade_cols <- c("gulls" = "blue", "plovers" = "orange", "sandpipers" = "darkgreen", "phylogeny" = "grey")
+## Plot the model
+covar.plot(chara_data, col = clade_cols, ellipses = mean, legend = TRUE, lwd = 2, points.cex = 0.5)
+```
+
+And this is what the variances look like:
+
+```{r, echo = FALSE}
+plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"))
+```
+
 # Model with different sample sizes
+
+This is the big model run just on 2D and only with the super orders levels.
+The differences with the true results are due to it:
+ * not being the same dataset!
+ * not having converged...
+
+```{r, echo = FALSE}
+## Load the models from the local files
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain1_20.rda")
+model_1_20 <- shapespace_allbirds_lvl_superorder_order_2D
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain21_40.rda")
+model_21_40 <- shapespace_allbirds_lvl_superorder_order_2D
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain41_60.rda")
+model_41_60 <- shapespace_allbirds_lvl_superorder_order_2D
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain61_80.rda")
+model_61_80 <- shapespace_allbirds_lvl_superorder_order_2D
+load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain81_100.rda")
+model_81_100 <-shapespace_allbirds_lvl_superorder_order_2D
+## Combine all the models
+model <- combine.mini.chains(unlist(list(model_1_20, model_21_40, model_41_60, model_61_80, model_81_100), recursive = FALSE))
+
+## Load the data
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
+## Select the data
+shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
+## Clean garbage
+rm(shapespace_allbirds_lvl_superorder_order)
+
+## Set up the dispRity object
+birds_data <- MCMCglmm.subsets(data       = shapespace[, -c(3:8)],
+                               posteriors = model,
+                               group      = MCMCglmm.levels(model)[-10],
+                               rename.groups = c(levels(shapespace$level1)[-1], "phylogeny"))
+## Set up the clade colours
+load("../Data/Processed/tip_colours_super_orders.rda")
+clade_cols <- c(tip_colours_super_orders, "phylogeny" = "grey")
+## Plot the model
+covar.plot(birds_data, col = clade_cols, ellipses = mean, legend = TRUE, lwd = 2, points.cex = 0.5)
+```
+
+```{r, echo = FALSE}
+## Plotting the variance with different data
+plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"))
+```
 
@@
 
-Here the model is highly dimensional and the terms sample sizes vary by two orders of magnitude.
-
-This script is a working script to run the chains on the SHARC Sheffield cluster.
-
-```{r load_packages, eval = TRUE, message = FALSE}
-set.seed(42)
-## Loading beer
-library(mcmcmcglmmm)
-library(MCMCglmm)
-library(dispRity)
-
-## Loading the parametrisation data
+```{r, echo = FALSE}
+## Load the data
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
-
 ## Select the data
 shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
-selected_dim <- shapespace_allbirds_lvl_superorder_order[[1]]$dimensions
-consensus_tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
-trees_list <- shapespace_allbirds_lvl_superorder_order[[1]]$trees_list
+## Clean garbage
+rm(shapespace_allbirds_lvl_superorder_order)
+
+## Load the posteriors
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
+model <- shapespace_allbirds_lvl_superorder_order[[1]]
+
+## Get the colourss
+load("../Data/Processed/tip_colours_orders.rda")
+clade_cols <- c(tip_colours_super_orders, tip_colours_orders, "phylogeny" = "grey")
+
+## Set up the dispRity object
+full_model <- MCMCglmm.subsets(data       = shapespace,
+                               posteriors = model,
+                               group      = MCMCglmm.levels(model)[-37],
+                               rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny"))
+
+covar.plot(full_model, col = clade_cols, ellipses = mean, legend = TRUE, lines.lty = c(rep(1, length(levels(shapespace$level1)[-1])), rep(2, length(levels(shapespace$level2)[-1])), 1), points = FALSE)
 ```
 
-# All birds shapespace analyses
-
-## Mini-chains parametrization
-
-Making a 2D model with just the super order random terms
-
-```{r, parametrise_the_model, eval = FALSE}
-## Set up the parametrising chains models
-param_MCMCglmm <- make.mini.chains(data         = shapespace,
-                                   dimensions   = selected_dim,
-                                   tree         = consensus_tree,
-                                   trait.family = "gaussian",
-                                   randoms      = c("global", "level1", "level2"),
-                                   residuals    = "global",
-                                   priors       = 0.02,
-                                   verbose      = TRUE,
-                                   parameters   = list(
-                                       nitt   = 20000,
-                                       thin   = 500,
-                                       burnin = 0))
-save(param_MCMCglmm, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
+```{r, echo = FALSE}
+## Plotting the variance with different data
+plot.decompose.var(full_model, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = 4)
 ```
-
-## Running the parametrisation chains (for the small model)
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
-## Running the three MCMCglmm models
-start <- Sys.time()
-params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_2D_9levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain1.rda")
-end <- Sys.time()
-end - start
-```
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
-## Running the three MCMCglmm models
-start <- Sys.time()
-params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_2D_9levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain2.rda")
-end <- Sys.time()
-end - start
-```
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k.rda")
-## Running the three MCMCglmm models
-start <- Sys.time()
-params_shapespace_2D_9levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_2D_9levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain3.rda")
-end <- Sys.time()
-end - start
-```
-
-## (for the big model)
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k.rda")
-## Running the three MCMCglmm models
-start <- Sys.time()
-params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_8D_32levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain1.rda")
-end <- Sys.time()
-end - start
-```
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k.rda")
-## Running the three MCMCglmm models
-start <- Sys.time()
-params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_8D_32levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain2.rda")
-end <- Sys.time()
-end - start
-```
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k.rda")
-## Running the three MCMCglmm models
-start <- Sys.time()
-params_shapespace_8D_32levels_20k <- run.mini.chains(param_MCMCglmm, replicates = 1)
-save(params_shapespace_8D_32levels_20k, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_8D_32levels_20k_chain3.rda")
-end <- Sys.time()
-end - start
-```
-
-From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
-We then extracted the posterior estimates of the R-Structure, G-Structure and fixed effects (mean and covariance) (ignoring the runs from the previously estimated burnin periods) to serve as our priors for our mini-chains.
-For those values, we took the highest burnin as the overall burnin and the median of the three posteriors as the overall priors.
-
-```{r, get_model_parameters, eval = FALSE}
-## Load the computed models
-param_chains <- list()
-for(i in 1:3) {
-    load(paste0("../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain", i, ".rda"))
-    param_chains[[i]] <- params_shapespace_2D_9levels_20k[[1]]
-}
-## Extract the parameters from the parameter chains
-est_params <- extract.parameters(param_chains)
-```
-
-### Running the mini-chains
-
-```{r, set_the_mini_chains, eval = FALSE}
-## The thinning parameter
-thin <- 200
-## The number of samples per chains
-samples <- 10
-
-## Set up the mini-chains models
-shapespace_allbirds_lvl_superorder_order_2D <- make.mini.chains(
-                                data         = shapespace,
-                                dimensions   = 1:2,
-                                tree         = trees_list[sample(1:1000, 1000)],
-                                trait.family = "gaussian",
-                                randoms      = c("global", "level1"),
-                                residuals    = "global",
-                                priors       = est_params$priors,
-                                verbose      = TRUE,
-                                parameters   = list(
-                                    nitt   = est_params$burnin + thin * samples,
-                                    thin   = thin,
-                                    burnin = est_params$burnin))
-## Save the mini.chains
-save(shapespace_allbirds_lvl_superorder_order_2D, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order_2D.mini.chains")
-```
-
-The chains are then run on the cluster
-TODO: details
-
-```{r}
-library(mcmcmcglmmm)
-load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/shapespace_allbirds_lvl_superorder_order_2D.mini.chains")
-## Running the three MCMCglmm models
-start <- Sys.time()
-shapespace_allbirds_lvl_superorder_order_2D <- run.mini.chains(shapespace_allbirds_lvl_superorder_order_2D, replicates = 20)
-save(shapespace_allbirds_lvl_superorder_order_2D, file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/param_chains/params_shapespace_2D_9levels_20k_chain1_20.rda")
-end <- Sys.time()
-end - start
-```
-
-### Combining the model data
-
-```{r, eval = FALSE}
-## Set up the variables
-models_path <- "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/cluster/"
-## Getting all the model files
-model_files <- list.files(models_path, pattern = ".rda")
-
-## Check if the model exists already
-if(!("shapespace_allbirds_lvl_superorder_order.MCMCglmm" %in% list.files("../Data/Processed/"))) {
-    ## Preparing the model files
-    model_list <- list()
-    class(model_list) <- c("beer", "mini.chains")
-    ## Get the first model
-    load(paste0(models_path, model_files[1], collapse = "/"))
-    ## Move it to the storage
-    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
-    print(model_files[1])
-    ## Clean the bits that are not going to be used (but are huge)
-    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
-    ## Update the model list
-    model_list[[length(model_list)+1]] <- model[[1]]
-    model <- NULL
-    model_files <- model_files[-1]
-} else {
-    ## Updating the model
-    load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-    model_list <- shapespace_allbirds_lvl_superorder_order
-}
-
-## Get the subsequent models and merge them as we go
-while(length(model_files) != 0) {
-    load(paste0(models_path, model_files[1], collapse = "/"))
-    ## Move it to storage
-    system(paste("mv", paste0(models_path, model_files[1]), paste0(models_path, "done/")))
-    print(model_files[1])
-    ## Clean the bits that are not going to be used (but are huge)
-    model[[1]][c("error.term", "family", "y.additional", "ginverse", "ZR")] <- NULL
-    ## Update the model list
-    model_list[[length(model_list)+1]] <- model[[1]]
-    ## Merge the models (overwrite the model list)
-    model_list <- list(combine.mini.chains(model_list))
-    ## DEBUG
-    print("model list size:")
-    print(format(object.size(model_list), units = "MB"))
-    model <- NULL
-    model_files <- model_files[-1]
-}
-
-## Combining the model files
-shapespace_allbirds_lvl_superorder_order <- model_list
-
-## Saving the results
-save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-
-```
-
-# Analysing the variance per terms
-
-```{r}
-## Get the variance for each level
-variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = TRUE)
-## Rename the columns
-colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
-reorder_clades <- c(9:35,1:8, 36, 37)
-# ## Pull by term
-# variance_per_levels<- cbind("orders" = apply(variance_per_clades[, 9:35], 1, sum),
-#                             "super-orders" = apply(variance_per_clades[, 1:8], 1, sum),
-#                             "phylogeny" = variance_per_clades[, 36],
-#                             "residuals" = variance_per_clades[, 37])
-# variance_per_terms <- cbind("clades" = apply(variance_per_clades[, 1:35], 1, sum),
-#                             "phylogeny" = variance_per_clades[, 36],
-#                             "residuals" = variance_per_clades[, 37])
-# ## Plot the results
-# pdf(file = "../Manuscript/Figures/variance_per_levels_whole_tree.pdf", height = 8, width = 14)
-# par(mar = c(5,8,2,2), mfrow = c(1, 3))
-# # boxplot(variance_per_terms, horizontal = TRUE, las = 1, main = "Scaled variance by terms", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-# # boxplot(variance_per_levels, horizontal = TRUE, las = 1, main = "Scaled variance by levels", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-# boxplot(variance_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled variance by clades", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-# boxplot(log(variance_per_clades[,reorder_clades])/log(sample_sizes), horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing sample size effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-# boxplot(log(variance_per_clades[,reorder_clades])/log(observed_variance)[names(variances)], horizontal = TRUE, las = 1, main = "Scaled variance by clades (removing variance effect)", xlab = "Scaled variance\n(sum of each VCV variance part (diagonal))")
-# dev.off()
-
-# ## Getting the variances per clades
-# variances <- apply(variance_per_clades[,reorder_clades], 2, mean)
-
-# ## Getting the sample sizes per clades (the one for the residuals is the whole tree size)
-# sample_sizes <- c("residuals" = size.subsets(order_shapespace)[28],
-#                   size.subsets(superorder_shapespace),
-#                   size.subsets(order_shapespace))[names(variances)]
-
-# ## Measuring the variance for each clade and for the residuals
-# ## the obs variance of the residuals is just the variance in the whole dataset (phylogeny)
-# observed_variance <- c("residuals" = unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[28], 
-#                        unlist(get.disparity(dispRity(superorder_shapespace, metric = sd))),
-#                        unlist(get.disparity(dispRity(order_shapespace, metric = sd)))[-28])^2
-
-# plot(log(variances)/log(sample_sizes), log(sample_sizes), xlab = "log variance / log n", ylab = "log n")
-
-```
-
-## Standard error per terms
-
-```{r}
-
-## Get the number of observations per groups
-variance_per_clades <- MCMCglmm.variance(shapespace_allbirds_lvl_superorder_order, scale = FALSE)
-colnames(variance_per_clades) <- c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny", "residuals")
-
-## Converting into standard errors
-get.se <- function(var, n) {
-    sqrt(var[[1]])/sqrt(n)
-}
-## Get the se per group
-se_per_clades <- mapply(get.se, apply(variance_per_clades, 2, list), as.list(c(size.subsets(superorder_shapespace)[-9], size.subsets(order_shapespace), "residuals" = 8748)))
-## Scale that
-se_per_clades <- se_per_clades/rowSums(se_per_clades)
-## Get the other levels
-se_per_levels<- cbind("orders" = apply(se_per_clades[, 9:35], 1, sum),
-                      "super-orders" = apply(se_per_clades[, 1:8], 1, sum),
-                      "phylogeny" = se_per_clades[, 36],
-                      "residuals" = se_per_clades[, 37])
-se_per_terms <- cbind("clades" = apply(se_per_clades[, 1:35], 1, sum),
-                      "phylogeny" = se_per_clades[, 36],
-                      "residuals" = se_per_clades[, 37])
-
-## Plot the results
-pdf(file = "../Manuscript/Figures/se_per_levels_whole_tree.pdf", height = 8, width = 14)
-par(mar = c(5,8,2,2), mfrow = c(1, 3))
-boxplot(se_per_terms, horizontal = TRUE, las = 1, main = "Scaled standard error by terms", xlab = "Scaled se ")
-boxplot(se_per_levels, horizontal = TRUE, las = 1, main = "Scaled standard error by levels", xlab = "Scaled se")
-boxplot(se_per_clades[,reorder_clades], horizontal = TRUE, las = 1, main = "Scaled standard error by clades", xlab = "Scaled variance")
-dev.off()
-```

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

@@
 ```
-
-# Variance explained in a simple model
-
-Here I'm just doing a one dimensional model of a one dimensional trait (BM) and two clades (`"s"` and `"t"`).
-
-```{r, echo = FALSE}
-## Setting the data
-set.seed(0)
-data_1 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
-data_2 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
-
-## Renaming the tips and data in data_2
-rownames(data_2$data)[match(data_2$tree$tip.label, rownames(data_2$data))] <- gsub("t", "s", data_2$tree$tip.label)
-data_2$tree$tip.label <- gsub("t", "s", data_2$tree$tip.label)
-data_1$tree$node.label <- data_2$tree$node.label <- NULL
-
-## Combine the trees
-ultrametric_tree <- chronos(multi2di(bind.tree(data_1$tree, data_2$tree)))
-class(ultrametric_tree) <- "phylo"
-tree <- slide.nodes(ultrametric_tree, node = 42, slide = 0.2)
-
-## Combine the data
-data <- c(data_1$data[match(data_1$tree$tip.label, rownames(data_1$data)), ], data_2$data[match(data_2$tree$tip.label, rownames(data_2$data)), ])
-data <- as.data.frame(matrix(data, ncol = 1, dimnames = list(c(names(data)), c("variable"))))
-data <- cbind(data, "var2" = rnorm(40), "clade" = as.factor(c(rep("t", 20), rep("s", 20))), "animal" = rownames(data))
-```
-
-We will now run it as the following model:
-
-```
-data ~ residuals + random.phylo + random.clade_s + random.clade_t
-```
-
-```{r, echo = FALSE, message = FALSE}
-## Setting up the model
-first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))
-
-## running the first model
-first_model_run <- run.mini.chains(first_model, replicates = 1)[[1]]
-```
-
-And this is what it looks like graphically
-
-```{r, echo = FALSE}
-## Model object
-first_model_results <- MCMCglmm.subsets(
-      data          = data,
-      posteriors    = first_model_run,
-      group         = MCMCglmm.levels(first_model_run)[c(1:3)],
-      rename.groups = c("clade_s", "clade_t", "phylogeny"))
-
-## Set the clade colours
-clade_cols <- c("clade_s" = "blue", "clade_t" = "orange", "phylogeny" = "grey")
-
-## Plotting the model
-covar.plot(first_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
-```
-We can then extract the variance of the model:
-
 ```{r, echo = FALSE}
@@
 # "log standard error" = logged standard error;
-
 plot.decompose.var <- function(data, posteriors, group.cols, obs.col = "red", relative = FALSE, plot = c("var boxplot", "var by n", "log var boxplot", "log var by n", "corrected var boxplot", "corrected var by n", "standard error", "log standard error"), ...) {
@@
     if(!relative) {
-        obs_var <- sd(unlist(data$matrix[[1]][, c(data$call$dimensions)]))^2
+        obs_var <- sum(apply(data$matrix[[1]][, c(data$call$dimensions)], 2, var))
     }
@@
 
+# Variance explained in a 1D model
+
+For this first very simple model, I will just simulate two trees of 20 taxa each with the two same independent Brownian Motion (BM) traits and combine them together to get a dataset with one variable (2 independent BMs) and two monophyletic clades (`"s"` and `"t"` because why not).
+
 ```{r, echo = FALSE}
-## Plotting all the variance elements
-plot.decompose.var(first_model_results, first_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = 1:8, relative = FALSE)
+## Setting the data
+set.seed(0)
+data_1 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
+data_2 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
+
+## Renaming the tips and data in data_2
+rownames(data_2$data)[match(data_2$tree$tip.label, rownames(data_2$data))] <- gsub("t", "s", data_2$tree$tip.label)
+data_2$tree$tip.label <- gsub("t", "s", data_2$tree$tip.label)
+data_1$tree$node.label <- data_2$tree$node.label <- NULL
+
+## Combine the trees
+ultrametric_tree <- chronos(multi2di(bind.tree(data_1$tree, data_2$tree)))
+class(ultrametric_tree) <- "phylo"
+tree <- slide.nodes(ultrametric_tree, node = 42, slide = 0.2)
+
+## Combine the data
+data <- c(data_1$data[match(data_1$tree$tip.label, rownames(data_1$data)), ], data_2$data[match(data_2$tree$tip.label, rownames(data_2$data)), ])
+data <- as.data.frame(matrix(data, ncol = 1, dimnames = list(c(names(data)), c("variable"))))
+data <- cbind(data, "var2" = rnorm(40), "clade" = as.factor(c(rep("t", 20), rep("s", 20))), "animal" = rownames(data))
 ```
 
-# More complex model with equal sample size
+I will run the simplest model which is the data (the 2 BMs) as a response variables, clades as explanatory ones and then two error terms: the data residuals and the phylogenetic randoms terms:
 
-For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (but equal-ish) but based on biological data.
+```
+data ~ clade + residuals + random(phylogeny)
+```
+
+```{r, echo = FALSE, message = FALSE}
+test <- MCMCglmm(fixed  = variable ~ clade,
+                 random = ~ animal,
+                 pedigree = tree,
+                 data = data, 
+                 nitt = 10000,
+                 burnin = 10,
+                 thin = 10)
+```
+
+I can then extract the variance for the error terms and plot the observed variance in the data (i.e. the variance from the trait)@
+
+```{r, echo = FALSE}
+par(mar = c(5, 6, 4, 2) + 0.1)
+var_table <- MCMCglmm.variance(test, scale = FALSE)
+colnames(var_table) <- c("phylogeny", "residuals")
+boxplot(var_table, horizontal = TRUE, las = 1, xlab = "Variance")
+abline(v = var(data$variable), col = "red", lwd = 2)
+legend("topright", col = "red", lwd = 2, lty = 1, legend = "Observed variance")
+```
+
+One first striking thing (to me) is that the residuals are below the observed variance, which is expected: the residuals cannot explain more than the variance in the datasets but this is not the case for the random terms (phylogeny)!
+
+# Variance explained in a 2D model with covariance
+
+We can then spice things up a bit more by looking at what does that variance extracted from the models mean in multi dimensions.
+Using the same data as before but adding a random variable (a normal distribution - not related to the phylogeny) we can then do the same model but in a multidimensional way:
+
+The model is now:
+
+```
+data(1D, 2D)~ clade + residuals + random(phylogeny)
+```
+
+```{r, echo = FALSE, message = FALSE}
+## Setting up the model
+first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global"), verbose = TRUE, parameters =  list(nitt = 20000))
+
+## running the first model
+first_model_run <- run.mini.chains(first_model, replicates = 1)[[1]]
+```
+
+```{r, echo = FALSE}
+par(mar = c(5, 6, 4, 2) + 0.1)
+var_table <- MCMCglmm.variance(first_model_run, scale = FALSE)
+colnames(var_table) <- c("phylogeny", "residuals")
+boxplot(var_table, horizontal = TRUE, las = 1, xlab = "Variance")
+abline(v = var(data$variable)+var(data$var2), col = "red", lwd = 2)
+legend("topright", col = "red", lwd = 2, lty = 1, legend = "Observed variance")
+```
+
+We now see something more funky with the variance of the residuals being sometimes higher than the variance available in the model.
+I suspect this is due to the multidimensionality but not sure.
+
+# Variance explained in a 2D model with covariance and nested terms
+
+We can then use the actual model type we used in our paper with a multidimensional variable and a random term for the phylogeny AND each clade:
+
+```
+data(1D, 2D) ~ residuals + random(phylogeny) + random(clade_s) + random(clade_t)
+```
+
+```{r, echo = FALSE, message = FALSE}
+## Setting up the model
+second_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))
+
+## running the first model
+second_model_run <- run.mini.chains(second_model, replicates = 1)[[1]]
+```
+
+And this is what it looks like graphically:
+
+```{r, echo = FALSE}
+## Model object
+second_model_results <- MCMCglmm.subsets(
+      data          = data,
+      posteriors    = second_model_run,
+      group         = MCMCglmm.levels(second_model_run)[c(1:3)],
+      rename.groups = c("clade_s", "clade_t", "phylogeny"))
+
+## Set the clade colours
+clade_cols <- c("clade_s" = "blue", "clade_t" = "orange", "phylogeny" = "grey")
+
+## Plotting the model
+covar.plot(second_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
+```
+
+We can then extract the variance of the model:
+
+```{r, echo = FALSE, fig.height = 16, fig.width = 4}
+## Plotting all the variance elements
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(second_model_results, second_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), relative = FALSE, plot = "var boxplot")
+```
+
+We can see now that the data consistently spreads over the total amount of variance (although the median values remain lower).
+To avoid this spread we can log the variance to see what it does:
+
+```{r, echo = FALSE, fig.height = 16, fig.width = 4}
+## Plotting all the variance elements
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(second_model_results, second_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), relative = FALSE, plot = "log var boxplot")
+```
+
+Bit clearer but still enigmatic to me (i.e. how can we have more variance explained than variance available).
+
+# Multidimensional nested model with different sample sizes.
+
+Another thing with our models is that they deal with different sample sizes.
+We can explore that using a mild example (a tree with three groups of roughly equal size) and a more contrasting example (with orders of magnitudes in differences in sample sizes).
+
+## Similar sample sizes
+
+For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (159, 98 and 102 species in each clade).
 
@@
 ```{r, echo = FALSE}
-plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"))
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = "var boxplot")
 ```
 
-# Model with different sample sizes
-
-This is the big model run just on 2D and only with the super orders levels.
-The differences with the true results are due to it:
- * not being the same dataset!
- * not having converged...
+It's interesting to see that the variance for the phylogeny is greater than for two of the groups (expected?) but not for the gulls group!
+Also, in these datasets, the residuals start to become very small (probably due to most variance being explained by the random terms).
+Here's a logged illustration of these variance differences:
 
 ```{r, echo = FALSE}
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = "log var boxplot")
+```
+
+We can then also plot these differences in variances relative to sample sizes (ignore the lm fit):
+
+```{r, echo = FALSE, fig.height = 4, fig.width = 8}
+plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("var by n", "log var by n"))
+```
+
+The red circle is the total variance (probably OK to ignore in these plots).
+
+## Different sample sizes
+
+We can then look at the effect of more wildly different sample sizes.
+For this test, I'm going to run the BIG whole shapespace dataset in 2D with all the super orders as nested levels.
+The sample sizes for the nest random terms then varies from 24 (Mirandortnithes) species to >6k (Telluraves) so we have three orders for magnitude here.
+
+```{r, echo = FALSE, fig.height = 6, fig.width = 6}
 ## Load the models from the local files
@@
 
-```{r, echo = FALSE}
+Note that the results (ellipses) look kind of different than the ones from the real model.
+This is interesting in highlighting the effect of using 8 rather than 2 dimensions and using the interactions of all nested terms (including the orders) rather than just the super orders.
+Also the calculations here are rough and the model hasn't converged.
+
+And then we can plot the variance and the logged variance per terms and by sample sizes:
+
+```{r, echo = FALSE, fig.height = 10, fig.width = 10}
 ## Plotting the variance with different data
-plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"))
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("var boxplot", "log var boxplot", "var by n", "log var by n"))
 ```
 
-# Model used in this paper (8D + 32 terms)
+Here there seems to be a clear relation between sample size and variance: big clades have on average lower variance than smaller ones (nicely illustrated on a log log scale).
+
+We can thus try to "correct" for these sample size issues by using the standard error (or log standard error):
+
+```{r, echo = FALSE, fig.height = 4, fig.width = 8}
+## Plotting the variance with different data
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("standard error", "log standard error"))
+```
+
+This shows maybe easier to interpret variations.
+
+Or, maybe more interpretable, we can just "correct" variance by dividing it by the log sample size (instead of the square root of the sample size in the standard error) assuming a log normal relation between variance and sample size (kind of clearer in the big model at the end).
+
+```{r, echo = FALSE, fig.height = 4, fig.width = 8}
+## Plotting the variance with different data
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("corrected var boxplot", "corrected var by n"))
+```
+
+# Variance on BIG models
+
+We can then apply these thingies to the big model (with 8 dimensions and 32 nested random terms.
+This is what it looks like:
 
@@
                                rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny"))
-
 covar.plot(full_model, col = clade_cols, ellipses = mean, legend = TRUE, lines.lty = c(rep(1, length(levels(shapespace$level1)[-1])), rep(2, length(levels(shapespace$level2)[-1])), 1), points = FALSE)
@@
 
-```{r, echo = FALSE}
+And these are all the different ways to illustrate this variance:
+
+```{r, echo = FALSE, fig.height = 32, fig.width = 8}
 ## Plotting the variance with different data
-plot.decompose.var(full_model, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = 4)
+par(mar = c(5, 6, 4, 2) + 0.1)
+plot.decompose.var(full_model, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("var boxplot", "log var boxplot", "var by n", "log var by n", "corrected var boxplot", "corrected var by n", "standard error", "log standard error"))
 ```
+
+Not sure what we get out of all of this but hopefully it gives us some insights on how to talk about the variance partition in the intro.

### 2022-10-18 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

@@
 
-```{r, echo = FALSE, fig.height = 16, fig.width = 4}
+```{r, echo = FALSE}
 ## Plotting all the variance elements
@@
 
-```{r, echo = FALSE, fig.height = 4, fig.width = 8}
+```{r, echo = FALSE, fig.height = 6, fig.width = 12}
 ## Plotting the variance with different data
@@
 plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("corrected var boxplot", "corrected var by n"))
+rm(birds_data)
+rm(model)
 ```

### 2022-10-18 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 %---------------------------------------------
-% QUESTIONS/narrative
-% How does phenotypic diversity arise?
-% "Disparity" through time = how is disparity within clades partitioned wrt overall disparity. Diversity accumulates in a boring way, but doesn't tell us about mechanisms, 
-% Lines of evolutionary least resistance theory. Very little testing 
-% Test this using new method: elaboration/innovation
-% How does this vary at different hierarchical levels - mega vs macro evolutionary?
-% Species-level analyses
-% Passerines
 
-% Bits from Andrew
-% A first insight into the evolutionary scale at which phenotypic diversity has arisen in birds comes from a simple exploration of the variance among major axes at different scales.
-If variance in the major axis is highest at the phylogenetic (mega) scale rather than the clade (macro) scale, it is evidence that diversity is very old.
-
-% ANSWER
-
-% Supporting this, we find overwhelming evidence that deep time evolutionary history does not constrain phenotypic variation.
-This is evidenced by the orientation of evolution within clades being markedly different from and frequently orthogonal to the major phylogenetic axis, indicating that deep evolutionary history does not constrain diversity.
-
-% DETAILS
-
-% Against this backdrop, we then estimated the relative importance of elaboration or innovation to driving this pattern….
-We made 3 specific investigations to do this....(Needs to be more precise about question, how we answer and answer - e.g.
-is elaboration most common at all evolutionary scales?
-Is innovation isolated to evolutionary scales?).
-
-% Trick to embed question in 'result'??
+% New narrative:
+% 1 - Novelty arises through elaboration and innovation
+% 2 - Theory suggests that it is easier to evolve along the major axis of variation (formely LOTR in this manuscript)
+% 3 - However this has never been tested in big datatests. We do that.
+% 4 - We measure elaboration and innovation in all birds
+% 5 - We find that there is more elaboration than innovation
+% 6 - But that asks another (more interesting) question: why do we have diversity if everything elaborates?
+% 7 - We then measure novelty by clades rather than by species
+% 8 - We find that clades innovate.
+% 9 - So basically species elaborate (in line with the micro evol theory of evolving along the line fo major axis of variation) and clades innovate (in line with the observations at a macro and mega level of evolutionary jumps leading to big diversity at high levels).
 
@@
 Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
-A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
+A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
 At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
-The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance \cite{endler2005animal, renaud2006conserved}.
-Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+The presence of distinct morphological forms suggests instead that species or clades may evolve away from the line of least resistance.
+At a macroevolutionary level, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal, renaud2006conserved}.
+The extent to which elaboration and innovation contribute to phenotypic diversity among species and clades is critical to our understanding of the origins and maintenance of biodiversity across scales from the micro- to macro and mega-evolutionary.
+Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks among species and clades.
 Our analyses show that elaboration is a common feature at all scales, consistent with theory.
@@
 % paragraph 1:  introducing the major axis of variation + novelty and innovation
-Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow lines of genetic least resistance \cite{schluter1996adaptive}.
-Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,Rombaut2022} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape \cite{jones2004evolution}.
-Indeed, evidence suggests that genetic constraints can extend over timescales spanning tens of millions of years leading to remarkable macroevolutionary stability in the orientation of phenotypic evolution \cite{mcglothlin2018adaptive}.
-This evolutionary major axis of phenotypic variation allows to describe the novelty of phenotypic traits made of two components: elaboration and innovation \cite{endler2005animal}.
-We can define phenotypic changes along the major axis of variation as elaboration and changes away from that line as innovation.
-And thus by definition, we would suspect that elaboration is a major driver of morphological evolution across the time (because it's equal to evolution along that line).
-In this paper we will test this hypothesis and measure whether elaboration is more common across the tree of life.
-%This is basically a fake hypothesis - although we're testing that, the plot twist happens later!
+Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
+Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
+Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
+Recent evidence even suggests that this stability can extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
+However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic divergence is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the directions of phenotypic evolution across scales where macroevolutionary divergence along lines of least resistance gives way to mega-evolutionary divergence leading to distinct phenotypes or directions of evolution.
+Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
+
+% paragraph 2:  introducing the Endler view.
+ Endler et al.
+\cite{endler2005animal}, writing in the specific context of signal evolution, describe two  routes to diversity that we suggest can help to illuminate this challenge.
+Specifically, Endler et al.
+refer to evolution of new phenotypes along the same axis as \textit{elaboration}.
+This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
+In contrast, they refer to evolution away from the major axis as \textit{innovation}.
+More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
+For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022} are consistent with specific innovation.
+Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
+These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of: i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of divergence (specific innovation), and ii) evolution along versus away from major axes of divergence.
+The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales.
+ Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.
 
 % paragraph 3: how we measure this (glmm) on the birds data + nestedness
+\subsection{Modelling nested covariances}
 Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
 1) large, multivariate datasets to allow exploration of trait covariances at different scales,
-2) reliable and efficient computational methods to estimate the lines of least resistance in a nested fashion,
-and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
-Here, we propose a novel analytical pipeline for measuring elaboration and innovation at the species- and higher taxonomic levels (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-We define the major axis of variation as the major axis of traits variance-covariance matrices for these bird beaks.
-And to test our hypothesis (is elaboration more common across the tree of life) we need to  measure bird beaks' novelty at different nested phylogenetic levels.
-We will this calculate this variance-covariances matrices in a nested way across all birds: first for all birds, then for all super-orders and then for each order.
-To measure these variance-covariance matrices, we fitted a phylogenetic generalised linear mixed effect model (pGLMM) with beak shape as an eight-dimensional response variable and treated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
-We regard the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
-The nested partitions of multivariate trait space provide the basis for all subsequent analyses.
+2) reliable and efficient computational methods to estimate the major axes of phenotypic divergence,
+and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale.
+To meet the first challenges we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
+To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
+In order to provide a baseline on which to calculate elaboration and innovation we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
+Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
+From these models we can calculate major axes of divergence as the leading eigenvector in trait space.
+We regard the major axis of the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
+The major axis can be defined at any scale (i.e.
+for all species in the phylogeny, or for all species within any consistent clade).
+In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of divergence) for i) the phylogenetic partition of all species, ii) all species within each of nine clades mapping approximately to superorders, and iii) all species within each of XX clades mapping to taxonomic orders.
+These nested partitions of multivariate trait space provide the basis for all subsequent analyses.
+
+\subsection{Elaboration dominates species-level divergence}
+% paragraph 4: results for elaboration and innovation for species
 We then measure innovation and elaboration as the multilinear algebra projection of each species onto their nested major axis of variation.
-
-% paragraph 4: results for elaboration and innovation for species
 Here are the results for each species. \ref{Fig:phylogeny}
@@
 
+\subsection{Multiple routes to specific innovation throughout avian evolutionary history}
 % Paragraph 5: paradox + methods for comparing ellipses
@@
 % Paragraph 7: putting these results together
+\subsection{Species elaborate, clades innovate}
 
@@
 % Paragraph 8: conculsion
+\subsection{Conclusions}
+
 Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.

### 2022-10-18 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 Recent evidence even suggests that this stability can extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
-However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic divergence is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}.
+However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}.
 The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the directions of phenotypic evolution across scales where macroevolutionary divergence along lines of least resistance gives way to mega-evolutionary divergence leading to distinct phenotypes or directions of evolution.
@@
 % paragraph 2:  introducing the Endler view.
- Endler et al.
-\cite{endler2005animal}, writing in the specific context of signal evolution, describe two  routes to diversity that we suggest can help to illuminate this challenge.
-Specifically, Endler et al.
-refer to evolution of new phenotypes along the same axis as \textit{elaboration}.
+Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
+Specifically, Endler et al. refer to evolution of new phenotypes along the same axis as \textit{elaboration}.
 This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
@@
 Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
-These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of: i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of divergence (specific innovation), and ii) evolution along versus away from major axes of divergence.
+These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of:
+i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of variation (specific innovation),
+and ii) evolution along versus away from major axes of variation.
 The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
@@
 % paragraph 3: how we measure this (glmm) on the birds data + nestedness
-\subsection{Modelling nested covariances}
+\subsection{Modelling nested trait covariances}
 Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
 1) large, multivariate datasets to allow exploration of trait covariances at different scales,
-2) reliable and efficient computational methods to estimate the major axes of phenotypic divergence,
+2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
+%TG: Gavin has been going with "major axes of phenotypic divergence" throughout but I've changed it to "major axes of phenotypic variation".
+It's a semantic point and we can easily toggle between one or the other but I would argue that the LOTR captures more the main direction of (co)variation (i.e.
+what is the most likely path of changes) rather than divergence (that in my head reads more as varying away from something).
 and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale.
@@
 To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-In order to provide a baseline on which to calculate elaboration and innovation we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
+In order to provide a baseline on which to calculate elaboration and innovation (i.e.
+the major axis of variation) we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
 Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components. 
-From these models we can calculate major axes of divergence as the leading eigenvector in trait space.
-We regard the major axis of the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
-The major axis can be defined at any scale (i.e.
-for all species in the phylogeny, or for all species within any consistent clade).
-In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of divergence) for i) the phylogenetic partition of all species, ii) all species within each of nine clades mapping approximately to superorders, and iii) all species within each of XX clades mapping to taxonomic orders.
+From these models we can calculate major axes of variation as the leading eigenvector of the variance-covariance matrix in trait space.
+We regard the major axis of variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
+The major axis can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any monophyletic clade).
+In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of variation) for
+i) the entire phylogeny of all species included in the species (hereafter the "whole tree" level),
+ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
+and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
 These nested partitions of multivariate trait space provide the basis for all subsequent analyses.
@@
 % paragraph 4: results for elaboration and innovation for species
-We then measure innovation and elaboration as the multilinear algebra projection of each species onto their nested major axis of variation.
+We then measure innovation and elaboration as the multilinear algebraic projection of each species onto their nested major axis of variation.
 Here are the results for each species. \ref{Fig:phylogeny}
-There seems to be a nested structure of the variation in novelty %TODO: needs to introduce nestedness; %TODO: needs to introduce novelty.
+There seems to be a nested structure of the variation in novelty (i.e. elaboration plus innovation).
 There is more variation the deeper you baseline comparison in the tree is, suggesting that novelty is relative to the baseline comparison.
@@
 
-% Paragraph 5+: Some elluipses stuff with bird names.
+% Paragraph 5+: Some ellipses stuff with bird names.
 However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).

### 2022-10-18 Manuscript/supplementaries.tex
Update supplementaries.tex

--- 2022-08-12 Manuscript/supplementaries.tex
@@
 
+\usepackage{lscape}
 \usepackage{hyperref}
@@
 
-\section{Supplementary material 1: additional results}
+\begin{landscape}
+\section{Supplementary material 1: glossary}
+
+\begin{table}[ht]
+\centering
+\scriptsize
+\begin{tabular}{p{0.2\textwidth}p{0.3\textwidth}p{0.3\textwidth}p{0.1\textwidth}p{0.1\textwidth}}
+  \hline
+Term in this paper         & definition in this paper & mathematical definition & in macroevolution & in microevoltion\\
+  \hline
+major axis of variation    & the major evolutionary ``direction'' of traits co-evolution & first eigenvector of the variance-covariance matrix  & allometry & line of genetic least resistance \\
+clade                      & a monophyletic group of \textbf{species} & a subset of \textbf{shapespace} & Birds & a group of species or a population\\
+species                    & the lowest evolutionary units in our analysis & a row in the \textbf{shapespace} & \textit{} & an individual\\
+shapespace                 & all the possible morphological combination (observed or not) & an ordinate and orthonormal matrix & the beak shapespace & a gene space \\
+novelty                    & morphological variation (either \textbf{innovation} or \textbf{elaboration}) & position of a \textbf{species} or a \textbf{clade} relative to the \textbf{major axis of variation} & & \\
+elaboration                & morphological variation along the \textbf{major axis of variation} & the algebraic projection on the eigenvector & & \\
+innovation                 & morphological variation away from the \textbf{major axis of variation} & the algebraic rejection from the eigenvector & & \\
+variance-covariance matrix & the multidimensional relation between traits & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) & & \\
+nested phylogenetic level  & the different \textbf{clades} used in this study (the whole phylogeny > super-orders > orders) & different nested \textbf{variance-covariance matrices} & & \\
+
+\end{tabular}
+\caption{Glossary of some specific terms in this manuscript}
+\label{tab_glossary}
+\end{table}
+\end{landscape}
+
+\section{Supplementary material 2: additional results}
 
@@
 
+\begin{equation}
+    \vec{a_{1}} = \frac{\vec{a} \cdot \vec{b}}{\|\sqrt{\vec{b} \cdot \vec{b}}\|}
+\end{equation}
+
 With $\|\vec{b}\| = \sqrt{\vec{b} \cdot \vec{b}}$ being the norm of $\vec{b}$.

### 2022-10-18 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 clade                      & a monophyletic group of \textbf{species} & a subset of \textbf{shapespace} & Birds & a group of species or a population\\
-species                    & the lowest evolutionary units in our analysis & a row in the \textbf{shapespace} & \textit{} & an individual\\
+species                    & the lowest evolutionary units in our analysis & a row in the \textbf{shapespace} & \textit{Pauxi pauxi} & an individual\\
 shapespace                 & all the possible morphological combination (observed or not) & an ordinate and orthonormal matrix & the beak shapespace & a gene space \\

### 2022-10-20 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

@@
 legend("topright", col = "red", lwd = 2, lty = 1, legend = "Observed variance")
+
+## Adding the cophenetic variance line?
+coph <- cophenetic(tree)
+coph_var <- sd(coph)^2
+scaled_var <- sd(coph * data$variable)^2
 ```

### 2022-10-20 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

--- 2022-10-18 Manuscript/main_flipped_narrative.tex
@@
 A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
-At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
+At macro- and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
 The presence of distinct morphological forms suggests instead that species or clades may evolve away from the line of least resistance.
@@
 % paragraph 1:  introducing the major axis of variation + novelty and innovation
-Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
+Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
 Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
 Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
-Recent evidence even suggests that this stability can extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
-However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic divergence is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}.
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the directions of phenotypic evolution across scales where macroevolutionary divergence along lines of least resistance gives way to mega-evolutionary divergence leading to distinct phenotypes or directions of evolution.
+Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
+However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
 Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
@@
 % paragraph 2:  introducing the Endler view.
- Endler et al.
-\cite{endler2005animal}, writing in the specific context of signal evolution, describe two  routes to diversity that we suggest can help to illuminate this challenge.
-Specifically, Endler et al.
-refer to evolution of new phenotypes along the same axis as \textit{elaboration}.
+Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
+Specifically, Endler et al. refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
 This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
-In contrast, they refer to evolution away from the major axis as \textit{innovation}.
+In contrast, they refer to evolution away from the major axis of phenotypic variation as \textit{innovation}.
 More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
@@
 Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
-These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of: i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of divergence (specific innovation), and ii) evolution along versus away from major axes of divergence.
+These core concepts provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
+i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of phenotypic variation (specific innovation),
+and ii) evolution along versus away from major axes of phenotypic variation.
 The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
@@
 % paragraph 3: how we measure this (glmm) on the birds data + nestedness
-\subsection{Modelling nested covariances}
+\subsection{Modelling nested trait covariances}
 Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
 1) large, multivariate datasets to allow exploration of trait covariances at different scales,
-2) reliable and efficient computational methods to estimate the major axes of phenotypic divergence,
+2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
 and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale.
-To meet the first challenges we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
+To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
 To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-In order to provide a baseline on which to calculate elaboration and innovation we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
+To provide a baseline from which to calculate elaboration and innovation (i.e.
+the major axis of phenotypic variation) we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
 Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components. 
-From these models we can calculate major axes of divergence as the leading eigenvector in trait space.
-We regard the major axis of the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
-The major axis can be defined at any scale (i.e.
-for all species in the phylogeny, or for all species within any consistent clade).
-In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of divergence) for i) the phylogenetic partition of all species, ii) all species within each of nine clades mapping approximately to superorders, and iii) all species within each of XX clades mapping to taxonomic orders.
+From these models we can calculate major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix in trait space.
+We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
+The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any monophyletic clade).
+To assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of phenotypic variation) for
+i) the entire phylogeny of all species included in the data (hereafter the "whole tree" level),
+ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
+and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
 These nested partitions of multivariate trait space provide the basis for all subsequent analyses.
@@
 % paragraph 4: results for elaboration and innovation for species
-We then measure innovation and elaboration as the multilinear algebra projection of each species onto their nested major axis of variation.
-Here are the results for each species. \ref{Fig:phylogeny}
-There seems to be a nested structure of the variation in novelty %TODO: needs to introduce nestedness; %TODO: needs to introduce novelty.
-There is more variation the deeper you baseline comparison in the tree is, suggesting that novelty is relative to the baseline comparison.
-Also, the majority of that novelty is through elaboration.
-We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
-However, we also found various combinations of elaboration and innovation across birds.
-For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
-Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
-Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
-The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
+We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species specific measure of random innovation and elaboration.
+To do this we estimated the multilinear algebraic projection of each species phenotype onto the three (whole tree, super-order, and order) major axis of phenotypic variation.
+Phylogenetic distributions of random innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or whole tree level (Fig.
 \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
-\textbf{But there is a paradox here:}
-If most organisms evolve incrementally along major axis of variation then how can we explain the staggering diversity of species morphologies across the tree of life?
+However, the extent of both elaboration and random innovation tends to reduce as we move from whole tree to within order comparison.
+Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
+The whole tree's major axis of phenotypic variation primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
+Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.
+
+Overall we found that, as expected, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}.
+We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration.
+%(Fig.
+S{#fig_correlations}).
+%TODO: maybe add to supp?
+However, we observe various combinations of elaboration and random innovation across birds.
+Examples of high innovation and/or elaboration are often clustered within clades.
+For example, compared to the whole tree major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
+In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
+Other clades such as the Apodidae (swifts) in the order Apodiformes and the distantly related but convergent Hirundidae (swallows and martins) have exceptionally wide (relative to length) beaks and display high levels of both elaboration and specific innovation.
+The dominance of elaboration when measured at the species level is consistent with idea of evolution along lines of least evolutionary resistance.
+However, the frequent clustering of species that elaborate, innovate, or both, again highlights the likelihood that major axes of phenotypic variation are not stable among clades.
+This leads us to ask: if most organisms evolve incrementally along major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?
 
@@
 \label{Fig:phylogeny}
-\end{figure}
-
-\begin{figure}[!htbp]
-\centering
-    \includegraphics[width=0.9\textwidth]{Figures/median_novelty_distributions.pdf}
-\caption{THIS FIGURE SHOULD BE EMBEDED IN THE ANUS FIGURE: showing the novelty grouped into two different ways}
-\label{Fig:phylogeny_2}
 \end{figure}
@@
 \subsection{Multiple routes to specific innovation throughout avian evolutionary history}
-% Paragraph 5: paradox + methods for comparing ellipses
-Although the adaptive radiation theory implies that phenotypic divergence arises primarily along the major axis of variation \cite{marroig2005size}, there has been many contrasting observations in the literature highlighting major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
-We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different clades compared to the global phylogenetic major axis.
-We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis (Fig.
-\ref{Fig:ellipses} - Fig.
-\ref{Fig:orthogonality}).
-Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis and each super order (Fig.
-\ref{Fig:ellipses}a) and order (Fig.
-\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
-\ref{Fig:orthogonality}).
-The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
-Comparisons of orientation among subclades (i.e.
-orders within superorders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
-This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
-These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
 
-% Paragraph 5+: Some elluipses stuff with bird names.
-However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
-While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
-To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
-As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
-Clades (or species) can be both elaborators and innovators.
-To measure innovation and elaboration we translated the major axes of each clade's (i.e.
-superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
-We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
-We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
+To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and specific innovation of different clades.
+To do this we translated the major axes of phenotypic variation of each clade's (i.e.
+super-orders and orders) phylogenetic variance-covariance matrices onto the whole tree's major axis of phenotypic variation so that they shared the same origin in the shapespace.
+We then used multilinear algebra to project a focal clade's major axis of phenotypic variation onto the whole tree's major axis of phenotypic variation  (Fig.S\ref{Fig:cheat_sheet}).
+We interpret the measured projection (distance onto the phylogenetic major axis of phenotypic variation) as the clade's elaboration score and the measured rejection (distance from the major axis of phenotypic variation) as the clade's innovation score.
 These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
-Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
-However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
+We found that for most clades the phenotypic major axis of phenotypic variation is not aligned with the whole tree's major axis of phenotypic variation , (Fig. \ref{Fig:ellipses}).
+Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, specific innovation is a more common generator of morphological diversity than elaboration in deep time.
+However, there is substantial variation in elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
 Cuculiformes; Fig.
@@
 \ref{Fig:ellipses}).
-These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
-This suggests that divergence away from line of least evolutionary resistance (i.e.
-the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
-
-This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
-Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig.
-\ref{Fig:ellipses}-bar plots).
-These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
-S\ref{fig_ellipses_passeriformes}).
-Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
-For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
-However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
+These patterns hold across scales including within super-orders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
+This suggests that there is no deep time analogue of the evolutionary line of least evolutionary resistance and that across scales there is a remarkable flexibility in the routes to innovation.
 
@@
 \noindent \textbf{Figure \ref{Fig:ellipses}:}
-Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
-We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
-The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
-Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
+We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all super-order ellipses relative to the all-bird phylogenetic ellipse.
+Subsequent inset plots show the results for each super-order. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
 The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
@@
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
+
+\bigskip
+
+This variability in the routes to innovation implies a distinct lack of deep-time evolutionary constraint in the dominant axes of evolution within clades.
+Although the whole tree's major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the whole tree's major axis of phenotypic variation is aligned on PC1; Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend Fig.
+\ref{Fig:ellipses}-bar plots).
+This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes PC1 (Fig.
+S\ref{fig_ellipses_passeriformes}).
+In addition to the variable alignment of clade major axes of phenotypic variation, we further found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the whole tree variance-covariance matrix.
+For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the whole tree but uniformity in their beak shape components within the clade.
+In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
+However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
+
+\subsection{Species elaborate, clades innovate}
+
+% Paragraph 5: paradox + methods for comparing ellipses
+Although adaptive radiation theory implies that phenotypic divergence arises primarily along the major axis of phenotypic variation \cite{marroig2005size}, there have been many contrasting observations in the literature highlighting major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
+We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the major axis of phenotypic variation within different clades compared to the whole tree's of phenotypic variation.
+We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis of phenotypic variation (Fig.
+\ref{Fig:ellipses} - Fig.
+\ref{Fig:orthogonality}).
+Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the whole tree's major axis of phenotypic variation and each super order (Fig.
+\ref{Fig:ellipses}a) and order (Fig.
+\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
+\ref{Fig:orthogonality}).
+The median angle of the major axis of phenotypic variation for subclades approaches orthogonality, differing from the phylogenetic major axis of phenotypic variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
+Comparisons of orientation among subclades (i.e.
+orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
+This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
+These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
 
@@
 \noindent \textbf{Figure \ref{Fig:orthogonality}:}
-Orthogonality of each clade's major axis compared to their parent or parent parent's clade.
+Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent parent's clade.
 Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
 Dots represent the clades median orthogonality and the lines their 95\% CI.
-No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
+No super-order's major axis of phenotypic variation is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
 These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
-Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
 For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
 The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
-Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
+Only two out of six super-orders; 6/25 orders relative to their super-order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
 This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
@@
 
+\bigskip
+
 % Paragraph 7: putting these results together
-\subsection{Species elaborate, clades innovate}
 
@@
 We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
-These eight dimensions contained at least 95\% of variance in each superorder, and 98.7\% of the variance in the whole dataset.
-We used a threshold of $>95$\% of variance explained for each superorder because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
-$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each superorder.
+These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
+We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
+$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each super-order.
 For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
-This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each superorder (see Fig.
+This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig.
 S\ref{Fig:axes_variance} and Table \ref{tab_variance_per_axis}).
@@
 We ran two separate nested models.
-One model using one random term for the whole phylogeny and 35 nested random terms for each superorder and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
+One model using one random term for the whole phylogeny and 35 nested random terms for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
 And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
@@
 We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
-In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that clade.
-In brief we can interpret where a species or a clade falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
+In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis of phenotypic variation onto that clade.
+In brief we can interpret where a species or a clade falls on the major axis of phenotypic variation of reference as its elaboration score and how far a species is from that major axis of phenotypic variation as its innovation score.
 We calculated the elaboration and innovation scores using two main approaches:
 1) by clades (i.e.
-to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes (Fig.
+to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes of phenotypic variation (Fig.
 \ref{Fig:ellipses});
 and 2) by species (i.e.
-to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes and b) the phylogenetic major axes of their respective clade (Fig.
+to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes of phenotypic variation and b) the whole tree's major axes of phenotypic variation of their respective clade (Fig.
 \ref{Fig:phylogeny}).
-To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
+To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of phenotypic variation of interest and made the elaboration values absolute.
 This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
@@
 For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
-To do so we measured the angles between the major axes of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade by randomly comparing major axes within a clade.
+To do so we measured the angles between the major axes of phenotypic variation of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade by randomly comparing major axes of phenotypic variation within a clade.
 We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).

### 2022-10-21 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 % 8 - We find that clades innovate.
-% 9 - So basically species elaborate (in line with the micro evol theory of evolving along the line fo major axis of variation) and clades innovate (in line with the observations at a macro and mega level of evolutionary jumps leading to big diversity at high levels).
+% 9 - So basically species elaborate (in line with the micro evol theory of evolving along the line for major axis of variation) and clades innovate (in line with the observations at a macro and mega level of evolutionary jumps leading to big diversity at high levels).
 
@@
 The extent to which elaboration and innovation contribute to phenotypic diversity among species and clades is critical to our understanding of the origins and maintenance of biodiversity across scales from the micro- to macro and mega-evolutionary.
-Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks among species and clades.
-Our analyses show that elaboration is a common feature at all scales, consistent with theory.
-We also find that innovation is a common and major contributor to avian morphological diversity among clades.
-Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
-These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+Here we apply new multi-trait methods to modelling the traits nested covariance structure to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks among species and clades.
+Our analyses show that elaboration dominates species-level phenotypic divergence which is consistent with micro-evolutionary theory.
+However, we also find that innovation is relatively common at the clade level suggesting multiple routes to specific phenotypic innovation throughout avian evolutionary hiostory.
+These results show that although phenotypic elaboration and innovation occur throughout evolutionary scales, elaboration seems to dominate at the micro-evolutionary scales and innovation is more prevalent at the macro-evolutionary scale.
+This interaction suggest that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and are a common evolutionary route that explain both the amazing avian diversity and phenotypic conservatism.
+%TG: bleuargh...
 
@@
 Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
+%TG: cite some of these numerous studies.
 Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
@@
 More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
-For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022} are consistent with specific innovation.
+For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022}
+%TG: potential add Marcy et al 2020 as a ref here too (hidden self-cite)?
+are consistent with specific innovation.
 Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
@@
 To assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of phenotypic variation) for
-i) the entire phylogeny of all species included in the data (hereafter the "whole tree" level),
+i) the entire phylogeny of all species included in the data (hereafter the global phylogenetic level),
 ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
@@
 % paragraph 4: results for elaboration and innovation for species
-We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species specific measure of random innovation and elaboration.
-To do this we estimated the multilinear algebraic projection of each species phenotype onto the three (whole tree, super-order, and order) major axis of phenotypic variation.
-Phylogenetic distributions of random innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or whole tree level (Fig.
+We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species specific measures of random innovation and elaboration.
+To do this we calculated the multilinear algebraic projection of each species phenotype onto the three major axis of phenotypic variation (global phylogeny, specific super-order, and specific order).
+Phylogenetic distributions of random innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig.
 \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-However, the extent of both elaboration and random innovation tends to reduce as we move from whole tree to within order comparison.
+However, the extent of both elaboration and random innovation tends to reduce as we move from global phylogeny to within order comparison.
 Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
-The whole tree's major axis of phenotypic variation primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
+The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
 Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.
@@
 Overall we found that, as expected, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}.
-We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration.
-%(Fig.
-S{#fig_correlations}).
-%TODO: maybe add to supp?
+We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
 However, we observe various combinations of elaboration and random innovation across birds.
 Examples of high innovation and/or elaboration are often clustered within clades.
-For example, compared to the whole tree major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
+For example, compared to global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
 In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
@@
 Blue and yellow colors highlight species with relatively low novelty whereas orange to red colours indicate high novelty.
-The novelty is represented as innovation (upper circle) and elaboration (lower circle) concentrically represented in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing novelty for each species at the order levels then at the super-order level and finally at the entire phylogeny level (@@@ THIS SHOULD POINT TO THE ADDITIONAL PLOTS TO BE ADDED IN THE FIGURE).
+The novelty is represented as innovation (upper circle) and elaboration (lower circle) concentrically represented in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing novelty for each species at the order levels then at the super-order level and finally at the entire phylogeny level.
 We can observe an increase of novelty at deeper phylogenetic with novelty consistently composed of more elaboration than innovation.
@@
 To do this we translated the major axes of phenotypic variation of each clade's (i.e.
-super-orders and orders) phylogenetic variance-covariance matrices onto the whole tree's major axis of phenotypic variation so that they shared the same origin in the shapespace.
-We then used multilinear algebra to project a focal clade's major axis of phenotypic variation onto the whole tree's major axis of phenotypic variation  (Fig.S\ref{Fig:cheat_sheet}).
-We interpret the measured projection (distance onto the phylogenetic major axis of phenotypic variation) as the clade's elaboration score and the measured rejection (distance from the major axis of phenotypic variation) as the clade's innovation score.
+super-orders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis of phenotypic variation so that they shared the same origin in the shapespace.
+We then used multilinear algebra to project a focal clade's major axis of phenotypic variation onto the global phylogenetic major axis of phenotypic variation (Fig.S\ref{Fig:cheat_sheet}).
+We interpret the measured projection (distance onto the phylogenetic major axis of phenotypic variation) as the clade's elaboration score and the measured rejection (distance from the major axis of phenotypic variation) as the clade's specific innovation score.
 These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-We found that for most clades the phenotypic major axis of phenotypic variation is not aligned with the whole tree's major axis of phenotypic variation , (Fig. \ref{Fig:ellipses}).
-Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, specific innovation is a more common generator of morphological diversity than elaboration in deep time.
+We found that for most clades the phenotypic major axis of phenotypic variation is not aligned with the global phylogenetic major axis of phenotypic variation (Fig. \ref{Fig:ellipses}). 
+Despite the limitations representing eight-dimensional space in 2D, our plots of the average elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis of phenotypic variation and each super order (Fig.
+\ref{Fig:ellipses}a) and order (Fig.
+\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
+\ref{Fig:orthogonality}).
+Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, specific innovation is a more common generator of morphological diversity than elaboration in deep time.
 However, there is substantial variation in elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
@@
 These patterns hold across scales including within super-orders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
-This suggests that there is no deep time analogue of the evolutionary line of least evolutionary resistance and that across scales there is a remarkable flexibility in the routes to innovation.
+Our observations of heterogeneity in the orientation of the major axis of phenotypic variation among clades is further supported by consistent evidence for orthogonality of clades relative to the major phylogenetic axis of phenotypic variation (Fig.
+\ref{Fig:ellipses} - Fig.
+\ref{Fig:orthogonality}).
+The median angle of the major axis of phenotypic variation for subclades approaches orthogonality, differing from the phylogenetic major axis of phenotypic variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig.
+\ref{Fig:orthogonality}).
+Comparisons of orientation among subclades (i.e.
+orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig.
+\ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
+This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
+This suggests that there is no deep time analogue of the evolutionary line of least evolutionary resistance and that across scales there is a remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
 
@@
 \noindent \textbf{Figure \ref{Fig:ellipses}:}
-Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
+Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
 We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
@@
 The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
-Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
+Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot.
+%TG: TODO: add figure with unscalled ellipses in the supplementaries
+Panel B) shows the results for each order.
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
@@
 \bigskip
-
-This variability in the routes to innovation implies a distinct lack of deep-time evolutionary constraint in the dominant axes of evolution within clades.
-Although the whole tree's major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the whole tree's major axis of phenotypic variation is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend Fig.
-\ref{Fig:ellipses}-bar plots).
-This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes PC1 (Fig.
-S\ref{fig_ellipses_passeriformes}).
-In addition to the variable alignment of clade major axes of phenotypic variation, we further found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the whole tree variance-covariance matrix.
-For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the whole tree but uniformity in their beak shape components within the clade.
-In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
-However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
-
-\subsection{Species elaborate, clades innovate}
-
-% Paragraph 5: paradox + methods for comparing ellipses
-Although adaptive radiation theory implies that phenotypic divergence arises primarily along the major axis of phenotypic variation \cite{marroig2005size}, there have been many contrasting observations in the literature highlighting major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
-We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the major axis of phenotypic variation within different clades compared to the whole tree's of phenotypic variation.
-We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis of phenotypic variation (Fig.
-\ref{Fig:ellipses} - Fig.
-\ref{Fig:orthogonality}).
-Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the whole tree's major axis of phenotypic variation and each super order (Fig.
-\ref{Fig:ellipses}a) and order (Fig.
-\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
-\ref{Fig:orthogonality}).
-The median angle of the major axis of phenotypic variation for subclades approaches orthogonality, differing from the phylogenetic major axis of phenotypic variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
-Comparisons of orientation among subclades (i.e.
-orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
-This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
-These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
 
@@
 Dots represent the clades median orthogonality and the lines their 95\% CI.
-No super-order's major axis of phenotypic variation is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
-These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
-Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
-For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
+No super-order's major axis of phenotypic variation is parallel to the global phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
+These results are even clearer when comparing them to the global phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
+Furthermore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the global phylogeny relative to their sample size and sd.
 The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
@@
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+%TG: TODO: very easy low hanging fruit (maybe?), but not for this paper is to plot orthogonality through time.
 
 \bigskip
+
+The inferred evolutionary innovation can arise in any direction in trait space.
+Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend Fig.
+\ref{Fig:ellipses}-bar plots).
+This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes PC1 (Fig.
+S\ref{fig_ellipses_passeriformes}).
+In addition to the variable alignment of clade major axes of phenotypic variation, we further found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic variance-covariance matrix.
+For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogeny but uniformity in their beak shape components within the clade.
+In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+These observations illustrate multiple routes to innovation and imply that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
+
+\subsection{Species elaborate, clades innovate OR Ubiquitous elaboration, clade innovation}
 
@@
 \ref{Fig:relative_EI}).
-This difference confirms our observation that innovation is more common at the clade level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
-However, we suggest that rather than a paradox the reorientation of trait space can instead be described as a gradual process occuring in response to shifting selection pressures acting on species evolving in similar adaptive zones.
-This implies that gradual directional evolution \cite{pagel2022general} rather than megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple} may be sufficient to explain diversity in avian beak morphology.
+This difference confirms our observation that while specific innovation is common and perhaps even dominant at the clade level, elaboration (rather than random innovation) dominates at the species level.
+So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
+However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely related species.
+This idea is similar in concept to models of multiple adaptive peak models where species evolve in a heterogenous adaptive landscape and share similar responses to selection pressures within adaptive zones.
+%GT: Tip-toeing into OU territory and needs citations but I think this is a fair suggestion. %TG: fair point!
+This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple}, may be sufficient to explain diversity in avian beak morphology.
 
@@
 
-Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.
+Our results show that although at the species level most bird beaks are elaborating along a mega-evolutionary major axis of phenotypic variation, at the macro-evolutionary level innovation way from the major axis of phenotypic variation is much more common.
 This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
-Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
+Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
+Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of microevolution on a shifting adaptive landscape.
 
@@
 We ran two separate nested models.
-One model using one random term for the whole phylogeny and 35 nested random terms for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
+One model using one random term for the global phylogeny and 35 nested random terms for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
 And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
@@
 To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
-Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the whole tree and the passeriformes ones).
+Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the global phylogeny and the passeriformes ones).
 Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.

### 2022-10-25 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
-At macro- and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
-The presence of distinct morphological forms suggests instead that species or clades may evolve away from the line of least resistance.
-At a macroevolutionary level, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal, renaud2006conserved}.
-The extent to which elaboration and innovation contribute to phenotypic diversity among species and clades is critical to our understanding of the origins and maintenance of biodiversity across scales from the micro- to macro and mega-evolutionary.
-Here we apply new multi-trait methods to modelling the traits nested covariance structure to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks among species and clades.
-Our analyses show that elaboration dominates species-level phenotypic divergence which is consistent with micro-evolutionary theory.
-However, we also find that innovation is relatively common at the clade level suggesting multiple routes to specific phenotypic innovation throughout avian evolutionary hiostory.
-These results show that although phenotypic elaboration and innovation occur throughout evolutionary scales, elaboration seems to dominate at the micro-evolutionary scales and innovation is more prevalent at the macro-evolutionary scale.
-This interaction suggest that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and are a common evolutionary route that explain both the amazing avian diversity and phenotypic conservatism.
+However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
+At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
+The extent to which species and clades elaborate or innovate is poorly understood.
+Of particular importance is the potential contribution of reorientations in trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
+Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
+However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
+These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macro-evolutionary scales (i.e.
+species diverging within clades) and innovation is more prevalent at the mega-evolutionary scale (nested patterns of clades diverging within clades).
+This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macro-evolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
 %TG: bleuargh...
@@
 More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
-For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022}
+%TG: just for info, here's the exact quote from Endler et al:
+% innovation: if new components fall off the line;
+% random innovation: new components fall off the line in random directions;
+% specific innovation: new components fall off the line in only one direction (or a few)
+Studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022}
 %TG: potential add Marcy et al 2020 as a ref here too (hidden self-cite)?
-are consistent with specific innovation.
-Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
+are consistent with specific innovation, whereas the multitude of adaptations exhibited by related species in adaptive radiations may be consistent with random innovation.
+We do not attempt to distinguish random from specific innovation, instead we use the broader terminology from Endler et al.
+and focus on the relative roles of elaboration versus innovation at the level of i) species diverging within clades, and ii) clades diverging within nested clades.
+Relative to elaboration, innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
 These core concepts provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
-i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of phenotypic variation (specific innovation),
+i) idiosyncratic species-level innovation against clade-level innovation (i.e. reorientation of the directions of phenotypic variation),
 and ii) evolution along versus away from major axes of phenotypic variation.
@@
 2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
-and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale.
+and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
 To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
+To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
 To provide a baseline from which to calculate elaboration and innovation (i.e.
-the major axis of phenotypic variation) we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
+the major axis of phenotypic variation) we use phylogenetic generalised linear mixed models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
 Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
 From these models we can calculate major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix in trait space.
-We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
-The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any monophyletic clade).
-To assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of phenotypic variation) for
+We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
+The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
+To assess the roles of elaboration and innovation across scales we fit our pGLMMs as nested models in which we define variance-covariance matrices, and therefore major axes of phenotypic variation, for
 i) the entire phylogeny of all species included in the data (hereafter the global phylogenetic level),
@@
 % paragraph 4: results for elaboration and innovation for species
-We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species specific measures of random innovation and elaboration.
-To do this we calculated the multilinear algebraic projection of each species phenotype onto the three major axis of phenotypic variation (global phylogeny, specific super-order, and specific order).
-Phylogenetic distributions of random innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig.
+We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species-specific measures of innovation and elaboration.
+If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall of the line (innovation).
+To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
+Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig.
 \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-However, the extent of both elaboration and random innovation tends to reduce as we move from global phylogeny to within order comparison.
+However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparison.
 Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
@@
 We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
-However, we observe various combinations of elaboration and random innovation across birds.
+However, we observe various combinations of elaboration and innovation across birds.
 Examples of high innovation and/or elaboration are often clustered within clades.
-For example, compared to global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
+For example, compared to the global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
 In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
-Other clades such as the Apodidae (swifts) in the order Apodiformes and the distantly related but convergent Hirundidae (swallows and martins) have exceptionally wide (relative to length) beaks and display high levels of both elaboration and specific innovation.
-The dominance of elaboration when measured at the species level is consistent with idea of evolution along lines of least evolutionary resistance.
-However, the frequent clustering of species that elaborate, innovate, or both, again highlights the likelihood that major axes of phenotypic variation are not stable among clades.
-This leads us to ask: if most organisms evolve incrementally along major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?
+Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
+The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
+The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
+This leads us to ask: if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?
 
@@
 
-\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
-Blue and yellow colors highlight species with relatively low novelty whereas orange to red colours indicate high novelty.
-The novelty is represented as innovation (upper circle) and elaboration (lower circle) concentrically represented in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing novelty for each species at the order levels then at the super-order level and finally at the entire phylogeny level.
-We can observe an increase of novelty at deeper phylogenetic with novelty consistently composed of more elaboration than innovation.
-%TODO: add density plots of the innovations and elaboration at different levels.
-The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
+\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
+Blue and yellow colors on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
+Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order-level, then at the super-order level, and finally at the global phylogenetic level.
+We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
 
-\subsection{Multiple routes to specific innovation throughout avian evolutionary history}
+\subsection{Multiple routes to innovation throughout avian evolutionary history}
 
-To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and specific innovation of different clades.
+To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.
 To do this we translated the major axes of phenotypic variation of each clade's (i.e.
-super-orders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis of phenotypic variation so that they shared the same origin in the shapespace.
+super-orders and orders) phylogenetic variance-covariance matrix onto the global phylogenetic major axis of phenotypic variation so that they shared the same origin in the shapespace.
 We then used multilinear algebra to project a focal clade's major axis of phenotypic variation onto the global phylogenetic major axis of phenotypic variation (Fig.S\ref{Fig:cheat_sheet}).
-We interpret the measured projection (distance onto the phylogenetic major axis of phenotypic variation) as the clade's elaboration score and the measured rejection (distance from the major axis of phenotypic variation) as the clade's specific innovation score.
+We interpret the measured projection (distance onto the phylogenetic major axis of phenotypic variation) as the clade's elaboration score and the measured rejection (distance from the major axis of phenotypic variation) as the clade's innovation score.
+A high innovation score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g.
+the direction of evolution of an order differs from its parent super-order).
 These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-We found that for most clades the phenotypic major axis of phenotypic variation is not aligned with the global phylogenetic major axis of phenotypic variation (Fig. \ref{Fig:ellipses}). 
-Despite the limitations representing eight-dimensional space in 2D, our plots of the average elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis of phenotypic variation and each super order (Fig.
+We found that for most clades the major axis of phenotypic variation is not aligned with the global phylogenetic major axis of phenotypic variation (Fig. \ref{Fig:ellipses}). 
+Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis of phenotypic variation and each super-order (Fig.
 \ref{Fig:ellipses}a) and order (Fig.
-\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
+\ref{Fig:ellipses}b) highlight the striking variation in the orientation of phenotypic major axes among clades (Fig.
 \ref{Fig:orthogonality}).
-Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, specific innovation is a more common generator of morphological diversity than elaboration in deep time.
+Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep time.
 However, there is substantial variation in elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
@@
 Coraciiformes; Fig.
-\ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g.
+\ref{Fig:ellipses}), and both relatively high elaboration and innovation (e.g.
 Podicipediformes; Fig.
@@
 These patterns hold across scales including within super-orders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
-Our observations of heterogeneity in the orientation of the major axis of phenotypic variation among clades is further supported by consistent evidence for orthogonality of clades relative to the major phylogenetic axis of phenotypic variation (Fig.
+Our observations of heterogeneity in the orientation of the major axis of phenotypic variation among clades is further supported by consistent evidence for orthogonality of clades relative to the global phylogenetic major axis of phenotypic variation (Fig.
 \ref{Fig:ellipses} - Fig.
 \ref{Fig:orthogonality}).
-The median angle of the major axis of phenotypic variation for subclades approaches orthogonality, differing from the phylogenetic major axis of phenotypic variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig.
+The median angle of the major axis of phenotypic variation for subclades approaches orthogonality, differing from the global phylogenetic major axis of phenotypic variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig.
 \ref{Fig:orthogonality}).
@@
 This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
-This suggests that there is no deep time analogue of the evolutionary line of least evolutionary resistance and that across scales there is a remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
+This suggests that there is no deep-time analogue of the genetic line of least resistance and that across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
 
@@
 \noindent \textbf{Figure \ref{Fig:ellipses}:}
-Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
-We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
-The first inset ellipse plot shows the positions of all super-order ellipses relative to the all-bird phylogenetic ellipse.
-Subsequent inset plots show the results for each super-order. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
+We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all super-order ellipses relative to the global phylogenetic ellipse.
+Subsequent inset plots show the results for each super-order. 
+Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
 The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
 The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
-Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot.
-%TG: TODO: add figure with unscalled ellipses in the supplementaries
+These scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation for each plot. %TG: TODO: add figure with unscalled ellipses in the supplementaries
 Panel B) shows the results for each order.
@@
 \noindent \textbf{Figure \ref{Fig:orthogonality}:}
-Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent parent's clade.
-Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
-Dots represent the clades median orthogonality and the lines their 95\% CI.
-No super-order's major axis of phenotypic variation is parallel to the global phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
-These results are even clearer when comparing them to the global phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
+Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent's parent clade.
+Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+Dots represent the median orthogonality of each clade and the lines their 95\% CI.
+No super-order's major axis of phenotypic variation is parallel to the global phylogenetic major axis (red lines and circles) and 12/25 orders are at least half orthogonal to the major axis of their super-order (green lines and circles).
+These results are even clearer when comparing them to the global phylogenetic major axis (green dashed lines and open circles) where only 3/25 are less than half orthogonal.
 Furthermore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
-For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the global phylogeny relative to their sample size and sd.
+For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the global phylogenetic major axis relative to their sample size and sd.
 The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
 Only two out of six super-orders; 6/25 orders relative to their super-order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
-This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
+This is the result of the variation in sample size (n) and standard variation (sd) among clades. 
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
@@
 
-The inferred evolutionary innovation can arise in any direction in trait space.
+Evolutionary innovation can arise in any direction in trait space.
 Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend (Fig.
 \ref{Fig:ellipses}-bar plots).
-This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes PC1 (Fig.
+This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
 S\ref{fig_ellipses_passeriformes}).
-In addition to the variable alignment of clade major axes of phenotypic variation, we further found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic variance-covariance matrix.
+In addition to the variable alignment of clade major axes of phenotypic variation, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic major axis of phenotypic variation.
 For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogeny but uniformity in their beak shape components within the clade.
 In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
+These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler et al's random innovation) or follow a single direction (akin to Endler et al's specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
 
-\subsection{Species elaborate, clades innovate OR Ubiquitous elaboration, clade innovation}
+\subsection{Species elaborate, clades innovate}
 
@@
 We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
-We then measured the difference between both areas and found that there was a significant difference in innovation between species and clades but no clear difference in elaboration (Fig.
+We then measured the difference between both areas and found that there was a significant difference in innovation among species and clades but no clear difference in elaboration (Fig.
 \ref{Fig:relative_EI}).
-This difference confirms our observation that while specific innovation is common and perhaps even dominant at the clade level, elaboration (rather than random innovation) dominates at the species level.
+This difference confirms our observation that while innovation is common and perhaps even dominant at the clade level, elaboration dominates at the species level.
 So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
-However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely related species.
+However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species.
 This idea is similar in concept to models of multiple adaptive peak models where species evolve in a heterogenous adaptive landscape and share similar responses to selection pressures within adaptive zones.
@@
 \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
-A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 clades.
-The scores for each clade is scaled by the maximum score within each clade (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
+A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each of the 35 clades.
+The scores for each clade are scaled by the maximum score within the clade (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
 This allows us to compare both types (species vs. clades) despite their different sample sizes.
@@
 
-Our results show that although at the species level most bird beaks are elaborating along a mega-evolutionary major axis of phenotypic variation, at the macro-evolutionary level innovation way from the major axis of phenotypic variation is much more common.
-This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
+Our results show that although at the species-level most bird beaks are elaborating along a mega-evolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
+This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic levels could thus explain the diversity of bird beaks we observe today.
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
-Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of microevolution on a shifting adaptive landscape.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
+Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
 
@@
 \textbf{Phylogenetic GLMM analysis.}
-We ran a phylogenetic Bayesian generalised linear mixed effects model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
+We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
 
@@
 We ran two separate nested models.
-One model using one random term for the global phylogeny and 35 nested random terms for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
-And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
-This resulted in two multidimensional GLMM.
-One with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
-And one with a $5229 \times 8$ response variable, one global residual term and 30 random terms.
+The first model used one random term for the global phylogeny and 35 nested random terms, one for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
+This resulted in a multidimensional GLMM with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
+The second model was fitted to a subset of the dataset containing only the 5229 Passeriformes species with 30 nested random terms, one for each sub-order and families containing more than 15 species in our dataset (Fig.
+S\ref{fig_ellipses_passeriformes} .
+This resulted in a multidimensional GLMM with a $5229 \times 8$ response variables, one global residual term and 30 random terms.
 
-To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
-Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the global phylogeny and the passeriformes ones).
+To account for phylogenetic uncertainty we ran the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
+Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each of the models, i.e.
+the global phylogeny and the Passeriformes models).
 Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
 
-\textbf{mcmcmcglmm.}
-This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
-This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
-1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
-2) extracting the following parameters from the parameterisation chains:
-a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
-b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
-We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
-This effectively resulted in getting 10 exploitable posterior samples per tree.
+\textbf{mcmcmcglmmm.}
+This method runs relatively short Monte-Carlo Markov chains (MCMC) across a large number of different phylogenetic trees and concatenates the resulting short posteriors into one larger posterior distribution that contains the phylogenetic uncertainty information.
+We performed the method using the following protocol (see Fig. S\ref{Fig:mcmcmcglmm}).
+First we ran three independent MCMCglmm chains (hereafter the parameterisation chains) for 50k generations with the trait data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\%, i.e.
+with a very low weight on the priors;
+Next we extracted the following parameters from the parameterisation chains:
+a) the conservative burnin average time across the three chains, defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
+b) the mini-chains priors, defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
+We then ran 400 mini chains in parallel with the estimated burnin time and priors to run 10 samples past the burnin.
+This resulted in 10 exploitable posterior samples for each tree.
 We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
 S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
-Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
-The whole procedure is reproducible in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
+Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain - a 400 fold improvement.
+Code to reproduce the procedure is available in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
 
-\textbf{Projection/rejection analyses.}
-We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
-In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis of phenotypic variation onto that clade.
-In brief we can interpret where a species or a clade falls on the major axis of phenotypic variation of reference as its elaboration score and how far a species is from that major axis of phenotypic variation as its innovation score.
+\textbf{Elaboration/innovation scores using projection/rejection analyses.}
+We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to obtain  elaboration and innovation scores for bird beaks.
+We can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade) and project either species or another major axis of phenotypic variation onto that clade.
+In brief we can then interpret where a species or a clade falls on the major axis of phenotypic variation of reference as its elaboration score, and how far a species is from that major axis of phenotypic variation as its innovation score.
 We calculated the elaboration and innovation scores using two main approaches:
 1) by clades (i.e.
-to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes of phenotypic variation (Fig.
+to measure the elaboration/innovation among clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axis of phenotypic variation (Fig.
 \ref{Fig:ellipses});
 and 2) by species (i.e.
-to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes of phenotypic variation and b) the whole tree's major axes of phenotypic variation of their respective clade (Fig.
+to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axis of phenotypic variation and b) the major axis of phenotypic variation of their specific clade (Fig.
 \ref{Fig:phylogeny}).
 To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of phenotypic variation of interest and made the elaboration values absolute.
-This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
-The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
-The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
+In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
+The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection}).
+Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
 and implemented in the \texttt{dispRity} package \cite{dispRity}.
 
-\textbf{Groups' orthogonality.}
-We measured the orthogonality of each random term in our models (clades) compared to their parent clade and their parent parent's.
-For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
-To do so we measured the angles between the major axes of phenotypic variation of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade by randomly comparing major axes of phenotypic variation within a clade.
+\textbf{Group orthogonality.}
+We measured the orthogonality of each random term (i.e. clade) in our models compared to their parent clade and their parent's parent clade.
+For example, the variance-covariance matrices of the Columbimorphae random terms against the phylogeny random terms, the Columbiformes random terms against the Columbimorphae random terms (parent) and the whole phylogeny random terms (parent's parent).
+To measure orthogonality we calculated the angles between the major axes of phenotypic variation of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade, by randomly comparing major axes of phenotypic variation within a clade.
 We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
-We then measured the posterior probability of the angles within a clade being different than the angles between clades.
+We then measured the posterior probability of the angles within a clade being different to the angles among clades.
 These results are available in Fig. \ref{Fig:orthogonality}.

### 2022-10-25 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 
+## Adding orders and families that don't have a higher level covered
+if(any(no_match <- is.na(match(orders_n2$Group, orders$Group)))) {
+    fill_in <- orders_n2[which(no_match), ]
+    fill_in[,c(3:8)] <- NA
+    orders <- rbind(orders, fill_in)
+}
+## Adding orders and families that don't have a higher level covered
+if(any(no_match <- is.na(match(families_n2$Group, families$Group)))) {
+    fill_in <- families_n2[which(no_match), ]
+    fill_in[,c(3:8)] <- NA
+    families <- rbind(families, fill_in)
+}
+
 ## Sort each level by orthogonality score
@@
 
-## Adding starts here and there
+## Adding stars and thingies here and there
 get.signif <- function(prob) {
@@
 add.stars <- function(data, counter, shift = 0, total, ...) {
+
     text(data[3] + 0.03, (total+1)-counter + shift + 0.2, labels = get.signif(data[6]), cex = 1, ...)   
@@
 add.line <- function(data, counter, shift = 0, total, pch = 19, lty = 1, ...) {
+
     lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), lty = lty, ...)
@@
 }
+convert.spaces <- function(x, max) {
 
-## Add the number of species and sd to the group names
-convert.spaces <- function(x, max) {
     # shifts <- max-nchar(as.character(x))
@@
 }
+
+## Set up the group names
 group_names <- paste0(plot_table$Group, " (n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
@@
 for(i in 1:nrow(plot_table)) {
+    if(!all(is.na(plot_table[plot_order[i], c(3:8)]))) {
         add.line (plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
         add.stars(plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
+    }
 }
@@
 for(i in 1:nrow(families_n2_plot)) {
+    if(!all(is.na(plot_table[plot_order[i], c(3:8)]))) {
         add.line(families_n2_plot[i, ], shift = -1/3, counter = i+nrow(sub_orders), col = adjustcolor(colours[i+nrow(sub_orders)], alpha.f = 2/3), total = nrow(plot_table), lty = 2, pch = 21)
         add.stars(families_n2_plot[i, ], shift = -1/3, counter = i+nrow(sub_orders), col = adjustcolor(colours[i+nrow(sub_orders)], alpha.f = 2/3), total = nrow(plot_table))
+    }
 }

### 2022-10-25 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

--- 2022-10-07 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
@@
 #TODO: make the groups nested in the fable
-pdf("../Manuscript/Figures/orthogonality_results.pdf", width = 9, height = 9)
 source("../Functions/ploting.utilities.R")
@@
 
-## Sort each level by orthogonality score
-ortho.sort <- function(data) {
-    return(data[order(data$orthogonality, decreasing = TRUE), , drop = FALSE])
+## Adding orders and families that don't have a higher level covered
+if(any(no_match <- is.na(match(orders_n2$Group, orders$Group)))) {
+    fill_in <- orders_n2[which(no_match), ]
+    fill_in[,c(3:8)] <- NA
+    orders <- rbind(orders, fill_in)
 }
-## Plotting table (with the sorted orthogonality)
-plot_table <- rbind(ortho.sort(super_orders),
-                    ortho.sort(orders))
-                    # ortho.sort(sub_orders),
-                    # ortho.sort(families))
-colours <- c(rep(gg.color.hue(4)[1], nrow(super_orders)),
-             rep(gg.color.hue(4)[2], nrow(orders)))
-             # rep(gg.color.hue(4)[3], nrow(sub_orders)),
-             # rep(gg.color.hue(4)[4], nrow(families)))
-
-## Adding starts here and there
-get.signif <- function(prob) {
-    if(prob >= 0.99) {
-        return("***")
-    } else {
-        if(prob >= 0.95) {
-            return("**")
-        } else {
-            if(prob >= 0.90) {
-                return("*")
-            } else {
-                if(prob >= 0.80) {
-                    return(".")
-                } else {
-                    return("")
-                }
-            }
-        }
-    }
-}
-add.stars <- function(data, counter, shift = 0, total, ...) {
-    text(data[3] + 0.03, (total+1)-counter + shift + 0.2, labels = get.signif(data[6]), cex = 1, ...)   
-}
-add.line <- function(data, counter, shift = 0, total, pch = 19, lty = 1, ...) {
-    lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), lty = lty, ...)
-    points(data[3], (total+1)-counter + shift, pch = pch,  ...)
+## Adding orders and families that don't have a higher level covered
+if(any(no_match <- is.na(match(families_n2$Group, families$Group)))) {
+    fill_in <- families_n2[which(no_match), ]
+    fill_in[,c(3:8)] <- NA
+    families <- rbind(families, fill_in)
 }
 
-## Add the number of species and sd to the group names
-convert.spaces <- function(x, max) {
-    # shifts <- max-nchar(as.character(x))
-    # if(shifts > 0) {
-    #     return(paste(c(x, rep(" ", shifts)), collapse = ""))
-    # } else {
-    #     return(x)
-    # }
-    return(x)
-}
-group_names <- paste0(plot_table$Group, " (n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
-
-plot_order <- 1:length(group_names)
-
-## Plot frame
-par(mar = c(5, 17, 4, 2))
-plot(NULL, yaxt = "n", xaxt = "n", ylab = "", xlab = "orthogonality", ylim = c(1,nrow(plot_table)), xlim = c(0,1))
-abline(v = 0.5, lwd = 0.5, col = "grey")
-abline(v = 0.75, lty = 2, col = "grey", lwd = 0.5)
-axis(2, at = nrow(plot_table):1, labels = group_names[plot_order], las = 2)
-axis(1, at = c(0, 0.5, 1), labels = c(0, 0.5, 1))
-## Adding the results
-for(i in 1:nrow(plot_table)) {
-    add.line (plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
-    add.stars(plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
-}
-
-## Adding the n2 results
-orders_n2_plot <- orders_n2[match(orders[order(orders$orthogonality, decreasing = TRUE), ]$Group, orders_n2$Group), ]
-# families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
-for(i in 1:nrow(orders_n2_plot)) {
-    add.line(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 2/3), total = nrow(plot_table), lty = 2, pch = 21)
-    add.stars(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 2/3), total = nrow(plot_table))
-}
-# for(i in 1:nrow(families_n2)) {
-#     add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
-#     add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
-# }
-
+pdf("../Manuscript/Figures/orthogonality_results.pdf", width = 9, height = 9)
+plot.orthogonality(super_orders, orders_n2, orders, col = gg.color.hue(4)[1:2])
 legend("topleft", legend = c("Super orders (vs.
 phylogeny)", "Orders (vs.
-super orders)", "Orders (vs.
-phylogeny)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
-
-# legend("topleft", legend = c("Super orders (vs.
 phylogeny)", "Orders (vs.
-super orders)", "Orders (vs.
-phylogeny)", "Suborders (vs.
-passeriformes)", "Families (vs.
-suborders)", "Families (vs.
-passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
-
+super orders)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
 dev.off()
@@
 pdf("../Manuscript/Figures/orthogonality_results_passeriformes.pdf", width = 9, height = 9)
-source("../Functions/ploting.utilities.R")
-## Plotting table (with the sorted orthogonality)
-plot_table <- rbind(ortho.sort(sub_orders),
-                    ortho.sort(families))
-colours <- c(rep(gg.color.hue(4)[3], nrow(sub_orders)),
-             rep(gg.color.hue(4)[4], nrow(families)))
-group_names <- paste0(plot_table$Group, " (n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
-plot_order <- 1:length(group_names)
-
-## Plot frame
-par(mar = c(5, 17, 4, 2))
-plot(NULL, yaxt = "n", xaxt = "n", ylab = "", xlab = "orthogonality", ylim = c(1,nrow(plot_table)), xlim = c(0,1))
-abline(v = 0.5, lwd = 0.5, col = "grey")
-abline(v = 0.75, lty = 2, col = "grey", lwd = 0.5)
-axis(2, at = nrow(plot_table):1, labels = group_names[plot_order], las = 2)
-axis(1, at = c(0, 0.5, 1), labels = c(0, 0.5, 1))
-## Adding the results
-for(i in 1:nrow(plot_table)) {
-    add.line (plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
-    add.stars(plot_table[plot_order[i], ], counter = i, col = colours[plot_order[i]], total = nrow(plot_table))
-}
-
-## Adding the n2 results
-families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
-for(i in 1:nrow(families_n2_plot)) {
-    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+nrow(sub_orders), col = adjustcolor(colours[i+nrow(sub_orders)], alpha.f = 2/3), total = nrow(plot_table), lty = 2, pch = 21)
-    add.stars(families_n2_plot[i, ], shift = -1/3, counter = i+nrow(sub_orders), col = adjustcolor(colours[i+nrow(sub_orders)], alpha.f = 2/3), total = nrow(plot_table))
-}
-
+plot.orthogonality(sub_orders, families_n2, families, col = gg.color.hue(4)[3:4])
 legend("topleft", legend = c("Sub-orders (vs.
@@
 passeriformes)"), col = c(gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
-
-# legend("topleft", legend = c("Super orders (vs.
-phylogeny)", "Orders (vs.
-super orders)", "Orders (vs.
-phylogeny)", "Suborders (vs.
-passeriformes)", "Families (vs.
-suborders)", "Families (vs.
-passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
-
 dev.off()

### 2022-10-25 Functions/ploting.utilities.R
Update ploting.utilities.R

@@ -536,3 +536,92 @@ plot.correlations <- function(cor.results, col.sub, col.fun = colour.palette, le
 
     return(invisible())
 }
+
+
+
+
+
+
+
+
+## Sort each level by orthogonality score
+ortho.sort <- function(data) {
+
+    return(data[order(data$orthogonality, decreasing = TRUE), , drop = FALSE])
+}
+## Adding stars and thingies here and there
+get.signif <- function(prob) {
+    if(prob >= 0.99) {
+        return("***")
+    } else {
+        if(prob >= 0.95) {
+            return("**")
+        } else {
+            if(prob >= 0.90) {
+                return("*")
+            } else {
+                if(prob >= 0.80) {
+                    return(".")
+                } else {
+                    return("")
+                }
+            }
+        }
+    }
+}
+add.stars <- function(data, counter, shift = 0, total, ...) {
+
+    text(data[3] + 0.03, (total+1)-counter + shift + 0.2, labels = get.signif(data[6]), cex = 1, ...)   
+}
+add.line <- function(data, counter, shift = 0, total, pch = 19, lty = 1, ...) {
+
+    lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), lty = lty, ...)
+    points(data[3], (total+1)-counter + shift, pch = pch,  ...)
+}
+convert.spaces <- function(x, max) {
+
+    # shifts <- max-nchar(as.character(x))
+    # if(shifts > 0) {
+    #     return(paste(c(x, rep(" ", shifts)), collapse = ""))
+    # } else {
+    #     return(x)
+    # }
+    return(x)
+}
+## Plotting orthogonality
+plot.orthogonality <- function(level1, level2.1, level2.2, col) {
+    ## Plotting table (with the sorted orthogonality)
+    plot_table <- rbind(ortho.sort(level1),
+                        ortho.sort(level2.1))
+    colours <- c(rep(col[1], nrow(level1)),
+                 rep(col[2], nrow(level2.1)))
+
+    ## Set up the group names
+    group_names <- paste0(plot_table$Group, " (n = ", sapply(plot_table$n, convert.spaces, max = 4), " ; sd = ", sapply(round(plot_table$`ellipse sd`, 2), convert.spaces, max = 5), ")")
+    plot_order <- 1:length(group_names)
[…]

### 2022-10-25 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

@@
 ## Quick load
+source("../Functions/stats.utilities.R")
+library(dispRity)
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
@@
 
-## Distribution of elaboration and innovation at the group level
+## Distribution of elaboration and innovation at the clade level
 scale.group <- function(X) return(X/max(X))
@@
 elaboration_orders       <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
-innovation_orders        <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
+innovation_orders        <- lapply(get.disparity(results$order_results_between$distance, concatenate = FALSE), scale.group)
+## The objects above are the ones that contain all the elaboration and innovation values by clade level
+
+## These bits are for the Bhatt coeff
 elaboration_group <- rbind(do.call(rbind, elaboration_super_orders), do.call(rbind, elaboration_orders))
@@
 
-## Measuring the area under the curves
+## Measuring the area under the curves (for the Bhatt coeff)
 elaboration_species_auc <- apply(elaboration_species, 2, area.density)

### 2022-10-25 Analysis/05-Figure_with_phylogeny.Rmd
Update 05-Figure_with_phylogeny.Rmd

@@
 ## Save as png and pdf
-ggsave("../Manuscript/Figures/InnovElabTree_main_text.png", innovelab_plot_final, width=12, height=12)
-ggsave("../Manuscript/Figures/InnovElabTree_main_text.pdf", innovelab_plot_final, width=12, height=12)
+ggsave("../Manuscript/Figures/InnovElabTree_main_text.png", innovelab_plot, width=12, height=12)
+ggsave("../Manuscript/Figures/InnovElabTree_main_text.pdf", innovelab_plot, width=12, height=12)
 

### 2022-11-04 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 The extent to which species and clades elaborate or innovate is poorly understood.
-Of particular importance is the potential contribution of reorientations in trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
+Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
 Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
@@
 However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
-These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macro-evolutionary scales (i.e.
-species diverging within clades) and innovation is more prevalent at the mega-evolutionary scale (nested patterns of clades diverging within clades).
-This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macro-evolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macroevolutionary scales (i.e.
+species diverging within clades) and innovation is more prevalent at the megaevolutionary scale (nested patterns of clades diverging within clades).
+This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
 %TG: bleuargh...
@@
 % paragraph 2:  introducing the Endler view.
-Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
-Specifically, Endler et al. refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
+Endler \textit{et al}. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
+Specifically, Endler \textit{et al}. refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
 This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
-In contrast, they refer to evolution away from the major axis of phenotypic variation as \textit{innovation}.
-More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
-%TG: just for info, here's the exact quote from Endler et al:
-% innovation: if new components fall off the line;
-% random innovation: new components fall off the line in random directions;
-% specific innovation: new components fall off the line in only one direction (or a few)
-Studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022}
-%TG: potential add Marcy et al 2020 as a ref here too (hidden self-cite)?
-are consistent with specific innovation, whereas the multitude of adaptations exhibited by related species in adaptive radiations may be consistent with random innovation.
-We do not attempt to distinguish random from specific innovation, instead we use the broader terminology from Endler et al.
-and focus on the relative roles of elaboration versus innovation at the level of i) species diverging within clades, and ii) clades diverging within nested clades.
+In contrast, they refer to evolution away from the major axis of phenotypic variation, either in random directions or along limited axes, as \textit{innovation} \cite{endler2005animal}.
 Relative to elaboration, innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
-These core concepts provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
+These core concepts can now be defined using robust multivariate statistics about trait covariance applied at multpiple phylogentic scales and provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
 i) idiosyncratic species-level innovation against clade-level innovation (i.e. reorientation of the directions of phenotypic variation),
 and ii) evolution along versus away from major axes of phenotypic variation.
-The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
+The contributions of these multivariate, evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
 Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.
@@
 and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
-To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
+To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric, morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
+To meet the second two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
 To provide a baseline from which to calculate elaboration and innovation (i.e.
@@
 Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
-From these models we can calculate major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix in trait space.
+From these models we can calculate and compare major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix in trait space.
 We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
@@
 and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
-These nested partitions of multivariate trait space provide the basis for all subsequent analyses.
+These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation at multiple scales.
 
@@
 \ref{Fig:orthogonality}).
-Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep time.
-However, there is substantial variation in elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
+Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep time (i.e evidence at the clade scale).
+However, there is substantial variation in elaboration and innovation among clades including relatively low elaboration and innovation (e.g.
 Cuculiformes; Fig.
@@
 Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend (Fig.
-\ref{Fig:ellipses}-bar plots).
-This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
+\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned (Fig.
+\ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
+This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
 S\ref{fig_ellipses_passeriformes}).
-In addition to the variable alignment of clade major axes of phenotypic variation, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic major axis of phenotypic variation.
+In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single (non primary?) axis) or higher-dimensional than expected (there is no dominant major axis) from the global phylogenetic major axis of phenotypic variation.
 For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogeny but uniformity in their beak shape components within the clade.
 In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler et al's random innovation) or follow a single direction (akin to Endler et al's specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
+These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
 
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/relative_EI.pdf}
+   \includegraphics[width=0.9\textwidth]{Figures/relative_EI_gg_style.pdf}
 \caption{.}
@@
 \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
-A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each of the 35 clades.
-The scores for each clade are scaled by the maximum score within the clade (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
-This allows us to compare both types (species vs. clades) despite their different sample sizes.
-The Bhattacharryya Coefficient represents the overlap of the area under the curve for species and clades (a BC of 0 represents no overlap).
-B - same as A but for the innovation score.
-We can see a clear difference in the amount of innovation across the posterior samples between species or clades.
+a) the elaboration (blue) for each species in each 4000 posterior samples compared to the elaboration for each of the 35 clades (yellow).
+The scores for each clade are scaled by the maximum score within the clade (corresponding to the scaled ellipses in figure \ref{Fig:ellipses}) and the scores for species are scaled by the maximum species elaboration score.
+This allows us to compare elaboration in species and clades despite their different sample sizes.
+We use the Bhattacharryya coefficient (BC) to quantify the overlap of the area under the curve for species and clades. 
+BC = 0 indicates no overlap.
+For elaboration BC = 0.217, indicating broad similarity in the amount of elaboration across the posterior samples between species or clades. 
+b) the same as a) but using innovation score. 
+For innovation BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.
 
-% Paragraph 8: conculsion
+% Paragraph 8: conclusion
 \subsection{Conclusions}
@@
 Our results show that although at the species-level most bird beaks are elaborating along a mega-evolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
-This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic levels could thus explain the diversity of bird beaks we observe today.
+This nested structure of elaboration at a higher (species) taxonomic level and innovation at a lower (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explaination of the mahooosive diversity of birdbeaks and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
 Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
@@
 The second model was fitted to a subset of the dataset containing only the 5229 Passeriformes species with 30 nested random terms, one for each sub-order and families containing more than 15 species in our dataset (Fig.
-S\ref{fig_ellipses_passeriformes} .
+S\ref{fig_ellipses_passeriformes}.
 This resulted in a multidimensional GLMM with a $5229 \times 8$ response variables, one global residual term and 30 random terms.
@@
 The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
-The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
+The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{GitHub} \cite{githubrepo}.
 
@@
 \bibliography{references}
+%TODO: add ref to ape.
 

### 2022-11-04 Analysis/01-Data_preparation.Rmd
renamed analyses

@@
 All the custom code and custom functions called in this document can be downloaded and installed in `R`.
+
+```{r packages_install, eval = FALSE}
+## Optional - this snippet installs all the packages required to run the analyses
+install.packages(c("devtools", "dispRity", "MCMCglmm", "xtable", "knitr", "ggplot2"))
+devtools::install_github("TGuillerme/mcmcmcglmmm")
+
+## The following packages are for ploting the phylogeny figure
+install.packages(c("phytools", "motmot", "viridis", "RColorBrewer", "colorspace", "ggnewscale", "treeplyr", "BiocManager"))
+BiocManager::install(c("ggtree", "ggtreeExtra", "treeio"))
+```
 
@@
 
-How many dimensions do we need to represent at least 95% of the variance in each group
+How many dimensions do we need to represent at least 95% of the variance in each group:
 
@@
 shape_axes <- select.axes(custom.subsets(shapespace, group = clades))
-form_axes  <- select.axes(custom.subsets(formspace, group = clades))
 
@@
 
-```{r, echo = FALSE}
-kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X3")
-```
-
 ```{r, echo = FALSE, fig.height = 18, fig.width = 18}
@@
 
-```{r, echo = FALSE, fig.height = 18, fig.width = 18}
-plot(form_axes)
-```
-Fig X1: variance per axes per superorder and order in the formspace
-
-By analysing the distribution of the variance within each group, we will need `r length(shape_axes$dimensions)` dimensions for the shape space and `r length(form_axes$dimensions)` for the form space.
+By analysing the distribution of the variance within each group, we will need `r length(shape_axes$dimensions)` dimensions for the shape space.
 
@@
 Here we are going to group the spaces, trees and levels for subsequent levels (nested).
-There are 5 levels: Super order (1), Order (2), Sub-order (3), Super families (4) and families (5).
-The script above creates the model for the two first levels, the subsequent models create the mini-chains for the following overlapping pairs of models (order + sub-order; sub-order + super families; super families + families).
+There are 2 levels: Super-order (1) and order (2).
+The script above creates the model for the two first levels.
 
@@
 
-## Getting the data ready for each bird super order
-shapespace_superorder_lvl_order_suborder <- prep.data(
-                                          level = 1, lvl.inc = c(2,3),
-                                          clades    = clades_list,
-                                          space     = shapespace,
-                                          dim       = shape_axes$dimensions,
-                                          consensus = consensus_tree,
-                                          trees     = trees_list,
-                                          verbose   = TRUE)
-save(shapespace_superorder_lvl_order_suborder, file = "../Data/Processed/shapespace_superorder_lvl_order_suborder.rda")
-
-## Getting the data ready for each bird order
-shapespace_order_lvl_suborder_superfam <- prep.data(
-                                          level = 2, lvl.inc = c(3,4),
-                                          clades    = clades_list,
-                                          space     = shapespace,
-                                          dim       = shape_axes$dimensions,
-                                          consensus = consensus_tree,
-                                          trees     = trees_list,
-                                          verbose   = TRUE)
-save(shapespace_order_lvl_suborder_superfam, file = "../Data/Processed/shapespace_order_lvl_suborder_superfam.rda")
-
-## Getting the data ready for each bird sub order
-shapespace_suborder_lvl_superfam_fam <- prep.data(
-                                          level = 3, lvl.inc = c(4,5),
-                                          clades    = clades_list,
-                                          space     = shapespace,
-                                          dim       = shape_axes$dimensions,
-                                          consensus = consensus_tree,
-                                          trees     = trees_list,
-                                          verbose   = TRUE)
-save(shapespace_suborder_lvl_superfam_fam, file = "../Data/Processed/shapespace_suborder_lvl_superfam_fam.rda")
-
-## Getting the data ready for each bird super family
-shapespace_superfam_lvl_fam <- prep.data(
-                                          level = 4, lvl.inc = 5,
-                                          clades    = clades_list,
-                                          space     = shapespace,
-                                          dim       = shape_axes$dimensions,
-                                          consensus = consensus_tree,
-                                          trees     = trees_list,
-                                          verbose   = TRUE)
-save(shapespace_superfam_lvl_fam, file = "../Data/Processed/shapespace_superfam_lvl_fam.rda")
-
 ## Get a list of all selected levels for the record (and number of elements per levels)
@@
 orders           <- shapespace_allbirds_lvl_superorder_order[[1]]$levels[[2]]
-sub_orders       <- lapply(shapespace_order_lvl_suborder_superfam, function(x) x$levels[[1]])
-names(sub_orders) <- NULL
-sub_orders       <- unlist(sub_orders, recursive = FALSE)
-super_families   <- lapply(shapespace_order_lvl_suborder_superfam, function(x) x$levels[[2]])
-names(super_families) <- NULL
-super_families   <- unlist(super_families, recursive = FALSE)
-families         <- lapply(shapespace_superfam_lvl_fam, function(x) x$levels[[1]])
-names(families) <- NULL
-families         <- unlist(families, recursive = FALSE)
 
@@
 levels_list <- list("super_orders"   = unlist(lapply(super_orders, length)),
-                    "orders"         = unlist(lapply(orders, length)),
-                    "sub_orders"     = unlist(lapply(sub_orders, length)),
-                    "super_families" = unlist(lapply(super_families, length)),
-                    "families"       = unlist(lapply(families, length)))
+                    "orders"         = unlist(lapply(orders, length)))
+
 save(levels_list, file = "../Data/Processed/levels_list.rda")
@@
 ## Passeriformes focus
+
+The following section repeats the procedure but focusing on Passeriformes only with two levels: the sub-orders and the families within Passeriformes.
 
@@
 ## Loading the Passeriformes clades
-load("../Data/Processed/clade.info_passerines.rda")
+load("../Data/Raw/clade_info_passerines.rda")
 ## Clean the levels

### 2022-11-04 Analysis/02-mcmcmcglmmm_analysis.Rmd
renamed analyses

@@
 
+Repeat this snippet three times (by changing the name of the file in the last line to `chain2` and `chain3`) to get the three parametrisation chains
+
 From these three chains, we first extracted the three burnin periods (defined as the generation when the chain reaches the median likelihood value times 1.25).
@@
 Each individual replicate runs for around 50 hours on a 2.4Ghz CPU and requires around 9GB of RAM.
-The whole process was completed in around 2 CPU years.
+The whole process was completed in around 2-3 CPU years.
 
@@
 save(shapespace_allbirds_lvl_superorder_order, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
-
 ```

### 2022-11-04 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd
renamed analyses

@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
+superorder_shapespace <- MCMCglmm.subsets(n = 4000,
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
+order_shapespace <- MCMCglmm.subsets(n = 4000,
       data          = shapespace,
@@
 
-<!-- TODO: add the pictures of the different tests here -->
-
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
+```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
 ## Run the analyses between groups
@@
 
-```{r, echo = FALSE, eval = FALSE}
+These sections are used for generated specific data for subsequent plots.
+
+```{r, echo = FALSE, eval = TRUE}
 ## Getting the plotting data
@@
               "Order level"       = c(elaborations_order, innovations_order)),
-        xlab = "Novelty per nested levels", ylab = "Novelty", col = colour.palette(5)[1:3])
+        xlab = "", ylab = "Elaboration/Innovation", col = colour.palette(5)[1:3])
 boxplot(cbind("Elaboration" = c(elaborations_tree, elaborations_super, elaborations_order),
               "Innovation"  = c(innovations_tree, innovations_super, innovations_order)),
-        xlab = "Novelty per type", ylab = "Novelty", col = colour.palette(5)[4:5])
+        xlab = "", ylab = "Elaboration/Innovation", col = colour.palette(5)[4:5])
 dev.off()
@@
 # Passeriformes shapespace analyses
+
+This runs the same analyses for the passeriformes analyses
 
@@
 
-```{r, echo = FALSE, eval = FALSE}
+```{r, echo = FALSE, eval = TRUE}
 load("../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")

### 2022-11-04 Analysis/07-Orthogonality_and_statistics.Rmd
renamed analyses

--- 2022-10-25 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
@@
 plot.orthogonality(super_orders, orders_n2, orders, col = gg.color.hue(4)[1:2])
-legend("topleft", legend = c("Super orders (vs.
+legend("topleft", legend = c("Super-orders (vs.
 phylogeny)", "Orders (vs.
 phylogeny)", "Orders (vs.
-super orders)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
+super-orders)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8, bty = "n")
 dev.off()
@@
 sub-orders)", "Families (vs.
-passeriformes)"), col = c(gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8)
+passeriformes)"), col = c(gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8, bty = "n")
 dev.off()

### 2022-11-04 Analysis/09-Figure_key.Rmd
renamed analyses

--- 2022-05-27 Analysis/08-Figure_key.Rmd
@@
 source("../Functions/covar.utilities.R")
-
 ```
@@
 rm(shapespace_allbirds_lvl_superorder_order)
-
 ```
@@
 ## Transform them into a colour gradient
-col.grad <- colorRamp(c("blue","orange"))
+col.grad <- colorRamp(rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#66c2a5", "#3288bd")))
 cols_elab <- col.grad(elaborations/max(elaborations))

### 2022-11-08 Manuscript/main_flipped_narrative.tex
updated MS

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
+\noindent {\normalsize \sc
+Thomas Guillerme$^{1,*}$, 
+Jen A. Bright$^{2}$,
+Christopher R. Cooney${^1}$,
+Emma C. Hughes${^1}$,
+Zoe K. Varley${^3}$,
+Natalie Cooper$^{3,+}$,
+Andrew P. Beckerman$^{1,+}$,
+and Gavin H. Thomas$^{1,4,+}$}\\
 \noindent {\small \it 
 $^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
+$^2$Department of Biological and Marine Sciences, University of Hull, Hull, HU6 7RX, United Kingdom.\\
+$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
+$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
 
@@
 \bigskip
-\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\ 
+\noindent{*\bf Corresponding author:} \textit{guillert@tcd.ie}; \noindent{+\bf contributed equally.}\\
 \vspace{1in}
@@
 Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
-A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
-However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
+A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.https://www.overleaf.com/project/62f128d3d7c36e4efd1e9b21
+However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
 At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
 The extent to which species and clades elaborate or innovate is poorly understood.
-Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
-Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+Of particular importance is the potential contribution of reorientations of the phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
+Here we apply new multi-trait methods that model the reorientations of the trait space via the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
 Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
@@
 This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
-%TG: bleuargh...
 
@@
 The contributions of these multivariate, evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
-Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.
+Here we estimate for the first time, using novel statistical methods, the relative contribution of elaboration and innovation to the remarkable global radiation and diversity of birds beaks.
 
@@
 To meet the second two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-To provide a baseline from which to calculate elaboration and innovation (i.e.
+% start: oldversion
+% To provide a baseline from which to calculate elaboration and innovation (i.e.
 the major axis of phenotypic variation) we use phylogenetic generalised linear mixed models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
-Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
-From these models we can calculate and compare major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix in trait space.
-We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
+% Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
+% From these models we can calculate and compare major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix of the trait space.
+% end: oldversion
+% start: newversion
+In order to estimate the major axis of phenotypic variation we fit pGLMM models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
+From the posterior distribution of the variance-covariance matrix at any scale, the major axis can be identified from the leading eigenvector of the variance-covariance matrix.
+We can estimate levels of elaboration and innovation based on that major axis of phenotypic variation (see Methods).
+% end: newversion
+We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
 The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
@@
 To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
+We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globaly more elaboration and innovation at the global phylogeny and the super-orders levels compared to the orders level.
 Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig.
@@
 
-Overall we found that, as expected, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}.
+Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}.
 We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
@@
 Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order-level, then at the super-order level, and finally at the global phylogenetic level.
+The boxplots at the bottom of the figure summarises the distribution of the median innovation and elaboration values for all levels (bottom left) showing that elaboration is more common than elaboration;
+and the pooled median innovation and elaboration values for all three different levels (bottom right) showing that elaboration and innovation is more common at higher taxonomic levels than at the orders level.
 We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
@@
 Evolutionary innovation can arise in any direction in trait space.
-Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned (Fig.
+Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig.
+\ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly with PC1 (Fig.
 \ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
@@
 S\ref{fig_ellipses_passeriformes}).
-In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single (non primary?) axis) or higher-dimensional than expected (there is no dominant major axis) from the global phylogenetic major axis of phenotypic variation.
+In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis; e.g.
+in Galliformes) or higher-dimensional than expected (there is no clear dominant major axis; e.g.
+in Podicipediformes) from the global phylogenetic major axis of phenotypic variation.
 For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogeny but uniformity in their beak shape components within the clade.
@@
 So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
-However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species.
+However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species \cite{hansen1997stabilizing}.
 This idea is similar in concept to models of multiple adaptive peak models where species evolve in a heterogenous adaptive landscape and share similar responses to selection pressures within adaptive zones.
@@
 We use the Bhattacharryya coefficient (BC) to quantify the overlap of the area under the curve for species and clades. 
-BC = 0 indicates no overlap.
-For elaboration BC = 0.217, indicating broad similarity in the amount of elaboration across the posterior samples between species or clades. 
+For elaboration the overlap is BC = 0.217, indicating no distinguishable dissimilarity in the amount of elaboration across the posterior samples between species or clades. 
 b) the same as a) but using innovation score. 
-For innovation BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.
+For innovation the overlap BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.
 
@@
 Our results show that although at the species-level most bird beaks are elaborating along a mega-evolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
-This nested structure of elaboration at a higher (species) taxonomic level and innovation at a lower (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
+This nested structure of elaboration at a lower (species) taxonomic level and innovation at a higher (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explaination of the mahooosive diversity of birdbeaks and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
 Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.

### 2022-11-09 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
-This script take around 6 minutes to compile.
+This script take around 4-5 hours to compile.
+Note that the bulk of the calculation time is spent on the snippets `projections_whole_tree` (lines 212-256) and `projections_passeriformes` (lines 444-489).
+This can be reduced by reducing the number of samples `n`.
+For the publication we used the full sample size (`n = 4000`) but lower numbers such as `n = 100` give correct estimates as well and reduce the calculation time to around 6 minutes! 
+For the detailed procedure on how the projection/rejection analysis works, see [this `dispRity` vignette](https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html).
 
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 4000,
+superorder_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 4000,
+order_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 
-Here's a visualisation of how the variance in VCVs affects the averaging artefact:
+Here's a visualisation of how the variance in VCVs affects the averaging artifact:
 
@@
 
-Or even more obvious (with an direction that does not really exist (due to averaging)):
+Or even more obvious (with an direction that does not really exist - due to averaging):
 
@@
 
-For the 2Dness, the effect is a bit less obvious (because the two first dimensions already have so much variance).
-Here is an example for the Columbimorphae.
-(to check different groups, you can easily modify the source code of the Rmd.
-TODO: this would be good in some shiny app innovation thingy).
+For the 2Dness, the effect is a bit less obvious (because the two first dimensions already have so much variance). Here is an example for the Columbimorphae:
 
@@
 
-We then calculate the projections/rejections as explained in more details in the [03-elaboration_exploration_analyses.Rmd]() script.
+We then calculate the projections/rejections as the elaboration/innovation scores/
 Really briefly we calculate three types of projections/rejections abbreviated as follows below:
@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
+```{r projections_whole_tree, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
 ## Run the analyses between groups
@@
 
-These sections are used for generated specific data for subsequent plots.
+These sections are used for generated median data for subsequent plots and figures.
+
+### Tip colours and informations
+
+```{r, echo = FALSE, eval = TRUE}
+## Getting the tip order for the ladderised super order tree
+tip_order <- c(
+    "Paleognathae" = 1,
+    "Galloanserae" = 2,
+    "Otidimorphae" = 5,
+    "Aequornithes" = 6,
+    "Mirandornithes" = 3,
+    "Columbimorphae" = 4,
+    "Telluraves" = 8,
+    "Strisores" = 7)
+
+## Plotting the tree and getting the tip orders
+tip_colours_super_orders <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
+save(tip_colours_super_orders, file = "../Data/Processed/tip_colours_super_orders.rda")
+```
+
+```{r, echo = FALSE, eval = TRUE}
+## Matching the order and the colours to the ladderised tree
+tip_order <- c(
+    "Tinamiformes"= 1,
+    "Galliformes"= 3,
+    "Anseriformes"= 2,
+    "Otidiformes"= 7,
+    "Cuculiformes" = 8,
+    "Gruiformes" = 9, 
+    "Musophagiformes" = 10,
+    "Procellariiformes" = 12,
+    "Sphenisciformes" = 11,
+    "Ciconiiformes" = 13,
+    "Pelecaniformes"= 15,
+    "Suliformes" = 14,
+    "Podicipediformes" = 4,
+    "Columbiformes" = 6,
+    "Pterocliformes" = 5,
+    "Charadriiformes" = 18,
+    "Trogoniformes" = 21,
+    "Bucerotiformes" = 22,
+    "Coraciiformes" = 23,
+    "Piciformes" = 24,
+    "Strigiformes" = 20,
+    "Accipitriformes" = 19, 
+    "Falconiformes" = 25,
+    "Psittaciformes" = 26,
+    "Passeriformes" = 27,
+    "Caprimulgiformes" = 16, 
+    "Apodiformes" = 17)
+
+## Plotting the tree and getting the tip orders
+tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
+```
+### Median values
+
+This section computes the median elaboration and innovation scores and the distances from the centre of the shapespace for each species at the three different levels (whole tree, super orders, orders).
 
@@
 orders <- lapply(results$order_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$order_results_phylo$position$matrix[[1]])
+
 ## Adding the colours
@@
 
-## Get the names
-names(clade_info$orders)
-
-## Get the colours
-unlist(lapply(clade_info$orders, `[[`, "col"))
-
-## Get the species in one specific group
-clade_info$orders$Cuculiformes$species
-
 ## Get the median elaboration/innovation for the groups
@@
 save(groups_median_EI, file = "../Data/Processed/groups_median_EI.rda")
-```
-
-```{r}
-## Plotting the novelty for the phylo picture:
-load("../Data/Processed/ploting_data.rda")
-load("../Data/Processed/ploting_data_superorder.rda")
-load("../Data/Processed/ploting_data_order.rda")
-
-## Get the values
-elaborations_tree  <- ploting_data$median_elaborations
-innovations_tree   <- ploting_data$median_innovations
-elaborations_super <- ploting_data_superorder$median_elaborations
-innovations_super  <- ploting_data_superorder$median_innovations
-elaborations_order <- ploting_data_order$median_elaborations
-innovations_order  <- ploting_data_order$median_innovations
-
-pdf("../Manuscript/Figures/median_novelty_distributions.pdf", height = 4, width = 8)
-par(mfrow = c(1,2), bty = "n")
-## Same but boxplot version
-boxplot(cbind("Whole tree level"  = c(elaborations_tree, innovations_tree),
-              "Super-order level" = c(elaborations_super, innovations_super),
-              "Order level"       = c(elaborations_order, innovations_order)),
-        xlab = "", ylab = "Elaboration/Innovation", col = colour.palette(5)[1:3])
-boxplot(cbind("Elaboration" = c(elaborations_tree, elaborations_super, elaborations_order),
-              "Innovation"  = c(innovations_tree, innovations_super, innovations_order)),
-        xlab = "", ylab = "Elaboration/Innovation", col = colour.palette(5)[4:5])
-dev.off()
 ```
@@
 ## Creating the subsets
-suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
+suborder_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 
-family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
+family_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 
-```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
+```{r projections_passeriformes, fig.height = 20, fig.width = 10, echo = FALSE, eval = TRUE, results = 'hide', message = FALSE, warning = FALSE}
 ## Run the analyses between groups
@@
 families <- lapply(results$family_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$family_results_phylo$position$matrix[[1]])
+
+## Getting the tip order for the ladderised super order tree
+tip_order <- c(
+    "Suboscines" = 1,
+    "Meliphagoidea" = 2,
+    "Corvides" = 3,
+    "Sylviida" = 4,
+    "Muscicapida" = 5,
+    "Passerida" = 6)
+
+## Plotting the tree and getting the tip orders
+tip_colours_passeriformes_suborder <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
+save(tip_colours_passeriformes_suborder, file = "../Data/Processed/tip_colours_passeriformes_suborder.rda")
+
+## Matching the order and the colours to the ladderised tree
+tip_order <- c(
+    "Aegithaloidea"= 12,
+    "Bombycilloidea"= 17,
+    "Cisticolidae"= 14,
+    "Corvoidea"= 7,
+    "Emberizoidea"= 23,
+    "Eurylaimides"= 1,
+    "Fringillidae"= 10,
+    "Furnariida"= 3,
+    "Hirundinidae"= 11,
+    "Locustelloidea"= 15,
+    "Malaconotoidea"= 5,
+    "Meliphagoidea"= 4,
+    "Motacillidae"= 22,
+    "Muscicapoidea"= 18,
+    "Nectariniidae"= 19,
+    "Orioloidea"= 6,
+    "Paridae"= 9,
+    "Passeridae"= 21,
+    "Petroicidae"= 8,
+    "PloceidaeEstrildidae"= 20,
+    "Pycnonotidae"= 13,
+    "Sylvioidea"= 16,
+    "Tyrannida" = 2)
+
+## Plotting the tree and getting the tip orders
+tip_colours_passeriformes_family <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+save(tip_colours_passeriformes_family, file = "../Data/Processed/tip_colours_passeriformes_family.rda")
+
 ## Adding the colours

### 2022-11-09 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
+superorder_shapespace <- MCMCglmm.subsets(n = 4000, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
+order_shapespace <- MCMCglmm.subsets(n = 4000, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 ## Creating the subsets
-suborder_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
+suborder_shapespace <- MCMCglmm.subsets(n = 4000, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,
@@
 
-family_shapespace <- MCMCglmm.subsets(n = 100, #TG: Toggle this to 100 for a faster calculation
+family_shapespace <- MCMCglmm.subsets(n = 4000, #TG: Toggle this to 100 for a faster calculation
       data          = shapespace,

### 2022-11-09 Analysis/04-Figure_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure_with_ellipse_scalling_and_8D_histograms.Rmd

--- 2022-07-15 Analysis/04-Figure3_with_ellipse_scalling_and_8D_histograms.Rmd
@@
 
-This section contains the code for reproducing the analyses. To get the details, refer to 03-elaboration_innovation_analyses.Rmd.
+This script reproduces the figure 2 in the main manuscript (the one with the ellipses) and takes around 5 minutes to compile.
+Note that the published figure is slightly modified in Inkscape (added headers and separation lines).
 
@@
 
-# All birds shapespace analyses
+# All birds clade level projections plot
 
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
+superorder_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
+order_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 ## Loading the results
-load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list_4k.rda")
+load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
 results <- shapespace_allbirds_lvl_superorder_order_results_list
@@
 ```{r, eval = TRUE, echo = FALSE}
+## Get the bar plots for the loadings of the phylogeny component (to be added separately in Inkscape to the main plot)
+# pdf(file = "../Manuscript/Figures/phylo_loadings.pdf", height = 4, width = 4)
+# plot(NULL, xlim = range(superorder_shapespace$matrix), ylim = range(superorder_shapespace$matrix))
+# add.dims(superorder_shapespace, name.col = c(phylogeny = "grey"), n = 1000)
+# dev.off()
+
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_ellipses.pdf", height = box_size*10, width = box_size*10)
+pdf(file = "../Manuscript/Figures/ellipses.pdf", height = box_size*10, width = box_size*10)
 
@@
         matrix(c(1,1, rep(73, 3), 0, 6:9), ncol = 10, nrow = 1), # line 8
-        matrix(c(78,78, 10:17), ncol = 10, nrow = 1) # line 8
-    )
+        matrix(c(78,78, 10:17), ncol = 10, nrow = 1)) # line 8)
 lay_mat_order <- rbind(
@@
         matrix(c(38:47), ncol = 10, nrow = 1),
-        matrix(c(48:55, 58:59), ncol = 10, nrow = 1)
-    )
-
+        matrix(c(48:55, 58:59), ncol = 10, nrow = 1))
 heights_vals <- rep(1, 9)
@@
 ## PLOTTING THE ORDERS
-
 par(bty = "n")
@@
            add = TRUE, n = 100)
-
 dev.off()
-
 ```
 
-# Passeriformes shapespace analyses
+# Passeriformes clade level projections plot
 
@@
 ## Creating the subsets
-suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
+suborder_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 
-family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
+family_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 ```{r}
+## Adding the bar plot for the phylogeny component (to be added separately in Inkscape)
+# pdf(file = "../Manuscript/Figures/phylo_loadings.pdf", height = 4, width = 4)
+# plot(NULL, xlim = range(suborder_shapespace$matrix), ylim = range(suborder_shapespace$matrix))
+# add.dims(suborder_shapespace, name.col = c(phylogeny = "grey"), n = 1000)
+# dev.off()
+
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/Figure3_passeriformes.pdf", height = box_size*9, width = box_size*10)
+pdf(file = "../Manuscript/Figures/ellipses_passeriformes.pdf", height = box_size*9, width = box_size*10)
 
@@
         matrix(c(1,1, rep(x, 3), 0, 6:9), ncol = 10, nrow = 1), # line 8
-        matrix(c(x+2,x+2, 10:13, rep(x+3, 4)), ncol = 10, nrow = 1) # line 8
-    )
+        matrix(c(x+2,x+2, 10:13, rep(x+3, 4)), ncol = 10, nrow = 1)) # line 8)
 y <- x - 13
@@
         matrix(c(28:37), ncol = 10, nrow = 1), 
-        matrix(c(38:47), ncol = 10, nrow = 1)
-    )
+        matrix(c(38:47), ncol = 10, nrow = 1))
 heights_vals <- rep(1, 8)
@@
 ## PLOTTING THE SUBORDERS
-
 ## Getting the tip order for the ladderised super order tree
@@
 ## PLOTTING THE FAMILIES
-
 ## Matching the order and the colours to the ladderised tree
@@
 dev.off()
-
 ```

### 2022-11-09 Analysis/01-Data_preparation.Rmd
Update 01-Data_preparation.Rmd

@@
 
-<!-- Compilation time: 30 minutes -->
-
-All the custom code and custom functions called in this document can be downloaded and installed in `R`.
+The following script takes around 5-10 minutes to compile. All the custom code and custom functions called in this document can be downloaded and installed in `R`.
 
@@
 ## Optional - this snippet installs all the packages required to run the analyses
-install.packages(c("devtools", "dispRity", "MCMCglmm", "xtable", "knitr", "ggplot2"))
+install.packages(c("devtools", "dispRity", "MCMCglmm", "xtable", "knitr", "ggplot2", "patchwork", "tidyverse"))
 devtools::install_github("TGuillerme/mcmcmcglmmm")

### 2022-11-09 Analysis/05-Figure_with_phylogeny.Rmd
Update 05-Figure_with_phylogeny.Rmd

@@
 
-## Plot the median elaboration/innovation on the phylogeny
+This script reproduces the figure 1 in the main manuscript (the phylogeny) and takes around 15 minutes to compile.
+Note that the published figure is slightly modified in Inkscape (boxplot added manually and changed the label position for passeriformes).
+
+# Plot the median elaboration/innovation on the phylogeny
 
@@
 
-clades <- read.csv("../Data/Processed/BirdClades.csv")
+clades <- read.csv("../Data/Raw/BirdClades.csv")
 passerines <- clades$TipLabel[which(clades$level2_order=="Passeriformes")]
@@
 ggsave("../Manuscript/Figures/InnovElabTree_passerine_supplement.pdf", innovelab_pass_plot, width=12, height=12)
+```
 
+## Adding a panel showing the differences in elaboration and innovation across the tree
+
+```{r}
+source("../Functions/ploting.utilities.R")
+colour.palette <- gg.color.hue
+## Plotting the novelty for the phylo plot:
+load("../Data/Processed/ploting_data.rda")
+load("../Data/Processed/ploting_data_superorder.rda")
+load("../Data/Processed/ploting_data_order.rda")
+
+## Get the values
+elaborations_tree  <- ploting_data$median_elaborations
+innovations_tree   <- ploting_data$median_innovations
+elaborations_super <- ploting_data_superorder$median_elaborations
+innovations_super  <- ploting_data_superorder$median_innovations
+elaborations_order <- ploting_data_order$median_elaborations
+innovations_order  <- ploting_data_order$median_innovations
+
+pdf("../Manuscript/Figures/median_novelty_distributions.pdf", height = 4, width = 8)
+par(mfrow = c(1,2), bty = "n")
+## Same but boxplot version
+boxplot(cbind("Whole tree level"  = c(elaborations_tree, innovations_tree),
+              "Super-order level" = c(elaborations_super, innovations_super),
+              "Order level"       = c(elaborations_order, innovations_order)),
+        xlab = "", ylab = "Elaboration/Innovation", col = colour.palette(5)[1:3])
+boxplot(cbind("Elaboration" = c(elaborations_tree, elaborations_super, elaborations_order),
+              "Innovation"  = c(innovations_tree, innovations_super, innovations_order)),
+        xlab = "", ylab = "Elaboration/Innovation", col = colour.palette(5)[4:5])
+dev.off()
 ```

### 2022-11-09 Analysis/06-Figure_with_correlations.Rmd
Update 06-Figure_with_correlations.Rmd

--- 2022-08-04 Analysis/06-Figure2_with_correlations.Rmd
@@
 
-This markdown contains the code to reproduce the figures containing the elaboration and innovation per species for the all birds and the passeriformes analyses.
-For the details on how to get the data objects called here, refer to `03-elaboration_innovation_analyses.Rmd`.
+This script reproduces a figure not present or used in the manuscript or the supplementary materials.
 
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+superorder_shapespace <- MCMCglmm.subsets(, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+order_shapespace <- MCMCglmm.subsets(, # Just run it on 1k matrices for now
       data          = shapespace,

### 2022-11-09 Analysis/07-Orthogonality_and_statistics.Rmd
Update 07-Orthogonality_and_statistics.Rmd

@@
 title: "Bird beak's elaboration and innovation"
-author: "Thomas Guillerme"
+author: "Thomas Guillerme and Gavin Thomas"
 date: "`r Sys.Date()`"
@@
 This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
-This script take around 6 minutes to compile.
+This script take around 35 minutes to compile.
 
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 ## Load the data for all birds
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 ## Creating the nested levels subgroups
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 source("../Functions/stats.utilities.R")
@@
 ```{r, echo = FALSE, eval = TRUE}
-## Quick load
 source("../Functions/stats.utilities.R")
 library(dispRity)
+library(ggplot2)
 load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
@@
 innovation_species  <- get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]]
+elaboration_species <- elaboration_species/max(elaboration_species)
+innovation_species  <- innovation_species/max(innovation_species)
 
@@
 innovation_orders        <- lapply(get.disparity(results$order_results_between$distance, concatenate = FALSE), scale.group)
+
 ## The objects above are the ones that contain all the elaboration and innovation values by clade level
-
 ## These bits are for the Bhatt coeff
@@
 
-pdf("../Manuscript/Figures/relative_EI.pdf", height = 6, width = 12)
-par(mfrow = c(1,2))
-plot.density(species = c(elaboration_species),
-             group   = c(unlist(elaboration_super_orders), unlist(elaboration_orders)),
-             xlab    = "relative elaboration", scale.x = TRUE,
-             ylab    = "density",
-             main    = "A - Total posterior elaboration per clade\n(scaled) and per species")
-legend("topright", lty = 1, legend = c("species", "clades"), col = c("blue", "orange"))
-text(x = 0.7, y = 0.5, paste0("Posterior area under the curve overlap\n(Bhattacharyya coefficient): ", round(bhatt.coeff(elaboration_species_auc, elaboration_group_auc), 3)))
-plot.density(species = c(innovation_species),
-             group   = c(unlist(innovation_super_orders), unlist(innovation_orders)),
-             xlab    = "relative innovation", scale.x = TRUE,
-             ylab    = "density",
-             main    = "B - Total posterior innovation per clade\n(scaled) and per species")
-text(x = 0.7, y = 0.5, paste0("Posterior area under the curve overlap\n(Bhattacharyya coefficient): ", round(bhatt.coeff(innovation_species_auc, innovation_group_auc), 3)))
-dev.off()
+library(tidyverse)
+library(patchwork)
+
+fake <- data.frame(elaboration = c(elaboration_species, c(unlist(elaboration_super_orders), unlist(elaboration_orders))), innovation = c(innovation_species, c(unlist(innovation_super_orders), unlist(innovation_orders))), Level = c(rep("Species", 34992000), rep("Clades", 140000)))
+
+elab <- 
+  ggplot(fake, aes(x = elaboration, colour=Level, fill = Level, ..scaled..)) +
+  geom_density(alpha = 0.5) +
+  theme_classic(base_size = 14) +
+  scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
+  scale_fill_manual(values = c("goldenrod", "#1E65A7")) +
+  xlab("Relative elaboration") +
+  ylab("Density") +
+  theme(legend.position = "none") +
+  labs(tags="a")
+
+innov <-
+  ggplot(fake, aes(x = innovation, colour = Level, fill = Level, ..scaled..)) +
+  geom_density(alpha = 0.5) +
+  theme_classic(base_size = 14) +
+  scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
+  scale_fill_manual(values = c("goldenrod", "#1E65A7")) +
+  xlab("Relative innovation") + 
+  theme(legend.position = c(0.8, 0.8)) +
+  labs(tags="b")
+
+remove_y <- 
+  theme(axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
+
+p <- elab + innov + remove_y
+
+ggsave("relative_EI_gg_style.pdf", plot=p, height=6, width=12)
 ```

### 2022-11-09 Analysis/08-Supplementary_figures_tables.Rmd
Update 08-Supplementary_figures_tables.Rmd

--- 2022-03-10 Analysis/06-Supplementary_figures_tables.Rmd
@@
 ## PC axis selection
+
+This parts looks at the number of dimensions necessary to describe at least 95% of the variance in each dimension of the trait space.
 
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+superorder_shapespace <- MCMCglmm.subsets(
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
+order_shapespace <- MCMCglmm.subsets(
       data          = shapespace,

### 2022-11-09 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 $^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$Department of Biological and Marine Sciences, University of Hull, Hull, HU6 7RX, United Kingdom.\\
+$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
 $^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
@@
 
-\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
+\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, major axis of phenotypic variation, GLMM)\\
 
@@
 Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
-A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.https://www.overleaf.com/project/62f128d3d7c36e4efd1e9b21
+A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
 However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
@@
 However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
-Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a confict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
+Resolving this confict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
 
@@
 We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species-specific measures of innovation and elaboration.
-If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall of the line (innovation).
+If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall off the line (innovation).
 To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
@@
 
-Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}.
+Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level (Fig. \ref{Fig:phylogeny}).
 We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
@@
 
-To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.
+To investigate the conflict implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.
 To do this we translated the major axes of phenotypic variation of each clade's (i.e.
@@
 Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
+Each ellipses represents a two dimensional projection of the average traits variance co-variance in the global bird beak shapespace.
+The ellipses are centred on the centroid of each clade they represent.
+Their direction represents the major direction of the traits covariance in two dimensions.
 We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
@@
 The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
-These scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation for each plot. %TG: TODO: add figure with unscalled ellipses in the supplementaries
+These scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation for each plot.
 Panel B) shows the results for each order.
@@
 \textbf{Reproducibility and repeatability.}
-The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
+The raw and processed data used in this analysis is available from \href{https://figshare.shef.ac.uk/account/articles/21526146}{figshare} \cite{fighsaredata}.
 The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{GitHub} \cite{githubrepo}.

### 2022-11-10 Manuscript/main_flipped_narrative.tex
Updated MS

@@
 % paragraph 1:  introducing the major axis of variation + novelty and innovation
-Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
+Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive,Rhoda2022}.
 Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
-Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
-%TG: cite some of these numerous studies.
+Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
 Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
 However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a confict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a confict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
 Resolving this confict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
@@
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general}.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales}.
 Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
@@
 We pruned these trees to contain only the 8748 species present in our beak shapespace.
+Phylogenetic related data was used in R using the \texttt{ape} package \cite{ape}.
 
@@
 
+\textbf{Author contributions.}
+A.P.B., N.C., T.G. and G.H.T designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. handled the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
+
 \bibliographystyle{naturemag}
 \bibliography{references}
-%TODO: add ref to ape.
 

### 2022-11-11 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 
-\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, major axis of phenotypic variation, GLMM)\\
+\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
 
@@
 The extent to which species and clades elaborate or innovate is poorly understood.
-Of particular importance is the potential contribution of reorientations of the phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
-Here we apply new multi-trait methods that model the reorientations of the trait space via the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
+Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
 Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
@@
 This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+%TG: bleuargh...
 
@@
 However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a confict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
-Resolving this confict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
+Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
 
@@
 
-To investigate the conflict implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.
+To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.
 To do this we translated the major axes of phenotypic variation of each clade's (i.e.
@@
 Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
-Each ellipses represents a two dimensional projection of the average traits variance co-variance in the global bird beak shapespace.
-The ellipses are centred on the centroid of each clade they represent.
-Their direction represents the major direction of the traits covariance in two dimensions.
 We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
@@
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales}.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales}\cite{pagel2022general}.
 Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
@@
 Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain - a 400 fold improvement.
+The total analysis took around three CPU years using the shARC cluster from the University of Sheffield.
 Code to reproduce the procedure is available in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
@@
 
+\textbf{Acknowledgements.}
+We acknowledge IT Services at The University of Sheffield for the provision of the High Performance Computing Service.
+
 \textbf{Funding.}
@@
 \textbf{Author contributions.}
-A.P.B., N.C., T.G. and G.H.T designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. handled the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
+G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
 

### 2022-11-11 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 The extent to which species and clades elaborate or innovate is poorly understood.
-Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
+Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to mega-evolutionary scale.
 Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
-Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
+Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with microevolutionary theory.
 However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
@@
 This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
-%TG: bleuargh...
 
-% TODO in abstract, add the "paradox" bit: However, if most organisms evolve incrementally along stable lines of least resistance then how can we explain the staggering diversity of species morphologies across the tree of life?
+%TG: cite Anjali somewhere
 
@@
 Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
-Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
+Numerous studies show that phenotypic divergence is biased along the evolutionary line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
 Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
-However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
-Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
+However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a conflict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
+Resolving this conflict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
 
@@
 Relative to elaboration, innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
-These core concepts can now be defined using robust multivariate statistics about trait covariance applied at multpiple phylogentic scales and provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
+These core concepts can now be defined using robust multivariate statistics about trait covariance applied at multiple phylogenetic scales and provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
 i) idiosyncratic species-level innovation against clade-level innovation (i.e. reorientation of the directions of phenotypic variation),
@@
 To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric, morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-To meet the second two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
+To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
 % start: oldversion
@@
 % start: newversion
-In order to estimate the major axis of phenotypic variation we fit pGLMM models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
+To estimate the major axis of phenotypic variation we fit Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
 From the posterior distribution of the variance-covariance matrix at any scale, the major axis can be identified from the leading eigenvector of the variance-covariance matrix.
@@
 % paragraph 4: results for elaboration and innovation for species
-We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species-specific measures of innovation and elaboration.
-If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall off the line (innovation).
+We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along evolutionary lines of least resistance by calculating species-specific measures of innovation and elaboration.
+If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than off the line (innovation).
 To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
-We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globaly more elaboration and innovation at the global phylogeny and the super-orders levels compared to the orders level.
+We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globally more elaboration and innovation at the global phylogenetic and super-order levels compared to the order level.
 Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig.
@@
  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparison.
+However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparisons.
 Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
-The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
+The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme, and long, shallow, and narrow beaks at the other.
 Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.
@@
 In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
-Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
+Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundinidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
 The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
 The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
-This leads us to ask: if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?
+This leads us to ask; if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?
 
@@
 \noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
-Blue and yellow colors on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
-Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order-level, then at the super-order level, and finally at the global phylogenetic level.
-The boxplots at the bottom of the figure summarises the distribution of the median innovation and elaboration values for all levels (bottom left) showing that elaboration is more common than elaboration;
-and the pooled median innovation and elaboration values for all three different levels (bottom right) showing that elaboration and innovation is more common at higher taxonomic levels than at the orders level.
+Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
+Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order level, then at the super-order level, and finally at the global phylogenetic level.
+The boxplots at the bottom of the figure summarise the distribution of the median innovation and elaboration values for all levels (bottom left), showing that elaboration is more common than elaboration,
+and the pooled median innovation and elaboration values for all three different levels (bottom right), showing that elaboration and innovation is more common at higher taxonomic levels than at the order level.
 We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
@@
 
-To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.
+To investigate the conflict implied by our observations of species-level elaboration and previously reported deep-time jumps in phenotype, we measured the elaboration and innovation of different clades.
 To do this we translated the major axes of phenotypic variation of each clade's (i.e.
@@
 \ref{Fig:orthogonality}).
-Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep time (i.e evidence at the clade scale).
+Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to microevolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep-time (i.e evidence at the clade scale).
 However, there is substantial variation in elaboration and innovation among clades including relatively low elaboration and innovation (e.g.
@@
 Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent's parent clade.
-Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+Orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
 Dots represent the median orthogonality of each clade and the lines their 95\% CI.
 No super-order's major axis of phenotypic variation is parallel to the global phylogenetic major axis (red lines and circles) and 12/25 orders are at least half orthogonal to the major axis of their super-order (green lines and circles).
-These results are even clearer when comparing them to the global phylogenetic major axis (green dashed lines and open circles) where only 3/25 are less than half orthogonal.
-Furthermore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+These results are even clearer when compared to the global phylogenetic major axis (green dashed lines and open circles) where only 3/25 are less than half orthogonal.
+We also indicate the number of species (n) and the standard deviation (sd) of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
 For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the global phylogenetic major axis relative to their sample size and sd.
@@
 Evolutionary innovation can arise in any direction in trait space.
-Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig.
+Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned with PC1; Fig.
 \ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly with PC1 (Fig.
 \ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
-This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
+This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
 S\ref{fig_ellipses_passeriformes}).
@@
 in Podicipediformes) from the global phylogenetic major axis of phenotypic variation.
-For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogeny but uniformity in their beak shape components within the clade.
+For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogenetic level but uniformity in their beak shape within the clade.
 In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
@@
 So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
-However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species \cite{hansen1997stabilizing}.
-This idea is similar in concept to models of multiple adaptive peak models where species evolve in a heterogenous adaptive landscape and share similar responses to selection pressures within adaptive zones.
+However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species.
+This idea is similar in concept to multiple adaptive peak models where species evolve in a heterogeneous adaptive landscape and share similar responses to selection pressures within adaptive zones \cite{hansen1997stabilizing}.
 %GT: Tip-toeing into OU territory and needs citations but I think this is a fair suggestion. %TG: fair point!
@@
 b) the same as a) but using innovation score. 
-For innovation the overlap BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.
+For innovation the overlap has BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.
 
@@
 
-Our results show that although at the species-level most bird beaks are elaborating along a mega-evolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
+Our results show that although at the species-level most bird beaks are elaborating along a megaevolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
 This nested structure of elaboration at a lower (species) taxonomic level and innovation at a higher (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
@@
 We pruned these trees to contain only the 8748 species present in our beak shapespace.
-Phylogenetic related data was used in R using the \texttt{ape} package \cite{ape}.
 
 \textbf{Phylogenetic GLMM analysis.}
-We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
+We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} and \texttt{ape} R packages \cite{MCMCglmm, ape} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
 
@@
 
-This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
+This model allows us to estimate a phenotypic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
 These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).

### 2022-11-14 Manuscript/main_flipped_narrative.tex
Updated and checked MS + supp

@@
 
-% New narrative:
-% 1 - Novelty arises through elaboration and innovation
-% 2 - Theory suggests that it is easier to evolve along the major axis of variation (formely LOTR in this manuscript)
-% 3 - However this has never been tested in big datatests. We do that.
-% 4 - We measure elaboration and innovation in all birds
-% 5 - We find that there is more elaboration than innovation
-% 6 - But that asks another (more interesting) question: why do we have diversity if everything elaborates?
-% 7 - We then measure novelty by clades rather than by species
-% 8 - We find that clades innovate.
-% 9 - So basically species elaborate (in line with the micro evol theory of evolving along the line for major axis of variation) and clades innovate (in line with the observations at a macro and mega level of evolutionary jumps leading to big diversity at high levels).
-
 \noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
@@
 However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
-At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
+At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from that major axis is termed innovation \cite{endler2005animal}. 
 The extent to which species and clades elaborate or innovate is poorly understood.
@@
 Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with microevolutionary theory.
-However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
-These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macroevolutionary scales (i.e.
-species diverging within clades) and innovation is more prevalent at the megaevolutionary scale (nested patterns of clades diverging within clades).
-This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
-
-%TG: cite Anjali somewhere
+However, innovation is surprisingly common at the clade level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
+Our results highlight the importance of scale in our understanding of deep time evolution.
+While we find that elaboration dominates at macroevolutionary scales (i.e.
+species diverging within clades), the major axis of phenotypic change at  megaevolutionary scales is an emergent property of innovation across clades.
+Taken together we suggest that innovation and elaboration are ubiquitous routes for phenotypic change but that their contribution to diversity is dependent on scale.
+The macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
 
@@
 
-% paragraph 1:  introducing the major axis of variation + novelty and innovation
 Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive,Rhoda2022}.
@@
 Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
-However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a conflict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
+However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{BurinWhales,jones2004evolution}.
+The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,pagel2022general,khabbazian2016fast,cooney2017mega,smaers2021evolution,goswami2022} suggests a conflict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
 Resolving this conflict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.
 
-% paragraph 2:  introducing the Endler view.
 Endler \textit{et al}. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
@@
 
-% paragraph 3: how we measure this (glmm) on the birds data + nestedness
 \subsection{Modelling nested trait covariances}
@@
 To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-% start: oldversion
-% To provide a baseline from which to calculate elaboration and innovation (i.e.
-the major axis of phenotypic variation) we use phylogenetic generalised linear mixed models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
-% Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
-% From these models we can calculate and compare major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix of the trait space.
-% end: oldversion
-% start: newversion
 To estimate the major axis of phenotypic variation we fit Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
@@
 We can estimate levels of elaboration and innovation based on that major axis of phenotypic variation (see Methods).
-% end: newversion
 We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
@@
 \subsection{Elaboration dominates species-level divergence}
-% paragraph 4: results for elaboration and innovation for species
 We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along evolutionary lines of least resistance by calculating species-specific measures of innovation and elaboration.
 If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than off the line (innovation).
-To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
+To assess this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
 We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globally more elaboration and innovation at the global phylogenetic and super-order levels compared to the order level.
@@
 Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level (Fig. \ref{Fig:phylogeny}).
-We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
+We find no consistent evidence that innovation is either positively correlated, or trades-off, with elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
 However, we observe various combinations of elaboration and innovation across birds.
@@
 In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
-Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundinidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
-The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
+Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundinidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length and display high levels of both elaboration and innovation.
+The dominance of elaboration when measured at the species-level is consistent with the idea of evolution along lines of least evolutionary resistance.
 The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
@@
 
-\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
+\noindent \textbf{Figure \ref{Fig:phylogeny}:}
+The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
 Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
@@
 \noindent \textbf{Figure \ref{Fig:orthogonality}:}
+\textbf{Clade's orthogonality.}
 Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent's parent clade.
@@
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
-%TG: TODO: very easy low hanging fruit (maybe?), but not for this paper is to plot orthogonality through time.
 
@@
 
-% Paragraph 7: putting these results together
-
 The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
@@
 This idea is similar in concept to multiple adaptive peak models where species evolve in a heterogeneous adaptive landscape and share similar responses to selection pressures within adaptive zones \cite{hansen1997stabilizing}.
-%GT: Tip-toeing into OU territory and needs citations but I think this is a fair suggestion. %TG: fair point!
-This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple}, may be sufficient to explain diversity in avian beak morphology.
+This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{venditti2011multiple,cooney2017mega}, may be sufficient to explain diversity in avian beak morphology.
 
@@
 However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales}\cite{pagel2022general}.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales,goswami2022}.
 Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
@@
 \textbf{Beak shape data.}
-We used the dataset from \cite{cooney2017mega,chira2020signature,hughes2022global} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
-The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,chira2020signature,hughes2022global} for detailed descriptions of data collection and processing).
+We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
+The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
 We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
@@
 For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
-This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig.
-S\ref{Fig:axes_variance} and Table \ref{tab_variance_per_axis}).
+This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. S\ref{Fig:axes_variance}).
 This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
@@
 
-\textbf{Group orthogonality.}
+\textbf{Clade orthogonality.}
 We measured the orthogonality of each random term (i.e. clade) in our models compared to their parent clade and their parent's parent clade.

### 2022-11-14 Manuscript/supplementaries.tex
Updated and checked MS + supp

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
+\noindent {\normalsize \sc
+Thomas Guillerme$^{1,*}$, 
+Jen A. Bright$^{2}$,
+Christopher R. Cooney${^1}$,
+Emma C. Hughes${^1}$,
+Zoe K. Varley${^3}$,
+Natalie Cooper$^{3,+}$,
+Andrew P. Beckerman$^{1,+}$,
+and Gavin H. Thomas$^{1,4,+}$}\\
 \noindent {\small \it 
 $^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
+$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
+$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
+$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
 
 \end{center}
-\medskip
-\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\ 
+\bigskip
+\noindent{*\bf Corresponding author:} \textit{guillert@tcd.ie}; \noindent{+\bf contributed equally.}\\
 \vspace{1in}
@@
 
-\noindent (Keywords: )\\
+\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
 
@@
 %---------------------------------------------
+
+\renewcommand{\thefigure}{S\arabic{figure}} 
+ \renewcommand{\thetable}{S\arabic{table}}
 
@@
 Species position on this axis elaborate and innovate at the macro-evolutionary level.
-On the top right panel (Figure 1 in the main text), we measured the elaboration and innovation at the entire clade level (cf. at the species level below).
+On the top right panel (Figure 1 in the main text), we measured the position of each species in terms of elaboration and innovation.
+A species elaboration score is their projection onto the phylogenetics' VCV major axis (i.e.
+their position on that axis) whereas their innovation score is their rejection from it (i.e.
+their distance away from that axis).
+On the bottom right panel (Figure 2 in the main text), we measured the elaboration and innovation at the entire clade level (cf. at the species level below).
 The ellipses are a scaled and centered in space representation of the group’s VCV ellipse relative to the overall phylogenetic one with a the relative length of the ellipse on each dimension represented on the barplots.
 The clade's innovation and elaboration scores are measured by projecting the clade's major axis onto the phylogenetic one and measuring the projection (elaboration) and rejection (innovation) as mentioned above.
-The distribution of innovation and elaboration scores comes from the projection of the 4000 pairs of posterior VCVs.
-On the bottom right panel (Figure 3 in the main text), we measured the position of each species in terms of elaboration and innovation.
-A species elaboration score is their projection onto the phylogenetics' VCV major axis (i.e.
-their position on that axis) whereas their innovation score is their rejection from it (i.e.
-their distance away from that axis).}
+The distribution of innovation and elaboration scores comes from the projection of the 4000 pairs of posterior VCVs.}
+
 \label{Fig:cheat_sheet}
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[htbp]
 \centering
@@
 
-\begin{landscape}
-\input{Tables/tab_variance_per_axis.tex}
-\end{landscape}
-
-\input{Tables/tab_ortho_results.tex}
-
+\clearpage
 \section{Supplementary material 3: Passeriformes results}
@@
 \centering
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement.pdf}
+\caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 1 in the main text for details.}
+\label{fig_phylogeny_passeriformes}
+\end{figure}
+
+\begin{figure}[!htbp]
+\centering
    \includegraphics[width=1\textwidth]{Figures/ellipses_passeriformes.pdf}
-\caption{Elaboration and innovation at the group level for each passeriformes sub-orders and families. See the caption of figure 1 in the main text for details.}
+\caption{Elaboration and innovation at the group level for each passeriformes sub-orders and families. See the caption of figure 2 in the main text for details.}
 \label{fig_ellipses_passeriformes}
@@
    \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results_passeriformes.pdf}
-\caption{Posterior results showing the orientation of phenotypic evolution (orthogonality) for each passeriformes sub-orders and families. See the caption of figure 2 in the main text for details.}
+\caption{Posterior results showing the orientation of phenotypic evolution (orthogonality) for each passeriformes sub-orders and families. See the caption of figure 3 in the main text for details.}
 \label{fig_orthogonality_passeriformes}
-\end{figure}
-
-\begin{figure}[!htbp]
-\centering
-    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement.pdf}
-\caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 3 in the main text for details.}
-\label{fig_phylogeny_passeriformes}
 \end{figure}

### 2022-11-14 Manuscript/supplementaries.tex
Updates

@@
 variance-covariance matrix & the multidimensional relation between traits & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) & & \\
-nested phylogenetic level  & the different \textbf{clades} used in this study (the whole phylogeny > super-orders > orders) & different nested \textbf{variance-covariance matrices} & & \\
+nested phylogenetic level  & the different \textbf{clades} used in this study (the whole phylogeny $>$ super-orders $>$ orders) & different nested \textbf{variance-covariance matrices} & & \\
 

### 2022-11-21 Manuscript/main_flipped_narrative.tex
Updates
Update main_flipped_narrative.tex

@@
 \usepackage{xr} %linking to supplementaries
+\usepackage{setspace} % double spacing
 \externaldocument{supplementaries}
@@
 \bigskip
+Short title: Innovation and elaboration in birds
+One sentence summary: Species elaborate; clades innovate, reconciling evolutionary theory from micro- to megaevolutionary scales.
 
@@
 
+\doublespacing
+
 %---------------------------------------------
@@
 
-Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
-A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
-However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
-At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from that major axis is termed innovation \cite{endler2005animal}. 
-The extent to which species and clades elaborate or innovate is poorly understood.
-Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to mega-evolutionary scale.
-Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
-Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with microevolutionary theory.
-However, innovation is surprisingly common at the clade level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
-Our results highlight the importance of scale in our understanding of deep time evolution.
-While we find that elaboration dominates at macroevolutionary scales (i.e.
-species diverging within clades), the major axis of phenotypic change at  megaevolutionary scales is an emergent property of innovation across clades.
-Taken together we suggest that innovation and elaboration are ubiquitous routes for phenotypic change but that their contribution to diversity is dependent on scale.
-The macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+Reconciling major changes in phenotypic diversity at macroevolutionary scales with gradual microevolutionary changes continues to perplex scientists.
+A major question is whether widely documented megaevolutionary jumps can emerge from gradual phenotypic elaboration and innovation.
+Here we apply new multi-trait models to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+We find that elaboration dominates at macroevolutionary scales whereas innovation becomes more prominent at megaevolutionary scales.
+The major axis of phenotypic change among species at megaevolutionary scales is an emergent property of innovation across clades.
+We suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
 
-\section{Main}
+% change Gustavo's paper to /in press in current biology/
+% Check if the ruminant paper is somewhere
+
+\section{Introduction}
+
+% TG: add these citations down there
+% However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
+% At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from that major axis is termed innovation \cite{endler2005animal}.
 
@@
 \centering
-    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text_with_box.pdf}
 \caption{.}
@@
 
-\textbf{Reproducibility and repeatability.}
+\bibliographystyle{naturemag}
+\bibliography{references}
+
+\textbf{Acknowledgements.}
+We acknowledge IT Services at The University of Sheffield for the provision of the High Performance Computing Service.
+We thank the museum in Tring.
+% TG: TODO: just take Chris' paper acknowledgments
+\textbf{Funding.}
+This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
+% TG: TODO: add ERC grant (ask Gavin)
+\textbf{Author contributions.}
+G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
+\textbf{Competing interests.}
+We declare no competing interests.
+\textbf{Data an materials availability.}
 The raw and processed data used in this analysis is available from \href{https://figshare.shef.ac.uk/account/articles/21526146}{figshare} \cite{fighsaredata}.
@@
 
-\textbf{Acknowledgements.}
-We acknowledge IT Services at The University of Sheffield for the provision of the High Performance Computing Service.
-
-\textbf{Funding.}
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
-
-\textbf{Author contributions.}
-G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
-
-\bibliographystyle{naturemag}
-\bibliography{references}
+% TODO: Add list of supplementary materials
+% All this comes after the references
+% Figures and captions after this
 

### 2022-11-22 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

@@
 
+\usepackage{scicite}
 \usepackage{epsfig}
@@
 \bigskip
-\bigskip
 \begin{center}
@@
 \bigskip
-Short title: Innovation and elaboration in birds
-One sentence summary: Species elaborate; clades innovate, reconciling evolutionary theory from micro- to megaevolutionary scales.
+
+% Short title: Innovation and elaboration in birds
+
+% One sentence summary: Species elaborate; clades innovate, reconciling evolutionary theory from micro- to megaevolutionary scales.
 
@@
 Emma C. Hughes${^1}$,
-Zoe K. Varley${^3}$,
+% Zoe K. Varley${^3}$,
 Natalie Cooper$^{3,+}$,
@@
 \noindent{*\bf Corresponding author:} \textit{guillert@tcd.ie}; \noindent{+\bf contributed equally.}\\
-\vspace{1in}
+% \vspace{1in}
 
@@
 
-\section{Abstract}
+\section*{Abstract}
 
-Reconciling major changes in phenotypic diversity at macroevolutionary scales with gradual microevolutionary changes continues to perplex scientists.
-A major question is whether widely documented megaevolutionary jumps can emerge from gradual phenotypic elaboration and innovation.
-Here we apply new multi-trait models to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+Widely documented, megaevolutionary jumps in diversity continue to perplex researchers because it remains unclear whether these dramatic changes can emerge from microevolutionary processes.
+Here we reconcile major changes in phenotypic diversity at macroevolutionary scales with gradual microevolutionary change using new multi-trait models to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
 We find that elaboration dominates at macroevolutionary scales whereas innovation becomes more prominent at megaevolutionary scales.
-The major axis of phenotypic change among species at megaevolutionary scales is an emergent property of innovation across clades.
-We suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
+Indeed, the major axis of phenotypic change among species at megaevolutionary scales is an emergent property of innovation across clades.
+Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
 
-% change Gustavo's paper to /in press in current biology/
-% Check if the ruminant paper is somewhere
+\section*{Introduction}
 
-\section{Introduction}
-
-% TG: add these citations down there
-% However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
-% At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from that major axis is termed innovation \cite{endler2005animal}.
-
-Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive,Rhoda2022}.
-Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
+Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
+Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,renaud2006conserved,fasanelli2022allometry}.
 Numerous studies show that phenotypic divergence is biased along the evolutionary line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
 Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
-However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{BurinWhales,jones2004evolution}.
+However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution,BurinWhales}.
 The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,pagel2022general,khabbazian2016fast,cooney2017mega,smaers2021evolution,goswami2022} suggests a conflict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
@@
 
-\subsection{Modelling nested trait covariances}
+\section*{Modelling nested trait covariances}
 Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
@@
 To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-To estimate the major axis of phenotypic variation we fit Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
+To estimate the major axis of phenotypic variation we fit Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable \cite{MCMCglmm}.
 From the posterior distribution of the variance-covariance matrix at any scale, the major axis can be identified from the leading eigenvector of the variance-covariance matrix.
@@
 
-\subsection{Elaboration dominates species-level divergence}
+\section*{Elaboration dominates species-level divergence}
 We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along evolutionary lines of least resistance by calculating species-specific measures of innovation and elaboration.
@@
 
-\begin{figure}[!htbp]
-\centering
-    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text_with_box.pdf}
-\caption{.}
-\label{Fig:phylogeny}
-\end{figure}
+% \begin{figure}[!htbp]
+% \centering
+%     \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text_with_box.pdf}
+% \caption{.}
+% \label{Fig:phylogeny}
+% \end{figure}
 
-\bigskip
+% \bigskip
 
-\noindent \textbf{Figure \ref{Fig:phylogeny}:}
-The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
-Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
-Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order level, then at the super-order level, and finally at the global phylogenetic level.
-The boxplots at the bottom of the figure summarise the distribution of the median innovation and elaboration values for all levels (bottom left), showing that elaboration is more common than elaboration,
-and the pooled median innovation and elaboration values for all three different levels (bottom right), showing that elaboration and innovation is more common at higher taxonomic levels than at the order level.
-We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
-A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
+% \noindent \textbf{Figure \ref{Fig:phylogeny}:}
+% The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
+% Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
+% Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order level, then at the super-order level, and finally at the global phylogenetic level.
+% The boxplots at the bottom of the figure summarise the distribution of the median innovation and elaboration values for all levels (bottom left), showing that elaboration is more common than elaboration,
+% and the pooled median innovation and elaboration values for all three different levels (bottom right), showing that elaboration and innovation is more common at higher taxonomic levels than at the order level.
+% We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
+% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
 
-\subsection{Multiple routes to innovation throughout avian evolutionary history}
+\section*{Multiple routes to innovation throughout avian evolutionary history}
 
@@
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
-\caption{.}
-\label{Fig:ellipses}
-\end{figure}
-\bigskip
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
+% \caption{.}
+% \label{Fig:ellipses}
+% \end{figure}
+% \bigskip
+
+% \noindent \textbf{Figure \ref{Fig:ellipses}:}
+% Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
+% We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
+% The first inset ellipse plot shows the positions of all super-order ellipses relative to the global phylogenetic ellipse.
+% Subsequent inset plots show the results for each super-order. 
+% Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+% The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
+% The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
+% These scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation for each plot.
+% Panel B) shows the results for each order.
+% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
+
+% \bigskip
+
+% \bigskip
+
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
+% \caption{.}
+% \label{Fig:orthogonality}
+% \end{figure}
+
+% \bigskip
+
+% \noindent \textbf{Figure \ref{Fig:orthogonality}:}
+% \textbf{Clade's orthogonality.}
+% Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent's parent clade.
+% Orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+% Dots represent the median orthogonality of each clade and the lines their 95\% CI.
+% No super-order's major axis of phenotypic variation is parallel to the global phylogenetic major axis (red lines and circles) and 12/25 orders are at least half orthogonal to the major axis of their super-order (green lines and circles).
+% These results are even clearer when compared to the global phylogenetic major axis (green dashed lines and open circles) where only 3/25 are less than half orthogonal.
+% We also indicate the number of species (n) and the standard deviation (sd) of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+% For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the global phylogenetic major axis relative to their sample size and sd.
+% The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
+% Only two out of six super-orders; 6/25 orders relative to their super-order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
+% This is the result of the variation in sample size (n) and standard variation (sd) among clades. 
+% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+
+% \bigskip
+
+Evolutionary innovation can arise in any direction in trait space.
+Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned with PC1; Fig.
+\ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly with PC1 (Fig.
+\ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
+This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
+S\ref{fig_ellipses_passeriformes}).
+In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis; e.g.
+in Galliformes) or higher-dimensional than expected (there is no clear dominant major axis; e.g.
+in Podicipediformes) from the global phylogenetic major axis of phenotypic variation.
+For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogenetic level but uniformity in their beak shape within the clade.
+In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
+
+\section*{Species elaborate, clades innovate}
+
+The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
+We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
+We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
+We then measured the difference between both areas and found that there was a significant difference in innovation among species and clades but no clear difference in elaboration (Fig.
+\ref{Fig:relative_EI}).
+This difference confirms our observation that while innovation is common and perhaps even dominant at the clade level, elaboration dominates at the species level.
+So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
+However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species.
+This idea is similar in concept to multiple adaptive peak models where species evolve in a heterogeneous adaptive landscape and share similar responses to selection pressures within adaptive zones \cite{hansen1997stabilizing}.
+This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{venditti2011multiple,cooney2017mega}, may be sufficient to explain diversity in avian beak morphology.
+
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=0.9\textwidth]{Figures/relative_EI_gg_style.pdf}
+% \caption{.}
+% \label{Fig:relative_EI}
+% \end{figure}
+
+% \bigskip
+
+% \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
+% a) the elaboration (blue) for each species in each 4000 posterior samples compared to the elaboration for each of the 35 clades (yellow).
+% The scores for each clade are scaled by the maximum score within the clade (corresponding to the scaled ellipses in figure \ref{Fig:ellipses}) and the scores for species are scaled by the maximum species elaboration score.
+% This allows us to compare elaboration in species and clades despite their different sample sizes.
+% We use the Bhattacharryya coefficient (BC) to quantify the overlap of the area under the curve for species and clades. 
+% For elaboration the overlap is BC = 0.217, indicating no distinguishable dissimilarity in the amount of elaboration across the posterior samples between species or clades. 
+% b) the same as a) but using innovation score. 
+% For innovation the overlap has BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.
+
+% Paragraph 8: conclusion
+\section*{Conclusions}
+
+Our results show that although at the species-level most bird beaks are elaborating along a megaevolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
+This nested structure of elaboration at a lower (species) taxonomic level and innovation at a higher (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
+However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales,goswami2022}.
+Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
+
+% \section*{Methods}
+
+% \textbf{Beak shape data.}
+% We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
+% The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
+% We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
+% These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
+% We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
+% $\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each super-order.
+% For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
+% This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. S\ref{Fig:axes_variance}).
+% This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
+
+% \textbf{Phylogenetic data.}
+% We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
+% We pruned these trees to contain only the 8748 species present in our beak shapespace.
+
+% \textbf{Phylogenetic GLMM analysis.}
+% We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} and \texttt{ape} R packages \cite{MCMCglmm, ape} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
+
+% \begin{equation}
+% \text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
+% \end{equation}
+
+% This model allows us to estimate a phenotypic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
+% These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
+% We ran two separate nested models.
+% The first model used one random term for the global phylogeny and 35 nested random terms, one for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
+% This resulted in a multidimensional GLMM with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
+% The second model was fitted to a subset of the dataset containing only the 5229 Passeriformes species with 30 nested random terms, one for each sub-order and families containing more than 15 species in our dataset (Fig.
+S\ref{fig_ellipses_passeriformes}.
+% This resulted in a multidimensional GLMM with a $5229 \times 8$ response variables, one global residual term and 30 random terms.
+
+% To account for phylogenetic uncertainty we ran the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
+% Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each of the models, i.e.
+the global phylogeny and the Passeriformes models).
+% Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
+
+% \textbf{mcmcmcglmmm.}
+% This method runs relatively short Monte-Carlo Markov chains (MCMC) across a large number of different phylogenetic trees and concatenates the resulting short posteriors into one larger posterior distribution that contains the phylogenetic uncertainty information.
+% We performed the method using the following protocol (see Fig. S\ref{Fig:mcmcmcglmm}).
+% First we ran three independent MCMCglmm chains (hereafter the parameterisation chains) for 50k generations with the trait data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\%, i.e.
+with a very low weight on the priors;
+% Next we extracted the following parameters from the parameterisation chains:
+% a) the conservative burnin average time across the three chains, defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
+% b) the mini-chains priors, defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
+% We then ran 400 mini chains in parallel with the estimated burnin time and priors to run 10 samples past the burnin.
+% This resulted in 10 exploitable posterior samples for each tree.
+% We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
+S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
+% Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain - a 400 fold improvement.
+% The total analysis took around three CPU years using the shARC cluster from the University of Sheffield.
+% Code to reproduce the procedure is available in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
+
+% \textbf{Elaboration/innovation scores using projection/rejection analyses.}
+% We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to obtain  elaboration and innovation scores for bird beaks.
+% We can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade) and project either species or another major axis of phenotypic variation onto that clade.
+% In brief we can then interpret where a species or a clade falls on the major axis of phenotypic variation of reference as its elaboration score, and how far a species is from that major axis of phenotypic variation as its innovation score.
+% We calculated the elaboration and innovation scores using two main approaches:
+% 1) by clades (i.e.
+to measure the elaboration/innovation among clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axis of phenotypic variation (Fig.
+\ref{Fig:ellipses});
+% and 2) by species (i.e.
+to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axis of phenotypic variation and b) the major axis of phenotypic variation of their specific clade (Fig.
+\ref{Fig:phylogeny}).
+% To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of phenotypic variation of interest and made the elaboration values absolute.
+% In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
+% The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection}).
+% Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
+% and implemented in the \texttt{dispRity} package \cite{dispRity}.
+
+% \textbf{Clade orthogonality.}
+% We measured the orthogonality of each random term (i.e. clade) in our models compared to their parent clade and their parent's parent clade.
+% For example, the variance-covariance matrices of the Columbimorphae random terms against the phylogeny random terms, the Columbiformes random terms against the Columbimorphae random terms (parent) and the whole phylogeny random terms (parent's parent).
+% To measure orthogonality we calculated the angles between the major axes of phenotypic variation of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade, by randomly comparing major axes of phenotypic variation within a clade.
+% We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
+% We then measured the posterior probability of the angles within a clade being different to the angles among clades.
+% These results are available in Fig. \ref{Fig:orthogonality}.
+
+\bibliographystyle{Science}
+\bibliography{references}
+
+\section*{Acknowledgements}
+We acknowledge IT Services at The University of Sheffield for the provision of the High Performance Computing Service.
+We thank M. Adams, H. van Grouw, R. Prys-Jones and and Zo\"{e} K. Varley from the Bird Group at the NHM, Tring.
+We are indebted to the volunteer citizen scientists at http://www.markmybird.org for helping to build the database of bird beak shape and contribute to our understanding of avian evolution. 
+\textbf{Funding.}
+This work was funded by the European Research Council (615709 Project ‘ToLERates’), UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to G.H.T.).
+\textbf{Author contributions.}
+G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., and G.H.T. % and Z.K.V.
+curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
+\textbf{Competing interests.}
+We declare no competing interests.
+\textbf{Data an materials availability.}
+The raw and processed data used in this analysis is available from \href{https://figshare.shef.ac.uk/account/articles/21526146}{figshare} \cite{fighsaredata}.
+The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{GitHub} \cite{githubrepo}.
+
+\section*{Supplementary materials}
+Methods\\
+Supplementary material 1: glossary\\
+Supplementary material 2: additional results\\
+Supplementary material 3: Passeriformes results\\
+Supplementary materials 4: detailed projection and rejection operations\\
+Figs. S1 to S8\\
+Table S1\\
+
+% TODO: Add list of supplementary materials
+% All this comes after the references
+% Figures and captions after this
+
+\noindent \textbf{Figure \ref{Fig:phylogeny}:}
+The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
+Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
+Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order level, then at the super-order level, and finally at the global phylogenetic level.
+The boxplots at the bottom of the figure summarise the distribution of the median innovation and elaboration values for all levels (bottom left), showing that elaboration is more common than elaboration,
+and the pooled median innovation and elaboration values for all three different levels (bottom right), showing that elaboration and innovation is more common at higher taxonomic levels than at the order level.
+We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
 
@@
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
-
-\bigskip
-
-\bigskip
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
-\caption{.}
-\label{Fig:orthogonality}
-\end{figure}
-
-\bigskip
 
@@
 
-\bigskip
-
-Evolutionary innovation can arise in any direction in trait space.
-Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned with PC1; Fig.
-\ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly with PC1 (Fig.
-\ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
-This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig.
-S\ref{fig_ellipses_passeriformes}).
-In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis; e.g.
-in Galliformes) or higher-dimensional than expected (there is no clear dominant major axis; e.g.
-in Podicipediformes) from the global phylogenetic major axis of phenotypic variation.
-For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogenetic level but uniformity in their beak shape within the clade.
-In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
-
-\subsection{Species elaborate, clades innovate}
-
-The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
-We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
-We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
-We then measured the difference between both areas and found that there was a significant difference in innovation among species and clades but no clear difference in elaboration (Fig.
-\ref{Fig:relative_EI}).
-This difference confirms our observation that while innovation is common and perhaps even dominant at the clade level, elaboration dominates at the species level.
-So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
-However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species.
-This idea is similar in concept to multiple adaptive peak models where species evolve in a heterogeneous adaptive landscape and share similar responses to selection pressures within adaptive zones \cite{hansen1997stabilizing}.
-This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{venditti2011multiple,cooney2017mega}, may be sufficient to explain diversity in avian beak morphology.
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/relative_EI_gg_style.pdf}
-\caption{.}
-\label{Fig:relative_EI}
-\end{figure}
-
-\bigskip
-
 \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
@@
 
-% Paragraph 8: conclusion
-\subsection{Conclusions}
+\begin{figure}[!htbp]
+\centering
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text_with_box.pdf}
+\caption{.}
+\label{Fig:phylogeny}
+\end{figure}
 
-Our results show that although at the species-level most bird beaks are elaborating along a megaevolutionary major axis of phenotypic variation, at the macroevolutionary level innovation away from the global phylogenetic major axis of phenotypic variation is much more common.
-This nested structure of elaboration at a lower (species) taxonomic level and innovation at a higher (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
-However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary reorientations in deep-time (innovation), coupled with elaboration at the species-level, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions that apparently abrupt phenotypic shifts observed in univariate analyses can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales,goswami2022}.
-Indeed, the global phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
+\caption{.}
+\label{Fig:ellipses}
+\end{figure}
 
-\section{Methods}
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
+\caption{.}
+\label{Fig:orthogonality}
+\end{figure}
 
-\textbf{Beak shape data.}
-We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
-The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
-We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
-These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
-We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
-$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each super-order.
-For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
-This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. S\ref{Fig:axes_variance}).
-This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
-
-\textbf{Phylogenetic data.}
-We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
-We pruned these trees to contain only the 8748 species present in our beak shapespace.
-
-\textbf{Phylogenetic GLMM analysis.}
-We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} and \texttt{ape} R packages \cite{MCMCglmm, ape} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
-
-\begin{equation}
-\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
-\end{equation}
-
-This model allows us to estimate a phenotypic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
-These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
-We ran two separate nested models.
-The first model used one random term for the global phylogeny and 35 nested random terms, one for each super-order and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
-This resulted in a multidimensional GLMM with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
-The second model was fitted to a subset of the dataset containing only the 5229 Passeriformes species with 30 nested random terms, one for each sub-order and families containing more than 15 species in our dataset (Fig.
-S\ref{fig_ellipses_passeriformes}.
-This resulted in a multidimensional GLMM with a $5229 \times 8$ response variables, one global residual term and 30 random terms.
-
-To account for phylogenetic uncertainty we ran the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
-Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each of the models, i.e.
-the global phylogeny and the Passeriformes models).
-Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
-
-\textbf{mcmcmcglmmm.}
-This method runs relatively short Monte-Carlo Markov chains (MCMC) across a large number of different phylogenetic trees and concatenates the resulting short posteriors into one larger posterior distribution that contains the phylogenetic uncertainty information.
-We performed the method using the following protocol (see Fig. S\ref{Fig:mcmcmcglmm}).
-First we ran three independent MCMCglmm chains (hereafter the parameterisation chains) for 50k generations with the trait data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\%, i.e.
-with a very low weight on the priors;
-Next we extracted the following parameters from the parameterisation chains:
-a) the conservative burnin average time across the three chains, defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
-b) the mini-chains priors, defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
-We then ran 400 mini chains in parallel with the estimated burnin time and priors to run 10 samples past the burnin.
-This resulted in 10 exploitable posterior samples for each tree.
-We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
-S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
-Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain - a 400 fold improvement.
-The total analysis took around three CPU years using the shARC cluster from the University of Sheffield.
-Code to reproduce the procedure is available in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
-
-\textbf{Elaboration/innovation scores using projection/rejection analyses.}
-We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to obtain  elaboration and innovation scores for bird beaks.
-We can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade) and project either species or another major axis of phenotypic variation onto that clade.
-In brief we can then interpret where a species or a clade falls on the major axis of phenotypic variation of reference as its elaboration score, and how far a species is from that major axis of phenotypic variation as its innovation score.
-We calculated the elaboration and innovation scores using two main approaches:
-1) by clades (i.e.
-to measure the elaboration/innovation among clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axis of phenotypic variation (Fig.
-\ref{Fig:ellipses});
-and 2) by species (i.e.
-to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axis of phenotypic variation and b) the major axis of phenotypic variation of their specific clade (Fig.
-\ref{Fig:phylogeny}).
-To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of phenotypic variation of interest and made the elaboration values absolute.
-In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
-The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection}).
-Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
-and implemented in the \texttt{dispRity} package \cite{dispRity}.
-
-\textbf{Clade orthogonality.}
-We measured the orthogonality of each random term (i.e. clade) in our models compared to their parent clade and their parent's parent clade.
-For example, the variance-covariance matrices of the Columbimorphae random terms against the phylogeny random terms, the Columbiformes random terms against the Columbimorphae random terms (parent) and the whole phylogeny random terms (parent's parent).
-To measure orthogonality we calculated the angles between the major axes of phenotypic variation of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade, by randomly comparing major axes of phenotypic variation within a clade.
-We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
-We then measured the posterior probability of the angles within a clade being different to the angles among clades.
-These results are available in Fig. \ref{Fig:orthogonality}.
-
-\bibliographystyle{naturemag}
-\bibliography{references}
-
-\textbf{Acknowledgements.}
-We acknowledge IT Services at The University of Sheffield for the provision of the High Performance Computing Service.
-We thank the museum in Tring.
-% TG: TODO: just take Chris' paper acknowledgments
-\textbf{Funding.}
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
-% TG: TODO: add ERC grant (ask Gavin)
-\textbf{Author contributions.}
-G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
-\textbf{Competing interests.}
-We declare no competing interests.
-\textbf{Data an materials availability.}
-The raw and processed data used in this analysis is available from \href{https://figshare.shef.ac.uk/account/articles/21526146}{figshare} \cite{fighsaredata}.
-The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{GitHub} \cite{githubrepo}.
-
-% TODO: Add list of supplementary materials
-% All this comes after the references
-% Figures and captions after this
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/relative_EI_gg_style.pdf}
+\caption{.}
+\label{Fig:relative_EI}
+\end{figure}
 

### 2023-04-04 Analysis/09-Figure_key.Rmd
Update 09-Figure_key.Rmd

@@
 ## Get the base vectors
-covars <- sample.n(space_paleo$covar, 1000)
+covars <- sample.n(space_paleo$covar, 4000)
 centres <- lapply(unlist(space_paleo$subsets, recursive = FALSE) ,function(group, data, fun) fun(data[c(group), ]), data = space_paleo$matrix[[1]], fun = colMeans)
@@
 base <- base_vectors$Paleognathae[[1]]
-base <- matrix(c(base)[c(1, 1, 4, 4)], 2, 2, byrow = FALSE) # or c(1, 4, 3, 2)
+# base <- matrix(c(base)[c(1, 1, 4, 4)], 2, 2, byrow = FALSE) # or c(1, 4, 3, 2)
 for(i in 1:nrow(matrix)) {
-      lines(get.proj.line(matrix[i, ], test))
+      lines(get.proj.line(matrix[i, ], base))
 }
@@
 ## Transform them into a colour gradient
-col.grad <- colorRamp(rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#66c2a5", "#3288bd")))
-cols_elab <- col.grad(elaborations/max(elaborations))
-cols_innov <- col.grad(innovations/max(innovations))
+## Change colours to grey -> blue innov. grey -> orange elab
+col.grad.inno <- colorRamp(c("grey", "blue"))
+col.grad.elab <- colorRamp(c("grey", "orange"))
+# col.grad <- colorRamp(rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#66c2a5", "#3288bd")))
+cols_elab <- col.grad.elab (elaborations/max(elaborations))
+cols_inno <- col.grad.inno(innovations/max(innovations))
 
@@
 tiplabels("", frame = "none", pch = 19, cex = 2, col = rgb(cols_elab, maxColorValue = 256), offset = -1)
-tiplabels("", frame = "none", pch = 19, cex = 2, col = rgb(cols_innov, maxColorValue = 256), offset = 4)
+tiplabels("", frame = "none", pch = 19, cex = 2, col = rgb(cols_inno, maxColorValue = 256), offset = 4)
+dev.off()
+
+pdf("../Manuscript/Figures/figure_key_tree_example2.pdf")
+plot(ladderize(tree_paleo), show.tip.label = FALSE)
 dev.off()

### 2023-04-10 Analysis/04-Figure_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure_with_ellipse_scalling_and_8D_histograms.Rmd

@@
 
+TODO:
+ - [ ] split in two figures
+ - [ ] make E/I boxes larger
+ - [ ] make E/I colours always orange/blue
+
 ```{r, eval = TRUE, echo = FALSE}
-## Get the bar plots for the loadings of the phylogeny component (to be added separately in Inkscape to the main plot)
-# pdf(file = "../Manuscript/Figures/phylo_loadings.pdf", height = 4, width = 4)
-# plot(NULL, xlim = range(superorder_shapespace$matrix), ylim = range(superorder_shapespace$matrix))
-# add.dims(superorder_shapespace, name.col = c(phylogeny = "grey"), n = 1000)
-# dev.off()
+## Matching colours (fast utility)
+match.col <- function(data, col) {
+    return(c(col[match(names(size.subsets(data)), names(col))[-length(size.subsets(data))]], "grey"))
+}
+```
 
+### Ellipses plot super orders
+
+```{r, eval = TRUE, echo = FALSE}
+## Plot dimensions and name
 box_size <- 1.5
-pdf(file = "../Manuscript/Figures/ellipses.pdf", height = box_size*10, width = box_size*10)
+pdf(file = "../Manuscript/Figures/ellipses_super_order.pdf", height = box_size*4, width = box_size*8)
 
-## SETTING THE WINDOW PARAMETERS
-## Global window parameters
-lay_mat_super <- rbind(
-        matrix(c(1,1, rep(73, 3), 0, 2:5), ncol = 10, nrow = 1), # line 7
-        matrix(c(1,1, rep(73, 3), 0, 6:9), ncol = 10, nrow = 1), # line 8
-        matrix(c(78,78, 10:17), ncol = 10, nrow = 1)) # line 8)
-lay_mat_order <- rbind(
-        matrix(c(1,1, rep(57, 3), 60, 2:5), ncol = 10, nrow = 1), 
-        matrix(c(1,1, rep(57, 3), 60, 6:9), ncol = 10, nrow = 1), 
-        matrix(c(1,1, 10:17), ncol = 10, nrow = 1),
-        matrix(c(18:27), ncol = 10, nrow = 1),
-        matrix(c(28:37), ncol = 10, nrow = 1), 
-        matrix(c(38:47), ncol = 10, nrow = 1),
-        matrix(c(48:55, 58:59), ncol = 10, nrow = 1))
-heights_vals <- rep(1, 9)
-widths_vals <- rep(c(1, 1/3), 5)
-layout_matrix <- layout(rbind(lay_mat_super, (lay_mat_order+17)), widths = widths_vals, heights = heights_vals, respect = TRUE)
+## New layout for super_order panel (separately)
+lay_mat_super_2 <- rbind(
+    matrix(c(1,1, rep(18, 3), 0, 2:3), ncol = 8, nrow = 1),
+    matrix(c(1,1, rep(18, 3), 0, 4:5), ncol = 8, nrow = 1),
+    matrix(c(6:13), ncol = 8, nrow = 1),
+    matrix(c(14:17, rep(0, 4)), ncol = 8, nrow = 1))
+widths_vals <- rep(c(1, 1/2), 4)
+heights_vals <- rep(1, 4)
+layout_super <- layout(lay_mat_super_2, widths = widths_vals, heights = heights_vals, respect = TRUE)
+# layout.show(layout_super)
 
@@
 
-## Plotting the ellipses
+## Plotting the ellipses for each order
 for(i in 1:length(tip_order)) {
     ## Plot each pannel
-    wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between)
+    wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between, ei.col = c("orange", "blue"))
 }
 
-## Matching colours (fast utility)
-match.col <- function(data, col) {
-    return(c(col[match(names(size.subsets(data)), names(col))[-length(size.subsets(data))]], "grey"))
-}
+## Adding the ellipses
+par(bty = "n")
+lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
+plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "n", ylab = "n")
+abline(v = 0, col = "grey", lwd = 0.5)
+abline(h = 0, col = "grey", lwd = 0.5)
+covar.plot(superorder_shapespace, ellipses = mean,
+           col = match.col(superorder_shapespace, tip_colours_super_orders),
+           legend = FALSE, points = FALSE, scale = "phylogeny",
+           lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
+           add = TRUE, n = 100)
+dev.off()
+```
 
-## Matching the order and the colours to the ladderised tree
+### Ellipses plot orders
+
+```{r, eval = TRUE, echo = FALSE}
+## Plot dimensions and name
+box_size <- 1.5
+pdf(file = "../Manuscript/Figures/ellipses_order.pdf", height = box_size*7, width = box_size*10)
+
+## New layout for order panel (separately)
+lay_mat_order_2 <- rbind(
+        matrix(c(1,1, rep(56, 3), 0, 2:5), ncol = 10, nrow = 1), 
+        matrix(c(1,1, rep(56, 3), 0, 6:9), ncol = 10, nrow = 1), 
+        matrix(c(1,1, 10:17), ncol = 10, nrow = 1),
+        matrix(c(18:27), ncol = 10, nrow = 1),
+        matrix(c(28:37), ncol = 10, nrow = 1), 
+        matrix(c(38:47), ncol = 10, nrow = 1),
+        matrix(c(48:55, rep(0, 2)), ncol = 10, nrow = 1))
+widths_vals <- rep(c(1, 1/2), 5)
+heights_vals <- rep(1, 7)
+layout_order <- layout(lay_mat_order_2, widths = widths_vals, heights = heights_vals, respect = TRUE)
+# layout.show(layout_order)
+
+## PLOTTING THE SUPERORDERS
+
+## Getting the tip order for the ladderised super order tree
+
+## Getting the tip order for the ladderised super order tree
 tip_order <- c(
@@
     ## Plot each pannel
-    wrap.plot.ellipses(i, tip_colours_orders, order_shapespace, results$order_results_between)
+    wrap.plot.ellipses(i, tip_colours_orders, order_shapespace, results$order_results_between, ei.col = c("orange", "blue"))
 }
 
-## Plotting the aligned and scale ellipses for the superorders
-par(bty = "n")
-lims <- c(-1/3, 1/3) # c(-1.25, 1.25)
-plot(NULL, xlim = lims, ylim = lims, xaxt = "n", yaxt = "n", xlab = "n", ylab = "n")
-abline(v = 0, col = "grey", lwd = 0.5)
-abline(h = 0, col = "grey", lwd = 0.5)
-covar.plot(superorder_shapespace, ellipses = mean,
-           col = match.col(superorder_shapespace, tip_colours_super_orders),
-           legend = FALSE, points = FALSE, scale = "phylogeny",
-           lines.lwd = c(rep(1, length(tip_colours_super_orders)), 3),
-           add = TRUE, n = 100)
-
-## PLOTTING THE ORDERS
+## Adding the ellipses
 par(bty = "n")
@@
 dev.off()
+
 ```

### 2023-04-11 Analysis/06-Figure_with_correlations.Rmd
Update 06-Figure_with_correlations.Rmd

@@
 ## Calculate the correlations for the projections onto the phylo axis
-super_phylo_cor <- get.correlations(results$super_results_phylo)
+super_phylo_cor <- get.correlations.plot(results$super_results_phylo)
 
 ## Calculate the correlations for the projections onto the group' axes
-super_group_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
+super_group_cor <- get.correlations.plot(results$super_results_within, exclude.phylo = TRUE)
+```
+
+## Get the correlation fable
+
+y axis = levels
+column 1 = correlation of e/i projected on the mega level
+column 2 = correlation of e/i projected on the macro level
+stars for significant correlation between means.
+
+```{r, echo = TRUE, eval = TRUE}
+## Get all the correlations for all the clades
+cor_mega_level <- get.correlations.scores(results$super_results_phylo)
+cor_macro_level <- get.correlations.scores(results$super_results_within)
+
+## Get the colours
+load("../Data/Processed/tip_colours_super_orders.rda")
+super_orders_col <- c(tip_colours_super_orders, "phylogeny" = "grey")
+
+## Plot the correlations
+pdf("../Manuscript/Figures/correlations_fable.pdf")
+par(mfrow = c(1,2))
+par(mar = c(5, 8, 4, 0) + 0.1, bty = "n") # c(bottom, left, top, right)
+plot.cor.scores(cor_mega_level, col = super_orders_col,
+                y.axis = TRUE, plot.phylo = TRUE,
+                main = "Elaboration and innovation\nat the mega level") 
+par(mar = c(5, 2, 4, 6) + 0.1, bty = "n") # c(bottom, left, top, right)
+plot.cor.scores(cor_macro_level, col = super_orders_col,
+                y.axis = FALSE, plot.phylo = FALSE,
+                main = "Elaboration and innovation\nat the macro level")
+dev.off()
 ```
@@
 ## Calculate the correlations for the projections onto the phylo axis
-suborder_phylo_cor <- get.correlations(results$suborder_results_phylo)
+suborder_phylo_cor <- get.correlations.plot(results$suborder_results_phylo)
 
 ## Calculate the correlations for the projections onto the group' axes
-suborder_group_cor <- get.correlations(results$suborder_results_within, exclude.phylo = TRUE)
+suborder_group_cor <- get.correlations.plot(results$suborder_results_within, exclude.phylo = TRUE)
 ```

### 2023-04-11 Functions/correlations.utilities.R
correlation fable bizz

@@ -1,4 +1,4 @@
-#' @name get.correlations
+#' @name get.correlations.plot
 #'
 #' @description extracts the point estimate (e.g. median) of a series of results for the correlation plots
 #' 
@@ -7,7 +7,7 @@
 #' @param exclude.phylo logical, whether to exclude the phylogeny component (level 0)
 #' 
 ## Extract the point estimate correlation from the results structure
-get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE) {
+get.correlations.plot <- function(res, cent.tend = median, exclude.phylo = FALSE) {
 
     ## Remove the phylo one (if needed)
     if(exclude.phylo) {
@@ -41,4 +41,41 @@ get.correlations <- function(res, cent.tend = median, exclude.phylo = FALSE) {
     }
 
     return(mapply(prep.cor.results, centrals_pos, centrals_dis, as.list(names(centrals_pos)), SIMPLIFY = FALSE))
+}
+
+#' @name get.correlations.scores
+#'
+#' @description calculates the distribution of correlation between elaboration and innovation
+#'
+#' @param res the results in a dispRity list structure (typically from dispRity.covar.projections)
+#' @param cent.tend the central tendency for testing the correlation
+#' @param exclude.phylo logical, whether to exclude the phylogeny component (level 0)
+#' @param test which test to apply to check correlation
+get.correlations.scores <- function(res, cent.tend = median, exclude.phylo = FALSE, test = cor.test) {
+    ## Remove the phylo one (if needed)
+    if(exclude.phylo) {
+        select_pos <- names(size.subsets(res$position))
+        select_dis <- names(size.subsets(res$distance))
+        select_pos <- select_pos[-which(select_pos %in% "phylogeny")]
+        select_dis <- select_dis[-which(select_dis %in% "phylogeny")]
+        res$position <- get.subsets(res$position, subsets = select_pos)
+        res$distance <- get.subsets(res$distance, subsets = select_dis)
+    }
+
+    ## Extracting all the results
+    disparity_pos <- lapply(res$position$disparity, function(X, type) return(X$elements))
+    disparity_dis <- lapply(res$distance$disparity, function(X, type) return(X$elements))
+    
+    ## Measuring the correlations
+    get.cor <- function(elaboration, innovation, cent.tend, test) {
+        correlations <- mapply(cor,
+                               unlist(apply(elaboration, 2, list), recursive = FALSE),
[…]

### 2023-04-11 Functions/ploting.utilities.R

@@ -146,9 +146,10 @@ add.dims <- function(data, name.col, n = 1000) {
 #' @param x logical, whether to plot the x axis
 #' @param y logical, whether to plot the y axis
 #' @param main the plot title
+#' @param ei.col optional, the colour for elaboration and innovation (overrides name.col)
 #' 
 ## Plot the projection/rejection
-plot.one.proj.rej <- function(data, name.col, x = TRUE, y = TRUE, main = "") {
+plot.one.proj.rej <- function(data, name.col, x = TRUE, y = TRUE, main = "", ei.col) {
 
     ## Get the proj/rej
     elaboration <- get.disparity(data$position, concatenate = FALSE)
@@ -163,6 +164,12 @@ plot.one.proj.rej <- function(data, name.col, x = TRUE, y = TRUE, main = "") {
 
     ## Change the colours
     col <- c(name.col, adjustcolor(name.col, alpha.f = 0.5))
+
+    ## Overiding colours
+    if(!missing(ei.col)) {
+        col <- ei.col
+    }
+
     
     ## Set the limits
     quant_95 <- quantile(c(elaboration, exploration), prob = 0.975)
@@ -254,9 +261,10 @@ plot.short.tree <- function(tree, shapespace, level, colour.palette = gg.color.h
 #' @param tip.colours the named vector with colours and group names
 #' @param shapespace the shapespace as a dispRity object
 #' @param results the list of dispRity objects containing the results (typically output from dispRity.covar.projections)
+#' @param ei.col optional, the colour for elaboration and innovation (overriding tip.colours)
 #
 ## Plotting the ellipses, the dimensions and the projection/rejection
-wrap.plot.ellipses <- function(i, tip.colours, shapespace, results) {
+wrap.plot.ellipses <- function(i, tip.colours, shapespace, results, ei.col) {
     ## Select the colour and name
     name.col <- tip.colours[i]
 
@@ -274,7 +282,7 @@ wrap.plot.ellipses <- function(i, tip.colours, shapespace, results) {
 
     ## Plot the elaboration/exploration
     par(mar = c(2, 1, 1, 1)+0.1)
-    plot.one.proj.rej(results, name.col, x = TRUE, y = FALSE)
+    plot.one.proj.rej(results, name.col, x = TRUE, y = FALSE, ei.col = ei.col)
 }
 
 #' @name scale rgb color
[…]

### 2023-04-12 Analysis/14-Elab_innov_through_time.Rmd
Create 14-Elab_innov_through_time.Rmd

--- 2022-11-09 Analysis/06-Figure_with_correlations.Rmd
@@
 
-This script reproduces a figure not present or used in the manuscript or the supplementary materials.
-
 ```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
@@
 ```
-
-# All birds analysis
 
@@
 
-Extract the median elaboration and innovation from the results for the mega level (projection on the phylogeny) and the macro level (projection on each group).
+# Checking elaboration and innovation values through time (clades)
 
 ```{r, eval = TRUE, echo = TRUE}
-## Calculate the correlations for the projections onto the phylo axis
-super_phylo_cor <- get.correlations(results$super_results_phylo)
+## Get all the statistics
+super_orders_elaboration <- summary(results$super_results_between$position)
+super_orders_innovation <- summary(results$super_results_between$distance)
+orders_elaboration <- summary(results$order_results_between$position)
+orders_innovation <- summary(results$order_results_between$distance)
 
-## Calculate the correlations for the projections onto the group' axes
-super_group_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
-```
+## Get all the tree ages
+load("../Data/Processed/ploting_data.rda")
+load("../Data/Processed/clade_info.rda")
 
-## Ploting the figure
-
-The following snippets plots the entire figure minus the panel IDs:
- * Panel A: the overal density figure
- * Panel B: the correlation plots for the projection on the mega level
- * Panel C: the correlation plots for the projection on the macro level
-
-Density plot of the correlation of the median elaboration and median exploration for all species.
-
-```{r, echo = TRUE, eval = TRUE}
-## Set the pdf parameters
-pdf("../Manuscript/Figures/correlations.pdf", height = 12, width = 8)
-
-## Set up the plotting layout
-layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
-layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
-plot_layout <- layout(layout_matrix)
-# layout.show(plot_layout)
-
-## Setting the colour pallette for the subsequent plots
-colour.palette <- gg.color.hue
-
-## --------
-## PANEL A
-## --------
-
-## Plotting the densities (with the quadrants) in the first panel
-par(mar = c(5, 4, 4, 2) + 0.1)
-plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)
-
-## --------
-## PANEL B
-## --------
-
-## Getting the subsets for the colouring
-col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
-
-## Set the vector of legend positions
-legends_pos <- c("topleft", "topright", "topright", "topright",
-                 "topleft", "topright", "topleft", "topright")
-## Create the vector for counting the legend numbers
-legend_counter <- 1
-
-## Plotting each individual correlation plots
-for(i in 1:8) {
-    par(mar = c(2,2.5,2,1.5))
-    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
+get.clade.age <- function(name, info, tree) {
+  return(list(age = max(dispRity::tree.age(keep.tip(tree, info[[name]]$species))$ages), col = info[[name]]$col))
 }
 
-## --------
-## PANEL C
-## --------
+super_age_info <- t(sapply(unlist(strsplit(super_orders_elaboration$subsets, split = ":phylogeny")), get.clade.age, info = clade_info$super_orders, tree = ploting_data$consensus_tree))
+order_age_info <- t(sapply(unlist(strsplit(orders_elaboration$subsets, split = ":phylogeny")), get.clade.age, info = clade_info$orders, tree = ploting_data$consensus_tree))
 
-## Set the vector of legend positions
-legends_pos <- c("bottomright", "topright", "topleft", "topleft",
-                 "topleft", "topleft", "topright", "bottomright")
-## Resetting the legend_counter
-legend_counter <- 1
+## Table that for easy plotting
+super_orders_elaboration <- cbind(super_orders_elaboration, super_age_info, cex = 2)
+super_orders_innovation <- cbind(super_orders_innovation, super_age_info, cex = 2)
+orders_elaboration <- cbind(orders_elaboration, order_age_info, cex = 1)
+orders_innovation <- cbind(orders_innovation, order_age_info, cex = 1)
+elaboration_plot_data <- rbind(super_orders_elaboration, orders_elaboration)
+innovation_plot_data <- rbind(super_orders_innovation, orders_innovation)
 
-## Plotting each individual correlation plots
-for(i in 1:8) {
-    par(mar = c(2,2.5,2,1.5))
-    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
+## Empty plot
+plot.age.data <- function(data, ylab, xlab) {
+  plot(NULL, xlab = xlab, ylab = ylab,
+       xlim = rev(range(pretty(range(data$age)))),
+       ylim = range(pretty(range(data[,c(4, 8)]))))
+  ## Adding the lines for each clade
+  for(i in 1:nrow(data)) {
+    lines(x = rep(data[i, "age"][[1]], 2), y = data[i, c("2.5%", "97.5%")], lty = 2, col = data[i, "col"][[1]])
+    lines(x = rep(data[i, "age"][[1]], 2), y = data[i, c("25%", "75%")], lty = 1, lwd = 2, col = data[i, "col"][[1]])
+    points(x = data[i, "age"][[1]], y = data[i, "obs.median"], cex = data[i, "cex"], col = data[i, "col"][[1]], pch = 19)
   }
+}
+pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_through_time.pdf")
+par(mfrow = c(2,1))
+plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "")
+legend("topleft", pch = 19, col = "grey", legend = c("super orders", "orders"), pt.cex = c(2, 1))
+plot.age.data(innovation_plot_data, ylab = "Innovation", xlab = "Time (Mya)")
 dev.off()
 ```
-
-Extract the median elaboration and innovation from the results for the macro level (projection on the passeriformes) and the macro level-1 (projection on each group).
-
-# Passeriformes analysis
-
-## Loading the data
-
-```{r, eval = TRUE, echo = TRUE}
-## Load the data
-load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
-## Select the data
-shapespace <- shapespace_passeriformes_two_levels[[1]]$space
-tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
-## Clean garbage
-rm(shapespace_passeriformes_two_levels)
-
-## Load the covar
-load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
-shapespace_passeriformes <- shapespace_passeriformes[[1]]
-
-## Get the levels
-phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
-subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
-famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))
-
-## Creating the subsets
-suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
-      data          = shapespace,
-      posteriors    = shapespace_passeriformes,
-      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
-      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""
-
-family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
-      data          = shapespace,
-      posteriors    = shapespace_passeriformes,
-      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
-      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
-## Clean garbage
-rm(shapespace_passeriformes)
-
-## Loading the results
-load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
-results <- shapespace_passeriformes_suborder_family_results_list
-rm(shapespace_passeriformes_suborder_family_results_list)
-```
-
-```{r, eval = TRUE, echo = FALSE}
-## Calculate the correlations for the projections onto the phylo axis
-suborder_phylo_cor <- get.correlations(results$suborder_results_phylo)
-
-## Calculate the correlations for the projections onto the group' axes
-suborder_group_cor <- get.correlations(results$suborder_results_within, exclude.phylo = TRUE)
-```
-
-## Ploting the figure
-
-The following snippets plots the entire figure minus the panel IDs:
- * Panel A: the overal density figure
- * Panel B: the correlation plots for the projection on the mega level
- * Panel C: the correlation plots for the projection on the macro level
-
-Density plot of the correlation of the median elaboration and median exploration for all species.
-
-```{r, echo = TRUE, eval = TRUE}
-## Set the pdf parameters
-pdf("../Manuscript/Figures/correlations_passeriformes.pdf", height = 12, width = 6)
-
-## Set up the plotting layout
-layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,1))
-layout_matrix <- rbind(rbind(layout_matrix, matrix(1:6+1, 2, 3, byrow = TRUE)), matrix(7:12+1, 2, 3,  byrow = TRUE))
-plot_layout <- layout(layout_matrix)
-# layout.show(plot_layout)
-
-## Setting the colour pallette for the subsequent plots
-colour.palette <- gg.color.hue
-
-## --------
-## PANEL A
-## --------
-
-## Plotting the densities (with the quadrants) in the first panel
-par(mar = c(5, 4, 4, 2) + 0.1)
-plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)
-
-## --------
-## PANEL B
-## --------
-
-## Getting the subsets for the colouring
-col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))
-
-## Set the vector of legend positions
-legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
-## Create the vector for counting the legend numbers
-legend_counter <- 1
-
-## Plotting each individual correlation plots
-for(i in 1:6) {
-    par(mar = c(2,2.5,2,1.5))
-    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
-}
-
-## --------
-## PANEL C
-## --------
-
-## Set the vector of legend positions
-legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
-
-## Resetting the legend_counter
-legend_counter <- 1
-
-## Plotting each individual correlation plots
-for(i in 1:6) {
-    par(mar = c(2,2.5,2,1.5))
-    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
-}
-dev.off()
-```
-
-Legend numbers are:
-1 - others (groups with < 15 species)
-2 - Corvoidea
-3 - Malaconotoidea
-4 - Orioloidea
-5 - Meliphagoidea
-6 - Bombycilloidea
-7 - Muscicapoidea
-8 - Emberizoidea
-9 - Motacillidae
-10 - Nectariniidae 
-11 - Passeridae
-12 - PloceidaeEstrildidae
-13 - Eurylaimides
-14 - Furnariida
-15 - Tyrannida 
-16 - Aegithaloidea
-17 - Cisticolidae
-18 - Fringillidae
-19 - Hirundinidae
-20 - Locustelloidea
-21 - Paridae
-22 - Pycnonotidae
-23 - Sylvioidea

### 2023-04-17 Analysis/07-Orthogonality_and_statistics.Rmd
Update 07-Orthogonality_and_statistics.Rmd
Working on PGLS (still buged)

@@
 
-library(tidyverse)
+# library(tidyverse)
 library(patchwork)
@@
 
-ggsave("relative_EI_gg_style.pdf", plot=p, height=6, width=12)
+ggsave("../Manuscript/Figures/relative_EI_gg_style_unscaled.pdf", plot=p, height=6, width=12)
 ```
@@
 ```
+
+## PGLS
+
+```{r}
+## Combining the model data as:
+    ## centroid
+    ## elaboration
+    ## innovation
+    ## elaboration order
+    ## innovation order
+    ## order_tree_elab/innov (from groups_median_EI$orders)
+        # model_data[match(spp, model_data$species), "order_tree_elab"] <- groups_median_EI$orders[[1]][i]
+    ## species names as row names
+load("../Data/Processed/ploting_data.rda")
+#load("../Data/Processed/ploting_data_superorder.rda")
+load("../Data/Processed/ploting_data_order.rda")
+load("../Data/Processed/groups_median_EI.rda")
+load("../Data/Processed/shapespace.rda")
+
+## Initialise the data
+model_data <- data.frame("species" = rownames(shapespace), "orders" = shapespace$level2)
+rownames(model_data) <- rownames(shapespace) 
+model_data <- cbind(model_data,
+                    centroid    = ploting_data$distances_to_centre[rownames(shapespace)],
+                    elaboration_order = ploting_data_order$median_elaborations[rownames(shapespace)],
+                    innovation_order  = ploting_data_order$median_innovations[rownames(shapespace)],
+                    order_tree_elab = NA, order_tree_inno = NA)
+## Attributing the median innov/elab value for each order
+for(order in unique(model_data$orders)) {
+    if(order != "") { # This is because a couple of species have no order
+        group_num <- grep(order, names(groups_median_EI$orders$innovation)) ## FIX THAT!
+        model_data[which(model_data$orders == order), "order_tree_elab"] <- groups_median_EI$orders$elaboration[group_num]
+        model_data[which(model_data$orders == order), "order_tree_inno"] <- groups_median_EI$orders$innovation[group_num]
+    }
+}
+
+consensus_tree <- ploting_data$consensus_tree
+trees <- list(consensus_tree, consensus_tree)
+
+clean_data <- model_data[!(apply(data, 1, function(x) any(is.na(x)))), ] ## NOT WORKING
+
+run.models <- function(tree, data) {
+    ## all predictors
+    all_preds <- phylolm::phylolm(species ~ elaboration_order + innovation_order + order_tree_elab + order_tree_inno,
+                         data = data, phy = tree, model = "lambda")
+    ## macro predictors
+    macro_preds <- phylolm::phylolm(species ~ elaboration_order + innovation_order,
+                           data = data, phy = tree, model = "lambda")
+    ## mega predictors
+    mega_preds <- phylolm::phylolm(species ~ order_tree_elab + order_tree_inno,
+                          data = data, phy = tree, model = "lambda")
+    return(list(all = all_preds, macro = macro_preds, mega = mega_preds))
+}
+
+## Run 100 models
+models_out <- lapply(trees, run.models, data = clean_data)
+
+## Summarise them
+summary_out <- lapply(models_out, lapply, summary)
+
+## Tree is: phy <- read.tree("Prum_merge_hackett_stage2_1K_mcc.tree") (consensus - but get from distribution)
+## Models are:
+
+# # Model for all predictors
+# mod1 <- phylolm(phylogeny ~ elaboration_order + innovation_order + order_tree_elab + order_tree_inno, data=model_data, phy=phy, model="lambda")
+# summary(mod1)
+
+```

### 2023-04-18 Analysis/07-Orthogonality_and_statistics.Rmd
Update 07-Orthogonality_and_statistics.Rmd

--- 2022-11-09 Analysis/07-Orthogonality_and_statistics.Rmd
@@
 ```
+
+## PGLS
+
+```{r}
+## Get the median estimates for elaboration and innovation
+load("../Data/Processed/ploting_data.rda")
+load("../Data/Processed/ploting_data_order.rda")
+load("../Data/Processed/groups_median_EI.rda")
+load("../Data/Processed/shapespace.rda")
+
+## Initialise the data
+model_data <- data.frame("orders" = shapespace$level2, "super_orders" = shapespace$level1)
+rownames(model_data) <- rownames(shapespace) 
+model_data <- cbind(model_data,
+                    centroid            = ploting_data$distances_to_centre[rownames(shapespace)],
+                    species_elaboration = ploting_data_order$median_elaborations[rownames(shapespace)],
+                    species_innovation  = ploting_data_order$median_innovations[rownames(shapespace)],
+                    orders_elaboration = NA, orders_innovation = NA)
+## Attributing the median innov/elab value for each order
+for(order in unique(model_data$orders)) {
+    if(order != "") { # This is because a couple of species have no order
+        group_num <- grep(order, names(groups_median_EI$orders$innovation))
+        model_data[which(model_data$orders == order), "orders_elaboration"] <- groups_median_EI$orders$elaboration[group_num]
+        model_data[which(model_data$orders == order), "orders_innovation"] <- groups_median_EI$orders$innovation[group_num]
+    }
+}
+
+## Remove the tips left over from the phylogeny
+trees_list <- read.tree("../Data/Raw/Tree_data/AllBirdsHackett1.tre")
+
+run.models <- function(tree, data) {
+    ## all predictors
+    cat(".")
+    all_preds <- phylolm::phylolm(centroid ~ species_elaboration + species_innovation + orders_elaboration + orders_innovation,
+                         data = data, phy = tree, model = "lambda")
+    ## macro predictors
+    macro_preds <- phylolm::phylolm(centroid ~ species_elaboration + species_innovation,
+                           data = data, phy = tree, model = "lambda")
+    ## mega predictors
+    mega_preds <- phylolm::phylolm(centroid ~ orders_elaboration + orders_innovation,
+                          data = data, phy = tree, model = "lambda")
+    return(list(all = all_preds, macro = macro_preds, mega = mega_preds))
+}
+
+## Run 100 models (9 sec per model - 15 minutes for 100 models)
+pgls_models_out <- lapply(trees_list[sample(1:length(trees_list), size = 100)], run.models, data = model_data)
+save(pgls_models_out, file = "../Data/Processed/pgls_models_out.rda")
+
+## Summarise the models
+summary_out <- lapply(pgls_models_out, lapply, summary)
+
+## Extract the slopes, p-values and r2
+extract.summary <- function(one_summary) {
+    ## Get the fits
+    fits <- unlist(lapply(one_summary, function(X) return(X$adj.r.squared)))
+    names(fits) <- paste0(names(fits), ".r2")
+    ## Get the coefficients and their p vals
+    coeffs <- lapply(one_summary, function(X) return(X$coefficient))
+
+    get.est <- function(one_coeffs) {
+        vals <- c(one_coeffs[-1, c("Estimate", "p.value")])
+        names(vals) <- c(paste0(rownames(one_coeffs[-1, ]), c(".slope")), paste0(rownames(one_coeffs[-1, ]), c(".pval")))
+        return(vals)
+    }
+    return(c(fits, unlist(lapply(lapply(one_summary, function(X) return(X$coefficient)), get.est))))
+}
+all_estimates <- do.call(rbind, lapply(summary_out, extract.summary))
+
+## Output table
+summary_table <- data.frame(matrix(NA, ncol = 8, nrow = 8))
+colnames(summary_table) <- c("model", "predictor", "median slope", "5% slope", "95% slope", "median p.val", "5% p.val", "95% p.val")
+summary_table$model <- c(rep("all", 4), rep("macro", 2), rep("mega", 2))
+summary_table$predictor <- rep(c("species elaboration", "species innovation", "orders elaboration", "orders innovation"), 2)
+summary_table[, 3:5] <- t(apply(all_estimates[, c(4:7,12,13,16,17)], 2, quantile, probs = c(0.5, 0.05, 0.95)))
+summary_table[, 6:8] <- t(apply(all_estimates[, -c(1:3,4:7,12,13,16,17)], 2, quantile, probs = c(0.5, 0.05, 0.95)))
+
+```

### 2023-04-19 Analysis/04-Figure_with_ellipse__alternative.Rmd
Create 04-Figure_with_ellipse__alternative.Rmd

--- 2023-04-10 Analysis/04-Figure_with_ellipse_scalling_and_8D_histograms.Rmd
@@
 
-TODO:
- - [ ] split in two figures
- - [ ] make E/I boxes larger
- - [ ] make E/I colours always orange/blue
-
 ```{r, eval = TRUE, echo = FALSE}
@@
 }
+```
+
+```{r}
+pdf(file = "test.pdf", height = 11.693, width = 8.268)
+## Plot splitting
+x <- 82
+super_orders <- rbind(
+      #"Paleognathae" = "Tinamiformes",  
+      matrix(c(rep(1+x,  1), rep(2+x,  1), rep(3+x,  1)), 1, 3),
+      #"Galloanserae" = c("Galliformes", "Anseriformes"),
+      matrix(c(rep(4+x,  2), rep(5+x,  2), rep(6+x,  2)), 2, 3),
+      #"Mirandornithes" = c("Podicipediformes"),
+      matrix(c(rep(7+x,  1), rep(8+x,  1), rep(9+x,  1)), 1, 3),
+      #"Columbimorphae" = c("Columbiformes", "Pterocliformes"),
+      matrix(c(rep(10+x, 2), rep(11+x, 2), rep(12+x, 2)), 2, 3),
+      #"Otidimorphae" = c("Otidiformes", "Cuculiformes", "Musophagiformes"),
+      matrix(c(rep(13+x, 3), rep(14+x, 3), rep(15+x, 3)), 3, 3),
+      #"Gruiformes" (NOT A SUPER ORDER)
+      matrix(0, 1, 3),
+      #"Aequornithes" = c("Procellariiformes", "Sphenisciformes", "Ciconiiformes", "Pelecaniformes", "Suliformes"),
+      matrix(c(rep(16+x, 5), rep(17+x, 5), rep(18+x, 5)), 5, 3),
+      #"Strisores" = c("Caprimulgiformes", "Apodiformes"),
+      matrix(c(rep(19+x, 2), rep(20+x, 2), rep(21+x, 2)), 2, 3),
+      #"Charadriiformes" (NOT A SUPER ORDER)
+      matrix(0, 1, 3),
+      #"Telluraves" = c("Trogoniformes", "Bucerotiformes", "Coraciiformes", "Piciformes", "Strigiformes", "Accipitriformes", "Falconiformes", "Psittaciformes", "Passeriformes")
+      matrix(c(rep(22+x, 9), rep(23+x, 9), rep(24+x, 9)), 9, 3))
+
+layout_matrix <- cbind( 
+                matrix(1, 27, 1), # first column
+                matrix(seq(1:(27*3))+1, 27, 3, byrow = TRUE), # second/third column
+                super_orders)
+
+widths_vals <- c(5, 1, 1, 1, 1, 1, 1)
+height_vals <- rep(1, 27)
+layout_spiro <- layout(layout_matrix, widths = widths_vals, heights = height_vals, respect = TRUE)
+# layout.show(layout_spiro)
+
+## Plotting the tree
+tip_order <- c(
+    "Tinamiformes"= 1,
+    "Galliformes"= 3,
+    "Anseriformes"= 2,
+    "Otidiformes"= 7,
+    "Cuculiformes" = 8,
+    "Gruiformes" = 9, 
+    "Musophagiformes" = 10,
+    "Procellariiformes" = 12,
+    "Sphenisciformes" = 11,
+    "Ciconiiformes" = 13,
+    "Pelecaniformes"= 15,
+    "Suliformes" = 14,
+    "Podicipediformes" = 4,
+    "Columbiformes" = 6,
+    "Pterocliformes" = 5,
+    "Charadriiformes" = 18,
+    "Trogoniformes" = 21,
+    "Bucerotiformes" = 22,
+    "Coraciiformes" = 23,
+    "Piciformes" = 24,
+    "Strigiformes" = 20,
+    "Accipitriformes" = 19, 
+    "Falconiformes" = 25,
+    "Psittaciformes" = 26,
+    "Passeriformes" = 27,
+    "Caprimulgiformes" = 16, 
+    "Apodiformes" = 17)
+
+## Plotting the tree and getting the tip orders
+tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
+
+## Plotting each ellipse
+for(i in 1:27) {
+    wrap.plot.ellipses(i, tip_colours_orders, order_shapespace, results$order_results_between, ei.col = c("orange", "blue"), phylo.plot = TRUE)
+}
+load(file = "../Data/Processed/tip_colours_super_orders.rda")
+for(i in 1:8) {
+    ## Plot each pannel
+    wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between, ei.col = c("orange", "blue"), phylo.plot = TRUE)
+}
+dev.off()
+
+# ## Plotting the ellipses for each order
+# for(i in 1:length(tip_order)) {
+#     ## Plot each pannel
+#     wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between, ei.col = c("orange", "blue"))
+# }
+
 ```

### 2023-04-21 Analysis/04-Figure_with_ellipse__alternative.Rmd
Update 04-Figure_with_ellipse__alternative.Rmd

--- 2023-04-10 Analysis/04-Figure_with_ellipse_scalling_and_8D_histograms.Rmd
@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(
+superorder_shapespace <- MCMCglmm.subsets(n = 100,
       data          = shapespace,
@@
 
-order_shapespace <- MCMCglmm.subsets(
+order_shapespace <- MCMCglmm.subsets(n = 100,
       data          = shapespace,
@@
 }
+```
+
+```{r}
+pdf(file = "../Manuscript/Figures/figure_phylo_spiro.pdf", height = 11.693, width = 8.268)
+## Plot splitting
+x <- 82
+super_orders <- rbind(
+      #"Paleognathae" = "Tinamiformes",  
+      matrix(c(rep(1+x,  1), rep(2+x,  1), rep(3+x,  1)), 1, 3),
+      #"Galloanserae" = c("Galliformes", "Anseriformes"),
+      matrix(c(rep(4+x,  2), rep(5+x,  2), rep(6+x,  2)), 2, 3),
+      #"Mirandornithes" = c("Podicipediformes"),
+      matrix(c(rep(7+x,  1), rep(8+x,  1), rep(9+x,  1)), 1, 3),
+      #"Columbimorphae" = c("Columbiformes", "Pterocliformes"),
+      matrix(c(rep(10+x, 2), rep(11+x, 2), rep(12+x, 2)), 2, 3),
+      #"Otidimorphae" = c("Otidiformes", "Cuculiformes", "Musophagiformes"),
+      matrix(c(rep(13+x, 3), rep(14+x, 3), rep(15+x, 3)), 3, 3),
+      #"Gruiformes" (NOT A SUPER ORDER)
+      matrix(0, 1, 3),
+      #"Aequornithes" = c("Procellariiformes", "Sphenisciformes", "Ciconiiformes", "Pelecaniformes", "Suliformes"),
+      matrix(c(rep(16+x, 5), rep(17+x, 5), rep(18+x, 5)), 5, 3),
+      #"Strisores" = c("Caprimulgiformes", "Apodiformes"),
+      matrix(c(rep(19+x, 2), rep(20+x, 2), rep(21+x, 2)), 2, 3),
+      #"Charadriiformes" (NOT A SUPER ORDER)
+      matrix(0, 1, 3),
+      #"Telluraves" = c("Trogoniformes", "Bucerotiformes", "Coraciiformes", "Piciformes", "Strigiformes", "Accipitriformes", "Falconiformes", "Psittaciformes", "Passeriformes")
+      matrix(c(rep(22+x, 9), rep(23+x, 9), rep(24+x, 9)), 9, 3))
+
+layout_matrix <- cbind( 
+                matrix(1, 27, 1), # first column
+                matrix(seq(1:(27*3))+1, 27, 3, byrow = TRUE), # second/third column
+                super_orders)
+
+widths_vals <- c(5, 1, 1, 1, 1, 1, 1)
+height_vals <- rep(1, 27)
+layout_spiro <- layout(layout_matrix, widths = widths_vals, heights = height_vals, respect = TRUE)
+# layout.show(layout_spiro)
+
+## Plotting the tree
+tip_order <- c(
+    "Tinamiformes"= 1,
+    "Galliformes"= 3,
+    "Anseriformes"= 2,
+    "Otidiformes"= 7,
+    "Cuculiformes" = 8,
+    "Gruiformes" = 9, 
+    "Musophagiformes" = 10,
+    "Procellariiformes" = 12,
+    "Sphenisciformes" = 11,
+    "Ciconiiformes" = 13,
+    "Pelecaniformes"= 15,
+    "Suliformes" = 14,
+    "Podicipediformes" = 4,
+    "Columbiformes" = 6,
+    "Pterocliformes" = 5,
+    "Charadriiformes" = 18,
+    "Trogoniformes" = 21,
+    "Bucerotiformes" = 22,
+    "Coraciiformes" = 23,
+    "Piciformes" = 24,
+    "Strigiformes" = 20,
+    "Accipitriformes" = 19, 
+    "Falconiformes" = 25,
+    "Psittaciformes" = 26,
+    "Passeriformes" = 27,
+    "Caprimulgiformes" = 16, 
+    "Apodiformes" = 17)
+
+## Plotting the tree and getting the tip orders
+tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
+
+## Plotting each ellipse
+for(i in 1:27) {
+    wrap.plot.ellipses(i, tip_colours_orders, order_shapespace, results$order_results_between, ei.col = c("orange", "blue"), phylo.plot = TRUE)
+}
+
+#TODO:
+ # - change the E/I thermos to scale with the amount of E/I <- maybe?
+ # - change the E/I thermos to look less clutered (remove the box and labels), one tick = 5 units
+
+load(file = "../Data/Processed/tip_colours_super_orders.rda")
+for(i in 1:8) {
+    ## Plot each pannel
+    wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between, ei.col = c("orange", "blue"), phylo.plot = TRUE)
+}
+dev.off()
+
+# ## Plotting the ellipses for each order
+# for(i in 1:length(tip_order)) {
+#     ## Plot each pannel
+#     wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between, ei.col = c("orange", "blue"))
+# }
+
 ```

### 2023-04-21 Functions/ploting.utilities.R
Update ploting.utilities.R

@@ -81,6 +81,8 @@ plot.one.ellipse <- function(data, name.col, x = TRUE, y = TRUE, main = "", lege
 
     ## Do the empty plot
     if(lims[1] == "dynamic") {
+
+
         ## detect the limits
         ellipse_data <- get.subsets(data, subsets = subsets)
         covars <- sample.n(ellipse_data$covar, n = 100)
@@ -100,6 +102,27 @@ plot.one.ellipse <- function(data, name.col, x = TRUE, y = TRUE, main = "", lege
         ## Get the scaled ranges       
         xlims <- range(c(all_ellipses[[1]][[1]][, 1], all_ellipses[[2]][[1]][, 1]))
         ylims <- range(c(all_ellipses[[1]][[1]][, 2], all_ellipses[[2]][[1]][, 2]))
+        # cat(paste0(names(name.col), "\n"))
+        # cat(paste0("xlims = c(", paste0(xlims, collapse = ", "), ")\ndiff = ", diff(xlims), "\n"))
+        # cat(paste0("ylims = c(", paste0(ylims, collapse = ", "), ")\ndiff = ", diff(ylims), "\n"))
+
+        ## hard coded range
+        hard_range <- 0.45
+        if(hard_range > 0) {
+            ## Spread the x range
+            x_range_diff <- hard_range - diff(xlims)
+            if(x_range_diff > 0) {
+                xlims[1] <- xlims[1] - (x_range_diff/2)
+                xlims[2] <- xlims[2] + (x_range_diff/2)
+            }
+            ## Spread the y range
+            y_range_diff <- hard_range - diff(ylims)
+            if(x_range_diff > 0) {
+                ylims[1] <- ylims[1] - (y_range_diff/2)
+                ylims[2] <- ylims[2] + (y_range_diff/2)
+            }
+        }
+
     } else {
         xlims <- ylims <- lims
     }

### 2023-04-24 Manuscript/revisions/linking_metrics.Rmd
Create linking_metrics.Rmd

---
title: "Linking different metrics to Elaboration and Innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

## Elaboration and Innovation definitions

For any species with coordinates $\vec{species}$ in a $n$ dimensional space, we can estimate $VCV_{superorder}$ and $VCV_{order}$, respectively two nested variance-covariance matrices representing the any order and super order that contains $\vec{species}$.
We can then calculate the eigen vector and eigenvalues from these matrices as:

$\textbf{v}_{superorder}$ and $\lambda_{superorder}$ defined as:
$$VCV_{superorder} \textbf{v}_{superorder} = \lambda_{superorder} \textbf{v}_{superorder}$$

and $\textbf{v}_{order}$ and $\lambda_{order}$ defined as:
$$VCV_{order} \textbf{v}_{order} = \lambda_{order} \textbf{v}_{order}$$

In practice, these are solved using a eigen decomposition in `R` `eigen`.

We can then define the followings:

$$\text{Elaboration}_{\text{species}} = \|\frac{\vec{species} \cdot \textbf{v}_{order}} {\sqrt{\textbf{v}_{order} \cdot \textbf{v}_{order}}}\|$$

$$\text{Innovation}_{\text{species}} = \| \vec{species} - \text{Elaboration}_{\text{species}} \|$$

$$\text{Elaboration}_{\text{clade}} = \| \frac{\textbf{v}_{order} \cdot \textbf{v}_{superorder}} {\sqrt{\textbf{v}_{superorder} \cdot \textbf{v}_{superorder}}}\|$$

$$\text{Innovation}_{\text{clade}} = \| \textbf{v}_{order} - \text{Elaboration}_{\text{clade}} \|$$

## Relative eigenvalues dispersion

$$V_{rel} = \frac{\sum_{i}^{n}(\lambda_{i} - \sum{\frac{\lambda_{i}}{n}})^2}{n(n-1)\sum{\frac{\lambda_{i}}{n}}^2}$$
Which is the dispersion of eigenvalues (squared difference between each eigen value and the average eigen value).

```{r, eval = FALSE}
## Relative eigenvalue dispersion
rel.eigenvalue.dispersion <- function(matrix) {
    ## Getting the eigen values, n and mean eigen value
    lambdas <- eigen(var(matrix))$values
    n <- length(lambdas)
    lambda_mean <- mean(lambdas)
    ## Calculating the relative eigen dispersion
    return(sum((lambdas - lambda_mean)^2)/n*(n-1)*lambda_mean)
}

## Calculating the Elaboration/Innovation for clades and species
data(charadriiformes)

## Creating a dispRity object with a covar component
my_covar <- MCMCglmm.subsets(
                 data       = charadriiformes$data,
                 posteriors = charadriiformes$posteriors,
                 tree       = charadriiformes$tree,
                 group      = MCMCglmm.levels(
                                 charadriiformes$posteriors)[1:4],
                 rename.groups = c("gulls", "plovers", "sandpipers", "phylo"))

## Clade level
clade   <- dispRity.covar.projections(my_covar, type = "groups", base = "phylo", n = 100,  output = c("position", "distance"))
species <- dispRity.covar.projections(my_covar, type = "elements", n = 100,  output = c("position", "distance"))
eigen_dispersion <- dispRity(my_covar, metric = rel.eigenvalue.dispersion)
```

dispersion of eigenvalues of a covariance or correlation matrix

 - [ ] link orthogonality to random skewers
 - [ ] link E/I to relative eigenvector dispersion (Pavlicev et al., 2009, Watanabe, 2022) 
 - [ ] link E/I to conditional evolvability (Hansen & Houle, 2008) 
 - [ ] link E/I to integration (Hansen & Houle, 2008)

### 2023-04-25 Manuscript/revisions/linking_metrics.Rmd
Update linking_metrics.Rmd

@@
 
-## Relative eigenvalues dispersion
-
-$$V_{rel} = \frac{\sum_{i}^{n}(\lambda_{i} - \sum{\frac{\lambda_{i}}{n}})^2}{n(n-1)\sum{\frac{\lambda_{i}}{n}}^2}$$
-Which is the dispersion of eigenvalues (squared difference between each eigen value and the average eigen value).
-
-```{r, eval = FALSE}
-## Relative eigenvalue dispersion
-rel.eigenvalue.dispersion <- function(matrix) {
-    ## Getting the eigen values, n and mean eigen value
-    lambdas <- eigen(var(matrix))$values
-    n <- length(lambdas)
-    lambda_mean <- mean(lambdas)
-    ## Calculating the relative eigen dispersion
-    return(sum((lambdas - lambda_mean)^2)/n*(n-1)*lambda_mean)
-}
+```{r}
+library(dispRity)
 
@@
 ## Clade level
-clade   <- dispRity.covar.projections(my_covar, type = "groups", base = "phylo", n = 100,  output = c("position", "distance"))
+clade   <- dispRity.covar.projections(my_covar, type = "groups", base = "phylo", n = 1000,  output = c("position", "distance"))
 species   <- dispRity.covar.projections(my_covar, type = "elements", n = 100,  output = c("position", "distance"))
-eigen_dispersion <- dispRity(my_covar, metric = rel.eigenvalue.dispersion)
 ```
 
-dispersion of eigenvalues of a covariance or correlation matrix
+## Relative eigenvalues dispersion (Watanabe, 2022)
+
+$$V_{rel} = \frac{\sum_{i}^{n}(\lambda_{i} - \sum{\frac{\lambda_{i}}{n}})^2}{n(n-1)\sum{\frac{\lambda_{i}}{n}}^2}$$
+Which is the dispersion of eigenvalues (squared difference between each eigen value and the average eigen value).
+
+```{r, eval = TRUE}
+## Relative eigenvalue dispersion
+rel.eigenvalue.dispersion <- function(matrix, get.var = FALSE) {
+    if(get.var) {
+        matrix <- var(matrix)
+    }
+    ## Getting the eigen values, n and mean eigen value
+    lambdas <- eigen(matrix)$values
+    n <- length(lambdas)
+    lambda_mean <- mean(lambdas)
+    ## Calculating the relative eigen dispersion
+    return(sum((lambdas - lambda_mean)^2)/n*(n-1)*lambda_mean)
+}
+
+## Calculate the eigen dispersions
+eigen_dispersion <- dispRity(my_covar, metric = as.covar(rel.eigenvalue.dispersion))
+
+## Get the values
+clade_disp <- get.disparity(eigen_dispersion, concatenate = FALSE)
+clade_elab <- get.disparity(clade$position, concatenate = FALSE)
+clade_inov <- get.disparity(clade$distance, concatenate = FALSE)
+
+par(mfrow = c(1,2))
+plot(NULL, xlim = range(unlist(clade_disp)), ylim = range(unlist(clade_elab)), xlab = "Eigen dispersion", ylab = "Elaboration")
+points(clade_disp$gulls, clade_elab$gulls, pch = 19, col = "darkgreen")
+points(clade_disp$sandpipers, clade_elab$sandpipers, pch = 19, col = "blue")
+points(clade_disp$plovers, clade_elab$plovers, pch = 19, col = "orange")
+legend("topright", pch = 19, col = c("darkgreen", "blue", "orange"), legend = c("gulls", "sandpipers", "plovers"))
+
+plot(NULL, xlim = range(unlist(clade_disp)), ylim = range(unlist(clade_inov)), xlab = "Eigen dispersion", ylab = "Innovation")
+points(clade_disp$gulls, clade_inov$gulls, pch = 19, col = "darkgreen")
+points(clade_disp$sandpipers, clade_inov$sandpipers, pch = 19, col = "blue")
+points(clade_disp$plovers, clade_inov$plovers, pch = 19, col = "orange")
+```
+
+## Conditional evolvability (Hansen & Houle, 2008)
+
+If species can be interpreted as "the response to selection in $n$ traits", then:
+
+$$\text{Evolvability} = \text{Elaboration}_{\text{species}}$$
+
+$$\text{Respondability} = \sqrt{\text{Elaboration}_{\text{species}}^2 + \text{Innovation}_{\text{species}}^2}$$
+
+Conditional evolvability = evolvability * autonomy
+Autonomy = fraction of genetic variation that is independent of potentially constraining characters = inverse of G matrix.
+
+```{r, eval = TRUE}
+## Relative eigenvalue dispersion
+respondability <- function(x, y) {
+    sqrt(x^2 + y^2)
+}
+
+## calculating the respondability
+species_elab <- get.disparity(species$position, concatenate = FALSE)
+species_inov <- get.disparity(species$distance, concatenate = FALSE)
+species_resp <- mapply(respondability, species_elab, species_inov, SIMPLIFY = FALSE)
+
+par(mfrow = c(1,2))
+plot(NULL, xlim = range(unlist(species_resp)), ylim = range(unlist(species_elab)), xlab = "Respondability", ylab = "Elaboration (evolvability)")
+points(species_resp$plovers, species_elab$plovers, pch = 19, col = "orange", cex = 0.5)
+points(species_resp$sandpipers, species_elab$sandpipers, pch = 19, col = "blue", cex = 0.5)
+points(species_resp$gulls, species_elab$gulls, pch = 19, col = "darkgreen", cex = 0.5)
+legend("topright", pch = 19, col = c("darkgreen", "blue", "orange"), legend = c("gulls", "sandpipers", "plovers"))
+
+plot(NULL, xlim = range(unlist(species_resp)), ylim = range(unlist(species_inov)), xlab = "Respondability", ylab = "Innovation")
+points(species_resp$plovers, species_inov$plovers, pch = 19, col = "orange", cex = 0.5)
+points(species_resp$sandpipers, species_inov$sandpipers, pch = 19, col = "blue", cex = 0.5)
+points(species_resp$gulls, species_inov$gulls, pch = 19, col = "darkgreen", cex = 0.5)
+```
 
  - [ ] link orthogonality to random skewers
- - [ ] link E/I to relative eigenvector dispersion (Pavlicev et al., 2009, Watanabe, 2022) 
- - [ ] link E/I to conditional evolvability (Hansen & Houle, 2008) 
+ - [x] link E/I to relative eigenvector dispersion (Pavlicev et al., 2009, Watanabe, 2022) 
+ - [x] link E/I to conditional evolvability (Hansen & Houle, 2008) 
  - [ ] link E/I to integration (Hansen & Houle, 2008)

### 2023-04-27 Analysis/07-Orthogonality_and_statistics.Rmd
Update 07-Orthogonality_and_statistics.Rmd

@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 source("../Functions/stats.utilities.R")
@@
 ```
+
+## Random skewers
+
+```{r}
+## Random skewers function wrapper for dispRity
+skewers.fun <- function(matrix, matrix2) {
+    return(phytools::skewers(matrix, matrix2)$r)
+}
+skewers.fun.p <- function(matrix, matrix2) {
+    return(phytools::skewers(matrix, matrix2)$p)
+}
+p.star <- function(x) {
+    if(x > 0.1) {
+        return("")
+    } else {
+        if(x > 0.05) {
+            return(".")
+        } else {
+            if(x > 0.01) {
+                return("*")
+            } else {
+                if(x > 0.001) {
+                    return("**")
+                } else {
+                    return("***")
+                }
+            }
+        }
+    }
+}
+
+## Getting the same order as the orthogonality data
+ortho_order <- c(
+    "Columbimorphae",
+    "Galloanserae",
+    "Mirandornithes",
+    "Telluraves",
+    "Strisores",
+    "Paleognathae",
+    "Aequornithes",
+    "Otidimorphae",
+    "Galliformes",
+    "Accipitriformes",
+    "Strigiformes",
+    "Musophagiformes",
+    "Trogoniformes",
+    "Columbimorphae",
+    "Caprimulgiformes",
+    "Bucerotiformes",
+    "Falconiformes",
+    "Pterocliformes",
+    "Sphenisciformes",
+    "Podicipediformes",
+    "Ciconiiformes",
+    "Tinamiformes",
+    "Suliformes",
+    "Anseriformes",
+    "Apodiformes",
+    "Otidiformes",
+    "Cuculiformes",
+    "Psittaciformes",
+    "Gruiformes",
+    "Pelecaniformes",
+    "Procellariiformes",
+    "Piciformes",
+    "Passeriformes",
+    "Coraciiformes",
+    "Charadriiformes")
+
+## Getting the group pairs
+group_pairs <- sapply(name.subsets(all_birds_shapespace)[-36], function(x) c(x, "phylogeny"), simplify = FALSE)
+
+## Calculating the random skewers for each group
+skewers_r <- dispRity(all_birds_shapespace, metric = as.covar(skewers.fun), between.groups = group_pairs, verbose = TRUE)
+skewers_p <- dispRity(all_birds_shapespace, metric = as.covar(skewers.fun.p), between.groups = group_pairs, verbose = TRUE)
+
+## Extract disparity for fancy plotting
+skewers_distrib <- lapply(get.disparity(skewers_r, concatenate = FALSE), quantile, probs = c(0.5, 0.025, 0.25, 0.75, 0.975))
+skewers_median_p <- lapply(get.disparity(skewers_p, concatenate = FALSE), median)
+plot_names <- unlist(strsplit(names(skewers_distrib), split = ":phylogeny"))
+names(skewers_distrib) <- names(skewers_median_p) <- plot_names
+skewers_distrib <- skewers_distrib[ortho_order]
+skewers_median_p <- skewers_median_p[ortho_order]
+plot_names <- names(skewers_median_p)
+
+col <- rev(c(rep(gg.color.hue(4)[1], 8), rep(gg.color.hue(4)[2], 27)))
+
+par(mar = c(5, 17, 4, 2))
+plot(NULL, xlim = c(0, 1), ylim = c(1,length(skewers_distrib)), xlab = "Skewers correlation", ylab = "", yaxt = "n")
+abline(v = 0.5, lty = 2, col = "grey", lwd = 0.5)
+axis(2, at = length(skewers_distrib):1, labels = plot_names, las = 2)
+## Skewers
+for(i in length(skewers_distrib):1) {
+    lines(x = skewers_distrib[i][[1]][c(2, 5)], y = rep(i, 2), lty = 2, lwd = 1, col = col[i])
+    lines(x = skewers_distrib[i][[1]][c(3, 4)], y = rep(i, 2), lty = 1, lwd = 2, col = col[i])
+    points(x = skewers_distrib[i][[1]][1], y = i, pch = 19, col = col[i])
+    text(x = skewers_distrib[i][[1]][1]+0.03, y = i+0.2, label = p.star(skewers_median_p[[i]]), col = col[i])
+}
+
+```

### 2023-05-03 Analysis/13-mcmcmcglmmm.ppca.Rmd
Create 13-mcmcmcglmmm.ppca.Rmd

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

# Comparing results of `mcmcmcglmmm` to phylo-pca results

## Comparing phylo pca to just pca

```{r}
library(phytools)
data(anoletree)
data(anole.data)

## Phylo PCA
phyl_pca <- phyl.pca(anoletree, anole.data)

## Normal PCA
norm_pca <- prcomp(anole.data)

## Comparing the loadings
scree_phyl <- apply(phyl_pca$S, 2, var)/sum(apply(phyl_pca$S, 2, var))
scree_norm <- apply(norm_pca$x, 2, var)/sum(apply(norm_pca$x, 2, var))

## Anolis example

### 2023-05-11 Analysis/04-Figure_with_ellipse__alternative.Rmd
Update 04-Figure_with_ellipse__alternative.Rmd

@@
       #"Galloanserae" = c("Galliformes", "Anseriformes"),
-      matrix(c(rep(4+x,  2), rep(5+x,  2), rep(6+x,  2)), 2, 3),
+      matrix(c(rep(4+x,  1), rep(5+x,  1), rep(6+x,  1)), 1, 3),
+      matrix(0, 1, 3),
       #"Mirandornithes" = c("Podicipediformes"),
@@
       #"Columbimorphae" = c("Columbiformes", "Pterocliformes"),
-      matrix(c(rep(10+x, 2), rep(11+x, 2), rep(12+x, 2)), 2, 3),
+      matrix(c(rep(10+x, 1), rep(11+x, 1), rep(12+x, 1)), 1, 3),
+      matrix(0, 1, 3),
       #"Otidimorphae" = c("Otidiformes", "Cuculiformes", "Musophagiformes"),
-      matrix(c(rep(13+x, 3), rep(14+x, 3), rep(15+x, 3)), 3, 3),
+      matrix(0, 1, 3),
+      matrix(c(rep(13+x, 1), rep(14+x, 1), rep(15+x, 1)), 1, 3),
+      matrix(0, 1, 3),
       #"Gruiformes" (NOT A SUPER ORDER)
@@
       #"Aequornithes" = c("Procellariiformes", "Sphenisciformes", "Ciconiiformes", "Pelecaniformes", "Suliformes"),
-      matrix(c(rep(16+x, 5), rep(17+x, 5), rep(18+x, 5)), 5, 3),
+      matrix(0, 2, 3),
+      matrix(c(rep(16+x, 1), rep(17+x, 1), rep(18+x, 1)), 1, 3),
+      matrix(0, 2, 3),
       #"Strisores" = c("Caprimulgiformes", "Apodiformes"),
-      matrix(c(rep(19+x, 2), rep(20+x, 2), rep(21+x, 2)), 2, 3),
+      matrix(c(rep(19+x, 1), rep(20+x, 1), rep(21+x, 1)), 1, 3),
+      matrix(0, 1, 3),
       #"Charadriiformes" (NOT A SUPER ORDER)
       matrix(0, 1, 3),
+      matrix(0, 4, 3),
       #"Telluraves" = c("Trogoniformes", "Bucerotiformes", "Coraciiformes", "Piciformes", "Strigiformes", "Accipitriformes", "Falconiformes", "Psittaciformes", "Passeriformes")
-      matrix(c(rep(22+x, 9), rep(23+x, 9), rep(24+x, 9)), 9, 3))
+      matrix(c(rep(22+x, 1), rep(23+x, 1), rep(24+x, 1)), 1, 3),
+      matrix(0, 4, 3))
 
@@
 tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
-save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
+# save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
 

### 2023-05-11 Analysis/05-Figure_with_phylogeny.Rmd
Update 05-Figure_with_phylogeny.Rmd

@@
 library(treeio)
+# install_github("uyedaj/treeplyr")
 library(treeplyr)
+library(dplyr)
 library(ggplot2)
@@
                       "Elab (order)", "Innov (order)")
+## Just orders
+heatmapdat <- beak[,c(1,7,8)]
+colnames(heatmapdat) <- c("species", "Elaboration", "Innovation")
 

### 2023-05-11 Analysis/07-Orthogonality_and_statistics.Rmd
Update 07-Orthogonality_and_statistics.Rmd

@@
   scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
-  scale_fill_manual(values = c("goldenrod", "#1E65A7")) +
+  scale_fill_manual(values = c("yellow", "purple")) +
   xlab("Relative elaboration") +
@@
   scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
-  scale_fill_manual(values = c("goldenrod", "#1E65A7")) +
+  scale_fill_manual(values = c("yellow", "purple")) +
   xlab("Relative innovation") + 

### 2023-05-11 Analysis/06-Figure_with_correlations.Rmd
Update 06-Figure_with_correlations.Rmd

@@
 
-y axis = levels
-column 1 = correlation of e/i projected on the mega level
-column 2 = correlation of e/i projected on the macro level
-stars for significant correlation between means.
-
 ```{r, echo = TRUE, eval = TRUE}
@@
                 y.axis = TRUE, plot.phylo = TRUE,
-                main = "Elaboration and innovation\nat the mega level") 
+                main = "Species projected onto\nthe whole phylogeny VCV") 
 par(mar = c(5, 2, 4, 6) + 0.1, bty = "n") # c(bottom, left, top, right)
@@
                 y.axis = FALSE, plot.phylo = FALSE,
-                main = "Elaboration and innovation\nat the macro level")
+                main = "Species projected onto\ntheir superorders' VCV")
 dev.off()

### 2023-05-11 Manuscript/first_submission.tex
Create first_submission.tex

--- 2022-11-21 Manuscript/main_flipped_narrative.tex
@@
 
-Reconciling major changes in phenotypic diversity at macroevolutionary scales with gradual microevolutionary changes continues to perplex scientists.
-A major question is whether widely documented megaevolutionary jumps can emerge from gradual phenotypic elaboration and innovation.
-Here we apply new multi-trait models to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+Widely documented, megaevolutionary jumps in diversity continue to perplex researchers because it remains unclear whether these dramatic changes can emerge from microevolutionary processes.
+Here we reconcile major changes in phenotypic diversity at macroevolutionary scales with gradual microevolutionary change using new multi-trait models to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
 We find that elaboration dominates at macroevolutionary scales whereas innovation becomes more prominent at megaevolutionary scales.
-The major axis of phenotypic change among species at megaevolutionary scales is an emergent property of innovation across clades.
-We suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
-
-% change Gustavo's paper to /in press in current biology/
-% Check if the ruminant paper is somewhere
+Indeed, the major axis of phenotypic change among species at megaevolutionary scales is an emergent property of innovation across clades.
+Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
 
@@
 
-Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive,Rhoda2022}.
-Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
+Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
+Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,renaud2006conserved,fasanelli2022allometry}.
 Numerous studies show that phenotypic divergence is biased along the evolutionary line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
 Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
-However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{BurinWhales,jones2004evolution}.
+However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution,BurinWhales}.
 The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,pagel2022general,khabbazian2016fast,cooney2017mega,smaers2021evolution,goswami2022} suggests a conflict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
@@
 We acknowledge IT Services at The University of Sheffield for the provision of the High Performance Computing Service.
-We thank the museum in Tring.
-% TG: TODO: just take Chris' paper acknowledgments
+We thank M. Adams, H. van Grouw, R. Prys-Jones and and Zo\"{e} K. Varley from the Bird Group at the NHM, Tring.
+We are indebted to the volunteer citizen scientists at http://www.markmybird.org for helping to build the database of bird beak shape and contribute to our understanding of avian evolution. 
 \textbf{Funding.}
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
-% TG: TODO: add ERC grant (ask Gavin)
+This work was funded by the European Research Council (615709 Project ‘ToLERates’), UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to G.H.T.).
 \textbf{Author contributions.}
-G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., G.H.T. and Z.K.V. curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
+G.H.T., A.P.B., N.C., and T.G. designed the study. J.A.B., C.R.C., E.C.H., and G.H.T. % and Z.K.V.
+curated the data. T.G. analysed the data. All authors contributed to the writing of the manuscript.
 \textbf{Competing interests.}

### 2023-05-11 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/cheat_sheet.pdf}
-\caption{\scriptsize Relation between the figures in the main text.
-The top left panel represents the two first dimensions of the 8 dimensional trait space with all bird beaks represented as grey circles and the Palaeognathes as red ones.
-The grey ellipse represents the average posterior variance covariance matrix (VCV) from the general phylogenetic random term of the MCMCglmm model (i.e.
-the overall phylogenetic axis of main variation).
-Species aligned with this axis tend to be elaborators at the mega-evolutionary level and species going away from that line are innovators at the mega-evolutionary level.
-The red ellipse and axis represents the average posterior VCV for the Palaeognathes only.
-Species position on this axis elaborate and innovate at the macro-evolutionary level.
-On the top right panel (Figure 1 in the main text), we measured the position of each species in terms of elaboration and innovation.
-A species elaboration score is their projection onto the phylogenetics' VCV major axis (i.e.
-their position on that axis) whereas their innovation score is their rejection from it (i.e.
-their distance away from that axis).
-On the bottom right panel (Figure 2 in the main text), we measured the elaboration and innovation at the entire clade level (cf. at the species level below).
-The ellipses are a scaled and centered in space representation of the group’s VCV ellipse relative to the overall phylogenetic one with a the relative length of the ellipse on each dimension represented on the barplots.
-The clade's innovation and elaboration scores are measured by projecting the clade's major axis onto the phylogenetic one and measuring the projection (elaboration) and rejection (innovation) as mentioned above.
-The distribution of innovation and elaboration scores comes from the projection of the 4000 pairs of posterior VCVs.}
-
-\label{Fig:cheat_sheet}
-\end{figure}
-
 \begin{figure}[htbp]
@@
 \label{Fig:model_ess_passeriformes}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/correlations.pdf}
+\caption{Correlation plots of $elaboration_{species}$ and $innovation_{species}$ at different focal levels.
+\textbf{A}: correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny.
+The colours of the dots correspond to their density in the plot.
+Blue and yellow colours corresponds to respectively low and high density of species.
+The coloured quadrant in the back corresponds to $elaboration_{species}$ and $innovation_{species}$ scores higher than 1.
+The blue and yellow quadrants shows species with noticeable high $elaboration_{species}$ and $innovation_{species}$ respectively.
+The species in the green quadrant have noticeable high levels of both $elaboration_{species}$ and $innovation_{species}$.
+\textbf{B}: correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny (mega-level projection) but plotted by super-order and coloured by order (see list below).
+\textbf{C} correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on their super-orders' major axis of variation (macro-level projection).
+The colours correspond to each order.
+For the distribution of correlation scores, see Fig.
+\ref{Fig:fable_correlations}.
+The orders numbers are:
+1 - others (orders with $<$ 15 species)
+2 - Corvoidea;
+3 - Malaconotoidea;
+4 - Orioloidea;
+5 - Meliphagoidea;
+6 - Bombycilloidea;
+7 - Muscicapoidea;
+8 - Emberizoidea;
+9 - Motacillidae;
+10 - Nectariniidae ;
+11 - Passeridae;
+12 - PloceidaeEstrildidae;
+13 - Eurylaimides;
+14 - Furnariida;
+15 - Tyrannida;
+16 - Aegithaloidea;
+17 - Cisticolidae;
+18 - Fringillidae;
+19 - Hirundinidae;
+20 - Locustelloidea;
+21 - Paridae;
+22 - Pycnonotidae;
+23 - Sylvioidea;}
+\label{Fig:figure_correlations}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/correlations_fable.pdf}
+\caption{Posterior distributions of the correlations scores between $elaboration_{species}$ and $innovation_{species}$ as displayed for each super-orders in Fig.
+\ref{Fig:figure_correlations}.
+Species' projection onto the whole phylogeny's major axis of phenotypic variance-covariance (VCV) corresponds to the correlation scores for the panels A and B in Fig.
+\ref{Fig:figure_correlations} and species' projections onto their own super-order's major axis of phenotypic variance-covariance (VCV) corresponds to the correlation scores for the panel C in Fig.
+\ref{Fig:figure_correlations}}
+\label{Fig:fable_correlations}
 \end{figure}

### 2023-05-11 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 \ref{Fig:figure_correlations} and species' projections onto their own super-order's major axis of phenotypic variance-covariance (VCV) corresponds to the correlation scores for the panel C in Fig.
-\ref{Fig:figure_correlations}}
+\ref{Fig:figure_correlations}.}
 \label{Fig:fable_correlations}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/Elaboration_and_innovation_through_time.pdf}
+\caption{Distribution of $elaboration_{clade}$ and $innovation_{clade}$ for each order and super-order through time.
+Refer to the colours in Fig.
+3 in the main text for the names of each order and super-order plotted.}
 \end{figure}

### 2023-05-12 Analysis/04-Figure_with_ellipse__alternative.Rmd
grey version

--- 2023-04-21 Analysis/04-Figure_with_ellipse__alternative.Rmd
@@
 ```{r}
-pdf(file = "../Manuscript/Figures/figure_phylo_spiro.pdf", height = 11.693, width = 8.268)
+pdf(file = "../Manuscript/Figures/figure_phylo_spiro_grey.pdf", height = 11.693, width = 8.268)
 ## Plot splitting
@@
 ## Plotting the tree and getting the tip orders
-tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
-save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
+# tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
+# save(tip_colours_orders, file = "../Data/Processed/tip_colours_orders.rda")
+
+## Non coloured version
+grey.palette <- function(n) {return(rep("grey", n))}
+tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order, colour.palette = black.palette)
+tip_colours_orders[1:length(tip_colours_orders)] <- "black"
 
@@
 load(file = "../Data/Processed/tip_colours_super_orders.rda")
+
+## Black version
+tip_colours_super_orders[1:length(tip_colours_super_orders)] <- "black"
+## Grey version
+# tip_colours_super_orders <- rep("darkgrey", length(tip_colours_super_orders))
+
 for(i in 1:8) {

### 2023-05-12 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
   \hline
+macroevolutionary scale    & analysis where the focal elements are the species & & \\
+megaevolutionary scale     & analysis where the focal elements are clades (i.e. groups of species) & & \\
 major axis of variation    & the major evolutionary ``direction'' of traits co-evolution & first eigenvector of the variance-covariance matrix  & allometry & line of genetic least resistance \\

### 2023-05-15 Analysis/14-Elab_innov_through_time.Rmd
Update 14-Elab_innov_through_time.Rmd

@@
 ## Table that for easy plotting
-super_orders_elaboration <- cbind(super_orders_elaboration, super_age_info, cex = 2)
-super_orders_innovation <- cbind(super_orders_innovation, super_age_info, cex = 2)
-orders_elaboration <- cbind(orders_elaboration, order_age_info, cex = 1)
-orders_innovation <- cbind(orders_innovation, order_age_info, cex = 1)
+super_orders_elaboration <- cbind(super_orders_elaboration, super_age_info, pch = 19)
+super_orders_innovation <- cbind(super_orders_innovation, super_age_info, pch = 19)
+orders_elaboration <- cbind(orders_elaboration, order_age_info, pch = 21)
+orders_innovation <- cbind(orders_innovation, order_age_info, pch = 21)
 elaboration_plot_data <- rbind(super_orders_elaboration, orders_elaboration)
@@
 ## Empty plot
-plot.age.data <- function(data, ylab, xlab) {
-  plot(NULL, xlab = xlab, ylab = ylab,
+plot.age.data <- function(data, ylab, xlab, main = "") {
+  plot(NULL, xlab = xlab, ylab = ylab, main = main,
        xlim = rev(range(pretty(range(data$age)))),
-       ylim = range(pretty(range(data[,c(4, 8)]))))
+       ylim = range(pretty(range(data[c("2.5%", "25%", "75%", "97.5%")]))))
   ## Adding the lines for each clade
@@
     lines(x = rep(data[i, "age"][[1]], 2), y = data[i, c("25%", "75%")], lty = 1, lwd = 2, col = data[i, "col"][[1]])
-    points(x = data[i, "age"][[1]], y = data[i, "obs.median"], cex = data[i, "cex"], col = data[i, "col"][[1]], pch = 19)
+    points(x = data[i, "age"][[1]], y = data[i, "obs.median"], pch = data[i, "pch"], col = data[i, "col"][[1]])
   }
 }
-pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_through_time.pdf")
+
+pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_clade_through_time.pdf")
 par(mfrow = c(2,1))
-plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "")
-legend("topleft", pch = 19, col = "grey", legend = c("super orders", "orders"), pt.cex = c(2, 1))
+plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "", main = "Clade's elaboration/innovation\nthrough time")
+legend("topleft", pch = 21, col = unlist(orders_elaboration$col), legend = rownames(orders_elaboration), cex = 0.7)
+legend("top", pch = 19, col = unlist(super_orders_elaboration$col), legend = rownames(super_orders_elaboration), cex = 0.7)
 plot.age.data(innovation_plot_data, ylab = "Innovation", xlab = "Time (Mya)")
@@
 ```
+
+# Checking elaboration and innovation values through time (species)
+
+```{r, eval = TRUE, echo = TRUE}
+## Get all the statistics
+super_orders_elaboration <- summary(results$super_results_within$position)[-9,]  #-9 is for removing the "phylogeny" element
+super_orders_innovation <- summary(results$super_results_within$distance)[-9,]
+orders_elaboration <- summary(results$order_results_within$position)[-28,] #-28 is for removing the "phylogeny" element
+orders_innovation <- summary(results$order_results_within$distance)[-28,]
+
+## Get all the tree ages
+load("../Data/Processed/ploting_data.rda")
+load("../Data/Processed/clade_info.rda")
+
+super_age_info <- t(sapply(unlist(strsplit(super_orders_elaboration$subsets, split = ":phylogeny"))[-9], get.clade.age, info = clade_info$super_orders, tree = ploting_data$consensus_tree)) #-9 is for removing the "phylogeny" element
+order_age_info <- t(sapply(unlist(strsplit(orders_elaboration$subsets, split = ":phylogeny"))[-28], get.clade.age, info = clade_info$orders, tree = ploting_data$consensus_tree)) #-28 is for removing the "phylogeny" element
+
+## Table that for easy plotting
+super_orders_elaboration <- cbind(super_orders_elaboration, super_age_info, pch = 19)
+super_orders_innovation <- cbind(super_orders_innovation, super_age_info, pch = 19)
+orders_elaboration <- cbind(orders_elaboration, order_age_info, pch = 21)
+orders_innovation <- cbind(orders_innovation, order_age_info, pch = 21)
+elaboration_plot_data <- rbind(super_orders_elaboration, orders_elaboration)
+innovation_plot_data <- rbind(super_orders_innovation, orders_innovation)
+
+pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_species_through_time.pdf")
+par(mfrow = c(2,1))
+plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "", main = "Species's elaboration/innovation\nthrough time (per clade)")
+legend("topright", pch = 21, col = unlist(orders_elaboration$col), legend = unlist(orders_elaboration$subsets), cex = 0.7)
+legend("top", pch = 19, col = unlist(super_orders_elaboration$col), legend = unlist(super_orders_elaboration$subsets), cex = 0.7)
+plot.age.data(innovation_plot_data, ylab = "Innovation", xlab = "Time (Mya)")
+dev.off()
+```

### 2023-05-16 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 dev.off()
+
+pdf(file = "../Manuscript/Figures/50_random_axes_superorders.pdf", height = 8, width = 8)
+load("../Data/Processed/tip_colours_super_orders.rda")
+tip_colours_super_orders <- c(tip_colours_super_orders, "phylogeny" = "grey")
+covar.plot(superorder_shapespace,
+           transparent.scale = 1/3,
+           n = 50,
+           col = tip_colours_super_orders[name.subsets(superorder_shapespace)],
+           main = "50 random scaled major axes per super order",
+           major.axes = TRUE,
+           legend = TRUE,
+           points = FALSE,
+           scale = "phylogeny",
+           cex = 1)
+dev.off()
+
 ```

### 2023-05-16 Analysis/15-linking_metrics.Rmd
Create 15-linking_metrics.Rmd

--- 2023-04-25 Manuscript/revisions/linking_metrics.Rmd
@@
 
+pdf("../Manuscript/Figures/elaboration_innovation_dispersion.pdf", width = 12, height = 6)
 par(mfrow = c(1,2))
@@
 points(clade_disp$plovers, clade_inov$plovers, pch = 19, col = "orange")
+dev.off()
 ```
@@
 
+pdf("../Manuscript/Figures/respondability.pdf", width = 12, height = 6)
 par(mfrow = c(1,2))
@@
 points(species_resp$gulls, species_elab$gulls, pch = 19, col = "darkgreen", cex = 0.5)
-legend("topright", pch = 19, col = c("darkgreen", "blue", "orange"), legend = c("gulls", "sandpipers", "plovers"))
+legend("topleft", pch = 19, col = c("darkgreen", "blue", "orange"), legend = c("gulls", "sandpipers", "plovers"))
 
@@
 points(species_resp$gulls, species_inov$gulls, pch = 19, col = "darkgreen", cex = 0.5)
+dev.off()
 ```
-
- - [ ] link orthogonality to random skewers
- - [x] link E/I to relative eigenvector dispersion (Pavlicev et al., 2009, Watanabe, 2022) 
- - [x] link E/I to conditional evolvability (Hansen & Houle, 2008) 
- - [ ] link E/I to integration (Hansen & Houle, 2008)

### 2023-05-16 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
    \includegraphics[width=0.9\textwidth]{Figures/axis_selection.pdf}
-\caption{variance and cumulative variance for each 8 axis per super-order.}
+\caption{Variance and cumulative variance for each 8 axis per super-order.
+Note that not all super orders have their own variance explained matching the variance explained by the entire ordination.
+For example, Strisores only need the first three dimensions to explain at least 95\% of their variance within the shapespace whilst Aequornithes need 8 dimensions to achieve at least the same amount of variance.}
 \label{Fig:axes_variance}
@@
 \begin{figure}[htpb]
-\centering
 \centering
@@
 \centering
-\centering
    \includegraphics[width=0.5\textwidth]{Figures/correlations_fable.pdf}
@@
 \centering
+   \includegraphics[width=0.5\textwidth]{Figures/Elaboration_and_innovation_clade_through_time.pdf}
+\caption{Distribution of $elaboration_{clade}$ and $innovation_{clade}$ for each order and super-order through time.}
+\label{Fig:figure_ei_clade_through_time}
+\end{figure}
+
+\begin{figure}[htpb]
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/Elaboration_and_innovation_through_time.pdf}
-\caption{Distribution of $elaboration_{clade}$ and $innovation_{clade}$ for each order and super-order through time.
-Refer to the colours in Fig.
-3 in the main text for the names of each order and super-order plotted.}
+   \includegraphics[width=0.5\textwidth]{Figures/Elaboration_and_innovation_species_through_time.pdf}
+\caption{Distribution of $elaboration_{species}$ and $innovation_{species}$ for each order and super-order through time.}
+\label{Fig:figure_ei_species_through_time}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/50_random_axes_superorders.pdf}
+\caption{Distribution of 50 random scaled major axes (scaled variance-covariance matrices eigen values) for each super order illustrating that individual major axes can span in many different directions with some clades have a higher spread of direction than others.}
+\label{Fig:mikado}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/random_skewers.pdf}
+\caption{Random skewers test for the differences between the super orders (in purple) and the orders (in gold) variance-covariance matrices against the whole phylogeny ones.
+And, in light gold, between the orders and their super-orders variance covariance matrices.
+The different line types (dashed and full) represent respectively the 95\% and the 75\% CI distribution and the dots represent the median skewer correlation estimate.
+The stars denote the median p-value estimate (0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1).}
+\label{Fig:random_skewers}
 \end{figure}
@@
 The detailed procedure was adapted from \href{https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331}{Zheyuan Li's} post on Stack Overflow and implemented in \texttt{dispRity::axis.covar}.
+
+The use of the scaled 95\% confidence interval hyper-sphere matrix $O'_{n}$ allows both (1) to standardise the resulting $M_{n}$ (the scaled version of $VCV_{n}$) so that the unit vector of the resulting space corresponds to the 95\% CI on every dimension (for any dimension $i$ in the matrix $M_{i,n}$, any vector with a norm smaller or greater than 1 is respectively inside or outside the 95\% CI boundary).
+And (2) greatly speed up the calculations of projection and rejection in an algorithm (see below) since the hypersphere $O'_{n}$ can be calculated only once for any number of matrix and the projection of rejection values are directly columns in the output transformed matrix (after the projection procedure, see below).
+This greatly reduces RAM and CPU usage and scaling for very large number of matrices.
 
@@
 
+\subsection{Definition of species and clade innovation and elaboration}
+
+From the definitions above, we can define $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$, the elaboration and innovation at the macroevolutionary (species level) and $\text{elaboration}_{\text{clade}}$ and $\text{innovation}_{\text{clade}}$, the elaboration and innovation at the megaevolutionary level (clade).
+
+For any species with coordinates $\vec{species}$ in a $n$ dimensional space, we can estimate $VCV_{clade}$, the variance-covariance matrices representing the clade that contains $\vec{species}$.
+We can then calculate the eigen vector and eigenvalues from these matrices as:
+
+$\textbf{v}_{clade}$ defined as:
+
+\begin{equation}
+VCV_{clade} \textbf{v}_{clade} = \lambda_{clade} \textbf{v}_{clade}
+\end{equation}
+
+In practice, these are solved using a eigen decomposition in \texttt{R} \texttt{base::eigen}.
+
+We can then define the followings for any species $x$:
+
+\begin{equation}
+\text{elaboration}_{\text{species}_{x}} = \|\frac{\vec{species}_{x} \cdot \textbf{v}_{clade}} {\sqrt{\textbf{v}_{clade} \cdot \textbf{v}_{clade}}}\|
+\end{equation}
+
+With $$\vec{species}_{x}$$ being the vector in $n$ dimensional space for species $x$. And $$\textbf{v}_{clade}$$ being the eigen vector of a specific clade (e.g. $$\textbf{v}_{clade_{c}}$$)
+
+\begin{equation}
+\text{innovation}_{\text{species}_{x}} = \| \vec{species}_{x} - \text{elaboration}_{\text{species}_{x}} \|
+\end{equation}
+
+And the followings for any clade $c$ (withing a parent clade $c-1$):
+
+\begin{equation}
+\text{elaboration}_{\text{clade}_{c}} = \| \frac{\textbf{v}_{clade_{c}} \cdot \textbf{v}_{clade_{c-1}}} {\sqrt{\textbf{v}_{clade_{c-1}} \cdot \textbf{v}_{clade_{c-1}}}}\|
+\end{equation}
+
+\begin{equation}
+\text{innovation}_{\text{clade}_{c}} = \| \textbf{v}_{clade_{c}} - \text{elaboration}_{\text{clade}_{c}} \|
+\end{equation}
+
+\subsection{Relation between elaboration and innovation and metrics in micro/macro-evolution}
+
+\subsubsection{Relative eigenvalues dispersion}
+
+\cite{vanvalen1974,watanabe2022} propose to describe a variance-covariance matrix in terms of its eigen dispersion $V_{rel}$ as:
+
+\begin{equation}
+V_{rel} = \frac{\sum_{i}^{n}(\lambda_{i} - \sum{\frac{\lambda_{i}}{n}})^2}{n(n-1)\sum{\frac{\lambda_{i}}{n}}^2}
+\end{equation}
+
+Which is the dispersion of eigenvalues (squared difference between each eigen value and the average eigen value) where $\lambda_{i}$ are each individual eigen values and $n$ is the number of dimensions considered.
+
+Using this definition and the definitions above of $\text{elaboration}_{\text{clade}_{c}}$ and $\text{innovation}_{\text{clade}_{c}}$ we can calculate the and plot the relation between elaboration and innovation and their relative eigen dispersion:
+
+\begin{figure}[htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/elaboration_innovation_dispersion.pdf}
+\caption{Relation between relative eigen dispersion \cite{vanvalen1974,watanabe2022} and $\text{elaboration}_{\text{clade}_{c}}$ and $\text{innovation}_{\text{clade}_{c}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
+\label{Fig:eigen_dispersion}
+\end{figure}
+
+\subsubsection{Evolvability and respondability}
+
+\cite{hansen2008measuring} propose to measure the evolvability and respondability of an individual as their ``response to selection in $n$ traits''.
+Here, some equivalent can be drawn (albeit cautiously) between the response of an individual or a group (population) to selection pressure and a species.
+Arguably, a species is the end product of this (and many other) selection pressures on a group of individuals.
+If so, then, we can relate $\text{evolvability}$ and $\text{respondability}$ \cite{hansen2008measuring} for a $species_{x}$ to it's $\text{elaboration}_{\text{species}_{x}}$ and $\text{innovation}_{\text{species}_{x}}$ as follows:
+
+\begin{equation}
+\text{evolvability}_{\text{species}_{x}} = \text{elaboration}_{\text{species}_{x}}
+\end{equation}
+
+\noindent and, using Pythagorean theorem in a triangle rectangle:
+
+\begin{equation}
+\text{respondability}_{\text{species}_{x}}  = \sqrt{\text{elaboration}_{\text{species}_{x}}^2 + \text{innovation}_{\text{species}_{x}}^2}
+\end{equation}
+
+\begin{figure}[htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/respondability.pdf}
+\caption{Relation between respondability \cite{hansen2008measuring} and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
+\label{Fig:respondability}
+\end{figure}
+
+Conditional evolvability however will be more complex to determine since it is defined as the relation between evolvability and autonomy which is specific defined in \cite{hansen2008measuring} as the inverse of the \textbf{G} matrix.
+In \cite{hansen2008measuring}, the inverse of \textbf{G} is defined as ``the fraction of the genetic variation that is independent of potentially constraining characters'' which in our case will be harder to define.
+Although it would technically be possible to use the inverse of the \textbf{R} matrix used here ($VCV_{clade}$), it is unclear what it would represent biologically.
+
 \bibliographystyle{naturemag}

### 2023-05-25 Analysis/04-Figure_with_ellipse_phyloview.Rmd
rename

--- 2023-05-12 Analysis/04-Figure_with_ellipse__alternative.Rmd
@@
 grey.palette <- function(n) {return(rep("grey", n))}
-tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order, colour.palette = black.palette)
-tip_colours_orders[1:length(tip_colours_orders)] <- "black"
+tip_colours_orders <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order, colour.palette = grey.palette)
+tip_colours_orders[1:length(tip_colours_orders)] <- "#191919"
 
@@
 ## Black version
-tip_colours_super_orders[1:length(tip_colours_super_orders)] <- "black"
+tip_colours_super_orders[1:length(tip_colours_super_orders)] <- "#191919"
 ## Grey version
-# tip_colours_super_orders <- rep("darkgrey", length(tip_colours_super_orders))
+# tip_colours_super_orders <- rep("#191919", length(tip_colours_super_orders))
 

### 2023-05-30 Analysis/06-Figure_with_correlations.Rmd
Update 06-Figure_with_correlations.Rmd

@@
 super_group_cor <- get.correlations.plot(results$super_results_within, exclude.phylo = TRUE)
+
+## Check for order_results_within
+order_correlations <- get.correlations.plot(results$order_results_within, exclude.phylo = TRUE)
 ```
@@
 ## Get all the correlations for all the clades
-cor_mega_level <- get.correlations.scores(results$super_results_phylo)
-cor_macro_level <- get.correlations.scores(results$super_results_within)
+cor_mega_super_orders <- get.correlations.scores(results$super_results_phylo)
+cor_macro_super_orders <- get.correlations.scores(results$super_results_within)
+cor_macro_orders <- get.correlations.scores(results$order_results_within)
 
@@
 ## Plot the correlations
-pdf("../Manuscript/Figures/correlations_fable.pdf")
+pdf("../Manuscript/Figures/correlations_fable_super_orders.pdf")
 par(mfrow = c(1,2))
 par(mar = c(5, 8, 4, 0) + 0.1, bty = "n") # c(bottom, left, top, right)
-plot.cor.scores(cor_mega_level, col = super_orders_col,
+plot.cor.scores(cor_mega_super_orders, col = super_orders_col,
                 y.axis = TRUE, plot.phylo = TRUE,
-                main = "Species projected onto\nthe whole phylogeny VCV") 
+                main = "Species projected onto\nthe whole phylogeny major axes") 
 par(mar = c(5, 2, 4, 6) + 0.1, bty = "n") # c(bottom, left, top, right)
-plot.cor.scores(cor_macro_level, col = super_orders_col,
+plot.cor.scores(cor_macro_super_orders, col = super_orders_col,
                 y.axis = FALSE, plot.phylo = FALSE,
-                main = "Species projected onto\ntheir superorders' VCV")
+                main = "Species projected onto\ntheir superorders' major axes")
+dev.off()
+
+## Correlation fable
+pdf(width = 5, height = 9, file = "../Manuscript/Figures/correlations_fable_orders.pdf")
+load("../Data/Processed/tip_colours_orders.rda")
+orders_col <- c(tip_colours_orders, "phylogeny" = "grey")
+orders_col[] <- "black"
+par(mar = c(5, 8, 4, 0) + 0.1, bty = "n") # c(bottom, left, top, right)
+plot.cor.scores(cor_macro_orders, col = orders_col,
+                y.axis = TRUE, plot.phylo = FALSE,
+                show.phylo = FALSE, xlim = c(-0.5, 1),
+                main = "Species projected onto\ntheir orders' major axes") 
 dev.off()
@@
     plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
+}
+dev.off()
+
+## Extra figure
+pdf("../Manuscript/Figures/correlations_2.pdf", height = 20/3*1.3, width = 8*1.3)
+par(mfrow = c(5,6))
+for(i in 1:26) {
+    par(mar = c(2,4,2.5,0)+0.1)
+    plot.correlations(order_correlations[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
 }

### 2023-05-30 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
    \includegraphics[width=0.5\textwidth]{Figures/correlations.pdf}
-\caption{Correlation plots of $elaboration_{species}$ and $innovation_{species}$ at different focal levels.
+\caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ at different focal levels.
 \textbf{A}: correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny.
@@
 \ref{Fig:fable_correlations}.
-The orders numbers are:
-1 - others (orders with $<$ 15 species)
+The groups numbers are:
+1 - others (groups with $<$ 15 species)
 2 - Corvoidea;
@@
 22 - Pycnonotidae;
-23 - Sylvioidea;}
+23 - Sylvioidea.}
 \label{Fig:figure_correlations}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/correlations_2.pdf}
+\caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ for each species projected on their order's major axes.
+Each point is the median $elaboration_{species}$ and $innovation_{species}$ for one species from the distribution of 4000 posterior variance-covariance matrices.
+The full distribution of the correlation between $elaboration_{species}$ and $innovation_{species}$ is available in the main text in figure @@@.}
+\label{Fig:figure_correlations2}
 \end{figure}
@@
 
-\noindent and, using Pythagorean theorem in a triangle rectangle:
+\noindent and, using Pythagorean theorem in a right angled triangle:
 

### 2023-05-31 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 
+We define three evolutionary scales based on the focal elements as follow:
+\begin{itemize}
+   \item \textbf{micro}evolutionary scale: the focal elements are \textbf{individual specimen};
+   \item \textbf{macro}evolutionary scale: the focal elements are \textbf{species} (i.e. groups of individuals);
+   \item \textbf{mega}evolutionary scale: the focal elements are \textbf{clades} (i.e. groups of species).
+\end{itemize}
+
 \begin{table}[ht]
@@
   \hline
-macroevolutionary scale    & analysis where the focal elements are the species & & \\
-megaevolutionary scale     & analysis where the focal elements are clades (i.e. groups of species) & & \\
 major axis of variation    & the major evolutionary ``direction'' of traits co-evolution & first eigenvector of the variance-covariance matrix  & allometry & line of genetic least resistance \\
@@
 variance-covariance matrix & the multidimensional relation between traits & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) & & \\
-nested phylogenetic level  & the different \textbf{clades} used in this study (the whole phylogeny $>$ super-orders $>$ orders) & different nested \textbf{variance-covariance matrices} & & \\
+nested phylogenetic level  & the different \textbf{clades} used in this study (the class-wide phylogeny $>$ super-orders $>$ orders) & different nested \textbf{variance-covariance matrices} & & \\
 
@@
 \section{Supplementary material 2: additional results}
+
+\subsection{Data preparation}
 
@@
 \end{figure}
+
+\subsection{Additional results}
+
+\subsubsection{Correlations between $elaboration_{species}$ and $innovation_{species}$}
 
@@
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/correlations_2.pdf}
+   \includegraphics[width=0.9\textwidth]{Figures/correlations_2.pdf}
 \caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ for each species projected on their order's major axes.
 Each point is the median $elaboration_{species}$ and $innovation_{species}$ for one species from the distribution of 4000 posterior variance-covariance matrices.
-The full distribution of the correlation between $elaboration_{species}$ and $innovation_{species}$ is available in the main text in figure @@@.}
+The full distribution of the correlation between $elaboration_{species}$ and $innovation_{species}$ is available in the main text in figure 3.}
 \label{Fig:figure_correlations2}
@@
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/correlations_fable.pdf}
+   \includegraphics[width=0.9\textwidth]{Figures/correlations_fable.pdf}
 \caption{Posterior distributions of the correlations scores between $elaboration_{species}$ and $innovation_{species}$ as displayed for each super-orders in Fig.
@@
 
+\subsubsection{Elaboration and innovation through time}
+
 \begin{figure}[htpb]
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/Elaboration_and_innovation_clade_through_time.pdf}
+   \includegraphics[width=1\textwidth]{Figures/Elaboration_and_innovation_clade_through_time.pdf}
 \caption{Distribution of $elaboration_{clade}$ and $innovation_{clade}$ for each order and super-order through time.}
@@
 \centering
-   \includegraphics[width=0.5\textwidth]{Figures/Elaboration_and_innovation_species_through_time.pdf}
+   \includegraphics[width=1\textwidth]{Figures/Elaboration_and_innovation_species_through_time.pdf}
 \caption{Distribution of $elaboration_{species}$ and $innovation_{species}$ for each order and super-order through time.}
 \label{Fig:figure_ei_species_through_time}
+\end{figure}
+
+\subsubsection{Orientations in $elaboration_{clade}$ and $innovation_{clade}$}
+
+\begin{figure}[htpb]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/50_random_axes_superorders.pdf}
+\caption{Distribution of 50 random scaled major axes (scaled variance-covariance matrices eigen values) for each super order illustrating that individual major axes can span in many different directions with some clades have a higher spread of direction than others.}
+\label{Fig:mikado}
+\end{figure}
+
+\begin{figure}[htpb]
+\centering
+   \includegraphics[width=0.8\textwidth]{Figures/ellipses_supplementary.pdf}
+\caption{\scriptsize{Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
+We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all super-order ellipses relative to the global phylogenetic ellipse.
+Subsequent inset plots show the results for each super-order. 
+Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+The inset boxplots correspond to the $elaboration_{clade}$ (E) and $innovation_{clade}$ (I) scores for all 4000 posterior samples.
+The dots represent the median $elaboration_{clade}$ and $innovation_{clade}$ values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
+These scores were calculated on the unscaled ellipses resulting in different scales of $elaboration_{clade}$ and $innovation_{clade}$ for each plot.
+Panel B) shows the results for each order.}}
+\label{Fig:ellipses_rainbow}
 \end{figure}

### 2023-06-01 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 \subsection{Additional results}
+
+\begin{figure}[htbp]
+\centering
+   \includegraphics[width=0.7\textwidth]{Figures/InnovElabTree_supp_revision.pdf}
+\caption{Avian phylogeny (n = 8748 species) showing distributions of species beak shape elaboration_{species} (inner circle, yellow scale), innovation_{species} (outer circle, blue scale), and Euclidean distance of species to the centroid of beak space (branches, cividis scale).}
+\label{Fig:fig_phylogeny_supplement}
+\end{figure}
 

### 2023-06-01 Manuscript/main.tex
Updated introduction. Big time!

--- 2022-09-06 Manuscript/main.tex
@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
+\noindent {\normalsize \sc
+Thomas Guillerme$^{1,*}$, 
+Jen A. Bright$^{2}$,
+Christopher R. Cooney${^1}$,
+Emma C. Hughes${^1}$,
+% Zoe K. Varley${^3}$,
+Natalie Cooper$^{3,+}$,
+Andrew P. Beckerman$^{1,+}$,
+and Gavin H. Thomas$^{1,4,+}$}\\
 \noindent {\small \it 
 $^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
+$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
+$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
+$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
 
@@
 
-\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
+\noindent (Keywords: macroevolution, elaboration, innovation, evolutionary rate matrix, line of least resistance, GLMM)\\
 
@@
 
-Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
-A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
-At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
-The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance \cite{endler2005animal, renaud2006conserved}.
-Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
-Our analyses show that elaboration is a common feature at all scales, consistent with theory.
-We also find that innovation is a common and major contributor to avian morphological diversity among clades.
-Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
-These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
+Widely documented, megaevolutionary jumps in phenotypic diversity continue to perplex researchers because it remains unclear whether these dramatic changes can emerge from microevolutionary processes.
+Here we tackle this question using new approaches for modeling multivariate traits %@L_multi_traits
+ to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
+We find that elaboration, evolution along the major axis of phenotypic change, is common at both macro- and megaevolutionary scales whereas innovation, evolution away from the major axis of phenotypic change, is more prominent at megaevolutionary scales.
+Indeed, the major axis of phenotypic change among species beak shapes at megaevolutionary scales is an emergent property of innovation across clades.
+Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
 
-\section{Main}
+\section{Introduction}
 
-Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow lines of genetic least resistance \cite{schluter1996adaptive}.
-Over longer timescales, the line of least evolutionary resistance is observable as the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry} and is an emergent property of genomic and developmental constraints interacting with selection on the adaptive landscape \cite{jones2004evolution}.
-Indeed, evidence suggests that genetic constraints can extend over timescales spanning tens of millions of years leading to remarkable macroevolutionary stability in the orientation of phenotypic evolution \cite{mcglothlin2018adaptive}.
-However, if most organisms evolve incrementally along stable lines of least resistance then how can we explain the staggering diversity of species morphologies across the tree of life?
-The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the evolvability of phenotypes across scales.
-This apparent tension implies deep-time reorientation of phenotypic lines of least resistance and points to alternative routes to the exploration of phenotypic space by evolving either along, or away from, the dominant phenotypic axis.
-We refer to these alternatives as elaboration - evolution along the phylogenetically constrained phenotypic major axis - and innovation - evolution away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
-The contributions of these two multivariate pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales.
-Here we estimate and compare major axes of beak shape evolution in birds across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.
+Understanding patterns and constraints in the adaptive evolution of species traits is a major goal of evolutionary research.
+At macroevolutionary scales, renewed attention has been given to variation in the tempo and mode of trait evolution among lineages \cite{venditti2011multiple, khabbazian2016fast, castiglione2018, hansen2022, larouche2022mosaic, pagel2022general, navalon2022environmental, basu2023tempo}.
+This has led to growing recognition of the importance of major jumps or discontinuities in evolutionary rates as drivers of diversification of species traits \cite{cooney2017mega, smaers2021evolution, goswami2022}.
+For example, in both bird beaks and mammal skulls there are major jumps in the rate of trait evolution early in clade history as morphological diversity expands \cite{cooney2017mega, goswami2022} with lower magnitude jumps towards the present.
+Major jumps in species trait values are consistent with the concept of megaevolution, describing major discontinuities in phenotypes observed from the fossil record, first proposed by Simpson.
 
-Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
-1) large, multivariate datasets to allow exploration of trait covariances at different scales,
-2) reliable and efficient computational methods to estimate the lines of least resistance in a nested fashion,
-and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
-Here, we propose a novel analytical pipeline for measuring elaboration and innovation at the species- and higher taxonomic levels (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
-All of our analyses use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-We fitted a phylogenetic generalised linear mixed effect model (pGLMM) with beak shape as an eight-dimensional response variable and treated phylogeny as a nested random term including a phylogenetic partition and multiple partitions for nested clades (see Methods; \cite{MCMCglmm}).
-We regard the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
-The nested partitions of multivariate trait space provide the basis for all subsequent analyses.
+Megaevolutionary dynamics appear to contrast with theory based on microevolutionary processes.
+Over microevolutionary timescales, phenotypic evolution is expected to follow lines of least resistance \cite{schluter1996adaptive}.
+The term ‘line of least resistance’ has been used both specifically to describe phenotypic evolution along the axis of greatest genetic variation \cite{schluter1996adaptive}, and more loosely to describe the direction of phenotypic evolution with greatest observable variation \cite{marroig2005size}.
+Regardless of the definition used, the microevolutionary line of least resistance represents the direction with the most potential for evolution because it contains the most variation on which selection can act \cite{hansen2016, Rhoda2022, jablonski2022evolvability}.
+This expectation is supported by numerous studies showing that phenotypic divergence is often biased along (i.e.
+aligned with) the line of least resistance \cite{marroig2005size, bolstad2014, walter2018, mcglothlin2018adaptive, tsuboi2018, Rhoda2022, mongle2022}.
+These studies suggest that alignment of phenotypic divergence with the line of least resistance may be common over comparatively short (1-2 million year) timescales.
+More generally, the alignment of macroevolutionary divergence with the microevolutionary line of least resistance is expected to decline over time \cite{mcglothlin2018adaptive}.
+However, there is evidence, for example from Anolis lizards, that stability in the direction of phenotypic evolution can extend over timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
 
-Adaptive radiation theory implies that phenotypic divergence arises primarily along a dominant phenotypic axis (the line of least evolutionary resistance; \cite{marroig2005size}).
-However, this contrasts with observations of major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
-We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different groups compared to the global phylogenetic major axis.
-We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis (Fig.
-\ref{Fig:ellipses} - Fig.
-\ref{Fig:orthogonality}).
-Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis and each super order (Fig.
-\ref{Fig:ellipses}a) and order (Fig.
-\ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig.
-\ref{Fig:orthogonality}).
-The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
-Comparisons of orientation among subclades (i.e.
-orders within superorders) show similar differences (median = $47.17^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
-This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
-These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.
+The juxtaposition of microevolutionary predictions to macro- and mega-evolutionary observations is striking and raises the question of how phenotypic divergence accumulates across evolutionary scales.
+In particular, how can megaevolutionary jumps emerge from gradual micro- or macro-evolutionary processes of phenotypic evolution implied by microevolutionary predictions?
+\cite{hansen2016,jablonski2022evolvability}
+The original descriptions of microevolutionary lines of least resistance explicitly refer to genetic constraints.
+However, at the macroevolutionary scale the major axis of phenotypic variation is an emergent property of broader set of genetic and developmental constraints interacting with selection on a moving adaptive landscape % GT: Masahito suggested a whole paragraph on adaptive landscapes and Simpon-Lande-Arnold ideas.
+I'm not sure that will help but hopefully the expanded intro overall works better.
+Possibly just needs another sentence or two here but it would break the flow.
+more citations of relevant papers here (the first 7 or so papers from Gustavo's paper would be relevant.
+\cite{BurinWhales,jones2004evolution}. % GT: More references needed? I'm not sure where the idea originates.
+At this scale the major axis of phenotypic variation among species \cite{marroig2005size, fasanelli2022allometry} can be estimated from a matrix of divergences in species mean phenotypic traits (e.g.
+the \textbf{D} matrix of \cite{BlowsHiggie2003, mcglothlin2018adaptive}.
+These matrices do not take into account phylogenetic relationships among species.
+In contrast, the \textbf{R} matrix of \cite{Houle2017} describes the rate of among-species divergence in species traits and explicitly incorporates phylogeny (also see the \textbf{B} matrix of \cite{Machado2020}) %@L_Rmatrix.
+We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}. %TG: TODO: change throughout (that's the VCV matrix now)
+The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales.
+
+By studying the phylogenetic line of least resistance and related properties of the evolutionary rate matrix \cite{caetano2017ratematrix, caetano2018estimating, slater2019hierarchy} at different scales we aim to gain new insights into the macro- and mega-evolutionary trajectories that lead to the diversity of beak shapes among extant bird species and clades.
+While there is no strict definition that separates macro- and mega-evolution, we define the macroevolutionary scale as species divergence within clades and the megaevolutionary scale as divergence of clades within larger clades.
+The diversity of bird beaks has been well-studied at broad scales, particularly in relation to dietary ecology \cite{olsen2017feeding, felice2019, bright2016shapes, navalon2019, pigot2020macroevolutionary}.
+Previous studies have assessed how variation in rates of evolution of beak shape has generated this diversity \cite{cooney2017mega,chira2018correlates}, revealing a general pattern of phylogenetically clustered evolutionary rates interspersed with major shifts deep in avian evolutionary history and in isolated lineages.
+Major shifts indicate lineages that occupy distinct areas of beak shape space relative to their close relatives.
+For example, a major shift in beak shape arises in the early divergence of Strisores separating hummingbirds (Trochilidae) from swifts (Apodidae), perhaps marking changes in ontogenetic trajectories \cite{navalon2021} as hummingbirds diverged from broad billed ancestors (e.g.
+the Eocene stem Trochilidae fossils \textit{Parargornis messelensis} and \textit{Eocypselus rowei} \cite{mayr2003new, ksepka2013fossil}).
+By assessing the multivariate directions of the evolution of beak shapes of species and clades we aim to deepen our understanding of how diversity in morphological form arises at macro- and mega-evolutionary scales in birds.
+To do this we use data on bird beaks derived from 3D scans of museum specimens for $>$8700 species ($\sim$85\% of all extant bird species) and generate evolutionary rate matrices describing the directions of multivariate beak shape evolution.
+The phylogenetic breadth of the data allows us to compare directions of beak evolution of subclades (e.g.
+orders) within the class Aves (the megaevolutionary scale) and of species within subclades (the macroevolutionary scale).%@L_macromega_definition2
+
+Various approaches have been developed to measure properties of trait matrices \cite{hansen2008measuring, pavlicev2009measuring, watanabe2022}, broadly focused on describing the size, shape, and orientation of trait divergence \cite{phillips1999}.
+Our specific focus is on divergence of species and clades along or away from the phylogenetic line of least resistance.
+We adopt concepts developed in the specific context of signal evolution to describe these alternative routes to divergence \cite{endler2005animal}.
+Specifically, Endler \textit{et al}. (2005) refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
+This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, is expected to be a major driver of phenotypic evolution.
+In contrast, Endler \textit{et al}.
+(2005) refer to evolution away from the major axis of phenotypic variation, either in random directions or along limited axes, as \textit{innovation} \cite{endler2005animal}.
+Relative to elaboration, innovation is more likely to generate phenotypic novelty, but, if directions of phenotypic evolution are constrained, would not be expected to be a primary driver of phenotypic diversity among closely related species.
+We apply these terms to i) species divergence within clades, and ii) clade divergence within larger clades (Figure 1).
+Both innovation and elaboration could in principle occur at the macro- or mega-evolutionary scale.
+We expect that elaboration is more important at macroevolutionary scales (species diverging within clades) and innovation more important at megaevolutionary scale (divergence among clades).
+For our bird beak data, macroevolutionary scales refer to the beak shapes of species elaborating or innovating relative to the phylogenetic line of least resistance of their taxonomic order, and megaevolutionary scales refer to avian orders elaborating or innovating relative to the phylogenetic line of least resistance for the entire class Aves.
+For clarity throughout the manuscript, we refer to elaboration and innovation as the general concept described above but use the following terms to distinguish macroevolutionary scale (species) and megaevolutionary scale (clade) elaboration and innovation: $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$.
+We describe the algebraic formulation of elaboration and innovation for species and clades in detail in the Methods section.
+Using these novel statistical tools, we estimate the relative contribution of elaboration and innovation to the remarkable global radiation and diversity of bird beaks.
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
+   \includegraphics[width=1\textwidth]{Figures/cheat_sheet.pdf}
 \caption{.}
-\label{Fig:ellipses}
+\label{Fig:cheatsheet}
 \end{figure}
@@
 
-\noindent \textbf{Figure \ref{Fig:ellipses}:}
-Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
-We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
-The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
-Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
-The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
-The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
-Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
-A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
+\noindent \textbf{Figure \ref{Fig:cheatsheet}}
+Figure 1:
+Relationships between key figures in the main text.
+A) represents the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
+The gray ellipse represents the overall phylogenetic major axis of beak variation in birds.
+Species aligned or away from this axis are respectively elaborators or innovators relative to the class Aves.
+The red ellipse and axis represents the phylogenetic major axis of beak variation in Tinamiformes only.
+Species aligned or away from this axis are respectively elaborators and innovators relative to the Tinamiformes order.
+B) In Figure 2, we measured elaboration and innovation at the macroevolutionary scale , i.e.
+the position of each species in terms of $elaboration_{species}$ and $innovation_{species}$ relative to their order's phylogenetic major axis of beak variation.
+A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation(i.
+e.
+their position on that axis) whereas their $innovation_{species}$ score is their rejection from it (i.e.
+their distance away from that axis).
+The colours on the tips of the phylogeny correspond to the median $elaboration_{species}$ (orange gradient) and $innovation_{species}$ (blue gradient).
+C) Figure 4, we measured elaboration and innovation at the megaevolutionary scale, i.e. $elaboration_{clade}$ and $innovation_{clade}$.
+The ellipses are scaled and centered and represent the clade's phylogenetic major axis of beak variation relative to the overall phylogenetic major axis of beak variation with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
+A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis of beak variation onto the overall phylogenetic major axis of beak variation and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
+The distribution elaboration and innovation scores are taken from the projection of the 4000 pairs of evolutionary rate matrices with each pair being the focal level, e.g.
+order,and the parent level, e.g.
+the whole bird phylogeny.
 
-\bigskip
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
+% \caption{.}
+% \label{Fig:ellipses}
+% \end{figure}
+% \bigskip
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
-\caption{.}
-\label{Fig:orthogonality}
-\end{figure}
+% \noindent \textbf{Figure \ref{Fig:ellipses}:}
+% Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
+% We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
+% The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
+% Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+% The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
+% The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
+% Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
+% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
 
-\bigskip
+% \bigskip
 
-\noindent \textbf{Figure \ref{Fig:orthogonality}:}
-Orthogonality of each clade's major axis compared to their parent or parent parent's clade.
-Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
-Dots represent the clades median orthogonality and the lines their 95\% CI.
-No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
-These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
-Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
-For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
-The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
-Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
-This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
-A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
+% \caption{.}
+% \label{Fig:orthogonality}
+% \end{figure}
 
-\bigskip
+% \bigskip
 
-However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
-While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
-To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
-As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
-Clades (or species) can be both elaborators and innovators.
-To measure innovation and elaboration we translated the major axes of each group's (i.e.
+% \noindent \textbf{Figure \ref{Fig:orthogonality}:}
+% Orthogonality of each clade's major axis compared to their parent or parent parent's clade.
+% Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+% Dots represent the clades median orthogonality and the lines their 95\% CI.
+% No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
+% These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
+% Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+% For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
+% The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
+% Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
+% This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
+% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+
+% \bigskip
+
+% However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
+% While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
+% To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
+% As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
+% Clades (or species) can be both elaborators and innovators.
+% To measure innovation and elaboration we translated the major axes of each clade's (i.e.
 superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
-We then used multilinear algebra to project a focal group's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
-We interpret the measured projection (distance onto the phylogenetic major axis) as the group's elaboration score and the measured rejection (distance from the major axis) as the group's innovation score.
-These scores were measured for each of 4000 posterior pairs of group vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-We found that for most groups the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
-Remarkably, more than half of the assessed groups (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
-However, there is substantial variation in terms of elaboration and innovation across groups including relatively low elaboration and innovation (e.g.
+% We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
+% We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
+% These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
+% We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
+% Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
+% However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
 Cuculiformes; Fig.
@@
 \ref{Fig:ellipses}).
-These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
-This suggests that divergence away from line of least evolutionary resistance (i.e.
+% These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
+% This suggests that divergence away from line of least evolutionary resistance (i.e.
 the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
 
-This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
-Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
+% This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
+% Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
 \ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig.
 \ref{Fig:ellipses}-bar plots).
-These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
+% These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
 S\ref{fig_ellipses_passeriformes}).
-Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
-For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
-However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
+% Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
+% For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+% These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
+% However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
 
-Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
-This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
-To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shapespace onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
-We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire groups.
-We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
-However, we also found various combinations of elaboration and innovation across birds.
-For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
-Other groups such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
-Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
-The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
+% Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
+% This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
+% To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shapespace onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
+% We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire clades.
+% We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
+% However, we also found various combinations of elaboration and innovation across birds.
+% For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
+% Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
+% Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
+% The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
 \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
- and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
+%  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
+% The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
 
-\begin{figure}[!htbp]
-\centering
-    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
-\caption{.}
-\label{Fig:phylogeny}
-\end{figure}
+% \begin{figure}[!htbp]
+% \centering
+%     \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
+% \caption{.}
+% \label{Fig:phylogeny}
+% \end{figure}
 
-\bigskip
+% \bigskip
 
-\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
-Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
-Innovations and elaboration are displayed as: a) comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
-The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
-A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
+% \noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
+% Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
+% Innovations and elaboration are displayed as: a) comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
+% The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
+% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
 
-\bigskip
+% \bigskip
 
-The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
-We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
-We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
-We then measured the difference between both areas and found that there was a significant difference in innovation between species and groups but no clear difference in elaboration (Fig.
+% The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
+% We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
+% We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
+% We then measured the difference between both areas and found that there was a significant difference in innovation between species and clades but no clear difference in elaboration (Fig.
 \ref{Fig:relative_EI}).
-This difference confirms our observation that innovation is more common at the group level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
-However, we suggest that rather than a paradox the reorientation of trait space can instead be described as a gradual process occuring in response to shifting selection pressures acting on species evolving in similar adaptive zones.
-This implies that gradual directional evolution \cite{pagel2022general} rather than megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple} may be sufficient to explain diversity in avian beak morphology.
+% This difference confirms our observation that innovation is more common at the clade level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
+% However, we suggest that rather than a paradox the reorientation of trait space can instead be described as a gradual process occuring in response to shifting selection pressures acting on species evolving in similar adaptive zones.
+% This implies that gradual directional evolution \cite{pagel2022general} rather than megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple} may be sufficient to explain diversity in avian beak morphology.
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/relative_EI.pdf}
-\caption{.}
-\label{Fig:relative_EI}
-\end{figure}
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=0.9\textwidth]{Figures/relative_EI.pdf}
+% \caption{.}
+% \label{Fig:relative_EI}
+% \end{figure}
 
-\bigskip
+% \bigskip
 
-\noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among groups (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
-A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 groups.
-The scores for each group is scaled by the maximum score within each group (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
-This allows us to compare both types (species vs.
-groups) despite their different sample sizes.
-B - same as A but for the innovation score.
-C - density curve of the area under the curve for each posterior sample relative elaboration score as a proxy for the overall amount of elaboration per sample and per type (species or group - respectively blues and orange).
-The Bhattacharyya coefficient indicates the probability of overlap between the posterior distributions.
-D - same as C but for innovation.
-We can see a clear difference in the amount of innovation across the posterior samples between species or groups.
+% \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
+% A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 clades.
+% The scores for each clade is scaled by the maximum score within each clade (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
+% This allows us to compare both types (species vs. clades) despite their different sample sizes.
+% The Bhattacharryya Coefficient represents the overlap of the area under the curve for species and clades (a BC of 0 represents no overlap).
+% B - same as A but for the innovation score.
+% We can see a clear difference in the amount of innovation across the posterior samples between species or clades.
 
-\bigskip
+% \bigskip
 
-Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.
-This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
-However, individual species-level variation in the past could also have led to a shift in a group's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
+% Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.
+% This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
+% However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
 Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
-Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
+% Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
 
-\section{Methods}
+% \section{Methods}
 
-\textbf{Beak shape data.}
-We used the dataset from \cite{cooney2017mega,chira2020signature,hughes2022global} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
-The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,chira2020signature,hughes2022global} for detailed descriptions of data collection and processing).
-We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
-These eight dimensions contained at least 95\% of variance in each superorder, and 98.7\% of the variance in the whole dataset.
-We used a threshold of $>95$\% of variance explained for each superorder because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
-$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each superorder.
-For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
-This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each superorder (see Fig.
+% \textbf{Beak shape data.}
+% We used the dataset from \cite{cooney2017mega,chira2020signature,hughes2022global} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
+% The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,chira2020signature,hughes2022global} for detailed descriptions of data collection and processing).
+% We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
+% These eight dimensions contained at least 95\% of variance in each superorder, and 98.7\% of the variance in the whole dataset.
+% We used a threshold of $>95$\% of variance explained for each superorder because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
+% $\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each superorder.
+% For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
+% This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each superorder (see Fig.
 S\ref{Fig:axes_variance} and Table \ref{tab_variance_per_axis}).
-This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
+% This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
 
-\textbf{Phylogenetic data.}
-We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
-We pruned these trees to contain only the 8748 species present in our beak shapespace.
+% \textbf{Phylogenetic data.}
+% We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
+% We pruned these trees to contain only the 8748 species present in our beak shapespace.
 
-\textbf{Phylogenetic GLMM analysis.}
-We ran a phylogenetic Bayesian generalised linear mixed effects model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
+% \textbf{Phylogenetic GLMM analysis.}
+% We ran a phylogenetic Bayesian generalised linear mixed effects model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
 
-\begin{equation}
-\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
-\end{equation}
+% \begin{equation}
+% \text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
+% \end{equation}
 
-This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
-These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
-We ran two separate nested models.
-One model using one random term for the whole phylogeny and 35 nested random terms for each superorder and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
-And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
-This resulted in two multidimensional GLMM.
-One with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
-And one with a $5229 \times 8$ response variable, one global residual term and 30 random terms.
+% This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
+% These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
+% We ran two separate nested models.
+% One model using one random term for the whole phylogeny and 35 nested random terms for each superorder and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
+% And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
+% This resulted in two multidimensional GLMM.
+% One with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
+% And one with a $5229 \times 8$ response variable, one global residual term and 30 random terms.
 
-To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
-Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the whole tree and the passeriformes ones).
-Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
+% To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
+% Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the whole tree and the passeriformes ones).
+% Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
 
-\textbf{mcmcmcglmm.}
-This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
-This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
-1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
-2) extracting the following parameters from the parameterisation chains:
-a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
-b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
-We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
-This effectively resulted in getting 10 exploitable posterior samples per tree.
-We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
+% \textbf{mcmcmcglmm.}
+% This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
+% This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
+% 1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
+% 2) extracting the following parameters from the parameterisation chains:
+% a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
+% b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
+% We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
+% This effectively resulted in getting 10 exploitable posterior samples per tree.
+% We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
 S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
-Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
-The whole procedure is reproducible in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
+% Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
+% The whole procedure is reproducible in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
 
-\textbf{Projection/rejection analyses.}
-We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
-In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group.
-In brief we can interpret where a species or a group falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
-We calculated the elaboration and innovation scores using two main approaches:
-1) by clades (i.e.
+% \textbf{Projection/rejection analyses.}
+% We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
+% In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that clade.
+% In brief we can interpret where a species or a clade falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
+% We calculated the elaboration and innovation scores using two main approaches:
+% 1) by clades (i.e.
 to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes (Fig.
 \ref{Fig:ellipses});
-and 2) by species (i.e.
+% and 2) by species (i.e.
 to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes and b) the phylogenetic major axes of their respective clade (Fig.
 \ref{Fig:phylogeny}).
-To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
-This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
-The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
-The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
-and implemented in the \texttt{dispRity} package \cite{dispRity}.
+% To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
+% This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
+% The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
+% The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
+% and implemented in the \texttt{dispRity} package \cite{dispRity}.
 
-\textbf{Groups' orthogonality.}
-We measured the orthogonality of each random term in our models (clades) compared to their parent group and their parent parent's.
-For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
-To do so we measured the angles between the major axes of the focal group and their parent group for each posterior variance-covariance matrix as well as within each group by randomly comparing major axes within a group.
-We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
-We then measured the posterior probability of the angles within a group being different than the angles between groups.
-These results are available in Fig. \ref{Fig:orthogonality}.
+% \textbf{Groups' orthogonality.}
+% We measured the orthogonality of each random term in our models (clades) compared to their parent clade and their parent parent's.
+% For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
+% To do so we measured the angles between the major axes of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade by randomly comparing major axes within a clade.
+% We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
+% We then measured the posterior probability of the angles within a clade being different than the angles between clades.
+% These results are available in Fig. \ref{Fig:orthogonality}.
 
-\textbf{Reproducibility and repeatability.}
-The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
-The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
+% \textbf{Reproducibility and repeatability.}
+% The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
+% The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
 

### 2023-06-01 Manuscript/main.tex
results discussion done!

@@
 \noindent \textbf{Figure \ref{Fig:cheatsheet}}
-Figure 1:
 Relationships between key figures in the main text.
@@
 
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
-% \caption{.}
-% \label{Fig:ellipses}
-% \end{figure}
-% \bigskip
+\section{Results and Discussion}
 
-% \noindent \textbf{Figure \ref{Fig:ellipses}:}
-% Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
-% We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
-% The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
-% Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
-% The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
-% The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
-% Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
-% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.
+\subsection{Modeling nested trait covariance}
+Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of phenotypic variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
+To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
+To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves) (Fig.
+ \ref{Fig:cheatsheet} and \ref{Fig:mcmcmcglmm}).
+To estimate the major axis of beak shape variation we estimate the evolutionary rate matrix \cite{Houle2017,Machado2020} by fitting Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable \cite{MCMCglmm}.
+The major axis can be identified from the leading eigenvector of the posterior distribution of evolutionary rate matrices.
+We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
+This major axis of beak variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
+We can then estimate elaboration and innovation based on that major axis: elaboration is defined as the position on the phylogenetic major axis of beak variation and innovation is defined as the distance from the phylogenetic major axis of beak variation (see Methods).
+%@L_ei_meth_precision.
+To assess the roles of elaboration and innovation in beak shape evolution across scales we fit our pGLMMs as nested models in which we define evolutionary rate matrices, and therefore major axes of beak shape variation, for
+i) the entire phylogeny of all species included in the data (hereafter the class-wide phylogenetic level),
+ii) all species within each of nine clades mapping approximately to super-orders (the super-order level), and
+iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
+These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation across scales.
 
-% \bigskip
+\subsection{Macroevolution: elaboration is a common route to divergence among species}
+We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along phylogenetic lines of least resistance by calculating species-specific measures of $elaboration_{species}$ and $innovation_{species}$.
+If lines of least resistance are stable among species then we expect that species typically fall on, or close to, a conserved phylogenetic major axis of beak variation ($elaboration_{species}$) rather than away from the axis ($innovation_{species}$).
+To assess this we projected the beak shape data for each species onto %@L_linear_algebra_stuff @L_beak_shape
+the major axes of evolutionary rate matrices of their orders (see supplementary materials \ref{Fig:phylogeny_supplement} for projection onto the major axes of beak shape variation from the class-wide phylogenetic major axis of beak variation or their super-order’s phylogenetic major axis of beak variation).
+We found typically higher values of $elaboration_{species}$ than $innovation_{species}$ (Fig. \ref{Fig:phylogeny}) %@L_rephrase_wefound
+and strong clustering of both $elaboration_{species}$ and $innovation_{species}$ (median $elaboration_{species}$ for all orders = 0.114, 95\%CI = 0.004-0.485;  $elaboration_{species}$ Pagel's $\lambda$ = 0.888; median $innovation_{species}$ for all orders = 0.071, 95\%CI = 0.018-0.224); $innovation_{species}$ Pagel's $\lambda$ = 0.848).%@L_ei_species_values.
+% GT: TODO: add code for reproducibility
 
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results.pdf}
-% \caption{.}
-% \label{Fig:orthogonality}
-% \end{figure}
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/InnovElabTree_main_text_revision.pdf}
+\caption{.}
+\label{Fig:phylogeny}
+\end{figure}
+\bigskip
 
-% \bigskip
+\noindent \textbf{Figure \ref{Fig:phylogeny}}
+Avian phylogeny (n = 8748 species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
+Elaboration and innovation scores represent comparisons of species at the order level.
+Additional comparisons to super-orders and at the class-wide phylogenetic level are shown in Fig.
+\ref{Fig:phylogeny_supplement} and for family and super-family comparisons of the order Passeriformes in Fig.
+\ref{fig_phylogeny_passeriformes}.
 
-% \noindent \textbf{Figure \ref{Fig:orthogonality}:}
-% Orthogonality of each clade's major axis compared to their parent or parent parent's clade.
-% Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
-% Dots represent the clades median orthogonality and the lines their 95\% CI.
-% No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
-% These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
-% Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
-% For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
-% The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
-% Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
-% This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
-% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
+$Innovation_{species}$ is more apparent within non-passerine, rather than passerine, clades.
+Most notably we see higher levels of innovation in orders with hook-shaped beaks (Pstittaciformes, Falconiformes, Accipitriformes), and the Apodiformes (swifts and hummingbirds).
+Clades with hooked beaks have highly integrated beaks and brain cases and strong allometric patterns of shape variation, particularly among raptorial clades \cite{bright2016shapes, bright2019multifactorial}.
+Apodiformes are also notable for having both high innovation and elaboration scores.
+The presence of extremes of elaboration and innovation might reflect the deep divergence and distinct ecologies of the two families, Trochilidae (hummingbirds) and Apodidae (swifts), that constitute the order.
+The major axes of beak divergence within the order follow a trend from swifts (short, wide-gaped beaks; aerial insectivores) to hummingbirds (long, narrow, and often curved beaks; primarily nectarivores) that is not present within either family.
+% GT: Maybe needs a bit more here, possibly linking from integration to expectations of evolvability.
+Also, need to work on the end of the paragraph to articulate why the family split might lead to odd E and I (something along the lines of major axis is not elaboration within the families but instead is a mega level innovation).
+This phylogenetic major axis of beak variation is likely to be representative of innovation between clades, rather than a line of least resistance within a clade.
 
-% \bigskip
+Phylogenetic distributions of $elaboration_{species}$ and $innovation_{species}$ broadly hold whether comparisons are made at the order, super-order ($elaboration_{species}$ Pagel's $\lambda$ = 0.853; $innovation_{species}$ Pagel's $\lambda$ = 0.914), or class-wide phylogenetic level ($elaboration_{species}$ Pagel's $\lambda$ = 0.935; $innovation_{species}$ Pagel’s $\lambda$ = 0.837; Fig.
+\ref{Fig:phylogeny_supplement}); \ref{Fig:figure_correlations} %@L_correlation_figure_link)
+and in further nested analysis of the order Passeriformes (Fig. \ref{fig_ellipses_passeriformes} and \ref{fig_phylogeny_passeriformes}).
+The extent of both elaboration and innovation tends to reduce as we move from class-wide phylogenetic to within-order comparisons.
+We regard within-order comparisons as the most informative level of comparison at the macroevolutionary scale
+% GT: I added this so that we are clear on our focal analysis. The second part of the sentence is taken directly from one of reviewer 1's comments. TG:TODO: add this to the response letter @@@
+because comparisons of species level values derived from the main axis of variation at the level of whole Neornithes might reflect global constraints in beak shape or an axis of innovation among clades (as with the Apodiformes discussed above).
+The class-wide phylogenetic major axis of beak variation aligns closely with the raw (non-phylogenetic) major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme, and long, shallow, and narrow beaks at the other.
+Our results point to the possibility that within clades the phylogenetic major axes of beak variation follow different directions.
 
-% However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
-% While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
-% To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
-% As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
-% Clades (or species) can be both elaborators and innovators.
-% To measure innovation and elaboration we translated the major axes of each clade's (i.e.
-superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
-% We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
-% We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
-% These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
-% We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
-% Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
-% However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g.
+Collectively, our results suggest that elaboration is the most common route by which species beak shape evolves.
+While $elaboration_{species}$ is the main mode of beak shape divergence at the macroevolutionary scale %@L_dominant
+(Fig. \ref{Fig:phylogeny}), it is also commonly positively correlated with $innovation_{species}$ (Fig. \ref{fable_correlations} \ref{fable_correlations_supplementary}. %@L_correlation_figure_link).
+In 14/27 clades we find significant positive correlations with non-significant but positive trends in a further 11 clades.
+Only two clades (Falconiformes and Columbiformes) hint at a negative correlation but are clearly non-significant.
+The apparent predominance of $elaboration_{species}$ when measured at the macroevolutionary scale is consistent with the idea of evolution along lines of least resistance % @L_genetic_lol
+yet the lack of trade-off between $elaboration_{species}$ and $innovation_{species}$ further suggests that evolvability, which is measured by elaboration along a single axis, may in fact be better regarded as multi-dimensional.
+Indeed, if divergence can arise simultaneously along multiple axes then this may help to resolve the paradox of apparent shifts in phenotype that have contributed to the diversity of species morphologies across the tree of life.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/correlations_fable_orders.pdf}
+\caption{.}
+\label{fable_correlations}
+\end{figure}
+\bigskip
+
+\noindent \textbf{Figure \ref{fable_correlations}}
+Posterior correlations between $elaboration_{species}$ and $innovation_{species}$ for each super-order.
+The dots, thick lines and dashed lines represent respectively the median, 50\% confidence interval (CI) and 95\% CI scores.
+14 out of 27 of the orders have a clear posterior correlation between $elaboration_{species}$ and $innovation_{species}$ (i.e.
+95\% CI does not overlap with 0), nine have a somewhat positive posterior correlation (i.e.
+50\% CI does not overlap with 0) and four have no clear posterior correlation.
+Note that no order has a clear negative posterior correlation.
+This trend suggests that there is no tradeoff between $elaboration_{species}$ and $innovation_{species}$ and that both can be common routes for beak shape evolution at the macroevolutionary scale.
+
+\subsection{Megaevolution: multiple routes to innovation throughout avian evolutionary history}
+If deep-time jumps in beak shape \cite{cooney2017mega} are the result of innovation we would expect to see changes in the orientation of trait space of clades (i.e.
+orders or super-orders) relative to the phylogenetic major axis of beak variation of the whole class Aves.
+To test this, we measured the $elaboration_{clade}$ and $innovation_{clade}$ by translating the phylogenetic major axis of beak variation of each clade's (i.e.
+super-orders and orders) evolutionary rate matrix onto the class-wide phylogenetic major axis of beak variation so that they shared the same origin in the shapespace.
+We then used linear algebra to project %@L_linear_algebra_stuff
+a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig.S\ref{Fig:cheatsheet}).
+We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ scor.
+% @L_euclidean.
+A high $innovation_{clade}$ score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g.
+the direction of evolution of an order differs from its parent super-order).
+These scores were calculated for each of 4000 posterior pairs of clade vs.
+evolutionary rate matrix (Fig. \ref{Fig:ellipses} boxplots).
+We found that for most clades the phylogenetic major axis of beak variation is not aligned with the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}).
+Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the evolutionary rate matrix for the class-wide phylogenetic major axis of beak variation and each super-order and order (Fig.
+\ref{Fig:ellipses}b) highlight the striking variation in the orientation of beak shape major axes among clades (Fig.
+\ref{Fig:orthogonality}).
+More than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median $innovation_{clade}$ than median $elaboration_{clade}$ scores implying that $innovation_{clade}$ is a more common generator of beak shape diversity than $elaboration_{clade}$ in deep-time (i.e evidence at the clade scale).
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
+\caption{.}
+\label{Fig:ellipses}
+\end{figure}
+\bigskip
+
+\noindent \textbf{Figure \ref{Fig:ellipses}:}
+$Elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale for each order and super-order in the bird phylogeny.
+The black and gray ellipses are the scaled average evolutionary rate matrices from the pGLMM models for respectively the clades (black) and the class-wide phylogeny (gray).
+The ellipses are centered on the position of the clade in the shapespace.
+The associated black bar plots represent the variance on each of the eight dimensions in the shapespace with the top gray bar plot representing the variance on each of the eight dimensions for the class-wide phylogeny.
+The orange and blue distributions represent respectively the distribution of the $elaboration_{clade}$ (orange) and the $innovation_{clade}$ (blue) scores for each clade and the dots are the median $elaboration_{clade}$ or $innovation_{clade}$ and the solid and the dashed lines representing the 50\% and the 95\% confidence intervales of the distribution of the scores across the 4000 posterior samples.
+The ticks on the y-axis always represent five arbitrary units of elaboration and innovation.
+An alternative visualization of the same information is available in the supplementary materials in Fig.
+\ref{Fig:ellipses_rainbow} along with a companion plot showing further nested structure within the Passeriformes in Fig.
+\ref{fig_ellipses_passeriformes}.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/orthogonality_results.pdf}
+\caption{.}
+\label{Fig:orthogonality}
+\end{figure}
+\bigskip
+
+\noindent \textbf{Figure \ref{Fig:orthogonality}:}
+Amount of orthogonality of each clade's phylogenetic major axis of beak variation compared to their parent or parent's parent clade.
+The amount of orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background gray and dashed gray lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+Dots represent the median orthogonality of each clade and the lines their 95\% confidence intervals (CI).
+No super-order's phylogenetic major axis of beak variation is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal (>  $45^\circ$) to the class-wide phylogenetic major axis of beak variation.
+These results are to be contrasted with the orientation of each order relative to their super-order where on average (median) only 14/27 are at least half orthogonal to the orientation of their super-orders’s phylogenetic major axis of beak variation.
+We also indicate the number of species (n) and the standard deviation (sd) of the orientation of their phylogenetic major axis of beak variation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+Note that noclades are clearly parallel to their parent’s or parent’s parent’s ellipses (i.e.
+no 95\% CI includes 0).
+For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the class-wide phylogenetic major axis of beak variation relative to their sample size and sd.
+The stars represent the posterior probability of the clade's orientation being different from the comparison clade (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
+Only 1/6 super-orders, 8/27 orders relative to class-wide phylogeny and 6/27 relative to their super-orders have a pp $> 0.99$.
+This is the result of the variation in sample size (n) and standard variation (sd) among clades.
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. \ref{fig_orthogonality_passeriformes}.
+
+Substantial variation in $elaboration_{clade}$ and $innovation_{clade}$ arises among clades including clades with relatively low $elaboration_{clade}$ and $innovation_{clade}$(e.g.
 Cuculiformes; Fig.
-\ref{Fig:ellipses}), low elaboration and high innovation (e.g.
+\ref{Fig:ellipses}), low $elaboration_{clade}$ and high $innovation_{clade}$ (e.g.
 Galliformes; Fig.
-\ref{Fig:ellipses}), high elaboration and low innovation (e.g.
+\ref{Fig:ellipses}), high $elaboration_{clade}$ and low $innovation_{clade}$ (e.g.
 Coraciiformes; Fig.
-\ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g.
+\ref{Fig:ellipses}), and both relatively high $elaboration_{clade}$ and $innovation_{clade}$ (e.g.
 Podicipediformes; Fig.
 \ref{Fig:ellipses}).
-% These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
-% This suggests that divergence away from line of least evolutionary resistance (i.e.
-the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
+These patterns hold across scales including within super-orders and at finer scales within the order Passeriformes (Fig.
+\ref{fig_ellipses_passeriformes}).
+Evolutionary theory predicts that lines of least resistance should break down with time since divergence \cite{schluter1996adaptive, marroig2005size, mcglothlin2018adaptive}.
+We find that the extent of both $innovation_{clade}$ and $elaboration_{clade}$ increase with time since divergence from the root of the tree.
+Similarly, when measured at the macroevolutionary scale, $innovation_{species}$ and $elaboration_{species}$ increase with species age (see supplementary Fig.
+\ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}).
+These results are consistent with both divergence in phenotype that is proportional to time (e.g.
+as expected under a Brownian motion model) and with the idea that changes in phenotypic correlations are expected to change over longer periods of time, potentially in response to moving adaptive landscapes \cite{BurinWhales}.
 
-% This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
-% Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig.
-\ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig.
-\ref{Fig:ellipses}-bar plots).
-% These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig.
-S\ref{fig_ellipses_passeriformes}).
-% Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
-% For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-% These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
-% However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.
+Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig.
+\ref{Fig:ellipses} - Fig.
+\ref{Fig:orthogonality}).
+The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig.
+\ref{Fig:orthogonality}).
+Comparisons of orientation among subclades (i.e.
+orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig.
+\ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary scale and are no more likely to occur along any one axis than another.
+This differs from previous inference from a subset of our data analysed  without incorporating phylogeny, that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
+This suggests that there is no megaevolutionary analogue of the genetic line of least resistance and instead that megaevolutionary shifts more likely reflect changes in the adaptive landscape.
+Across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is sometimes assumed \cite{venditti2011multiple}.
+% @L_occasionaly
 
-% Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
-% This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
-% To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shapespace onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
-% We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire clades.
-% We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
-% However, we also found various combinations of elaboration and innovation across birds.
-% For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
-% Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
-% Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
-% The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig.
-\ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
-%  and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
-% The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
+Evolutionary $innovation_{clade}$ can arise in many directions in trait space (see examples in supplementary Fig. \ref{Fig:mikado}). %@L_any_dir
+Although the class-wide phylogenetic major axis of beak variation aligns closely with the first dimension of the raw beak shapespace (70.96\% of the class-wide phylogenetic major axis of beak variation is aligned with the first dimension of the shape space; Fig.
+\ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly PC1 (Fig.
+\ref{Fig:ellipses}-bar plots); $innovation_{clade}$ dominates at this scale.
+This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes phylogenetic major axis of beak variation (Fig.
+\ref{fig_ellipses_passeriformes}).
+In addition to the lack of alignment of clade phylogenetic major axis of beak variation with the class-wide phylogenetic major axis of beak variation for birds, we also found that within clades, beak shape variation is either highly constrained (varying almost entirely along a single axis; e.g.
+in Galliformes) or higher-dimensional than expected (there is no clear dominant phylogenetic major axis of beak variation; e.g.
+in Podicipediformes) from the class-wide phylogenetic major axis of beak variation.
+For example, Accipitriformes (hawks and allies) have phylogenetic major axes of beak variation that mainly align with the second dimension of the shapespace suggesting distinct directions of beak evolution for the clade relative to the class-wide phylogenetic level but uniformity in their beak shape within the clade.
+In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
+These observations illustrate multiple routes to $innovation_{clade}$ and imply that within clades, species phenotypes can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
 
-% \begin{figure}[!htbp]
-% \centering
-%     \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_main_text.pdf}
-% \caption{.}
-% \label{Fig:phylogeny}
-% \end{figure}
+\subsection{Species elaborate \textit{and} innovate, clades innovate} %TG: TODO: note in the response letter that we've now nuanced this
+The disconnect between observations of reorientation of trait space at the macroevolutionary scale, and an apparent predominance of $elaboration_{species}$ could be viewed as a paradox.
+We formalize this paradox by testing whether elaboration is more common at higher taxonomic levels (megaevolutionary scale) than at the species level (macroevolutionary scale).
+We compared the area under the curve of the scaled density of $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$.
+We then measured the difference between both areas and found that there was a significant difference in $innovation_{species}$ and $innovation_{clade}$ but no clear difference in $elaboration_{species}$ and $elaboration_{clade}$ (Fig.
+\ref{Fig:relative_EI}).
+This difference confirms our observation that innovation is indeed common and dominant at the megaevolutionary scale, however, the contributions of elaboration and innovation are shown to be indistinguishable at the macroevolutionary scale.
+So, although species often evolve preferentially along a shared phylogenetic major axes of beak variation within clades, there is frequent deviation from these major axes, and the phylogenetic major axes of beak variation change among clades.
 
-% \bigskip
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/relative_EI_gg_style_unscaled.pdf}
+\caption{.}
+\label{Fig:relative_EI}
+\end{figure}
+\bigskip
 
-% \noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n=8748 species) coloured by species beak shape innovation and elaboration level.
-% Blue and yellow colors highlight species with relatively low elaboration or low innovation, whereas orange to red colours indicate high innovation or elaboration.
-% Innovations and elaboration are displayed as: a) comparisons of species to the phylogenetic major axis (inner two bands); b) comparisons of species to their superorder major axis (middle two bands); and c) comparisons of species to their order major axis (outer two bands).
-% The branches are coloured by the species distance from the centroid of the shapespace and the estimation of their ancestors' position in that space.
-% A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.
+\noindent \textbf{Figure \ref{Fig:relative_EI}:}
+Comparisons of relative $elaboration_{species}$ and $innovation_{species}$ (Fig. \ref{Fig:phylogeny}) and $elaboration_{clade}$ and $innovation_{clade}$ (Fig. \ref{Fig:ellipses}).
+a) the $elaboration_{species}$ (purple) for each species in each 4000 posterior samples compared to the $elaboration_{clade}$ for each of the 35 clades (yellow).
+The $elaboration_{clade}$ scores are scaled by the maximum score within the clade (corresponding to the scaled ellipses in Fig.
+\ref{Fig:ellipses}) and the scores for species are scaled by the maximum $elaboration_{species}$ score.
+This allows us to compare $elaboration_{species}$ in and $elaboration_{clade}$ despite their different sample sizes.
+We use the Bhattacharryya Coefficient to quantify the overlap of the area under the curve for species and clades.
+For elaboration the overlap is Bhattacharryya Coefficient = 0.217, indicating no distinguishable dissimilarity in the amount of $elaboration_{species}$ and $elaboration_{clade}$ across the posterior samples.
+b) as a) but using $innovation_{species}$ and $innovation_{clade}$ scores.
+For innovation the overlap has Bhattacharryya Coefficient = 0.039, which indicates a clear difference in the amount of $innovation_{species}$ and $innovation_{clade}$ across the posterior samples.
 
-% \bigskip
+We further tested the contributions of elaboration and innovation at different scales by examining the consequences of elaboration and innovation for the observed divergence of bird beak shapes.
+Specifically, we calculated the distance to centroid of avian beak morphospace for each species (Fig. \ref{Fig:phylogeny}).
+We then used phylogenetic generalised least squares (PGLS; \cite{phylom}) to model distance to centroid as a function of $elaboration_{species}$, $innovation_{species}$ and $elaboration_{clade}$ and $innovation_{clade}$.
+We ran the PGLS analysis using both single and multiple predictors.
+Our analyses indicate that the majority of variation (r2 = 0.888; Table \ref{Tab:pgls}) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
+When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table 1).
+In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table 1).
+However, $innovation_{clade}$ is nonetheless an important contributor to total beak shape space.
+We reach this conclusion because models with interactions between species and clade metrics provide by far the best model fit overall and removal of terms indicate that the most important interactions are between $innovation_{clade}$  and $innovation_{species}$ followed by $innovation_{clade}$ and $elaboration_{species}$.
+The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table 1 \ref{Tab:pgls})
+Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space.
 
-% The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
-% We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
-% We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
-% We then measured the difference between both areas and found that there was a significant difference in innovation between species and clades but no clear difference in elaboration (Fig.
-\ref{Fig:relative_EI}).
-% This difference confirms our observation that innovation is more common at the clade level rather than at the species level and that although species evolve preferentially along a line of least evolutionary resistance, this line of least resistance changes frequently among clades.
-% However, we suggest that rather than a paradox the reorientation of trait space can instead be described as a gradual process occuring in response to shifting selection pressures acting on species evolving in similar adaptive zones.
-% This implies that gradual directional evolution \cite{pagel2022general} rather than megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple} may be sufficient to explain diversity in avian beak morphology.
+\begin{table}
+\begin{tabular}{ c c c }
+ So & many & stars \\ 
+ *** & *** & *** \\  
+ *** & *** & *** \\
+\end{tabular}
+\caption{Elaboration and innovation as predictors of beak shape distance from centroid.
+The table shows parameter estimates of predictors from  13 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms.
+Models 1-6 are multiple regressions including all variables and one or more interaction terms; models 7-9 are are multiple regressions including two or more variables and no interaction terms; models 10-13 are single predictor models.
+Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e.
+assuming a Brownian motion model of evolution).
+Fixing $\lambda$ is necessary to allow model comparison with AIC.
+We also fitted the same models fixing $\lambda$ to 0.7 (the lowest value of $\lambda$lambda from any individual model) and relative model fit and interpretation of parameters was unaffected.
+E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
+\label{Tab:pgls}
+\end{table}
 
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=0.9\textwidth]{Figures/relative_EI.pdf}
-% \caption{.}
-% \label{Fig:relative_EI}
-% \end{figure}
+Taken together, these results suggest that rather than a paradox, the reorientation of trait space can instead arise as a result of species-level innovations within clades arising along common directions in phenotypic space.
+This idea is similar in concept to multiple adaptive peak models where species evolve in a heterogeneous adaptive landscape and share similar responses to selection pressures within adaptive zones \cite{hansen1997stabilizing}.
+This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{venditti2011multiple,cooney2017mega}, may be sufficient to explain diversity in avian beak morphology.
+In other words, populations can always change morphologically within the shapespace with minimal amounts of innovation over long periods of time. %@L_shapespacetravel
 
-% \bigskip
+\subsection{Conclusions}
+Our results show that although at the macroevolutionary scale most bird beaks are elaborating along a class-wide or sub-clade phylogenetic major axis of beak variation, at the megaevolutionary scale, innovation away from the class-wide phylogenetic major axis of beak variation is much more common.
+This nested structure of elaboration at a lower (species) taxonomic level and innovation at a higher (clade) taxonomic levels could thus explain the diversity of bird beaks we observe today.
+However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
+Taken together, our results suggest that the signature of evolutionary reorientations in deep-time ($innovation_{clade}$), coupled with $elaboration_{species}$, is a robust explanation of the massive diversity of bird beaks we observe today and is consistent with recent suggestions from univariate analysis that observation of apparently abrupt phenotypic shifts can be explained by gradual Darwinian processes \cite{pagel2022general, BurinWhales,goswami2022}.
+Indeed, the class-wide phylogenetic major axis of phenotypic variation is an emergent property of reorientation of trait space among clades that requires no special evolutionary process: megaevolutionary patterns appear to be an inevitable outcome of evolution on a shifting adaptive landscape.
 
-% \noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure  \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
-% A - In blue, the elaboration for each species in each 4000 posterior samples compared to the elaboration for each 35 clades.
-% The scores for each clade is scaled by the maximum score within each clade (corresponding to the scaled ellipses in figure  \ref{Fig:ellipses}) and the score for the species is scaled by the maximum species elaboration score.
-% This allows us to compare both types (species vs. clades) despite their different sample sizes.
-% The Bhattacharryya Coefficient represents the overlap of the area under the curve for species and clades (a BC of 0 represents no overlap).
-% B - same as A but for the innovation score.
-% We can see a clear difference in the amount of innovation across the posterior samples between species or clades.
-
-% \bigskip
-
-% Our results show that although at the species level most bird beaks are elaborating along a megaevolutionary line of least resistance, at the macroevolutionary level innovation way from the megaevolutionary line of least resistance is much more common.
-% This nested structure of elaboration at a higher taxonomic level and innovation at a lower taxonomic level could thus explain the diversity of bird beaks we observe today.
-% However, individual species-level variation in the past could also have led to a shift in a clade's line of least evolutionary resistance through different evolutionary routes on a dynamic adaptive landscape.
-Taken together, our results suggest that the signature of evolutionary orientations in deep-time, coupled with elaboration at the species-level, is consistent with recent suggestions that the apparently abrupt phenotypic shifts can be explained by normal Darwinian processes \cite{pagel2022general}.
-% Our results help to bridge the gap between micro- and macro-evolution by proposing a multi-scale framework for  phenotypic evolution that allows species to evolve along lines of least evolutionary resistance while accommodating frequent shifts in the direction of evolutionary change.
-
-% \section{Methods}
-
-% \textbf{Beak shape data.}
-% We used the dataset from \cite{cooney2017mega,chira2020signature,hughes2022global} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of  the beaks of 8748 bird species.
-% The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,chira2020signature,hughes2022global} for detailed descriptions of data collection and processing).
-% We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
-% These eight dimensions contained at least 95\% of variance in each superorder, and 98.7\% of the variance in the whole dataset.
-% We used a threshold of $>95$\% of variance explained for each superorder because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
-% $\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each superorder.
-% For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig. \ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
-% This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each superorder (see Fig.
-S\ref{Fig:axes_variance} and Table \ref{tab_variance_per_axis}).
-% This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace).
-
-% \textbf{Phylogenetic data.}
-% We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
-% We pruned these trees to contain only the 8748 species present in our beak shapespace.
-
-% \textbf{Phylogenetic GLMM analysis.}
-% We ran a phylogenetic Bayesian generalised linear mixed effects model (pGLMM) using the \texttt{MCMCglmm} R package \cite{MCMCglmm} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
-
-% \begin{equation}
-% \text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
-% \end{equation}
-
-% This model allows us to estimate a phenetic variance-covariance matrix for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
-% These variance-covariance matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
-% We ran two separate nested models.
-% One model using one random term for the whole phylogeny and 35 nested random terms for each superorder and order containing more than 15 species in our dataset (Fig. \ref{Fig:ellipses}).
-% And one model on a subset of the dataset containing only the 5229 passeriformes species with 30 nested random terms.
-% This resulted in two multidimensional GLMM.
-% One with a $8748 \times 8$ response variables, one global residual term and 36 random terms.
-% And one with a $5229 \times 8$ response variable, one global residual term and 30 random terms.
-
-% To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
-% Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the whole tree and the passeriformes ones).
-% Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
-
-% \textbf{mcmcmcglmm.}
-% This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
-% This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
-% 1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
-% 2) extracting the following parameters from the parameterisation chains:
-% a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
-% b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
-% We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
-% This effectively resulted in getting 10 exploitable posterior samples per tree.
-% We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
-S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
-% Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
-% The whole procedure is reproducible in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
-
-% \textbf{Projection/rejection analyses.}
-% We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
-% In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that clade.
-% In brief we can interpret where a species or a clade falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
-% We calculated the elaboration and innovation scores using two main approaches:
-% 1) by clades (i.e.
-to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes (Fig.
-\ref{Fig:ellipses});
-% and 2) by species (i.e.
-to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes and b) the phylogenetic major axes of their respective clade (Fig.
-\ref{Fig:phylogeny}).
-% To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
-% This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
-% The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
-% The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
-% and implemented in the \texttt{dispRity} package \cite{dispRity}.
-
-% \textbf{Groups' orthogonality.}
-% We measured the orthogonality of each random term in our models (clades) compared to their parent clade and their parent parent's.
-% For example, the Columbimorphae random terms' variance-covariance matrices against the phylogeny ones, the Columbiformes ones against the Columbimorphae ones (parent) and the whole phylogeny ones (parent parents).
-% To do so we measured the angles between the major axes of the focal clade and their parent clade for each posterior variance-covariance matrix as well as within each clade by randomly comparing major axes within a clade.
-% We converted each angle measurement into an orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
-% We then measured the posterior probability of the angles within a clade being different than the angles between clades.
-% These results are available in Fig. \ref{Fig:orthogonality}.
-
-% \textbf{Reproducibility and repeatability.}
-% The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
-% The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
+\textbf{Reproducibility and repeatability.}
+The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
+The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
 

### 2023-06-01 Manuscript/supplementaries.tex
results discussion done!

--- 2023-05-31 Manuscript/supplementaries.tex
@@
 
+\begin{figure}[htbp]
+\centering
+   \includegraphics[width=0.7\textwidth]{Figures/InnovElabTree_supp_revision.pdf}
+\caption{Avian phylogeny (n = 8748 species) showing distributions of species beak shape $elaboration_{species}$ (inner circle, yellow scale), $innovation_{species}$ (outer circle, blue scale), and Euclidean distance of species to the centroid of beak space (branches, cividis scale).}
+\label{Fig:phylogeny_supplement}
+\end{figure}
+
 \subsubsection{Correlations between $elaboration_{species}$ and $innovation_{species}$}
@@
 \ref{Fig:figure_correlations}.}
-\label{Fig:fable_correlations}
+\label{fable_correlations_supplementary}
 \end{figure}
@@
 \label{Fig:ellipses_rainbow}
-\end{figure}
-
-\begin{figure}[htpb]
-\centering
-   \includegraphics[width=0.5\textwidth]{Figures/50_random_axes_superorders.pdf}
-\caption{Distribution of 50 random scaled major axes (scaled variance-covariance matrices eigen values) for each super order illustrating that individual major axes can span in many different directions with some clades have a higher spread of direction than others.}
-\label{Fig:mikado}
 \end{figure}

### 2023-06-01 Manuscript/main.tex
Done! Pre-check

@@
 
+\section{Materials and Methods}
+
+\textbf{Beak shape data.} We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of the beaks of 8748 bird species.
+The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
+We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
+These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
+We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
+$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each super-order.
+For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig.
+\ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
+This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. \ref{Fig:axes_variance}).
+This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace). %@L_shapespace_description
+
+\textbf{Phylogenetic data.} We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and one randomly selected tree for the figures (Fig.
+\ref{Fig:phylogeny}).
+We pruned these trees to contain only the 8748 species present in our beak shapespace.
+
+\textbf{Phylogenetic GLMM analysis.} We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} and \texttt{ape} R packages \cite{MCMCglmm, ape} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:
+
+\begin{equation}
+\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
+\end{equation}
+
+This model allows us to estimate an evolutionary rate matrix, here defined as a matrix containing the rate of among-species divergence in species traits that explicitly incorporates phylogeny (\textbf{R} matrix of \cite{Houle2017} and \textbf{B} matrix of \cite{ Machado2020})) for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
+These evolutionary rate matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
+We ran two separate nested models.
+The first model used one random term for the class-wide phylogeny and 35 nested random terms, one for each super-order and order containing more than 15 species in our dataset (Fig.
+\ref{Fig:ellipses}).
+This resulted in a multidimensional GLMM with a $8748 \times 8$ response variables, one class-wide residual term and 36 random terms.
+The second model was fitted to a subset of the dataset containing only the 5229 Passeriformes species with 30 nested random terms, one for each sub-order and families containing more than 15 species in our dataset (Fig.
+\ref{fig_ellipses_passeriformes}.
+This resulted in a multidimensional GLMM with a $5229 \times 8$ response variables, one order-wide residual term and 30 random terms.
+To account for phylogenetic uncertainty we ran the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
+Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each of the models, i.e.
+the class-wide phylogeny and the Passeriformes models).
+Instead, we developed and implemented a ''mini-chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.
+
+\textbf{mcmcmcglmmm.} This method runs relatively short Monte-Carlo Markov chains (MCMC) across a large number of different phylogenetic trees and concatenates the resulting short posteriors into one larger posterior distribution that contains the phylogenetic uncertainty information.
+We performed the method using the following protocol (see Fig. \ref{Fig:mcmcmcglmm}).
+First we ran three independent MCMCglmm chains, hereafter the parameterisation chains, for 50,000 generations with the trait data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\%, i.e.
+with a very low weight on the priors.
+Next we extracted the following parameters from the parameterisation chains: a) the conservative burnin average time across the three chains, defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and b) the mini-chains priors, defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
+We then ran 400 mini chains in parallel with the estimated burnin time and priors to run 10 samples past the burnin.
+This resulted in 10 exploitable posterior samples for each tree.
+We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig.
+\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
+Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain - a 400 fold improvement.
+The total analysis took around three CPU years using the shARC cluster from the University of Sheffield.
+Code to reproduce the procedure is available in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.
+
+\textbf{Elaboration/innovation scores using projection/rejection analyses.} We used the distributions of the 4000 posterior evolutionary rate matrices to run the projection/rejection analyses to obtain elaboration and innovation scores for bird beaks.
+We used linear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the evolutionary rate matrices (referred to throughout the text as the phylogenetic major axis of beak variation) as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade).
+We projected either species or another phylogenetic major axis of beak variation onto that clade.
+We can then interpret where a species or a clade falls on the phylogenetic major axis of beak variation of reference as its $elaboration_{species}$ score, and how far a species is from that phylogenetic major axis of beak variation as its $innovation_{species}$ score.
+We calculated the elaboration and innovation scores using two main approaches: 1) by clades (i.e.
+to measure the $elaboration_{clade}$/$innovation_{clade}$ at the megaevolutionary scale)where we calculated the projections of the phylogenetic major axis of beak variation of each clade onto the class-wide phylogenetic major axis of beak variation (Fig.
+\ref{Fig:ellipses}); and 2) by species (i.e.
+to measure the $elaboration_{species}$/$innovation_{species}$ at the macroevolutionary scale) where we calculated the projections of each species onto a) the class-wide phylogenetic major axis of beak variation and b) the phylogenetic major axis of beak variation of their parent clade (Fig.
+\ref{Fig:phylogeny}).
+To make the results easier to interpret, we centered and scaled elaboration and innovation scores onto the center of the phylogenetic major axis of beak variation of interest and made the elaboration values absolute.
+In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional, i.e within the 95\% confidence interval of the evolutionary rate matrix.
+The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection} - note that the procedure described in the section \ref{supp_projection} is applied to each individual evolutionary rate matrix for each term in the model, e.g.
+for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score %@L_individual matrices).
+Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.
+
+\textbf{Clade orthogonality.} We measured the amount of orthogonality of each random term (i.e.
+clade) in our models compared to their parent clade and their parent's parent clade.
+For example for Columbiformes, we measured the amount of orthogonality in the evolutionary ratematrices of (i) the Columbimorphae random terms against the phylogeny random terms, (ii) the Columbiformes random terms against the Columbimorphae random terms (parent) and (iii) the Columbiformes random terms the whole phylogeny random terms (parent's parent).
+To measure the amount of orthogonality we calculated the angles between the phylogenetic major axis of beak variation of the focal clade and their parent clade for each posterior evolutionary rate matrix and within each clade, by randomly comparing phylogenetic major axes of beak variation within a clade.
+We converted each angle measurement into a degree of orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
+We then measured the posterior probability of the angles within a clade being different to the angles among clades.
+These results are available in Fig. \ref{Fig:orthogonality}.
+Note that we did not use the random skewers method here \cite{cheverud2007research} because we were interested in measuring the amount of orthogonality between pairs of evolutionary rate matrices rather than simply measuring whether they were different (but see supplementary materials for random skewers correlation results \ref{Fig:random_skewers}).
+%@L_noskewers
+
+\textbf{Phylogenetic generalised least squares analysis.} We ran phylogenetic generalised least squares models (PGLS) to test the effects of elaboration and innovation on beak shape divergence.
+Beak shape divergence was calculated as the Euclidean distance of each species beak shape from the centroid of beak shape of all birds.
+We then fitted a series of nested models with distance as the response and $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$, and $innovation_{clade}$ as predictors.
+The most complex model included all two-way interaction terms:
+
+\begin{equation}
+\text{distance} \mathtt{\sim} (elaboration_{species} + innovation_{species}) \times (elaboration_{clade} + innovation_{clade})
+\end{equation}
+
+We then sequentially removed terms to examine effects on model fit.
+We initially estimated Pagel's $\lambda$ \cite{pagel1997inferring} for all models.
+However, model comparison based on AIC scores can be misleading when estimating $\lambda$ (i.e.
+model differences can arise because of different $\lambda$ fits, rather than due to effects of the parameters of interest).
+We therefore repeated all models a further two times, once fixing $\lambda$ to 1 (equivalent to assuming Brownian motion model of evolution) and once fixing $\lambda$ to the lowest value estimated from any of the models.
+The results were qualitatively similar regardless of $\lambda$ and we report only models based on pure Brownian motion in the main text.
+In addition, we estimated Pagel’s $\lambda$ for $elaboration_{species}$ and $innovation_{species}$ as univariate traits.
+All models were fitted using the phylolm library using a single phylogeny \cite{phylom}.
+
 \textbf{Reproducibility and repeatability.}
@@
 The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
+
+\textbf{Acknowledgements.}
+We thank Masahito Tusboi and three other anonymous reviewers for their suggestions that greatly helped improving the manuscript.
 

### 2023-06-01 Manuscript/main.tex
Update main.tex

@@
 Relative to elaboration, innovation is more likely to generate phenotypic novelty, but, if directions of phenotypic evolution are constrained, would not be expected to be a primary driver of phenotypic diversity among closely related species.
-We apply these terms to i) species divergence within clades, and ii) clade divergence within larger clades (Figure 1).
+We apply these terms to i) species divergence within clades, and ii) clade divergence within larger clades (Fig. \ref{Fig:cheatsheet}).
 Both innovation and elaboration could in principle occur at the macro- or mega-evolutionary scale.
@@
 \centering
-   \includegraphics[width=1\textwidth]{Figures/correlations_fable_orders.pdf}
+   \includegraphics[width=0.5\textwidth]{Figures/correlations_fable_orders.pdf}
 \caption{.}
@@
 \centering
-   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
+   \includegraphics[width=1\textwidth]{Figures/figure_phylo_spiro_dark.pdf}
 \caption{.}

### 2023-06-01 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 The stars denote the median p-value estimate (0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1).}
-\label{Fig:random_skewers}
+\label{/home/tguillerme/Projects/elaboration_exploration_bird_beaks/Manuscript/Figures/others/figure_phylo_spiro_dark.png
+/home/tguillerme/Projects/elaboration_exploration_bird_beaks/Manuscript/Figures/others/orthogonality_results.png
+/home/tguillerme/Projects/elaboration_exploration_bird_beaks/Manuscript/Figures/others/relative_EI_gg_style.png}
 \end{figure}
@@
 \centering
-    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement.pdf}
-\caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 1 in the main text for details.}
+    \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement_revised.pdf}
+\caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 2 in the main text for details.}
 \label{fig_phylogeny_passeriformes}

### 2023-06-02 Analysis/06-Figure_with_correlations.Rmd
Volume + pp for Gustova and Travis

@@
                 show.phylo = FALSE, xlim = c(-0.5, 1),
-                main = "Species projected onto\ntheir orders' major axes") 
+                main = "") 
 dev.off()

### 2023-06-02 Manuscript/main.tex
Update main.tex

@@
 For example, in both bird beaks and mammal skulls there are major jumps in the rate of trait evolution early in clade history as morphological diversity expands \cite{cooney2017mega, goswami2022} with lower magnitude jumps towards the present.
-Major jumps in species trait values are consistent with the concept of megaevolution, describing major discontinuities in phenotypes observed from the fossil record, first proposed by Simpson.
+Major jumps in species trait values are consistent with the concept of megaevolution, describing major discontinuities in phenotypes observed from the fossil record, first proposed by Simpson \cite{simpson1944tempo,simpson1953major}.
 
@@
 the Eocene stem Trochilidae fossils \textit{Parargornis messelensis} and \textit{Eocypselus rowei} \cite{mayr2003new, ksepka2013fossil}).
+%@L_strisores_example
 By assessing the multivariate directions of the evolution of beak shapes of species and clades we aim to deepen our understanding of how diversity in morphological form arises at macro- and mega-evolutionary scales in birds.
@@
    \includegraphics[width=1\textwidth]{Figures/cheat_sheet.pdf}
-\caption{.}
-\label{Fig:cheatsheet}
-\end{figure}
-\bigskip
-
-\noindent \textbf{Figure \ref{Fig:cheatsheet}}
-Relationships between key figures in the main text.
+\caption{\scriptsize{Relationships between key figures in the main text.
 A) represents the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
@@
 order,and the parent level, e.g.
-the whole bird phylogeny.
+the whole bird phylogeny.}}
+\label{Fig:cheatsheet}
+\end{figure}
+\bigskip
+
+% \noindent \textbf{Figure \ref{Fig:cheatsheet}}
 
@@
    \includegraphics[width=1\textwidth]{Figures/InnovElabTree_main_text_revision.pdf}
-\caption{.}
+\caption{Avian phylogeny (n = 8748 species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
+Elaboration and innovation scores represent comparisons of species at the order level.
+Additional comparisons to super-orders and at the class-wide phylogenetic level are shown in Fig.
+\ref{Fig:phylogeny_supplement} and for family and super-family comparisons of the order Passeriformes in Fig.
+\ref{fig_phylogeny_passeriformes}.
+}
 \label{Fig:phylogeny}
@@
 
-\noindent \textbf{Figure \ref{Fig:phylogeny}}
-Avian phylogeny (n = 8748 species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
-Elaboration and innovation scores represent comparisons of species at the order level.
-Additional comparisons to super-orders and at the class-wide phylogenetic level are shown in Fig.
-\ref{Fig:phylogeny_supplement} and for family and super-family comparisons of the order Passeriformes in Fig.
-\ref{fig_phylogeny_passeriformes}.
+% \noindent \textbf{Figure \ref{Fig:phylogeny}}
 
@@
    \includegraphics[width=0.5\textwidth]{Figures/correlations_fable_orders.pdf}
-\caption{.}
-\label{fable_correlations}
-\end{figure}
-\bigskip
-
-\noindent \textbf{Figure \ref{fable_correlations}}
-Posterior correlations between $elaboration_{species}$ and $innovation_{species}$ for each super-order.
+\caption{Posterior correlations between $elaboration_{species}$ and $innovation_{species}$ for each super-order.
 The dots, thick lines and dashed lines represent respectively the median, 50\% confidence interval (CI) and 95\% CI scores.
@@
 This trend suggests that there is no tradeoff between $elaboration_{species}$ and $innovation_{species}$ and that both can be common routes for beak shape evolution at the macroevolutionary scale.
+}
+\label{fable_correlations}
+\end{figure}
+\bigskip
+
+% \noindent \textbf{Figure \ref{fable_correlations}}
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{Figures/figure_phylo_spiro_dark.pdf}
-\caption{.}
-\label{Fig:ellipses}
-\end{figure}
-\bigskip
-
-\noindent \textbf{Figure \ref{Fig:ellipses}:}
-$Elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale for each order and super-order in the bird phylogeny.
+   \includegraphics[width=0.8\textwidth]{Figures/figure_phylo_spiro_dark.pdf}
+\caption{\scriptsize{$Elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale for each order and super-order in the bird phylogeny.
 The black and gray ellipses are the scaled average evolutionary rate matrices from the pGLMM models for respectively the clades (black) and the class-wide phylogeny (gray).
@@
 \ref{Fig:ellipses_rainbow} along with a companion plot showing further nested structure within the Passeriformes in Fig.
-\ref{fig_ellipses_passeriformes}.
+\ref{fig_ellipses_passeriformes}.}
+}
+\label{Fig:ellipses}
+\end{figure}
+\bigskip
+
+% \noindent \textbf{Figure \ref{Fig:ellipses}:}
 
@@
    \includegraphics[width=1\textwidth]{Figures/orthogonality_results.pdf}
-\caption{.}
-\label{Fig:orthogonality}
-\end{figure}
-\bigskip
-
-\noindent \textbf{Figure \ref{Fig:orthogonality}:}
-Amount of orthogonality of each clade's phylogenetic major axis of beak variation compared to their parent or parent's parent clade.
+\caption{\scriptsize{Amount of orthogonality of each clade's phylogenetic major axis of beak variation compared to their parent or parent's parent clade.
 The amount of orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background gray and dashed gray lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
 Dots represent the median orthogonality of each clade and the lines their 95\% confidence intervals (CI).
-No super-order's phylogenetic major axis of beak variation is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal (>  $45^\circ$) to the class-wide phylogenetic major axis of beak variation.
+No super-order's phylogenetic major axis of beak variation is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal ($>$  $45^\circ$) to the class-wide phylogenetic major axis of beak variation.
 These results are to be contrasted with the orientation of each order relative to their super-order where on average (median) only 14/27 are at least half orthogonal to the orientation of their super-orders’s phylogenetic major axis of beak variation.
@@
 This is the result of the variation in sample size (n) and standard variation (sd) among clades.
-A companion plot showing further nested structure within the Passeriformes is shown in Fig. \ref{fig_orthogonality_passeriformes}.
+A companion plot showing further nested structure within the Passeriformes is shown in Fig. \ref{fig_orthogonality_passeriformes}.}}
+\label{Fig:orthogonality}
+\end{figure}
+\bigskip
+
+% \noindent \textbf{Figure \ref{Fig:orthogonality}:}
 
@@
    \includegraphics[width=1\textwidth]{Figures/relative_EI_gg_style_unscaled.pdf}
-\caption{.}
-\label{Fig:relative_EI}
-\end{figure}
-\bigskip
-
-\noindent \textbf{Figure \ref{Fig:relative_EI}:}
-Comparisons of relative $elaboration_{species}$ and $innovation_{species}$ (Fig. \ref{Fig:phylogeny}) and $elaboration_{clade}$ and $innovation_{clade}$ (Fig. \ref{Fig:ellipses}).
+\caption{Comparisons of relative $elaboration_{species}$ and $innovation_{species}$ (Fig. \ref{Fig:phylogeny}) and $elaboration_{clade}$ and $innovation_{clade}$ (Fig. \ref{Fig:ellipses}).
 a) the $elaboration_{species}$ (purple) for each species in each 4000 posterior samples compared to the $elaboration_{clade}$ for each of the 35 clades (yellow).
@@
 b) as a) but using $innovation_{species}$ and $innovation_{clade}$ scores.
-For innovation the overlap has Bhattacharryya Coefficient = 0.039, which indicates a clear difference in the amount of $innovation_{species}$ and $innovation_{clade}$ across the posterior samples.
+For innovation the overlap has Bhattacharryya Coefficient = 0.039, which indicates a clear difference in the amount of $innovation_{species}$ and $innovation_{clade}$ across the posterior samples.}
+\label{Fig:relative_EI}
+\end{figure}
+\bigskip
+
+% \noindent \textbf{Figure \ref{Fig:relative_EI}:}
 
@@
 We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
-$\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each super-order.
+$\sigma$ PC1 $>$ $\sigma$ PC2 $>$ $\sigma$ PC3), this is not always true for each super-order.
 For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig.

### 2023-06-02 Manuscript/supplementaries.tex

@@
-\documentclass[12pt,letterpaper]{article}
+\documentclass[12pt,a4paper]{article}
 
@@
 
+\usepackage[margin=2cm]{geometry}
 \usepackage{lscape}
@@
 \usepackage{fullpage}
-% \usepackage{float}
+\usepackage{float}
 
@@
 
+\renewcommand{\thefigure}{S\arabic{figure}} 
+ \renewcommand{\thetable}{S\arabic{table}}
+
+%-------------------------------------------------------------------------------
+% Title page information
+%-------------------------------------------------------------------------------
+
+\title{Supplementary Materials from: \textit{Innovation and elaboration on the avian tree of life}; Science Advances, in review}
+
+\author{Thomas Guillerme, 
+Jen A. Bright,
+Christopher R. Cooney\\
+Emma C. Hughes,
+Zoe K. Varley,
+Natalie Cooper,
+Andrew P. Beckerman,\\
+and Gavin H. Thomas}
+
+\date{}
 %---------------------------------------------
@@
 \begin{document}
-%Running head
-\begin{flushright}
-Version dated: \today
-\end{flushright}
 
-\bigskip
-\medskip
-\begin{center}
+\maketitle
 
-\noindent{\Large \bf Innovation and elaboration on the avian tree of life - supplementary materials}
-\bigskip
+\tableofcontents
 
-\noindent {\normalsize \sc
-Thomas Guillerme$^{1,*}$, 
-Jen A. Bright$^{2}$,
-Christopher R. Cooney${^1}$,
-Emma C. Hughes${^1}$,
-Zoe K. Varley${^3}$,
-Natalie Cooper$^{3,+}$,
-Andrew P. Beckerman$^{1,+}$,
-and Gavin H. Thomas$^{1,4,+}$}\\
-\noindent {\small \it 
-$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
-$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
-
-\end{center}
-\bigskip
-\noindent{*\bf Corresponding author:} \textit{guillert@tcd.ie}; \noindent{+\bf contributed equally.}\\
-\vspace{1in}
-
-%Line numbering
-\modulolinenumbers[1]
-\linenumbers
-
-\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\
-
-\vspace{1.5in}
+\parindent = 1.5em
+\addtolength{\parskip}{.3em}
 
@@
 %
-%       INTRODUCTION
+%   Glossary   % NC: I think this is not really needed now the text is clearer, also some of the definitions are not 100% correct...
 %
@@
 
-\renewcommand{\thefigure}{S\arabic{figure}} 
- \renewcommand{\thetable}{S\arabic{table}}
+%\begin{landscape}
+%\section{Supplementary material 1: glossary}
 
-\begin{landscape}
-\section{Supplementary material 1: glossary}
+%We define three evolutionary scales based on the focal elements as follow:
+%\begin{itemize}
+%   \item \textbf{micro}evolutionary scale: the focal elements are \textbf{individual specimen};
+ %  \item \textbf{macro}evolutionary scale: the focal elements are \textbf{species} (i.e. groups of individuals);
+  % \item \textbf{mega}evolutionary scale: the focal elements are \textbf{clades} (i.e. groups of species).
+%\end{itemize}
 
-We define three evolutionary scales based on the focal elements as follow:
-\begin{itemize}
-   \item \textbf{micro}evolutionary scale: the focal elements are \textbf{individual specimen};
-   \item \textbf{macro}evolutionary scale: the focal elements are \textbf{species} (i.e. groups of individuals);
-   \item \textbf{mega}evolutionary scale: the focal elements are \textbf{clades} (i.e. groups of species).
-\end{itemize}
+%\begin{table}[ht]
+%\scriptsize
+%\caption{Glossary of some specific terms in this manuscript}
+%\begin{tabular}{p{0.2\textwidth}p{0.3\textwidth}p{0.3\textwidth}p{0.2\textwidth}p{0.2\textwidth}}
+%  \hline
+% Term in this paper         & definition in this paper & mathematical definition & in macroevolution & in microevolution\\
+%   \hline
+% evolutionary rate matrix &  matrix of rates of among-species divergence in species traits that explicitly incorporates phylogeny  &  & (\textbf{R} matrix of \cite{Houle2017}; \textbf{B} matrix of \cite{Machado2020}) & \\
+% major axis of variation    & the major evolutionary ``direction'' of trait co-evolution & first eigenvector of the evolutionary rate matrix  & allometry & line of genetic least resistance \\
+% clade                      & a monophyletic group of \textbf{species} & a subset of \textbf{shapespace} & Birds & a group of species or a population\\
+% species                    & the lowest evolutionary units in our analysis & a row in the \textbf{shapespace} & \textit{Pauxi pauxi} & an individual\\
+% shapespace                 & all the possible morphological combination (observed or not) & an ordinate and orthonormal matrix & the beak shapespace & a gene space \\
+% novelty                    & morphological variation (either \textbf{innovation} or \textbf{elaboration}) & position of a \textbf{species} or a \textbf{clade} relative to the \textbf{major axis of variation} & & \\
+% elaboration                & morphological variation along the \textbf{major axis of variation} & the algebraic projection on the eigenvector & & \\
+% innovation                 & morphological variation away from the \textbf{major axis of variation} & the algebraic rejection from the eigenvector & & \\
+% variance-covariance matrix & the multidimensional relation between traits & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) & & \\
+% nested phylogenetic level  & the different \textbf{clades} used in this study (the class-wide phylogeny $>$ super-orders $>$ orders) & different nested \textbf{variance-covariance matrices} & & \\
 
-\begin{table}[ht]
-\centering
-\scriptsize
-\begin{tabular}{p{0.2\textwidth}p{0.3\textwidth}p{0.3\textwidth}p{0.1\textwidth}p{0.1\textwidth}}
-  \hline
-Term in this paper         & definition in this paper & mathematical definition & in macroevolution & in microevoltion\\
-  \hline
-major axis of variation    & the major evolutionary ``direction'' of traits co-evolution & first eigenvector of the variance-covariance matrix  & allometry & line of genetic least resistance \\
-clade                      & a monophyletic group of \textbf{species} & a subset of \textbf{shapespace} & Birds & a group of species or a population\\
-species                    & the lowest evolutionary units in our analysis & a row in the \textbf{shapespace} & \textit{Pauxi pauxi} & an individual\\
-shapespace                 & all the possible morphological combination (observed or not) & an ordinate and orthonormal matrix & the beak shapespace & a gene space \\
-novelty                    & morphological variation (either \textbf{innovation} or \textbf{elaboration}) & position of a \textbf{species} or a \textbf{clade} relative to the \textbf{major axis of variation} & & \\
-elaboration                & morphological variation along the \textbf{major axis of variation} & the algebraic projection on the eigenvector & & \\
-innovation                 & morphological variation away from the \textbf{major axis of variation} & the algebraic rejection from the eigenvector & & \\
-variance-covariance matrix & the multidimensional relation between traits & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) & & \\
-nested phylogenetic level  & the different \textbf{clades} used in this study (the class-wide phylogeny $>$ super-orders $>$ orders) & different nested \textbf{variance-covariance matrices} & & \\
+% \end{tabular}
+% \label{tab_glossary}
+% \end{table}
+% \end{landscape}
 
-\end{tabular}
-\caption{Glossary of some specific terms in this manuscript}
-\label{tab_glossary}
-\end{table}
-\end{landscape}
+%---------------------------------------------
+%
+%   additional methods 
+%
+%---------------------------------------------
 
-\section{Supplementary material 2: additional results}
+\section{Additional methods details}
 
-\subsection{Data preparation}
-
-\begin{figure}[htbp]
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.9\textwidth]{Figures/mini-chains_diagram.pdf}
-\caption{mcmcmcglmmm: mini-chains MCMCglmm method diagram}
+\caption{mcmcmcglmmm: mini-chains MCMCglmm method diagram. See Methods for more details.}
 \label{Fig:mcmcmcglmm}
@@
 
-\begin{figure}[htbp]
+\newpage
+
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.9\textwidth]{Figures/axis_selection.pdf}
-\caption{Variance and cumulative variance for each 8 axis per super-order.
-Note that not all super orders have their own variance explained matching the variance explained by the entire ordination.
-For example, Strisores only need the first three dimensions to explain at least 95\% of their variance within the shapespace whilst Aequornithes need 8 dimensions to achieve at least the same amount of variance.}
+\caption{Variance and cumulative variance for each of the eight principal component (PC) axes for each super-order.
+Note that not all super-orders have their own variance explained matching the variance explained by the entire ordination.
+For example, Strisores only need the first three dimensions to explain at least 95\% of their variance within the shapespace whilst Aequornithes need all eight dimensions to achieve at least the same amount of variance explained.}
 \label{Fig:axes_variance}
@@
 
-\begin{figure}[htbp]
+\newpage
+
+\begin{figure}[H]
 \centering
@@
 \caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for all birds for the 4000 posteriors.
-\textbf{solution} is the solution of the model; \textbf{Gcovariance} are the covariances for the random terms (the phylogeny and the clades); \textbf{Rcovariance} are the residuals error term.}
+\textbf{solutions} is the solution of the model; \textbf{Gcovariances} is the covariance for the random terms (the phylogeny and the clades); \textbf{Rcovariances} is the residual error term.}
 \label{Fig:model_ess_all_birds}
@@
 \caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for the passeriformes for the 1870 posteriors.
-\textbf{solution} is the solution of the model; \textbf{Gcovariance} are the covariances for the random terms (the phylogeny and the clades); \textbf{Rcovariance} are the residuals error term.}
+\textbf{solutions} is the solution of the model; \textbf{Gcovariances} is the covariance for the random terms (the phylogeny and the clades); \textbf{Rcovariances} is the residual error term.}
 \label{Fig:model_ess_passeriformes}
@@
 
-\subsection{Additional results}
+%---------------------------------------------
+%
+%   additional results
+%
+%---------------------------------------------
 
-\begin{figure}[htbp]
+\newpage
+\section{Additional results}
+\subsection{Elaboration and innovation at the super-order or clade-wide phylogenetic level}
+
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.7\textwidth]{Figures/InnovElabTree_supp_revision.pdf}
-\caption{Avian phylogeny (n = 8748 species) showing distributions of species beak shape $elaboration_{species}$ (inner circle, yellow scale), $innovation_{species}$ (outer circle, blue scale), and Euclidean distance of species to the centroid of beak space (branches, cividis scale).}
+\caption{Avian phylogeny (n = 8748 species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
+Elaboration and innovation scores represent comparisons of species at the order, super-order and class-wide phylogenetic level.}
 \label{Fig:phylogeny_supplement}
@@
 
-\subsubsection{Correlations between $elaboration_{species}$ and $innovation_{species}$}
+%\newpage
 
-\begin{figure}[htpb]
+\subsection{Correlations between $elaboration_{species}$ and $innovation_{species}$}
+
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.5\textwidth]{Figures/correlations.pdf}
-\caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ at different focal levels.
-\textbf{A}: correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny.
+\caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ at different scales. 
+\textbf{(A)} Correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny. 
 The colours of the dots correspond to their density in the plot. 
 Blue and yellow colours corresponds to respectively low and high density of species. 
-The coloured quadrant in the back corresponds to $elaboration_{species}$ and $innovation_{species}$ scores higher than 1.
-The blue and yellow quadrants shows species with noticeable high $elaboration_{species}$ and $innovation_{species}$ respectively.
-The species in the green quadrant have noticeable high levels of both $elaboration_{species}$ and $innovation_{species}$.
-\textbf{B}: correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny (mega-level projection) but plotted by super-order and coloured by order (see list below).
+The coloured quadrant in the background corresponds to $elaboration_{species}$ and $innovation_{species}$ scores higher than 1. 
+The blue and yellow quadrants shows species with noticeably high $elaboration_{species}$ and $innovation_{species}$ respectively. 
+The species in the green quadrant have noticeably high levels of both $elaboration_{species}$ and $innovation_{species}$.
+\textbf{(B)} Correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on the whole phylogeny (mega-level projection) but plotted by super-order and coloured by order (see list below).
 \textbf{C} correlation between $elaboration_{species}$ and $innovation_{species}$ for each species when projected on their super-orders' major axis of variation (macro-level projection). 
 The colours correspond to each order. 
-For the distribution of correlation scores, see Fig.
-\ref{Fig:fable_correlations}.
-The groups numbers are:
-1 - others (groups with $<$ 15 species)
-2 - Corvoidea;
-3 - Malaconotoidea;
-4 - Orioloidea;
-5 - Meliphagoidea;
-6 - Bombycilloidea;
-7 - Muscicapoidea;
-8 - Emberizoidea;
-9 - Motacillidae;
-10 - Nectariniidae ;
-11 - Passeridae;
-12 - PloceidaeEstrildidae;
-13 - Eurylaimides;
-14 - Furnariida;
-15 - Tyrannida;
-16 - Aegithaloidea;
-17 - Cisticolidae;
-18 - Fringillidae;
-19 - Hirundinidae;
-20 - Locustelloidea;
-21 - Paridae;
-22 - Pycnonotidae;
-23 - Sylvioidea.}
+For the distribution of correlation scores, see Fig. \ref{Fig:fable_correlations}. 
+The group numbers are:
+1) others (groups with $<$ 15 species)
+2) Corvoidea;
+3) Malaconotoidea;
+4) Orioloidea;
+5) Meliphagoidea;
+6) Bombycilloidea;
+7) Muscicapoidea;
+8) Emberizoidea;
+9) Motacillidae;
+10) Nectariniidae;
+11) Passeridae;
+12) PloceidaeEstrildidae;
+13) Eurylaimides;
+14) Furnariida;
+15) Tyrannida;
+16) Aegithaloidea;
+17) Cisticolidae;
+18) Fringillidae;
+19) Hirundinidae;
+20) Locustelloidea;
+21) Paridae;
+22) Pycnonotidae;
+23) Sylvioidea.}
 \label{Fig:figure_correlations}
@@
 
-\begin{figure}[htpb]
+\newpage
+
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.9\textwidth]{Figures/correlations_2.pdf}
-\caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ for each species projected on their order's major axes.
-Each point is the median $elaboration_{species}$ and $innovation_{species}$ for one species from the distribution of 4000 posterior variance-covariance matrices.
-The full distribution of the correlation between $elaboration_{species}$ and $innovation_{species}$ is available in the main text in figure 3.}
+\caption{Median correlation plots of $elaboration_{species}$ and $innovation_{species}$ for each species projected on their order's phylogenetic major axis of beak variation.
+Each point is the median $elaboration_{species}$ and $innovation_{species}$ for one species from the distribution of 4000 posterior evolutionary rate matrices.
+The full distribution of the correlations between $elaboration_{species}$ and $innovation_{species}$ is available in the main text in Figure 3.}
 \label{Fig:figure_correlations2}
@@
 
-\begin{figure}[htpb]
+\newpage
+
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.9\textwidth]{Figures/correlations_fable.pdf}
-\caption{Posterior distributions of the correlations scores between $elaboration_{species}$ and $innovation_{species}$ as displayed for each super-orders in Fig.
+\caption{Posterior distributions of the correlation between $elaboration_{species}$ and $innovation_{species}$ for each super-orders in Fig.
 \ref{Fig:figure_correlations}.
-Species' projection onto the whole phylogeny's major axis of phenotypic variance-covariance (VCV) corresponds to the correlation scores for the panels A and B in Fig.
-\ref{Fig:figure_correlations} and species' projections onto their own super-order's major axis of phenotypic variance-covariance (VCV) corresponds to the correlation scores for the panel C in Fig.
-\ref{Fig:figure_correlations}.}
-\label{fable_correlations_supplementary}
+Species' projection onto the clade-wide phylogenetic major axis of beak variation corresponds to the correlation scores in Fig.
+\ref{Fig:figure_correlations}A-B and species' projections onto their own super-order's phylogenetic major axis of beak variation corresponds to the correlation scores in Fig.
+\ref{Fig:figure_correlations}C.}
+\label{Fig:fable_correlations}
 \end{figure}
 
-\subsubsection{Elaboration and innovation through time}
+\subsection{Elaboration and innovation through time}
 
-\begin{figure}[htpb]
+\begin{figure}[H]
 \centering
    \includegraphics[width=1\textwidth]{Figures/Elaboration_and_innovation_clade_through_time.pdf}
-\caption{Distribution of $elaboration_{clade}$ and $innovation_{clade}$ for each order and super-order through time.}
+\caption{Distribution of $elaboration_{clade}$ and $innovation_{clade}$ for each order and super-order through time in millions of years (Mya).}
 \label{Fig:figure_ei_clade_through_time}
@@
 
-\begin{figure}[htpb]
+\begin{figure}[H]
 \centering
    \includegraphics[width=1\textwidth]{Figures/Elaboration_and_innovation_species_through_time.pdf}
-\caption{Distribution of $elaboration_{species}$ and $innovation_{species}$ for each order and super-order through time.}
+\caption{Distribution of $elaboration_{species}$ and $innovation_{species}$ for each order and super-order through time in millions of years (Mya).}
 \label{Fig:figure_ei_species_through_time}
@@
 
-\subsubsection{Orientations in $elaboration_{clade}$ and $innovation_{clade}$}
+\newpage
 
-\begin{figure}[htpb]
+\subsection{Orientations in $elaboration_{clade}$ and $innovation_{clade}$}
+
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.5\textwidth]{Figures/50_random_axes_superorders.pdf}
-\caption{Distribution of 50 random scaled major axes (scaled variance-covariance matrices eigen values) for each super order illustrating that individual major axes can span in many different directions with some clades have a higher spread of direction than others.}
+\caption{Distribution of 50 random scaled major axes (scaled evolutionary rate matrices eigenvalues) for each super-order illustrating that individual phylogenetic major axes of beak variation can span many different directions and that some clades have a higher spread of directions than others.}
 \label{Fig:mikado}
@@
 
-\begin{figure}[htpb]
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.8\textwidth]{Figures/ellipses_supplementary.pdf}
-\caption{\scriptsize{Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
-We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
-The first inset ellipse plot shows the positions of all super-order ellipses relative to the global phylogenetic ellipse.
+\caption{\scriptsize{\textbf{(A)} Ellipses representing the scaled average posterior evolutionary rate matrix from the pGLMM models for each super-order (coloured ellipses) compared to the clade-wide phylogenetic component of the models (grey ellipses).
+We scaled the ellipses so the length of the phylogenetic major axis of beak variation of the clade ellipse is the same length as that of the clade-wide phylogenetic major axis ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all super-order ellipses relative to the clade-wide phylogenetic ellipse.
 Subsequent inset plots show the results for each super-order. 
-Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
+Inset bar plots show the proportion of variance associated with each of the eight principal component (PC) axes in shape morphospace.
 The inset boxplots correspond to the $elaboration_{clade}$ (E) and $innovation_{clade}$ (I) scores for all 4000 posterior samples.
@@
 These scores were calculated on the unscaled ellipses resulting in different scales of $elaboration_{clade}$ and $innovation_{clade}$ for each plot.
-Panel B) shows the results for each order.}}
+\textbf{(B)} Results for each order.}}
 \label{Fig:ellipses_rainbow}
@@
 
-\begin{figure}[htpb]
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.5\textwidth]{Figures/random_skewers.pdf}
-\caption{Random skewers test for the differences between the super orders (in purple) and the orders (in gold) variance-covariance matrices against the whole phylogeny ones.
-And, in light gold, between the orders and their super-orders variance covariance matrices.
+\caption{Random skewers test for the differences between the super-orders (in purple) and the orders (in gold) evolutionary rate matrices against the clade-wide matrices.
+And, in light gold, between the orders and their super-orders evolutionary rate matrices.
 The different line types (dashed and full) represent respectively the 95\% and the 75\% CI distribution and the dots represent the median skewer correlation estimate.
@@
 
+%---------------------------------------------
+%
+%   passeriformes results
+%
+%---------------------------------------------
+
+\newpage
 \clearpage
-\section{Supplementary material 3: Passeriformes results}
+\section{Passeriformes results}
 
-\begin{figure}[!htbp]
+\subsection{Passeriformes $elaboration_{species}$ and $innovation_{species}$}
+
+\begin{figure}[H]
 \centering
     \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement_revised.pdf}
-\caption{Passeriformes phylogeny coloured by beak shape innovation and elaboration level. See the caption of figure 2 in the main text for details.}
+\caption{Passeriformes phylogeny (n = ??
+species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
+Elaboration and innovation scores represent comparisons of species at the order, sub-order and family level.}
 \label{fig_phylogeny_passeriformes}
-\end{figure}
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{Figures/ellipses_passeriformes.pdf}
-\caption{Elaboration and innovation at the group level for each passeriformes sub-orders and families. See the caption of figure 2 in the main text for details.}
-\label{fig_ellipses_passeriformes}
-\end{figure}
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results_passeriformes.pdf}
-\caption{Posterior results showing the orientation of phenotypic evolution (orthogonality) for each passeriformes sub-orders and families. See the caption of figure 3 in the main text for details.}
-\label{fig_orthogonality_passeriformes}
 \end{figure}
@@
 
-\section{Supplementary materials 4: detailed projection and rejection operations}
+\subsection{Passeriformes $elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale}
+
+\begin{figure}[H]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/ellipses_passeriformes.pdf}
+\caption{$Elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale for each Passeriformes sub-order and family.
+The black and gray ellipses are the scaled average evolutionary rate matrices from the pGLMM models for respectively the clades (black) and the class-wide phylogeny (gray).
+The ellipses are centered on the position of the clade in the shapespace.
+The associated black bar plots represent the variance on each of the eight dimensions in the shapespace with the top gray bar plot representing the variance on each of the eight dimensions for the class-wide phylogeny.
+The orange and blue distributions represent respectively the distribution of the $elaboration_{clade}$ (orange) and the $innovation_{clade}$ (blue) scores for each clade and the dots are the median $elaboration_{clade}$ or $innovation_{clade}$ and the solid and the dashed lines representing the 50\% and the 95\% confidence intervales of the distribution of the scores across the 4000 posterior samples.
+The ticks on the y-axis always represent five arbitrary units of elaboration and innovation.}
+\label{fig_ellipses_passeriformes}
+\end{figure}
+
+\newpage
+
+\subsection{Passeriformes orthogonality}
+
+\begin{figure}[H]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/orthogonality_results_passeriformes.pdf}
+\caption{Amount of orthogonality of each Passeriformes sub-order's and family's phylogenetic major axis of beak variation compared to their parent or parent's parent clade.
+The amount of orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background gray and dashed gray lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
+Dots represent the median orthogonality of each clade and the lines their 95\% confidence intervals (CI).
+We also indicate the number of species (n) and the standard deviation (sd) of the orientation of their phylogenetic major axis of beak variation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
+For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the class-wide phylogenetic major axis of beak variation relative to their sample size and sd.
+The stars represent the posterior probability of the clade's orientation being different from the comparison clade (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).}
+\label{fig_orthogonality_passeriformes}
+\end{figure}
+
+%---------------------------------------------
+%
+%   maths
+%
+%---------------------------------------------
+
+\newpage
+
+\section{Detailed projection and rejection operations for measuring elaboration and innovation}
 \label{supp_projection}
 
-The following section contains the detailed definition and procedure of how we measured elaboration and innovation
+The following section contains the detailed definition and procedure of how we measured elaboration and innovation.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[H]
 \centering
@@
 
-\begin{figure}[htbp]
+\begin{figure}[H]
 \centering
    \includegraphics[width=0.9\textwidth]{Figures/respondability.pdf}
-\caption{Relation between respondability \cite{hansen2008measuring} and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
+\caption{Relation between respondability \cite{hansen2008measuring} and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{Charadriiformes} example dataset from the \texttt{dispRity} package.}
 \label{Fig:respondability}
@@
 
+%---------------------------------------------
+%
+%   references
+%
+%---------------------------------------------
+
 \bibliographystyle{naturemag}

### 2023-06-05 Manuscript/main.tex
Update main.tex
PGLS table
Added pgls table for BM results. Formatting needs fixing.

@@
 Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
+%@L_reworded_abstract
 
@@
 
-\begin{table}
-\begin{tabular}{ c c c }
- So & many & stars \\ 
- *** & *** & *** \\  
- *** & *** & *** \\
+\begin{landscape}
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrllllllllrr}
+  \hline
+ Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & delta AIC & Adj.
+R2 \\
+  \hline
+1 & 0.04*** & -0.139*** & 0.002 & -0.002 & 0.064*** & 0.045*** & 0.341*** & 0.093*** & 0.00 & 0.92 \\ 
+2 & 0.257*** & 0.169*** & 0.01 & 0.001 & 0.045*** &  & 0.312*** & 0.034*** & -691.48 & 0.92 \\ 
+3 & 0.108*** & 0.236*** & 0.012 & -0.002 & 0.055*** & 0.034*** & 0.343*** &  & -284.74 & 0.92 \\ 
+4 & 0.254*** & 0.104*** & 0.002 & 0.004 &  & 0.027*** & 0.311*** & 0.052*** & -877.85 & 0.92 \\ 
+5 & 0.1*** & 0.571*** & -0.018* & 0.02*** & 0.054*** & 0.036*** &  & 0.096*** & -1498.36 & 0.91 \\ 
+6 & 0.364*** & 0.973*** & -0.007 & 0.023*** &  &  &  &  & -2230.10 & 0.90 \\ 
+7 & 0.364*** & 0.972*** &  &  &  &  &  &  & -2280.08 & 0.90 \\ 
+8 &  &  & -0.025 & 0.011 &  &  &  &  & -22270.55 & -0.00 \\ 
+9 & 0.286*** &  &  &  &  &  &  &  & -18986.63 & 0.32 \\ 
+10 &  & 0.798*** &  &  &  &  &  &  & -17714.99 & 0.41 \\ 
+11 &  &  & -0.008 &  &  &  &  &  & -22269.34 & -0.00 \\ 
+12 &  &  &  & 0.002 &  &  &  &  & -22269.51 & -0.00 \\ 
+   \hline
 \end{tabular}
@@
 Fixing $\lambda$ is necessary to allow model comparison with AIC.
-We also fitted the same models fixing $\lambda$ to 0.7 (the lowest value of $\lambda$lambda from any individual model) and relative model fit and interpretation of parameters was unaffected.
+We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$lambda from any individual model) and relative model fit and interpretation of parameters was unaffected.
 E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
@@
 \end{table}
+\end{landscape}
 
@@
 
+\textbf{Author contributions.} G.H.T., A.P.B., N.C., and T.
+G.
+designed the study.
+J.A.B., C.R.
+C., E.
+C.H.,  and G.H.T.
+curated the data.
+T.G.
+analysed the data.
+All authors contributed to the writing of the manuscript.
+
+\textbf{Competing interests.} We declare no competing interests.
+
 \textbf{Funding.}

### 2023-06-05 Manuscript/main.tex
Fixing pgls table legend.

@@
   \hline
- Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & delta AIC & Adj.
-R2 \\
+ Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & $\Delta$AIC & Adj.
+$r^2$ \\
   \hline
@@
 \caption{Elaboration and innovation as predictors of beak shape distance from centroid.
-The table shows parameter estimates of predictors from  13 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms.
-Models 1-6 are multiple regressions including all variables and one or more interaction terms; models 7-9 are are multiple regressions including two or more variables and no interaction terms; models 10-13 are single predictor models.
+The table shows parameter estimates of predictors from  12 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms.
+Models 1-5 are multiple regressions including all variables and one or more interaction terms; models 6-8 are are multiple regressions including two or more variables and no interaction terms; models 9-12 are single predictor models.
 Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e.

### 2023-06-06 Manuscript/supplementaries.tex
Added table...

@@
 \usepackage{float}
+\usepackage{ltablex}
 
@@
    \includegraphics[width=0.5\textwidth]{Figures/parameters_ESS_passeriformes.pdf}
-\caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for the passeriformes for the 1870 posteriors.
+\caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for the passeriformes for the 4000 posteriors.
 \textbf{solutions} is the solution of the model; \textbf{Gcovariances} is the covariance for the random terms (the phylogeny and the clades); \textbf{Rcovariances} is the residual error term.}
@@
 
+\newpage
+
+\input{Tables/tab_ortho_results.tex}
+
 \subsection{Elaboration and innovation through time}
@@
    \includegraphics[width=0.5\textwidth]{Figures/50_random_axes_superorders.pdf}
-\caption{Distribution of 50 random scaled major axes (scaled evolutionary rate matrices eigenvalues) for each super-order illustrating that individual phylogenetic major axes of beak variation can span many different directions and that some clades have a higher spread of directions than others.}
+\caption{Distribution of 50 random scaled major axes (scaled evolutionary rate matrices eigenvectors) for each super-order illustrating that individual phylogenetic major axes of beak variation can span many different directions and that some clades have a higher spread of directions than others.}
 \label{Fig:mikado}
@@
 Subsequent inset plots show the results for each super-order. 
-Inset bar plots show the proportion of variance associated with each of the eight principal component (PC) axes in shape morphospace.
+Inset bar plots show the proportion of variance associated with each of the eight principal component (PC) axes in shapespace.
 The inset boxplots correspond to the $elaboration_{clade}$ (E) and $innovation_{clade}$ (I) scores for all 4000 posterior samples.
@@
     \includegraphics[width=0.9\textwidth]{Figures/InnovElabTree_passerine_supplement_revised.pdf}
-\caption{Passeriformes phylogeny (n = ??
-species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
+\caption{Passeriformes phylogeny (n = 5229 species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
 Elaboration and innovation scores represent comparisons of species at the order, sub-order and family level.}
@@
    \includegraphics[width=1\textwidth]{Figures/ellipses_passeriformes.pdf}
-\caption{$Elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale for each Passeriformes sub-order and family.
-The black and gray ellipses are the scaled average evolutionary rate matrices from the pGLMM models for respectively the clades (black) and the class-wide phylogeny (gray).
-The ellipses are centered on the position of the clade in the shapespace.
-The associated black bar plots represent the variance on each of the eight dimensions in the shapespace with the top gray bar plot representing the variance on each of the eight dimensions for the class-wide phylogeny.
-The orange and blue distributions represent respectively the distribution of the $elaboration_{clade}$ (orange) and the $innovation_{clade}$ (blue) scores for each clade and the dots are the median $elaboration_{clade}$ or $innovation_{clade}$ and the solid and the dashed lines representing the 50\% and the 95\% confidence intervales of the distribution of the scores across the 4000 posterior samples.
-The ticks on the y-axis always represent five arbitrary units of elaboration and innovation.}
+\caption{\textbf{(A)} Ellipses representing the scaled average posterior evolutionary rate matrix from the pGLMM models for each passeriformes sub-orders and families (coloured ellipses) compared to the passeriformes phylogenetic component of the models (grey ellipses).
+We scaled the ellipses so the length of the phylogenetic major axis of beak variation of the clade ellipse is the same length as that of the passeriformes phylogenetic major axis ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all sub-order ellipses relative to the passeriformes phylogenetic ellipse.
+Subsequent inset plots show the results for each sub-order. 
+Inset bar plots show the proportion of variance associated with each of the eight principal component (PC) axes in shapespace.
+The inset boxplots correspond to the $elaboration_{clade}$ (left) and $innovation_{clade}$ (right) scores for all 4000 posterior samples.
+The dots represent the median $elaboration_{clade}$ and $innovation_{clade}$ values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
+These scores were calculated on the unscaled ellipses resulting in different scales of $elaboration_{clade}$ and $innovation_{clade}$ for each plot.
+\textbf{(B)} Results for each families.}
 \label{fig_ellipses_passeriformes}
@@
    \includegraphics[width=0.9\textwidth]{Figures/respondability.pdf}
-\caption{Relation between respondability \cite{hansen2008measuring} and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{Charadriiformes} example dataset from the \texttt{dispRity} package.}
+\caption{Relation between respondability \cite{hansen2008measuring} and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
 \label{Fig:respondability}

### 2023-06-06 Manuscript/main.tex
Update main.tex

@@
 Emma C. Hughes${^1}$,
-% Zoe K. Varley${^3}$,
+Zoe K. Varley${^3}$,
 Natalie Cooper$^{3,+}$,
@@
 The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales.
+%@L_Rmatrix_full
 
@@
 \subsection{Modeling nested trait covariance}
-Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of phenotypic variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
+Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of beak shape variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
 To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
@@
 \ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}).
+%@L_ei_through_time_plots
 These results are consistent with both divergence in phenotype that is proportional to time (e.g.
@@
 \ref{Fig:orthogonality}).
-Comparisons of orientation among subclades (i.e.
+Comparisons of orientation am%@L_ei_through_time_plotsong subclades (i.e.
 orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig.
@@
 In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
-These observations illustrate multiple routes to $innovation_{clade}$ and imply that within clades, species phenotypes can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
+These observations illustrate multiple routes to $innovation_{clade}$ and imply that within clades, species beak shapes can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.
 
-\subsection{Species elaborate \textit{and} innovate, clades innovate} %TG: TODO: note in the response letter that we've now nuanced this
+\subsection{Species elaborate \textit{and} innovate, clades innovate} %TG: @L_TLDR_nuance
 The disconnect between observations of reorientation of trait space at the macroevolutionary scale, and an apparent predominance of $elaboration_{species}$ could be viewed as a paradox.
@@
 Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space.
+%@L_PGLS_results
 
@@
 \centering
+\small
    \begin{tabular}{rrllllllllrr}
@@
 The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection} - note that the procedure described in the section \ref{supp_projection} is applied to each individual evolutionary rate matrix for each term in the model, e.g.
-for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score %@L_individual matrices).
+for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score.
+%@L_individual_matrices).
 Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.

### 2023-06-06 Manuscript/main.tex
Update main.tex

@@
 \subsection{Macroevolution: elaboration is a common route to divergence among species}
-We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along phylogenetic lines of least resistance by calculating species-specific measures of $elaboration_{species}$ and $innovation_{species}$.
+We tested the expectation, derived from adaptive radiation theory \cite{schluter1996adaptive,mcglothlin2018adaptive,slater2019hierarchy} %@L_cite_adaptive
+, that species divergence is biased along phylogenetic lines of least resistance by calculating species-specific measures of $elaboration_{species}$ and $innovation_{species}$.
 If lines of least resistance are stable among species then we expect that species typically fall on, or close to, a conserved phylogenetic major axis of beak variation ($elaboration_{species}$) rather than away from the axis ($innovation_{species}$).

### 2023-06-07 Manuscript/main.tex
Update from revisions

@@
 While there is no strict definition that separates macro- and mega-evolution, we define the macroevolutionary scale as species divergence within clades and the megaevolutionary scale as divergence of clades within larger clades.
+%@L_mega_macro_def
 The diversity of bird beaks has been well-studied at broad scales, particularly in relation to dietary ecology \cite{olsen2017feeding, felice2019, bright2016shapes, navalon2019, pigot2020macroevolutionary}.
@@
 The major axis can be identified from the leading eigenvector of the posterior distribution of evolutionary rate matrices.
-We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
+We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{steppan2002comparative, robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
 This major axis of beak variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
@@
 \ref{Fig:ellipses} - Fig.
-\ref{Fig:orthogonality}).
+\ref{Fig:orthogonality}; note that these comparisons are based on the evolutionary rate matrix).
+%@L_orientation_phylo
 The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig.
 \ref{Fig:orthogonality}).
-Comparisons of orientation am%@L_ei_through_time_plotsong subclades (i.e.
+Comparisons of orientation among subclades (i.e.
 orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig.
@@
 So, although species often evolve preferentially along a shared phylogenetic major axes of beak variation within clades, there is frequent deviation from these major axes, and the phylogenetic major axes of beak variation change among clades.
+This could be explained in part because the evolutionary rate matrix (\textbf{R}) is more influenced by shifts in the adaptive landscape through time and thus reflecting natural selection in deeper nodes more than in younger nodes (Fig.
+\ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}).
+%@L_R_selection
 
@@
 
-\textbf{Phylogenetic generalised least squares analysis.} We ran phylogenetic generalised least squares models (PGLS) to test the effects of elaboration and innovation on beak shape divergence.
+\textbf{Phylogenetic generalised least squares analysis.} %@L_pgls_methods
+We ran phylogenetic generalised least squares models (PGLS) to test the effects of elaboration and innovation on beak shape divergence.
 Beak shape divergence was calculated as the Euclidean distance of each species beak shape from the centroid of beak shape of all birds.

### 2023-06-07 Manuscript/main.tex
minor updates

@@
 \textbf{Acknowledgements.}
+We thank M. Adams, H. van Grouw and R. Prys-Jones from the Bird Group at the NHM, Tring and H. McGhie at the Manchester Museum for providing access to and expertise in the collections.
+We are indebted to the volunteer citizen scientists at \url{http://www.markmybird.org} for helping to build the database of bird bill shape and contribute to our understanding of avian evolution. 
 We thank Masahito Tusboi and three other anonymous reviewers for their suggestions that greatly helped improving the manuscript.
@@
 designed the study.
-J.A.B., C.R.
-C., E.
-C.H.,  and G.H.T.
+J.A.B., C.R.C., E.C.H., Z.K.V., and G.H.T.
 curated the data.

### 2023-06-07 Manuscript/supplementaries.tex
minor updates

@@
 \ref{Fig:figure_correlations}C.}
-\label{Fig:fable_correlations}
+\label{Fig:fable_correlations_supplementary}
 \end{figure}
@@
 The stars denote the median p-value estimate (0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1).}
-\label{/home/tguillerme/Projects/elaboration_exploration_bird_beaks/Manuscript/Figures/others/figure_phylo_spiro_dark.png
-/home/tguillerme/Projects/elaboration_exploration_bird_beaks/Manuscript/Figures/others/orthogonality_results.png
-/home/tguillerme/Projects/elaboration_exploration_bird_beaks/Manuscript/Figures/others/relative_EI_gg_style.png}
+\label{Fig:random_skewers}
 \end{figure}

### 2023-06-07 Manuscript/main.tex

@@
 To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
-To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves) (Fig.
+To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves; Fig.
  \ref{Fig:cheatsheet} and \ref{Fig:mcmcmcglmm}).
@@
 Phylogenetic distributions of $elaboration_{species}$ and $innovation_{species}$ broadly hold whether comparisons are made at the order, super-order ($elaboration_{species}$ Pagel's $\lambda$ = 0.853; $innovation_{species}$ Pagel's $\lambda$ = 0.914), or class-wide phylogenetic level ($elaboration_{species}$ Pagel's $\lambda$ = 0.935; $innovation_{species}$ Pagel’s $\lambda$ = 0.837; Fig.
-\ref{Fig:phylogeny_supplement}); \ref{Fig:figure_correlations} %@L_correlation_figure_link)
+\ref{Fig:phylogeny_supplement} and \ref{Fig:figure_correlations}) %@L_correlation_figure_link)
 and in further nested analysis of the order Passeriformes (Fig. \ref{fig_ellipses_passeriformes} and \ref{fig_phylogeny_passeriformes}).
@@
 While $elaboration_{species}$ is the main mode of beak shape divergence at the macroevolutionary scale %@L_dominant
-(Fig. \ref{Fig:phylogeny}), it is also commonly positively correlated with $innovation_{species}$ (Fig. \ref{fable_correlations} \ref{fable_correlations_supplementary}. %@L_correlation_figure_link).
+(Fig. \ref{Fig:phylogeny}), it is also commonly positively correlated with $innovation_{species}$ (Fig. \ref{fable_correlations} and S8). %@L_correlation_figure_link).
 In 14/27 clades we find significant positive correlations with non-significant but positive trends in a further 11 clades.
@@
 We then used linear algebra to project %@L_linear_algebra_stuff
-a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig.S\ref{Fig:cheatsheet}).
+a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. S\ref{Fig:cheatsheet}).
 We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ scor.
@@
 Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig.
-\ref{Fig:ellipses} - Fig.
+\ref{Fig:ellipses}; Fig.
 \ref{Fig:orthogonality}; note that these comparisons are based on the evolutionary rate matrix).
 %@L_orientation_phylo
-The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig.
+The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$; Fig.
 \ref{Fig:orthogonality}).
 Comparisons of orientation among subclades (i.e.
-orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig.
+orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$; Fig.
 \ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary scale and are no more likely to occur along any one axis than another.
@@
 We ran the PGLS analysis using both single and multiple predictors.
-Our analyses indicate that the majority of variation (r2 = 0.888; Table \ref{Tab:pgls}) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
-When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table 1).
-In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table 1).
+Our analyses indicate that the majority of variation (r$^2$ = 0.888; Table \ref{Tab:pgls}) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
+When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table \ref{Tab:pgls}).
+In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table \ref{Tab:pgls}).
 However, $innovation_{clade}$ is nonetheless an important contributor to total beak shape space.
 We reach this conclusion because models with interactions between species and clade metrics provide by far the best model fit overall and removal of terms indicate that the most important interactions are between $innovation_{clade}$  and $innovation_{species}$ followed by $innovation_{clade}$ and $elaboration_{species}$.
-The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table 1 \ref{Tab:pgls})
+The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table \ref{Tab:pgls}).
 Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space.
@@
 Fixing $\lambda$ is necessary to allow model comparison with AIC.
-We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$lambda from any individual model) and relative model fit and interpretation of parameters was unaffected.
+We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$ from any individual model) and relative model fit and interpretation of parameters was unaffected.
 E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
@@
 These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
-We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
+We used a threshold of at least 95\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e.
 $\sigma$ PC1 $>$ $\sigma$ PC2 $>$ $\sigma$ PC3), this is not always true for each super-order.
 For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig.
-\ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
+\ref{Fig:ellipses} and \ref{Fig:axes_variance}).
 This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. \ref{Fig:axes_variance}).
@@
 In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional, i.e within the 95\% confidence interval of the evolutionary rate matrix.
-The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection} - note that the procedure described in the section \ref{supp_projection} is applied to each individual evolutionary rate matrix for each term in the model, e.g.
-for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score.
+The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection};  note that the procedure described is applied to each individual evolutionary rate matrix for each term in the model, e.g.
+for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score).
 %@L_individual_matrices).
-Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.
+Code to reproduce the analyses is available in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.
 
@@
 These results are available in Fig. \ref{Fig:orthogonality}.
-Note that we did not use the random skewers method here \cite{cheverud2007research} because we were interested in measuring the amount of orthogonality between pairs of evolutionary rate matrices rather than simply measuring whether they were different (but see supplementary materials for random skewers correlation results \ref{Fig:random_skewers}).
+Note that we did not use the random skewers method here \cite{cheverud2007research} because we were interested in measuring the amount of orthogonality between pairs of evolutionary rate matrices rather than simply measuring whether they were different (but see supplementary materials Fig.
+S13 for random skewers correlation results).
 %@L_noskewers
@@
 
-\textbf{Reproducibility and repeatability.}
+\section*{Reproducibility and repeatability}
 The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
@@
 
-\textbf{Acknowledgements.}
+\section*{Acknowledgements}
 We thank M. Adams, H. van Grouw and R. Prys-Jones from the Bird Group at the NHM, Tring and H. McGhie at the Manchester Museum for providing access to and expertise in the collections.
@@
 
-\textbf{Author contributions.} G.H.T., A.P.B., N.C., and T.
+\section*{Author contributions} G.H.T., A.P.B., N.C., and T.
 G.
@@
 
-\textbf{Competing interests.} We declare no competing interests.
+\section*{Competing interests} We declare no competing interests.
 
-\textbf{Funding.}
+\section*{Funding}
 This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).

### 2023-06-08 Manuscript/supplementaries.tex
Fix long table
Update supplementaries.tex

@@
 \usepackage{ltablex}
+\usepackage{longtable}
+\usepackage[tablewithin=none,figurewithin=none,singlelinecheck=false]{caption}
 
@@
 
+ % GT Figure legend needs explanation of whole_space and no_calde
 \begin{figure}[H]
@@
    \includegraphics[width=0.5\textwidth]{Figures/parameters_ESS_passeriformes.pdf}
-\caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for the passeriformes for the 4000 posteriors.
+\caption{Histogram of the effective sample sizes (ESS) for each class of parameters in the MCMCglmm model for the Passeriformes for the 4000 posteriors.
 \textbf{solutions} is the solution of the model; \textbf{Gcovariances} is the covariance for the random terms (the phylogeny and the clades); \textbf{Rcovariances} is the residual error term.}
@@
 
+\begin{landscape}
 \input{Tables/tab_ortho_results.tex}
+\end{landscape}
+
+\newpage
 
@@
    \includegraphics[width=1\textwidth]{Figures/ellipses_passeriformes.pdf}
-\caption{\textbf{(A)} Ellipses representing the scaled average posterior evolutionary rate matrix from the pGLMM models for each passeriformes sub-orders and families (coloured ellipses) compared to the passeriformes phylogenetic component of the models (grey ellipses).
-We scaled the ellipses so the length of the phylogenetic major axis of beak variation of the clade ellipse is the same length as that of the passeriformes phylogenetic major axis ellipse (in eight dimensions).
-The first inset ellipse plot shows the positions of all sub-order ellipses relative to the passeriformes phylogenetic ellipse.
+\caption{\textbf{(A)} Ellipses representing the scaled average posterior evolutionary rate matrix from the pGLMM models for each Passeriformes sub-orders and families (coloured ellipses) compared to the Passeriformes phylogenetic component of the models (grey ellipses).
+We scaled the ellipses so the length of the phylogenetic major axis of beak variation of the clade ellipse is the same length as that of the Passeriformes phylogenetic major axis ellipse (in eight dimensions).
+The first inset ellipse plot shows the positions of all sub-order ellipses relative to the Passeriformes phylogenetic ellipse.
 Subsequent inset plots show the results for each sub-order. 

### 2023-06-08 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/correlations_fable.pdf}
-\caption{Posterior distributions of the correlation between $elaboration_{species}$ and $innovation_{species}$ for each super-orders in Fig.
-\ref{Fig:figure_correlations}.
-Species' projection onto the clade-wide phylogenetic major axis of beak variation corresponds to the correlation scores in Fig.
-\ref{Fig:figure_correlations}A-B and species' projections onto their own super-order's phylogenetic major axis of beak variation corresponds to the correlation scores in Fig.
-\ref{Fig:figure_correlations}C.}
+   \includegraphics[width=0.9\textwidth]{Figures/correlations_fable_super_orders.pdf}
+\caption{Posterior distributions of the correlation between $elaboration_{species}$ and $innovation_{species}$ for each super-orders in Fig. \ref{Fig:figure_correlations}.
+A) Species' projection onto the class-wide phylogenetic major axes of beak variation (corresponding to the correlation scores in Fig.
+\ref{Fig:figure_correlations}A-B); B) Species' projection onto the super-orders' phylogenetic major axes of beak variation (corresponding to the correlation scores in Fig.
+\ref{Fig:figure_correlations}C).}
 \label{Fig:fable_correlations_supplementary}

### 2023-06-08 Analysis/06-Figure_with_correlations.Rmd
Update 06-Figure_with_correlations.Rmd

@@
                 y.axis = TRUE, plot.phylo = TRUE,
-                main = "Species projected onto\nthe whole phylogeny major axes") 
+                main = "A") 
 par(mar = c(5, 2, 4, 6) + 0.1, bty = "n") # c(bottom, left, top, right)
@@
                 y.axis = FALSE, plot.phylo = FALSE,
-                main = "Species projected onto\ntheir superorders' major axes")
+                main = "B")
 dev.off()

### 2023-06-08 Manuscript/reponse_letter.tex
Create reponse_letter.tex

\documentclass[12pt,letterpaper]{article}

%Packages
% \usepackage{textcomp}
% \usepackage{latexsym}
% \usepackage{url}
% \usepackage{amssymb}
% \usepackage{amsmath}
% \usepackage{mathtools}
% \usepackage{bm}
% \usepackage{array}
% \usepackage[version=3]{mhchem}
% \usepackage{ifthen}
% \usepackage{amsthm}
% \usepackage{amstext}
% \usepackage{enumerate}
% \usepackage{dcolumn}

\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{pdflscape}
\usepackage{mathtools}
\usepackage[osf]{mathpazo}
\usepackage{fullpage}
\usepackage{float}
\usepackage{xr} %linking to supplementaries
\usepackage{xcolor}
\externaldocument{supplementaries}

\pagenumbering{arabic}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}
\noindent \today
\bigskip

\noindent Dear editor,
\bigskip

We are grateful for the detailed and useful comments from all four reviewers. We have taken all their suggestions on board, integrated changes associated with the suggestion into the manuscript and responded to them in detail below. For clarity we have displayed \textcolor{blue}{the reviewers’ comments in blue} and our own responses in black.

We have revised  the manuscript taking advantage of the extra space and number of figures allowed for Science Advances. Briefly, the major changes are:

\begin{itemize}
    \item We now clearly distinguish the two types of elaboration and innovation: $elaboration_{species}$ and $innovation_{species}$ that focuses on the macro-evolutionary scale and $elaboration_{clades}$ and $innovation_{clades}$ that focuses on the mega-evolutionary scale;
    \item For the analyses on elaboration and innovation at the macro scale ($elaboration_{species}$ and $innovation_{species}$), we now only focus on the projection of species onto their order’s major axes;
    \item We added two new figures and a table in the main text: the cheat sheet figure (previously in the supplementaries, now in the introduction), the correlation figure and a table from the new PGLS results; 
    \item We have added several new analyses to the supplementary materials including: correlations between elaboration/innovation, elaboration and innovation through time and random skewers analysis;
    \item We have linked our metrics definitions to previous work in the supplementary materials;
    \item We have made the vocabulary more consistent throughout the manuscript;
    \item We fully rewrote the introduction and most of the results/discussion section.
\end{itemize}

We fully describe each change below.
\bigskip
Best regards,
\bigskip
Thomas Guillerme, on behalf of all co-authors.

\section*{\textcolor{blue}{Reviewer 1}}

\textcolor{blue}{This study is an important contribution to the field of phenotypic macroevolution, and I strongly recommend its publication. The conceptualisation is very interesting, and I am aware that it represents the culmination of many years of research by the group/s involved, whose contributions I have seen several times along the years in several conferences. The authors devise a smart new method to evaluate the contributions of elaboration (i.e., evolutionary change along the major axis of variation of a clade) against innovation (i.e., evolutionary change away from this major axis).}

\textcolor{blue}{However, I have one major problem with the interpretation of the results. I will use an example to illustrate my point:}

\textcolor{blue}{Lines 108-110 go: ‘For example, compared to the global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation’.}

\textcolor{blue}{One could argue that the beak shape of hummingbirds represents one of the most exceptional cases of niche expansion and beak shape innovation in birds (Cooney, Bright et al., 2017, Nature). We know from the fossil record that hummingbirds likely evolved from broad-billed ancestors (i.e., stem trochilid Parargornis) so in reality, this comparison might not faithfully represent the actual evolutionary process undergone by crown group hummingbirds (high innovation as it departs from the main axis of evolutionary variation within strisoreans). Therefore, I am not sure I see what species values of ‘innovation’ and ‘elaboration’ in this context (i.e., compared to the global axis of whole-Neornithes variation) are telling us about the actual underlying innovation and elaboration as real evolutionary forces.}

\textcolor{blue}{I think the appropriate comparison for species values of innovation and elaboration might be with the main axes of variation of the immediately superior phylogenetic scale (order level in this case). Comparisons of species level values derived from the main axis of variation at the level of whole Neornithes might reflect global constraints in beak shape. For instance, coming back to the hummingbirds, there is some evidence that the divergence in cranial shape between hummingbirds and the remaining strisoreans might be related with a reversal to an ontogenetic trajectory that is pervasive among non- strisorean birds (Navalon et al., 2021, Proc. B). The similarities in beak shape between hummingbird species and the main axis of beak shape variation (long and slender to short and stout) in this study might reflect general constraints in beak shape variation rather than anything else. I think therefore conflating species level data from the whole-Neornithes axis with species level data from the order or parvorder axes might be confounding two kinds of data, different in nature, although, admittedly, it is intriguing that the same pattern (elaboration higher than innovation) holds at all the different phylogenetic levels.}

\textcolor{blue}{I suggest therefore keeping only the species level values as compared with the main axis of variation at the order/parvorder level as indicating ‘true’ innovation and elaboration. The comparisons of the main axis between different clades, in my opinion, indicate deep divergences in the main axes of beak shape variation. So I suggest keeping those too. Then either ditching the comparisons of species level dispersion values with whole- Neornithes or reinterpreting them.}

\textcolor{blue}{As a result, I am not entirely convinced with the authors conclusion that accumulated gradual elaboration creates deep divergences among main axes of variation in clades. I might be wrong, but my reading of the presented data is that:
\begin{enumerate}
\item Deep divergences among clades create different patterns of beak shape variation in different orders.
\item Species level beak shape variation within clades is characterised by more elaboration than innovation – evolution along lines of least resistance.
\item Species level variation as compared with the main axis of beak shape variation in whole Neornithes indicates that beak shapes are affected by whole-clade constraints – possibly of developmental nature. Calling this innovation and elaboration and equating it to the concepts in bullet point 2 is confusing in my opinion because it is several levels detached from the actual evolutionary process creating ‘elaboration’ or ‘innovation’.
\end{enumerate}
I reckon my interpretation is aligned with what earlier work by the authors (Cooney, Bright et al., 2017 and more by other authors!) suggest and adds to the discussion by showing the very interesting result that evolutionary variation in beak shape is largely constrained at several levels. But I don’t think it really contributes to ascertaining why deep divergences among clades occur unless further analyses can be carried out.}

We are grateful to the reviewer for pointing out this concern so clearly. We have now clarified what we mean by innovation and elaboration throughout the manuscript. We now consider the following:
\begin{itemize}
    \item Elaboration and innovation at the macroevolutionary scale for each species ($elaboration_{species}$ and $innovation_{species}$) as the projection/rejection of individual species on their order’s major axis of trait variance-covariance. We have moved the projections/rejection of species onto higher taxonomic levels (super-orders and the whole phylogeny) into the supplementary materials.
    \item Elaboration and innovation at the megaevolutionary scale (sensu Simpson 1944, 1953) $elaboration_{clade}$ and $innovation_{clade}$) as the projection/rejection of entire clades (orders or super-orders) onto their higher clades (super-order or the whole phylogeny).
\end{itemize}
We have reframed the whole manuscript and analyses with this in mind and have completely rewritten the introduction. Note that our previous version of the manuscript was a submission to Science and was directly transferred to Science Advances. We have now fully taken advantage of the extra space available which hopefully increases the clarity of the manuscript. This re-framing should remove the issues raised by the reviewer regarding the interpretations of our results.

We now also directly refer to the example suggested by this reviewers on lines 73-77:%@L_strisores_example:

\noindent\textit{For example, a major shift in beak shape arises in the early divergence of Strisores separating hummingbirds (Trochilidae) from swifts (Apodidae), perhaps marking changes in ontogenetic trajectories \cite{navalon2021} as hummingbirds diverged from broad billed ancestors (e.g. the Eocene stem Trochilidae fossils \textit{Parargornis messelensis} and \textit{Eocypselus rowei} \cite{mayr2003new, ksepka2013fossil}).} %@L_strisores_example

\textbf{\textcolor{blue}{More detailed comments:}}

\textcolor{blue}{I think Figure S5 – or a simplified version of it - should be Figure 1 in the main text to guide the readers through the methods and the theoretical framework used in the study. This will be beneficial too for the published paper to be noticed by the community – which I think it might be desirable because it is a very interesting contribution.}

Thanks for the suggestion. We have now reworked Figure S5 and use it as a “cheat sheet” in the introduction to guide the reader through the results of the original figures 1 and 2 (now figures 2 and 4).

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.6\textwidth]{Figures/cheat_sheet.pdf}
\caption{\scriptsize{Relationships between key figures in the main text.
A) represents the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
The gray ellipse represents the overall phylogenetic major axis of beak variation in birds.
Species aligned or away from this axis are respectively elaborators or innovators relative to the class Aves.
The red ellipse and axis represents the phylogenetic major axis of beak variation in Tinamiformes only.
Species aligned or away from this axis are respectively elaborators and innovators relative to the Tinamiformes order.
B) In Figure 2, we measured elaboration and innovation at the macroevolutionary scale , i.e. the position of each species in terms of $elaboration_{species}$ and $innovation_{species}$ relative to their order's phylogenetic major axis of beak variation.
A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation(i. e. their position on that axis) whereas their $innovation_{species}$ score is their rejection from it (i.e. their distance away from that axis).
The colours on the tips of the phylogeny correspond to the median $elaboration_{species}$ (orange gradient) and $innovation_{species}$ (blue gradient).
C) Figure 4, we measured elaboration and innovation at the megaevolutionary scale, i.e. $elaboration_{clade}$ and $innovation_{clade}$.
The ellipses are scaled and centered and represent the clade's phylogenetic major axis of beak variation relative to the overall phylogenetic major axis of beak variation with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis of beak variation onto the overall phylogenetic major axis of beak variation and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
The distribution elaboration and innovation scores are taken from the projection of the 4000 pairs of evolutionary rate matrices with each pair being the focal level, e.g. order,and the parent level, e.g. the whole bird phylogeny.}}
\label{Fig:cheatsheet}
\end{figure}

\newpage

\textcolor{blue}{The abstract could benefit from briefly stating (e.g., a parenthesis) what innovation and elaboration are in this context. Similarly, the introduction would benefit from defining very clearly what is meant by micro- macro- and mega-evolution. I guess mega- in this context refers to divergence among major avian clades (order level or similar). Also bear in mind microevolution is within species so it is of limited utility here.}

We have now completely rewritten the abstract and introduction, ensuring we clearly introduce the concepts of innovation and elaboration, and mega-, macro- and micro- evolution. We defined the mega-, macro- and micro- evolutionary scales as the scales focusing respectively on the clades, species and individuals. We also link these concepts to the literature more thoroughly. Note that our previous version was a direct transfer from a journal with a more limited page count, but having more space to explain our approach has improved the manuscript substantially (lines 5-16):

\noindent\textit{Widely documented, megaevolutionary jumps in phenotypic diversity continue to perplex researchers because it remains unclear whether these dramatic changes can emerge from microevolutionary processes.
Here we tackle this question using new approaches for modeling multivariate traits %@L_multi_traits
 to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
We find that elaboration, evolution along the major axis of phenotypic change, is common at both macro- and megaevolutionary scales whereas innovation, evolution away from the major axis of phenotypic change, is more prominent at megaevolutionary scales.
Indeed, the major axis of phenotypic change among species beak shapes at megaevolutionary scales is an emergent property of innovation across clades.
Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.}

\textcolor{blue}{Lines 104-105: ‘We find no consistent evidence that innovation is either positively correlated, or trades-off, with elaboration’. Have you formally tested this? Maybe it would be good to show a dot plot of this for species values.}

We have now added a new analysis to the supplementary materials where we show the correlation between $innovation_{species}$ and $elaboration_{species}$ and we link to this figure in the main text (Figure 3). We find some variability in this relationship depending whether innovation and elaboration are calculated from comparisons to orders, suborders, or the whole tree. At the order level (our main focus in the new version of the manuscript) we find that positive correlations are common and negative correlations (trade-offs) are absent.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/correlations_fable_orders.pdf}
\caption{Posterior correlations between $elaboration_{species}$ and $innovation_{species}$ for each super-order. The dots, thick lines and dashed lines represent respectively the median, 50\% confidence interval (CI) and 95\% CI scores. 14 out of 27 of the orders have a clear posterior correlation between $elaboration_{species}$ and $innovation_{species}$ (i.e. 95\% CI does not overlap with 0), nine have a somewhat positive posterior correlation (i.e. 50\% CI does not overlap with 0) and four have no clear posterior correlation. Note that no order has a clear negative posterior correlation. This trend suggests that there is no tradeoff between $elaboration_{species}$ and $innovation_{species}$ and that both can be common routes for beak shape evolution at the macroevolutionary scale.
}
\label{fable_correlations}
\end{figure}
\bigskip

\newpage

\textcolor{blue}{Overall, I think the Results are written in a slightly disorganised and convoluted way (e.g., lines 91-94) and that could be easily improved.}

We have re-written the results, and reframed the manuscript to only focus on $elaboration_{species}$ and $innovation_{species}$ projected on their order’s major axis. e.g. lines 158-162:

\noindent{We found typically higher values of $elaboration_{species}$ than $innovation_{species}$ (Fig. \ref{Fig:phylogeny}) %@L_rephrase_wefound
and strong clustering of both $elaboration_{species}$ and $innovation_{species}$ (median $elaboration_{species}$ for all orders = 0.114, 95\%CI = 0.004-0.485;  $elaboration_{species}$ Pagel's $\lambda$ = 0.888; median $innovation_{species}$ for all orders = 0.071, 95\%CI = 0.018-0.224); $innovation_{species}$ Pagel's $\lambda$ = 0.848).%@L_ei_species_values. }

\textcolor{blue}{Figures:
I think the figures can be significantly improved. I personally feel it would be a pity that this important study will go unnoticed because the figures are less intuitive or aesthetic than they could be, so, I suggest that the authors could put a bit more emphasis on trying to make them easier to read.}

We have improved and modified our figures as suggested below, so hopefully they are now more intuitive.

\textcolor{blue}{Figure 1:Considering colours carry meaning in this figure I suggest removing clade colours and using alternate shades of grey or of one single hue as per Cooney, Bright et al., 2017, for instance. Some silhouettes for each major clade could better guide the readership.
I also suggest using one palette for innovation and another one for elaboration. These palettes would ideally go from grey for low values to a single hue for high values (e.g., orange for innovation; blue for elaboration).}

We tried adding silhouettes to the former figure 1 (now figure 2; circular phylogeny) however, due to the imbalance in clade sizes, the silhouettes are either mostly clumped in one portion ofthe tree due to the dominance of the passerines which looks unpleasant, or if they are spread evenly they are not next to the correct clade which is misleading. Therefore we have chosen to omit silhouettes but have overall simplified the figure. We do this by removing the innovation and elaboration scores for the whole phylogeny and the super-orders and standardised innovation and elaboration colour gradients as suggested (this was a really good suggestion, thanks). We have now changed the colour coding throughout the manuscript using orange for elaboration and blue for innovation. When showing gradients of either elaboration and innovation we are now using a gradient from grey (0) to orange/blue (highest value of elaboration/innovation) as suggested.

\textcolor{blue}{Some typos in the caption. Fix please. There is some text that is not very clearly written – check please.}

Fixed.

\textcolor{blue}{‘Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid’. In your figure the branch colours go from blue to yellow.}

Fixed.

\textcolor{blue}{Finally, I really do not understand the left boxplot inset.}

We have removed these boxplots altogether since we now only focus on the distribution of elaboration and innovation at the order level for the main text.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/InnovElabTree_main_text_revision.pdf}
\caption{Avian phylogeny (n = 8748 species) showing Euclidean distance of species to the centroid of beak space (branches, cividis scale), and distributions of species beak shape $elaboration_{species}$ (inner circle, orange scale), and $innovation_{species}$ (outer circle, blue scale).
Elaboration and innovation scores represent comparisons of species at the order level.
Additional comparisons to super-orders and at the class-wide phylogenetic level are shown in Fig. \ref{Fig:phylogeny_supplement} and for family and super-family comparisons of the order Passeriformes in Fig. \ref{fig_phylogeny_passeriformes}.
}
\label{Fig:phylogeny}
\end{figure}
\bigskip

\textcolor{blue}{Figure 2: I think you could use another colour that shows better contrast among each other perhaps using a palette which is not the default from ggplot. Also please label elaboration and innovation in the boxplots by each clades plot.}

The colour palette used here is not from ggplot (although we agree it looks very similar) and was chosen to offer the maximum contrast gradient among lineages. However, we agree that it is difficult to discern the different colours (it’s hard to find a palette for 35 nested variables!), so we have removed the colours from this plot. We have also completely reworked figure 2 (now figure 3) so it is now presented in a phylogenetic format with the super-orders overlapping the orders. We have also standardised the colours for elaboration (orange) and innovation (blue) as suggested in the comment about figure 1 above.

\newpage

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.6\textwidth]{Figures/figure_phylo_spiro_dark.pdf}
\caption{\scriptsize{$Elaboration_{clade}$ and $innovation_{clade}$ at the megaevolutionary scale for each order and super-order in the bird phylogeny.
The black and gray ellipses are the scaled average evolutionary rate matrices from the pGLMM models for respectively the clades (black) and the class-wide phylogeny (gray).
The ellipses are centered on the position of the clade in the shapespace.
The associated black bar plots represent the variance on each of the eight dimensions in the shapespace with the top gray bar plot representing the variance on each of the eight dimensions for the class-wide phylogeny.
The orange and blue distributions represent respectively the distribution of the $elaboration_{clade}$ (orange) and the $innovation_{clade}$ (blue) scores for each clade and the dots are the median $elaboration_{clade}$ or $innovation_{clade}$ and the solid and the dashed lines representing the 50\% and the 95\% confidence intervales of the distribution of the scores across the 4000 posterior samples.
The ticks on the y-axis always represent five arbitrary units of elaboration and innovation.
An alternative visualization of the same information is available in the supplementary materials in Fig. \ref{Fig:ellipses_rainbow} along with a companion plot showing further nested structure within the Passeriformes in Fig. \ref{fig_ellipses_passeriformes}.}
}
\label{Fig:ellipses}
\end{figure}

\newpage

\textcolor{blue}{Figure S5 caption: ‘The distribution of innovation and elaboration scores comes from the projection of the 4000 pairs of posterior VCVs.’ It would be good to briefly state where these 4k pairs are.}

We have now integrated Figure S5 into the main text (as figure 1) and have completely rewritten the caption (including the precision of what these pairs are). See comment above referring to now figure 1:

\noindent\textit{Relationships between key figures in the main text.
A) represents the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
The gray ellipse represents the overall phylogenetic major axis of beak variation in birds.
Species aligned or away from this axis are respectively elaborators or innovators relative to the class Aves.
The red ellipse and axis represents the phylogenetic major axis of beak variation in Tinamiformes only.
Species aligned or away from this axis are respectively elaborators and innovators relative to the Tinamiformes order.
B) In Figure 2, we measured elaboration and innovation at the macroevolutionary scale , i.e. the position of each species in terms of $elaboration_{species}$ and $innovation_{species}$ relative to their order's phylogenetic major axis of beak variation.
A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation(i. e. their position on that axis) whereas their $innovation_{species}$ score is their rejection from it (i.e. their distance away from that axis).
The colours on the tips of the phylogeny correspond to the median $elaboration_{species}$ (orange gradient) and $innovation_{species}$ (blue gradient).
C) Figure 4, we measured elaboration and innovation at the megaevolutionary scale, i.e. $elaboration_{clade}$ and $innovation_{clade}$.
The ellipses are scaled and centered and represent the clade's phylogenetic major axis of beak variation relative to the overall phylogenetic major axis of beak variation with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis of beak variation onto the overall phylogenetic major axis of beak variation and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
The distribution elaboration and innovation scores are taken from the projection of the 4000 pairs of evolutionary rate matrices with each pair being the focal level, e.g. order,and the parent level, e.g. the whole bird phylogeny.}

\bibliographystyle{naturemag}
\bibliography{references}

\end{document}

### 2023-06-08 Manuscript/main.tex
Update main.tex

@@
 a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. S\ref{Fig:cheatsheet}).
-We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ scor.
+We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ score.
 % @L_euclidean.
@@
 In addition to the lack of alignment of clade phylogenetic major axis of beak variation with the class-wide phylogenetic major axis of beak variation for birds, we also found that within clades, beak shape variation is either highly constrained (varying almost entirely along a single axis; e.g.
-in Galliformes) or higher-dimensional than expected (there is no clear dominant phylogenetic major axis of beak variation; e.g.
+in Galliformes) or higher-dimensional than in some orders or superorder than across the whole class Aves (for example, there is no clear dominant phylogenetic major axis of beak variation; e.g.
 in Podicipediformes) from the class-wide phylogenetic major axis of beak variation.
+%@L_highdimensions
 For example, Accipitriformes (hawks and allies) have phylogenetic major axes of beak variation that mainly align with the second dimension of the shapespace suggesting distinct directions of beak evolution for the clade relative to the class-wide phylogenetic level but uniformity in their beak shape within the clade.

### 2023-06-08 Manuscript/reponse_letter.tex
reviewer 2 (!) done

@@
 
+\section*{\textcolor{blue}{Reviewer 2}}
+
+\textcolor{blue}{I appreciate what the paper is trying to quantify, and find the topic quite interesting.
+Obviously the dataset is stupendous.
+I support continued consideration of the ms.
+However, at present it is very confusing.
+Terms are used inconsistently and the link between analysis and hypotheses is not laid out with sufficient clarity (other than the authors stating that their method is 'powerful').
+The text is too short, and there does not seem to be any reason why it must be so concise.
+It does not place the work in a broader macroevolution context, or speak to the fact that birds are real biological entities that are a field of study in their own right.
+I am also concerned that the ms as currently written is very confusing, will not speak to a wide audience, and has a great deal of opacity around the methods and concepts.
+In its current form therefore it is a highly technical piece for a more specialist journal.
+However, the authors have a lot of scope to improve on this.
+The work has a great deal of potential and I would like to see the review process continue.}
+
+\textcolor{blue}{I hope that my comments below are useful to the authors in pursuing this.
+Sorry that I have written so much.
+I have also made suggestions that may result in changes to the analytical approach.
+My intention is to be as clear and constructive as possible and I hope that the authors find this helpful.}
+
+Thanks for these comments.
+Note that in our previous version was written for Science and was transferred to Science Advances without modifications.
+We fully take advantage of having more space to explain our approach which has improved the manuscript substantially.
+We detail our changes below.
+
+\subsection*{\textcolor{blue}{Major comments}}
+
+\textcolor{blue}{(1) The methods used assume that the evolutionary lines of least resistance are univariate, and can be derived from data at high taxonomic levels, with some biological meaning.
+The interpretations start are on soft ground if we thought there might be causes for bivariate or multivariate lines of least resistance at high taxonomic levels.
+Or even that lines of least resistance inferred at higher taxonomic levels did not have a straightforward biological interpretation.
+Myself, I'd be a lot more confident in the hypotheses that there are evolutionary lines of least resistance at low taxonomic levels (i.e.
+among closely related species), but I'm not sure what to make of this hypothesis at high taxonomic levels.
+Therefore it is challenging for me that the analytical procedure starts with higher-level patterns and interprets lower level pattern in light of those, and not vice versa.
+The procedure in which major axes of variation for subclades or phenotypic data for species are compared to the whole clade of all birds seems to run backward compared to the reasoning derived from theory.
+It isn't clear that this means what the authors say it does.
+The underlying theory seems to be that there are evolutionary lines of least resistance at more local phylogenetic scales (e.g.
+within a group of closely-related species), and that the morphological variation resulting from shifts in these could be interpreted as innovation.
+This suggests a procedure that compares patterns among similar-level groups, or compares the patterns at larger phylogenetic scales to the predictions of extrapolating what was found at smaller scales.
+It may be that the procedure can be done backwards and yield equivalent, interpretable, results.
+But this needs to be unpacked and explained clearly.
+I'm concerned about this (see point (1) above).
+Within the current set of analyses, the analysis I find the most useful is the one that compared major axes ('lines of least resistance') among subgroups of approximately equivalent rank, finding them to be generally different to each other.
+However, this needs development (see my detailed comments) because currently there are unsupported statements such as the statement that evolution can happen in 'any' direction.}
+
+Some of this is related to misunderstandings about our method and how it builds on the previous literature.
+This was our fault for not fully explaining things in the introduction and we hope we have now explained things more clearly.
+
+First, we chose to work with univariate evolutionary lines of least resistance rather than surfaces or volume because we believe they are more correctly interpreted as directional vectors in any number of dimensions.
+For example, classically, adaptive landscapes are represented in two dimensions with a third dimension being the theoretical fitness advantage from the covariance between the two said dimensions (e.g.
+as illustrated here: \url{https://en.wikipedia.org/wiki/Fitness_landscape}).
+In this context the line of least resistance is the vector pointing towards peaks in the landscape.
+This has the advantage of working with any number of dimensions (whereas a surface or volume are bound to at least n+1 dimensions) and remains interpretable by biologists (i.e.
+the ``evolutionary direction'').
+
+Second, evolutionary lines of least resistance are by their nature designed to investigate patterns across higher-levels.
+However, we agree with the reviewers’ comment that this might be confusing in this specific manuscript.
+We have thus narrowed down the taxonomic levels presented in the main text and are now only focusing on the $elaboration_{species}$ and $innovation_{species}$ projected on the the eigenvalues of the order’s evolutionary rate matrix which we interpret as similar to lines of least resistance at the micro-evolutionary scale.
+We have now clarified this throughout the manuscript.
+E.g.
+on lines 52-60:
+
+\noindent\textit{At this scale the major axis of phenotypic variation among species \cite{marroig2005size, fasanelli2022allometry} can be estimated from a matrix of divergences in species mean phenotypic traits (e.g.
+the \textbf{D} matrix of \cite{BlowsHiggie2003, mcglothlin2018adaptive}.
+These matrices do not take into account phylogenetic relationships among species.
+In contrast, the \textbf{R} matrix of \cite{Houle2017} describes the rate of among-species divergence in species traits and explicitly incorporates phylogeny (also see the \textbf{B} matrix of \cite{Machado2020}) %@L_Rmatrix.
+We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}. %TG: TODO: change throughout (that's the VCV matrix now)
+The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales.
+}
+
+\textcolor{blue}{(2) The paper is notionally quantitative.
+And is quite abstract and complicated in the way it was written.
+But actually, under the surface, a lot of the interpretations are actually qualitative (see several of my comments where I have requested to report a qualitative test of hypothesised interpretations).
+This should be avoided where possible.}
+
+We acknowledge that some interpretations are qualitative.
+However, these are based on detailed quantitive analysis generating large posterior distributions of parameter estimates.
+We have now added references to more supplementary figures providing the exact values from our analyses where appropriate.
+In the revised main text, Figures 3, 4 (part), 5, and 6, and the new Table 1 provide explicitly quantitative analysis.
+See details below for further analyses added to the text.
+We have also improved the clarity of the writing which hopefully helps highlight our quantitative findings.
+
+\textcolor{blue}{(3) A major conclusion is that "Species elaborate, clades innovate".
+But the difference between species and clade is an arbitrary one.
+Species are just the arbitrary subset of evolving lineages that cross the present-day time horizon.
+They have the attribute of being highly over-sampled compared to lineages that occur deeper in the phylogeny.
+Please envision an unfolding evolutionary process generating phenotypic variation through time.
+Let's say that 'innovation' by your definition is rare, but makes a directional and general unreversed change (these off-axis changes should be difficult to reverse by chance because we have a wide range of off-axis trait changes we could make so the chance of randomly drawing the 'right' one is low).
+Int he same process, 'elaboration is common', and is also 'easy' to reverse because it occurs along one univariate axis.
+If we sample information about a high proportion of lineages alive at some particular time (the present) then we expect to find that the differences between species phenotypes mainly reflect elaboration.
+But if we sampled fewer, then the difference between species phenotypes will reflect an increasing contribution of innovation.
+I don't think this a.
+Big problem for the analyses oft he paper.
+But I think ti is a problem for the interpretation that arbitrarily- designated taxonomic ranks are important.
+This rank-based, interpretation makes the present-day into a 'special' time and has little meaning in terms of the evolutionary process.}
+
+We have now nuanced this subtitle by changing it to the following (line 293):
+
+\noindent{\textit{Species elaborate \textit{and} innovate, clades innovate}}
+
+Furthermore, we agree with the reviewers comment on how clades and species are arbitrary cuts of a continuum, they do, in our view, however remain a common and useful tool in research to understand evolutionary biology \cite{baker2021nothing}, particularly in well studied taxa such as birds where high taxa map meaningfully to phylogeny and trait variation.
+As the reviewer notes, it is possible that the major axes change depending on the specific bracketing of taxa (as would be true using arbitrary timescales).
+However, here we use two different higher taxonomic levels for comparison and find broadly consistent results when taken together.
+One additional strength of a taxonomy based approach is that it doesn’t force arbitrary time constraints on interpretation.
+Indeed, the higher taxa that we focus on span a wide range of timescales.
+This has enabled us to conduct new analysis (inspired by the reviewers comments) in the supplementary materials showing the distribution of $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$ through time.
+This new analysis indicates that the magnitude of both elaboration and innovation increases with time since divergence, as would be expected under the commonly used Brownian motion model of evolution.
+lines 242-243: %@L_ei_through_time_plots
+
+\noindent\textit{Similarly, when measured at the macroevolutionary scale, $innovation_{species}$ and $elaboration_{species}$ increase with species age (see supplementary Fig.
+\ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}).
+}%@L_ei_through_time_plots}
+
+\textcolor{blue}{(4) The ms includes very few citations and is broadly not in conversation with the wider literature.
+The approach and its conclusions come across as being very abstract in nature.
+Let's keep a sense of perspective here, the work is about birds, but only a very selective subset of work on bird evolution is cited, primarily by the group who wrote this paper.
+A lot more could be done to make this work accessible to a wide audience, not only for biologists interested in birds, but also for wider macroevolutionary study.
+For example, how do the approaches used here connect with other forms of inference?
+The conceptual framework used here maid numerous references to Endler et al (2005), as though nearly two decade since then provided no useful ideas.
+It may be true that there has not been much specific, directed work that formalises the concepts of innovation or elaboration.
+Nevertheless, many other works contain relevant analyses and interpretations and the authors here show little attempt to integrate those works in their synthesis.
+The contributes to the impression of an opaque and abstract study with little reach.}
+
+We have now updated and added many more references throughout the whole manuscript (from 28 to 63), expanding on the background and giving more context to the Endler et al study.
+Our previous version was a direct transfer from a journal with a more limited page count so we have been able to fully make use of the available space to expand our ideas, citations and links to the existing literature.
+
+\textcolor{blue}{(5) There are some confusing and inconsistent uses of terms (see my detailed comments).
+Sometimes it seems like unnecessarily elaborate language is used, compounding the issues.
+For example:}
+
+\textcolor{blue}{ -'Shapespace' would just be 'shape' in most studies.
+'Shape' is more appropriate in the statistical model formula because each taxon has 'shape', but does not have 'shapespace'.For example, you wouldn't write out "skull size range $\sim$ body size range", you'd write "skull size $\sim$ body size".}
+
+We respectfully disagree with this comment.
+Although the reviewer is right that most studies would use the term “shape”, we do like the precision that the term “shapespace” adds to the formula (the correct most precise term would be “position of species in the shapespace”).
+We do make the distinction that a single species beak has a shape (the raw data from the bird beaks, described using classic gmm approach) and that the shapes of multiple species beaks occupy a shapespace (the transformed and usable data for our analyses after Procrustes superimposition, ordination and dimensionality reduction - lines 363-378).
+In the skull/body size example provided by the reviewer, naming the terms “skull size range” might actually bring a useful methodological precision (what is the skull or body size) while simply using “skull size” gives the more general idea (comparing the skull to the body size).
+In our case, since the formula is in the methods section, we chose to emphasise the methodological aspect rather than the general aspect.
+
+\textcolor{blue}{-'Multilinear algebra', is more typically called 'linear algebra'.
+In fact, it would be better to say what computation was performed e.g.
+taking the dot product or cross product, or whatever.
+Here, little attempt has been made to clearly communicate what is being done.}
+
+We have changed multilinear algebra to linear algebra throughout the manuscript.
+The whole mathematical procedure can be found in the supplementary materials.
+We have now specified that in the main manuscript along with the previous brief description of what computation was performed (projection).
+lines 154-158:
+
+\noindent\textit{To assess this we projected the beak shape data for each species onto %@L_linear_algebra_stuff @L_beak_shape
+the major axes of evolutionary rate matrices of their orders (see supplementary materials \ref{Fig:phylogeny_supplement} for projection onto the major axes of beak shape variation from the class-wide phylogenetic major axis of beak variation or their super-order’s phylogenetic major axis of beak variation).}
+
+\textcolor{blue}{-"Species phenotype' is used in many places where the 'beak shape' is actually being referred to.
+The 'phenotype' is all the morphological and behavioural traits of an organism.
+Not just the beak shape.
+It is OK to call it 'phenotypic data', but not 'species phenotype'.}
+
+We have changed “species phenotype” to “species beak shape” throughout.
+
+\textcolor{blue}{(6) Choice of nodes to analyse - order and suborders represent an arbitrary subset of nodes on the tree.
+Is there a justification for not analysing patterns across all pairs of ancestor and descendent nodes in the full tree?
+This is important because something the patterns for e.g.
+orders are compared to each other as though they were meaningful biological units that could be expected to share some features.
+In fact, they cannot.}
+
+We agree that the choice of nodes to analyse is ultimately an arbitrary decision (see our comment above).
+We had originally considered analysing every node and agree this would be great.
+However we decided to use the named nodes (orders and super orders) for the following reasons:
+\begin{itemize}
+\item We estimate that it would take between 5 CPU centuries + 1ZB of RAM (using the normal MCMCglmm framework) or 12 CPU millenia + 2.5 GB of RAM (using the mcmcmcglmmm framework).
+Both are currently not achievable with modern computers.
+\item Named nodes represent divergence points of interest to biologists (especially ornithologists) and are based on common biological features shared by living species within the orders (namely their beaks).
+\end{itemize}
+
+We’ve added a plot of $elaboration_{clade}$/$innovation_{clade}$ through time per clade (orders and super-orders) to the supplementary materials to give an idea of the time span of the named nodes used in the analysis and how it might affect our results.
+This plot gives a rough idea on how the nodes are distributed through time (lines 242-243).
+
+\noindent\textit{Similarly, when measured at the macroevolutionary scale, $innovation_{species}$ and $elaboration_{species}$ increase with species age (see supplementary Fig.
+\ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}).
+}%@L_ei_through_time_plots}
+
+\textcolor{blue}{(7) The interpretations emphasise 'levels' of innovation and elaboration.
+But these absolute amounts cannot be interpreted easily, independent of the total amount of phenotypic variation for a (arbitrarily delimited) subgroup.
+I'd like to see more transparency about variation the relative importance of each of these effects among groups.}
+
+We are not sure about what the reviewer means by “transparency” in this context.
+Figure 3 shows the distribution (including the 50 and 95\% CI) of the posterior distribution of the $innovation_{clade}$ and $elaboration_{clade}$ scores.
+The code and the data to reproduce this figure is made available on GitHub \url{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}.
+We have now also added a whole new analysis where we test how the amounts of elaboration and innovation relate to overall beak divergence using PGLS models presented in the new Table 1.
+(lines 311-324)
+
+\noindent\textit{We further tested the contributions of elaboration and innovation at different scales by examining the consequences of elaboration and innovation for the observed divergence of bird beak shapes.
+Specifically, we calculated the distance to centroid of avian beak morphospace for each species (Fig.
+\ref{Fig:phylogeny}).
+We then used phylogenetic generalised least squares (PGLS; \cite{phylom}) to model distance to centroid as a function of $elaboration_{species}$, $innovation_{species}$ and $elaboration_{clade}$ and $innovation_{clade}$.
+We ran the PGLS analysis using both single and multiple predictors.
+Our analyses indicate that the majority of variation (r2 = 0.888; Table 1) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
+When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table 1).
+In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table 1).
+However, $innovation_{clade}$ is nonetheless an important contributor to total beak shape space.
+We reach this conclusion because models with interactions between species and clade metrics provide by far the best model fit overall and removal of terms indicate that the most important interactions are between $innovation_{clade}$  and $innovation_{species}$ followed by $innovation_{clade}$ and $elaboration_{species}$.
+The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table 1)
+Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space.
+}
+
+\textcolor{blue}{(8) The authors choose a computationally-intensive method to find the major axis of variation, and then say they can't do more because of computational intensity.
+But the decision to take this approach in the first place is not explained.
+It also isn't clear that the mini-chains or the aggregate result from them has converged because no info about convergence is given.
+Could their be come comparison to a similar method that requesters less computation power such as phylogenetic PCA, that may validate that the major axes returned are sensible?}
+
+We use the mini-chains approach to account for phylogenetic uncertainty.
+Our phylogenetic hypothesis comes from a Bayesian analysis so it is not appropriate to perform analyses on just one tree.
+Therefore a posterior distribution of trees needs to be used.
+This is not possible within our basic MCMCglmm format without a huge amount of computational time, thus we developed the mini chains approach to make it tractable.
+Phylogenetic PCA would not provide a solution to this issue as it also uses one tree, nor can it provide results in a nested way as our method does.
+We have now added a clearer explanation of why the mini chains were used.
+
+We do not provide convergence information between mini-chains since they are considered in our framework as different samples from the same chain.
+We are thus effectively running single chains for each of the 35 nested clades.
+Running pairs of chains would just double the amount of computational time and not provide useful information.
+In fact, this mini-chains method highly favors accuracy over precision, therefore mini-chains with a high level of uncertainty (e.g.
+because of a small clade size) will always lead to imprecise but accurate results (i.e.
+two blurry posteriors).
+This means that convergence will never be useful information as some of the chains will be not converging on anything by design (i.e.
+the “true” precise solution is unknown but the 95\% posterior accurately contains the “true” solution).
+We do however include the distribution of effective sample sizes for all chains (supplementary information S3 and S4) which gives a good indication of the posterior distribution containing enough samples to be used as the “true” solution.
+Note that for all analytical parts in the analysis, we use the entire posterior distribution - we only use the averages from these distributions for plotting and in the new PGLS analyses.
+We have added this precision in the manuscript (lines 450-454).
+
+\noindent\textit{The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection} - note that the procedure described in the section \ref{supp_projection} is applied to each individual evolutionary rate matrix for each term in the model, e.g.
+for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score.
+%@L_individual_matrices).
+}
+
+\subsection*{\textcolor{blue}{Detailed comments}}
+
+\textcolor{blue}{Abstract}
+
+\textcolor{blue}{Lines 7-8: "Here we reconcile major changes in phenotypic diversity at macroevolutionary scales with gradual microevolutionary change"
+>>I read the ms a few times, considering whether the approach used here reconciles the patterns at smaller scales with those at larger scales.
+I think it is a descriptive method rather than a model of varying processes at different scales.
+So I'm not comfortable with the statement that it 'reconciles' these things.
+That reconciliation is done int h msusing verbal argument - and this verbal arguments are based on a much older literature.
+Consider re-phrasing here.}
+
+We have now reworded the abstract without “reconcile”.
+
+\textcolor{blue}{Lines 8-9: "using new multi-trait models"
+$>>$It isn't clear that these are 'multi-trait models' rather than 'approaches to modelling multivariate traits'.
+The subject of the current study is beak shape, which arguably is a single multivariate trait rather than multiple traits (implied by 'multi-trait').}
+
+We’ve changed “new multi-trait models” to “new approaches to modelling multivariate traits” as suggested. (line 8)
+
+\textcolor{blue}{Lines 10-11: "We find that elaboration dominates at macroevolutionary scales whereas innovation becomes more prominent at megaevolutionary scales"
+$>>$I quite like these terms 'mega' and 'macro'.
+But we should probably acknowledge that the distinction is qualitative and arbitrary.
+Here, it would be more useful to directly state the taxonomic scales of the analyses.
+That is avoids going too far down the road of abstraction.}
+
+Following this reviewer’s previous comment on arbitrariness, we do agree but find this comment inapplicable (everything we do is arbitrary).
+Also we disagree on the absence of distinction between “mega” and “macro”, we refer here to “macro” as the species focal level (i.e.
+using species rather than individuals as the unit of our analyses) and “mega” as the clade level (i.e.
+using groups of species rather than individual species as the unit of our analyses).
+We have now specified this one lines 82-83
+
+\noindent\textit{The phylogenetic breadth of the data allows us to compare directions of beak evolution of subclades (e.g.
+orders) within the class Aves (the megaevolutionary scale) and of species within subclades (the macroevolutionary scale).}%@L_macromega_definition2
+
+\textcolor{blue}{Introduction}
+
+\textcolor{blue}{Lines 24-27: "However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape"
+$>>$Some inconsistency of language in these statements.
+The first clause says 'genetic' and the second clause say 'genomic and developmental'.
+I think that the authors are referring tot he same identical thing twice, but using different words each time.
+If so, please make this consistent so it is easy for a reader to follow your meaning.}
+
+We have now entirely rewritten the introduction (also following the other reviewers comments).
+
+\textcolor{blue}{Lines 31-33: "Resolving this conflict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades"
+$>>$"Balance" here is a metaphor and is potentially confusing. Consider something like "the relative contributions of ..."
+$>>$I also think it is important to say that what they are contributions towards.
+I suspect you are asking about the relative contributions of these conceptual processes to the total disparity of a large group such as all birds.
+If so, can you clarify this here, and potentially in other places where this important goal of the study is stated.Lines 41-43: "Relative to elaboration, innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales"
+$>>$There is first-pass evidence relevant to this across many fields of evolutionary research (e.g.
+paleontological study of major evolutionary transitions; island biogeography; developmental biology; many others).
+Only one area is mentioned (signal evolution), and that relies on one citation from 2005.
+It seems strange to me that the authors have this pinpoint focus on one item of literature.
+It makes the while topic seem a lot more esoteric and erudite the it actually is.
+Could the authors please contemplate broadening their introduction into a wider biological context.
+The work should be integrative rather than striking out in a unique direction with undue emphasis on one paper from 2005.}
+
+We have now entirely rewritten the introduction (also following the other reviewers comments) to make our motivations clearer.
+We used the Endler et al.
+(2005) paper as inspiration for the ideas of innovation and elaboration (hence the focus on a single paper) but acknowledge that this resulted in ignoring a much broader body of research.
+We have added more references, explained our motivations and how our work sits in the current literature and spent more time explaining the key concepts for this work.
+
+\textcolor{blue}{Lines 43-46: "These core concepts can now be defined using robust multivariate statistics about trait covariance applied at multiple phylogenetic scales and provide a powerful theoretical framework with which to study macro- to megaevolutionary divergence and in particular to determine the relative importance of"
+$>>$These statements are a bit confusing to read.
+They say that the concepts can "now be defined...".
+But this is not clear as to whether they are presenting a new method or using an established one.
+Please write either "We present a new framework for..., which defines X and Y as...", or "Recent work established a framework for...".
+Also, please avoid describing your own method as 'powerful'.
+This has many possible meaning.
+For example, statistical power, explanatory power, or power to test hypotheses that previously couldn't be tested.
+Not only that (ambiguity), but it also seems self-congratulatory.
+It is for others to determine if the method is powerful or not (or simulations, depending what is meant by 'power').
+Line 47: "idiosyncratic species-level innovation against clade-level innovation"
+$>>$Are these different things?
+The 'species-level' in a dataset like this is the arbitrary subset of all the lineages in the phylogeny that happen to cross the present-day time horizon.
+We have no way of knowing whether those innovations are 'idiosyncratic', or are just the early evidence of what will be in future a clade-level innovation.
+Also, 'idiosyncratic' is anthropomorphic.
+Lines 49-50: "The contributions of these multivariate, evolutionary pathways"
+$>>$Say contributions to what (e.g. to total clade disparity, or 'phenotypic variation' as generally used in the current work).}
+
+We are grateful for this reviewers’ specific and justified comments.
+We have now entirely rewritten the introduction (also following the other reviewers comments) and believe that we have clarified the points above.
+
+\textcolor{blue}{Methods}
+
+\textcolor{blue}{Lines 68-69: "can estimate levels of elaboration and innovation based on that major axis of phenotypic variation (see Methods)"
+$>>$This is a really important sentence.
+But it isn't working hard enough, and the word 'levels' is vague where it needs to be specific.
+This is the main place it the ms, when read in sequence, where a reader gets some idea of your approach.
+It needs to be super- clear what you're doing and how it related to these concepts.
+Currently it is not.
+I get to the end of the paragraph and I have a few clues only e.g.
+(1) it is an analogue of the G matrix - which is not a direct explanation, (2) It can be defined at any phylogenetic level.
+(3) It provides the basis of all subsequent estimates.
+Missing is a direct explanation that would allow even a qualitative reader to understand the business end of what you're doing without skipping to the Methods.}
+
+We have moved this line further down in this paragraph and added the following precisions to make it easier to grasp without having to read through the methods section.(lines 135-138)
+
+\noindent\textit{We can then estimate elaboration and innovation based on that major axis: elaboration is defined as the position on the phylogenetic major axis of beak variation and innovation is defined as the distance from the phylogenetic major axis of beak variation (see Methods).
+%@L_ei_meth_precision.
+}
+
+\textcolor{blue}{$>>$The word 'level' is really important in the ms for referring to e.g.
+'species-level', 'clade- level'.
+But you also use it in an entirely different context to refer to 'levels of innovation'.
+This is naturally confusing.
+I strongly advise to change this to something like 'contribution of'.}
+
+We have now standardised the use of the word “level” throughout the manuscript. It is now only referring to taxonomic levels (i.e. species, orders, super-orders, the whole phylogeny).
+
+\textcolor{blue}{Section beginning line 82
+$>>$This section reports the relative contributions of elaboration and innovation to beak shape variation based on comparisons of lower-level data to major axes beak shape variation among higher taxa.
+I read it a few time and I'm struck that the major axes of variation in higher taxa are not described.
+This makes it very abstract and detached from reality.
+It also make it difficult to make sense of statements about hummingbirds and hornbills.
+The ?mean?
+beak shapes of these groups are described verbally, but it isn't possible to use this information really without knowing what the large-scale patterns of shape variation are.}
+
+Describing the distribution of the major axes and the elaboration and innovation for each clade would make a really long list which we believe would distract the reader from our key results.
+This is outside of the scope of this paper.
+We have added however a list of the distributions of the ellipses in the supplementary materials (Table S1).
+
+\textcolor{blue}{Line 83: "We tested the expectation, derived from adaptive radiation theory"
+$>>$Citations...? There is a rich literature on this that is not really cited closely. This is one example of the scarcity of citations across the whole ms.}
+
+We now cite the following references for this sentence (lines 14-151):
+
+\noindent\textit{We tested the expectation, derived from adaptive radiation theory \cite{schluter1996adaptive,mcglothlin2018adaptive,slater2019hierarchy} %@L_cite_adaptive
+, that species divergence is biased along phylogenetic lines of least resistance by calculating species-specific measures of $elaboration_{species}$ and $innovation_{species}$.}
+
+\textcolor{blue}{Lines 87-88: "To assess this we calculated the multilinear algebraic projection of each species phenotype onto three major axes".
+$>>$Is there a reason to call this 'multilinear algebraic projection of each species phenotype' instead of 'we projected the beak shape data for each species onto...'.
+I think this is a fairly commonplace projection using conventional linear algebra, and in some other field of science we wouldn't feel the need to use so many syllables.
+$>>$I also don't like this approach to say 'species phenotype', which is a very broad and encompassing term, when you mean 'beak shape'.}
+
+We changed the sentence to the reviewer’s suggestion (see major comments above and lines 154-158 in the manuscript):
+
+\noindent\textit{To assess this we projected the beak shape data for each species onto %@L_linear_algebra_stuff @L_beak_shape
+the major axes of evolutionary rate matrices of their orders (see supplementary materials \ref{Fig:phylogeny_supplement} for projection onto the major axes of beak shape variation from the class-wide phylogenetic major axis of beak variation or their super-order’s phylogenetic major axis of beak variation).}
+
+\textcolor{blue}{Lines 89-91: "We found higher values of elaboration than innovation overall across all levels (Fig 1) and globally more elaboration and innovation at the global phylogenetic and super-order levels compared to the order level"
+$>>$I found this confusing on the first read.
+It says 'more elaboration...
+at the global...
+level'.
+But the elaboration and innovation are estimated from comparisons of lower-level data to the higher-level major axis.
+Maybe add a few words of clarification.
+I think you could mean something like" "great contribution of elaboration to patterns of phenotypic variation at the global..." or something.
+I'd also myself prefer "beak shape variation" to "phenotypic variation".
+This is an analysis of beak shape among birds, not of the generalised concept of phenotypic variation among birds.
+It does not even include, for example, the relative beak size, let alone all phenotype.}
+
+This sentence has been changed due to the overall change of the notation of elaboration and innovation and is hopefully now clearer. (lines 158-159)
+
+\noindent\textit{We found typically higher values of $elaboration_{species}$ than $innovation_{species}$ (Fig. \ref{Fig:phylogeny}) %@L_rephrase_wefound
+}
+
+\textcolor{blue}{Lines 94-95: "However, the extent of both elaboration and innovation tends to reduce as we move from"
+$>$I've been trying to figure out what is so confusing about this.
+I think the difficulty I've had is that you use a rotation of different terms to refer to the same thing.
+Here, you discuss 'the extent of' elaboration and innovation.
+But in other places it was 'levels of'.
+I strongly advocate that you go through the whole ms and make sure that you already use the same terms so that the reader doesn't get lost and confused.
+For this specific thing, I think you should go with 'relative contributions of...
+to phenotypic variation' (or 'beak shape variation'), or 'contributions of...' (Not 'levels of' or 'extent of').}
+
+Following this reviewers’ previous comment, we have now removed any mention of the word “level” if it doesn’t refer to phylogenetic levels (clade, species, etc.).
+
+\textcolor{blue}{Lines 103-104: "Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level"
+$>>$'Divergence' here is ambiguous.
+It could mean population genetic divergence giving rise to speciation.
+Or it could mean phenotypic divergence.
+Consider clarifying this.
+For example, 'dominant mode of phenotypic evolution at the species level' or 'dominant mode of phenotypic divergence at the species level'.}
+
+We have now rephrased it to: “the dominant mode of phenotypic divergence at the species level” as suggested. (lines 192-195)
+
+\noindent\textit{Collectively, our results suggest that elaboration is the most common route by which species beak shape evolves.
+While $elaboration_{species}$ is the main mode of beak shape divergence at the macroevolutionary scale %@L_dominant
+(Fig. \ref{Fig:phylogeny}), it is also commonly positively correlated with $innovation_{species}$ (Fig. \ref{fable_correlations} and S8).} %@L_correlation_figure_link).
+
+\textcolor{blue}{Lines 104-105: "We find no consistent evidence that innovation is either positively correlated, or trades-off, with elaboration"
+$>>$Please refer to the specific test of this with "...as indicated by the [test and results]".}
+
+We have now added two figures in the supplementaries displaying either the median correlation between elaboration and innovation for each species within each super-orders and an extra figure summarising the correlation statistics for all the posterior elaboration and innovation calculations in the supplementary Fig.
+S6.
+
+\textcolor{blue}{Lines 113-114: "have exceptionally wide beaks relative to length and display high levels of both elaboration"
+$>>$I'm sensing a problem of interpretation here.
+When I read the text, I can see in context of the total disparity of some subgroup, that a portion of this can be attributed to innovation, and a portion can be attributed to alb oration.
+All good so far.
+There is a thirdvariable of interest though, and that is the total phenotypic variation within the subclass.
+So a group like Hirundinidae could only have high levels of both elaboration and innovation if it also had high total variation.
+The in-text description at the moment then is missing a piece of information that would allow us to meaningfully compare patterns in Hirundinidae to other subgroups.
+The comparison can only be meaningful if we can compare both eh absolute amounts and -also- the relative contributions.}
+
+We have now also provided the amounts of species per group (Figure 4), and standard deviation in ellipses orientation per group (Figure 5).
+
+\textcolor{blue}{$>>$In this context I can see that your description here is focused on 'levels' (absolute amounts).
+But we also need to know the relative contributions independent of the total disparity of a clade.
+oOne reason or this is that a clade could have high absolute amounts of both simply by having more time to evolve.
+This is important given that the clades are chosen arbitrarily based on taxonomic rank, without attempts to standardise for the amount of branch length (or rate of evolution( in the group.}
+
+We have now added a new analysis in the supplementary materials showing the $elaboration_{clade}$ and $innovation_{clade}$ as well as $elaboration_{species}$ and $innovation_{species}$ for each clade through time (Figures S7 and S8).
+This shows the results somewhat scaled by branch length and put in the evolutionary context of each clade (rather than more arbitrarily referring to named nodes).
+
+\textcolor{blue}{Line 115: "The dominance of elaboration when measured at the species-level"
+$>>$Here you talk about elaboration at the 'species-level'.
+Ut on lines 89-91 you say "we found higher values of elaboration than innovation overall across all levels (Fig 1) and globally more elaboration and innovation at the global phylogenetic and super-order levels compared to the order level".
+I'm a bit confused about what is being described each time.
+This partly results from the ambiguity that there are pairwise comparisons (e.g.
+between species beak shape and the major axis of global phylogenetic variation among all birds), so 'species-level' doesn't uniquely refer to a particular pairwise comparison.
+Please thing seriously about how to clarify this.
+I think the interpretability of your manuscript rests upon making this really clear.}
+
+Following reviewers’ comments we have now standardised many terms throughout the manuscript (see responses to comments above) which hopefully greatly clarifies the manuscript.
+
+\textcolor{blue}{Lines 116-117: "The frequent phylogenetic clustering of species that elaborate, innovate, or both"
+$>>$Please say what hypothesis test (=analysis) this is based on.
+I can see ways that it might just reflect phylogenetic autocorrelation.
+Also, the approach to making this inference is not outlined in the Methods.
+So I haven't got quite enough information from which to decide if it is an interesting inference or just an artefact of method construction.}
+
+We have now removed this sentence from the manuscript to not add another extra analysis here.
+
+\textcolor{blue}{$>>$Also, strictly-speaking, -all- groups do both.
+There are presumably zero species that have absolutely no 'innovation', or no 'elaboration; as defined here.
+So sharpen this section top a bit please.}
+
+We have now clarified the actual values of $elaboration_{species}$ and $innovation_{species}$ we are referring to making it clearer that no groups have no elaboration or no innovation. (lines 158-162)
+
+\textcolor{blue}{Lines 124-126: "To investigate the conflict implied by our observations of species-level elaboration and previously reported deep-time jumps in phenotype, we measured the elaboration and innovation of different clades.
+"
+$>>$Thanks for being patient with me.
+But is it possible for the difference between the results for species compared to global phylogenetic variation is different to the results for subgroups compared to global phylogenetic variation only because 'elaboration' and 'innovation' are defined differently for these types of analysis?
+This deserves some attention, if only to clarify.}
+
+We have now rewritten the manuscript to clarify that innovation and elaboration are different depending on whether we focus on species (macro) or clades (mega).
+However, they are mathematically and theoretically similar, i.e.
+they are both describing changes along and away from an axis of major phenotypic differences.
+In the same way, we can also link them to elaboration and innovation at the individual level (micro) as discussed in more details with our responses to reviewer 4).
+Ultimately all these are the same measurement applied on different focal levels of the same data (specimen data $->$ species data $->$ clade data).
+Therefore the differences in $elaboration$/$innovation_{species}$ and $elaboration$/$innovation_{clade}$ are not unexpected biologically (i.e.
+biologists have observed discontinuities and emergent properties at higher levels above the individual) but are also not expected mathematically (i.e.
+the same measurement on nested subsets should give the same or similar results).
+
+\textcolor{blue}{Lines 133-136: "measured rejection (distance from the major axis of phenotypic variation) as the clade’s innovation score.
+A high innovation score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g.
+the direction of evolution of an order differs from its parent super-order)"
+$>>$Usually 'distance' means e.g.
+the Euclidean distance / Procrustes distance in multivariate shape space, or some other measure of distance.
+Here, from eh context, it sounds like 'distance' in the first sentence is actually a different between angles (implied by the text about the 'direction' of evolution).
+Please clarify this.}
+
+We have now specified that the distances are indeed Euclidean (lines 218).
+The relation between elaboration (projection) and innovation (rejection) and the direction comes from the Pythagorean relationship between projection and rejection in a right angle triangle:
+
+\begin{equation}
+\text{direction (angle/orientation)} = arcos\left(\frac{\text{projection}^2 - \text{rejection}^2 + \sqrt{\text{rejection}^2+\text{projection}^2}  }{2\times\text{projection}\times\text{rejection}}\right)
+\end{equation}
+
+\textcolor{blue}{$>>$The term 'rejection' is not defined in the introduction, differing from other terms. It isn't immediately clear what it relates to in macroevolutionary theory.}
+
+We have now specified that we mean linear algebraic projection and linear algebraic rejection (lines 218).
+
+\textcolor{blue}{Line 168: "Evolutionary innovation can arise in any direction in trait space"
+$>>$This sentence is followed by a statement of evidence, which says that none fo the big subgroups are aligned to the major axis of all birds.
+However that does not indicate that innovation can arise in -any- (literally any) direction.
+Please clarify this.
+Ideally present some kind go figure that illustrates the range of angular differences form all bird that are present among subgroups.}
+
+We have changed “any” to “many” and now provide an illustrative example in the supplementary materials Fig S9 (lines 265-269):
+
+\noindent\textit{Evolutionary $innovation_{clade}$ can arise in many directions in trait space (see examples in supplementary Fig. \ref{Fig:mikado}). %@L_any_dir
+}
+
+\textcolor{blue}{Line 174: "This pattern of orthogonality..."
+$>>$This statement follows after an explanation that subgroups have a range of different angles.
+They can have different angles without being orthogonal, given that orthogonal has a precise geometric meaning.
+SO it isn't clear what is meant by 'this pattens of orthogonality' - either the word 'orthogonal' is misapplied here, or the pattern was mis- described in the preceding text.}
+
+We have changed “orthogonality” to “amounts of orthogonality” throughout the manuscript with 1 being completely orthogonal and 0 being not orthogonal at all (parallel).
+
+\textcolor{blue}{Lines 180-182: "either highly constrained (varying almost entirely along a single axis; e.g.
+in Galliformes) or higher-dimensional than expected (there is no clear dominant major axis; e.g.
+in Podicipediformes)"
+$>>$There is no analysis that explains how much dimensionality is expected.
+So I'm not clear on what is meant here.
+Is it the qualitative expectation of the authors?
+It is important to explicitly differentiate between this and quantitative analysis.
+You will note that several fo the comments I made above also seem to be qualitative interpretations that do not refer to quantitative analysis.
+But in each case, a quantitative analysis is possible, and is arguably important to test the hypothesised interpretations.}
+
+We apologise for sending an earlier version of figure 2 (now figure 3).
+We have now added the representation dimensionality for each order and super order as intended in figure 3 and rephrased the following section on lines 278-282:
+
+\noindent\textit{beak shape variation is either highly constrained (varying almost entirely along a single axis; e.g.
+in Galliformes) or higher-dimensional than in some orders or superorder than across the whole class Aves (for example, there is no clear dominant phylogenetic major axis of beak variation; e.g.
+in Podicipediformes) from the class-wide phylogenetic major axis of beak variation.} %@L_highdimensions
+
+\textcolor{blue}{Lines 183-186: "For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogenetic level but uniformity in their beak shape within the clade"
+$>>$This interpretation rests not he assumption that variation among all bird sis adequately captured by a univariate evolutionary line of least resistance.
+If not, then it seems relevant that accipitriforms along to 'PC2' of all birds}
+
+We are not sure how to respond to this comment.
+Here we assume that this reviewer is wondering why we chose to use an axis of major phenotypic variation (1D) rather than a surface (equivalent of an evolutionary landscape - 2D) or something else (a n-1D “surface”).
+Although there have been great publications using surfaces rather than axes to describe evolutionary patterns, here we decided to focus on one dimension to describe the major axes rather than a multidimensional solution.
+This is because, although multidimensional tools can tell maybe more precise information, there is no good way to decide how much to reduce the space between 1D and n-1D.
+Therefore we chose to use the 1D reduction which we believe is less arbitrary (although still arbitrary) and more consistent for all clades, i.e.
+different clades can have different numbers of dimensions required to describe their n-1D “surface” but all clades have at least 1 dimension that describe most of their variation.
+
 \bibliographystyle{naturemag}

### 2023-06-08 Manuscript/reponse_letter.tex
reviewer 3 done.

@@
 
+\section*{\textcolor{blue}{Reviewer 3}}
+
+\textcolor{blue}{The study presents a quantitative framework for investigating the relative contributions of "elaborative" and "innovative" trait evolution on phenotypic diversification, using geometric morphometric data of bird beaks as a case study.
+Overall, I thought the study provided a compelling synthesis of methodological and empirical advancements, all presented beautifully.
+In my view, the study would of great interest to evolutionary biologists, particularly those who focus on comparative methods, phenotypic diversification, and avian biology.
+Admittedly, I did not have a chance to run the code on the data within the period of time I was given to review the paper.
+However, both the code and data are uploaded to publicly accessible repositories.
+I have few general comments that I believe should be addressed, followed by more minor ones.}
+
+Thanks for these positive comments!
+
+\subsection*{\textcolor{blue}{General Comments}}
+
+\textcolor{blue}{1.
+Since the term "megaevolution" is used frequently in the text and is a crucial aspect of the study, I recommend clearly and directly defining "megaevolution" near the beginning of the main text.
+Based on the methodology, it seems like the study considers megaevolution to desscribe patterns that span the entire avian phylogeny, but clear definition (that is generalizable beyond birds) would be helpful.
+Related to this point, the discussion section could touch on how the results of this study help define or refine the definition of "megaevolution" beyond the a priori distinction between bird-wide vs.
+"macroevolutionary" levels.
+Essentially, I'm curious to understand how "macro" and "mega" evolutionary scales are distinguished conceptually and methodologically.}
+
+We have now clarified our distinction between mega and macro evolutionary level.
+Throughout this manuscript we follow GG Simpsons’ idea of megaevolution being evolution focusing on clades as evolutionary units and macroevolution being evolution focusing on species as evolutionary units.
+We have now clarified this in the text lines 64-67:
+
+\noindent\textit{While there is no strict definition that separates macro- and mega-evolution, we define the macroevolutionary scale as species divergence within clades and the megaevolutionary scale as divergence of clades within larger clades.
+}%@L_mega_macro_def
+
+\textcolor{blue}{2.
+Does the comparisons of orientation among subclades (Ln 159–162) take into account shifts occurring along a phylogenetic tree; and are the sub-clades compared taking account the phylogenetic distance between these clades?
+After reading through the manuscript, I wasn't certain if the comparisons were phylogenetically informed.
+This would be a good information to state clearly in the main text since we would initially expect more closely related clades to exhibit similar orientations.}
+
+Yes, these comparisons take phylogeny (i.e. distance between clades) into account. We have now made this clear by mentioning it in the main text (lines 248-252):  %@L_orientation_phylo
+
+\noindent\textit{Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig.
+4; Fig.
+5; note that these comparisons are based on the evolutionary rate matrix).} %@L_orientation_phylo
+
+\textcolor{blue}{3.
+To more fully explore the major aims of this study (i.e., relative contributions of "mega" vs "macro"evolutionary phenoma to the overall phenotypic diversity), the rate of beak evolution between "elaboration" and "innovation" should be incorporated into the analysis and reporting.
+For example, do the data show that "innovations" occur at a faster pace than "elaborations"?
+In my view, this is a key missing piece of the study that would form a more holistic answer to the main questions being investigated by the study.}
+
+We agree that this would be interesting.
+Following this reviewer’s comment, we attempted to run analyses using the rate heterogeneity model of \cite{venditti2011multiple} in BayesTraits and ridge regression model of \cite{castiglione2018} in  RRphylo on a single topology.
+From these we could extract evolutionary rates per species and explore correlations with elaboration and innovation.
+However, while initial runs in BayesTraits looked promising, we found that only local optima had been explored in the MCMC chain.
+We identified this from examining three independent analyses each of 500 millions generations in which initially all three chains appeared to converge in parameter space.
+However, after approximately 400 million generations, one chain moved to new parameter space with a higher likelihood.
+Further runs are currently stuck on the lower-likelihood parameter space.
+Because of the size of the dataset, these analyses have not converged after two months since we received the reviewers comments (2nd of April 2023).
+We are not confident that we will see adequate convergence in a reasonable timeframe based on previous experience on a smaller version of the data set that took >6 months and 2 billion generations to converge (\cite{chira2018correlates}).
+Similarly, we have not been able to examine output from RRPhylo because the analyses have not completed.
+Because of the time it is taking to run on one single tree (eventually it will have to be run on the tree distributions) we decided that this extra analysis is out of the scope of this revision as it could take more than a year to complete.
+
+\textcolor{blue}{4.
+While this study is primarily a methods-focused one, the study would be strengthened by further discussing the biological explanations and implications of the results.
+For example, how do the results align or contrast with what we know about our mechanistic (e.g., genetic, developmental) understanding of how the beak forms.
+Do the elaborative vs.
+innovative changes to the beak correspond to morphological differences linked to certain genes and developmental pathways?
+Since there is a good corpus of work on beakdevelopment, it could be used to strengthen the biological aspects and implications of the study.}
+
+We agree that linking elaboration and innovation to genes, developmental processes, or biomechanical constraints is an excellent idea and would add biological depth.
+However, we do not think that there is yet sufficient breadth of understanding of these processes across a wide taxonomic breadth of birds to make meaningful inferences.
+This would be a fascinating direction to explore in the future by collaborating with experts on such processes.
+
+\subsection*{\textcolor{blue}{Minor Comments}}
+
+\textcolor{blue}{$>$ Ln 69-70: A reference to the G matrix is provided, but would a reference to P matrix be more appropriate here?
+A paper from 2013 is referenced but the concept of G (and P) matrices go back further, so I suggest citing additional papers, e.g.: Steppan, S.
+J., Phillips, P.
+C., \& Houle, D.
+(2002).
+Comparative quantitative genetics: evolution of the G matrix.
+Trends in ecology \& evolution, 17(7), 320-327.}
+
+We have now added the suggested reference.
+
+\textcolor{blue}{$>$ Fig. S5: "Palaeognathes" should be "Palaeognathae" or "palaeognaths"}
+
+Figure S5 has now been moved as figure one in the main manuscript and we have now anglicized the clade name in the caption.
+
 \bibliographystyle{naturemag}

### 2023-06-08 Manuscript/reponse_letter.tex
Done! (TODO: needs checks)

@@
 
+\subsection*{\textcolor{blue}{Reviewer 4}}
+
+\textcolor{blue}{Dear authors,}
+
+\textcolor{blue}{Here I summarize my review of a manuscript entitled “Innovation and elaboration on the avian tree of life”.
+In this article, authors present a suite of results obtained through herculean effort to estimate multi-level variance-covariance matrix that represents rates of phenotypic evolution in beak morphology using deep and massive (> 8,000 species) avian phylogeny.
+Authors devised an interesting method to quantify phenotypic evolution in directions of elaboration and innovation to provide multiple lines of evidence showing that species elaborate and clade innovate.}
+
+\textcolor{blue}{I am very excited about this study, both in terms of its findings and its method being devised and used.
+Reflecting my interests, there are many comments, as you see below.
+Should you have any questions for my comments, please do not hesitate to contact me.
+I hope my comments are helpful for further development of this research.}
+
+\textcolor{blue}{Best wishes, Masahito Tsuboi}
+
+Thanks for these positive comments!
+
+\subsection*{\textcolor{blue}{Major comments}}
+
+\textcolor{blue}{1. Framing
+The framing of this research using Endler’s “elaboration and innovation” (here I referred to as e/i framework) should be critically reconsidered.
+I find this framing interesting and useful in some ways, but I also find some problems too.
+I am not suggesting changing this, but I do suggest authors to think again based on what I think is problematic.
+First e/i framework has nothing to do with macroevolution.
+This mismatch will make it difficult for readers to anticipate what this paper is about when reading the introduction.
+Second, e/i framework has little, if not at all, connection to evolutionary quantitively genetics.
+This makes it hard for readers to link this study to its conceptual basis, which is the adaptive landscape framework of Simpson-Lande-Arnold.
+Third, e/i framework has rarely been discussed in evolutionary quantitively genetics literature.
+I am not really a fun of populism, but it is generally a good idea to stick to typical framing of the field.
+This is particularly so when a manuscript is written in a short format.
+Fourth, e/i framework does not seem to help the discussion.
+There are some passing sentences referring to this framework (lines 188-191) but the same sentences will work fine pretty much without referring to it.
+Overall, my suggestion, which is related to my thoughts that I expressed in comment 5, is to devote the second paragraph of introduction to place this research in the context ofshifting adaptive landscape.
+This will help readers to associate this study with conceptually related studies, which will make it easier for introducing methods to readers as well.
+}
+
+We have now completely rewritten the introduction following several reviewer’s comments.
+We now start by introducing concepts from quantitative genetics and macroevolution and introduce the e/i framework only at the end of the introduction as a tool to answer our hypotheses.
+
+\textcolor{blue}{2. Clade orthogonality
+When angles of high dimensional matrices are compared, observed angles should be compared to expected angles from random matrices of the same dimensionality and not to 0 and 90 degrees.
+A method that provides you with null distribution of angles is called “random-skewer”, which is described in Marroig and Cheverud (2001) and implemented in several R-packages (evolvability and evolqg, to my knowledge).
+I suggest authors to test statistical significance of angles between matrices obtained at different levels (Figure 3) using random skewers.}
+
+The random skewers methods has been used and designed to compare covariance matrices and to provide a parametric test whether their orientations are different.
+They are routinely used to compare phenotypic and genotypic matrices (\textbf{P} and \textbf{G}) for example.
+However, here our objective was not to use a parametric test to compare the matrices \textit{per se} but just to see their differences in orientation.
+This is because:
+\begin{itemize}
+\item Using our Bayesian approach, we do use the whole posterior distribution which is approximating to the whole statistical population rather than a sample of it (no need to use a parametric test).
+\item Here we are interested in understanding the change in orientation between pairs of matrices (e.g.
+super-order vs.
+phylogeny) to approximate the potential of innovation of a matrix.
+For two VCV matrices with their own covariance scores equal to 1, the amount of orthogonality is a proxy for their $innovation_{clade}$ score: if they are completely orthogonal, then they will have an $elaboration_{clade}$ score of 0 and a positive $innovation_{clade}$ score (of 1 if they share the same variance).
+\end{itemize}
+We have now specified this in the main text and in the methods sections.
+We have however also added the random skewers methods results in the supplementary materials Figure S13 to have an idea of how different the pairs of matrices are: (lines 469-473)
+
+\noindent\textit{Note that we did not use the random skewers method here \cite{cheverud2007research} because we were interested in measuring the amount of orthogonality between pairs of evolutionary rate matrices rather than simply measuring whether they were different (but see supplementary materials Fig.
+S13 for random skewers correlation results).} %@L_noskewers
+
+\textcolor{blue}{3. Innovation/elaboration analysis
+Study of high dimensional matrices needs to hit a fine balance to explain abstract methods to general readers without compromising conceptual and methodological details.
+I commend authors’ effort in directing this manuscript towards general readers using the innovation/elaboration metaphor, which I think is an excellent idea.
+I however see some rooms to improve descriptions of technical details of this approach.}
+
+\textcolor{blue}{First, I suggest authors to add some descriptions of how innovation/elaboration metric is related to existing approaches in the field.
+Relative eigenvector dispersion (Pavlicev et al., 2009, Watanabe, 2022), conditional evolvability and integration (Hansen \& Houle, 2008) are three metric comes on top of my head that are likely relevant.
+For instance, high elaboration relative to innovation means that the matrix forms a cigar shape than a circle, and that can also be represented by high relative eigenvector dispersion, high integration, and low conditional evolvability.
+Statements of this kind will allow connecting this method to a more standard methods available in the field.
+Descriptions of those metrics are also found in Haber (2016), which might be a useful reference for this article.}
+
+We have made an attempt to link our elaboration and innovation metrics (for clade and species) to the metrics suggested by this reviewer and put our attempts in the supplementary materials (supplementary materials section 4).
+Although some are relatively straightforward (e.g.
+relative eigen value, evolvability which is directly equal to $elaboration_{species}$ or respondability which can be easily calculated using Pythagorean theorems) we are left wondering whether the exercise is actually not beyond the scope of this paper.
+If we can interpret our approach as using the \textbf{R} matrix (see new intro lines 44-60), then relating elaboration and innovation (for clade or species) depending on the \textbf{R} matrix to metrics based on the \textbf{G} matrix is not straightforward.
+In fact, even though it seems feasible mathematically, linking these concepts applied at the macro level (focal = species) and mega level (focal = clades) to ones at the individual or population level (micro) is by no means easy in terms of what these concepts represent biologically (as we know this reviewer is well aware!).
+
+Basically, we believe that even though it would be possible to link these metrics, we are not sure what would be the interpretation of it.
+We illustrate that in the supplementary materials section 4 where we show how linking $elaboration_{species}$ to evolvability (Hanse \& Houle 2008) is trivial ($elaboration_{species}$ = evolvability), linking it to conditional evolvability is going to be very complicated since the concept of autonomy involved in conditional evolvability is directly linked to individuals genetic variation which is hard to mathematically relate to species phenotypic variation (let alone clade’s phenotypic variation).
+We believe however that digging more into this in the future would be a great potential future direction of research.
+
+\textcolor{blue}{Second, I suggest authors to use one term for innovation/elaboration across the entire phylogeny and another term for those at the clade-specific levels.
+For instance, consider using something like Innovation(global as subscript), elaboration(global), Innovation(clade), elaboration(clade).
+This way, readers will always be informed which levels of innovation or elaboration are referred to throughput this study.
+This is related to my next point regarding the comparability of i/e scores.
+That is, Innovation(global) and elaboration(global) should always be comparable because they are defined in the same phenotype space, but Innovation(clade) and elaboration(clade) may not be comparable among clades because they are basically different traits.}
+
+As mentioned in the responses to reviewers above, we have now standardised the distinctions between the different types of elaboration and innovation as $elaboration_{species}$/$innovation_{species}$ for the macroevolutionary scale and $elaboration_{clade}$/$innovation_{clade}$ for the megaevolutionary scale.
+For the main manuscript we have also now standardised the baseline comparisons that are always based on the parent clade (e.g.
+species on orders, orders on super orders and super orders on the whole phylogeny) rather than the parent’s parent clade.
+
+\textcolor{blue}{4. Is Innovation(global) and Innovation(clade) comparable?
+To my understanding, Figure 4 compares i/e scores evaluated at the clade level (not the global level).
+This raises the question of how comparable they are because i/e scores defined by different evolutionary rate matrices are different traits.
+Arguably, Figure 2 and 3 show considerable variation in the first two leading eigenvectors of clade-specific covariance matrices, meaning that i/e scores defined at clade levels are variable across clades.
+Thus, it is an open question whether i/e scores defined by a variety of clades are comparable, calling into some validations for results presented in Figure 4.
+For instance, how Figure 4 look if it plots innovation(global) and elaboration(global) instead of innovation(clade) and elaboration(clade)?}
+
+We have now clarified throughout the manuscript the different comparisons by naming them differently (see response just above).
+Hopefully this does remove potential confusion about figure 4 (now figure 5).
+And regarding the potential problems between comparing $elaboration$/$innovation_{clade}$ to $elaboration$/$innovation_{species}$ we have added the direct link between both metrics in the supplementary materials (section 4).
+We would argue that although they represent quite distinctive biological things, they are both solely based on the same variance-covariance matrices and represent nuances in the different focal levels of these matrices (macro vs.
+mega).
+Since our manuscript now more clearly compares the beak shape variation at the macro and mega level, we feel that the direct comparison between both is still justified.
+
+\textcolor{blue}{5. Shifting adaptive landscape
+I am in principle in favor of authors’ conclusion (lines 213-228).
+However, based on the current presentation, readers need to parse pieces of information spread across the manuscript and re-interpret them in light of some uncited references to reach this conclusion.
+So, I encourage authors to make additional efforts to help readers to find relevant information and arguments that may lead to conclusion that authors allude to.
+What I consider particularly relevant are the following:}
+
+\textcolor{blue}{- citing original formulation of the adaptive zone by G.
+G.
+Simpson, either 1944 (Tempo and Mode) or 1953 (Major Features), will tremendously help framing this manuscript in the context of sifting adaptive landscape.
+Authors could additionally consider referring to Arnold et al.
+(2001), Hansen (2012), Estes and Arnold (2007) and Uyeda et al.
+(2011) as more recent effort in rejuvenating the adaptive zone metaphor in evolutionary quantitate genetics and phylogenetic comparative study}
+
+We have now added the references to Simpson in the relevant places as well as more recent papers discussing the adaptive zone metaphors.
+
+\textcolor{blue}{- rates of evolution matrices (R-matrix) are the sum of effects of underlying variational property (genetic co/variance, developmental bias, line 26) and selection (or the dynamics of adaptive landscape, line 27).
+Because feeding ecology likely exhibit phylogenetic signal, and because adaptive landscape of beak shape should reflect this, and assuming that R is influenced by the shifted adaptive landscape through adaptation to different feeding niche, R of higher taxa likely reflects selection more than R of lower taxa.
+This may explain Figure 4b.}
+
+We have added a sentence along this reviewer’s reflection on lines 307-310: %@L_R_selection
+
+\noindent\textit{This could be explained in part because the evolutionary rate matrix (\textbf{R}) is more influenced by shifts in the adaptive landscape through time and thus reflecting natural selection in deeper nodes more than in younger nodes (Fig.
+\ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}).} %@L_R_selection
+
+\textcolor{blue}{- innovation is by definition lower than elaboration, but there is no evidence for a complete lack of evolution along the direction of innovation, regardless of how this direction is defined (I suppose, and authors could show that explicitly).
+This means that populations could always traverse innovative directions in the morphospace, even though these directions were traveled less often than elaborative directions.
+This helps arguments in line 209-211.}
+
+We have now added the following line to the paragraph (lines 341-343):
+
+\noindent\textit{In other words, populations can always change morphologically within the shapespace with minimal amounts of innovation over long periods of time.} %@L_shapespacetravel
+
+\subsection*{\textcolor{blue}{Technical comments}}
+
+\textcolor{blue}{Lines 63-74: Variance covariance matrices estimated from hierarchical phylogenetic mixed models are better termed as “rates of evolution matrix” based on multivariate Brownian motion model of evolution.
+This is the same as R-matrix in Houle et al.
+(2017).
+R-matrix is a better estimate of rates of phenotypic evolution than phenotypic divergence matrix (D-matrix, used in (McGlothlin et al., 2018) for instance).
+I suggest authors to make it explicit that estimates from pMCMC are rates of macroevolution and not a simple divergence matrix.
+This is one of the core methodological strengths of this method and should be highlighted more.}
+
+We have now added a direct reference to this point in the introduction (lines 52-60):
+
+\noindent\textit{At this scale the major axis of phenotypic variation among species \cite{marroig2005size, fasanelli2022allometry} can be estimated from a matrix of divergences in species mean phenotypic traits (e.g.
+the \textbf{D} matrix of \cite{BlowsHiggie2003, mcglothlin2018adaptive}.
+These matrices do not take into account phylogenetic relationships among species.
+In contrast, the \textbf{R} matrix of \cite{Houle2017} describes the rate of among-species divergence in species traits and explicitly incorporates phylogeny (also see the \textbf{B} matrix of \cite{Machado2020}) %@L_Rmatrix.
+We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}. %TG: TODO: change throughout (that's the VCV matrix now)
+The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales.
+}
+
+\textcolor{blue}{Lines 98-99, 108: Following my previous point, this is the direction in beak morphospace in which the rate of evolution is the fastest (again assuming multivariate BM).
+Perhaps this could better be stated here as such in the first instance, instead of “global phylogenetic major axis”.
+I would then call it in the rest of the manuscript “the major axis in rates of beak shape evolution”.
+Lines 127-137: this part motivate me even more strongly to suggest using “rates of evolution matrix”.
+Here, using the term “phylogenetic” co/variance matrices should be avoided, because this term is normally used to represent the matrix that express shared branch lengths among taxa.
+I suggest authors to explain this part using the term “rates of evolution matrix”.}
+
+We have now standardised the mentions to the eigenvectors of the VCV matrices as the “phylogenetic major axis of beak variation” throughout the manuscript.
+We also now specify that it is based on the “evolutionary rates matrix” (aka the \textbf{R} matrix).
+Hopefully this will clarify it for the readers.
+
+\textcolor{blue}{Lines 287-289: This description seems like authors took summed-distance of each species from the global centroid, which must not have been the case because they should all be close to 1 as a result of Generalized Procrustes analysis (lines 354-356).
+My guess is that they are Euclidian distances from the global mean beak shape.
+In any cases, description here must have some errors and this needs to be corrected.}
+
+We have now clarified this sentence by specifying it was the Euclidean distance.
+
+\textcolor{blue}{Lines 341-343: I do not recommend scaling scores by maximum values because maximum/minimum are highly influenced by sample size.
+An alternative scaling method is variance standardization (i.e.
+divide traits by its own variance).
+It could also be argued that you do not need to scale these values because they are all scaled to a unit centroid size as part of Generalized Procrustes analyses, which is equivalent to mean-standardization of ratio-scale traits.
+It would be good to eliminate the possibility that results shown in Figure 4 are artifact of scaling procedure by comparing presented results with those based on non-scaled traits and variance-scaled traits.}
+
+We agree with this reviewer that it would be better to scale these distributions using a variance-standardisation but unfortunately this is the standard procedure for data that are a sample of a statistical population and are normally distributed.
+This is not the case for us: they are not normally distributed with homoscedasticity of variance and represent the full statistical population.
+We thus kept the figure displaying the maximum scaled $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$, $innovation_{clade}$ scores.
+
+\textcolor{blue}{Projection and rejection: I have a couple of suggestions and questions.}
+
+\textcolor{blue}{1 – according to section 4.1, it seems that authors essentially did an eigen analysis of a shrunken variance covariance matrix.
+What is the purpose of shrinking a matrix using O’n?
+I guess this is something to do with descriptions in line 433-435.
+An explanation relating this part and section 4.1 would be very helpful.}
+
+We have now added the following specific explanation in the supplementary materials (last paragraph of section 4.1):
+
+\noindent\textit{The use of the scaled 95\% confidence interval hyper-sphere matrix $O'_{n}$ allows both (1) to standardise the resulting $M_{n}$ (the scaled version of $VCV_{n}$) so that the unit vector of the resulting space corresponds to the 95\% CI on every dimension (for any dimension $i$ in the matrix $M_{i,n}$, any vector with a norm smaller or greater than 1 is respectively inside or outside the 95\% CI boundary).
+And (2) greatly speed up the calculations of projection and rejection in an algorithm (see below) since the hypersphere $O'_{n}$ can be calculated only once for any number of matrix and the projection of rejection values are directly columns in the output transformed matrix (after the projection procedure, see below).
+This greatly reduces RAM and CPU usage and scaling for very large number of matrices.}
+
+\textcolor{blue}{2 – based on line 476, the best estimate was used to perform this analysis but in line 136 (and in some other places), there are indications that 4000 posterior samples are used instead.
+When does models\$Sol used and when does posterior samples used?}
+
+This procedure is applied for each VCV matrix individually (i.e. 4000 times per random term in our model). We have now specified this in the methods section (lines 450-454):
+
+\noindent\textit{The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection};  note that the procedure described is applied to each individual evolutionary rate matrix for each term in the model, e.g.
+for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score).}
+
+\textcolor{blue}{3 – I do not recommend mentioning Stack Overflow as a reference (line 481) because (a) this will not credit original references to this solution and (b) solutions in are Stack Overflow not critically examined as peer-reviewed journal articles did.}
+
+We didn’t find any specific reference for the linear algebra solution provided in the Stack Overflow post but (1) we still want to give credit to the author of the solution (although, yes, unfortunately it will not credit them directly on a classic citation counter); (2) we want to highlight that we did not come up with this solution on our own but that we are standing on the shoulders of a helpful and knowledgeable community.
+And regarding the critical examination of Stack Overflow solutions, in a sense, we feel that they are much more critically and openly examined than many papers published in peer reviewed journals.
+We are actually very grateful for this reviewer to have spent time critically reviewing the supplementary materials of this paper which is very rarely done!
+
+\textcolor{blue}{4 - although it eventually is a matter of taste, I do not think the word “rejection” is appropriate.
+The “rejected” values are simply another projection (hence a use of another term can be misleading) of high-dimensional trait onto the second major axis (hence a use of a term that could be associated with insignificance can be misleading) that is orthogonal to the first major axis.}
+
+We are not sure about how to deal with this comment, here we used what we believe is the standard terminology to refer to vector projection (e.g.
+\url{https://en.wikipedia.org/wiki/Vector_projection}) were for any vector $a$ and $b$ with $a'$ the projection of a onto b, we call $a'' = a - a'$ the rejection of $a$ onto $b$.
+If c is $a$ vector orthogonal to $b$, then, indeed, $a''$ is just the projection of $a$ onto $c$ and $a'$ is the rejection of $a$ onto $c$.
+
+\subsection*{\textcolor{blue}{Minor comments and suggestions}}
+
+\textcolor{blue}{First paragraph: The first few sentences of this paragraph include quite a few technical terms without definitions.
+Brief definition statement and a usage of a more common term would help readers from outside of the field.
+For instance
+- instead of using “evolutionary lines of genetic least resistance” right at the first sentence, it could be said as “a direction in the phenotype space that exhibit the highest amount of additive genetic variance, which sometimes is referred to as the genetic line of least resistance”}
+
+We have now entirely rewritten the introduction.
+
+\textcolor{blue}{- in line 21 and line 24 “evolutionary line” could be replaced by “genetic line” simply because the latter is more well-defined in the literature.}
+
+We have now entirely rewritten the introduction.
+
+\textcolor{blue}{Line 24: Houle et al. (2017) should be cited here.}
+
+We have now entirely rewritten the introduction (and cite Houle et al. 2017 on line 55).
+
+\textcolor{blue}{Line 26: Perhaps replacing “genomic” to “genetic” will make it easier for readers to associate the framing of this work to quantitative genetics.}
+
+Done.
+
+\textcolor{blue}{Line 116: Based on similar reason as I stated above, replacing “lines of least evolutionary resistance” with “genetic lines of least resistance” would be more consistent with popular usage of this term in the literature}
+
+We have changed this as suggested on line 200.
+
+\textcolor{blue}{Line 167: I’d say that it is often assumed that such constraints are relaxed in deep time.
+So, either using “occasionally” instead of “often”, or replacing “than” with “as” would flow more naturally (to me).}
+
+We changed the sentence to (lines 262-264):
+
+\noindent\textit{Across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is sometimes assumed \cite{venditti2011multiple}.
+% @L_occasionaly
+}
+
+\textcolor{blue}{Line 168: I think “any” is too strong. “many” should be more appropriate.}
+
+We have changed this and added a link to an illustration in the supplementary materials. (lines 265-269)
+
+\textcolor{blue}{Line 193: I love this header!}
+
+Thanks!
+
+\textcolor{blue}{Figure 1: I suggest authors to increase the font size of bottom-left and bottom-right panels.}
+
+We have now removed these panels from what is now figure 2.
+
+\textcolor{blue}{Also, if I understand correctly, the bottom-right panel shows median of pooled elaboration and innovation scores (lines 295-296).
+If I am correct, pooling does not make sense because innovation and elaboration are two different traits .
+Instead of pooling these, why not show 6 box plots, showing innovation and elaboration at 3 levels?}
+
+We have now removed the boxes from this figure since we now only focus on the $elaboration_{species}$ and $innovation_{species}$ at the order level in the main manuscript.
+
+\textcolor{blue}{Figure S7: Why does proportion of variance sometimes does not follow the PC rank?
+Are these proportion of variance in PCs defined by the global R-matrix projected on each clade’s D-matrix?}
+
+This because in dark grey we are displaying the amount of variance for different parts of the shapespace for different groups.
+The point here is that not all groups are distributed as a random subsample of the shapespace (if it would be the case, each group would have the same pattern of variance as suspected by the reviewer), it’s the opposite and thus our justification for using 8 dimensions.
+
+\noindent\textit{Note that not all super orders have their own variance explained matching the variance explained by the entire ordination.
+For example, Strisores only need the first three dimensions to explain 95\% of their variance within the shapespace whilst Aequornithes need 8 dimensions to achieve the same amount of variance.}
+
+\textcolor{blue}{A suggestion: One question that remains to be addressed, and that can readily be addressed by results presented in this study is how much elaboration/innovation matters for the actual evolution in beak shape.
+Authors can simply test this by regressing “distance” shown in Figure 1 (which I believe is the Euclidian distances from global mean shape.
+If not, this analysis should use this) against any of elaboration/innovation scores.
+The most interesting is perhaps the global innovation/elaboration scores because they will show how much extreme shapes are generated by innovation/elaboration.
+By the look of Figure 1, there is a clear relationship between distance and elaboration (all).
+Another interesting index to correlate with Euclidian distance is the angle between clade-specific variance covariance matrix and the global variance covariance matrix.
+One could even run a multiple regression model of something like:}
+
+\textcolor{blue}{(distance $\sim$ elaboration score + innovation score + angle, random = phylogeny)}
+
+\textcolor{blue}{In MCMCglmm or PGLS to partial out the effect of different matrix features to the evolution of beak shape.
+Those analyses might be useful to show some of presented results in a more succinct manner.}
+
+We have now added an entire along this reviewer’s comment in the main text.
+We now analyse and show the PGLS results which interestingly give results similar to our Bhattacharyya approach showing that elaboration and innovation is common at the macroevolutionary scale and that innovation is more common at the megaevolutionary scale (lines 311-324 and dedicated methods section):
+
+\noindent\textit{We further tested the contributions of elaboration and innovation at different scales by examining the consequences of elaboration and innovation for the observed divergence of bird beak shapes.
+Specifically, we calculated the distance to centroid of avian beak morphospace for each species (Fig.
+\ref{Fig:phylogeny}).
+We then used phylogenetic generalised least squares (PGLS; \cite{phylom}) to model distance to centroid as a function of $elaboration_{species}$, $innovation_{species}$ and $elaboration_{clade}$ and $innovation_{clade}$.
+We ran the PGLS analysis using both single and multiple predictors.
+Our analyses indicate that the majority of variation (r2 = 0.888; Table 1) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
+When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table 1).
+In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table 1).
+However, $innovation_{clade}$ is nonetheless an important contributor to total beak shape space.
+We reach this conclusion because models with interactions between species and clade metrics provide by far the best model fit overall and removal of terms indicate that the most important interactions are between $innovation_{clade}$  and $innovation_{species}$ followed by $innovation_{clade}$ and $elaboration_{species}$.
+The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table 1)
+Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space.
+}
+
 \bibliographystyle{naturemag}

### 2023-06-09 Manuscript/supplementaries.tex
fixed typo in Zoe's name
Final bits resubmitted

@@
 Emma C. Hughes,
-Zoe K. Varley,
+Zo\"{e} K. Varley,
 Natalie Cooper,
@@
 
-With $$\vec{species}_{x}$$ being the vector in $n$ dimensional space for species $x$. And $$\textbf{v}_{clade}$$ being the eigen vector of a specific clade (e.g. $$\textbf{v}_{clade_{c}}$$)
+With $\vec{species}_{x}$ being the vector in $n$ dimensional space for species $x$. And $\textbf{v}_{clade}$ being the eigen vector of a specific clade (e.g. $\textbf{v}_{clade_{c}}$)
 

### 2023-06-09 Manuscript/second_submission.tex
Create second_submission.tex

--- 2023-06-08 Manuscript/main.tex
@@
 Emma C. Hughes${^1}$,
-Zoe K. Varley${^3}$,
+Zo\"{e} K. Varley${^3,4}$,
 Natalie Cooper$^{3,+}$,
@@
 We then used linear algebra to project %@L_linear_algebra_stuff
-a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. S\ref{Fig:cheatsheet}).
+a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:cheatsheet}).
 We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ score.
@@
 the direction of evolution of an order differs from its parent super-order).
-These scores were calculated for each of 4000 posterior pairs of clade vs.
-evolutionary rate matrix (Fig. \ref{Fig:ellipses} boxplots).
+These scores were calculated for each of 4000 posterior pairs of clade \textit{vs.} evolutionary rate matrix (Fig. \ref{Fig:ellipses} blue and orange lines).
 We found that for most clades the phylogenetic major axis of beak variation is not aligned with the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}).
 Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the evolutionary rate matrix for the class-wide phylogenetic major axis of beak variation and each super-order and order (Fig.
-\ref{Fig:ellipses}b) highlight the striking variation in the orientation of beak shape major axes among clades (Fig.
+\ref{Fig:ellipses}) highlight the striking variation in the orientation of beak shape major axes among clades (Fig.
 \ref{Fig:orthogonality}).
@@
 \centering
-\small
+\footnotesize
    \begin{tabular}{rrllllllllrr}
      \hline
-    Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & $\Delta$AIC & Adj.
+    Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ $\times$ $I_{clade}$ & $E_{species}$ $\times$ $E_{clade}$ & $I_{species}$ $\times$ $I_{clade}$ & $I_{species}$ $\times$ $E_{clade}$ & $\Delta$AIC & Adj.
 $r^2$ \\
@@
 The table shows parameter estimates of predictors from 12 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms.
-Models 1-5 are multiple regressions including all variables and one or more interaction terms; models 6-8 are are multiple regressions including two or more variables and no interaction terms; models 9-12 are single predictor models.
+Models 1-5 are multiple regressions including all variables and one or more interaction terms (e.g.
+model 1 contains all the terms and their interactions); models 6-8 are are multiple regressions including two or more variables and no interaction terms (e.g.
+model 6 contains all the terms but no interactions); models 9-12 are single predictor models (e.g.
+model 9 contains only one term).
 Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e.
@@
 We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$ from any individual model) and relative model fit and interpretation of parameters was unaffected.
-E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
+$E$ = elaboration; $I$ = innovation; AIC = Akaike Information Criterion; *** = p-value $<$ 0.001.}
 \label{Tab:pgls}

### 2023-06-19 Analysis/07-Orthogonality_and_statistics.Rmd
reinstall

@@
 ```{r, eval = TRUE, echo = FALSE, fig.height = 20, fig.width = 8}
-#TODO: make the groups nested in the fable
 source("../Functions/ploting.utilities.R")
@@
 pdf("../Manuscript/Figures/orthogonality_results.pdf", width = 9, height = 9)
-plot.orthogonality(super_orders, orders_n2, orders, col = gg.color.hue(4)[1:2])
+plot.orthogonality(super_orders, orders_n2, orders, col = c("purple", "#FBB117"))
 legend("topleft", legend = c("Super-orders (vs.
@@
 phylogeny)", "Orders (vs.
-super-orders)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8, bty = "n")
+super-orders)"), col = c("purple", "#FBB117", adjustcolor("#FBB117", alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8, bty = "n")
 dev.off()
@@
 pdf("../Manuscript/Figures/orthogonality_results_passeriformes.pdf", width = 9, height = 9)
-plot.orthogonality(sub_orders, families_n2, families, col = gg.color.hue(4)[3:4])
+plot.orthogonality(sub_orders, families_n2, families, col =c("purple", "#FBB117"))
 legend("topleft", legend = c("Sub-orders (vs.
@@
 sub-orders)", "Families (vs.
-passeriformes)"), col = c(gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8, bty = "n")
+passeriformes)"), col = c("purple", "#FBB117", adjustcolor("#FBB117", alpha.f = 2/3)), pch = c(19, 19, 21), lty = c(1,1,2), cex = 0.8, bty = "n")
 dev.off()
@@
 ## Distribution of elaboration and innovation at the clade level
-scale.group <- function(X) return(X/max(X))
+scale.group <- function(X) return(X/quantile(X, prob = 0.975)[1])
 elaboration_super_orders <- lapply(get.disparity(results$super_results_between$position, concatenate = FALSE), scale.group)
@@
 
+## RAM cleaning
+rm(results)
+
 ## The objects above are the ones that contain all the elaboration and innovation values by clade level
 ## These bits are for the Bhatt coeff
-elaboration_group <- rbind(do.call(rbind, elaboration_super_orders), do.call(rbind, elaboration_orders))
-innovation_group <- rbind(do.call(rbind, innovation_super_orders), do.call(rbind, innovation_orders))
+# elaboration_group <- rbind(do.call(rbind, elaboration_super_orders), do.call(rbind, elaboration_orders))
+# innovation_group <- rbind(do.call(rbind, innovation_super_orders), do.call(rbind, innovation_orders))
 
 ## Measuring the area under the curves (for the Bhatt coeff)
-elaboration_species_auc <- apply(elaboration_species, 2, area.density)
-innovation_species_auc  <- apply(innovation_species, 2, area.density)
-elaboration_group_auc   <- apply(elaboration_group, 2, area.density)
-innovation_group_auc    <- apply(innovation_group, 2, area.density)
+# elaboration_species_auc <- apply(elaboration_species, 2, area.density)
+# innovation_species_auc  <- apply(innovation_species, 2, area.density)
+# elaboration_group_auc   <- apply(elaboration_group, 2, area.density)
+# innovation_group_auc    <- apply(innovation_group, 2, area.density)
 
@@
   scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
-  scale_fill_manual(values = c("yellow", "purple")) +
+  scale_fill_manual(values = c("#FBB117", "purple")) +
   xlab("Relative elaboration") +
@@
   scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
-  scale_fill_manual(values = c("yellow", "purple")) +
+  scale_fill_manual(values = c("#FBB117", "purple")) +
   xlab("Relative innovation") + 
@@
 
-ggsave("../Manuscript/Figures/relative_EI_gg_style_unscaled.pdf", plot=p, height=6, width=12)
+ggsave("../Manuscript/Figures/relative_EI_gg_style_95CIscaled.pdf", plot=p, height=6, width=12)
 ```
@@
 group_pairs <- sapply(name.subsets(all_birds_shapespace)[-36], function(x) c(x, "phylogeny"), simplify = FALSE)
+group_pairs_orders <- list(
+    "Galliformes" = c("Galliformes", "Galloanserae"),
+    "Accipitriformes" = c("Accipitriformes", "Telluraves"),
+    "Strigiformes" = c("Strigiformes", "Telluraves"),
+    "Musophagiformes" = c("Musophagiformes", "Aequornithes"),
+    "Trogoniformes" = c("Trogoniformes", "Telluraves"),
+    "Columbiformes" = c("Columbiformes", "Columbimorphae"),
+    "Caprimulgiformes" = c("Caprimulgiformes", "Strisores"),
+    "Bucerotiformes" = c("Bucerotiformes", "Telluraves"),
+    "Falconiformes" = c("Falconiformes", "Telluraves"),
+    "Pterocliformes" = c("Pterocliformes", "Columbimorphae"),
+    "Sphenisciformes" = c("Sphenisciformes", "Aequornithes"),
+    "Podicipediformes" = c("Podicipediformes", "Mirandornithes"),
+    "Ciconiiformes" = c("Ciconiiformes", "Aequornithes"),
+    "Tinamiformes" = c("Tinamiformes", "Paleognathae"),
+    "Suliformes" = c("Suliformes", "Aequornithes"),
+    "Anseriformes" = c("Anseriformes", "Galloanserae"),
+    "Apodiformes" = c("Apodiformes", "Strisores"),
+    "Otidiformes" = c("Otidiformes", "Otidimorphae"),
+    "Cuculiformes" = c("Cuculiformes", "Otidimorphae"),
+    "Psittaciformes" = c("Psittaciformes", "Telluraves"),
+    "Pelecaniformes" = c("Pelecaniformes", "Aequornithes"),
+    "Procellariiformes" = c("Procellariiformes", "Aequornithes"),
+    "Piciformes" = c("Piciformes", "Telluraves"),
+    "Passeriformes" = c("Passeriformes", "Telluraves"),
+    "Coraciiformes" = c("Coraciiformes", "Telluraves"))
 
@@
 skewers_p <- dispRity(all_birds_shapespace, metric = as.covar(skewers.fun.p), between.groups = group_pairs, verbose = TRUE)
+skewers_r_order <- dispRity(all_birds_shapespace, metric = as.covar(skewers.fun), between.groups = group_pairs_orders, verbose = TRUE)
+skewers_p_order <- dispRity(all_birds_shapespace, metric = as.covar(skewers.fun.p), between.groups = group_pairs_orders, verbose = TRUE)
 
@@
 
-col <- rev(c(rep(gg.color.hue(4)[1], 8), rep(gg.color.hue(4)[2], 27)))
+## Same but for the orders
+skewers_distrib_order <- lapply(get.disparity(skewers_r_order, concatenate = FALSE), quantile, probs = c(0.5, 0.025, 0.25, 0.75, 0.975))
+skewers_median_p_order <- lapply(get.disparity(skewers_p_order, concatenate = FALSE), median)
+plot_names_order <- unlist(lapply(strsplit(names(skewers_distrib_order), split = ":"), `[[`, 1))
+names(skewers_distrib_order) <- names(skewers_median_p_order) <- plot_names_order
 
+## Set the colours to match the ones from the ortho figure
+col <- c(rep("purple", 8), rep("#FBB117", 27))
+
+pdf(height = 12, width = 7, file = "../Manuscript/Figures/random_skewers.pdf")
 par(mar = c(5, 17, 4, 2))
-plot(NULL, xlim = c(0, 1), ylim = c(1,length(skewers_distrib)), xlab = "Skewers correlation", ylab = "", yaxt = "n")
+plot(NULL, xlim = c(0, 1), ylim = c(length(skewers_distrib),1), xlab = "Skewers correlation", ylab = "", yaxt = "n")
 abline(v = 0.5, lty = 2, col = "grey", lwd = 0.5)
-axis(2, at = length(skewers_distrib):1, labels = plot_names, las = 2)
+axis(2, at = 1:length(skewers_distrib), labels = plot_names, las = 2)
 ## Skewers
@@
     points(x = skewers_distrib[i][[1]][1], y = i, pch = 19, col = col[i])
-    text(x = skewers_distrib[i][[1]][1]+0.03, y = i+0.2, label = p.star(skewers_median_p[[i]]), col = col[i])
+    text(x = skewers_distrib[i][[1]][1]+0.03, y = i-0.2, label = p.star(skewers_median_p[[i]]), col = col[i])
+    ## Add orders (if available)
+    if(names(skewers_distrib)[i] %in% plot_names_order) {
+        n <- which(names(skewers_distrib)[i] == plot_names_order)
+        col_n <- adjustcolor(col[i], alpha = 1/2)
+        ## Plotting the order-superorder level results
+        lines(x = skewers_distrib_order[n][[1]][c(2, 5)], y = rep(i, 2)+0.2, lty = 3, lwd = 1, col = col_n)
+        lines(x = skewers_distrib_order[n][[1]][c(3, 4)], y = rep(i, 2)+0.2, lty = 1, lwd = 2, col = col_n)
+        points(x = skewers_distrib_order[n][[1]][1], y = i+0.2, pch = 21, col = col_n)
+        text(x = skewers_distrib_order[n][[1]][1]+0.03, y = i, label = p.star(skewers_median_p_order[[n]]), col = col_n)
     }
-
+}
+dev.off()
 ```

### 2023-06-19 Analysis/09-Figure_key.Rmd
reinstall

@@
 ## Creating the subsets
-superorder_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
+superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
       data          = shapespace,
@@
 ## Get projection lines
-get.proj.line <- function(point, base) {
+get.proj.line <- function(point, base, just.proj = FALSE) {
   point <- t(point)
@@
   proj_y <- ( (point[1]*base[1] + point[2]*base[2]) / (base[1]*base[1] + base[2]*base[2]) ) * base[2];
+  if(just.proj) {
+      return(cbind(proj_x, proj_y))
+  } else {
       return(rbind(cbind(proj_x, proj_y), point))
+  }
+}
+get.any.proj.lines <- function(points, base) {
+      ## 1 - Get the coordinates of the points
+      points_coords <- points[, ]
+            # points(points_coords, col = "grey")
+      ## 2 - Get the coordinates of the base
+      base_coords <- base
+            # lines(base_coords, col = "grey")
+
+      ## 3 - translate the points and base onto the origin of the base
+      translation_vector <- c(0, 0) - base_coords[2,]
+      points_coords_translated <- t(t(points_coords) + translation_vector)
+      base_coords_translated <- t(t(base_coords) + translation_vector)
+            # lines(base_coords_translated)
+            # points(points_coords_translated)
+
+      ## 4 - project all the translated points onto the translated base
+      lines_out_translated <- t(apply(points_coords_translated, 1, get.proj.line, base = base_coords_translated, just.proj = TRUE))
+            # lines(rbind(lines_out_translated[1, ], points_coords_translated[1, ]))
+
+      ## 5 - untranslate the points
+      lines_out <- t(t(lines_out_translated) - translation_vector)
+            # lines(rbind(lines_out[1, ], points_coords[1, ]), col = "red")
+      return(lines_out)
 }
@@
 plot(NULL, xlab = "PC1", ylab = "PC2", xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5))
-# lines(base_vector)
+## Project all the points on the base line
+proj_points <- get.any.proj.lines(matrix, base = base_vectors$phylogeny[[1]])
+## Plot all these lines
 for(i in 1:nrow(matrix)) {
-      lines(get.proj.line(matrix[i, ], base_vectors$phylogeny[[1]]))
+    lines(rbind(matrix[i, ], proj_points[i, ]), col = "grey")
 }
+## Add the points
 points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)
+## Add the ellipses
 covar.plot(space_paleo,
@@
 plot(NULL, xlab = "PC1", ylab = "PC2", xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5))
+## Project all the points on the base line
+proj_points <- get.any.proj.lines(matrix, base = base_vectors$Paleognathae[[1]])
+## Plot all these lines
+for(i in 1:nrow(matrix)) {
+    lines(rbind(matrix[i, ], proj_points[i, ]), col = gg.color.hue(1))
+}
+## Add the points
 points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)
-
-base <- base_vectors$Paleognathae[[1]]
-# base <- matrix(c(base)[c(1, 1, 4, 4)], 2, 2, byrow = FALSE) # or c(1, 4, 3, 2)
-for(i in 1:nrow(matrix)) {
-      lines(get.proj.line(matrix[i, ], base))
-}
-
-points(shapespace[which(shapespace$level1 == "Paleognathae"), c(1,2)], col = gg.color.hue(1), pch = 19)
+## Add the ellipses
 covar.plot(space_paleo,
@@
 ## Get the data (in the right order)
-load("../Data/Processed/ploting_data.rda")
-elaborations <- ploting_data$median_elaborations[tree_paleo$tip.label]
-innovations <- ploting_data$median_innovations[tree_paleo$tip.label]
+load("../Data/Processed/ploting_data_order.rda")
+elaborations <- ploting_data_order$median_elaborations[tree_paleo$tip.label]
+innovations <- ploting_data_order$median_innovations[tree_paleo$tip.label]
 
@@
 
-pdf("../Manuscript/Figures/figure_key_tree_example2.pdf")
-plot(ladderize(tree_paleo), show.tip.label = FALSE)
-dev.off()
-
+# Focal species is Dromaius
 ```

### 2023-06-19 Analysis/11-sensitivity_analysis.Rmd
reinstall

--- 2022-06-16 Analysis/10-sensitivity_analysis.Rmd
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 ## Random space
@@
 
-<!-- Figure 9: BMised shapespace mcmcmcglmmm diagnosis -->
+Figure 9: BMised shapespace mcmcmcglmmm diagnosis
 
@@
 
+Figure 10: the VCV matrices from the original shapespace.
+
 ```{r, eval = TRUE, echo = FALSE}
@@
 ```
+
+Figure 11: the VCV matrices from the balled shapespace.
 
@@
 
-```{r, eval = FALSE, echo = FALSE}
+Figure 12: the VCV matrices from the randomised shapespace.
+
+```{r, eval = TRUE, echo = FALSE}
 ## Get the shapespace
@@
 ```
+
+Figure 13: the VCV matrices from the original shapespace.
+
+## Projection analyses
+
+```{r, eval = FALSE, echo = FALSE}
+## Getting all the projections
+get.all.projetion.analyses <- function(space) {
+      all_results <- list()
+      all_results$groups <-  dispRity.covar.projections(space,
+                                            type = "groups",
+                                            base = "phylogeny",
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+      all_results$proj_phylo <- dispRity.covar.projections(space,
+                                            type = "elements",
+                                            base = "phylogeny",
+                                            inc.base = TRUE,
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+      all_results$proj_group <- dispRity.covar.projections(space,
+                                            type = "elements",
+                                            verbose = TRUE,
+                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
+      return(all_results)
+}
+
+shapespace_results <- get.all.projetion.analyses(chara_shapespace)
+ballspace_results <- get.all.projetion.analyses(chara_ballspace)
+randospace_results <- get.all.projetion.analyses(chara_randospace)
+BMspace_results <- get.all.projetion.analyses(chara_BMspace)
+
+sensitivity_results <- list("original"   = shapespace_results,
+                            "balled"     = ballspace_results,
+                            "randomised" = randospace_results,
+                            "BMised"     = BMspace_results)
+save(sensitivity_results, file = "../Data/Processed/sensitivity_results.rda")
+```
+
+```{r, eval = TRUE, echo = FALSE}
+## Loading the results
+library(dispRity)
+source("../Functions/ploting.utilities.R")
+source("../Functions/correlations.utilities.R")
+load("../Data/Processed/sensitivity_results.rda")
+
+## Getting the average projection/rejections
+original_proj_phylo <- get.correlations(sensitivity_results$original$proj_phylo)
+ballspace_proj_phylo <- get.correlations(sensitivity_results$balled$proj_phylo)
+randospace_proj_phylo <- get.correlations(sensitivity_results$randomised$proj_phylo)
+BMspace_proj_phylo <- get.correlations(sensitivity_results$BMised$proj_phylo)
+original_proj_group <- get.correlations(sensitivity_results$original$proj_group)
+ballspace_proj_group <- get.correlations(sensitivity_results$balled$proj_group)
+randospace_proj_group <- get.correlations(sensitivity_results$randomised$proj_group)
+BMspace_proj_group <- get.correlations(sensitivity_results$BMised$proj_group)
+
+## Plotting the original space
+plot.cor <- function(data, col = col_vector, ...) {
+      data <- lapply(data, function(x) return(x[c(1,2)]))
+      range <- range(unlist(data))
+      plot(NULL, xlim = range, ylim = range, xlab = "elaboration", ylab = "innovation", ...)
+      abline(v = 1, lty = 2)
+      abline(h = 1, lty = 2)
+      points(data[[1]], col = col[1], pch = 19)
+      points(data[[2]], col = col[2], pch = 19)
+      points(data[[3]], col = col[3], pch = 19)
+}
+
+col_vector <- rev(c("blue", "darkgreen", "orange"))
+par(mfrow = c(2,2))
+plot.cor(original_proj_phylo, main = "Original space")
+plot.cor(ballspace_proj_phylo, main = "Balled space")
+plot.cor(randospace_proj_phylo, main = "Random space")
+plot.cor(BMspace_proj_phylo, main = "BM space")
+
+```
+
+Figure 14: comparison between the different space modifications
+
+```{r, eval = TRUE, echo = FALSE}
+## Loading the results
+par(mfrow = c(2,2))
+plot.cor(original_proj_group, main = "Original space (macro level)")
+plot.cor(ballspace_proj_group, main = "Balled space (macro level)")
+plot.cor(randospace_proj_group, main = "Random space (macro level)")
+plot.cor(BMspace_proj_group, main = "BM space (macro level)")
+```
+
+Figure 15: comparison between the different space modifications (when projecting onto their ownb group)
+
+```{r, eval = FALSE, echo = FALSE}
+par(mfrow = c(2,2))
+## Get the scaling the variance of each dimension
+ballspace <- shapespace
+ballspace[, selected_dim] <- apply(shapespace[, selected_dim], 2, scale)
+
+## Ballspace var
+ball_var_prop <-  apply(ballspace[, selected_dim], 2, var)
+ball_var_prop <- ball_var_prop/sum(ball_var_prop)*100
+
+## Comparing D1 and D3 for the contrast
+png("balling_effect.png", height = 8, width = 8)
+par(mfrow = c(2,2))
+plot(shapespace[, c(1,3)], xlab = write.pc(1, shape_var_prop), ylab = write.pc(3, shape_var_prop),
+      xlim = range.axis(shapespace, selected_dim), ylim = range.axis(shapespace, selected_dim),
+      main = "Original shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+legend("bottomrigh", pch = 19, legend = c("gulls", "plovers", "sandpipers"), col= c("orange", "blue", "darkgreen"))
+
+plot(ballspace[, c(1,3)], xlab = write.pc(1, ball_var_prop), ylab = write.pc(3, ball_var_prop),
+      xlim = c(-4, 4), ylim = c(-4, 4),
+      main = "Balled shapespace", pch = 19,
+      col = c("orange", "blue", "darkgreen")[as.numeric(shapespace$level)])
+plot.cor(original_proj_phylo, main = "Original space")
+plot.cor(ballspace_proj_phylo, main = "Balled space")
+dev.off()
+```

### 2023-06-19 Analysis/14-Elab_innov_through_time.Rmd
reinstall

@@
 
-pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_clade_through_time.pdf")
-par(mfrow = c(2,1))
-plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "", main = "Clade's elaboration/innovation\nthrough time")
+pdf(height = 7, width = 14, file = "../Manuscript/Figures/Elaboration_and_innovation_clade_through_time.pdf")
+par(mfrow = c(1,2))
+plot.age.data(elaboration_plot_data, ylab = expression('elaboration'['clade']), xlab = "Time (Mya)", main = "")
 legend("topleft", pch = 21, col = unlist(orders_elaboration$col), legend = rownames(orders_elaboration), cex = 0.7)
 legend("top", pch = 19, col = unlist(super_orders_elaboration$col), legend = rownames(super_orders_elaboration), cex = 0.7)
-plot.age.data(innovation_plot_data, ylab = "Innovation", xlab = "Time (Mya)")
+plot.age.data(innovation_plot_data, ylab = expression('innovation'['clade']), xlab = "Time (Mya)")
 dev.off()
@@
 
-pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_species_through_time.pdf")
-par(mfrow = c(2,1))
-plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "", main = "Species's elaboration/innovation\nthrough time (per clade)")
+pdf(height = 7, width = 14, file = "../Manuscript/Figures/Elaboration_and_innovation_species_through_time.pdf")
+par(mfrow = c(1,2))
+plot.age.data(elaboration_plot_data, ylab = expression('elaboration'['species']), xlab = "Time (Mya)", main = "")
 legend("topright", pch = 21, col = unlist(orders_elaboration$col), legend = unlist(orders_elaboration$subsets), cex = 0.7)
 legend("top", pch = 19, col = unlist(super_orders_elaboration$col), legend = unlist(super_orders_elaboration$subsets), cex = 0.7)
-plot.age.data(innovation_plot_data, ylab = "Innovation", xlab = "Time (Mya)")
+plot.age.data(innovation_plot_data, ylab = expression('innovation'['species']), xlab = "Time (Mya)")
 dev.off()

### 2023-06-19 Analysis/16-ElabInovScriptPGLS.Rmd
reinstall

---
title: "Linking different metrics to Elaboration and Innovation"
author: "Gavin Thomas"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

```{r}
library(dplyr)
library(ape)
library(phylolm)
library(motmot)
library(xtable)

## Read data for elaboration and innovation scores for species and clades, and read in phylogeny ##
model_data <- read.csv("../Data/Raw/ElabInovDataPGLS.csv", row.names=1)
phy <- read.tree("../Data/Raw/Prum_merge_hackett_stage2_1K_mcc.tree")

#############################################
## Runs all models while estimating lambda ##
#############################################
mod1 <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno +
                  order_tree_inno:elaboration_order +
                  order_tree_elab:elaboration_order +
                  order_tree_inno:innovation_order +
                  order_tree_elab:innovation_order
                , data=model_data, phy=phy, model="lambda")

mod2 <- phylolm(phylogeny ~ elaboration_order + 
                   innovation_order +
                   order_tree_elab +
                   order_tree_inno +
                   order_tree_inno:elaboration_order +
                   # order_tree_elab:elaboration_order +
                   order_tree_inno:innovation_order +
                   order_tree_elab:innovation_order
                 , data=model_data, phy=phy, model="lambda")

mod3 <- phylolm(phylogeny ~ elaboration_order + 
                   innovation_order +
                   order_tree_elab +
                   order_tree_inno +
                   order_tree_inno:elaboration_order +
                   order_tree_elab:elaboration_order +
                   order_tree_inno:innovation_order
                 # order_tree_elab:innovation_order
                 , data=model_data, phy=phy, model="lambda")

mod4 <- phylolm(phylogeny ~ elaboration_order + 
                   innovation_order +
                   order_tree_elab +
                   order_tree_inno +
                   # order_tree_inno:elaboration_order +
                   order_tree_elab:elaboration_order +
                   order_tree_inno:innovation_order +
                   order_tree_elab:innovation_order
                 , data=model_data, phy=phy, model="lambda")

mod5 <- phylolm(phylogeny ~ elaboration_order + 
                   innovation_order +
                   order_tree_elab +
                   order_tree_inno +
                   order_tree_inno:elaboration_order +
                   order_tree_elab:elaboration_order +
                   # order_tree_inno:innovation_order +
                   order_tree_elab:innovation_order
                 , data=model_data, phy=phy, model="lambda")

mod6 <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno
                , data=model_data, phy=phy, model="lambda")

mod7 <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order
                , data=model_data, phy=phy, model="lambda")

mod8 <- phylolm(phylogeny ~ order_tree_elab +
                  order_tree_inno
                , data=model_data, phy=phy, model="lambda")

mod9 <- phylolm(phylogeny ~ elaboration_order
                , data=model_data, phy=phy, model="lambda")

mod10 <- phylolm(phylogeny ~ 
                  innovation_order
                , data=model_data, phy=phy, model="lambda")

mod11 <- phylolm(phylogeny ~ 
                  order_tree_elab 
                , data=model_data, phy=phy, model="lambda")

mod12 <- phylolm(phylogeny ~ 
                  order_tree_inno
                , data=model_data, phy=phy, model="lambda")

#############################################
## Runs all models under Brownian motion   ##
#############################################
bm_model_list <- vector(mode="list", length=12)

names(bm_model_list) <- c("mod_bm1", "mod_bm2", "mod_bm3", "mod_bm4", "mod_bm5", "mod_bm6",
                          "mod_bm7", "mod_bm8", "mod_bm9", "mod_bm10", "mod_bm11", "mod_bm12")

bm_model_list[[1]] <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno +
                  order_tree_inno:elaboration_order +
                  order_tree_elab:elaboration_order +
                  order_tree_inno:innovation_order +
                  order_tree_elab:innovation_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[2]]  <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno +
                  order_tree_inno:elaboration_order +
                  # order_tree_elab:elaboration_order +
                  order_tree_inno:innovation_order +
                  order_tree_elab:innovation_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[3]]  <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno +
                  order_tree_inno:elaboration_order +
                  order_tree_elab:elaboration_order +
                  order_tree_inno:innovation_order
                # order_tree_elab:innovation_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[4]]  <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno +
                  # order_tree_inno:elaboration_order +
                  order_tree_elab:elaboration_order +
                  order_tree_inno:innovation_order +
                  order_tree_elab:innovation_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[5]]  <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno +
                  order_tree_inno:elaboration_order +
                  order_tree_elab:elaboration_order +
                  # order_tree_inno:innovation_order +
                  order_tree_elab:innovation_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[6]] <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order +
                  order_tree_elab +
                  order_tree_inno
                , data=model_data, phy=phy, model="BM")

bm_model_list[[7]] <- phylolm(phylogeny ~ elaboration_order + 
                  innovation_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[8]] <- phylolm(phylogeny ~ order_tree_elab +
                  order_tree_inno
                , data=model_data, phy=phy, model="BM")

bm_model_list[[9]] <- phylolm(phylogeny ~ elaboration_order
                , data=model_data, phy=phy, model="BM")

bm_model_list[[10]] <- phylolm(phylogeny ~ 
                   innovation_order
                 , data=model_data, phy=phy, model="BM")

bm_model_list[[11]] <- phylolm(phylogeny ~ 
                   order_tree_elab 
                 , data=model_data, phy=phy, model="BM")

bm_model_list[[12]] <- phylolm(phylogeny ~ 
                   order_tree_inno
                 , data=model_data, phy=phy, model="BM")

#### Collate Brownian motion model results ####
bm_results <- matrix(NA, nrow=12, ncol=11)
bm_results <- data.frame(bm_results)
colnames(bm_results) <- c("Model", "Especies", "Ispecies", "Eclade", "Iclade",
                          "Especies x Iclade", "Especies x Eclade",
                          "Ispecies x Iclade", "Ispecies x Eclade",
                          "delta AIC", "Adj. R2")

## Assign model number
bm_results[,1] <- 1:12

## Collate AIC scores
aic_scores <- rep(NA, 12)
for (i in 1:12) { aic_scores[i] <- bm_model_list[[i]]$aic}

for (i in 1:12){
  curr_model <- bm_model_list[[i]]
  params <- summary(curr_model)$coefficients

  if ("elaboration_order" %in% rownames(params)) {
    if (params["elaboration_order",4] <0.001) {try(bm_results[i,"Especies"] <- paste(round(params["elaboration_order",1], 3), "***", sep=""))}  
    if (params["elaboration_order",4] >0.001 & params["elaboration_order",4] <0.01) {try(bm_results[i,"Especies"] <- paste(round(params["elaboration_order",1], 3), "**", sep=""))}  
    if (params["elaboration_order",4] >0.01 & params["elaboration_order",4] <0.05) {try(bm_results[i,"Especies"] <- paste(round(params["elaboration_order",1], 3), "*", sep=""))}  
    if (params["elaboration_order",4] >0.05) {try(bm_results[i,"Especies"] <- round(params["elaboration_order",1], 3))}  
    }

  if ("innovation_order" %in% rownames(params)) {
    if (params["innovation_order",4] <0.001) {try(bm_results[i,"Ispecies"] <- paste(round(params["innovation_order",1], 3), "***", sep=""))}  
    if (params["innovation_order",4] >0.001 & params["innovation_order",4] <0.01) {try(bm_results[i,"Ispecies"] <- paste(round(params["innovation_order",1], 3), "**", sep=""))}  
    if (params["innovation_order",4] >0.01 & params["innovation_order",4] <0.05) {try(bm_results[i,"Ispecies"] <- paste(round(params["innovation_order",1], 3), "*", sep=""))}  
    if (params["innovation_order",4] >0.05) {try(bm_results[i,"Ispecies"] <- round(params["innovation_order",1], 3))} 
    }

  if ("order_tree_elab" %in% rownames(params)) {
    if (params["order_tree_elab",4] <0.001) {try(bm_results[i,"Eclade"] <- paste(round(params["order_tree_elab",1], 3), "***", sep=""))}  
    if (params["order_tree_elab",4] >0.001 & params["order_tree_elab",4] <0.01) {try(bm_results[i,"Eclade"] <- paste(round(params["order_tree_elab",1], 3), "**", sep=""))}  
    if (params["order_tree_elab",4] >0.01 & params["order_tree_elab",4] <0.05) {try(bm_results[i,"Eclade"] <- paste(round(params["order_tree_elab",1], 3), "*", sep=""))}  
    if (params["order_tree_elab",4] >0.05) {try(bm_results[i,"Eclade"] <- round(params["order_tree_elab",1], 3))} 
    }

  if ("order_tree_inno" %in% rownames(params)) {
    if (params["order_tree_inno",4] <0.001) {try(bm_results[i,"Iclade"] <- paste(round(params["order_tree_inno",1], 3), "***", sep=""))}  
    if (params["order_tree_inno",4] >0.001 & params["order_tree_inno",4] <0.01) {try(bm_results[i,"Iclade"] <- paste(round(params["order_tree_inno",1], 3), "**", sep=""))}  
    if (params["order_tree_inno",4] >0.01 & params["order_tree_inno",4] <0.05) {try(bm_results[i,"Iclade"] <- paste(round(params["order_tree_inno",1], 3), "*", sep=""))}  
    if (params["order_tree_inno",4] >0.05) {try(bm_results[i,"Iclade"] <- round(params["order_tree_inno",1], 3))} 
    }

  if ("elaboration_order:order_tree_inno" %in% rownames(params)) {
    if (params["elaboration_order:order_tree_inno",4] <0.001) {try(bm_results[i,"Especies x Iclade"] <- paste(round(params["elaboration_order:order_tree_inno",1], 3), "***", sep=""))}  
    if (params["elaboration_order:order_tree_inno",4] >0.001 & params["elaboration_order:order_tree_inno",4] <0.01) {try(bm_results[i,"Especies x Iclade"] <- paste(round(params["elaboration_order:order_tree_inno",1], 3), "**", sep=""))}  
    if (params["elaboration_order:order_tree_inno",4] >0.01 & params["elaboration_order:order_tree_inno",4] <0.05) {try(bm_results[i,"Especies x Iclade"] <- paste(round(params["elaboration_order:order_tree_inno",1], 3), "*", sep=""))}  
    if (params["elaboration_order:order_tree_inno",4] >0.05) {try(bm_results[i,"Especies x Iclade"] <- round(params["elaboration_order:order_tree_inno",1], 3))} 
    }

  if ("elaboration_order:order_tree_elab" %in% rownames(params)) {
    if (params["elaboration_order:order_tree_elab",4] <0.001) {try(bm_results[i,"Especies x Eclade"] <- paste(round(params["elaboration_order:order_tree_elab",1], 3), "***", sep=""))}  
    if (params["elaboration_order:order_tree_elab",4] >0.001 & params["elaboration_order:order_tree_elab",4] <0.01) {try(bm_results[i,"Especies x Eclade"] <- paste(round(params["elaboration_order:order_tree_elab",1], 3), "**", sep=""))}  
    if (params["elaboration_order:order_tree_elab",4] >0.01 & params["elaboration_order:order_tree_elab",4] <0.05) {try(bm_results[i,"Especies x Eclade"] <- paste(round(params["elaboration_order:order_tree_elab",1], 3), "*", sep=""))}  
    if (params["elaboration_order:order_tree_elab",4] >0.05) {try(bm_results[i,"Especies x Eclade"] <- round(params["elaboration_order:order_tree_elab",1], 3))} 
    }

  if ("innovation_order:order_tree_inno" %in% rownames(params)) {
    if (params["innovation_order:order_tree_inno",4] <0.001) {try(bm_results[i,"Ispecies x Iclade"] <- paste(round(params["innovation_order:order_tree_inno",1], 3), "***", sep=""))}  
    if (params["innovation_order:order_tree_inno",4] >0.001 & params["innovation_order:order_tree_inno",4] <0.01) {try(bm_results[i,"Ispecies x Iclade"] <- paste(round(params["innovation_order:order_tree_inno",1], 3), "**", sep=""))}  
    if (params["innovation_order:order_tree_inno",4] >0.01 & params["innovation_order:order_tree_inno",4] <0.05) {try(bm_results[i,"Ispecies x Iclade"] <- paste(round(params["innovation_order:order_tree_inno",1], 3), "*", sep=""))}  
    if (params["innovation_order:order_tree_inno",4] >0.05) {try(bm_results[i,"Ispecies x Iclade"] <- round(params["innovation_order:order_tree_inno",1], 3))} 
    }

  if ("innovation_order:order_tree_elab" %in% rownames(params)) {
    if (params["innovation_order:order_tree_elab",4] <0.001) {try(bm_results[i,"Ispecies x Eclade"] <- paste(round(params["innovation_order:order_tree_elab",1], 3), "***", sep=""))}  
    if (params["innovation_order:order_tree_elab",4] >0.001 & params["innovation_order:order_tree_elab",4] <0.01) {try(bm_results[i,"Ispecies x Eclade"] <- paste(round(params["innovation_order:order_tree_elab",1], 3), "**", sep=""))}  
    if (params["innovation_order:order_tree_elab",4] >0.01 & params["innovation_order:order_tree_elab",4] <0.05) {try(bm_results[i,"Ispecies x Eclade"] <- paste(round(params["innovation_order:order_tree_elab",1], 3), "*", sep=""))}  
    if (params["innovation_order:order_tree_elab",4] >0.05) {try(bm_results[i,"Ispecies x Eclade"] <- round(params["innovation_order:order_tree_elab",1], 3))} 
  }

  bm_results[i,"delta AIC"] <- round(min(aic_scores)-curr_model$aic, 3)
  bm_results[i,"Adj. R2"] <- round(summary(curr_model)$adj.r.squared, 3)

}

####################################################
## Runs all models under lowest lambda estimate   ##
####################################################

## Find minimum lambda value
all_lambda <- c(mod1$optpar, mod2$optpar, mod3$optpar, mod4$optpar, mod5$optpar,
                mod6$optpar, mod7$optpar, mod8$optpar, mod9$optpar, mod10$optpar,
                mod11$optpar, mod12$optpar)

fix_lambda <- min(all_lambda)

## Transform phylogeny with lowest lambda value
phy_lambda <- transformPhylo(phy, model="lambda", lambda=fix_lambda)

min_lambda_model_list <- vector(mode="list", length=12)
names(min_lambda_model_list) <- c("mod_min_lambda1", "mod_min_lambda2", "mod_min_lambda3", "mod_min_lambda4", "mod_min_lambda5", "mod_min_lambda6",
  "mod_min_lambda7", "mod_min_lambda8", "mod_min_lambda9", "mod_min_lambda10", "mod_min_lambda11", "mod_min_lambda12")

## Run the models
min_lambda_model_list[[1]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order +
                     order_tree_elab +
                     order_tree_inno +
                     order_tree_inno:elaboration_order +
                     order_tree_elab:elaboration_order +
                     order_tree_inno:innovation_order +
                     order_tree_elab:innovation_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[2]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order +
                     order_tree_elab +
                     order_tree_inno +
                     order_tree_inno:elaboration_order +
                     # order_tree_elab:elaboration_order +
                     order_tree_inno:innovation_order +
                     order_tree_elab:innovation_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[3]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order +
                     order_tree_elab +
                     order_tree_inno +
                     order_tree_inno:elaboration_order +
                     order_tree_elab:elaboration_order +
                     order_tree_inno:innovation_order
                   # order_tree_elab:innovation_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[4]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order +
                     order_tree_elab +
                     order_tree_inno +
                     # order_tree_inno:elaboration_order +
                     order_tree_elab:elaboration_order +
                     order_tree_inno:innovation_order +
                     order_tree_elab:innovation_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[5]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order +
                     order_tree_elab +
                     order_tree_inno +
                     order_tree_inno:elaboration_order +
                     order_tree_elab:elaboration_order +
                     # order_tree_inno:innovation_order +
                     order_tree_elab:innovation_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[6]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order +
                     order_tree_elab +
                     order_tree_inno
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[7]] <- phylolm(phylogeny ~ elaboration_order + 
                     innovation_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[8]] <- phylolm(phylogeny ~ order_tree_elab +
                     order_tree_inno
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[9]] <- phylolm(phylogeny ~ elaboration_order
                   , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[10]] <- phylolm(phylogeny ~ 
                      innovation_order
                    , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[11]] <- phylolm(phylogeny ~ 
                      order_tree_elab 
                    , data=model_data, phy=phy_lambda, model="BM")

min_lambda_model_list[[12]] <- phylolm(phylogeny ~ 
                      order_tree_inno
                    , data=model_data, phy=phy_lambda, model="BM")

#### Collate Brownian motion model results ####
min_lambda_results <- matrix(NA, nrow=12, ncol=11)
min_lambda_results <- data.frame(min_lambda_results)
colnames(min_lambda_results) <- c("Model", "Especies", "Ispecies", "Eclade", "Iclade",
                          "Especies x Iclade", "Especies x Eclade",
                          "Ispecies x Iclade", "Ispecies x Eclade",
                          "delta AIC", "Adj. R2")

## Assign model number
min_lambda_results[,1] <- 1:12

## Collate AIC scores
aic_scores <- rep(NA, 12)
for (i in 1:12) { aic_scores[i] <- min_lambda_model_list[[i]]$aic}

for (i in 1:12){
  curr_model <- min_lambda_model_list[[i]]
  params <- summary(curr_model)$coefficients

  if ("elaboration_order" %in% rownames(params)) {
    if (params["elaboration_order",4] <0.001) {try(min_lambda_results[i,"Especies"] <- paste(round(params["elaboration_order",1], 3), "***", sep=""))}  
    if (params["elaboration_order",4] >0.001 & params["elaboration_order",4] <0.01) {try(min_lambda_results[i,"Especies"] <- paste(round(params["elaboration_order",1], 3), "**", sep=""))}  
    if (params["elaboration_order",4] >0.01 & params["elaboration_order",4] <0.05) {try(min_lambda_results[i,"Especies"] <- paste(round(params["elaboration_order",1], 3), "*", sep=""))}  
    if (params["elaboration_order",4] >0.05) {try(min_lambda_results[i,"Especies"] <- round(params["elaboration_order",1], 3))}  
  }

  if ("innovation_order" %in% rownames(params)) {
    if (params["innovation_order",4] <0.001) {try(min_lambda_results[i,"Ispecies"] <- paste(round(params["innovation_order",1], 3), "***", sep=""))}  
    if (params["innovation_order",4] >0.001 & params["innovation_order",4] <0.01) {try(min_lambda_results[i,"Ispecies"] <- paste(round(params["innovation_order",1], 3), "**", sep=""))}  
    if (params["innovation_order",4] >0.01 & params["innovation_order",4] <0.05) {try(min_lambda_results[i,"Ispecies"] <- paste(round(params["innovation_order",1], 3), "*", sep=""))}  
    if (params["innovation_order",4] >0.05) {try(min_lambda_results[i,"Ispecies"] <- round(params["innovation_order",1], 3))} 
  }

  if ("order_tree_elab" %in% rownames(params)) {
    if (params["order_tree_elab",4] <0.001) {try(min_lambda_results[i,"Eclade"] <- paste(round(params["order_tree_elab",1], 3), "***", sep=""))}  
    if (params["order_tree_elab",4] >0.001 & params["order_tree_elab",4] <0.01) {try(min_lambda_results[i,"Eclade"] <- paste(round(params["order_tree_elab",1], 3), "**", sep=""))}  
    if (params["order_tree_elab",4] >0.01 & params["order_tree_elab",4] <0.05) {try(min_lambda_results[i,"Eclade"] <- paste(round(params["order_tree_elab",1], 3), "*", sep=""))}  
    if (params["order_tree_elab",4] >0.05) {try(min_lambda_results[i,"Eclade"] <- round(params["order_tree_elab",1], 3))} 
  }

  if ("order_tree_inno" %in% rownames(params)) {
    if (params["order_tree_inno",4] <0.001) {try(min_lambda_results[i,"Iclade"] <- paste(round(params["order_tree_inno",1], 3), "***", sep=""))}  
    if (params["order_tree_inno",4] >0.001 & params["order_tree_inno",4] <0.01) {try(min_lambda_results[i,"Iclade"] <- paste(round(params["order_tree_inno",1], 3), "**", sep=""))}  
    if (params["order_tree_inno",4] >0.01 & params["order_tree_inno",4] <0.05) {try(min_lambda_results[i,"Iclade"] <- paste(round(params["order_tree_inno",1], 3), "*", sep=""))}  
    if (params["order_tree_inno",4] >0.05) {try(min_lambda_results[i,"Iclade"] <- round(params["order_tree_inno",1], 3))} 
  }

  if ("elaboration_order:order_tree_inno" %in% rownames(params)) {
    if (params["elaboration_order:order_tree_inno",4] <0.001) {try(min_lambda_results[i,"Especies x Iclade"] <- paste(round(params["elaboration_order:order_tree_inno",1], 3), "***", sep=""))}  
    if (params["elaboration_order:order_tree_inno",4] >0.001 & params["elaboration_order:order_tree_inno",4] <0.01) {try(min_lambda_results[i,"Especies x Iclade"] <- paste(round(params["elaboration_order:order_tree_inno",1], 3), "**", sep=""))}  
    if (params["elaboration_order:order_tree_inno",4] >0.01 & params["elaboration_order:order_tree_inno",4] <0.05) {try(min_lambda_results[i,"Especies x Iclade"] <- paste(round(params["elaboration_order:order_tree_inno",1], 3), "*", sep=""))}  
    if (params["elaboration_order:order_tree_inno",4] >0.05) {try(min_lambda_results[i,"Especies x Iclade"] <- round(params["elaboration_order:order_tree_inno",1], 3))} 
  }

  if ("elaboration_order:order_tree_elab" %in% rownames(params)) {
    if (params["elaboration_order:order_tree_elab",4] <0.001) {try(min_lambda_results[i,"Especies x Eclade"] <- paste(round(params["elaboration_order:order_tree_elab",1], 3), "***", sep=""))}  
    if (params["elaboration_order:order_tree_elab",4] >0.001 & params["elaboration_order:order_tree_elab",4] <0.01) {try(min_lambda_results[i,"Especies x Eclade"] <- paste(round(params["elaboration_order:order_tree_elab",1], 3), "**", sep=""))}  
    if (params["elaboration_order:order_tree_elab",4] >0.01 & params["elaboration_order:order_tree_elab",4] <0.05) {try(min_lambda_results[i,"Especies x Eclade"] <- paste(round(params["elaboration_order:order_tree_elab",1], 3), "*", sep=""))}  
    if (params["elaboration_order:order_tree_elab",4] >0.05) {try(min_lambda_results[i,"Especies x Eclade"] <- round(params["elaboration_order:order_tree_elab",1], 3))} 
  }

  if ("innovation_order:order_tree_inno" %in% rownames(params)) {
    if (params["innovation_order:order_tree_inno",4] <0.001) {try(min_lambda_results[i,"Ispecies x Iclade"] <- paste(round(params["innovation_order:order_tree_inno",1], 3), "***", sep=""))}  
    if (params["innovation_order:order_tree_inno",4] >0.001 & params["innovation_order:order_tree_inno",4] <0.01) {try(min_lambda_results[i,"Ispecies x Iclade"] <- paste(round(params["innovation_order:order_tree_inno",1], 3), "**", sep=""))}  
    if (params["innovation_order:order_tree_inno",4] >0.01 & params["innovation_order:order_tree_inno",4] <0.05) {try(min_lambda_results[i,"Ispecies x Iclade"] <- paste(round(params["innovation_order:order_tree_inno",1], 3), "*", sep=""))}  
    if (params["innovation_order:order_tree_inno",4] >0.05) {try(min_lambda_results[i,"Ispecies x Iclade"] <- round(params["innovation_order:order_tree_inno",1], 3))} 
  }

  if ("innovation_order:order_tree_elab" %in% rownames(params)) {
    if (params["innovation_order:order_tree_elab",4] <0.001) {try(min_lambda_results[i,"Ispecies x Eclade"] <- paste(round(params["innovation_order:order_tree_elab",1], 3), "***", sep=""))}  
    if (params["innovation_order:order_tree_elab",4] >0.001 & params["innovation_order:order_tree_elab",4] <0.01) {try(min_lambda_results[i,"Ispecies x Eclade"] <- paste(round(params["innovation_order:order_tree_elab",1], 3), "**", sep=""))}  
    if (params["innovation_order:order_tree_elab",4] >0.01 & params["innovation_order:order_tree_elab",4] <0.05) {try(min_lambda_results[i,"Ispecies x Eclade"] <- paste(round(params["innovation_order:order_tree_elab",1], 3), "*", sep=""))}  
    if (params["innovation_order:order_tree_elab",4] >0.05) {try(min_lambda_results[i,"Ispecies x Eclade"] <- round(params["innovation_order:order_tree_elab",1], 3))} 
  }

  min_lambda_results[i,"delta AIC"] <- round(min(aic_scores)-curr_model$aic, 3)
  min_lambda_results[i,"Adj. R2"] <- round(summary(curr_model)$adj.r.squared, 3)

}

##### Latexify tables

min_lambda_results_latex <- xtable(min_lambda_results)
bm_results_latex <- xtable(bm_results)
```

### 2023-06-19 Manuscript/main.tex
fixed typo in Zoe's name
Update main.tex

@@
 
-\bibliographystyle{naturemag}
+\bibliographystyle{Science}
 \bibliography{references}

### 2023-08-21 Manuscript/main.tex
Update main.tex

@@
 $^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
+$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\
+$^{+}$Contributed equally.\\}
 
@@
 The original descriptions of microevolutionary lines of least resistance explicitly refer to genetic constraints.
-However, at the macroevolutionary scale the major axis of phenotypic variation is an emergent property of broader set of genetic and developmental constraints interacting with selection on a moving adaptive landscape % GT: Masahito suggested a whole paragraph on adaptive landscapes and Simpon-Lande-Arnold ideas.
-I'm not sure that will help but hopefully the expanded intro overall works better.
-Possibly just needs another sentence or two here but it would break the flow.
-more citations of relevant papers here (the first 7 or so papers from Gustavo's paper would be relevant.
-\cite{BurinWhales,jones2004evolution}. % GT: More references needed? I'm not sure where the idea originates.
+However, at the macroevolutionary scale the major axis of phenotypic variation is an emergent property of broader set of genetic and developmental constraints interacting with selection on a moving adaptive landscape \cite{BurinWhales,jones2004evolution}.
 At this scale the major axis of phenotypic variation among species \cite{marroig2005size, fasanelli2022allometry} can be estimated from a matrix of divergences in species mean phenotypic traits (e.g.
@@
 In contrast, the \textbf{R} matrix of \cite{Houle2017} describes the rate of among-species divergence in species traits and explicitly incorporates phylogeny (also see the \textbf{B} matrix of \cite{Machado2020}) %@L_Rmatrix.
-We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}. %TG: TODO: change throughout (that's the VCV matrix now)
+We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}.
 The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales.
@@
 \caption{\scriptsize{Relationships between key figures in the main text.
-A) represents the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
-The gray ellipse represents the overall phylogenetic major axis of beak variation in birds.
+A) the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
+The gray ellipse represents the overall phylogenetic major axis of variation.
 Species aligned or away from this axis are respectively elaborators or innovators relative to the class Aves.
-The red ellipse and axis represents the phylogenetic major axis of beak variation in Tinamiformes only.
-Species aligned or away from this axis are respectively elaborators and innovators relative to the Tinamiformes order.
-B) In Figure 2, we measured elaboration and innovation at the macroevolutionary scale , i.e.
-the position of each species in terms of $elaboration_{species}$ and $innovation_{species}$ relative to their order's phylogenetic major axis of beak variation.
-A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation(i.
-e.
+The red ellipse and axis represents the phylogenetic major axis of variation in Tinamiformes only.
+Species aligned or away from this axis are elaborators/innovators relative to the Tinamiformes order.
+B) Figure 2 shows elaboration and innovation at the macroevolutionary scale: $elaboration_{species}$ and $innovation_{species}$.
+A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation (i.e.
 their position on that axis) whereas their $innovation_{species}$ score is their rejection from it (i.e.
@@
 The colours on the tips of the phylogeny correspond to the median $elaboration_{species}$ (orange gradient) and $innovation_{species}$ (blue gradient).
-C) Figure 4, we measured elaboration and innovation at the megaevolutionary scale, i.e. $elaboration_{clade}$ and $innovation_{clade}$.
-The ellipses are scaled and centered and represent the clade's phylogenetic major axis of beak variation relative to the overall phylogenetic major axis of beak variation with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
-A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis of beak variation onto the overall phylogenetic major axis of beak variation and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
+C) Figure 4 shows elaboration and innovation at the megaevolutionary scale: $elaboration_{clade}$ and $innovation_{clade}$.
+The ellipses are scaled and centered and represent the clade's phylogenetic major axis relative to the overall phylogenetic major axis with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
+A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis onto the overall phylogenetic major axis and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
 The distribution elaboration and innovation scores are taken from the projection of the 4000 pairs of evolutionary rate matrices with each pair being the focal level, e.g.
@@
 Dots represent the median orthogonality of each clade and the lines their 95\% confidence intervals (CI).
-No super-order's phylogenetic major axis of beak variation is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal ($>$  $45^\circ$) to the class-wide phylogenetic major axis of beak variation.
-These results are to be contrasted with the orientation of each order relative to their super-order where on average (median) only 14/27 are at least half orthogonal to the orientation of their super-orders’s phylogenetic major axis of beak variation.
-We also indicate the number of species (n) and the standard deviation (sd) of the orientation of their phylogenetic major axis of beak variation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
-Note that noclades are clearly parallel to their parent’s or parent’s parent’s ellipses (i.e.
+No super-order's phylogenetic major axis is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal ($>$  $45^\circ$) to the class-wide phylogenetic major axis.
+The orientation of each order relative to their super-order where on average (median) at least half orthogonal to the orientation of their super-orders’s phylogenetic major axis of beak variation for 14/27 of the orders.
+We also indicate the number of species (n) and the standard deviation (sd; expressed in degrees) of the orientation of their phylogenetic major axis over the 4000 variance-covariance posteriors).
+Note that no clades are clearly parallel to their parent’s or parent’s parent’s ellipses (i.e.
 no 95\% CI includes 0).
-For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the class-wide phylogenetic major axis of beak variation relative to their sample size and sd.
-The stars represent the posterior probability of the clade's orientation being different from the comparison clade (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
+For each clade we measured the posterior probability of each clade's orientation being different from their parent's clade or the class-wide phylogenetic major axis of beak variation relative to their sample size and sd (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; .
+= pp $> 0.8$).
 Only 1/6 super-orders, 8/27 orders relative to class-wide phylogeny and 6/27 relative to their super-orders have a pp $> 0.99$.
-This is the result of the variation in sample size (n) and standard variation (sd) among clades.
 A companion plot showing further nested structure within the Passeriformes is shown in Fig. \ref{fig_orthogonality_passeriformes}.}}
@@
 
-\section*{Reproducibility and repeatability}
+\section*{Data and Materials Availability}
 The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.

### 2023-08-21 Manuscript/reponse_letter.tex
Update reponse_letter.tex

@@
     \item For the analyses on elaboration and innovation at the macro scale ($elaboration_{species}$ and $innovation_{species}$), we now only focus on the projection of species onto their order’s major axes;
-    \item We added two new figures and a table in the main text: the cheat sheet figure (previously in the supplementaries, now in the introduction), the correlation figure and a table from the new PGLS results;
+    \item We added two new figures and a table in the main text: the cheat sheet figure (previously in the supplementaries, now in the introduction), the correlation figure and a table from the new PGLS results and revamped all the previous figures.
     \item We have added several new analyses to the supplementary materials including: correlations between elaboration/innovation, elaboration and innovation through time and random skewers analysis;
@@
 
-We fully describe each change below.
+\noindent We fully describe each change below.
+
 \bigskip
-Best regards,
+
+\noindent Best regards,
+
 \bigskip
-Thomas Guillerme, on behalf of all co-authors.
+
+\noindent Thomas Guillerme, on behalf of all co-authors.
 
@@
 
+We now reflect that in the supplementary materials section 4.5:
+
+\noindent\textit{In \cite{hansen2008measuring}, the inverse of \textbf{G} is defined as ``the fraction of the genetic variation that is independent of potentially constraining characters'' which in our case will be harder to define.
+Although it would technically be possible to use the inverse of the \textbf{R} matrix used here ($VCV_{clade}$), it is unclear what it would represent biologically.}
+
 \textcolor{blue}{Second, I suggest authors to use one term for innovation/elaboration across the entire phylogeny and another term for those at the clade-specific levels.

### 2023-08-21 Manuscript/supplementaries.tex
Update supplementaries.tex

@@
 %
-%   Glossary   % NC: I think this is not really needed now the text is clearer, also some of the definitions are not 100% correct...
-%
-%---------------------------------------------
-
-%\begin{landscape}
-%\section{Supplementary material 1: glossary}
-
-%We define three evolutionary scales based on the focal elements as follow:
-%\begin{itemize}
-%   \item \textbf{micro}evolutionary scale: the focal elements are \textbf{individual specimen};
- %  \item \textbf{macro}evolutionary scale: the focal elements are \textbf{species} (i.e. groups of individuals);
-  % \item \textbf{mega}evolutionary scale: the focal elements are \textbf{clades} (i.e. groups of species).
-%\end{itemize}
-
-%\begin{table}[ht]
-%\scriptsize
-%\caption{Glossary of some specific terms in this manuscript}
-%\begin{tabular}{p{0.2\textwidth}p{0.3\textwidth}p{0.3\textwidth}p{0.2\textwidth}p{0.2\textwidth}}
-%  \hline
-% Term in this paper         & definition in this paper & mathematical definition & in macroevolution & in microevolution\\
-%   \hline
-% evolutionary rate matrix &  matrix of rates of among-species divergence in species traits that explicitly incorporates phylogeny  &  & (\textbf{R} matrix of \cite{Houle2017}; \textbf{B} matrix of \cite{Machado2020}) & \\
-% major axis of variation    & the major evolutionary ``direction'' of trait co-evolution & first eigenvector of the evolutionary rate matrix  & allometry & line of genetic least resistance \\
-% clade                      & a monophyletic group of \textbf{species} & a subset of \textbf{shapespace} & Birds & a group of species or a population\\
-% species                    & the lowest evolutionary units in our analysis & a row in the \textbf{shapespace} & \textit{Pauxi pauxi} & an individual\\
-% shapespace                 & all the possible morphological combination (observed or not) & an ordinate and orthonormal matrix & the beak shapespace & a gene space \\
-% novelty                    & morphological variation (either \textbf{innovation} or \textbf{elaboration}) & position of a \textbf{species} or a \textbf{clade} relative to the \textbf{major axis of variation} & & \\
-% elaboration                & morphological variation along the \textbf{major axis of variation} & the algebraic projection on the eigenvector & & \\
-% innovation                 & morphological variation away from the \textbf{major axis of variation} & the algebraic rejection from the eigenvector & & \\
-% variance-covariance matrix & the multidimensional relation between traits & the variation of the dimension within the \textbf{shapespace} taking into account the distance matrix between the observations (the \textbf{phylogeny} or the \textbf{clades}) & & \\
-% nested phylogenetic level  & the different \textbf{clades} used in this study (the class-wide phylogeny $>$ super-orders $>$ orders) & different nested \textbf{variance-covariance matrices} & & \\
-
-% \end{tabular}
-% \label{tab_glossary}
-% \end{table}
-% \end{landscape}
-
-%---------------------------------------------
-%
 %   additional methods 
@@
 
- % GT Figure legend needs explanation of whole_space and no_calde
 \begin{figure}[H]
@@
 Note that not all super-orders have their own variance explained matching the variance explained by the entire ordination.
-For example, Strisores only need the first three dimensions to explain at least 95\% of their variance within the shapespace whilst Aequornithes need all eight dimensions to achieve at least the same amount of variance explained.}
+For example, Strisores only need the first three dimensions to explain at least 95\% of their variance within the shapespace whilst Aequornithes need all eight dimensions to achieve at least the same amount of variance explained.
+$no clade$ is the cumulative variance for species not present in the other displayed clades and $whole\_space$ is the cumulative for all species.}
 \label{Fig:axes_variance}
@@
 We can define the major axis from the variance-covariance matrices and then project and reject each element of interest in the space onto this axis.
-The following steps are generalised to $n$ dimensions and detailed below (as well as the algorithm used in \texttt{dispRity} \cite{dispRity} to perform the transformations):
+The following steps are generalised to $n$ dimensions and detailed below (as well as the algorithm used in \texttt{dispRity} [64]%\cite{dispRity}
+to perform the transformations):
 
@@
 
-Finally, we can centre the matrix $M_{n}$ on the estimated solution of each GLMM (\texttt{model\$Sol}, in \texttt{MCMCglmm} \cite{MCMCglmm}) corresponding to the estimation of the position of the variance-covariance matrix in the trait-space.
+Finally, we can centre the matrix $M_{n}$ on the estimated solution of each GLMM (\texttt{model\$Sol}, in \texttt{MCMCglmm} [48]% \cite{MCMCglmm}
+) corresponding to the estimation of the position of the variance-covariance matrix in the trait-space.
 $M_{n}$ then contains all the major axes of the 0.95 hyper-ellipse fitting the variance-covariance matrix.
@@
 
-In practice we followed \href{https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions}{this procedure} and applied a modification of \href{https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385}{this implementation} (see \cite{aguilera2004} for the formal generalisation of this algorithm in $n$ dimensions) using the following algorithm implemented in \texttt{dispRity::projections}:
+In practice we followed \href{https://math.stackexchange.com/questions/598750/finding-the-rotation-matrix-in-n-dimensions}{this procedure} and applied a modification of \href{https://stackoverflow.com/questions/42520301/find-rotation-matrix-of-one-vector-to-another-using-r/42542385#42542385}{this implementation} (see [65]% \cite{aguilera2004}
+ for the formal generalisation of this algorithm in $n$ dimensions) using the following algorithm implemented in \texttt{dispRity::projections}:
 
@@
 
-\cite{vanvalen1974,watanabe2022} propose to describe a variance-covariance matrix in terms of its eigen dispersion $V_{rel}$ as:
+[44, 66] %\cite{vanvalen1974,watanabe2022}
+propose to describe a variance-covariance matrix in terms of its eigen dispersion $V_{rel}$ as:
 
@@
    \includegraphics[width=0.9\textwidth]{Figures/elaboration_innovation_dispersion.pdf}
-\caption{Relation between relative eigen dispersion \cite{vanvalen1974,watanabe2022} and $\text{elaboration}_{\text{clade}_{c}}$ and $\text{innovation}_{\text{clade}_{c}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
+\caption{Relation between relative eigen dispersion [44, 66] %\cite{vanvalen1974,watanabe2022}
+and $\text{elaboration}_{\text{clade}_{c}}$ and $\text{innovation}_{\text{clade}_{c}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
 \label{Fig:eigen_dispersion}
@@
 
-\cite{hansen2008measuring} propose to measure the evolvability and respondability of an individual as their ``response to selection in $n$ traits''.
+[42] %\cite{hansen2008measuring}
+propose to measure the evolvability and respondability of an individual as their ``response to selection in $n$ traits''.
 Here, some equivalent can be drawn (albeit cautiously) between the response of an individual or a group (population) to selection pressure and a species.
 Arguably, a species is the end product of this (and many other) selection pressures on a group of individuals.
-If so, then, we can relate $\text{evolvability}$ and $\text{respondability}$ \cite{hansen2008measuring} for a $species_{x}$ to it's $\text{elaboration}_{\text{species}_{x}}$ and $\text{innovation}_{\text{species}_{x}}$ as follows:
+If so, then, we can relate $\text{evolvability}$ and $\text{respondability}$ [42] %\cite{hansen2008measuring}
+for a $species_{x}$ to it's $\text{elaboration}_{\text{species}_{x}}$ and $\text{innovation}_{\text{species}_{x}}$ as follows:
 
@@
    \includegraphics[width=0.9\textwidth]{Figures/respondability.pdf}
-\caption{Relation between respondability \cite{hansen2008measuring} and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
+\caption{Relation between respondability [42] %\cite{hansen2008measuring}
+and $\text{elaboration}_{\text{species}}$ and $\text{innovation}_{\text{species}}$ on the \texttt{charadriiformes} example dataset from the \texttt{dispRity} package.}
 \label{Fig:respondability}
@@
 
-Conditional evolvability however will be more complex to determine since it is defined as the relation between evolvability and autonomy which is specific defined in \cite{hansen2008measuring} as the inverse of the \textbf{G} matrix.
-In \cite{hansen2008measuring}, the inverse of \textbf{G} is defined as ``the fraction of the genetic variation that is independent of potentially constraining characters'' which in our case will be harder to define.
+Conditional evolvability however will be more complex to determine since it is defined as the relation between evolvability and autonomy which is specific defined in [42] %\cite{hansen2008measuring}
+ as the inverse of the \textbf{G} matrix.
+In [42] %\cite{hansen2008measuring}
+, the inverse of \textbf{G} is defined as ``the fraction of the genetic variation that is independent of potentially constraining characters'' which in our case will be harder to define.
 Although it would technically be possible to use the inverse of the \textbf{R} matrix used here ($VCV_{clade}$), it is unclear what it would represent biologically.
@@
 
-\bibliographystyle{naturemag}
-\bibliography{references}
+% \bibliographystyle{naturemag}
+% \bibliography{references}
 

### 2023-09-11 Functions/Cluster/login.sh
cleaned functions

@@ -1,1 +1,0 @@
-ssh -X bo1tgu@sharc.shef.ac.uk

### 2023-09-11 Data/Cluster/shapespace_passeriformes/make.minichains.jobs.sh
cleaning up commits

@@ -1,0 +1,91 @@
+#!/bin/sh
+##########################
+# Script for initialising the minichains
+##########################
+#SYNTAX: sh make.minichains.jobs.sh -c <chain_name> -r <replicates> -p <path (optional)>
+#with:
+#-c <chain_name> the name of the chain
+#-r <number_of_replicates> the number of replicates
+#-p <path> optional, the path to the chain
+##########################
+#guillert(at)tcd.ie - 2021/10/05
+##########################
+
+#INPUT
+## Input values
+while [[ $# -gt 1 ]]
+do
+key="$1"
+
+case $key in
+    -c|--chain)
+        CHAINNAME="$2"
+        shift
+        ;;
+    -r|--replicates)
+        REPLICATES="$2"
+        shift
+        ;;        
+    -p|--path)
+        LOCALPATH="$2"
+        ;;
+        *)
+        ;;
+esac
+shift
+done
+
+## Creating the batch job template
+cp job_template.job ${LOCALPATH}${CHAINNAME}_template.job
+echo "" >> ${LOCALPATH}${CHAINNAME}_template.job
+echo "## Run the script" >> ${LOCALPATH}${CHAINNAME}_template.job
+echo "Rscript ${LOCALPATH}${CHAINNAME}_<REPLICATE>.R" >> ${LOCALPATH}${CHAINNAME}_template.job
+
+## Creating the Rscript template
+echo ".libPaths(\"~/homelib\")" > ${LOCALPATH}${CHAINNAME}_template.R
+echo "library(MCMCglmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "library(mcmcmcglmmm)" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "load(\"${LOCALPATH}${CHAINNAME}.mini.chains\")" >> ${LOCALPATH}${CHAINNAME}_template.R
+echo "model <- run.mini.chains(${CHAINNAME}, replicates = 2, record.tree = TRUE)" >> ${LOCALPATH}${CHAINNAME}_template.R
+# echo "save(model, file = \"${LOCALPATH}${CHAINNAME}_<REPLICATE>.rda\")" >> ${LOCALPATH}${CHAINNAME}_template.R
[…]

### 2023-09-14 Manuscript/main.tex
Minor updates (should be in the proofs stage)

@@
 Indeed, the major axis of phenotypic change among species beak shapes at megaevolutionary scales is an emergent property of innovation across clades.
-Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up new avenues for evolution to explore.
+Our analyses suggest that the reorientation of phenotypes via innovation is a ubiquitous route for divergence that can arise through gradual change alone, opening up further avenues for evolution to explore.
 %@L_reworded_abstract
@@
 
-By studying the phylogenetic line of least resistance and related properties of the evolutionary rate matrix \cite{caetano2017ratematrix, caetano2018estimating, slater2019hierarchy} at different scales we aim to gain new insights into the macro- and mega-evolutionary trajectories that lead to the diversity of beak shapes among extant bird species and clades.
+By studying the phylogenetic line of least resistance and related properties of the evolutionary rate matrix \cite{caetano2017ratematrix, caetano2018estimating, slater2019hierarchy} at different scales we aim to gain further insights into the macro- and mega-evolutionary trajectories that lead to the diversity of beak shapes among extant bird species and clades.
 While there is no strict definition that separates macro- and mega-evolution, we define the macroevolutionary scale as species divergence within clades and the megaevolutionary scale as divergence of clades within larger clades.
@@
 \section*{Data and Materials Availability}
-The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
-The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
+All data needed to evaluate the conclusions in the paper are present in the paper, the Supplementary Materials and/or are deposited in Figshare \url{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355} \cite{fighsaredata}.
+The code is available on Zenodo \url{https://zenodo.org/record/6984934} \cite{githubrepo} and additionally available on GitHub \url{https://github.com/TGuillerme/elaboration_exploration_bird_beaks/tree/v1.0}.
 

### 2024-02-01 Analysis/update_procrustes_and_PCA.Rmd
Create update_procrustes_and_PCA.Rmd

---
title: "Updates a Procrustes and a PCA"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

Trying to update a Procrustes superimposition by adding new species to it (i.e. run the Procrustes and then add a new species following the same deformation).
And doing the same for a PCA (i.e. adding some new data to a previously calculated eigen transformation).

```{r}
library(Morpho)
require(landvR) # devtools::install_github("TGuillerme/landvR")
## Get some 3D data from Morpho
data(boneData)

## Run a procrustes and PCA
proc <- procSym(boneLM)
## Create a new specimens: the average, the median and the minimum, of all bones)
new_spec <- array(data = c(apply(boneLM, c(1,2), mean),
                           apply(boneLM, c(1,2), median),
                           unlist(landvR::select.procrustes(proc$rotated, selector = min))),
                           dim = c(10, 3, 3))
## Add them to the procrustes + PCA
new_proc <- align2procSym(proc, new_spec)
dimnames(new_proc)[[3]] <- c("mean", "median", "min")
## Transforming the procrustes arrays into matrices
proc_matrix <- landvR::array.to(proc$rotated, to = "matrix")
new_proc_matrix <- landvR::array.to(new_proc, to = "matrix")

## Run a PCA on the original data
pca <- prcomp(proc_matrix)
plot(pca$x[,c(1,2)])
## Note that this is equal to the pca from Morpho but rotated (*-1)
# plot(proc$PCscores[, c(1,2)]*-1)

## Run the PCA with the added procrustes
new_pca <- rbind(pca$x, predict(pca, newdata = new_proc_matrix))
points(new_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5)
points(new_pca[c(81:83), c(1,2)], pch = 19, col = "red")
legend("topright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
```

### 2024-12-16 Analysis/17-update_procrustes_and_PCA.Rmd
updated projections of chaffinches

---
title: "Updates a Procrustes and a PCA"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

In this script we will be looking at how to update both a Procrustes superimposition (GPA) and a PCA by projecting new specimens in it.
The goal is to use the GPA and PCA from a previous dataset (published with Guillerme et al 2023 - Science Advances) and to project new specimens onto it.

# The pipeline (testing)

Here is an example on how to do it using the `Morpho` package for the base GPA and PCA:

```{r}
library(Morpho)

## Get some 3D data from Morpho
data(boneData)
original_landmarks <- boneLM

## Run the GPA for the original data
proc_orig <- Morpho::procSym(original_landmarks)

## Run the PCA for the original data
proc_matrix_orig <- landvR::array.to(proc_orig$rotated, to = "matrix")
pca_orig <- prcomp(proc_matrix_orig)

## Plot the results
plot(pca_orig$x[,c(1,2)])
```

And now let's generate some extra data out of the pipeline.
Here we're just creating a mean, median and minimum landmark set using the mean, median and minimum landmark coordinates values.

```{r}
## Create a new specimens: the average, the median and the minimum, of all bones)
require(landvR) # devtools::install_github("TGuillerme/landvR")
new_spec <- array(data = c(apply(original_landmarks, c(1,2), mean),
                           apply(original_landmarks, c(1,2), median),
                           unlist(landvR::select.procrustes(proc_orig$rotated, selector = min))),
                           dim = c(10, 3, 3))
dimnames(new_spec)[[3]] <- paste0("new_", c("mean", "median", "min"))

## Here is what the landmark positions look like in 3D
op <- par(mfrow = c(2,2))
plot(original_landmarks[, c(1,2), 1], main = "First specimen\n(reference from real dataset)")
plot(new_spec[, c(1,2), 1], main = "New specimen 1\n(average)")
plot(new_spec[, c(1,2), 2], main = "New specimen 2\n(median)")
plot(new_spec[, c(1,2), 3], main = "New specimen 3\n(weird)")
par(op)
```

We can then add these specimens to the GPA and PCA.

```{r}
## Add them to the procrustes + PCA
new_proc <- Morpho::align2procSym(proc_orig, new_spec)
dimnames(new_proc)[[3]] <- c("mean", "median", "min")

## Transforming the procrustes arrays into matrices
new_proc_matrix <- landvR::array.to(new_proc, to = "matrix")

## Run the PCA with the added procrustes
new_pca <- rbind(pca_orig$x, predict(pca_orig, newdata = new_proc_matrix))

## Plot all the results
plot(pca_orig$x[,c(1,2)])
points(new_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5)
points(new_pca[c(81:83), c(1,2)], pch = 19, col = "red")
legend("topright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
```

Seems to work quiet well!
All the positions from previous specimens are respected.
And the two new median and average specimens plot in the centre of the traitspace (as expected) with the weird one plotting on some edge of the traitspace.

The function that does it automatically is now implemented in `landvR`.

# Mapping the chaffinches on the whole traitspace

```{r}
## Loading the entire bird GPA
library(Morpho)
whole_gpa <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")

## Reading in the chaffinches
land_pairs <- read.table("../Data/Raw/Beak_data/landpairs.txt")

## Read all the chaffinches
path <- "../Data/Raw/Beak_data/Chaffinch_data/"
individuals <- list.files(path)
all_individuals <- list()
while(length(individuals) != 0) {
    file <- paste0(path, individuals[1], "/", list.files(paste0(path, individuals[1]))) 
    if(grepl(".nts", file)) { 
        ## Does contain a file
        data <- readland.nts(paste0(path, individuals[1], "/", list.files(paste0(path, individuals[1]))))
        name <- paste0(strsplit(individuals[1], split = "_")[[1]][1:4], collapse = "_")
        all_individuals[[length(all_individuals) + 1]] <- data[, , 1]
        names(all_individuals)[length(all_individuals)] <- name
    } else {
        print(paste0(file, ": empty"))
    }
    individuals <- individuals[-1]
}

## The following are missing:
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_14_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_19_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_coelebs_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_21_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_5_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_moreletti_11_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_palmae_13_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_polatzeki_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_teydea_10_intraspecific/: empty"

## Converting into an array
individuals_array <- simplify2array(all_individuals)

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, individuals_array)*-1

## BUG IN THE NUMBER OF DIMENSIONS!
dim(updated_pca)[1] == dim(whole_gpa$PCscores)[1] + length(all_individuals)
dim(updated_pca)[2] == dim(whole_gpa$PCscores)[2]

## SOME SMALL SHIFT!
plot(whole_gpa$PCscores[,c(1,2)])
points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5)
points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "red")
legend("bottomright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
```

### 2024-12-17 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 And now let's generate some extra data out of the pipeline.
-Here we're just creating a mean, median and minimum landmark set using the mean, median and minimum landmark coordinates values.
+Here we're just creating a mean, median and minimum landmark set using the mean and median landmark coordinates values as well as one of a specific specimen (42).
 
@@
 ## Create a new specimens: the average, the median and the minimum, of all bones)
-require(landvR) # devtools::install_github("TGuillerme/landvR")
 new_spec <- array(data = c(apply(original_landmarks, c(1,2), mean),
                            apply(original_landmarks, c(1,2), median),
-                           unlist(landvR::select.procrustes(proc_orig$rotated, selector = min))),
-                           dim = c(10, 3, 3))
-dimnames(new_spec)[[3]] <- paste0("new_", c("mean", "median", "min"))
+                           original_landmarks[,,42]), dim = c(10, 3, 3))
+dimnames(new_spec)[[3]] <- paste0("new_", c("mean", "median", "spec42"))
 
@@
 plot(new_spec[, c(1,2), 2], main = "New specimen 2\n(median)")
-plot(new_spec[, c(1,2), 3], main = "New specimen 3\n(weird)")
+plot(new_spec[, c(1,2), 3], main = "New specimen 3\n(spec 42)")
 par(op)
@@
 ```{r}
-## Add them to the procrustes + PCA
-new_proc <- Morpho::align2procSym(proc_orig, new_spec)
-dimnames(new_proc)[[3]] <- c("mean", "median", "min")
-
-## Transforming the procrustes arrays into matrices
-new_proc_matrix <- landvR::array.to(new_proc, to = "matrix")
-
-## Run the PCA with the added procrustes
-new_pca <- rbind(pca_orig$x, predict(pca_orig, newdata = new_proc_matrix))
+## Updating the gpa and pca
+updated_pca <- update.gpa.pca(proc_orig, new_spec)
 
 ## Plot all the results
-plot(pca_orig$x[,c(1,2)])
-points(new_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5)
-points(new_pca[c(81:83), c(1,2)], pch = 19, col = "red")
+plot(proc_orig$PCscores[,c(1,2)])
+points(updated_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5)
+points(updated_pca[c(81:83), c(1,2)], pch = 19, col = "red")
 legend("topright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
@@
         names(all_individuals)[length(all_individuals)] <- name
-    } else {
-        print(paste0(file, ": empty"))
-    }
+    } #else {
+        # print(paste0(file, ": empty"))
+    # }
     individuals <- individuals[-1]
@@
 ## Project them! (*-1 is to flip it around)
-updated_pca <- update.gpa.pca(whole_gpa, individuals_array)*-1
+updated_pca <- update.gpa.pca(whole_gpa, individuals_array, scale.pca = FALSE)
 
@@
 ## SOME SMALL SHIFT!
-plot(whole_gpa$PCscores[,c(1,2)])
-points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5)
-points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "red")
-legend("bottomright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
+plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
+points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
+## Plot the chaffinches to see where they lay:
+fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
+points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
+points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
+legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
 ```

### 2025-04-07 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 
+## Getting distributions
+elaborations_global <- get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]][, sample(1:4000, 100)]
+innovations_global <- get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]][, sample(1:4000, 100)]
+
+write.csv(elaborations_global, file = "../Data/Processed/elaborations_distributions.csv")
+write.csv(innovations_global, file = "../Data/Processed/elaborations_distributions.csv")
+
 ## Get the data at the macro level (with no missing data)
@@
 save(ploting_data_order, file = "../Data/Processed/ploting_data_order.rda")
+
+## Different scales
+elaboration_innovation_scale <- data.frame("median_elaboration_superorder" = ploting_data_superorder$median_elaborations,
+                                           "median_elaboration_order" = ploting_data_order$median_elaborations,
+                                           "median_innovation_superorder" = ploting_data_superorder$median_innovations,
+                                           "median_innovation_order" = ploting_data_order$median_innovations,
+                                           "median_distance_centre_superorder" = ploting_data_superorder$distances_to_centre,
+                                           "median_distance_centre_order" = ploting_data_order$distances_to_centre)
+write.csv(elaboration_innovation_scale, file = "../Data/Processed/elaboration_innovation_scale.csv")
 

### 2025-04-14 Analysis/03-Elaboration_innovation_analysis.Rmd
Update 03-Elaboration_innovation_analysis.Rmd

@@
 write.csv(elaborations_global, file = "../Data/Processed/elaborations_distributions.csv")
-write.csv(innovations_global, file = "../Data/Processed/elaborations_distributions.csv")
+write.csv(innovations_global, file = "../Data/Processed/innovations_distributions.csv")
 

### 2025-04-14 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 library(Morpho)
+library(landvR) # install from devtools::install_github("TGuillerme/landvR")
 
@@
 
-The function that does it automatically is now implemented in `landvR`.
+But does it work with the same specimens? I.e. does it place previous specimens in the correct position?
+
+```{r}
+## Old specimens
+old_spec <- array(data = c(original_landmarks[,,1],
+                           original_landmarks[,,2],
+                           original_landmarks[,,3]), dim = c(10, 3, 3))
+dimnames(old_spec)[[3]] <- paste0("old_", c("spec1", "spec2", "spec3"))
+
+## Updating the gpa and pca
+updated_pca <- update.gpa.pca(proc_orig, old_spec, center.pca = TRUE, scale.pca = FALSE)
+
+## Plot all the results
+plot(proc_orig$PCscores[,c(1,2)], col = "grey")
+points(updated_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5, col = "grey")
+
+points(proc_orig$PCscores[c(1,2,3),c(1,2)], col = "blue", pch = 21)
+points(updated_pca[c(81:83),c(1,2)], col = "orange", pch = 19)
+
+legend("topright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "old_spec (target)", "old_spec (results)"))
+```
+
+This is not ideal!
+The reason why it doesn't recover them is because it slightly offsets the superimposition which cascades to big changes in the PCA.
+
+For example, when using the same specimens:
+
+```{r}
+## Plotting the landmarks before superimposition
+op <- par(mfrow = c(2,2))
+plot(old_spec[, c(1,2), 1], main = "2D landmarks for spec 1", xlab = "", ylab = "")
+plot(old_spec[, c(1,2), 2], main = "2D landmarks for spec 2", xlab = "", ylab = "")
+plot(old_spec[, c(1,2), 3], main = "2D landmarks for spec 3", xlab = "", ylab = "")
+par(op)
+
+## Project the new specimens on the original GPA
+new_gpa <- Morpho::align2procSym(proc_orig, old_spec, orp = TRUE)
+
+## Slightly off?
+op <- par(mfrow = c(2,2))
+## Spec 1
+plot.offset <- function(id, new_gpa, old_gpa) {
+    main <- paste0("spec ", id, " reprojected onto itself\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
+    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
+    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
+}
+
+plot.offset(id = 1, new_gpa, proc_orig$rotated)
+plot.offset(id = 2, new_gpa, proc_orig$rotated)
+plot.offset(id = 3, new_gpa, proc_orig$rotated)
+
+main <- paste0("spec 1 projected onto spec 2 (for ref.)\nTotal offset = ", sum(sqrt((proc_orig$rotated[, 1, 1] - proc_orig$rotated[, 1, 2])^2 + (proc_orig$rotated[, 2, 1] - proc_orig$rotated[, 2, 2])^2)))
+plot(proc_orig$rotated[, c(1,2), 1], main = main, xlab = "", ylab = "")
+points(proc_orig$rotated[, c(1,2), 2], pch = 19, cex = 0.5)
+
+par(op)
+```
+
+This behaviour is actually highlighted in the notes for `align2procSym`: "this will never yield the same result as a pooled Procrustes analysis because the sample mean is iteratively updated and new data would change the mean."
+
+> TL;DR: it's not working until we find a way to not update the mean.
 
@@
 ## Project them! (*-1 is to flip it around)
-updated_pca <- update.gpa.pca(whole_gpa, individuals_array, scale.pca = FALSE)
+updated_pca <- update.gpa.pca(whole_gpa, whole_gpa$rotated[,,c(2819, 2820)], scale.pca = FALSE)
 
-## BUG IN THE NUMBER OF DIMENSIONS!
-dim(updated_pca)[1] == dim(whole_gpa$PCscores)[1] + length(all_individuals)
-dim(updated_pca)[2] == dim(whole_gpa$PCscores)[2]
-
-## SOME SMALL SHIFT!
+## Some big shift!
 plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
@@
 ```
+
+```{r}
+library(Morpho)
+
+## Get some 3D data from Morpho
+data(boneData)
+original_landmarks <- boneLM
+
+## Run the GPA for the original data
+proc_orig <- Morpho::procSym(original_landmarks)
+
+## Creating some new specimens (two old ones, for benchmarking the method)
+new_spec <- array(c(original_landmarks[,,1],
+                    original_landmarks[,,2]), dim = c(10, 3, 2))
+
+## Project the new specimens on the original GPA
+new_gpa <- Morpho::align2procSym(proc_orig, new_spec, orp = TRUE)
+
+## Slightly off?
+plot.offset <- function(id, new_gpa, old_gpa, main) {
+    main <- paste0(main, "\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
+    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
+    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
+}
+
+op <- par(mfrow = c(2,1))
+plot.offset(id = 1, new_gpa, proc_orig$rotated, main = "Spec 1 projected onto itself")
+plot.offset(id = 2, new_gpa, proc_orig$rotated, main = "Spec 2 projected onto itself")
+par(op)
+
+data(boneData)
+# run procSym on entire data set
+proc <- procSym(boneLM)
+# this is the training data
+array1 <- boneLM[,,1:60]
+proc1 <- procSym(array1)
+newdata <- boneLM[,,1:2]
+newalign <- align2procSym(x = proc1, newdata = newdata)
+
+plot.offset(id = 1, newalign, proc1$rotated, main = "Spec 1 projected onto itself")
+
+```
+
+TODO: try doing the alignement of some species on the edges of the traitspace and see if it falls back on the edges or if it goes to the middle again.
+
+TODO: check with the `morphospace` package?

### 2025-04-28 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 # The pipeline (testing)
+
+TODO: TRY CENTERING THE DATA BEFORE DOING THE PCA
 
@@
 legend("topright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "old_spec (target)", "old_spec (results)"))
+
 ```
@@
 par(op)
+```
+
+Checking all over again with `princomp`
+
+```{r}
+## New test
+library(Morpho)
+library(landvR) # install from devtools::install_github("TGuillerme/landvR")
+
+## Get some 3D data from Morpho
+data(boneData)
+original_landmarks <- boneLM
+
+## Run the GPA for the original data
+proc_orig <- Morpho::procSym(original_landmarks)
+
+## New species (three to anchor and two new ones)
+new_spec <- array(data = c(original_landmarks[,,1],
+                           original_landmarks[,,2],
+                           original_landmarks[,,3],
+                           apply(original_landmarks, c(1,2), median),
+                           apply(original_landmarks, c(1,2), min)),  dim = c(10, 3, 5))
+dimnames(new_spec)[[3]] <- c(paste0("old_", c("spec1", "spec2", "spec3")), "median", "min")
+
+## Run the PCA for the original data
+proc_matrix_orig <- landvR::array.to(proc_orig$rotated, to = "matrix")
+pca_orig <- prcomp(proc_matrix_orig)
+princomp_orig <- princomp(landvR::array.to(proc_orig$rotated, to = "matrix"))
+
+## THIS PIPELINE SEEMS TO WORK:
+
+## Test the reprojection using princomp
+new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
+new_points <- stats::predict(princomp_orig, newdata = array.to(new_gpa, to = "matrix"))
+plot(original_pca$scores[,c(1,2)], col = "grey")
+points(new_points[,c(1,2)], col = "orange", pch = 19)
+points(original_pca$scores[c(1,2,3),c(1,2)], col = "blue", pch = 21)
+
+## NOW NEEDS TO CHANGE FROM THE MORPHO PCA to PRINCOMP FORMAT AND BACK
 ```

### 2025-04-28 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

--- 2025-04-14 Analysis/17-update_procrustes_and_PCA.Rmd
@@
 
+Checking all over again with `princomp`
+
+```{r}
+## New test
+library(Morpho)
+library(landvR) # install from devtools::install_github("TGuillerme/landvR")
+
+## Get some 3D data from Morpho
+data(boneData)
+original_landmarks <- boneLM
+
+## Run the GPA for the original data
+proc_orig <- Morpho::procSym(original_landmarks)
+
+## New species (three to anchor and two new ones)
+new_spec <- array(data = c(original_landmarks[,,1],
+                           original_landmarks[,,2],
+                           original_landmarks[,,3],
+                           apply(original_landmarks, c(1,2), median),
+                           apply(original_landmarks, c(1,2), min)),  dim = c(10, 3, 5))
+dimnames(new_spec)[[3]] <- c(paste0("old_", c("spec1", "spec2", "spec3")), "median", "min")
+
+## Run the PCA for the original data
+pca_orig <- prcomp(landvR::array.to(proc_orig$rotated, to = "matrix"))
+princomp_orig <- princomp(landvR::array.to(proc_orig$rotated, to = "matrix"))
+
+## THIS PIPELINE SEEMS TO WORK:
+
+## Test the reprojection using princomp
+new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
+new_points <- stats::predict(princomp_orig, newdata = array.to(new_gpa, to = "matrix"))
+plot(princomp_orig$scores[,c(1,2)], col = "grey")
+points(new_points[,c(1,2)], col = "orange", pch = 19)
+points(princomp_orig$scores[c(1,2,3),c(1,2)], col = "blue", pch = 21)
+
+## Thest using prcomp
+new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
+new_points <- stats::predict(pca_orig, newdata = array.to(new_gpa, to = "matrix"))
+plot(pca_orig$x[,c(1,2)], col = "grey")
+points(new_points[,c(1,2)], col = "orange", pch = 19)
+points(pca_orig$x[c(1,2,3),c(1,2)], col = "blue", pch = 21)
+
+## Test with pipeline
+updated_pca <- update.gpa.pca(proc_orig, new_spec)*-1 # TODO: fix the -1
+plot(proc_orig$PCscores[,c(1,2)])
+points(updated_pca[,c(1,2)], col = "grey", pch = 19)
+points(updated_pca[c(81:85),c(1,2)], col = "orange", pch = 19)
+points(updated_pca[c(81:83),c(1,2)], col = "blue", pch = 21)
+
+## Not really equal though!
+expect_equal(updated_pca[-c(81:85),c(1,2)], proc_orig$PCscores[,c(1,2)])
+
+```
+
 This behaviour is actually highlighted in the notes for `align2procSym`: "this will never yield the same result as a pooled Procrustes analysis because the sample mean is iteratively updated and new data would change the mean."
@@
 
-## The following are missing:
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_14_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_19_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_coelebs_1_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_1_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_21_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_5_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_moreletti_11_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_palmae_13_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_polatzeki_1_intraspecific/: empty"
-# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_teydea_10_intraspecific/: empty"
-
 ## Converting into an array
-individuals_array <- simplify2array(all_individuals)
+new_individuals <- simplify2array(all_individuals)
 
 ## Project them! (*-1 is to flip it around)
-updated_pca <- update.gpa.pca(whole_gpa, whole_gpa$rotated[,,c(2819, 2820)], scale.pca = FALSE)
+updated_pca <- update.gpa.pca(whole_gpa, new_individuals)*-1
+
+# No pipeline
+# Get the CPA
+original_data <- whole_gpa
+new_landmarks <- new_individuals
+original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+# original_pca <- prcompfast(original_data$tan) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+## Project the new specimens on the original GPA
+new_data <- Morpho::align2procSym(original_data, new_landmarks)
+## Project the new GPAed specimens onto the PCA
+new_pca <- stats::predict(original_pca, newdata = array.to(new_data, to = "matrix"))
+## Combine the new and old matrices for the output PCA
+updated_pca <- rbind(original_pca$x, new_pca) *-1
 
@@
 ```
-
-```{r}
-library(Morpho)
-
-## Get some 3D data from Morpho
-data(boneData)
-original_landmarks <- boneLM
-
-## Run the GPA for the original data
-proc_orig <- Morpho::procSym(original_landmarks)
-
-## Creating some new specimens (two old ones, for benchmarking the method)
-new_spec <- array(c(original_landmarks[,,1],
-                    original_landmarks[,,2]), dim = c(10, 3, 2))
-
-## Project the new specimens on the original GPA
-new_gpa <- Morpho::align2procSym(proc_orig, new_spec, orp = TRUE)
-
-## Slightly off?
-plot.offset <- function(id, new_gpa, old_gpa, main) {
-    main <- paste0(main, "\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
-    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
-    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
-}
-
-op <- par(mfrow = c(2,1))
-plot.offset(id = 1, new_gpa, proc_orig$rotated, main = "Spec 1 projected onto itself")
-plot.offset(id = 2, new_gpa, proc_orig$rotated, main = "Spec 2 projected onto itself")
-par(op)
-
-data(boneData)
-# run procSym on entire data set
-proc <- procSym(boneLM)
-# this is the training data
-array1 <- boneLM[,,1:60]
-proc1 <- procSym(array1)
-newdata <- boneLM[,,1:2]
-newalign <- align2procSym(x = proc1, newdata = newdata)
-
-plot.offset(id = 1, newalign, proc1$rotated, main = "Spec 1 projected onto itself")
-
-```
-
-TODO: try doing the alignement of some species on the edges of the traitspace and see if it falls back on the edges or if it goes to the middle again.
-
-TODO: check with the `morphospace` package?

### 2025-05-01 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 # The pipeline (testing)
-
-TODO: TRY CENTERING THE DATA BEFORE DOING THE PCA
 
@@
 ## Test with pipeline
-updated_pca <- update.gpa.pca(proc_orig, new_spec)*-1 # TODO: fix the -1
+updated_pca <- update.gpa.pca(proc_orig, new_spec)#*-1 # TODO: fix the -1
 plot(proc_orig$PCscores[,c(1,2)])
@@
 
+## From Morpho:
+eigenPCA <- function(data,tol=.Machine$double.eps) {
+    center1 <- colMeans(data,2,mean)
+    datascale <- scale(data,2,scale=FALSE)
+    eigData <- eigen(cov(datascale),symmetric=TRUE)
+    eigData$values <- eigData$values[which(eigData$values > tol)] 
+    sdev <- sqrt(eigData$values)
+    rotation <- eigData$vectors[,1:length(sdev)]
+    x <- t(t(rotation)%*%t(datascale))
+    return(list(x=x,rotation=rotation,sdev=sdev,center=center1))
+}
+
 # No pipeline
@@
 original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
-# original_pca <- prcompfast(original_data$tan) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+original_pca <- eigenPCA(original_data$tan)
+
+    center1 <- colMeans(data,2,mean)
+    datascale <- scale(data,2,scale=FALSE)
+    eigData <- eigen(cov(datascale),symmetric=TRUE)
+
+landvR::array.to(original_data$rotated, to = "matrix")
+
 ## Project the new specimens on the original GPA

### 2025-05-02 Analysis/17-update_procrustes_and_PCA.Rmd
Working-ish?

@@
 ## Project them! (*-1 is to flip it around)
-updated_pca <- update.gpa.pca(whole_gpa, new_individuals)*-1
-
-## From Morpho:
-eigenPCA <- function(data,tol=.Machine$double.eps) {
-    center1 <- colMeans(data,2,mean)
-    datascale <- scale(data,2,scale=FALSE)
-    eigData <- eigen(cov(datascale),symmetric=TRUE)
-    eigData$values <- eigData$values[which(eigData$values > tol)] 
-    sdev <- sqrt(eigData$values)
-    rotation <- eigData$vectors[,1:length(sdev)]
-    x <- t(t(rotation)%*%t(datascale))
-    return(list(x=x,rotation=rotation,sdev=sdev,center=center1))
-}
+# updated_pca <- update.gpa.pca(whole_gpa, new_individuals)*-1
 
@@
 new_landmarks <- new_individuals
-original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+# original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
 original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
-original_pca <- eigenPCA(original_data$tan)
+#original_pca <- eigenPCA(landvR::array.to(original_data$rotated, to = "matrix"))
 
-    center1 <- colMeans(data,2,mean)
-    datascale <- scale(data,2,scale=FALSE)
-    eigData <- eigen(cov(datascale),symmetric=TRUE)
-
-landvR::array.to(original_data$rotated, to = "matrix")
+    # center1 <- colMeans(data,2,mean)
+    # datascale <- scale(data,2,scale=FALSE)
+    # eigData <- eigen(cov(datascale),symmetric=TRUE)
 
@@
 ## Project the new GPAed specimens onto the PCA
-new_pca <- stats::predict(original_pca, newdata = array.to(new_data, to = "matrix"))
+
+## New data needs to be projected on the tangent space because of symmetry.
+# new_procrustes <- array.to(new_data, to = "matrix")
+
+new_proj_symtan <- sweep(new_data, 1:2, whole_gpa$mshape)
+new_proc_tan <- vecx(new_proj_symtan)
+# if (sizeshape) {
+#     meanlogCS <- mean(log(CS))
+#     CSlog <- log(CS) - meanlogCS
+#     tan <- cbind(CSlog,tan)
+# }
+
+new_pca <- stats::predict(original_pca, newdata = new_proc_tan)#array.to(new_proc_tan, to = "matrix"))
 ## Combine the new and old matrices for the output PCA
-updated_pca <- rbind(original_pca$x, new_pca) *-1
+updated_pca <- rbind(original_pca$x*-1, new_pca)
 
@@
 fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
+points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
 points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
-points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
 legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))

### 2025-05-02 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 
-# The pipeline (testing)
+# Mapping the chaffinches on the whole traitspace
 
-Here is an example on how to do it using the `Morpho` package for the base GPA and PCA:
+```{r, eval = FALSE, echo = FALSE}
+## Get some random shapes for cross-validation
 
-```{r}
-library(Morpho)
-library(landvR) # install from devtools::install_github("TGuillerme/landvR")
+# Read in landmark data and landmark pairs file (necessary for enforcing symmetry later)
+beaks <- readRDS("../Data/Raw/Beak_data/2020_06_19_MMB_LANDMARKS_ALL.rds")
+pairs <- read.table("../Data/Raw/Beak_data/landpairs.txt")
 
-## Get some 3D data from Morpho
-data(boneData)
-original_landmarks <- boneLM
+# Remove some annoying extra labelling on individual beaks
+spp <- gsub("_1|_2|_3|_4|_5|_6", "", beaks$species)
 
-## Run the GPA for the original data
-proc_orig <- Morpho::procSym(original_landmarks)
+# Get unique names
+uspp <- unique(spp)
 
-## Run the PCA for the original data
-proc_matrix_orig <- landvR::array.to(proc_orig$rotated, to = "matrix")
-pca_orig <- prcomp(proc_matrix_orig)
+# Create empty array and assign species names to each element in the array
+spp_mshape_all <- array(NA, dim=c(79, 3, length(uspp)))
+dimnames(spp_mshape_all)[[3]] <- uspp
 
-## Plot the results
-plot(pca_orig$x[,c(1,2)])
+# Loop over all species, read in individual user landmarks, apply procrustes alignment, forcing symmetry (other than for species with true asymmetry)
+# Store mean shape configuration for each species as the species levels set of landmarks for downstream analyses
+
+for (i in 1:length(uspp)) {
+    idx <- which(spp==uspp[i])
+    tmp_beaks <- beaks[[1]][,,idx]
+    spp_alig <- procSym(tmp_beaks, CSinit=FALSE, scale=FALSE, pairedLM=as.matrix(pairs))
+
+    if (sum(grep("Loxia_", uspp[i]),  grep("Anarhynchus_", uspp[i]))>0) {
+        spp_alig <- procSym(tmp_beaks, CSinit=FALSE, scale=FALSE)
+    }
+    spp_mshape <- spp_alig$mshape
+    spp_mshape_all[,,i] <- spp_mshape
+    print(i)
+}
+
+## Selecting 20 random bird
+randoms = c(sample(1:8748, 17), 753, 6077, 918) # hummingbird = 753, flamingo = 6077, shoebill = 918
+testing_projection_shapes <- spp_mshape_all[,, randoms]
+save(testing_projection_shapes, file = "../Data/Processed/testing_projection_shapes.rda")
 ```
 
-And now let's generate some extra data out of the pipeline.
-Here we're just creating a mean, median and minimum landmark set using the mean and median landmark coordinates values as well as one of a specific specimen (42).
+Here are the new chaffinches projected onto the former traitspace.
+The blue circles are the old chaffinches that were present in the dataset before.
 
-```{r}
-## Create a new specimens: the average, the median and the minimum, of all bones)
-new_spec <- array(data = c(apply(original_landmarks, c(1,2), mean),
-                           apply(original_landmarks, c(1,2), median),
-                           original_landmarks[,,42]), dim = c(10, 3, 3))
-dimnames(new_spec)[[3]] <- paste0("new_", c("mean", "median", "spec42"))
-
-## Here is what the landmark positions look like in 3D
-op <- par(mfrow = c(2,2))
-plot(original_landmarks[, c(1,2), 1], main = "First specimen\n(reference from real dataset)")
-plot(new_spec[, c(1,2), 1], main = "New specimen 1\n(average)")
-plot(new_spec[, c(1,2), 2], main = "New specimen 2\n(median)")
-plot(new_spec[, c(1,2), 3], main = "New specimen 3\n(spec 42)")
-par(op)
-```
-
-We can then add these specimens to the GPA and PCA.
-
-```{r}
-## Updating the gpa and pca
-updated_pca <- update.gpa.pca(proc_orig, new_spec)
-
-## Plot all the results
-plot(proc_orig$PCscores[,c(1,2)])
-points(updated_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5)
-points(updated_pca[c(81:83), c(1,2)], pch = 19, col = "red")
-legend("topright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
-```
-
-Seems to work quiet well!
-All the positions from previous specimens are respected.
-And the two new median and average specimens plot in the centre of the traitspace (as expected) with the weird one plotting on some edge of the traitspace.
-
-But does it work with the same specimens? I.e. does it place previous specimens in the correct position?
-
-```{r}
-## Old specimens
-old_spec <- array(data = c(original_landmarks[,,1],
-                           original_landmarks[,,2],
-                           original_landmarks[,,3]), dim = c(10, 3, 3))
-dimnames(old_spec)[[3]] <- paste0("old_", c("spec1", "spec2", "spec3"))
-
-## Updating the gpa and pca
-updated_pca <- update.gpa.pca(proc_orig, old_spec, center.pca = TRUE, scale.pca = FALSE)
-
-## Plot all the results
-plot(proc_orig$PCscores[,c(1,2)], col = "grey")
-points(updated_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5, col = "grey")
-
-points(proc_orig$PCscores[c(1,2,3),c(1,2)], col = "blue", pch = 21)
-points(updated_pca[c(81:83),c(1,2)], col = "orange", pch = 19)
-
-legend("topright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "old_spec (target)", "old_spec (results)"))
-```
-
-This is not ideal!
-The reason why it doesn't recover them is because it slightly offsets the superimposition which cascades to big changes in the PCA.
-
-For example, when using the same specimens:
-
-```{r}
-## Plotting the landmarks before superimposition
-op <- par(mfrow = c(2,2))
-plot(old_spec[, c(1,2), 1], main = "2D landmarks for spec 1", xlab = "", ylab = "")
-plot(old_spec[, c(1,2), 2], main = "2D landmarks for spec 2", xlab = "", ylab = "")
-plot(old_spec[, c(1,2), 3], main = "2D landmarks for spec 3", xlab = "", ylab = "")
-par(op)
-
-## Project the new specimens on the original GPA
-new_gpa <- Morpho::align2procSym(proc_orig, old_spec, orp = TRUE)
-
-## Slightly off?
-op <- par(mfrow = c(2,2))
-## Spec 1
-plot.offset <- function(id, new_gpa, old_gpa) {
-    main <- paste0("spec ", id, " reprojected onto itself\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
-    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
-    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
-}
-
-plot.offset(id = 1, new_gpa, proc_orig$rotated)
-plot.offset(id = 2, new_gpa, proc_orig$rotated)
-plot.offset(id = 3, new_gpa, proc_orig$rotated)
-
-main <- paste0("spec 1 projected onto spec 2 (for ref.)\nTotal offset = ", sum(sqrt((proc_orig$rotated[, 1, 1] - proc_orig$rotated[, 1, 2])^2 + (proc_orig$rotated[, 2, 1] - proc_orig$rotated[, 2, 2])^2)))
-plot(proc_orig$rotated[, c(1,2), 1], main = main, xlab = "", ylab = "")
-points(proc_orig$rotated[, c(1,2), 2], pch = 19, cex = 0.5)
-
-par(op)
-```
-
-Checking all over again with `princomp`
-
-```{r}
-## New test
-library(Morpho)
-library(landvR) # install from devtools::install_github("TGuillerme/landvR")
-
-## Get some 3D data from Morpho
-data(boneData)
-original_landmarks <- boneLM
-
-## Run the GPA for the original data
-proc_orig <- Morpho::procSym(original_landmarks)
-
-## New species (three to anchor and two new ones)
-new_spec <- array(data = c(original_landmarks[,,1],
-                           original_landmarks[,,2],
-                           original_landmarks[,,3],
-                           apply(original_landmarks, c(1,2), median),
-                           apply(original_landmarks, c(1,2), min)),  dim = c(10, 3, 5))
-dimnames(new_spec)[[3]] <- c(paste0("old_", c("spec1", "spec2", "spec3")), "median", "min")
-
-## Run the PCA for the original data
-pca_orig <- prcomp(landvR::array.to(proc_orig$rotated, to = "matrix"))
-princomp_orig <- princomp(landvR::array.to(proc_orig$rotated, to = "matrix"))
-
-## THIS PIPELINE SEEMS TO WORK:
-
-## Test the reprojection using princomp
-new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
-new_points <- stats::predict(princomp_orig, newdata = array.to(new_gpa, to = "matrix"))
-plot(princomp_orig$scores[,c(1,2)], col = "grey")
-points(new_points[,c(1,2)], col = "orange", pch = 19)
-points(princomp_orig$scores[c(1,2,3),c(1,2)], col = "blue", pch = 21)
-
-## Thest using prcomp
-new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
-new_points <- stats::predict(pca_orig, newdata = array.to(new_gpa, to = "matrix"))
-plot(pca_orig$x[,c(1,2)], col = "grey")
-points(new_points[,c(1,2)], col = "orange", pch = 19)
-points(pca_orig$x[c(1,2,3),c(1,2)], col = "blue", pch = 21)
-
-## Test with pipeline
-updated_pca <- update.gpa.pca(proc_orig, new_spec)#*-1 # TODO: fix the -1
-plot(proc_orig$PCscores[,c(1,2)])
-points(updated_pca[,c(1,2)], col = "grey", pch = 19)
-points(updated_pca[c(81:85),c(1,2)], col = "orange", pch = 19)
-points(updated_pca[c(81:83),c(1,2)], col = "blue", pch = 21)
-
-## Not really equal though!
-expect_equal(updated_pca[-c(81:85),c(1,2)], proc_orig$PCscores[,c(1,2)])
-
-```
-
-This behaviour is actually highlighted in the notes for `align2procSym`: "this will never yield the same result as a pooled Procrustes analysis because the sample mean is iteratively updated and new data would change the mean."
-
-> TL;DR: it's not working until we find a way to not update the mean.
-
-# Mapping the chaffinches on the whole traitspace
-
-```{r}
+```{r, echo = FALSE}
 ## Loading the entire bird GPA
@@
         ## Does contain a file
-        data <- readland.nts(paste0(path, individuals[1], "/", list.files(paste0(path, individuals[1]))))
+        data <- geomorph::readland.nts(paste0(path, individuals[1], "/", list.files(paste0(path, individuals[1]))))
         name <- paste0(strsplit(individuals[1], split = "_")[[1]][1:4], collapse = "_")
@@
 
+## Get some random beak shapes for cross validating
+load(file = "../Data/Processed/testing_projection_shapes.rda")
+
 ## Converting into an array
 new_individuals <- simplify2array(all_individuals)
+n_new <- dim(new_individuals)[[3]] 
+## Add the cross validation specimens
+new_individuals <- abind::abind(new_individuals, testing_projection_shapes)
 
@@
 new_landmarks <- new_individuals
-# original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+
+## Running the pca on the tangent space data
 original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
-#original_pca <- eigenPCA(landvR::array.to(original_data$rotated, to = "matrix"))
-
-    # center1 <- colMeans(data,2,mean)
-    # datascale <- scale(data,2,scale=FALSE)
-    # eigData <- eigen(cov(datascale),symmetric=TRUE)
 
@@
 new_data <- Morpho::align2procSym(original_data, new_landmarks)
-## Project the new GPAed specimens onto the PCA
 
 ## New data needs to be projected on the tangent space because of symmetry.
-# new_procrustes <- array.to(new_data, to = "matrix")
-
 new_proj_symtan <- sweep(new_data, 1:2, whole_gpa$mshape)
 new_proc_tan <- vecx(new_proj_symtan)
-# if (sizeshape) {
-#     meanlogCS <- mean(log(CS))
-#     CSlog <- log(CS) - meanlogCS
-#     tan <- cbind(CSlog,tan)
-# }
 
@@
 
-## Some big shift!
-plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
+## Good locations
+plot(whole_gpa$PCscores[,c(1,2)], col = "grey", main = "New chaffinches projected onto the former space")
 points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
+
 ## Plot the chaffinches to see where they lay:
 fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
-points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
+points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):(dim(updated_pca)[1]-20)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
 points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
-legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
+legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New chaffinches"))
 ```
+
+Looks all good but then I thought of doing a sanity check: do specimens that were present in the traitspace before get reprojected in the exact same space when running through the pipeline?
+I.e. if the "new" specimens are actually old ones, do they fall back in their exact location (we want the answer to be yes for the results to be exact and reusable).
+Here are twenty spread specimens:
+
+```{r, echo = FALSE}
+## But problem
+## Good locations
+plot(whole_gpa$PCscores[,c(1,2)], col = "grey", main = "Sanity check")
+points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
+
+## Test testing ones
+points(whole_gpa$PCscores[dimnames(testing_projection_shapes)[[3]], c(1,2)], pch = 21, col = "darkgreen")
+points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1+n_new):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "darkgreen")
+legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "darkgreen", "darkgreen"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Original spec", "Reprojected spec"))
+```
+
+Unfortunately they don't project where they were supposed to and there is some kind of "pull" to the center of the trait space.
+This is because of the method to reproject the procrustes actually recalculates the mean shape each time.
+I still need to tinker with `Morpho::align2procSym` so that it doesn't.

### 2025-05-02 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 The goal is to use the GPA and PCA from a previous dataset (published with Guillerme et al 2023 - Science Advances) and to project new specimens onto it.
+
+# The pseudo math:
+
+In pseudo math, the whole procedure is something like:
+
+ 1. Procrustes = multiplying the landmark data `D` by some estimated transformation matrix `M1`.
+ 2. PCA = multiplying the procrustes data `M1*D` by some transformation matrix (the eigvenctors) `M2`.
+
+So we should technically be able to get the numerical values from both transformation matrices and use them to feed in the new data to them to have something like `M2*M1*D'` where `D'` are the new landmarks.
+In that case if `M2*M1*D = traitspace`, then if `D' = D`, we should get the exact same results (a traispace with _n_ species with all the same coordinates).
+And if `D' = D + x` where `x` are _m_ new specimens, we should have a traitspace with the exact same results for `D` and new results of `x` that are "correctly" positioned into that space.
 
@@
 ## Running the pca on the tangent space data
-original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
+original_pca <- prcompfast(original_data$tan*-1)
 
@@
 I still need to tinker with `Morpho::align2procSym` so that it doesn't.
+
+```{r, eval = FALSE, echo = FALSE}
+orp <- function(A, mshape=NULL) { 
+  p <- dim(A)[1]
+  k <- dim(A)[2]
+  n <- dim(A)[3]
+  if (is.null(mshape))
+      mshape <- arrMean3(A)
+
+  m.size <- cSize(mshape)
+  Xc <- as.vector(mshape/m.size)
+  X <- vecx(A)/m.size
+  ##direction along mshape onto plane
+  XtoPlane <- t(apply(X,1,function(x){x <- t(c(crossprod(x,Xc))*Xc)}))
+  X1 <- X-XtoPlane
+  X1 <- t(X1)+Xc
+  proj <- array(X1, dim=c(p, k, n))*m.size
+  return(proj)
+}
+
+## Testing align2procSym without changing the mean
+new_individuals <- testing_projection_shapes
+original_data <- whole_gpa
+new_landmarks <- new_individuals
+
+## Internal logic for align2procSym
+n <- dim(new_landmarks)[3]
+atts <- attributes(whole_gpa)
+newdatarot <- new_landmarks
+
+## Centroid Sizing
+mysize <- apply(new_landmarks,3,cSize)
+for (i in 1:n){
+    new_landmarks[,,i] <- new_landmarks[,,i]/mysize[i]
+}
+
+for (i in 1:n) {
+    newdatarot[,,i] <- rotonto(x = whole_gpa$mshape, y =new_landmarks[,,i],scale=atts$scale,reflection=atts$reflect,centerweight=atts$centerweight,weights=atts$weights)$yrot
+}
+
+orpdata <- orp(newdatarot,whole_gpa$mshape)
+noorpda <- newdatarot
+
+## New data needs to be projected on the tangent space because of symmetry.
+symtan_orpda <- sweep(orpdata, 1:2, whole_gpa$mshape)
+symtan_noorp <- sweep(noorpda, 1:2, whole_gpa$mshape)
+proctan_ordat <- vecx(symtan_orpda)
+proctan_norop <- vecx(symtan_noorp)
+
+## Running the pca on the tangent space data
+original_pca <- prcompfast(original_data$tan*-1)
+
+## Getting the new PCAs
+new_pca_orpda <- stats::predict(original_pca, newdata = proctan_ordat)#array.to(new_proc_tan, to = "matrix"))
+new_pca_norop <- stats::predict(original_pca, newdata = proctan_norop)#array.to(new_proc_tan, to = "matrix"))
+
+## Combine the new and old matrices for the output PCA
+updated_pca <- original_pca$x*-1
+
+plot(whole_gpa$PCscores[,c(1,2)], col = "grey", main = "Sanity check")
+points(updated_pca[, c(1,2)], pch = 19, cex = 0.5, col = "grey")
+points(whole_gpa$PCscores[dimnames(testing_projection_shapes)[[3]], c(1,2)], pch = 21, col = "darkgreen")
+points(new_pca_orpda[, c(1,2)], col = "blue", pch = 19, cex = 0.5)
+points(new_pca_norop[, c(1,2)], col = "orange", pch = 19, cex = 0.5)
+
+```

### 2025-07-23 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 
-```{r, eval = FALSE, echo = FALSE}
+```r #, eval = FALSE, echo = FALSE}
 ## Get some random shapes for cross-validation
-
 # Read in landmark data and landmark pairs file (necessary for enforcing symmetry later)
@@
 
-```{r, echo = FALSE}
+```r #, echo = FALSE}
 ## Loading the entire bird GPA
@@
 
-```{r, echo = FALSE}
+```r #, echo = FALSE}
 ## But problem
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```r #, eval = FALSE, echo = FALSE}
 orp <- function(A, mshape=NULL) { 
@@
 
-## Centroid Sizing
+# Centroid Sizing -> skipped (cSizes all equal 1 already) -> maybe cSize needs to be a bit smaller? 0.95 seems to be the correct amount to get them to the right dimension.
 mysize <- apply(new_landmarks,3,cSize)
@@
 
+## The mean shape from the previously calculated gpa
+
+## This value needs to be multiplied by 1.048-ish for some reason
+prev_mean_shape <- whole_gpa$mshape # * 1.048
+
 for (i in 1:n) {
-    newdatarot[,,i] <- rotonto(x = whole_gpa$mshape, y =new_landmarks[,,i],scale=atts$scale,reflection=atts$reflect,centerweight=atts$centerweight,weights=atts$weights)$yrot
+    newdatarot[,,i]  <- rotonto(x = prev_mean_shape, y = new_landmarks[,,i], scale = atts$scale, reflection = atts$reflect, centerweight = atts$centerweight, weights = atts$weights)$yrot
 }
@@
 ## Getting the new PCAs
-new_pca_orpda <- stats::predict(original_pca, newdata = proctan_ordat)#array.to(new_proc_tan, to = "matrix"))
-new_pca_norop <- stats::predict(original_pca, newdata = proctan_norop)#array.to(new_proc_tan, to = "matrix"))
+new_pca_orpda <- stats::predict(original_pca, newdata = proctan_ordat)
+new_pca_norop <- stats::predict(original_pca, newdata = proctan_norop)
 
@@
 
+points(new_pca_orpda[, c(1,2)]/0.954, col = "blue", pch = 19, cex = 0.5)
+points(new_pca_norop[, c(1,2)]*1.048, col = "orange", pch = 19, cex = 0.5)
+
+## Something with 1.048 (or /0.95) scaling somewhere? Maybe cSize?
+
 ```

### 2025-10-24 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@
 
-## This value needs to be multiplied by 1.048-ish for some reason
-prev_mean_shape <- whole_gpa$mshape # * 1.048
+## This value needs to be multiplied by 1.048-ish for some reason or 0.954
+prev_mean_shape <- whole_gpa$mshape # / 0.954 # * 1.048
+
+## n Previous specimens = 8748
+## n New specimens = 20
+
+## Kind of works.
+scalars <- seq(from = 0.9538, to = 0.9552, length.out = 20)
 
 for (i in 1:n) {
+
+    ## New shape should be updated each time?
+    prev_mean_shape <- whole_gpa$mshape / scalars[i]
+    ## Test solution: update gpa (and mshape) one spec at the time? Like:
+    # updated_gpa <- previous_gpa
+    # for (i in 1:new_spec) {
+    #     updated_gpa <- update.gpa(updated_gpa, new_specimen)
+    # }
+
     newdatarot[,,i]  <- rotonto(x = prev_mean_shape, y = new_landmarks[,,i], scale = atts$scale, reflection = atts$reflect, centerweight = atts$centerweight, weights = atts$weights)$yrot
@@
 new_pca_norop <- stats::predict(original_pca, newdata = proctan_norop)
+        # -> No difference between orp or not
 
@@
 points(new_pca_orpda[, c(1,2)], col = "blue", pch = 19, cex = 0.5)
-points(new_pca_norop[, c(1,2)], col = "orange", pch = 19, cex = 0.5)
+# points(new_pca_norop[, c(1,2)], col = "orange", pch = 19, cex = 0.5)
+
+ratios_orp <- new_pca_orpda[, c(1,2)] / whole_gpa$PCscores[dimnames(testing_projection_shapes)[[3]], c(1,2)]
+ratios_noorp <- new_pca_norop[, c(1,2)] / whole_gpa$PCscores[dimnames(testing_projection_shapes)[[3]], c(1,2)]
+
+plot(sort(ratios_orp[, 1]), type = "l")
+lines(sort(ratios_orp[, 2]), col = "red")
+
+points(new_pca_orpda[, c(1,2)] / ratios, col = "blue", pch = 19, cex = 0.5)
 
